### 2020-09-04 draft/intro.tex
working latex

\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usepackage{array,multirow}
\usepackage{subcaption}
\usepackage{subfiles}
\usepackage{url}
\usepackage{amsmath}
\usepackage{float}
\usepackage{graphicx}
\graphicspath{{sections/figs/}{.figs/}}
\usepackage[backend=bibtex, style=numeric-comp]{biblatex}
\bibliography{glasslab_viz}

\title{Make any stupid plot you want}
\author{Hannah Aizenman}
\committee{Dr. Michael Grossberg (Advisor), Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo}
\submitted{}
\abstract{Often the question in data exploration is not if a pattern occurs, but why. The search for the why is often the search for a conditionally dependent relationship in the dataset. While there are visualization methods that can be used to uncover a potential causal link between variables in the dataset, many of these techniques only work when the causal variables are categorical or can only be applied between two variables at a time or discard the complex, often spatiotemporal, structure of the dataset. Instead structure preserving methods for visualizing the distribution of observations provide a way of understanding some of the probabilistic relationships within and between the observations in these complex datasets. Using the type and structure of the variables in a dataset as a guide, this survey presents methods for visualizing conditional dependencies for multivariate datasets with discrete observations and visualizing probability distributions for datasets where observations are drawn from a continuous (functional) space. Because there is limited literature on explicit ways to visualize conditional dependency, we include techniques that suggest or can be adapted to show these dependencies.}

\begin{document}
\makefrontmatter

\section{Introduction}
\label{sec:introduction}

\printbibliography
\end{document}

### 2020-09-04 draft/intro.tex
latex + bulk notes in code

@@
 \usepackage{graphicx}
-\graphicspath{{sections/figs/}{.figs/}}
+\graphicspath{{figures}{.figures}}
 \usepackage[backend=bibtex, style=numeric-comp]{biblatex}
@@
 \submitted{}
-\abstract{Often the question in data exploration is not if a pattern occurs, but why.
-The search for the why is often the search for a conditionally dependent relationship in the dataset.
-While there are visualization methods that can be used to uncover a potential causal link between variables in the dataset, many of these techniques only work when the causal variables are categorical or can only be applied between two variables at a time or discard the complex, often spatiotemporal, structure of the dataset.
-Instead structure preserving methods for visualizing the distribution of observations provide a way of understanding some of the probabilistic relationships within and between the observations in these complex datasets.
-Using the type and structure of the variables in a dataset as a guide, this survey presents methods for visualizing conditional dependencies for multivariate datasets with discrete observations and visualizing probability distributions for datasets where observations are drawn from a continuous (functional) space.
-Because there is limited literature on explicit ways to visualize conditional dependency, we include techniques that suggest or can be adapted to show these dependencies.}
-
+\abstract{}
 \begin{document}
@@
 
+\cite{bertinSemiologyGraphicsDiagrams2011}
+
+\begin{figure}
+    \includegraphics{figures/intro/munzner_datatypes}
+    \label{fig:munzner_datatypes}
+    \caption{Keys are unique lookup values used to find individual observations in the dataset.
+Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube.
+Image modified from a diagram from Munzner's website\cite{munznerChDataAbstraction}}
+\end{figure}
+
+\cite{munznerWhatDataAbstraction2014}
+
 \printbibliography

### 2020-09-04 code/matplottoy/artists/core.py

@@ -28,7 +28,7 @@ def check_constraints(required, optional, encodings):
             <= optional):
         raise ValueError(f"Valid optional encodings: {optional}")
 
-
+"""
 def draw_alt(...):
     # do the external monad
     data_view = self.data.view(self.axes)
@@ -45,3 +45,4 @@ def _sekret_draw(self, transformed, renderer):
     gc = rendere.get_new_gc()
     gc.set_foreground_color(transform['color'])
     renderer.draw_path(transformed['x'], transformed['y'], )
+"""

### 2020-09-04 code/matplottoy/artists/line.py

@@ -8,7 +8,7 @@ import matplotlib.collections as mcollections
 from matplottoy.artists.core import check_constraints
 class Line(mcollections.LineCollection):
     required = {'y'}
-    optional = {'x', 'edges'}
+    optional = {'x'}
 
     def __init__(self, datasource, transfrom_functions=None, *args, **kwargs):
         """
@@ -24,11 +24,20 @@ class Line(mcollections.LineCollection):
         check_constraints(self.required, self.optional, 
                           datasource.encodings.keys())
              
+
+
         #how to refactor this into all the thing
         self.opt_encodings = (datasource.encodings.keys() - 
                                 Line.required)
         self.data = datasource
 
+        assert (self.keys.encodings==self.transforms.keys())
+
+        for key in encodings:
+            assert self.transforms[key].validate(data.info[key])
+
+        # check transforms via signature matching
+        # check transforms via validation
 
     def draw(self, renderer, *args, **kwargs):
         data_view = self.data.view(self.axes) 
@@ -36,8 +45,22 @@ class Line(mcollections.LineCollection):
         if 'x' in self.opt_encodings:
             x = partial(data_view.get, 'x')   
         else:
-            x = partial(np.arange, self.data.info['y']['length'])
-            
-        self.set_segments([np.vstack([x(), data_view.get('y')]).T])
+            x = partial(np.arange, self.data.view.m['observations'])
+
+        'xy' =  np.vstack([x(), data_view.get('y')]).T
+
+        for key in opt_encodings:
+            c[key] = append(self.transfroms.key.transform(...))
+    
+        # maps from idealised to insanity
+        # down the line don't want the 
+        # __draw(**c) 
+        
+    def __draw(components):
+        # make_composites(c) <- 
+        """collect all the like parameters
+        markerstyle, markerface, etc
+        """"
+        self.set_segments(components('xy'))
                            
         super().draw(renderer, *args, **kwargs)

### 2020-09-04 code/matplottoy/artists/point.py

@@ -45,10 +45,28 @@ class Point(mcollections.Collection):
         x = aes.x(data_view_get(''))
         y = aes.y(data.view_get(''))
 
+        """
+        #curry/partialized the mpath.Path.circle
+        c = {}
+        c['marker'] = mpath.Path.circle
+        c['size'] = 5
+        c['color'] = 'red'
+        """
         for (x, y, r) in zip(data_view.get('x'), 
                              data_view.get('y'), 
                              radius):
+            paths.append(c)
+
+      
+        self.__draw(renderer, *args, **kwargs)
+    """
+    def __draw(self):
+            ## to __draw   
+        for p in paths:
+            func = p.pop('marker')
+            p2.append(func(**p))
             paths.append(mpath.Path.circle(center=(x,y), radius=r))                         
         self._paths = aes.geo.get_paths('???')
         self.set_edgecolors('k')
-        super().draw(renderer, *args, **kwargs)
+        pass
+    """

### 2020-09-04 code/matplottoy/datasources/dataframe.py

@@ -9,6 +9,24 @@ from typing import Protocol
 
 import numpy as np 
 from matplottoy.datasources.core import  DataSource, Projection
+"""
+plot(x,y, ....transfroms)
+sataSource
+scatter(x,y,c, s...)
+artist:
+....s, markerstyle/edgestyle, ...
+"""
+class ArrayContainer:
+    """does the unpacking
+    {'x':0}...
+    """
+    DataSourcePoint(bundleofdata, [....])
+
+{'x':dataframe[0], 'y'= array[2], 
+'z': graph[j]}
+
+C1, C2, C3 -> {(DataSource , Transfroms) -> Artist}
+              {Line2D(**)}
 
 class View:
     # rows are selected, data is all the same number of rows
@@ -21,7 +39,7 @@ class View:
 
     def get(self, visual_var): # axis seperation of concern
         return getattr(self, visual_var)
-
+        
 def make_info(data, m, encodings):
     info = {'encodings': len(encodings)} 
     for key, col in encodings.items():
@@ -31,6 +49,27 @@ def make_info(data, m, encodings):
                     'max': var.max(), 
                     'name': col}
     return info
+
+{'x':0}, {('x', 'y'): LineEncoder}
+{'x':(0, LineEncoder)}
+{0:LineEncoder} <- signature checking
+{'x':LineEcoder} <-Datasource
+{'x': 0} <-DataContainer Level
+
+# passing in a literal means using 
+# default mpl transforms
+# identity transform
+# pass in two tuple w/ validate, transform
+
+Artist: {'marker': markertransformer...}
+{'markerfacecolor':colortransformer...}
+{'markeredgecolor':colortransformer,
+'color': 'orange', 'cmap': 'RdBu', 
+'fec': TransformTuple(lambda *args, **kwargs: 'red', lambda *arg, **kwargs: pass)}
+
+# encodings dict gets baked into data
+# encodings_meta = {'x':{name:}}
+
 class DataFramePoint:
     def __init__(self, data, m=None, encodings=None):
         """ 

### 2020-09-04 code/matplottoy/datasources/toplogy.py

@@ -1,0 +1,62 @@
+class View: # want an atomic snapshot
+# get returns part of coherant dataset
+# rows are selected, data is all the same number of rows
+    def __init__(self, data, m, ax=None):
+        """contract w/ draw method, just the things the draw needs
+        """
+        self.ax = ax # everything has the same axes
+        
+        for key, data in encodings.items():
+            setattr(self, key, data)
+    # get can't happen w/o view
+    def get(self, visual_var): # axis seperation of concern
+        return getattr(self, visual_var)
+
+
+df = pd.DataFrame(np.cumsum(
+                  np.tile([1., 2., .03, 5.], reps=100).reshape(100,4), axis=0), 
+                  columns = ['odds', 'evens','threes', 'fives'])
+
+encodings_meta({'x': {'min', 'max', 'shape'} )
+Data({'x': df['odds']}, self.m, encodings_meta=None)
+
+M = {'edges': tuple pairs,
+    'verts' : N or list((vertices),
+    'k':,}
+
+F = ['heterogenous', 'homogenous', 'univariate']
+# N + edges = 0 has to be discrete
+# N + edges 1 has to be continuous 1D
+
+class Data:
+    def __init__(self, encodings, encodings_meta = None):
+        """ 
+        """
+        # should axes go here - 
+        # (since there's a top level controller anyway)
+        # can only draw on the axes it's drawing on?
+        self.data = data
+        self.encodings = encodings if encodings else {}
+
+        if encodings_meta.keys()>= encodings.keys():
+            raise Warning(f"Meta provided for unencoded keys 
+                            {encodings_meta.keys() - encoded.keys()}")
+        self._info = {'encodings':encodings.keys()}
+
+        if 'm' in encodings_meta:
+            raise ValueError("Invalid Key: m")
+        self._info[m] = self.m
+
+    @property
+    def info(self):
+        """Read only attribute providing information about the data 
+        that can be used for labeling and axes formatting
+        """
+        if self._info['m'] != self.m:
+            # somehow data should provide a way to update into
+            raise Warning("key no longer valid") 
+        return self._info
+
+    def view(self, ax=None):
+        return View(self.data, self.m, ax=ax)
+    

### 2020-09-09 draft/intro.tex
du bois + more citations

@@
 \label{sec:introduction}
+\begin{figure}
+    \includegraphics[width=.25\textwidth]{figures/intro/dubois_bookplate}
+    \caption{This collection of visualizations is the inside cover art of W. E. B. Du Bois's Data Portraits\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018}}
+\end{figure}
+\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018}
 
+\begin{verbatim}
+make any stupid plot you want in robust & rigourous way
+        * need rich description langauge + right choice of paradigm (functional) 
+        * ability to mathemetaically formalize/conceptual framework for doing this
+        * data -> artist, need to preserved the structure of the data moreso than the data 
+            * how points are connected to each other
+    * targeted implementation rather than protocal (numpy)
+        * visualization libraries bound to the datastructures (concrete implementation)
+            * encode a lot of assumptions about data in the data structure
+            * MPL assume x/y plotted in order you want them in 
+            * topology makes the assumptions explicit
+            * explicit->math->functional  
+    * spell out the layers of visualization libaries:
+        * Data + computation -> visualization composites -> drawing library
+        * domain specific library
+        * utility library <- formalize this piece (SciPy Diagram)
+    * viz - Munzner + Bertin
+    * functional programming makes sense 
+\end{verbatim}
+
+\begin{figure}
+    \includegraphics{figures/intro/retinal_variables}
+    \label{fig:retinal_variable}
+    \caption{\cite{krygierMakingMapsVisual2005}}
+\end{figure}
 \cite{bertinSemiologyGraphicsDiagrams2011}
@@
 \end{figure}
-
 \cite{munznerWhatDataAbstraction2014}

### 2020-09-09 draft/intro.tex
introish
encoding start

@@
 \begin{figure}
-    \includegraphics[width=.25\textwidth]{figures/intro/dubois_bookplate}
+    \includegraphics[width=.5\textwidth]{figures/intro/dubois_bookplate}
     \caption{This collection of visualizations is the inside cover art of W. E. B. Du Bois's Data Portraits\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018}}
+    \label{fig:dubois_bookplate}
 \end{figure}
-\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018}
+
+Why am I starting with DuBois?
+
+ Some of DuBois visualizations are common chart types most modern visualization tools (cite excel, tableu, matplotlib, ggplot, maybe high chart) support, while others are far more custom and likely need tools with drawing capabilities (cite: matplotlib, base r, d3) \cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018}.
+This intentionality in what charts the tool supports and the flexibility it gives the user underpins why we care about this problem.
+Drawing programs don't really have to care about the data structure because everything is explicit - the user chooses the shape, line, color, they are by hand manually doing all the encoding.
+Visualization libraries on the other hand allow users to specify which bits of the data they want encoded and how, but there's a lot of implicit mapping of the raw variables in the data to the encodings.
+We want some confidence that the visualization tool is making the mapping we intended.
+We often do this by implictely encoding the data structure in the grammar of the visualiztion tool (grammaer of graphics, vega, ggplot, altair), but a goal of Matplotlib \cite{huntermatplotlib2007} is to be fairly data structure agnostic.
+We are proposing an architecture that converts these implicit assumptions into explicit contracts between the data and the chart types based on the assumptions visualization types make about the topology of the data.
+
+ Why topology?
+
+ The difference between a line plot and a scatter plot is the former assumes that the data is continous, the latter that it is discrete (cite line and scatter - friendly?) In concrete implementation terms, matplotlib's imshow assumes that the data is continous and therefore does implementation, the matshow assumes it is discrete so it does not.
+And there are visualization types like area charts that allow even more information to be encoded bewteen the lines.
+When redesigning the architecture, we needed a way to articulate the differences between the different plot types and found that a topological approach allowed us to encode connectivity (discrete versus continous), dimensionality (point, line, area), and how much information the visualization encodes.
+
+ What even is encoding?
+ \begin{figure}
+    \includegraphics{figures/intro/retinal_variables}
+    \label{fig:retinal_variable}
+    \caption{This codification of Bertin's retinal variables \cite{bertinSemiologyGraphicsDiagrams2011} that lays out the geometry (columns) versus the aesthetics (rows) with a recommendation based on data type (qualitative versus quantative) comes from Krygier and Wood's Making Maps \cite{krygierMakingMapsVisual2005}}
+\end{figure}
+
+When talking about encodings, we are referering to Bertin's codification of the properties of the graphic system \cite{bertinSemiologyGraphicsDiagrams2011}.
+
+\cite{bertinSemiologyGraphicsDiagrams2011}
+
+ By encoding, we mean
+
+\begin{figure}
+    \includegraphics{figures/intro/munzner_datatypes}
+    \label{fig:munzner_datatypes}
+    \caption{Keys are unique lookup values used to find individual observations in the dataset.
+Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube.
+Image modified from a diagram from Munzner's website\cite{munznerChDataAbstraction}}
+\end{figure}
+Munzner's key/value semantics\cite{munznerWhatDataAbstraction2014}
+provide a way to identify variables that in effect act as metadata for other variables, such as how when we are interested in the temperature at a time the temperature is the lookup value. But
+
+ that did not generalize well for complex heterogenous datasets. In this proposal, we refine those ideas by using topology to formally describe the connectivity between the measurements.
 
@@
 
-\begin{figure}
-    \includegraphics{figures/intro/retinal_variables}
-    \label{fig:retinal_variable}
-    \caption{\cite{krygierMakingMapsVisual2005}}
-\end{figure}
-\cite{bertinSemiologyGraphicsDiagrams2011}
-
-\begin{figure}
-    \includegraphics{figures/intro/munzner_datatypes}
-    \label{fig:munzner_datatypes}
-    \caption{Keys are unique lookup values used to find individual observations in the dataset.
-Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube.
-Image modified from a diagram from Munzner's website\cite{munznerChDataAbstraction}}
-\end{figure}
-\cite{munznerWhatDataAbstraction2014}
-
 \printbibliography

### 2020-09-10 draft/intro.tex
munzner encoders

@@
 \usepackage{graphicx}
-\graphicspath{{figures}{.figures}}
 \usepackage[backend=bibtex, style=numeric-comp]{biblatex}
@@
 
+ What even is data?
+ Remake this graph notated w/ graph semantics:
+ \begin{figure}
+    \includegraphics{figures/intro/data_formatting}
+    \caption{Modified from Munzner's Visualizarion Analysis and Design}
+ \end{figure}
+
  What even is encoding?
@@
     \label{fig:retinal_variable}
-    \caption{This codification of Bertin's retinal variables \cite{bertinSemiologyGraphicsDiagrams2011} that lays out the geometry (columns) versus the aesthetics (rows) with a recommendation based on data type (qualitative versus quantative) comes from Krygier and Wood's Making Maps \cite{krygierMakingMapsVisual2005}}
+    \caption{This organization of Bertin's retinal variables \cite{bertinSemiologyGraphicsDiagrams2011} that lays out the geometry(marks) and corresponding aesthetic variations (channels) with a recommendation based on data type (qualitative versus quantative) comes from Krygier and Wood's Making Maps \cite{krygierMakingMapsVisual2005}}
 \end{figure}
 
-When talking about encodings, we are referering to Bertin's codification of the properties of the graphic system \cite{bertinSemiologyGraphicsDiagrams2011}.
+When talking about encodings, we are referring to Bertin's codification of the properties of the graphic system \cite{bertinIIPropertiesGraphic2011} seen in fig~\ref{fig:retinal_variable} as visual variables.
+In this system, there are the marks, which are the point, line, and area geometric primatives on the visual plane that each represent a single instance of the data being visualized \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels2014}.
+These marks can be varied along different visual channels  size, value, texture, color, orientation, shape, and position \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels2014}.
+As shown in fig~\ref{fig:retinal_variable}, the marks encode topological connectivity such that they have the following graph semantics:
 
-\cite{bertinSemiologyGraphicsDiagrams2011}
+\begin{description} 
+    \item[Points] - 0d, no edges
+    \item[Lines] - 1d, every vertex has <=2 edges (1d simplex)
+    \item[Area] - 2d,  every vertex has <=4 edges (2d simplex)
+\end{description}
 
- By encoding, we mean
+The visual channels in fig~\ref{fig:retinal_variable} are recommended based on the measurement type, quantative versus qualitativem  and Munzner generalizes the quantative channels as magnitude channels for ordered attributes (measurements drawn from an ordered set of possible values) and the qualitative channels as identity channels for categorical attributes (random set of possible values).
 
@@
 Munzner's key/value semantics\cite{munznerWhatDataAbstraction2014}
-provide a way to identify variables that in effect act as metadata for other variables, such as how when we are interested in the temperature at a time the temperature is the lookup value. But
-
- that did not generalize well for complex heterogenous datasets. In this proposal, we refine those ideas by using topology to formally describe the connectivity between the measurements.
+provide a way to identify variables that in effect act as metadata for other variables, such as how when we are interested in the temperature at a time the temperature is the lookup value.
+But that did not generalize well for complex heterogenous datasets.
+In this proposal, we refine those ideas by using topology to formally describe the connectivity between the measurements.
 

### 2020-09-10 draft/intro.tex
gglplot references
funcitonal fill in

@@
 We often do this by implictely encoding the data structure in the grammar of the visualiztion tool (grammaer of graphics, vega, ggplot, altair), but a goal of Matplotlib \cite{huntermatplotlib2007} is to be fairly data structure agnostic.
-We are proposing an architecture that converts these implicit assumptions into explicit contracts between the data and the chart types based on the assumptions visualization types make about the topology of the data.
+We are proposing an architecture that converts these implicit assumptions into explicit contracts between the data and the visualization.
+
+ Why functional?
+ because what we care about is the interface, the contract between the data and the visualization.
+We need the minimal amount of information that fulfills our contract that we can then put on the screen.
+(insert react notes here).
+Our theoretical framework is that that visualization is a transformation from one topological space to another to a CW complex (this probably needs to be explained more), and a functional programming paradigm allows us to directly express that in code.
+
+ \subsection{What will be the work}
+ %%this comes from the grant
+ Matplotlib needs to support use cases across a vast range of science domains, enabling complex visualizations [1].
+At the same time, common visualizations in a domain need to be fluid for the end-practitioners, tuned to the domain's standard data structures, and with domain-specific customization options exposed.
+To achieve both of these goals, we need to continue to foster a two-layer ecosystem with a shared core (Matplotlib) and many domain-specific libraries (seaborn, nipy, ...).
+
+In the current grant cycle we have been developing a new architecture that is heavily invested in cleanly separating the three steps in a visualization pipeline: data representation, encoding the data as visual elements, and rendering those elements to screen.
+We believe that a better delineation of these steps will allow domain practitioners to more easily implement extensions.
+Following on the work done this year to design consistent data and artist abstractions, we will develop simpler and more expressive user-level APIs in core Matplotlib and in collaboration with domain-specific libraries.
 
@@
 
+\section{Library Review}
+\subsection{Grammar of Graphics \& ggplot}
+\cite{wilkinsonGrammarGraphicsStatistics2005}
+charts - words 
+    - instances of much more general objects (geometric primatives)
+    - histogram =/= bar chart
+graphics - statements 
+expliceltly OO
+
+\cite{wickhamGgplot2ElegantGraphics2016}
+
 \printbibliography

### 2020-09-10 draft/intro.tex
more gog

@@
  What even is encoding?
+
+ \begin{figure}
+    %% Bertin's decomposition of an image
+ \end{figure}
  \begin{figure}
@@
 \subsection{Grammar of Graphics \& ggplot}
+\begin{figure}
+    \includegraphics{figures/intro/grammar_chart_composition.png}
+    \caption{page 10 (introduction)}
+\end{figure}
+\begin{figure}
+    \includegraphics{figures/intro/grammar_example.png}
+    \caption{Wilkenson decomposed a graphic into ...} 
+\end{figure}
 \cite{wilkinsonGrammarGraphicsStatistics2005}
@@
 expliceltly OO
+    1. specification
+        data - operations/computations
+        trans - variable transforms (rank)
+        scale - scale transforms
+        coord - coordinate system
+        element - marks and channels
+        guide - meta elements - (axes, legends, etc)
+    2. assembly - kinda what happens in artist (spec->something that can go to renderer) 
+    3. display - rendering
+
+How is our proposal different? seperation in spec between what's data, aesthetics, and render specific, and what part of the architecture owns those operations
 

### 2020-09-11 draft/intro.tex
notes from meeting

@@
 Why am I starting with DuBois? 
+Visualization library - spells out/regularizes the ways you can aesthetically encode data, preserving structure of data, scatter - magnitude between values preserved - some ways an accurate representation of the data, concerned w/ as data is changed, graph has to change accordingly, graph needs to change accordingly - owns responsibility for preserving measurement integrity when specified by encodings
+    - specify what you want to preserve
+        - (choose which) preserve relations in a strict math/set theory sense
+    - functional is a good fit b/c it spells out these constraints 
+Drawing program - user is responsible for preserving these measurements
 
@@
 The visual channels in fig~\ref{fig:retinal_variable} are recommended based on the measurement type, quantative versus qualitativem  and Munzner generalizes the quantative channels as magnitude channels for ordered attributes (measurements drawn from an ordered set of possible values) and the qualitative channels as identity channels for categorical attributes (random set of possible values).
+topology in measurement types, and that topology informs the channels appropriate for those types:
+    quantative continouse, quantative discrete, qualitative catorical ordinal, qualitative categorical
+    marks - topology of measurements (people wearing dresses)
+    channels - topology of space from which measurements are drawn (possible dress sizes)
+    categorical - discrete points w/o anything to do 
+    categorical - poset topology?
+
+    base space - topology of observations (people wearing dresses)
+    fiber topology - topology of measurements (possible dress sizes)
+
+functional 'cause focus on transforms - if transforms preserve linearity than whole pipeline does. Dev gets guarantee on calling composite.
+
+Add table of which sorts of invariance a channel will preserve
 

### 2020-09-11 draft/intro.tex
cites + line or two

@@
  What even is encoding?
-
+Bertin table on 69 rewrite here clearly  
  \begin{figure}
@@
 
-Add table of which sorts of invariance a channel will preserve
+Add table of which sorts of invariance a channel will preserve....
 
@@
 \subsection{Grammar of Graphics \& ggplot}
+Specifies the end chart (graphic)
+we want to specify the transformations ....
 \begin{figure}
@@
 
+ggplot - strictly hierarchal components (layers), matplotlib - transforms that mostly happen indepently 
+"designing and producing statistical graphics is not an art" - ties in w/ difference between drawing program and visualization library, preservation of properties of measurement type and observation type
+
 \cite{wickhamGgplot2ElegantGraphics2016}

### 2020-09-17 draft/intro.tex
algebra of gog
gog pipeline description

@@
 
+history of viz - collins, friendly
+
+Wilkenson suggests that the graphics pipeline could be implemented functionally but does not push it in any direction.
+
+ordering: processed before plotted, 
+scales before stats, marks before channels
+\begin{figure}
+\includegraphics[]{figures/intro/gog_pathway.png}
+\caption{Diagram of Grammar of Graphics stages of visualization. Fig 2.2 from chapter 2 of The Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}}
+\label{fig:gog_pathway}
+\end{figure}
+In grammar of graphics, as shown in figure~\ref{fig:gog_pathway} data is extracted from the database one variable at a time with an associated index (primary key).
+The algebra stage joins the variables to become the varset, which is a flat table where the columns are the variables/attributes, each row is an observation/item, and each cell contains a measurement \cite{munznerChDataAbstraction}.
+Wilkenson then describes the constraints of the measurement space for each variable through scales \cite{wilkinsonGrammarGraphics2005} with implicit assumptions of the data constraints.
+After this step, data is transformed computationally in service of the visualization.
+Some of the variables are then mapped into geometric marks (symbols).
+Position based visual mappings such as x, y, or height are applied at this stage.
+These mappings are then transformed into the coordinate system of the target graph.
+
+Wilkenson's thoughts on coherancy are equivalent to invariance but less? formally stated \dots
+
+"To call these charts meainingul, defenders must falsify specific assumptions of GoG" \cite{wilkinsonMathematicalFoundationAnalytic2010}
+
+gog implementationss:
+\begin{enumerate}
+    \item SPSS nViZn
+    \item tableu
+    \item ggplot
+    \item protoviz\d3
+    \item vega (interactive GoG)
+\end{enumerate}
+
 \printbibliography

### 2020-09-22 draft/intro.tex
thesis statement

@@
 \begin{figure}
-    \includegraphics[width=.5\textwidth]{figures/intro/dubois_bookplate}
-    \caption{This collection of visualizations is the inside cover art of W. E. B. Du Bois's Data Portraits\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018}}
-    \label{fig:dubois_bookplate}
+    \includegraphics[width=.5\textwidth]{figures/intro/viz_same.png}
+    \caption[]{Implicit in visualization is the assumption that these three representations of data are equivalent, specifically that the measurements within a variable and relations of the measurements of each variable are preserved.
+}
+    \label{fig:viz_same}
 \end{figure}
 
-Why am I starting with DuBois? 
-Visualization library - spells out/regularizes the ways you can aesthetically encode data, preserving structure of data, scatter - magnitude between values preserved - some ways an accurate representation of the data, concerned w/ as data is changed, graph has to change accordingly, graph needs to change accordingly - owns responsibility for preserving measurement integrity when specified by encodings
-    - specify what you want to preserve
-        - (choose which) preserve relations in a strict math/set theory sense
-    - functional is a good fit b/c it spells out these constraints 
-Drawing program - user is responsible for preserving these measurements
+\section{Thesis statement}
+We define a visualization as a transform from data to graphic that preserves the topology of the data and the properties of the measurement type.
+In fig~\ref{fig:viz_same}, we implicitly assume that the translation from table to heatmap has preserved the order of observations (the rows) and that the perceptually uniform sequential colormap has been applied such that the ordering relation on floats matches the ordering on the colormap (darker colors map to larger numbers).
+We also make this assumption about color in the scatter map, and that the translation to size and position on screen also respect the ordering on floats.
+In this work, we propose to mathematically describe the transform of data to visual space such that we can make explicit the implicit topology and types visualizations preserve.
+We then propose a new architecture for the Python visualization library Matplotlib \cite{hunterMatplotlib2DGraphics2007} based on these descriptions because the Matplolib artist layer is analogous to the transforms.
 
- Some of DuBois visualizations are common chart types most modern visualization tools (cite excel, tableu, matplotlib, ggplot, maybe high chart) support, while others are far more custom and likely need tools with drawing capabilities (cite: matplotlib, base r, d3) \cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018}.
-This intentionality in what charts the tool supports and the flexibility it gives the user underpins why we care about this problem.
-Drawing programs don't really have to care about the data structure because everything is explicit - the user chooses the shape, line, color, they are by hand manually doing all the encoding.
-Visualization libraries on the other hand allow users to specify which bits of the data they want encoded and how, but there's a lot of implicit mapping of the raw variables in the data to the encodings.
-We want some confidence that the visualization tool is making the mapping we intended.
-We often do this by implictely encoding the data structure in the grammar of the visualiztion tool (grammaer of graphics, vega, ggplot, altair), but a goal of Matplotlib \cite{huntermatplotlib2007} is to be fairly data structure agnostic.
-We are proposing an architecture that converts these implicit assumptions into explicit contracts between the data and the visualization.
+\section{Notation \& Definitions}
 
- Why functional?
- because what we care about is the interface, the contract between the data and the visualization.
-We need the minimal amount of information that fulfills our contract that we can then put on the screen.
-(insert react notes here).
-Our theoretical framework is that that visualization is a transformation from one topological space to another to a CW complex (this probably needs to be explained more), and a functional programming paradigm allows us to directly express that in code.
-
- \subsection{What will be the work}
- %%this comes from the grant
- Matplotlib needs to support use cases across a vast range of science domains, enabling complex visualizations [1].
-At the same time, common visualizations in a domain need to be fluid for the end-practitioners, tuned to the domain's standard data structures, and with domain-specific customization options exposed.
-To achieve both of these goals, we need to continue to foster a two-layer ecosystem with a shared core (Matplotlib) and many domain-specific libraries (seaborn, nipy, ...).
-
-In the current grant cycle we have been developing a new architecture that is heavily invested in cleanly separating the three steps in a visualization pipeline: data representation, encoding the data as visual elements, and rendering those elements to screen.
-We believe that a better delineation of these steps will allow domain practitioners to more easily implement extensions.
-Following on the work done this year to design consistent data and artist abstractions, we will develop simpler and more expressive user-level APIs in core Matplotlib and in collaboration with domain-specific libraries.
-
- Why topology?
-
- The difference between a line plot and a scatter plot is the former assumes that the data is continous, the latter that it is discrete (cite line and scatter - friendly?) In concrete implementation terms, matplotlib's imshow assumes that the data is continous and therefore does implementation, the matshow assumes it is discrete so it does not.
-And there are visualization types like area charts that allow even more information to be encoded bewteen the lines.
-When redesigning the architecture, we needed a way to articulate the differences between the different plot types and found that a topological approach allowed us to encode connectivity (discrete versus continous), dimensionality (point, line, area), and how much information the visualization encodes.
-
- What even is data?
- Remake this graph notated w/ graph semantics:
- \begin{figure}
-    \includegraphics{figures/intro/data_formatting}
-    \caption{Modified from Munzner's Visualizarion Analysis and Design}
- \end{figure}
-
- What even is encoding?
-Bertin table on 69 rewrite here clearly  
- \begin{figure}
-    %% Bertin's decomposition of an image
- \end{figure}
- \begin{figure}
-    \includegraphics{figures/intro/retinal_variables}
-    \label{fig:retinal_variable}
-    \caption{This organization of Bertin's retinal variables \cite{bertinSemiologyGraphicsDiagrams2011} that lays out the geometry(marks) and corresponding aesthetic variations (channels) with a recommendation based on data type (qualitative versus quantative) comes from Krygier and Wood's Making Maps \cite{krygierMakingMapsVisual2005}}
-\end{figure}
-
-When talking about encodings, we are referring to Bertin's codification of the properties of the graphic system \cite{bertinIIPropertiesGraphic2011} seen in fig~\ref{fig:retinal_variable} as visual variables.
-In this system, there are the marks, which are the point, line, and area geometric primatives on the visual plane that each represent a single instance of the data being visualized \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels2014}.
-These marks can be varied along different visual channels  size, value, texture, color, orientation, shape, and position \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels2014}.
-As shown in fig~\ref{fig:retinal_variable}, the marks encode topological connectivity such that they have the following graph semantics:
-
-\begin{description} 
-    \item[Points] - 0d, no edges
-    \item[Lines] - 1d, every vertex has <=2 edges (1d simplex)
-    \item[Area] - 2d,  every vertex has <=4 edges (2d simplex)
-\end{description}
-
-The visual channels in fig~\ref{fig:retinal_variable} are recommended based on the measurement type, quantative versus qualitativem  and Munzner generalizes the quantative channels as magnitude channels for ordered attributes (measurements drawn from an ordered set of possible values) and the qualitative channels as identity channels for categorical attributes (random set of possible values).
-topology in measurement types, and that topology informs the channels appropriate for those types:
-    quantative continouse, quantative discrete, qualitative catorical ordinal, qualitative categorical
-    marks - topology of measurements (people wearing dresses)
-    channels - topology of space from which measurements are drawn (possible dress sizes)
-    categorical - discrete points w/o anything to do 
-    categorical - poset topology?
-
-    base space - topology of observations (people wearing dresses)
-    fiber topology - topology of measurements (possible dress sizes)
-
-functional 'cause focus on transforms - if transforms preserve linearity than whole pipeline does. Dev gets guarantee on calling composite.
-
-Add table of which sorts of invariance a channel will preserve....
-
-\begin{figure}
-    \includegraphics{figures/intro/munzner_datatypes}
-    \label{fig:munzner_datatypes}
-    \caption{Keys are unique lookup values used to find individual observations in the dataset.
-Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube.
-Image modified from a diagram from Munzner's website\cite{munznerChDataAbstraction}}
-\end{figure}
-Munzner's key/value semantics\cite{munznerWhatDataAbstraction2014}
-provide a way to identify variables that in effect act as metadata for other variables, such as how when we are interested in the temperature at a time the temperature is the lookup value.
-But that did not generalize well for complex heterogenous datasets.
-In this proposal, we refine those ideas by using topology to formally describe the connectivity between the measurements.
-
-\begin{verbatim}
-make any stupid plot you want in robust & rigourous way
-        * need rich description langauge + right choice of paradigm (functional) 
-        * ability to mathemetaically formalize/conceptual framework for doing this
-        * data -> artist, need to preserved the structure of the data moreso than the data 
-            * how points are connected to each other
-    * targeted implementation rather than protocal (numpy)
-        * visualization libraries bound to the datastructures (concrete implementation)
-            * encode a lot of assumptions about data in the data structure
-            * MPL assume x/y plotted in order you want them in 
-            * topology makes the assumptions explicit
-            * explicit->math->functional  
-    * spell out the layers of visualization libaries:
-        * Data + computation -> visualization composites -> drawing library
-        * domain specific library
-        * utility library <- formalize this piece (SciPy Diagram)
-    * viz - Munzner + Bertin
-    * functional programming makes sense 
-\end{verbatim}
+\section{Background}
+Bertin's proposed an organization of visual variables \cite{bertinIIPropertiesGraphic2011}
 
@@
 After this step, data is transformed computationally in service of the visualization.
-Some of the variables are then mapped into geometric marks (symbols).
-Position based visual mappings such as x, y, or height are applied at this stage.
+Some of the variables are then mapped into geometric marks (symbols) and are scaled (height, width) as appropriate.
 These mappings are then transformed into the coordinate system of the target graph.
+Then the aesthethic attributes (Bertin's retinal variables, also called channels) such as color or  texture) are applied.
+Finally
 

### 2020-09-22 draft/intro.tex
wilk line up

@@
 
-\section{Thesis statement}
+\subsection{Thesis statement}
 We define a visualization as a transform from data to graphic that preserves the topology of the data and the properties of the measurement type.
@@
 
-\section{Background}
+\subsection{visual variables}
 Bertin's proposed an organization of visual variables \cite{bertinIIPropertiesGraphic2011}
 
-\section{Library Review}
 \subsection{Grammar of Graphics \& ggplot}

### 2020-09-24 draft/intro.tex
start

@@
 \section{Notation \& Definitions}
+System of transforms:
+input: data w/ structure on fiber bundle + measurement type
+artist: data -> visualization
+(what structure of input is preserved in output)
+we can provide list of
+artist : topology it preserves
+point/line/face parameters: measurement type properties
+
+ex: 
+Line2D preserves continuity
+Linestyle: categorical
+
+%%define basic properties of measurement of types in a theoretical way or C&P from a paper
+
+Miniumum assumption a transform will make - is this a valid transform for the slot?
+
+Set of questions/rules to apply such that see a new artist can be defined which lets us maybe simplify optimizations.
+
+M, Fiber Bundle, Section, operatons on \{data, visual\}, transform data space \& visual space,
+
+topology gives us that m is preserved
+spivak gives us that v is preserved
+
+temperature on globe:
+\begin{equation}
+data = [[t1], [t2_{la1, lo2}]
+fiber = {lat, lon, temp}
+m = the data is 2d continous
+\end{equation}
+artist has implict m
+aesthetic paremeters transform in fiber that preserves type operations \& relations in the fiber
+lat - relative distance + ordering
+
+\subsubsection{Graphical Elements}
+There is a set of transform functions $T$ that maps from the data space $D$ to the visual space composed of geometric marks and aesthetic channels $V$ \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels}.
+We propose that a visualization is
+\begin{equation}
+    T: D \rightarrow V
+\end{equation}
+
+Wilkenson proposes  that visualization is  \cite{wilkinsonGrammarGraphics2005, wilkinsonMathematicalFoundationAnalytic2010}
+
+\begin{align}
+\label{eq:gog_data_range}
+G &= {(x, f(x)): x \in R and f(x)=e^{-x^2}}\\
+F &= [-3,3] x [0,1]
+\end{align}
+
+\begin{equation}
+\label{eq:gog_aesthetic_mapping}
+A: x \mapsto x_{position}, f(x) \mapsto y_{position}
+\end{equation}
+
+\begin{equation}
+G_{A} = A(F \cap G) 
+\end{equation}
+
+\subsection{What is the artist}
+\begin{equation}
+A: D \mapsto CW\\
+\end{equation}
+
+An artist is a function F from X to Y
+that satisfyies properties P1, ... PN
+Step 1: Define X
+step 2: Define Y
+Spefiy Pi in terms of equations.
 
@@
 \end{figure}
+
 \begin{figure}
@@
 \end{figure}
-\cite{wilkinsonGrammarGraphicsStatistics2005}
+
 charts - words 

### 2020-09-24 draft/intro.tex
I don't even know

@@
 
+Wilkenson formulates plotting in terms of varsets, which are sets of variables where variables are:
+\begin{equation}
+V:O\mapsto V    
+\end{equation}
+Wilkenson's data algebra is equivalent to data reshapes and join and therefore \cite{wickhamLayeredGrammarGraphics2010,wilkinsonGrammarGraphics2005}
+
 \subsection{What is the artist}
+Visualization is a two step process where the artists $A$ transforms the data into visual encodings $V$ and then the renderer $R$ transfroms the encodings $V$ to the set of pixels in $P$.
 \begin{equation}
-A: D \mapsto CW\\
+    \label{eq:artist}
+    A: D \mapsto V
+\end{equation}
+
+\begin{equation}
+    \label{eg:renderer}
+    R: V \mapsto P
 \end{equation}
@@
 ggplot - strictly hierarchal components (layers), matplotlib - transforms that mostly happen indepently 
-"designing and producing statistical graphics is not an art" - ties in w/ difference between drawing program and visualization library, preservation of properties of measurement type and observation type
-
-\cite{wickhamGgplot2ElegantGraphics2016}
+"designing and producing statistical graphics is not an art" - ties in w/ difference between drawing program and visualization library, preservation of properties of measurement type and observation type \cite{wickhamGgplot2ElegantGraphics2016}
 

### 2020-09-25 draft/intro.tex
setting up symmetry tables

@@
     \label{eq:artist}
-    A: D \mapsto V
+    A: D \mapsto M
 \end{equation}
+TYPES: A(D) = M
+Spivak A is analogous DT
 
+D is the type  which means it's indexexer K(M/indexer) + V (fiber bundle)
+d is a section of a bundle is the data is the values given keys on m + vars on V (c in C)
+a(d) = M
+
+$M$ is a composition of the visual channels applied as ....
+Is also a CW/simplicial complex/simplacial set
+
+Channels are ducktyping the variable types
+Marks are encoding seperability
+
+Marks are mapping of simplacial complex into RGB space
+
+(find paper on complex marks)
+
+for every key value in M, have coordinates in the fiber (space)
+section: keys -> values
+
+Scatter Plot:
+K - 0D disconnected points
+
+A: (C1, C2, C3...)\mapsto (M....)
+t \in A
+-> preserve type operation and relation
+
+M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order. M is a bunch of disconnected Marks
+
+Line Plot:
+L - 1D, can be a function on an interval [a,b]
+functions in the column...f(key) -> values
+
+invariance: \mapto -> preserves operations/relations types
+K might have symmetry -> move back and forth in time, preserved in K
+symmatry in fibers -> scaled up/down, categorical,
+preserved in the visual variables
+
+groups on K and fibers and subgroups w/ different types and temperatures (Steven's fundemental data types-measurment scales and types)
+
+intervals can transalte, ratio can only scale
+group of fibers has to do w/ measurment types
+channels themselves independent of data have their own symmatery
+
+(make table of what symmateries are preserved)
+visual transformation groups
+
+instead of inputting parameter into input, get to all the other parameters - base thing is a square, 
+preserved: operation(D) <-> operation (M)
+if A(g*d) = A(d)*g
+
+normalizations of scales in mpl - devision usually leaves same picture but axis is affected - visually length got uniformally smaller but we changed ticks to compensate
+Axes has rules about how transform affects it - inside graph invariance is preserved
+
+groups:
+symmteries of K
+symmetries of cartesian product of columns
+each visual channel has symmtery group associated w/
+orbit- space of possible visualiztions by changing the constants of each aesthetic (or f_{red}->f_{green})  
+visual type might be orbit of that (orbit of function in space)->single function that can be transformed in that way
+each visualization type is an orbit & then try to set up equivalence on these...
+
+artist will respect that w/ equivariance, knob on one side will tweak knob on another
+
+\begin{table}
+    \begin{tabular}{|r|r|}
+
+    \end{tabular}
+\caption{Symmetrys one the data preserved with each type of mark\cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
+\label{tab:symmetry_mark}
+\end{table}
+
+\begin{table}
+\begin{tabular}{|r|r|}
+
+\end{tabular}
+\caption{Symmetries on variable columns C \caption{Symmetrys one the data preserved with each type of channel \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
+\label{tab:symmetry_channel}
+\end{table}
+Combining 
 \begin{equation}
     \label{eg:renderer}
-    R: V \mapsto P
+    R: M \mapsto P
 \end{equation}
+Where $P$ is the set of pixels rendered to screen ${p_{i,j} \in P}$
 

### 2020-09-25 draft/intro.tex
clean up latex + remove tables I don't need 'cause found Stevens

@@
 K - 0D disconnected points
+\begin{multline}
+A: (C1, C2, C3...)\mapsto (M....)\\
+t \in A
+\end{multline}
 
-A: (C1, C2, C3...)\mapsto (M....)
-t \in A
 -> preserve type operation and relation
 
-M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order. M is a bunch of disconnected Marks
+M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order.
 
@@
 
-invariance: \mapto -> preserves operations/relations types
+invariance: The symmetry group $A$ preserves invariance when the transforms $a \in A$ are symmetric such that $operation(column) = operation(channel)$ and $operation on data -> operation (mark)$
+
 K might have symmetry -> move back and forth in time, preserved in K
@@
 each visual channel has symmtery group associated w/
-orbit- space of possible visualiztions by changing the constants of each aesthetic (or f_{red}->f_{green})  
+orbit- space of possible visualiztions by changing the constants of each aesthetic (or $f_{red} \rightarrow f_{green}$)  
 visual type might be orbit of that (orbit of function in space)->single function that can be transformed in that way
-each visualization type is an orbit & then try to set up equivalence on these...
+each visualization type is an orbit \& then try to set up equivalence on these...
 
@@
 
+Steven's defines the group
+
+Wiki: Symmetry group is the group of all transformations under which the object is invariant. 
+The aesthetic transfroms $a \in A$ are a symmetry group that preserves transforms:
+
 \begin{table}
+    \label{tab:sym_mark}
     \begin{tabular}{|r|r|}
-
     \end{tabular}
 \caption{Symmetrys one the data preserved with each type of mark\cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
-\label{tab:symmetry_mark}
 \end{table}
@@
 \begin{table}
+    \label{tab:sym_mark}
     \begin{tabular}{|r|r|}
+    \end{tabular}
+\caption{Symmetrys one the data preserved with each type of mark\cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
+\end{table}
 
-\end{tabular}
-\caption{Symmetries on variable columns C \caption{Symmetrys one the data preserved with each type of channel \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
-\label{tab:symmetry_channel}
-\end{table}
+Scatter Plot:
+K - 0D disconnected points
+\begin{multline}
+A: (C1, C2, C3...)\mapsto (M....)\\
+ t \in A 
+\end{multline}
+
+-> preserve type operation and relation
+
+M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order. M is a bunch of disconnected Marks
+
+Line Plot:
+L - 1D, can be a function on an interval [a,b]
+functions in the column...f(key) -> values
+
+Box chart: area + scatter
+
 Combining 

### 2020-09-25 draft/intro.tex
finish out math + tables

@@
 
-Steven's defines the group
+Steven's identifies the group symmetries for the measurement scales\cite{stevensTheoryScalesMeasurement1946} as follows:
+
+\begin{table}
+    \begin{tabular}{l l l l}
+        nominal & permutation group & one to one substitutions & $=, \neq $\\
+        ordinal & isotonic group & monotonic increasing functions & $\leq, \geq$\\
+        interval & general linear group & & $+, -$, matrix multiplication\\
+        ratio & similarity group & & \\
+    \end{tabular}
+\end{table}
 

### 2020-09-29 draft/intro.tex
scale table?

--- 2020-09-25 draft/intro.tex
@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{float}
@@
 
-Miniumum assumption a transform will make - is this a valid transform for the slot?
+Minimum assumption a transform will make - is this a valid transform for the slot?
 
@@
 
-Steven's defines the group
+Steven's measurement scales have the following structure \cite{stevensTheoryScalesMeasurement1946}:
+%the big point of disagreement on the scales seems to be the stats column that we don't care about
+\begin{table}
+    \begin{tabular}{l l l l}
+        nominal & permutation group & one to one substitutions & $=, \neq $\\
+        ordinal & isotonic group & monotonic increasing functions &  $=, \neq $, $\leq, \geq$\\
+        interval & general linear group & matrix operations &  $=, \neq, \leq, \geq, +, -, $\\
+        ratio & similarity group & $ \mathbb{Z}^+ $ & 
+    \end{tabular}
+\end{table}
 
-Wiki: Symmetry group is the group of all transformations under which the object is invariant. 
-The aesthetic transfroms $a \in A$ are a symmetry group that preserves transforms:
+The aesthetic transfroms $a \in A$ are a symmetry group that preserves the group structure of the measurment scales as follows:
 
@@
     \label{tab:sym_mark}
-    \begin{tabular}{|r|r|}
-    \end{tabular}
-\caption{Symmetrys one the data preserved with each type of mark\cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
-\end{table}
-
-\begin{table}
-    \label{tab:sym_mark}
-    \begin{tabular}{|r|r|}
+    \begin{tabular}{|r|r|r|r|r|}
+             & nominal & ordinal & interval & ratio \\
+    position &         &         &          &       \\
+    size     & &         &          &       \\
+    shape    &&         &          &       \\
+    texture  &&         &          &       \\
+    color hue &&         &          &       \\
+    color value &&         &          &       \\
     \end{tabular}
@@
 ggplot - strictly hierarchal components (layers), matplotlib - transforms that mostly happen indepently 
-"designing and producing statistical graphics is not an art" - ties in w/ difference between drawing program and visualization library, preservation of properties of measurement type and observation type \cite{wickhamGgplot2ElegantGraphics2016}
-
+"designing and producing statistical graphics is not an art" - ties in w/ difference between drawing program and visualization library, preservation of properties of measurement type and observation type \cite{wickhamGgplot2ElegantGraphics2016a}
 history of viz - collins, friendly

### 2020-09-30 draft/intro.tex
notes on tau

@@
 
+categorical & shapres have the group structure and so can do 
+a $a: c \mapsto \s$
+Optimal transform $a_{i}$ is one from a measurment space to a visual channel where both sides have the same group transforms-
+because equivaraince requires having the same transforms on the same side.
+
+artists take in class of visuals +
+
+A - matplotlib artists 
+$A: FB \mapsto CW$
+$\tau: fb \mapsto vv $visual variable transformers
+$\{\tau_{1}, \tau_{2}...\tau_{n}}(K)\mapsto {CW_{1}, CW_{2}, ...CW_{n}}$ 
+$\mapsto$ is what embeds the topology k 
+functions that know how to embed the k (mark)
+functions that work on the fibers $\tau$
+
+%% flip this to the otherside so it's an adjacency matrix
+
 \begin{table}

### 2020-10-01 draft/intro.tex
table

--- 2020-09-29 draft/intro.tex
@@
 \usepackage{float}
+\usepackage{diagbox}
 \usepackage{graphicx}
@@
 \section{Notation \& Definitions}
+\label{sec:hunterMatplotlib2DGraphics2007}
 System of transforms:
@@
 K might have symmetry -> move back and forth in time, preserved in K
-symmatry in fibers -> scaled up/down, categorical,
+symmatery in fibers -> scaled up/down, categorical,
 preserved in the visual variables
@@
 visual type might be orbit of that (orbit of function in space)->single function that can be transformed in that way
+
 each visualization type is an orbit \& then try to set up equivalence on these...
@@
 
+categorical \& shapres have the group structure and so can do a $a: c \mapsto s$
+Optimal transform $a_{i}$ is one from a measurment space to a visual channel where both sides have the same group transforms-
+because equivaraince requires having the same transforms on the same side.
+
+artists take in class of visuals +
+
+A - matplotlib artists 
+$A: D \mapsto CW$
+
+$\tau: fb \mapsto vv $visual variable transformers
+$\{\tau_{1}, \tau_{2}...\tau_{n}\}(K)\mapsto \{CW_{1}, CW_{2}, ...CW_{n}\}$ 
+$\mapsto$ is what embeds the topology k 
+functions that know how to embed the k (mark)
+functions that work on the fibers $\tau$
+
+%% distance along edge to transformed physical space, s function 
+%% colormap_nrom - t: d \mapsto RGBA, continouts t: d(s) \mapsto 
+%% 1d has to be parameterized  t: d(s), really want t: d\ mapsto RGBA
+%%  d(s) \mapsto {d....} on interval [s_{0}..s_{n}]
+%% flip this to the otherside so it's an adjacency matrix
+%% how do we represent invariance/equivariance  A: D \mapsto \I
+%% D - columns + topology (section + topology)
+%% artist is the visual idiom 
+%% {D cartesian product Tau} \mapsto I
+%% Artist already has everything in VV space and composing into idiom
+%% functions on simplex crossed wuth transformers - everything in visual space w/ same topology
+%% A: visual \mapsto idiom
+%% Artist (raw data + transforms )
+%% carry topology over from VV space to I space -> singular CW complex 
+%% (has no inherent structure) 
+%% invaraince - shuffle_sq on tick labels = shuffle_sq on data 
+%% how do you preserve invariance on topology->Idiom?
+%% \tau_{2} is a new \tau function taht's really tau + 2
+%% taus is a group of transform functions under composition 
+%% tau gos data to visual variable
+%% group of functions + 
+%% measurement space groups is what class of functions don't destroy your data
+%% \tau: d \mapsto vv - group of mapto functions
+%% set of allowable \tau is where symmtery d/v are preserved 
+%% d\v isomorphisms where tau takes you from one to another
+%% d\v is  isomorphisms under group of taus where tau is the permutation group
+%%
+%% math: structure is same as type
+
 \begin{table}
-    \label{tab:sym_mark}
-    \begin{tabular}{|r|r|r|r|r|}
-             & nominal & ordinal & interval & ratio \\
-    position &         &         &          &       \\
-    size     & &         &          &       \\
-    shape    &&         &          &       \\
-    texture  &&         &          &       \\
-    color hue &&         &          &       \\
-    color value &&         &          &       \\
+    \label{tab:taus}
+    \begin{tabular}{l*{6}{c}} 
+                              & position &  size  & shape & texture & color hue & color value  \\
+    nominal                   & $\tau$   & $\tau$ & $\tau$& $\tau$  & $\tau$    & $\tau$       \\
+    ordinal                   & $\tau$   & $\tau$ &       &         &           & $\tau$  \\
+    interval                  & $\tau$   & $\tau$ &       &         &           & $\tau$  \\
+    ratio                     & $\tau$   & $\tau$ &       &         &           & $\tau$ \\  
     \end{tabular}
-\caption{Symmetrys one the data preserved with each type of mark\cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
+\caption{Group of $\{t_{1}....t_{n}\}$ where $tau: d \mapsto v$ preserving equivariance such that $d$ and $v$ are isomorphic.
+$d$ is data of type \{nominal, ordinal, interval, ratio\} and $v$ are the visual channels.
+\cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
 \end{table}
+
+Table~\ref{tab:taus} shows under which $\tau: d maps to v$, $d$ and $v$ can be isomorphisms.
 
@@
 
-\subsection{visual variables}
-Bertin's proposed an organization of visual variables \cite{bertinIIPropertiesGraphic2011}
-
-\subsection{Grammar of Graphics \& ggplot}
-Specifies the end chart (graphic)
-we want to specify the transformations ....
-\begin{figure}
-    \includegraphics{figures/intro/grammar_chart_composition.png}
-    \caption{page 10 (introduction)}
-\end{figure}
-
-\begin{figure}
-    \includegraphics{figures/intro/grammar_example.png}
-    \caption{Wilkenson decomposed a graphic into ...} 
-\end{figure}
-
-charts - words 
-    - instances of much more general objects (geometric primatives)
-    - histogram =/= bar chart
-graphics - statements 
-expliceltly OO
-    1. specification
-        data - operations/computations
-        trans - variable transforms (rank)
-        scale - scale transforms
-        coord - coordinate system
-        element - marks and channels
-        guide - meta elements - (axes, legends, etc)
-    2. assembly - kinda what happens in artist (spec->something that can go to renderer) 
-    3. display - rendering
-
-How is our proposal different? seperation in spec between what's data, aesthetics, and render specific, and what part of the architecture owns those operations
-
-ggplot - strictly hierarchal components (layers), matplotlib - transforms that mostly happen indepently 
-"designing and producing statistical graphics is not an art" - ties in w/ difference between drawing program and visualization library, preservation of properties of measurement type and observation type \cite{wickhamGgplot2ElegantGraphics2016a}
-history of viz - collins, friendly
-
-Wilkenson suggests that the graphics pipeline could be implemented functionally but does not push it in any direction.
-
-ordering: processed before plotted, 
-scales before stats, marks before channels
-\begin{figure}
-\includegraphics[]{figures/intro/gog_pathway.png}
-\caption{Diagram of Grammar of Graphics stages of visualization. Fig 2.2 from chapter 2 of The Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}}
-\label{fig:gog_pathway}
-\end{figure}
-In grammar of graphics, as shown in figure~\ref{fig:gog_pathway} data is extracted from the database one variable at a time with an associated index (primary key).
-The algebra stage joins the variables to become the varset, which is a flat table where the columns are the variables/attributes, each row is an observation/item, and each cell contains a measurement \cite{munznerChDataAbstraction}.
-Wilkenson then describes the constraints of the measurement space for each variable through scales \cite{wilkinsonGrammarGraphics2005} with implicit assumptions of the data constraints.
-After this step, data is transformed computationally in service of the visualization.
-Some of the variables are then mapped into geometric marks (symbols) and are scaled (height, width) as appropriate.
-These mappings are then transformed into the coordinate system of the target graph.
-Then the aesthethic attributes (Bertin's retinal variables, also called channels) such as color or  texture) are applied.
-Finally
-
-Wilkenson's thoughts on coherancy are equivalent to invariance but less? formally stated \dots
-
-"To call these charts meainingul, defenders must falsify specific assumptions of GoG" \cite{wilkinsonMathematicalFoundationAnalytic2010}
-
-gog implementationss:
-\begin{enumerate}
-    \item SPSS nViZn
-    \item tableu
-    \item ggplot
-    \item protoviz\d3
-    \item vega (interactive GoG)
-\end{enumerate}
-
 \printbibliography

### 2020-10-15 draft/intro.tex
:shrugs: dunno what changed

@@
 
-Steven's measurement scales have the following structure \cite{stevensTheoryScalesMeasurement1946}:
-%the big point of disagreement on the scales seems to be the stats column that we don't care about
-\begin{table}
-    \begin{tabular}{l l l l}
-        nominal & permutation group & one to one substitutions & $=, \neq $\\
-        ordinal & isotonic group & monotonic increasing functions &  $=, \neq $, $\leq, \geq$\\
-        interval & general linear group & matrix operations &  $=, \neq, \leq, \geq, +, -, $\\
-        ratio & similarity group & $ \mathbb{Z}^+ $ & 
-    \end{tabular}
-\end{table}
-
 The aesthetic transfroms $a \in A$ are a symmetry group that preserves the group structure of the measurment scales as follows:
@@
 %% math: structure is same as type
+Steven's measurement scales have the following structure \cite{stevensTheoryScalesMeasurement1946}:
+%the big point of disagreement on the scales seems to be the stats column that we don't care about
+\begin{table}
+    \lable{tab:measument_type}
+    \begin{tabular}{l l l l}
+        nominal & permutation group & one to one substitutions & $=, \neq $\\
+        ordinal & isotonic group & monotonic increasing functions &  $=, \neq $, $\leq, \geq$\\
+        interval & general linear group & matrix operations &  $=, \neq, \leq, \geq, +, -, $\\
+        ratio & similarity group & $ \mathbb{Z}^+ $ & 
+    \end{tabular}
+\end{table}
+
+%%% make a second table later of what MPL currently supports
+%%% redraw as graph
+%%% set of all possible taus (operations can possibly be defined on both sides)
+
+Split is discrete/ vs continous measurment types 
+It is possible to write taus to ....
+
+A: (v...v) -> I
+%% tau is when we have data/visual lineups for labeling colorbar \ legend
+%% colorbar vs. legend - continous vs discrete
 
@@
     \begin{tabular}{l*{6}{c}} 
-                              & position &  size  & shape & texture & color hue & color value  \\
-    nominal                   & $\tau$   & $\tau$ & $\tau$& $\tau$  & $\tau$    & $\tau$       \\
-    ordinal                   & $\tau$   & $\tau$ &       &         &           & $\tau$  \\
-    interval                  & $\tau$   & $\tau$ &       &         &           & $\tau$  \\
-    ratio                     & $\tau$   & $\tau$ &       &         &           & $\tau$ \\  
+                              & position      &  size        & shape        & texture     & color \\
+    nominal                   &  1            &  1           & 1            & 1           & 1     \\ 
+    ordinal                   &  1            &  1           & 1            & 1           & 1     \\
+    interval                  &  1            &  1           &              &             & 1     \\   
+    ratio                     &  1            &  1           &              &             & 1     \\  
     \end{tabular}
@@
 
+%%% draw what the function graph looks like for scatter versus linesg
 Scatter Plot:

### 2020-10-16 code/matplottoy/scraps/encoders/core.py
clear canvas

@@ -1,0 +1,24 @@
+
+def transform(single_bertan_variable_series : ArrayLike) -> ArrayLike:
+    ...
+
+def transform(x, y, singles_bertan_variable_series):
+    ...
+
+def transform(M, singles_bertan_verabile_series):
+    ...
+
+class MySN(SimpleNampsace):
+    def __repr_html__(...):
+        ...
+
+    def __init__(self, validate, transform):
+        super().__init__(validate, transform)
+
+colorobj = SimpleNamespace(validate=f1, transfrom=f2)
+colordict = {'validate': f1, 'transfrom': f2}
+
+coleobj = NamedTuple('colorobj', ['validate', 'transform'])
+
+#init/validate = info/meta <- mapping/encoding phase, needs to know if possible
+#draw/transform = get('x') <- step to render /needs the things it needs to render

### 2020-11-02 code/matplottoy/artists/bar.py
getting things working again

@@ -7,7 +7,7 @@ from matplotlib import rcParams
 import matplotlib.collections as mcollections
 import matplotlib.path as mpath
 
-import channels
+from matplottoy.artists import utils
 
 def _make_bars(xvalues, xoffsets, yvalues, yoffsets):
     return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] for 
@@ -15,7 +15,7 @@ def _make_bars(xvalues, xoffsets, yvalues, yoffsets):
 
 class Bar(mcollections.Collection):
     required = {'position', 'length'}
-    optional = {'width', 'floor', 'groups'}
+    optional = {'width', 'floor'}
     def __init__(self, data, transforms, *args, **kwargs):
         """
         Parameters
@@ -33,13 +33,9 @@ class Bar(mcollections.Collection):
         """
   
         assert 'vertex' in data.FB.K['tables']
-        assert Bar.required <= transforms.keys()
-        assert ((transforms.keys()-Bar.required) 
-                            <= Bar.optional) 
-        groups = transforms.pop("groups", [])
-        print(transforms)
-        assert all(tau.validate(data.FB.F[column]) 
-                    for (column, tau) in transforms.values())
+        utils.check_constraints(Bar, transforms)
+        assert utils.validate_transforms(data, transforms)
+
         self.orientation = kwargs.pop('orientation', 'v')
      
         super().__init__(*args, **kwargs)
@@ -73,6 +69,10 @@ class Bar(mcollections.Collection):
         self.set_edgecolors('k')
         super().draw(renderer, *args, **kwargs)
         return      
+
+class MultiBar:
+    pass
+
 
 
 def stacked_bar(ax, data, transforms, orientation):

### 2020-11-02 code/matplottoy/artists/point.py

@@ -7,8 +7,6 @@ from matplotlib import rcParams
 import matplotlib.collections as mcollections
 import matplotlib.path as mpath
 
-import channels
-
 class Point(mcollections.Collection):
     required = {'x', 'y'}
     optional = {'facecolors'} 

### 2020-11-02 code/matplottoy/artists/utils.py

@@ -1,0 +1,20 @@
+def check_constraints(artist, transforms):
+    """
+    required: set
+        Encodings required by the artists
+    optional: set
+        Encodings the artist supports
+    encodings:
+        Set of encodings the datasource is providing
+    """
+    # check required encodings are there
+    if not (artist.required <= transforms.keys()):
+        raise ValueError(f"Required encodings {required}")
+    # check optional encodings 
+    if not ((transforms.keys() - artist.required) 
+            <= artist.optional):
+        raise ValueError(f"Valid optional encodings: {optional}")
+
+def validate_transforms(data, transforms):
+    return all(tau.validate(data.FB.F[column]) 
+               for (column, tau) in transforms.values())

### 2020-11-02 code/matplottoy/data/iris.py

@@ -1,8 +1,8 @@
-import data
-import mtypes
+from matplottoy.data.simplex import FiberBundle
+from matplottoy.encoders import mtypes
 
 class Iris:
-    FB = data.FiberBundle({'tables': ['vertex']},
+    FB = FiberBundle({'tables': ['vertex']},
                      {'species': mtypes.Nominal({'setosa', 'versicolor', 'virginica'}),
                       'sepal_length':  mtypes.IntervalRatio(),
                       'sepal_width': mtypes.IntervalRatio(),
@@ -32,7 +32,7 @@ class FrequencyBar:
         F = {'category': mtypes.Nominal(set(categories)), 
              'count': mtypes.PositiveInteger()}
         
-        self.FB = data.FiberBundle(self.K, F)
+        self.FB = FiberBundle(self.K, F)
 
         self.categories = categories
         self.counts = counts        

### 2020-11-02 code/matplottoy/data/simplex.py

@@ -3,7 +3,7 @@ import itertools
 
 import numpy as np
 
-import mtypes
+from matplottoy.encoders import mtypes
  
 class FiberBundle:
     def __init__(self, K, F):

### 2020-11-02 code/matplottoy/encoders/color.py

@@ -1,4 +1,8 @@
-class IdentityColor:
+import numpy as np
+
+import matplotlib.colors as mcolors
+
+class Identity:
     @staticmethod
     def convert(value):
         values = np.atleast_1d(np.array(value, dtype=object))
@@ -7,8 +11,8 @@ class IdentityColor:
     @staticmethod
     def validate(mtype):
         return all(mcolors.is_color_like(v) for v in mtype.categories)
-
-class NominalColor:
+        
+class Nominal:
     def __init__(self, mapping):
         """goal of init is to store parameters that would otherwise be
         curried higher level function"""

### 2020-11-02 code/matplottoy/encoders/position.py

@@ -1,22 +1,17 @@
 import functools
 
 import numpy as np 
-
-import matplotlib.colors as mcolors
 #taus
 
 #TODO: write atleast_1d wrapper
-class Position:
+class Identity:
     @staticmethod
     def convert(value):
         return value
     @staticmethod
     def validate(mtype):
         return mtype.mtype in ['nominal', 'ordinal', 'interval']
-
-class DataFramePosition:
-    
-class NominalPosition:
+class Nominal:
     def __init__(self, mapping):
         self._mapping = mapping
         
@@ -27,29 +22,3 @@ class NominalPosition:
     def validate(self, value):
         values = np.atleast_1d(np.array(value, dtype=object))
         return True
-
-###write at least1d wrapper
-class IdentityColor:
-    @staticmethod
-    def convert(value):
-        values = np.atleast_1d(np.array(value, dtype=object))
-        return [mcolors.to_rgba(v) for v in values]
-
-    @staticmethod
-    def validate(mtype):
-        return all(mcolors.is_color_like(v) for v in mtype.categories)
-
-class NominalColor:
-    def __init__(self, mapping):
-        """goal of init is to store parameters that would otherwise be
-        curried higher level function"""
-        self._mapping = mapping
-
-    def convert(self, value):
-        values = np.atleast_1d(np.array(value, dtype=object))
-        return [mcolors.to_rgba(self._mapping[v]) for v in values]
-
-    def validate(self, mtype):
-        return (set(mtype.categories) == self._mapping.keys() and
-        all(mcolors.is_color_like(color) for color in self._mapping.values()))
-

### 2020-11-02 draft/intro.tex
math

--- 2020-10-01 draft/intro.tex
@@
 
-\section{Notation \& Definitions}
-\label{sec:hunterMatplotlib2DGraphics2007}
-System of transforms:
-input: data w/ structure on fiber bundle + measurement type
-artist: data -> visualization
-(what structure of input is preserved in output)
-we can provide list of
-artist : topology it preserves
-point/line/face parameters: measurement type properties
-
-ex: 
-Line2D preserves continuity
-Linestyle: categorical
-
-%%define basic properties of measurement of types in a theoretical way or C&P from a paper
-
-Minimum assumption a transform will make - is this a valid transform for the slot?
-
-Set of questions/rules to apply such that see a new artist can be defined which lets us maybe simplify optimizations.
-
-M, Fiber Bundle, Section, operatons on \{data, visual\}, transform data space \& visual space,
-
-topology gives us that m is preserved
-spivak gives us that v is preserved
-
-temperature on globe:
-\begin{equation}
-data = [[t1], [t2_{la1, lo2}]
-fiber = {lat, lon, temp}
-m = the data is 2d continous
-\end{equation}
-artist has implict m
-aesthetic paremeters transform in fiber that preserves type operations \& relations in the fiber
-lat - relative distance + ordering
-
-\subsubsection{Graphical Elements}
-There is a set of transform functions $T$ that maps from the data space $D$ to the visual space composed of geometric marks and aesthetic channels $V$ \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels}.
-We propose that a visualization is
-\begin{equation}
-    T: D \rightarrow V
-\end{equation}
-
-Wilkenson proposes  that visualization is  \cite{wilkinsonGrammarGraphics2005, wilkinsonMathematicalFoundationAnalytic2010}
-
-\begin{align}
-\label{eq:gog_data_range}
-G &= {(x, f(x)): x \in R and f(x)=e^{-x^2}}\\
-F &= [-3,3] x [0,1]
-\end{align}
-
-\begin{equation}
-\label{eq:gog_aesthetic_mapping}
-A: x \mapsto x_{position}, f(x) \mapsto y_{position}
-\end{equation}
-
-\begin{equation}
-G_{A} = A(F \cap G) 
-\end{equation}
-
-Wilkenson formulates plotting in terms of varsets, which are sets of variables where variables are:
-\begin{equation}
-V:O\mapsto V    
-\end{equation}
-Wilkenson's data algebra is equivalent to data reshapes and join and therefore \cite{wickhamLayeredGrammarGraphics2010,wilkinsonGrammarGraphics2005}
-
-\subsection{What is the artist}
-Visualization is a two step process where the artists $A$ transforms the data into visual encodings $V$ and then the renderer $R$ transfroms the encodings $V$ to the set of pixels in $P$.
-\begin{equation}
-    \label{eq:artist}
-    A: D \mapsto M
-\end{equation}
-TYPES: A(D) = M
-Spivak A is analogous DT
-
-D is the type  which means it's indexexer K(M/indexer) + V (fiber bundle)
-d is a section of a bundle is the data is the values given keys on m + vars on V (c in C)
-a(d) = M
-
-$M$ is a composition of the visual channels applied as ....
-Is also a CW/simplicial complex/simplacial set
-
-Channels are ducktyping the variable types
-Marks are encoding seperability
-
-Marks are mapping of simplacial complex into RGB space
-
-(find paper on complex marks)
-
-for every key value in M, have coordinates in the fiber (space)
-section: keys -> values
-
-Scatter Plot:
-K - 0D disconnected points
-\begin{multline}
-A: (C1, C2, C3...)\mapsto (M....)\\
-t \in A
-\end{multline}
-
--> preserve type operation and relation
-
-M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order.
-
-Line Plot:
-L - 1D, can be a function on an interval [a,b]
-functions in the column...f(key) -> values
-
-invariance: The symmetry group $A$ preserves invariance when the transforms $a \in A$ are symmetric such that $operation(column) = operation(channel)$ and $operation on data -> operation (mark)$
-
-K might have symmetry -> move back and forth in time, preserved in K
-symmatery in fibers -> scaled up/down, categorical,
-preserved in the visual variables
-
-groups on K and fibers and subgroups w/ different types and temperatures (Steven's fundemental data types-measurment scales and types)
-
-intervals can transalte, ratio can only scale
-group of fibers has to do w/ measurment types
-channels themselves independent of data have their own symmatery
-
-(make table of what symmateries are preserved)
-visual transformation groups
-
-instead of inputting parameter into input, get to all the other parameters - base thing is a square, 
-preserved: operation(D) <-> operation (M)
-if A(g*d) = A(d)*g
-
-normalizations of scales in mpl - devision usually leaves same picture but axis is affected - visually length got uniformally smaller but we changed ticks to compensate
-Axes has rules about how transform affects it - inside graph invariance is preserved
-
-groups:
-symmteries of K
-symmetries of cartesian product of columns
-each visual channel has symmtery group associated w/
-orbit- space of possible visualiztions by changing the constants of each aesthetic (or $f_{red} \rightarrow f_{green}$)  
-visual type might be orbit of that (orbit of function in space)->single function that can be transformed in that way
-
-each visualization type is an orbit \& then try to set up equivalence on these...
-
-artist will respect that w/ equivariance, knob on one side will tweak knob on another
-
-Steven's measurement scales have the following structure \cite{stevensTheoryScalesMeasurement1946}:
-%the big point of disagreement on the scales seems to be the stats column that we don't care about
-\begin{table}
-    \begin{tabular}{l l l l}
-        nominal & permutation group & one to one substitutions & $=, \neq $\\
-        ordinal & isotonic group & monotonic increasing functions &  $=, \neq $, $\leq, \geq$\\
-        interval & general linear group & matrix operations &  $=, \neq, \leq, \geq, +, -, $\\
-        ratio & similarity group & $ \mathbb{Z}^+ $ & 
-    \end{tabular}
-\end{table}
-
-The aesthetic transfroms $a \in A$ are a symmetry group that preserves the group structure of the measurment scales as follows:
-
-categorical \& shapres have the group structure and so can do a $a: c \mapsto s$
-Optimal transform $a_{i}$ is one from a measurment space to a visual channel where both sides have the same group transforms-
-because equivaraince requires having the same transforms on the same side.
-
-artists take in class of visuals +
-
-A - matplotlib artists 
-$A: D \mapsto CW$
-
-$\tau: fb \mapsto vv $visual variable transformers
-$\{\tau_{1}, \tau_{2}...\tau_{n}\}(K)\mapsto \{CW_{1}, CW_{2}, ...CW_{n}\}$ 
-$\mapsto$ is what embeds the topology k 
-functions that know how to embed the k (mark)
-functions that work on the fibers $\tau$
-
-%% distance along edge to transformed physical space, s function 
-%% colormap_nrom - t: d \mapsto RGBA, continouts t: d(s) \mapsto 
-%% 1d has to be parameterized  t: d(s), really want t: d\ mapsto RGBA
-%%  d(s) \mapsto {d....} on interval [s_{0}..s_{n}]
-%% flip this to the otherside so it's an adjacency matrix
-%% how do we represent invariance/equivariance  A: D \mapsto \I
-%% D - columns + topology (section + topology)
-%% artist is the visual idiom 
-%% {D cartesian product Tau} \mapsto I
-%% Artist already has everything in VV space and composing into idiom
-%% functions on simplex crossed wuth transformers - everything in visual space w/ same topology
-%% A: visual \mapsto idiom
-%% Artist (raw data + transforms )
-%% carry topology over from VV space to I space -> singular CW complex 
-%% (has no inherent structure) 
-%% invaraince - shuffle_sq on tick labels = shuffle_sq on data 
-%% how do you preserve invariance on topology->Idiom?
-%% \tau_{2} is a new \tau function taht's really tau + 2
-%% taus is a group of transform functions under composition 
-%% tau gos data to visual variable
-%% group of functions + 
-%% measurement space groups is what class of functions don't destroy your data
-%% \tau: d \mapsto vv - group of mapto functions
-%% set of allowable \tau is where symmtery d/v are preserved 
-%% d\v isomorphisms where tau takes you from one to another
-%% d\v is  isomorphisms under group of taus where tau is the permutation group
-%%
-%% math: structure is same as type
-
-\begin{table}
-    \label{tab:taus}
-    \begin{tabular}{l*{6}{c}} 
-                              & position &  size  & shape & texture & color hue & color value  \\
-    nominal                   & $\tau$   & $\tau$ & $\tau$& $\tau$  & $\tau$    & $\tau$       \\
-    ordinal                   & $\tau$   & $\tau$ &       &         &           & $\tau$  \\
-    interval                  & $\tau$   & $\tau$ &       &         &           & $\tau$  \\
-    ratio                     & $\tau$   & $\tau$ &       &         &           & $\tau$ \\  
-    \end{tabular}
-\caption{Group of $\{t_{1}....t_{n}\}$ where $tau: d \mapsto v$ preserving equivariance such that $d$ and $v$ are isomorphic.
-$d$ is data of type \{nominal, ordinal, interval, ratio\} and $v$ are the visual channels.
-\cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
-\end{table}
-
-Table~\ref{tab:taus} shows under which $\tau: d maps to v$, $d$ and $v$ can be isomorphisms.
-
-Scatter Plot:
-K - 0D disconnected points
-\begin{multline}
-A: (C1, C2, C3...)\mapsto (M....)\\
-t \in A 
-\end{multline}
-
--> preserve type operation and relation
-
-M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order. M is a bunch of disconnected Marks
-
-Line Plot:
-L - 1D, can be a function on an interval [a,b]
-functions in the column...f(key) -> values
-
-Box chart: area + scatter
-
-Combining 
-\begin{equation}
-    \label{eg:renderer}
-    R: M \mapsto P
-\end{equation}
-Where $P$ is the set of pixels rendered to screen ${p_{i,j} \in P}$
-
-An artist is a function F from X to Y
-that satisfyies properties P1, ... PN
-Step 1: Define X
-step 2: Define Y
-Spefiy Pi in terms of equations.
-
+\subfile{sections/math}
 \printbibliography

### 2020-11-02 draft/sections/math.tex
math

\section{Notation \& Definitions}
\begin{figure}
    \includegraphics{figures/sections/dar.png}
    \caption{Data with semantic structure (such as tables and images) is mapped to visual encodings (color, position) that are composited into visual idioms (scatter, line) that are rendered into graphics (rasters, vectors). 
\end{figure}

\subsection{Data}

\subsection{Artist}

\subsection{Render}

\label{sec:hunterMatplotlib2DGraphics2007}
System of transforms:
input: data w/ structure on fiber bundle + measurement type
artist: data -> visualization
(what structure of input is preserved in output)
we can provide list of
artist : topology it preserves
point/line/face parameters: measurement type properties

ex: 
Line2D preserves continuity
Linestyle: categorical

%%define basic properties of measurement of types in a theoretical way or C&P from a paper

Minimum assumption a transform will make - is this a valid transform for the slot?

Set of questions/rules to apply such that see a new artist can be defined which lets us maybe simplify optimizations.

M, Fiber Bundle, Section, operatons on \{data, visual\}, transform data space \& visual space,

topology gives us that m is preserved
spivak gives us that v is preserved

temperature on globe:
\begin{equation}
data = [[t1], [t2_{la1, lo2}]
fiber = {lat, lon, temp}
m = the data is 2d continous
\end{equation}
artist has implict m
aesthetic paremeters transform in fiber that preserves type operations \& relations in the fiber
lat - relative distance + ordering

\subsubsection{Graphical Elements}
There is a set of transform functions $T$ that maps from the data space $D$ to the visual space composed of geometric marks and aesthetic channels $V$ \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels}.  We propose that a visualization is 
\begin{equation}
    T: D \rightarrow V
\end{equation}

Wilkenson proposes  that visualization is  \cite{wilkinsonGrammarGraphics2005, wilkinsonMathematicalFoundationAnalytic2010}

\begin{align}
\label{eq:gog_data_range}
G &= {(x, f(x)): x \in R and f(x)=e^{-x^2}}\\
F &= [-3,3] x [0,1]
\end{align}

\begin{equation}
\label{eq:gog_aesthetic_mapping}
A: x \mapsto x_{position}, f(x) \mapsto y_{position}
\end{equation}

\begin{equation}
G_{A} = A(F \cap G) 
\end{equation}

Wilkenson formulates plotting in terms of varsets, which are sets of variables where variables are:
\begin{equation}
V:O\mapsto V    
\end{equation}
Wilkenson's data algebra is equivalent to data reshapes and join and therefore \cite{wickhamLayeredGrammarGraphics2010,wilkinsonGrammarGraphics2005}

\subsection{What is the artist}
Visualization is a two step process where the artists $A$ transforms the data into visual encodings $V$ and then the renderer $R$ transfroms the encodings $V$ to the set of pixels in $P$.
\begin{equation}
    \label{eq:artist}
    A: D \mapsto M
\end{equation}
TYPES: A(D) = M
Spivak A is analogous DT

D is the type  which means it's indexexer K(M/indexer) + V (fiber bundle)
d is a section of a bundle is the data is the values given keys on m + vars on V (c in C)
a(d) = M

$M$ is a composition of the visual channels applied as ....
Is also a CW/simplicial complex/simplacial set

Channels are ducktyping the variable types
Marks are encoding seperability

Marks are mapping of simplacial complex into RGB space

(find paper on complex marks)

for every key value in M, have coordinates in the fiber (space)
section: keys -> values

Scatter Plot:
K - 0D disconnected points
\begin{multline}
A: (C1, C2, C3...)\mapsto (M....)\\
t \in A
\end{multline}

-> preserve type operation and relation

M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order.

Line Plot:
L - 1D, can be a function on an interval [a,b]
functions in the column...f(key) -> values

invariance: The symmetry group $A$ preserves invariance when the transforms $a \in A$ are symmetric such that $operation(column) = operation(channel)$ and $operation on data -> operation (mark)$

K might have symmetry -> move back and forth in time, preserved in K
symmatery in fibers -> scaled up/down, categorical,
preserved in the visual variables

groups on K and fibers and subgroups w/ different types and temperatures (Steven's fundemental data types-measurment scales and types)

intervals can transalte, ratio can only scale
group of fibers has to do w/ measurment types
channels themselves independent of data have their own symmatery

(make table of what symmateries are preserved)
visual transformation groups

instead of inputting parameter into input, get to all the other parameters - base thing is a square, 
preserved: operation(D) <-> operation (M)
if A(g*d) = A(d)*g

normalizations of scales in mpl - devision usually leaves same picture but axis is affected - visually length got uniformally smaller but we changed ticks to compensate
Axes has rules about how transform affects it - inside graph invariance is preserved

groups:
symmteries of K
symmetries of cartesian product of columns
each visual channel has symmtery group associated w/
orbit- space of possible visualiztions by changing the constants of each aesthetic (or $f_{red} \rightarrow f_{green}$)  
visual type might be orbit of that (orbit of function in space)->single function that can be transformed in that way

each visualization type is an orbit \& then try to set up equivalence on these...

artist will respect that w/ equivariance, knob on one side will tweak knob on another

The aesthetic transfroms $a \in A$ are a symmetry group that preserves the group structure of the measurment scales as follows:

categorical \& shapres have the group structure and so can do a $a: c \mapsto s$
Optimal transform $a_{i}$ is one from a measurment space to a visual channel where both sides have the same group transforms-
because equivaraince requires having the same transforms on the same side.

artists take in class of visuals +

A - matplotlib artists 
$A: D \mapsto CW$

$\tau: fb \mapsto vv $visual variable transformers
$\{\tau_{1}, \tau_{2}...\tau_{n}\}(K)\mapsto \{CW_{1}, CW_{2}, ...CW_{n}\}$ 
$\mapsto$ is what embeds the topology k 
functions that know how to embed the k (mark)
functions that work on the fibers $\tau$

%% distance along edge to transformed physical space, s function 
%% colormap_nrom - t: d \mapsto RGBA, continouts t: d(s) \mapsto 
%% 1d has to be parameterized  t: d(s), really want t: d\ mapsto RGBA
%%  d(s) \mapsto {d....} on interval [s_{0}..s_{n}]
%% flip this to the otherside so it's an adjacency matrix
%% how do we represent invariance/equivariance  A: D \mapsto \I
%% D - columns + topology (section + topology)
%% artist is the visual idiom 
%% {D cartesian product Tau} \mapsto I
%% Artist already has everything in VV space and composing into idiom
%% functions on simplex crossed wuth transformers - everything in visual space w/ same topology
%% A: visual \mapsto idiom
%% Artist (raw data + transforms )
%% carry topology over from VV space to I space -> singular CW complex 
%% (has no inherent structure) 
%% invaraince - shuffle_sq on tick labels = shuffle_sq on data 
%% how do you preserve invariance on topology->Idiom?
%% \tau_{2} is a new \tau function taht's really tau + 2
%% taus is a group of transform functions under composition 
%% tau gos data to visual variable
%% group of functions + 
%% measurement space groups is what class of functions don't destroy your data
%% \tau: d \mapsto vv - group of mapto functions
%% set of allowable \tau is where symmtery d/v are preserved 
%% d\v isomorphisms where tau takes you from one to another
%% d\v is  isomorphisms under group of taus where tau is the permutation group
%%
%% math: structure is same as type
Steven's measurement scales have the following structure \cite{stevensTheoryScalesMeasurement1946}:
%the big point of disagreement on the scales seems to be the stats column that we don't care about
\begin{table}
    \lable{tab:measument_type}
    \begin{tabular}{l l l l}
        nominal & permutation group & one to one substitutions & $=, \neq $\\
        ordinal & isotonic group & monotonic increasing functions &  $=, \neq $, $\leq, \geq$\\
        interval & general linear group & matrix operations &  $=, \neq, \leq, \geq, +, -, $\\
        ratio & similarity group & $ \mathbb{Z}^+ $ & 
    \end{tabular}
\end{table}

%%% make a second table later of what MPL currently supports
%%% redraw as graph
%%% set of all possible taus (operations can possibly be defined on both sides)

Split is discrete/ vs continous measurment types 
It is possible to write taus to ....

A: (v...v) -> I
%% tau is when we have data/visual lineups for labeling colorbar \ legend
%% colorbar vs. legend - continous vs discrete

\begin{table}
    \label{tab:taus}
    \begin{tabular}{l*{6}{c}} 
                              & position      &  size        & shape        & texture     & color \\
    nominal                   &  1            &  1           & 1            & 1           & 1     \\ 
    ordinal                   &  1            &  1           & 1            & 1           & 1     \\
    interval                  &  1            &  1           &              &             & 1     \\   
    ratio                     &  1            &  1           &              &             & 1     \\  
    \end{tabular}
\caption{Group of $\{t_{1}....t_{n}\}$ where $tau: d \mapsto v$ preserving equivariance such that $d$ and $v$ are isomorphic. $d$ is data of type \{nominal, ordinal, interval, ratio\} and $v$ are the visual channels. \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
\end{table}

Table~\ref{tab:taus} shows under which $\tau: d maps to v$, $d$ and $v$ can be isomorphisms.

%%% draw what the function graph looks like for scatter versus linesg
Scatter Plot:
K - 0D disconnected points
\begin{multline}
A: (C1, C2, C3...)\mapsto (M....)\\
t \in A 
\end{multline}

-> preserve type operation and relation

M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order. M is a bunch of disconnected Marks

Line Plot:
L - 1D, can be a function on an interval [a,b]
functions in the column...f(key) -> values

Box chart: area + scatter

Combining 
\begin{equation}
    \label{eg:renderer}
    R: M \mapsto P
\end{equation}
Where $P$ is the set of pixels rendered to screen ${p_{i,j} \in P}$

An artist is a function F from X to Y
that satisfyies properties P1, ... PN
Step 1: Define X
step 2: Define Y
Spefiy Pi in terms of equations.

### 2020-11-03 draft/sections/math.tex
i dunno notation for cw complex

@@
+\documentclass[../intro.tex]{subfiles}
+\begin{document}
 \section{Notation \& Definitions}
@@
     \includegraphics{figures/sections/dar.png}
-    \caption{Data with semantic structure (such as tables and images) is mapped to visual encodings (color, position) that are composited into visual idioms (scatter, line) that are rendered into graphics (rasters, vectors).
+    \caption{Data with semantic structure (such as tables and images) is mapped to visual encodings (color, position) that are composited into visual idioms (scatter, line) that are rendered into graphics (rasters, vectors).}
+    \label{fig:artists}
 \end{figure}
+We propose that the mapping of a precise subset of data to a visual idiom ($I$) can be encapsulated in a class of functors called artists.
+\begin{equation}
+    A: \Gamma(V) \mapsto \{x \in \mathbb{R}^{7}\}
+\end{equation}
+%% cw versus list of pixels for the connectivity?
+
+Figure~\ref{fig:artists} shows transformation from data into rendered graphical object, wherein there is an intermediate artist stage where the data gets transformed into a visual form.
+We argue that a faithful representation of the data is one where:
+invariance:
+functor
+* types: variables measurement groups are  with visual encoding groups
+* topology: visual idioms preserve the topology/connectivity of the data
 
 \subsection{Data}
+We use fiberbundles as our underlying data representation because it concisely encodes topology\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} and type \cite{spivakSIMPLICIALDATABASES}.
+Butler has
 
-\subsection{Artist}
+For example, the iris dataset \cite{UCIMachineLearning} has a topology of disconnected points (k=0) and variables of type nominal (species type), and ratio (sepal and petal length and width).
+To encode this in a formal way, we rely on Butler's
 
-\subsection{Render}
+Types and topology are (statement is line of equation)
 
-\label{sec:hunterMatplotlib2DGraphics2007}
-System of transforms:
-input: data w/ structure on fiber bundle + measurement type
-artist: data -> visualization
-(what structure of input is preserved in output)
-we can provide list of
-artist : topology it preserves
-point/line/face parameters: measurement type properties
-
-ex: 
-Line2D preserves continuity
-Linestyle: categorical
-
-%%define basic properties of measurement of types in a theoretical way or C&P from a paper
-
-Minimum assumption a transform will make - is this a valid transform for the slot?
-
-Set of questions/rules to apply such that see a new artist can be defined which lets us maybe simplify optimizations.
-
-M, Fiber Bundle, Section, operatons on \{data, visual\}, transform data space \& visual space,
-
-topology gives us that m is preserved
-spivak gives us that v is preserved
-
-temperature on globe:
-\begin{equation}
-data = [[t1], [t2_{la1, lo2}]
-fiber = {lat, lon, temp}
-m = the data is 2d continous
-\end{equation}
-artist has implict m
-aesthetic paremeters transform in fiber that preserves type operations \& relations in the fiber
-lat - relative distance + ordering
-
-\subsubsection{Graphical Elements}
-There is a set of transform functions $T$ that maps from the data space $D$ to the visual space composed of geometric marks and aesthetic channels $V$ \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels}.
-We propose that a visualization is
-\begin{equation}
-    T: D \rightarrow V
-\end{equation}
-
-Wilkenson proposes  that visualization is  \cite{wilkinsonGrammarGraphics2005, wilkinsonMathematicalFoundationAnalytic2010}
-
-\begin{align}
-\label{eq:gog_data_range}
-G &= {(x, f(x)): x \in R and f(x)=e^{-x^2}}\\
-F &= [-3,3] x [0,1]
-\end{align}
-
-\begin{equation}
-\label{eq:gog_aesthetic_mapping}
-A: x \mapsto x_{position}, f(x) \mapsto y_{position}
-\end{equation}
-
-\begin{equation}
-G_{A} = A(F \cap G) 
-\end{equation}
-
-Wilkenson formulates plotting in terms of varsets, which are sets of variables where variables are:
-\begin{equation}
-V:O\mapsto V    
-\end{equation}
-Wilkenson's data algebra is equivalent to data reshapes and join and therefore \cite{wickhamLayeredGrammarGraphics2010,wilkinsonGrammarGraphics2005}
-
-\subsection{What is the artist}
-Visualization is a two step process where the artists $A$ transforms the data into visual encodings $V$ and then the renderer $R$ transfroms the encodings $V$ to the set of pixels in $P$.
-\begin{equation}
-    \label{eq:artist}
-    A: D \mapsto M
-\end{equation}
-TYPES: A(D) = M
-Spivak A is analogous DT
-
-D is the type  which means it's indexexer K(M/indexer) + V (fiber bundle)
-d is a section of a bundle is the data is the values given keys on m + vars on V (c in C)
-a(d) = M
-
-$M$ is a composition of the visual channels applied as ....
-Is also a CW/simplicial complex/simplacial set
-
-Channels are ducktyping the variable types
-Marks are encoding seperability
-
-Marks are mapping of simplacial complex into RGB space
-
-(find paper on complex marks)
-
-for every key value in M, have coordinates in the fiber (space)
-section: keys -> values
-
-Scatter Plot:
-K - 0D disconnected points
-\begin{multline}
-A: (C1, C2, C3...)\mapsto (M....)\\
-t \in A
-\end{multline}
-
--> preserve type operation and relation
-
-M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order.
-
-Line Plot:
-L - 1D, can be a function on an interval [a,b]
-functions in the column...f(key) -> values
-
-invariance: The symmetry group $A$ preserves invariance when the transforms $a \in A$ are symmetric such that $operation(column) = operation(channel)$ and $operation on data -> operation (mark)$
-
-K might have symmetry -> move back and forth in time, preserved in K
-symmatery in fibers -> scaled up/down, categorical,
-preserved in the visual variables
-
-groups on K and fibers and subgroups w/ different types and temperatures (Steven's fundemental data types-measurment scales and types)
-
-intervals can transalte, ratio can only scale
-group of fibers has to do w/ measurment types
-channels themselves independent of data have their own symmatery
-
-(make table of what symmateries are preserved)
-visual transformation groups
-
-instead of inputting parameter into input, get to all the other parameters - base thing is a square, 
-preserved: operation(D) <-> operation (M)
-if A(g*d) = A(d)*g
-
-normalizations of scales in mpl - devision usually leaves same picture but axis is affected - visually length got uniformally smaller but we changed ticks to compensate
-Axes has rules about how transform affects it - inside graph invariance is preserved
-
-groups:
-symmteries of K
-symmetries of cartesian product of columns
-each visual channel has symmtery group associated w/
-orbit- space of possible visualiztions by changing the constants of each aesthetic (or $f_{red} \rightarrow f_{green}$)  
-visual type might be orbit of that (orbit of function in space)->single function that can be transformed in that way
-
-each visualization type is an orbit \& then try to set up equivalence on these...
-
-artist will respect that w/ equivariance, knob on one side will tweak knob on another
-
-The aesthetic transfroms $a \in A$ are a symmetry group that preserves the group structure of the measurment scales as follows:
-
-categorical \& shapres have the group structure and so can do a $a: c \mapsto s$
-Optimal transform $a_{i}$ is one from a measurment space to a visual channel where both sides have the same group transforms-
-because equivaraince requires having the same transforms on the same side.
-
-artists take in class of visuals +
-
-A - matplotlib artists 
-$A: D \mapsto CW$
-
-$\tau: fb \mapsto vv $visual variable transformers
-$\{\tau_{1}, \tau_{2}...\tau_{n}\}(K)\mapsto \{CW_{1}, CW_{2}, ...CW_{n}\}$ 
-$\mapsto$ is what embeds the topology k 
-functions that know how to embed the k (mark)
-functions that work on the fibers $\tau$
-
-%% distance along edge to transformed physical space, s function 
-%% colormap_nrom - t: d \mapsto RGBA, continouts t: d(s) \mapsto 
-%% 1d has to be parameterized  t: d(s), really want t: d\ mapsto RGBA
-%%  d(s) \mapsto {d....} on interval [s_{0}..s_{n}]
-%% flip this to the otherside so it's an adjacency matrix
-%% how do we represent invariance/equivariance  A: D \mapsto \I
-%% D - columns + topology (section + topology)
-%% artist is the visual idiom 
-%% {D cartesian product Tau} \mapsto I
-%% Artist already has everything in VV space and composing into idiom
-%% functions on simplex crossed wuth transformers - everything in visual space w/ same topology
-%% A: visual \mapsto idiom
-%% Artist (raw data + transforms )
-%% carry topology over from VV space to I space -> singular CW complex 
-%% (has no inherent structure) 
-%% invaraince - shuffle_sq on tick labels = shuffle_sq on data 
-%% how do you preserve invariance on topology->Idiom?
-%% \tau_{2} is a new \tau function taht's really tau + 2
-%% taus is a group of transform functions under composition 
-%% tau gos data to visual variable
-%% group of functions + 
-%% measurement space groups is what class of functions don't destroy your data
-%% \tau: d \mapsto vv - group of mapto functions
-%% set of allowable \tau is where symmtery d/v are preserved 
-%% d\v isomorphisms where tau takes you from one to another
-%% d\v is  isomorphisms under group of taus where tau is the permutation group
-%%
-%% math: structure is same as type
-Steven's measurement scales have the following structure \cite{stevensTheoryScalesMeasurement1946}:
-%the big point of disagreement on the scales seems to be the stats column that we don't care about
-\begin{table}
-    \lable{tab:measument_type}
-    \begin{tabular}{l l l l}
-        nominal & permutation group & one to one substitutions & $=, \neq $\\
-        ordinal & isotonic group & monotonic increasing functions &  $=, \neq $, $\leq, \geq$\\
-        interval & general linear group & matrix operations &  $=, \neq, \leq, \geq, +, -, $\\
-        ratio & similarity group & $ \mathbb{Z}^+ $ & 
-    \end{tabular}
-\end{table}
-
-%%% make a second table later of what MPL currently supports
-%%% redraw as graph
-%%% set of all possible taus (operations can possibly be defined on both sides)
-
-Split is discrete/ vs continous measurment types 
-It is possible to write taus to ....
-
-A: (v...v) -> I
-%% tau is when we have data/visual lineups for labeling colorbar \ legend
-%% colorbar vs. legend - continous vs discrete
-
-\begin{table}
-    \label{tab:taus}
-    \begin{tabular}{l*{6}{c}} 
-                              & position      &  size        & shape        & texture     & color \\
-    nominal                   &  1            &  1           & 1            & 1           & 1     \\ 
-    ordinal                   &  1            &  1           & 1            & 1           & 1     \\
-    interval                  &  1            &  1           &              &             & 1     \\   
-    ratio                     &  1            &  1           &              &             & 1     \\  
-    \end{tabular}
-\caption{Group of $\{t_{1}....t_{n}\}$ where $tau: d \mapsto v$ preserving equivariance such that $d$ and $v$ are isomorphic.
-$d$ is data of type \{nominal, ordinal, interval, ratio\} and $v$ are the visual channels.
-\cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}}
-\end{table}
-
-Table~\ref{tab:taus} shows under which $\tau: d maps to v$, $d$ and $v$ can be isomorphisms.
-
-%%% draw what the function graph looks like for scatter versus linesg
-Scatter Plot:
-K - 0D disconnected points
-\begin{multline}
-A: (C1, C2, C3...)\mapsto (M....)\\
-t \in A 
-\end{multline}
-
--> preserve type operation and relation
-
-M is the disjoint collection of the output of all these transformers where the transformers preserve the relation and order. M is a bunch of disconnected Marks
-
-Line Plot:
-L - 1D, can be a function on an interval [a,b]
-functions in the column...f(key) -> values
-
-Box chart: area + scatter
-
-Combining 
-\begin{equation}
-    \label{eg:renderer}
-    R: M \mapsto P
-\end{equation}
-Where $P$ is the set of pixels rendered to screen ${p_{i,j} \in P}$
-
-An artist is a function F from X to Y
-that satisfyies properties P1, ... PN
-Step 1: Define X
-step 2: Define Y
-Spefiy Pi in terms of equations.
+\end{document}

### 2020-11-03 code/matplottoy/artists/bar.py

@@ -71,8 +71,62 @@ class Bar(mcollections.Collection):
         return      
 
 class MultiBar:
-    pass
+class Bar(mcollections.Collection):
+    required = {'position', 'length'}
+    optional = {'width', 'floor'}
+    def __init__(self, data, transforms, *args, **kwargs):
+        """
+        Parameters
+        ----------
+        datasource: matplottoy.datasources.DataSource like
+            data object containing data to be plotted
+        stacked: bool, optional
+            True: bar chart is stacked (groups are added)
+            False: bar chart is unstacked (groups are interspersed)
+        orientation: str, optional
+            vertical: bars aligned along x axis, heights on y
+            horizontal: bars aligned along y axis, heights on x
+        **kwargs:
+            kwargs passed through 
+        """
+  
+        assert 'vertex' in data.FB.K['tables']
+        utils.check_constraints(Bar, transforms)
+        assert utils.validate_transforms(data, transforms)
 
+        self.orientation = kwargs.pop('orientation', 'v')
+     
+        super().__init__(*args, **kwargs)
+        self.data = data
+        self.transforms = transforms
+
+        
+    
+    def draw(self, renderer, *args, **kwargs):
+       
+        view = self.data.view('vertex')
+        visual = dict([(t, tau.convert(view[var]))
+            for (t, (var, tau)) in self.transforms.items()])
+
+
+        
+        if 'width' not in visual:
+            visual['width'] = itertools.repeat(0.8)
+        
+      
+            visual['floor'] = itertools.repeat(0)
+     
+        if self.orientation in {'vertical', 'v'}:
+            verts = _make_bars(visual['position'], visual['width'], 
+                              visual['floor'], visual['length'])
+        elif self.orientation in {'horizontal', 'h'}:
+            verts = _make_bars(visual['floor'], visual['length'], 
+                               visual['position'], visual['width'])
+
+        self._paths = [mpath.Path(xy, closed=True) for xy in verts]
+        self.set_edgecolors('k')
+        super().draw(renderer, *args, **kwargs)
+        return      
 
 
 def stacked_bar(ax, data, transforms, orientation):

### 2020-11-03 code/matplottoy/artists/line.py

@@ -21,6 +21,7 @@ class Line(mcollections.LineCollection):
         """
         super().__init__(None, *args, **kwargs)
         assert 'edge' in data.FB.K['tables']
+        utils.check_
         assert Line.required <= transforms.keys()
         assert ((transforms.keys()-Line.required) 
                             <= Line.optional) 

### 2020-11-04 draft/sections/math.tex
groups are a mess

@@
 \end{figure}
-We propose that the mapping of a precise subset of data to a visual idiom ($I$) can be encapsulated in a class of functors called artists.
-\begin{equation}
-    A: \Gamma(V) \mapsto \{x \in \mathbb{R}^{7}\}
-\end{equation}
-%% cw versus list of pixels for the connectivity?
-
 Figure~\ref{fig:artists} shows transformation from data into rendered graphical object, wherein there is an intermediate artist stage where the data gets transformed into a visual form.
@@
 * types: variables measurement groups are  with visual encoding groups
-* topology: visual idioms preserve the topology/connectivity of the data
+* topology: visual idioms preserve the topology/connectivity of the dat
 
-\subsection{Data}
+We propose that the mapping of a precise subset of data to a visual idiom ($I$) can be encapsulated in a class of functors called artists.
+
+\begin{equation}
+    A: \Gamma(V) \mapsto 
+\end{equation}
+
+%% cw versus list of pixels for the connectivity?
+    %% toplogical spaces allow you to encode connectvity/continuity 
+%% anypoint on the cw complex I can hook into R7 to get back the {color, alpha, position}
+%% has functions that take it into R7 (lookup table for the spec (raster, vector))
+%% vector represention is the CW 
+%% f:T\mapsto \mathbb(R^{7}), T \in CW
+%% this function is never called w/ elements in T
+%% T is embedded in f, is the topology of the thing on the screen (scene graph?)
+%% output of artist has no semantic meaning, is the line representation of the visual idiom 
+%% in what sense are we preserving continuity on dotted lines? 
+\subsection{data}
 We use fiberbundles as our underlying data representation because it concisely encodes topology\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} and type \cite{spivakSIMPLICIALDATABASES}.
-Butler has
-
 For example, the iris dataset \cite{UCIMachineLearning} has a topology of disconnected points (k=0) and variables of type nominal (species type), and ratio (sepal and petal length and width).

### 2020-11-04 code/matplottoy/artists/bar.py

@@ -9,9 +9,13 @@ import matplotlib.path as mpath
 
 from matplottoy.artists import utils
 
-def _make_bars(xvalues, xoffsets, yvalues, yoffsets):
-    return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] for 
-            (x, xo, y, yo) in zip(xvalues, xoffsets, yvalues, yoffsets)]
+def _make_bars(orientation, position, width, floor, length):
+    if orientation in {'vertical', 'v'}:
+        xval, xoff, yval, yoff = position, width, floor, length
+    elif orientation in {'horizontal', 'h'}:
+        xval, xoff, yval, yoff = floor, length, position, width
+    return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
+            for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
 
 class Bar(mcollections.Collection):
     required = {'position', 'length'}
@@ -49,29 +53,22 @@ class Bar(mcollections.Collection):
         view = self.data.view('vertex')
         visual = dict([(t, tau.convert(view[var]))
             for (t, (var, tau)) in self.transforms.items()])
-
-
         
         if 'width' not in visual:
             visual['width'] = itertools.repeat(0.8)
         
-      
+        if 'floor' not in visual:
             visual['floor'] = itertools.repeat(0)
      
-        if self.orientation in {'vertical', 'v'}:
-            verts = _make_bars(visual['position'], visual['width'], 
-                              visual['floor'], visual['length'])
-        elif self.orientation in {'horizontal', 'h'}:
-            verts = _make_bars(visual['floor'], visual['length'], 
-                               visual['position'], visual['width'])
+        verts = _make_bars(self.orientation, visual['position'], 
+                   visual['width'], visual['floor'], visual['length'])
 
         self._paths = [mpath.Path(xy, closed=True) for xy in verts]
         self.set_edgecolors('k')
         super().draw(renderer, *args, **kwargs)
         return      
 
-class MultiBar:
-class Bar(mcollections.Collection):
+class MultiBar(mcollections.Collection):
     required = {'position', 'length'}
     optional = {'width', 'floor'}
     def __init__(self, data, transforms, *args, **kwargs):
@@ -87,59 +84,53 @@ class Bar(mcollections.Collection):
             vertical: bars aligned along x axis, heights on y
             horizontal: bars aligned along y axis, heights on x
[…]

### 2020-11-04 code/matplottoy/artists/utils.py

@@ -1,3 +1,4 @@
+import numpy as np
 def check_constraints(artist, transforms):
     """
     required: set
@@ -7,14 +8,21 @@ def check_constraints(artist, transforms):
     encodings:
         Set of encodings the datasource is providing
     """
+    print(transforms.keys)
     # check required encodings are there
     if not (artist.required <= transforms.keys()):
-        raise ValueError(f"Required encodings {required}")
+        raise ValueError(f"Required encodings {artist.required}")
     # check optional encodings 
     if not ((transforms.keys() - artist.required) 
             <= artist.optional):
-        raise ValueError(f"Valid optional encodings: {optional}")
+        raise ValueError(f"Valid optional encodings: {artist.optional}")
 
 def validate_transforms(data, transforms):
-    return all(tau.validate(data.FB.F[column]) 
-               for (column, tau) in transforms.values())
+     return all(tau.validate(data.FB.F[column]) 
+                for (column, tau) in transforms.values())
+
+def validate_columns(data, columns, tau):
+    return all(tau.validate(data.FB.F[column])
+                  for column in np.atleast_1d(columns))
+
+

### 2020-11-05 code/matplottoy/artists/bar.py
grouped bar w/ fully vectorized color

@@ -3,6 +3,8 @@ import itertools
 
 import numpy as np 
 
+from cycler import cycler
+
 from matplotlib import rcParams
 import matplotlib.collections as mcollections
 import matplotlib.path as mpath
@@ -19,7 +21,7 @@ def _make_bars(orientation, position, width, floor, length):
 
 class Bar(mcollections.Collection):
     required = {'position', 'length'}
-    optional = {'width', 'floor'}
+    optional = {'width', 'floor', 'color'}
     def __init__(self, data, transforms, *args, **kwargs):
         """
         Parameters
@@ -38,7 +40,7 @@ class Bar(mcollections.Collection):
   
         assert 'vertex' in data.FB.K['tables']
         utils.check_constraints(Bar, transforms)
-        assert utils.validate_transforms(data, transforms)
+        utils.validate_transforms(data, transforms)
 
         self.orientation = kwargs.pop('orientation', 'v')
      
@@ -51,8 +53,7 @@ class Bar(mcollections.Collection):
     def draw(self, renderer, *args, **kwargs):
        
         view = self.data.view('vertex')
-        visual = dict([(t, tau.convert(view[var]))
-            for (t, (var, tau)) in self.transforms.items()])
+        visual = utils.convert_transforms(view, self.transforms)  
         
         if 'width' not in visual:
             visual['width'] = itertools.repeat(0.8)
@@ -70,7 +71,7 @@ class Bar(mcollections.Collection):
 
 class MultiBar(mcollections.Collection):
     required = {'position', 'length'}
-    optional = {'width', 'floor'}
+    optional = {'width', 'floor', 'color'}
     def __init__(self, data, transforms, *args, **kwargs):
         """
         Parameters
@@ -87,49 +88,60 @@ class MultiBar(mcollections.Collection):
             kwargs passed through         """
   
         assert 'vertex' in data.FB.K['tables']
-        self.transforms = transforms.copy()
-        utils.check_constraints(MultiBar, self.transforms)
-        length = self.transforms.pop('length') # allowed to be >1
-        assert utils.validate_transforms(data, self.transforms)
-        assert utils.validate_columns(data, length[0], length[-1])
+        utils.check_constraints(MultiBar, transforms)
+        utils.validate_transforms(data, transforms)
+       
 
         self.orientation = kwargs.pop('orientation', 'v')
         self.stacked = kwargs.pop('stacked', False)     
         self.width = kwargs.pop('width', .8)
+       
[…]

### 2020-11-05 code/matplottoy/artists/utils.py

@@ -8,7 +8,6 @@ def check_constraints(artist, transforms):
     encodings:
         Set of encodings the datasource is providing
     """
-    print(transforms.keys)
     # check required encodings are there
     if not (artist.required <= transforms.keys()):
         raise ValueError(f"Required encodings {artist.required}")
@@ -18,11 +17,19 @@ def check_constraints(artist, transforms):
         raise ValueError(f"Valid optional encodings: {artist.optional}")
 
 def validate_transforms(data, transforms):
-     return all(tau.validate(data.FB.F[column]) 
-                for (column, tau) in transforms.values())
+    for column, tau in transforms.values():
+        for col in np.atleast_1d(column):
+            if not tau.validate(data.FB.F[col]):
+                raise ValueError(f"Invalid transform {tau} for {col}")
 
-def validate_columns(data, columns, tau):
-    return all(tau.validate(data.FB.F[column])
-                  for column in np.atleast_1d(columns))
-
-
+def convert_transforms(view, transforms, exclude=None):
+    exclude = [] if exclude is None else exclude
+    visual = {}
+    for (t, (var, tau)) in transforms.items():
+        if t in exclude:
+            continue
+        if isinstance(var, (str, bytes )):
+            visual[t] = tau.convert(view[var])
+        else:
+            visual[t] = [tau.convert(view[v]) for v in var]
+    return visual

### 2020-11-05 code/matplottoy/data/iris.py

@@ -8,10 +8,15 @@ class Iris:
                       'sepal_width': mtypes.IntervalRatio(),
                       'petal_length': mtypes.IntervalRatio(),
                       'petal_width': mtypes.IntervalRatio(),
+                      'sepal_length_color': mtypes.Color(),
+                      'sepal_width_color': mtypes.Color(),
+                      'petal_length_color': mtypes.Color(),
+                      'petal_width_color': mtypes.Color()
                       })
     def __init__(self, dataframe):
         for column in dataframe:
-            assert all(self.FB.F[column].validate(v) for v in dataframe[column])
+            if not all(self.FB.F[column].validate(v) for v in dataframe[column]):
+                raise ValueError(f'{column} has values that are invalid {self.FB.F[column]}')
         
         self.sigma = dataframe.iloc
         self._view = dataframe

### 2020-11-05 code/matplottoy/encoders/color.py

@@ -2,6 +2,8 @@ import numpy as np
 
 import matplotlib.colors as mcolors
 
+from matplottoy.encoders import mtypes
+
 class Identity:
     @staticmethod
     def convert(value):
@@ -10,7 +12,7 @@ class Identity:
 
     @staticmethod
     def validate(mtype):
-        return all(mcolors.is_color_like(v) for v in mtype.categories)
+        return isinstance(mtype, mtypes.Color)
         
 class Nominal:
     def __init__(self, mapping):

### 2020-11-05 code/matplottoy/encoders/mtypes.py

@@ -4,6 +4,7 @@ to check against visual channels
 # if validate overwrites is instance, then we can 
 # use float/int/etc...
 
+import matplotlib.colors as mcolors
 class Nominal: 
     mtype = 'nominal'
     shape = 'scaler'
@@ -11,6 +12,13 @@ class Nominal:
         self.categories = categories
     def validate(self, value):
         return value in self.categories
+
+class Color:
+    mtype = 'nominal'
+    @staticmethod
+    def validate(value):
+        return mcolors.is_color_like(value)
+
 
 class Ordinal:
     mtype = 'ordinal'

### 2020-11-05 code/matplottoy/encoders/position.py

@@ -7,18 +7,27 @@ import numpy as np
 class Identity:
     @staticmethod
     def convert(value):
-        return value
+        return np.array(value)
     @staticmethod
     def validate(mtype):
         return mtype.mtype in ['nominal', 'ordinal', 'interval']
+
+class Log:
+    def __init__(self, base=10):
+        self.base = base
+
+    def convert(self, value):
+        return np.log(value)/np.log(self.base)
+
+    def validate(self, mtype):
+        return True
 class Nominal:
     def __init__(self, mapping):
         self._mapping = mapping
         
     def convert(self, value):
         values = np.atleast_1d(np.array(value, dtype=object))
-        return [self._mapping[v] for v in values]
+        return np.array([self._mapping[v] for v in values])
 
-    def validate(self, value):
-        values = np.atleast_1d(np.array(value, dtype=object))
+    def validate(self, mtype):
         return True

### 2020-11-05 code/matplottoy/artists/line.py
step function missing risers?

@@ -7,7 +7,7 @@ from matplotlib import rcParams
 import matplotlib.collections as mcollections
 import matplotlib.path as mpath
 
-import channels
+from matplottoy.artists import utils
 
 class Line(mcollections.LineCollection):
     required = {'x', 'y'}
@@ -21,12 +21,8 @@ class Line(mcollections.LineCollection):
         """
         super().__init__(None, *args, **kwargs)
         assert 'edge' in data.FB.K['tables']
-        utils.check_
-        assert Line.required <= transforms.keys()
-        assert ((transforms.keys()-Line.required) 
-                            <= Line.optional) 
-        assert all(tau.validate(data.FB.F[column]) 
-                    for (column, tau) in transforms.values())
+        utils.check_constraints(Line, transforms)
+        utils.validate_transforms(data, transforms)
      
         self.data = data
         self.transforms = transforms
@@ -34,9 +30,7 @@ class Line(mcollections.LineCollection):
     def draw(self, renderer, *args, **kwargs):
       
         view = self.data.view('edge')
-        visual = dict([(t, tau.convert(view[var]))
-            for (t, (var, tau)) in self.transforms.items()])
-           
+        visual = utils.convert_transforms(view, self.transforms)
         # dictionary here in place of visual(parameter) function
         if 'facecolor' not in visual:
             visual['facecolor'] = "C0"

### 2020-11-05 code/matplottoy/data/simplex.py

@@ -157,7 +157,6 @@ def is_continuous(vertices):
     # intervals along the 1D number line
     return (adj_mat == np.diag(np.diagonal(adj_mat))).all()
 
-
 class ContinuousLine:
     FB = FiberBundle({'tables': ['vertex','edge']},
                      {'x' : mtypes.IntervalClosed([ 0,4]),
@@ -186,10 +185,53 @@ class ContinuousLine:
         color = self.edges['color'][k](self.distances) 
         if simplex=='vertex':
             color = np.repeat(color, self.num_samples) 
-        return (k, (x, y, color))
-
+            return (k, (x, y, color))
 
         return (k, (self.edges[c][k](self.distances) for c in self.FB.F.keys()))
+
+    def view(self, simplex='edge'):
+        """walk the edge_vertex table to return the edge value
+        """
+        # contuinity test asserted one source, one sink 
+        # sort here on sources (can also sort on sink)
+        table = defaultdict(list)
+        #since intervals lie along number line and are ordered pair neighbors
+        #sort on start or end should yield ordering
+        for (i, (start, end)) in sorted(zip(self.ids, self.vertices), key=lambda v:v[1][0]):
+            table['index'].append(i)
+            for (name, value) in zip(self.FB.F.keys(), self.sigma(i, simplex)[1]):
+                table[name].append(value)
+        
+        if simplex =='vertex':
+            table['index'] = [(k,d) for d in self.distances for k in table['index']]
+            for name in self.FB.F.keys():
+                table[name] = list(itertools.chain.from_iterable(table[name]))
+        return table
+
+
+class DiscontinousLine:
+    FB = FiberBundle({'tables': ['vertex','edge']},
+                     {'x' : mtypes.IntervalClosed([ -1, 13]),
+                      'y':  mtypes.IntervalClosed([-1,3]),
+                      })
+    def __init__(self, edge_table, vertex_table, num_samples=2):
+        self.num_samples = num_samples
+        self.distances = np.linspace(0,1, num_samples)
+        for  (column, functions) in edge_table.items():
+            for f in functions:
+                #check using random point along [0,1]
+                self.FB.F[column].validate(f(np.random.random()))
+
+        self.ids = range(len(vertex_table))
+        self.vertices = vertex_table        
+        self.edges = edge_table
+
+
+    def sigma(self, k, simplex='edge'):
+        """this function knows that there are functions on the fibers defined in F"""
[…]

### 2020-11-05 code/matplottoy/encoders/color.py

@@ -12,7 +12,8 @@ class Identity:
 
     @staticmethod
     def validate(mtype):
-        return isinstance(mtype, mtypes.Color)
+        return (isinstance(mtype, mtypes.Color) or 
+               all(mcolors.is_color_like(v) for v in mtype.categories))
         
 class Nominal:
     def __init__(self, mapping):

### 2020-11-06 draft/sections/math.tex
step

@@
 \begin{equation}
-    A: \Gamma(V) \mapsto 
+    A: \Gamma(V) \mapsto f:\T \mapsto R^{7}
 \end{equation}
+diagram here: gamma is a dataframe, t is like xkcd line (clear abstract), R7 is nice pixelated version 
+\begin{definition}
+    \item[$F$] fiber, which is the space of all possible values encoded in the fiber bundle
+    \item[$K$] base space, tells how the values in the fibers are connected, is analogous to the indexer in data containers.
+    \item[$V$] $F$ = $K \X F$ total space of the fiber bundle, which is space of all possible values on the index 
+    \item[$\sigma$]
+    \item[$\Gamma$]
+    \item[$f$] analogous to spec (such as svg or raster) that render will rely on to translate image to screen
+    \item[$\T \in CW$] cw simplex representing topology of visual elements on screen (embedded in code rather than explicit) 
+    \item[$R^{7}$]  {R,G,B,A, x, y, z} pixel in display space
+\end{definition}
 
-%% cw versus list of pixels for the connectivity?
-    %% toplogical spaces allow you to encode connectvity/continuity 
-%% anypoint on the cw complex I can hook into R7 to get back the {color, alpha, position}
-%% has functions that take it into R7 (lookup table for the spec (raster, vector))
-%% vector represention is the CW 
-%% f:T\mapsto \mathbb(R^{7}), T \in CW
-%% this function is never called w/ elements in T
-%% T is embedded in f, is the topology of the thing on the screen (scene graph?)
-%% output of artist has no semantic meaning, is the line representation of the visual idiom 
-%% in what sense are we preserving continuity on dotted lines? 
 \subsection{data}
@@
 
+\subsubsection{Data}
+\begin{equation}
+\Gamma(V) = 
+\end{equation}
+
+\subsection{artist}
+%%category theory diagrams to show artist is a functor, which is how you get invariance
+%% artist is parameterized on taus
+%% artists are curried collection of taus , orbits of artists are the visual idiomsW
+\begin{equation}
+    A: \Gamma(V), \Taus \mapsto f:\T \mapsto R^{7}
+\end{equation}
+
+%%%in practuce this has currently same topology as gamma V
+\begin{equation}%%pull in spivak
+    \tau: one type in the fiber \mapsto different type in the T fiber 
+\begin{equation}
+
+\subsection{render}
+\begin{equation}
+    f: \T \mapsto R^{7}
+\end{equation}
+
 \end{document}

### 2020-11-06 code/matplottoy/artists/line.py

@@ -42,3 +42,4 @@ class Line(mcollections.LineCollection):
         self.set_color(visual['facecolor'])
 
         super().draw(renderer, *args, **kwargs)
+

### 2020-11-06 code/matplottoy/data/simplex.py

@@ -214,7 +214,7 @@ class DiscontinousLine:
                      {'x' : mtypes.IntervalClosed([ -1, 13]),
                       'y':  mtypes.IntervalClosed([-1,3]),
                       })
-    def __init__(self, edge_table, vertex_table, num_samples=2):
+    def __init__(self, edge_table, vertex_table, num_samples=2, connect=False):
         self.num_samples = num_samples
         self.distances = np.linspace(0,1, num_samples)
         for  (column, functions) in edge_table.items():
@@ -225,7 +225,7 @@ class DiscontinousLine:
         self.ids = range(len(vertex_table))
         self.vertices = vertex_table        
         self.edges = edge_table
-
+        self.connect = connect
 
     def sigma(self, k, simplex='edge'):
         """this function knows that there are functions on the fibers defined in F"""
@@ -236,7 +236,7 @@ class DiscontinousLine:
     def view(self, simplex='edge'):
         """walk the edge_vertex table to return the edge value
         """
-        # contuinity test asserted one source, one sink 
+        # continuity test asserted one source, one sink 
         # sort here on sources (can also sort on sink)
         table = defaultdict(list)
         #since intervals lie along number line and are ordered pair neighbors
@@ -245,9 +245,10 @@ class DiscontinousLine:
             table['index'].append(i)
             for (name, value) in zip(self.FB.F.keys(), self.sigma(i, simplex)[1]):
                 table[name].append(value)
-        
+
         if simplex =='vertex':
             table['index'] = [(k,d) for d in self.distances for k in table['index']]
             for name in self.FB.F.keys():
                 table[name] = list(itertools.chain.from_iterable(table[name]))
+        
         return table

### 2020-11-06 code/matplottoy/encoders/mtypes.py

@@ -19,7 +19,6 @@ class Color:
     def validate(value):
         return mcolors.is_color_like(value)
 
-
 class Ordinal:
     mtype = 'ordinal'
     def __init__(self, categories):

### 2020-11-11 draft/sections/math.tex
math?

@@
 We propose that the mapping of a precise subset of data to a visual idiom ($I$) can be encapsulated in a class of functors called artists.
+Gluing keys along a base space
+
+| Fiber Bundle | Spivak |
+|--------------|--------|
+| B | K |
+| F | U |
+| embedded in F  | (C, \sigma), \pi |
+| E/V |  |
+|   | \tau:  K-\Gamma(sigma) |
+|   |
 
 \begin{equation}
-    A: \Gamma(V) \mapsto f:\T \mapsto R^{7}
+    A: {\sigma \in \Gamma(V)}\mapsto f:\T \mapsto R^{7}
 \end{equation}
@@
     \item[$F$] fiber, which is the space of all possible values encoded in the fiber bundle
-    \item[$K$] base space, tells how the values in the fibers are connected, is analogous to the indexer in data containers.
+    \item[$K$] base space, tells how the values in the fibers are connected, is analogous to the indexer in data containers
     \item[$V$] $F$ = $K \X F$ total space of the fiber bundle, which is space of all possible values on the index 
-    \item[$\sigma$]
-    \item[$\Gamma$]
+    \item[$\sigma(k)$] section of the fiber at a single k, (record on the  r) (dataframe)
+    \item[$\Gamma(V)$] set of all possible sections (Spivak - $\Gamma(\sigma))$ - is space of all dataframes (we use this in artists to generalize- class of functions/sigmas that the artist can work on)
     \item[$f$] analogous to spec (such as svg or raster) that render will rely on to translate image to screen
@@
 
-\subsection{data}
-We use fiberbundles as our underlying data representation because it concisely encodes topology\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} and type \cite{spivakSIMPLICIALDATABASES}.
-For example, the iris dataset \cite{UCIMachineLearning} has a topology of disconnected points (k=0) and variables of type nominal (species type), and ratio (sepal and petal length and width).
-To encode this in a formal way, we rely on Butler's
-
-Types and topology are (statement is line of equation)
+\begin{equation}
+    A: {\sigma \in \Gamma(V)}\mapsto f:\T \mapsto R^{7}
+\end{equation}
 
 \subsubsection{Data}
+We use Butler's visualization model built on the mathematics of fiber bundles \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} and Spivak's schema like description of fibers \cite{spivakSIMPLICIALDATABASES} as the basis of our data model.
+The components of the fiber bundle are:
+\begin{definition}
+    \item[$K$] base space that describes the topology of the data, is the indexer in data contain
+\end{definition}
+
+\begin{center}
+    \begin{tabular}{ c c c }
+     term & mathematical definition & data container analogue\\
+      $K$ & base space along which the data lies & indexer \\
+      $F$ & fiber space from which possible values are pulled & schema \\
+      $E$ & trivial: $F= K \times F$, non-trivial \dots &
+
+    \end{tabular}
+\end{center}
+
 \begin{equation}
-\Gamma(V) = 
+
 \end{equation}

### 2020-11-12 draft/sections/math.tex
mostly notes from yesterday

@@
 Gluing keys along a base space
-
-| Fiber Bundle | Spivak |
-|--------------|--------|
-| B | K |
-| F | U |
-| embedded in F  | (C, \sigma), \pi |
-| E/V |  |
-|   | \tau:  K-\Gamma(sigma) |
-|   |
 
@@
 The components of the fiber bundle are:
-\begin{definition}
-    \item[$K$] base space that describes the topology of the data, is the indexer in data contain
-\end{definition}
 
@@
     \begin{tabular}{ c c c }
-     term & mathematical definition & data container analogue\\
+     term & mathematical definition & data container analog\\
       $K$ & base space along which the data lies & indexer \\
@@
       $E$ & trivial: $F= K \times F$, non-trivial \dots & 
-
     \end{tabular}
@@
 
-\begin{equation}
-
-\end{equation}
+trivial: 
+locally trivial: take small enough regions, it becomes trivial, cutting it up into simplixes is one way to do so...
+gluing maps are transition functions <- interesting things are in transition functions <- currently writing about pieces, they would write transition function validators and sections that we would check against
+sheaf: association of u (point on eaerth) w/ its section (sheafs associates queries of data fibers over particular u)
+pre-sheaf - just need space of sections for u
+sheaf - if on section of u and section of v and they agree on intersection than their should be a section of union
 
@@
 \begin{equation}
-    A: \Gamma(V), \Taus \mapsto f:\T \mapsto R^{7}
+    A: \sigma \in \Gamma(V), \Taus \mapsto f:\T \mapsto R^{7}
 \end{equation}
-
+$\sigma$ evaluated on $k \in K$
 %%%in practuce this has currently same topology as gamma V
@@
     \tau: one type in the fiber \mapsto different type in the T fiber
-\begin{equation}
 

### 2020-11-12 draft/sections/math.tex
more words

@@
 \subsubsection{Data}
-We use Butler's visualization model built on the mathematics of fiber bundles \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} and Spivak's schema like description of fibers \cite{spivakSIMPLICIALDATABASES} as the basis of our data model.
-The components of the fiber bundle are:
+We are using Butler's visualization model built on the mathematics of fiber bundles \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} and Spivak's schema like description of fibers \cite{spivakSIMPLICIALDATABASES} as the basis of our data model.
+Fiber bundles give us a way to disentangle the connectivity of the data (is it a timeseries, a map?) from the variables in the dataset (temperature).
+Building on Munzner's formulation of the former as keys and the latter as values \cite{munznerWhatDataAbstraction2014}, the data has implicit keys that lie in the $K$ basespace of the fiberbundle, while the labels for the keys (such as latitude and longitude) are variables of the dataset that lie in the fiber $F$.
 
+%%%maybe this comes in a later API section....
 \begin{center}
+    \table{tab:fiberbundle}
     \begin{tabular}{ c c c }
@@
 \end{center}
+
+\begin{figure}
+    \lable{fig:simplex}
+    \includegraphics{figures/sections/math/simplex.png}
+    \caption{Examples of simplexes that can be connected to make up a trivial total space $E = K\times F$.
+    EX: data that both has a vertex & edge - find & visualize}
+\end{figure}
+
+In our data model, we encode the indexer as its simplex representation so that we can keep all variables of the fiber encoded on a simplex together \cite{butlerVisualizationModelBased1989}
 

### 2020-11-17 draft/sections/math.tex
pull through figure for architecture
shred and restart math

--- 2020-11-11 draft/sections/math.tex
@@
 \documentclass[../intro.tex]{subfiles}
+
 \begin{document}
 \section{Notation \& Definitions}
-\begin{figure}
-    \includegraphics{figures/sections/dar.png}
-    \caption{Data with semantic structure (such as tables and images) is mapped to visual encodings (color, position) that are composited into visual idioms (scatter, line) that are rendered into graphics (rasters, vectors).}
-    \label{fig:artists}
-\end{figure}
-Figure~\ref{fig:artists} shows transformation from data into rendered graphical object, wherein there is an intermediate artist stage where the data gets transformed into a visual form.
-We argue that a faithful representation of the data is one where:
-invariance:
-functor
-* types: variables measurement groups are  with visual encoding groups
-* topology: visual idioms preserve the topology/connectivity of the dat
-
-We propose that the mapping of a precise subset of data to a visual idiom ($I$) can be encapsulated in a class of functors called artists.
-Gluing keys along a base space
-
-| Fiber Bundle | Spivak |
-|--------------|--------|
-| B | K |
-| F | U |
-| embedded in F  | (C, \sigma), \pi |
-| E/V |  |
-|   | \tau:  K-\Gamma(sigma) |
-|   |
 
 \begin{equation}
-    A: {\sigma \in \Gamma(V)}\mapsto f:\T \mapsto R^{7}
-\end{equation}
-diagram here: gamma is a dataframe, t is like xkcd line (clear abstract), R7 is nice pixelated version 
-\begin{definition}
-    \item[$F$] fiber, which is the space of all possible values encoded in the fiber bundle
-    \item[$K$] base space, tells how the values in the fibers are connected, is analogous to the indexer in data containers
-    \item[$V$] $F$ = $K \X F$ total space of the fiber bundle, which is space of all possible values on the index 
-    \item[$\sigma(k)$] section of the fiber at a single k, (record on the  r) (dataframe)
-    \item[$\Gamma(V)$] set of all possible sections (Spivak - $\Gamma(\sigma))$ - is space of all dataframes (we use this in artists to generalize- class of functions/sigmas that the artist can work on)
-    \item[$f$] analogous to spec (such as svg or raster) that render will rely on to translate image to screen
-    \item[$\T \in CW$] cw simplex representing topology of visual elements on screen (embedded in code rather than explicit) 
-    \item[$R^{7}$]  {R,G,B,A, x, y, z} pixel in display space
-\end{definition}
-
-\begin{equation}
-    A: {\sigma \in \Gamma(V)}\mapsto f:\T \mapsto R^{7}
+    A: \Gamma(E) \rightarrow \Gamma(S)
 \end{equation}
 
-\subsubsection{Data}
-We use Butler's visualization model built on the mathematics of fiber bundles \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} and Spivak's schema like description of fibers \cite{spivakSIMPLICIALDATABASES} as the basis of our data model.
-The components of the fiber bundle are:
-\begin{definition}
-    \item[$K$] base space that describes the topology of the data, is the indexer in data contain
-\end{definition}
-
-\begin{center}
-    \begin{tabular}{ c c c }
-     term & mathematical definition & data container analogue\\
-      $K$ & base space along which the data lies & indexer \\
-      $F$ & fiber space from which possible values are pulled & schema \\
-      $E$ & trivial: $F= K \times F$, non-trivial \dots &
-
-    \end{tabular}
-\end{center}
-
+%%% mobias figure to show what K space and F space are? 
 \begin{equation}
-
+ E = F \times K
 \end{equation}
 
-\subsection{artist}
-%%category theory diagrams to show artist is a functor, which is how you get invariance
-%% artist is parameterized on taus
-%% artists are curried collection of taus , orbits of artists are the visual idiomsW
-\begin{equation}
-    A: \Gamma(V), \Taus \mapsto f:\T \mapsto R^{7}
-\end{equation}
+\begin{description}
+\item[K] base space, connectivity of observations (discrete, continuous, etc)
+\item[F] fiber space, measurement type/group (nominal, ordinal, interval, ratio)
+\end{description}
 
-%%%in practuce this has currently same topology as gamma V
-\begin{equation}%%pull in spivak
-    \tau: one type in the fiber \mapsto different type in the T fiber 
-\begin{equation}
-
-\subsection{render}
-\begin{equation}
-    f: \T \mapsto R^{7}
-\end{equation}
+\begin{tikzcd}
+    F \arrow[r, hook] E
+    %%     \arrow[d, bend left , "\pi", from=E] & K   
+    %% \arrow[u, bend right, "\sigma", from=K, to = E]
+\end{tikzcd}
 

### 2020-11-18 draft/sections/math.tex
most of the data bit

@@
 \begin{equation}
-    A: \Gamma(E) \rightarrow \Gamma(S)
+    A: \Gamma(E) \rightarrow \Gamma(H)
 \end{equation}
 
+\subsection{Data Space}
 %%% mobias figure to show what K space and F space are?
-\begin{equation}
- E = F \times K
-\end{equation}
 
+We use a fiber bundle model to represent the data, as proposed by Butler \cite{BuildSoftwareBetter}.
+
+\begin{tikzcd}
+    F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right ] \\
+                      & K \arrow[u, "\sigma" description, bend right]
+\end{tikzcd}
+
+%%This should probably be a paragraph
 \begin{description}
@@
     \item[F] fiber space, measurement type/group (nominal, ordinal, interval, ratio)
+    \item[$\pi$] map from total space E to base space K
+    \item[$\sigma$] a map from base space to values in the fiber
+    \item[$\sigma(k)| k \in K $] section of values at the point k in the base space
+    \item[$\Gamma$] space of sections 
 \end{description}
 
+Spivak's \cite{spivakSIMPLICIALDATABASES} formulation of the fibers as a union of named types gives a way to separate the values in the fiber by type:
+
+\begin{figure}[h]
+    \label{fig:simplex}
+    \includegraphics{figures/sections/math/simplex.png}
+    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
+Higher order simplicies are outside the scope of this paper.}
+\end{figure}
+
+One way to represent the topological space K is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
+Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another:
+
+\begin{description}
+    \item[0-simplex] discrete observations (inventory records)
+    \item[1-simplex] 1D continuos data (timeseries)
+    \item[2-simplex] 2D continuos data (map)
+    \item[3-simplex] 3D continuos data (video)
+\end{description}
+
+In a locally trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
+A fiber bundle can be made locally trivial by approximating the total space E as a simplacial complex.
+
+\subsubsection{Subset \& Streaming}
+i for inclusion 
+%% insert diagram of U, E, etc
+
+%%% sheaves & presheaves
+
+\subsection{Visual Space}
+%%render image
 \begin{tikzcd}
-    F \arrow[r, hook] E
-    %%     \arrow[d, bend left , "\pi", from=E] & K   
-    %% \arrow[u, bend right, "\sigma", from=K, to = E]
+    \mathbb{R}^{7} \arrow[r, hook] & H \arrow[d, "\zeta" description, bend right] \\
+                                   & S \arrow[u, "\rho" description, bend right] 
 \end{tikzcd}
 
+%%% R7 in H, S base space, mostly will be trivial case H = S\times R^{7}
+
+%% region on screen that corresponds to selection on H that corresponds to selection on S
+
+\subsection{Artist}
+
+\begin{equation}
+    A: \Gamma(E) \rightarrow \Gamma(H)
+\end{equation}
+
+%%\psi map from S to topological K...
+\subsubsection{Visual Idioms: Equivalance class of artists}
+Visual Idioms
+
+\subsubsection{Visual Variable Invariance}
+Tau can preserves the measurement type properties (group scales)
+
+Will implement in code such that this happens.
+
 \end{document}

### 2020-11-19 draft/sections/math.tex
words around E

@@
 \documentclass[../intro.tex]{subfiles}
+\begin{document}
 
-\begin{document}
 \section{Notation \& Definitions}
@@
 \subsection{Data Space}
-%%% mobias figure to show what K space and F space are?
+\begin{figure}[ht]
+    \label{fig:mobius}
+    \includegraphics[width=0.5\linewidth]{figures/sections/math/mobius.png}
+    \caption{write up some words here}
+\end{figure}
 
-We use a fiber bundle model to represent the data, as proposed by Butler \cite{BuildSoftwareBetter}.
+We use a fiber bundle model to represent the data, as proposed by Butler 
+\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
+
+One example of a fiber bundle is the mobius band shown in figure~\ref{fig:mobius}.
+A fiber bundle is a topological total space $E$ with an embedded fiber space $F$, a base space on which the fibers lie $K$ and the $\pi$ and $\sigma$ mappings between $E$ and $K$.
 
@@
 
-%%This should probably be a paragraph
-\begin{description}
-    \item[K] base space, connectivity of observations (discrete, continuous, etc)
-    \item[F] fiber space, measurement type/group (nominal, ordinal, interval, ratio)
-    \item[$\pi$] map from total space E to base space K
-    \item[$\sigma$] a map from base space to values in the fiber
-    \item[$\sigma(k)| k \in K $] section of values at the point k in the base space
-    \item[$\Gamma$] space of sections 
-\end{description}
+As illustrated by the mobius band example in figure~\ref{fig:mobius}, the vertical lines $F$ are the range of possible values embedded in $E$ and the circle $K$ is a representation of the connectivity of the points in $E$.
+The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$, and the section $\sigma$ is the mapping from a location $k$ on $K$ to a point on $F_{k}$ in $E$.
+The line $\Gamma$ is a space of sections
 
-Spivak's \cite{spivakSIMPLICIALDATABASES} formulation of the fibers as a union of named types gives a way to separate the values in the fiber by type:
+\subsubsection{Base Space $K$}
 
-\begin{figure}[h]
+\begin{figure}[ht]
     \label{fig:simplex}
@@
 
+\subsubsection{Fiber Space $F$}
+The fibers encode the set of all possible values each observation can take. 
+Spivak's \cite{spivakSIMPLICIALDATABASES} formulation of the fibers as a union of named types gives a way to separate the values in the fiber by type:
+
 \subsubsection{Subset \& Streaming}
-i for inclusion 
-%% insert diagram of U, E, etc
+$\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U$ embedded in $K$ with the same fiber $\iota^*E$ and $\iota^*\sigma$.
 
+\begin{tikzcd}
+    \iota^\ast E \arrow[r, hook] \arrow[d] & E \arrow[d]\\
+    U \arrow[r, "\iota" description, hook] \arrow[u, "\iota^\ast \sigma" description, bend right,] & K \arrow[u, "\sigma" description, bend right]
+\end{tikzcd}
 %%% sheaves & presheaves
@@
 %% region on screen that corresponds to selection on H that corresponds to selection on S
+The arrow in figure~\ref{fig:pullback} is an illustration of the lookup function $$f:\tau \rightarrow \mathbb{R}^7$$ returned by $$\mathbb{A}$$.
+The screen (represented by ~\ref{fig:pullback}(d)) queries the topological space $\mathbb{T}$ that underlies the ideal (realworld coordinate free) visual idiom (~\ref{fig:pullback}(c)) of the image.
+This idealized representations toplogy can be encoded as a 3 simplex since screens are ultimaltely 3 dimensional
 
-\subsection{Artist}
+%%A: \Gamma(V) \rightarrow \Gamma(S)
+
+%%f: \S \rightarrow R^{7}
+
+%%make bounding box bigger, pull in more rows (for averaging), bounding box on triangle, table is integral on bounding box - R,G,B, A = $$ \integral f(a,b)$$  (R, G, B, A functions )
+
+draw functions return the renderer - T simplex 
+set segments etc are currying f
+(sampled, would be a table, visualizer can create different triangles/triangles based on renderer, invariances would hold - curremtly it's a png/r7 lookup table)
+
+f to rnedring is lossy process
+S - (face/collection of faces) <- simplicial complex, details encoded many ways dependent on renderer, connectivity is in line or edge (can also show multiple triangles...)
+
+in practice w/ a renderer we have one triangle
+
+(change f to rho)
 

### 2020-11-23 draft/sections/math.tex
stopped at marks

--- 2020-11-18 draft/sections/math.tex
@@
 \subsection{Data Space}
-%%% mobias figure to show what K space and F space are?
+\begin{figure}[ht]
+    \includegraphics[width=0.5\linewidth]{figures/sections/math/mobius.png}
+    \caption{write up some words here}
+    \label{fig:mobius}
+\end{figure}
 
-We use a fiber bundle model to represent the data, as proposed by Butler \cite{BuildSoftwareBetter}.
+We use a fiber bundle model to represent the data, as proposed by Butler 
+\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
+
+One example of a fiber bundle is the mobius band shown in figure~\ref{fig:mobius}.
+A fiber bundle is a topological total space $E$ with an embedded fiber space $F$, a base space on which the fibers lie $K$ and the $\pi$ and $\sigma$ mappings between $E$ and $K$.
+The space of all $\sigma$ is $\Gamma$
 
@@
 
-%%This should probably be a paragraph
-\begin{description}
-    \item[K] base space, connectivity of observations (discrete, continuous, etc)
-    \item[F] fiber space, measurement type/group (nominal, ordinal, interval, ratio)
-    \item[$\pi$] map from total space E to base space K
-    \item[$\sigma$] a map from base space to values in the fiber
-    \item[$\sigma(k)| k \in K $] section of values at the point k in the base space
-    \item[$\Gamma$] space of sections 
-\end{description}
+As illustrated by the mobius band example in figure~\ref{fig:mobius}, the vertical lines $F$ are the range of possible values embedded in $E$.
+The circle $K$ is a representation of the connectivity of the points in $E$.
+The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$.?
+The section $\sigma$ is the mapping from locations $k$ on $K$ to points on $F_{k}$ in $E$.  (Pull this back into more specific about fig, the general is making this more confusing.)
 
-Spivak's \cite{spivakSIMPLICIALDATABASES} formulation of the fibers as a union of named types gives a way to separate the values in the fiber by type:
+\subsubsection{Base Space $K$}
 
-\begin{figure}[h]
+\begin{figure}[ht]
     \label{fig:simplex}
@@
 
+\subsubsection{Fiber Space $F$}
+The fibers encode the set of all possible values each observation can take. 
+Spivak's \cite{spivakSIMPLICIALDATABASES} formulation of the fibers as a union of named types gives a way to separate the values in the fiber by type:
+
 \subsubsection{Subset \& Streaming}
-i for inclusion 
-%% insert diagram of U, E, etc
+$\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U$ embedded in $K$ with the same fiber $\iota^*E$ and $\iota^*\sigma$.
 
+\begin{tikzcd}
+    \iota^\ast E \arrow[r, hook] \arrow[d]                                                                       & E \arrow[d, bend right]                       \\
+    U \arrow[r, "\iota" description, hook] \arrow[u, "\iota^\ast \sigma" description, bend right, shift right=2] & K \arrow[u, "\sigma" description, bend right]
+\end{tikzcd}
 %%% sheaves & presheaves
 
-\subsection{Visual Space}
-%%render image
+\subsection{Display Space}
+\begin{figure}
+    \includegraphics{figures/sections/math/render.png}
+    \caption{}
+    \label{fig:render}
+\end{figure}
+
+A physical display space can be thought of sets of $\mathbb{R}^{7}$ tuples, where 
+\begin{equation}
+    \mathbb{R}^{7} = {X, Y, Z, R, G, B, A}
+\end{equation}
+
+and the sets correspond to the sections on $\S$, which is the topology of the output of the artist $A$.
+The space $H$ is a total space representing the predisplay space, with a fiber of $\mathbb{R}^7$ and a base space of $\S$:
+
 \begin{tikzcd}
-    \mathbb{R}^{7} \arrow[r, hook] & H \arrow[d, "\zeta" description, bend right] \\
+    \mathbb{R}^{7} \arrow[r, hook] & H \arrow[d, "\pi" description, bend right] \\
                                    & S \arrow[u, "\rho" description, bend right] 
@@
 
-%%% R7 in H, S base space, mostly will be trivial case H = S\times R^{7}
+In the case of 2D screens, the predisplay space is a trivial fiber bundle $$H=\mathbb{R}^{7}\times S$$.
+As illustrated in figure~\ref{fig:render}, a region on the screen defined by the corners $$(x_1, y_1)$$ and $$(x_2, y_2)$$ maps into a region on a 2-simplex in $S$ defined by $$(\alpha_1, \beta_1)$$ and $$(\alpha_2, \beta_2)$$.
+The function on the simplex $f$ returns the (R, G, B, A) value for that $(\alpha, \beta)$ pair.
+For a region,
 
-%% region on screen that corresponds to selection on H that corresponds to selection on S
+\begin{equation*}
+\rho(\Gamma(S)) = \int_{\alpha_1}^{\alpha_2}\int_{\beta_1}^{\beta_2}\int_{z_1}^{z_2}{R, G, B, A}  
+\end{equation*}
+
+where the R,G,B,A values are derived from the how the data values are mapped to visual characteristics.
+The z component of the mapping to $$\mathbb{R}^7$$ is moved to the integration because this is a trivial space representing a 2D screen; $\rho$ varies depending on $H$.
 
 \subsection{Artist}
-
 \begin{equation}
@@
 
-%%\psi map from S to topological K...
+\subsubsection{Screen to Data}
+%%diagram: [data] -Tau->[RGVXYZ]
+%%%            \ /\ / 
+\begin{tikzcd}
+    E \arrow[d, "\pi" description]                & H \arrow[d, "\pi" description]                                                 \\
+    K \arrow[u, "\sigma" description, bend right] & S \arrow[l, "\xi" description] \arrow[u, "\rho" description, bend right]
+\end{tikzcd}
+
+The pullback $\xi$ on $S \rightarrow K$ means that the values in $E$ can be directly mapped to a simplex in $S$, which means there's a mapping from screen space back to the values.
+
+\begin{tikzcd}
+    \xi E \arrow[rr, "\tau" description] \arrow[rd, "\xi \sigma" description, bend right] &     & H \arrow[ld] \\
+    & S \arrow[lu, bend right] &             
+    \end{tikzcd}
+
+\subsubsection{Marks}
+
+\subsubsection{Visual Characteristics}
+Tau can preserves the measurement type properties (group scales)
+
 \subsubsection{Visual Idioms: Equivalance class of artists}
@@
 
-\subsubsection{Visual Variable Invariance}
-Tau can preserves the measurement type properties (group scales)
-
-Will implement in code such that this happens.
-
 \end{document}

### 2020-11-23 draft/sections/math.tex
mess of marks

@@
     \label{fig:mobius}
+    %%possibly replace w/ simpler deconstructed band image notes-11-19-20
 \end{figure}
@@
 \begin{equation*}
-\rho(\Gamma(S)) = \int_{\alpha_1}^{\alpha_2}\int_{\beta_1}^{\beta_2}\int_{z_1}^{z_2}{R, G, B, A}  
+\rho(S) = \int_{\alpha_1}^{\alpha_2}\int_{\beta_1}^{\beta_2}\int_{z_1}^{z_2}{R, G, B, A}  
 \end{equation*}
@@
 \begin{tikzcd}
-    \xi E \arrow[rr, "\tau" description] \arrow[rd, "\xi \sigma" description, bend right] &     & H \arrow[ld] \\
+    \xi E \arrow[rr, "\Tau" description] \arrow[rd, "\xi \sigma" description, bend right] &     & H \arrow[ld] \\
     & S \arrow[lu, bend right] &             
@@
 \subsubsection{Marks}
+%% diagram of connected components/line thing
+Bertin defines the physical symbols that on screen represent the data as marks 
+\cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}. For each point on the screen $s$, the mark it belongs to is
+
+%%
+The others fit perfectly.
+So it sounds like for each point s on the screen we can define the "mark it belongs" by mapping it back to K via lookup to S to get s, then \xi(s) back to K, and then take the connected component $J \subset containing k\in J \sub K $ and then taking the inverse image of $J \in S, \xi^(-1)(J)$ and then map into $R^7  \rho(\xi^(-1)(J))$ and project into the x-y plane.
+So thats the mark and the thing the mark "signifies" is J.
 
@@
 \subsubsection{Visual Idioms: Equivalance class of artists}
-Visual Idioms
+Two artists are equivalent when given data containers $\Gamma(E)$ of the same type, they output the same type of prerender $\Gamma(S)$ 
+\begin{tikzcd}
+    A_{\tau_2}: \arrow[d, shift right=2] & \Gamma(E) \arrow[r] & \Gamma(H) &                                                \\
+    A_{\tau_1}: \arrow[u, shift right]   & \Gamma(E) \arrow[r] & \Gamma(H) & {} \arrow[loop, distance=2em, in=305, out=235]
+\end{tikzcd}
 

### 2020-11-23 draft/sections/math.tex
marks prose

--- 2020-11-23 draft/sections/math.tex
@@
 \subsection{Display Space}
-\begin{figure}
-    \includegraphics{figures/sections/math/render.png}
+\label{sec:display}
+\begin{figure}[h]
+    \includegraphics[width=.4\linewidth]{figures/sections/math/render.png}
     \caption{}
@@
 \begin{equation}
-    \mathbb{R}^{7} = {X, Y, Z, R, G, B, A}
+    \mathbb{R}^{7} = \{X, Y, Z, R, G, B, A\}
 \end{equation}
@@
 
-In the case of 2D screens, the predisplay space is a trivial fiber bundle $$H=\mathbb{R}^{7}\times S$$.
-As illustrated in figure~\ref{fig:render}, a region on the screen defined by the corners $$(x_1, y_1)$$ and $$(x_2, y_2)$$ maps into a region on a 2-simplex in $S$ defined by $$(\alpha_1, \beta_1)$$ and $$(\alpha_2, \beta_2)$$.
+In the case of 2D screens, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$.
+As illustrated in figure~\ref{fig:render}, a region on the screen defined by the corners $(x_1, y_1)$ and $(x_2, y_2)$ maps into a region on a 2-simplex in $S$ defined by $(\alpha_1, \beta_1)$ and $(\alpha_2, \beta_2)$.
 The function on the simplex $f$ returns the (R, G, B, A) value for that $(\alpha, \beta)$ pair.
@@
 where the R,G,B,A values are derived from the how the data values are mapped to visual characteristics.
-The z component of the mapping to $$\mathbb{R}^7$$ is moved to the integration because this is a trivial space representing a 2D screen; $\rho$ varies depending on $H$.
+The z component of the mapping to $\mathbb{R}^7$ is moved to the integration because this is a trivial space representing a 2D screen; $\rho$ varies depending on $H$.
 
@@
 \subsubsection{Marks}
+%% diagram of connected components/line thing 11-19-20 notes
+Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space these are marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
+\begin{equation}
+\begin{tikzcd}
+    H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \Gamma \text{ s.t.
+} \Gamma(0)=k \text{ and }\Gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
+\end{tikzcd}
+\label{eq:mark}
+\end{equation}
+
+Each point $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on S described in section~\ref{sec:display} then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
+To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
 
@@
 \subsubsection{Visual Idioms: Equivalance class of artists}
-Visual Idioms
+Two artists are equivalent when given data containers $\Gamma(E)$ of the same type, they output the same type of prerender $\Gamma(S)$ 
+\begin{tikzcd}
+    A_{\tau_2}: \arrow[d, shift right=2] & \Gamma(E) \arrow[r] & \Gamma(H) &                                                \\
+    A_{\tau_1}: \arrow[u, shift right]   & \Gamma(E) \arrow[r] & \Gamma(H) 
+\end{tikzcd}
 

### 2020-11-23 draft/sections/math.tex
start of fibers

@@
 The fibers encode the set of all possible values each observation can take.
-Spivak's \cite{spivakSIMPLICIALDATABASES} formulation of the fibers as a union of named types gives a way to separate the values in the fiber by type:
+Spivak's formulation is that the fibers encode the union of the types of measurements\cite{spivakSIMPLICIALDATABASES}.
+For example, given the section $\Gamma(mobuis strip) =
+\{\sigma_1=\sin, \sigma_2=\cos\}$:
+%%fill in more spivak
+the fiber $F = \{float, float\}$.
+
+%%% mapping from measurement back to type is the basis for gaurateening invariance
 
@@
 \subsubsection{Visual Idioms: Equivalance class of artists}
-Two artists are equivalent when given data containers $\Gamma(E)$ of the same type, they output the same type of prerender $\Gamma(S)$ 
+Two artists are equivalent when given data containers $\Gamma(E)$ of the same type, they output the same type of prerender $\Gamma(S)$:
+\begin{equation}
     \begin{tikzcd}
@@
     \end{tikzcd}
+\end{equation}
 

### 2020-11-24 draft/sections/math.tex
msotly notes on what I need to add/make

@@
 \section{Notation \& Definitions}
+%%% add some words around
 
@@
 
-\subsection{Data Space}
+%% E is a locally trivial fiber bundle over K where K is triangulizable space representing data space & H is a fiber bundle over S which is a simplacial complex of triangles representing visual space and artists in matplotlib will go between the two.
+E is a trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
+
+\subsection{Data Model}
 \begin{figure}[ht]
-    \includegraphics[width=0.5\linewidth]{figures/sections/math/mobius.png}
     \caption{write up some words here}
-    \label{fig:mobius}
+    \label{fig:fiberbundle}
     %%possibly replace w/ simpler deconstructed band image notes-11-19-20
+    %% or replace w/ box w/ thing that's kind of convoluted 
+    %% simpler still replace w/ timeseries of temp, 1D fiber
 \end{figure}
 
+%%% preface w/ we care about topology & types. 
 We use a fiber bundle model to represent the data, as proposed by Butler 
@@
 
-One example of a fiber bundle is the mobius band shown in figure~\ref{fig:mobius}.
+One example of a fiber bundle is shown in figure~\ref{fig:mobius}.
 A fiber bundle is a topological total space $E$ with an embedded fiber space $F$, a base space on which the fibers lie $K$ and the $\pi$ and $\sigma$ mappings between $E$ and $K$.
-The space of all $\sigma$ is $\Gamma$
 
@@
     F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right ] \\
-                      & K \arrow[u, "\sigma" description, bend right]
+                      & K \arrow[u, description, bend right]
 \end{tikzcd}
 
-As illustrated by the mobius band example in figure~\ref{fig:mobius}, the vertical lines $F$ are the range of possible values embedded in $E$.
-The circle $K$ is a representation of the connectivity of the points in $E$.
-The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$.?
-The section $\sigma$ is the mapping from locations $k$ on $K$ to points on $F_{k}$ in $E$.  (Pull this back into more specific about fig, the general is making this more confusing.)
+As illustrated by figure~\ref{fig:mobius}, the vertical lines $F$ are the range of possible values embedded in the total space $E$.
+The topol $K$ is a representation of the connectivity of the points in $E$.
+The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$.
+The section $\sigma$ is the mapping from locations $k$ on $K$ to points on $F_{k}$ in $E$.
+%%% include diagram here of formal definition of sigma
+
+The space of all $\sigma$ is $\Gamma$  (Pull this back into more specific about fig, the general is making this more confusing.)
+%% locally trivial - for every point in K there exists an open set neighborhood $U$ such that if you restrict E to that point, then  \iota*E = F\times U [maybe do mobius strip diagram here?
+U is subsection of the ring, part where it's not twisted.
+there exists a triangulation & a way to do so is simplicies
 
 \subsubsection{Base Space $K$}
+%% general statement about topology, examples of 0, 1, ....Ks [none triangular things, square/torus, ....], condition on K is that it can be represent as a triangulation of K (can be realized as triangle on K - extra structure that may be arbitrary
 
@@
 
+%% k is a triangularizable topological space 
 One way to represent the topological space K is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
-Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another:
+Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another: %%formula for simplacial complexes/simplacial sets
 
+% replace this table w/ a diagram that shows this 
 \begin{description}
@@
 
-In a locally trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
-A fiber bundle can be made locally trivial by approximating the total space E as a simplacial complex.
+I
 
 \subsubsection{Fiber Space $F$}
+\begin{figure}
+    \includegraphics{}
+    \label{fig:movie}
+\end{figure}
+%%% might 
 The fibers encode the set of all possible values each observation can take.
-Spivak's formulation is that the fibers encode the union of the types of measurements\cite{spivakSIMPLICIALDATABASES}.
-For example, given the section $\Gamma(mobuis strip) =
-\{\sigma_1=\sin, \sigma_2=\cos\}$:
-%%fill in more spivak
-the fiber $F = \{float, float\}$.
+For example, the fiber bundle in figure~\ref{fig:movie} is the set of all possible movies, where each $k \in K$ is a frame in the movie.
+Each \sigma
 
-%%% mapping from measurement back to type is the basis for gaurateening invariance
+which in Spivak's formulation means the fibers encode the union of the types of measurements\cite{spivakSIMPLICIALDATABASES}. Spivak also provides a way to map from
 
-\subsubsection{Subset \& Streaming}
+\subsubsection{Subset}
 $\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U$ embedded in $K$ with the same fiber $\iota^*E$ and $\iota^*\sigma$.
@@
 
-\subsection{Display Space}
+\subsection{Prerender Space}
 \label{sec:display}
+
+% preamble type intro to displays, examples - displays can include screen, 3D printer, sphereical display, H is the total space of the screen. 11
 \begin{figure}[h]
@@
 
+%%%pho is paramterized over alpha, beta, which is obtained via pullback lookup from (Hx,Hy)-?(alpha, beta)
+
 \subsection{Artist}
+%% include some words about motivation 
 \begin{equation}

### 2020-11-25 draft/sections/implementation.tex
clean up, moving around

--- 2020-11-23 draft/sections/math.tex
@@
-\documentclass[../intro.tex]{subfiles}
+\documentclass[../main.tex]{subfiles}
 \begin{document}
-
-\section{Notation \& Definitions}
-
-\begin{equation}
-    A: \Gamma(E) \rightarrow \Gamma(H)
-\end{equation}
-
-\subsection{Data Space}
-\begin{figure}[ht]
-    \includegraphics[width=0.5\linewidth]{figures/sections/math/mobius.png}
-    \caption{write up some words here}
-    \label{fig:mobius}
-    %%possibly replace w/ simpler deconstructed band image notes-11-19-20
-\end{figure}
-
-We use a fiber bundle model to represent the data, as proposed by Butler 
-\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
-
-One example of a fiber bundle is the mobius band shown in figure~\ref{fig:mobius}.
-A fiber bundle is a topological total space $E$ with an embedded fiber space $F$, a base space on which the fibers lie $K$ and the $\pi$ and $\sigma$ mappings between $E$ and $K$.
-The space of all $\sigma$ is $\Gamma$
-
-\begin{tikzcd}
-    F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right ] \\
-                      & K \arrow[u, "\sigma" description, bend right]
-\end{tikzcd}
-
-As illustrated by the mobius band example in figure~\ref{fig:mobius}, the vertical lines $F$ are the range of possible values embedded in $E$.
-The circle $K$ is a representation of the connectivity of the points in $E$.
-The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$.?
-The section $\sigma$ is the mapping from locations $k$ on $K$ to points on $F_{k}$ in $E$.  (Pull this back into more specific about fig, the general is making this more confusing.)
-
-\subsubsection{Base Space $K$}
 
@@
 \end{description}
-
-In a locally trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
-A fiber bundle can be made locally trivial by approximating the total space E as a simplacial complex.
-
-\subsubsection{Fiber Space $F$}
-The fibers encode the set of all possible values each observation can take. 
-Spivak's \cite{spivakSIMPLICIALDATABASES} formulation of the fibers as a union of named types gives a way to separate the values in the fiber by type:
-
-\subsubsection{Subset \& Streaming}
-$\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U$ embedded in $K$ with the same fiber $\iota^*E$ and $\iota^*\sigma$.
-
-\begin{tikzcd}
-    \iota^\ast E \arrow[r, hook] \arrow[d]                                                                       & E \arrow[d, bend right]                       \\
-    U \arrow[r, "\iota" description, hook] \arrow[u, "\iota^\ast \sigma" description, bend right, shift right=2] & K \arrow[u, "\sigma" description, bend right]
-\end{tikzcd}
-%%% sheaves & presheaves
-
-\subsection{Display Space}
-\label{sec:display}
-\begin{figure}[h]
-    \includegraphics[width=.4\linewidth]{figures/sections/math/render.png}
-    \caption{}
-    \label{fig:render}
-\end{figure}
-
-A physical display space can be thought of sets of $\mathbb{R}^{7}$ tuples, where 
-\begin{equation}
-    \mathbb{R}^{7} = \{X, Y, Z, R, G, B, A\}
-\end{equation}
-
-and the sets correspond to the sections on $\S$, which is the topology of the output of the artist $A$.
-The space $H$ is a total space representing the predisplay space, with a fiber of $\mathbb{R}^7$ and a base space of $\S$:
-
-\begin{tikzcd}
-    \mathbb{R}^{7} \arrow[r, hook] & H \arrow[d, "\pi" description, bend right] \\
-                                   & S \arrow[u, "\rho" description, bend right] 
-\end{tikzcd}
-
-In the case of 2D screens, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$.
-As illustrated in figure~\ref{fig:render}, a region on the screen defined by the corners $(x_1, y_1)$ and $(x_2, y_2)$ maps into a region on a 2-simplex in $S$ defined by $(\alpha_1, \beta_1)$ and $(\alpha_2, \beta_2)$.
-The function on the simplex $f$ returns the (R, G, B, A) value for that $(\alpha, \beta)$ pair.
-For a region,
-
-\begin{equation*}
-\rho(S) = \int_{\alpha_1}^{\alpha_2}\int_{\beta_1}^{\beta_2}\int_{z_1}^{z_2}{R, G, B, A}  
-\end{equation*}
-
-where the R,G,B,A values are derived from the how the data values are mapped to visual characteristics.
-The z component of the mapping to $\mathbb{R}^7$ is moved to the integration because this is a trivial space representing a 2D screen; $\rho$ varies depending on $H$.
-
-\subsection{Artist}
-\begin{equation}
-    A: \Gamma(E) \rightarrow \Gamma(H)
-\end{equation}
-
-\subsubsection{Screen to Data}
-%%diagram: [data] -Tau->[RGVXYZ]
-%%%            \ /\ / 
-\begin{tikzcd}
-    E \arrow[d, "\pi" description] & H \arrow[d, "\pi" description]                                                 \\
-    K \arrow[u, "\sigma" description, bend right] & S \arrow[l, "\xi" description] \arrow[u, "\rho" description, bend right]
-\end{tikzcd}
-
-The pullback $\xi$ on $S \rightarrow K$ means that the values in $E$ can be directly mapped to a simplex in $S$, which means there's a mapping from screen space back to the values.
-
-\begin{tikzcd}
-    \xi E \arrow[rr, "\tau" description] \arrow[rd, "\xi \sigma" description, bend right] &     & H \arrow[ld] \\
-    & S \arrow[lu, bend right] &             
-\end{tikzcd}
-
-\subsubsection{Marks}
-%% diagram of connected components/line thing 11-19-20 notes
-Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space these are marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
-\begin{equation}
-\begin{tikzcd}
-    H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \Gamma \text{ s.t.
-} \Gamma(0)=k \text{ and }\Gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
-\end{tikzcd}
-\label{eq:mark}
-\end{equation}
-
-Each point $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on S described in section~\ref{sec:display} then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
-To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
-
-\subsubsection{Visual Characteristics}
-Tau can preserves the measurement type properties (group scales)
-
-\subsubsection{Visual Idioms: Equivalance class of artists}
-Two artists are equivalent when given data containers $\Gamma(E)$ of the same type, they output the same type of prerender $\Gamma(S)$ 
-\begin{tikzcd}
-    A_{\tau_2}: \arrow[d, shift right=2] & \Gamma(E) \arrow[r] & \Gamma(H) &                                                \\
-    A_{\tau_1}: \arrow[u, shift right]   & \Gamma(E) \arrow[r] & \Gamma(H) 
-\end{tikzcd}
-
 \end{document}

### 2020-11-25 draft/sections/math.tex
clean up, moving around

@@
-\documentclass[../intro.tex]{subfiles}
+\documentclass[../main.tex]{subfiles}
 \begin{document}
@@
 
+%% locally trival bundle is map from one topoligical space E to another K such that around any point in K there exists an open set containing that point such that if we restrict E to that part, it's quivalent to (homoemorphic to) open set cross fiber.
+%%  
 \subsection{Data Model}
@@
     %% or replace w/ box w/ thing that's kind of convoluted 
-    %% simpler still replace w/ timeseries of temp, 1D fiber
+    %% simpler still replace w/ timeseries of temp, 1D fiber -> inspired by butler 1989 FIgG 1
 \end{figure}
@@
 
-As illustrated by figure~\ref{fig:mobius}, the vertical lines $F$ are the range of possible values embedded in the total space $E$.
-The topol $K$ is a representation of the connectivity of the points in $E$.
+As illustrated by figure~\ref{fig:fiberbundle}, the vertical lines $F$ are the range of possible temperature values embedded in the total space $E$.
+The base space $K$ of the fiber bundle describes the connectivity of the points in $E$; in figure~\ref{fig:fiberbundle} the connectivty of the timeseries is encoded in the line representation of $K$.
 The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$.
@@
 \subsubsection{Base Space $K$}
+
 %% general statement about topology, examples of 0, 1, ....Ks [none triangular things, square/torus, ....], condition on K is that it can be represent as a triangulation of K (can be realized as triangle on K - extra structure that may be arbitrary
-
-\begin{figure}[ht]
-    \label{fig:simplex}
-    \includegraphics{figures/sections/math/simplex.png}
-    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
-Higher order simplicies are outside the scope of this paper.}
-\end{figure}
-
-%% k is a triangularizable topological space 
-One way to represent the topological space K is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
-Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another: %%formula for simplacial complexes/simplacial sets
-
-% replace this table w/ a diagram that shows this 
-\begin{description}
-    \item[0-simplex] discrete observations (inventory records)
-    \item[1-simplex] 1D continuos data (timeseries)
-    \item[2-simplex] 2D continuos data (map)
-    \item[3-simplex] 3D continuos data (video)
-\end{description}
-
-I
+%% k is a topological space ....equation soup
 
@@
 The fibers encode the set of all possible values each observation can take.
-For example, the fiber bundle in figure~\ref{fig:movie} is the set of all possible movies, where each $k \in K$ is a frame in the movie.
-Each \sigma
+For example, the fiber in figure~\ref{fig:temp} is the space of possible temperature values in $\circ$ celsius, ranging from [start, end].
+In figure~\ref{fig:temp_time}, time is encoded as a second dimension and since the data is from 2019 the value t
 
@@
 
+Tau is fully flexible and can do whatever; knows about fiber & neighborhood of fiber. Can in theory approximate hatching/dashing/etc can be approximated w/ functions and neighborhood of k. 
 \subsubsection{Visual Idioms: Equivalance class of artists}

### 2020-11-30 draft/sections/math.tex
pictures & not so many words
rewrote intro to math section w/ more context

@@
 \section{Notation \& Definitions}
-%%% add some words around
+In this section we introduce a mathematical formulation of the visualization pipeline where artist $A$ functions transform data $\Gamma(E)$ to an intermediate representation in prerendered display space $Gamme(H)$:
 
@@
     A: \Gamma(E) \rightarrow \Gamma(H)
+    \lab{eq:artist}
 \end{equation}
 
-%% E is a locally trivial fiber bundle over K where K is triangulizable space representing data space & H is a fiber bundle over S which is a simplacial complex of triangles representing visual space and artists in matplotlib will go between the two.
-E is a trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
+$E$ is a locally trivial fiber bundle over $K$ representing data space; $K$ is a triangulizable space encoding the connectivity of the observations in the data.
+$H$ is a fiber bundle over $S$ representing visual space;  $S$ is a simplacial complex of triangles encoding the connectivity of the visual output of each artist $A$.
+When E is a trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
 
-%% locally trival bundle is map from one topoligical space E to another K such that around any point in K there exists an open set containing that point such that if we restrict E to that part, it's quivalent to (homoemorphic to) open set cross fiber.
-%%  
+\begi
+
 \subsection{Data Model}
 \begin{figure}[ht]
+    \label{fig:fiberbundle}
+    \includegraphics{figures/sections/math/fiberbundle.png}
     \caption{write up some words here}
-    \label{fig:fiberbundle}
-    %%possibly replace w/ simpler deconstructed band image notes-11-19-20
-    %% or replace w/ box w/ thing that's kind of convoluted 
+
     %% simpler still replace w/ timeseries of temp, 1D fiber -> inspired by butler 1989 FIgG 1
@@
 
-One example of a fiber bundle is shown in figure~\ref{fig:mobius}.
 A fiber bundle is a topological total space $E$ with an embedded fiber space $F$, a base space on which the fibers lie $K$ and the $\pi$ and $\sigma$ mappings between $E$ and $K$.
@@
 \subsubsection{Base Space $K$}
-
+\begin{figure}
+    \includegraphics{figures/sections/math/temp_1k.png}
+\end{figure}
+\begin{figure}
+    \includegraphics{figures/sections/math/temp_2k.png}
+\end{figure}
 %% general statement about topology, examples of 0, 1, ....Ks [none triangular things, square/torus, ....], condition on K is that it can be represent as a triangulation of K (can be realized as triangle on K - extra structure that may be arbitrary
@@
 \begin{figure}
-    \includegraphics{}
-    \label{fig:movie}
+    \includegraphics{figures/sections/math/temp_2f.png}
+    \label{fig:}
 \end{figure}
+\begin{figure}
+    \includegraphics{figures/sections/math/temp_3f.png}
+\end{figures}
 %%% might 

### 2020-11-30 draft/sections/math.tex
start of more worked through F

@@
 \section{Notation \& Definitions}
-In this section we introduce a mathematical formulation of the visualization pipeline where artist $A$ functions transform data $\Gamma(E)$ to an intermediate representation in prerendered display space $Gamme(H)$:
+In this section we introduce a mathematical description of the visualization pipeline where artist $A$ functions transform data of type $\Gamma(E)$ to an intermediate representation in prerendered display space of type $\Gamma(H)$:
 
@@
 
-$E$ is a locally trivial fiber bundle over $K$ representing data space; $K$ is a triangulizable space encoding the connectivity of the observations in the data.
-$H$ is a fiber bundle over $S$ representing visual space;  $S$ is a simplacial complex of triangles encoding the connectivity of the visual output of each artist $A$.
+\begin{enumerate}
+\item $A$ is the function that converts an instance of data $\Gamma(E)$ to an instance of a visual representation $\Gamma(H)$ 
+\item $E$ is a locally trivial fiber bundle over $K$ representing data space.
+\item $K$ is a triangulizable space encoding the connectivity of the observations in the data. 
+\item $H$ is a fiber bundle over $S$ representing visual space
+\item $S$ is a simplacial complex of triangles encoding the connectivity of the visualization of $\Gamma(E)$
+\end{enumerate}
+
 When E is a trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
-
-\begi
+Fiber bundles are product spaces of toplological spaces, which are a set of points with a set of neighborhoods for each point\cite{some topology text book, really wikipedia}.
 
@@
 %%% might 
-The fibers encode the set of all possible values each observation can take.
-For example, the fiber in figure~\ref{fig:temp} is the space of possible temperature values in $\circ$ celsius, ranging from [start, end].
-In figure~\ref{fig:temp_time}, time is encoded as a second dimension and since the data is from 2019 the value t
+The fibers $F$ are a topological space embedded in $E$ on which lie the set of all possible values.
+For example, if $F$ is the interval $[0, 1]$, then $g(k)$ from equation~\ref{eq:sigma} returns a single measurement $x$ in the interval $F$:
+\begin{equation}
+    \label{eg:goff}
+    g(k) = x, \text(where} 0\leq x \leq 1
+\end{equation}
 
-which in Spivak's formulation means the fibers encode the union of the types of measurements\cite{spivakSIMPLICIALDATABASES}. Spivak also provides a way to map from
+The fiber in figure~\ref{fig:temp} is the space of possible temperature values in $\circ$ celsius, ranging from [start, end], similar to the interval $F$ in equation~\ref{eg:goff}.
+F can be any number of dimensions, for example in figure~\ref{fig:temp_time} time is encoded as a second dimension.
+Given:
+\begin{enumerate}
+\item interval of all possible temperarture values $\[T_{min}, T_{max}\]$ 
+\item interval of all possible  time values $\[t_{min}, t_{max}\]$
+\end{enumerate}
+
+then $F$ is the cross product $F= $\[T_{min}, T_{max}\]$  \times
+
+$g(k) = (temp, time)$.  
+\begin{equation}
+\end{equation}
+
+As an alternative to listing the intervals or discrete values
 

### 2020-12-01 draft/sections/math.tex
more on f

@@
 
-%%% preface w/ we care about topology & types. 
 We use a fiber bundle model to represent the data, as proposed by Butler 
 \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
-
 A fiber bundle is a topological total space $E$ with an embedded fiber space $F$, a base space on which the fibers lie $K$ and the $\pi$ and $\sigma$ mappings between $E$ and $K$.
@@
 
-then $F$ is the cross product $F= $\[T_{min}, T_{max}\]$  \times
-
-$g(k) = (temp, time)$.  
+then $F$ is the cross product $F= \[T_{min}, T_{max}\] \times \[t_{min}, t_{max}\]$, and $g(k)$ listed in equation~\ref{eq:sigma} is:
 \begin{equation}
+g(k) = (x_0, x_1) \text{where} x_0 \in \[T_{min}, T_{max}\], x_1 \in \[t_{min}, t_{max}\]
 \end{equation}
 
-As an alternative to listing the intervals or discrete values
+Given $F = X_{0} \times X_{1} \times \ldots \times X_{n}$, where $X_{0}, X_{1}, \ldots X_{n}$ are measurement spaces (e.g. ints, floats, colors).
+\begin{equation}
+g(k) = (x_0, x_1, \ldots x_n) \text{where} x_0 \in X_0, x_1 \in X_1, \ldots x_n \in X_n
+\end{equation}
+
+These measurement spaces $X$ are each variables and have the properties of measurement scales, such as Steven's nominal, ordinal, interval, and ratio \cite{stevensTheoryScalesMeasurement1946}
+
+\cite{spivakSIMPLICIALDATABASES}.
 
@@
 \end{tikzcd}
-%%% sheaves & presheaves
 

### 2020-12-01 draft/sections/math.tex
unpacking U

--- 2020-11-30 draft/sections/math.tex
@@
     A: \Gamma(E) \rightarrow \Gamma(H)
-    \lab{eq:artist}
+    \label{eq:artist}
 \end{equation}
 
-\begin{enumerate}
+\begin{itemize}
 \item $A$ is the function that converts an instance of data $\Gamma(E)$ to an instance of a visual representation $\Gamma(H)$ 
@@
 \item $S$ is a simplacial complex of triangles encoding the connectivity of the visualization of $\Gamma(E)$
-\end{enumerate}
+\end{itemize}
 
-When E is a trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
+When $E$ is a trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
 Fiber bundles are product spaces of toplological spaces, which are a set of points with a set of neighborhoods for each point\cite{some topology text book, really wikipedia}.
+We will be discussing total spaces $E$ that are locally trivial, meaning that for every point in K there exists an open set neighborhood $U$ such that if you restrict $E$ to that point, then  $\iota*E = F\times U$.
+[there's a missing sentance here...]
 
@@
     \label{fig:fiberbundle}
-    \includegraphics{figures/sections/math/fiberbundle.png}
+    \includegraphics[width=.2\linewidth]{figures/sections/math/fiberbundle.png}
     \caption{write up some words here}
-
     %% simpler still replace w/ timeseries of temp, 1D fiber -> inspired by butler 1989 FIgG 1
@@
 The space of all $\sigma$ is $\Gamma$  (Pull this back into more specific about fig, the general is making this more confusing.)
-%% locally trivial - for every point in K there exists an open set neighborhood $U$ such that if you restrict E to that point, then  \iota*E = F\times U [maybe do mobius strip diagram here?
-U is subsection of the ring, part where it's not twisted.
-there exists a triangulation & a way to do so is simplicies
 
 \subsubsection{Base Space $K$}
-\begin{figure}
-    \includegraphics{figures/sections/math/temp_1k.png}
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
 \end{figure}
-\begin{figure}
-    \includegraphics{figures/sections/math/temp_2k.png}
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
 \end{figure}
@@
 \subsubsection{Fiber Space $F$}
-\begin{figure}
-    \includegraphics{figures/sections/math/temp_2f.png}
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2f.png}
     \label{fig:}
 \end{figure}
-\begin{figure}
-    \includegraphics{figures/sections/math/temp_3f.png}
-\end{figures}
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_3f.png}
+\end{figure}
 %%% might 
@@
     \label{eg:goff}
-    g(k) = x, \text(where} 0\leq x \leq 1
+    g(k) = x, \text{ where } 0\leq x \leq 1
 \end{equation}
@@
 Given:
-\begin{enumerate}
-\item interval of all possible temperarture values $\[T_{min}, T_{max}\]$ 
-\item interval of all possible  time values $\[t_{min}, t_{max}\]$
-\end{enumerate}
+\begin{itemize}
+\item interval of all possible temperature values $[T_{min}, T_{max}]$ 
+\item interval of all possible  time values $[t_{min}, t_{max}]$
+\end{itemize}
 
-then $F$ is the cross product $F= $\[T_{min}, T_{max}\]$  \times
+then $F$ is the cross product $F= [T_{min}, T_{max}] \times [t_{min}, t_{max}]$, and $g(k)$ listed in equation~\ref{eq:sigma} is:
 
-$g(k) = (temp, time)$.  
 \begin{equation}
+g(k) = (x_0, x_1) \text{ where } x_0 \in [T_{min}, T_{max}], x_1 \in [t_{min}, t_{max}]
 \end{equation}
 
-As an alternative to listing the intervals or discrete values
+Given $F = X_{0} \times X_{1} \times \ldots \times X_{n}$, where $X_{0}, X_{1}, \ldots X_{n}$ are measurement spaces (e.g. ints, floats, colors): 
+\begin{equation}
+g(k) = (x_0, x_1, \ldots x_n) \text{ where } x_0 \in X_0, x_1 \in X_1, \ldots x_n \in X_n
+\end{equation}
+
+The measurement spaces $X$ are each variables and have the properties of measurement scales, such as Steven's nominal, ordinal, interval, and ratio \cite{stevensTheoryScalesMeasurement1946}. 
+Spivak provides a formal way of attaching a name to each $X \in F$ and mapping from the value back to the variable type by formulating $E$ in terms of databases \cite{spivakSIMPLICIALDATABASES}.
+He defines a total space $U$, analogous to $E$, that is the disjoint union of all possible elements of all possible types
+
+For example, the interval $[T_{min}, T_{max}]$ is a restriction on $\mathbb{R}$ that is part of the definition of a type named $\mathrm{Temperature}$ and  $[t_{min}, t_{max}]$ is a restriction on $\mathbb{R}$ that is part of a definition of a type named $\mathrm{Time}$.
+As in the case of $E$ embedding the $F$ on which these values live, $U$ contains all pairs of all values of these types $U \coloneqq (\mathrm{Temperature}, \mathrm{Time})$.
+Spivak defines U as being over the space of $\mathrm{DT}$, which is the space of data types $\mathrm{DT} = \{\mathrm{Temperature}, \mathrm{Time}}$.
 
@@
 
-Tau is fully flexible and can do whatever; knows about fiber & neighborhood of fiber. Can in theory approximate hatching/dashing/etc can be approximated w/ functions and neighborhood of k. 
+Tau is fully flexible and can do whatever; knows about fiber \& neighborhood of fiber. Can in theory approximate hatching/dashing/etc can be approximated w/ functions and neighborhood of k. 
 \subsubsection{Visual Idioms: Equivalance class of artists}

### 2020-12-03 draft/sections/math.tex
putting in rest of fiber bundle defintion

@@
     \label{eq:artist}
+\end{equation}
+
+\begin{equation}
+    A: \sigma \rightarrow \rho
 \end{equation}
@@
 \subsection{Data Model}
+
+We use a fiber bundle model to represent the data, as proposed by Butler 
+\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
+A fiber bundle is a structure $(E, K, \pi, F)$  consisting of topological spaces $E, K, F$ and the map from total space to base space $\pi: E \rightarrow K$.
+
+\begin{tikzcd}
+    F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right ] \\
+                      & K \arrow[u, description, bend right]
+\end{tikzcd}
+
+By definition, every point in the base space $k \in K$ has a local neighborhood $U$ such that \pi()
+\cite{rowlandFiberBundle,FiberBundle2020}
+
+\begin{tikzcd}
+    \pi^{-1}(U) \arrow[r, "\varphi" description] \arrow[d, "\pi" description] & U \times F \arrow[ld, "\mathrm{proj}_U"] \\
+    U                                                                         &                                         
+\end{tikzcd}
+
+There is a bijection from the fiber $F$ to the fiber $F_k$ over the point $k \in K$ such that 
+\begin{equation}
+\end{equation}
+
+The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$.
+The section $\sigma$ is the mapping from locations $k$ on $K$ to points on $F_{k}$ in $E$
+
+\begin{tikzcd}
+    F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right ] \\
+                      & K \arrow[u, "\sigma", description, bend right]
+\end{tikzcd}
+
+such that it is the right inverse of $\pi$:
+\begin{equation}
+    \pi(\sigma(k)) = k \text{ for all } \k \in K 
+\end{equation}
+
+In a locally trivial fiber bundle, $\sigma = K \times E$   \cite{weissteinFiberBundle}
+\begin{equation}
+\sigma(k) = (k, g(k))
+\end{equation}
+
+where the fiber at $k$, $F_k$,  is the domain of $g(k)$.  The space of global sections $\sigma$ of $E$ is $\Gamma(E)$:
+
 \begin{figure}[ht]
@@
 
-We use a fiber bundle model to represent the data, as proposed by Butler 
-\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
-A fiber bundle is a topological total space $E$ with an embedded fiber space $F$, a base space on which the fibers lie $K$ and the $\pi$ and $\sigma$ mappings between $E$ and $K$.
-
-\begin{tikzcd}
-    F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right ] \\
-                      & K \arrow[u, description, bend right]
-\end{tikzcd}
-
 As illustrated by figure~\ref{fig:fiberbundle}, the vertical lines $F$ are the range of possible temperature values embedded in the total space $E$.
 The base space $K$ of the fiber bundle describes the connectivity of the points in $E$; in figure~\ref{fig:fiberbundle} the connectivty of the timeseries is encoded in the line representation of $K$.
-The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$.
-The section $\sigma$ is the mapping from locations $k$ on $K$ to points on $F_{k}$ in $E$.
-%%% include diagram here of formal definition of sigma
-
-The space of all $\sigma$ is $\Gamma$  (Pull this back into more specific about fig, the general is making this more confusing.)
 
@@
 
+K is a triangulizable topological space.
+
 \subsubsection{Fiber Space $F$}
+The fiver ver is the set of elements $e \in F$ that lies over 
+$F_{k}$ is a fiber over $K$ that lo
+
+Spivak provides a formalization for the values $x \in X$ whereby they can have specialized types. He defines the space of datatypes $\mathrm{DT}$
+
+DT = set of strings names of types
+U = set of single elements of type in DT
+pi (U-DT)
+Also need to take column names (U cross C)
+
+# C is more about naming types to reuse primatives
+
+Given $F = X_{0} \times X_{1} \times \ldots \times X_{n}$, where $X_{0}, X_{1}, \ldots X_{n}$ are measurement spaces (e.g. ints, floats, colors): 
+\begin{equation}
+g(k) = (x_0, x_1, \ldots x_n) \text{ where } x_0 \in X_0, x_1 \in X_1, \ldots x_n \in X_n
+\end{equation}
+
 \begin{figure}[ht]
@@
 
-Given $F = X_{0} \times X_{1} \times \ldots \times X_{n}$, where $X_{0}, X_{1}, \ldots X_{n}$ are measurement spaces (e.g. ints, floats, colors): 
-\begin{equation}
-g(k) = (x_0, x_1, \ldots x_n) \text{ where } x_0 \in X_0, x_1 \in X_1, \ldots x_n \in X_n
-\end{equation}
+% datetime type, columns = start, end
 
-The measurement spaces $X$ are each variables and have the properties of measurement scales, such as Steven's nominal, ordinal, interval, and ratio \cite{stevensTheoryScalesMeasurement1946}. 
-Spivak provides a formal way of attaching a name to each $X \in F$ and mapping from the value back to the variable type by formulating $E$ in terms of databases \cite{spivakSIMPLICIALDATABASES}.
-He defines a total space $U$, analogous to $E$, that is the disjoint union of all possible elements of all possible types
-
-For example, the interval $[T_{min}, T_{max}]$ is a restriction on $\mathbb{R}$ that is part of the definition of a type named $\mathrm{Temperature}$ and  $[t_{min}, t_{max}]$ is a restriction on $\mathbb{R}$ that is part of a definition of a type named $\mathrm{Time}$.
-As in the case of $E$ embedding the $F$ on which these values live, $U$ contains all pairs of all values of these types $U \coloneqq (\mathrm{Temperature}, \mathrm{Time})$.
-Spivak defines U as being over the space of $\mathrm{DT}$, which is the space of data types $\mathrm{DT} = \{\mathrm{Temperature}, \mathrm{Time}}$.
+% schema is ordered list of Column, type is approx F, U cross C over sigma, 
+% space of rows/records f:k -> r, sigma:k -> r, is the table function (dataframe as instance)
 
@@
 
-\subsubsection{Visual Characteristics}
+\subsubsection{Visual Encodings}
 Tau can preserves the measurement type properties (group scales)
 
+%%The measurement spaces $X$ are each variables and have the properties of measurement scales, such as Steven's nominal, ordinal, interval, and ratio \cite{stevensTheoryScalesMeasurement1946}.
+Stevens talks about measurements and this is how we define it here:....
+can check symmatry under stevens via carry through on tau
+
 Tau is fully flexible and can do whatever; knows about fiber \& neighborhood of fiber. Can in theory approximate hatching/dashing/etc can be approximated w/ functions and neighborhood of k.
+
+%%hatches can either grow the R space or function
+
 \subsubsection{Visual Idioms: Equivalance class of artists}

### 2020-12-03 draft/sections/math.tex
fleshing out locally trivial, making all disgrams equations so I can have references

@@
 
-When $E$ is a trivial fiber bundle $E = F \times K$, it can be assumed that all $F_{k}$ for $k \in K$ are equal.
-Fiber bundles are product spaces of toplological spaces, which are a set of points with a set of neighborhoods for each point\cite{some topology text book, really wikipedia}.
-We will be discussing total spaces $E$ that are locally trivial, meaning that for every point in K there exists an open set neighborhood $U$ such that if you restrict $E$ to that point, then  $\iota*E = F\times U$.
-[there's a missing sentance here...]
+When $E$ is a trivial fiber bundle $E = F \times K$, it can be assumed that all fibers $F_{k}$ over $k \in K$ are equal.
+Fiber bundles are product spaces of toplological spaces, which are a set of points with a set of neighborhoods for each point\cite{FiberBundle2020, rowlandFiberBundle}.
 
@@
 \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
-A fiber bundle is a structure $(E, K, \pi, F)$  consisting of topological spaces $E, K, F$ and the map from total space to base space $\pi: E \rightarrow K$.
+A fiber bundle is a structure $(E, K, \pi, F)$  consisting of topological spaces $E, K, F$ and the map from total space to base space:
 
+\begin{equation}
     \begin{tikzcd}
@@
     \end{tikzcd}
+\end{equation}
 
-By definition, every point in the base space $k \in K$ has a local neighborhood $U$ such that \pi()
-\cite{rowlandFiberBundle,FiberBundle2020}
+where there is a bijection from $F$ to every fiber $F_k$ over point $k \in K$ in $E$ and the function $\pi: E \rightarrow K$ is the map into the $K$ quotient space of $E$.
 
+Every point in the base space $k \in K$ has a local neighborhood $U$ \cite{FiberBundle2020, rowlandFiberBundle}
+
+\begin{equation}
     \begin{tikzcd}
@@
     \end{tikzcd}
+\end{equation}
+$\varphi: \pi^{-1}(U) \rightarrow U \times F$ is a homeomorphism such that $\pi$ and $\mathrm{proj}_U$ both map to $U$, the fiber over $k$ $F_k = \pi^{-1}({k \in K}) $ is homomorphic to the fiber $F$.
+The base space $K$ carries the quotient topology determined by $\pi$, meaning it is
 
-There is a bijection from the fiber $F$ to the fiber $F_k$ over the point $k \in K$ such that 
+meaning that for every point $k in K$ there exists an open set neighborhood $U$ such that if $E$ is restricted to $k$ , then  $\iota*E = F\times U$. 
 \begin{equation}
+abc
 \end{equation}
@@
 The section $\sigma$ is the mapping from locations $k$ on $K$ to points on $F_{k}$ in $E$
-
+\begin{equation}
     \begin{tikzcd}
@@
 \begin{equation}
-    \pi(\sigma(k)) = k \text{ for all } \k \in K 
+    \pi(\sigma(k)) = k \text{ for all } k \in K 
 \end{equation}
 
-In a locally trivial fiber bundle, $\sigma = K \times E$   \cite{weissteinFiberBundle}
+In a locally trivial fiber bundle, $\sigma = K \times E$ \cite{rowlandFiberBundle,FiberBundle2020}:
 \begin{equation}
@@
 \subsubsection{Base Space $K$}
+
+K is a triangulizable topological space.
+
 \begin{figure}[ht]
@@
 %% k is a topological space ....equation soup
-
-K is a triangulizable topological space.
 
@@
 
-# C is more about naming types to reuse primatives
+%% C is more about naming types to reuse primatives
 
@@
 $\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U$ embedded in $K$ with the same fiber $\iota^*E$ and $\iota^*\sigma$.
-
+\begin{equation}
     \begin{tikzcd}
@@
     \end{tikzcd}    
+\end{equation}
 
@@
 The space $H$ is a total space representing the predisplay space, with a fiber of $\mathbb{R}^7$ and a base space of $\S$:
-
+\begin{equation}
     \begin{tikzcd}
@@
 %%%            \ /\ / 
+\begin{equation}
     \begin{tikzcd}
@@
     \end{tikzcd}
-
+\end{equation}
 The pullback $\xi$ on $S \rightarrow K$ means that the values in $E$ can be directly mapped to a simplex in $S$, which means there's a mapping from screen space back to the values. 
-
+\begin{equation}
     \begin{tikzcd}

### 2020-12-03 draft/sections/math.tex
spivak! up to schemas

@@
 \item $S$ is a simplacial complex of triangles encoding the connectivity of the visualization of $\Gamma(E)$
+\item $\sigma: K\rightarrow E$ is the data being visualized
+\item $\rho: S \rightarrow H$ is the render map
 \end{itemize}
@@
 where there is a bijection from $F$ to every fiber $F_k$ over point $k \in K$ in $E$ and the function $\pi: E \rightarrow K$ is the map into the $K$ quotient space of $E$.
-
-Every point in the base space $k \in K$ has a local neighborhood $U$ \cite{FiberBundle2020, rowlandFiberBundle}
+Every point in the base space $k \in K$ has a local open set neighborhood $U$ \cite{FiberBundle2020, rowlandFiberBundle}
 
@@
 \end{equation}
-$\varphi: \pi^{-1}(U) \rightarrow U \times F$ is a homeomorphism such that $\pi$ and $\mathrm{proj}_U$ both map to $U$, the fiber over $k$ $F_k = \pi^{-1}({k \in K}) $ is homomorphic to the fiber $F$.
-The base space $K$ carries the quotient topology determined by $\pi$, meaning it is
+such that $\varphi: \pi^{-1}(U) \rightarrow U \times F$ is a homeomorphism where $\pi$ and $\mathrm{proj}_U$ both map to $U$ and the fiber over $k$ $F_k = \pi^{-1}({k \in K}) $ is homomorphic to the fiber $F$.
 
-meaning that for every point $k in K$ there exists an open set neighborhood $U$ such that if $E$ is restricted to $k$ , then  $\iota*E = F\times U$. 
-\begin{equation}
-abc
-\end{equation}
-
-The function $\pi$ is the mapping from a point on a specific fiber $F_{k}|k\in K$ in $E$ to a location $k \in K$.
-The section $\sigma$ is the mapping from locations $k$ on $K$ to points on $F_{k}$ in $E$
+The section $\sigma$ is the mapping $\sigma: K\rightarrow E$ 
 \begin{equation}
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right ] \\
-                        & K \arrow[u, "\sigma", description, bend right]
+        F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right]    \\
+                          & K \arrow[u, "\sigma" description, bend right]
         \end{tikzcd}
@@
 
-such that it is the right inverse of $\pi$:
+such that it is the right inverse of $\pi$
 \begin{equation}
@@
 
-where the fiber at $k$, $F_k$,  is the domain of $g(k)$.  The space of global sections $\sigma$ of $E$ is $\Gamma(E)$:
+where the domain of $g(k)$ is $F_k$.
+The space of sections over $U \subset K$ is called a sheaf and the space of all possible sections $\sigma$ of $E$ is $\Gamma(E)$.
+All datasets $\sigma \in \Gamma(E)$ have the same variables $F$ and connectivity $K$ but can have different values such that $\sigma_{i}\neq\sigma_{j}$.
 
@@
     \caption{write up some words here}
-    %% simpler still replace w/ timeseries of temp, 1D fiber -> inspired by butler 1989 FIgG 1
 \end{figure}
@@
 \subsubsection{Base Space $K$}
-
-K is a triangulizable topological space.
+K is a triangulizable quotient topological space.
+%%insert stuff about qotient spaces....
 
@@
 \end{figure}
-%% general statement about topology, examples of 0, 1, ....Ks [none triangular things, square/torus, ....], condition on K is that it can be represent as a triangulation of K (can be realized as triangle on K - extra structure that may be arbitrary
-%% k is a topological space ....equation soup
 
 \subsubsection{Fiber Space $F$}
-The fiver ver is the set of elements $e \in F$ that lies over 
-$F_{k}$ is a fiber over $K$ that lo
+Spivak models the fiber space $F$ as schema and the data as sheafs (localized $\sigma$ functions) on the schema \cite{spivakSIMPLICIALDATABASES}. He defines the type specification 
+\begin{equation}
+\pi: U \rightarrow DT
+\end{equation}
 
-Spivak provides a formalization for the values $x \in X$ whereby they can have specialized types. He defines the space of datatypes $\mathrm{DT}$
-
-DT = set of strings names of types
-U = set of single elements of type in DT
-pi (U-DT)
-Also need to take column names (U cross C)
-
-%% C is more about naming types to reuse primatives
+where $DT$ is the set of data types (as identified by their names) and $U$ is the set of all possible objects $x$ of all types in $DT$.
+This means that for each type $T\in DT$, the preimage $\pi^{1}(T)\subset U $ is the domain of $T$, and a value $x \in \pi^{1}(T)\subset U$ is an object of type $T$.
+Spivak then defines a schema $(C, \sigma)$ of type $\pi$, where $\pi$ is the universe of all types, such that
+\begin{equation}
+\sigma: C \rightarrow DT
+\end{equation}
+where $C$ is a finite ordered set of columns in a database. We extend C to refer to the names of all distinct
 
@@
 \subsubsection{Subset}
-$\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U$ embedded in $K$ with the same fiber $\iota^*E$ and $\iota^*\sigma$.
+$\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U \subset K$ with the same fiber $F$
 \begin{equation}
@@
 \end{equation}
+where $\iota^*E$ and $\iota^*\sigma$ are $E$ and $\sigma$ restricted to points $k \in U \subset K$.
 

### 2020-12-07 draft/sections/math.tex
k bits of math, feel super circular

@@
 \subsubsection{Base Space $K$}
-K is a triangulizable quotient topological space.
-%%insert stuff about qotient spaces....
+
+\begin{figure}
+    \includegraphics{figures/sections/math/k_qspace.png}
+    \label{fig:kquotient}
+\end{figure}
+
+$K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $p$ in of $E$ defined via the map $\pi: E \rightarrow K$ that sends each point $p \in E$ to its equivalence class in $[p] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+As shown in figure~\ref{fig:kquotient}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
+This subdivision is projected down to the toplology $\Tau_k$
+
+\begin{equation}
+\tau_K = {U\subseteqq K: {p \in E: [p] \in U}\in \Tau_E}
+\end{equation}
+
+where $[p] \in U$ is the point $k \in K$ with an open set around it that has an open  preimage in $E$ under the surjective map $\pi: p \rightarrow [p]$.
 
@@
 \end{equation}
-where $C$ is a finite ordered set of columns in a database. We extend C to refer to the names of all distinct
+where $C$ is a finite set of names of columns in a database or a list of attributes.
 
-Given $F = X_{0} \times X_{1} \times \ldots \times X_{n}$, where $X_{0}, X_{1}, \ldots X_{n}$ are measurement spaces (e.g. ints, floats, colors): 
-\begin{equation}
-g(k) = (x_0, x_1, \ldots x_n) \text{ where } x_0 \in X_0, x_1 \in X_1, \ldots x_n \in X_n
-\end{equation}
+In the fiber bundle, $F$ is the cartesian product of all sets in the disjoint union $U_{\sigma}$
 
@@
 \end{figure}
-%%% might 
+
 The fibers $F$ are a topological space embedded in $E$ on which lie the set of all possible values.
@@
 % space of rows/records f:k -> r, sigma:k -> r, is the table function (dataframe as instance)
+When $E$ is trivial, then we can decompose $E$ so that each $E$
 
@@
 
-\subsubsection{Visual Encodings}
+\subsubsection{Channels}
+Acts on different parts of F, types means measurement groups , can be broken out so tau(E_{u} in E)
 Tau can preserves the measurement type properties (group scales)

### 2020-12-07 draft/sections/math.tex
coherency maybe

@@
 As illustrated by figure~\ref{fig:fiberbundle}, the vertical lines $F$ are the range of possible temperature values embedded in the total space $E$.
-The base space $K$ of the fiber bundle describes the connectivity of the points in $E$; in figure~\ref{fig:fiberbundle} the connectivty of the timeseries is encoded in the line representation of $K$.
+The base space $K$ of the fiber bundle describes the connectivity of the points in $E$; in figure~\ref{fig:fiberbundle} the connectivity of the timeseries is encoded in the line representation of $K$.
 
@@
 
-\begin{figure}
-    \includegraphics{figures/sections/math/k_qspace.png}
+\begin{figure}[ht]
+    \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
     \label{fig:kquotient}
@@
 As shown in figure~\ref{fig:kquotient}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
-This subdivision is projected down to the toplology $\Tau_k$
+This subdivision is projected down to the toplology $\tau_k$
 
 \begin{equation}
-\tau_K = {U\subseteqq K: {p \in E: [p] \in U}\in \Tau_E}
+\tau_K = \{U\subseteqq K: \{p \in E: [p] \in U\}\in \tau_E\}
 \end{equation}
@@
 \begin{figure}[ht]
+    \label{fig:k_data}
     \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
+    %% add box around neighboring P and Map
 \end{figure}
@@
 \end{figure}
+
+We use $K$ to encode the connectivity of the points $p$, such as the ones in figure~\ref{fig:k_data} that only store temperature data.
+In a timeseries, the temperature $p$ at timme $t$ is dependent on the value at $p_{t-1}$ and $\p_{t+1}$ is dependent on the value in $p_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
+In the case of the map, every point $p$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
+$K$ does not know the time or latitude or longitude of the point - those are metadata variables potentially encoded in $p$ because they are ways of describing the connectivity rather than the connectivity itself.
+The mapping $sigma: K \rightarrow E$ provides the binding between the key on $K$ and the value $p$ in $E$ \cite{munznerChDataAbstraction}
 
@@
 \subsubsection{Channels}
-Acts on different parts of F, types means measurement groups , can be broken out so tau(E_{u} in E)
+Acts on different parts of F, types means measurement groups , can be broken out so 
+%%tau(E_{u} in E)
 Tau can preserves the measurement type properties (group scales)

### 2020-12-07 draft/sections/math.tex
is U_sigma singular

@@
 \begin{figure}[ht]
-    \label{fig:fiberbundle}
     \includegraphics[width=.2\linewidth]{figures/sections/math/fiberbundle.png}
     \caption{write up some words here}
+    \label{fig:fiberbundle}
 \end{figure}
@@
     \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
-    \label{fig:kquotient}
+    \label{fig:kquote}
 \end{figure}
@@
 $K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $p$ in of $E$ defined via the map $\pi: E \rightarrow K$ that sends each point $p \in E$ to its equivalence class in $[p] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
-As shown in figure~\ref{fig:kquotient}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
+As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
 This subdivision is projected down to the toplology $\tau_k$
@@
 \begin{figure}[ht]
-    \label{fig:k_data}
     \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
     %% add box around neighboring P and Map
+    \label{fig:k_data}
 \end{figure}
@@
 
-We use $K$ to encode the connectivity of the points $p$, such as the ones in figure~\ref{fig:k_data} that only store temperature data.
-In a timeseries, the temperature $p$ at timme $t$ is dependent on the value at $p_{t-1}$ and $\p_{t+1}$ is dependent on the value in $p_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
+We use $K$ to encode the connectivity of the points $p$.
+In figure~\ref{fig:k_data}, there is only one data field in $p$, temperature, but the points $p$ are connected differently.
+In a timeseries, the temperature $p$ at time $t$ is dependent on the value at $p_{t-1}$ and $p_{t+1}$ is dependent on the value in $p_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
 In the case of the map, every point $p$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
 $K$ does not know the time or latitude or longitude of the point - those are metadata variables potentially encoded in $p$ because they are ways of describing the connectivity rather than the connectivity itself.
-The mapping $sigma: K \rightarrow E$ provides the binding between the key on $K$ and the value $p$ in $E$ \cite{munznerChDataAbstraction}
+The mapping $sigma: K \rightarrow E$ provides the binding between the key on $K$ and the value $p$ in $E$ \cite{munznerChDataAbstraction}.
 
@@
 
-where $DT$ is the set of data types (as identified by their names) and $U$ is the set of all possible objects $x$ of all types in $DT$.
-This means that for each type $T\in DT$, the preimage $\pi^{1}(T)\subset U $ is the domain of $T$, and a value $x \in \pi^{1}(T)\subset U$ is an object of type $T$.
+where $DT$ is the set of data types (as identified by their names) and $U$ is the disjoint set of all possible objects $x$ of all types in $DT$.
+This means that for each type $T\in DT$, the preimage $\pi^{-1}(T)\subset U $ is the domain of $T$, and $x \in \pi^{1}(T)\subset U$ is an object of type $T$.
 Spivak then defines a schema $(C, \sigma)$ of type $\pi$, where $\pi$ is the universe of all types, such that
@@
 \end{equation}
-where $C$ is a finite set of names of columns in a database or a list of attributes.
+where $C$ is the finite set of names of data fields in $E$. The set of all values restricted to the datatypes in $DT$
+
+\begin{equation}
+    \begin{tikzcd}
+            U_{\sigma} \arrow[d, "\pi_{\sigma}" description] \arrow[r] & U \arrow[d, "\pi" description] \\
+            C \arrow[r, "\sigma" description]                          & DT                            
+    \end{tikzcd}
+\end{equation}
 
 In the fiber bundle, $F$ is the cartesian product of all sets in the disjoint union $U_{\sigma}$
+
+The record function is the sigma function
 

### 2020-12-07 draft/sections/math.tex
maybe possibily finall grok pullbacks maybe

@@
 \end{equation}
-where $C$ is the finite set of names of data fields in $E$. The set of all values restricted to the datatypes in $DT$
+where $C$ is the finite set of names of data fields in $E$. The set of all values restricted to the datatypes in $DT$ is $U_{\sigma}$
 
@@
 \end{equation}
+The pullback $U_{\sigma} \coloneqq \sigma^{-1}(U)$ restricts $U$ to the datatypes of the fields in $C$ such that $U_{\sigma}$ is the fiber product $U \times_{DT} C$, and the pullback $\pi_{\sigma}:U_{\sigma} \rightarrow C$ defines a domain bundle $U_{\sigma}$ over $C$ induced by $\sigma$.
 

### 2020-12-07 draft/sections/math.tex
too many sigma tau

--- 2020-12-07 draft/sections/math.tex
@@
 \end{equation}
+The pullback $U_{\sigma} \coloneqq \sigma^{-1}(U)$ restricts $U$ to the datatypes of the fields in $C$ such that $U_{\sigma}$ is the fiber product $U \times_{DT} C$, and the pullback $\pi_{\sigma}:U_{\sigma} \rightarrow C$ specifies the domain bundle $U_{\sigma}$ over $C$ induced by $\sigma$.
+This domain bundle backs the fiber $F$ in the data total space $E$
+\begin{equation}
+    F = \prod_{i \in I} U_{\sigma_i} = 
+    \label{eq:fiber}
+\end{equation}
 
-In the fiber bundle, $F$ is the cartesian product of all sets in the disjoint union $U_{\sigma}$
+where $F$ is the cartesian product of all sets in the disjoint union $U_{\sigma}$
 

### 2020-12-08 draft/sections/math.tex
switched out notation and filled in blanks for start/k/f

@@
 \begin{equation}
-    A: \sigma \rightarrow \rho
+    A: \tau \rightarrow \rho
 \end{equation}
@@
 \item $S$ is a simplacial complex of triangles encoding the connectivity of the visualization of $\Gamma(E)$
-\item $\sigma: K\rightarrow E$ is the data being visualized
+\item $\tau: K\rightarrow E$ is the data being visualized
 \item $\rho: S \rightarrow H$ is the render map
@@
 
-The section $\sigma$ is the mapping $\sigma: K\rightarrow E$ 
+The section $\tau$ is the mapping $\tau: K\rightarrow E$ 
 \begin{equation}
@@
         F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right]    \\
-                          & K \arrow[u, "\sigma" description, bend right]
+                          & K \arrow[u, "\tau" description, bend right]
         \end{tikzcd}
@@
 \begin{equation}
-    \pi(\sigma(k)) = k \text{ for all } k \in K 
+    \pi(\tau(k)) = k \text{ for all } k \in K 
 \end{equation}
 
-In a locally trivial fiber bundle, $\sigma = K \times E$ \cite{rowlandFiberBundle,FiberBundle2020}:
+In a locally trivial fiber bundle, $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
 \begin{equation}
-\sigma(k) = (k, g(k))
+\tau(k) = (k, g(k))
 \end{equation}
 
-where the domain of $g(k)$ is $F_k$.
-The space of sections over $U \subset K$ is called a sheaf and the space of all possible sections $\sigma$ of $E$ is $\Gamma(E)$.
-All datasets $\sigma \in \Gamma(E)$ have the same variables $F$ and connectivity $K$ but can have different values such that $\sigma_{i}\neq\sigma_{j}$.
+where the domain of $g(k)$ is $F_k$ and returns a data point $r$.
+The space of all possible sections $\tau$ of $E$ is $\Gamma(E)$.
+All datasets $\tau \in \Gamma(E)$ have the same variables $F$ and connectivity $K$ but can have different values such that $\tau_{i}\neq\tau_{j}$.
 
@@
 As illustrated by figure~\ref{fig:fiberbundle}, the vertical lines $F$ are the range of possible temperature values embedded in the total space $E$.
-The base space $K$ of the fiber bundle describes the connectivity of the points in $E$; in figure~\ref{fig:fiberbundle} the connectivity of the timeseries is encoded in the line representation of $K$.
+The base space $K$ of the fiber bundle is a line because the data points $r$ in $E$ are on a space that is  continous in one dimension.
 
@@
 
-$K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $p$ in of $E$ defined via the map $\pi: E \rightarrow K$ that sends each point $p \in E$ to its equivalence class in $[p] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+$K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $r$ in $E$ defined via the map $\pi: E \rightarrow K$ that sends each $r \in E$ to its equivalence class in $[r] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+
 As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
-This subdivision is projected down to the toplology $\tau_k$
+This subdivision is projected down to the toplology $\mathcal{T}$
 
 \begin{equation}
-\tau_K = \{U\subseteqq K: \{p \in E: [p] \in U\}\in \tau_E\}
+\mathcal{T}_k = \{U\subseteqq K: \{r \in E: [r] \in U\}\in \mathcal{T}_E\}
 \end{equation}
 
-where $[p] \in U$ is the point $k \in K$ with an open set around it that has an open  preimage in $E$ under the surjective map $\pi: p \rightarrow [p]$.
+where $[r] \in U$ is the point $k \in K$ with an open set surrounding it that has an open preimage in $E$ under the surjective map $\pi: r \rightarrow [r]$.
 
@@
 
-We use $K$ to encode the connectivity of the points $p$.
-In figure~\ref{fig:k_data}, there is only one data field in $p$, temperature, but the points $p$ are connected differently.
-In a timeseries, the temperature $p$ at time $t$ is dependent on the value at $p_{t-1}$ and $p_{t+1}$ is dependent on the value in $p_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
-In the case of the map, every point $p$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
-$K$ does not know the time or latitude or longitude of the point - those are metadata variables potentially encoded in $p$ because they are ways of describing the connectivity rather than the connectivity itself.
-The mapping $sigma: K \rightarrow E$ provides the binding between the key on $K$ and the value $p$ in $E$ \cite{munznerChDataAbstraction}.
+in figure~\ref{fig:k_data}, temperature is the only one data field in $r$ but the $K$ base spaces are different.
+subfig[1] is a timeseries, so the temperature in $r$ at time $t$ is dependent on the temperature in $r_{t-1}$ and the temperature in $r_{t+1}$ is dependent on  $r_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
+In the case of the map, every temperature in $r$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
+$K$ does not know the time or latitude or longitude of the point as those are metadata variables describing the $k$ rather than the value of $k$.
+The mapping $\tau: K \rightarrow E$ provides the binding between the key $k \in K$ and the value $r$ in $E$ \cite{munznerChDataAbstraction}.
 
 \subsubsection{Fiber Space $F$}
-Spivak models the fiber space $F$ as schema and the data as sheafs (localized $\sigma$ functions) on the schema \cite{spivakSIMPLICIALDATABASES}. He defines the type specification 
+We use Spivak's formalization of data base schemas as the basis of our fiber space $F$ \cite{spivakSIMPLICIALDATABASES}. He defines the type specification 
 \begin{equation}
@@
 \end{equation}
-where $C$ is the finite set of names of data fields in $E$. The set of all values restricted to the datatypes in $DT$ is $U_{\sigma}$
+where $C$ is the finite set of names of columuns, which we generalize to data fields in $E$. The set of all values restricted to the datatypes in $DT$ is $U_{\sigma}$
 
@@
 The pullback $U_{\sigma} \coloneqq \sigma^{-1}(U)$ restricts $U$ to the datatypes of the fields in $C$ such that $U_{\sigma}$ is the fiber product $U \times_{DT} C$, and the pullback $\pi_{\sigma}:U_{\sigma} \rightarrow C$ specifies the domain bundle $U_{\sigma}$ over $C$ induced by $\sigma$.
-This domain bundle backs the fiber $F$ in the data total space $E$
-\begin{equation}
-    F = \prod_{i \in I} U_{\sigma_i} = 
-    \label{eq:fiber}
-\end{equation}
+The fiber $F$ is the cartesian product of all sets in the disjoint union $U_{\sigma}$.
 
-where $F$ is the cartesian product of all sets in the disjoint union $U_{\sigma}$
-
-The record function is the sigma function
+For each field $c \in C$, the record function $r: C \rightarrow U_{\sigma}$ returns an object of type $\sigma(c) \in DT$.
+The set of all records $\Gamma(\sigma)$ is the set of all sections on $U_\sigma$ is $F$ over $E$.
+Spivak defines the $\tau$ mapping from an index of databases $K$ to records $\Gamma(\sigma)$ as $\tau: K \rightarrow \Gamma(\sigma)$.
+This is equivalent to $\tau: k \rightarrow E$ since $F = \Gamma(\sigma)$ and $F$ is the embedding in $E$ on which the records $r$ lie.
 
@@
 
-The fibers $F$ are a topological space embedded in $E$ on which lie the set of all possible values.
-For example, if $F$ is the interval $[0, 1]$, then $g(k)$ from equation~\ref{eq:sigma} returns a single measurement $x$ in the interval $F$:
+%% pull out measurement
+The fiber in figure~\ref{fig:temp} is the space of possible temperature values in degrees celsius, such that $F=[temp_{min}, temp_{max}]$ and each $r$ is a record with a single element $temp$.
+$F$ can be any number of dimensions, for example in figure~\ref{fig:temp_time} time is encoded as a second dimension.
+This means that the set of possible values $F$ is
+
 \begin{equation}
-    \label{eg:goff}
-    g(k) = x, \text{ where } 0\leq x \leq 1
+F = [temp_{min}, temp_{max}] \times [time_{min}, time_{max}]
 \end{equation}
 
-The fiber in figure~\ref{fig:temp} is the space of possible temperature values in $\circ$ celsius, ranging from [start, end], similar to the interval $F$ in equation~\ref{eg:goff}.
-F can be any number of dimensions, for example in figure~\ref{fig:temp_time} time is encoded as a second dimension.
-Given:
-\begin{itemize}
-\item interval of all possible temperature values $[T_{min}, T_{max}]$ 
-\item interval of all possible time values $[t_{min}, t_{max}]$
-\end{itemize}
+and the function $\tau$ that retrieves records from $F$ is
 
-then $F$ is the cross product $F= [T_{min}, T_{max}] \times [t_{min}, t_{max}]$, and $g(k)$ listed in equation~\ref{eq:sigma} is:
+\begin{gather}
+\tau(k) =(k, (temp, time))\\
+temp \in [temp_{min}, temp_{max}], time \in [time_{min}, time_{max}])
+\end{gather}
 
-\begin{equation}
-g(k) = (x_0, x_1) \text{ where } x_0 \in [T_{min}, T_{max}], x_1 \in [t_{min}, t_{max}]
-\end{equation}
+Since $\tau(k)=(k, r)$, $\sigma$ binds the $temp$ to a named temperature data field and a temperature data type and $time$ to a named time field and a time data type.
 
-% datetime type, columns = start, end
-
-% schema is ordered list of Column, type is approx F, U cross C over sigma, 
-% space of rows/records f:k -> r, sigma:k -> r, is the table function (dataframe as instance)
-When $E$ is trivial, then we can decompose $E$ so that each $E$
-
-\subsubsection{Subset}
+\subsubsection{Sheaf: Subset}
+Something something this is how we get that subsets on E retain propertoes of transforms on E.
 $\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U \subset K$ with the same fiber $F$
@@
 where $\iota^*E$ and $\iota^*\sigma$ are $E$ and $\sigma$ restricted to points $k \in U \subset K$.
+
+% add U ^ V + intersection to get section on union which is condition of sheafs.
 
@@
 
-A physical display space can be thought of sets of $\mathbb{R}^{7}$ tuples, where 
+The sets correspond to the sections on $\S$, which is the topology of the output of the artist $A$.
+The space $H$ is a total space representing the predisplay space, with a fiber of   which is dependent on the render space and a base space of $\S$:
+\begin{equation}
+    \begin{tikzcd}
+        D \arrow[r, hook] & H \arrow[d, "\pi" description, bend right] \\
+                                    & S \arrow[u, "\rho" description, bend right] 
+    \end{tikzcd}
+\end{equation}
+
+We are primarily concerned with physical display space. For the renderers currently defined in Matplotlib, we can define the fiber as the $\mathbb{R}^{7}$ tuple
 \begin{equation}
@@
 \end{equation}
-
-and the sets correspond to the sections on $\S$, which is the topology of the output of the artist $A$.
-The space $H$ is a total space representing the predisplay space, with a fiber of $\mathbb{R}^7$ and a base space of $\S$:
-\begin{equation}
-    \begin{tikzcd}
-        \mathbb{R}^{7} \arrow[r, hook] & H \arrow[d, "\pi" description, bend right] \\
-                                    & S \arrow[u, "\rho" description, bend right] 
-    \end{tikzcd}
-\end{equation}
+where each tuple is representative of a pixel.
 
@@
 
-\begin{equation*}
-\rho(S) = \int_{\alpha_1}^{\alpha_2}\int_{\beta_1}^{\beta_2}\int_{z_1}^{z_2}{R, G, B, A}  
-\end{equation*}
+\begin{equation}
+ab=23
+\end{equation}
 
@@
 The z component of the mapping to $\mathbb{R}^7$ is moved to the integration because this is a trivial space representing a 2D screen; $\rho$ varies depending on $H$.
-
-%%%pho is paramterized over alpha, beta, which is obtained via pullback lookup from (Hx,Hy)-?(alpha, beta)
 

### 2020-12-08 draft/sections/intro.tex
clean up, moving around
rewrote intro to math section w/ more context
taus

--- 2020-11-02 draft/intro.tex
@@
-\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}
-\usepackage[utf8]{inputenc}
-\usepackage{tikz}
-\usepackage{array,multirow}
-\usepackage{subcaption}
-\usepackage{subfiles}
-\usepackage{url}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{float}
-\usepackage{diagbox}
-\usepackage{graphicx}
-\usepackage[backend=bibtex, style=numeric-comp]{biblatex}
-\bibliography{glasslab_viz}
-
-\title{Make any stupid plot you want}
-\author{Hannah Aizenman}
-\committee{Dr. Michael Grossberg (Advisor), Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo}
-\submitted{}
-\abstract{}
+\documentclass[../main.tex]{subfiles}
 \begin{document}
-\makefrontmatter
-
 \section{Introduction}
@@
     \label{fig:viz_same}
+    %% change up image to dubois or bertin or
+
 \end{figure}
@@
 \subsection{Thesis statement}
-We define a visualization as a transform from data to graphic that preserves the topology of the data and the properties of the measurement type.
+We define a visualization as a transform from data to graphic that preserves the topology of the data and faithfully map the properties of the measurement type.
 In fig~\ref{fig:viz_same}, we implicitly assume that the translation from table to heatmap has preserved the order of observations (the rows) and that the perceptually uniform sequential colormap has been applied such that the ordering relation on floats matches the ordering on the colormap (darker colors map to larger numbers).
@@
 
-\subfile{sections/math}
-\printbibliography
+\subsection{ What is a viz}?
+Acquired codes of meaning
+
+\section{Not all data are tables}
+
+Tables, images (Lev), graphs (network X)
+
 \end{document}

### 2020-12-08 draft/sections/math.tex
taus

@@
 For each field $c \in C$, the record function $r: C \rightarrow U_{\sigma}$ returns an object of type $\sigma(c) \in DT$.
-The set of all records $\Gamma(\sigma)$ is the set of all sections on $U_\sigma$ is $F$ over $E$.
+The set of all records $\Gamma(\sigma)$ is the set of all sections on $U_\sigma$.
 Spivak defines the $\tau$ mapping from an index of databases $K$ to records $\Gamma(\sigma)$ as $\tau: K \rightarrow \Gamma(\sigma)$.
@@
 %% pull out measurement
-The fiber in figure~\ref{fig:temp} is the space of possible temperature values in degrees celsius, such that $F=[temp_{min}, temp_{max}]$ and each $r$ is a record with a single element $temp$.
-$F$ can be any number of dimensions, for example in figure~\ref{fig:temp_time} time is encoded as a second dimension.
-This means that the set of possible values $F$ is
+The fiber in figure~\ref{fig:temp} is the space of possible temperature values in degrees celsius, such that $F=[temp_{min}, temp_{max}]$ and is named \textrm{Temp}.
+In figure~\ref{fig:temp_time} \textrm{time} is encoded as a second dimension.
+This means that the set of possible values $F$ with $C=\{\textrm{Temp}, \textrm{Time}\}$:
 
@@
 
-\begin{gather}
-\tau(k) =(k, (temp, time))\\
+\begin{equation}
+\tau(k) =(k, (r: \textrm{Temp}\rightarrow temp, r: \textrm{Time}\rightarrow time))\\
 temp \in [temp_{min}, temp_{max}], time \in [time_{min}, time_{max}])
-\end{gather}
+\end{equation}
 
-Since $\tau(k)=(k, r)$, $\sigma$ binds the $temp$ to a named temperature data field and a temperature data type and $time$ to a named time field and a time data type.
+Since $\tau(k)=(k, r)$, $temp$ is bound to a named data field and $sigma$ binds $temp$ to a temperature data type.
 
 \subsubsection{Sheaf: Subset}
-Something something this is how we get that subsets on E retain propertoes of transforms on E.
 $\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U \subset K$ with the same fiber $F$
@@
 % preamble type intro to displays, examples - displays can include screen, 3D printer, sphereical display, H is the total space of the screen. 11
-\begin{figure}[h]
-    \includegraphics[width=.4\linewidth]{figures/sections/math/render.png}
-    \caption{}
-    \label{fig:render}
-\end{figure}
 
 The sets correspond to the sections on $\S$, which is the topology of the output of the artist $A$.
-The space $H$ is a total space representing the predisplay space, with a fiber of   which is dependent on the render space and a base space of $\S$:
+The space $H$ is a total space representing the predisplay space, with a fiber dependent on the render space and a base space of $\S$:
 \begin{equation}
@@
 
-We are primarily concerned with physical display space. For the renderers currently defined in Matplotlib, we can define the fiber as the $\mathbb{R}^{7}$ tuple
+where $\rho: H \rightarrow S$ is mapping from a mathematical encoding of the image to pixels on the screen.
+
+For a physical screen display, the fiber $D$ encodes position and color of the visual elements as $\mathbb{R}^{7}$:
+
 \begin{equation}
@@
 \end{equation}
-where each tuple is representative of a pixel.
+where the pixel $p \in H$ has the fields $\{x, y, z, r, g, b\}$. In the case of 2D screens, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$.
 
-In the case of 2D screens, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$.
-As illustrated in figure~\ref{fig:render}, a region on the screen defined by the corners $(x_1, y_1)$ and $(x_2, y_2)$ maps into a region on a 2-simplex in $S$ defined by $(\alpha_1, \beta_1)$ and $(\alpha_2, \beta_2)$.
-The function on the simplex $f$ returns the (R, G, B, A) value for that $(\alpha, \beta)$ pair.
-For a region,
-
+To draw an image, a region $p \in H$ is inverse mapped into a region $s \in S$
 \begin{equation}
-ab=23
+s = (\rho_{x}^{-1}(X(p)), \rho_{y}^{-1}(Y(p)))
 \end{equation}
 
-where the R,G,B,A values are derived from the how the data values are mapped to visual characteristics.
-The z component of the mapping to $\mathbb{R}^7$ is moved to the integration because this is a trivial space representing a 2D screen; $\rho$ varies depending on $H$.
+such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $S$ to yield the remaining fields in $p$
+\begin{align}
+    R(p) &= \oint_s \rho_R(s)ds^{2}\\
+    G(p) &= \oint_s \rho G(s)ds^{2}\\
+    B(p) &= \oint_s \rho B(s)ds^{2}
+\end{align}
+
+Here we assume a single opaque 2D image such that the $z$ and $alpha$ fields can be omitted.
+
+\begin{figure}[h]
+    \includegraphics[width=.4\linewidth]{figures/sections/math/render.png}
+    \caption{}
+    \label{fig:render}
+\end{figure}
+
+As illustrated in figure~\ref{fig:render}, ....
 
 \subsection{Artist}
-%% include some words about motivation 
+The artist is a mapping from the $\Gamma(E)$ space of all datasets with a shared $F$ and $K$ to a pre-render space $H$ with display target properties $D$ and visualization structure $S$.
+
 \begin{equation}
@@
 \end{equation}
-The pullback $\xi$ on $S \rightarrow K$ means that the values in $E$ can be directly mapped to a simplex in $S$, which means there's a mapping from screen space back to the values. 
+The pullback $\xi$ on $S \rightarrow K$ means that $ r \in E$ can be mapped to from a simplex in $S$
 \begin{equation}
     \begin{tikzcd}
-        \xi E \arrow[rr, "\tau" description] \arrow[rd, "\xi \sigma" description, bend right] &     & H \arrow[ld] \\
+        \xi E \arrow[rr, "\tau" description] \arrow[rd, "\xi \sigma" description, bend right] &                          & H \arrow[ld, "\rho^{-1}"] \\
                                                                                               & S \arrow[lu, bend right] &                          

### 2020-12-09 draft/sections/math.tex
what is visual mapping?

@@
 \subsubsection{Sheaf: Subset}
-$\Gamma(E)$ is the space of all points in $F$ returned by $\sigma$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U \subset K$ with the same fiber $F$
+
+$\Gamma(E)$ is the space of all points in $F$ returned by $\tau$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U \subset K$ with the same fiber $F$
 \begin{equation}
@@
         \iota^\ast E \arrow[r, hook] \arrow[d]                                                                       & E \arrow[d, bend right]                       \\
-        U \arrow[r, "\iota" description, hook] \arrow[u, "\iota^\ast \sigma" description, bend right, shift right=2] & K \arrow[u, "\sigma" description, bend right]
+        U \arrow[r, "\iota" description, hook] \arrow[u, "\iota^\ast \tau" description, bend right, shift right=2] & K \arrow[u, "\tau" description, bend right]
     \end{tikzcd}    
 \end{equation}
-where $\iota^*E$ and $\iota^*\sigma$ are $E$ and $\sigma$ restricted to points $k \in U \subset K$.
+where $\iota^*E$ and $\iota^*\tau$ are $E$ and $\tau$ restricted to points $k \in U \subset K$.
 
+%not sure what's supposed to happen w/ sheafs here
 % add U ^ V + intersection to get section on union which is condition of sheafs.
@@
 
-where $\rho: H \rightarrow S$ is mapping from a mathematical encoding of the image to pixels on the screen.
+where $\rho: H \rightarrow S$ is mapping from a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to pixel space.
 
-For a physical screen display, the fiber $D$ encodes position and color of the visual elements as $\mathbb{R}^{7}$:
+For a physical screen display, the fiber $D$ encodes position and color of the visual elements as $\mathbb{R}^{7} = \{X, Y, Z, R, G, B, A\}$, where the patch $xy$ is a region in $H$ that we query into $S$ to get values that we integrate into an $\{r, g, b\}$ value.
+In the case of 2D screens, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$.
 
-\begin{equation}
-    \mathbb{R}^{7} = \{X, Y, Z, R, G, B, A\}
-\end{equation}
-where the pixel $p \in H$ has the fields $\{x, y, z, r, g, b\}$. In the case of 2D screens, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$.
-
-To draw an image, a region $p \in H$ is inverse mapped into a region $s \in S$
-\begin{equation}
-s = (\rho_{x}^{-1}(X(p)), \rho_{y}^{-1}(Y(p)))
-\end{equation}
-
-such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $S$ to yield the remaining fields in $p$
+To draw an image, a region  $H$ is inverse mapped into a region $s \in S$ where $s$ is the inverse mapping from pixel to region on $s = \rho^{-1}_XY(xy)$ such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields in $p$
 \begin{align}
     R(p) &= \oint_s \rho_R(s)ds^{2}\\
-    G(p) &= \oint_s \rho G(s)ds^{2}\\
-    B(p) &= \oint_s \rho B(s)ds^{2}
+    G(p) &= \oint_s \rho_G(s)ds^{2}\\
+    B(p) &= \oint_s \rho_B(s)ds^{2}
 \end{align}
@@
 \subsection{Artist}
+What is A supposed to preserve?
+
 The artist is a mapping from the $\Gamma(E)$ space of all datasets with a shared $F$ and $K$ to a pre-render space $H$ with display target properties $D$ and visualization structure $S$.
@@
     A: \Gamma(E) \rightarrow \Gamma(H)
-\end{equation}
-
-\subsubsection{Screen to Data}
-%%diagram: [data] -Tau->[RGVXYZ]
-%%%            \ /\ / 
-\begin{equation}
-    \begin{tikzcd}
-        E \arrow[d, "\pi" description] & H \arrow[d, "\pi" description]                                                 \\
-        K \arrow[u, "\sigma" description, bend right] & S \arrow[l, "\xi" description] \arrow[u, "\rho" description, bend right]
-    \end{tikzcd}
-\end{equation}
-The pullback $\xi$ on $S \rightarrow K$ means that $ r \in E$ can be mapped to from a simplex in $S$
-\begin{equation}
-    \begin{tikzcd}
-        \xi E \arrow[rr, "\tau" description] \arrow[rd, "\xi \sigma" description, bend right] &                          & H \arrow[ld, "\rho^{-1}"] \\
-                                                                                              & S \arrow[lu, bend right] &                          
-        \end{tikzcd}
 \end{equation}
@@
 
-Each point $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on S described in section~\ref{sec:display} then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
+Each point $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on S
+
+then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
 To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
@@
 \subsubsection{Channels}
-Acts on different parts of F, types means measurement groups , can be broken out so 
-%%tau(E_{u} in E)
-Tau can preserves the measurement type properties (group scales)
+The function $rho: S \rightarrow H$ is composed of subfunctions that are dependent on the prerender fiber $D$. In the case of a screen
+\begin{equation}
+    \rho (\alpha, \beta) = (\rho_X(\alpha, \beta), \rho_Y(\alpha, \beta), \rho_Z(\alpha, \beta), \rho_B(\alpha, \beta), \rho_G(\alpha, \beta), \rho_B(\alpha, \beta), \rho_A(\alpha, \beta))
+\end{equation}
+
+where $(\alpha, \beta)$ are coordinates on $S$. The function $Q$
+
+\begin{equation}
+    \begin{tikzcd}
+        \Gamma(\theta) \arrow[rr, "{Q(\nu_{0}(c),\ldots,\nu_n(c))}"] &  & {\rho_X, \rho_Y, \rho_Z, \rho_R, \rho_B, \rho_G, \rho_A}
+    \end{tikzcd}
+\end{equation}
+
+% complex columns are combined to \nu
+
+%% \nu is structure preserving if  E->E  & H->H under nu. \phi can be one of the scales in Stevens
+
+% concrete example of \nu as one of our encodings, if artist is faithfully preserving 
+% if was fiber, trans on fiber bundle.
+we require that nu goes to visual side that can support same translation  - work through some examples : if an artist wanted to preserve, this is what it would have to do - for example relations, interval
 
@@
 
-Tau is fully flexible and can do whatever; knows about fiber \& neighborhood of fiber. Can in theory approximate hatching/dashing/etc can be approximated w/ functions and neighborhood of k.
-
-%%hatches can either grow the R space or function
+ Can in theory approximate hatching/dashing/etc can be approximated w/ functions and neighborhood of k.
 

### 2020-12-10 draft/sections/math.tex
start of nu

@@
 
+As described in section~\ref{sec:display}, $rho: S \rightarrow H$ is composed of subfunctions that are dependent on the prerender fiber $D$. In the case of a screen
+\begin{equation}
+    \rho (\alpha, \beta) = (\rho_X(\alpha, \beta), \rho_Y(\alpha, \beta), \rho_Z(\alpha, \beta), \rho_B(\alpha, \beta), \rho_G(\alpha, \beta), \rho_B(\alpha, \beta), \rho_A(\alpha, \beta))
+\end{equation}
+
+where $(\alpha, \beta)$ are coordinates on $S$.
+The mapping $Q: \Gamma(\tau) \rightarrow D$ builds $\rho$ functions on $S$ by compositing $\nu: c \mapsto v \subset {x,y,z,r,g,b,a}$ functions that map data in some field $c$ to a subset of display variables.
+
+\begin{equation}
+    \begin{tikzcd}
+        \Gamma(\tau) \arrow[rr, "{Q(\nu_{0}(c),\ldots,\nu_n(c))}"] &  & {\rho_X, \rho_Y, \rho_Z, \rho_R, \rho_B, \rho_G, \rho_A}
+    \end{tikzcd}
+    \label{eq:Q}
+\end{equation}
+
+$Q$ composites $\nu$ functions because multiple mappings can target the same field in $D$, for example y position and linewidth will concurrently affect where a pixel is drawn on screen.
+
 \subsubsection{Marks}
-%% diagram of connected components/line thing 11-19-20 notes
 Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space these are marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
@@
 \subsubsection{Channels}
-The function $rho: S \rightarrow H$ is composed of subfunctions that are dependent on the prerender fiber $D$. In the case of a screen
-\begin{equation}
-    \rho (\alpha, \beta) = (\rho_X(\alpha, \beta), \rho_Y(\alpha, \beta), \rho_Z(\alpha, \beta), \rho_B(\alpha, \beta), \rho_G(\alpha, \beta), \rho_B(\alpha, \beta), \rho_A(\alpha, \beta))
-\end{equation}
-
-where $(\alpha, \beta)$ are coordinates on $S$. The function $Q$
-
-\begin{equation}
-    \begin{tikzcd}
-        \Gamma(\theta) \arrow[rr, "{Q(\nu_{0}(c),\ldots,\nu_n(c))}"] &  & {\rho_X, \rho_Y, \rho_Z, \rho_R, \rho_B, \rho_G, \rho_A}
-    \end{tikzcd}
-\end{equation}
+Visual channels are mappings from data into visual space, for example distinguishing categories by coloring them differently  \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}.
+Channels are an example of $\nu: c \mapsto d \in D$, which are the mappings from data $r \in \Gamma(\tau)$ to some part of the prerender fiber $D$.
+the $\nu$ functions can target only one field in $D$, such as $X$, or multiple such as $R,G,B$.
 

### 2020-12-10 draft/sections/math.tex
latex is borked

@@
 \end{figure}
+
 \begin{figure}[ht]
@@
 % preamble type intro to displays, examples - displays can include screen, 3D printer, sphereical display, H is the total space of the screen. 11
-
-The sets correspond to the sections on $\S$, which is the topology of the output of the artist $A$.
+Every point $k \in K$ maps to a space $S_{k} \in S$, which is the topology of the output of the artist $A$.
 The space $H$ is a total space representing the predisplay space, with a fiber dependent on the render space and a base space of $\S$:
@@
 
-\subsection{Artist}
-What is A supposed to preserve?
+\subsection{Artist)
 
@@
 
-As described in section~\ref{sec:display}, $rho: S \rightarrow H$ is composed of subfunctions that are dependent on the prerender fiber $D$. In the case of a screen
+As described in section~\ref{sec:display}, $rho: S \rightarrow H$ is composed of subfunctions that are dependent on the prerender fiber $D$. In the case of a screen where $D=\mathbb{R}^{7}$
+
 \begin{equation}
-    \rho (\alpha, \beta) = (\rho_X(\alpha, \beta), \rho_Y(\alpha, \beta), \rho_Z(\alpha, \beta), \rho_B(\alpha, \beta), \rho_G(\alpha, \beta), \rho_B(\alpha, \beta), \rho_A(\alpha, \beta))
+    \rho (u, v) = (\rho_X(u, v), \rho_Y (u, v), \rho_Z (u, v), \rho_B (u, v), \rho_G (u, v), \rho_B (u, v), \rho_A (u, v))
 \end{equation}
 
-where $(\alpha, \beta)$ are coordinates on $S$.
-The mapping $Q: \Gamma(\tau) \rightarrow D$ builds $\rho$ functions on $S$ by compositing $\nu: c \mapsto v \subset {x,y,z,r,g,b,a}$ functions that map data in some field $c$ to a subset of display variables.
+where $(u,v)$ are coordinates of a point on $S$.
+The map reduce function $Q: F\rightarrow \rho$ builds $\rho$ functions on $S$ by compositing $\nu_{param}: c \mapsto \rho_{d \in D}$ functions that map data in some field $c$ to a subset of visual variables such as position or color\cite{bertinIIPropertiesGraphic2011, munznerChDataAbstraction} that are identified as parameters to the artist function.
 
 \begin{equation}
-    \begin{tikzcd}
-        \Gamma(\tau) \arrow[rr, "{Q(\nu_{0}(c),\ldots,\nu_n(c))}"] &  & {\rho_X, \rho_Y, \rho_Z, \rho_R, \rho_B, \rho_G, \rho_A}
-    \end{tikzcd}
+     \rho(u, v) = Q(\nu_param_{0}(\tau_{c_0}), \ldots \nu_param_{n}(\tau_{c_n}))(u,v)
     \label{eq:Q}
@@
 
-$Q$ composites $\nu$ functions because multiple mappings can target the same field in $D$, for example y position and linewidth will concurrently affect where a pixel is drawn on screen.
+$Q$ composites $\nu$ functions because multiple parameters can target the same field in $D$, for example $\nu_{ypos}$ and $\nu_{linewidth}$ together determine the y position of a line drawn on screen such that $Q: \nu_{ypos}, \nu_{linewidth} \rightarrow \rho_{Y}$
 

### 2020-12-14 draft/sections/math.tex
compiles again
stalks

@@
 \begin{equation}
-    A: \Gamma(E) \rightarrow \Gamma(H)
+    A: O(E) \rightarrow O(H)
     \label{eq:artist}
@@
 \item $H$ is a fiber bundle over $S$ representing visual space
-\item $S$ is a simplacial complex of triangles encoding the connectivity of the visualization of $\Gamma(E)$
+\item $S$ is a simplacial complex of triangles encoding the connectivity of the visualization of the data in $E$
 \item $\tau: K\rightarrow E$ is the data being visualized
@@
     \end{tikzcd}
+    \label{eq:local_trivial}
 \end{equation}
@@
 
-\subsubsection{Sheaf: Subset}
-
-$\Gamma(E)$ is the space of all points in $F$ returned by $\tau$; therefore the points being visualized in a streaming or animation example can be considered a subset that lives on base space $U \subset K$ with the same fiber $F$
+\subsubsection{Sheaf and Stalk}
+As described in equation~\ref{eq:local_trivial}, there is a local space $U \subset K$ around every $k$.
+The inclusion map $\iota: U \rightarrow K$ can be pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $U$.
 \begin{equation}
@@
 \end{equation}
-where $\iota^*E$ and $\iota^*\tau$ are $E$ and $\tau$ restricted to points $k \in U \subset K$.
+
+The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$. The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_k$ \cite{StalkSheaf2019}
+
+\begin{equation}
+    \iota^{-1}\mathscr{F}(\{k\}) = \varinjlim_{k\subseteq U}\mathscr{F}(U) =  \varinjlim_{k \in U} = \mathscr{F}_k 
+\end{equation}
+
+which through $\iota$ gets the data in $E$ at and near to $k$.
 
@@
 
-As illustrated in figure~\ref{fig:render}, ....
+As illustrated in figure~\ref{fig:render}, words.
 
-\subsection{Artist)
+\subsection{Artist}
 

### 2020-12-14 draft/sections/math.tex
sheaves?

@@
 % add U ^ V + intersection to get section on union which is condition of sheafs. 
-
 \subsection{Prerender Space}
@@
 
-The artist is a mapping from the $\Gamma(E)$ space of all datasets with a shared $F$ and $K$ to a pre-render space $H$ with display target properties $D$ and visualization structure $S$.
+The artist is a mapping from the sheaf $O(E)$ which a dataset to a pre-render space $O(H)$ with display target properties $D$ and visualization structure $S$.
 
@@
 
-As described in section~\ref{sec:display}, $rho: S \rightarrow H$ is composed of subfunctions that are dependent on the prerender fiber $D$. In the case of a screen where $D=\mathbb{R}^{7}$
-
-\begin{equation}
-    \rho (u, v) = (\rho_X(u, v), \rho_Y (u, v), \rho_Z (u, v), \rho_B (u, v), \rho_G (u, v), \rho_B (u, v), \rho_A (u, v))
-\end{equation}
-
-where $(u,v)$ are coordinates of a point on $S$.
-The map reduce function $Q: F\rightarrow \rho$ builds $\rho$ functions on $S$ by compositing $\nu_{param}: c \mapsto \rho_{d \in D}$ functions that map data in some field $c$ to a subset of visual variables such as position or color\cite{bertinIIPropertiesGraphic2011, munznerChDataAbstraction} that are identified as parameters to the artist function.
+where $A$ is composed of an asthethic mapping $\Xi = Q \circ \nu$ that builds the prerender render function $rho: S \rightarrow H$ functions described in section~\ref{sec:display}.
+$\XI$ builds $\rho$ functions on $S$ by compositing $\nu_{param}: c \mapsto \rho_{d \in D}$ functions that map data in some field $c$ to a subset of visual variables such as position or color\cite{bertinIIPropertiesGraphic2011, munznerChDataAbstraction} that are identified as parameters to the artist function.
 
@@
 
-$Q$ composites $\nu$ functions because multiple parameters can target the same field in $D$, for example $\nu_{ypos}$ and $\nu_{linewidth}$ together determine the y position of a line drawn on screen such that $Q: \nu_{ypos}, \nu_{linewidth} \rightarrow \rho_{Y}$
+$Q$ composites $\nu$ functions because multiple parameters can target the same field in $D$, for example $\nu_{ypos}$ and $\nu_{linewidth}$ together determine the y position of a line drawn on screen such that $Q: \nu_{ypos}, \nu_{linewidth} \rightarrow \rho_{Y}$.
+The artist is
+\begin{equation}
+\begin{tikzcd}
+    \iota^{*}E \arrow[rd, "\Xi" description] & H \arrow[d, "\rho^{-1}" description, bend right]                         \\
+    U \arrow[u, "\tau" description]          & S \arrow[u, "\rho" description, bend right] \arrow[l, "\xi" description]
+\end{tikzcd}
+\end{equation}
 

### 2020-12-15 draft/sections/math.tex
artist

@@
 \begin{equation}
-    \begin{tikzcd}
-        \iota^\ast E \arrow[r, hook] \arrow[d]                                                                       & E \arrow[d, bend right]                       \\
-        U \arrow[r, "\iota" description, hook] \arrow[u, "\iota^\ast \tau" description, bend right, shift right=2] & K \arrow[u, "\tau" description, bend right]
-    \end{tikzcd}    
+
 \end{equation}
 
-The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$. The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_k$ \cite{StalkSheaf2019}
+The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$.
+The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_k$ \cite{StalkSheaf2019,spanier1989algebraic}
 
@@
 
-% preamble type intro to displays, examples - displays can include screen, 3D printer, sphereical display, H is the total space of the screen. 11
 Every point $k \in K$ maps to a space $S_{k} \in S$, which is the topology of the output of the artist $A$.
@@
 
-where $A$ is composed of an asthethic mapping $\Xi = Q \circ \nu$ that builds the prerender render function $rho: S \rightarrow H$ functions described in section~\ref{sec:display}.
-$\XI$ builds $\rho$ functions on $S$ by compositing $\nu_{param}: c \mapsto \rho_{d \in D}$ functions that map data in some field $c$ to a subset of visual variables such as position or color\cite{bertinIIPropertiesGraphic2011, munznerChDataAbstraction} that are identified as parameters to the artist function.
+where $A$ is composed of an asthethic mapping $\Xi: \xi^*\tau \rightarrow \rho$ that builds the prerender render functions $\rho: S \rightarrow H$ described in section~\ref{sec:display}.
+Every region on $H$ maps back through $S$ to $K$ through $\xi: S_k \rightarrow K$ and $\xi^*k\tau$ pulls back to the data on $\xi^*E$ which is restricted over the visualization space $S$ .
 
 \begin{equation}
-     \rho(u, v) = Q(\nu_param_{0}(\tau_{c_0}), \ldots \nu_param_{n}(\tau_{c_n}))(u,v)
+    \begin{tikzcd}
+        F \arrow[r, hook] & E \arrow[d, "\pi" description]                           & \xi^*E \arrow[l] \arrow[d, "\pi" description]                                              &  & \xi^*E \arrow[rr, "\Xi" description] &                                                                                               & H \arrow[ld, "\rho^{-1}" description] \\
+                          & K \arrow[u, "\tau" description, bend right, shift right] & S \arrow[l, "\xi" description] \arrow[u, "\xi^*\tau" description, bend right, shift right] &  &                                      & S \arrow[ru, "\rho" description, bend right, shift right] \arrow[lu, "\xi^*\tau" description] &
+        \end{tikzcd}
+\end{equation}
+
+The aesthetic property mapper $\Xi=Q\circ\nu$  builds $\rho$ functions via a function $Q$ that compose $\nu_{param}( \xi^*\tau)$ functions that map data in some field $c$ to a subset of $D$, such as x, or RGB.
+These $\rho$ functions are then mapped back into $S$ to provide the values for the region in $H$.
+$Q$ is analogous to a visualization type such as scatter or line plot, and $\nu$ to aesthetic parameters such as line color or marker shape \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels}.
+
+\begin{equation}
+     \rho(u, v) = Q(\nu_(\tau_{c_0}), \ldots \nu_param_{n}(\tau_{c_n}))(u,v)
     \label{eq:Q}
@@
 $Q$ composites $\nu$ functions because multiple parameters can target the same field in $D$, for example $\nu_{ypos}$ and $\nu_{linewidth}$ together determine the y position of a line drawn on screen such that $Q: \nu_{ypos}, \nu_{linewidth} \rightarrow \rho_{Y}$.
-The artist is
-\begin{equation}
-\begin{tikzcd}
-    \iota^{*}E \arrow[rd, "\Xi" description] & H \arrow[d, "\rho^{-1}" description, bend right]                         \\
-    U \arrow[u, "\tau" description]          & S \arrow[u, "\rho" description, bend right] \arrow[l, "\xi" description]
-\end{tikzcd}
-\end{equation}
 

### 2020-12-16 draft/sections/math.tex
word soup about nus

@@
 The inclusion map $\iota: U \rightarrow K$ can be pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $U$.
-\begin{equation}
-
-\end{equation}
-
 The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$.
@@
 
-%not sure what's supposed to happen w/ sheafs here
-% add U ^ V + intersection to get section on union which is condition of sheafs. 
 \subsection{Prerender Space}
@@
 \begin{equation}
-     \rho(u, v) = Q(\nu_(\tau_{c_0}), \ldots \nu_param_{n}(\tau_{c_n}))(u,v)
+     \rho(u, v) = Q(\nu_{param_{n}}(\tau_{c_0}), \ldots \nu_{param_{n}}(\tau_{c_n}))(u,v)
     \label{eq:Q}
@@
 
-$Q$ composites $\nu$ functions because multiple parameters can target the same field in $D$, for example $\nu_{ypos}$ and $\nu_{linewidth}$ together determine the y position of a line drawn on screen such that $Q: \nu_{ypos}, \nu_{linewidth} \rightarrow \rho_{Y}$.
+$Q$ composites $\nu$ functions because multiple parameters can target the same field in $D$, for example $\nu_{ypos}$ and $\nu_{linewidth}$ together determine the output of the $\rho_y$ y position of a line on screen.
 
@@
 \subsubsection{Channels}
-Visual channels are mappings from data into visual space, for example distinguishing categories by coloring them differently  \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}.
-Channels are an example of $\nu: c \mapsto d \in D$, which are the mappings from data $r \in \Gamma(\tau)$ to some part of the prerender fiber $D$.
-the $\nu$ functions can target only one field in $D$, such as $X$, or multiple such as $R,G,B$.
+For the aesthetic mapping function $\nu$ to faithfully transform the data in $\xi^*\tau$, the $\nu$ and $\tau$ must be symmetrical under all transformations such that
+\begin{equation}
+    \nu_{param}(\tau_{})
+\end{equation}
 

### 2020-12-16 draft/sections/math.tex
revised render while working out nu

@@
 
-\subsubsection{Base Space $K$}
-
-\begin{figure}[ht]
-    \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
-    \label{fig:kquote}
-\end{figure}
-
-$K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $r$ in $E$ defined via the map $\pi: E \rightarrow K$ that sends each $r \in E$ to its equivalence class in $[r] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
-
-As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
-This subdivision is projected down to the toplology $\mathcal{T}$
-
-\begin{equation}
-\mathcal{T}_k = \{U\subseteqq K: \{r \in E: [r] \in U\}\in \mathcal{T}_E\}
-\end{equation}
-
-where $[r] \in U$ is the point $k \in K$ with an open set surrounding it that has an open preimage in $E$ under the surjective map $\pi: r \rightarrow [r]$.
-
-\begin{figure}[ht]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
-    %% add box around neighboring P and Map
-    \label{fig:k_data}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
-\end{figure}
-
-in figure~\ref{fig:k_data}, temperature is the only one data field in $r$ but the $K$ base spaces are different.
-subfig[1] is a timeseries, so the temperature in $r$ at time $t$ is dependent on the temperature in $r_{t-1}$ and the temperature in $r_{t+1}$ is dependent on  $r_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
-In the case of the map, every temperature in $r$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
-$K$ does not know the time or latitude or longitude of the point as those are metadata variables describing the $k$ rather than the value of $k$.
-The mapping $\tau: K \rightarrow E$ provides the binding between the key $k \in K$ and the value $r$ in $E$ \cite{munznerChDataAbstraction}.
-
 \subsubsection{Fiber Space $F$}
@@
 
-\begin{equation}
+\begin{multline}
 \tau(k) =(k, (r: \textrm{Temp}\rightarrow temp, r: \textrm{Time}\rightarrow time))\\
 temp \in [temp_{min}, temp_{max}], time \in [time_{min}, time_{max}])
-\end{equation}
+\end{multline}
 
@@
 
+\subsubsection{Base Space $K$}
+
+\begin{figure}[ht]
+    \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
+    \label{fig:kquote}
+\end{figure}
+
+$K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $r$ in $E$ defined via the map $\pi: E \rightarrow K$ that sends each $r \in E$ to its equivalence class in $[r] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+
+As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
+This subdivision is projected down to the toplology $\mathcal{T}$
+
+\begin{equation}
+\mathcal{T}_k = \{U\subseteqq K: \{r \in E: [r] \in U\}\in \mathcal{T}_E\}
+\end{equation}
+
+where $[r] \in U$ is the point $k \in K$ with an open set surrounding it that has an open preimage in $E$ under the surjective map $\pi: r \rightarrow [r]$.
+
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
+    %% add box around neighboring P and Map
+    \label{fig:k_data}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
+\end{figure}
+
+in figure~\ref{fig:k_data}, temperature is the only one data field in $r$ but the $K$ base spaces are different.
+subfig[1] is a timeseries, so the temperature in $r$ at time $t$ is dependent on the temperature in $r_{t-1}$ and the temperature in $r_{t+1}$ is dependent on  $r_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
+In the case of the map, every temperature in $r$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
+$K$ does not know the time or latitude or longitude of the point as those are metadata variables describing the $k$ rather than the value of $k$.
+The mapping $\tau: K \rightarrow E$ provides the binding between the key $k \in K$ and the value $r$ in $E$ \cite{munznerChDataAbstraction}.
+
 \subsection{Prerender Space}
@@
 
-where $\rho: H \rightarrow S$ is mapping from a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to pixel space.
+where $\rho: S \rightarrow H$ is mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to pixel space.
+For a physical screen display, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$ such that $\rho$ is
+\begin{equation}
+    \rho(s)  = \{x, y, z, g, b, a\}
+    \label{eq:rho}
+\end{equation}
 
-For a physical screen display, the fiber $D$ encodes position and color of the visual elements as $\mathbb{R}^{7} = \{X, Y, Z, R, G, B, A\}$, where the patch $xy$ is a region in $H$ that we query into $S$ to get values that we integrate into an $\{r, g, b\}$ value.
-In the case of 2D screens, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$.
+To draw an image, a region, $H$ is inverse mapped into a region $s \in S$ where
+\begin{equation}
+s = \rho^{-1}_{XY}(xy)
+\end{equation}
+such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields in $p$
 
-To draw an image, a region  $H$ is inverse mapped into a region $s \in S$ where $s$ is the inverse mapping from pixel to region on $s = \rho^{-1}_XY(xy)$ such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields in $p$
 \begin{align}
@@
 
-The aesthetic property mapper $\Xi=Q\circ\nu$  builds $\rho$ functions via a function $Q$ that compose $\nu_{param}( \xi^*\tau)$ functions that map data in some field $c$ to a subset of $D$, such as x, or RGB.
-These $\rho$ functions are then mapped back into $S$ to provide the values for the region in $H$.
-$Q$ is analogous to a visualization type such as scatter or line plot, and $\nu$ to aesthetic parameters such as line color or marker shape \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels}.
+\subsubsection{Channels}
+
+The visualization specification function $\Xi=Q\circ\nu$  builds $\rho$ functions via a function $Q$ that compose $\nu_{param}$ visual characteristic mappers
+\begin{equation}
+    \nu_{param}(\tau_{param}) = \mu_{param}
+\end{equation}
+where $\tau_{param}$ is a field in $\tau$ bound to the paremeter and $\mu_param$ is an intermediate visual representation such that $\nu_shape(\tau_shape)$ returns a $\mu_shape$ function that returns a shape for each $k$.
+$Q$ then composites the $\nu$ functions to build $\rho$ as described in equation~\ref{eq:rho}
 
 \begin{equation}
-     \rho(u, v) = Q(\nu_{param_{n}}(\tau_{c_0}), \ldots \nu_{param_{n}}(\tau_{c_n}))(u,v)
+    Q(\mu_{param_0}, \ldots, \mu_{param_n}(s) = \rho(s)
     \label{eq:Q}
@@
 
-$Q$ composites $\nu$ functions because multiple parameters can target the same field in $D$, for example $\nu_{ypos}$ and $\nu_{linewidth}$ together determine the output of the $\rho_y$ y position of a line on screen.
+$Q$ is analogous to a visualization type such as scatter or line plot, and $\nu$ to aesthetic parameters such as line color or marker shape \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels} and composites the $\nu$ mappings into $\rho$ functions as
 
@@
 
-Each point $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on S
+Each point $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$
 
@@
 To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
-
-\subsubsection{Channels}
-For the aesthetic mapping function $\nu$ to faithfully transform the data in $\xi^*\tau$, the $\nu$ and $\tau$ must be symmetrical under all transformations such that
-\begin{equation}
-    \nu_{param}(\tau_{})
-\end{equation}
 

### 2020-12-16 draft/sections/math.tex
germs

@@
 
+\subsubsection{Base Space $K$}
+
+\begin{figure}[ht]
+    \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
+    \label{fig:kquote}
+\end{figure}
+
+$K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $r$ in $E$ defined via the map $\pi: E \rightarrow K$ that sends each $r \in E$ to its equivalence class in $[r] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+
+As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
+This subdivision is projected down to the toplology $\mathcal{T}$
+
+\begin{equation}
+\mathcal{T}_k = \{U\subseteqq K: \{r \in E: [r] \in U\}\in \mathcal{T}_E\}
+\end{equation}
+
+where $[r] \in U$ is the point $k \in K$ with an open set surrounding it that has an open preimage in $E$ under the surjective map $\pi: r \rightarrow [r]$.
+
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
+    %% add box around neighboring P and Map
+    \label{fig:k_data}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
+\end{figure}
+
+in figure~\ref{fig:k_data}, temperature is the only one data field in $r$ but the $K$ base spaces are different.
+subfig[1] is a timeseries, so the temperature in $r$ at time $t$ is dependent on the temperature in $r_{t-1}$ and the temperature in $r_{t+1}$ is dependent on  $r_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
+In the case of the map, every temperature in $r$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
+$K$ does not know the time or latitude or longitude of the point as those are metadata variables describing the $k$ rather than the value of $k$.
+The mapping $\tau: K \rightarrow E$ provides the binding between the key $k \in K$ and the value $r$ in $E$ \cite{munznerChDataAbstraction}.
+
 \subsubsection{Fiber Space $F$}
@@
 
-\begin{multline}
-\tau(k) =(k, (r: \textrm{Temp}\rightarrow temp, r: \textrm{Time}\rightarrow time))\\
-temp \in [temp_{min}, temp_{max}], time \in [time_{min}, time_{max}])
-\end{multline}
+\begin{align}
+\tau(k) =(k, (r: \textrm{Temp}\rightarrow &temp, r: \textrm{Time}\rightarrow time))\\
+&temp \in [temp_{min}, temp_{max}], time \in [time_{min}, time_{max}])
+\end{align}
 
@@
 The inclusion map $\iota: U \rightarrow K$ can be pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $U$.
-The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$.
+\begin{equation}
+    \begin{tikzcd}
+        \xi^*E \arrow[d, "\pi" description]              & E \arrow[d, "\pi" description] \arrow[l, "\xi^*" description]            \\
+        U \arrow[u, "\xi^*\tau" description, bend right] & K \arrow[u, "\tau" description, bend right] \arrow[l, "\xi" description]
+    \end{tikzcd}
+\end{equation}
+
+The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$ with germ $\xi^*\tau$.
 The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_k$ \cite{StalkSheaf2019,spanier1989algebraic}
@@
 which through $\iota$ gets the data in $E$ at and near to $k$.
-
-\subsubsection{Base Space $K$}
-
-\begin{figure}[ht]
-    \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
-    \label{fig:kquote}
-\end{figure}
-
-$K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $r$ in $E$ defined via the map $\pi: E \rightarrow K$ that sends each $r \in E$ to its equivalence class in $[r] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
-
-As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
-This subdivision is projected down to the toplology $\mathcal{T}$
-
-\begin{equation}
-\mathcal{T}_k = \{U\subseteqq K: \{r \in E: [r] \in U\}\in \mathcal{T}_E\}
-\end{equation}
-
-where $[r] \in U$ is the point $k \in K$ with an open set surrounding it that has an open preimage in $E$ under the surjective map $\pi: r \rightarrow [r]$.
-
-\begin{figure}[ht]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
-    %% add box around neighboring P and Map
-    \label{fig:k_data}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
-\end{figure}
-
-in figure~\ref{fig:k_data}, temperature is the only one data field in $r$ but the $K$ base spaces are different.
-subfig[1] is a timeseries, so the temperature in $r$ at time $t$ is dependent on the temperature in $r_{t-1}$ and the temperature in $r_{t+1}$ is dependent on  $r_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
-In the case of the map, every temperature in $r$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
-$K$ does not know the time or latitude or longitude of the point as those are metadata variables describing the $k$ rather than the value of $k$.
-The mapping $\tau: K \rightarrow E$ provides the binding between the key $k \in K$ and the value $r$ in $E$ \cite{munznerChDataAbstraction}.
+Restricting the artist to the sheaf means the artist knows the data in $F$ and also has access to derivatives of the data.
+This property is useful for some visual transformations.
 

### 2020-12-17 draft/sections/math.tex
artists?

@@
     \begin{tikzcd}
-        \xi^*E \arrow[d, "\pi" description]              & E \arrow[d, "\pi" description] \arrow[l, "\xi^*" description]            \\
-        U \arrow[u, "\xi^*\tau" description, bend right] & K \arrow[u, "\tau" description, bend right] \arrow[l, "\xi" description]
+        \xi^*E \arrow[d, "\pi"']              & E \arrow[d, "\pi" description] \arrow[l, "\xi^*" description]            \\
+        U \arrow[u, "\xi^*\tau"', bend right] & K \arrow[u, "\tau" description, bend right] \arrow[l, "\xi" description]
         \end{tikzcd}
@@
 
-The artist is a mapping from the sheaf $O(E)$ which a dataset to a pre-render space $O(H)$ with display target properties $D$ and visualization structure $S$.
-
+The artist is a mapping from the sheaf $O(E)$ representing the data to a pre-render space $O(H)$. 
 \begin{equation}
-    A: \Gamma(E) \rightarrow \Gamma(H)
+    A: O(E) \rightarrow O(H)
 \end{equation}
 
-where $A$ is composed of an asthethic mapping $\Xi: \xi^*\tau \rightarrow \rho$ that builds the prerender render functions $\rho: S \rightarrow H$ described in section~\ref{sec:display}.
-Every region on $H$ maps back through $S$ to $K$ through $\xi: S_k \rightarrow K$ and $\xi^*k\tau$ pulls back to the data on $\xi^*E$ which is restricted over the visualization space $S$ .
+The artist acts on a sheaf because some visual characterstics, such as line thickness, need more information than just the data at a point.
+The artist is the composition of visual variable mappers $F: E \rightarrow V$ and assembly function  $Q: V \rightarrow H$ that transforms $V$ into a visual idiom\cite{munznerMarksChannels} in pre-render space:
 
@@
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[d, "\pi" description]                           & \xi^*E \arrow[l] \arrow[d, "\pi" description]                                              &  & \xi^*E \arrow[rr, "\Xi" description] &                                                                                               & H \arrow[ld, "\rho^{-1}" description] \\
-                          & K \arrow[u, "\tau" description, bend right, shift right] & S \arrow[l, "\xi" description] \arrow[u, "\xi^*\tau" description, bend right, shift right] &  &                                      & S \arrow[ru, "\rho" description, bend right, shift right] \arrow[lu, "\xi^*\tau" description] &
+        O(E) \arrow[r, "F" description] \arrow[rd, "A"'] \arrow[rd] & O(V) \arrow[d, "Q" description] \\
+                                                                    & O(H)                           
         \end{tikzcd}
@@
 
+The mapping from data space to visual variable  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
+
+\begin{equation}
+    \begin{tikzcd}
+        E \arrow[r, "\nu" description]                                & V \\
+        K \arrow[u, "\tau" description] \arrow[ru, "\mu" description] &  
+    \end{tikzcd}
+\end{equation}
+
+such that $\nu: \tau \mapsto \mu$ is a homorphism where $\tau$ and $\mu$ are equivariant such that the properties of the field type persist in the visual representation.
+The map $\xi: S \rightarrow K$ goes from a region $s$ to it associated $k$
+\begin{equation}
+    \begin{tikzcd}
+        E \arrow[d, "\pi" description]              & H \arrow[d, "\pi" description]                                    \\
+        K \arrow[u, "\tau" description, bend right] & S \arrow[l, "\xi" description] \arrow[lu, "\xi^*\tau" description]
+        \end{tikzcd}
+\end{equation}
+and can be pulled back up to $E$ (or $V$) such that for a region $s$ there is an associated record $\tau$ and visual mapping $\mu$.
+The visual fiber bundle $V$ gets pulled back over $S$ via $\xi$ such that
+\begin{equation}
+    \begin{tikzcd}
+        V \arrow[r, "q" description]                                       & H \\
+        S \arrow[u, "\xi^*\mu" description] \arrow[ru, "\rho" description] &  
+    \end{tikzcd}
+\end{equation}
+the composition $q \circ \xi^*\mu$ generates the $\rho$ function described in section~\ref{sec:display}.
+
 \subsubsection{Channels}
+\begin{equation}
+    \begin{tikzcd}
+        E_1 \arrow[r, "F" description] \arrow[d, "f" description] & V_1 \arrow[d, "f" description] \\
+        E_2 \arrow[r, "F" description]                            & V_2                           
+    \end{tikzcd}
+\end{equation}
 
-The visualization specification function $\Xi=Q\circ\nu$  builds $\rho$ functions via a function $Q$ that compose $\nu_{param}$ visual characteristic mappers
+%nu is functor
+% examples of nu \tau\mu pairs - ordinal
+%maybe param is the wrong thing 
+Define a channel in terms of nus
+This is what we mean by equivariance: 
+step thrpoughh measurment scale groups
+doesn't matter where computation lives
+
+Here is the thing you need to preserve for {X} when writing a transform:
+
+compose $\nu_{param}$ visual characteristic mappers
 \begin{equation}
-    \nu_{param}(\tau_{param}) = \mu_{param}
+    \nu(\tau) = \mu
 \end{equation}
 where $\tau_{param}$ is a field in $\tau$ bound to the paremeter and $\mu_param$ is an intermediate visual representation such that $\nu_shape(\tau_shape)$ returns a $\mu_shape$ function that returns a shape for each $k$.
-$Q$ then composites the $\nu$ functions to build $\rho$ as described in equation~\ref{eq:rho}
 
-\begin{equation}
-    Q(\mu_{param_0}, \ldots, \mu_{param_n}(s) = \rho(s)
-    \label{eq:Q}
-\end{equation}
+$\nu$ is Hom set? 
+1) must have identity
+2) is composable
+3) associativity
 
-$Q$ is analogous to a visualization type such as scatter or line plot, and $\nu$ to aesthetic parameters such as line color or marker shape \cite{bertinIIPropertiesGraphic2011, munznerMarksChannels} and composites the $\nu$ mappings into $\rho$ functions as
+homomorphism in land of categories:
+objects in C, objects in D
+C arrow D
+ A-B for example stephens group scale
+
+every $\nu$ is a functor that
+
+put all the functorial things together is still functorial 
+%argument into  is $\[\nu_{0}(\tau_{0}), \ldots \nu_{n}(\tau_{n})\]$
 
 \subsubsection{Marks}
-Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space these are marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
+Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space are the point, line, and area marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
+For each region $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
 \begin{equation}
@@
 \end{equation}
-
-Each point $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$
-
-then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
 To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
@@
 \subsubsection{Visual Idioms: Equivalance class of artists}
-Two artists are equivalent when given data containers $\Gamma(E)$ of the same type, they output the same type of prerender $\Gamma(S)$:
-\begin{equation}
-    \begin{tikzcd}
-        A_{\tau_2}: \arrow[d, shift right=2] & \Gamma(E) \arrow[r] & \Gamma(H) & \\
-        A_{\tau_1}: \arrow[u, shift right]   & \Gamma(E) \arrow[r] & \Gamma(H) 
-    \end{tikzcd}
-\end{equation}
+n $O(E)$ of the same type, they output the same type of prerender $O(H)$:
+
+Natural transformation + composition is partial ordering? Back and forth is equivalent
 

### 2020-12-17 draft/sections/math.tex
artist?

@@
 The artist acts on a sheaf because some visual characterstics, such as line thickness, need more information than just the data at a point.
-The artist is the composition of visual variable mappers $F: E \rightarrow V$ and assembly function  $Q: V \rightarrow H$ that transforms $V$ into a visual idiom\cite{munznerMarksChannels} in pre-render space:
-
-\begin{equation}
-    \begin{tikzcd}
-        O(E) \arrow[r, "F" description] \arrow[rd, "A"'] \arrow[rd] & O(V) \arrow[d, "Q" description] \\
-                                                                    & O(H)                           
-        \end{tikzcd}
-\end{equation}
-
-The mapping from data space to visual variable  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
+The artist through $F: E \rightarrow V$ maps from data space to visual variable  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
 
@@
 \begin{equation}
-    \begin{tikzcd}
-        V \arrow[r, "q" description]                                       & H \\
-        S \arrow[u, "\xi^*\mu" description] \arrow[ru, "\rho" description] &  
-    \end{tikzcd}
+
 \end{equation}
@@
 \subsubsection{Channels}
+Each $\nu$ function maps a field $\tau_c$ to a paramterized visual field $\mu$, for example the color or shape visual channels \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}.
+The set of $\nus$ is a functor from data to visual space such that any $\nu$ is an equivariant map if
+
 \begin{equation}
     \begin{tikzcd}
-        E_1 \arrow[r, "F" description] \arrow[d, "f" description] & V_1 \arrow[d, "f" description] \\
+        E_1 \arrow[r, "F" description] \arrow[d, "" description] & V_1 \arrow[d, "f" description] \\
         E_2 \arrow[r, "F" description]                            & V_2                           

### 2020-12-18 draft/sections/math.tex
artists + pdf
moved mrks out

@@
 
-The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$ with germ $\xi^*\tau$.
+The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$ with germ of $\xi^*\tau$.
 The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_k$ \cite{StalkSheaf2019,spanier1989algebraic}
@@
 This property is useful for some visual transformations.
-
+%% gives all direvatives but we need the jet which is Add stuff about jets
 \subsection{Prerender Space}
@@
 such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields in $p$
-
+%%change the integral to a double integral
 \begin{align}
-    R(p) &= \oint_s \rho_R(s)ds^{2}\\
-    G(p) &= \oint_s \rho_G(s)ds^{2}\\
-    B(p) &= \oint_s \rho_B(s)ds^{2}
+    R(p) &= \iint_s \rho_R(s)ds^{2}\\
+    G(p) &= \iint_s \rho_G(s)ds^{2}\\
+    B(p) &= \iint_s \rho_B(s)ds^{2}
 \end{align}
@@
 \end{equation}
+The artist is composed of two stages, data $E$ to visual mappings $V$ and $V$ mappings composited to a visual in prerender space $H$:
+\begin{equation}
+insert triangle here
+\end{equation}
 
-The artist acts on a sheaf because some visual characterstics, such as line thickness, need more information than just the data at a point.
-The artist through $F: E \rightarrow V$ maps from data space to visual variable  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
-
+The first stage of the artist $F: E \rightarrow V$ maps from data space to visual variable  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
+%%include the $
 \begin{equation}
@@
 such that $\nu: \tau \mapsto \mu$ is a homorphism where $\tau$ and $\mu$ are equivariant such that the properties of the field type persist in the visual representation.
-The map $\xi: S \rightarrow K$ goes from a region $s$ to it associated $k$
+The map $\xi: S \rightarrow K$ goes from a region $s$ to it associated $k$ %%remove the taus from the diagram
 \begin{equation}
@@
 \end{equation}
-and can be pulled back up to $E$ (or $V$) such that for a region $s$ there is an associated record $\tau$ and visual mapping $\mu$.
-The visual fiber bundle $V$ gets pulled back over $S$ via $\xi$ such that
+such that when $\xi(s) = k$, there is a mapping from screen space to k  $\xi^*\tau(s) = \tau(k)$. The visual fiber bundle $V$ gets pulled back over $S$ via $\xi$ such that 
 \begin{equation}
-
+    \begin{tikzcd}
+        \xi^*V \arrow[r, "q"']                      & H \\
+        S \arrow[u, "\xi^*\mu"] \arrow[ru, "\rho"'] &  
+        \end{tikzcd}
 \end{equation}
@@
 Each $\nu$ function maps a field $\tau_c$ to a paramterized visual field $\mu$, for example the color or shape visual channels \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}.
-The set of $\nus$ is a functor from data to visual space such that any $\nu$ is an equivariant map if
+The set of $\nu$ is a functor from data to visual space such that any $\nu$ is an equivariant map if
 
@@
 
+$E$ or $J(E)$
 %nu is functor
@@
 where $\tau_{param}$ is a field in $\tau$ bound to the paremeter and $\mu_param$ is an intermediate visual representation such that $\nu_shape(\tau_shape)$ returns a $\mu_shape$ function that returns a shape for each $k$.
+
+%For example, given a categorical field $c$ carry through temperature, maybe picture?
 

### 2020-12-21 draft/sections/math.tex
pulled general fiber bundle explaination out into its own section

@@
 
-\subsection{Data Model}
-
-We use a fiber bundle model to represent the data, as proposed by Butler 
-\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
-A fiber bundle is a structure $(E, K, \pi, F)$  consisting of topological spaces $E, K, F$ and the map from total space to base space:
+\subsection{Fiber Bundles}
+We provide a brief introduction to fiber bundles because we model data, visual transformations, and a prerendered visual graphic as fiber bundles.
+A fiber bundle is a structure $(E, B, \pi, F)$  consisting of topological total space $E$, base space $K$, fiber space $F$ and the map from total space to base space:
 
@@
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right ] \\
-                        & K \arrow[u, description, bend right]
+        F \arrow[r, hook] & E \arrow[r, "\pi" description] & B  \\
     \end{tikzcd}
@@
 
-where there is a bijection from $F$ to every fiber $F_k$ over point $k \in K$ in $E$ and the function $\pi: E \rightarrow K$ is the map into the $K$ quotient space of $E$.
-Every point in the base space $k \in K$ has a local open set neighborhood $U$ \cite{FiberBundle2020, rowlandFiberBundle}
+where there is a bijection from $F$ to every fiber $F_b$ over point $b \in B$ in $E$ and the function $\pi: E \rightarrow B$ is the map into the $B$ quotient space of $E$.
+
+\subsubsection{Base space}
+\begin{figure}[ht]
+    \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
+    \label{fig:kquote}
+\end{figure}
+
+$B$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $p$ in $E$ defined via the map $\pi: E \rightarrow B$ that sends each $p \in E$ to its equivalence class in $[p] \in B$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+
+As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
+This subdivision is projected down to the toplology $\mathcal{T}$
+
+\begin{equation}
+\mathcal{T}_b = \{U\subseteqq B: \{p \in E: [p] \in U\}\in \mathcal{T}_E\}
+\end{equation}
+
+where $[p] \in U$ is the point $b \in B$ with an open set surrounding it that has an open preimage in $E$ under the surjective map $\pi: p \rightarrow [p]$.
+
+\subsubsection{Fiber}
+Every point in the base space $b \in B$ has a local open set neighborhood $U$ \cite{FiberBundle2020, rowlandFiberBundle}
 
@@
 \end{equation}
-such that $\varphi: \pi^{-1}(U) \rightarrow U \times F$ is a homeomorphism where $\pi$ and $\mathrm{proj}_U$ both map to $U$ and the fiber over $k$ $F_k = \pi^{-1}({k \in K}) $ is homomorphic to the fiber $F$.
+such that $\varphi: \pi^{-1}(U) \rightarrow U \times F$ is a homeomorphism where $\pi$ and $\mathrm{proj}_U$ both map to $U$ and the fiber over $k$ $F_b = \pi^{-1}({b \in B}) $ is homomorphic to the fiber $F$.
 
-The section $\tau$ is the mapping $\tau: K\rightarrow E$ 
+The section $f$ is the mapping from base space to total space $f: B\rightarrow E$ 
 \begin{equation}
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[d, "\pi" description, bend right]    \\
-                          & K \arrow[u, "\tau" description, bend right]
+        F \arrow[r, hook] & E \arrow[d, "\pi" description]    \\
+                          & B \arrow[u, "f" description, bend right]
         \end{tikzcd}
@@
 \begin{equation}
-    \pi(\tau(k)) = k \text{ for all } k \in K 
+    \pi(f(b)) = b \text{ for all } b \in B 
 \end{equation}
 
-In a locally trivial fiber bundle, $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
+In a locally trivial fiber bundle, $E = B \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
 \begin{equation}
-\tau(k) = (k, g(k))
+f(b) = (b, g(b))
 \end{equation}
 
-where the domain of $g(k)$ is $F_k$ and returns a data point $r$.
-The space of all possible sections $\tau$ of $E$ is $\Gamma(E)$.
-All datasets $\tau \in \Gamma(E)$ have the same variables $F$ and connectivity $K$ but can have different values such that $\tau_{i}\neq\tau_{j}$.
+where the domain of $g(b)$ is $F_b$ and returns a point $p$ in $F_b$.
+The space of all possible sections $f$ of $E$ is $\Gamma(E)$.
+All sections $f \in \Gamma(E)$ have the same fibers $F$ and connectivity $B$.
 
+\subsubsection{Sheaf and Stalk}
+As described in equation~\ref{eq:local_trivial}, there is a local space $U \subset B$ around every $b$.
+The inclusion map $\iota: U \rightarrow B$ can be pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $U$.
+\begin{equation}
+    \begin{tikzcd}
+        \iota^*E \arrow[d, "\pi"']           & E \arrow[d, "\pi"'] \arrow[l, "\iota^*" description]                    \\
+        U \arrow[u, "\iota^*f"', bend right] & B \arrow[u, "f" description, bend right] \arrow[l, "\iota" description]
+        \end{tikzcd}
+\end{equation}
+
+The localized section of fibers $\iota^*\f: U \rightarrow \iota^*E$ is the sheaf $O(E)$ with germ of $\xi^*f$.
+The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}
+
+\begin{equation}
+    \iota^{-1}\mathscr{F}(\{b\}) = \varinjlim_{b\subseteq U}\mathscr{F}(U) =  \varinjlim_{b \in U} = \mathscr{F}_b 
+\end{equation}
+
+which through $\iota$ gets the data in $E$ at and near to $b$.
+Restricting the artist to the sheaf means the artist knows the data in $F$ and also has access to derivatives of the data.
+This property is useful for some visual transformations.
+%% gives all direvatives but we need the jet which is Add stuff about jets
+
+\subsection{Data Model}
+
+We use a fiber bundle model to represent the data, as proposed by Butler.  
+\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
 \begin{figure}[ht]
@@
 \subsubsection{Base Space $K$}
-
-\begin{figure}[ht]
-    \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
-    \label{fig:kquote}
-\end{figure}
-
-$K$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $r$ in $E$ defined via the map $\pi: E \rightarrow K$ that sends each $r \in E$ to its equivalence class in $[r] \in K$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
-
-As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
-This subdivision is projected down to the toplology $\mathcal{T}$
-
-\begin{equation}
-\mathcal{T}_k = \{U\subseteqq K: \{r \in E: [r] \in U\}\in \mathcal{T}_E\}
-\end{equation}
-
-where $[r] \in U$ is the point $k \in K$ with an open set surrounding it that has an open preimage in $E$ under the surjective map $\pi: r \rightarrow [r]$.
 
@@
 
-\subsubsection{Sheaf and Stalk}
-As described in equation~\ref{eq:local_trivial}, there is a local space $U \subset K$ around every $k$.
-The inclusion map $\iota: U \rightarrow K$ can be pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $U$.
-\begin{equation}
-    \begin{tikzcd}
-        \xi^*E \arrow[d, "\pi"']              & E \arrow[d, "\pi" description] \arrow[l, "\xi^*" description]            \\
-        U \arrow[u, "\xi^*\tau"', bend right] & K \arrow[u, "\tau" description, bend right] \arrow[l, "\xi" description]
-        \end{tikzcd}
-\end{equation}
-
-The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf $O(E)$ with germ of $\xi^*\tau$.
-The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_k$ \cite{StalkSheaf2019,spanier1989algebraic}
-
-\begin{equation}
-    \iota^{-1}\mathscr{F}(\{k\}) = \varinjlim_{k\subseteq U}\mathscr{F}(U) =  \varinjlim_{k \in U} = \mathscr{F}_k 
-\end{equation}
-
-which through $\iota$ gets the data in $E$ at and near to $k$.
-Restricting the artist to the sheaf means the artist knows the data in $F$ and also has access to derivatives of the data.
-This property is useful for some visual transformations.
-%% gives all direvatives but we need the jet which is Add stuff about jets
 \subsection{Prerender Space}
@@
 
+An artist is the choice of ..... constraints are maps that preserve
+\subsubsection{Compositing}
+
+$Q$ is the assembly function for the visual idiom step something something something something something
+
 \subsubsection{Channels}
@@
 
-\subsubsection{Marks}
-Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space are the point, line, and area marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
-For each region $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
-\begin{equation}
-\begin{tikzcd}
-    H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \Gamma \text{ s.t.
-} \Gamma(0)=k \text{ and }\Gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
-\end{tikzcd}
-\label{eq:mark}
-\end{equation}
-To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
-
 % complex columns are combined to \nu
@@
 
-\subsubsection{Visual Idioms: Equivalance class of artists}
-n $O(E)$ of the same type, they output the same type of prerender $O(H)$:
-
-Natural transformation + composition is partial ordering? Back and forth is equivalent
-
 \end{document}

### 2020-12-21 draft/sections/math.tex
rearranging/changing/cleaning up diagrams

@@
     \label{eq:artist}
-\end{equation}
-
-\begin{equation}
-    A: \tau \rightarrow \rho
 \end{equation}
@@
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[r, "\pi" description] & B  \\
+        F \arrow[r, hook] & E \arrow[r, "\pi"] & B
     \end{tikzcd}
@@
     \begin{tikzcd}
-        \pi^{-1}(U) \arrow[r, "\varphi" description] \arrow[d, "\pi" description] & U \times F \arrow[ld, "\mathrm{proj}_U"] \\
+        \pi^{-1}(U) \arrow[r, "\varphi"] \arrow[d, "\pi"'] & U \times F \arrow[ld, "\mathrm{proj}_U"] \\
         U                                                  &                                         
@@
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[d, "\pi" description]    \\
-                          & B \arrow[u, "f" description, bend right]
+        F \arrow[r, hook] & E \arrow[d, "\pi"']           \\
+        & B \arrow[u, "f"', bend right]
     \end{tikzcd}
@@
     \begin{tikzcd}
-        \iota^*E \arrow[d, "\pi"']           & E \arrow[d, "\pi"'] \arrow[l, "\iota^*" description]                    \\
-        U \arrow[u, "\iota^*f"', bend right] & B \arrow[u, "f" description, bend right] \arrow[l, "\iota" description]
+        \iota^*E \arrow[d, "\pi"']           & E \arrow[d, "\pi"'] \arrow[l, "\iota^*"']         \\
+        U \arrow[u, "\iota^*f"', bend right] & B \arrow[u, "f"', bend right] \arrow[l, "\iota"']
         \end{tikzcd}
@@
 
-The localized section of fibers $\iota^*\f: U \rightarrow \iota^*E$ is the sheaf $O(E)$ with germ of $\xi^*f$.
+The localized section of fibers $\iota^*f: U \rightarrow \iota^*E$ is the sheaf $O(E)$ with germ of $\xi^*f$.
 The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}
@@
 
-\subsubsection{Base Space $K$}
+\subsubsection{Base Space}
 
@@
     \begin{tikzcd}
-            U_{\sigma} \arrow[d, "\pi_{\sigma}" description] \arrow[r] & U \arrow[d, "\pi" description] \\
-            C \arrow[r, "\sigma" description]                          & DT                            
+        U_{\sigma} \arrow[d, "\pi_{\sigma}"'] \arrow[r] & U \arrow[d, "\pi"] \\
+        C \arrow[r, "\sigma"]                           & DT                
     \end{tikzcd}
@@
 Every point $k \in K$ maps to a space $S_{k} \in S$, which is the topology of the output of the artist $A$.
-The space $H$ is a total space representing the predisplay space, with a fiber dependent on the render space and a base space of $\S$:
+The space $H$ is a total space representing the predisplay space, with a fiber dependent on the render space and a base space of $S$:
 \begin{equation}
     \begin{tikzcd}
-        D \arrow[r, hook] & H \arrow[d, "\pi" description, bend right] \\
-                                    & S \arrow[u, "\rho" description, bend right] 
+        D \arrow[r, hook] & H \arrow[r, "\pi"] &S\\   
         \end{tikzcd}
@@
 
-where $\rho: S \rightarrow H$ is mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to pixel space.
+Where the section $\rho: S \rightarrow H$
+\begin{equation}
+    \begin{tikzcd}
+        D \arrow[r, hook] & H \arrow[d, "\pi"']              \\
+                          & S \arrow[u, "\rho"', bend right]
+        \end{tikzcd}
+\end{equation}
+
+ is mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to pixel space.
 For a physical screen display, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$ such that $\rho$ is
 \begin{equation}
-    \rho(s)  = \{x, y, z, g, b, a\}
+    \rho(s)  = \{x, y, z, r, g, b, a\}
     \label{eq:rho}
@@
 \begin{equation}
-s = \rho^{-1}_{XY}(xy)
+s = \rho^{-1}_{\tiny{XY}}(xy)
 \end{equation}
-such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields in $p$
+such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields:
 %%change the integral to a double integral
 \begin{align}
-    R(p) &= \iint_s \rho_R(s)ds^{2}\\
-    G(p) &= \iint_s \rho_G(s)ds^{2}\\
-    B(p) &= \iint_s \rho_B(s)ds^{2}
+    r &= \iint\limits_s \rho_{\tiny{R}}(s)ds^{2}\\
+    g &= \iint\limits_s \rho_{\tiny{G}}(s)ds^{2}\\
+    b &= \iint\limits_s \rho_{\tiny{B}}(s)ds^{2}
 \end{align}
@@
     \begin{tikzcd}
-        E \arrow[r, "\nu" description]                                & V \\
-        K \arrow[u, "\tau" description] \arrow[ru, "\mu" description] &  
+        E \arrow[d, "\pi"'] \arrow[r, "\nu"]                                        & V \arrow[ld, "\pi"'] \\
+        K \arrow[u, "\tau"', bend right] \arrow[ru, "\mu"', bend right, shift left] &                     
         \end{tikzcd}
@@
 such that $\nu: \tau \mapsto \mu$ is a homorphism where $\tau$ and $\mu$ are equivariant such that the properties of the field type persist in the visual representation.
-The map $\xi: S \rightarrow K$ goes from a region $s$ to it associated $k$ %%remove the taus from the diagram
+The map $\xi: S \rightarrow K$ goes from a region $s$ to it associated $k$
 \begin{equation}
     \begin{tikzcd}
-        E \arrow[d, "\pi" description]              & H \arrow[d, "\pi" description]                                    \\
-        K \arrow[u, "\tau" description, bend right] & S \arrow[l, "\xi" description] \arrow[lu, "\xi^*\tau" description]
+        E \arrow[d, "\pi"'] & H \arrow[d, "\pi"'] \\
+        K                   & S \arrow[l, "\xi"']
         \end{tikzcd}

### 2020-12-21 draft/sections/math.tex
what even is xi?

@@
 \begin{equation}
+    \label{eq:base_space}
 \mathcal{T}_b = \{U\subseteqq B: \{p \in E: [p] \in U\}\in \mathcal{T}_E\}
@@
 \subsubsection{Fiber}
-Every point in the base space $b \in B$ has a local open set neighborhood $U$ \cite{FiberBundle2020, rowlandFiberBundle}
+As shown in equation!\ref{eg:base_space}, every point $b \in B$ has a local open set neighborhood $U$ \cite{FiberBundle2020, rowlandFiberBundle}
 
@@
 \subsubsection{Fiber Space $F$}
+\label{sec:data_fiber}
 We use Spivak's formalization of data base schemas as the basis of our fiber space $F$ \cite{spivakSIMPLICIALDATABASES}. He defines the type specification 
@@
 \end{equation}
-The artist is composed of two stages, data $E$ to visual mappings $V$ and $V$ mappings composited to a visual in prerender space $H$:
+The artist is composed of two stages, data $E$ to visual mappings $V$ and $V$ mappings composited to a visual in prerender space $H$. The visual fiber bundle $V$ shares basespace $K$ with $E$:
+
 \begin{equation}
-insert triangle here
+    \begin{tikzcd}
+        P \arrow[r, hook] & V \arrow[r, "\pi"] & K\\   
+    \end{tikzcd}
 \end{equation}
 
-The first stage of the artist $F: E \rightarrow V$ maps from data space to visual variable  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
-%%include the $
+with fiber $P$ that encodes visual paramters such as position or color.
+As in section~\ref{sec:data_fiber}, the fields in $P$ are strongly typed.
+The function $\nu: E \rightarrow V$ maps from data space to visual space  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
 \begin{equation}
@@
         E \arrow[d, "\pi"'] \arrow[r, "\nu"] & V \arrow[ld, "\pi"'] \\
-        K \arrow[u, "\tau"', bend right] \arrow[ru, "\mu"', bend right, shift left] &                     
+        K                                    &                     
         \end{tikzcd}
@@
 
-such that $\nu: \tau \mapsto \mu$ is a homorphism where $\tau$ and $\mu$ are equivariant such that the properties of the field type persist in the visual representation.
-The map $\xi: S \rightarrow K$ goes from a region $s$ to it associated $k$
+such that the visual section $\mu: K \rightarrow V$ 
+\begin{equation}
+    \begin{tikzcd}
+        E \arrow[r, "\nu"]                      & V \\
+        K \arrow[ru, "\mu"'] \arrow[u, "\tau"'] &  
+        \end{tikzcd}
+
+\end{equation}
+
+where visual transformers $\nu$ applied to data $\tau$ is equivalent to $\mu$ such that $\mu(k) = \nu\circ\tau(k)$.
+To obtain the equivalent visual values in screen space $S$, the look up map $\xi: S \rightarrow K$ goes from a region $s$ to it associated $k$
 \begin{equation}
@@
     \begin{tikzcd}
-        \xi^*V \arrow[r, "q"']                      & H \\
-        S \arrow[u, "\xi^*\mu"] \arrow[ru, "\rho"'] &  
+        \xi^*V \arrow[r, "Q"] \arrow[rd, "\xi^*\pi"'] & H \arrow[d, "\pi"] \\
+                                                      & S                 
     \end{tikzcd}
 \end{equation}
-the composition $q \circ \xi^*\mu$ generates the $\rho$ function described in section~\ref{sec:display}.
+the function $Q:\xi^*V\rightarrow H$ composites points in $V$ into $\mathbb{R^7}$ tuples. The section $\mu$ is pulled over $s$
 
-An artist is the choice of ..... constraints are maps that preserve
+\begin{equation}
+    \begin{tikzcd}
+        \xi^*V \arrow[r, "Q"] & H                                           \\
+                              & S \arrow[lu, "\xi^*\mu"] \arrow[u, "\rho"']
+    \end{tikzcd}
+\end{equation}
+
+such that the composition $Q$ of $\mu$ is equivalent to the render $\rho(s) = Q\circ\mu(s)$. The full visualization path is
+\begin{equation} 
+    \begin{tikzcd}
+        E \arrow[r, "\nu=transforms"] & V                                                              & \xi^*V \arrow[r, "Q = \text{set attrs}"]                             & H \\
+                                  & K \arrow[lu, "\tau = data"] \arrow[u, "{\mu = \{\tau,\nu\}}"'] & S \arrow[l, "\xi - init?draw"] \arrow[u, "\xi^*\mu = \nu\circ\tau"'] &  
+    \end{tikzcd}
+\end{equation}
+
 \subsubsection{Compositing}
@@
 \subsubsection{Channels}
+
+%% the \mu [oeve gpes here maybe]
 Each $\nu$ function maps a field $\tau_c$ to a paramterized visual field $\mu$, for example the color or shape visual channels \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}.

### 2020-12-21 draft/sections/math.tex
fleshing out p

@@
  is mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to pixel space.
+
+ \subsubsection{Section $\rho$}
  For a physical screen display, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$ such that $\rho$ is
@@
 
+\subsubsection{{Example}}
 \begin{figure}[h]
@@
 As in section~\ref{sec:data_fiber}, the fields in $P$ are strongly typed.
+Inside the artist is also the equivalance class map $\xi: S \rightarrow K$ that goes from a region $s$ to its associated point $k$
+\begin{equation}
+    \begin{tikzcd}
+        E \arrow[d, "\pi"'] & H \arrow[d, "\pi"'] \\
+        K                   & S \arrow[l, "\xi"']
+        \end{tikzcd}
+\end{equation}
+such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
+
+\subsubsection{Visual Mapping}
+
 The function $\nu: E \rightarrow V$ maps from data space to visual space  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
@@
 
-such that the visual section $\mu: K \rightarrow V$ 
+such that the visual section $\mu: K \rightarrow V$  is equivalent to the visual transformation of the data $\mu(k) = \nu\circ\tau(k)$
 \begin{equation}
@@
         \end{tikzcd}
-
 \end{equation}
 
-where visual transformers $\nu$ applied to data $\tau$ is equivalent to $\mu$ such that $\mu(k) = \nu\circ\tau(k)$.
-To obtain the equivalent visual values in screen space $S$, the look up map $\xi: S \rightarrow K$ goes from a region $s$ to it associated $k$
-\begin{equation}
-    \begin{tikzcd}
-        E \arrow[d, "\pi"'] & H \arrow[d, "\pi"'] \\
-        K                   & S \arrow[l, "\xi"']
-        \end{tikzcd}
-\end{equation}
-such that when $\xi(s) = k$, there is a mapping from screen space to k  $\xi^*\tau(s) = \tau(k)$. The visual fiber bundle $V$ gets pulled back over $S$ via $\xi$ such that 
+The constraint on $\nu: \tau\rightarrow \mu$ is that $\tau$ and $\mu$ must be equivariant.
+
+%%set up proofs here
+
+\subsubsection{Visual Composition}
+The visual fiber bundle $V$ gets pulled back over $S$ via $\xi$ such that 
 \begin{equation}
@@
 
-such that the composition $Q$ of $\mu$ is equivalent to the render $\rho(s) = Q\circ\mu(s)$. The full visualization path is
+such that the composition $Q$ of $\mu$ is equivalent to the render $\rho(s) = Q\circ\mu(s)$.
+
+\subsubsection{Pipeline}
+The full visualization path is
 \begin{equation} 
@@
 
-\subsubsection{Compositing}
-
-$Q$ is the assembly function for the visual idiom step something something something something something
-
-\subsubsection{Channels}
-
-%% the \mu [oeve gpes here maybe]
-Each $\nu$ function maps a field $\tau_c$ to a paramterized visual field $\mu$, for example the color or shape visual channels \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels}.
-The set of $\nu$ is a functor from data to visual space such that any $\nu$ is an equivariant map if
-
-\begin{equation}
-    \begin{tikzcd}
-        E_1 \arrow[r, "F" description] \arrow[d, "" description] & V_1 \arrow[d, "f" description] \\
-        E_2 \arrow[r, "F" description]                            & V_2                           
-    \end{tikzcd}
-\end{equation}
-
-$E$ or $J(E)$
-%nu is functor
-% examples of nu \tau\mu pairs - ordinal
-%maybe param is the wrong thing 
-Define a channel in terms of nus
-This is what we mean by equivariance: 
-step thrpoughh measurment scale groups
-doesn't matter where computation lives
-
-Here is the thing you need to preserve for {X} when writing a transform:
-
-compose $\nu_{param}$ visual characteristic mappers
-\begin{equation}
-    \nu(\tau) = \mu
-\end{equation}
-where $\tau_{param}$ is a field in $\tau$ bound to the paremeter and $\mu_param$ is an intermediate visual representation such that $\nu_shape(\tau_shape)$ returns a $\mu_shape$ function that returns a shape for each $k$.
-
-%For example, given a categorical field $c$ carry through temperature, maybe picture?
-
-$\nu$ is Hom set? 
-1) must have identity
-2) is composable
-3) associativity
-
-homomorphism in land of categories:
-objects in C, objects in D
-C arrow D
- A-B for example stephens group scale
-
-every $\nu$ is a functor that
-
-put all the functorial things together is still functorial 
-%argument into  is $\[\nu_{0}(\tau_{0}), \ldots \nu_{n}(\tau_{n})\]$
-
-% complex columns are combined to \nu
-
-%% \nu is structure preserving if  E->E  & H->H under nu. \phi can be one of the scales in Stevens
-
-% concrete example of \nu as one of our encodings, if artist is faithfully preserving 
-% if was fiber, trans on fiber bundle.
-we require that nu goes to visual side that can support same translation  - work through some examples : if an artist wanted to preserve, this is what it would have to do - for example relations, interval
-
-%%The measurement spaces $X$ are each variables and have the properties of measurement scales, such as Steven's nominal, ordinal, interval, and ratio \cite{stevensTheoryScalesMeasurement1946}.
-Stevens talks about measurements and this is how we define it here:....
-can check symmatry under stevens via carry through on tau
-
- Can in theory approximate hatching/dashing/etc can be approximated w/ functions and neighborhood of k.
-
 \end{document}

### 2020-12-22 draft/sections/math.tex
sorting out notation + fleshing out

@@
 \begin{equation}
-    A: O(E) \rightarrow O(H)
     \label{eq:artist}
+    A: \mathcal{O}(E) \rightarrow \mathcal{O}(H)
 \end{equation}
@@
 \item $E$ is a locally trivial fiber bundle over $K$ representing data space.
-\item $K$ is a triangulizable space encoding the connectivity of the observations in the data. 
+\item $K$ is a triangulizable space encoding the connectivity of the points in $E$
 \item $H$ is a fiber bundle over $S$ representing visual space
-\item $S$ is a simplacial complex of triangles encoding the connectivity of the visualization of the data in $E$
-\item $\tau: K\rightarrow E$ is the data being visualized
-\item $\rho: S \rightarrow H$ is the render map
+\item $S$ is a simplacial complex encoding the visualization
 \end{itemize}
@@
 \begin{equation}
+    \label{eq:fiber_bundle}
     \begin{tikzcd}
@@
 where there is a bijection from $F$ to every fiber $F_b$ over point $b \in B$ in $E$ and the function $\pi: E \rightarrow B$ is the map into the $B$ quotient space of $E$.
+By defintion of a fiber bundle, $\pi$ is always a mapping from total space to base space, independent of the points $p \in E$, and therefore we call this mapping $\pi$ for all the fiber bundles in the model.
 
@@
 \begin{figure}[ht]
+    \label{fig:base_space}
     \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
-    \label{fig:kquote}
 \end{figure}
@@
 \begin{equation}
+    \label{eq:fiber_local_trivial}
     \begin{tikzcd}
@@
     \end{tikzcd}
-    \label{eq:local_trivial}
 \end{equation}
@@
 \begin{equation}
+    \label{eq:section}
     \begin{tikzcd}
@@
 \begin{equation}
+    \label{eq:section_domain}
     \pi(f(b)) = b \text{ for all } b \in B 
@@
 \begin{equation}
+    \label{eq:section_return}
 f(b) = (b, g(b))
@@
 \begin{equation}
+    \label{eq:sheaf}
     \begin{tikzcd}
@@
 
-The localized section of fibers $\iota^*f: U \rightarrow \iota^*E$ is the sheaf $O(E)$ with germ of $\xi^*f$.
-The neighborhood of points the sheaf lies over is the stalk $\mathscr{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}
+The localized section of fibers $\iota^*f: U \rightarrow \iota^*E$ is the sheaf $\mathcal{O}(E)$ with germ of $\xi^*f$.
+The neighborhood of points the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}
 
 \begin{equation}
-    \iota^{-1}\mathscr{F}(\{b\}) = \varinjlim_{b\subseteq U}\mathscr{F}(U) =  \varinjlim_{b \in U} = \mathscr{F}_b 
+    \label{eq:stalk}
+    \iota^{-1}\mathcal{F}(\{b\}) = \varinjlim_{b\subseteq U}\mathcal{F}(U) =  \varinjlim_{b \in U} = \mathcal{F}_b 
 \end{equation}
@@
 \subsection{Data Model}
+As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle:
+\begin{equation}
+    \label{eq:data_fiber_bundle}
+    \begin{tikzcd}
+        F \arrow[r, hook] & E \arrow[r, "\pi"] & B
+    \end{tikzcd}
+\end{equation}
 
-We use a fiber bundle model to represent the data, as proposed by Butler.  
-\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}.
+with a total space $E$, fiber space $F$, and base space $K$ which can be thought of as a set of keys $k$.
+The section $\tau:K \rightarrow E$ is an instance of the data that lies in $E$ and is discussed in section~\ref{sec:data_fiber}.
+
+\subsubsection{Example}
+
 \begin{figure}[ht]
+    \label{fig:data_fiber_bundle}
     \includegraphics[width=.2\linewidth]{figures/sections/math/fiberbundle.png}
     \caption{write up some words here}
-    \label{fig:fiberbundle}
 \end{figure}
@@
 \subsubsection{Base Space}
+The base space $K$ is a representation of the connectivity of the data, specifically whether the points in $E$ are discrete or sampled from a continous space.
+The same dataset can be expressed with different $K$.
 
+\subsubsection{Example}
 \begin{figure}[ht]
@@
 
-\subsubsection{Fiber Space $F$}
+\subsubsection{Fiber and Sections}
 \label{sec:data_fiber}
@@
 
+\subsubsection{Example}
 \begin{figure}[ht]
@@
 
-Where the section $\rho: S \rightarrow H$
+\subsubsection{Section $\rho$}
+
+A section $\rho: S \rightarrow H$
 \begin{equation}
@@
 
- is mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to pixel space.
-
- \subsubsection{Section $\rho$}
+ is a mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to visual space as defined in D.
 For a physical screen display, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$ such that $\rho$ is
@@
 such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields:
-%%change the integral to a double integral
 \begin{align}
@@
 
-The artist is a mapping from the sheaf $O(E)$ representing the data to a pre-render space $O(H)$. 
+In this section we will define the artist as a mapping from the sheaf $O(E)$ representing the data to a pre-render space $O(H)$. 
 \begin{equation}
-    A: O(E) \rightarrow O(H)
+    A: \mathcal{O}(E) \rightarrow \mathcal{O}(H)
 \end{equation}
-The artist is composed of two stages, data $E$ to visual mappings $V$ and $V$ mappings composited to a visual in prerender space $H$. The visual fiber bundle $V$ shares basespace $K$ with $E$:
+The artist decomposes to data $E$ to visual mappings $V$ and $V$ mappings composited to a visual in prerender space $H$.
+In order to define $A$, we first define the visual fiber bundle $V$ that shares basespace $K$ with $E$:
 
@@
 
-with fiber $P$ that encodes visual paramters such as position or color.
-As in section~\ref{sec:data_fiber}, the fields in $P$ are strongly typed.
-Inside the artist is also the equivalance class map $\xi: S \rightarrow K$ that goes from a region $s$ to its associated point $k$
+with fiber $P$ that encodes visual parameters such as position or color. The section $\mu: V \rightarrow K$
+
+\begin{equation}
+    \begin{tikzcd}
+        P \arrow[r, hook] & V \arrow[d, "\pi"']              \\
+                          & K \arrow[u, "\mu"', bend right]
+        \end{tikzcd}
+\end{equation}
+
+returns a record in $P$. The fiber $P$ is the space of visual parameters %% defined like P, dependent on Q, P is targeting our library
+
+\subsubsection{Data Screen Mapping}
+The data bases pace $K$ can be considered a subspace of the screen base space $S$ such that $\xi: S \rightarrow K$ is a deformation retraction
 \begin{equation}
@@
 \end{equation}
-such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
 
-\subsubsection{Visual Mapping}
+that goes from a region $s \in S_{k}$ to its associated point $k$, such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
 
-The function $\nu: E \rightarrow V$ maps from data space to visual space  \cite{carpendaleVisualRepresentationSemiology,bertinIIPropertiesGraphic2011,munznerWhatDataAbstraction2014} space
+\subsubsection{Visual Channels}
+
+The function $\nu: E \rightarrow V$ is a mapping from the data bundle to the visual bundle 
 \begin{equation}
@@
 
-such that the visual section $\mu: K \rightarrow V$  is equivalent to the visual transformation of the data $\mu(k) = \nu\circ\tau(k)$
-\begin{equation}
-    \begin{tikzcd}
-        E \arrow[r, "\nu"]                      & V \\
-        K \arrow[ru, "\mu"'] \arrow[u, "\tau"'] &  
-        \end{tikzcd}
-\end{equation}
+such that a visual section $\mu: K \rightarrow V$  is equivalent to the visual transformation of the data $\mu(k) = \nu\circ\tau(k)$
 
-The constraint on $\nu: \tau\rightarrow \mu$ is that $\tau$ and $\mu$ must be equivariant.
+%%throw out this diagram
+%%something about pushforward ...
+%% \mu is pushfowwards of \tau = \mu = \nu(lowercase * ,\tau), \nu: *\tau \mapsto \mu
+%% \nu acts from subspace E to subspace V
+The constraint on $\nu: \tau\rightarrow \mu$
 
-%%set up proofs here
+%%set up proofs here, are x&y transfereble here? (x,y coupled together, Es bundled by what nu is promosing, maps over to V in indepdent ways 
+%E ^ V decomposed in one to one correspondances, \nu_0 is map from E_0 to V_0...\nu_n from E_n, V_n...., \nu = {\nu_0....\nu_n}
 
-\subsubsection{Visual Composition}
+\paragraph{Allowable Maps}
+For example, stevens: preserve structure ....
+Some examples of allowable maps are described in Bertin/Munzner:
+nus are constrained to preserve..., expect functorial nus - transform on input side to transform on output side (E1-E1, V1-V1, nu) <- condition
+that's how we're constraining nu. 
+functorial in $\nu$ component
+
+\subsubsection{Constructing Marks}
 The visual fiber bundle $V$ gets pulled back over $S$ via $\xi$ such that 
@@
 such that the composition $Q$ of $\mu$ is equivalent to the render $\rho(s) = Q\circ\mu(s)$. 
+(In practice $\mu \approxeq \xi^*\mu$)
+$Q$ is the constructor for the mark, 
+one $Q$ per artist, A = sum of artists
+$Q$ as there's default line, point, etc that's stretched and tweaked and whatever else by the $\mu$
+visual variable + structure preserving, Q must have map that commutes such that a change in $\mu$ carries over somehow to $\rho$, functorial in its arguments - commutative diagram - into rho
 

### 2020-12-22 draft/sections/math.tex
what is P"

@@
 \subsection{Fiber Bundles}
-We provide a brief introduction to fiber bundles because we model data, visual transformations, and a prerendered visual graphic as fiber bundles.
+We provide a brief description of fiber bundles because we model data, visual transformations, and a prerendered visual graphic as fiber bundles.
 A fiber bundle is a structure $(E, B, \pi, F)$  consisting of topological total space $E$, base space $K$, fiber space $F$ and the map from total space to base space:
@@
 
+\subsubsection{Section}
 The section $f$ is the mapping from base space to total space $f: B\rightarrow E$ 
@@
 
-such that it is the right inverse of $\pi$
+such that $f$ is the right inverse of $\pi$
 \begin{equation}
@@
 \subsection{Data Model}
-As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle:
-\begin{equation}
-    \label{eq:data_fiber_bundle}
-    \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[r, "\pi"] & B
-    \end{tikzcd}
-\end{equation}
-
-with a total space $E$, fiber space $F$, and base space $K$ which can be thought of as a set of keys $k$.
-The section $\tau:K \rightarrow E$ is an instance of the data that lies in $E$ and is discussed in section~\ref{sec:data_fiber}.
+As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle (E, K, $\pi$, F) with $\pi: E \rightarrow K$ where $K$ which can be thought of as a set of keys $k$.
+A section $\tau:K \rightarrow E$ is an instance of the data that lies in $E$ and is discussed in section~\ref{sec:data_fiber}.
 
@@
 \label{sec:display}
+We model the prerender space on which lives on ideal version of the visulaization as a fiber bundle (H, S, $\pi$, D).
+$H$ is the predisplay space, with a fiber $D$ dependent on the target physical display and a base space of $S$.
 
-Every point $k \in K$ maps to a space $S_{k} \in S$, which is the topology of the output of the artist $A$.
-The space $H$ is a total space representing the predisplay space, with a fiber dependent on the render space and a base space of $S$:
+\subsubsection{Base space}
+$K$ can be considered a subspace of the screen base space $S$ such that $\xi: S \rightarrow K$ is a deformation retraction \cite{RetractionTopology2020}
 \begin{equation}
     \begin{tikzcd}
-        D \arrow[r, hook] & H \arrow[r, "\pi"] &S\\   
+        E \arrow[d, "\pi"'] & H \arrow[d, "\pi"'] \\
+        K                   & S \arrow[l, "\xi"']
         \end{tikzcd}
@@
 
-\subsubsection{Section $\rho$}
+that goes from a region $s \in S_{k}$ to its associated point $k$, such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
 
-A section $\rho: S \rightarrow H$
-\begin{equation}
-    \begin{tikzcd}
-        D \arrow[r, hook] & H \arrow[d, "\pi"']              \\
-                          & S \arrow[u, "\rho"', bend right]
-        \end{tikzcd}
-\end{equation}
-
- is a mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to visual space as defined in D.
+\subsubsection{Fiber and Section}
+A section $\rho: S \rightarrow H$ is a mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to visual space as defined in D.
 For a physical screen display, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$ such that $\rho$ is
@@
 The artist decomposes to data $E$ to visual mappings $V$ and $V$ mappings composited to a visual in prerender space $H$.
-In order to define $A$, we first define the visual fiber bundle $V$ that shares basespace $K$ with $E$:
+In order to define $A$, we first define the visual fiber bundle $V$ that shares basespace $K$ with $E$ as ($V$,$K$, $\pi$, $P$) with section $\mu: V \rightarrow K$.
+The fiber $P$ is the space of visual variables $\mu$ \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} that can be encoded in the visualization being generated by $A$.
 
-\begin{equation}
-    \begin{tikzcd}
-        P \arrow[r, hook] & V \arrow[r, "\pi"] & K\\   
-    \end{tikzcd}
-\end{equation}
-
-with fiber $P$ that encodes visual parameters such as position or color. The section $\mu: V \rightarrow K$
-
-\begin{equation}
-    \begin{tikzcd}
-        P \arrow[r, hook] & V \arrow[d, "\pi"']              \\
-                          & K \arrow[u, "\mu"', bend right]
-        \end{tikzcd}
-\end{equation}
-
-returns a record in $P$. The fiber $P$ is the space of visual parameters %% defined like P, dependent on Q, P is targeting our library
-
-\subsubsection{Data Screen Mapping}
-The data bases pace $K$ can be considered a subspace of the screen base space $S$ such that $\xi: S \rightarrow K$ is a deformation retraction
-\begin{equation}
-    \begin{tikzcd}
-        E \arrow[d, "\pi"'] & H \arrow[d, "\pi"'] \\
-        K                   & S \arrow[l, "\xi"']
-        \end{tikzcd}
-\end{equation}
-
-that goes from a region $s \in S_{k}$ to its associated point $k$, such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
+\subsubsection{Example}
+If for example, a line artist as defined in \cite{Matplotlib} only supported the visual mappings xposition, yposition, width, color, style; then the $P$ for this line artist would be:
+\begin{itemize}
+    \item $C = \{\text{xposition, yposition, linewidth, color, linestyle}\}$ 
+    \item $DT =\{\mathbb{R}, \mathbb{R}, \mathbb{R}, \mathbb{R}^3, f\}$ 
+    \item $\sigma: C \rightarrow DT$ 
+    \item $P = \Gamma(\sigma) = \Gamma(\mu)$ 
+\end{itemize}
 

### 2020-12-22 draft/sections/math.tex
more p

@@
 
-In this section we will define the artist as a mapping from the sheaf $O(E)$ representing the data to a pre-render space $O(H)$. 
+In this section we will define the artist as a mapping from a sheaf $\mathcal{O}(E)$  to $\mathcal{O}(H)$. 
 \begin{equation}
@@
 \end{equation}
-The artist decomposes to data $E$ to visual mappings $V$ and $V$ mappings composited to a visual in prerender space $H$.
-In order to define $A$, we first define the visual fiber bundle $V$ that shares basespace $K$ with $E$ as ($V$,$K$, $\pi$, $P$) with section $\mu: V \rightarrow K$.
-The fiber $P$ is the space of visual variables $\mu$ \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} that can be encoded in the visualization being generated by $A$.
+The artist decomposes to data $E$ to visual mappings $V$ and $V$ mappings pulled back along $\xi$ to $\xi^*V$ which is then composited to a visual in prerender space $H$.
+
+\begin{equation}
+    \label{eq:artist}
+\begin{tikzcd}
+    E \arrow[r, "\nu"] \arrow[rd, "\pi"'] & V \arrow[d, "\pi"] & \xi^*V \arrow[r, "Q"] \arrow[d, "\xi^*\pi"'] & H \arrow[ld, "\pi"] \\
+                                          & K                  & S \arrow[l, "\xi"']                          &                    
+\end{tikzcd}
+\end{equation}
+
+The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable $\mu$ \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber $P$.
+The space of visual variables types in $P$ is restricted to the types $Q$ takes as input.
 
 \subsubsection{Example} 
-If for example, a line artist as defined in \cite{Matplotlib} only supported the visual mappings xposition, yposition, width, color, style; then the $P$ for this line artist would be:
+Given a  $Q$ that manufactored lines with signature $Q(\mu_{xpos}, \mu_{ypos}, \mu_{width}, \mu_{color}, \mu_{style})$ then the $P=\Gamma(\mu)$ for this line artist would be:
 \begin{itemize}
-    \item $C = \{\text{xposition, yposition, linewidth, color, linestyle}\}$ 
-    \item $DT =\{\mathbb{R}, \mathbb{R}, \mathbb{R}, \mathbb{R}^3, f\}$ 
-    \item $\sigma: C \rightarrow DT$ 
-    \item $P = \Gamma(\sigma) = \Gamma(\mu)$ 
+    \item $C = \{\text{xpos, ypos, width, color, style}\}$ 
+    \item $DT =\{\mathbb{R}, \mathbb{R}, \mathbb{R}, \mathbb{R}^3, f:s\mapsto\mathbb{R}^3 \}$ 
+    \item $P = U \times_{DT} C$
 \end{itemize}
@@
 \subsubsection{Visual Channels}
-
 The function $\nu: E \rightarrow V$ is a mapping from the data bundle to the visual bundle 
@@
 
-\subsubsection{Pipeline}
-The full visualization path is
-\begin{equation} 
-    \begin{tikzcd}
-        E \arrow[r, "\nu=transforms"] & V                                                              & \xi^*V \arrow[r, "Q = \text{set attrs}"]                             & H \\
-                                  & K \arrow[lu, "\tau = data"] \arrow[u, "{\mu = \{\tau,\nu\}}"'] & S \arrow[l, "\xi - init?draw"] \arrow[u, "\xi^*\mu = \nu\circ\tau"'] &  
-    \end{tikzcd}
-\end{equation}
-
 \end{document}

### 2020-12-23 draft/sections/math.tex
further down the defining types in P rabbit hole

@@
 The space of visual variables types in $P$ is restricted to the types $Q$ takes as input.
+\subsubsection{Retinal Variables}
+Visual Variables need to ultimately target some aspect of D
 
 \subsubsection{Example} 
-Given a  $Q$ that manufactored lines with signature $Q(\mu_{xpos}, \mu_{ypos}, \mu_{width}, \mu_{color}, \mu_{style})$ then the $P=\Gamma(\mu)$ for this line artist would be:
+Given a $Q$ with signature $Q(\mu_{xpos}, \mu_{ypos}, \mu_{width}, \mu_{color}, \mu_{style})$ then the $P=\Gamma(\mu)$ for this artist would be:
 \begin{itemize}

### 2020-12-23 draft/sections/math.tex
table of types in P for mpl

--- 2020-12-22 draft/sections/math.tex
@@
 The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable $\mu$ \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber $P$.
-The space of visual variables types in $P$ is restricted to the types $Q$ takes as input.
+$P$ is the cartesian product of types $U \times_{DT} C$ where $C$ is the set of parameter $\mu_param$ arguments into $Q$ and $DT$ is defined in terms of valid intermediate types.
 
-\subsubsection{Example} 
-Given a  $Q$ that manufactored lines with signature $Q(\mu_{xpos}, \mu_{ypos}, \mu_{width}, \mu_{color}, \mu_{style})$ then the $P=\Gamma(\mu)$ for this line artist would be:
-\begin{itemize}
-    \item $C = \{\text{xpos, ypos, width, color, style}\}$ 
-    \item $DT =\{\mathbb{R}, \mathbb{R}, \mathbb{R}, \mathbb{R}^3, f:s\mapsto\mathbb{R}^3 \}$ 
-    \item $P = U \times_{DT} C$
-\end{itemize}
+\subsubsection {Example P}
+For example, for Matplotlib \cite{hunterMatplotlib2DGraphics2007}:
+\begin{table}[]
+    \begin{tabular}{llll}
+    retinal variable         & $T \in DT$  & $\pi^{-1}(T) \subset U$                                 & example $C$                                                     \\
+    position                 & position    & $\mathbb{R}$                                              & x, y, z                                                         \\
+    size                     & size        & $\mathbb{R}$                                              & linewidth, markersize                                            \\
+    shape                    & markerstyle & $(\textrm{Strings})$                                      & markerstyle                                                     \\
+    color                    & color       & $(\mathbb{R}^3 \sqcup \mathbb{R}\sqcup \textrm{Strings})$ & color, colors, linecolor, facecolor, markerfacecolor, edgecolor \\
+    \multirow{2}{*}{texture} & hatch       & $(\textrm{Strings})$                                      & hatch                                                           \\
+                             & linestyle   & $(\textrm{Strings}\sqcup(\mathbb{ZS}, (\mathbb{Z}^*)))$                           & linestyle                                                      
+    \end{tabular}
+\end{table}
 

### 2020-12-24 draft/sections/math.tex
specifying the domain of $nu$

--- 2020-12-22 draft/sections/math.tex
@@
 \end{equation}
-The artist decomposes to data $E$ to visual mappings $V$ and $V$ mappings pulled back along $\xi$ to $\xi^*V$ which is then composited to a visual in prerender space $H$.
+
+The artist decomposes to mapping data to visual $\nu:E\rightarrow V$, then  compositing $V$ pulled back along $\xi$ to $\xi^*V$ to a visual mark in prerender space $Q:\xi^*V\rightarrow H$.
 
@@
 
-The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable $\mu$ \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber $P$.
-The space of visual variables types in $P$ is restricted to the types $Q$ takes as input.
+The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber $P$.
+The visual transformer $\nu$ is a set of functions each targeting a different $\mu$
+\begin{equation}
+    \label{eq:nu_expanded}
+    \nu:\{\nu_{0}, \ldots, \nu_{n}\} \rightarrow \{\mu_{0}, \ldots, \mu_{n}\}
+\end{equation}
 
-\subsubsection{Example} 
-Given a  $Q$ that manufactored lines with signature $Q(\mu_{xpos}, \mu_{ypos}, \mu_{width}, \mu_{color}, \mu_{style})$ then the $P=\Gamma(\mu)$ for this line artist would be:
-\begin{itemize}
-    \item $C = \{\text{xpos, ypos, width, color, style}\}$ 
-    \item $DT =\{\mathbb{R}, \mathbb{R}, \mathbb{R}, \mathbb{R}^3, f:s\mapsto\mathbb{R}^3 \}$ 
-    \item $P = U \times_{DT} C$
-\end{itemize}
+where $\mu_{i}$ are the visual parameters in the assembly function $Q(\mu_{0}, \ldots, \mu_{n})(s) = \rho(s)$.
+
+\subsubsection {Example: Matplotlib Visual Fiber}
+For example, for Matplotlib \cite{hunterMatplotlib2DGraphics2007}, some of the possible types in $P$ are:
+\begin{table}[ht]
+    \renewcommand{\arraystretch}{2}
+    \begin{tabulary}{\textwidth}{|l|C|L|}\hline
+     $\nu_{i}$                      & $\mu_{i}$                                                            & $dom(\nu_{i})$  \\ \hline                                              
+    position                    & x, y, z                                                          & $\mathbb{R}$   \\ \hline
+    size                        & linewidth, markersize                                            & $\mathbb{R}$   \\ \hline
+    shape                       & markerstyle                                                      & $((\mathbb{R}, \mathbb{R}), x \in \{0,1,2,3,4,79\})^{n}^{\dag}$ \\ \hline
+    color                       & color, colors, linecolor, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^4$ \\ \hline
+    \multirow{2}{*}{texture}    & hatch                                                            & $/^*\backslash^*-^*+^*x^*o^*O^*.^**^* \in \Sigma^{+}$ \newline $\Sigma=\{/, \backslash, |, -, +, x, o, O, ., *\}^{\ddag}$\\\cline{2-3}
+                                & linestyle                                                        & $(\mathbb{Z}, (\mathbb{Z}^n))^{\dag}$ \\ \hline              
+    \end{tabulary}
+    \caption{\\
+        \footnotesize{\dag $n>0$}\\
+        \footnotesize{\ddag$\Sigma^+=\cup_{n\in\mathbb{Z}^+}\Sigma^n$}
+
+    }
+\end{table}
 
@@
 \begin{equation}
+    \label{eq:nu_categorical}
 \begin{tikzcd}
-        E \arrow[d, "\pi"'] \arrow[r, "\nu"] & V \arrow[ld, "\pi"'] \\
-        K                                    &                     
+    E_1 \arrow[r] \arrow[r, "\nu"] \arrow[d, "f"'] & V_1 \arrow[d, "f"] \\
+    E_2 \arrow[r, "\nu"]                           & V_2               
 \end{tikzcd}
@@
 such that a visual section $\mu: K \rightarrow V$  is equivalent to the visual transformation of the data $\mu(k) = \nu\circ\tau(k)$
+
++2 on data side means +2 on visual side
+
+.2 = E1 - lambda e -> .2 V1 = .2
++.2                              +.2
+.4 = E2 - lambda e -> .2 V2 = .2
+
+.2 = E1  - lambda e-> .2 V1 = .2
+
+                        V2 = .2
+
+1 = E1 - lambda e->e/5 = .25
+
++ 1                      +.25
+
+2 = E2 - lambda e->e/5 = .5
+
+preserve unique mapping? does $\nu$ need to be invertable?
+
+shifts in either space need to be mirrored in the other space, this is not valid  here's why:
+$\lambda e: .2$ 
+\begin{equation}
+    \label{eq:nu_equation_bad}
+    \begin{tikzcd}
+        E_1 \arrow[r] \arrow[r, "\lambda:e\mapsto.5"] \arrow[d, "e+2"'] & V_1 \arrow[d, "h+2"] \\
+        E_2 \arrow[r, "\lambda"]                                        & V_2                 
+    \end{tikzcd}
+\end{equation}
+
+{x,y} = {1,2}
+{y,x} = {2,1}
+{y,x} = {1,2}
 

### 2020-12-24 draft/sections/math.tex
forget new packages + formatting

@@
 \begin{table}[ht]
+    \label{tab:mpl_visual_variable_fiber}
+\begin{threeparttable}
     \renewcommand{\arraystretch}{2}
-    \begin{tabulary}{\textwidth}{|l|C|L|}\hline
-     $\nu_{i}$                      & $\mu_{i}$                                                            & $dom(\nu_{i})$  \\ \hline                                              
+    \begin{tabulary}{\textwidth}{|l|L|L|}\hline
+     $\bm{\nu_{i}}$                      & $\bm{\mu_{i}}$                                                            & $\bm{dom(\nu_{i})}$  \\ \hline                                              
     position                    & x, y, z                                                          & $\mathbb{R}$   \\ \hline
     size                        & linewidth, markersize                                            & $\mathbb{R}$   \\ \hline
-    shape                       & markerstyle                                                      & $((\mathbb{R}, \mathbb{R}), x \in \{0,1,2,3,4,79\})^{n}^{\dag}$ \\ \hline
+    shape                       & markerstyle                                                      & $((\mathbb{R}, \mathbb{R}), x \in \{0,1,2,3,4,79\})^{n}$\tnote{\dag} \\ \hline
     color                       & color, colors, linecolor, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^4$ \\ \hline
-    \multirow{2}{*}{texture}    & hatch                                                            & $/^*\backslash^*-^*+^*x^*o^*O^*.^**^* \in \Sigma^{+}$ \newline $\Sigma=\{/, \backslash, |, -, +, x, o, O, ., *\}^{\ddag}$\\\cline{2-3}
-                                & linestyle                                                        & $(\mathbb{Z}, (\mathbb{Z}^n))^{\dag}$ \\ \hline              
+    \multirow{2}{*}{texture}    & hatch                                                            & $/^*\backslash^*-^*+^*x^*o^*O^*.^**^* \in \Sigma^{+}$ \newline $\Sigma=\{/, \backslash, |, -, +, x, o, O, ., *\}$ \tnote{\ddag}\\\cline{2-3}
+                                & linestyle                                                        & $(\mathbb{Z}, (\mathbb{Z}^n))$\tnote{\dag} \\ \hline              
     \end{tabulary}
-    \caption{\\
-        \footnotesize{\dag $n>0$}\\
-        \footnotesize{\ddag$\Sigma^+=\cup_{n\in\mathbb{Z}^+}\Sigma^n$}
+    \begin{tablenotes}\footnotesize
+        \item[\dag] $n>0$
+        \item[\ddag]$\Sigma^+=\cup_{n\in\mathbb{Z}^+}\Sigma^n$
+    \end{tablenotes}
 
-    }
+\end{threeparttable}
 \end{table}

### 2020-12-29 draft/sections/math.tex
trying to pin action versus versus group

--- 2020-12-24 draft/sections/math.tex
@@
 \begin{table}[ht]
+    \label{tab:mpl_visual_variable_fiber}
     \renewcommand{\arraystretch}{2}
-    \begin{tabulary}{\textwidth}{|l|C|L|}\hline
-     $\nu_{i}$                      & $\mu_{i}$                                                            & $dom(\nu_{i})$  \\ \hline                                              
-    position                    & x, y, z                                                          & $\mathbb{R}$   \\ \hline
-    size                        & linewidth, markersize                                            & $\mathbb{R}$   \\ \hline
-    shape                       & markerstyle                                                      & $((\mathbb{R}, \mathbb{R}), x \in \{0,1,2,3,4,79\})^{n}^{\dag}$ \\ \hline
+    \begin{tabulary}{\textwidth}{|l|L|l|}\hline
+     $\bm{\nu_{i}}$                      & $\bm{\mu_{i}}$                                                            & $\bm{codomain(\nu_{i})}$  \\ \hline                                              
+    position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
+    size                        & linewidth, markersize                                            & $\mathbb{R}^{+}$   \\ \hline
+    shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
     color                       & color, colors, linecolor, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^4$ \\ \hline
-    \multirow{2}{*}{texture}    & hatch                                                            & $/^*\backslash^*-^*+^*x^*o^*O^*.^**^* \in \Sigma^{+}$ \newline $\Sigma=\{/, \backslash, |, -, +, x, o, O, ., *\}^{\ddag}$\\\cline{2-3}
-                                & linestyle                                                        & $(\mathbb{Z}, (\mathbb{Z}^n))^{\dag}$ \\ \hline              
+    \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
+                                & linestyle                                                        & $(\mathbb{N}, \mathbb{N}^{n})$ \\ \hline              
     \end{tabulary}
-    \caption{\\
-        \footnotesize{\dag $n>0$}\\
-        \footnotesize{\ddag$\Sigma^+=\cup_{n\in\mathbb{Z}^+}\Sigma^n$}
-
-    }
 \end{table}
@@
 \subsubsection{Visual Channels}
+
+A monoid is a set with a closed operation
+
 The function $\nu: E \rightarrow V$ is a mapping from the data bundle to the visual bundle 
@@
 \begin{tikzcd}
-    E_1 \arrow[r] \arrow[r, "\nu"] \arrow[d, "f"'] & V_1 \arrow[d, "f"] \\
-    E_2 \arrow[r, "\nu"]                           & V_2               
+    E_i \arrow[r] \arrow[r, "\nu_i"] \arrow[d, "f"'] & V_i \arrow[d, "f"] \\
+    E_i \arrow[r, "\nu_i"]                           & V_i               
 \end{tikzcd}
@@
 such that a visual section $\mu: K \rightarrow V$  is equivalent to the visual transformation of the data $\mu(k) = \nu\circ\tau(k)$
+
+The Stevens measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} are examples of some of the group actions th
 
@@
 
-%%throw out this diagram
-%%something about pushforward ...
-%% \mu is pushfowwards of \tau = \mu = \nu(lowercase * ,\tau), \nu: *\tau \mapsto \mu
-%% \nu acts from subspace E to subspace V
-The constraint on $\nu: \tau\rightarrow \mu$
-
-%%set up proofs here, are x&y transfereble here? (x,y coupled together, Es bundled by what nu is promosing, maps over to V in indepdent ways 
-%E ^ V decomposed in one to one correspondances, \nu_0 is map from E_0 to V_0...\nu_n from E_n, V_n...., \nu = {\nu_0....\nu_n}
-
-\paragraph{Allowable Maps}
-For example, stevens: preserve structure ....
-Some examples of allowable maps are described in Bertin/Munzner:
-nus are constrained to preserve..., expect functorial nus - transform on input side to transform on output side (E1-E1, V1-V1, nu) <- condition
-that's how we're constraining nu. 
-functorial in $\nu$ component
-
 \subsubsection{Constructing Marks}

### 2020-12-29 draft/sections/visual.tex
trying to pin action versus versus group

\documentclass[../main.tex]{subfiles}
\begin{document}
\subsubsection{Marks}
Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space are the point, line, and area marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}. For each region $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$. 
\begin{equation}
\begin{tikzcd}
    H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \Gamma \text{ s.t. } \Gamma(0)=k \text{ and }\Gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
\end{tikzcd}
\label{eq:mark}
\end{equation}
To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.

\subsubsection{Visual Idioms: Equivalance class of artists}
n $O(E)$ of the same type, they output the same type of prerender $O(H)$:

Natural transformation + composition is partial ordering? Back and forth is equivalent 
\end{document}

### 2020-12-30 draft/sections/math.tex
start of scratching out \nu

@@
 \subsubsection{Visual Channels}
-
-A monoid is a set with a closed operation
-
-The function $\nu: E \rightarrow V$ is a mapping from the data bundle to the visual bundle 
+$\nu: \tau \mapsto \mu$ is an equivariant map such that there is a homomorphism from monoid actions on $E_{i}$ to manoid actions on $V_{i}$  
 \begin{equation}
@@
 \end{equation}
+where the diagram commutes such that  $\nu g(\tau_{i}) = g(\nu(tau_{i})$. Manoid actions are:
+1) set of functions including identity, composition operator, closed, associative
 
-such that a visual section $\mu: K \rightarrow V$  is equivalent to the visual transformation of the data $\mu(k) = \nu\circ\tau(k)$
+To construct a valid $\nu$:
+\begin{enumerate}
+    \item identify monoid actions on $E_i$
+    \item choose $V_i$ that can support same monoid actions
+    \item implement $\nu$ such that composition holds
+\end{enumerate}
 
-The Stevens measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} are examples of some of the group actions th
+\paragraph{Example: Translation}
+When $tau_{i}$ of $E_1$ returns interval data, Steven's defines the group actions on the data as the general linear group \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}; therefore to construct a valid $\nu$, we choose a target variable in $P$ that is
 
-+2 on data side means +2 on visual side
+$\nu_i(\tau_i) = \tau + 2$
 
-.2 = E1 - lambda e -> .2 V1 = .2
-+.2                              +.2
-.4 = E2 - lambda e -> .2 V2 = .2
-
-.2 = E1  - lambda e-> .2 V1 = .2
-
-                        V2 = .2
-
-1 = E1 - lambda e->e/5 = .25
-
-+ 1                      +.25
-
-2 = E2 - lambda e->e/5 = .5
-
-preserve unique mapping? does $\nu$ need to be invertable?
-
+$\lambda e: .2$ 
+\paragraph{Example: Invalid $\nu$}
 shifts in either space need to be mirrored in the other space, this is not valid  here's why:
-$\lambda e: .2$ 
 \begin{equation}
@@
 
-{x,y} = {1,2}
-{y,x} = {2,1}
-{y,x} = {1,2}
+\paragraph{Example: Partial Order}
+$\nu$ induces a monoid homomorphism, rather than a group homomorphism, to allow for partial ordering. For example if E1 is the set '{}
 

### 2020-12-31 draft/sections/abstract.tex
abstract

The Python Matplotlib visualization library grew organically such that the visualization functions are deeply entangled with the data the functions expect in such a way that it has become hard to adapt the library for modern needs. To rearchtecture the library while not sacrificing its current very high degree of flexiblity, we developed a model that preserves the structure of the data while also putting very few constraints on the data it can encode.

We propose that afunction that converts data into a graphic is the composition of two functions: a set of equivariant maps from data to visual channels and an equivariant map that assembles the visual channels into a graphic renderable by a physical display. We encode data as a fiber bundle to separately consider the fields of the data (the variables) as the fiber and the connectivity as the base space. We also represent prerender display spaces as a fiber bundle with a base space that is a continous map into the data base space, thereby preserving the connectivity of the data in the graphic.

We then illustrate this model through case studies of a scatter plot, line plot, and heatmap. Finally we show that this model can be realized in software with a small prototype. We encode the topological base spaces using triangulation, make use of programming types for the fiber, and build Matplotlib's existing infrastructure for the rendering.

In future work, we propose to develop the math further such that we can represent graphics that are composites of primitives and implement those ideas as user facing composite visualization functions.

### 2020-12-31 draft/sections/implementation.tex
pictures & not so many words
abstract

@@
 
-\begin{figure}[ht]
-    \label{fig:simplex}
-    \includegraphics{figures/sections/math/simplex.png}
-    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
-Higher order simplicies are outside the scope of this paper.}
-\end{figure}
+\subsection{Data Model}
 
-%% k is a triangularizable topological space 
-One way to represent the topological space K is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
-Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another: %%formula for simplacial complexes/simplacial sets
-
-% replace this table w/ a diagram that shows this 
-\begin{description}
-    \item[0-simplex] discrete observations (inventory records)
-    \item[1-simplex] 1D continuos data (timeseries)
-    \item[2-simplex] 2D continuos data (map)
-    \item[3-simplex] 3D continuos data (video)
-\end{description}
 \end{document}

### 2020-12-31 draft/sections/math.tex
abstract

@@
 The same dataset can be expressed with different $K$.
+
+In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
+
+\begin{figure}[ht]
+    \label{fig:simplex}
+    \includegraphics{figures/sections/math/simplex.png}
+    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
+Higher order simplicies are outside the scope of this paper.}
+\end{figure}
+
+%% k is a triangularizable topological space 
+One way to represent the topological space K is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
+Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another: %%formula for simplacial complexes/simplacial sets
+
+% replace this table w/ a diagram that shows this 
+\begin{description}
+    \item[0-simplex] discrete observations (inventory records)
+    \item[1-simplex] 1D continuos data (timeseries)
+    \item[2-simplex] 2D continuos data (map)
+    \item[3-simplex] 3D continuos data (video)
+\end{description}
 
@@
 \end{equation}
-where the diagram commutes such that  $\nu g(\tau_{i}) = g(\nu(tau_{i})$. Manoid actions are:
+where the diagram commutes such that  $\nu g(\tau_{i}) = g(\nu(tau_{i})$. Monoid actions are:
 1) set of functions including identity, composition operator, closed, associative

### 2020-12-31 draft/sections/abstract.tex
slight rewording

@@
 The Python Matplotlib visualization library grew organically such that the visualization functions are deeply entangled with the data the functions expect in such a way that it has become hard to adapt the library for modern needs.
-To rearchtecture the library while not sacrificing its current very high degree of flexiblity, we developed a model that preserves the structure of the data while also putting very few constraints on the data it can encode.
+To rearchtecture the library while not sacrificing its current very high degree of flexiblity, we developed a model of visualization that preserves the structure of the data while also putting very few constraints on the data it can encode.
 
-We propose that afunction that converts data into a graphic is the composition of two functions: a set of equivariant maps from data to visual channels and an equivariant map that assembles the visual channels into a graphic renderable by a physical display.
+We propose that a function that converts data into a graphic is the composition of two functions: a set of equivariant maps from data to visual channels and an equivariant map that assembles the visual channels into a graphic renderable by a physical display.
 We encode data as a fiber bundle to separately consider the fields of the data (the variables) as the fiber and the connectivity as the base space.

### 2020-12-31 draft/sections/abstract.tex
revised abstract

--- 2020-12-31 draft/sections/abstract.tex
@@
-The Python Matplotlib visualization library grew organically such that the visualization functions are deeply entangled with the data the functions expect in such a way that it has become hard to adapt the library for modern needs.
-To rearchtecture the library while not sacrificing its current very high degree of flexiblity, we developed a model that preserves the structure of the data while also putting very few constraints on the data it can encode.
+The difference between drawing and data visualization is that visualization implies that some structure of the data is preserved in the visual presentation.
+We formalize what structure is preserved and how via a mathematical topological artist model.
+In our model, data and graphics can be viewed as sections of fiber bundles; this allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that encode the connectivity of the points in the dataset and elements of the graphic.
+Furthermore, this model supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
+We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap.
+We show that this model can be implemented with a small prototype.
 
-We propose that afunction that converts data into a graphic is the composition of two functions: a set of equivariant maps from data to visual channels and an equivariant map that assembles the visual channels into a graphic renderable by a physical display.
-We encode data as a fiber bundle to separately consider the fields of the data (the variables) as the fiber and the connectivity as the base space.
-We also represent prerender display spaces as a fiber bundle with a base space that is a continous map into the data base space, thereby preserving the connectivity of the data in the graphic.
-
-We then illustrate this model through case studies of a scatter plot, line plot, and heatmap.
-Finally we show that this model can be realized in software with a small prototype.
-We encode the topological base spaces using triangulation, make use of programming types for the fiber, and build Matplotlib's existing infrastructure for the rendering.
-
-In future work, we propose to develop the math further such that we can represent graphics that are composites of primitives and implement those ideas as user facing composite visualization functions.
+We call the function mapping sections of the data bundle to the graphical bundle the "Artist" based on analogous architecture elements in the Python visualization library Matplotlib.
+To demonstrate the practical value of the model, we propose a model driven re-architecture of the artist layer of the Matplotlib.
+We represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
+In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
+The thesis will follow through on this proposal to explore how to further develop the model, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.

### 2021-01-01 draft/sections/abstract.tex

@@
-The difference between drawing and data visualization is that visualization implies that some structure of the data is preserved in the visual presentation.
-We formalize what structure is preserved and how via a mathematical topological artist model.
-In our model, data and graphics can be viewed as sections of fiber bundles; this allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that encode the connectivity of the points in the dataset and elements of the graphic.
+The difference between drawing and data visualization is that visualization implies that some structure of the data is preserved in the visual presentation.* We formalize what structure is preserved and how via a mathematical topological artist model.
+In our model, data and graphics can be viewed as sections of fiber bundles; this allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that encode the preserving the dataset connectivity into graphical elements.
 Furthermore, this model supports an algebraic sum operation such that more complex visualizations can be built from simple ones.

### 2021-01-01 draft/sections/abstract.tex
expressiveness

@@
-The difference between drawing and data visualization is that visualization implies that some structure of the data is preserved in the visual presentation.* We formalize what structure is preserved and how via a mathematical topological artist model.
-In our model, data and graphics can be viewed as sections of fiber bundles; this allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that encode the preserving the dataset connectivity into graphical elements.
-Furthermore, this model supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
+Potentially the most important property of a data visualization is its expressiveness.
+We propose to formalize expressiveness via a topological artist model (TAM).
+In this model, data and graphics can be viewed as sections of fiber bundles; this allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
+Furthermore, TAM supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
 We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap.
-We show that this model can be implemented with a small prototype.
+We show that TAM can be implemented with a small prototype.
 
-We call the function mapping sections of the data bundle to the graphical bundle the "Artist" based on analogous architecture elements in the Python visualization library Matplotlib.
-To demonstrate the practical value of the model, we propose a model driven re-architecture of the artist layer of the Matplotlib.
+We call the function mapping sections of the data bundle to the graphical bundle the "Artist" based on the analogous architecture elements in the Python visualization library Matplotlib.
+To demonstrate the practical value of our model, we propose a TAM driven re-architecture of the artist layer of Matplotlib.
 We represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
 In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
-The thesis will follow through on this proposal to explore how to further develop the model, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.
+The thesis will follow through on this proposal to explore how to further develop TAM, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.

### 2021-01-01 draft/sections/abstract.tex
naturalness

@@
-Potentially the most important property of a data visualization is its expressiveness.
-We propose to formalize expressiveness via a topological artist model (TAM).
-In this model, data and graphics can be viewed as sections of fiber bundles; this allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
+ A critical aspect of data visualization is that the graphical representation of data match the properties of the data, as stated in Norman's Naturalness Principal.
+One failure for numerical data is the Lie Factor, another for categorical data is ordering not being preserved.
+In this work, we propose that the mathematical notion of equivariance formalizes the naturalness principle.
+We developed a topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles.
+This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
 Furthermore, TAM supports an algebraic sum operation such that more complex visualizations can be built from simple ones.

### 2021-01-03 draft/sections/abstract.tex
abstracts

@@
- A critical aspect of data visualization is that the graphical representation of data match the properties of the data, as stated in Norman's Naturalness Principal.
-One failure for numerical data is the Lie Factor, another for categorical data is ordering not being preserved.
-In this work, we propose that the mathematical notion of equivariance formalizes the naturalness principle.
-We developed a topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles.
+ A critical aspect of data visualization is that the graphical representation of data match the properties of the data; this fails when order is not preserved in representations of ordinal data or scale for numerical data.
+In this work, we propose that the mathematical notion of equivariance formalizes the expectation that graphics match the data.
+We developed a topological model in which data and graphics can be viewed as sections of fiber bundles.
 This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
-Furthermore, TAM supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
+Furthermore, our model supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
 We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap.
-We show that TAM can be implemented with a small prototype.
+We show that this model can be implemented with a small prototype.
 
-We call the function mapping sections of the data bundle to the graphical bundle the "Artist" based on the analogous architecture elements in the Python visualization library Matplotlib.
-To demonstrate the practical value of our model, we propose a TAM driven re-architecture of the artist layer of Matplotlib.
+ To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of Matplotlib.
 We represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.

### 2021-01-04 draft/sections/abstract.tex
Update abstract.tex

@@
 In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
-The thesis will follow through on this proposal to explore how to further develop TAM, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.
+The thesis will follow through on this proposal to explore how to further develop our model, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.

### 2021-01-04 draft/sections/abstract.tex
Update abstract.tex

@@
 In this work, we propose that the mathematical notion of equivariance formalizes the expectation that graphics match the data.
-We developed a topological model in which data and graphics can be viewed as sections of fiber bundles.
+We developed a model we call the topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles.
 This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
@@
 
- To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of Matplotlib.
+To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
 We represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.

### 2021-01-04 draft/sections/math.tex
math

@@
     shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
-    color                       & color, colors, linecolor, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^4$ \\ \hline
+    color                       & color, linecolor, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
     \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
-                                & linestyle                                                        & $(\mathbb{N}, \mathbb{N}^{n})$ \\ \hline              
+                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
     \end{tabulary}
@@
 \end{equation}
-where the diagram commutes such that  $\nu g(\tau_{i}) = g(\nu(tau_{i})$. Monoid actions are:
-1) set of functions including identity, composition operator, closed, associative
-
-To construct a valid $\nu$:
-\begin{enumerate}
-    \item identify monoid actions on $E_i$
-    \item choose $V_i$ that can support same monoid actions
-    \item implement $\nu$ such that composition holds
-\end{enumerate}
+where the diagram commutes such that  $\nu_{i}\circ f(\tau_{i}) = f \circ \nu_{i}(\tau_{i})$. A monoid is a set $M$ with a closed associative binary operation ($\*$) and an identity element $e$.
 
 \paragraph{Example: Translation}
-When $tau_{i}$ of $E_1$ returns interval data, Steven's defines the group actions on the data as the general linear group \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}; therefore to construct a valid $\nu$, we choose a target variable in $P$ that is
+When $\tau_{i}$ of $E_1$ returns interval data, Steven's defines the group actions on the data as the general linear group \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}; therefore to construct a valid $\nu$, we choose a target variable in $P$ that is
 
@@
 \paragraph{Example: Partial Order}
-$\nu$ induces a monoid homomorphism, rather than a group homomorphism, to allow for partial ordering. For example if E1 is the set '{}
+$\nu$ induces a monoid homomorphism to allow for partial ordering. For example if E1 is the set '{}
 
@@
 
+Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space are the point, line, and area marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
+For each region $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
+\begin{equation}
+\begin{tikzcd}
+    H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \Gamma \text{ s.t.
+} \Gamma(0)=k \text{ and }\Gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
+\end{tikzcd}
+\label{eq:mark}
+\end{equation}
+To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
+
+\subsubsection{Visual Idioms: Equivalance class of artists}
+n $O(E)$ of the same type, they output the same type of prerender $O(H)$:
+
+Natural transformation + composition is partial ordering? Back and forth is equivalent 
 \end{document}

### 2021-01-06 draft/sections/math.tex
monoids

@@
     \label{eq:artist}
-    A: \mathcal{O}(E) \rightarrow \mathcal{O}(H)
+    A: \Gamma(E) \rightarrow \Gamma(H)
 \end{equation}
@@
 \item $S$ is a simplacial complex encoding the visualization
+\item $\Gamma$ is the global space of sections
 \end{itemize}
@@
 \subsubsection{Base space}
-\begin{figure}[ht]
+\begin{figure}[H]
     \label{fig:base_space}
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
     \label{fig:data_fiber_bundle}
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
     \label{fig:simplex}
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
     \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
@@
 \subsubsection{Example}
-\begin{figure}[ht]
+\begin{figure}[H]
     \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2f.png}
@@
 \end{figure}
-\begin{figure}[ht]
+\begin{figure}[H]
     \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_3f.png}
@@
 \subsubsection{{Example}}
-\begin{figure}[h]
+\begin{figure}[H]
     \includegraphics[width=.4\linewidth]{figures/sections/math/render.png}
@@
     \label{eq:nu_expanded}
-    \nu:\{\nu_{0}, \ldots, \nu_{n}\} \rightarrow \{\mu_{0}, \ldots, \mu_{n}\}
+    \{\nu_{0}, \ldots, \nu_{n}\}: \{\tau_{0}, \ldots, \tau_{n}\} \mapsto \{\mu_{0}, \ldots, \mu_{n}\}
 \end{equation}
@@
     shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
-    color                       & color, linecolor, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
+    color                       & color, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
     \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
-                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
+                                & linestyle                                                        & $\{f_{0}, \ldots, f_{n}\} \times (\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
     \end{tabulary}
@@
 \subsubsection{Visual Channels}
-$\nu: \tau \mapsto \mu$ is an equivariant map such that there is a homomorphism from monoid actions on $E_{i}$ to manoid actions on $V_{i}$  
+$\nu: E \rightarrow V$ is an equivariant map such that there is a homomorphism from left monoid actions on $E_{i}$ to left monoid actions on $V_{i}$ where $i$ identifies a field in the fiber.
+A validly constructed $\nu$ is one where the  diagram
 \begin{equation}
@@
 \begin{tikzcd}
-    E_i \arrow[r] \arrow[r, "\nu_i"] \arrow[d, "f"'] & V_i \arrow[d, "f"] \\
+    E_i \arrow[r] \arrow[r, "\nu_i"] \arrow[d, "m"'] & V_i \arrow[d, "m"] \\
     E_i \arrow[r, "\nu_i"]                           & V_i               
@@
 \end{equation}
-where the diagram commutes such that  $\nu_{i}\circ f(\tau_{i}) = f \circ \nu_{i}(\tau_{i})$. A monoid is a set $M$ with a closed associative binary operation ($\*$) and an identity element $e$.
+ commutes such that $\nu_i(m(E_i)) = m(\nu_i(E_i))$ where $m$ is a monoid action. $E_i$ and $V_{i}$ each contain a set of values as defined in $F$ and $P$ respectively.
+
+\paragraph{Definition}
+A monoid\cite{Monoid2021} $M$ is a set that is closed under an associative binary operator $\ast$ and has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $X$ with an operation $\cdot$ with the monoid properties:
+\begin{description}
+    \item[closure] $\cdot: M\times X \rightarrow X$
+    \item[associativity] $\text{for all} m,t \in M \text{and} x\in X, m\cdot(t\cdot x) = (m\ast t)\cdot x$ 
+    \item[identity] $\text{for all} x\in X, e\in M,  e\cdot x = x$ 
+\end{description}
 
@@
 
-\subsubsection{Constructing Marks}
-The visual fiber bundle $V$ gets pulled back over $S$ via $\xi$ such that 
+\subsubsection{Assembling Marks}
+
+\begin{figure}[H]
+    \includegraphics[width=0.4\linewidth]{figures/sections/math/path_of_q}
+    \label{fig:q}
+    \caption{change $\rho$ to input J}
+\end{figure}
+
+The assembly function $Q: \Gamma(V) \rightarrow \Gamma(H)$ composites the visual variables in $V$ into an element in $H$.
+Given a monoid action on a set in $V$, there should be a monoid action on the corresponding $Q(\Gamma(V))$.
+<why gamma?
+
+Lets call $\Gamma(V) = X$ and $Q(\Gamma(V))$
+
+For each region $s$ in the display space $H$, the mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology} it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the path connected component $J \subset K$.
 \begin{equation}
     \begin{tikzcd}
-        \xi^*V \arrow[r, "Q"] \arrow[rd, "\xi^*\pi"'] & H \arrow[d, "\pi"] \\
-                                                      & S                 
-    \end{tikzcd}
-\end{equation}
-the function $Q:\xi^*V\rightarrow H$ composites points in $V$ into $\mathbb{R^7}$ tuples. The section $\mu$ is pulled over $s$
-
-\begin{equation}
-    \begin{tikzcd}
-        \xi^*V \arrow[r, "Q"] & H                                           \\
-                              & S \arrow[lu, "\xi^*\mu"] \arrow[u, "\rho"']
-    \end{tikzcd}
-\end{equation}
-
-such that the composition $Q$ of $\mu$ is equivalent to the render $\rho(s) = Q\circ\mu(s)$. 
-(In practice $\mu \approxeq \xi^*\mu$)
-$Q$ is the constructor for the mark, 
-one $Q$ per artist, A = sum of artists
-$Q$ as there's default line, point, etc that's stretched and tweaked and whatever else by the $\mu$
-visual variable + structure preserving, Q must have map that commutes such that a change in $\mu$ carries over somehow to $\rho$, functorial in its arguments - commutative diagram - into rho
-
-Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space are the point, line, and area marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
-For each region $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
-\begin{equation}
-\begin{tikzcd}
-    H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \Gamma \text{ s.t.
-} \Gamma(0)=k \text{ and }\Gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
+        H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \gamma \text{ s.t.
+} \gamma(0)=k \text{ and }\gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
     \end{tikzcd}
@@
 \end{equation}
-To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
+where $\gamma$ is a path where input is a point on the interval [0,1] and
+
+To get back to the display space $H$ from $J$, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then $\rho(\xi^{-1}(J))$ maps it into $R^{7}$ such that $Q\circ\xi^*\mu(\xi^{-1}(J))$ generates a graphical mark.
+
+%%abs: R \rightarrow R^+, 
+%%f: R \rightarrow R, 
+%%abs:[real->real] -> [real->real+]
+%%fmap?, fmap(abs)->ufunc on all functions
+%%lifting: decorator that only looks at arguments
+
+Q goes from one to the other, could need more information than just the return values...
+map on global sections means I might need to know the whole section, map V to H says I only need to know points, restrict to the case where defined on the stalk..but will look at particular cases - sections to sections is most general case that is true, but can be defined V to H or or O(V) to O(H) depending on the visualization.
 

### 2021-01-06 draft/sections/math.tex
math

@@
 $\nu: E \rightarrow V$ is an equivariant map such that there is a homomorphism from left monoid actions on $E_{i}$ to left monoid actions on $V_{i}$ where $i$ identifies a field in the fiber.
+$E_i$ and $V_{i}$ each contain a set of values as defined in $F$ and $P$ respectively.
 A validly constructed $\nu$ is one where the  diagram
@@
 \begin{tikzcd}
-    E_i \arrow[r] \arrow[r, "\nu_i"] \arrow[d, "m"'] & V_i \arrow[d, "m"] \\
+    E_i \arrow[r] \arrow[r, "\nu_i"] \arrow[d, "m_e"'] & V_i \arrow[d, "m_v"] \\
     E_i \arrow[r, "\nu_i"]                           & V_i               
@@
 \end{equation}
- commutes such that $\nu_i(m(E_i)) = m(\nu_i(E_i))$ where $m$ is a monoid action. $E_i$ and $V_{i}$ each contain a set of values as defined in $F$ and $P$ respectively.
+commutes such that $\nu_i(m_e(E_i)) = m_v(\nu_i(E_i))$.
 
@@
 A monoid\cite{Monoid2021} $M$ is a set that is closed under an associative binary operator $\ast$ and has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $X$ with an operation $\cdot$ with the monoid properties:
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $X$ with an action $\bullet$ with the properties:
 \begin{description}
-    \item[closure] $\cdot: M\times X \rightarrow X$
-    \item[associativity] $\text{for all} m,t \in M \text{and} x\in X, m\cdot(t\cdot x) = (m\ast t)\cdot x$ 
-    \item[identity] $\text{for all} x\in X, e\in M,  e\cdot x = x$ 
+    \item[closure] $\bullet: M\times X \rightarrow X$,
+    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in X, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
+    \item[identity] $\text{for all } x\in X, e\in M,  e\bullet x = x$ 
 \end{description}
 
+\paragraph{Example: Partial Order}
+To preserve ordering of elements in $E_i$, $\nu$ must be a monotonic function such that given $e_1, e_2 \in E_{i}$
+\begin{equation}
+\text{ if } e_1 \leq e_2 \text{ then } \nu(e_1) \leq \nu(e_2)
+\end{equation}
+
 \paragraph{Example: Translation}
-When $\tau_{i}$ of $E_1$ returns interval data, Steven's defines the group actions on the data as the general linear group \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}; therefore to construct a valid $\nu$, we choose a target variable in $P$ that is
+According to Stevens, interval data jas general linear group actions \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
+Position is a visual variable that can support translation \cite{bertinIIPropertiesGraphic2011,krygierMakingMapsVisual2005,munznerMarksChannels}.
 
-$\nu_i(\tau_i) = \tau + 2$
+For example, here $\nu$ is a direct map $nu(e) = v$ and the transform function is a multiplicative shift $t(x) = x + x$ applied twice
 
-$\lambda e: .2$ 
+\begin{align}
+    \nu(t(t(e))) &= t_v(t(\nu(e)))\\
+    \nu(t(e+e)) &= t(t_v(v))\\
+    \nu((e+e) + (e+e)) &= t(v+v)\\
+    \nu(4e) &= (v_v) + (v+v)\\
+    \nu(4e)&=4v
+\end{align}
+
+Here $\nu$ is valid because scaling the data by a factor of 4 causes a scaling of the equivalent visualization by 4.
+
 \paragraph{Example: Invalid $\nu$}
@@
 
-\paragraph{Example: Partial Order}
-$\nu$ induces a monoid homomorphism to allow for partial ordering. For example if E1 is the set '{}
-
 \subsubsection{Assembling Marks}
-
 \begin{figure}[H]
@@
     \label{fig:q}
-    \caption{change $\rho$ to input J}
+    \caption{The $\nu$ functions convert data $E$ to visual $V$.
+$Q$ assembles the different types of visual parameters $V_{i}$ into a graphic in $H$.
+$Q\circ\mu(\xi^{-1}J)$ forms a visual mark by applying $Q$ to a region mapped to connected components $J \subset K$.}
 \end{figure}
@@
 Given a monoid action on a set in $V$, there should be a monoid action on the corresponding $Q(\Gamma(V))$.
-<why gamma?
 

### 2021-01-06 draft/sections/math.tex
q

@@
 \paragraph{Example: Translation}
-According to Stevens, interval data jas general linear group actions \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
+\textbf{fairly certain I lost the thread here}
+According to Stevens, interval data is a set with general linear group actions \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
 Position is a visual variable that can support translation \cite{bertinIIPropertiesGraphic2011,krygierMakingMapsVisual2005,munznerMarksChannels}.
 
-For example, here $\nu$ is a direct map $nu(e) = v$ and the transform function is a multiplicative shift $t(x) = x + x$ applied twice
+For example, here $\nu$ is a direct map $\nu(\tau_{i,k}) = \mu{i,k}$ and the transform function is a multiplicative shift $t(x) = x + x$ applied twice
 
 \begin{align}
-    \nu(t(t(e))) &= t_v(t(\nu(e)))\\
-    \nu(t(e+e)) &= t(t_v(v))\\
-    \nu((e+e) + (e+e)) &= t(v+v)\\
-    \nu(4e) &= (v_v) + (v+v)\\
-    \nu(4e)&=4v
+    \nu(t(t(\tau_{i,k}))) &= t(t(\nu(\tau_{i,k})))\\
+    \nu(t(2\tau_{i,k})) &= t(t_v(\mu_{i,k}))\\
+    \nu(2\tau_{i,k} + 2\tau_{i,k}) &= t(2\mu_{i,k})\\
+    \nu(4\tau_{i,k}) &= 2\mu_{i,k} + 2\mu_{i,k} \\
+    \nu(4\tau_{i,k})&=4\mu_{i,k}\\
+    4\mu_{i,k}&=4\mu_{i,k}\\
 \end{align}
-
-Here $\nu$ is valid because scaling the data by a factor of 4 causes a scaling of the equivalent visualization by 4.
+$\nu$ is valid because scaling the data by a factor of 4 causes a scaling of the equivalent visualization by 4, which since the constant factor can be pulled out means they are equivalent.
 
 \paragraph{Example: Invalid $\nu$}
-shifts in either space need to be mirrored in the other space, this is not valid  here's why:
+Given a transform $t(x) = 2*x$, we construct a $\nu$ that always takes data to .5: 
 \begin{equation}
@@
     \begin{tikzcd}
-        E_1 \arrow[r] \arrow[r, "\lambda:e\mapsto.5"] \arrow[d, "e+2"'] & V_1 \arrow[d, "h+2"] \\
-        E_2 \arrow[r, "\lambda"]                                        & V_2                 
+        E_1 \arrow[r] \arrow[r, "\lambda:e\mapsto.5"] \arrow[d, "2e"'] & V_i \arrow[d, "2v"] \\
+        E_1 \arrow[r, "\lambda"]                                        & V_1                 
     \end{tikzcd}
 \end{equation}
+
+This $\nu$ is invalid because the graph does not commute for $t$:
+\begin{align}
+    \nu(t(e)) & \overset{?}{=} t(\nu(e))\\
+    .5 & \overset{?}{=} t(.5)\\
+    .5 & \neq 2*.5
+\end{align}
+
+To consutruct a valid $\nu$, the diagram must commute for all monoid actions on the sets $E_i, V_i$.
 
@@
 Given a monoid action on a set in $V$, there should be a monoid action on the corresponding $Q(\Gamma(V))$.
+While $\Gamma(V)$ holds for all cases, we can specialize to the bundle $V$ or the sheaf $\mathcal(V)$ depending on the specific visualization.
 
-Lets call $\Gamma(V) = X$ and $Q(\Gamma(V))$
+\paragraph{Proposition}
+If $\forall g \in M$ and $\forall x_1, x_2 \in \Gamma(V)$ then  $Q(x_1) = Q(x_2) \\implies Q(g\circ x_1) = Q(g\circ x_2)$; therefore we can define a group action on $Q(\Gamma(V))=Y$ as $g \circ y = y'$ where $y' = Q(g\circ x)$ with $x \in f^{-1}(y)$
 
@@
     \begin{tikzcd}
-        H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \gamma \text{ s.t.
+        H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K \text{ s.
+t.
+} \exists \gamma \text{ s.t.
 } \gamma(0)=k \text{ and }\gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
@@
 \end{equation}
-where $\gamma$ is a path where input is a point on the interval [0,1] and
-
+where the path\cite{ConnectedSpace2020}  $\gamma$ from $k$ to $j$ is a continous function from the interval [0,1].
 To get back to the display space $H$ from $J$, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then $\rho(\xi^{-1}(J))$ maps it into $R^{7}$ such that $Q\circ\xi^*\mu(\xi^{-1}(J))$ generates a graphical mark.
-
-%%abs: R \rightarrow R^+, 
-%%f: R \rightarrow R, 
-%%abs:[real->real] -> [real->real+]
-%%fmap?, fmap(abs)->ufunc on all functions
-%%lifting: decorator that only looks at arguments
-
-Q goes from one to the other, could need more information than just the return values...
-map on global sections means I might need to know the whole section, map V to H says I only need to know points, restrict to the case where defined on the stalk..but will look at particular cases - sections to sections is most general case that is true, but can be defined V to H or or O(V) to O(H) depending on the visualization.
 

### 2021-01-06 draft/sections/visual.tex
q

@@
 \begin{document}
-\subsubsection{Marks}
-Bertin describes a location on the plane as the signifying characteristic of a point, measurable length as the signifying characteristic of a line, and measurable size as the signifying characteristic of an area and that in display (pixel) space are the point, line, and area marks \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}.
-For each region $s$ in the display space $H$, the mark it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the connected component $J \subset K$.
-\begin{equation}
-\begin{tikzcd}
-    H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K| \exists \Gamma \text{ s.t.
-} \Gamma(0)=k \text{ and }\Gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
-\end{tikzcd}
-\label{eq:mark}
-\end{equation}
-To got back to the display space $H$  from the simplacial complex $J$ of the signifier implanted in the mark, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then  $\rho(\xi^{-1}(J))$ maps it into $R^{7}$.
 
-\subsubsection{Visual Idioms: Equivalance class of artists}
-n $O(E)$ of the same type, they output the same type of prerender $O(H)$:
-
-Natural transformation + composition is partial ordering? Back and forth is equivalent 
 \end{document}

### 2021-01-07 draft/sections/math.tex
going in circles here

@@
 
+%% convention is base space \times fiber bundle
+%% change B to K
 \section{Notation \& Definitions}
+%% intro placeholder of eq in intro
+%% use let or where before you introduce symbols
+%% more notations w/ general words 
+%% tone down - mathcal A, mathcal E, and mathcal H - space, will formalized in this section 
+%% give the why of the math after every mention of why
 In this section we introduce a mathematical description of the visualization pipeline where artist $A$ functions transform data of type $\Gamma(E)$ to an intermediate representation in prerendered display space of type $\Gamma(H)$:
@@
 \item $E$ is a locally trivial fiber bundle over $K$ representing data space.
-\item $K$ is a triangulizable space encoding the connectivity of the points in $E$
 \item $H$ is a fiber bundle over $S$ representing visual space
-\item $S$ is a simplacial complex encoding the visualization
-\item $\Gamma$ is the global space of sections
+\item $K$ and $S$ is a triangulizable topological space or CW complex encoding the connectivity of the points in $E$
 \end{itemize}
+All the fiber bundles mentioned in this work is assumed to be locally trivial \cite{rowlandFiberBundle,FiberBundle2020}.
 
-When $E$ is a trivial fiber bundle $E = F \times K$, it can be assumed that all fibers $F_{k}$ over $k \in K$ are equal.
+\subsection{Data Fiber Bundles}
+\begin{figure}[H]
+    \label{fig:data_fiber_bundle}
+    \includegraphics[width=.2\linewidth]{figures/sections/math/fiberbundle.png}
+    \caption{write up some words here}
+\end{figure}
+
+As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle (E, K, $\pi$, F) with $\pi: E \rightarrow K$ where $K$ which can be thought of as a set of keys $k$.
+A section $\tau:K \rightarrow E$ is an instance of the data that lies in $E$ and is discussed in section~\ref{sec:data_fiber}.
+
+We provide a brief description of fiber bundles because we model data, visual transformations, and a prerendered visual graphic as fiber bundles.
 Fiber bundles are product spaces of toplological spaces, which are a set of points with a set of neighborhoods for each point\cite{FiberBundle2020, rowlandFiberBundle}.
-
-\subsection{Fiber Bundles}
-We provide a brief description of fiber bundles because we model data, visual transformations, and a prerendered visual graphic as fiber bundles.
 A fiber bundle is a structure $(E, B, \pi, F)$  consisting of topological total space $E$, base space $K$, fiber space $F$ and the map from total space to base space:
@@
     \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
+    %%include second figure where fiber goes otherway
 \end{figure}
+
+%%% add in motivation - encodes the connectiveity of the points, choice of split is intrinsinc to how we want to view the data
 
@@
 
-As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood around $F$.
-This subdivision is projected down to the toplology $\mathcal{T}$
+As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood \cite{FiberBundle2020, rowlandFiberBundle}  around $F$.
+%%so why topology? 
+\paragraph{Example}
+%%% pull up data structures like discrete, continous, etc w/ base space 
+%%% little drawing dots, line, plane, cube
 
-\begin{equation}
-    \label{eq:base_space}
-\mathcal{T}_b = \{U\subseteqq B: \{p \in E: [p] \in U\}\in \mathcal{T}_E\}
-\end{equation}
+\begin{figure}[ht]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
+    %% add box around neighboring P and Map
+    \label{fig:k_data}
+\end{figure}
 
-where $[p] \in U$ is the point $b \in B$ with an open set surrounding it that has an open preimage in $E$ under the surjective map $\pi: p \rightarrow [p]$.
+\begin{figure}[H]
+    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
+\end{figure}
+
+in figure~\ref{fig:k_data}, temperature is the only one data field in $r$ but the $K$ base spaces are different.
+subfig[1] is a timeseries, so the temperature in $r$ at time $t$ is dependent on the temperature in $r_{t-1}$ and the temperature in $r_{t+1}$ is dependent on  $r_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
+In the case of the map, every temperature in $r$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
+$K$ does not know the time or latitude or longitude of the point as those are metadata variables describing the $k$ rather than the value of $k$.
+The mapping $\tau: K \rightarrow E$ provides the binding between the key $k \in K$ and the value $r$ in $E$ \cite{munznerChDataAbstraction}.
 
 \subsubsection{Fiber}
-As shown in equation!\ref{eg:base_space}, every point $b \in B$ has a local open set neighborhood $U$ \cite{FiberBundle2020, rowlandFiberBundle}
+%% pull fiber stuff up here? what is it? why? fiber as cartesian product of sets
+%% actual values are in the fiber 
+%% gonna view spaces as simple piece put together 
+%% Our fibers are cross products of (column name, domain) as described by Spivak's \cite{spivakSIMPLICIALDATABASES}.
 
@@
 \end{equation}
-such that $\varphi: \pi^{-1}(U) \rightarrow U \times F$ is a homeomorphism where $\pi$ and $\mathrm{proj}_U$ both map to $U$ and the fiber over $k$ $F_b = \pi^{-1}({b \in B}) $ is homomorphic to the fiber $F$.
+such that $\varphi: \pi^{-1}(U) \rightarrow U \times F$ is a homeomorphism where $\pi$ and $\mathrm{proj}_U$ both map to $U$ and the fiber over $k$
+
+%% rewrite this section - what are the goals here and why should we care? 
+%% devoid of any context
+
+$F_b = \pi^{-1}({b \in B}) $ is homomorphic to the fiber $F$.
+%% add a sentence for why we care
+%% fibers are cartesian products of these typles sets (name, domain, monoidal action)
+
+%%the break out more about how fibers are cross products and what that means and how that's the structure - pull the monoids definitions here
+
+\paragraph{Example}
+%% stevens scales are examples of this 
+%% one ecample of F1 is (names, strings, concat), (ordinal, Z,  ordering), ()
 
 \subsubsection{Section}
+%% add context
 The section $f$ is the mapping from base space to total space $f: B\rightarrow E$ 
@@
 All sections $f \in \Gamma(E)$ have the same fibers $F$ and connectivity $B$.
-
-\subsubsection{Sheaf and Stalk}
-As described in equation~\ref{eq:local_trivial}, there is a local space $U \subset B$ around every $b$.
-The inclusion map $\iota: U \rightarrow B$ can be pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $U$.
-\begin{equation}
-    \label{eq:sheaf}
-    \begin{tikzcd}
-        \iota^*E \arrow[d, "\pi"']           & E \arrow[d, "\pi"'] \arrow[l, "\iota^*"']         \\
-        U \arrow[u, "\iota^*f"', bend right] & B \arrow[u, "f"', bend right] \arrow[l, "\iota"']
-        \end{tikzcd}
-\end{equation}
-
-The localized section of fibers $\iota^*f: U \rightarrow \iota^*E$ is the sheaf $\mathcal{O}(E)$ with germ of $\xi^*f$.
-The neighborhood of points the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}
-
-\begin{equation}
-    \label{eq:stalk}
-    \iota^{-1}\mathcal{F}(\{b\}) = \varinjlim_{b\subseteq U}\mathcal{F}(U) =  \varinjlim_{b \in U} = \mathcal{F}_b 
-\end{equation}
-
-which through $\iota$ gets the data in $E$ at and near to $b$.
-Restricting the artist to the sheaf means the artist knows the data in $F$ and also has access to derivatives of the data.
-This property is useful for some visual transformations.
-%% gives all direvatives but we need the jet which is Add stuff about jets
-
-\subsection{Data Model}
-As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle (E, K, $\pi$, F) with $\pi: E \rightarrow K$ where $K$ which can be thought of as a set of keys $k$.
-A section $\tau:K \rightarrow E$ is an instance of the data that lies in $E$ and is discussed in section~\ref{sec:data_fiber}.
-
-\subsubsection{Example}
-
-\begin{figure}[H]
-    \label{fig:data_fiber_bundle}
-    \includegraphics[width=.2\linewidth]{figures/sections/math/fiberbundle.png}
-    \caption{write up some words here}
-\end{figure}
-
-As illustrated by figure~\ref{fig:fiberbundle}, the vertical lines $F$ are the range of possible temperature values embedded in the total space $E$.
-The base space $K$ of the fiber bundle is a line because the data points $r$ in $E$ are on a space that is  continous in one dimension.
-
-\subsubsection{Base Space}
-The base space $K$ is a representation of the connectivity of the data, specifically whether the points in $E$ are discrete or sampled from a continous space.
-The same dataset can be expressed with different $K$.
-
-In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
-
-\begin{figure}[H]
-    \label{fig:simplex}
-    \includegraphics{figures/sections/math/simplex.png}
-    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
-Higher order simplicies are outside the scope of this paper.}
-\end{figure}
-
-%% k is a triangularizable topological space 
-One way to represent the topological space K is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
-Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another: %%formula for simplacial complexes/simplacial sets
-
-% replace this table w/ a diagram that shows this 
-\begin{description}
-    \item[0-simplex] discrete observations (inventory records)
-    \item[1-simplex] 1D continuos data (timeseries)
-    \item[2-simplex] 2D continuos data (map)
-    \item[3-simplex] 3D continuos data (video)
-\end{description}
-
-\subsubsection{Example}
-\begin{figure}[ht]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
-    %% add box around neighboring P and Map
-    \label{fig:k_data}
-\end{figure}
-
-\begin{figure}[H]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
-\end{figure}
-
-in figure~\ref{fig:k_data}, temperature is the only one data field in $r$ but the $K$ base spaces are different.
-subfig[1] is a timeseries, so the temperature in $r$ at time $t$ is dependent on the temperature in $r_{t-1}$ and the temperature in $r_{t+1}$ is dependent on  $r_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
-In the case of the map, every temperature in $r$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
-$K$ does not know the time or latitude or longitude of the point as those are metadata variables describing the $k$ rather than the value of $k$.
-The mapping $\tau: K \rightarrow E$ provides the binding between the key $k \in K$ and the value $r$ in $E$ \cite{munznerChDataAbstraction}.
-
-\subsubsection{Fiber and Sections}
-\label{sec:data_fiber}
-We use Spivak's formalization of data base schemas as the basis of our fiber space $F$ \cite{spivakSIMPLICIALDATABASES}. He defines the type specification 
-\begin{equation}
-\pi: U \rightarrow DT
-\end{equation}
-
-where $DT$ is the set of data types (as identified by their names) and $U$ is the disjoint set of all possible objects $x$ of all types in $DT$.
-This means that for each type $T\in DT$, the preimage $\pi^{-1}(T)\subset U $ is the domain of $T$, and $x \in \pi^{1}(T)\subset U$ is an object of type $T$.
-Spivak then defines a schema $(C, \sigma)$ of type $\pi$, where $\pi$ is the universe of all types, such that
-\begin{equation}
-\sigma: C \rightarrow DT
-\end{equation}
-where $C$ is the finite set of names of columuns, which we generalize to data fields in $E$. The set of all values restricted to the datatypes in $DT$ is $U_{\sigma}$
-
-\begin{equation}
-    \begin{tikzcd}
-        U_{\sigma} \arrow[d, "\pi_{\sigma}"'] \arrow[r] & U \arrow[d, "\pi"] \\
-        C \arrow[r, "\sigma"]                           & DT                
-    \end{tikzcd}
-\end{equation}
-The pullback $U_{\sigma} \coloneqq \sigma^{-1}(U)$ restricts $U$ to the datatypes of the fields in $C$ such that $U_{\sigma}$ is the fiber product $U \times_{DT} C$, and the pullback $\pi_{\sigma}:U_{\sigma} \rightarrow C$ specifies the domain bundle $U_{\sigma}$ over $C$ induced by $\sigma$.
-The fiber $F$ is the cartesian product of all sets in the disjoint union $U_{\sigma}$.
+\paragraph{Example}
+%%talk about table and all that jazz
 
@@
 
+\subsubsection{Sheaf and Stalk}
+%% add motivation - zoom, subsection, streaming, also for realizing visualization, restricting to that piece - need to know near point w/o 
+As described in equation~\ref{eq:local_trivial}, there is a local space $U \subset B$ around every $b$.
+The inclusion map $\iota: U \rightarrow B$ can be pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $U$.
+\begin{equation}
+    \label{eq:sheaf}
+    \begin{tikzcd}
+        \iota^*E \arrow[d, "\pi"']           & E \arrow[d, "\pi"'] \arrow[l, "\iota^*"']         \\
+        U \arrow[u, "\iota^*f"', bend right] & B \arrow[u, "f"', bend right] \arrow[l, "\iota"']
+        \end{tikzcd}
+\end{equation}
+
+The localized section of fibers $\iota^*f: U \rightarrow \iota^*E$ is the sheaf $\mathcal{O}(E)$ with germ of $\xi^*f$ which is the localized section.
+The neighborhood of points the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic} which is a ...$\dots$
+. Restricting the artist to the sheaf means the artist knows the data in $F$ and also has access to derivatives of the data. This property is useful for some visual transformations. 
+%% gives all direvatives but we need the jet which is Add stuff about jets
+
+\subsubsection{Triangulization}
+%%drop down to after the fiber section
+The base space $K$ is a representation of the connectivity of the data, specifically whether the points in $E$ are discrete or sampled from a continous space.
+The same dataset can be expressed with different $K$.
+
+In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
+
+\begin{figure}[H]
+    \label{fig:simplex}
+    \includegraphics{figures/sections/math/simplex.png}
+    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
+Higher order simplicies are outside the scope of this paper.}
+\end{figure}
+
+%% k is a triangularizable topological space 
+One way to represent the topological space K is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
+Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another: %%formula for simplacial complexes/simplacial sets
+
+\paragraph{Example}
+chopping up a torus maybe? talk about how that gets unpacked into triangles and then into vertices
+
 \subsection{Prerender Space}
 \label{sec:display}
+
 We model the prerender space on which lives on ideal version of the visulaization as a fiber bundle (H, S, $\pi$, D).
@@
 \subsubsection{Base space}
+%%pull up to the very top that dim(k) < screen space, need to thicken therefore S
 $K$ can be considered a subspace of the screen base space $S$ such that $\xi: S \rightarrow K$ is a deformation retraction \cite{RetractionTopology2020}
@@
 A section $\rho: S \rightarrow H$ is a mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to visual space as defined in D.
-For a physical screen display, the predisplay space is a trivial fiber bundle $H=\mathbb{R}^{7}\times S$ such that $\rho$ is
+
+\paragraph{Example}
+For a physical screen display, we can consider a predisplay space that is a trivial fiber bundle $H=\mathbb{R}^{5}\times S$ such that $\rho$ is
 \begin{equation}
-    \rho(s)  = \{x, y, z, r, g, b, a\}
+    \rho(s)  = [x(s), y(s), r(s), g(s), b(s)]
     \label{eq:rho}
@@
 
-Here we assume a single opaque 2D image such that the $z$ and $alpha$ fields can be omitted.
+Here we assume a single opaque 2D image such that the $z$ and $alpha$ fields can be omitted. To support overplotting and transparency, we consider $D=R^{7}$
+%%H can be other spaces
 
@@
 \subsection{Artist}
+%%introduce O(E) up in previous section, then refer back to that section
 
@@
     \begin{tikzcd}
-    E \arrow[r, "\nu"] \arrow[rd, "\pi"'] & V \arrow[d, "\pi"] & \xi^*V \arrow[r, "Q"] \arrow[d, "\xi^*\pi"'] & H \arrow[ld, "\pi"] \\
+        E \arrow[r, "\nu"] \arrow[rd, "\pi"'] & V \arrow[d, "\pi"] & \xi^*V \arrow[r, "Q"] \arrow[d, "\xi^*\pi"'] \arrow[l, "\xi^*"'] & H \arrow[ld, "\pi"] \\
                                               & K                  & S \arrow[l, "\xi"']                                              &                    
@@
 \end{equation}
+%%map backwords via pullback can copy values over k over and over to corresponding S_k
 
@@
 \end{table}
-
+%%needs context, pulling monoids up gives it
 \subsubsection{Visual Channels}
@@
 Position is a visual variable that can support translation \cite{bertinIIPropertiesGraphic2011,krygierMakingMapsVisual2005,munznerMarksChannels}.
-
-For example, here $\nu$ is a direct map $\nu(\tau_{i,k}) = \mu{i,k}$ and the transform function is a multiplicative shift $t(x) = x + x$ applied twice
-
-\begin{align}
-    \nu(t(t(\tau_{i,k}))) &= t(t(\nu(\tau_{i,k})))\\
-    \nu(t(2\tau_{i,k})) &= t(t_v(\mu_{i,k}))\\
-    \nu(2\tau_{i,k} + 2\tau_{i,k}) &= t(2\mu_{i,k})\\
-    \nu(4\tau_{i,k}) &= 2\mu_{i,k} + 2\mu_{i,k} \\
-    \nu(4\tau_{i,k})&=4\mu_{i,k}\\
-    4\mu_{i,k}&=4\mu_{i,k}\\
-\end{align}
-$\nu$ is valid because scaling the data by a factor of 4 causes a scaling of the equivalent visualization by 4, which since the constant factor can be pulled out means they are equivalent.
+\begin{equation}
+\nu(x + c) = \nu(x) + \nu(c)
+\end{equation}
 
 \paragraph{Example: Invalid $\nu$}
-Given a transform $t(x) = 2*x$, we construct a $\nu$ that always takes data to .5: 
+Given a transform $t(x) = x+2$, we construct a $\nu$ that always takes data to .5: 
 \begin{equation}
@@
 For each region $s$ in the display space $H$, the mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology} it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the path connected component $J \subset K$.
+
+%% defined J_{k} first, then pull down to here, then can do \rho S_j
 \begin{equation}

### 2021-01-07 draft/sections/implementation.tex
going in circles here
start of implementation

@@
 \begin{document}
+\section{Matplottoy}
+\begin{figure}
+\includegraphics[width=\linewidth]{figures/sections/math/path_of_q.png}
+\end{figure}
+%%% objects: Data, V, H
+%%% functions: \nu, q
 
-\subsection{Data Model}
+We build on top of the existing Matplotlib architecture \cite{hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and rely on Matplotlib for the other graphical elements of the visualization and the rendering.
+For a primitive graph such as a bar, scatter, line, our path is:
+\begin{minted}{python}
+
+E = Section()
+V = {'parameter': ('field': transformer)}
+
+fig, ax = plt.subplots()
+artist = primative_Q(E, V)
+ax.add_artist(artist)
+\end{minted}
+
+\subsection{Data: $\mathcal{E}$}
+We propose a semantic markup of the fiber and connectivity that we will later use to check the constraints of the $\nu$ and $Q$. 
+\begin{minted}{python}
+class FiberBundle:
+    def __init__(self, K, F):
+            """
+            K: {'tables':[]}
+            F: {'field name': (type, monoid action)}
+            """
+
+    def is_section(self, section):
+        """checks if a section is from a given fiber bundle:
+        are values in F, are keys in K"""
+\end{minted}
+We add a monoid field to the schema like structure of the fiber since the monoid actions are the constraints on the $\nu$ and $Q$.
+We implement $K$ using the triangulization scheme described in section~\ref{sec:triangulization}.
+This means we expect data to be provided as tables where the name encodes the connectivity:
+
+\begin{description}
+    \item[vertex] - disconnected points
+    \item[edge] - 1D continuity along the edge
+    \item[face] - 2D continuity along the face
+\end{description}
+Nested continuity is encoded in the fiber. For example, one way to encode a movie is a 1D timeseries of \mathrm{Frame} objects, where each \mathrm{Frame} is a $n\x m$ continuous array.
+
+\subsubsection{Sections}
+As described in section~\ref{sec:section_data}, the data values are encode as the section of the fiber bundle. We implement this as as a wrapper class around a data container object of the form:
+\begin{minted}
+class Section:
+    FB = FiberBundle(K, F)
+
+    def __init__(self, *args):
+        pass
+    def view(self, simplex="vertex"):
+        """"converts data into atomic column order for get method
+        # maybe also pass in the columns?
+        """
+        return {fieldname: data}
+\end{minted}
+
+The view function is the closest analog to $\tau$, as it returns a a data container type object.
+We specify field based top level indexing so that we can pair the fields with transformer $\nu$ objects.
+
+\subsection{Encoding: $\mathcal{V}$}
+We implement $\nu$ as encoder objects with an optional `__init__` method that can be used to specify the range of target visual variables.
+
+\begin{minted}{python}
+class Encoder:
+    def __init__(self, args):
+        pass
+    def validate(self, monoid):
+        pass
+    def convert(self, value):
+        return converted_value
+\end{minted}
+
+The \mathrm{validate} method specifies the monoids for which this a valid $\nu$ and checks against the specification in the fiber.
+The \mathrm{convert} method converts a value from data space to an internalized normal form as described in table~\ref{tab:mpl_visual_variable_fiber}.
+
+\subsection{Graphic: $\mathcal{H}$}
+
+\subsection{Case Study: Iris(Penguins)}
 

### 2021-01-08 draft/sections/implementation.tex
refactored vector simplex + point w/ new fiber structure

@@
 fig, ax = plt.subplots()
-artist = primative_Q(E, V)
+artist = Q(E, V)
 ax.add_artist(artist)
@@
         are values in F, are keys in K"""
+        pass
 \end{minted}
@@
 As described in section~\ref{sec:section_data}, the data values are encode as the section of the fiber bundle. We implement this as as a wrapper class around a data container object of the form:
-\begin{minted}
+\begin{minted}{python}
 class Section:
@@
     def view(self, simplex="vertex"):
-        """"converts data into atomic column order for get method
-        # maybe also pass in the columns?
-        """
+        #convert data into atomic column order 
         return {fieldname: data}
@@
     def validate(self, monoid):
+        #check if transform supports monoid action
         pass
     def convert(self, value):
-        return converted_value
+        # convert value to libraries internal normalized form
+        pass
 \end{minted}
@@
 \subsection{Graphic: $\mathcal{H}$}
+We inherent from the current Matplotlib artists for our $Q$, which is responsible for constructing a curried fully specified internal representation of the graphic.
+\begin{minted}{python}
+class Q(BaseClass):
+    required = {}
+    optional = {}
+    def __init__(self, data, *args, kwargs):
+        # check that required and optional visual parameters are passed in
+        # check that the transforms can be applied to the data 
+        pass
+    def draw(self, render, *args, **kwargs):
+        # get frozen atomic section of the data 
+        # apply  transforms data
+        # set internal attributes of the graphic q generates
+        #hands off to backends to draw on screen
+        super().draw(renderer, *args, **kwargs)
+\end{minted}
+While equation~\ref{eq:artist} specifies that the values of $\mu$ can be generated before they are input to $Q$, we curry the opertaions of obtaining the data and performing the transforms to the draw method because that allows us to more easily propogate updates to the data to the screen.
 

### 2021-01-08 code/matplottoy/artists/line.py

@@ -9,6 +9,7 @@ import matplotlib.path as mpath
 
 from matplottoy.artists import utils
 
+## should probably remove transforms and spell out *args, **kwargs?
 class Line(mcollections.LineCollection):
     required = {'x', 'y'}
     optional = {'facecolor'} 

### 2021-01-08 code/matplottoy/artists/point.py

@@ -7,6 +7,8 @@ from matplotlib import rcParams
 import matplotlib.collections as mcollections
 import matplotlib.path as mpath
 
+from matplottoy.artists import utils
+
 class Point(mcollections.Collection):
     required = {'x', 'y'}
     optional = {'facecolors'} 
@@ -23,16 +25,10 @@ class Point(mcollections.Collection):
         # has a way to provide vertex data  
         assert 'vertex' in data.FB.K['tables']
         # check that you've given the required parameters
-        assert Point.required <= transforms.keys()
-        # this one might not be necessary 
-        # (checks that rest of keys are valid)
-        assert ((transforms.keys()-Point.required) 
-                            <= Point.optional) 
+        utils.check_constraints(Point, transforms)
+        utils.validate_transforms(data.FB.F, transforms)
+
         
-        # checking that transform can take that type
-        # group symmetry steo
-        assert all(tau.validate(data.FB.F[column]) 
-                    for (column, tau) in transforms.values())
         self.data = data
         self.transforms = transforms
 

### 2021-01-08 code/matplottoy/artists/utils.py

@@ -16,10 +16,10 @@ def check_constraints(artist, transforms):
             <= artist.optional):
         raise ValueError(f"Valid optional encodings: {artist.optional}")
 
-def validate_transforms(data, transforms):
+def validate_transforms(fibers, transforms):
     for column, tau in transforms.values():
         for col in np.atleast_1d(column):
-            if not tau.validate(data.FB.F[col]):
+            if not tau.validate(fibers[col]['monoid']):
                 raise ValueError(f"Invalid transform {tau} for {col}")
 
 def convert_transforms(view, transforms, exclude=None):

### 2021-01-08 code/matplottoy/data/simplex.py

@@ -5,6 +5,7 @@ import numpy as np
 
 from matplottoy.encoders import mtypes
  
+ 
 class FiberBundle:
     def __init__(self, K, F):
         """
@@ -14,14 +15,14 @@ class FiberBundle:
             attributes of the base space of the fiber bundle, 
             describes domain of base space + the type of simplexs
             on the base space
-            {'simplexes':{}, domain: type }
-            k is a tuple (which simplex)
-                         (which simplex, which line)
-                         (which simplex, face x, face y)
+            {'tables': []}
+
         F: dict
             the fiber, which is the space of all possible observations 
             type is used here to describe the space 
             {variable name: type}
+
+            # rework as ('field name': (type, monoid action))
         """
         self.K = K 
         self.F = F
@@ -39,9 +40,12 @@ class VertexSimplex: #maybe change name to something else
     some k's in K
     """
     FB = FiberBundle({'tables': ['vertex']},  
-                     {'v1': mtypes.Ordinal(range(1,6)), 
-                      'v2': mtypes.IntervalClosed([0,10]),
-                      'v3': mtypes.Nominal(['true', 'false'])})
+                     {'v1': {'type': int, 'monoid': 'ordinal', 
+                             'range': [1,2,3,4,5]},
+                      'v2': {'type': float,'monoid':'interval',
+                             'range': [0,1]},
+                      'v3': {'type': str, 'monoid':'nominal', 
+                             'range':['true', 'false']}})
 
     def __init__(self, sid = 45, size=1000, max_key=10**10):
         self.section_id = sid
@@ -51,15 +55,15 @@ class VertexSimplex: #maybe change name to something else
         rng = np.random.default_rng(sid)
         self.keys = rng.integers(0,max_key, size)
 
-    def sigma(self, k):
+    def tau(self, k):
         """Returns the (k, obs) at k"""
         # set of functions for choosing values from F at a given k
         # is usually the data structure
         rng = np.random.default_rng(k*self.section_id)
-        return (k, (rng.choice(self.FB.F['v1'].categories), 
-                    rng.uniform(self.FB.F['v2'].min, 
-                                self.FB.F['v2'].max),
-                    rng.choice(self.FB.F['v3'].categories)))
+        e1 = rng.choice(self.FB.F['v1']['range'])
+        e2 = rng.uniform(*self.FB.F['v2']['range'])
+        e3 = rng.choice(self.FB.F['v3']['range'])
[…]

### 2021-01-08 code/matplottoy/encoders/color.py

@@ -6,19 +6,19 @@ from matplottoy.encoders import mtypes
 
 class Identity:
     @staticmethod
+    def validate(mtype):
+        return mtype in ['nominal']
+        
+    @staticmethod
     def convert(value):
         values = np.atleast_1d(np.array(value, dtype=object))
         return [mcolors.to_rgba(v) for v in values]
 
-    @staticmethod
-    def validate(mtype):
-        return (isinstance(mtype, mtypes.Color) or 
-               all(mcolors.is_color_like(v) for v in mtype.categories))
-        
-class Nominal:
+class Categorical:
     def __init__(self, mapping):
         """goal of init is to store parameters that would otherwise be
         curried higher level function"""
+        assert(mcolors.is_color_like(color) for color in mapping.values())
         self._mapping = mapping
 
     def convert(self, value):
@@ -26,6 +26,5 @@ class Nominal:
         return [mcolors.to_rgba(self._mapping[v]) for v in values]
 
     def validate(self, mtype):
-        return (set(mtype.categories) == self._mapping.keys() and
-        all(mcolors.is_color_like(color) for color in self._mapping.values()))
+        return mtype in ['nominal']
 

### 2021-01-08 code/matplottoy/encoders/position.py

@@ -10,7 +10,7 @@ class Identity:
         return np.array(value)
     @staticmethod
     def validate(mtype):
-        return mtype.mtype in ['nominal', 'ordinal', 'interval']
+        return mtype in ['nominal', 'ordinal', 'interval', 'ratio']
 
 class Log:
     def __init__(self, base=10):

### 2021-01-08 code/matplottoy/artists/line.py
switched all the simplex code to the new fiber model

@@ -12,7 +12,7 @@ from matplottoy.artists import utils
 ## should probably remove transforms and spell out *args, **kwargs?
 class Line(mcollections.LineCollection):
     required = {'x', 'y'}
-    optional = {'facecolor'} 
+    optional = {'color'} 
     def __init__(self, data, transforms, *args, **kwargs):
         """
         Parameters
@@ -23,7 +23,7 @@ class Line(mcollections.LineCollection):
         super().__init__(None, *args, **kwargs)
         assert 'edge' in data.FB.K['tables']
         utils.check_constraints(Line, transforms)
-        utils.validate_transforms(data, transforms)
+        utils.validate_transforms(data.FB.F, transforms)
      
         self.data = data
         self.transforms = transforms
@@ -33,14 +33,14 @@ class Line(mcollections.LineCollection):
         view = self.data.view('edge')
         visual = utils.convert_transforms(view, self.transforms)
         # dictionary here in place of visual(parameter) function
-        if 'facecolor' not in visual:
-            visual['facecolor'] = "C0"
+        if 'color' not in visual:
+            visual['color'] = "C0"
 
         segments = [np.vstack((vx, vy)).T for vx, vy 
                     in zip(visual['x'], visual['y'])]
 
         self.set_segments(segments)
-        self.set_color(visual['facecolor'])
+        self.set_color(visual['color'])
 
         super().draw(renderer, *args, **kwargs)
 

### 2021-01-08 code/matplottoy/data/simplex.py

@@ -90,10 +90,11 @@ class EdgeSimplex: # ToDo: generalize to take list of functions as input
     """
 
     FB = FiberBundle({'tables': ['vertex','edge']},
-                     {'x' : mtypes.IntervalClosed([-1,1]),
-                      'y':  mtypes.IntervalClosed([-1,1]),
-                      'color': mtypes.Nominal(['red', 'green', 'orange', 'blue'])
-                      })
+                     {'x' : {'type:':float, 'monoid':'interval', 'range':[-1,1]},
+                      'y':  {'type:':float, 'monoid':'interval', 'range':[-1,1]},
+                      'color': {'type':mtypes.Color(), 'monoid':'nominal', 
+                                'range': ['red', 'green', 'orange', 'blue']}
+                    })
 
     def __init__(self, num_verts=4, num_samples=1000):
         """which simplex am I on and distance""" 
@@ -123,7 +124,7 @@ class EdgeSimplex: # ToDo: generalize to take list of functions as input
         return np.cos(angles), np.sin(angles)
 
         
-    def sigma(self, k, simplex='edge'):
+    def tau(self, k, simplex='edge'):
         """arbitrary k, even between vertices, will give back
         an observation, so all functions must target back to vertice 
         table
@@ -141,7 +142,7 @@ class EdgeSimplex: # ToDo: generalize to take list of functions as input
         table = defaultdict(list)
         for k in self.keys:
             table['index'].append(k)
-            for (name, value) in zip(self.FB.F.keys(), self.sigma(k, simplex)[1]):
+            for (name, value) in zip(self.FB.F.keys(), self.tau(k, simplex)[1]):
                 table[name].append(value)
         
         if simplex =='vertex':
@@ -163,26 +164,34 @@ def is_continuous(vertices):
 
 class ContinuousLine:
     FB = FiberBundle({'tables': ['vertex','edge']},
-                     {'x' : mtypes.IntervalClosed([ 0,4]),
-                      'y':  mtypes.IntervalClosed([-1,1]),
-                      'color': mtypes.Nominal(['red', 'green', 'orange', 'blue'])
-                      })
+                     {'x' : {'type:':float, 'monoid':'interval', 'range':[0,4]},
+                      'y': {'type:':float, 'monoid':'interval', 'range':[-1,1]},
+                      'color': {'type':str, 'monoid':'nominal', 
+                                'range': ['red', 'green', 'orange', 'blue']}
+                    })
+                    
     def __init__(self, edge_table, vertex_table, num_samples=1000):
[…]

### 2021-01-08 code/matplottoy/tests/utils.py

@@ -1,0 +1,3 @@
+import numpy as np
+def mnx(d):
+    return np.min(d)-np.std(d) , np.max(d) + np.std(d)

### 2021-01-11 draft/sections/implementation.tex
need to resort fiber section

@@
             K: {'tables':[]}
-            F: {'field name': (type, monoid action)}
+            F: {'field name': {'type': data type, 'monoid': measurement type, 
+                                'range': restrictions on data type }}
             """
@@
 We add a monoid field to the schema like structure of the fiber since the monoid actions are the constraints on the $\nu$ and $Q$.
-We implement $K$ using the triangulization scheme described in section~\ref{sec:triangulization}.
+These will usually be identified by a measurment scale such as 'nominal', 'ordinal', 'interval', and 'ratio'.
+The optional?
+range field can be used to specify restrictions on the type such as set of string categorical values or fixed interval of float values.
+We implement $K$ using the triangleization scheme described in section~\ref{sec:triangulization}.
 This means we expect data to be provided as tables where the name encodes the connectivity:

### 2021-01-11 draft/sections/math.tex
refactored vector simplex + point w/ new fiber structure
need to resort fiber section

@@
 \section{Notation \& Definitions}
-%% intro placeholder of eq in intro
-%% use let or where before you introduce symbols
-%% more notations w/ general words 
-%% tone down - mathcal A, mathcal E, and mathcal H - space, will formalized in this section 
+
 %% give the why of the math after every mention of why
-In this section we introduce a mathematical description of the visualization pipeline where artist $A$ functions transform data of type $\Gamma(E)$ to an intermediate representation in prerendered display space of type $\Gamma(H)$:
+
+In this section we introduce a mathematical description of the visualization pipeline where artist $\mathcal{A}$ functions transform data  $\mathcal{E}$ to an intermediate representation in a prerendered display space $\mathcal{H}$.
 
@@
     \label{eq:artist}
-    A: \Gamma(E) \rightarrow \Gamma(H)
+    \mathcal{A}: \mathcal{E} \rightarrow \mathcal{H}
 \end{equation}
 
+We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they are products of a topological total space that encodes points and a topological base space that can encode the connectivity of those points:
 \begin{itemize}
-\item $A$ is the function that converts an instance of data $\Gamma(E)$ to an instance of a visual representation $\Gamma(H)$ 
 \item $E$ is a locally trivial fiber bundle over $K$ representing data space.
 \item $H$ is a fiber bundle over $S$ representing visual space
-\item $K$ and $S$ is a triangulizable topological space or CW complex encoding the connectivity of the points in $E$
+\item $K$ and $S$ are a triangulizable topological space or a CW complex encoding the connectivity of points in $E$ and $H$ respectively
 \end{itemize}
-All the fiber bundles mentioned in this work is assumed to be locally trivial \cite{rowlandFiberBundle,FiberBundle2020}.
+
+The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}.
 
 \subsection{Data Fiber Bundles}
-\begin{figure}[H]
-    \label{fig:data_fiber_bundle}
-    \includegraphics[width=.2\linewidth]{figures/sections/math/fiberbundle.png}
-    \caption{write up some words here}
-\end{figure}
-
-As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle (E, K, $\pi$, F) with $\pi: E \rightarrow K$ where $K$ which can be thought of as a set of keys $k$.
-A section $\tau:K \rightarrow E$ is an instance of the data that lies in $E$ and is discussed in section~\ref{sec:data_fiber}.
-
-We provide a brief description of fiber bundles because we model data, visual transformations, and a prerendered visual graphic as fiber bundles.
-Fiber bundles are product spaces of toplological spaces, which are a set of points with a set of neighborhoods for each point\cite{FiberBundle2020, rowlandFiberBundle}.
-A fiber bundle is a structure $(E, B, \pi, F)$  consisting of topological total space $E$, base space $K$, fiber space $F$ and the map from total space to base space:
+As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle ($E$, $K$, $\pi$, $F$)
 
@@
 
-where there is a bijection from $F$ to every fiber $F_b$ over point $b \in B$ in $E$ and the function $\pi: E \rightarrow B$ is the map into the $B$ quotient space of $E$.
-By defintion of a fiber bundle, $\pi$ is always a mapping from total space to base space, independent of the points $p \in E$, and therefore we call this mapping $\pi$ for all the fiber bundles in the model.
+with topological total space $E$, base space $K$, fiber space $F$, and the map from total space to base space $\pi: E \rightarrow K$.
+Maps from $K$ to $E$ are called sections and select specific points in $K$.
+The global space of sections in $E$ is $\Gamma(E)$.
 
-\subsubsection{Base space}
+\paragraph{example}
+
 \begin{figure}[H]
-    \label{fig:base_space}
-    \includegraphics[width=0.4\linewidth]{figures/sections/math/k_qspace.png}
+    \label{fig:data_fiber_bundle}
+    \includegraphics[width=1\linewidth]{figures/sections/math/fiberbundle.png}
+    \caption{$F$ is an $n\times m$ space of points.
+The section $\tau_1$ returns the blue points,  while $\tau_2$ returns the purple points.
+$\Gamma(E)$ is the set of all sections, including $\tau_1$ and $\tau_2$}.
+\end{figure}
+
+The fibers $F$ in figure~\ref{fig:data_fiber_bundle} lie in some total space $E$ and are points that lie in a 2D space.
+For example, this space can be $\mathrm{temperature}\times \mathrm{pressure}$ or $\mathrm{time} \times \mathrm{precipitation}$.
+$K$ is a line on the interval [0,1] and indicates that the points lie on a 1D continous space, but we derive whether the data is a distribution or a timeseries or a function from the fields in $F$.
+The sections $\tau_1$ and $\tau_2$ are distinct maps from $k$ to the blue and purple points in $E$ respectively.
+
+\subsubsection{Base Space}
+\label{sec:base_data}
+\begin{figure}[H]
+    %%\includegraphics[\textwidth]{}
+    \label{fig:base_space_types}
+    %%% pull up data structures like discrete, continous, etc w/ base space 
+    %%% little drawing dots, line, plane, cube
+    \caption{The topological base space $K$ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
+\end{figure}
+
+Data can be discrete observations, timeseries, maps, fields \cite{munznerWhatDataAbstraction2014} and $K$ is a set of points $k$ that can act as keys from a representative space, such as seen in figure~\ref{fig:base_space_types}, to the data values in $E$ \cite{munznerWhatDataAbstraction2014}.
+
+\begin{figure}[H]
+    \label{fig:base_space_div}
+    \includegraphics[width=.5\linewidth]{figures/sections/math/k_qspace.png}
     %%include second figure where fiber goes otherway
@@
 
-%%% add in motivation - encodes the connectiveity of the points, choice of split is intrinsinc to how we want to view the data
+$K$ and $F$ are not intrinsic to $E$, rather they are choices in how $E$ is subdivided\cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+In figure~\ref{fig:base_space_div} we can divide a rectangular base space such that there is a short fiber and long base space or a long fiber and short base space.
+This is analogous to long and wide forms of the same table \cite{wickham2014tidy}.
 
-$B$ is the quotient space of $E$, meaning it is the set of equivalence classes of elements $p$ in $E$ defined via the map $\pi: E \rightarrow B$ that sends each $p \in E$ to its equivalence class in $[p] \in B$ \cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
-
-As shown in figure~\ref{fig:kquote}, the fibers $F$ divide $E$ into smaller spaces consisisting of $F$ and an open set neighborhood \cite{FiberBundle2020, rowlandFiberBundle}  around $F$.
-%%so why topology? 
 \paragraph{Example}
-%%% pull up data structures like discrete, continous, etc w/ base space 
-%%% little drawing dots, line, plane, cube
-
 \begin{figure}[ht]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_1k.png}
+    \includegraphics[width=\textwidth]{figures/sections/math/temp_1k.png}
     %% add box around neighboring P and Map
@@
 \begin{figure}[H]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2k.png}
+    \includegraphics[width=\textwidth]{figures/sections/math/temp_2k.png}
 \end{figure}
@@
 
-\subsubsection{Fiber}
+\subsubsection{Triangulization}
+\label{sec:triangulization}
+
+The base space $K$ is a representation of the connectivity of the data, specifically whether the points in $E$ are discrete or sampled from a continous space.
+The same dataset can be expressed with different $K$.
+
+In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
+
+\begin{figure}[H]
+    \label{fig:simplex}
+    \includegraphics{figures/sections/math/simplex.png}
+    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
+Higher order simplicies are outside the scope of this paper.}
+\end{figure}
+
+K is a triangularizable topological space; one triangularization scheme is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
+
+\paragraph{Example}
+chopping up a torus maybe? talk about how that gets unpacked into triangles and then into vertices
+
+\subsubsection{Fiber Space}
+\label{sec:fiber_data}
 %% pull fiber stuff up here? what is it? why? fiber as cartesian product of sets
@@
 
+\paragraph{Definition}
+A monoid\cite{Monoid2021} $M$ is a set that is closed under an associative binary operator $\ast$ and has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $X$ with an action $\bullet$ with the properties:
+\begin{description}
+    \item[closure] $\bullet: M\times X \rightarrow X$,
+    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in X, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
+    \item[identity] $\text{for all } x\in X, e\in M,  e\bullet x = x$ 
+\end{description}
+
 \paragraph{Example}
@@
 \subsubsection{Section}
+\label{sec:section_data}
 %% add context
-The section $f$ is the mapping from base space to total space $f: B\rightarrow E$ 
+The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$ 
 \begin{equation}
@@
 
-In a locally trivial fiber bundle, $E = B \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
+In a trivial fiber bundle, $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
 \begin{equation}
@@
 %% add motivation - zoom, subsection, streaming, also for realizing visualization, restricting to that piece - need to know near point w/o
+
 As described in equation~\ref{eq:local_trivial}, there is a local space $U \subset B$ around every $b$.
@@
 . Restricting the artist to the sheaf means the artist knows the data in $F$ and also has access to derivatives of the data. This property is useful for some visual transformations. 
-%% gives all direvatives but we need the jet which is Add stuff about jets
-
-\subsubsection{Triangulization}
-%%drop down to after the fiber section
-The base space $K$ is a representation of the connectivity of the data, specifically whether the points in $E$ are discrete or sampled from a continous space.
-The same dataset can be expressed with different $K$.
-
-In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
-
-\begin{figure}[H]
-    \label{fig:simplex}
-    \includegraphics{figures/sections/math/simplex.png}
-    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
-Higher order simplicies are outside the scope of this paper.}
-\end{figure}
-
-%% k is a triangularizable topological space 
-One way to represent the topological space K is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
-Simplices are a way of encoding the connectivity of each observation ($\sigma(k)$) to another: %%formula for simplacial complexes/simplacial sets
-
-\paragraph{Example}
-chopping up a torus maybe? talk about how that gets unpacked into triangles and then into vertices
+%% gives all direvatives but we need the jet which is add stuff about jets
 
@@
 \label{sec:display}
+We define a graphic space $H$ such that we do not have to assume the physical output space of the renderer. This means that the graphic in $H$ can be output to a screen or 3D printed space or a dome.
 
-We model the prerender space on which lives on ideal version of the visulaization as a fiber bundle (H, S, $\pi$, D).
-$H$ is the predisplay space, with a fiber $D$ dependent on the target physical display and a base space of $S$.
+We model the prerender space as a fiber bundle (H, S, $\pi$, D). $H$ is the predisplay space, with a fiber $D$ dependent on the target display and a base space of $S$.
 
 \subsubsection{Base space}
-%%pull up to the very top that dim(k) < screen space, need to thicken therefore S
-$K$ can be considered a subspace of the screen base space $S$ such that $\xi: S \rightarrow K$ is a deformation retraction \cite{RetractionTopology2020}
+The underlying topology $S$ of a graphic often needs more dimensions than the data topology $K$ because of the specifications of the display space.
+For example, a line plot on a plane (such as a screen or a piece of paper) by necessity needs to also have a thickness so that it is visible, which maps back to a set of connected points in $H$.
+The topology of these connected points is therefore the region $s \subset S$ such that $\xi: S \rightarrow K$ is a deformation retraction \cite{RetractionTopology2020}
 \begin{equation}
@@
 \end{equation}
-%%map backwords via pullback can copy values over k over and over to corresponding S_k
+The pullback map $\xi^*$ copies each value in $V$ over $k$ to $s$ in corresponding $S_k$ such that $\xi^*V$ can have multiple values that map to one value in $V$.
 
@@
 
-\paragraph{Definition}
-A monoid\cite{Monoid2021} $M$ is a set that is closed under an associative binary operator $\ast$ and has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $X$ with an action $\bullet$ with the properties:
-\begin{description}
-    \item[closure] $\bullet: M\times X \rightarrow X$,
-    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in X, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
-    \item[identity] $\text{for all } x\in X, e\in M,  e\bullet x = x$ 
-\end{description}
-
 \paragraph{Example: Partial Order}
@@
 \begin{figure}[H]
-    \includegraphics[width=0.4\linewidth]{figures/sections/math/path_of_q}
-    \label{fig:q}
+    \includegraphics[width=\textwidth]{figures/sections/math/path_of_q}
+    \label{fig:artist_q}
     \caption{The $\nu$ functions convert data $E$ to visual $V$.
@@
 
-The assembly function $Q: \Gamma(V) \rightarrow \Gamma(H)$ composites the visual variables in $V$ into an element in $H$.
-Given a monoid action on a set in $V$, there should be a monoid action on the corresponding $Q(\Gamma(V))$.
-While $\Gamma(V)$ holds for all cases, we can specialize to the bundle $V$ or the sheaf $\mathcal(V)$ depending on the specific visualization.
+As shown in figure~\ref{fig:artist_q}, $Q$ takes the individual fields in $V$ as input and outputs a single piece of a graphic on $H$.
+As with $\nu$, the constraint on $Q$ is that for every monoid actions on the input in $V$ there is a corresponding monoid action on the output in $H$.
 
@@
 
-For each region $s$ in the display space $H$, the mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology} it belongs to can be found by mapping $s$ back to $K$ via the lookup on $S$ then taking $\xi(s)$ back to a point on $k \in K$ which lies on the path connected component $J \subset K$.
+To output a mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}, $Q$ is called with all the regions $s$ that map back to a set of connected components $J \subset K$:
+\begin{equation}
+J = \{j \in K \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=k \text{ and }\gamma(1)=j\}
+\end{equation}
+where the path\cite{ConnectedSpace2020}  $\gamma$ from $k$ to $j$ is a continous function from the interval [0,1].
 
-%% defined J_{k} first, then pull down to here, then can do \rho S_j
+We define the mark as the graphic generated by $Q(S_j)$:
+
 \begin{equation}
     \begin{tikzcd}
-        H \arrow[r, shift left] & S \arrow[l, "\rho(\xi^{-1}(J))", shift left] \arrow[rr, "\xi(s)", shift left] &  & J_{k} =  \{j \in K \text{ s.
-t.
-} \exists \gamma \text{ s.t.
-} \gamma(0)=k \text{ and }\gamma(1)=j\} \arrow[ll, "\xi^{-1}(J)", shift left]
+        H \arrow[r, shift left] & S_j \arrow[rr, "\xi(s)", shift left] \arrow[l, "\rho(S_j)"] &  & J_{k} \arrow[ll, "\xi^{-1}(J)"]
         \end{tikzcd}
@@
 \end{equation}
-where the path\cite{ConnectedSpace2020}  $\gamma$ from $k$ to $j$ is a continous function from the interval [0,1].
-To get back to the display space $H$ from $J$, the inverse image of $J \in S, \xi^{-1}(J)$ is pushed back to $S$, and then $\rho(\xi^{-1}(J))$ maps it into $R^{7}$ such that $Q\circ\xi^*\mu(\xi^{-1}(J))$ generates a graphical mark.
+
+where the set $j \subset J$ is the set of marks in the graphic.
 

### 2021-01-13 draft/sections/math.tex
start of nu constraint check functions + words and words

@@
 
-In this section we introduce a mathematical description of the visualization pipeline where artist $\mathcal{A}$ functions transform data  $\mathcal{E}$ to an intermediate representation in a prerendered display space $\mathcal{H}$.
+In this section we introduce a mathematical description of the visualization pipeline where artist $\mathfrak{A}$ functions transform data space $\mathfrak{E}$ to an intermediate representation in a prerendered display space $\mathfrak{H}$.
 
@@
 
-We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they are products of a topological total space that encodes points and a topological base space that can encode the connectivity of those points:
+We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow to seperate concerns
+
+%%find word to put here topological total space that encodes points and a topological base space that can encode the connectivity of those points:
 \begin{itemize}
@@
 
-\subsection{Data Fiber Bundles}
+\subsection{Data Space $E$}
 As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle ($E$, $K$, $\pi$, $F$)
@@
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[r, "\pi"] & B
+        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
     \end{tikzcd}
@@
 
-\paragraph{example}
+\paragraph{Example}
 
@@
 
-The fibers $F$ in figure~\ref{fig:data_fiber_bundle} lie in some total space $E$ and are points that lie in a 2D space.
-For example, this space can be $\mathrm{temperature}\times \mathrm{pressure}$ or $\mathrm{time} \times \mathrm{precipitation}$.
-$K$ is a line on the interval [0,1] and indicates that the points lie on a 1D continous space, but we derive whether the data is a distribution or a timeseries or a function from the fields in $F$.
-The sections $\tau_1$ and $\tau_2$ are distinct maps from $k$ to the blue and purple points in $E$ respectively.
+The fiber bundle in figure~\ref{fig:data_fiber_bundle} can encode different types of data.
+$F$ is a cartesian product of the codomains of 2 variables and $K$ is a line on the interval [0,1].
+The sections $\tau_{1}$ and $\tau_{2}$ are different sets of data values in the space.
 
-\subsubsection{Base Space}
+\begin{description}
+    \item[distribution] $F$ is the cartesian cross product of temperature values and the range of probabilities [0,1] such that every point in $E$ is a (temperature, probabilty) pair.
+$\tau_1$ and $\tau_2$ each return a different distribution of temperatures.
+    \item[timeseries]  $F = \{\text{temperature}\} \times \{\text{timestamps}\}$, such that every point is a (temperature, time) pair.
+$\tau_1$ and $\tau_2$ each return different timeseries of temperature.
+\end{description}
+
+\subsubsection{Base Space $\mathcal{K}$}
 \label{sec:base_data}
@@
 
-\subsubsection{Triangulization}
+\paragraph{Triangulization}
 \label{sec:triangulization}
@@
 
-\paragraph{Example}
+\subparagraph{Example}
 chopping up a torus maybe? talk about how that gets unpacked into triangles and then into vertices
@@
 \label{sec:fiber_data}
+
 %% pull fiber stuff up here? what is it? why? fiber as cartesian product of sets
@@
 
-$F_b = \pi^{-1}({b \in B}) $ is homomorphic to the fiber $F$.
+$F_k = \pi^{-1}({k}) $ is homomorphic to the fiber $F$.
 %% add a sentence for why we care
@@
 \end{description}
+
+$M = M_{1} \times M_{2} \times \ldots \times M_{n}$
 
@@
     \label{eq:section_domain}
-    \pi(f(b)) = b \text{ for all } b \in B 
+    \pi(f(k)) = k \text{ for all } k \in K 
 \end{equation}
@@
 The space of all possible sections $f$ of $E$ is $\Gamma(E)$.
-All sections $f \in \Gamma(E)$ have the same fibers $F$ and connectivity $B$.
+All sections $f \in \Gamma(E)$ have the same fibers $F$ and connectivity $K$.
 \paragraph{Example}
@@
 
-As described in equation~\ref{eq:local_trivial}, there is a local space $U \subset B$ around every $b$.
-The inclusion map $\iota: U \rightarrow B$ can be pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $U$.
+Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(E)$:
+
 \begin{equation}
@@
         \iota^*E \arrow[d, "\pi"']           & E \arrow[d, "\pi"'] \arrow[l, "\iota^*"']         \\
-        U \arrow[u, "\iota^*f"', bend right] & B \arrow[u, "f"', bend right] \arrow[l, "\iota"']
+        U \arrow[u, "\iota^*\tau"', bend right] & K \arrow[u, "\tau"', bend right] \arrow[l, "\iota"']
         \end{tikzcd}
@@
 
-The localized section of fibers $\iota^*f: U \rightarrow \iota^*E$ is the sheaf $\mathcal{O}(E)$ with germ of $\xi^*f$ which is the localized section.
-The neighborhood of points the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic} which is a ...$\dots$
-. Restricting the artist to the sheaf means the artist knows the data in $F$ and also has access to derivatives of the data. This property is useful for some visual transformations. 
-%% gives all direvatives but we need the jet which is add stuff about jets
+As shown in equation~\ref{eq:fiber_local_trivial}, there is a local space $U \subset K$ around every $k$.
+The inclusion map $\iota: U \rightarrow K$ is pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $K$.
+The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf with a germ of $\xi^*\tau$.
+The neighborhood of points $k_i$ surrounding the point $k$ the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
 
-\subsection{Prerender Space}
+The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is a type of sheaf that allows for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.
+
+\subsection{Prerender Space $H$}
 \label{sec:display}
@@
 
-Here we assume a single opaque 2D image such that the $z$ and $alpha$ fields can be omitted. To support overplotting and transparency, we consider $D=R^{7}$
-%%H can be other spaces
+Here we assume a single opaque 2D image such that the $z$ and $alpha$ fields can be omitted. To support overplotting and transparency, we can consider $D=R^{7}$
 
@@
 \subsection{Artist}
-%%introduce O(E) up in previous section, then refer back to that section
 
@@
 
-The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber $P$.
+The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels,} in fiber $P$.
 The visual transformer $\nu$ is a set of functions each targeting a different $\mu$
@@
 %%needs context, pulling monoids up gives it
+Table~\ref{tab:mpl_visual_variable_fiber} is a
+
 \subsubsection{Visual Channels}
@@
 
-\paragraph{Example: Partial Order}
+\paragraph{Example: Ordering}
 To preserve ordering of elements in $E_i$, $\nu$ must be a monotonic function such that given $e_1, e_2 \in E_{i}$
@@
 \paragraph{Example: Translation}
-\textbf{fairly certain I lost the thread here}
 According to Stevens, interval data is a set with general linear group actions \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
-Position is a visual variable that can support translation \cite{bertinIIPropertiesGraphic2011,krygierMakingMapsVisual2005,munznerMarksChannels}.
+Position is a visual variable that can support translation
 \begin{equation}
@@
 
-To consutruct a valid $\nu$, the diagram must commute for all monoid actions on the sets $E_i, V_i$.
+To consutruct a valid $\nu$, the diagram must commute for all monoid actions on the sets in $E_i, V_i$.
 
@@
 As shown in figure~\ref{fig:artist_q}, $Q$ takes the individual fields in $V$ as input and outputs a single piece of a graphic on $H$.
-As with $\nu$, the constraint on $Q$ is that for every monoid actions on the input in $V$ there is a corresponding monoid action on the output in $H$.
+As with $\nu$, the constraint on $Q$ is that for every monoid actions on the input in $V$ there is a corresponding monoid action on the output in $H$:
+\begin{equation}
+    Q: \Gamma(V) \rightarrow \Gamma(H)
+\end{equation}
 
-\paragraph{Proposition}
-If $\forall g \in M$ and $\forall x_1, x_2 \in \Gamma(V)$ then  $Q(x_1) = Q(x_2) \\implies Q(g\circ x_1) = Q(g\circ x_2)$; therefore we can define a group action on $Q(\Gamma(V))=Y$ as $g \circ y = y'$ where $y' = Q(g\circ x)$ with $x \in f^{-1}(y)$
+$Q: \mu \mapsto \rho$
+
+We want a monoid/group action on $Q(\Gamma(V))$, do not need an action on all of $\Gamma(H)$
 
@@
 
+$M = M_{1} \times M_{2} \times \ldots \times M_{n}$
+color X xpos (Cross product of monoid actions over V/E), cartesian product b/c independent. $M$ over $E$ was translated to acting over $\Gamma(V)$ of these act on $\Gamma(V)$
+$\nu$ translates the sets, constraint on $\nu$ is equivariance of M. Is limited to the $Q(\Gamma(V))\subset of \Gamma(H)$ because not all visual tramsforms ($\mu$) are supported by all  $\rho$.
+%%insert little thickness sketch maybe?O->(O)
+So we define the actions $M$ on the image of $Q$, $Q(\Gamma(V)) = Y$. We can backward define our actions?
+
+When the visual atrtribute in $\mu$ is some kind of direct property of D, then we can define $M$ on $\Gamma(H)$ and require that $Q$ preserves it.
+But we have graphical parameters that do not apply to the whole glyph and therefore are not direct mappings on D, such as facecolor or line thickness.
+For these types of properties, we need to define an action on the target graphic $Q(\Gamma(V)) \in \Gamma(H)$.
+%%in a direct transformation of D, the map on s to d will fall out. Need the eqiuvariance in cases where there are not direct maps to D but instead are partial/compositional/
+
+If can check this on $Q$, then you can define an equivariant $M$. Use action of $M$ on $V$ to define action $M$ on $Y$.
+
+Let ....
+
+If $\forall g \in M$ and $\forall \mu, \mu\prime \in X$, 
+\begin{equation}
+Q(\mu) = Q(x2) 
+\end{equation}
+If true, then we can define a group action on $Y$ is defined as $g\circ \rho = \rho\prime$ where $\rho\prime=Q(g\circ \mu)$ with $\mu \in Q^{-1}(\rho)$.
+
+tuple of figure above Is ... %%mu is first circle, mu' is second circle, broken mu is squggle%%
+if two $\mu$ go to the same graphic(glyph), then if we transform them the same way they need to generate the same graphic in H.
+two sections of V if they map to the same thing, they need to map to the same thing if you transform they need to map to the same thing.
+
+change g to  m
+
+%% m is a cartesian product of group actions in M, for example {identity, ...., identity, thickens}
+%% \mu and \mu' are different sections of \Gamma(V)
+
+\paragraph{Example: Invalid Q}
+Check a well defined map $M\times Y \rightarrow Y$.
+
+constraint: inputs go to same output means changes to inputs mean same changes to output
+
 \subsubsection{Visual Idioms: Equivalance class of artists}
-n $O(E)$ of the same type, they output the same type of prerender $O(H)$:
+Given $O(E)$ of the same type that output to the same type of graphic $O(H)$, the
 

### 2021-01-13 code/matplottoy/artists/constraint.py

@@ -1,0 +1,17 @@
+
+def is_partial_order(values, transform):
+    return True
+
+def is_nominal(values, transform):
+    return True
+
+def is_ordinal(values, transform):
+    return True
+
+def is_interval(values, transform):
+    return True
+
+def is_ratio(values, transform):
+    return True
+
+

### 2021-01-13 draft/sections/math.tex
reworked\reorganized example

@@
 
-In this section we introduce a mathematical description of the visualization pipeline where artist $\mathfrak{A}$ functions transform data space $\mathfrak{E}$ to an intermediate representation in a prerendered display space $\mathfrak{H}$.
+In this section we introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.
 
@@
     \label{eq:artist}
-    \mathcal{A}: \mathcal{E} \rightarrow \mathcal{H}
+    \mathscr{A}: \mathscr{E} \rightarrow \mathscr{H}
 \end{equation}
 
-We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow to seperate concerns
-
-%%find word to put here topological total space that encodes points and a topological base space that can encode the connectivity of those points:
+We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the fields in a dataset from how the values in those fields are connected to each other:
 \begin{itemize}
@@
 
-\subsection{Data Space $E$}
+\subsection{Data Space}
 As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle ($E$, $K$, $\pi$, $F$)
@@
 
-\paragraph{Example}
-
-\begin{figure}[H]
-    \label{fig:data_fiber_bundle}
-    \includegraphics[width=1\linewidth]{figures/sections/math/fiberbundle.png}
-    \caption{$F$ is an $n\times m$ space of points.
-The section $\tau_1$ returns the blue points,  while $\tau_2$ returns the purple points.
-$\Gamma(E)$ is the set of all sections, including $\tau_1$ and $\tau_2$}.
-\end{figure}
-
-The fiber bundle in figure~\ref{fig:data_fiber_bundle} can encode different types of data.
-$F$ is a cartesian product of the codomains of 2 variables and $K$ is a line on the interval [0,1].
-The sections $\tau_{1}$ and $\tau_{2}$ are different sets of data values in the space.
-
-\begin{description}
-    \item[distribution] $F$ is the cartesian cross product of temperature values and the range of probabilities [0,1] such that every point in $E$ is a (temperature, probabilty) pair.
-$\tau_1$ and $\tau_2$ each return a different distribution of temperatures.
-    \item[timeseries]  $F = \{\text{temperature}\} \times \{\text{timestamps}\}$, such that every point is a (temperature, time) pair.
-$\tau_1$ and $\tau_2$ each return different timeseries of temperature.
-\end{description}
-
 \subsubsection{Base Space $\mathcal{K}$}
@@
 \begin{figure}[H]
-    %%\includegraphics[\textwidth]{}
+    \includegraphics[width=\textwidth]{figures/sections/math/k_different_types.png}
     \label{fig:base_space_types}
-    %%% pull up data structures like discrete, continous, etc w/ base space 
-    %%% little drawing dots, line, plane, cube
     \caption{The topological base space $K$ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
@@
 
-Data can be discrete observations, timeseries, maps, fields \cite{munznerWhatDataAbstraction2014} and $K$ is a set of points $k$ that can act as keys from a representative space, such as seen in figure~\ref{fig:base_space_types}, to the data values in $E$ \cite{munznerWhatDataAbstraction2014}.
+Datasets have a semantic topology such as the values are interpreted as discrete observations or part of a timeseries or map, or nodes in a networks \cite{munznerWhatDataAbstraction2014,geveci2012vtk}.
+As illustrated in fig~\ref{fig:base_space_div}, $K$ is this underlying structure.
+$K$ does not directly know the values; instead it is the sections that define the lookup between keys $k \in K$ and the corresponding values in $E$.
 
@@
 \end{figure}
-
-$K$ and $F$ are not intrinsic to $E$, rather they are choices in how $E$ is subdivided\cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+The topology $K$ and the fields $F$ are determined by how $E$ is subdivided\cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
 In figure~\ref{fig:base_space_div} we can divide a rectangular base space such that there is a short fiber and long base space or a long fiber and short base space.
 This is analogous to long and wide forms of the same table \cite{wickham2014tidy}.
-
-\paragraph{Example}
-\begin{figure}[ht]
-    \includegraphics[width=\textwidth]{figures/sections/math/temp_1k.png}
-    %% add box around neighboring P and Map
-    \label{fig:k_data}
-\end{figure}
-
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/sections/math/temp_2k.png}
-\end{figure}
-
-in figure~\ref{fig:k_data}, temperature is the only one data field in $r$ but the $K$ base spaces are different.
-subfig[1] is a timeseries, so the temperature in $r$ at time $t$ is dependent on the temperature in $r_{t-1}$ and the temperature in $r_{t+1}$ is dependent on  $r_t$; this connectivity is expressed as a one dimensional $K$ where $K$ is the number line.
-In the case of the map, every temperature in $r$ is dependent on its nearest neighbors on the plane, and one way to express this is by encoding $K$ as a plane.
-$K$ does not know the time or latitude or longitude of the point as those are metadata variables describing the $k$ rather than the value of $k$.
-The mapping $\tau: K \rightarrow E$ provides the binding between the key $k \in K$ and the value $r$ in $E$ \cite{munznerChDataAbstraction}.
 
@@
 
-\subsubsection{Example}
-\begin{figure}[H]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_2f.png}
-    \label{fig:}
-\end{figure}
-\begin{figure}[H]
-    \includegraphics[width=0.2\linewidth]{figures/sections/math/temp_3f.png}
-\end{figure}
+\subsubsection{Sheaf and Stalk}
 
-%% pull out measurement
-The fiber in figure~\ref{fig:temp} is the space of possible temperature values in degrees celsius, such that $F=[temp_{min}, temp_{max}]$ and is named \textrm{Temp}.
-In figure~\ref{fig:temp_time} \textrm{time} is encoded as a second dimension.
-This means that the set of possible values $F$ with $C=\{\textrm{Temp}, \textrm{Time}\}$:
-
-\begin{equation}
-F = [temp_{min}, temp_{max}] \times [time_{min}, time_{max}]
-\end{equation}
-
-and the function $\tau$ that retrieves records from $F$ is
-
-\begin{align}
-\tau(k) =(k, (r: \textrm{Temp}\rightarrow &temp, r: \textrm{Time}\rightarrow time))\\
-&temp \in [temp_{min}, temp_{max}], time \in [time_{min}, time_{max}])
-\end{align}
-
-Since $\tau(k)=(k, r)$, $temp$ is bound to a named data field and $sigma$ binds $temp$ to a temperature data type.
-
-\subsubsection{Sheaf and Stalk}
-%% add motivation - zoom, subsection, streaming, also for realizing visualization, restricting to that piece - need to know near point w/o
-
-Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(E)$:
+Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(E)$
 
@@
 \end{equation}
-
-As shown in equation~\ref{eq:fiber_local_trivial}, there is a local space $U \subset K$ around every $k$.
-The inclusion map $\iota: U \rightarrow K$ is pulled back such that $\iota^{*}E$ is the space of $E$ restricted over $K$.
-The localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ is the sheaf with a germ of $\xi^*\tau$.
+which is the localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ pulled back via the inclusion map $\iota: U \rightarrow K$.
+The localized section is the germ $\xi^*\tau$.
 The neighborhood of points $k_i$ surrounding the point $k$ the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
@@
 The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is a type of sheaf that allows for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.
+
+\subsubsection{Example: Temperature}
+\textcolor{red}{Moved \& walked through b/c was getting clunky to not have terms yet}\\
+The fiber bundle model is flexible enough to express some of the many different forms that temperature data can come in.
+
+\begin{figure}[H]
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/sections/math/temp_1k.png}
+        %% add box around neighboring P and Map
+        \label{fig:base_example_line}
+    \end{subfigure}
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/sections/math/temp_2k.png}
+        \label{fig:base_example_plane}
+    \end{subfigure}
+    \label{fig:base_example}
+    \caption{These two datasets have the same fiber of temperature but different base spaces.
+In figure~\ref{fig:base_example_line} the temperature values are 1D continuous, while in figure~\ref{fig:base_example_plane} the temperature values are 2D continuous.}
+\end{figure}
+
+The datasets in figure~\ref{fig:base_example} have identical fibers that encode a set of temperature values.
+In figure~\ref{fig:base_example_line} the temperatures lie on a line such that a section could return a timeseries or a distribution.
+In figure~\ref{fig:base_example_plane}, the temperatures lie on a 2D continuous plane; a section could return a map or contour.
+Because the fiber is 1D, it does not encode metadata
+
+\begin{figure}[H]
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/sections/math/temp_2f.png}
+        \label{fig:fiber_example_plane}
+    \end{subfigure}
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/sections/math/temp_3f.png}
+        \label{fig:fiber_example_cube}
+    \end{subfigure}
+    \label{fig:fiber_example}
+    \caption{The fiber is expanded to include metadata fields that describe the semantics of $K$.
+In figure~\ref{fig:fiber_example_plane} the fiber is \textrm{temperature} $\times$ \textrm{time} and in figure~\ref{fig:fiber_example_cube} the fiber is \textrm{temperature}$\times$ \textrm{latitude} $\times$ {longitude}}.
+\end{figure}
+
+To encode the metadata, the fiber is expanded as illustrated in figure~\ref{fig:fiber_example}.
+The fiber in figure~\ref{fig:fiber_example_plane} is the cartesian product of the space of possible temperature values in degrees celsius and space of possible time values
+\begin{equation}
+F = [temp_{min}, temp_{max}] \times [time_{min}, time_{max}]
+\label{eq:fiber_plane}
+\end{equation}
+
+while the fiber in figure~\ref{fig:fiber_example_cube} is the cartesian product of temperature, latitude, and longitude
+\begin{equation}
+F = [temp_{min}, temp_{max}] \times [-90, 90] \times [-180, 180]
+\label{eg:fiber_cube}
+\end{equation}
+
+such that $E$ is the space of all possible points in $F$.
+
+\begin{figure}[H]
+    \includegraphics[width=1\linewidth]{figures/sections/math/fiberbundle.png}
+    \label{fig:fiber_example_section}
+    \caption{The section $\tau_1$ returns the blue points,  while $\tau_2$ returns the purple points. $\Gamma(E)$ is the set of all sections, including $\tau_1$ and $\tau_2$}.  
+\end{figure}
+Given the fiber described in equation~\ref{eq:fiber_plane}, the sections $\tau_{1}$ and $\tau_{2}$ in figure~\ref{fig:fiber_example_section} return tuples of the form
+\begin{equation}
+\tau(k) = (k, (temperature, time))
+\end{equation}
+such that sections with the constraint that time is monotonic return a timeseries.
 
@@
 $\nu$ translates the sets, constraint on $\nu$ is equivariance of M. Is limited to the $Q(\Gamma(V))\subset of \Gamma(H)$ because not all visual tramsforms ($\mu$) are supported by all  $\rho$.
-%%insert little thickness sketch maybe?O->(O)
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{figures/sections/math/diff_type_q.png}
+\end{figure}
 So we define the actions $M$ on the image of $Q$, $Q(\Gamma(V)) = Y$. We can backward define our actions?

### 2021-01-13 draft/sections/math.tex
fibers w/ monoids

@@
 \label{sec:triangulization}
-
-The base space $K$ is a representation of the connectivity of the data, specifically whether the points in $E$ are discrete or sampled from a continous space.
-The same dataset can be expressed with different $K$.
-
 In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
@@
     \includegraphics{figures/sections/math/simplex.png}
-    \caption{Simplices encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations.
-Higher order simplicies are outside the scope of this paper.}
+    \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations}
 \end{figure}
 
-K is a triangularizable topological space; one triangularization scheme is as a set composed of simplices, such as those shown in figure~\ref{fig:simplex}.
+K is a triangularizable topological space; one triangularization scheme is as a set composed of simplices\cite{geometricaSimplicialComplexes}, such as those shown in figure~\ref{fig:simplex}.
 
 \subparagraph{Example}
+\textcolor{blue}{Need to sketch and write this example}\\
 chopping up a torus maybe? talk about how that gets unpacked into triangles and then into vertices
@@
 \label{sec:fiber_data}
+Datasets can be univariate or multivariate, homogenous or heteregenous, and fibers are where we encode this information.
+Each field in the dataset $f_{i}$ contains a range of values that come from the codomain of $f_{i}$.
+The fiber is the cartesian product of the codomains of fields:
+\begin{align}
+F &= codomain(f_{0})\times \ldots \times codomain(f_{n})\\
+F &= F_{0} \times \ldots \times F_{n}
+\end{align}
 
-%% pull fiber stuff up here? what is it? why? fiber as cartesian product of sets
-%% actual values are in the fiber 
-%% gonna view spaces as simple piece put together 
-%% Our fibers are cross products of (column name, domain) as described by Spivak's \cite{spivakSIMPLICIALDATABASES}.
+We formulate the fiber such that it has a mapping between the codomain and the field name, as discussed in Spivak's simplacial description of databases \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010}.
+An important property of data fields are the measurement type, which we generalize to identifying the left monoid action such that the monoid actions $M$
+\begin{equation}
+M = M_{0} \times \ldots\times M_{i}
+\end{equation}
+is the cartesion cross product of monoids applied to each $F_{i}\in F$.
+A monoid \cite{Monoid2021} $M$ is a set that is closed under an associative binary operator $\ast$ and has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
 
-\begin{equation}
-    \label{eq:fiber_local_trivial}
-    \begin{tikzcd}
-        \pi^{-1}(U) \arrow[r, "\varphi"] \arrow[d, "\pi"'] & U \times F \arrow[ld, "\mathrm{proj}_U"] \\
-        U                                                  &                                         
-    \end{tikzcd}
-\end{equation}
-such that $\varphi: \pi^{-1}(U) \rightarrow U \times F$ is a homeomorphism where $\pi$ and $\mathrm{proj}_U$ both map to $U$ and the fiber over $k$
-
-%% rewrite this section - what are the goals here and why should we care? 
-%% devoid of any context
-
-$F_k = \pi^{-1}({k}) $ is homomorphic to the fiber $F$.
-%% add a sentence for why we care
-%% fibers are cartesian products of these typles sets (name, domain, monoidal action)
-
-%%the break out more about how fibers are cross products and what that means and how that's the structure - pull the monoids definitions here
-
-\paragraph{Definition}
-A monoid\cite{Monoid2021} $M$ is a set that is closed under an associative binary operator $\ast$ and has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $X$ with an action $\bullet$ with the properties:
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet$ with the properties:
 \begin{description}
-    \item[closure] $\bullet: M\times X \rightarrow X$,
-    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in X, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
-    \item[identity] $\text{for all } x\in X, e\in M,  e\bullet x = x$ 
+    \item[closure] $\bullet: M\times F \rightarrow X$,
+    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
+    \item[identity] $\text{for all } x\in F, e\in M,  e\bullet x = x$ 
 \end{description}
 
-$M = M_{1} \times M_{2} \times \ldots \times M_{n}$
-
-\paragraph{Example}
-%% stevens scales are examples of this 
-%% one ecample of F1 is (names, strings, concat), (ordinal, Z,  ordering), ()
+The Steven's measurement scales are identified via the group actions on $F_{i}$  \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale} they support.
+For example, nominal variables are permutable and interval values are translatable.
+We identify monoid actions, rather than group, to support an extended variaty of scales such as partial order
 
@@
 \label{sec:display}
-We define a graphic space $H$ such that we do not have to assume the physical output space of the renderer. This means that the graphic in $H$ can be output to a screen or 3D printed space or a dome.
-
-We model the prerender space as a fiber bundle (H, S, $\pi$, D). $H$ is the predisplay space, with a fiber $D$ dependent on the target display and a base space of $S$.
+We define a graphic space $H$ such that we do not have to assume the physical output space of the renderer.
+This means that the graphic in $H$ can be output to a screen or 3D printed space or a dome.
+We model the prerender space as a fiber bundle (H, S, $\pi$, D).
+$H$ is the predisplay space, with a fiber $D$ dependent on the target display and a base space of $S$.
 
@@
 \subsection{Artist}
+\label{sec:artist}
 
@@
 \begin{table}[ht]
-    \label{tab:mpl_visual_variable_fiber}
     \renewcommand{\arraystretch}{2}
@@
     \end{tabulary}
+    \label{tab:mpl_visual_variable_fiber}
 \end{table}
-%%needs context, pulling monoids up gives it
-Table~\ref{tab:mpl_visual_variable_fiber} is a
+
+Table~\ref{tab:mpl_visual_variable_fiber} is an example of the visual fiber defined in terms of common parameters to plots in Matplotlib.
+The range of $\mu_{i}$ determine the monoid actions on $\mu_{i}$.
+A section of V $\mu$ is a tuple of visual values that specifies the visual characteristics of a glyph.
+For example, given a fiber of $\{xpos, ypos, color\}$ one section is $\{.5, .5, (255, 20,147)\}$.
+$Q$ determines how this section is applied to a graphic.
 
@@
 
-To consutruct a valid $\nu$, the diagram must commute for all monoid actions on the sets in $E_i, V_i$.
+To construct a valid $\nu$, the diagram must commute for all monoid actions on the sets in $E_i, V_i$.
 
@@
 As shown in figure~\ref{fig:artist_q}, $Q$ takes the individual fields in $V$ as input and outputs a single piece of a graphic on $H$.
-As with $\nu$, the constraint on $Q$ is that for every monoid actions on the input in $V$ there is a corresponding monoid action on the output in $H$:
+As with $\nu$, the constraint on $Q$ is that for every monoid actions on the input in $V$ there is a corresponding monoid action on the output in $H$
 \begin{equation}
@@
 \end{equation}
+When $Q: \mu \mapsto \rho$ yields a $\rho$ that maps to the same values in $D$ over all $S_k$, then $M$ can be defined over $\Gamma(H)$ such that a constraint on $Q$ is that it must be equivariant.
+For example, when $\mu_{i}$ is the color of the glyph, it maps directly to (R,G,B) in D.
 
-$Q: \mu \mapsto \rho$
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/sections/math/diff_type_q.png}
+    \label{fig:artist_mark_change}
+\end{figure}
 
-We want a monoid/group action on $Q(\Gamma(V))$, do not need an action on all of $\Gamma(H)$
+Many $\mu_{i}$ are graphical parameters that do not apply to the whole glyph, such as edge thickness in figure~\ref{fig:artist_mark_change}.
+In these situations, not all  $\rho$ in $\Gamma(H)$ will support these parameters; instead we define an action on the output graphic $Q(\Gamma(V)) \in \Gamma(H)$ since by definition every section $\mu$ will have a corresponding $\rho$.
 
+We then define the constraint on $Q$ such that if $Q$ applied to two sections $\mu, \mu\prime$ generate the same graphic $\rho$, then the output of both sections acted on by the same monoid $m$ must also be the same.
+
+Lets call the visual encodings $\Gamma(V)=X$ and the graphic $Q(\Gamma(V))=Y$. If $\forall m \in M$ and $\forall \mu, \mu\prime \in X$, 
+\begin{equation}
+Q(\mu) = Q(\mu\prime)\implies Q(m\circ\mu) = Q(m\circ\mu\prime)
+\end{equation}
+then a group action on $Y$ can be defined as $m\circ \rho = \rho\prime$ where $\rho\prime=Q(g\circ \mu)$ with $\mu \in Q^{-1}(\rho)$.
+
+Given  
+\begin{itemize}
+    \item $P = \{xpos, ypos, color, thickness\}$
+    \item $\mu = {0,0,0, 1}$
+    \item  $Q(\mu)=\rho$ generates a piece of the thin circle in figure~\ref{fig:artist_mark_change}
+\end{itemize}
+
+the constraint on $Q$ means that the translation $m=\{e, e, e, x+2\}$ applied to $\mu$ such that $\mu\prime=\{0,0,0,3\}$ has an equivalent action on $\rho$ that causes $Q(\mu\prime)$ to be equivalent to the thicker circle in figure~\ref{fig:artist_mark_change}.
+
+\paragraph{Example: Invalid Q}
+\textcolor{blue}{Insert some degenerate Q that generates an inconsistent glyph}
+
+Check a well defined map $M\times Y \rightarrow Y$.
+
+constraint: inputs go to same output means changes to inputs mean same changes to output
+
+\paragraph{Graphical Marks}
 To output a mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}, $Q$ is called with all the regions $s$ that map back to a set of connected components $J \subset K$:
@@
 
-We define the mark as the graphic generated by $Q(S_j)$:
+We define the mark as the graphic generated by $Q(S_j)$
 
@@
 
-where the set $j \subset J$ is the set of marks in the graphic.
-
-$M = M_{1} \times M_{2} \times \ldots \times M_{n}$
-color X xpos (Cross product of monoid actions over V/E), cartesian product b/c independent. $M$ over $E$ was translated to acting over $\Gamma(V)$ of these act on $\Gamma(V)$
-$\nu$ translates the sets, constraint on $\nu$ is equivariance of M. Is limited to the $Q(\Gamma(V))\subset of \Gamma(H)$ because not all visual tramsforms ($\mu$) are supported by all  $\rho$.
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{figures/sections/math/diff_type_q.png}
-\end{figure}
-So we define the actions $M$ on the image of $Q$, $Q(\Gamma(V)) = Y$. We can backward define our actions?
-
-When the visual atrtribute in $\mu$ is some kind of direct property of D, then we can define $M$ on $\Gamma(H)$ and require that $Q$ preserves it.
-But we have graphical parameters that do not apply to the whole glyph and therefore are not direct mappings on D, such as facecolor or line thickness.
-For these types of properties, we need to define an action on the target graphic $Q(\Gamma(V)) \in \Gamma(H)$.
-%%in a direct transformation of D, the map on s to d will fall out. Need the eqiuvariance in cases where there are not direct maps to D but instead are partial/compositional/
-
-If can check this on $Q$, then you can define an equivariant $M$. Use action of $M$ on $V$ to define action $M$ on $Y$.
-
-Let ....
-
-If $\forall g \in M$ and $\forall \mu, \mu\prime \in X$, 
-\begin{equation}
-Q(\mu) = Q(x2) 
-\end{equation}
-If true, then we can define a group action on $Y$ is defined as $g\circ \rho = \rho\prime$ where $\rho\prime=Q(g\circ \mu)$ with $\mu \in Q^{-1}(\rho)$.
-
-tuple of figure above Is ... %%mu is first circle, mu' is second circle, broken mu is squggle%%
-if two $\mu$ go to the same graphic(glyph), then if we transform them the same way they need to generate the same graphic in H.
-two sections of V if they map to the same thing, they need to map to the same thing if you transform they need to map to the same thing.
-
-change g to  m
-
-%% m is a cartesian product of group actions in M, for example {identity, ...., identity, thickens}
-%% \mu and \mu' are different sections of \Gamma(V)
-
-\paragraph{Example: Invalid Q}
-Check a well defined map $M\times Y \rightarrow Y$.
-
-constraint: inputs go to same output means changes to inputs mean same changes to output
+in terms of $K$ because mark is a semantic term denoting the graphic representation of the data.
 

### 2021-01-13 draft/sections/math.tex
sections

@@
 \label{sec:section_data}
-%% add context
+The sections of the fiber bundles are the instances of the data.
+They are the functions that map from a key on $K$ to a set of values in $F$.
+For example, if we have a database, the section is the table that yields a row given an index.
 The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$
@@
         F \arrow[r, hook] & E \arrow[d, "\pi"']           \\
-        & B \arrow[u, "f"', bend right]
+        & K \arrow[u, "\tau"', bend right]
     \end{tikzcd}
@@
 
-such that $f$ is the right inverse of $\pi$
-\begin{equation}
-    \label{eq:section_domain}
-    \pi(f(k)) = k \text{ for all } k \in K 
-\end{equation}
-
-In a trivial fiber bundle, $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
+such that in a trivial fiber bundle, $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
 \begin{equation}
     \label{eq:section_return}
-f(b) = (b, g(b))
+    \tau(k) = (k, (x_{F_{0}}, \ldots, x_{F_{n}}))
 \end{equation}
+which we can also decompose such that
+\begin{align}
+\tau_{0}(k) &= (k, (x_{F_{0}}))\\
+            & \vdotswithin{=}\\
+\tau_{n}(k) &= (k, (x_{F_{n}}))
+\end{align}
 
-where the domain of $g(b)$ is $F_b$ and returns a point $p$ in $F_b$.
-The space of all possible sections $f$ of $E$ is $\Gamma(E)$.
-All sections $f \in \Gamma(E)$ have the same fibers $F$ and connectivity $K$.
-\paragraph{Example}
-%%talk about table and all that jazz
-
-For each field $c \in C$, the record function $r: C \rightarrow U_{\sigma}$ returns an object of type $\sigma(c) \in DT$.
-The set of all records $\Gamma(\sigma)$ is the set of all sections on $U_\sigma$.
-Spivak defines the $\tau$ mapping from an index of databases $K$ to records $\Gamma(\sigma)$ as $\tau: K \rightarrow \Gamma(\sigma)$.
-This is equivalent to $\tau: k \rightarrow E$ since $F = \Gamma(\sigma)$ and $F$ is the embedding in $E$ on which the records $r$ lie.
+which allows us to have field wise access to the data so long as there is a shared $K$.
 

### 2021-01-14 draft/sections/math.tex
feedback + local

@@
 
-We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the fields in a dataset from how the values in those fields are connected to each other:
-\begin{itemize}
-\item $E$ is a locally trivial fiber bundle over $K$ representing data space.
-\item $H$ is a fiber bundle over $S$ representing visual space
-\item $K$ and $S$ are a triangulizable topological space or a CW complex encoding the connectivity of points in $E$ and $H$ respectively
-\end{itemize}
-
+We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the variables in a dataset from how the values in those fields are connected to each other.
 The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}.
 
+%% we will define data, visual, and graphic spaces in terms of fiberbundles and then describe equivariant maps from E to V to H
+
 \subsection{Data Space}
-As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, we model data as a fiber bundle ($E$, $K$, $\pi$, $F$)
+% introduce what this is gonna happen in this section - we will talk about, fiber and connectivity 
+% motivate w/ data/ example - have weather station. %tidydata 
+% we will introduce the fiber bundle and its components base and total space to explain the fiber
 
+As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}
+
+%%% might get moved into basespace discussion:
+
+We model data as a fiber bundle ($E$, $K$, $\pi$, $F$)
 \begin{equation}
@@
 
-\subsubsection{Base Space $\mathcal{K}$}
+\subsubsection{Variable Field in F }
+\label{sec:fiber_data}
+ %%%first sentence should be a paragraph
+%%% butler's paper, no notion of distance on K, can only map F to the visual fibers, so need an X to map to Xpos
+% variable, measurement, pull in tidy here, 
+% % motivate by example, temperature \ time 
+%% F_0 = {temp_0, temp_n}, 
+%% then adding in a second set of variables time: 
+%% F_1 = {time_0, time_1}
+% break out into F first as set of values, then F=F_{0}\times F_{1} as component
+
+The fiber is the cartesian product of the variable spaces:
+\begin{align}
+F &= F_{0} \times \ldots \times F_{n}
+\end{align}
+
+We formulate the fiber such that it has a mapping between the somethine thing and the field name, as discussed in Spivak's simplacial description of databases \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010}.
+
+Fibers do not need to be 1D.
+
+\subsubsection{Measurement Type}
+%% maybe move somewhere else w/ some more motivation
+%% do this section in terms of $M0, F_0$
+An important property of data fields are the measurement type, which we generalize to identifying the left monoid action such that the monoid actions $M$ 
+\begin{equation}
+M = M_{0} \times \ldots\times M_{i}
+\end{equation}
+is the cartesion cross product of monoids applied to each $F_{i}\in F$.
+A monoid \cite{Monoid2021} $M$ is a set with an associative binary operator $\ast:M \times M\rightarrow M$.
+A monoid has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
+\begin{description}
+    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
+    \item[identity] $\text{for all } x\in F, e\in M,  e\bullet x = x$ 
+\end{description}
+
+The Steven's measurement scales are identified via the group actions on $F_{i}$  \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale} they support.
+For example, nominal variables are permutable and interval values are translatable.
+We identify monoid actions, rather than group, to support an extended variety of scales such as partial order.
+
+%%close out the fiber bundle section with the cartesian products, which gives us heterogenous data , univariate to multivariate.
+We can decompose the fiber bundle to components bundles \pi:E_1\rightarrow K, ....
+\pi: E_n\rightarrow K
+\subsubsection{Connectivity of the Data} 
+%% flows out of the fiber section into, now gluing observations together into a thing
+%% can then put functions on edges/planes/etc so we have true continouity
 \label{sec:base_data}
@@
 Datasets have a semantic topology such as the values are interpreted as discrete observations or part of a timeseries or map, or nodes in a networks \cite{munznerWhatDataAbstraction2014,geveci2012vtk}.
+%%turn it into a paragraph
+
 As illustrated in fig~\ref{fig:base_space_div}, $K$ is this underlying structure.
@@
 \end{figure}
-The topology $K$ and the fields $F$ are determined by how $E$ is subdivided\cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
+The topology $K$ and the
+
+%%this paragraph isn't working
+$F$ are determined by how $E$ is subdivided\cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
 In figure~\ref{fig:base_space_div} we can divide a rectangular base space such that there is a short fiber and long base space or a long fiber and short base space.
@@
 
-\paragraph{Triangulization}
-\label{sec:triangulization}
-In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
-
-\begin{figure}[H]
-    \label{fig:simplex}
-    \includegraphics{figures/sections/math/simplex.png}
-    \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations}
-\end{figure}
-
-K is a triangularizable topological space; one triangularization scheme is as a set composed of simplices\cite{geometricaSimplicialComplexes}, such as those shown in figure~\ref{fig:simplex}.
-
-\subparagraph{Example}
-\textcolor{blue}{Need to sketch and write this example}\\
-chopping up a torus maybe? talk about how that gets unpacked into triangles and then into vertices
-
-\subsubsection{Fiber Space}
-\label{sec:fiber_data}
-Datasets can be univariate or multivariate, homogenous or heteregenous, and fibers are where we encode this information.
-Each field in the dataset $f_{i}$ contains a range of values that come from the codomain of $f_{i}$.
-The fiber is the cartesian product of the codomains of fields:
-\begin{align}
-F &= codomain(f_{0})\times \ldots \times codomain(f_{n})\\
-F &= F_{0} \times \ldots \times F_{n}
-\end{align}
-
-We formulate the fiber such that it has a mapping between the codomain and the field name, as discussed in Spivak's simplacial description of databases \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010}.
-An important property of data fields are the measurement type, which we generalize to identifying the left monoid action such that the monoid actions $M$
-\begin{equation}
-M = M_{0} \times \ldots\times M_{i}
-\end{equation}
-is the cartesion cross product of monoids applied to each $F_{i}\in F$.
-A monoid \cite{Monoid2021} $M$ is a set that is closed under an associative binary operator $\ast$ and has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
-
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet$ with the properties:
-\begin{description}
-    \item[closure] $\bullet: M\times F \rightarrow X$,
-    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
-    \item[identity] $\text{for all } x\in F, e\in M,  e\bullet x = x$ 
-\end{description}
-
-The Steven's measurement scales are identified via the group actions on $F_{i}$  \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale} they support.
-For example, nominal variables are permutable and interval values are translatable.
-We identify monoid actions, rather than group, to support an extended variaty of scales such as partial order
-
-\subsubsection{Section}
+\subsubsection{Data: Section}
 \label{sec:section_data}
 The sections of the fiber bundles are the instances of the data.
-They are the functions that map from a key on $K$ to a set of values in $F$.
+They are the functions that map from a point $k \in K$ to a set of values in $F$.
 For example, if we have a database, the section is the table that yields a row given an index.
@@
 
+%%% change to b\c locally trivial, can represent 
 such that in a trivial fiber bundle, $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
@@
 \end{equation}
+
+%%%can decompose bundles into Ei such that \tau: E_{i} \rightarrow K, where F_{i} is the fiber of E_{i}
 which we can also decompose such that
@@
 which allows us to have field wise access to the data so long as there is a shared $K$.  
+%% we will use this property later for visual mapping such that we can seperate each variable into its own bundle
 
 \subsubsection{Sheaf and Stalk}
-
+%%% sheaf is how we break up the base space 
 Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(E)$
@@
 
-The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is a type of sheaf that allows for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.
+%% the stalk contains all the local information, typically we only need some of the information, such as the firtst or second direivative.
+in practice we only need the jet which is...in practive will extend the fiber
+The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is a for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.
 
@@
 \textcolor{red}{Moved \& walked through b/c was getting clunky to not have terms yet}\\
+
+% can't make a 1D fiber - can't make a 1D fiber, switch the picture to plane over 0D and 1D and then can talk about disconnected or connected
+
+% break out the mbox/clean up the typesetting
 The fiber bundle model is flexible enough to express some of the many different forms that temperature data can come in.
@@
 Many $\mu_{i}$ are graphical parameters that do not apply to the whole glyph, such as edge thickness in figure~\ref{fig:artist_mark_change}.
+
+%%% rework this sentence
 In these situations, not all  $\rho$ in $\Gamma(H)$ will support these parameters; instead we define an action on the output graphic $Q(\Gamma(V)) \in \Gamma(H)$ since by definition every section $\mu$ will have a corresponding $\rho$.
@@
 
-Lets call the visual encodings $\Gamma(V)=X$ and the graphic $Q(\Gamma(V))=Y$. If $\forall m \in M$ and $\forall \mu, \mu\prime \in X$, 
+Lets call the visual encodings $\Gamma(V)=X$ and the graphic $Q(\Gamma(V))=Y$. If $\forall m \in M$ and $\forall \mu, \mu^\prime \in X$, 
 \begin{equation}
-Q(\mu) = Q(\mu\prime)\implies Q(m\circ\mu) = Q(m\circ\mu\prime)
+Q(\mu) = Q(\mu^\prime)\implies Q(m\circ\mu) = Q(m\circ\mu^\prime)
 \end{equation}
-then a group action on $Y$ can be defined as $m\circ \rho = \rho\prime$ where $\rho\prime=Q(g\circ \mu)$ with $\mu \in Q^{-1}(\rho)$.
+then a group action on $Y$ can be defined as $m\circ \rho = \rho^\prime$ where $\rho^\prime=Q(g\circ \mu)$ with $\mu \in Q^{-1}(\rho)$.
 
@@
 
-the constraint on $Q$ means that the translation $m=\{e, e, e, x+2\}$ applied to $\mu$ such that $\mu\prime=\{0,0,0,3\}$ has an equivalent action on $\rho$ that causes $Q(\mu\prime)$ to be equivalent to the thicker circle in figure~\ref{fig:artist_mark_change}.
+the constraint on $Q$ means that the translation $m=\{e, e, e, x+2\}$ applied to $\mu$ such that $\mu^\prime=\{0,0,0,3\}$ has an equivalent action on $\rho$ that causes $Q(\mu\prime)$ to be equivalent to the thicker circle in figure~\ref{fig:artist_mark_change}.
 
@@
 Natural transformation + composition is partial ordering? Back and forth is equivalent
+
+\subsection{Making the fiber bundle computable}
+\label{sec:triangulization}
+%% we can talk about triangulazation backwards from the mark
+%% To implement this, we 
+In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
+
+\begin{figure}[H]
+    \label{fig:simplex}
+    \includegraphics{figures/sections/math/simplex.png}
+    \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations}
+\end{figure}
+
+K is a triangularizable topological space; one triangularization scheme is as a set composed of simplices\cite{geometricaSimplicialComplexes}, such as those shown in figure~\ref{fig:simplex}.
+
+\subparagraph{Example}
+\textcolor{blue}{Need to sketch and write this example}\\
+chopping up a torus maybe? talk about how that gets unpacked into triangles and then into vertices
 \end{document}

### 2021-01-14 code/matplottoy/data/local.py

@@ -1,0 +1,50 @@
+class SimplexPoint:
+    # whole shape
+    def __init__(self, coord):
+        """one point in triangle or edge or whatever, class is whole simplex
+        """
+        coord = np.array(coord)
+        # Needs to be postive and add to one
+        assert (coord>=0).prod() > 0
+        assert coord.sum() == 1
+        self.coord = coord
+
+class LocalFiberComponent: # F_i - type, spec, U\sigma
+    def __init__(self, fiber_component=None):
+        """
+        fiber_componenet: value in F_i
+        """
+        self.fiber_component = fiber_component
+
+class LocalFiberPoint: # F = F_0 X F
+    def __init__(self, local_fiber_point=[LocalFiberComponent()]):
+        """ return of \tau at one point in K - no base point yet
+        is a point in F
+        """
+        self.local_fiber_point = local_fiber_point
+
+class LocalSectionComponent: #\tau_i on an F_i (column wise indexing)
+    """defining the bundle in terms of the datastructure that is holding it
+    """
+    def __init__(self, funct):
+        self.funct = funct
+    def __call__(self, base_point): 
+        """
+        k in K
+        """
+        return LocalFiberComponent(funct(base_point))
+
+class LocalSection:# \tau on F at base_point
+    def __init__(self, local_section_components=[lambda base_point: None]):
+        """\tau = {\tau_o, \ldots ,\tau_n}
+        """
+        self.local_section_components = local_section_components
+
+    def __call__(self,base_points):
+        """ \tau 
+            construct fiber based on data 
+        """
+        return [(base_point, 
+                LocalFiberPoint([local_section_component(base_point)
+                            for local_section_component in local_section_components]) 
+                                for base_point in base_points]

### 2021-01-14 code/matplottoy/data/simplex.py
draft\sections\math.tex

@@ -45,7 +45,10 @@ class VertexSimplex: #maybe change name to something else
                       'v2': {'type': float,'monoid':'interval',
                              'range': [0,1]},
                       'v3': {'type': str, 'monoid':'nominal', 
-                             'range':['true', 'false']}})
+                             'range':['true', 'false']},
+                      'v4': {'type': float, 'monoid':'interval', 
+                             'range':[2,3]}       
+                        })
 
     def __init__(self, sid = 45, size=1000, max_key=10**10):
         self.section_id = sid
@@ -63,7 +66,8 @@ class VertexSimplex: #maybe change name to something else
         e1 = rng.choice(self.FB.F['v1']['range'])
         e2 = rng.uniform(*self.FB.F['v2']['range'])
         e3 = rng.choice(self.FB.F['v3']['range'])
-        return (k, (e1, e2, e3))
+        e4 = rng.uniform(*self.FB.F['v4']['range'])
+        return (k, (e1, e2, e3, e4))
 
     def view(self, simplex="vertex"):
         """"converts data into atomic column order for get method

### 2021-01-14 code/matplottoy/artists/point.py
thicken v to include scaler parameters

@@ -10,8 +10,9 @@ import matplotlib.path as mpath
 from matplottoy.artists import utils
 
 class Point(mcollections.Collection):
+    # this is the visual fiber 
     required = {'x', 'y'}
-    optional = {'facecolors'} 
+    optional = {'facecolors', 's'} 
     def __init__(self, data, transforms, *args, **kwargs):
         """
         Parameters
@@ -25,10 +26,9 @@ class Point(mcollections.Collection):
         # has a way to provide vertex data  
         assert 'vertex' in data.FB.K['tables']
         # check that you've given the required parameters
-        utils.check_constraints(Point, transforms)
+        utils.check_constraints(Point, transforms.keys())
         utils.validate_transforms(data.FB.F, transforms)
 
-        
         self.data = data
         self.transforms = transforms
 
@@ -36,15 +36,13 @@ class Point(mcollections.Collection):
         view = self.data.view('vertex') #resolve to size
 
         # call tau step
-        visual = dict([(t, tau.convert(view[var]))
-            for (t, (var, tau)) in self.transforms.items()])
+        visual = utils.convert_transforms(view, self.transforms)
            
         # assembles taus to generate idiom
         # dictionary here in place of visual(parameter) function
-        if 's' not in visual:
-            visual['s'] = itertools.repeat(0.05)
-        if 'facecolors' not in visual:
-            visual['facecolors'] = "C0"
+       
+        visual['s'] = itertools.repeat(visual.get('s', 0.05))
+        visual['facecolors'] = visual.get('facecolors', "C0")
         #switch out to a marker 
         self._paths = [mpath.Path.circle(center=(x,y), radius=s)  
                         for (x, y, s) 

### 2021-01-14 code/matplottoy/artists/utils.py

@@ -1,5 +1,6 @@
 import numpy as np
-def check_constraints(artist, transforms):
+
+def check_constraints(artist, transforms_keys):
     """
     required: set
         Encodings required by the artists
@@ -9,27 +10,38 @@ def check_constraints(artist, transforms):
         Set of encodings the datasource is providing
     """
     # check required encodings are there
-    if not (artist.required <= transforms.keys()):
+    if not (artist.required <= transforms_keys):
         raise ValueError(f"Required encodings {artist.required}")
     # check optional encodings 
-    if not ((transforms.keys() - artist.required) 
+    if not ((transforms_keys - artist.required) 
             <= artist.optional):
         raise ValueError(f"Valid optional encodings: {artist.optional}")
 
 def validate_transforms(fibers, transforms):
-    for column, tau in transforms.values():
-        for col in np.atleast_1d(column):
-            if not tau.validate(fibers[col]['monoid']):
-                raise ValueError(f"Invalid transform {tau} for {col}")
+    # this is where could tie into GSOD
+    for param, transform in transforms.items():
+        if np.isscalar(transform):
+            # check that input value is valid for the parameter,
+            assert True
+        else: 
+            column, tau  = transform
+            for col in np.atleast_1d(column):
+                if not tau.validate(fibers[col]['monoid']):
+                    raise ValueError(f"Invalid transform {tau} for {col}")
 
 def convert_transforms(view, transforms, exclude=None):
     exclude = [] if exclude is None else exclude
     visual = {}
-    for (t, (var, tau)) in transforms.items():
-        if t in exclude:
+    for (param, value) in transforms.items():
+        if param in exclude:
             continue
-        if isinstance(var, (str, bytes )):
-            visual[t] = tau.convert(view[var])
+        if np.isscalar(value):
+            visual[param] = value
         else:
-            visual[t] = [tau.convert(view[v]) for v in var]
+            (var, tau) = value
+            if isinstance(var, (str, bytes )):
+                visual[param] = tau.convert(view[var])
+            else:
+                visual[param] = [tau.convert(view[v]) for v in var]
+
     return visual

### 2021-01-14 code/matplottoy/tests/utils.py

@@ -1,3 +1,3 @@
 import numpy as np
 def mnx(d):
-    return np.min(d)-np.std(d) , np.max(d) + np.std(d)
+    return np.min(d) , np.max(d)

### 2021-01-15 draft/sections/math.tex
trying to rework to get a good scatter + gotta encode them all as motivation
changed all the file paths

@@
 \pi: E_n\rightarrow K
-\subsubsection{Connectivity of the Data} 
+\subsubsection{Connectivity: Base Space} 
 %% flows out of the fiber section into, now gluing observations together into a thing
 %% can then put functions on edges/planes/etc so we have true continouity
+
 \label{sec:base_data}
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/sections/math/k_different_types.png}
+    \includegraphics[width=\textwidth]{figures/math/k_different_types.png}
     \label{fig:base_space_types}
@@
     \label{fig:base_space_div}
-    \includegraphics[width=.5\linewidth]{figures/sections/math/k_qspace.png}
+    \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
     %%include second figure where fiber goes otherway
@@
 \subsubsection{Data: Section}
+%%% visual variable encoding in a consistent/atomic way in a complex strcutrue 
 \label{sec:section_data}
@@
     \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/sections/math/temp_1k.png}
+        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
         %% add box around neighboring P and Map
@@
     \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/sections/math/temp_2k.png}
+        \includegraphics[width=\textwidth]{figures/math/temp_2k.png}
         \label{fig:base_example_plane}
@@
     \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/sections/math/temp_2f.png}
+        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
         \label{fig:fiber_example_plane}
@@
     \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/sections/math/temp_3f.png}
+        \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
         \label{fig:fiber_example_cube}
@@
 \begin{figure}[H]
-    \includegraphics[width=1\linewidth]{figures/sections/math/fiberbundle.png}
+    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
     \label{fig:fiber_example_section}
@@
 \begin{figure}[H]
-    \includegraphics[width=.4\linewidth]{figures/sections/math/render.png}
+    \includegraphics[width=.4\linewidth]{figures/math/render.png}
     \caption{}
@@
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/sections/math/path_of_q}
+    \includegraphics[width=\textwidth]{figures/math/path_of_q}
     \label{fig:artist_q}
@@
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/sections/math/diff_type_q.png}
+    \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
     \label{fig:artist_mark_change}
@@
     \label{fig:simplex}
-    \includegraphics{figures/sections/math/simplex.png}
+    \includegraphics{figures/math/simplex.png}
     \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations}

### 2021-01-15 draft/sections/implementation.tex
feedback + local
moving files, start of implementation, fightes with minted

--- 2021-01-07 draft/sections/implementation.tex
@@
 \begin{document}
+
 \section{Matplottoy}
-\begin{figure}
-\includegraphics[width=\linewidth]{figures/sections/math/path_of_q.png}
+
+We build on the existing Matplotlib architecture \cite{hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and rely on Matplotlib for the other graphical elements of the visualization and the rendering.
+We first introduce an implementation of scatter, line, and bar charts because they map to the fundemental marks of point, line, and area.
+We then introduce aggregated bar charts to show how to build more complex graphics.
+
+\subsection{Scatter}
+ A scatter plot \cite{friendlyBriefHistoryData2006a,tukeyExploratoryDataAnalysis1977} is a chart type for plotting discrete values against each other.
+Minimally a scatter plot requires and x or y position, but the other variables in our dataset can be mapped to visual aspects of the scatter graphical mark.
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
+    \label{fig:scatter}
 \end{figure}
-%%% objects: Data, V, H
-%%% functions: \nu, q
 
-We build on top of the existing Matplotlib architecture \cite{hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and rely on Matplotlib for the other graphical elements of the visualization and the rendering.
-For a primitive graph such as a bar, scatter, line, our path is:
+For the scatter plot in figure~\ref{fig:scatter}, we define $Q$ as 
+ \begin{equation}
+    Q(\mu_{xpos}, \mu_{ypos}, \mu_{facecolor}, \mu_{s})
+ \end{equation}
+
+which we implement building on the \texttt{matplotlib.collections} API:
+
 \begin{minted}{python}
+    class Point(mcollections.Collection):
+        # this is the visual fiber 
+        required = {'x', 'y'}
+        optional = {'facecolors', 's'} 
+        def __init__(self, data, transforms, *args, **kwargs):
+            """
+            Parameters
+            ----------
+            data: sections of the fiber bundle
+            transforms
+            """
+            super().__init__(*args, **kwargs)
 
-E = Section()
-V = {'parameter': ('field': transformer)}
+            # check that the data you're trying to transform 
+            # has a way to provide vertex data  
+            assert 'vertex' in data.FB.K['tables']
+            # check that you've given the required parameters
+            utils.check_constraints(Point, transforms.keys())
+            utils.validate_transforms(data.FB.F, transforms)
 
-fig, ax = plt.subplots()
-artist = primative_Q(E, V)
-ax.add_artist(artist)
+            self.data = data
+            self.transforms = transforms
+
+        def draw(self, renderer, *args, **kwargs):
+            view = self.data.view('vertex') #resolve to size
+
+            # call tau step
+            visual = utils.convert_transforms(view, self.transforms)
+
+            # assembles taus to generate idiom
+            # dictionary here in place of visual(parameter) function
+
+            visual['s'] = itertools.repeat(visual.get('s', 0.05))
+            visual['facecolors'] = visual.get('facecolors', "C0")
+            #switch out to a marker 
+            self._paths = [mpath.Path.circle(center=(x,y), radius=s)  
+                            for (x, y, s) 
+                            in zip(visual['x'],visual['y'], visual['s'])]
+
+            self.set_facecolors(visual['facecolors'])
+            super().draw(renderer, *args, **kwargs)
     \end{minted}
 
-\subsection{Data: $\mathcal{E}$}
-We propose a semantic markup of the fiber and connectivity that we will later use to check the constraints of the $\nu$ and $Q$. 
-\begin{minted}{python}
-class FiberBundle:
-    def __init__(self, K, F):
-            """
-            K: {'tables':[]}
-            F: {'field name': (type, monoid action)}
-            """
-
-    def is_section(self, section):
-        """checks if a section is from a given fiber bundle:
-        are values in F, are keys in K"""
-\end{minted}
-We add a monoid field to the schema like structure of the fiber since the monoid actions are the constraints on the $\nu$ and $Q$.
-We implement $K$ using the triangulization scheme described in section~\ref{sec:triangulization}.
-This means we expect data to be provided as tables where the name encodes the connectivity:
-
-\begin{description}
-    \item[vertex] - disconnected points
-    \item[edge] - 1D continuity along the edge
-    \item[face] - 2D continuity along the face
-\end{description}
-Nested continuity is encoded in the fiber. For example, one way to encode a movie is a 1D timeseries of \mathrm{Frame} objects, where each \mathrm{Frame} is a $n\x m$ continuous array.
-
-\subsubsection{Sections}
-As described in section~\ref{sec:section_data}, the data values are encode as the section of the fiber bundle. We implement this as as a wrapper class around a data container object of the form:
-\begin{minted}
-class Section:
-    FB = FiberBundle(K, F)
-
-    def __init__(self, *args):
-        pass
-    def view(self, simplex="vertex"):
-        """"converts data into atomic column order for get method
-        # maybe also pass in the columns?
-        """
-        return {fieldname: data}
-\end{minted}
-
-The view function is the closest analog to $\tau$, as it returns a a data container type object.
-We specify field based top level indexing so that we can pair the fields with transformer $\nu$ objects.
-
-\subsection{Encoding: $\mathcal{V}$}
-We implement $\nu$ as encoder objects with an optional `__init__` method that can be used to specify the range of target visual variables.
-
-\begin{minted}{python}
-class Encoder:
-    def __init__(self, args):
-        pass
-    def validate(self, monoid):
-        pass
-    def convert(self, value):
-        return converted_value
-\end{minted}
-
-The \mathrm{validate} method specifies the monoids for which this a valid $\nu$ and checks against the specification in the fiber.
-The \mathrm{convert} method converts a value from data space to an internalized normal form as described in table~\ref{tab:mpl_visual_variable_fiber}.
-
-\subsection{Graphic: $\mathcal{H}$}
-
-\subsection{Case Study: Iris(Penguins)}
-
 \end{document}

### 2021-01-15 draft/sections/implementation.tex
scatter Q and nu, not sure I need data

@@
 We then introduce aggregated bar charts to show how to build more complex graphics.
+
+Because we do not expect the user to specify every visual parameter, we define a class of required visual variables that are intrinsic to S. %%need to revise for formality but
+What I mean here is that you cannot construct the mark w/o this - position for scatter \& line, values for heatmap.
 
@@
     Q(\mu_{xpos}, \mu_{ypos}, \mu_{facecolor}, \mu_{s})
+    \label{eq:scatter}
  \end{equation}
@@
         def __init__(self, data, transforms, *args, **kwargs):
-            """
-            Parameters
-            ----------
-            data: sections of the fiber bundle
-            transforms
-            """
+            # check that the constraints on Q can be met  
+        def draw(self, renderer, *args, **kwargs):
+            # assemble the glyph
+            super().draw(renderer, *args, **kwargs)
+    \end{minted}
+
+(subject to change).
+Unlike equation~\ref{eq:scatter}, we pass also pass a reference to the data so that getting the data can be fully curried until absolutely necessary.
+The \texttt{*args} and \texttt{**kwargs} are artifacts of using Matplotlib as a base.
+We pass in the $\mu$ as the dictionary "transforms" mostly for readability, which is also why we define the visual fiber as class attributes.
+
+Inside the \texttt{__init__} constructor, is a check if there is a valid mapping between $S$ and $K$.
+Line 7-8 check if the user has passed in the required $\mu$ and whether the $\nu$ functions are valid for the input data.
+
+\begin{minted}[lineno]{python}
+def __init__(self, data, transforms, *args, **kwargs):
     super().__init__(*args, **kwargs)
-
     # check that the data you're trying to transform 
@@
     self.transforms = transforms
+\end{minted}
 
+Lines 9-10 attach the data and transforms to the $Point$ object as a concession to the Matplotlib architecture that seperates drawing from creation.
+In \texttt{draw}, the attributes of the graphic are set to the different sections of the visual fiber bundle.
+
+\begin{minted}[lineno]{python}
 def draw(self, renderer, *args, **kwargs):
@@
 
-            # call tau step
+    # \nu(\tau)
     visual = utils.convert_transforms(view, self.transforms)
@@
     # assembles taus to generate idiom
-            # dictionary here in place of visual(parameter) function
-
     visual['s'] = itertools.repeat(visual.get('s', 0.05))
@@
 \end{minted}
+Line 1 pulls back into the data bundle and returns $\tau = \{\tau_0, \ldots, tau_{n}\}$.
+Line 3 applies the transforms $\nu$ to $\tau$ to build the visual bundle $V$.
+Line 8 sets the abstract segment marks with the visual characteristics.
+Line 10 passes this information to the renderer, acting like a $rho$.
+
+The visual fiber bundle is specified as a dictionary $\{\mbox{parameter}: (\mbox{variable}, \nu), \mbox{parameter}:\mbox{value}\}$ 
+\begin{minted}[lineno]{python}
+
+transforms = {'y': ('v2', position.Identity()),       
+              'x': ('v4', position.Identity()),
+              'facecolors': ('v3', color.Categorical( {'true':'deeppink', 'false':'deepskyblue'})), 
+              's':.01}
+\end{minted}
+
+where the $\nu$ functions are represented as classes so that parameters can be curried.
+For example, \texttt{color.Categorical} is a whole set of $\nu$ functions explicitly tuned to specific categorical mappings.
+A pure functional approach would mean the user would have to write new functions
+
+\begin{minted}[lineno]{python}
+def color_true_deepink_false_deepink(value):
+    return {'true':'deeppink', 'false':'deepskyblue'}[value]
+
+\end{minted}
+
+for every categorical colormapping, which would be somewhat untenable. An alternative approach could be to partially curry:
+
+\begin{minted}[lineno]{python}
+     'facecolors': ('v3', (color.categorical, {'true':'deeppink', 'false':'deepskyblue'}))
+\end{minted}
+
+but that doesn't allow for as easy reuse? (I should probably actually just switch to this form) The \textt{color.categorical} function is implemented as 
+\begin{minted}[lineno]{python}
+class Categorical:
+    def __init__(self, mapping):
+        """goal of init is to store parameters that would otherwise be
+        curried higher level function"""
+        assert(mcolors.is_color_like(color) for color in mapping.values())
+        self._mapping = mapping
+
+    def convert(self, value):
+        values = np.atleast_1d(np.array(value, dtype=object))
+        return [mcolors.to_rgba(self._mapping[v]) for v in values]
+
+    def validate(self, mtype):
+        return mtype in ['nominal']
+\end{minted}
+
+where the \texttt{convert} function converts the input value to the internal normalized form Matplotlib expects.
+The `\texttt{validate}' function checks which monoid actions the transform supports; optimally this method should be replaced by functions that check the properties, for example a \texttt{is_ordinal} method that checks that monotonicity is preserved.
+
+#does the data even matter?
 

### 2021-01-18 draft/sections/implementation.tex
started fleshing out new figures

@@
 \documentclass[../main.tex]{subfiles}
+
 \begin{document}
@@
 
-Because we do not expect the user to specify every visual parameter, we define a class of required visual variables that are intrinsic to S. %%need to revise for formality but
-What I mean here is that you cannot construct the mark w/o this - position for scatter \& line, values for heatmap.
+Because it is impractical for the user to specify every possible visual parameter, we define a required set of visual variables that must be explicitly specified to generate distinguishable marks.
 
@@
 
-which we implement building on the \texttt{matplotlib.collections} API:
+which we implement building on the \mintinline{python}{matplotlib.collections} API:
+%%rewrite that Point is a fully parameterized A  with the transforms dict - callable objects are a way to do tractable currying, can use __call__ same as convert - don't mutate internal state of object
 
-\begin{minted}{python}
+\begin{minted}[linenos]{python}
     class Point(mcollections.Collection):
@@
 Unlike equation~\ref{eq:scatter}, we pass also pass a reference to the data so that getting the data can be fully curried until absolutely necessary.
-The \texttt{*args} and \texttt{**kwargs} are artifacts of using Matplotlib as a base.
+The \mintinline{python}{*args} and \mintinline{python}{**kwargs} are artifacts of using Matplotlib as a base.
 We pass in the $\mu$ as the dictionary "transforms" mostly for readability, which is also why we define the visual fiber as class attributes.
 
-Inside the \texttt{__init__} constructor, is a check if there is a valid mapping between $S$ and $K$.
+Inside the \mintinline{python}{__init__} constructor, is a check if there is a valid mapping between $S$ and $K$.
 Line 7-8 check if the user has passed in the required $\mu$ and whether the $\nu$ functions are valid for the input data.
 
-\begin{minted}[lineno]{python}
+%% semi functional psuedocode - strip down to don't care about broadcasting & simplest form of validation -
+- assume that the $\nu$ are inherently valid - no error checking
+- don't try to exactly encode monoid - is it a key feature you want to encode
+- What is the important bits?
+
+\begin{minted}[linenos]{python}
 def __init__(self, data, transforms, *args, **kwargs):
@@
     # has a way to provide vertex data  
-    assert 'vertex' in data.FB.K['tables']
+    assert 'vertex' in data.FB.K['tables'] #there is no check here
+
     # check that you've given the required parameters
@@
 Lines 9-10 attach the data and transforms to the $Point$ object as a concession to the Matplotlib architecture that seperates drawing from creation.
-In \texttt{draw}, the attributes of the graphic are set to the different sections of the visual fiber bundle.
+In \mintinline{python}{draw}, the attributes of the graphic are set to the different sections of the visual fiber bundle.
 
-\begin{minted}[lineno]{python}
+\begin{minted}[linenos]{python}
 def draw(self, renderer, *args, **kwargs):
-    view = self.data.view('vertex') #resolve to size
+    # use \xi^{-1} because we want to pull forward the data
+    # propogate information of the downselection to do that 
+    # restrict sheaf to subset of data over current view - feedback at draw restriction
+    view = self.data.view('vertex') #resolve to size \xi^{-1}
 
@@
 \end{minted}
+
 Line 1 pulls back into the data bundle and returns $\tau = \{\tau_0, \ldots, tau_{n}\}$.
@@
 The visual fiber bundle is specified as a dictionary $\{\mbox{parameter}: (\mbox{variable}, \nu), \mbox{parameter}:\mbox{value}\}$ 
-\begin{minted}[lineno]{python}
+\begin{minted}[linenos]{python}
 
@@
               'x': ('v4', position.Identity()),
-              'facecolors': ('v3', color.Categorical( {'true':'deeppink', 'false':'deepskyblue'})), 
+              'facecolors': ('v3', color.Categorical(
+                                    {'true':'deeppink', 
+                                     'false':'deepskyblue'})), 
               's':.01}
@@
 where the $\nu$ functions are represented as classes so that parameters can be curried.
-For example, \texttt{color.Categorical} is a whole set of $\nu$ functions explicitly tuned to specific categorical mappings.
+For example, \mintinline{python}{color.Categorical} is a whole set of $\nu$ functions explicitly tuned to specific categorical mappings.
 A pure functional approach would mean the user would have to write new functions
 
-\begin{minted}[lineno]{python}
+\begin{minted}[linenos]{python}
 def color_true_deepink_false_deepink(value):
@@
 
-\begin{minted}[lineno]{python}
+\begin{minted}[linenos]{python}
      'facecolors': ('v3', (color.categorical, {'true':'deeppink', 'false':'deepskyblue'}))
@@
 
-but that doesn't allow for as easy reuse? (I should probably actually just switch to this form) The \textt{color.categorical} function is implemented as 
-\begin{minted}[lineno]{python}
+but that doesn't allow for as easy reuse? (I should probably actually just switch to this form) The \mintinline{python}{color.categorical} function is implemented as
+
+%%interface important, not implementation
+\begin{minted}[linenos]{python}
 class Categorical:
@@
 
-where the \texttt{convert} function converts the input value to the internal normalized form Matplotlib expects.
-The `\texttt{validate}' function checks which monoid actions the transform supports; optimally this method should be replaced by functions that check the properties, for example a \texttt{is_ordinal} method that checks that monotonicity is preserved.
+where the \mintinline{python}{convert} function converts the input value to the internal normalized form Matplotlib expects.
+The \mintinline{python}{validate} function checks which monoid actions the transform supports; optimally this method should be replaced by functions that check the properties, for example a \mintinline{python}{is_ordinal} method that checks that monotonicity is preserved.
 
-#does the data even matter?
+%does the data even matter?
 

### 2021-01-18 draft/sections/math.tex
started fleshing out new figures

@@
 \subsubsection{Sheaf and Stalk}
+\label{sec:sheaf_stalk}
 %%% sheaf is how we break up the base space 
@@
 
-constraint: inputs go to same output means changes to inputs mean same changes to output
+constraint: inputs go to same o
+utput means changes to inputs mean same changes to output
 
@@
 
+\subsubsection{Sample Qs}
+In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and a simple heatmap.
+
+\paragraph{Q: scatter plot}
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/scatter.png}
+    \label{fig:artist_scatter}
+\end{figure}
+
+\begin{equation}
+Q(xpos, ypos)(\alpha_{1}, \alpha_{2}) 
+\end{equation}
+
+Given a defaut color of black, $\rho_{RGB} = (0,0,0)$. The position of this swatch of color can be computed relative to the location on the disc $S_{i}$ as shown in figure~\ref{fig:artist_scatter}:
+\begin{align}
+x &= size\bullet \alpha_1 \bullet \cos(\alpha_2) + xpos\\
+y &= size\bullet \alpha_1 \bullet \sin(\alpha_2) + ypos
+\end{align}
+
+such that $\rho(s) = (x, y, 0, 0, 0)$ where $s$ is the region in $H$.
+
+\paragraph{Q: line plot}
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/line.png}
+    \label{fig:artist_line}
+\end{figure}
+
+The line plot shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet discussed in section~\ref{sec:sheaf_stalk}
+
+\begin{equation}
+    Q(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha_1, \alpha_2)
+\end{equation}
+
+where the magnitude of the thickness is 
+\begin{equation}
+    \lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}
+\end{equation}
+such that the components are 
+\begin{equation}
+    \hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}
+\end{equation}
+
+which yields components of $\rho(s)$:
+\begin{align}
+ x = xpos(\xi(\alpha_1)) &+ \alpha_2\hat(n_1)(\xi(\alpha_1)) \\
+ y = ypos(\xi(\alpha_1)) &+ \alpha_2\hat(n_2)(\xi(\alpha_2)) 
+\end{align}
+
+\paragraph{Q: heatmap}
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/heatmap.png}
+    \label{fig:artist_heatmap}
+\end{figure}
+The heatmap in figure~\ref{fig:artist_heatmap} 
+\begin{equation}
+Q(xpos, ypos, color)
+\end{equation}
+has in the simple case a direct lookup into $K$ to obtain the $\mu = (x,y,c)$ values that are mapped into 
+\begin{align}
+D_{RGB} &= color(\xi(\alpha_1, \alpha_2))
+D_x & = xpos(\xi(\alpha_1, \alpha_2))
+D_y &= ypos(\xi(\alpha_1, \alpha_2))
+\end{align}
+through $\rho$.
+
 \subsubsection{Visual Idioms: Equivalance class of artists}
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
     \label{fig:simplex}
@@
 \subparagraph{Example}
-\textcolor{blue}{Need to sketch and write this example}\\
-chopping up a torus maybe? talk about how that gets unpacked into triangles and then into vertices
+\begin{figure}[h!]
+    \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
+    \label{fig:triangle_torus}
+\end{figure}
+
+Given data that lies on a toroidal space as shown in figure~\ref{fig:triangle_torus}, in our model this structure can be implemented as a simplacial complex of two 2-simplexes.
+We unfold the torus into the two triangles that compose the square in figure~\ref{fig:artist_heatmap} and encode that as K.
+We also constrain the functions on $A$, $A^\prime$, $B$ and $B^\prime$ such that $A$ can be glued to $A^\prime$ and $B$ to $B^\prime$ to reconstruct the torus.
+
 \end{document}

### 2021-01-18 draft/sections/math.tex
triangularization section

@@
 \label{sec:base_data}
-\begin{figure}[H]
+\begin{figure}[h!]
     \includegraphics[width=\textwidth]{figures/math/k_different_types.png}
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
     \label{fig:base_space_div}
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
     \begin{subfigure}{.5\textwidth}
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
     \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
@@
 \subsubsection{{Example}}
-\begin{figure}[H]
+\begin{figure}[h!]
     \includegraphics[width=.4\linewidth]{figures/math/render.png}
@@
 \subsubsection{Assembling Marks}
-\begin{figure}[H]
+\begin{figure}[h!]
     \includegraphics[width=\textwidth]{figures/math/path_of_q}
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
     \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
@@
 \paragraph{Q: scatter plot}
-\begin{figure}[H]
+\begin{figure}[h!]
     \includegraphics[width=\textwidth]{figures/math/scatter.png}
@@
 \paragraph{Q: line plot}
-\begin{figure}[H]
+\begin{figure}[h!]
     \includegraphics[width=\textwidth]{figures/math/line.png}
@@
 \paragraph{Q: heatmap}
-\begin{figure}[H]
+\begin{figure}[h!]
     \includegraphics[width=\textwidth]{figures/math/heatmap.png}
@@
 \label{sec:triangulization}
-%% we can talk about triangulazation backwards from the mark
-%% To implement this, we 
-In our draft implementation of the data as fiber bundle model, we represent $K$ as a simplacial complex.
 
 \begin{figure}[h!]
-    \label{fig:simplex}
     \includegraphics{figures/math/simplex.png}
     \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations}
+    \label{fig:triangle_simplex}
 \end{figure}
 
-K is a triangularizable topological space; one triangularization scheme is as a set composed of simplices\cite{geometricaSimplicialComplexes}, such as those shown in figure~\ref{fig:simplex}.
+One way to build reusable and flexible $\xi$ is to choose a consistent way of representing$K$.
+In our draft implementation of the data as fiber bundle model, we triangularize $K$ using complexes of the simplicies shown in figure~\ref{fig:triangle_simplex} such that $\xi$ consistently yields some combination of vertexes, edges, and faces.
+This gives a common data indexing structure on which to build the rest of $Q$.
 
@@
 
-Given data that lies on a toroidal space as shown in figure~\ref{fig:triangle_torus}, in our model this structure can be implemented as a simplacial complex of two 2-simplexes.
+Given data that lies on the toroidal space shown in figure~\ref{fig:triangle_torus}, the torus $K$ can be implemented as a simplacial complex of two 2-simplexes.
 We unfold the torus into the two triangles that compose the square in figure~\ref{fig:artist_heatmap} and encode that as K.

### 2021-01-18 draft/sections/math.tex
words for visual idioms, will come back to it later

@@
 
-\subsubsection{Visual Idioms: Equivalance class of artists}
-Given $O(E)$ of the same type that output to the same type of graphic $O(H)$, the
-
-Natural transformation + composition is partial ordering? Back and forth is equivalent
-
 \subsection{Making the fiber bundle computable}
@@
 
-One way to build reusable and flexible $\xi$ is to choose a consistent way of representing$K$.
+One way to build flexible $\xi$ is to choose a consistent way of representing$K$.
 In our draft implementation of the data as fiber bundle model, we triangularize $K$ using complexes of the simplicies shown in figure~\ref{fig:triangle_simplex} such that $\xi$ consistently yields some combination of vertexes, edges, and faces.
-This gives a common data indexing structure on which to build the rest of $Q$.
+This gives a common data indexing structure on which to build components that could potentially be reused across $Q$.
 
@@
 
+\subsubsection{Visual Idioms: Equivalance class of artists}
+As formulated above, every artist function $A$ has fixed $\nu$ and generates a distinct graphic $\rho$.
+It is unfeasible to implement $A$ for every single graphic; instead we implement the equivalence class of artists $\{A \in A^\prime: A_{1} \equiv A_{2}\}$ which is $Q:\Gamma(V) \rightarrow \Gamma(H)$.
+
 \end{document}

### 2021-01-18 draft/sections/math.tex
roadmap

@@
 
-%% convention is base space \times fiber bundle
-%% change B to K
 \section{Notation \& Definitions}
@@
 
-We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the variables in a dataset from how the values in those fields are connected to each other.
+We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the variables in a dataset from how the values  are connected to each other.
 The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}.
 
-%% we will define data, visual, and graphic spaces in terms of fiberbundles and then describe equivariant maps from E to V to H
+In this section, we first describe the fiber bundle spaces of data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}).
+We then discuss the equivariant maps between data and visual characteristics (~\ref{sec:artist_nu}) and visual characteristics and graphics (~\ref{sec:artist_q}) that make up the artist.
 
 \subsection{Data Space}
+\label{sec:space_data}
 % introduce what this is gonna happen in this section - we will talk about, fiber and connectivity 
@@
 \subsubsection{Variable Field in F }
-\label{sec:fiber_data}
+\label{sec:data}
  %%%first sentence should be a paragraph
@@
 \subsection{Prerender Space $H$}
-\label{sec:display}
+\label{sec:graphic}
 We define a graphic space $H$ such that we do not have to assume the physical output space of the renderer.
@@
 \subsubsection{Visual Channels}
+\label{sec:artist_nu}
 $\nu: E \rightarrow V$ is an equivariant map such that there is a homomorphism from left monoid actions on $E_{i}$ to left monoid actions on $V_{i}$ where $i$ identifies a field in the fiber.
@@
 \subsubsection{Assembling Marks}
+\label{sec:artist_q}
 \begin{figure}[h!]
@@
 It is unfeasible to implement $A$ for every single graphic; instead we implement the equivalence class of artists $\{A \in A^\prime: A_{1} \equiv A_{2}\}$ which is $Q:\Gamma(V) \rightarrow \Gamma(H)$.
-
 \end{document}

### 2021-01-18 draft/sections/math.tex
dropped fb down to base space

@@
 
-In this section, we first describe the fiber bundle spaces of data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}).
+In this section, we describe the fiber bundle spaces of data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}).
 We then discuss the equivariant maps between data and visual characteristics (~\ref{sec:artist_nu}) and visual characteristics and graphics (~\ref{sec:artist_q}) that make up the artist.

### 2021-01-18 draft/sections/intro.tex
q
raided my second exam for data + rv intro to lead into math data

@@
 
+set up: dubois
+
+theorists:
+bertin, munzner, mackinlay
+
+talk about:
+matplotlib arch paper, excel/matlab arch, 
+vtk \& ggplot (compare/contrast, we're blending these things)
+
+\subsection{Terminology}
+\begin{figure}[h!]
+%%remake properly at some point
+\includegraphics[width=\textwidth]{figures/intro/flowchart.png} 
+\caption{To go from a dataset to a visualization, the data is subset based on a set of constraints (the invariant).
+The resulting subset becomes the components that are visualized, but the choice of visualization is dependent on the type and structure of the component variables.}
+\label{fig:flowchart}
+\end{figure}
+
+Given a dataset, we need to decide what subset of the data to visualize.
+Bertin describes the set of constraints used to subset the data as the \textit{invariant}.
+Formally, the \textit{invariant} is the set of shared characteristics of the data being visualized.
+When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertin_semiology_2011}.
+In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables.
+As shown in figure~\ref{fig:flowchart}, the final step in creating a visualization is choosing how to encode the
+components using retinal (visual) variables.
+
+\subsubsection{Retinal (Visual) Variables}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
+\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the \textit{components} of a visualization.
+This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{_information_????} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} \cite{krygier_making_2005}}
+\label{fig:retinal_variables}
+\end{figure}
+
+Figure~\ref{fig:retinal_variables} illustrates common guidelines for encoding \textit{components}, derived from what Bertin terms a retinal variable and most other visualization theorists call visual variables \cite{bertin_semiology_2011,krygier_making_2005,chambers_graphical_1983,wilkinson_grammar_2005,munzner_visualization_2014}.
+The columns of figure~\ref{fig:retinal_variables} correspond to the type of observation: discrete points, continuous events (e.g.
+a timeseries), two dimensional continuous events (e.g.
+a vector map).
+The rows of figure~\ref{fig:retinal_variables} describe ways to visualize variations in the \textit{components}; generally, quantitative components are represented by retinal variables that change quantitatively and categorical components are represented by retinal variables that vary qualitatively.
+In figure~\ref{fig:iris_scatter}, the hue of the marker is used to encode differentiation in species and the position of the marker is used to show variation in petal length and sepal length.
+The retinal variables suggest that any single visualization is limited to encoding at most about 8 or 9 components.
+Retinal variables provide guidelines for encoding \textit{components}, but the choice of graph is based on the type and structure of the data.
+
+\subsubsection{Data Type and Structure}
+
+\begin{figure}[h!]
+ \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
+\caption{Keys are unique lookup values used to find individual observations in the dataset.
+Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube.
+Image modified from a diagram from Munzner's website \cite{_visualization_????}}
+\label{fig:munzner_datatypes}
+\end{figure}
+
+As shown in figure~\ref{fig:flowchart}, there are multiple ways to translate data into pictures.
+A map is always an option, except when the observations do not have associated coordinates in a physical plane.
+Tamara Munzner provides a way to distinguish between these datasets using $\{key, value\}$ designations \cite{munzner_visualization_2014}.
+Munzner defines \textit{values} as measurements of interest in the dataset, analogous to dependent variables in statistics.
+She defines \textit{keys} as indexes that can be used to look up values, analogous to independent variables in statistics and dimensions in computer science.
+Figure~\ref{fig:munzner_datatypes} illustrates how these keys are used to look up variables in a dataset:
+\begin{itemize}
+	\item map: keys are the coordinates of the points
+	\item table: row index, database primary key
+	\item data cube: row, column, etc. (.e.g. $i,j,k$) index
+\end{itemize}
+
+Expanding on Munzner's key and value semantics, in many datasets the keys are discrete variables like time or geophysical locations sampled from a continuous curve, surface, or field.
+While these observations are discrete samples from the continuous space, often the continuous (functional) characteristic\cite{ramsay_functional_2006,muller_functional_2006} of the observational space is also of interest.
+Besides quantitative discrete, quantitative continuous, or categorical measurement type considerations, the choice of visualization is also influenced by the measurement being on an interval, ratio, nominal, or categorical scale.
+
 \end{document}

### 2021-01-18 draft/sections/math.tex
raided my second exam for data + rv intro to lead into math data

@@
 
-In this section, we describe the fiber bundle spaces of data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}).
-We then discuss the equivariant maps between data and visual characteristics (~\ref{sec:artist_nu}) and visual characteristics and graphics (~\ref{sec:artist_q}) that make up the artist.
+We first describe the fiber bundle spaces of data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}).
+We then discuss the equivariant maps between data and visual characteristics (\ref{sec:artist_nu}) and visual characteristics and graphics (\ref{sec:artist_q}) that make up the artist.
 
 \subsection{Data Space}
-\label{sec:space_data}
+\label{sec:data}
+Generally when we discuss data, we are interested in two aspects: the variables in the dataset
+
 % introduce what this is gonna happen in this section - we will talk about, fiber and connectivity 
@@
 
-%%% might get moved into basespace discussion:
-
-We model data as a fiber bundle ($E$, $K$, $\pi$, $F$)
-\begin{equation}
-    \label{eq:fiber_bundle}
-    \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
-    \end{tikzcd}
-\end{equation}
-
-with topological total space $E$, base space $K$, fiber space $F$, and the map from total space to base space $\pi: E \rightarrow K$.
-Maps from $K$ to $E$ are called sections and select specific points in $K$.
-The global space of sections in $E$ is $\Gamma(E)$.
-
 \subsubsection{Variable Field in F }
-\label{sec:data}
+\label{sec:data_fiber}
  %%%first sentence should be a paragraph
@@
 \subsubsection{Connectivity: Base Space} 
+\label{sec:data_base}
 %% flows out of the fiber section into, now gluing observations together into a thing
 %% can then put functions on edges/planes/etc so we have true continouity
+We model data as a fiber bundle ($E$, $K$, $\pi$, $F$)
+\begin{equation}
+    \label{eq:fiber_bundle}
+    \begin{tikzcd}
+        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
+    \end{tikzcd}
+\end{equation}
 
-\label{sec:base_data}
+with topological total space $E$, base space $K$, fiber space $F$, and the map from total space to base space $\pi: E \rightarrow K$.
+Maps from $K$ to $E$ are called sections and select specific points in $K$.
+The global space of sections in $E$ is $\Gamma(E)$.
+
 \begin{figure}[h!]
@@
 \subsubsection{Data: Section}
-%%% visual variable encoding in a consistent/atomic way in a complex strcutrue 
+%%% visual variable encoding in a consistent/atomic way in a complex structure 
 \label{sec:section_data}
@@
 It is unfeasible to implement $A$ for every single graphic; instead we implement the equivalence class of artists $\{A \in A^\prime: A_{1} \equiv A_{2}\}$ which is $Q:\Gamma(V) \rightarrow \Gamma(H)$.
+
+%% Well maybe.
+I think this is where required components comes in.
+There is some minimal V where if you didn't have those variables you wouldn't call it a scatter plot ....
+something like xpos,ypos with K just a descrete set.
+%% so V probably can be any V that contains that sub-bundle
+%% so if bubble plot is not the same as scatter plot, and something where the colors change based on data is not scatter plot then it is the equivalent class of something with that minimal bundle.
+%% But then if you want to include ANYTHING that also maps with X-Y ... then you hit "direct limits" where you limit over all possivle V_big containing V_minimal
+%% You still get a map on some A_minimal -> A_big
+%% If you know how to visualize with more variables .. you can always fix them.
 \end{document}

### 2021-01-18 draft/sections/math.tex
new pdf

@@
 
-We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the variables in a dataset from how the values  are connected to each other.
+We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the variables in a dataset from how the values  are connected to each other \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
 The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}.
@@
 \label{sec:data}
-Generally when we discuss data, we are interested in two aspects: the variables in the dataset
+Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a global daily record of weather station  such as rainfall and temperature.
+Depending on which aspects of the data we want to visualize, we can encode it as a timeseries or map or both or even represent the measurements as completely disconnected.
 
-% introduce what this is gonna happen in this section - we will talk about, fiber and connectivity 
-% motivate w/ data/ example - have weather station. %tidydata 
-% we will introduce the fiber bundle and its components base and total space to explain the fiber
-
-As proposed by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}
+Using a fiber bundle allows us to encode the properties of the heterogenous variables (\ref{sec:data_fiber}), the connectivity of the measurements (\ref{sec:data_base}), and functions between these spaces that yield the measurements (\ref{sec:data_section}).
 
@@
 \subsubsection{Data: Section}
+\label{sec:data_section}
 %%% visual variable encoding in a consistent/atomic way in a complex structure 
@@
 \subsubsection{Sheaf and Stalk}
-\label{sec:sheaf_stalk}
+\label{sec:data_sheaf_stalk}
 %%% sheaf is how we break up the base space 

### 2021-01-19 draft/sections/intro.tex
new pdf
removed minted, fleshed out fiber bundles

@@
 
-\subsection{ What is a viz}?
-Acquired codes of meaning
+\subsection{ What are the the basic requirements of a visualization library?}
+Acquired codes of meaning, dubois
 
@@
 \subsection{Terminology}
+\textcolor{blue}{Is currently C\&P'ed from my second exam so needs some messaging, but }
 \begin{figure}[h!]
@@
 Formally, the \textit{invariant} is the set of shared characteristics of the data being visualized.
-When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertin_semiology_2011}.
-In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables.
+When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertinSemiologyGraphicsDiagrams2011a}.
+%%rework this statement for french labor data
+%%In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables.
 As shown in figure~\ref{fig:flowchart}, the final step in creating a visualization is choosing how to encode the
@@
 \caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the \textit{components} of a visualization.
-This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{_information_????} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} \cite{krygier_making_2005}}
+This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS}
+\cite{krygierMakingMapsVisual2005}}
 \label{fig:retinal_variables}
@@
 \subsubsection{Data Type and Structure}
-
+\label{sec:intro_data_structure}
 \begin{figure}[h!]
@@
 Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube.
-Image modified from a diagram from Munzner's website \cite{_visualization_????}}
+Image modified from a diagram from Munzner's \textit{Visualization Analysis and Design} \cite{munznerVisualizationAnalysisDesign2014}}
 \label{fig:munzner_datatypes}
@@
 A map is always an option, except when the observations do not have associated coordinates in a physical plane.
-Tamara Munzner provides a way to distinguish between these datasets using $\{key, value\}$ designations \cite{munzner_visualization_2014}.
+Tamara Munzner provides a way to distinguish between these datasets using $\{key, value\}$ designations \cite{munznerVisualizationAnalysisDesign2014}.
 Munzner defines \textit{values} as measurements of interest in the dataset, analogous to dependent variables in statistics.
@@
 Expanding on Munzner's key and value semantics, in many datasets the keys are discrete variables like time or geophysical locations sampled from a continuous curve, surface, or field.
-While these observations are discrete samples from the continuous space, often the continuous (functional) characteristic\cite{ramsay_functional_2006,muller_functional_2006} of the observational space is also of interest.
+While these observations are discrete samples from the continuous space, often the continuous (functional) characteristic\cite{ramsayFunctionalDataAnalysis2006a,mullerFunctionalVarianceProcesses2006a} of the observational space is also of interest.
 Besides quantitative discrete, quantitative continuous, or categorical measurement type considerations, the choice of visualization is also influenced by the measurement being on an interval, ratio, nominal, or categorical scale.

### 2021-01-19 draft/sections/math.tex
removed minted, fleshed out fiber bundles

@@
 %% give the why of the math after every mention of why 
-
 In this section we introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.
@@
 \label{sec:data}
-Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a global daily record of weather station  such as rainfall and temperature.
-Depending on which aspects of the data we want to visualize, we can encode it as a timeseries or map or both or even represent the measurements as completely disconnected.
-
-Using a fiber bundle allows us to encode the properties of the heterogenous variables (\ref{sec:data_fiber}), the connectivity of the measurements (\ref{sec:data_base}), and functions between these spaces that yield the measurements (\ref{sec:data_section}).
+Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations.
+Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
+Using a fiber bundle allows us to express the properties of the variables(\ref{sec:data_fiber}), the connectivity of the observations (\ref{sec:data_base}), and tables of records (\ref{sec:data_section}).
 
@@
 \label{sec:data_fiber}
- %%%first sentence should be a paragraph
-%%% butler's paper, no notion of distance on K, can only map F to the visual fibers, so need an X to map to Xpos
-% variable, measurement, pull in tidy here, 
-% % motivate by example, temperature \ time 
-%% F_0 = {temp_0, temp_n}, 
-%% then adding in a second set of variables time: 
-%% F_1 = {time_0, time_1}
-% break out into F first as set of values, then F=F_{0}\times F_{1} as component
+The fiber is a topological space containing the set of all observations that could possible be in the dataset, even values that do not actually exist in the dataset.
+For example, lets construct a fiber bundle that is just the temperature column from GHCN.
+Then the fiber $F$ is the range of physically possible temperature values $\text{Temp} \subset \mathbb{R}$.
+Lets rename this
+\begin{equation}
+F_0 = \{\text{Temp}\}
+\end{equation}
+so that more variables can be added to the fiber. For example, lets add the time from GHCN $F_1 =\text{Time}$
+\begin{equation}
+F =  \{\text{Temp}\} \times \text{Time}
+\end{equation}
 
-The fiber is the cartesian product of the variable spaces:
+The elements in a given fiber set do not need to be scalers; for example the location can be encoded as $F_3=\text{Loc}$ where $(\text{lat},\text{lon{}) \in \text{Loc}$.
+We add the location to the fiber by adding the set to the cartesian product
+\begin{equation}
+    F =  \{\text{Temp}\} \times \text{Time} \times \text{Loc}
+\end{equation}
+
+We also add the variable name and measurement scale $M$ monoid to the fiber $F_i$ such that 
 \begin{align}
-F &= F_{0} \times \ldots \times F_{n}
+    $F_0$ &= (\mathrm{temperature}, \text{Temp}, $M_0$)\\
+    $F_1$ &= (\mathrm{time}, \text{Time}, $M_1$)\\
+    $F_2$ &= (\mathrm{location}, \text{Loc}, $M_2$)
 \end{align}
 
-We formulate the fiber such that it has a mapping between the somethine thing and the field name, as discussed in Spivak's simplacial description of databases \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010}.
-
-Fibers do not need to be 1D.
+as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
+\begin{equation}
+F= F_{0} \times \ldots \times F_{n}
+\end{equation}
+while also retaining the individual sets $F_{i}$ that we will later apply visual variables to.
 
 \subsubsection{Measurement Type}
+\label{sec:data_monoid}
 %% maybe move somewhere else w/ some more motivation
@@
 
-Datasets have a semantic topology such as the values are interpreted as discrete observations or part of a timeseries or map, or nodes in a networks \cite{munznerWhatDataAbstraction2014,geveci2012vtk}.
-%%turn it into a paragraph
-
-As illustrated in fig~\ref{fig:base_space_div}, $K$ is this underlying structure.
-$K$ does not directly know the values; instead it is the sections that define the lookup between keys $k \in K$ and the corresponding values in $E$.
+As mentioned in section~\ref{sec:intro_data_structure}, datasets have a semantic topology wherein the values are interpreted as discrete observations or part of a timeseries or map, or nodes in a networks \cite{munznerWhatDataAbstraction2014,geveci2012vtk}.
+one way of encoding this underlying structure are the $K$s illustrated in figure~\ref{fig:base_space_div}, $K$ does not directly know the values; instead it is the sections that define the lookup between keys $k \in K$ and the corresponding values in $E$.
 
@@
 \paragraph{Q: scatter plot}
+\\
 \begin{figure}[h!]
@@
 \paragraph{Q: line plot}
+\\
 \begin{figure}[h!]
@@
 The line plot shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet discussed in section~\ref{sec:sheaf_stalk}
+(needs the normal to push up/down against the normal)
 
@@
 
+%% only alpha1 exists on the data side
 which yields components of $\rho(s)$:
@@
 \paragraph{Q: heatmap}
+\\
 \begin{figure}[h!]
@@
 has in the simple case a direct lookup into $K$ to obtain the $\mu = (x,y,c)$ values that are mapped into 
+(arrays are upside down, x is y, is x) - $\xi$ is about translating indices from data to visual alignment
 \begin{align}
@@
 
-One way to build flexible $\xi$ is to choose a consistent way of representing$K$.
+One way to build flexible $\xi$ is to choose a consistent way of representing $K$.
 In our draft implementation of the data as fiber bundle model, we triangularize $K$ using complexes of the simplicies shown in figure~\ref{fig:triangle_simplex} such that $\xi$ consistently yields some combination of vertexes, edges, and faces.

### 2021-01-19 draft/sections/math.tex
next section

@@
 \begin{align}
-    $F_0$ &= (\mathrm{temperature}, \text{Temp}, $M_0$)\\
-    $F_1$ &= (\mathrm{time}, \text{Time}, $M_1$)\\
-    $F_2$ &= (\mathrm{location}, \text{Loc}, $M_2$)
+    F_0 &= (\mathrm{temperature}, \text{Temp}, M_0)\\
+    F_1 &= (\mathrm{time}, \text{Time}, M_1)\\
+    F_2 &= (\mathrm{location}, \text{Loc}, M_2)
 \end{align}
-
+ or more generally (name, set, monoid)
+ \begin{equation}
+    F_i = (\mathrm{name}, \{x_{i}, \ldots, \x_{n}\}, M_{i})
+ \end{equation}
 as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
@@
 \end{equation}
-while also retaining the individual sets $F_{i}$ that we will later apply visual variables to.
+while also retaining the individual sets $F_{i}$ that we will transform into visual characteristics of the graphic.
 

### 2021-01-19 draft/sections/next.tex
next section

\documentclass[../main.tex]{subfiles}

\begin{document}

%% proposed work:
%% transition functions, how it'd work on bundles
%% compositing + algebra
%% interactivity
%% implementation

\end{document}

### 2021-01-19 draft/sections/math.tex
fleshed out fiber further, unpacked into set notation

@@
 For example, lets construct a fiber bundle that is just the temperature column from GHCN.
-Then the fiber $F$ is the range of physically possible temperature values $\text{Temp} \subset \mathbb{R}$.
+Then the fiber $F$ is the range of physically possible temperature values $\{tp_0, \ldots, tp_n\} \subset \mathbb{R}$.
 Lets rename this
 \begin{equation}
-F_0 = \{\text{Temp}\}
+F_0 = \{tp_0, \ldots, tp_n\} 
 \end{equation}
-so that more variables can be added to the fiber. For example, lets add the time from GHCN $F_1 =\text{Time}$
+so that more variables can be added to the fiber. Then lets add the time from GHCN $F_1 =\{t_0, \ldots, t_n\}$
 \begin{equation}
-F =  \{\text{Temp}\} \times \text{Time}
+F =  \{tp_0, \ldots, tp_n\}  \times \{ti_0, \ldots, ti_n\} 
 \end{equation}
 
-The elements in a given fiber set do not need to be scalers; for example the location can be encoded as $F_3=\text{Loc}$ where $(\text{lat},\text{lon{}) \in \text{Loc}$.
-We add the location to the fiber by adding the set to the cartesian product
+The elements in a given fiber set do not need to be scalers; for example the location can be encoded as the set of tuples $(\textit{lat},\textit{lon})$.  We add the location to the fiber 
 \begin{equation}
-    F =  \{\text{Temp}\} \times \text{Time} \times \text{Loc}
+    F = \{tp_0, \ldots, tp_n\}  \times \{ti_0, \ldots, ti_n\} \times 
+    \{(la_0, ln_0),\ldots, (la_j, ln_k), \ldots (la_n,ln_n)\}
 \end{equation}
-
-We also add the variable name and measurement scale $M$ monoid to the fiber $F_i$ such that 
+such that one record $r$ in the fiber would have the form $(tp, t, (la,ln))$. We also add the variable name and measurement scale $M$ monoid to the fiber $F_i$ such that 
 \begin{align}
-    F_0 &= (\mathrm{temperature}, \text{Temp}, M_0)\\
-    F_1 &= (\mathrm{time}, \text{Time}, M_1)\\
-    F_2 &= (\mathrm{location}, \text{Loc}, M_2)
+    F_0 &= (\textit{temperature}, \{tp_0, \ldots, tp_n\}, M_0)\\
+    F_1 &= (\textit{time}, \{ti_0, \ldots, ti_n\}, M_1)\\
+    F_2 &= (\textit{location},   \{(la_0, ln_0),\ldots, (la_j, ln_k), \ldots (la_n,ln_n)\}, M_2)
 \end{align}
@@
  \begin{equation}
-    F_i = (\mathrm{name}, \{x_{i}, \ldots, \x_{n}\}, M_{i})
+    F_i = (\textit{name}, \{x_{i}, \ldots, x_{n}\}, M_{i})
  \end{equation}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \includegraphics[width=\textwidth]{figures/math/k_different_types.png}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \label{fig:base_space_div}
@@
 
-\subsubsection{Example: Temperature}
+\subsubsection{Case Study: Temperature}
 \textcolor{red}{Moved \& walked through b/c was getting clunky to not have terms yet}\\
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \begin{subfigure}{.5\textwidth}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
@@
 \subsubsection{{Example}}
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \includegraphics[width=.4\linewidth]{figures/math/render.png}
@@
 \begin{equation}
-    \label{eq:artist}
+    \label{eq:artist_daigram}
     \begin{tikzcd}
@@
 \label{sec:artist_q}
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \includegraphics[width=\textwidth]{figures/math/path_of_q}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
@@
 
-\paragraph{Q: scatter plot}
-\\
-\begin{figure}[h!]
+\paragraph{Q: scatter plot}\mbox{} \\
+\label{sec:artist_example_scatter}
+\begin{figure}[ht!]
     \includegraphics[width=\textwidth]{figures/math/scatter.png}
@@
 
-\paragraph{Q: line plot}
-\\
-\begin{figure}[h!]
+\paragraph{Q: line plot}\mbox{} \\
+\label{sec:artist_example_line}
+\begin{figure}[ht!]
     \includegraphics[width=\textwidth]{figures/math/line.png}
@@
 
-\paragraph{Q: heatmap}
-\\
-\begin{figure}[h!]
+\paragraph{Q: heatmap}\mbox{} \\
+\label{sec:artist_example_heatmap}
+\begin{figure}[ht!]
     \includegraphics[width=\textwidth]{figures/math/heatmap.png}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \includegraphics{figures/math/simplex.png}
@@
 \subparagraph{Example}
-\begin{figure}[h!]
+\begin{figure}[ht!]
     \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}

### 2021-01-19 draft/sections/math.tex
more on fibers + monoids

@@
 
-\subsubsection{Variable Field in F }
+\subsubsection{Variables: Fiber Space $F$}
 \label{sec:data_fiber}
 The fiber is a topological space containing the set of all observations that could possible be in the dataset, even values that do not actually exist in the dataset.
-For example, lets construct a fiber bundle that is just the temperature column from GHCN.
-Then the fiber $F$ is the range of physically possible temperature values $\{tp_0, \ldots, tp_n\} \subset \mathbb{R}$.
+For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $\{tp_0, \ldots, tp_n\} \subset \mathbb{R}$.
 Lets rename this
@@
 \end{equation}
-such that one record $r$ in the fiber would have the form $(tp, t, (la,ln))$. We also add the variable name and measurement scale $M$ monoid to the fiber $F_i$ such that 
+such that one record $r$ in the fiber would have the form $(tp, t, (la,ln))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
 \begin{align}
@@
 \end{equation}
-while also retaining the individual sets $F_{i}$ that we will transform into visual characteristics of the graphic.
+and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.
 
-\subsubsection{Measurement Type}
+\subsubsection{Measurement Scale: Monoid Actions}
 \label{sec:data_monoid}
+The monoid actions $M_i$ is the set of transformations on $F_i$ that we expect to be preserved in the visual transform. For example, the temperature $F_0$ with units $\degree C$ 
+Steven's measurement scales define group actions \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale}
+
 %% maybe move somewhere else w/ some more motivation
@@
 \pi: E_n\rightarrow K
-\subsubsection{Connectivity: Base Space} 
+\subsubsection{Connectivity: Base Space $K$} 
 \label{sec:data_base}
@@
 
-\subsubsection{Data: Section}
+\subsubsection{Values: Sections $\tau$}
 \label{sec:data_section}
@@
 
-\subsubsection{Case Study: Temperature}
+\subsubsection{Case Study: GHCN}
 \textcolor{red}{Moved \& walked through b/c was getting clunky to not have terms yet}\\

### 2021-01-19 draft/sections/math.tex
monoid section rewrite

@@
 \begin{equation}
-F= F_{0} \times \ldots \times F_{n}
+F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
 \end{equation}
@@
 
-\subsubsection{Measurement Scale: Monoid Actions}
+\subsubsection{Measurement Scales: Monoid Actions}
 \label{sec:data_monoid}
-The monoid actions $M_i$ is the set of transformations on $F_i$ that we expect to be preserved in the visual transform. For example, the temperature $F_0$ with units $\degree C$ 
-Steven's measurement scales define group actions \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale}
+A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or rotating a shape; this model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
+One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
 
-%% maybe move somewhere else w/ some more motivation
-%% do this section in terms of $M0, F_0$
-An important property of data fields are the measurement type, which we generalize to identifying the left monoid action such that the monoid actions $M$ 
-\begin{equation}
-M = M_{0} \times \ldots\times M_{i}
-\end{equation}
-is the cartesion cross product of monoids applied to each $F_{i}\in F$.
+We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$.
 A monoid \cite{Monoid2021} $M$ is a set with an associative binary operator $\ast:M \times M\rightarrow M$.
@@
 A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
-\begin{description}
-    \item[associativity] $\text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x$
-    \item[identity] $\text{for all } x\in F, e\in M,  e\bullet x = x$ 
-\end{description}
+\begin{align*}
+    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x\\
+    \textbf{identity}\;& \text{for all } x\in F, e\in M,  e\bullet x = x 
+\end{align*}
 
-The Steven's measurement scales are identified via the group actions on $F_{i}$  \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale} they support.
-For example, nominal variables are permutable and interval values are translatable.
-We identify monoid actions, rather than group, to support an extended variety of scales such as partial order.
+For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $F_0$ are the set of all transformations $tp_i + c$ such that 
+\begin{equation}
+F_0 = (\textit{temperature}, \{tp_0, \ldots, tp_n\}, \{tp_0 + c_0, \ldots, tp_0 + c_n, \ldots, tp_n + c_0, \ldots, tp_n + c_n\} ) 
+\end{equation}
 
-%%close out the fiber bundle section with the cartesian products, which gives us heterogenous data , univariate to multivariate.
-We can decompose the fiber bundle to components bundles \pi:E_1\rightarrow K, ....
-\pi: E_n\rightarrow K
+As with the fiber $F$, the total monoid space $M$ is the cartesian product
+\begin{equation}
+M = M_{0} \times \ldots \times M_{i}\times \ldots \times\ldots M_{n}
+\end{equation}
+of the monoids $M_{i}$ on each $F_{i}\in F$.
+Because each $F_i, M_i$ can be independent of the other fibers, we can formally express the constraints on heterogenous multivariate datasets in the same manner as univariate datasets.
+
 \subsubsection{Connectivity: Base Space $K$} 
 \label{sec:data_base}
+%We can decompose the fiber bundle to components bundles \pi:E_1\rightarrow K, .... \pi: E_n\rightarrow K
 %% flows out of the fiber section into, now gluing observations together into a thing

### 2021-01-19 draft/sections/math.tex
finished out k

@@
 Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
-Using a fiber bundle allows us to express the properties of the variables(\ref{sec:data_fiber}), the connectivity of the observations (\ref{sec:data_base}), and tables of records (\ref{sec:data_section}).
+Using a fiber bundle allows us to express the properties of the variables(\ref{sec:data_fiber}), the connectivity of the records (\ref{sec:data_base}), and tables of records (\ref{sec:data_section}).
 
@@
 \label{sec:data_fiber}
-The fiber is a topological space containing the set of all observations that could possible be in the dataset, even values that do not actually exist in the dataset.
+The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset.
 For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $\{tp_0, \ldots, tp_n\} \subset \mathbb{R}$.
-Lets rename this
+Lets rename this $F$ to a component
 \begin{equation}
@@
 \label{sec:data_base}
-%We can decompose the fiber bundle to components bundles \pi:E_1\rightarrow K, .... \pi: E_n\rightarrow K
-%% flows out of the fiber section into, now gluing observations together into a thing
-%% can then put functions on edges/planes/etc so we have true continouity
-We model data as a fiber bundle ($E$, $K$, $\pi$, $F$)
+The fiber bundle is the gluing map $\pi$ between the fiber space included in the total space $E$ and the base space $K$
 \begin{equation}
@@
 
-with topological total space $E$, base space $K$, fiber space $F$, and the map from total space to base space $\pi: E \rightarrow K$.
-Maps from $K$ to $E$ are called sections and select specific points in $K$.
-The global space of sections in $E$ is $\Gamma(E)$.
+The base space $K$ provides a way to encode how the records are connected to each other; for example the GHCN data can be expressed as discrete observations, many timeseries, maps, a network of stations, etc.
 
@@
 
-As mentioned in section~\ref{sec:intro_data_structure}, datasets have a semantic topology wherein the values are interpreted as discrete observations or part of a timeseries or map, or nodes in a networks \cite{munznerWhatDataAbstraction2014,geveci2012vtk}.
-one way of encoding this underlying structure are the $K$s illustrated in figure~\ref{fig:base_space_div}, $K$ does not directly know the values; instead it is the sections that define the lookup between keys $k \in K$ and the corresponding values in $E$.
+$K$ has no knowledge of the records.
+As illustrated in figure~\ref{fig:base_space_div}, $K$ is more of an indexing space into $E$ that describes the structure of $E$ and itself can be any number of dimensions.
 
@@
     \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
-    %%include second figure where fiber goes otherway
+    \caption{ The base space $E$ is divided into fiber segments $F$. The base space $K$ acts as an index into the fibers.}
 \end{figure}
-The topology $K$ and the
 
-%%this paragraph isn't working
-$F$ are determined by how $E$ is subdivided\cite{QuotientSpaceTopology2020,QuotientSpaceTopology2020}.
-In figure~\ref{fig:base_space_div} we can divide a rectangular base space such that there is a short fiber and long base space or a long fiber and short base space.
-This is analogous to long and wide forms of the same table \cite{wickham2014tidy}.
+Formally, $K$ is the quotient space \cite{QuotientSpaceTopology2020} of $E$, meaning it is the finest space such that every $k \in K$ has a corresponding fiber $F_k$ in $E$.
+In figure~\ref{fig:base_space_div}, $E$ is a rectangle divided by vertical fibers $F$, so the minimal $K$ for which there is always a mapping $\pi: F\rightarrow K$ is the straight line.
+
+We encode indexing from $K$ to $E$ as sections $\tau$ of the fiber bundle that return corresponding records. For example when $K$ has
+\begin{align*}
+    \textbf{vertexes}\;& \tau(k) \text{k= vertex id}
+    \textbf{edges}\; \tau(k,\alpha), \text{k = edge id, $\alpha$ = distance along edge} 
+    \textbf{faces}\; \tau(k,\alpha_1, \alpha_2), \text{k = face id, $\alpha_1$ = x on face, $\alpha_2$ = y on face}
+\end{align*}
+such that we can check for continuity of sections $\tau$ by verifying that the edges align such that $\tau(k_{i},1) = \tau(k_{j},0)$.
+
+As with fibers and monoids, we can decompose the total space into components 
+\begin{equation}
+    \pi: E_1 \rightarrow K, \ldots, \pi: E_i \rightarrow K, \ldots \pi: E_n \rightarrow K
+\end{equation}
+
+which is a decomposition of $F$. The $K$ remains the same because the connectivity of records does not change just because there are fewer variables in each record.
 
@@
     \includegraphics{figures/math/simplex.png}
-    \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) observations to all observations are connected to at least 3 other observations}
+    \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) records to all records are connected to at least 3 others}
     \label{fig:triangle_simplex}

### 2021-01-20 draft/sections/math.tex
rotation group sketch

@@
 The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset.
-For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $\{tp_0, \ldots, tp_n\} \subset \mathbb{R}$.
+For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ \mathscr{}\subset \mathbb{R}$.
 Lets rename this $F$ to a component
+%%use spivaks U w/ subset data {x \in R | x >= 0}
 \begin{equation}
-F_0 = \{tp_0, \ldots, tp_n\} 
+F_0 = \{T_0, \ldots, T_n\} 
 \end{equation}
@@
 \begin{equation}
-F =  \{tp_0, \ldots, tp_n\}  \times \{ti_0, \ldots, ti_n\} 
+F =  \{T_0, \ldots, T_n\}  \times \{t_0, \ldots, t_n\} 
 \end{equation}
@@
 The elements in a given fiber set do not need to be scalers; for example the location can be encoded as the set of tuples $(\textit{lat},\textit{lon})$.  We add the location to the fiber 
+%lat,lon in U
 \begin{equation}
-    F = \{tp_0, \ldots, tp_n\}  \times \{ti_0, \ldots, ti_n\} \times 
-    \{(la_0, ln_0),\ldots, (la_j, ln_k), \ldots (la_n,ln_n)\}
+    F = \{T_0, \ldots, T_n\}  \times \{t_0, \ldots, t_n\} \times 
+    \{(a_0, o_0),\ldots, (a_j, o_k), \ldots (a_n, o_n)\}
 \end{equation}
-such that one record $r$ in the fiber would have the form $(tp, t, (la,ln))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
+such that one record $r$ in the fiber would have the form $(T, t, (a, o))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
 \begin{align}
@@
 \label{sec:data_monoid}
-A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or rotating a shape; this model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
+A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or enlarging a shape.
+This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
+    \label{fig:data_monoid_rotation}
+\end{figure}
+One such set of transforms is the rotation group showin in figure~\ref{fig:}
 One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
 
+% change to Mi 
 We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$.
@@
 A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
-\begin{align*}
+\begin{align*}%lost the \ast
     \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x\\
@@
 
-For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $F_0$ are the set of all transformations $tp_i + c$ such that 
+For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $F_0$ are the set of all transformations $tp_i + c$ such that %%generalize out to x in \U_{temp}, c in \R
+%%the action of the interval is the + 
+%% c is the element that shifts from x1 to x2
 \begin{equation}
@@
 \begin{figure}[ht!]
-    \includegraphics[width=\textwidth]{figures/math/k_different_types.png}
+    \includegraphics[width=.5\textwidth]{figures/math/k_different_types.png}
     \label{fig:base_space_types}
@@
 
-We encode indexing from $K$ to $E$ as sections $\tau$ of the fiber bundle that return corresponding records. For example when $K$ has
+We encode indexing from $K$ to $E$ as sections $\tau$ of the fiber bundle that return corresponding records.
+
+%%if we had a coordinate on
+
+%%pull this down to triangulization section 
+For example when $K$ has
 \begin{align*}
-    \textbf{vertexes}\;& \tau(k) \text{k= vertex id}
-    \textbf{edges}\; \tau(k,\alpha), \text{k = edge id, $\alpha$ = distance along edge} 
-    \textbf{faces}\; \tau(k,\alpha_1, \alpha_2), \text{k = face id, $\alpha_1$ = x on face, $\alpha_2$ = y on face}
+    \textbf{vertexes}\;& \tau(k) \text{k= vertex id}\\
+    \textbf{edges}\;& \tau(k,\alpha), \text{k = edge id, $\alpha$ = distance along edge} \\
+    \textbf{faces}\;& \tau(k,\alpha_1, \alpha_2), \text{k = face id, $\alpha_1$ = x on face, $\alpha_2$ = y on face}
 \end{align*}
 such that we can check for continuity of sections $\tau$ by verifying that the edges align such that $\tau(k_{i},1) = \tau(k_{j},0)$.
+
+\begin{figure}
+    %%include transition function sketch
+\end{figure}
+
+We can represent more complex spaces with transition functions that are maps from the fiber to the fiber along the edges. can also fold in if monoid is commutative on transition function:
 
@@
 
-which is a decomposition of $F$. The $K$ remains the same because the connectivity of records does not change just because there are fewer variables in each record.
+which is a decomposition of $F$. The $K$ remains the same because the connectivity of records does not change just because there are fewer elements in each record.
 
@@
 \label{sec:data_section}
+%%maybe pull down/redo figure
 %%% visual variable encoding in a consistent/atomic way in a complex structure 
@@
 \end{equation}
-
+such that %%insrt math equiality here 
 %%% change to b\c locally trivial, can represent 
-such that in a trivial fiber bundle, $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
+If the fiber bundle is trivial, then $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
 \begin{equation}
@@
 \end{equation}
-
+which is also true when the fiber bundle is locally trivial over $U \subset K$.
 %%%can decompose bundles into Ei such that \tau: E_{i} \rightarrow K, where F_{i} is the fiber of E_{i}
@@
 %% the stalk contains all the local information, typically we only need some of the information, such as the firtst or second direivative.
-in practice we only need the jet which is...in practive will extend the fiber
-The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is a for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.
+in practice we only need the jet which is...in practive will extend the fiber /fatten the E , notatoe E' = E + Jet
+The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.
 
@@
 
+%%define a region b = [top, bottom, right, left]
+%% inverse of b is capital S_b
 \subsubsection{Base space}
@@
 \begin{equation}
-    \rho(s)  = [x(s), y(s), r(s), g(s), b(s)]
+    \rho(s)  = (D_x(s), D_y(s), D_r(s), D_g(s), D_b(s))
     \label{eq:rho}
@@
 
-To draw an image, a region, $H$ is inverse mapped into a region $s \in S$ where
+To draw an image, a region, $H$ is inverse mapped into a region $s \in S$ where %%% 
 \begin{equation}
@@
 \end{equation}
-such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields:
+such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields
+%% redefining notation 
 \begin{align}
-    r &= \iint\limits_s \rho_{\tiny{R}}(s)ds^{2}\\
-    g &= \iint\limits_s \rho_{\tiny{G}}(s)ds^{2}\\
-    b &= \iint\limits_s \rho_{\tiny{B}}(s)ds^{2}
+    R_b &= \iint\limits_s D_r(s)ds^{2}\\
+    G_b &= \iint\limits_s D_g(s)ds^{2}\\
+    B_b &= \iint\limits_s D_b(s)ds^{2}
 \end{align}
@@
 
+%% E' not E
 In this section we will define the artist as a mapping from a sheaf $\mathcal{O}(E)$  to $\mathcal{O}(H)$. 
@@
 \begin{equation}
-    \label{eq:artist_daigram}
+    \label{eq:artist_diagram}
     \begin{tikzcd}
@@
 $Q\circ\mu(\xi^{-1}J)$ forms a visual mark by applying $Q$ to a region mapped to connected components $J \subset K$.}
+    %%maybe change to \tau, \mu, \rho  
 \end{figure}

### 2021-01-20 draft/sections/math.tex
transitions

@@
 \begin{figure}
-    %%include transition function sketch
+    \label{fig:data_base_transition}
+    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
+    \caption{Many non-trivial spaces can be made locally trivial by dividing $E$ into subspaces and defining transition functions between the edges on K for how to glue the two subspaces such that the sigmas are continuous.}
 \end{figure}
 
-We can represent more complex spaces with transition functions that are maps from the fiber to the fiber along the edges. can also fold in if monoid is commutative on transition function:
+We can represent some spaces that are not trivial as
+
+with transition functions that are maps from the fiber to the fiber along the edges. can also fold in if monoid is commutative on transition function:
 

### 2021-01-20 draft/sections/math.tex
transition function

@@
 The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset.
-For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ \mathscr{}\subset \mathbb{R}$.
-Lets rename this $F$ to a component
-%%use spivaks U w/ subset data {x \in R | x >= 0}
+For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$.
 \begin{equation}
-F_0 = \{T_0, \ldots, T_n\} 
+    F = U_{temp}
 \end{equation}
-so that more variables can be added to the fiber. Then lets add the time from GHCN $F_1 =\{t_0, \ldots, t_n\}$
+Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb({Z}^+ \mid x \geq 1011832\}$ 
 \begin{equation}
-F =  \{T_0, \ldots, T_n\}  \times \{t_0, \ldots, t_n\} 
+F =  U_{temp} \times U_{time}
 \end{equation}
-
-The elements in a given fiber set do not need to be scalers; for example the location can be encoded as the set of tuples $(\textit{lat},\textit{lon})$.  We add the location to the fiber 
-%lat,lon in U
+such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$.
+The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_loc = \{y, x \in \mathbb{R}^{2} \mid -90 \leq y \leq 90, -180 \leq x -180\}$:
 \begin{equation}
-    F = \{T_0, \ldots, T_n\}  \times \{t_0, \ldots, t_n\} \times 
-    \{(a_0, o_0),\ldots, (a_j, o_k), \ldots (a_n, o_n)\}
+    F = U_{temp} \times U_{time} \times U_{loc}
 \end{equation}
-such that one record $r$ in the fiber would have the form $(T, t, (a, o))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
+such that one record $r$ in the fiber would have the form $(temperature,\, time,\, (latitude,\, longitude))$.
+We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that
 \begin{align}
-    F_0 &= (\textit{temperature}, \{tp_0, \ldots, tp_n\}, M_0)\\
-    F_1 &= (\textit{time}, \{ti_0, \ldots, ti_n\}, M_1)\\
-    F_2 &= (\textit{location},   \{(la_0, ln_0),\ldots, (la_j, ln_k), \ldots (la_n,ln_n)\}, M_2)
+    F_0 &= (\textit{temperature}, U_{temp}, M_0)\\
+    F_1 &= (\textit{time}, U_{time}, M_1)\\
+    F_2 &= (\textit{location}, U_{loc}, M_2)
 \end{align}
- or more generally (name, set, monoid)
+ or more generally (name, type, monoid)
  \begin{equation}
-    F_i = (\textit{name}, \{x_{i}, \ldots, x_{n}\}, M_{i})
+    F_i = (\textit{name}, U_{type}, M_{i})
  \end{equation}
-as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
+where type is the set of values defined by the datatype as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}.
+We describe $M$ in more detail in section~\ref{sec:data_monoid}.
+We can keep growing the fiber
 \begin{equation}
@@
     \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
-    \caption{Many non-trivial spaces can be made locally trivial by dividing $E$ into subspaces and defining transition functions between the edges on K for how to glue the two subspaces such that the sigmas are continuous.}
+    \caption{Many non-trivial spaces can be made locally trivial by dividing $E$ into locally trivial subspaces and defining transition functions between the edges on K for how to glue the two subspaces such that the sigmas are continuous.}
 \end{figure}
 
-We can represent some spaces that are not trivial as
+Many
 

### 2021-01-20 draft/sections/math.tex
monoids, w stars!

@@
 such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$.
-The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_loc = \{y, x \in \mathbb{R}^{2} \mid -90 \leq y \leq 90, -180 \leq x -180\}$:
+The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90 \leq y \leq 90, -180 \leq x -180\}$:
 \begin{equation}
@@
 \begin{align}
-    F_0 &= (\textit{temperature}, U_{temp}, M_0)\\
-    F_1 &= (\textit{time}, U_{time}, M_1)\\
-    F_2 &= (\textit{location}, U_{loc}, M_2)
+    F_0 &= (\textit{temperature},\, U_{temp},\, M_0)\\
+    F_1 &= (\textit{time},\, U_{time},\, M_1)\\
+    F_2 &= (\textit{location},\, U_{loc},\, M_2)
 \end{align}
- or more generally (name, type, monoid)
+ or more generally (name, type, monoid action)
  \begin{equation}
-    F_i = (\textit{name}, U_{type}, M_{i})
+    F_i = (\textit{name},\, U_{type},\, M_{i})
  \end{equation}
@@
 \label{sec:data_monoid}
-A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or enlarging a shape.
-This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
 
-\begin{figure}
+\begin{figure}[ht!]
     \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
     \label{fig:data_monoid_rotation}
+    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.}.
 \end{figure}
-One such set of transforms is the rotation group showin in figure~\ref{fig:}
-One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
 
-% change to Mi 
+A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or rotating a shape such as in figure~\ref{fig:data_monoid_rotation}.
+This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
+One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
+
 We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$.
-A monoid \cite{Monoid2021} $M$ is a set with an associative binary operator $\ast:M \times M\rightarrow M$.
-A monoid has an identity element $e\in M$ such that $e\ast a= a \ast e = a$ for all $a \in M$.
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M$ is a set $F$ with an action $\bullet: M\times F \rightarrow F$ with the properties:
-\begin{align*}%lost the \ast
-    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in F, m\bullet(t\bullet x) = (m\bullet t) \bullet x\\
-    \textbf{identity}\;& \text{for all } x\in F, e\in M,  e\bullet x = x 
+A monoid \cite{Monoid2021} $M_i$ is a set with an associative binary operator $\ast:M_i \times M_i\rightarrow M_i$.
+A monoid has an identity element $e\in M_i$ such that $e\ast a= a \ast e = a$ for all $a \in M_i$.
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $U_{type}$ with an action $\bullet: M\times U_{type} \rightarrow U_{type}$ with the properties:
+\begin{align*}
+    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in U_{type}, m\bullet(t\bullet x) = (m\ast t) \bullet x\\
+    \textbf{identity}\;& \text{for all } x\in U_{type}, e\in M,  e\bullet x = x 
 \end{align*}
 
-For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $F_0$ are the set of all transformations $tp_i + c$ such that %%generalize out to x in \U_{temp}, c in \R
-%%the action of the interval is the + 
-%% c is the element that shifts from x1 to x2
-\begin{equation}
-F_0 = (\textit{temperature}, \{tp_0, \ldots, tp_n\}, \{tp_0 + c_0, \ldots, tp_0 + c_n, \ldots, tp_n + c_0, \ldots, tp_n + c_n\} ) 
-\end{equation}
+For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $U_{temp}$ are the set of transformations $\{x + c \in U_{temp} \mid x,c \in U_temp\}$ where $c$ is a valid interval between two temperatures $c=x_i-x_j$.
 
@@
 \end{equation}
-of the monoids $M_{i}$ on each $F_{i}\in F$.
+of the monoid $M_{i}$ on each $F_{i}\in F$.
 Because each $F_i, M_i$ can be independent of the other fibers, we can formally express the constraints on heterogenous multivariate datasets in the same manner as univariate datasets.

### 2021-01-20 draft/sections/math.tex
triangularization

@@
 We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the variables in a dataset from how the values  are connected to each other \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}.
+The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified.
 
@@
 \label{sec:data_base}
-The fiber bundle is the gluing map $\pi$ between the fiber space included in the total space $E$ and the base space $K$
+The fiber provides the schema of the dataset, but the fiber bundle is what glues that schema to a structure.
+The fiber bundle is the map $\pi$ between the fiber space included in the total space $E$ and the base space $K$
 \begin{equation}
@@
 
-Formally, $K$ is the quotient space \cite{QuotientSpaceTopology2020} of $E$, meaning it is the finest space such that every $k \in K$ has a corresponding fiber $F_k$ in $E$.
-In figure~\ref{fig:base_space_div}, $E$ is a rectangle divided by vertical fibers $F$, so the minimal $K$ for which there is always a mapping $\pi: F\rightarrow K$ is the straight line.
-
-We encode indexing from $K$ to $E$ as sections $\tau$ of the fiber bundle that return corresponding records.
-
-%%if we had a coordinate on
-
-%%pull this down to triangulization section 
-For example when $K$ has
-\begin{align*}
-    \textbf{vertexes}\;& \tau(k) \text{k= vertex id}\\
-    \textbf{edges}\;& \tau(k,\alpha), \text{k = edge id, $\alpha$ = distance along edge} \\
-    \textbf{faces}\;& \tau(k,\alpha_1, \alpha_2), \text{k = face id, $\alpha_1$ = x on face, $\alpha_2$ = y on face}
-\end{align*}
-such that we can check for continuity of sections $\tau$ by verifying that the edges align such that $\tau(k_{i},1) = \tau(k_{j},0)$.
+Formally $K$ is the quotient space \cite{QuotientSpaceTopology2020} of $E$, meaning it is the finest space such that every $k \in K$ has a corresponding fiber $F_k$ in $E$.
+In figure~\ref{fig:base_space_div}, $E$ is a rectangle divided by vertical fibers $F$, so the minimal $K$ for which there is always a mapping $\pi: F\rightarrow K$ is the line.
 
@@
 \subsubsection{Sample Qs}
+%rewrite as alpha/beta 
 In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and a simple heatmap.
@@
 
+By using simplicies, we can then encode the following indexing structure on $\tau$
+\begin{align*}
+    \textbf{vertexes}\;& \tau(k) \text{k= vertex id}\\
+    \textbf{edges}\;& \tau(k,\alpha), \text{k = edge id, $\alpha$ = distance along edge} \\
+    \textbf{faces}\;& \tau(k,\alpha, \beta), \text{k = face id, $\alpha$ = x on face, $\beta$ = y on face}
+\end{align*}
+
+which $\xi$ can use as part of its mapping from visual space back to data space. Path connected components are then sections where $\tau(k_{i},1) = \tau(k_{j},0)$ or the edges of the faces align.
+
 \subparagraph{Example}
 \begin{figure}[ht!]
+    \label{fig:triangle_torus}
     \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
-    \label{fig:triangle_torus}
+    \caption{Representation of the base space $K$ of a toridal total space $E$ as two connected 2-simplices.}
 \end{figure}
 
-Given data that lies on the toroidal space shown in figure~\ref{fig:triangle_torus}, the torus $K$ can be implemented as a simplacial complex of two 2-simplexes.
-We unfold the torus into the two triangles that compose the square in figure~\ref{fig:artist_heatmap} and encode that as K.
-We also constrain the functions on $A$, $A^\prime$, $B$ and $B^\prime$ such that $A$ can be glued to $A^\prime$ and $B$ to $B^\prime$ to reconstruct the torus.
+Given data that lies on the toroidal space shown in figure~\ref{fig:triangle_torus}, the torus $E$ base space $K$ can be implemented as a simplacial complex of two 2-simplexes.
+We unfold the torus into the two triangles that compose the square; the sections on these triangles are $\tau(triangle id k, \alpha, \beta)$.
+We also put transition functions on the edges such that $A$ can be glued to $A^\prime$ and $B$ to $B^\prime$ to reconstruct the torus.
 

### 2021-01-20 draft/sections/math.tex
closing out k

@@
 
-Many
-
-with transition functions that are maps from the fiber to the fiber along the edges. can also fold in if monoid is commutative on transition function:
+A fiber bundle is trivial if $E = K\times F$, which means that the fibers $F$ are the same over every $k$.
+We can make many non-trivial spaces locally trivial by breaking up the total space $E$ over local neighborhoods $U \subset K$ such that the fibers are equivalant.
+An example of a non-trivial fiber bundle is a mobius strip, where because of the twist some fibers are $\left[0,1\right]$ and others are $\left[1,0\right]$ such that we cannot construct a global map of which edge is 0 and which edge is 1.
+In figure ~\ref{fig:data_base_transition}, the mobius strip is cut into two rectangular base spaces over $K=\{e_1, e_2\}$.
+We define transition functions from the fiber to the fiber along the edges of the rectangles such that we can glue these rectangles together to reform the mobius strip.
+A constraint we impose on the transition functions is that the monoid actions are commutative
+-% need more detail please on commutativty
 

### 2021-01-20 draft/sections/math.tex
sections

@@
 \end{equation}
-such that one record $r$ in the fiber would have the form $(temperature,\, time,\, (latitude,\, longitude))$.
+such that one record $r$ in the fiber would have the form $r=(temperature,\, time,\, (latitude,\, longitude))$.
 We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that
@@
 
-A fiber bundle is trivial if $E = K\times F$, which means that the fibers $F$ are the same over every $k$.
+A fiber bundle is trivial if $E = K\times F$, which means that the fibers $F$ are the same over every $k$ \cite{rowlandFiberBundle,FiberBundle2020}.
 We can make many non-trivial spaces locally trivial by breaking up the total space $E$ over local neighborhoods $U \subset K$ such that the fibers are equivalant.
@@
 \label{sec:data_section}
-%%maybe pull down/redo figure
-%%% visual variable encoding in a consistent/atomic way in a complex structure 
-\label{sec:section_data}
-The sections of the fiber bundles are the instances of the data.
-They are the functions that map from a point $k \in K$ to a set of values in $F$.
-For example, if we have a database, the section is the table that yields a row given an index.
+\begin{figure}[ht!]
+    \label{fig:base_space_div}
+    \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
+    \caption{ A section $\tau$ is a mapping from $K$ to $F$ that returns a record for each $k \in K$.}
+\end{figure}
+
+The sections of the fiber bundle are maps from points $k$ on $K$ to records $r$ in $F$.
+When we map our data to visual variables, the $\tau$ returns the set of values that are mapped together into a single graphic element.
+For GHCN, the $\tau$ functions return a table in a database with the schema defined in $F$ and the connectivity defined in $K$: this means the sections can be defined such that all values have the same year or location, e.g each table returned by a $\tau$ is for a different station.
+
 The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$ 
@@
 %%% change to b\c locally trivial, can represent 
-If the fiber bundle is trivial, then $E = K \times F$ \cite{rowlandFiberBundle,FiberBundle2020}:
+If the fiber bundle is trivial, then :
 \begin{equation}

### 2021-01-20 draft/sections/math.tex
extended E

@@
 
-As with fibers and monoids, we can decompose the total space into components 
+As with fibers and monoids, we can decompose the total space into components $\pi:E_i\rightarrow K$ where
 \begin{equation}
-    \pi: E_1 \rightarrow K, \ldots, \pi: E_i \rightarrow K, \ldots \pi: E_n \rightarrow K
+    \pi:E_1\oplus\ldots\oplus E_i \oplus\ldots \oplus E_n \rightarrow K
 \end{equation}
@@
 \end{equation}
-such that %%insrt math equiality here 
-%%% change to b\c locally trivial, can represent 
-If the fiber bundle is trivial, then :
+such that there is $\pi(\tau(k)) = k$ map back to $K$.
+
+If the fiber bundle is locally trivial, then
 \begin{equation}
     \label{eq:section_return}
-    \tau(k) = (k, (x_{F_{0}}, \ldots, x_{F_{n}}))
+    \tau(k) = (k, (g_{F_{0}}(k), \ldots, g_{F_{n}}(k))
 \end{equation}

### 2021-01-21 draft/sections/math.tex
reshuffling data/moving around and kicking out continuity:

@@
 
-We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the variables in a dataset from how the values  are connected to each other \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified.
-
 We first describe the fiber bundle spaces of data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}).
@@
 \label{sec:data}
-Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations.
-Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
-Using a fiber bundle allows us to express the properties of the variables(\ref{sec:data_fiber}), the connectivity of the records (\ref{sec:data_base}), and tables of records (\ref{sec:data_section}).
+%%% maybe change this to a generic representational data set
+We build on Butler's fiber bundle driven abstraction of data formats\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} because fiber bundles can express the properties of the variables(\ref{sec:data_fiber}), the continuity of the records (\ref{sec:data_base}), and distinct instances of records (\ref{sec:data_section}).
+The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified.
 
@@
 The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset.
+
+%%push straight math generalization up
+%% We define w/ mapping, as Spivak
+
+%%push down to an example, knock it back to semi concrete
 For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
@@
 \end{equation}
-Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb({Z}^+ \mid x \geq 1011832\}$ 
+Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb{R}^+\}$ 
 \begin{equation}
@@
 such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$.
-The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90 \leq y \leq 90, -180 \leq x -180\}$:
+The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90^\circ \leq y \leq 90^\circ, -180^\circ \leq x -180^\circ\}$: %add comment about latitude and longitude
 \begin{equation}
@@
     F_2 &= (\textit{location},\, U_{loc},\, M_2)
-\end{align}
+\end{align}%the fiber is the set, variable is the tuple such that there is a map from the index i in F to the tuple (name, M_0)  for every U_{type} we can construct a map to the name and the type as described by Spivak
  or more generally (name, type, monoid action)
@@
 
+%%invert this section, start w/ general math than GHCN
+
 \subsubsection{Measurement Scales: Monoid Actions}
@@
     \label{fig:data_monoid_rotation}
-    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.}.
+    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.
+    %% dataset that is symmetric, fiber is x,y, rotatation action
+    %% rotate + plot or plot + rotate
+    %%emoji shuffle 
+    }  
 \end{figure}
@@
 This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
-One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
 
@@
 \end{align*}
-
-For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $U_{temp}$ are the set of transformations $\{x + c \in U_{temp} \mid x,c \in U_temp\}$ where $c$ is a valid interval between two temperatures $c=x_i-x_j$.
-
 As with the fiber $F$, the total monoid space $M$ is the cartesian product
@@
 of the monoid $M_{i}$ on each $F_{i}\in F$.
-Because each $F_i, M_i$ can be independent of the other fibers, we can formally express the constraints on heterogenous multivariate datasets in the same manner as univariate datasets.
 
-\subsubsection{Connectivity: Base Space $K$} 
+Steven's defined the measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions permissible on the measurments; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
+
+For example, the temperature in \textdegree C is on an interval scale, which means that the monoid actions $M_0$ on $U_{temp}$ are the set of translations $\{f: x\mapsto x + c \in U_{temp} \mid x,c \in U_temp\}$ where $c$ is a valid interval between two temperatures $c=x_i-x_j$.
+
+\subsubsection{continuity: Base Space $K$} 
+%%fold butlder
 \label{sec:data_base}
@@
 
-\begin{figure}
-    \label{fig:data_base_transition}
-    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
-    \caption{Many non-trivial spaces can be made locally trivial by dividing $E$ into locally trivial subspaces and defining transition functions between the edges on K for how to glue the two subspaces such that the sigmas are continuous.}
-\end{figure}
-
-A fiber bundle is trivial if $E = K\times F$, which means that the fibers $F$ are the same over every $k$ \cite{rowlandFiberBundle,FiberBundle2020}.
-We can make many non-trivial spaces locally trivial by breaking up the total space $E$ over local neighborhoods $U \subset K$ such that the fibers are equivalant.
-An example of a non-trivial fiber bundle is a mobius strip, where because of the twist some fibers are $\left[0,1\right]$ and others are $\left[1,0\right]$ such that we cannot construct a global map of which edge is 0 and which edge is 1.
-In figure ~\ref{fig:data_base_transition}, the mobius strip is cut into two rectangular base spaces over $K=\{e_1, e_2\}$.
-We define transition functions from the fiber to the fiber along the edges of the rectangles such that we can glue these rectangles together to reform the mobius strip.
-A constraint we impose on the transition functions is that the monoid actions are commutative
--% need more detail please on commutativty
-
 As with fibers and monoids, we can decompose the total space into components $\pi:E_i\rightarrow K$ where
@@
 When we map our data to visual variables, the $\tau$ returns the set of values that are mapped together into a single graphic element.
-For GHCN, the $\tau$ functions return a table in a database with the schema defined in $F$ and the connectivity defined in $K$: this means the sections can be defined such that all values have the same year or location, e.g each table returned by a $\tau$ is for a different station.
-
 The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$
@@
 
-If the fiber bundle is locally trivial, then
+A fiber bundle is trivial if $E = K\times F$\cite{rowlandFiberBundle,FiberBundle2020}.
+All fiber bundles are locally trivial, which means if you restrict $E$ over a small enough neighborhood $U \subset K$ then it is a locally trivial bundle over $U$.
+When the fiber bundle is locally trivial, then
 \begin{equation}
@@
 \end{equation}
-which is also true when the fiber bundle is locally trivial over $U \subset K$.
-%%%can decompose bundles into Ei such that \tau: E_{i} \rightarrow K, where F_{i} is the fiber of E_{i}
-which we can also decompose such that
+the section $\tau$ is the record in the $F$. Because we can decompose the bundle and the fiber, we can formulate $\tau$ as 
 \begin{align}
-\tau_{0}(k) &= (k, (x_{F_{0}}))\\
-            & \vdotswithin{=}\\
-\tau_{n}(k) &= (k, (x_{F_{n}}))
+\tau = (\tau_0,\ldots, \tau_i, \dots, \tau_n) 
 \end{align}
+where each section $\tau_i$ is a variable or set of variables.
 
-which allows us to have field wise access to the data so long as there is a shared $K$.  
-%% we will use this property later for visual mapping such that we can seperate each variable into its own bundle
+\paragraph{Example?}
+For GHCN, the $\tau$ functions return a table in a database with the schema defined in $F$ and the continuity defined in $K$: this means the sections can be defined such that all values have the same year or location, e.g each table returned by a $\tau$ is for a different station.
 
@@
 
+Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations.
+Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
 % can't make a 1D fiber - can't make a 1D fiber, switch the picture to plane over 0D and 1D and then can talk about disconnected or connected
@@
 
-\subparagraph{Example}
+\subparagraph{Example: Mobius Strip}\mbox{}\\
+\begin{figure}[ht!]
+    \label{fig:data_base_transition}
+    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
+    \caption{Many non-trivial spaces can be made locally trivial by dividing $E$ into locally trivial subspaces and defining transition functions between the edges on K for how to glue the two subspaces such that the sigmas are continuous.}
+\end{figure}
+
+ An example of a non-trivial fiber bundle is a mobius strip\cite{find the citation for why};
+
+In figure ~\ref{fig:data_base_transition}, the mobius strip is cut into two rectangular base spaces over $K=\{e_1, e_2\}$.
+We define transition functions from the fiber to the fiber along the edges of the rectangles such that we can glue these rectangles together to reform the mobius strip.
+A constraint we impose on the transition functions is that the monoid actions are commutative
+%monoid action m*f\mapsto delta(mf)= m*delta(f)
+
+\subparagraph{Example: Torus}\mbox{}\\
 \begin{figure}[ht!]

### 2021-01-22 draft/sections/math.tex
pulled general FB intro back up to set context for unpacking

@@
 
-We first describe the fiber bundle spaces of data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}).
+We first describe how we model data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles.
 We then discuss the equivariant maps between data and visual characteristics (\ref{sec:artist_nu}) and visual characteristics and graphics (\ref{sec:artist_q}) that make up the artist.
@@
 \label{sec:data}
-%%% maybe change this to a generic representational data set
-We build on Butler's fiber bundle driven abstraction of data formats\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} because fiber bundles can express the properties of the variables(\ref{sec:data_fiber}), the continuity of the records (\ref{sec:data_base}), and distinct instances of records (\ref{sec:data_section}).
-The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified.
+%%% 
+We build on Butler's fiber bundle driven abstraction of data formats\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989},  ....
+
+Fiber bundles are structures $(E,\,B,\,\pi ,\,F)$ where $E$, $F$, and $K$ are topological spaces that encode 
+\begin{itemize}
+\item the properties of the variables in the fiber $F$ (\ref{sec:data_fiber})
+\item the continuity of the records in the base space $K$ (\ref{sec:data_base})
+\item collections of records $\tau$ (\ref{sec:data_section}). 
+\end{itemize}
+
+where $E$ is the total space of data that $F$ lives in. The bundle is the projection map $\pi$
+\begin{equation}
+    \label{eq:fiber_bundle}
+    \begin{tikzcd}
+        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
+    \end{tikzcd}
+\end{equation}
+
+that binds the variable space $F$ to the base space $K$.
+
+%%probbaly need some more words here maybe, maybe pull that first picture back up now
+
+The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified.
+In the trivial bundle $E=K\times F$ such that the total space $E$ falls out and discussion can be limited to the $K$ and $F$ spaces.
 
@@
 \label{sec:data_base}
-The fiber provides the schema of the dataset, but the fiber bundle is what glues that schema to a structure.
-The fiber bundle is the map $\pi$ between the fiber space included in the total space $E$ and the base space $K$
-\begin{equation}
-    \label{eq:fiber_bundle}
-    \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
-    \end{tikzcd}
-\end{equation}
-
 The base space $K$ provides a way to encode how the records are connected to each other; for example the GHCN data can be expressed as discrete observations, many timeseries, maps, a network of stations, etc.

### 2021-01-22 draft/sections/intro.tex
finished out k
transition function
pulled more spivak back in

@@
 We then propose a new architecture for the Python visualization library Matplotlib \cite{hunterMatplotlib2DGraphics2007} based on these descriptions because the Matplolib artist layer is analogous to the transforms.
+
+\subsection{what are we doing}
+constraints: toplogy of dataset + structure on the variables
+implementation: be fully feature
+we can haz compose the things into larger structure. \& has framwework to do larger things w/ algebraic formalizations.
 
@@
 \subsubsection{Data Type and Structure}
+%% mention tidy/spivak first, set up that we use variable, record, key/index notation
 \label{sec:intro_data_structure}
@@
 
+Throughout this form we will be using the following conventions to discuss data:
+
+\begin{description}
+    \item[variable]
+    \item[index]  
+    \item[record]
+    \item[value]  
+\end{description}
+
 \end{document}

### 2021-01-22 draft/sections/math.tex
pulled more spivak back in

@@
 \label{sec:data}
-%%% 
-We build on Butler's fiber bundle driven abstraction of data formats\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989},  ....
-
-Fiber bundles are structures $(E,\,B,\,\pi ,\,F)$ where $E$, $F$, and $K$ are topological spaces that encode 
+We build on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} because fiber bundles are mathematical structures that are flexible enough to encode all the types of data described in section~\ref{sec:intro_data_structure} and many more.
+We model data as the fiber bundle $(E,\,B,\,\pi ,\,F)$, where $E$, $F$, and $K$ are topological spaces that encode
 \begin{itemize}
@@
 
-where $E$ is the total space of data that $F$ lives in. The bundle is the projection map $\pi$
+and $E$ is the total space of data that $F$ lives in. The bundle is the projection map $\pi$
 \begin{equation}
@@
 that binds the variable space $F$ to the base space $K$.
+The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified, because the trivial bundle is $E=K\times F$ such that extra structure in the total space $E$ falls out and discussion can be focused on the fiber and base space.
 
-%%probbaly need some more words here maybe, maybe pull that first picture back up now
-
-The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified.
-In the trivial bundle $E=K\times F$ such that the total space $E$ falls out and discussion can be limited to the $K$ and $F$ spaces.
+\begin{figure}[ht!]
+    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
+    \label{fig:data_base_space}
+    \caption{}.  
+\end{figure}
+ We propose fiber bundles as an underlying data representation because they provide a way to decouple the components of a dataset.
+In figure~\ref{fig:}, we have two datasets $\tau_1$ and $\tau_2$ that have the same indexing structure $K$ and and the same schema $F$ and we can independently codify which properties of each component need to be preserved in the visualization.
 
@@
 \label{sec:data_fiber}
-The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset.
+We use Spivak's description of schemas in his categorical formulation of databases \cite{spivakSIMPLICIALDATABASES} as our fiber space.
+He describes the variables in a dataset as having names $C$ and datetypes $D$.
+Spivak then constructs a set $U_{\sigma}$ that is the universal set of values
+\begin{tikzcd}
+    U_{\sigma} \arrow[r] \arrow[d, "\pi_{\sigma}"'] & U \arrow[d, "\pi"] \\
+    C \arrow[r, "\sigma"']                          & \textbf{DT}       
+\end{tikzcd}
 
-%%push straight math generalization up
-%% We define w/ mapping, as Spivak
+restricted to values of types $\mathbf{DT}$ bound to variable names $C$. Therefore the fiber is for a one variable dataset is
+\begin{equation}
+    F = U_{\sigma}
+\end{equation}
+where $\sigma$ is the schema binding variable name to variable type.
+Therefore a dataset with multiple variables $c \in C$ has a fiber that is the cartesian cross product of $U_{\sigma}$ restricted over $c$ spaces:
 
-%%push down to an example, knock it back to semi concrete
+\begin{equation}
+F = U_{\sigma(c_1)}\times \ldots U{\sigma{c_i}} \ldots\timesU_{sigma(c_n)}
+\end{equation}
+
+which is equivalent to
+
+\begin{equation}
+    F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
+\end{equation}
+
+%%somehoe close out this piece.
+
 For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
@@
     F_2 &= (\textit{location},\, U_{loc},\, M_2)
-\end{align}%the fiber is the set, variable is the tuple such that there is a map from the index i in F to the tuple (name, M_0)  for every U_{type} we can construct a map to the name and the type as described by Spivak
- or more generally (name, type, monoid action)
- \begin{equation}
-    F_i = (\textit{name},\, U_{type},\, M_{i})
- \end{equation}
-where type is the set of values defined by the datatype as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}.
-We describe $M$ in more detail in section~\ref{sec:data_monoid}.
-We can keep growing the fiber
-\begin{equation}
-F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
-\end{equation}
-and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.
+\end{align}
 
@@
 The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.
-
-\subsubsection{Case Study: GHCN}
-\textcolor{red}{Moved \& walked through b/c was getting clunky to not have terms yet}\\
-
-Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations.
-Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
-% can't make a 1D fiber - can't make a 1D fiber, switch the picture to plane over 0D and 1D and then can talk about disconnected or connected
-
-% break out the mbox/clean up the typesetting
-The fiber bundle model is flexible enough to express some of the many different forms that temperature data can come in.
-
-\begin{figure}[ht!]
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-        %% add box around neighboring P and Map
-        \label{fig:base_example_line}
-    \end{subfigure}
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_2k.png}
-        \label{fig:base_example_plane}
-    \end{subfigure}
-    \label{fig:base_example}
-    \caption{These two datasets have the same fiber of temperature but different base spaces.
-In figure~\ref{fig:base_example_line} the temperature values are 1D continuous, while in figure~\ref{fig:base_example_plane} the temperature values are 2D continuous.}
-\end{figure}
-
-The datasets in figure~\ref{fig:base_example} have identical fibers that encode a set of temperature values.
-In figure~\ref{fig:base_example_line} the temperatures lie on a line such that a section could return a timeseries or a distribution.
-In figure~\ref{fig:base_example_plane}, the temperatures lie on a 2D continuous plane; a section could return a map or contour.
-Because the fiber is 1D, it does not encode metadata
-
-\begin{figure}[ht!]
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
-        \label{fig:fiber_example_plane}
-    \end{subfigure}
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
-        \label{fig:fiber_example_cube}
-    \end{subfigure}
-    \label{fig:fiber_example}
-    \caption{The fiber is expanded to include metadata fields that describe the semantics of $K$.
-In figure~\ref{fig:fiber_example_plane} the fiber is \textrm{temperature} $\times$ \textrm{time} and in figure~\ref{fig:fiber_example_cube} the fiber is \textrm{temperature}$\times$ \textrm{latitude} $\times$ {longitude}}.
-\end{figure}
-
-To encode the metadata, the fiber is expanded as illustrated in figure~\ref{fig:fiber_example}.
-The fiber in figure~\ref{fig:fiber_example_plane} is the cartesian product of the space of possible temperature values in degrees celsius and space of possible time values
-\begin{equation}
-F = [temp_{min}, temp_{max}] \times [time_{min}, time_{max}]
-\label{eq:fiber_plane}
-\end{equation}
-
-while the fiber in figure~\ref{fig:fiber_example_cube} is the cartesian product of temperature, latitude, and longitude
-\begin{equation}
-F = [temp_{min}, temp_{max}] \times [-90, 90] \times [-180, 180]
-\label{eg:fiber_cube}
-\end{equation}
-
-such that $E$ is the space of all possible points in $F$.
-
-\begin{figure}[ht!]
-    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
-    \label{fig:fiber_example_section}
-    \caption{The section $\tau_1$ returns the blue points,  while $\tau_2$ returns the purple points. $\Gamma(E)$ is the set of all sections, including $\tau_1$ and $\tau_2$}.  
-\end{figure}
-Given the fiber described in equation~\ref{eq:fiber_plane}, the sections $\tau_{1}$ and $\tau_{2}$ in figure~\ref{fig:fiber_example_section} return tuples of the form
-\begin{equation}
-\tau(k) = (k, (temperature, time))
-\end{equation}
-such that sections with the constraint that time is monotonic return a timeseries.
 

### 2021-01-22 draft/sections/scraps.tex
pulled more spivak back in

\subsubsection{Case Study: GHCN}
\textcolor{red}{Moved \& walked through b/c was getting clunky to not have terms yet}\\

Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations. Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
% can't make a 1D fiber - can't make a 1D fiber, switch the picture to plane over 0D and 1D and then can talk about disconnected or connected

% break out the mbox/clean up the typesetting
The fiber bundle model is flexible enough to express some of the many different forms that temperature data can come in.

\begin{figure}[ht!]
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
        %% add box around neighboring P and Map
        \label{fig:base_example_line}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/temp_2k.png}
        \label{fig:base_example_plane}
    \end{subfigure}
    \label{fig:base_example}
    \caption{These two datasets have the same fiber of temperature but different base spaces. In figure~\ref{fig:base_example_line} the temperature values are 1D continuous, while in figure~\ref{fig:base_example_plane} the temperature values are 2D continuous.}
\end{figure}

The datasets in figure~\ref{fig:base_example} have identical fibers that encode a set of temperature values. In figure~\ref{fig:base_example_line} the temperatures lie on a line such that a section could return a timeseries or a distribution. In figure~\ref{fig:base_example_plane}, the temperatures lie on a 2D continuous plane; a section could return a map or contour. Because the fiber is 1D, it does not encode metadata

\begin{figure}[ht!]
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
        \label{fig:fiber_example_plane}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
        \label{fig:fiber_example_cube}
    \end{subfigure}
    \label{fig:fiber_example}
    \caption{The fiber is expanded to include metadata fields that describe the semantics of $K$. In figure~\ref{fig:fiber_example_plane} the fiber is \textrm{temperature} $\times$ \textrm{time} and in figure~\ref{fig:fiber_example_cube} the fiber is \textrm{temperature}$\times$ \textrm{latitude} $\times$ {longitude}}.
\end{figure}

To encode the metadata, the fiber is expanded as illustrated in figure~\ref{fig:fiber_example}. The fiber in figure~\ref{fig:fiber_example_plane} is the cartesian product of the space of possible temperature values in degrees celsius and space of possible time values
\begin{equation}
F = [temp_{min}, temp_{max}] \times [time_{min}, time_{max}]
\label{eq:fiber_plane}
\end{equation}

while the fiber in figure~\ref{fig:fiber_example_cube} is the cartesian product of temperature, latitude, and longitude
\begin{equation}
F = [temp_{min}, temp_{max}] \times [-90, 90] \times [-180, 180]
\label{eg:fiber_cube}
\end{equation}

such that $E$ is the space of all possible points in $F$.

\begin{figure}[ht!]
    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
    \label{fig:fiber_example_section}
    \caption{The section $\tau_1$ returns the blue points,  while $\tau_2$ returns the purple points. $\Gamma(E)$ is the set of all sections, including $\tau_1$ and $\tau_2$}.  
\end{figure}
Given the fiber described in equation~\ref{eq:fiber_plane}, the sections $\tau_{1}$ and $\tau_{2}$ in figure~\ref{fig:fiber_example_section} return tuples of the form
\begin{equation}
\tau(k) = (k, (temperature, time))
\end{equation}
such that sections with the constraint that time is monotonic return a timeseries.

### 2021-01-24 draft/sections/math.tex
translation monoid

@@
 
-\section{Notation \& Definitions}
+\section{Topological Artist Model}
 
-%% give the why of the math after every mention of why 
 In this section we introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.
@@
 \label{sec:data}
-We build on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} because fiber bundles are mathematical structures that are flexible enough to encode all the types of data described in section~\ref{sec:intro_data_structure} and many more.
+We build on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989} because fiber bundles are mathematical structures that are flexible enough express all the types of data described in section~\ref{sec:intro_data_structure}.
+\textcolor{blue}{potentially more words on why fiber bundles stolen from like the scipy proposal but maybe this goes in the intro}
+
 We model data as the fiber bundle $(E,\,B,\,\pi ,\,F)$, where $E$, $F$, and $K$ are topological spaces that encode 
@@
 \end{figure}
- We propose fiber bundles as an underlying data representation because they provide a way to decouple the components of a dataset.
-In figure~\ref{fig:}, we have two datasets $\tau_1$ and $\tau_2$ that have the same indexing structure $K$ and and the same schema $F$ and we can independently codify which properties of each component need to be preserved in the visualization.
+In figure~\ref{fig:}, $\tau_1$ and $\tau_2$ encode two different timeseries of temperature.
+Both these timeseries live in the fiber $F$ which is an $\n \times \m$ array because it is the set of all possible (time, temperature) values.
+The base space $K$ encodes that all datasets in this fiberbundle are 1D continous.
 
@@
 \label{sec:data_fiber}
-We use Spivak's description of schemas in his categorical formulation of databases \cite{spivakSIMPLICIALDATABASES} as our fiber space.
-He describes the variables in a dataset as having names $C$ and datetypes $D$.
-Spivak then constructs a set $U_{\sigma}$ that is the universal set of values
+The fiber is a topological space that is the set of possible values of the data; the values themselves can be any dimension and type and have any continuity.
+We use Spivak's description of simplicial database schemas \cite{spivakSIMPLICIALDATABASES} as the basis of our fiber space because he binds the values in the fiber to variable names and types.
+Spivak constructs a set $U$ that is the disjoint union of all possible objects of the types $T\in DT$ that are in the datasets embedded in the fiber bundle.
+He then defines the single variable set $U_\sigma$
+\begin{equation}
+    \label{eq:data_types}
 \begin{tikzcd}
@@
 \end{tikzcd}
-
-restricted to values of types $\mathbf{DT}$ bound to variable names $C$. Therefore the fiber is for a one variable dataset is
+\end{equation}
+which is $U$ restricted to objects of type $T$ bound to variable name $c$. Given $\U_{\sigma}$, the fiber for a one variable dataset is
 \begin{equation}
-    F = U_{\sigma}
+    F = U_{\sigma(c)} = U_{T} 
 \end{equation}
-where $\sigma$ is the schema binding variable name to variable type.
-Therefore a dataset with multiple variables $c \in C$ has a fiber that is the cartesian cross product of $U_{\sigma}$ restricted over $c$ spaces:
-
+where $\sigma$ is the schema binding variable name $c$ to its datatype $T$.
+A dataset with multiple variables has a fiber that is the cartesian cross product of $U_{\sigma}$ applied to all the columns:
 \begin{equation}
-F = U_{\sigma(c_1)}\times \ldots U{\sigma{c_i}} \ldots\timesU_{sigma(c_n)}
+F = U_{T_1}\times \ldots U{T_i} \ldots\timesU_{T_n}
 \end{equation}
@@
 
-%%somehoe close out this piece.
+which allows us to decouple $F$ into its component fibers $F_i$. For every variable
+\begin{equation}
+V_i = (c_i,\, T_i,\, U_{T_i}\)
+\end{equation} 
+we keep track of its name $c$, type $T$, and component in the fiber $U_{T}$.
 
-For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
+\begin{figure}[ht!]
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
+        \label{fig:fiber_example_plane}
+    \end{subfigure}
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
+        \label{fig:fiber_example_cube}
+    \end{subfigure}
+    \label{fig:data_fiber_example}
+    \caption{}}.
+\end{figure}
+For example, the data in figure~\ref{fig:fiber_example_plane} is a set of temperature measurements in \textdegree C and the times those measurements were taken at.
+Temperature values are floating point numbers that come from the set of real numbers $U_{float} = \mathbb{R}$ and times that are always positive $U_{datetime}= \mathbb{R}^+$, so the fiber is
 \begin{equation}
-    F = U_{temp}
+    F = \mathbb{R} \times \mathbb{R}^{+}
 \end{equation} 
-Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb{R}^+\}$ 
+where $V_0 = (temperature,\, \texttt{float},\, \mathbb{R})$ and $V_1 = (time,\, \texttt{datetime},\, \mathbb{R}^+)$
+In figure~\ref{fig:fiber_example_cube}, a third variable wind is added to the dataset.
+This wind variable is of type \texttt{wind} and has two components speed and direction $\{(m,d) \in \mathbb{R}^{2} \mid  0\leq m,\, 0 \leq d \leq 360\}$ .
+The new variable is added to the fiber
 \begin{equation}
-F =  U_{temp} \times U_{time}
+    F = \mathbb{R} \times \mathbb{R} \times \mathbb{R}^2
 \end{equation} 
-such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$.
-The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90^\circ \leq y \leq 90^\circ, -180^\circ \leq x -180^\circ\}$: %add comment about latitude and longitude
-\begin{equation}
-    F = U_{temp} \times U_{time} \times U_{loc}
-\end{equation}
-such that one record $r$ in the fiber would have the form $r=(temperature,\, time,\, (latitude,\, longitude))$.
-We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that
-\begin{align}
-    F_0 &= (\textit{temperature},\, U_{temp},\, M_0)\\
-    F_1 &= (\textit{time},\, U_{time},\, M_1)\\
-    F_2 &= (\textit{location},\, U_{loc},\, M_2)
-\end{align}
-
-%%invert this section, start w/ general math than GHCN
+such that $V_2 = (wind,\, \texttt{wind},\, \mathbb{R}^{2})$
 
@@
 \label{sec:data_monoid}
-
-\begin{figure}[ht!]
-    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
-    \label{fig:data_monoid_rotation}
-    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.
-    %% dataset that is symmetric, fiber is x,y, rotatation action
-    %% rotate + plot or plot + rotate
-    %%emoji shuffle 
-    }  
-\end{figure}
-
-A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or rotating a shape such as in figure~\ref{fig:data_monoid_rotation}.
-This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
-
-We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$.
+We formally describe the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
+We generalize this set of allowable transformations to the monoid actions $M_i$ on $F_i$.
 A monoid \cite{Monoid2021} $M_i$ is a set with an associative binary operator $\ast:M_i \times M_i\rightarrow M_i$.
 A monoid has an identity element $e\in M_i$ such that $e\ast a= a \ast e = a$ for all $a \in M_i$.
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $U_{type}$ with an action $\bullet: M\times U_{type} \rightarrow U_{type}$ with the properties:
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $U_{T_i}$ with an action $\bullet: M\times U_{T_i} \rightarrow U_{T_i}$ with the properties:
 \begin{align*}
-    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in U_{type}, m\bullet(t\bullet x) = (m\ast t) \bullet x\\
-    \textbf{identity}\;& \text{for all } x\in U_{type}, e\in M,  e\bullet x = x 
+    \textbf{associativity}\;& \text{for all } f,g \in M_i \text{ and } x\in U_{T_i}, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
+    \textbf{identity}\;& \text{for all } x\in U_{T_i}, e\in M_i,  e\bullet x = x 
 \end{align*}
@@
 \end{equation}
-of the monoid $M_{i}$ on each $F_{i}\in F$.
+of the monoid $M_{i}$ on each $F_{i}\in F$. We expand our definition of variables 
+\begin{equation}
+        V_i = (c_i,\, T_i,\, U_{T_i}\, M_i)
+\end{equation}
+to include the monoid actions that act on $U_{T_i}$.
 
-Steven's defined the measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions permissible on the measurments; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-
-For example, the temperature in \textdegree C is on an interval scale, which means that the monoid actions $M_0$ on $U_{temp}$ are the set of translations $\{f: x\mapsto x + c \in U_{temp} \mid x,c \in U_temp\}$ where $c$ is a valid interval between two temperatures $c=x_i-x_j$.
+Steven's defined the measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions permissible on the measurments; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
+For example, given
+\begin{itemize}
+    \item temperature $x \in \mathbb{R}$
+    \item monoid addition operator $\ast = +$
+    \item with monoid operations: $f: x\mapsto x + c_1 $, $g: x\mapsto x + c_2$
+    \item monoid action operator $\bullet = \circ$
+\end{itemize}
+then the translation monoid actions on temperature satisfy the condition
+\begin{equation}
+    \begin{tikzcd}
+        \mathbb{R} \arrow[rd, "(x+c_1)\circ(x+c_2)"] \arrow[d, "x+c_1"'] &            \\
+        \mathbb{R} \arrow[r, "x+c_2"']                                   & \mathbb{R}
+    \end{tikzcd}
+\end{equation}
 

### 2021-01-25 draft/sections/math.tex
filling in captions

@@
 We build on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989} because fiber bundles are mathematical structures that are flexible enough express all the types of data described in section~\ref{sec:intro_data_structure}.
-\textcolor{blue}{potentially more words on why fiber bundles stolen from like the scipy proposal but maybe this goes in the intro}
+The advantage of fiber bundles is they provide a way to encode the continuity in a dataset without making assumptions as to what that continuity is.
+In turn this representation of continuity can then be used to keep track of how the data fits together, for example if a visualization of a very large dataset calls for parallelization.
 
@@
     \label{fig:data_base_space}
-    \caption{}.  
+    \caption{ The fiber $F$ is a continuous plane encoding the set of all possible (time, temperature) values.
+The base space $K$ expresses that the fibers lie in continous space and therefore the data is continuous.
+The mapping $\pi:E\rightarrow K$ from total space to base space is what defines this structure as a fiber bundle.
+The continuous function $\tau_1$ returns a record in each fiber for each point on $K$; the set of all values $\tau_1$ returns is a dataset that lies in this fiberbundle.
+\textcolor{MidnightBlue}{remove one of the sections from this figure}}
 \end{figure}
-In figure~\ref{fig:}, $\tau_1$ and $\tau_2$ encode two different timeseries of temperature.
-Both these timeseries live in the fiber $F$ which is an $\n \times \m$ array because it is the set of all possible (time, temperature) values.
-The base space $K$ encodes that all datasets in this fiberbundle are 1D continous.
+The fiber bundle in figure~\ref{fig:data_base_space} is a representation of continuous (temperature, time) data.
+The fiber bundle abstraction facilitates seperately representing the time and temperature variables in the fiber $F$ and how these variables are connected in the base space $K$.
+The function $\tau_1$ returns a record (temperature, time) for ever point on the interval $K$ and there can be many $\tau$ functions in the fiber bundle.
 
@@
 \end{equation}
-which is $U$ restricted to objects of type $T$ bound to variable name $c$. Given $\U_{\sigma}$, the fiber for a one variable dataset is
+which is $U$ restricted to objects of type $T$ bound to variable name $c$. Given $U_{\sigma}$, the fiber for a one variable dataset is
 \begin{equation}
@@
 \begin{equation}
-F = U_{T_1}\times \ldots U{T_i} \ldots\timesU_{T_n}
+F = U_{T_1}\times \ldots U{T_i} \ldots\times U_{T_n}
 \end{equation}
@@
 \begin{equation}
-V_i = (c_i,\, T_i,\, U_{T_i}\)
+V_i = (c_i,\, T_i,\, U_{T_i})
 \end{equation} 
@@
     \label{fig:data_fiber_example}
-    \caption{}}.
+    \caption{}.
 \end{figure}
+
 For example, the data in figure~\ref{fig:fiber_example_plane} is a set of temperature measurements in \textdegree C and the times those measurements were taken at.
@@
 Steven's defined the measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions permissible on the measurments; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-For example, given
+
+For example, given  $V_0 = (temperature,\, \texttt{float},\, \mathbb{R})$ with
 \begin{itemize}
-    \item temperature $x \in \mathbb{R}$
-    \item monoid addition operator $\ast = +$
-    \item with monoid operations: $f: x\mapsto x + c_1 $, $g: x\mapsto x + c_2$
-    \item monoid action operator $\bullet = \circ$
+    \item monoid operator addition $\ast = +$
+    \item monoid operations: $f: x\mapsto x + c_1 $, $g: x\mapsto x + c_2$
+    \item monoid action operator composition $\bullet = \circ$
 \end{itemize}
@@
 \end{equation}
+where $c_1, c_2$ are intervals $x_i-x_j$.
 

### 2021-01-25 draft/sections/math.tex
first example of total fiber that I'm gonna throw out

@@
 \section{Topological Artist Model}
-
+\note{can change this all to 3rd person passive later}\\
 In this section we introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.
@@
 \begin{figure}[ht!]
-    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
-    \label{fig:data_base_space}
+    \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
+
     \caption{ The fiber $F$ is a continuous plane encoding the set of all possible (time, temperature) values.
@@
 The continuous function $\tau_1$ returns a record in each fiber for each point on $K$; the set of all values $\tau_1$ returns is a dataset that lies in this fiberbundle.
-\textcolor{MidnightBlue}{remove one of the sections from this figure}}
+\note{remove one of the sections from this figure/maybe kill this example}}
+    \label{fig:data_base_space} 
 \end{figure}

### 2021-01-25 draft/sections/math.tex
fiber w/ 2D dimensions

@@
 We build on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989} because fiber bundles are mathematical structures that are flexible enough express all the types of data described in section~\ref{sec:intro_data_structure}.
-The advantage of fiber bundles is they provide a way to encode the continuity in a dataset without making assumptions as to what that continuity is.
-In turn this representation of continuity can then be used to keep track of how the data fits together, for example if a visualization of a very large dataset calls for parallelization.
 
@@
 
-\begin{figure}[ht!]
-    \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
-
-    \caption{ The fiber $F$ is a continuous plane encoding the set of all possible (time, temperature) values.
-The base space $K$ expresses that the fibers lie in continous space and therefore the data is continuous.
-The mapping $\pi:E\rightarrow K$ from total space to base space is what defines this structure as a fiber bundle.
-The continuous function $\tau_1$ returns a record in each fiber for each point on $K$; the set of all values $\tau_1$ returns is a dataset that lies in this fiberbundle.
-\note{remove one of the sections from this figure/maybe kill this example}}
-    \label{fig:data_base_space} 
-\end{figure}
-The fiber bundle in figure~\ref{fig:data_base_space} is a representation of continuous (temperature, time) data.
-The fiber bundle abstraction facilitates seperately representing the time and temperature variables in the fiber $F$ and how these variables are connected in the base space $K$.
-The function $\tau_1$ returns a record (temperature, time) for ever point on the interval $K$ and there can be many $\tau$ functions in the fiber bundle.
-
 \subsubsection{Variables: Fiber Space $F$}
@@
 The fiber is a topological space that is the set of possible values of the data; the values themselves can be any dimension and type and have any continuity.
-We use Spivak's description of simplicial database schemas \cite{spivakSIMPLICIALDATABASES} as the basis of our fiber space because he binds the values in the fiber to variable names and types.
-Spivak constructs a set $U$ that is the disjoint union of all possible objects of the types $T\in DT$ that are in the datasets embedded in the fiber bundle.
+We use Spivak's description of simplicial database schemas \cite{spivakSIMPLICIALDATABASES} as the basis of our fiber space because he binds the components of the fiber to variable names and types.
+Spivak constructs a set $U$ that is the disjoint union of all possible objects of types $\{T_0, \ldots, T_n\} \in DT$ where $DT$ are the data types of the variables in the dataset.
 He then defines the single variable set $U_\sigma$
@@
 \end{equation} 
-we keep track of its name $c$, type $T$, and component in the fiber $U_{T}$.
+we keep track of its name $c$, type $T$, and values $U_{T}$.
 
-\begin{figure}[ht!]
+\begin{figure}[H]
     \begin{subfigure}{.5\textwidth}
         \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
+        \caption{}
         \label{fig:fiber_example_plane}
@@
         \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
+        \caption{}
         \label{fig:fiber_example_cube}
     \end{subfigure}
+    \caption{These two datasets have the same base space $K$, but figure~\ref{fig:fiber_example_plane} has fiber  $F=\mathbb{R}\times\mathbb{R}^+$ which is (time, temperature) while figure~\ref{fig:fiber_example_cube} has fiber $\mathbb{R}^{+}\times\mathbb{R}^2$ which is (time, wind=(speed, direction))}
     \label{fig:data_fiber_example}
-    \caption{}.
 \end{figure}
 
-For example, the data in figure~\ref{fig:fiber_example_plane} is a set of temperature measurements in \textdegree C and the times those measurements were taken at.
-Temperature values are floating point numbers that come from the set of real numbers $U_{float} = \mathbb{R}$ and times that are always positive $U_{datetime}= \mathbb{R}^+$, so the fiber is
+For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree C temperature measurements taken at those times.
+Time is a positive number ot type \textt{detetime} which can be resolved to positive floats $U_{\texttt{datetime}}= \mathbb{R}^+$.
+Temperature values are real numbers $U_{\texttt{float}} = \mathbb{R}$.
+The fiber is
 \begin{equation}
-    F = \mathbb{R} \times \mathbb{R}^{+}
+    F =  \mathbb{R}^{+} \times \mathbb{R} 
 \end{equation} 
-where $V_0 = (temperature,\, \texttt{float},\, \mathbb{R})$ and $V_1 = (time,\, \texttt{datetime},\, \mathbb{R}^+)$
-In figure~\ref{fig:fiber_example_cube}, a third variable wind is added to the dataset.
-This wind variable is of type \texttt{wind} and has two components speed and direction $\{(m,d) \in \mathbb{R}^{2} \mid  0\leq m,\, 0 \leq d \leq 360\}$ .
-The new variable is added to the fiber
+where $V_0 = (time,\, \texttt{datetime},\, \mathbb{R}^+)$ and $V_1 = (temperature,\, \texttt{float},\, \mathbb{R})$.
+In figure~\ref{fig:fiber_example_cube}, temperature is replaced with wind.
+This wind variable is of type \texttt{wind} and has two components speed and direction $\{(s,d) \in \mathbb{R}^{2} \mid  0\leq s,\, 0 \leq d \leq 360\}$ .
+Therefore, the fiber is
 \begin{equation}
-    F = \mathbb{R} \times \mathbb{R} \times \mathbb{R}^2
+    F = \mathbb{R}^{+} \times \mathbb{R}^2
 \end{equation} 
-such that $V_2 = (wind,\, \texttt{wind},\, \mathbb{R}^{2})$
+such that $V_1 = (wind,\, \texttt{wind},\, \mathbb{R}^{2})$
 
@@
 We formally describe the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
-We generalize this set of allowable transformations to the monoid actions $M_i$ on $F_i$.
+Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of these transformations: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
+
+We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$.
 A monoid \cite{Monoid2021} $M_i$ is a set with an associative binary operator $\ast:M_i \times M_i\rightarrow M_i$.
@@
 
-Steven's defined the measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions permissible on the measurments; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-
 For example, given  $V_0 = (temperature,\, \texttt{float},\, \mathbb{R})$ with
@@
 
-\subsubsection{continuity: Base Space $K$} 
-%%fold butlder
+\subsubsection{Continuity: Base Space $K$} 
 \label{sec:data_base}
-The base space $K$ provides a way to encode how the records are connected to each other; for example the GHCN data can be expressed as discrete observations, many timeseries, maps, a network of stations, etc.
+The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the basespace $K$ without making assumptions as to what that continuity is.
+In turn this representation of continuity can then be used to keep track of how the data fits together, for example if a visualization of a very large dataset calls for parallelization.
 
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics[width=.5\textwidth]{figures/math/k_different_types.png}
+    \caption{The topological base space $K$ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
     \label{fig:base_space_types}
-    \caption{The topological base space $K$ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
+\end{figure}
+As illustrated in figure~\ref{fig:base_space_types}, $K$ is akin to an indexing space into $E$ that describes the structure of $E$.
+$K$ can have any number of dimensions and be continuous or discrete.
+
+\begin{figure}[H]
+    \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
+    \caption{The base space $E$ is divided into fiber segments $F$. The base space $K$ acts as an index into the records in the fibers.
+    \note{this figure might be good all the way up top to lay out the components of fb}}
+    \label{fig:base_space_div}
 \end{figure}
 
-$K$ has no knowledge of the records.
-As illustrated in figure~\ref{fig:base_space_div}, $K$ is more of an indexing space into $E$ that describes the structure of $E$ and itself can be any number of dimensions.
-
-\begin{figure}[ht!]
-    \label{fig:base_space_div}
-    \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
-    \caption{ The base space $E$ is divided into fiber segments $F$. The base space $K$ acts as an index into the fibers.}
-\end{figure}
-
-Formally $K$ is the quotient space \cite{QuotientSpaceTopology2020} of $E$, meaning it is the finest space such that every $k \in K$ has a corresponding fiber $F_k$ in $E$.
+Formally $K$ is the quotient space \cite{QuotientSpaceTopology2020} of $E$, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $k \in K$ has a corresponding fiber $F_k$ in $E$\cite{QuotientSpaceTopology2020}.
 In figure~\ref{fig:base_space_div}, $E$ is a rectangle divided by vertical fibers $F$, so the minimal $K$ for which there is always a mapping $\pi: F\rightarrow K$ is the line.
@@
 
-which is a decomposition of $F$. The $K$ remains the same because the connectivity of records does not change just because there are fewer elements in each record.
+which is a decomposition of $F$. The $K$ remains the same because the connectivity of records 
+does not change just because there are fewer elements in each record.
 
-\subsubsection{Values: Sections $\tau$}
+\begin{figure}[ht!]
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
+        %% add box around neighboring P and Map
+        \caption{}
+        \label{fig:base_example_discrete}
+    \end{subfigure}
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/math/temp_2k.png}
+        \caption{}
+        \label{fig:base_example_continuous}
+    \end{subfigure}
+    \caption{These two datasets have the same (time, temperature) fiber, but in figure~\ref{fig:base_example_line} the records are discrete, while in figure~\ref{fig:base_example_plane} the records are sampled from the continuous interval $K$
+    \note{revamp figure: F=Plane, k1 = dots, k2=line}}
+    \label{fig:base_example}
+\end{figure}
+
+The datasets in figure~\ref{fig:base_example} have the same fiber of (temperature, time) which is $F=\mathbb{R}\times \mathbb{R}^+$.
+In figure~\ref{fig:base_example_discrete} the fibers lie over discrete $K$ such that the records in the datasets in the fiber bundles are discrete.
+The same fiber in figure~\ref{fig:base_example_continuous} lies over a continuous interval $K$ such that the records are samples from a continuous function defined on $K$.
+
+\subsubsection{Data: Sections $\tau$}
 \label{sec:data_section}
-\begin{figure}[ht!]
+\begin{figure}[H]
     \label{fig:base_space_div}
@@
 \subsubsection{{Example}}
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics[width=.4\linewidth]{figures/math/render.png}
@@
 \label{sec:artist_nu}
-$\nu: E \rightarrow V$ is an equivariant map such that there is a homomorphism from left monoid actions on $E_{i}$ to left monoid actions on $V_{i}$ where $i$ identifies a field in the fiber.
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/equivariance_nu.png}
+    \caption{An equivariant $\nu$ preserves monoid actions.
+Here $\nu$ maps the words to the emojis and is constructed such that a permutation of the words is reflected in a permutation of the symbols.
+\note{will switch to a figure table thing but this is the idea}}
+    \label{fig:artist_nu}
+\end{figure}
+The expectation of a visual transformation is that it will convert data to visual variables in a consistent manner.
+For example in figure~\ref{fig:artist_nu}, the $\nu$ maps words to emojies.
+Our constraint on $\nu$ is that if we shuffle the words, the emojis should be shuffled such that the map defined by $\nu$ is still valid.
+Formally, this means that $\nu: E \rightarrow V$ is an equivariant map such that there is a homomorphism from left monoid actions on $E_{i}$ to left monoid actions on $V_{i}$ where $i$ identifies a field in the fiber.
 $E_i$ and $V_{i}$ each contain a set of values as defined in $F$ and $P$ respectively.
@@
 \paragraph{Example: Ordering}
-To preserve ordering of elements in $E_i$, $\nu$ must be a monotonic function such that given $e_1, e_2 \in E_{i}$
+To preserve ordering of elements, $\nu$ must be a monotonic function such that given $e_1, e_2 \in E_{i}$
 \begin{equation}
@@
 \paragraph{Example: Translation}
-According to Stevens, interval data is a set with general linear group actions \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
-Position is a visual variable that can support translation
+A $\nu$ is equivariant under translation monoid actions if 
 \begin{equation}
@@
 \label{sec:artist_q}
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/math/path_of_q}
@@
 
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
@@
 \label{sec:artist_example_scatter}
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/math/scatter.png}
@@
 \label{sec:artist_example_line}
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/math/line.png}
@@
 \label{sec:artist_example_heatmap}
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/math/heatmap.png}
@@
 
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics{figures/math/simplex.png}
@@
 \subparagraph{Example: Mobius Strip}\mbox{}\\
-\begin{figure}[ht!]
+\begin{figure}[H]
     \label{fig:data_base_transition}
@@
 \subparagraph{Example: Torus}\mbox{}\\
-\begin{figure}[ht!]
+\begin{figure}[H]
     \label{fig:triangle_torus}

### 2021-01-25 draft/sections/scraps.tex
fiber w/ 2D dimensions

@@
 such that sections with the constraint that time is monotonic return a timeseries.
+
+\begin{figure}[ht!]
+    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
+    \label{fig:data_monoid_rotation}
+    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.
+    %% dataset that is symmetric, fiber is x,y, rotatation action
+    %% rotate + plot or plot + rotate
+    %%emoji shuffle 
+    }  
+\end{figure}
+
+\begin{figure}[ht!]
+    \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
+
+    \caption{ The fiber $F$ is a continuous plane encoding the set of all possible (time, temperature) values.
+The base space $K$ expresses that the fibers lie in continous space and therefore the data is continuous.
+The mapping $\pi:E\rightarrow K$ from total space to base space is what defines this structure as a fiber bundle.
+The continuous function $\tau_1$ returns a record in each fiber for each point on $K$; the set of all values $\tau_1$ returns is a dataset that lies in this fiberbundle.
+\note{remove one of the sections from this figure/maybe kill this example}}
+    \label{fig:data_base_space} 
+\end{figure}
+The fiber bundle in figure~\ref{fig:data_base_space} is a representation of continuous (temperature, time) data.
+The fiber bundle abstraction facilitates seperately representing the time and temperature variables in the fiber $F$ and how these variables are connected in the base space $K$.
+The function $\tau_1$ returns a record (temperature, time) for ever point on the interval $K$ and there can be many $\tau$ functions in the fiber bundle.

### 2021-01-25 draft/sections/math.tex
sections

@@
 \label{sec:data_monoid}
-We formally describe the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.
-Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of these transformations: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
+After specifying the variables $V$ in the dataset, we next describe ways in which we can transform the variables by identifying the monoid actions $M_i$ on the $F_i$.
+We use monoids as the abstraction because they encode composibilty, which maps well to the data transformation process in a software library \cite{yorgeyMonoidsThemeVariations}.
 
-We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$.
 A monoid \cite{Monoid2021} $M_i$ is a set with an associative binary operator $\ast:M_i \times M_i\rightarrow M_i$.
 A monoid has an identity element $e\in M_i$ such that $e\ast a= a \ast e = a$ for all $a \in M_i$.
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $U_{T_i}$ with an action $\bullet: M\times U_{T_i} \rightarrow U_{T_i}$ with the properties:
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $F_i$ with an action $\bullet: M\times F_i \rightarrow F_i$ with the properties:
 \begin{align*}
-    \textbf{associativity}\;& \text{for all } f,g \in M_i \text{ and } x\in U_{T_i}, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-    \textbf{identity}\;& \text{for all } x\in U_{T_i}, e\in M_i,  e\bullet x = x 
+    \textbf{associativity}\;& \text{for all } f,g \in M_i \text{ and } x\in F_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
+    \textbf{identity}\;& \text{for all } x\in F_i, e\in M_i,\,  e\bullet x = x 
 \end{align*}
@@
 \begin{equation}
-        V_i = (c_i,\, T_i,\, U_{T_i}\, M_i)
+        V_i = (c_i,\, T_i,\, F_i\, M_i)
 \end{equation}
-to include the monoid actions that act on $U_{T_i}$.
+to include the monoid actions that act on $F_i$.
 
-For example, given  $V_0 = (temperature,\, \texttt{float},\, \mathbb{R})$ with
+Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
+For example, given  $V_0 = (temperature,\, \txtett{float},\, \mathbb{R})$ which is interval data:
 \begin{itemize}
@@
 \label{sec:data_section}
-\begin{figure}[H]
-    \label{fig:base_space_div}
-    \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
-    \caption{ A section $\tau$ is a mapping from $K$ to $F$ that returns a record for each $k \in K$.}
-\end{figure}
-
-The sections of the fiber bundle are maps from points $k$ on $K$ to records $r$ in $F$.
-When we map our data to visual variables, the $\tau$ returns the set of values that are mapped together into a single graphic element.
-The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$
+While the fiber and base space describe the general structure of all data that lives in the fiber bundle, the sections $\tau: K\rightarrow E$ define the datasets that live in the fiber.
+We generalize Spivak's section as table $\tau$\cite{spivakSIMPLICIALDATABASES} to any sort of structured dataset such that
 \begin{equation}
@@
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[d, "\pi"']           \\
+        F \arrow[r, hook] & E \arrow[d, "\pi"']           
         & K \arrow[u, "\tau"', bend right]
@@
 \end{equation}
-such that there is $\pi(\tau(k)) = k$ map back to $K$.
-
-A fiber bundle is trivial if $E = K\times F$\cite{rowlandFiberBundle,FiberBundle2020}.
-All fiber bundles are locally trivial, which means if you restrict $E$ over a small enough neighborhood $U \subset K$ then it is a locally trivial bundle over $U$.
-When the fiber bundle is locally trivial, then
+such that there is $\pi(\tau(k)) = k$ map back to $K$. There can be many sections $\tau$, and the space of global sections is $\Gamma(E)$. For a trivial fiber bundle
 \begin{equation}
@@
 \end{equation}
-the section $\tau$ is the record in the $F$. Because we can decompose the bundle and the fiber, we can formulate $\tau$ as 
+where $g: K \rightarrow F$ is the index function into the fiber. Because we can decompose the bundle and the fiber, we can formulate $\tau$ as 
 \begin{align}
@@
 
-\paragraph{Example?}
-For GHCN, the $\tau$ functions return a table in a database with the schema defined in $F$ and the continuity defined in $K$: this means the sections can be defined such that all values have the same year or location, e.g each table returned by a $\tau$ is for a different station.
+\begin{figure}[ht!]
+    \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
+    \caption{ Fiber (time, temperature) with an interval $K$ basespace.
+The sections $\tau_1$ and $\tau_2$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of temperature values.
+They are included in the global set of sections  $\tau_1, \tau_2 \in \Gamma(E)$}
+    \label{fig:data_sections}
+\end{figure}
+
+In the example in figure~\ref{fig:data_sections}, the fiber is $\mathbb{R}^+\times \mathbb{R}$ as described in figure~\ref{fig:data_fiber_example} and the base space is the interval $K$.
+The section $\tau_1$ resolves to a series of monotonically increasing in time records of temperature.
+Section $\tau_2$ returns a different timeseries of temperature values.
+Both sections are included in the global set of sections $\tau_1, \tau_2 \in \Gamma(E)$.
 
@@
 %%% sheaf is how we break up the base space
+
+all fiber bundles are locally trivial, which means that $E$ restricted over a small enough neighborhood $U \subset K$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre},
 Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(E)$
@@
 Our constraint on $\nu$ is that if we shuffle the words, the emojis should be shuffled such that the map defined by $\nu$ is still valid.
-Formally, this means that $\nu: E \rightarrow V$ is an equivariant map such that there is a homomorphism from left monoid actions on $E_{i}$ to left monoid actions on $V_{i}$ where $i$ identifies a field in the fiber.
+We construct $\nu: E \rightarrow V$ as an equivariant map such that there is a homomorphism from left monoid actions on $E_{i}$ to left monoid actions on $V_{i}$ where $i$ identifies a field in the fiber.
 $E_i$ and $V_{i}$ each contain a set of values as defined in $F$ and $P$ respectively.

### 2021-01-25 draft/sections/scraps.tex
sections

@@
-\subsubsection{Case Study: GHCN}
-\textcolor{red}{Moved \& walked through b/c was getting clunky to not have terms yet}\\
-
-Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations.
-Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
-% can't make a 1D fiber - can't make a 1D fiber, switch the picture to plane over 0D and 1D and then can talk about disconnected or connected
-
-% break out the mbox/clean up the typesetting
-The fiber bundle model is flexible enough to express some of the many different forms that temperature data can come in.
-
-\begin{figure}[ht!]
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-        %% add box around neighboring P and Map
-        \label{fig:base_example_line}
-    \end{subfigure}
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_2k.png}
-        \label{fig:base_example_plane}
-    \end{subfigure}
-    \label{fig:base_example}
-    \caption{These two datasets have the same fiber of temperature but different base spaces.
-In figure~\ref{fig:base_example_line} the temperature values are 1D continuous, while in figure~\ref{fig:base_example_plane} the temperature values are 2D continuous.}
-\end{figure}
-
-The datasets in figure~\ref{fig:base_example} have identical fibers that encode a set of temperature values.
-In figure~\ref{fig:base_example_line} the temperatures lie on a line such that a section could return a timeseries or a distribution.
-In figure~\ref{fig:base_example_plane}, the temperatures lie on a 2D continuous plane; a section could return a map or contour.
-Because the fiber is 1D, it does not encode metadata
-
-\begin{figure}[ht!]
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
-        \label{fig:fiber_example_plane}
-    \end{subfigure}
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
-        \label{fig:fiber_example_cube}
-    \end{subfigure}
-    \label{fig:fiber_example}
-    \caption{The fiber is expanded to include metadata fields that describe the semantics of $K$.
-In figure~\ref{fig:fiber_example_plane} the fiber is \textrm{temperature} $\times$ \textrm{time} and in figure~\ref{fig:fiber_example_cube} the fiber is \textrm{temperature}$\times$ \textrm{latitude} $\times$ {longitude}}.
-\end{figure}
-
-To encode the metadata, the fiber is expanded as illustrated in figure~\ref{fig:fiber_example}.
-The fiber in figure~\ref{fig:fiber_example_plane} is the cartesian product of the space of possible temperature values in degrees celsius and space of possible time values
-\begin{equation}
-F = [temp_{min}, temp_{max}] \times [time_{min}, time_{max}]
-\label{eq:fiber_plane}
-\end{equation}
-
-while the fiber in figure~\ref{fig:fiber_example_cube} is the cartesian product of temperature, latitude, and longitude
-\begin{equation}
-F = [temp_{min}, temp_{max}] \times [-90, 90] \times [-180, 180]
-\label{eg:fiber_cube}
-\end{equation}
-
-such that $E$ is the space of all possible points in $F$.
-
-\begin{figure}[ht!]
-    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
-    \label{fig:fiber_example_section}
-    \caption{The section $\tau_1$ returns the blue points,  while $\tau_2$ returns the purple points. $\Gamma(E)$ is the set of all sections, including $\tau_1$ and $\tau_2$}.  
-\end{figure}
-Given the fiber described in equation~\ref{eq:fiber_plane}, the sections $\tau_{1}$ and $\tau_{2}$ in figure~\ref{fig:fiber_example_section} return tuples of the form
-\begin{equation}
-\tau(k) = (k, (temperature, time))
-\end{equation}
-such that sections with the constraint that time is monotonic return a timeseries.
-
 \begin{figure}[ht!]

### 2021-01-25 draft/sections/math.tex
i has finished data section again

@@
 \label{sec:data_fiber}
+\note{yes, I know i may need to change this U to like \mathcal{U}, spivak uses U}
 The fiber is a topological space that is the set of possible values of the data; the values themselves can be any dimension and type and have any continuity.
@@
 \label{sec:data_sheaf_stalk}
-%%% sheaf is how we break up the base space
-
-all fiber bundles are locally trivial, which means that $E$ restricted over a small enough neighborhood $U \subset K$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre},
-Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(E)$
+Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(E)$.
+All fiber bundles are locally trivial, which means that $E$ restricted over a small enough neighborhood $U \subset K$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
+The sheaf $O(E)$ is the localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$
 
@@
 \end{equation}
-which is the localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$ pulled back via the inclusion map $\iota: U \rightarrow K$.
+pulled back over the neighborhood $U$ via the inclusion map $\iota: U \rightarrow K$.
 The localized section is the germ $\xi^*\tau$.
 The neighborhood of points $k_i$ surrounding the point $k$ the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
-
-%% the stalk contains all the local information, typically we only need some of the information, such as the firtst or second direivative.
-in practice we only need the jet which is...in practive will extend the fiber /fatten the E , notatoe E' = E + Jet
-The jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} is for writing differential equations on sections of fiber bundles; this information is required for some visual characteristics, such as line thickness.
+While $E$ is only the records in a fiber $F_k$ over a specific $k$, because the stalk includes the neighborhood $U$ it also includes nearby records.
+While this can be useful for visual transforms, often the extra needed information can be found in the jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016}.
+For example, line thickness requires the derivative of the given position to be rendered, which can be found in the fattened $E^\prime=E+\mathcal{J}(E)$
 

### 2021-01-25 draft/sections/math.tex
cleaning up some language about data while I try to figure out why anyone shoudl care about H

@@
     \end{subfigure}
-    \caption{These two datasets have the same (time, temperature) fiber, but in figure~\ref{fig:base_example_line} the records are discrete, while in figure~\ref{fig:base_example_plane} the records are sampled from the continuous interval $K$
+    \caption{These two datasets have the same (time, temperature) fiber.
+In figure~\ref{fig:base_example_line} the total space $E$ is discrete over points $k \in K$, meaning the records in the fiber are also discrete.
+In figure~\ref{fig:base_example_plane} $E$ lies over the continuous interval $K$, meaning the records in the fiber are sampled from a continous space.
     \note{revamp figure: F=Plane, k1 = dots, k2=line}}
@@
 While this can be useful for visual transforms, often the extra needed information can be found in the jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016}.
-For example, line thickness requires the derivative of the given position to be rendered, which can be found in the fattened $E^\prime=E+\mathcal{J}(E)$
+For example, line thickness requires the derivative of the given position to be rendered, which can be found in $E^\prime=E+\mathcal{J}(E)$
 
@@
 \label{sec:graphic}
+
 We define a graphic space $H$ such that we do not have to assume the physical output space of the renderer.

### 2021-01-25 draft/sections/math.tex
fleshing out H

@@
 \label{sec:data_fiber}
-\note{yes, I know i may need to change this U to like \mathcal{U}, spivak uses U}
+\note{yes, I know i may need to change this U to like $\mathcal{U}$, spivak uses U}
 The fiber is a topological space that is the set of possible values of the data; the values themselves can be any dimension and type and have any continuity.
@@
 For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree C temperature measurements taken at those times.
-Time is a positive number ot type \textt{detetime} which can be resolved to positive floats $U_{\texttt{datetime}}= \mathbb{R}^+$.
+Time is a positive number ot type \texttt{detetime} which can be resolved to positive floats $U_{\texttt{datetime}}= \mathbb{R}^+$.
 Temperature values are real numbers $U_{\texttt{float}} = \mathbb{R}$.
@@
 Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-For example, given  $V_0 = (temperature,\, \txtett{float},\, \mathbb{R})$ which is interval data:
+For example, given  $V_0 = (temperature,\, \texttt{float},\, \mathbb{R})$ which is interval data:
 \begin{itemize}
@@
 
-\subsection{Prerender Space $H$}
+\subsection{Graphic: $H$}
 \label{sec:graphic}  
+We can separate the structure of the graphic from the properties of the output format by modeling the space of graphics as a fiber bundle  $(H, S, \pi, D)$.
+As with data, the fiber bundle is for a class of graphics with shared base space $S$ (~\ref{sec:graphic_fiber}) and fiber $D$ (~\ref{sec:graphic_base}) and the sections $\rho$ (~\ref{sec:graphic_section}) encode the fully specified graphics.
 
-We define a graphic space $H$ such that we do not have to assume the physical output space of the renderer.
-This means that the graphic in $H$ can be output to a screen or 3D printed space or a dome.
-We model the prerender space as a fiber bundle (H, S, $\pi$, D).
-$H$ is the predisplay space, with a fiber $D$ dependent on the target display and a base space of $S$.
+\subsubsection{Display $D$}
+\label{sec:graphic_fiber}
+The fiber $D$ is specified in terms of the target display space, for example a 2D screen or 3D printer.
+In this work, we assume a simple 2D opaque image $D=\mathbb{R}^5$ with a record $d \in D$ of the form
+\begin{equation}
+    d = (x, y, R, G, B)
+\end{equation}
+such that a graphic only has 2D position and color. To support overplotting and transparency, the fiber would be $D=R^{7}$ such that $d=(x,y,z, R, G, B, A)$,
 
@@
 %% inverse of b is capital S_b
-\subsubsection{Base space}
-The underlying topology $S$ of a graphic often needs more dimensions than the data topology $K$ because of the specifications of the display space.
-For example, a line plot on a plane (such as a screen or a piece of paper) by necessity needs to also have a thickness so that it is visible, which maps back to a set of connected points in $H$.
-The topology of these connected points is therefore the region $s \subset S$ such that $\xi: S \rightarrow K$ is a deformation retraction \cite{RetractionTopology2020}
+
+\subsubsection{Continuity of the Graphic $S$} 
+\label{sec:graphic_base}
+An assumption of graphical representations is that they match the continuity of the data, for example discrete data is mapped to discrete elements\cite{tufteVisualDisplayQuantitative2001,friendlyBriefHistoryData2008}, but the underlying topology $S$ of a graphic may need more dimensions than the data topology $K$ because of the specifications of the display space.
+Therefore we define the base space mapping from graphic $S$ to data $K$
 \begin{equation}
@@
 
-that goes from a region $s \in S_{k}$ to its associated point $k$, such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
+ as the deformation retraction \cite{RetractionTopology2020} $\xi: S \rightarrow K$ that goes from a region $s \in S_{k}$ to its associated point $k$, such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
 
-\subsubsection{Fiber and Section}
+\begin{figure}
+    \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
+    \caption{The scatter and line graphic base spaces $S$ are larger than $K$ so that they can encode physical aspects of the glyph, such as shape (a circle) or thickness.
+The heatmap base $S$ is the same as $K$ because a heatmap is directly mapping the values in the data to the graphic.
+    \note{add $\alpha, \beta$ coordinates}}
+    \label{fig:graphic_retraction_map}
+\end{figure}
+Figure~\ref{fig:graphic_retraction_map} illustrates some examples of $\xi$.
+The base space $S$ is mapping to a in $D$.
+For example, in the scatter plot, $S$ encodes what the glyph looks like such that each piece of a single glyph maps to a different region on $S$.
+The line plot $S$ can encode position in the $\alpha$ dimension, but the $\beta$ dimension describes
+
+The  line plot $S$ encodes on a plane  by necessity needs to have a thickness to be visible; therefore we construct a rectangular base space $S$ that mirrors the plane on which we can encode the position in one dimension and thickness (should this pixel be colored?) in the other.
+The map $xi$ then goes from a rectangle to
+
+\subsubsection{Renderable Glyph $\rho$}
+\label{sec:graphic_section}
 A section $\rho: S \rightarrow H$ is a mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to visual space as defined in D.

### 2021-01-25 draft/sections/math.tex
can has $S$

@@
 \end{equation}
-such that a graphic only has 2D position and color. To support overplotting and transparency, the fiber would be $D=R^{7}$ such that $d=(x,y,z, R, G, B, A)$,
-
-%%define a region b = [top, bottom, right, left]
-%% inverse of b is capital S_b
+such that a graphic only has 2D position and color.
+To support overplotting and transparency, the fiber would be $D=R^{7}$ such that $d=(x,y,z, R, G, B, A)$.
+The location coordinates $x$ and $y$ are defined in terms of the display, while the \textrm{R,G,B} values are filled in via a lookup on $S$.
 
@@
 \label{sec:graphic_base}
-An assumption of graphical representations is that they match the continuity of the data, for example discrete data is mapped to discrete elements\cite{tufteVisualDisplayQuantitative2001,friendlyBriefHistoryData2008}, but the underlying topology $S$ of a graphic may need more dimensions than the data topology $K$ because of the specifications of the display space.
+An assumption of graphical representations is that they match the continuity of the data\cite{tufteVisualDisplayQuantitative2001,friendlyBriefHistoryData2008}, but the underlying topology $S$ of a graphic may need more dimensions than the data topology $K$ so that the glyph can be defined in $D$.
 Therefore we define the base space mapping from graphic $S$ to data $K$
@@
  as the deformation retraction \cite{RetractionTopology2020} $\xi: S \rightarrow K$ that goes from a region $s \in S_{k}$ to its associated point $k$, such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
+While dimensions can be added to $S$, it retains the same continuity as $K$.
 
@@
 \end{figure}
-Figure~\ref{fig:graphic_retraction_map} illustrates some examples of $\xi$.
-The base space $S$ is mapping to a in $D$.
-For example, in the scatter plot, $S$ encodes what the glyph looks like such that each piece of a single glyph maps to a different region on $S$.
-The line plot $S$ can encode position in the $\alpha$ dimension, but the $\beta$ dimension describes
 
-The  line plot $S$ encodes on a plane  by necessity needs to have a thickness to be visible; therefore we construct a rectangular base space $S$ that mirrors the plane on which we can encode the position in one dimension and thickness (should this pixel be colored?) in the other.
-The map $xi$ then goes from a rectangle to
+In figure~\ref{fig:graphic_retraction_map} each disk $S_k$ indexes how the records in $D$ are held together to generate a single glyph that is the visual representation of a single point in $F_k$.
+For the line, the region $\beta$ over a point $\alpha_i$ specifies the thickness of the line in $D$ for the corresponding section on the interval $K$.
+The heatmap has the same continuity in data space and graphic space such that no extra dimensions are needed.
 
@@
 A section $\rho: S \rightarrow H$ is a mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to visual space as defined in D.
+
+We define a section of the displayed graphic as a region on D defined by the bounding box $b=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$
 

### 2021-01-26 draft/sections/math.tex
finished out graphics section

@@
 \begin{equation}
-    \label{eq:section}
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[d, "\pi"']           
+        F \arrow[r, hook] & E \arrow[d, "\pi"']              \\
                           & K \arrow[u, "\tau"', bend right]
@@
 \begin{equation}
-    d = (x, y, R, G, B)
+    d = (x, y, r, g, b)
 \end{equation}
 such that a graphic only has 2D position and color.
-To support overplotting and transparency, the fiber would be $D=R^{7}$ such that $d=(x,y,z, R, G, B, A)$.
-The location coordinates $x$ and $y$ are defined in terms of the display, while the \textrm{R,G,B} values are filled in via a lookup on $S$.
+To support overplotting and transparency, the fiber would be $D=R^{7}$ such that $d=(x, y, z, r, g, b, a)$.
+The location coordinates $x$ and $y$ are defined in terms of the display, while the $(r,g,b)$ values are filled in via a lookup on $S$.
 
@@
 \label{sec:graphic_section}
-A section $\rho: S \rightarrow H$ is a mapping from a region $s$ on a mathematical encoding of the image to a region $xy$ on the screen that the renderer then maps to visual space as defined in D.
 
-We define a section of the displayed graphic as a region on D defined by the bounding box $b=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$
+A section $\rho: S\rightarrow H$ defines a piece of the graphical representation of the data.
+The physical display coordinates in $D$ are not equivalent to the mathematical coordinates in $S$ such that many points $s \in S$ might map to a single $d\in D$ and many $d$ might go into a visible section of a displayed graphic.
+Therefore, we define the bounding box $b=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ in the display space $D$ that is a piece of the rendered graphic.
 
-\paragraph{Example}
-For a physical screen display, we can consider a predisplay space that is a trivial fiber bundle $H=\mathbb{R}^{5}\times S$ such that $\rho$ is
-\begin{equation}
-    \rho(s)  = (D_x(s), D_y(s), D_r(s), D_g(s), D_b(s))
-    \label{eq:rho}
-\end{equation}
+We inverse map this bounding box $S_b =\rho^{-1}(b)$ into a region $S_b \subset S$  
+\begin{align}
+    R_b &= \iint\limits_{S_b} \rho_r(s)ds^{2}\\
+    G_b &= \iint\limits_{S_b} \rho_g(s)ds^{2}\\
+    B_b &= \iint\limits_{S_b} \rho_b(s)ds^{2}
+\end{align}
+to then compute the remaining $(R, G, B)$ components of the bounded display region $D_b$.
 
-To draw an image, a region, $H$ is inverse mapped into a region $s \in S$ where %%% 
-\begin{equation}
-s = \rho^{-1}_{\tiny{XY}}(xy)
-\end{equation}
-such that the rest of the fields in $\mathbb{R}^{7}$ are then integrated over $s$ to yield the remaining fields
-%% redefining notation 
-\begin{align}
-    R_b &= \iint\limits_s D_r(s)ds^{2}\\
-    G_b &= \iint\limits_s D_g(s)ds^{2}\\
-    B_b &= \iint\limits_s D_b(s)ds^{2}
-\end{align}
-
-Here we assume a single opaque 2D image such that the $z$ and $alpha$ fields can be omitted. To support overplotting and transparency, we can consider $D=R^{7}$
-
-\subsubsection{{Example}}
-\begin{figure}[H]
-    \includegraphics[width=.4\linewidth]{figures/math/render.png}
-    \caption{}
-    \label{fig:render}
+\begin{figure}
+    \includegraphics[width=\textwidth]{figures/math/graphic_render.png}
+    \caption{To render a graphic, a region $b$ is selected in the display space.
+This region is in coordinates $D$ so the inverse mapping $\rho^{-1}(b)$ maps to an equivalent region $S_b \subset S$.
+We then take the sections $\rho(S_b)$ which are the collection of records $d$ that combine together to make the piece of the graphic in $b$.}
+    \label{fig:graphic_rho_lookup}
 \end{figure}
-
-As illustrated in figure~\ref{fig:render}, words.
+As shown in figure~\ref{fig:graphic_rho_lookup}, the output space queries into the graphic bundle to render the image.
+We select a region $b$ in the output space, inverse map the region into $S_b \subset S$, then take the section $\rho$ over the region $S_b$ which yields a set of corresponding records $d$.
+These records specify the $(r, g, b)$ values corresponding to the coordinates in $b$.
+These records are then combined to render the piece of the graphic in $b$.
 

### 2021-01-26 draft/sections/math.tex
artist picture

@@
 The fiber $D$ is specified in terms of the target display space, for example a 2D screen or 3D printer.
-In this work, we assume a simple 2D opaque image $D=\mathbb{R}^5$ with a record $d \in D$ of the form
+In this work, we assume a simple 2D opaque image $D=\mathbb{R}^5$ with an element $d \in D$ of the form
 \begin{equation}
@@
 
-In figure~\ref{fig:graphic_retraction_map} each disk $S_k$ indexes how the records in $D$ are held together to generate a single glyph that is the visual representation of a single point in $F_k$.
+In figure~\ref{fig:graphic_retraction_map} each disk $S_k$ indexes how elements in $D$ are held together to generate a single glyph that is the visual representation of a single point in $F_k$.
 For the line, the region $\beta$ over a point $\alpha_i$ specifies the thickness of the line in $D$ for the corresponding section on the interval $K$.
@@
 This region is in coordinates $D$ so the inverse mapping $\rho^{-1}(b)$ maps to an equivalent region $S_b \subset S$.
-We then take the sections $\rho(S_b)$ which are the collection of records $d$ that combine together to make the piece of the graphic in $b$.}
+We then take the sections $\rho(S_b)$ which are the elements $\{d_0, \ldots, d_n\}$ that make the piece of the graphic in $b$.}
     \label{fig:graphic_rho_lookup}
@@
 As shown in figure~\ref{fig:graphic_rho_lookup}, the output space queries into the graphic bundle to render the image.
-We select a region $b$ in the output space, inverse map the region into $S_b \subset S$, then take the section $\rho$ over the region $S_b$ which yields a set of corresponding records $d$.
-These records specify the $(r, g, b)$ values corresponding to the coordinates in $b$.
-These records are then combined to render the piece of the graphic in $b$.
+We select a region $b$ in the output space, inverse map the region into $S_b \subset S$, then take the section $\rho$ over the region $S_b$.
+The section yields the set of elements $\{d_0, \ldots, d_n\}$ that specify the $(r, g, b)$ values corresponding to the coordinates in $b$.
 
@@
 \label{sec:artist}
+The artist is the composite function that converts data into graphics; its name is taken from the analogous part of the existing Matplotlib architecture.
+The binding of $\xi:S \rightarrow K$ keeps track of which data goes to which part of the graphic such that the artist can be trivially parallelized or formulated as a delayed computer graph or otherwise not need to immediately work on all the data.
 
-%% E' not E
-In this section we will define the artist as a mapping from a sheaf $\mathcal{O}(E)$  to $\mathcal{O}(H)$. 
+The artist is a mapping from $E$ padded with data from $\mathcal{J}(E)$
 \begin{equation}
-    A: \mathcal{O}(E) \rightarrow \mathcal{O}(H)
+    A: E^\prime \rightarrow \Gamma(H)
 \end{equation}
-
+\note{this is soup}
+to a graphic in $\Gamma(H)$.
 The artist decomposes to mapping data to visual $\nu:E\rightarrow V$, then  compositing $V$ pulled back along $\xi$ to $\xi^*V$ to a visual mark in prerender space $Q:\xi^*V\rightarrow H$.
@@
 
+\begin{figure}
+    \includegraphics[width=\textwidth]{figures/math/artist_stages.png}
+    \caption{The artist $A$ converts data into the visual variables in $V$ and then composes those visual variables into a graphic in $H$}
+    \label{fig:artist_stages}
+\end{figure}
 \subsubsection {Example: Matplotlib Visual Fiber}

### 2021-01-26 draft/sections/math.tex
rewrote nus

@@
 
-\subsubsection{Display $D$}
+\subsubsection{Abstract Display $D$}
 \label{sec:graphic_fiber}
-The fiber $D$ is specified in terms of the target display space, for example a 2D screen or 3D printer.
-In this work, we assume a simple 2D opaque image $D=\mathbb{R}^5$ with an element $d \in D$ of the form
+The fiber $D$ is specified in terms of the target display space, for example a 2D screen or 3D printer. In this work, we assume a 2D opaque image $D=\mathbb{R}^5$ with elements in $D$ of the form
 \begin{equation}
@@
 \end{equation}
-such that a graphic only has 2D position and color.
-To support overplotting and transparency, the fiber would be $D=R^{7}$ such that $d=(x, y, z, r, g, b, a)$.
+such that a rendered graphic only consists of 2D position and color.
+To support overplotting and transparency, the fiber could be $D=R^{7}$ such that $d=(x, y, z, r, g, b, a)$.
 The location coordinates $x$ and $y$ are defined in terms of the display, while the $(r,g,b)$ values are filled in via a lookup on $S$.
@@
 The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels,} in fiber $P$.
-The visual transformer $\nu$ is a set of functions each targeting a different $\mu$
-\begin{equation}
-    \label{eq:nu_expanded}
-    \{\nu_{0}, \ldots, \nu_{n}\}: \{\tau_{0}, \ldots, \tau_{n}\} \mapsto \{\mu_{0}, \ldots, \mu_{n}\}
-\end{equation}
-
-where $\mu_{i}$ are the visual parameters in the assembly function $Q(\mu_{0}, \ldots, \mu_{n})(s) = \rho(s)$.
-
 \begin{figure}
@@
 \end{figure}
-\subsubsection {Example: Matplotlib Visual Fiber}
+\subsubsection {Visual Fiber Bundle}
+
 For example, for Matplotlib \cite{hunterMatplotlib2DGraphics2007}, some of the possible types in $P$ are:
@@
 \end{figure}
-The expectation of a visual transformation is that it will convert data to visual variables in a consistent manner.
-For example in figure~\ref{fig:artist_nu}, the $\nu$ maps words to emojies.
-Our constraint on $\nu$ is that if we shuffle the words, the emojis should be shuffled such that the map defined by $\nu$ is still valid.
-We construct $\nu: E \rightarrow V$ as an equivariant map such that there is a homomorphism from left monoid actions on $E_{i}$ to left monoid actions on $V_{i}$ where $i$ identifies a field in the fiber.
-$E_i$ and $V_{i}$ each contain a set of values as defined in $F$ and $P$ respectively.
-A validly constructed $\nu$ is one where the  diagram
+As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers $\nu$ as the set of conversion functions 
+\begin{equation}
+    \label{eq:nu_expanded}
+    \{\nu_{0}, \ldots, \nu_{n}\}: \{\tau_{0}, \ldots, \tau_{n}\} \mapsto \{\mu_{0}, \ldots, \mu_{n}\}
+\end{equation}
+such that $\nu_i: \tau_i \mapsto \mu_i$ is an equivariant map such that there is a monoid homomorphism from $F_{i}$ to $V_{i}$. A validly constructed $\nu$ is one where the  diagram 
 \begin{equation}
@@
 commutes such that $\nu_i(m_e(E_i)) = m_v(\nu_i(E_i))$.
-
-\paragraph{Example: Ordering}
-To preserve ordering of elements, $\nu$ must be a monotonic function such that given $e_1, e_2 \in E_{i}$
-\begin{equation}
-\text{ if } e_1 \leq e_2 \text{ then } \nu(e_1) \leq \nu(e_2)
-\end{equation}
-
-\paragraph{Example: Translation}
-A $\nu$ is equivariant under translation monoid actions if 
-\begin{equation}
-\nu(x + c) = \nu(x) + \nu(c)
-\end{equation}
-
-\paragraph{Example: Invalid $\nu$}
-Given a transform $t(x) = x+2$, we construct a $\nu$ that always takes data to .5: 
-\begin{equation}
-    \label{eq:nu_equation_bad}
-    \begin{tikzcd}
-        E_1 \arrow[r] \arrow[r, "\lambda:e\mapsto.5"] \arrow[d, "2e"'] & V_i \arrow[d, "2v"] \\
-        E_1 \arrow[r, "\lambda"]                                        & V_1                 
-    \end{tikzcd}
-\end{equation}
-
-This $\nu$ is invalid because the graph does not commute for $t$:
+This equivariance constraint yields guidance on what makes for an invalid transform.
+For example, the single value conversion $\nu_{i}(x) = .5$ does not commute under translation  $t(x) = x+2$
 \begin{align}
-    \nu(t(e)) & \overset{?}{=} t(\nu(e))\\
-    .5 & \overset{?}{=} t(.5)\\
-    .5 & \neq 2*.5
+    \nu(t(e + 2)) & \overset{?}{=} \nu(e) + \nu(2)\\
+    .5 &\neq .5 + .5
 \end{align}
 
-To construct a valid $\nu$, the diagram must commute for all monoid actions on the sets in $E_i, V_i$.
+On the other hand figure~\ref{fig:artist_nu} illustrates a valid $\nu$ mapping from \textbf{Strings} to symbols.
+The group action on these sets is permutation, so shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
+To preserve ordinal and partial order monoid actions, $\nu$ must be a monotonic function such that given $e_1, e_2 \in E_{i}$, $\text{ if } e_1 \leq e_2 \text{ then } \nu(e_1) \leq \nu(e_2)$.
+For interval scale data, $\nu$ is equivariant under translation monoid actions if $\nu(x + c) = \nu(x) + \nu(c)$.
+For ratio data, there must be equivalent scaling $ \nu(xc) = \nu(x)\nu(c)$.
 

### 2021-01-26 draft/sections/math.tex
artist _ fiber _nu

@@
 \label{sec:artist}
-The artist is the composite function that converts data into graphics; its name is taken from the analogous part of the existing Matplotlib architecture.
-The binding of $\xi:S \rightarrow K$ keeps track of which data goes to which part of the graphic such that the artist can be trivially parallelized or formulated as a delayed computer graph or otherwise not need to immediately work on all the data.
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/dar.png}
+    \caption{The artist $A$ converts an section of data $\tau$ into a section of visual equivalents $\mu$ that are then composed into a piece of a glyph defined by the graphic section $\rho$
+    \note{might need to label this as E', V, H}}
+    \label{fig:artist_stages}
+\end{figure}
 
-The artist is a mapping from $E$ padded with data from $\mathcal{J}(E)$
-\begin{equation}
-    A: E^\prime \rightarrow \Gamma(H)
-\end{equation}
-\note{this is soup}
-to a graphic in $\Gamma(H)$.
-The artist decomposes to mapping data to visual $\nu:E\rightarrow V$, then  compositing $V$ pulled back along $\xi$ to $\xi^*V$ to a visual mark in prerender space $Q:\xi^*V\rightarrow H$.
-
+The artist is the function that converts data into graphics; its name is taken from the analogous part of the existing Matplotlib architecture that builds visual elements to pass off to the renderer.
+The artist is a mapping from $E$ padded with data from $\mathcal{J}(E)$ to a graphic that is a section $\rho$ in  $\Gamma(H)$
 \begin{equation}
@@
     \begin{tikzcd}
-        E \arrow[r, "\nu"] \arrow[rd, "\pi"'] & V \arrow[d, "\pi"] & \xi^*V \arrow[r, "Q"] \arrow[d, "\xi^*\pi"'] \arrow[l, "\xi^*"'] & H \arrow[ld, "\pi"] \\
+        E' \arrow[r, "\nu"] \arrow[rd, "\pi"'] & V \arrow[d, "\pi"] & \xi^*V \arrow[r, "Q"] \arrow[d, "\xi^*\pi"'] \arrow[l, "\xi^*"'] & \Gamma(H) \arrow[ld, "\pi"] \\
                                               & K                  & S \arrow[l, "\xi"']                                              &                    
@@
 \end{equation}
-The pullback map $\xi^*$ copies each value in $V$ over $k$ to $s$ in corresponding $S_k$ such that $\xi^*V$ can have multiple values that map to one value in $V$.
+with an intermediate fiber bundle $V$ to hold visual representations and stages
+\begin{enumerate}
+    \item $\xi$ binding the continuity in the graphic to the continuity in the data (~\ref{sec:graphic_base})
+    \item $\nu$ conversion of data into visual characteristics (~\ref{sec:artist_nu})
+    \item $Q$ assembly of visual variables into a glyph (~\ref{sec:artist_q})
+\end{enumerate}
 
+of the visual transformation illustrated in figure~\ref{fig:artist_stages}.
+The functions $\xi$, $\nu$ and $Q$ are constructed such that the input can be a single section on $K$ or $S$, which allows the artist to be trivially parallelized or formulated as a delayed computer graph or otherwise not need to immediately work on all the data.
+
+\subsubsection {Visual Fiber Bundle $V$}
 The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels,} in fiber $P$.
-\begin{figure}
-    \includegraphics[width=\textwidth]{figures/math/artist_stages.png}
-    \caption{The artist $A$ converts data into the visual variables in $V$ and then composes those visual variables into a graphic in $H$}
-    \label{fig:artist_stages}
-\end{figure}
-\subsubsection {Visual Fiber Bundle}
+The fiber space $P$ is defined in terms of the parameters of the visualization specification and the values those parameters resolve to - for example APT\cite{mackinlayAutomatingDesignGraphical1986} or Vega\cite{satyanarayanDeclarativeInteractionDesign2014}.
 
-For example, for Matplotlib \cite{hunterMatplotlib2DGraphics2007}, some of the possible types in $P$ are:
-\begin{table}[ht]
+\begin{table}[H]
     \renewcommand{\arraystretch}{2}
@@
     \end{tabulary}
+    \caption{Some possible components of the fiber $P$ for a visualization implemented in Matplotlib}
     \label{tab:mpl_visual_variable_fiber}
@@
 
-Table~\ref{tab:mpl_visual_variable_fiber} is an example of the visual fiber defined in terms of common parameters to plots in Matplotlib.
-The range of $\mu_{i}$ determine the monoid actions on $\mu_{i}$.
-A section of V $\mu$ is a tuple of visual values that specifies the visual characteristics of a glyph.
-For example, given a fiber of $\{xpos, ypos, color\}$ one section is $\{.5, .5, (255, 20,147)\}$.
-$Q$ determines how this section is applied to a graphic.
+Table~\ref{tab:mpl_visual_variable_fiber} is a sample of the fiber space for Matplotlib \cite{hunterMatplotlib2DGraphics2007}.  
+A section $\mu$ of $V$ is a tuple of visual values that specifies the visual characteristics of a part of the graphic.
+For example, given a fiber of $\{xpos, ypos, color\}$ one possible section could be  $\{.5, .5, (255, 20,147)\}$.
+The $codomain(\nu_i)$ determines the monoid actions on $\mu_i)$.
+These fiber components are implicit in the library, by making them explicit as components of the fiber we can build consistent definitions and expectations of how these parameters behave.
 
@@
 \subsubsection{Assembling Marks}
+\note{HAVE NOT REACHED HERE TO EDIT YET}
 \label{sec:artist_q}
@@
 \label{sec:triangulization}
+\note{HAVE NOT REACHED HERE TO EDIT YET}
 
@@
 \subsubsection{Visual Idioms: Equivalance class of artists}
+\note{HAVE NOT REACHED HERE TO EDIT YET}
 As formulated above, every artist function $A$ has fixed $\nu$ and generates a distinct graphic $\rho$.

### 2021-01-27 draft/sections/math.tex
rewrote intro again

@@
 \section{Topological Artist Model}
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/dar.png}
+    \caption{Visualizaiton is the process of converting data into visual characteristics that are then assembled into a graphic}
+    \label{fig:artist_stages}
+\end{figure}
+
 \note{can change this all to 3rd person passive later}\\
+%% add sentence or two about visualization as structure preserving map
 In this section we introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.
@@
 \begin{equation}
-F = U_{T_1}\times \ldots U{T_i} \ldots\times U_{T_n}
+F = U_{T_1}\times \ldots U_{T_i} \ldots\times U_{T_n}
 \end{equation}
@@
 For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree C temperature measurements taken at those times.
-Time is a positive number ot type \texttt{detetime} which can be resolved to positive floats $U_{\texttt{datetime}}= \mathbb{R}^+$.
+Time is a positive number of type \texttt{detetime} which can be resolved to positive floats $U_{\texttt{datetime}}= \mathbb{R}^+$.
 Temperature values are real numbers $U_{\texttt{float}} = \mathbb{R}$.
@@
 
-\subsubsection{Abstract Display $D$}
+\subsubsection{Idealized Display $D$}
 \label{sec:graphic_fiber}
+% infinite resolution version of D
 The fiber $D$ is specified in terms of the target display space, for example a 2D screen or 3D printer. In this work, we assume a 2D opaque image $D=\mathbb{R}^5$ with elements in $D$ of the form
@@
 \label{sec:graphic_section}
-
+%pick a pixel which is the bounding box
 A section $\rho: S\rightarrow H$ defines a piece of the graphical representation of the data.
@@
 \subsection{Artist}
-\label{sec:artist}
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/dar.png}
-    \caption{The artist $A$ converts an section of data $\tau$ into a section of visual equivalents $\mu$ that are then composed into a piece of a glyph defined by the graphic section $\rho$
-    \note{might need to label this as E', V, H}}
-    \label{fig:artist_stages}
+    \includegraphics[width=\textwidth]{figures/math/path_of_q}
+
+    \caption{The $\nu$ functions convert data $E$ to visual $V$.
+$Q$ assembles the different types of visual parameters $V_{i}$ into a graphic in $H$.
+$Q\circ\mu(\xi^{-1}J)$ forms a visual mark by applying $Q$ to a region mapped to connected components $J \subset K$.
+\note{maybe change to $\tau, \mu, \rho$ }}
+    \label{fig:artist_q}
 \end{figure}
@@
 The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels,} in fiber $P$.
-The fiber space $P$ is defined in terms of the parameters of the visualization specification and the values those parameters resolve to - for example APT\cite{mackinlayAutomatingDesignGraphical1986} or Vega\cite{satyanarayanDeclarativeInteractionDesign2014}.
-
+The fiber space $P$ is defined in terms of the parameters of the visualization specification and the values those parameters resolve to - for example ggplot \cite{wickhamGgplot2ElegantGraphics2016a}, D3\cite{bostockDataDrivenDocuments2011
 \begin{table}[H]
@@
 \subsubsection{Assembling Marks}
-\note{HAVE NOT REACHED HERE TO EDIT YET}
 \label{sec:artist_q}
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/path_of_q}
-    \label{fig:artist_q}
-    \caption{The $\nu$ functions convert data $E$ to visual $V$.
-$Q$ assembles the different types of visual parameters $V_{i}$ into a graphic in $H$.
-$Q\circ\mu(\xi^{-1}J)$ forms a visual mark by applying $Q$ to a region mapped to connected components $J \subset K$.}
-    %%maybe change to \tau, \mu, \rho  
-\end{figure}
 

### 2021-01-27 draft/sections/math.tex
compiles again after lost }

@@
 \section{Topological Artist Model}
+\label{sec:tam}
 \begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/math/dar.png}
-    \caption{Visualizaiton is the process of converting data into visual characteristics that are then assembled into a graphic}
+    \caption{Visualizaiton is the process of mapping data into visual characteristics that are then assembled into a graphic}
     \label{fig:artist_stages}
@@
 \note{can change this all to 3rd person passive later}\\
-%% add sentence or two about visualization as structure preserving map
-In this section we introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.
+Visualization is generally thought of as structure preserving maps from data into graphics, and in this section we formally define that structure and how it is preserved via equivariant maps.
+We can then specify that a faithful visual mapping is structure preserving, and apply these constraints to visualizations we may want to develop or implement.
+We model the data, visual characteristic, and graphic stages of visualization, shown in figure~\ref{fig:artist_stages}, as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
+
+We introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.
 
@@
 In figure~\ref{fig:base_example_line} the total space $E$ is discrete over points $k \in K$, meaning the records in the fiber are also discrete.
-In figure~\ref{fig:base_example_plane} $E$ lies over the continuous interval $K$, meaning the records in the fiber are sampled from a continous space.
+In figure~\ref{fig:base_example_plane} $E$ lies over the continuous interval $K$, meaning the records in the fiber are sampled from a continuous space.
     \note{revamp figure: F=Plane, k1 = dots, k2=line}}
@@
 
-\begin{figure}[ht!]
+\begin{figure}[H]
     \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
@@
 \subsection{Artist}
+\label{sec:artist}
 \begin{figure}[H]
@@
 \subsubsection {Visual Fiber Bundle $V$}
-The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels,} in fiber $P$.
-The fiber space $P$ is defined in terms of the parameters of the visualization specification and the values those parameters resolve to - for example ggplot \cite{wickhamGgplot2ElegantGraphics2016a}, D3\cite{bostockDataDrivenDocuments2011
+The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber $P$.
+The fiber space $P$ is defined in terms of the parameters of the visualization specification and the values those parameters resolve to - for example ggplot \cite{wickhamGgplot2ElegantGraphics2016a} or D3\cite{bostockDataDrivenDocuments2011}.
 \begin{table}[H]

### 2021-01-27 draft/sections/math.tex
switched data notation to standin + cleaned up language

@@
 
-\note{can change this all to 3rd person passive later}\\
+\note{should this be in 3rd person passive?}\\
 Visualization is generally thought of as structure preserving maps from data into graphics, and in this section we formally define that structure and how it is preserved via equivariant maps.
@@
 
-We introduce a mathematical description of the visualization pipeline where artist $\mathscr{A}$ functions transform data space $\mathscr{E}$ to an intermediate representation in a prerendered graphic space $\mathscr{H}$.
+We introduce a mathematical description of the visualization pipeline where artist $\mathscr{\vartist}$ functions transform data space $\mathscr{\dtotal}$ to an intermediate representation in a prerendered graphic space $\mathscr{\gtotal}$.
 
@@
     \label{eq:artist}
-    \mathscr{A}: \mathscr{E} \rightarrow \mathscr{H}
+    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
 \end{equation}
@@
 
-We model data as the fiber bundle $(E,\,B,\,\pi ,\,F)$, where $E$, $F$, and $K$ are topological spaces that encode 
-\begin{itemize}
-\item the properties of the variables in the fiber $F$ (\ref{sec:data_fiber})
-\item the continuity of the records in the base space $K$ (\ref{sec:data_base})
-\item collections of records $\tau$ (\ref{sec:data_section}). 
-\end{itemize}
+We model data as the fiber bundle $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$, where \dtotal, \dfiber, and \dbase are topological spaces that encode 
+\begin{description}
+\item[\dfiber] the properties of the variables in the fiber (\ref{sec:data_fiber})
+\item[\dbase] the continuity of the records in the base space (\ref{sec:data_base})
+\item[\dsection] collections of records (\ref{sec:data_section}). 
+\end{description}
 
-and $E$ is the total space of data that $F$ lives in. The bundle is the projection map $\pi$
+and \dtotal is the total space of data that \dfiber lives in. The bundle is the projection map $\pi$
 \begin{equation}
@@
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
+        \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
     \end{tikzcd}
@@
 
-that binds the variable space $F$ to the base space $K$.
-The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified, because the trivial bundle is $E=K\times F$ such that extra structure in the total space $E$ falls out and discussion can be focused on the fiber and base space.
+that binds the variables \dfiber continuity \dbase.
+The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified, because the trivial bundle is $\dtotal=\dbase\times \dfiber$ such that extra structure in the total space \dtotal falls out and discussion can be focused on the fiber and base space.
 
-\subsubsection{Variables: Fiber Space $F$}
+\subsubsection{Variables: Fiber Space \dfiber}
 \label{sec:data_fiber}
-\note{yes, I know i may need to change this U to like $\mathcal{U}$, spivak uses U}
 The fiber is a topological space that is the set of possible values of the data; the values themselves can be any dimension and type and have any continuity.
 We use Spivak's description of simplicial database schemas \cite{spivakSIMPLICIALDATABASES} as the basis of our fiber space because he binds the components of the fiber to variable names and types.
-Spivak constructs a set $U$ that is the disjoint union of all possible objects of types $\{T_0, \ldots, T_n\} \in DT$ where $DT$ are the data types of the variables in the dataset.
-He then defines the single variable set $U_\sigma$
+Spivak constructs a set \ftotal that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_n\} \in \ftypes$, where \ftypes are the data types of the variables in the dataset.
+He then defines the single variable set \fttype
 \begin{equation}
@@
     \begin{tikzcd}
-    U_{\sigma} \arrow[r] \arrow[d, "\pi_{\sigma}"'] & U \arrow[d, "\pi"] \\
-    C \arrow[r, "\sigma"']                          & \textbf{DT}       
+        \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] & \ftotal \arrow[d, "\pi"] \\
+        \fnames \arrow[r, "\fsection"']                          & \ftypes       
     \end{tikzcd}
 \end{equation}
-which is $U$ restricted to objects of type $T$ bound to variable name $c$. Given $U_{\sigma}$, the fiber for a one variable dataset is
+which is \ftotal restricted to objects of type \ftype bound to variable name \fname. Given \fsection, the fiber for a one variable dataset is
 \begin{equation}
-    F = U_{\sigma(c)} = U_{T} 
+    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
 \end{equation}
-where $\sigma$ is the schema binding variable name $c$ to its datatype $T$.
-A dataset with multiple variables has a fiber that is the cartesian cross product of $U_{\sigma}$ applied to all the columns:
+where \fsection is the schema binding variable name \fname to its datatype \ftype.
+A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
 \begin{equation}
-F = U_{T_1}\times \ldots U_{T_i} \ldots\times U_{T_n}
+F = \ftotal_{\ftype_1}\times \ldots \ftotal_{\ftype_i} \ldots\times \ftotal_{\ftype_n}
+\end{equation}
+which is equivalent to 
+\begin{equation}
+    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
 \end{equation}
 
-which is equivalent to
-
-\begin{equation}
-    F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
-\end{equation}
-
-which allows us to decouple $F$ into its component fibers $F_i$. For every variable
-\begin{equation}
-V_i = (c_i,\, T_i,\, U_{T_i})
-\end{equation} 
-we keep track of its name $c$, type $T$, and values $U_{T}$.
+which allows us to decouple \dfiber into components $\dfiber_i$.
 
@@
     \end{subfigure}
-    \caption{These two datasets have the same base space $K$, but figure~\ref{fig:fiber_example_plane} has fiber  $F=\mathbb{R}\times\mathbb{R}^+$ which is (time, temperature) while figure~\ref{fig:fiber_example_cube} has fiber $\mathbb{R}^{+}\times\mathbb{R}^2$ which is (time, wind=(speed, direction))}
+    \caption{These two datasets have the same base space \dbase, but figure~\ref{fig:fiber_example_plane} has fiber  $\dfiber=\reals\times\reals$ which is (time, temperature) while figure~\ref{fig:fiber_example_cube} has fiber $\realsp\times\reals^2$ which is (time, wind=(speed, direction))}
     \label{fig:data_fiber_example}
@@
 For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree C temperature measurements taken at those times.
-Time is a positive number of type \texttt{detetime} which can be resolved to positive floats $U_{\texttt{datetime}}= \mathbb{R}^+$.
-Temperature values are real numbers $U_{\texttt{float}} = \mathbb{R}$.
+Time is a positive number of type \texttt{datetime} which can be resolved to positive floats $\ftotal_{\texttt{datetime}}= \realsp$.
+Temperature values are real numbers $\ftotal_{\texttt{float}} = \reals$.
 The fiber is
 \begin{equation}
-    F =  \mathbb{R}^{+} \times \mathbb{R} 
+    \ftotal =  \realsp \times \reals 
 \end{equation} 
-where $V_0 = (time,\, \texttt{datetime},\, \mathbb{R}^+)$ and $V_1 = (temperature,\, \texttt{float},\, \mathbb{R})$.
+where the first component $F_0$ is the set of values specified by $(\fname_0=time,\, \ftype_0=\texttt{datetime},\, \fttype=\mathbb{R}^+)$ and $F_1$ is specified by $(\fname_1=temperature,\, \ftype_1=\texttt{float},\, \fttype=\reals)$.
 In figure~\ref{fig:fiber_example_cube}, temperature is replaced with wind.
-This wind variable is of type \texttt{wind} and has two components speed and direction $\{(s,d) \in \mathbb{R}^{2} \mid  0\leq s,\, 0 \leq d \leq 360\}$ .
+This wind variable is of type \texttt{wind} and has two components speed and direction $\{(s,d) \in \reals^{2} \mid  0\leq s,\, 0 \leq d \leq 360\}$ .
 Therefore, the fiber is
 \begin{equation}
-    F = \mathbb{R}^{+} \times \mathbb{R}^2
+    \dfiber = \realsp \times \reals^{2}
 \end{equation} 
-such that $V_1 = (wind,\, \texttt{wind},\, \mathbb{R}^{2})$
+such that $F_1$ is specified by $(\fname_1 = wind,\, \ftype_1=\texttt{wind},\, \fttype=\reals^{2})$
 
@@
 \label{sec:data_monoid}
-After specifying the variables $V$ in the dataset, we next describe ways in which we can transform the variables by identifying the monoid actions $M_i$ on the $F_i$.
+After specifying \dfiber, we next describe the ways in which we can transform the values by identifying the monoid actions \monoid on the \dfiber.
 We use monoids as the abstraction because they encode composibilty, which maps well to the data transformation process in a software library \cite{yorgeyMonoidsThemeVariations}.
 
-A monoid \cite{Monoid2021} $M_i$ is a set with an associative binary operator $\ast:M_i \times M_i\rightarrow M_i$.
-A monoid has an identity element $e\in M_i$ such that $e\ast a= a \ast e = a$ for all $a \in M_i$.
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $F_i$ with an action $\bullet: M\times F_i \rightarrow F_i$ with the properties:
+A monoid \cite{Monoid2021} $\monoid_i$ is a set with an associative binary operator $\ast:\monoid_i \times \monoid_i\rightarrow \monoid_i$.
+A monoid has an identity element $e\in \monoid_i$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid_i$.
+A left monoid action \cite{SemigroupAction2021,ActionNLab} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
 \begin{align*}
-    \textbf{associativity}\;& \text{for all } f,g \in M_i \text{ and } x\in F_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-    \textbf{identity}\;& \text{for all } x\in F_i, e\in M_i,\,  e\bullet x = x 
+    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
+    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
 \end{align*}
-As with the fiber $F$, the total monoid space $M$ is the cartesian product
+As with the fiber \dfiber, the total monoid space \monoid is the cartesian product
 \begin{equation}
-M = M_{0} \times \ldots \times M_{i}\times \ldots \times\ldots M_{n}
+\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
 \end{equation}
-of the monoid $M_{i}$ on each $F_{i}\in F$. We expand our definition of variables 
-\begin{equation}
-        V_i = (c_i,\, T_i,\, F_i\, M_i)
-\end{equation}
-to include the monoid actions that act on $F_i$.
+of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$
 
 Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-For example, given  $V_0 = (temperature,\, \texttt{float},\, \mathbb{R})$ which is interval data:
+For example, given  the fiber $(\fname=temperature,\, \ftype=\texttt{float},\,\fttype=\reals)$ which is interval data:
 \begin{itemize}
     \item monoid operator addition $\ast = +$
-    \item monoid operations: $f: x\mapsto x + c_1 $, $g: x\mapsto x + c_2$
+    \item monoid operations: $f: x\mapsto x + \delta_1 $, $g: x\mapsto x + \delta_2  $
     \item monoid action operator composition $\bullet = \circ$
@@
     \begin{tikzcd}
-        \mathbb{R} \arrow[rd, "(x+c_1)\circ(x+c_2)"] \arrow[d, "x+c_1"'] &            \\
-        \mathbb{R} \arrow[r, "x+c_2"']                                   & \mathbb{R}
+        \reals \arrow[rd, "(x+\delta_1)\circ(x+\delta_2)"] \arrow[d, "x+\delta_2"'] &            \\
+        \reals \arrow[r, "x+\delta_2"']                                   & \reals
     \end{tikzcd}
 \end{equation}
-where $c_1, c_2$ are intervals $x_i-x_j$.
+where $\delta_1$ and $\delta_2$ are intervals between two temperatures $x$.
 
@@
 \label{sec:data_base}
-The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the basespace $K$ without making assumptions as to what that continuity is.
+The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the base space \dbase without making assumptions as to what that continuity is.
 In turn this representation of continuity can then be used to keep track of how the data fits together, for example if a visualization of a very large dataset calls for parallelization.
@@
     \includegraphics[width=.5\textwidth]{figures/math/k_different_types.png}
-    \caption{The topological base space $K$ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
+    \caption{The topological base space \dbase encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
     \label{fig:base_space_types}
 \end{figure}
-As illustrated in figure~\ref{fig:base_space_types}, $K$ is akin to an indexing space into $E$ that describes the structure of $E$.
-$K$ can have any number of dimensions and be continuous or discrete.
+As illustrated in figure~\ref{fig:base_space_types}, \dbase is akin to an indexing space into \dtotal that describes the structure of \dtotal.
+\dbase can have any number of dimensions and can be continuous or discrete.
 
@@
     \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
-    \caption{The base space $E$ is divided into fiber segments $F$. The base space $K$ acts as an index into the records in the fibers.
+    \caption{The base space \dtotal is divided into fiber segments \dfiber. The base space \dbase acts as an index into the records in the fibers.
     \note{this figure might be good all the way up top to lay out the components of fb}}
@@
 
-Formally $K$ is the quotient space \cite{QuotientSpaceTopology2020} of $E$, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $k \in K$ has a corresponding fiber $F_k$ in $E$\cite{QuotientSpaceTopology2020}.
-In figure~\ref{fig:base_space_div}, $E$ is a rectangle divided by vertical fibers $F$, so the minimal $K$ for which there is always a mapping $\pi: F\rightarrow K$ is the line.
+Formally \dbase is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
+In figure~\ref{fig:base_space_div}, \dtotal is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the line.
 
-As with fibers and monoids, we can decompose the total space into components $\pi:E_i\rightarrow K$ where
+As with fibers and monoids, we can decompose the total space into components $\pi:\dtotal_i\rightarrow \dbase$ where
 \begin{equation}
-    \pi:E_1\oplus\ldots\oplus E_i \oplus\ldots \oplus E_n \rightarrow K
+    \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
 \end{equation}
 
-which is a decomposition of $F$. The $K$ remains the same because the connectivity of records 
-does not change just because there are fewer elements in each record.
+which is a decomposition of \dfiber. The \dbase remains the same because the connectivity of records does not change just because there are fewer elements in each record.
 
@@
     \caption{These two datasets have the same (time, temperature) fiber.
-In figure~\ref{fig:base_example_line} the total space $E$ is discrete over points $k \in K$, meaning the records in the fiber are also discrete.
-In figure~\ref{fig:base_example_plane} $E$ lies over the continuous interval $K$, meaning the records in the fiber are sampled from a continuous space.
+In figure~\ref{fig:base_example_line} the total space \dtotal is discrete over points $\dbasepoint \in \dbase$, meaning the records in the fiber are also discrete.
+In figure~\ref{fig:base_example_plane} \dtotal lies over the continuous interval \dbase, meaning the records in the fiber are sampled from a continuous space.
     \note{revamp figure: F=Plane, k1 = dots, k2=line}}
@@
 
-The datasets in figure~\ref{fig:base_example} have the same fiber of (temperature, time) which is $F=\mathbb{R}\times \mathbb{R}^+$.
-In figure~\ref{fig:base_example_discrete} the fibers lie over discrete $K$ such that the records in the datasets in the fiber bundles are discrete.
-The same fiber in figure~\ref{fig:base_example_continuous} lies over a continuous interval $K$ such that the records are samples from a continuous function defined on $K$.
+The datasets in figure~\ref{fig:base_example} have the same fiber of (temperature, time).
+In figure~\ref{fig:base_example_discrete} the fibers lie over discrete \dbase such that the records in the datasets in the fiber bundles are discrete.
+The same fiber in figure~\ref{fig:base_example_continuous} lies over a continuous interval \dbase such that the records are samples from a continuous function defined on \dbase.
 
-\subsubsection{Data: Sections $\tau$}
+\subsubsection{Data: Sections \dsection}
 \label{sec:data_section}
-While the fiber and base space describe the general structure of all data that lives in the fiber bundle, the sections $\tau: K\rightarrow E$ define the datasets that live in the fiber.
-We generalize Spivak's section as table $\tau$\cite{spivakSIMPLICIALDATABASES} to any sort of structured dataset such that
+While the fiber and base space describe the general structure of all data that lives in the fiber bundle, the sections $\dsection: \dbase\rightarrow \dtotal$ define the datasets that live in the fiber.
+We generalize Spivak's description of the section as a table of records \cite{spivakSIMPLICIALDATABASES} to any sort of structured dataset such that
 \begin{equation}
     \begin{tikzcd}
-        F \arrow[r, hook] & E \arrow[d, "\pi"']              \\
-                          & K \arrow[u, "\tau"', bend right]
+        \dfiber \arrow[r, hook] & \dtotal \arrow[d, "\pi"'] \\
+                          & \dbase \arrow[u, "\dsection"', bend right]
     \end{tikzcd}
 \end{equation}
-such that there is $\pi(\tau(k)) = k$ map back to $K$. There can be many sections $\tau$, and the space of global sections is $\Gamma(E)$. For a trivial fiber bundle
+such that there is always a map $\pi(\dsection(\dbasepoint)) = \dbasepoint$.
+There can be many sections \dsection; the space of global sections is $\Gamma(\dtotal)$.
+For a trivial fiber bundle, the section is
 \begin{equation}
     \label{eq:section_return}
-    \tau(k) = (k, (g_{F_{0}}(k), \ldots, g_{F_{n}}(k))
+    \dsection(\dbasepoint) = (\dbasepoint, (g_{\dfiber_{0}}(\dbasepoint), \ldots, g_{\dfiber_{n}}(\dbasepoint)))
 \end{equation}
-where $g: K \rightarrow F$ is the index function into the fiber. Because we can decompose the bundle and the fiber, we can formulate $\tau$ as 
+where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber. Because we can decompose the bundle and the fiber, we can formulate $\dsection$ as 
 \begin{align}
-\tau = (\tau_0,\ldots, \tau_i, \dots, \tau_n) 
+\dsection= (\dsection_0,\ldots, \dsection_i, \dots, \dsection_n) 
 \end{align}
-where each section $\tau_i$ is a variable or set of variables.
+where each section $\dsection_i$ is a variable or set of variables.
 
@@
     \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
-    \caption{ Fiber (time, temperature) with an interval $K$ basespace.
-The sections $\tau_1$ and $\tau_2$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of temperature values.
-They are included in the global set of sections  $\tau_1, \tau_2 \in \Gamma(E)$}
+    \caption{ Fiber (time, temperature) with an interval \dbase basespace.
+The sections $\dsection_i$ and $\dsection_j$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of temperature values.
+They are included in the global set of sections  $\dsection_1, \dsection_2 \in \Gamma(\dtotal)$}
     \label{fig:data_sections}
@@
 
-In the example in figure~\ref{fig:data_sections}, the fiber is $\mathbb{R}^+\times \mathbb{R}$ as described in figure~\ref{fig:data_fiber_example} and the base space is the interval $K$.
-The section $\tau_1$ resolves to a series of monotonically increasing in time records of temperature.
-Section $\tau_2$ returns a different timeseries of temperature values.
-Both sections are included in the global set of sections $\tau_1, \tau_2 \in \Gamma(E)$.
+In the example in figure~\ref{fig:data_sections}, the fiber is $(time, \, temperature)$ as described in figure~\ref{fig:data_fiber_example} and the base space is the interval \dbase.
+The section $\dsection_i$ resolves to a series of monotonically increasing in time records of temperature.
+Section $\dsection_j$ returns a different timeseries of temperature values.
+Both sections are included in the global set of sections $\dsection_1, \dsection_2 \in \Gamma(\dtotal)$.
 
@@
 \label{sec:data_sheaf_stalk}
-Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(E)$.
-All fiber bundles are locally trivial, which means that $E$ restricted over a small enough neighborhood $U \subset K$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
-The sheaf $O(E)$ is the localized section of fibers $\iota^*\tau: U \rightarrow \iota^*E$
+Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(\dtotal)$.
+All fiber bundles are locally trivial, which means that \dtotal restricted over a small enough neighborhood $U \subset \dbase$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
+The sheaf $O(\dtotal)$ is the localized section of fibers $\iota^*\section: U \rightarrow \iota^*\dtotal$
 
@@
     \begin{tikzcd}
-        \iota^*E \arrow[d, "\pi"']           & E \arrow[d, "\pi"'] \arrow[l, "\iota^*"']         \\
-        U \arrow[u, "\iota^*\tau"', bend right] & K \arrow[u, "\tau"', bend right] \arrow[l, "\iota"']
+        \iota^*\dtotal \arrow[d, "\pi"']           & \dtotal \arrow[d, "\pi"'] \arrow[l, "\iota^*"']         \\
+        U \arrow[u, "\iota^*\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right] \arrow[l, "\iota"']
     \end{tikzcd}
 \end{equation}
-pulled back over the neighborhood $U$ via the inclusion map $\iota: U \rightarrow K$.
-The localized section is the germ $\xi^*\tau$.
-The neighborhood of points $k_i$ surrounding the point $k$ the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
-While $E$ is only the records in a fiber $F_k$ over a specific $k$, because the stalk includes the neighborhood $U$ it also includes nearby records.
-While this can be useful for visual transforms, often the extra needed information can be found in the jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016}.
-For example, line thickness requires the derivative of the given position to be rendered, which can be found in $E^\prime=E+\mathcal{J}(E)$
+pulled back over the neighborhood $U$ via the inclusion map $\iota: U \rightarrow \dbase$.
+The localized section is the germ $\xi^*\dsection$.
+The neighborhood of points $\dbasepoint_i$ surrounding the point \dbasepoint the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
+While \dtotal is only the fiber $\dfiber_k$ over a specific \dbasepoint , the stalk includes nearby records because the sheaf lies over the neighborhood $U$.
+While this can be useful for visual transforms, often the extra needed information can be found in the smaller jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016}.
+For example, line thickness requires the derivative of the given position to be rendered, which can be found in $\dtotal^\prime=\dtotal+\mathcal{J}(\dtotal)$
 

### 2021-01-27 draft/sections/math.tex
finished out render section

@@
 
-We model data as the fiber bundle $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$, where \dtotal, \dfiber, and \dbase are topological spaces that encode 
+We model data as the fiber bundle $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$, where \dtotal\, \dfiber\, and \dbase\ are topological spaces that encode 
 \begin{description}
@@
 
-and \dtotal is the total space of data that \dfiber lives in. The bundle is the projection map $\pi$
+and \dtotal\ is the total space of data that \dfiber\ lives in. The bundle is the projection map $\pi$
 \begin{equation}
@@
 
-that binds the variables \dfiber continuity \dbase.
-The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified, because the trivial bundle is $\dtotal=\dbase\times \dfiber$ such that extra structure in the total space \dtotal falls out and discussion can be focused on the fiber and base space.
+that binds the variables \dfiber\ continuity \dbase.
+The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified, because the trivial bundle is $\dtotal=\dbase\times \dfiber$ such that extra structure in the total space \dtotal\ falls out and discussion can be focused on the fiber and base space.
 
@@
 We use Spivak's description of simplicial database schemas \cite{spivakSIMPLICIALDATABASES} as the basis of our fiber space because he binds the components of the fiber to variable names and types.
-Spivak constructs a set \ftotal that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_n\} \in \ftypes$, where \ftypes are the data types of the variables in the dataset.
-He then defines the single variable set \fttype
+Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_n\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
+He then defines the single variable set \fttype\
 \begin{equation}
@@
 \end{equation}
-which is \ftotal restricted to objects of type \ftype bound to variable name \fname. Given \fsection, the fiber for a one variable dataset is
+which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. Given \fsection, the fiber for a one variable dataset is
 \begin{equation}
@@
 \end{equation}
-where \fsection is the schema binding variable name \fname to its datatype \ftype.
+where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype.
 A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
@@
 
-which allows us to decouple \dfiber into components $\dfiber_i$.
+which allows us to decouple \dfiber\ into components $\dfiber_i$.
 
@@
     \end{subfigure}
-    \caption{These two datasets have the same base space \dbase, but figure~\ref{fig:fiber_example_plane} has fiber  $\dfiber=\reals\times\reals$ which is (time, temperature) while figure~\ref{fig:fiber_example_cube} has fiber $\realsp\times\reals^2$ which is (time, wind=(speed, direction))}
+    \caption{These two datasets have the same base space \dbase\, but figure~\ref{fig:fiber_example_plane} has fiber  $\dfiber=\reals\times\reals$ which is (time, temperature) while figure~\ref{fig:fiber_example_cube} has fiber $\realsp\times\reals^2$ which is (time, wind=(speed, direction))}
     \label{fig:data_fiber_example}
@@
 \label{sec:data_monoid}
-After specifying \dfiber, we next describe the ways in which we can transform the values by identifying the monoid actions \monoid on the \dfiber.
+After specifying \dfiber\, we next describe the ways in which we can transform the values by identifying the monoid actions \monoid\ on the \dfiber.
 We use monoids as the abstraction because they encode composibilty, which maps well to the data transformation process in a software library \cite{yorgeyMonoidsThemeVariations}.
@@
 \end{align*}
-As with the fiber \dfiber, the total monoid space \monoid is the cartesian product
+As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
 \begin{equation}
@@
 \label{sec:data_base}
-The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the base space \dbase without making assumptions as to what that continuity is.
+The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the base space \dbase\ without making assumptions as to what that continuity is.
 In turn this representation of continuity can then be used to keep track of how the data fits together, for example if a visualization of a very large dataset calls for parallelization.
@@
     \includegraphics[width=.5\textwidth]{figures/math/k_different_types.png}
-    \caption{The topological base space \dbase encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
+    \caption{The topological base space \dbase\ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
     \label{fig:base_space_types}
 \end{figure}
-As illustrated in figure~\ref{fig:base_space_types}, \dbase is akin to an indexing space into \dtotal that describes the structure of \dtotal.
-\dbase can have any number of dimensions and can be continuous or discrete.
+As illustrated in figure~\ref{fig:base_space_types}, \dbase\ is akin to an indexing space into \dtotal\ that describes the structure of \dtotal.
+\dbase\ can have any number of dimensions and can be continuous or discrete.
 
@@
     \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
-    \caption{The base space \dtotal is divided into fiber segments \dfiber. The base space \dbase acts as an index into the records in the fibers.
+    \caption{The base space \dtotal\ is divided into fiber segments \dfiber. The base space \dbase\ acts as an index into the records in the fibers.
     \note{this figure might be good all the way up top to lay out the components of fb}}
@@
 
-Formally \dbase is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
-In figure~\ref{fig:base_space_div}, \dtotal is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the line.
+Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
+In figure~\ref{fig:base_space_div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the line.
 
@@
 
-which is a decomposition of \dfiber. The \dbase remains the same because the connectivity of records does not change just because there are fewer elements in each record.
+which is a decomposition of \dfiber. The \dbase\ remains the same because the connectivity of records does not change just because there are fewer elements in each record.
 
@@
     \caption{These two datasets have the same (time, temperature) fiber.
-In figure~\ref{fig:base_example_line} the total space \dtotal is discrete over points $\dbasepoint \in \dbase$, meaning the records in the fiber are also discrete.
-In figure~\ref{fig:base_example_plane} \dtotal lies over the continuous interval \dbase, meaning the records in the fiber are sampled from a continuous space.
+In figure~\ref{fig:base_example_line} the total space \dtotal\ is discrete over points $\dbasepoint \in \dbase$, meaning the records in the fiber are also discrete.
+In figure~\ref{fig:base_example_plane} \dtotal\ lies over the continuous interval \dbase, meaning the records in the fiber are sampled from a continuous space.
     \note{revamp figure: F=Plane, k1 = dots, k2=line}}
@@
 The datasets in figure~\ref{fig:base_example} have the same fiber of (temperature, time).
-In figure~\ref{fig:base_example_discrete} the fibers lie over discrete \dbase such that the records in the datasets in the fiber bundles are discrete.
+In figure~\ref{fig:base_example_discrete} the fibers lie over discrete \dbase\ such that the records in the datasets in the fiber bundles are discrete.
 The same fiber in figure~\ref{fig:base_example_continuous} lies over a continuous interval \dbase such that the records are samples from a continuous function defined on \dbase.
@@
     \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
-    \caption{ Fiber (time, temperature) with an interval \dbase basespace.
+    \caption{ Fiber (time, temperature) with an interval \dbase\ basespace.
 The sections $\dsection_i$ and $\dsection_j$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of temperature values.
@@
 In the example in figure~\ref{fig:data_sections}, the fiber is $(time, \, temperature)$ as described in figure~\ref{fig:data_fiber_example} and the base space is the interval \dbase.
-The section $\dsection_i$ resolves to a series of monotonically increasing in time records of temperature.
-Section $\dsection_j$ returns a different timeseries of temperature values.
+The section $\dsection_i$ resolves to a series of monotonically increasing in time records of (time, temperature) values.
+Section $\dsection_j$ returns a different timeseries of (time, temperature) values.
 Both sections are included in the global set of sections $\dsection_1, \dsection_2 \in \Gamma(\dtotal)$.
@@
 Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(\dtotal)$.
-All fiber bundles are locally trivial, which means that \dtotal restricted over a small enough neighborhood $U \subset \dbase$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
-The sheaf $O(\dtotal)$ is the localized section of fibers $\iota^*\section: U \rightarrow \iota^*\dtotal$
+All fiber bundles are locally trivial, which means that \dtotal\ restricted over a small enough neighborhood $U \subset \dbase$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
+The sheaf $O(\dtotal)$ is the localized section of fibers $\iota^*\dsection: U \rightarrow \iota^*\dtotal$
 
@@
 The localized section is the germ $\xi^*\dsection$.
-The neighborhood of points $\dbasepoint_i$ surrounding the point \dbasepoint the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
-While \dtotal is only the fiber $\dfiber_k$ over a specific \dbasepoint , the stalk includes nearby records because the sheaf lies over the neighborhood $U$.
+The neighborhood of points $\dbasepoint_i$ surrounding the point \dbasepoint\ the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
+While \dtotal\ is only the fiber $\dfiber_k$ over a specific \dbasepoint, the stalk includes nearby records because the sheaf lies over the neighborhood $U$.
 While this can be useful for visual transforms, often the extra needed information can be found in the smaller jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016}.
@@
 
-\subsection{Graphic: $H$}
+\subsection{Graphic: \gtotal}
 \label{sec:graphic}  
-We can separate the structure of the graphic from the properties of the output format by modeling the space of graphics as a fiber bundle  $(H, S, \pi, D)$.
-As with data, the fiber bundle is for a class of graphics with shared base space $S$ (~\ref{sec:graphic_fiber}) and fiber $D$ (~\ref{sec:graphic_base}) and the sections $\rho$ (~\ref{sec:graphic_section}) encode the fully specified graphics.
+We can separate the structure of the graphic from the properties of the output format by modeling the space of graphics as a fiber bundle  $(\gtotal, \gbase, \pi, \gfiber)$.
+As with data, the fiber bundle is for a class of graphics with shared base space \gbase (~\ref{sec:graphic_fiber}) and fiber \gfiber (~\ref{sec:graphic_base}) and the sections \gsection (~\ref{sec:graphic_section}) encode a graphic where the visual characteristics are fully specified.
 
-\subsubsection{Idealized Display $D$}
+\subsubsection{Idealized Display \gfiber}
 \label{sec:graphic_fiber}
-% infinite resolution version of D
-The fiber $D$ is specified in terms of the target display space, for example a 2D screen or 3D printer. In this work, we assume a 2D opaque image $D=\mathbb{R}^5$ with elements in $D$ of the form
+The fiber \gfiber\ is an idealized infinite resolution version of the target display space, for example a 2D screen or 3D printer.
+In this work, we assume a 2D opaque image $\dfiber=\reals^5$ with elements
 \begin{equation}
-    d = (x, y, r, g, b)
+(x,\, y,\, r,\, g,\, b) \in \gfiber
 \end{equation}
 such that a rendered graphic only consists of 2D position and color.
-To support overplotting and transparency, the fiber could be $D=R^{7}$ such that $d=(x, y, z, r, g, b, a)$.
-The location coordinates $x$ and $y$ are defined in terms of the display, while the $(r,g,b)$ values are filled in via a lookup on $S$.
+To support overplotting and transparency, the fiber could be $\dfiber=\reals^{7}$ such that $(x, y, z, r, g, b, a) \in \gfiber$ specifies the target display.
+The location coordinates $x$ and $y$ are defined in terms of the display, while the $(r,g,b)$ values are filled in via a lookup on \gbase.
 
-\subsubsection{Continuity of the Graphic $S$} 
+\subsubsection{Continuity of the Graphic \gbase} 
 \label{sec:graphic_base}
-An assumption of graphical representations is that they match the continuity of the data\cite{tufteVisualDisplayQuantitative2001,friendlyBriefHistoryData2008}, but the underlying topology $S$ of a graphic may need more dimensions than the data topology $K$ so that the glyph can be defined in $D$.
-Therefore we define the base space mapping from graphic $S$ to data $K$
+An assumption of graphical representations is that they match the continuity of the data\cite{tufteVisualDisplayQuantitative2001,friendlyBriefHistoryData2008}, but the underlying topology \gbase\ of a graphic may need more dimensions than the data topology \dbase\ so that the glyph can be defined in \dfiber.
+Therefore we define the base space mapping from graphic \gbase to data \dbase
 \begin{equation}
     \begin{tikzcd}
-        E \arrow[d, "\pi"'] & H \arrow[d, "\pi"'] \\
-        K                   & S \arrow[l, "\xi"']
+        \dtotal \arrow[d, "\pi"'] & \gtotal \arrow[d, "\pi"'] \\
+        \dbase                   & \gbase \arrow[l, "\vindex"']
         \end{tikzcd}
@@
 
- as the deformation retraction \cite{RetractionTopology2020} $\xi: S \rightarrow K$ that goes from a region $s \in S_{k}$ to its associated point $k$, such that when $\xi(s) = k$, $\xi^*\tau(s) = \tau(k)$.
-While dimensions can be added to $S$, it retains the same continuity as $K$.
+ as the deformation retraction \cite{RetractionTopology2020} $\vindex: \gbase \rightarrow \dbase$ that goes from a region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$, such that when $\vindex(\gbasepoint) = \dbasepoint$,\, $\vindex^*\dsection(\gbasepoint) = \dsection(\gbasepoint)$.
+While dimensions can be added to \gbase, it retains the same continuity as \dbase.
 
@@
     \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
-    \caption{The scatter and line graphic base spaces $S$ are larger than $K$ so that they can encode physical aspects of the glyph, such as shape (a circle) or thickness.
-The heatmap base $S$ is the same as $K$ because a heatmap is directly mapping the values in the data to the graphic.
-    \note{add $\alpha, \beta$ coordinates}}
+    \caption{The scatter and line graphic base spaces have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
+The heatmap has the same continuity in the graphic \gbase\ as in the data \dbase.
+\note{add $\alpha, \beta$ coordinates to figures}}
     \label{fig:graphic_retraction_map}
@@
 
-In figure~\ref{fig:graphic_retraction_map} each disk $S_k$ indexes how elements in $D$ are held together to generate a single glyph that is the visual representation of a single point in $F_k$.
-For the line, the region $\beta$ over a point $\alpha_i$ specifies the thickness of the line in $D$ for the corresponding section on the interval $K$.
+In figure~\ref{fig:graphic_retraction_map} each disk $\gbase_{\dbasepoint}$ indexes how elements in \gfiber\ are glued together to generate a single glyph that is the visual representation of a single record in $\dfiber_{\dbasepoint}$.
+For the line, the region \gy\ over a point $\gx_i$ specifies the thickness of the line in \gbase\ for the corresponding \dsection\ on \dbase.
 The heatmap has the same continuity in data space and graphic space such that no extra dimensions are needed.
@@
 %pick a pixel which is the bounding box
-A section $\rho: S\rightarrow H$ defines a piece of the graphical representation of the data.
-The physical display coordinates in $D$ are not equivalent to the mathematical coordinates in $S$ such that many points $s \in S$ might map to a single $d\in D$ and many $d$ might go into a visible section of a displayed graphic.
-Therefore, we define the bounding box $b=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ in the display space $D$ that is a piece of the rendered graphic.
-
-We inverse map this bounding box $S_b =\rho^{-1}(b)$ into a region $S_b \subset S$  
+A section $\gsection: \gbase\rightarrow \gtotal$ defines a piece of the graphical representation of the data.
+Evaluated on a single \gbasepoint\, \gsection\ returns a single element in \gfiber\.
+For a 2D screen, the pixel is defined as a region $p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ of the rendered graphic.
+Since the x and y in $p$ are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box $\gbase_{p} ={\gsection_{xy}}^{-1}(p)$ is a region $\gbase_p \subset \gbase$.
+Integrating over this region on \gbase\
 \begin{align}
-    R_b &= \iint\limits_{S_b} \rho_r(s)ds^{2}\\
-    G_b &= \iint\limits_{S_b} \rho_g(s)ds^{2}\\
-    B_b &= \iint\limits_{S_b} \rho_b(s)ds^{2}
+    r_p &= \iint\limits_{S_p} \rho_r(s)ds^{2}\\
+    g_p &= \iint\limits_{S_p} \rho_g(s)ds^{2}\\
+    b_p &= \iint\limits_{S_p} \rho_b(s)ds^{2}
 \end{align}
-to then compute the remaining $(R, G, B)$ components of the bounded display region $D_b$.
+yields the color of the pixel $p$.
 
@@
     \includegraphics[width=\textwidth]{figures/math/graphic_render.png}
-    \caption{To render a graphic, a region $b$ is selected in the display space.
-This region is in coordinates $D$ so the inverse mapping $\rho^{-1}(b)$ maps to an equivalent region $S_b \subset S$.
-We then take the sections $\rho(S_b)$ which are the elements $\{d_0, \ldots, d_n\}$ that make the piece of the graphic in $b$.}
+    \caption{To render a graphic, a pixel $p$ is selected in the display space, which is defined in the same coordinates as the x and y components in \gfiber\.
+The inverse mapping ${\gsection_{xy}}^(p)$ returns a region $\gbase_{p} \subset \gbase$.
+$\gsection(\gbase_{p})$ returns the list of elements $(x,\,y,\,r,\,g,\,b) \in \gfiber$ that lie over $\gbase_{p}$.
+The integral over the $(r,\,g,\,b)$ elements is the color of the pixel.}
     \label{fig:graphic_rho_lookup}
@@
 As shown in figure~\ref{fig:graphic_rho_lookup}, the output space queries into the graphic bundle to render the image.
-We select a region $b$ in the output space, inverse map the region into $S_b \subset S$, then take the section $\rho$ over the region $S_b$.
-The section yields the set of elements $\{d_0, \ldots, d_n\}$ that specify the $(r, g, b)$ values corresponding to the coordinates in $b$.
+We select a pixel $p$ in the output space, inverse map the region of the pixel into $\gbase_p \subset \gbase$, then compute the section $gsection$ over the region $\gbase_p$.
+The section yields the set of elements in \gfiber\ that specify the $(r, g, b)$ values corresponding to the region $p$.
+The color of the pixel is then obtained by taking the integral of $\gsection_{rgb}(\gbase_p)$.
+\note{probably needs some conclusion}
 

### 2021-01-28 draft/sections/math.tex
finished conversion through nu + added elements

@@
     \includegraphics[width=\textwidth]{figures/math/dar.png}
-    \caption{Visualizaiton is the process of mapping data into visual characteristics that are then assembled into a graphic}
+    \caption{Visualization is equivariant maps between data and visual encoding of the variables and assembly of those encodings into a graphic.
+    \note{not gonna name these bubbles tau, mu, rho, but might keep the same basic structure of different types of data and encodings }}
     \label{fig:artist_stages}
@@
 
-\subsection{Data Space}
+\subsection{Data Space\dtotal}
 \label{sec:data}
@@
 
-\subsubsection{Renderable Glyph $\rho$}
+\subsubsection{Renderable Glyph \gsection}
 \label{sec:graphic_section}
@@
 The color of the pixel is then obtained by taking the integral of $\gsection_{rgb}(\gbase_p)$.
-\note{probably needs some conclusion}
 
@@
     \includegraphics[width=\textwidth]{figures/math/path_of_q}
-
-    \caption{The $\nu$ functions convert data $E$ to visual $V$.
-$Q$ assembles the different types of visual parameters $V_{i}$ into a graphic in $H$.
-$Q\circ\mu(\xi^{-1}J)$ forms a visual mark by applying $Q$ to a region mapped to connected components $J \subset K$.
-\note{maybe change to $\tau, \mu, \rho$ }}
+    \caption{\vchannel\ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.
+\gsection\ applied to a region of connected components $\gbase_J$ generates a graphical mark.}
     \label{fig:artist_q}
@@
 
-The artist is the function that converts data into graphics; its name is taken from the analogous part of the existing Matplotlib architecture that builds visual elements to pass off to the renderer.
-The artist is a mapping from $E$ padded with data from $\mathcal{J}(E)$ to a graphic that is a section $\rho$ in  $\Gamma(H)$
+The artist is the function that converts data into graphics; its name is taken from the analogous part of Matplotlib\cite{hunterArchitectureOpenSource} that builds visual elements to pass off to the renderer.
+The artist \vartist\ is a mapping from \dtotal\ padded with data from $\mathcal{J}(\dtotal)$ to a graphic that is a section \gsection\ in  $\Gamma(\gtotal)$
 \begin{equation}
@@
     \begin{tikzcd}
-        E' \arrow[r, "\nu"] \arrow[rd, "\pi"'] & V \arrow[d, "\pi"] & \xi^*V \arrow[r, "Q"] \arrow[d, "\xi^*\pi"'] \arrow[l, "\xi^*"'] & \Gamma(H) \arrow[ld, "\pi"] \\
-                                              & K                  & S \arrow[l, "\xi"']                                              &                    
+        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \Gamma(\gtotal) \arrow[ld, "\pi"] \\
+                                              & \dbase                  & \gbase \arrow[l, "\vindex"']                                              &                    
         \end{tikzcd}
 \end{equation}
-with an intermediate fiber bundle $V$ to hold visual representations and stages
+with an intermediate fiber bundle \vtotal\ to hold visual representations and stages
 \begin{enumerate}
-    \item $\xi$ binding the continuity in the graphic to the continuity in the data (~\ref{sec:graphic_base})
-    \item $\nu$ conversion of data into visual characteristics (~\ref{sec:artist_nu})
-    \item $Q$ assembly of visual variables into a glyph (~\ref{sec:artist_q})
+    \item \vindex\ binding the continuity in the graphic to the continuity in the data (~\ref{sec:graphic_base})
+    \item \vchannel\ conversion of data into visual characteristics (~\ref{sec:artist_nu})
+    \item \vmark\ assembly of visual variables into a glyph (~\ref{sec:artist_q})
 \end{enumerate}
@@
 of the visual transformation illustrated in figure~\ref{fig:artist_stages}.
-The functions $\xi$, $\nu$ and $Q$ are constructed such that the input can be a single section on $K$ or $S$, which allows the artist to be trivially parallelized or formulated as a delayed computer graph or otherwise not need to immediately work on all the data.
+The functions \vindex\, \vchannel\ and \vmark\ are defined such that they can be evaluated on a single section \dsection, which allows the artist to be implemented such that it does not need all the data.
+This allows for artists tuned to distributed and streaming data.
 
-\subsubsection {Visual Fiber Bundle $V$}
-The visual fiber bundle ($V$, $K$, $\pi$, $P$) has section $\mu: V \rightarrow K$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber $P$.
-The fiber space $P$ is defined in terms of the parameters of the visualization specification and the values those parameters resolve to - for example ggplot \cite{wickhamGgplot2ElegantGraphics2016a} or D3\cite{bostockDataDrivenDocuments2011}.
+\subsubsection {Visual Fiber Bundle \vtotal}
+The visual fiber bundle (\vtotal, \dbase, $\pi$, \vfiber) has section $\vsection: \vtotal \rightarrow \dbase$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber \vfiber.
+The fiber space \vfiber\ is defined in terms of the parameters of the visualization specification- for example aesthetics  in ggplot \cite{wickhamGgplot2ElegantGraphics2016a}, channels in vega\cite{satyanarayanDeclarativeInteractionDesign2014} or parameters in VTK\cite{hanwellVisualizationToolkitVTK2015} and Matplotlib.
+
 \begin{table}[H]
@@
     \begin{tabulary}{\textwidth}{|l|L|l|}\hline
-     $\bm{\nu_{i}}$                      & $\bm{\mu_{i}}$                                                            & $\bm{codomain(\nu_{i})}$  \\ \hline                                              
+     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i})}$  \\ \hline
     position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
@@
     \end{tabulary}
-    \caption{Some possible components of the fiber $P$ for a visualization implemented in Matplotlib}
+    \caption{Some possible components of the fiber \vfiber\ for a visualization function implemented in Matplotlib}
     \label{tab:mpl_visual_variable_fiber}
@@
 Table~\ref{tab:mpl_visual_variable_fiber} is a sample of the fiber space for Matplotlib \cite{hunterMatplotlib2DGraphics2007}.
-A section $\mu$ of $V$ is a tuple of visual values that specifies the visual characteristics of a part of the graphic.
+A section \vsection is a tuple of visual values that specifies the visual characteristics of a part of the graphic.
 For example, given a fiber of $\{xpos, ypos, color\}$ one possible section could be  $\{.5, .5, (255, 20,147)\}$.
-The $codomain(\nu_i)$ determines the monoid actions on $\mu_i)$.
+The $codomain(\vchannel_i)$ determines the monoid actions on $\vsection_i$.
 These fiber components are implicit in the library, by making them explicit as components of the fiber we can build consistent definitions and expectations of how these parameters behave.
@@
     \includegraphics[width=\textwidth]{figures/math/equivariance_nu.png}
-    \caption{An equivariant $\nu$ preserves monoid actions.
-Here $\nu$ maps the words to the emojis and is constructed such that a permutation of the words is reflected in a permutation of the symbols.
+    \caption{An equivariant \vchannel preserves monoid actions.
+Here \vchannel\ maps the words to the emojis and is constructed such that a permutation of the words is reflected in a permutation of the symbols.
 \note{will switch to a figure table thing but this is the idea}}
@@
 \end{figure}
-As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers $\nu$ as the set of conversion functions 
+As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers \vchannel\ as the set of conversion functions 
 \begin{equation}
     \label{eq:nu_expanded}
-    \{\nu_{0}, \ldots, \nu_{n}\}: \{\tau_{0}, \ldots, \tau_{n}\} \mapsto \{\mu_{0}, \ldots, \mu_{n}\}
+    \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
 \end{equation}
-such that $\nu_i: \tau_i \mapsto \mu_i$ is an equivariant map such that there is a monoid homomorphism from $F_{i}$ to $V_{i}$. A validly constructed $\nu$ is one where the  diagram 
+where $\vchannel_i: \dsection_i \mapsto \vsection_i$ is an equivariant map such that there is a monoid homomorphism from $\dfiber_{i}$ to $vfiber_{i}$.
+A validly constructed \vchannel\ is one where the  diagram of the monoid transform $m$
 \begin{equation}
@@
 \begin{tikzcd}
-    E_i \arrow[r] \arrow[r, "\nu_i"] \arrow[d, "m_e"'] & V_i \arrow[d, "m_v"] \\
-    E_i \arrow[r, "\nu_i"]                           & V_i               
+    \dtotal_i \arrow[r] \arrow[r, "\vchannel_i"] \arrow[d, "m_{\delement}"'] & \vtotal_i \arrow[d, "m_{\velement}"] \\
+    \dtotal_i \arrow[r, "\vchannel_i"]                           & \vtotal_i               
 \end{tikzcd}
 \end{equation}
-commutes such that $\nu_i(m_e(E_i)) = m_v(\nu_i(E_i))$.
+commutes such that $\vchannel_i(m_{\delement}(\dtotal_i)) = m_{\velement}(\vchannel_i(\dtotal_i))$.
 This equivariance constraint yields guidance on what makes for an invalid transform.
-For example, the single value conversion $\nu_{i}(x) = .5$ does not commute under translation  $t(x) = x+2$
+For example, the conversion $\vchannel_{i}(x) = .5$ does not commute under translation monoid action $t(x) = x+2$
 \begin{align}
-    \nu(t(e + 2)) & \overset{?}{=} \nu(e) + \nu(2)\\
+    \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + \vchannel(2)\\
     .5 &\neq .5 + .5
@@
 
-On the other hand figure~\ref{fig:artist_nu} illustrates a valid $\nu$ mapping from \textbf{Strings} to symbols.
+On the other hand figure~\ref{fig:artist_nu} illustrates a valid \vchannel\ mapping from \textbf{Strings} to symbols.
 The group action on these sets is permutation, so shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
-To preserve ordinal and partial order monoid actions, $\nu$ must be a monotonic function such that given $e_1, e_2 \in E_{i}$, $\text{ if } e_1 \leq e_2 \text{ then } \nu(e_1) \leq \nu(e_2)$.
-For interval scale data, $\nu$ is equivariant under translation monoid actions if $\nu(x + c) = \nu(x) + \nu(c)$.
-For ratio data, there must be equivalent scaling $ \nu(xc) = \nu(x)\nu(c)$.
+To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$, $\text{ if } delement_1 \leq delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$.
+For interval scale data, \vchannel\ is equivariant under translation monoid actions if $\vchannel (x + c) = \vchannel(x) + \vchannel(c)$.
+For ratio data, there must be equivalent scaling $ \vchannel(xc) = \vchannel(x)\vchannel(c)$.
 

### 2021-01-28 draft/sections/math.tex
triangulization + some clean ups

@@
     \item monoid operator addition $\ast = +$
-    \item monoid operations: $f: x\mapsto x + \delta_1 $, $g: x\mapsto x + \delta_2  $
+    \item monoid operations: $f: x\mapsto x + 1 $, $g: x\mapsto x + 2  $
     \item monoid action operator composition $\bullet = \circ$
@@
     \begin{tikzcd}
-        \reals \arrow[rd, "(x+\delta_1)\circ(x+\delta_2)"] \arrow[d, "x+\delta_2"'] &            \\
-        \reals \arrow[r, "x+\delta_2"']                                   & \reals
+        \reals \arrow[rd, "(x+ 1\degree)\circ(x+2\degree)"] \arrow[d, "x+ 1\degree"'] &            \\
+        \reals \arrow[r, "x+ 2\degree"']                                   & \reals
     \end{tikzcd}
 \end{equation}
-where $\delta_1$ and $\delta_2$ are intervals between two temperatures $x$.
+where $1\degree$ and $2\degree$ are valid distances between two temperatures $x$.
 
@@
 \label{sec:triangulization}
-\note{HAVE NOT REACHED HERE TO EDIT YET}
-
 \begin{figure}[H]
@@
 \end{figure}
+One way of expressing the connectivity of records in a dataset is to implement \dbase\ as a simplacial complex, which is a set of simplicies such as those shown in figure~\ref{fig:triangle_simplex}.
+The advantage of triangulization is that it is general enough to work for more complex topology based visualization methods \cite{heineSurveyTopologybasedMethods2016} while also providing a consistent interface of vertices, edges, and faces for \vindex\ to map into.
+When triangulated, the simplices encode the continuity in the data
 
-One way to build flexible $\xi$ is to choose a consistent way of representing $K$.
-In our draft implementation of the data as fiber bundle model, we triangularize $K$ using complexes of the simplicies shown in figure~\ref{fig:triangle_simplex} such that $\xi$ consistently yields some combination of vertexes, edges, and faces.
-This gives a common data indexing structure on which to build components that could potentially be reused across $Q$.
+\begin{table}[H]
+    \begin{center}
+        \begin{tabular}{|l|l|l|}\hline
+        \textbf{simplex} & \textbf{continuity} & \textbf{\dsection}   \\ \hline
+        vertex  & discrete   & \dsection(\dbasepoint)                  \\ \hline
+        edge    & 1D         & $\dsection(\dbasepoint,\, \alpha)$        \\ \hline
+        face    & 2D         & $\dsection(\dbasepoint,\, \alpha,\, \beta)$\\ \hline
+        \end{tabular}
+        \caption{}
+        \label{tab:triangulization}
+    \end{center}
+\end{table}
 
-By using simplicies, we can then encode the following indexing structure on $\tau$
-\begin{align*}
-    \textbf{vertexes}\;& \tau(k) \text{k= vertex id}\\
-    \textbf{edges}\;& \tau(k,\alpha), \text{k = edge id, $\alpha$ = distance along edge} \\
-    \textbf{faces}\;& \tau(k,\alpha, \beta), \text{k = face id, $\alpha$ = x on face, $\beta$ = y on face}
-\end{align*}
+such that each section is bound to a simplex $\dbasepoint \in \dbase$.
+As shown in table~\ref{tab:triangulization}, in a 1D continuous spaces each \dsection\ lies distance $\alpha$ along edge \dbasepoint, while in a 2D continuous space each \dsection\ lies at coordinate $\alpha, \beta$ on the face \dbasepoint.
+This is directly analogous to indexing to express connectivity in N-D arrays, while also natively supporting graphs and trees as they are simplacial complices of nodes and edges.
+Path connected components are then sections where edges or faces meet.
 
-which $\xi$ can use as part of its mapping from visual space back to data space. Path connected components are then sections where $\tau(k_{i},1) = \tau(k_{j},0)$ or the edges of the faces align.
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
+    \caption{The torus \dtotal is unraveled into a simplacial complex of 2 faces \dbase. Transition functions are defined on the edges of \dbase such that surface can be glued back into the torus.
+    \note{add cross sections a and b to ring and color same as edges in complex}}
+    \label{fig:triangle_torus}
+\end{figure}
+One way of encoding the torus in figure~\ref{fig:triangle_torus} while retaining the continuity of both cross sections $a, b$ is to unravel it into a simplacial complex of two triangles with labeled edges.
+Transition functions $\delta$ are defined on the edges such that $a$ can be glued to $a^\prime$ and $b$ to $b^\prime$ to reconstruct the torus.
+This simplacial complex is then used as the base space encoding the continuity of data that lies in the torus.
+A constraint on the transition functions is that the monoid actions on the fibers on the edges of \dtotal\ are commutative $\monoid*\dfiber\mapsto \delta(\monoid\dfiber)= \monoid*\delta(\dfiber)$
 
-\subparagraph{Example: Mobius Strip}\mbox{}\\
 \begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
+    \caption{Many non-trivial spaces can be made locally trivial by dividing \dtotal into locally trivial subspaces and defining transition functions between the edges on \dbase for how to glue the two subspaces such that the \dsection are continuous.}
     \label{fig:data_base_transition}
-    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
-    \caption{Many non-trivial spaces can be made locally trivial by dividing $E$ into locally trivial subspaces and defining transition functions between the edges on K for how to glue the two subspaces such that the sigmas are continuous.}
 \end{figure}
-
- An example of a non-trivial fiber bundle is a mobius strip\cite{find the citation for why};
-
-In figure ~\ref{fig:data_base_transition}, the mobius strip is cut into two rectangular base spaces over $K=\{e_1, e_2\}$.
-We define transition functions from the fiber to the fiber along the edges of the rectangles such that we can glue these rectangles together to reform the mobius strip.
-A constraint we impose on the transition functions is that the monoid actions are commutative
-%monoid action m*f\mapsto delta(mf)= m*delta(f)
-
-\subparagraph{Example: Torus}\mbox{}\\
-\begin{figure}[H]
-    \label{fig:triangle_torus}
-    \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
-    \caption{Representation of the base space $K$ of a toridal total space $E$ as two connected 2-simplices.}
-\end{figure}
-
-Given data that lies on the toroidal space shown in figure~\ref{fig:triangle_torus}, the torus $E$ base space $K$ can be implemented as a simplacial complex of two 2-simplexes.
-We unfold the torus into the two triangles that compose the square; the sections on these triangles are $\tau(triangle id k, \alpha, \beta)$.
-We also put transition functions on the edges such that $A$ can be glued to $A^\prime$ and $B$ to $B^\prime$ to reconstruct the torus.
+Another advantages of triangulization is that it provides a way to encode non-trivial structures such as the mobius strip\cite{MobiusStripNLab}.
+As shown in figure ~\ref{fig:data_base_transition}, one way of making the mobius strip trivial is to seperate it into two spaces $\dtotal_1$ and $\dtotal_2$ and then define transition functions that specify that the edges of $\dtotal_1$ need to be reversed to line up with $\dtotal_2$ such that the sections along the edges meet.
+As with the torus, the transition functions must preserve monoid commutativity.
 

### 2021-01-28 draft/sections/math.tex
equivalence class

@@
 
+\subsubsection{Equivalance class of artists}
+\note{HAVE NOT REACHED HERE TO EDIT YET}
+As formulated above, every artist function \vartist\ has fixed \vchannel\ and \vmark\ generates a distinct graphic \gsection\.
+It is impractical to implement an artist for every single graphic; instead we implement the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$.
+Equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\, but act on different sections \vsection.
+To further simplify implementation, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data\note{needs citation, maybe friendlys history or acquired codes of meaning}.
+For example, a scatter plot of red circles is one artist, a scatter plot of green squares is another.
+These two artists are equivalent since their only difference is in the literal visual encodings (color, shape).
+Shape and color could also be defined in \vmark\, but the position must come from the fiber $\vfiber=(xpos,ypos)$ since fundementally a scatter plot is the plotting of one position against another\cite{friendlyBriefHistoryData2008}.
+We also use this criteria to identify derivative types, for example the bubble chart\cite{tufteVisualDisplayQuantitative2001} is a type of scatter where by definition the glyph size is mapped from the data.
+
 \subsection{Making the fiber bundle computable}
@@
 
-\subsubsection{Visual Idioms: Equivalance class of artists}
-\note{HAVE NOT REACHED HERE TO EDIT YET}
-As formulated above, every artist function $A$ has fixed $\nu$ and generates a distinct graphic $\rho$.
-It is unfeasible to implement $A$ for every single graphic; instead we implement the equivalence class of artists $\{A \in A^\prime: A_{1} \equiv A_{2}\}$ which is $Q:\Gamma(V) \rightarrow \Gamma(H)$.
-
-%% Well maybe.
-I think this is where required components comes in.
-There is some minimal V where if you didn't have those variables you wouldn't call it a scatter plot ....
-something like xpos,ypos with K just a descrete set.
-%% so V probably can be any V that contains that sub-bundle
-%% so if bubble plot is not the same as scatter plot, and something where the colors change based on data is not scatter plot then it is the equivalent class of something with that minimal bundle.
-%% But then if you want to include ANYTHING that also maps with X-Y ... then you hit "direct limits" where you limit over all possivle V_big containing V_minimal
-%% You still get a map on some A_minimal -> A_big
-%% If you know how to visualize with more variables .. you can always fix them.
 \end{document}

### 2021-01-28 draft/sections/math.tex
graphic marks

@@
 \label{sec:artist}
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/path_of_q}
-    \caption{\vchannel\ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.
-\gsection\ applied to a region of connected components $\gbase_J$ generates a graphical mark.}
-    \label{fig:artist_q}
-\end{figure}
-
 The artist is the function that converts data into graphics; its name is taken from the analogous part of Matplotlib\cite{hunterArchitectureOpenSource} that builds visual elements to pass off to the renderer.
@@
 \label{sec:artist_nu}
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/equivariance_nu.png}
-    \caption{An equivariant \vchannel preserves monoid actions.
-Here \vchannel\ maps the words to the emojis and is constructed such that a permutation of the words is reflected in a permutation of the symbols.
-\note{will switch to a figure table thing but this is the idea}}
-    \label{fig:artist_nu}
-\end{figure}
-As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers \vchannel\ as the set of conversion functions 
+As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers \vchannel\ as the set of independent conversion functions 
 \begin{equation}
@@
 \label{sec:artist_q}
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/path_of_q}
+    \caption{\vchannel\ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.
+\gsection\ applied to a region of connected components $\gbase_{\dbasepathpoint}$  generates a graphical mark.}
+    \label{fig:artist_q}
+\end{figure}
 
-As shown in figure~\ref{fig:artist_q}, $Q$ takes the individual fields in $V$ as input and outputs a single piece of a graphic on $H$.
-As with $\nu$, the constraint on $Q$ is that for every monoid actions on the input in $V$ there is a corresponding monoid action on the output in $H$
+As shown in figure~\ref{fig:artist_q}, the assembly function \vmark\ combines the fiber $\dfiber_i$ wise \vchannel\ transforms into a single glyph.
+Together, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings, reduce the encodings into a glyph.
+As with \vchannel\, the constraint on \vmark\ is that for every monoid actions on the input \vsection\ there is a corresponding monoid action on the output \gsection.
+
+Since we define the equivariant map as  $\vmark: \vsection \mapsto \gsection$, we define an action on the subset of graphics $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ that \vmark\ can generate.
+We then define the constraint on \vmark such that if \vmark\ is applied to $\vsection, \vsection^{\prime}$ that generate the same \gsection\, then the output of both sections acted on by the same monoid $m$ must be the same.
+
+Lets call the visual encodings $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$.
+If for all monoids $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, the output is equivalent
 \begin{equation}
-    Q: \Gamma(V) \rightarrow \Gamma(H)
+\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
 \end{equation}
-When $Q: \mu \mapsto \rho$ yields a $\rho$ that maps to the same values in $D$ over all $S_k$, then $M$ can be defined over $\Gamma(H)$ such that a constraint on $Q$ is that it must be equivariant.
-For example, when $\mu_{i}$ is the color of the glyph, it maps directly to (R,G,B) in D.
+then a group action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
+The transformed graphic $\gsection^{\prime}$ is equivariant to a transform on the visual bundle $\gsection^{\prime}=\vmark(m\circ \vsection)$ on a section that $\vsection \in \vmark^{-1}(\gsection)$ that must be part of generating \gsection.
 
@@
     \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
+    \caption{These two glyphs are generated by the same \vmark\ function, but differ in the value of the edge thickness parameter $\vsection_i$.
+A valid \vmark\ is one where a shift in $\vsection_i$ is reflected in the glyph generated by \gsection.}
     \label{fig:artist_mark_change}
@@
 
-Many $\mu_{i}$ are graphical parameters that do not apply to the whole glyph, such as edge thickness in figure~\ref{fig:artist_mark_change}.
-
-%%% rework this sentence
-In these situations, not all  $\rho$ in $\Gamma(H)$ will support these parameters; instead we define an action on the output graphic $Q(\Gamma(V)) \in \Gamma(H)$ since by definition every section $\mu$ will have a corresponding $\rho$.
-
-We then define the constraint on $Q$ such that if $Q$ applied to two sections $\mu, \mu\prime$ generate the same graphic $\rho$, then the output of both sections acted on by the same monoid $m$ must also be the same.
-
-Lets call the visual encodings $\Gamma(V)=X$ and the graphic $Q(\Gamma(V))=Y$. If $\forall m \in M$ and $\forall \mu, \mu^\prime \in X$, 
-\begin{equation}
-Q(\mu) = Q(\mu^\prime)\implies Q(m\circ\mu) = Q(m\circ\mu^\prime)
-\end{equation}
-then a group action on $Y$ can be defined as $m\circ \rho = \rho^\prime$ where $\rho^\prime=Q(g\circ \mu)$ with $\mu \in Q^{-1}(\rho)$.
-
-Given  
-\begin{itemize}
-    \item $P = \{xpos, ypos, color, thickness\}$
-    \item $\mu = {0,0,0, 1}$
-    \item  $Q(\mu)=\rho$ generates a piece of the thin circle in figure~\ref{fig:artist_mark_change}
-\end{itemize}
-
-the constraint on $Q$ means that the translation $m=\{e, e, e, x+2\}$ applied to $\mu$ such that $\mu^\prime=\{0,0,0,3\}$ has an equivalent action on $\rho$ that causes $Q(\mu\prime)$ to be equivalent to the thicker circle in figure~\ref{fig:artist_mark_change}.
+The glyph in figure~\ref{fig:artist_mark_change} has the following characterstics \vfiber specified by  $(xpos,\, ypos,\, color,\, thickness)$ such that one section is $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
+The equivariance constraint on \vmark is that the action $m=(e, e, e, x+2)$, where e is identity, applied to \vsection\ such that $\vsection^{\prime}=(e,e,e,3)$ has an equivalent action on \gsection that causes $\vmark(\vsection^{\prime})$ to be equivalent to the thicker circle in figure~\ref{fig:artist_mark_change}.
 
 \paragraph{Example: Invalid Q}
-\textcolor{blue}{Insert some degenerate Q that generates an inconsistent glyph}
+\note{DEGENERATE Q - drawing a blank if this is necessry and if so how}
+Check a well defined map $\monoid\times Y \rightarrow Y$.
 
-Check a well defined map $M\times Y \rightarrow Y$.
-
-constraint: inputs go to same o
-utput means changes to inputs mean same changes to output
-
-\paragraph{Graphical Marks}
-To output a mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}, $Q$ is called with all the regions $s$ that map back to a set of connected components $J \subset K$:
+To output a mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}, \vmark\ is called with all the regions \gbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$:
 \begin{equation}
-J = \{j \in K \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=k \text{ and }\gamma(1)=j\}
+\dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
 \end{equation}
-where the path\cite{ConnectedSpace2020}  $\gamma$ from $k$ to $j$ is a continous function from the interval [0,1].
-
-We define the mark as the graphic generated by $Q(S_j)$
+where the path\cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1].
+We define the mark as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$
 
@@
     \begin{tikzcd}
-        H \arrow[r, shift left] & S_j \arrow[rr, "\xi(s)", shift left] \arrow[l, "\rho(S_j)"] &  & J_{k} \arrow[ll, "\xi^{-1}(J)"]
+        \gtotal \arrow[r, shift left] & \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] &  & \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
         \end{tikzcd}
@@
 \end{equation}
-
-in terms of $K$ because mark is a semantic term denoting the graphic representation of the data.
+such that for every mark there is at least one corresponding section on \dbase.
 
@@
 
-\subsubsection{Equivalance class of artists}
-\note{HAVE NOT REACHED HERE TO EDIT YET}
-As formulated above, every artist function \vartist\ has fixed \vchannel\ and \vmark\ generates a distinct graphic \gsection\.
+\subsubsection{Equivalance class of artists $\vartist^{\prime}$}
+As formulated above, every artist function \vartist\ has fixed \vchannel\ and \vmark\ which  generates a distinct graphic \gsection.
 It is impractical to implement an artist for every single graphic; instead we implement the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$.
@@
 To further simplify implementation, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data\note{needs citation, maybe friendlys history or acquired codes of meaning}.
-For example, a scatter plot of red circles is one artist, a scatter plot of green squares is another.
+\begin{figure}
+    \caption{Each of these graphics is generated by a different artist \vartist which is the equivalance class of scatter plots $\vartist^{\prime}$\note{this is gonna be a whole bunch of scatter plots}}
+    \label{fig:artist_equivalence}
+\end{figure}
+For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another, as are the rest of the graphs in figure~\ref{fig:artist_equivalance}.
 These two artists are equivalent since their only difference is in the literal visual encodings (color, shape).

### 2021-01-28 draft/sections/math.tex
math

@@
 
-\paragraph{Example: Invalid Q}
-\note{DEGENERATE Q - drawing a blank if this is necessry and if so how}
-Check a well defined map $\monoid\times Y \rightarrow Y$.
+\note{DEGENERATE Q - drawing a blank if this is necessry and if so how\\
+Check a well defined map $\monoid\times Y \rightarrow Y$}
 
@@
 %rewrite as alpha/beta 
-In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and a simple heatmap.
-
-\paragraph{Q: scatter plot}\mbox{} \\
+In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and a heatmap. 
 \label{sec:artist_example_scatter}
@@
     \includegraphics[width=\textwidth]{figures/math/scatter.png}
+    \caption{The data is discrete points (temperature, time).
+Via \vchannel\ these are converted to (xpos, ypos) and pulled over discrete \gbase.
+These values are then used to parameterize \gsection which returns a color based on the parameters (xpos,ypos) and position $\alpha, \beta$ on $\gbase_{\dbasepoint}$ that \gsection\ is evaluated on.
+    }
     \label{fig:artist_scatter}
 \end{figure}
-
-\begin{equation}
-Q(xpos, ypos)(\alpha_{1}, \alpha_{2}) 
-\end{equation}
-
-Given a defaut color of black, $\rho_{RGB} = (0,0,0)$. The position of this swatch of color can be computed relative to the location on the disc $S_{i}$ as shown in figure~\ref{fig:artist_scatter}:
+The scatter plot in figure~\ref{fig:scatter} can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
+The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in figure~\ref{fig:artist_scatter}:
 \begin{align}
-x &= size\bullet \alpha_1 \bullet \cos(\alpha_2) + xpos\\
-y &= size\bullet \alpha_1 \bullet \sin(\alpha_2) + ypos
+x &= size\bullet \alpha \bullet \cos(\beta) + xpos\\
+y &= size\bullet \alpha \bullet \sin(\beta) + ypos
 \end{align}
 
-such that $\rho(s) = (x, y, 0, 0, 0)$ where $s$ is the region in $H$.
+such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
 
-\paragraph{Q: line plot}\mbox{} \\
 \label{sec:artist_example_line}
@@
     \includegraphics[width=\textwidth]{figures/math/line.png}
+    \caption{The line fiber $(time,\, temp)$ is thickend with the derivative $(time^{\prime},\, temperature^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a thick line.
+This is because the line needs to be pushed perpendicular to the tangent of (xpos, ypos).
+\note{this is gonna move once this gets regenerated w/ labels} The data is converted to visual characteristics (xpos, ypos).
+The $\alpha$ coordinates on \gbase\ specifies the position of the line, the $\beta$ coordinate specifies thickness.}
     \label{fig:artist_line}
@@
 
-The line plot shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet discussed in section~\ref{sec:sheaf_stalk}
-(needs the normal to push up/down against the normal)
-
-\begin{equation}
-    Q(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha_1, \alpha_2)
-\end{equation}
-
-where the magnitude of the thickness is 
+The line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet discussed in section~\ref{sec:sheaf_stalk}.
+The line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
+The magnitude of the thickness is
 \begin{equation}
@@
 \end{equation}
-such that the components are 
+such that the normal is  
 \begin{equation}
@@
 
-%% only alpha1 exists on the data side
-which yields components of $\rho(s)$:
+which yields components of \gsection
 \begin{align}
- x = xpos(\xi(\alpha_1)) &+ \alpha_2\hat(n_1)(\xi(\alpha_1)) \\
- y = ypos(\xi(\alpha_1)) &+ \alpha_2\hat(n_2)(\xi(\alpha_2)) 
+ x = xpos(\vindex(\alpha)) &+ \beta\hat(n_1)(\vindex(\alpha)) \\
+ y = ypos(\vindex(\alpha)) &+ \beta\hat(n_2)(\vindex(\alpha)) 
 \end{align}
+
+where (x,y) look up the position $\vindex(\alpha)$ on the data and then apply thickness $\beta$ at that location.
 
@@
     \includegraphics[width=\textwidth]{figures/math/heatmap.png}
+    \caption{The only visual parameter a heatmap requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
     \label{fig:artist_heatmap}
 \end{figure}
-The heatmap in figure~\ref{fig:artist_heatmap} 
-\begin{equation}
-Q(xpos, ypos, color)
-\end{equation}
-has in the simple case a direct lookup into $K$ to obtain the $\mu = (x,y,c)$ values that are mapped into 
-(arrays are upside down, x is y, is x) - $\xi$ is about translating indices from data to visual alignment
+
+The heatmap $\vmark(color)$ in figure~\ref{fig:artist_heatmap} is a direct lookup $\vindex:\gbase\rightarrow\dbase$ such that
+
 \begin{align}
-D_{RGB} &= color(\xi(\alpha_1, \alpha_2))
-D_x & = xpos(\xi(\alpha_1, \alpha_2))
-D_y &= ypos(\xi(\alpha_1, \alpha_2))
+R &= R(\vindex(\alpha, \beta))\\
+G &= G(\vindex(\alpha, \beta))\\
+B &= B(\vindex(\alpha, \beta))
 \end{align}
-through $\rho$.
+where \vindex\ may do some translating to a convention expected by \vmark\, for example reorientng the array such that the first row in the data is at the bottom of the graphic.
 

### 2021-01-29 draft/sections/math.tex
put back in missing figure

@@
 
-\subsection{Data Space\dtotal}
+\subsection{Data Space \dtotal}
 \label{sec:data}
@@
 \label{sec:artist_nu}
+\begin{figure}
+    \includegraphics[width=\textwidth]{figures/math/equivariance_nu.png}
+    \caption{In this artis, \vchannel\ maps the strings to the emojis.
+For \vchannel\ to be equivariant, a shuffle in the words should have an equivalent shuffle in the emojies, and a shuffle in the emojies should have an equivalent shuffle in the words.}
+    \label{fig:artist_nu}
+\end{figure}
 As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers \vchannel\ as the set of independent conversion functions 

### 2021-01-29 draft/sections/implementation.tex
start of reworked implementation

--- 2021-01-15 draft/sections/implementation.tex
@@
 \documentclass[../main.tex]{subfiles}
+
 \begin{document}
-
-\section{Matplottoy}
-
-We build on the existing Matplotlib architecture \cite{hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and rely on Matplotlib for the other graphical elements of the visualization and the rendering.
-We first introduce an implementation of scatter, line, and bar charts because they map to the fundemental marks of point, line, and area.
-We then introduce aggregated bar charts to show how to build more complex graphics.
-
-\subsection{Scatter}
- A scatter plot \cite{friendlyBriefHistoryData2006a,tukeyExploratoryDataAnalysis1977} is a chart type for plotting discrete values against each other.
-Minimally a scatter plot requires and x or y position, but the other variables in our dataset can be mapped to visual aspects of the scatter graphical mark.
+%what, why, how
+\section{Prototype Implementation: Matplottoy}
+\label{sec:implementation}
 \begin{figure}[H]
+    \begin{subfigure}{0.5\textwidth}
         \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
-    \label{fig:scatter}
+    \end{subfigure}
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/line_1.png}
+    \end{subfigure}
+    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
+    \label{fig:code_scatter_line}
 \end{figure}
 
-For the scatter plot in figure~\ref{fig:scatter}, we define $Q$ as 
- \begin{equation}
-    Q(\mu_{xpos}, \mu_{ypos}, \mu_{facecolor}, \mu_{s})
- \end{equation}
+We implemented a proof of concept of the topological model to demonstrate its feasibility.
+We implemented the artist classes for the scatter, line, and bar plots shown in figure~\ref{fig:code_scatter_line} because scatter and line have different continuities \gbase\ and visual bundles \vtotal\.
+We built our prototype on top of the existing Matplotlib architecture \cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} so that we can initially focus on the data to graphic transformations and rely on Matplotlib for the rendering.
 
-which we implement building on the \texttt{matplotlib.collections} API:
+To generate figure~\ref{fig:code_scatter_line} in our prototype, we set up a canvas \mintinline{python}{fig} and an axes to draw on \mintinline{python}{ax} and 
+\begin{multicols*}{2}
+\begin{minted}[linenos]{python}
+    fig, ax = plt.subplots()
+    artist = Point(table, transforms)
+    ax.add_artist(artist)
+\end{minted}
+\columnbreak
+\begin{minted}[linenos]{python}
+    fig, ax = plt.subplots()
+    artist = Line(table, transforms)
+    ax.add_artist(artist)
+\end{minted}
+\end{multicols*}
+ and then pass in \dsection=\mintinline{python}{table} and \vchannel=\mintinline{python}{transforms} into equivalence classes \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{python}{Line}.
 
-\begin{minted}{python}
-    class Point(mcollections.Collection):
-        # this is the visual fiber 
-        required = {'x', 'y'}
-        optional = {'facecolors', 's'} 
+\subsubsection{Artist Class $\vartist^{\prime}$}
+What, why, how>
+We implement the \vartist\ functions that transform data to visual representations as the equivalence class of artists such that the user configure the visual sections \vsection\ as a parameter \mintinline{python}{transforms}
+
+\begin{minted}[linenos]{python}
+    class ArtistClass(matplotlib.artist.Artist):
         def __init__(self, data, transforms, *args, **kwargs):
-            """
-            Parameters
-            ----------
-            data: sections of the fiber bundle
-            transforms
-            """
             super().__init__(*args, **kwargs)
-
-            # check that the data you're trying to transform 
-            # has a way to provide vertex data  
-            assert 'vertex' in data.FB.K['tables']
-            # check that you've given the required parameters
-            utils.check_constraints(Point, transforms.keys())
-            utils.validate_transforms(data.FB.F, transforms)
-
-            self.data = data
-            self.transforms = transforms
-
         def draw(self, renderer, *args, **kwargs):
-            view = self.data.view('vertex') #resolve to size
-
-            # call tau step
-            visual = utils.convert_transforms(view, self.transforms)
-
-            # assembles taus to generate idiom
-            # dictionary here in place of visual(parameter) function
-
-            visual['s'] = itertools.repeat(visual.get('s', 0.05))
-            visual['facecolors'] = visual.get('facecolors', "C0")
-            #switch out to a marker 
-            self._paths = [mpath.Path.circle(center=(x,y), radius=s)  
-                            for (x, y, s) 
-                            in zip(visual['x'],visual['y'], visual['s'])]
-
-            self.set_facecolors(visual['facecolors'])
             super().draw(renderer, *args, **kwargs)
@@
 
+As mentioned in section~\ref{sec:artist_equivalance}, it would be impractical to implement graph types as \vartist\ functions where all the visual parameters are specified.
+
+\subsubsection{Encoders \vchannel}
+what, why, how?
+\subsubsection{Data \dtotal}
+what, why, how?
+
+\subsubsection{Case Study: Iris}
+
 \end{document}

### 2021-01-29 draft/sections/math.tex
start of reworked implementation

@@
 \subsubsection{Equivalance class of artists $\vartist^{\prime}$}
+\label{sec:artist_equivalance}
 As formulated above, every artist function \vartist\ has fixed \vchannel\ and \vmark\ which  generates a distinct graphic \gsection.

### 2021-01-29 code/matplottoy/artists/line.py

@@ -22,7 +22,7 @@ class Line(mcollections.LineCollection):
         """
         super().__init__(None, *args, **kwargs)
         assert 'edge' in data.FB.K['tables']
-        utils.check_constraints(Line, transforms)
+        utils.check_constraints(Line, transforms.keys())
         utils.validate_transforms(data.FB.F, transforms)
      
         self.data = data

### 2021-02-01 draft/sections/implementation.tex
veen rereading the aos and hunter papers

@@
 
-To generate figure~\ref{fig:code_scatter_line} in our prototype, we set up a canvas \mintinline{python}{fig} and an axes to draw on \mintinline{python}{ax} and 
+To generate the images in figure~\ref{fig:code_scatter_line}, we use the figure \mintinline{python}{fig}  artist and the axes \mintinline{python}{ax} artist that lies inside the figure as the canvas for our graphic.
+
 \begin{multicols*}{2}
@@
 \end{multicols*}
- and then pass in \dsection=\mintinline{python}{table} and \vchannel=\mintinline{python}{transforms} into equivalence classes \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{python}{Line}.
+ We then add to the axes our prototype artist equivalance classes \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{python}{Line}.
+We pass in a section of the data \dsection=\mintinline{python}{table} and  paramterize the artist using the \vchannel=\mintinline{python}{transforms} dictionary.
 

### 2021-02-01 draft/sections/math.tex
veen rereading the aos and hunter papers

@@
 
+%% whats the right model/abstraction for a visualization library? 
+%% why? the role this math is serving - static type checking/ reflection
+%% tracking/automatically inferring
+%% bookend w/ how this math does the why 
+%% if we deleted this, what would we lose? 
+%% can express continuous functions & not just discrete points
 \note{should this be in 3rd person passive?}\\
@@
 \label{sec:data_fiber}
+%%% so we can do units, static + dynamic type checking, so that we can tak about variables 
 The fiber is a topological space that is the set of possible values of the data; the values themselves can be any dimension and type and have any continuity.
@@
 \label{sec:data_monoid}
+%% pin down the structure we're trying to preserve, this is what we mean by equivarance, most general useful formalism is a monoid, most specifically, more commonly it's a group but sometimes it's a mpnoid
 After specifying \dfiber\, we next describe the ways in which we can transform the values by identifying the monoid actions \monoid\ on the \dfiber.
@@
 \label{sec:data_base}
+% scatter and line, heatmap express the continuity of the data, deeply implicit in every visualization system but equivariance of continuity of the records, explicit  topplogy/structure preserving - contuinity is the topology
 The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the base space \dbase\ without making assumptions as to what that continuity is.
@@
 which is a decomposition of \dfiber. The \dbase\ remains the same because the connectivity of records does not change just because there are fewer elements in each record.
-
 \begin{figure}[ht!]
@@
 \label{sec:data_section}
+%% how do we represent data
 While the fiber and base space describe the general structure of all data that lives in the fiber bundle, the sections $\dsection: \dbase\rightarrow \dtotal$ define the datasets that live in the fiber.
@@
 
+%% here are some of our real data structures and here is how it works: dataframe, movies, etc...
+%% all the same, special versions of fiber bundles
+
 \subsubsection{Sheaf and Stalk}
@@
 \label{sec:graphic_fiber}
+%% bookends sentences
+%% first we need a  target output abstraction 
+%% this allows us to w/ some level of detail specify a target display
 The fiber \gfiber\ is an idealized infinite resolution version of the target display space, for example a 2D screen or 3D printer.
@@
 To support overplotting and transparency, the fiber could be $\dfiber=\reals^{7}$ such that $(x, y, z, r, g, b, a) \in \gfiber$ specifies the target display.
-The location coordinates $x$ and $y$ are defined in terms of the display, while the $(r,g,b)$ values are filled in via a lookup on \gbase.
 
@@
 \label{sec:graphic_base}
+%% structure preservation on the graphics side 
 An assumption of graphical representations is that they match the continuity of the data\cite{tufteVisualDisplayQuantitative2001,friendlyBriefHistoryData2008}, but the underlying topology \gbase\ of a graphic may need more dimensions than the data topology \dbase\ so that the glyph can be defined in \dfiber.
@@
 
-\subsubsection{Renderable Glyph \gsection}
+\subsubsection{Renderering \gsection}
 \label{sec:graphic_section}
-%pick a pixel which is the bounding box
+% couple nu + Q , visual encodings 
+% pick a pixel which is the bounding box
 A section $\gsection: \gbase\rightarrow \gtotal$ defines a piece of the graphical representation of the data.
@@
 
+%% pdf/svg/etc are different ways of implementing rhos
+
 \subsection{Artist}
 \label{sec:artist}
+%%% line things up such that they're an equivariant maps 
+%%% make a little clearer what E' is 
 The artist is the function that converts data into graphics; its name is taken from the analogous part of Matplotlib\cite{hunterArchitectureOpenSource} that builds visual elements to pass off to the renderer.
 The artist \vartist\ is a mapping from \dtotal\ padded with data from $\mathcal{J}(\dtotal)$ to a graphic that is a section \gsection\ in  $\Gamma(\gtotal)$
+%%% \Gamma is total
+%%% H is point wise -
+%%% this map means we cna do this pointwise 
+%%% explain the arrows, chase through here
+%%% new disagram w/ section
+
 \begin{equation}
@@
     \begin{tikzcd}
-        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \Gamma(\gtotal) \arrow[ld, "\pi"] \\
+        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \gtotal \arrow[ld, "\pi"] \\
                                               & \dbase                  & \gbase \arrow[l, "\vindex"']                                              &                    
@@
 \end{equation}
+%%% Q \circ \xi  \nu = \rho, and nu is on tau, and sections are weaker 
 with an intermediate fiber bundle \vtotal\ to hold visual representations and stages
@@
 \subsubsection {Visual Fiber Bundle \vtotal}
+%% structure preserving on the visual side/target space, visualization library 
+%% equivariant, latent space of visual parameters, intermediate 
+%% don't formally need, let it more cleanly pin down the the fiber equivarance , expresisng the concept that is already in literature,
+
 The visual fiber bundle (\vtotal, \dbase, $\pi$, \vfiber) has section $\vsection: \vtotal \rightarrow \dbase$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber \vfiber.
@@
 \label{sec:artist_nu}
+%% equivariant maps 
 \begin{figure}
@@
 
+\subsubsection{+ union}
+\begin{figure}
+    \caption{components of axes + glyphs }
+\end{figure}
+disjoin union of fibers so we can have an axes/real figure, - diagram operator w/o overplotting, library won't auto stack, have to bake that into data, haskell data/depth, 
+hover, mouse over curve, picked a point selecting a little k, + operator/disjoint union in different k, same D
+%% but what if it;s supposed to be Bertin invariant? will do w/ inclusion of K
+
 \subsubsection{Equivalance class of artists $\vartist^{\prime}$}

### 2021-02-02 draft/sections/implementation.tex
artist + scatter

--- 2021-01-29 draft/sections/implementation.tex
@@
 
-We implemented a proof of concept of the topological model to demonstrate its feasibility.
-We implemented the artist classes for the scatter, line, and bar plots shown in figure~\ref{fig:code_scatter_line} because scatter and line have different continuities \gbase\ and visual bundles \vtotal\.
-We built our prototype on top of the existing Matplotlib architecture \cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} so that we can initially focus on the data to graphic transformations and rely on Matplotlib for the rendering.
+To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute of our artist: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.
+We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations.
 
-To generate figure~\ref{fig:code_scatter_line} in our prototype, we set up a canvas \mintinline{python}{fig} and an axes to draw on \mintinline{python}{ax} and 
+To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primative objects we implemented based on our topology model.
+
 \begin{multicols*}{2}
-\begin{minted}[linenos]{python}
+\begin{minted}{python}
     fig, ax = plt.subplots()
-    artist = Point(table, transforms)
+    artist = Point(data, transforms)
     ax.add_artist(artist)
@@
 \columnbreak
-\begin{minted}[linenos]{python}
+\begin{minted}{python}
     fig, ax = plt.subplots()
-    artist = Line(table, transforms)
+    artist = Line(data, transforms)
     ax.add_artist(artist)
@@
 \end{multicols*}
- and then pass in \dsection=\mintinline{python}{table} and \vchannel=\mintinline{python}{transforms} into equivalence classes \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{python}{Line}.
+
+We then add the \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{Line} artists that construct the scatter and line graphics.
+The arguments to the artist are the data \dtotal=\mintinline{data} that is to be plotted and the aesthetic configuration \vchannel=\mintinline{transforms}.
+We implement the artists as equivalance classes \vartisteq\ because it would be impractical to implement a new artist for every aesthetic setting, such as one artist for red lines and another for green.
 
@@
 What, why, how>
-We implement the \vartist\ functions that transform data to visual representations as the equivalence class of artists such that the user configure the visual sections \vsection\ as a parameter \mintinline{python}{transforms}
-
-\begin{minted}[linenos]{python}
+%q \circ nu (\xi^{-1}\tau(s)) = rho (s)
+The artist is the piece of the matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic.
+In the prototype artist, the \mintinline{transform} argument is a collection of  visual encoders \vchannel\ and the data bundle \vtotal\ is passed in as \mintinline{data}.
+By binding both parameters to \vartisteq\ inside \mintinline{__init__}, the \mintinline{draw} method is a fully specified artist \vartist.
+\begin{minted}{python}
 class ArtistClass(matplotlib.artist.Artist):
     def __init__(self, data, transforms, *args, **kwargs):
+        # check that constraints are met such that the artist can operate on data with the given transforms
+        self.data = data 
+        self.transforms = transforms
         super().__init__(*args, **kwargs)
+
     def draw(self, renderer, *args, **kwargs):
+        # returns K, indexed on fiber then key 
+        view = self.data.view() 
+        # visual channel encoding applied fiberwise 
+        visual = [nu(view[var]) for var, nu in 
+                    self.transforms]
+        # reindex from k to s
+        # assemble glyph              
+        # pass configurations off to the render
         super().draw(renderer, *args, **kwargs)
 \end{minted}
+We fetch the data via a \mintinline{python}{view} method on the data inside \mintinline{python}{draw} so that we do not have to construct a new artist object for every update of the data.
+We then apply the \vchannel\ maps to the data to generate the \vsection=\mintinline{python}{visual} input to \vmark.
+We could also apply the \vindex\ mapping such that $\mintinline{python}{visual}=\vindex^*\vsection$, but choose not to because here there is no benefit to either choice and \vsection\ is slightly easier to work with.
+We then assemble the visual variables to assemble the glyph, currently in a way that sets the attributes of the Matplotlib primitive artist it inherits from.
 
-As mentioned in section~\ref{sec:artist_equivalance}, it would be impractical to implement graph types as \vartist\ functions where all the visual parameters are specified.
+For example, the point artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a set of glyphs.
+The \mintinline{python}{view} method repackages the data in a triangulated form as described in section~\ref{sec:triangulization}.
+
+\begin{minted}{python}
+class Point(Collection):
+    ## init as described above
+    def draw(self, renderer, *args, **kwargs):
+        # query data for a vertex table K
+        view = self.data.view('vertex') 
+        visual = [nu(view[var]) for var, nu in self.transforms]
+        # no explicit \xi, k and s have the same indexing
+        # construct geometries of the circle glyphs in visual coordinates
+        circles = [circle(center=(x,y), radius=s) for (x, y, s) 
+                in zip(visual['x'],visual['y'], visual['s'])] 
+        # set attributes of glyphs, these are vectorized 
+        # circles and facecolors are lists of the same size
+        self._paths = circles
+        self.set_facecolors(visual['facecolors'])
+        # call the renderer that will draw based on properties
+        super().draw(renderer, *args, **kwargs)
+\end{minted}
 

### 2021-02-02 draft/sections/implementation.tex
redoing encoders to make translation from math cleaner

@@
 
-To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primative objects we implemented based on our topology model.
+To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primitive objects we implemented based on our topology model.
 
@@
 
-We then add the \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{Line} artists that construct the scatter and line graphics.
-The arguments to the artist are the data \dtotal=\mintinline{data} that is to be plotted and the aesthetic configuration \vchannel=\mintinline{transforms}.
-We implement the artists as equivalance classes \vartisteq\ because it would be impractical to implement a new artist for every aesthetic setting, such as one artist for red lines and another for green.
+We then add the \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{python}{Line} artists that construct the scatter and line graphics.
+The arguments to the artist are the data \dtotal=\mintinline{python}{data} that is to be plotted and the aesthetic configuration \vchannel=\mintinline{python}{transforms}.
+We implement the artists as equivalence classes \vartisteq\ because it would be impractical to implement a new artist for every aesthetic setting, such as one artist for red lines and another for green.
 
@@
 The artist is the piece of the matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic.
-In the prototype artist, the \mintinline{transform} argument is a collection of  visual encoders \vchannel\ and the data bundle \vtotal\ is passed in as \mintinline{data}.
-By binding both parameters to \vartisteq\ inside \mintinline{__init__}, the \mintinline{draw} method is a fully specified artist \vartist.
+In the prototype artist, \mintinline{python}{transform} is a dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber,  and the \vchannel\ encoders are passed in as functions or callable objects.
+The data bundle \vtotal\ is passed in as a \mintinline{python}{data} object.
+By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist.
 \begin{minted}{python}
@@
     def __init__(self, data, transforms, *args, **kwargs):
-        # check that constraints are met such that the artist can operate on data with the given transforms
+        # check that constraints are met such that the artist 
+        # can operate on data with the given transforms
         self.data = data 
@@
         # visual channel encoding applied fiberwise 
-        visual = [nu(view[var]) for var, nu in 
-                    self.transforms]
+        visual = {p: encoder(view.get(f, None)) for 
+                     p, (f, encoder) in self.transforms.items()}
         # reindex from k to s
@@
         view = self.data.view('vertex') 
-        visual = [nu(view[var]) for var, nu in self.transforms]
+        visual = {p: encoder(view.get(f, None)) for
+                     p, (f, encoder) in self.transforms.items()}
         # no explicit \xi, k and s have the same indexing
         # construct geometries of the circle glyphs in visual coordinates
-        circles = [circle(center=(x,y), radius=s) for (x, y, s) 
+        circles = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
                 in zip(visual['x'],visual['y'], visual['s'])] 

### 2021-02-02 code/matplottoy/artists/line.py

@@ -22,7 +22,7 @@ class Line(mcollections.LineCollection):
         """
         super().__init__(None, *args, **kwargs)
         assert 'edge' in data.FB.K['tables']
-        utils.check_constraints(Line, transforms.keys())
+        utils.check_constraints(Line, transforms)
         utils.validate_transforms(data.FB.F, transforms)
      
         self.data = data

### 2021-02-02 code/matplottoy/artists/point.py

@@ -24,29 +24,25 @@ class Point(mcollections.Collection):
         
         # check that the data you're trying to transform 
         # has a way to provide vertex data  
-        assert 'vertex' in data.FB.K['tables']
+        # assert 'vertex' in data.FB.K['tables']
         # check that you've given the required parameters
-        utils.check_constraints(Point, transforms.keys())
-        utils.validate_transforms(data.FB.F, transforms)
+        #utils.check_constraints(Point, transforms.keys())
+        #utils.validate_transforms(data.FB.F, transforms)
 
         self.data = data
         self.transforms = transforms
 
     def draw(self, renderer, *args, **kwargs):
-        view = self.data.view('vertex') #resolve to size
-
-        # call tau step
-        visual = utils.convert_transforms(view, self.transforms)
-           
-        # assembles taus to generate idiom
-        # dictionary here in place of visual(parameter) function
-       
-        visual['s'] = itertools.repeat(visual.get('s', 0.05))
-        visual['facecolors'] = visual.get('facecolors', "C0")
-        #switch out to a marker 
-        self._paths = [mpath.Path.circle(center=(x,y), radius=s)  
-                        for (x, y, s) 
-                        in zip(visual['x'],visual['y'], visual['s'])] 
-       
+        # query data for a vertex table K
+        view = self.data.view('vertex') 
+        visual = {p: encoder(view.get(f, None)) for p, (f, encoder) in self.transforms.items()}
+        # no explicit \xi, k and s have the same indexing
+        # construct geometries of the circle glyphs in visual coordinates
+        circles = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
+                in zip(visual['x'],visual['y'], visual['s'])] 
+        # set attributes of glyphs, these are vectorized 
+        # circles and facecolors are lists of the same size
+        self._paths = circles
         self.set_facecolors(visual['facecolors'])
+        # call the renderer that will draw based on properties
         super().draw(renderer, *args, **kwargs)

### 2021-02-02 code/matplottoy/encoders/color.py

@@ -10,7 +10,7 @@ class Identity:
         return mtype in ['nominal']
         
     @staticmethod
-    def convert(value):
+    def __call__(value):
         values = np.atleast_1d(np.array(value, dtype=object))
         return [mcolors.to_rgba(v) for v in values]
 
@@ -21,7 +21,7 @@ class Categorical:
         assert(mcolors.is_color_like(color) for color in mapping.values())
         self._mapping = mapping
 
-    def convert(self, value):
+    def __call__(self, value):
         values = np.atleast_1d(np.array(value, dtype=object))
         return [mcolors.to_rgba(self._mapping[v]) for v in values]
 

### 2021-02-02 code/matplottoy/encoders/position.py

@@ -6,7 +6,7 @@ import numpy as np
 #TODO: write atleast_1d wrapper
 class Identity:
     @staticmethod
-    def convert(value):
+    def __call__(value):
         return np.array(value)
     @staticmethod
     def validate(mtype):
@@ -16,16 +16,17 @@ class Log:
     def __init__(self, base=10):
         self.base = base
 
-    def convert(self, value):
+    def __call__(self, value):
         return np.log(value)/np.log(self.base)
 
     def validate(self, mtype):
         return True
+    
 class Nominal:
     def __init__(self, mapping):
         self._mapping = mapping
         
-    def convert(self, value):
+    def __call__(self, value):
         values = np.atleast_1d(np.array(value, dtype=object))
         return np.array([self._mapping[v] for v in values])
 

### 2021-02-03 draft/sections/implementation.tex
q hat + filter out glyph

@@
 
-\subsubsection{Artist Class $\vartist^{\prime}$}
-What, why, how>
-%q \circ nu (\xi^{-1}\tau(s)) = rho (s)
+\subsection{Artist Class $\vartist^{\prime}$}
 The artist is the piece of the matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic.
 In the prototype artist, \mintinline{python}{transform} is a dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber,  and the \vchannel\ encoders are passed in as functions or callable objects.
-The data bundle \vtotal\ is passed in as a \mintinline{python}{data} object.
+The data bundle \dtotal\ is passed in as a \mintinline{python}{data} object.
 By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist.
+
 \begin{minted}{python}
@@
     def __init__(self, data, transforms, *args, **kwargs):
-        # check that constraints are met such that the artist 
-        # can operate on data with the given transforms
+        # properties that are specific to the graphic but not the channels
         self.data = data 
@@
         super().__init__(*args, **kwargs)
+
+    def assemble(self, visual):
+        # set the properties of each glyph
 
@@
                      p, (f, encoder) in self.transforms.items()}
-        # reindex from k to s
-        # assemble glyph              
+        self.assemble(visual)
         # pass configurations off to the render
@@
 \end{minted}
-We fetch the data via a \mintinline{python}{view} method on the data inside \mintinline{python}{draw} so that we do not have to construct a new artist object for every update of the data.
-We then apply the \vchannel\ maps to the data to generate the \vsection=\mintinline{python}{visual} input to \vmark.
-We could also apply the \vindex\ mapping such that $\mintinline{python}{visual}=\vindex^*\vsection$, but choose not to because here there is no benefit to either choice and \vsection\ is slightly easier to work with.
-We then assemble the visual variables to assemble the glyph, currently in a way that sets the attributes of the Matplotlib primitive artist it inherits from.
 
-For example, the point artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a set of glyphs.
-The \mintinline{python}{view} method repackages the data in a triangulated form as described in section~\ref{sec:triangulization}.
+The data is fetched in section \dsection\ via a \mintinline{python}{view} method n the data because the input to the artist is a section on \dtotal.
+The return view object has a \mintinline{python}{get} method to support querying for components that are not in \dfiber\, which we exploit to support parameters in the visual fiber that are not bound to fiber components in \dfiber.
+The \vchannel\ functions are then applied to the data to generate the \vsection=\mintinline{python}{visual} input to \vmark.
+An explicit \vindex\ is not implemented since that would mean copying a single \vsection on \dbasepoint to all the associated \gbasepoint, as illustrated in figure~\ref{fig:graphic_retraction_map}, and that is unnecessary overhead for these scatter and line plots.
+In \vmarkd=\mintinline{python}{assemble} the artist generates instructions for the render by setting  the attributes that are related to the graphic.
+These are the settings that would have to be serialized in order to recreate a static version of the graphic.
+The \vchannel\ functions could be evaluated in this function to avoid passing over \dbase twice.
+The last step in the artist function is handing itself off to the renderer.
 
+The \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of simple glyphs.
+In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker.
 \begin{minted}{python}
-class Point(Collection):
+class Point(mcollections.Collection):
     ## init as described above
@@
                      p, (f, encoder) in self.transforms.items()}
-        # no explicit \xi, k and s have the same indexing
+        # no explicit  S->K since they have the same indexing
         # construct geometries of the circle glyphs in visual coordinates
-        circles = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
+        self._paths = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
                 in zip(visual['x'],visual['y'], visual['s'])] 
@@
         # circles and facecolors are lists of the same size
-        self._paths = circles
         self.set_facecolors(visual['facecolors'])
@@
 \end{minted} 
+The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices, as described in section~\ref{sec:triangulization}; even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$ such that all the data on one vertex maps to one marker.
+To ensure the integrity of the section, \mintinline{python}{view} must be atomic, meaning that the values cannot change after the method is called in draw until a new call in draw.
+
+This table is converted to a table of visual variables; it is then used to individually construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
+Since \mintinline{python}{view} returns a \desection\, all these operations could be applied on one element in the data or multiple.
+\note{this might be better suited for discussion/future work} We could also rewrite the base \vmark\ such that \mintinline{python}{draw} acts on only one point on \dbase\ or even \gbase\, but for small datasets it is more efficient to vectorize these operations.
+
+The only difference between the \mintinline{python}{Point} and \mintinline{python}{Line} objects is in the \mintinline{python}{view} and in the \vmark\ section because line has different continuity from scatter.
+
+\begin{minted}{python}
+class Line(mcollections.LineCollection):
+    def draw(self, renderer, *args, **kwargs):
+        view = self.data.view('edge')
+        visual = {p: encoder(view.get(f, None)) for 
+                      p, (f, encoder) in self.transforms.items()}
+        # glyph assembly, where the glyph is a line
+        segments = [np.vstack((vx, vy)).T for vx, vy 
+                    in zip(visual['x'], visual['y'])]
+        self.set_segments(segments)
+        self.set_color(visual['color'])
+
+        super().draw(renderer, *args, **kwargs)
+\end{minted}
+In the \mintinline{python}{Line} artist, \mintinline{python}{view} returns a table of edges.
+Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge.
+This information is converted into visual variables that are then composed into colored line segments.
 

### 2021-02-03 code/matplottoy/artists/line.py

@@ -21,9 +21,9 @@ class Line(mcollections.LineCollection):
         transforms
         """
         super().__init__(None, *args, **kwargs)
-        assert 'edge' in data.FB.K['tables']
-        utils.check_constraints(Line, transforms)
-        utils.validate_transforms(data.FB.F, transforms)
+        #assert 'edge' in data.FB.K['tables']
+        #utils.check_constraints(Line, transforms)
+        #utils.validate_transforms(data.FB.F, transforms)
      
         self.data = data
         self.transforms = transforms
@@ -31,14 +31,11 @@ class Line(mcollections.LineCollection):
     def draw(self, renderer, *args, **kwargs):
       
         view = self.data.view('edge')
-        visual = utils.convert_transforms(view, self.transforms)
-        # dictionary here in place of visual(parameter) function
-        if 'color' not in visual:
-            visual['color'] = "C0"
-
+        visual = {p: encoder(view.get(f, None)) for 
+                      p, (f, encoder) in self.transforms.items()}
+    
         segments = [np.vstack((vx, vy)).T for vx, vy 
                     in zip(visual['x'], visual['y'])]
-
         self.set_segments(segments)
         self.set_color(visual['color'])
 

### 2021-02-03 code/matplottoy/artists/point.py

@@ -38,11 +38,10 @@ class Point(mcollections.Collection):
         visual = {p: encoder(view.get(f, None)) for p, (f, encoder) in self.transforms.items()}
         # no explicit \xi, k and s have the same indexing
         # construct geometries of the circle glyphs in visual coordinates
-        circles = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
+        self._paths = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
                 in zip(visual['x'],visual['y'], visual['s'])] 
         # set attributes of glyphs, these are vectorized 
         # circles and facecolors are lists of the same size
-        self._paths = circles
         self.set_facecolors(visual['facecolors'])
         # call the renderer that will draw based on properties
         super().draw(renderer, *args, **kwargs)

### 2021-02-03 draft/sections/implementation.tex
qhat for point, line

@@
     def assemble(self, visual):
-        # set the properties of each glyph
+        # set the properties of the graphic
 
@@
         self.assemble(visual)
-        # pass configurations off to the render
+        # pass configurations off to the renderer
         super().draw(renderer, *args, **kwargs)
@@
 
-The data is fetched in section \dsection\ via a \mintinline{python}{view} method n the data because the input to the artist is a section on \dtotal.
+The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
 The return view object has a \mintinline{python}{get} method to support querying for components that are not in \dfiber\, which we exploit to support parameters in the visual fiber that are not bound to fiber components in \dfiber.
@@
 These are the settings that would have to be serialized in order to recreate a static version of the graphic.
-The \vchannel\ functions could be evaluated in this function to avoid passing over \dbase twice.
+Although \mintinline{python}{assemble} could be implemented outside the class such that it returns an object the artist could then parse to set attributes, the attributes are directly set here to reduce indirection.
+The \vchannel\ functions could be evaluated in this function to avoid passing over \dbase twice but are not done so here to demonstrate the seperability of \vchannel\ and \vmarkd\.
 The last step in the artist function is handing itself off to the renderer.
 
-The \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of simple glyphs.
+The \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of primitive point and area marks.
 In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker.
@@
 class Point(mcollections.Collection):
-    ## init as described above
+    def __init__(self, data, transforms, *args, **kwargs):
+        super().__init__(*args, **kwargs)
+        self.data = data
+        self.transforms = transforms
+
+    def assemble(self, visual):
+        # construct geometries of the circle marks in visual coordinates
+        self._paths = [mpath.Path.circle(center=(x,y), radius=s) 
+                       for (x, y, s) in zip(visual['x'],visual['y'], visual['s'])] 
+        # set attributes of marks, these are vectorized 
+        # circles and facecolors are lists of the same size
+        self.set_facecolors(visual['facecolors'])
+
     def draw(self, renderer, *args, **kwargs):
@@
                      p, (f, encoder) in self.transforms.items()}
-        # no explicit  S->K since they have the same indexing
-        # construct geometries of the circle glyphs in visual coordinates
-        self._paths = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
-                in zip(visual['x'],visual['y'], visual['s'])] 
-        # set attributes of glyphs, these are vectorized 
-        # circles and facecolors are lists of the same size
-        self.set_facecolors(visual['facecolors'])
+        self.assemble(visual)
         # call the renderer that will draw based on properties
         super().draw(renderer, *args, **kwargs)
+
 \end{minted} 
@@
 To ensure the integrity of the section, \mintinline{python}{view} must be atomic, meaning that the values cannot change after the method is called in draw until a new call in draw.
+This table is converted to a table of visual variables.
+It is then passed into \mintinline{python}{assemble}, where it is used to individually construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
+Since \mintinline{python}{view} returns a \dsection\, all these operations could be applied on a section on one \dbasepoint\ or a subset of \dbase.
 
-This table is converted to a table of visual variables; it is then used to individually construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
-Since \mintinline{python}{view} returns a \desection\, all these operations could be applied on one element in the data or multiple.
-\note{this might be better suited for discussion/future work} We could also rewrite the base \vmark\ such that \mintinline{python}{draw} acts on only one point on \dbase\ or even \gbase\, but for small datasets it is more efficient to vectorize these operations.
-
-The only difference between the \mintinline{python}{Point} and \mintinline{python}{Line} objects is in the \mintinline{python}{view} and in the \vmark\ section because line has different continuity from scatter.
+The only difference between the \mintinline{python}{Point} and \mintinline{python}{Line} objects is in the \mintinline{python}{view} and \mintinline{python}{assemble} function because line has different continuity from scatter and is represented by a different type of graphical mark.
 
@@
 class Line(mcollections.LineCollection):
-    def draw(self, renderer, *args, **kwargs):
-        view = self.data.view('edge')
-        visual = {p: encoder(view.get(f, None)) for 
-                      p, (f, encoder) in self.transforms.items()}
-        # glyph assembly, where the glyph is a line
+    def assemble(self, visual):
+        #assemble line marks as set of segments 
         segments = [np.vstack((vx, vy)).T for vx, vy 
@@
 
+    def draw(self, renderer, *args, **kwargs):
+        # query data source for edge table
+        view = self.data.view('edge')
+        visual = {p: encoder(view.get(f, None)) for 
+                     p, (f, encoder) in self.transforms.items()}
+        self.assemble(visual)
         super().draw(renderer, *args, **kwargs)
+
 \end{minted}
@@
 Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge.
-This information is converted into visual variables that are then composed into colored line segments.
+As with \mintinline{python}{Point}, the data is then converted into visual data.
+In \mintinline{python}{assemble}, this visual representation is composed into a set of line segments and then the colors of each line segment are set.
+The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on view.
 

### 2021-02-03 code/matplottoy/artists/line.py

@@ -28,16 +28,18 @@ class Line(mcollections.LineCollection):
         self.data = data
         self.transforms = transforms
 
-    def draw(self, renderer, *args, **kwargs):
-      
-        view = self.data.view('edge')
-        visual = {p: encoder(view.get(f, None)) for 
-                      p, (f, encoder) in self.transforms.items()}
-    
+    def assemble(self, visual):
+        #assemble line marks as set of segments 
         segments = [np.vstack((vx, vy)).T for vx, vy 
                     in zip(visual['x'], visual['y'])]
         self.set_segments(segments)
         self.set_color(visual['color'])
-
+        
+    def draw(self, renderer, *args, **kwargs):
+        # query data source for edge table
+        view = self.data.view('edge')
+        visual = {p: encoder(view.get(f, None)) for 
+                     p, (f, encoder) in self.transforms.items()}
+        self.assemble(visual)
         super().draw(renderer, *args, **kwargs)
 

### 2021-02-03 code/matplottoy/artists/point.py

@@ -32,16 +32,18 @@ class Point(mcollections.Collection):
         self.data = data
         self.transforms = transforms
 
-    def draw(self, renderer, *args, **kwargs):
-        # query data for a vertex table K
-        view = self.data.view('vertex') 
-        visual = {p: encoder(view.get(f, None)) for p, (f, encoder) in self.transforms.items()}
-        # no explicit \xi, k and s have the same indexing
+    def assemble(self, visual):
         # construct geometries of the circle glyphs in visual coordinates
         self._paths = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
                 in zip(visual['x'],visual['y'], visual['s'])] 
         # set attributes of glyphs, these are vectorized 
         # circles and facecolors are lists of the same size
         self.set_facecolors(visual['facecolors'])
+        
+    def draw(self, renderer, *args, **kwargs):
+        # query data for a vertex table K
+        view = self.data.view('vertex') 
+        visual = {p: encoder(view.get(f, None)) for p, (f, encoder) in self.transforms.items()}
+        self.assemble(visual)
         # call the renderer that will draw based on properties
         super().draw(renderer, *args, **kwargs)

### 2021-02-03 draft/sections/implementation.tex
implementation up to encoding

@@
 Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge.
-As with \mintinline{python}{Point}, the data is then converted into visual data.
+As with \mintinline{python}{Point}, the data is then converted into visual variables.
 In \mintinline{python}{assemble}, this visual representation is composed into a set of line segments and then the colors of each line segment are set.
@@
 \subsubsection{Encoders \vchannel}
-what, why, how?
+As mentioned above, the encoding dictionary is specified by the visual fiber component, the corresponding data fiber component, and the mapping function.
+The visual parameter serves as the dictionary key because the visual representation is constructed from the encoding applied to the data  $\vsection = \vchannel \circ \dsection$.
+For the scatter plot, the mappings for the visual fiber components $\vfiber=(x,y, facecolors, s)$ are defined as
+\begin{minted}{python}
+cmap =  color.Categorical({'true':'deeppink', 'false':'deepskyblue'})
+transforms = {'y': ('v2', lambda x: x),       
+              'x': ('v4', lambda x: x),
+             'facecolors': ('v3', cmap), 
+             's':(None ,lambda _: itertools.repeat(.02))}
+\end{minted}
+where the position \textit{(x,y)} \vchannel\ transformers are identity functions.
+The size \textit{s} transformer is not acting on a component of \dfiber\, instead it is a \vchannel\ that returns a constant value.
+While size could be embedded inside the \mintinline{python}{assembly} function, it is added to the transformers to illustrate user configured visual parameters that could either be constant or mapped to a component in \dfiber.
+The identity and constant \vchannels\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they could be optimized away.
+More complex encoders can be implemented as callable classes, such as
+\begin{minted}{python}
+class Categorical:
+    def __init__(self, mapping):
+        # check that the conversion is to valid colors
+        assert(mcolors.is_color_like(color) for color in mapping.values())
+        self._mapping = mapping
+
+    def __call__(self, value):
+        # convert value to a color
+        return [mcolors.to_rgba(self._mapping[v]) for v in values]
+\end{minted}
+
+where \mintinline{python}{__init__} can validate that the output of the \vchannel\ is a valid element of the \vfiber\ component the \vchannel\ function is targeting.
+Creating a callable class also provides a simple way to swap out the specific (data, value) mapping without having to reimplement the validation or conversion logic.
+
+The check for equivariance can be implemented trivially such that it is independent of data or encoder.
+\begin{minted}{python}
+def is_nominal(values, encoder):
+    m1 = list(zip(values, encoder(values)))
+    random.shuffle(values)
+    m2 = list(zip(values, encoder(values)))
+    return sorted(m1) == sorted(m2)
+\end{minted}
+In this example, \mintinline{python}{is_nominal} checks for equivariance of permutation group actions by applying the encoder to a set of values, shuffling values, and checking that (value, encoding) pairs remain the same.
+\note{not sure where/how these checks should be incorporated, maybe as part of transform dict?}
+
 \subsubsection{Data \dtotal}

### 2021-02-03 code/matplottoy/encoders/color.py

@@ -25,6 +25,4 @@ class Categorical:
         values = np.atleast_1d(np.array(value, dtype=object))
         return [mcolors.to_rgba(self._mapping[v]) for v in values]
 
-    def validate(self, mtype):
-        return mtype in ['nominal']
 

### 2021-02-04 draft/sections/implementation.tex
code data section + refactor continous and discontinous line into one source object

@@
 
-To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute of our artist: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.
+To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.
 We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations.
@@
 \subsection{Artist Class $\vartist^{\prime}$}
+\label{sec:code_artist}
 The artist is the piece of the matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic.
@@
         # query data for a vertex table K
-        view = self.data.view('vertex') 
+        view = self.data.view() 
         visual = {p: encoder(view.get(f, None)) for 
@@
         # query data source for edge table
-        view = self.data.view('edge')
+        view = self.data.view()
         visual = {p: encoder(view.get(f, None)) for 
@@
 
-\subsubsection{Encoders \vchannel}
+\subsection{Encoders \vchannel}
+\label{sec:code_channels}
 As mentioned above, the encoding dictionary is specified by the visual fiber component, the corresponding data fiber component, and the mapping function.
@@
 cmap =  color.Categorical({'true':'deeppink', 'false':'deepskyblue'})
-transforms = {'y': ('v2', lambda x: x),       
-              'x': ('v4', lambda x: x),
-             'facecolors': ('v3', cmap), 
+transforms = {'y': ('v1', lambda x: x),       
+              'x': ('v3', lambda x: x),
+             'facecolors': ('v2', cmap), 
              's':(None ,lambda _: itertools.repeat(.02))}
@@
 where the position \textit{(x,y)} \vchannel\ transformers are identity functions.
-The size \textit{s} transformer is not acting on a component of \dfiber\, instead it is a \vchannel\ that returns a constant value.
+The size \textit{s} transformer is not acting on a component of \dfiber, instead it is a \vchannel\ that returns a constant value.
 While size could be embedded inside the \mintinline{python}{assembly} function, it is added to the transformers to illustrate user configured visual parameters that could either be constant or mapped to a component in \dfiber.
-The identity and constant \vchannels\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they could be optimized away.
+The identity and constant \vchannel\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they could be optimized away.
 More complex encoders can be implemented as callable classes, such as
@@
 
-The check for equivariance can be implemented trivially such that it is independent of data or encoder.
+A test for equivariance can be implemented trivially such that it is independent of data or encoder.
 \begin{minted}{python}
-def is_nominal(values, encoder):
+def test_nominal(values, encoder):
     m1 = list(zip(values, encoder(values)))
@@
     m2 = list(zip(values, encoder(values)))
-    return sorted(m1) == sorted(m2)
+    assert sorted(m1) == sorted(m2)
 \end{minted}
 In this example, \mintinline{python}{is_nominal} checks for equivariance of permutation group actions by applying the encoder to a set of values, shuffling values, and checking that (value, encoding) pairs remain the same.
-\note{not sure where/how these checks should be incorporated, maybe as part of transform dict?}
+This equivariance test can be implemented as part of the artist or encoder, but for minimal overhead, the equivariant it is implemented as part of the library tests.
 
-\subsubsection{Data \dtotal}
-what, why, how?
+\subsection{Data \dtotal}
+\label{sec:code_data}
+The data input into the \mintinline{pythin}{Artist} will often be a wrapper class around an existing data structure, but must meet the following criteria:
+\begin{enumerate}
+\item specify the fiber components \dfiber\ and connectivity \dbase\
+\item have a \mintinline{view} method that returns an atomic object that encapsulates \dsection\
+\item the view object must have \mintinline{get} method that returns a fiber component
+\end{enumerate}
+To support specifying the fiber bundle, we define an optional \mintinline{python}{FiberBundle} class
 
-\subsubsection{Case Study: Iris}
+\begin{minted}{python}
+class FiberBundle:
+    def __init__(self, base, fiber):
+        """
+        base:  {'tables': ['vertex', 'edge', 'face']}
+        fiber: {'component name': {'type':, 'monoid':,  'range':}}
+        """
+        self.base = base
+        self.fiber = fiber
+
+    def is_section(self, section):
+        """checks if a section is from a given fiber bundle:
+        are values in F, are keys in K"""
+\end{minted}
+
+that asks the user to specify how \dbase is triangulated and the attributes of \dfiber.
+The \mintinline{python}{assembly} functions expect tables that match the continuity of the graphic; scatter expects a vertex table because it is discontinuous, line expects an edge table because it is 1D continuous.
+The fiber informs appropriate choice of \vchannel\, therefore it is a dictionary of attributes of the fiber components.
+\note{I've basically stripped this out of the artists above so should I just ditch this section?}
+
+To generate the scatter plot in figure~\ref{fig:code_scatter_line}, we fully specify a dataset with random keys and values in a section chosen at random form the corresponding fiber component.
+The fiberbundle \mintinline{python}{FB} is a class level attribute since all instances of \mintinline{python}{code}{VertexSimplex} come from the same fiberbundle.
+\begin{minted}{python}
+class VertexSimplex: #maybe change name to something else
+    """Fiberbundle is consistent across all sections
+    """
+    FB = FiberBundle({'tables': ['vertex']},  
+            {'v1': {'type': float,'monoid':'interval','range': [0,1]},
+             'v2': {'type': str, 'monoid':'nominal', 'range':['true', 'false']},
+             'v3': {'type': float, 'monoid':'interval', 'range':[2,3]}})
+
+    def __init__(self, sid = 45, size=1000, max_key=10**10):
+        # create random list of keys
+    def tau(self, k):
+        # e1 is sampled from F1, e2 from F2, etc...
+        return (k, (e1, e2, e3, e4))
+
+    def view(self):
+        table = defaultdict(list)
+        for k in self.keys:
+            table['index'] = k
+            # on each iteration, add one (name, value) pair per component
+            for (name, value) in zip(self.FB.fiber.keys(), self.tau(k)[1]):
+                table[name].append(value)
+        return table
+\end{minted}
+The view method returns a dictionary where the key is a fiber component name and the value is a list of values in the fiber component.
+The table is built one call to \mintinline{python}{tau} at a time, guaranteeing that all the fiber component values are over the same \dbasepoint.
+Table has a \mintinline{python}{get} method as it is a method on Python dictionaries.
+In contrast, the line in \mintinline{python}{EdgeSimplex} is defined as the functions \mintinline{python}{_color,_xy} on each edge.
+
+\begin{minted}{python}
+class EdgeSimplex: 
+    # assign a class level FB attribute
+    def __init__(self, num_edges=4, num_samples=1000): 
+        self.keys = range(num_edge) #edge id
+        # distance along edge
+        self.distances = np.linspace(0,1, num_samples)
+        # half generlized representation of arcs on a circle
+        self.angle_samples = np.linspace(0, 2*np.pi, len(self.keys)+1)
+
+    @staticmethod
+    def _color(edge):
+        colors = ['red','orange', 'green','blue']
+        return colors[edge%len(colors)]
+
+    @staticmethod
+    def _xy(edge, distances, start=0, end=2*np.pi):
+        # start and end are parameterizations b/c really there is 
+        angles = (distances *(end-start)) + start
+        return np.cos(angles), np.sin(angles)
+
+    def tau(self, k): #will fix location on page on revision
+        x, y = self._xy(k, self.distances, 
+                        self.angle_samples[k], self.angle_samples[k+1]) 
+        color = self._color(k) 
+        return (k, (x, y, color))
+
+    def view(self, simplex):
+        table = defaultdict(list)
+        for k in self.keys:
+            table['index'].append(k)
+            # (name, value) pair, value is [x0, ..., xn] for x, y
+            for (name, value) in zip(self.FB.fiber.keys(), self.tau(k, simplex)[1]):
+                table[name].append(value)
+\end{minted}
+Unlike scatter, the line \mintinline{python}{tau} method returns the functions on the edge evaluated on the interval [0,1].
+By default these means each \mintinline{python}{tau} returns a list of 1000 x and y points and the associated color.
+As with scatter, \mintinline{python}{view} builds a table by calling \mintinline{python}{tau} for each \dbasepoint\.
+Unlike scatter, the line table is a list where each item contains a list of points.
+This bookkeeping of which data is on an edge is used by the \mintinline{python}{assembly} functions to bind segments to their visual properties.
+
+\begin{figure}[H]
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/linec_1.png}
+    \end{subfigure}
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/lined_1.png}
+    \end{subfigure}
+    \caption{Continuous and discontinuous lines as defined by different data models, but generated with the same \vartisteq=\mintinline{python}{artist}}
+    \label{fig:circle_step}
+\end{figure}
+
+The graphics in figure~\ref{fig:circle_step} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{Graphline} data source
+
+\begin{minted}{python}
+class GraphLine:
+    def __init__(self, FB, edge_table, vertex_table, num_samples=1000, connect=False):
+        #s set args as attributes and generate distance
+        if connect: # test connectivity if edges are continuous
+            assert edge_table.keys() == self.FB.F.keys()
+            assert is_continuous(vertex_table)
+
+    def tau(self, k, simplex='edge'):
+        # evaluates functions defined in edge table
+        return(k, (self.edges[c][k](self.distances) for c in self.FB.F.keys()))
+
+    def view(self, simplex='edge'):
+        """walk the edge_vertex table to return the edge function
+        """
+        table = defaultdict(list)
+        #sort since intervals lie along number line and are ordered pair neighbors
+        for (i, (start, end)) in sorted(zip(self.ids, self.vertices), key=lambda v:v[1][0]):
+            table['index'].append(i)
+            # same as view for line, returns nested list
+            for (name, value) in zip(self.FB.F.keys(), self.tau(i, simplex)[1]):
+                table[name].append(value)
+        return table
+\end{minted}
+where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix.
+The multicolored line is a connected graph of edges with each edge function evaluated on 1000 samples
+\begin{minted}{python}
+simplex.GraphLine(FB, edge_table, vertex_table, connect=True)
+\end{minted}
+while the stair chart is discontinuous and only needs to be evaluated at the edges of the interval 
+\begin{minted}{python}
+simplex.GraphLine(FB, edge_table, vertex_table, num_samples=2, connect=False)
+\end{minted}
+such that one advantage of this model is it helps differentiate graphics that have different artists from graphics that have the same artist but make different assumptions about the source data.
+
+\subsection{Case Study: Iris}
 

### 2021-02-04 draft/sections/math.tex
artist + scatter
q hat + filter out glyph
code data section + refactor continous and discontinous line into one source object

@@
 \label{sec:artist}
+%%somewhere in this I need to add q \circ nu (\xi^{-1}\tau(s)) = rho (s)
 %%% line things up such that they're an equivariant maps 
@@
 %%% H is point wise -
-%%% this map means we cna do this pointwise 
+%%% this map means we can do this pointwise 
 %%% explain the arrows, chase through here
-%%% new disagram w/ section
+%%% new disagram w/ section?
+% so you aren't passing around \tau
+% thos maps make sense for \tau(k)\in E -> V
+% and then you have \mu(k) 
+% and then when you are in \xi*V .... well the arrow goes the WONG way so you didn't just pass \mu(k) inot \xi*V
+% so if you want to go from \xi*V to H via Q you need to do it over a PARTICULAR s
+% thats what I mean by Q (in this diagram) hiding a choice of s. 
+% and then IF you pick an s ...... its is just fine. because it returns \rho(s)
+% so types match \rho(s) is in D
 
@@
     \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
-                                & linestyle                                                        & $\{f_{0}, \ldots, f_{n}\} \times (\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
+                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
     \end{tabulary}
@@
 \end{figure}
+%%big data
 One way of expressing the connectivity of records in a dataset is to implement \dbase\ as a simplacial complex, which is a set of simplicies such as those shown in figure~\ref{fig:triangle_simplex}.

### 2021-02-04 code/matplottoy/data/simplex.py

@@ -100,10 +100,10 @@ class EdgeSimplex: # ToDo: generalize to take list of functions as input
                                 'range': ['red', 'green', 'orange', 'blue']}
                     })
 
-    def __init__(self, num_verts=4, num_samples=1000):
+    def __init__(self, num_edges=4, num_samples=1000):
         """which simplex am I on and distance""" 
         # define the k and distance 
-        self.keys = range(num_verts)
+        self.keys = range(num_)
         self.num_samples = num_samples
         self.distances = np.linspace(0,1, num_samples)
         # concession to this is a half generlized representation of 
@@ -166,75 +166,14 @@ def is_continuous(vertices):
     # intervals along the 1D number line
     return (adj_mat == np.diag(np.diagonal(adj_mat))).all()
 
-class ContinuousLine:
-    FB = FiberBundle({'tables': ['vertex','edge']},
-                     {'x' : {'type:':float, 'monoid':'interval', 'range':[0,4]},
-                      'y': {'type:':float, 'monoid':'interval', 'range':[-1,1]},
-                      'color': {'type':str, 'monoid':'nominal', 
-                                'range': ['red', 'green', 'orange', 'blue']}
-                    })
-                    
-    def __init__(self, edge_table, vertex_table, num_samples=1000):
+class GraphLine:
+    def __init__(self, FB, edge_table, vertex_table, num_samples=1000, connect=False):
+        self.FB = FB
         self.num_samples = num_samples
         self.distances = np.linspace(0,1, num_samples)
-        assert edge_table.keys() == self.FB.F.keys()
-        assert is_continuous(vertex_table)
-        for  (column, functions) in edge_table.items():
-            fiber_set = self.FB.F[column]['range']
-            for f in functions:
-                #check using random point along [0,1]
-                val = f(np.random.random())
-                if self.FB.F[column]['monoid'] == 'nominal':
-                    assert val in fiber_set
-                else:
-                    assert all([min(fiber_set)<=v<=max(fiber_set) 
-                                    for v in np.atleast_1d(val)])
-
-        self.ids = range(len(vertex_table))
-        self.vertices = vertex_table        
-        self.edges = edge_table
-
-
-    def tau(self, k, simplex='edge'):
-        """this function knows that there are functions on the fibers defined in F"""
-        x = self.edges['x'][k](self.distances)
-        y = self.edges['y'][k](self.distances)
-        color = self.edges['color'][k](self.distances) 
[…]

### 2021-02-04 draft/sections/implementation.tex
implementation close out + revision of bar

@@
 
-\subsection{Case Study: Iris}
+\subsection{Case Study: Penguins}
+For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020} since it is multivariate and has a varying number of penguins.
+We use a version of the data packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021, jeff_reback_2020_3715232} since that is a very commonly used Python labled data structure.
+The wrapper is very thin since here there is explicitly only one section.
+\begin{minted}{python}
+class DataFrameSection:
+    def __init__(self, dataframe):
+        self._tau = dataframe.iloc
+        self._view = dataframe
+    def view(self):
+        return self._view
+\end{minted}
+The pandas indexer is a key valued set of discrete vertices, so there is no need to repackage for triangulation.
+As with the previous examples, there is no need to implement an explicit \mintinline{python}{get} method since the \mintinline{python}{dataframe} object has a get method.
 
+\begin{figure}[H]
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
+    \end{subfigure}
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
+    \end{subfigure}
+    \caption{Frequency of Penguin types visualized as discrete bars. }
+    \label{fig:code_bar_simple}
+\end{figure}
+The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist.
+The have the same required \vfiber\ components of (position, length).
+In \mintinline{pyton}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which only applies holistically to the graphic and never to individual data parameters.
+Explicitly differentiate between parameters in \vtotal\ and ones that are only in \vmarkd\ is another way this model allows for cleaner separation of roles in the code.
+
+\begin{minted}{python}
+class Bar(mcollections.Collection):
+    def __init__(self, data, transforms, *args, **kwargs):
+        # parameter of the graphic
+        self.orientation = kwargs.pop('orientation', 'v')
+
+        super().__init__(*args, **kwargs)
+        self.data = data
+        self.transforms = transforms
+
+    @staticmethod    
+    def _make_bars(orientation, position, width, floor, length):
+        if orientation in {'vertical', 'v'}:
+            xval, xoff, yval, yoff = position, width, floor, length
+        elif orientation in {'horizontal', 'h'}:
+            xval, xoff, yval, yoff = floor, length, position, width
+        return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
+                for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
+
+    def assemble(self, visual):
+        #set some defaults
+        visual['width'] = visual.get('width', itertools.repeat(0.8))
+        visual['floor'] = visual.get('floor', itertools.repeat(0))
+        visual['facecolors'] = visual.get('facecolors', 'C0')
+        #build bar glyphs based on graphic parameter
+        verts = self._make_bars(self.orientation, visual['position'], 
+                   visual['width'], visual['floor'], visual['length'])
+        self._paths = [mpath.Path(xy, closed=True) for xy in verts]
+        self.set_edgecolors('k')
+        self.set_facecolors(visual['facecolors'])
+
+    def draw(self, renderer, *args, **kwargs):
+        view = self.data.view()
+        visual = utils.convert_transforms(view, self.transforms)  
+        self.assemble(visual)
+        super().draw(renderer, *args, **kwargs)
+        return      
+\end{minted}
+The \mintinline{python}{draw} method identical to the ones above, but here the visual transformations are factored out into a separate function.
+The \mintinline{python}{assemble} function sets some defaults, constructs bars, and sets their edge color to black.
+The \mintinline{python}{_make_bars} function is somewhat factored out because this is an operation that may be used by other bar making functions that may not be able to make use of bars assemble or draw.
+
+\begin{figure}[H]
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/bar_stacked.png}
+    \end{subfigure}
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/bar_grouped.png}
+    \end{subfigure}
+    \caption{Penguin count disaggregated by island and species}
+    \label{fig:code_bar_multi}
+\end{figure}
+
+For example, the \mintinline{python}{MultiBar} artist that makes figure~\ref{fig:code_bar_multi} reuses \mintinline{python}{_make_bars} but does not reuse the assemble function because the composition of elements forces fundemental differences in glyph construction.
+As demonstrated in the \mintinline{python}{init}, the composite bar chart has orientation and whether it is stacked or not.
+While the stacked bar chart and the grouped bar chart could be seperate artists, as demonstrated they share so much overlapping code that it is far less redundant to implement them together.
+\note{looking at the mess that is this code, I'm a) not convinced these should be combined b) no longer convinced this provides anything over just bar if it isn't rewritten to use bar more}
+
+\begin{minted}{python}
+class MultiBar(mcollections.Collection):
+    def __init__(self, data, transforms, *args, **kwargs):
+        #set the orientation of the graphic  
+        self.orientation = kwargs.pop('orientation', 'v')
+        # set how the bar glyphs are put together to create the graphic
+        self.stacked = kwargs.pop('stacked', False)     
+        # rest is same as other artist __init__s
+
+        #this needs to be factored out but just want to finish now
+        self.width = kwargs.pop('width', .8)
+
+    def assemble(self, visual, view):
+        (groups, gencoder) = self.transforms['length']
+        ngroups = len(np.atleast_1d(groups))
+        visual['floor'] = visual.get('floor', np.empty(len(view[groups[0]])))
+        visual['facecolors'] = visual.get('facecolors', 'C0')
+        # make equal width stacked columns 
+        if 'width' not in visual and self.stacked:
+            visual['width'] = itertools.repeat(self.width)
+
+        # make equal with groups
+        if not self.stacked:
+            visual['width'] = itertools.repeat(self.width/ngroups)
+            offset = (np.arange(ngroups) /ngroups) * self.width
+        else:
+            offset = itertools.repeat(0)
+
+        # make the bars and arrange them 
+        verts = []
+        for group, off in zip(groups, offset):
+            verts.extend(Bar._make_bars(self.orientation, visual['position'] + off, 
+                          visual['width'], visual['floor'], view[group]))
+            if self.stacked: #add stacked bar to previous bar
+                visual['floor'] += view[group]
+
+        # convert lengths after all calculations are made and reorient if needed
+        # here or in transform machinery?
+        if self.orientation in {'v', 'vertical'}:
+            tverts = [[(x, gencoder(y)) for (x, y) in vert] 
+                            for vert in verts]
+        elif self.orientation in {'h', 'horizontal'}:
+            tverts = [[(gencoder(x), y) for (x, y) in vert] 
+                            for vert in verts]
+        self._paths = [mpath.Path(xy, closed=True) for xy in tverts]
+        #flatted columns of colors to match list of bars
+        self.set_facecolor(list(itertools.chain.from_iterable(visual['facecolors'])))
+        self.set_edgecolors('k')
+
+    def draw(self, renderer, *args, **kwargs):
+        view = self.data.view()
+        #exclude converting the group visual length, special cased in assemble
+        visual = utils.convert_transforms(view, self.transforms, exclude=['length'])
+        # pass in view because nu is not distributable so may need to apply it 
+        # after visual assembly
+        self.assemble(visual, view)
+        super().draw(renderer, *args, **kwargs)
+        return      
+\end{minted}
+
+In the \mintinline{python}{__draw__}, a utility function is used for conversions, but the length transforms are held until after assembly because the length is computed by adding the current length to the previous and many transforms are not distributable such that $\nu(x_0 + x_1 + x_2) = \nu(x_0) + \nu(x_1) + \nu(x_2)$.
+Inside \mintinline{python}{assemble}, the glyphs are either shifted vertically (\texttt{stacked}) or horizontally (\texttt{grouped}) such that the positions are recorded and added to with the next group.
+This function allows multiple columns to be mapped to a visual parameter, but it must be equal numbers of columns
+\begin{minted}{python}
+{'position': ('island', lambda x: {'Biscoe':0, 'Dream':1, 'Torgersen':2}[x]),
+ 'length':(['Adelie', 'Chinstrap', 'Gentoo'], lambda x: x),
+ 'facecolors': (['Adelie_s', 'Chinstrap_s', 'Gentoo_s'], 
+        color.Categorical({'Adelie':'#FF8C00',
+                           'Gentoo':'#159090', 
+                           'Chinstrap':'#A034F0'}))}
+\end{minted}
+such as in this example where for each column contributing to a segment of the bar there is a corresponding column of colors for this segment.
+The reason the multibar can work with such a trasnformer is because it is relying on the data model to do most of the bookkeeping of which values get mapped to which bars.
+This also yields a much simpler function call to the artist
+
+\begin{minted}{python}
+fig, ax = plt.subplots()
+artist = bar.MultiBar(table, trans, orientation='h', stacked=True)
+ax.add_artist(artist)
+\end{minted}
+
+where \mintinline{python}{trans} is the same dictionary for both stacked and grouped version, as is the \mintinline{python}{DataFrameSection} object \mintinline{python}{table}.
+The only difference between the two versions is the \texttt{stacked} flag, and the only difference between figures~\ref{fig:code_bar_simple} is the \texttt{orientation} argument.
+By decomposing the architecture into data, visual encoding, and assembly steps, we are able to build components that are more flexible and also more self contained than the existing code base.
+
+\note{This API may want to be redesigned such that there's a way to clearly couple the columns when doing multindex broadcasting}
 \end{document}

### 2021-02-04 code/matplottoy/artists/bar.py

@@ -11,17 +11,8 @@ import matplotlib.path as mpath
 
 from matplottoy.artists import utils
 
-def _make_bars(orientation, position, width, floor, length):
-    if orientation in {'vertical', 'v'}:
-        xval, xoff, yval, yoff = position, width, floor, length
-    elif orientation in {'horizontal', 'h'}:
-        xval, xoff, yval, yoff = floor, length, position, width
-    return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
-            for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
 
 class Bar(mcollections.Collection):
-    required = {'position', 'length'}
-    optional = {'width', 'floor', 'color'}
     def __init__(self, data, transforms, *args, **kwargs):
         """
         Parameters
@@ -38,9 +29,9 @@ class Bar(mcollections.Collection):
             kwargs passed through 
         """
   
-        assert 'vertex' in data.FB.K['tables']
-        utils.check_constraints(Bar, transforms)
-        utils.validate_transforms(data, transforms)
+        #assert 'vertex' in data.FB.K['tables']
+        #utils.check_constraints(Bar, transforms)
+        #utils.validate_transforms(data, transforms)
 
         self.orientation = kwargs.pop('orientation', 'v')
      
@@ -48,30 +39,36 @@ class Bar(mcollections.Collection):
         self.data = data
         self.transforms = transforms
 
-        
+    @staticmethod    
+    def _make_bars(orientation, position, width, floor, length):
+        if orientation in {'vertical', 'v'}:
+            xval, xoff, yval, yoff = position, width, floor, length
+        elif orientation in {'horizontal', 'h'}:
+            xval, xoff, yval, yoff = floor, length, position, width
+        return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
+                for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
+
     
-    def draw(self, renderer, *args, **kwargs):
-       
-        view = self.data.view('vertex')
-        visual = utils.convert_transforms(view, self.transforms)  
-        
-        if 'width' not in visual:
-            visual['width'] = itertools.repeat(0.8)
-        
-        if 'floor' not in visual:
-            visual['floor'] = itertools.repeat(0)
+    def assemble(self, visual): 
+        visual['width'] = visual.get('width', itertools.repeat(0.8))
[…]

### 2021-02-04 code/matplottoy/artists/utils.py

@@ -40,8 +40,8 @@ def convert_transforms(view, transforms, exclude=None):
         else:
             (var, tau) = value
             if isinstance(var, (str, bytes )):
-                visual[param] = tau.convert(view[var])
+                visual[param] = tau(view[var])
             else:
-                visual[param] = [tau.convert(view[v]) for v in var]
+                visual[param] = [tau(view[v]) for v in var]
 
     return visual

### 2021-02-04 code/matplottoy/data/iris.py

@@ -1,6 +1,15 @@
 from matplottoy.data.simplex import FiberBundle
 from matplottoy.encoders import mtypes
 
+class DataFrame:
+    def __init__(self, dataframe):
+        self._tau = dataframe.iloc
+        self._view = dataframe
+
+    def view(self, simplex="vertex"):
+        return self._view
+    
+    
 class Iris:
     FB = FiberBundle({'tables': ['vertex']},
                      {'species': mtypes.Nominal({'setosa', 'versicolor', 'virginica'}),

### 2021-02-08 draft/sections/intro.tex
artist _ fiber _nu
switched data notation to standin + cleaned up language
finished out render section
most relevant bits of how to write a paper
start of reworked implementation
intro, circles many circles

@@
 \begin{figure}
-    \includegraphics[width=.5\textwidth]{figures/intro/viz_same.png}
-    \caption[]{Implicit in visualization is the assumption that these three representations of data are equivalent, specifically that the measurements within a variable and relations of the measurements of each variable are preserved.
-}
-    \label{fig:viz_same}
-    %% change up image to dubois or bertin or
 
@@
 
-\subsection{Thesis statement}
-We define a visualization as a transform from data to graphic that preserves the topology of the data and faithfully map the properties of the measurement type.
-In fig~\ref{fig:viz_same}, we implicitly assume that the translation from table to heatmap has preserved the order of observations (the rows) and that the perceptually uniform sequential colormap has been applied such that the ordering relation on floats matches the ordering on the colormap (darker colors map to larger numbers).
-We also make this assumption about color in the scatter map, and that the translation to size and position on screen also respect the ordering on floats.
-In this work, we propose to mathematically describe the transform of data to visual space such that we can make explicit the implicit topology and types visualizations preserve.
-We then propose a new architecture for the Python visualization library Matplotlib \cite{hunterMatplotlib2DGraphics2007} based on these descriptions because the Matplolib artist layer is analogous to the transforms.
+\subsection{what is the problem we're trying to solve}
+% build better viz libaries
+%formally define what constraints on libraries are
+% make something seperable and scalable
+
+\begin{figure}
+    \begin{subfigure}{.20\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_spinny.png}
+        \caption{}
+        \label{fig:intro_dpa}
+    \end{subfigure}
+    \begin{subfigure}{.20\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_bar.png}
+        \caption{}
+        \label{fig:intro_dpb}
+    \end{subfigure}
+    \begin{subfigure}{.20\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_country.png}
+        \caption{}
+        \label{fig:intro_dbc}
+    \end{subfigure}
+    \begin{subfigure}{.20\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_heat.png}
+        \caption{}
+        \label{fig:intro_dbd}
+    \end{subfigure}
+    \caption{Du Bois' data portraits\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018} of post reconstruction Black American life exemplify that the fundemental characteristics of data visualization is that the visual elements vary in proportion to the source data.
+In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the bar charts are intersected to show the number of property owners and how much they own in a given year; in figure~\ref{fig:intro_dpc} the countries are scaled to population size; and figure~\ref{fig:intro_dpd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
+The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900, duboisGeorgiaNegroCity1900,duboisSeriesStatisticalCharts,duboisSeriesStatisticalChartsa}}
+    \label{fig:intro_dubois}
+\end{figure}
+
+This work is motivated by a rearchitecture of the Matplotlib visualization library that aims to disentangle the transformation from data to visualization.
+
+\subsection{What even is a data visualization library?}
+
+For example, the Du Bois data portraits in figure~\ref{fig:intro_dubois} are not scatter plots or line plots or really maps, but are clearly identifiable as data visualizations because there are some properties of the original data they preserve.
+
+Mackinlay, expressiveness...Tufte lie principal\cite{mackinlayAutomatingDesignGraphical1986}
+
+"The boundary between scientific visualization and infovis has been explored by Tory and Moller [37] who note ways that discrete and continuous data and the attribute types form a design space that contains both infovis and scientific visualization." \cite{pousmanCasualInformation2007}
 
@@
 \subsubsection{Retinal (Visual) Variables}
+\label{sec:intro_visual_variables}
 \begin{figure}[h!]
@@
 %% mention tidy/spivak first, set up that we use variable, record, key/index notation
+%% define continuity in the math, connectivity in the data
+%% topological spaces are frame in terms of continuity
+% the implementation of $K$ embeds the connectivity
 \label{sec:intro_data_structure}
@@
 
+\subsection{how does this tackle the problem? what does it leave on the table?}
+% build on butler, gog
+% differ from ggplot (d3, vega), vtk, imageJ
+%% lets get on the same page about words
+% data is topology + variables
+% visual variables are sparkly
+
+\subsection{contribution of this work is}
+ % topological model (abstractions for continuity, formal definition of structure preservation as monoidal action equivarance)
+ % prototype
 \end{document}

### 2021-02-08 draft/sections/intro.tex
prototype

@@
 
-\subsection{what is the problem we're trying to solve}
-% build better viz libaries
-%formally define what constraints on libraries are
-% make something seperable and scalable
-
 \begin{figure}
-    \begin{subfigure}{.20\textwidth}
+    \begin{subfigure}{.24\textwidth}
         \includegraphics[width=1\textwidth]{figures/intro/du_bois_spinny.png}
@@
     \end{subfigure}
-    \begin{subfigure}{.20\textwidth}
+    \begin{subfigure}{.24\textwidth}
         \includegraphics[width=1\textwidth]{figures/intro/du_bois_bar.png}
@@
     \end{subfigure}
-    \begin{subfigure}{.20\textwidth}
+    \begin{subfigure}{.24\textwidth}
         \includegraphics[width=1\textwidth]{figures/intro/du_bois_country.png}
@@
     \end{subfigure}
-    \begin{subfigure}{.20\textwidth}
+    \begin{subfigure}{.24\textwidth}
         \includegraphics[width=1\textwidth]{figures/intro/du_bois_heat.png}
@@
 
-This work is motivated by a rearchitecture of the Matplotlib visualization library that aims to disentangle the transformation from data to visualization.
+\subsection{what is the problem we're trying to solve}
+% build better viz libaries
+%formally define what constraints on libraries are
+% make something seperable and scalable
+
+\subsection{what are we doing}
+constraints: toplogy of dataset + structure on the variables
+implementation: be fully feature
+we can haz compose the things into larger structure. \& has framwework to do larger things w/ algebraic formalizations.
 
@@
 
+Intuitively, data visualizations are drawings where elements of the figure are mapped 
 For example, the Du Bois data portraits in figure~\ref{fig:intro_dubois} are not scatter plots or line plots or really maps, but are clearly identifiable as data visualizations because there are some properties of the original data they preserve.
@@
 
-\subsection{what are we doing}
-constraints: toplogy of dataset + structure on the variables
-implementation: be fully feature
-we can haz compose the things into larger structure. \& has framwework to do larger things w/ algebraic formalizations.
-
-\subsection{ What are the the basic requirements of a visualization library?}
 Acquired codes of meaning, dubois
@@
 
-\subsection{contribution of this work is}
+\subsection{contribution}
  % topological model (abstractions for continuity, formal definition of structure preservation as monoidal action equivarance)
  % prototype
+
+\begin{enumerate}
+    \item fiber bundle representation of data and graphics
+    \item formalization of variable type structure preservation as monoid action equivariant maps between fibers 
+    \item formalization of continuity preservation as equivariant map between base space
+    \item algebraic sum for composition of graphics
+    \item functional approach to force seperation of scope and concerns
+    \item prototype  
+\end{enumerate}
+
 \end{document}

### 2021-02-08 draft/sections/intro.tex
axquired codes of meeting:

@@
 In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the bar charts are intersected to show the number of property owners and how much they own in a given year; in figure~\ref{fig:intro_dpc} the countries are scaled to population size; and figure~\ref{fig:intro_dpd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
-The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900, duboisGeorgiaNegroCity1900,duboisSeriesStatisticalCharts,duboisSeriesStatisticalChartsa}}
+The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroNegro1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
     \label{fig:intro_dubois}
@@
 \subsection{What even is a data visualization library?}
+Data visualizations originated as maps of places and stars \cite{friendlyBriefHistoryData2008}; direct graphical representations of physical items in space.
+While often decorative, to be useful these visualizations at least somewhat had to preserve some notion of relative distance between the objects being graphed.
+Byrne et al.
+deconstruct  visualizations as having either graphic or figurative components; graphic representations are maps from data values to visual encodings, while figurative representations have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+For example, the Du Bois data portraits in figure~\ref{fig:intro_dubois} are mostly graphic representations where quantitative variables are mapped to glyph segment length or area.
+In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from.
 
-Intuitively, data visualizations are drawings where elements of the figure are mapped 
-For example, the Du Bois data portraits in figure~\ref{fig:intro_dubois} are not scatter plots or line plots or really maps, but are clearly identifiable as data visualizations because there are some properties of the original data they preserve.
+Mackinlay expressive/exp
 
-Mackinlay, expressiveness...Tufte lie principal\cite{mackinlayAutomatingDesignGraphical1986}
+\section{Not all data are tables}
 
 "The boundary between scientific visualization and infovis has been explored by Tory and Moller [37] who note ways that discrete and continuous data and the attribute types form a design space that contains both infovis and scientific visualization." \cite{pousmanCasualInformation2007}
-
-Acquired codes of meaning, dubois
-
-\section{Not all data are tables}
 
@@
 \subsection{how does this tackle the problem? what does it leave on the table?}
-% build on butler, gog
-% differ from ggplot (d3, vega), vtk, imageJ
-%% lets get on the same page about words
-% data is topology + variables
-% visual variables are sparkly
+\begin{itemize}
+    \item butler's topological formulation
+    \item GoG's mathematics of visualizations
+    \item vtk's qausi topological
+\end{itemize}
 
 \subsection{contribution}
- % topological model (abstractions for continuity, formal definition of structure preservation as monoidal action equivarance)
- % prototype
 

### 2021-02-09 draft/sections/intro.tex
rearranging, words

@@
 \subsection{What even is a data visualization library?}
+ Byrne et al.
+describe visualizations as having graphic and figurative components; graphic representations are maps from data values to visual encodings, while figurative representations have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+For example, the Du Bois data portraits in figure~\ref{fig:intro_dubois} are mostly graphic representations where quantitative variables are mapped to glyph segment length or area.
+Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from in figure~\ref{fig:intro_dbc}.
+
  Data visualizations originated as maps of places and stars \cite{friendlyBriefHistoryData2008}; direct graphical representations of physical items in space.
-While often decorative, to be useful these visualizations at least somewhat had to preserve some notion of relative distance between the objects being graphed.
-Byrne et al.
-deconstruct  visualizations as having either graphic or figurative components; graphic representations are maps from data values to visual encodings, while figurative representations have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
-For example, the Du Bois data portraits in figure~\ref{fig:intro_dubois} are mostly graphic representations where quantitative variables are mapped to glyph segment length or area.
-In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from.
+While often decorative, to be useful these visualizations at least somewhat had to preserve some notion of relative distance between the objects being graphed, which Tufte termed the Lie principal\cite{tufteVisualDisplayQuantitative2001}.
 
@@
 
-\section{Not all data are tables}
-
-"The boundary between scientific visualization and infovis has been explored by Tory and Moller [37] who note ways that discrete and continuous data and the attribute types form a design space that contains both infovis and scientific visualization." \cite{pousmanCasualInformation2007}
-
-Tables, images (Lev), graphs (network X)
-
-set up: dubois
-
-theorists:
-bertin, munzner, mackinlay
-
-talk about:
-matplotlib arch paper, excel/matlab arch, 
-vtk \& ggplot (compare/contrast, we're blending these things)
-
-\subsection{Terminology}
-\textcolor{blue}{Is currently C\&P'ed from my second exam so needs some messaging, but }
 \begin{figure}[h!]
@@
 
+\section{Not all data are tables}
+
+"The boundary between scientific visualization and infovis has been explored by Tory and Moller [37] who note ways that discrete and continuous data and the attribute types form a design space that contains both infovis and scientific visualization." \cite{pousmanCasualInformation2007}
+
+Tables, images (Lev), graphs (network X)
+
+set up: dubois
 \subsubsection{Data Type and Structure}
@@
 
+\subsection{How do other visualization libraries tackle this?}
+matplotlib arch paper, excel/matlab arch, 
+vtk \& ggplot (compare/contrast, we're blending these things)
+
 \subsection{contribution}

### 2021-02-09 draft/sections/intro.tex
graphics: data -> visual

@@
 Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from in figure~\ref{fig:intro_dbc}.
+So given graphics by definition have an underlying data
 
@@
 
-Mackinlay expressive/exp
+Mackinlay defines two criteria for
 
@@
 \begin{itemize}
+
+    \item GoG's mathematics of visualizations
+    This book focuses instead on rules for constructing graphs mathematically
+and then representing them as graphics aesthetically.
+The title of this book also recalls Bertin’s Semiology of Graphics (1967),
+the first and most influential structural theory of statistical graphics. Bertin’s
+work has pervaded our thinking. Semiology deals with signs. Although Bertin
+put his signs on paper, his work applies as well to virtual space.
+    \item vtk's qausi topological
     \item butler's topological formulation
-    \item GoG's mathematics of visualizations
-    \item vtk's qausi topological
 \end{itemize}

### 2021-02-10 draft/sections/intro.tex
criteria

@@
 \subsection{What even is a data visualization library?}
+This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:}; to do so, we first discuss the criteria by which a visualization is evaluated.
+
 Byrne et al.
-describe visualizations as having graphic and figurative components; graphic representations are maps from data values to visual encodings, while figurative representations have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
-For example, the Du Bois data portraits in figure~\ref{fig:intro_dubois} are mostly graphic representations where quantitative variables are mapped to glyph segment length or area.
-Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from in figure~\ref{fig:intro_dbc}.
-So given graphics by definition have an underlying data
+describe visualizations as having graphic and figurative components; graphic representations are maps from data values to visual encodings\cite{bertinSemiologyGraphicsDiagrams2011a}, while figurative representations have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic nominal and numerical properties of the data are preserved in the graph.
+Norman states that when the properties of the representation match the properties of the information being represented\cite{norman_things_smart, NaturalnessPrincipleInfoVis} then the visualizations are more understandable
+, but does not give criteria for what it means for the properties to match.
+Tufte specifies that visual representations must be in proportion to the quantitative data being represented, but otherwise his notion of graphic integrity is heavily context dependent\cite{tufteVisualDisplayQuantitative2001}.
 
- Data visualizations originated as maps of places and stars \cite{friendlyBriefHistoryData2008}; direct graphical representations of physical items in space.
-While often decorative, to be useful these visualizations at least somewhat had to preserve some notion of relative distance between the objects being graphed, which Tufte termed the Lie principal\cite{tufteVisualDisplayQuantitative2001}.
-
-Mackinlay defines two criteria for
+Mackinlay
 

### 2021-02-10 draft/sections/intro.tex
contribution + intro section

@@
 
-\subsection{what is the problem we're trying to solve}
-% build better viz libaries
-%formally define what constraints on libraries are
-% make something seperable and scalable
+In data visualization it is assumed the graphical representation of data match the properties of the data, and in this work we propose that the mathematical notion of equivariance formalizes this expectation.
+We meld the infoviz communities interest in heterogenous often discrete datasets with the scientific visualization communities emphasis on continuous and sometimes topologically complex datasets.
+To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
+In addition to providing a way to ensure the library preserves structure, we propose a functional approach to  improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
 
-\subsection{what are we doing}
-constraints: toplogy of dataset + structure on the variables
-implementation: be fully feature
-we can haz compose the things into larger structure. \& has framwework to do larger things w/ algebraic formalizations.
-
-\subsection{What even is a data visualization library?}
-This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:}; to do so, we first discuss the criteria by which a visualization is evaluated.
-
-Byrne et al.
-describe visualizations as having graphic and figurative components; graphic representations are maps from data values to visual encodings\cite{bertinSemiologyGraphicsDiagrams2011a}, while figurative representations have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+\subsection{Background}
+This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:intro_dubois}; to do so, we first discuss the criteria by which a visualization is evaluated.
+As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design.
+He defines the graphical design as the set encoding relations from data to visual representation, and the design appied on an idealized space is what throughout this paper we will refer to as graphics \cite{mackinlayAutomatingDesignGraphical1986}.
+Mackinlay  proposes that a visualization tool can be evaluated by its expressiveness and effectiveness; A visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a}.
+Mackinlay's definition of expressiveness is formalized at the variable level, for example that ordinal variables must map to ordinal positions.
+He uses a scatter chart for discrete data states that a line chart must have continuous data, but does not describe how to prove this.
+While Mackinlay is strictly focusing on the graphical components of a visualization, Byrne et al.
+propose that visualizations also have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
 In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic nominal and numerical properties of the data are preserved in the graph.
-Norman states that when the properties of the representation match the properties of the information being represented\cite{norman_things_smart, NaturalnessPrincipleInfoVis} then the visualizations are more understandable
-, but does not give criteria for what it means for the properties to match.
-Tufte specifies that visual representations must be in proportion to the quantitative data being represented, but otherwise his notion of graphic integrity is heavily context dependent\cite{tufteVisualDisplayQuantitative2001}.
-
-Mackinlay
+Tufte specifies that visual representations must be in proportion to the quantitative data being represented for a chart to be faithful and that there should be no extra information in the graphic or figurative elements of the graph, but otherwise his notion of graphic integrity is heavily context dependent\cite{tufteVisualDisplayQuantitative2001}.
+As is Norman's Naturalness Principal, which states that visualizations are more understandable  when the properties of the representation match the properties of the information being represented\cite{norman_things_smart}.
+Bertin takes it as a given that data properties match visual properties, so much so that Munzner argues it is inherently built into his classification system \cite{munznerVisualizationAnalysisDesign2014}.
+In this work, we propose that this property matching can be formally defined as equivariant maps from data to graphic representation.
 
@@
 \subsection{contribution}
-
-\begin{enumerate}
-    \item fiber bundle representation of data and graphics
-    \item formalization of variable type structure preservation as monoid action equivariant maps between fibers 
-    \item formalization of continuity preservation as equivariant map between base space
-    \item algebraic sum for composition of graphics
-    \item functional approach to force seperation of scope and concerns
-    \item prototype  
-\end{enumerate}
-
+The contribition of this work is a model we call the topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles.
+This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
+Furthermore, this model supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
+To demonstrate the practical value of the model, we built a prototype where we represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
 \end{document}

### 2021-02-10 draft/sections/intro.tex
start of vv section

@@
 \label{sec:introduction}
-\begin{figure}
+In data visualization it is assumed the graphical representation of data match the properties of the data, and in this work we propose that the mathematical notion of equivariance formalizes this expectation.
+We meld the infoviz communities interest in heterogenous often discrete datasets with the scientific visualization communities emphasis on continuous and sometimes topologically complex datasets.
+To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
+In addition to providing a way to ensure the library preserves structure, we propose a functional approach to  improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
 
-\end{figure}
-
+\section{Background}
 \begin{figure}
@@
 
-In data visualization it is assumed the graphical representation of data match the properties of the data, and in this work we propose that the mathematical notion of equivariance formalizes this expectation.
-We meld the infoviz communities interest in heterogenous often discrete datasets with the scientific visualization communities emphasis on continuous and sometimes topologically complex datasets.
-To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
-In addition to providing a way to ensure the library preserves structure, we propose a functional approach to  improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
-
-\subsection{Background}
 This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:intro_dubois}; to do so, we first discuss the criteria by which a visualization is evaluated.
 As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design.
-He defines the graphical design as the set encoding relations from data to visual representation, and the design appied on an idealized space is what throughout this paper we will refer to as graphics \cite{mackinlayAutomatingDesignGraphical1986}.
+He defines the graphical design as the set encoding relations from data to visual representation, and the design evaluated on an idealized space is what throughout this paper we will refer to as graphics \cite{mackinlayAutomatingDesignGraphical1986}.
 Mackinlay  proposes that a visualization tool can be evaluated by its expressiveness and effectiveness; A visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a}.
@@
 In this work, we propose that this property matching can be formally defined as equivariant maps from data to graphic representation.
-
-\begin{figure}[h!]
-%%remake properly at some point
-\includegraphics[width=\textwidth]{figures/intro/flowchart.png} 
-\caption{To go from a dataset to a visualization, the data is subset based on a set of constraints (the invariant).
-The resulting subset becomes the components that are visualized, but the choice of visualization is dependent on the type and structure of the component variables.}
-\label{fig:flowchart}
-\end{figure}
-
-Given a dataset, we need to decide what subset of the data to visualize.
-Bertin describes the set of constraints used to subset the data as the \textit{invariant}.
-Formally, the \textit{invariant} is the set of shared characteristics of the data being visualized.
-When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertinSemiologyGraphicsDiagrams2011a}.
-%%rework this statement for french labor data
-%%In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables.
-As shown in figure~\ref{fig:flowchart}, the final step in creating a visualization is choosing how to encode the
-components using retinal (visual) variables.
 
@@
 \subsection{How do other visualization libraries tackle this?}
+begin{figure}[h!]
+%%remake properly at some point
+\includegraphics[width=\textwidth]{figures/intro/flowchart.png} 
+\caption{To go from a dataset to a visualization, the data is subset based on a set of constraints (the invariant).
+The resulting subset becomes the components that are visualized, but the choice of visualization is dependent on the type and structure of the component variables.}
+\label{fig:flowchart}
+\end{figure}
+
+Given a dataset, we need to decide what subset of the data to visualize.
+Bertin describes the set of constraints used to subset the data as the \textit{invariant}.
+Formally, the \textit{invariant} is the set of shared characteristics of the data being visualized.
+When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertinSemiologyGraphicsDiagrams2011a}.
+%%rework this statement for french labor data
+%%In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables.
+As shown in figure~\ref{fig:flowchart}, the final step in creating a visualization is choosing how to encode the
+components using retinal (visual) variables.
 matplotlib arch paper, excel/matlab arch, 

### 2021-02-11 draft/sections/intro.tex
data and gog

@@
 \section{Background}
-\begin{figure}
+\begin{figure}[H]
     \begin{subfigure}{.24\textwidth}
@@
 Mackinlay  proposes that a visualization tool can be evaluated by its expressiveness and effectiveness; A visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a}.
-Mackinlay's definition of expressiveness is formalized at the variable level, for example that ordinal variables must map to ordinal positions.
-He uses a scatter chart for discrete data states that a line chart must have continuous data, but does not describe how to prove this.
+Mackinlay's definition of expressiveness is formalized at the component level, for example that ordinal values must map to ordinal positions.
+He uses a scatter chart for discrete data and states that a line chart must have continuous data, but does not describe how to prove this.
 While Mackinlay is strictly focusing on the graphical components of a visualization, Byrne et al.
@@
 \end{figure}
-
-Figure~\ref{fig:retinal_variables} illustrates common guidelines for encoding \textit{components}, derived from what Bertin terms a retinal variable and most other visualization theorists call visual variables \cite{bertin_semiology_2011,krygier_making_2005,chambers_graphical_1983,wilkinson_grammar_2005,munzner_visualization_2014}.
+Figure~\ref{fig:retinal_variables} illustrates common guidelines for encoding \textit{components} of the data as components of the graphics.
+Bertin names these graphic components retinal variables; they are also called visual variables \cite{bertin_semiology_2011,krygier_making_2005,chambers_graphical_1983,wilkinson_grammar_2005,munzner_visualization_2014} or visual characteristics\cite{carpendaleVisualRepresentationSemiology}.
 The columns of figure~\ref{fig:retinal_variables} correspond to the type of observation: discrete points, continuous events (e.g.
@@
 
+Tory and Moller
+
 "The boundary between scientific visualization and infovis has been explored by Tory and Moller [37] who note ways that discrete and continuous data and the attribute types form a design space that contains both infovis and scientific visualization." \cite{pousmanCasualInformation2007}
-
-Tables, images (Lev), graphs (network X)
 
@@
 \begin{description}
-    \item[variable]
+    \item[component]
     \item[index]  
@@
 \subsection{how does this tackle the problem? what does it leave on the table?}
+Most visualization tools are designed around some notion of the structure of the data being visualized. For example Grammer of Graphics asusmes ythat
+
+Tables, images (Lev), graphs (network X)
+
 \begin{itemize}

### 2021-02-11 draft/sections/intro.tex
Mackinlay + rework of visual channels section

@@
 This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:intro_dubois}; to do so, we first discuss the criteria by which a visualization is evaluated.
-As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design.
-He defines the graphical design as the set encoding relations from data to visual representation, and the design evaluated on an idealized space is what throughout this paper we will refer to as graphics \cite{mackinlayAutomatingDesignGraphical1986}.
-Mackinlay  proposes that a visualization tool can be evaluated by its expressiveness and effectiveness; A visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a}.
-Mackinlay's definition of expressiveness is formalized at the component level, for example that ordinal values must map to ordinal positions.
-He uses a scatter chart for discrete data and states that a line chart must have continuous data, but does not describe how to prove this.
-While Mackinlay is strictly focusing on the graphical components of a visualization, Byrne et al.
-propose that visualizations also have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
-In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic nominal and numerical properties of the data are preserved in the graph.
+Byrne et al.
+propose that visualizations have graphic representations that are mappings from data to visuals and figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph.
 Tufte specifies that visual representations must be in proportion to the quantitative data being represented for a chart to be faithful and that there should be no extra information in the graphic or figurative elements of the graph, but otherwise his notion of graphic integrity is heavily context dependent\cite{tufteVisualDisplayQuantitative2001}.
 As is Norman's Naturalness Principal, which states that visualizations are more understandable  when the properties of the representation match the properties of the information being represented\cite{norman_things_smart}.
-Bertin takes it as a given that data properties match visual properties, so much so that Munzner argues it is inherently built into his classification system \cite{munznerVisualizationAnalysisDesign2014}.
-In this work, we propose that this property matching can be formally defined as equivariant maps from data to graphic representation.
+Bertin takes it as a given that data properties match visual properties, so much so that Munzner argues it is inherently built into his classification system \cite{munznerVisualizationAnalysisDesign2014} which is displayed in figure~\ref{fig:intro_retinal_variables}.
 
-\subsubsection{Retinal (Visual) Variables}
 \label{sec:intro_visual_variables}
-\begin{figure}[h!]
+\begin{figure}[H]
 \includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
-\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the \textit{components} of a visualization.
+\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the components of a visualization.
+The best to show column describes which types of information can be expressed in the corresponding visual encoding.
 This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS}
 \cite{krygierMakingMapsVisual2005}}
-\label{fig:retinal_variables}
+\label{fig:intro_retinal_variables}
 \end{figure}
-Figure~\ref{fig:retinal_variables} illustrates common guidelines for encoding \textit{components} of the data as components of the graphics.
-Bertin names these graphic components retinal variables; they are also called visual variables \cite{bertin_semiology_2011,krygier_making_2005,chambers_graphical_1983,wilkinson_grammar_2005,munzner_visualization_2014} or visual characteristics\cite{carpendaleVisualRepresentationSemiology}.
-The columns of figure~\ref{fig:retinal_variables} correspond to the type of observation: discrete points, continuous events (e.g.
-a timeseries), two dimensional continuous events (e.g.
-a vector map).
-The rows of figure~\ref{fig:retinal_variables} describe ways to visualize variations in the \textit{components}; generally, quantitative components are represented by retinal variables that change quantitatively and categorical components are represented by retinal variables that vary qualitatively.
-In figure~\ref{fig:iris_scatter}, the hue of the marker is used to encode differentiation in species and the position of the marker is used to show variation in petal length and sepal length.
-The retinal variables suggest that any single visualization is limited to encoding at most about 8 or 9 components.
-Retinal variables provide guidelines for encoding \textit{components}, but the choice of graph is based on the type and structure of the data.
+
+As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design.
+He defines the graphical design as the set encoding relations from data to visual representation\cite{mackinlayAutomatingDesignGraphical1986}, and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.
+Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tools effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
+Mackinlay's definition of expressiveness is formalized at the visual encoding level, which as shown in figure~\ref{fig:intro_retinal_variables} refers to the components of a graphic such as the color or position of a glyph.
+Bertin first classified these graphic components as retinal variables and discussed which types of data they can express \cite{bertinIIPropertiesGraphic2011} and how they are composited on point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond.
+Mackinlay's expressiveness criteria is well defined for the visual variables, such that he suggests the viability of a strict encoding relation that is a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
+We expand on this suggestion by proposing that monoid action equivariance is a strict condition of building valid encoders.
 

### 2021-02-11 draft/sections/intro.tex
connectivity

@@
     \caption{Du Bois' data portraits\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018} of post reconstruction Black American life exemplify that the fundemental characteristics of data visualization is that the visual elements vary in proportion to the source data.
-In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the bar charts are intersected to show the number of property owners and how much they own in a given year; in figure~\ref{fig:intro_dpc} the countries are scaled to population size; and figure~\ref{fig:intro_dpd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
+In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the bar charts are intersected to show the number of property owners and how much they own in a given year; in figure~\ref{fig:intro_dbc} the countries are scaled to population size; and figure~\ref{fig:intro_dbd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
 The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroNegro1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
@@
 We expand on this suggestion by proposing that monoid action equivariance is a strict condition of building valid encoders.
+Mackinlay does not provide a generalized criteria for plot types, instead embedding the requirements that scatter be discrete and line continuous within the definition of the charts.
 
-\section{Not all data are tables}
+\subsection{Not all data are tables}
 
-Tory and Moller
+Mackinlay was primarily concerned with visualizing relational data of the type found in databases.
+as do the tools for which Mackinlay's APT (A Presentation Tool) serves as the basis.
+These tools include grammar of graphics \cite{wilkinsonGrammarGraphics2005} and its derivatives ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+In contrast, tools such as ImageJ\cite{schneiderNIHImageImageJ2012} and the ImagePlot\cite{studiesCulturevisImageplot2021}macro were designed with images in mind.
+For more flexibility, vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as  MayaVi\cite{ramachandranMayaVi2011} support all sorts of data types.
+These tools are not domain specific, as they are used for scientific, information, and even artistic visualizations \cite{pousmanCasualInformation2007}; therefore Tory and Möller
+
+\begin{figure}
+    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
+    \caption{One way to describe data is by the connectivity of the points in the dataset.
+A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid.
+This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
+    \label{fig:intro_data_format}
+\end{figure}
+
+As shown in figure~\ref{fig:intro_data_format}, one method of generalizing data
+
+Tory and Moller propose that
 
@@
 
-set up: dubois
-\subsubsection{Data Type and Structure}
+\subsection{Data Type and Structure}
 %% mention tidy/spivak first, set up that we use variable, record, key/index notation
@@
 \subsection{How do other visualization libraries tackle this?}
-begin{figure}[h!]
+\begin{figure}[h!]
 %%remake properly at some point

### 2021-02-11 draft/sections/intro.tex
intro

--- 2021-02-11 draft/sections/intro.tex
@@
 \section{Introduction}
-\label{sec:introduction}
+\label{sec:intro}
 In data visualization it is assumed the graphical representation of data match the properties of the data, and in this work we propose that the mathematical notion of equivariance formalizes this expectation.
@@
 
-\label{sec:intro_visual_variables}
 \begin{figure}[H]
@@
 Bertin first classified these graphic components as retinal variables and discussed which types of data they can express \cite{bertinIIPropertiesGraphic2011} and how they are composited on point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond.
+Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information}.
 Mackinlay's expressiveness criteria is well defined for the visual variables, such that he suggests the viability of a strict encoding relation that is a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
 We expand on this suggestion by proposing that monoid action equivariance is a strict condition of building valid encoders.
+Mackinlay does not provide a generalized criteria for plot types, instead embedding the requirements within the definition of the charts.
 
-\section{Not all data are tables}
+\subsection{Data}
+\label{sec:intro_data}
+Tory and Möller propose that assumptions about the structure of data are built into the visual algorithms that display that information \cite{toryRethinkingVisualizationHighLevel2004}.
+Specifically they note that discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}.
 
-Tory and Moller
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
+    \caption{One way to describe data is by the connectivity of the points in the dataset.
+A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid.
+This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
+    \label{fig:intro_data_format}
+\end{figure}
 
-"The boundary between scientific visualization and infovis has been explored by Tory and Moller [37] who note ways that discrete and continuous data and the attribute types form a design space that contains both infovis and scientific visualization." \cite{pousmanCasualInformation2007}
-
-set up: dubois
-\subsubsection{Data Type and Structure}
-%% mention tidy/spivak first, set up that we use variable, record, key/index notation
-%% define continuity in the math, connectivity in the data
-%% topological spaces are frame in terms of continuity
-% the implementation of $K$ embeds the connectivity
-\label{sec:intro_data_structure}
-\begin{figure}[h!]
+As shown in figure~\ref{fig:intro_data_format}, there are many types of continuity in data.
+A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph.
+In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector.
+Each \textit{component} of the record is a single object, such as a temperature, a color, or an image.
+The way in which these records are connected is the \textit{connectivity}, \textit{continuity}, or more generally \textit{topology}.
+\begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
@@
 Image modified from a diagram from Munzner's \textit{Visualization Analysis and Design} \cite{munznerVisualizationAnalysisDesign2014}}
-\label{fig:munzner_datatypes}
+   \label{fig:intro_munzner_datatypes}
    \end{figure}
 
-As shown in figure~\ref{fig:flowchart}, there are multiple ways to translate data into pictures.
-A map is always an option, except when the observations do not have associated coordinates in a physical plane.
-Tamara Munzner provides a way to distinguish between these datasets using $\{key, value\}$ designations \cite{munznerVisualizationAnalysisDesign2014}.
-Munzner defines \textit{values} as measurements of interest in the dataset, analogous to dependent variables in statistics.
-She defines \textit{keys} as indexes that can be used to look up values, analogous to independent variables in statistics and dimensions in computer science.
-Figure~\ref{fig:munzner_datatypes} illustrates how these keys are used to look up variables in a dataset:
-\begin{itemize}
-	\item map: keys are the coordinates of the points
-	\item table: row index, database primary key
-	\item data cube: row, column, etc. (.e.g. $i,j,k$) index
-\end{itemize}
+Often this topology has metadata associated with it, as shown in figure~\ref{fig:intro_munzner_datatypes}.
+Munzner denotes this metadata as \textit{keys} that can be used to locate the record, and further to locate the specific \textit{value} \cite{munznerChDataAbstraction}.
+In figure~\ref{fig:intro_munzner_datatypes} tables have keys identifying the row and column and sometimes also the depth, while maps have a location key.
+We propose that information rich keys such as location are additional components of the record, and that instead there are coordinate free structural keys that identify the location of the record within a dataset of any continuity.
 
-Expanding on Munzner's key and value semantics, in many datasets the keys are discrete variables like time or geophysical locations sampled from a continuous curve, surface, or field.
-While these observations are discrete samples from the continuous space, often the continuous (functional) characteristic\cite{ramsayFunctionalDataAnalysis2006a,mullerFunctionalVarianceProcesses2006a} of the observational space is also of interest.
-Besides quantitative discrete, quantitative continuous, or categorical measurement type considerations, the choice of visualization is also influenced by the measurement being on an interval, ratio, nominal, or categorical scale.
+In this work, we extend Butler's topology driven representation of visualization data \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Butler proposes that fiber bundles are a good model for visualization data because it allows for encoding the connectivity seperately from the records and supports discrete and ND continuous datasets.
+Since Butler's model lacks a robust way of describing the components of the record, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like discription of the data in the fiber bundle.
+We then propose criteria on expressivity that take into account both the components and the continuity of the data.
 
-Throughout this form we will be using the following conventions to discuss data:
+\subsection{Tools}
+\label{sec:intro_data_tools}
 
-\begin{description}
-    \item[component]
-    \item[index]  
-    \item[record]
-    \item[value]  
-\end{description}
-
-\subsection{how does this tackle the problem? what does it leave on the table?}
-Most visualization tools are designed around some notion of the structure of the data being visualized. For example Grammer of Graphics asusmes ythat
-
-Tables, images (Lev), graphs (network X)
-
-\begin{itemize}
-
-    \item GoG's mathematics of visualizations
-    This book focuses instead on rules for constructing graphs mathematically
-and then representing them as graphics aesthetically.
-The title of this book also recalls Bertin’s Semiology of Graphics (1967),
-the first and most influential structural theory of statistical graphics. Bertin’s
-work has pervaded our thinking. Semiology deals with signs. Although Bertin
-put his signs on paper, his work applies as well to virtual space.
-    \item vtk's qausi topological
-    \item butler's topological formulation
-\end{itemize}
-
-\subsection{How do other visualization libraries tackle this?}
-begin{figure}[h!]
-%%remake properly at some point
-\includegraphics[width=\textwidth]{figures/intro/flowchart.png} 
-\caption{To go from a dataset to a visualization, the data is subset based on a set of constraints (the invariant).
-The resulting subset becomes the components that are visualized, but the choice of visualization is dependent on the type and structure of the component variables.}
-\label{fig:flowchart}
-\end{figure}
-
-Given a dataset, we need to decide what subset of the data to visualize.
-Bertin describes the set of constraints used to subset the data as the \textit{invariant}.
-Formally, the \textit{invariant} is the set of shared characteristics of the data being visualized.
-When these constraints are applied to the dataset, the resulting subset is what will become the \textit{components} of the visualization \cite{bertinSemiologyGraphicsDiagrams2011a}.
-%%rework this statement for french labor data
-%%In figure~\ref{fig:iris_scatter}, the \textit{invariant} common to all the data being visualized is "sepal length", "petal length", and "species" and the \textit{components} are the measurements of these variables.
-As shown in figure~\ref{fig:flowchart}, the final step in creating a visualization is choosing how to encode the
-components using retinal (visual) variables.
-matplotlib arch paper, excel/matlab arch, 
-vtk \& ggplot (compare/contrast, we're blending these things)
+A motivator for this work is that currently Matplotlib carries implicit assumptions about data continuity in how each function interfaces with the input data.
+This work proposes a unified internal representation that encodes connectivity in a common interface.
+Matplotlib aims to natively support data of varying connectivities, so tools primarily concerned with visualizing relational data of the type found in databases are insufficient models.
+Many of these tools are built on top of Wilkenson's Grammar of Graphics \cite{wilkinsonGrammarGraphics2005} which itself incorporates Mackinlay's  A Presentation Tool (APT) display algebra; GoG derivative include ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+While many of these tools support images, the first class data container is a table like object of discrete records.
+Tools that primarily support images are also insufficient.
+For example ImageJ\cite{schneiderNIHImageImageJ2012} and the ImagePlot\cite{studiesCulturevisImageplot2021}macro have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized.
+Plugins exist, but must work around the everything is an image data model\cite{WritingPlugins}.
+There are also visualization tools that do not carry implicit assumptions about structure, for example vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011}.
+\note{Not totally positive but I think VTK is deeply coupled to the renderer and that's why we're not using it as a model, but not positive and this is literally what Marc worked on...}
 

### 2021-02-11 draft/sections/intro.tex
changed a sentence to remove sci/info seperation

@@
 In data visualization it is assumed the graphical representation of data match the properties of the data, and in this work we propose that the mathematical notion of equivariance formalizes this expectation.
-We meld the infoviz communities interest in heterogenous often discrete datasets with the scientific visualization communities emphasis on continuous and sometimes topologically complex datasets.
+We propose a topological model that can encode heterogenous components of topologically complex datasets.
 To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
@@
 \end{figure}
-
+%% find a way to work friendly in
 This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:intro_dubois}; to do so, we first discuss the criteria by which a visualization is evaluated.

### 2021-02-12 draft/sections/intro.tex
I can has todo list

@@
 \label{sec:intro}
+"We have a unified mathematical description of understanding visualization.
+Lots of people have worked on this and each has identified and described important aspects, but each has specialized in some way and it is not clear how to align them.
+We can understand all of the prior work via our framework that pulls from {established math}"
+that way you make clear what your contribution is, gives you a context to bring up all of the citations, a frame to contrast them, and then a nice lead into the math
+I like your framing
+and "we validate the framework by using it to re-design artist and data access layer of Maptlotlib, a general purpose visualization tool"
+
 In data visualization it is assumed the graphical representation of data match the properties of the data, and in this work we propose that the mathematical notion of equivariance formalizes this expectation.

### 2021-02-12 draft/sections/next.tex
I can has todo list

@@
 \begin{document}
+\section{Discussion}
 
-%% proposed work:
-%% transition functions, how it'd work on bundles
-%% compositing + algebra
-%% interactivity
-%% implementation
+\subsection{What are the limitations?}
+Is not useful for much yet.
+
+\subsection{What else needs to be implemented?}
+\begin{itemize}
+    \item heatmaps/ proof of concept of face simplex
+    \item composition algebra
+    \item boxplot - multiglyph composition
+    \item debatable: non-trivial bundle like data on mobius backend
+    \item interactivity 
+    \item concurrent artist
+    \item numpy wrapper
+\end{itemize}
+\begin{itemize}
+    \item provisional implementation using new architecture in Matplotlib
+    \item developer-aimed documentation (API + narrative)
+    \item proof of concept 3rd party user-facing library with a biology partner
+    \item academic paper about building visualization tools on top of the core architecture
+\end{itemize}
+
+\subsection{What are the lessons learned?}
 

### 2021-02-12 draft/sections/scraps.tex
I can has todo list

@@
-\begin{figure}[ht!]
-    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
-    \label{fig:data_monoid_rotation}
-    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.
-    %% dataset that is symmetric, fiber is x,y, rotatation action
-    %% rotate + plot or plot + rotate
-    %%emoji shuffle 
-    }  
-\end{figure}
+s demonstrate that this framework of artists (functor between sheafs) works for all or possibly almost all the pyplot module as a composition of primitive artists of this framework.
+It's kind of like your table but we will need to walk through Q carefully and look at how sampling will function in quiver plots,
+and contour plots. You will absolutely not have to have this worked out for the proposal, I'm just saying, composition, and demonstration of feasibility for the rest of pyplot will be important.
+You will have the code for many/some of the primitives (eg scatter, line, bar, imshow, spine), and a description of the composition operator (which is a union/join/+).
+You will also need to realize K as a triangulated space (possibly CW-Complex) for the purposes of the computation.
 
-\begin{figure}[ht!]
-    \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
+The fun part for you can go through the ideas in Munsner, Bertin, Heer, Wilkonson interpreting these things in terms of the prior work.
+That sounds to me like a bookchapter.
+We will have to see if there is an opertunity to get that sent somewhere.
+You won't have much time for this for you proposal but if we have published a coherent model, this could be a nice desert.
 
-    \caption{ The fiber $F$ is a continuous plane encoding the set of all possible (time, temperature) values.
-The base space $K$ expresses that the fibers lie in continous space and therefore the data is continuous.
-The mapping $\pi:E\rightarrow K$ from total space to base space is what defines this structure as a fiber bundle.
-The continuous function $\tau_1$ returns a record in each fiber for each point on $K$; the set of all values $\tau_1$ returns is a dataset that lies in this fiberbundle.
-\note{remove one of the sections from this figure/maybe kill this example}}
-    \label{fig:data_base_space} 
-\end{figure}
-The fiber bundle in figure~\ref{fig:data_base_space} is a representation of continuous (temperature, time) data.
-The fiber bundle abstraction facilitates seperately representing the time and temperature variables in the fiber $F$ and how these variables are connected in the base space $K$.
-The function $\tau_1$ returns a record (temperature, time) for ever point on the interval $K$ and there can be many $\tau$ functions in the fiber bundle.
+Other things that may or may not play a big role would be working out the details of the S->K interaction for dashboards.
+By this I particularly mean when S is a composite object spanning all the figures shown in the dashboard and a component of K may have different compoents of S mapping into it slicing it in different ways.
+
+*********************
+One thought I had that might be a blocker we hit several times, and we already discussed this, is the difference between the object, and how it is represented.
+The number 6 is called "perfect" because it is the sum of its proper divisors 1,2, and 3.
+If we wrote 6 in binary, 110, it would still be the sum of its proper divisors, 1, 10, 11.
+"6" as a postive number is independent of how you represent it in binary, hex, octal or whatever.
+The fact 111 has 3 digits which are "1" is no a property of the number unless you add in the representation in base 10.
+In may (and physics and other sciences) we are constantly keeping track of what things are artifacts of our implementation, and what is invariant to ANY implementation.
+In CS I often run into students getting this confused.
+If you don't implement in a very particular way, constrained by the math, you end up breaking the invariance.

### 2021-02-14 draft/sections/next.tex
- existential crisis on what is next?
I can has no idea

@@
 
+\subsection{What have we done}
+presented mathematical framework w/ topological models for the things we have done. 
+ \begin{itemize}
+    \item fiberbundle -> continuity + variables
+    \item fiberbundle -> graphic continuity + display abstraction
+    \item formalizing equivariance on both sides 
+    \item prototype for scatter, line, bar
+    \item simple composition into multibar
+    \item commonly used container - dataframe
+ \end{itemize}
+
 \subsection{What are the limitations?}
-Is not useful for much yet.
+This framework is aimed at developers, not end users which is why it's equation soup only understood by physicits and mathematicians.
+
+.  It is a way to describe the pathway from data to graphic in a
 
@@
 
+Other things that may or may not play a big role would be working out the details of the S->K interaction for dashboards.
+By this I particularly mean when S is a composite object spanning all the figures shown in the dashboard and a component of K may have different compoents of S mapping into it slicing it in different ways.
 \subsection{What are the lessons learned?}

### 2021-02-14 draft/sections/next.tex
lapsed into reading Mackinlay for notes on limitations

@@
 \subsection{What are the limitations?}
-This framework is aimed at developers, not end users which is why it's equation soup only understood by physicits and mathematicians.
-
-.  It is a way to describe the pathway from data to graphic in a
-
+\begin{itemize}
+\item is at arc level
+\item is compose(encodes) and ggplot is plot(encodes, glyph)
+\item can't handle figerative/semantic equivariance (this actually could be done as extension of fber)
+\item is only developed against Matplotlib
+\item is written against AGG/SVG
+\end{itemize}
 \subsection{What else needs to be implemented?}

### 2021-02-14 draft/sections/next.tex
wrote words, need sleep

@@
 
+%"This chapter steps back from the details and considers the larger issues that have been raised.
+In particular, it considers how this research helps to accomplish the motivating goal of this dissertation, which is the development of computer-based graphical communication." - Mackinlay
+"We have a unified mathematical description of understanding visualization.
+Lots of people have worked on this and each has identified and described important aspects, but each has specialized in some way and it is not clear how to align them.
+We can understand all of the prior work via our framework that pulls from {established math}"
+that way you make clear what your contribution is, gives you a context to bring up all of the citations, a frame to contrast them, and then a nice lead into the math
+ "we validate the framework by using it to re-design artist and data access layer of Maptlotlib, a general purpose visualization tool"
+
 \subsection{What have we done}
-presented mathematical framework w/ topological models for the things we have done. 
+This work contributes a mathematical description of the transformation from data to visual representation.
+Combining Butler's fiber bundle model of data with Spivaks formalism of data schemas provides a way of decoupling topology from variability such that the model can support a very large variety of datasets, including discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks.
+Modeling the graphic as a fiber bundle provides a way to seperate the target display space from the topology of the graphic.
+By decomposing the mapping from data to visual representation as encoding \vchannel, assembly \vmark, and a mapping between data and graphic topologies \vindex and formalizing what equivariance each stage needs to preserved, we can derive constraints that visualization library authors could embed in their code to guaranteee visualizations that are equivariant transforms of the input data.
+While Mackinlay defined the visual encodings as equivariant maps, a new contribution is the generalization of topology preservation as an equivariant map from graphic space to data space.
+
  \begin{itemize}
-    \item fiberbundle -> continuity + variables
-    \item fiberbundle -> graphic continuity + display abstraction
+    \item Munzner glyph I can has parallisim
     \item formalizing equivariance on both sides
+    \item supports multivariate components/encoders 
     \item prototype for scatter, line, bar
@@
 \subsection{What are the limitations?}
+%%%An interesting control problem will be how to keep the text and graphics consistent.
+If the rendering exchanges the horizontal and vertical axes of a bar chart, the text must also exchange its references to the axes.
 \begin{itemize}
 \item is at arc level
+\item is at one dataset level - needs invariants/composites
+\item static (what happens with interactive - but this is just update )
 \item is compose(encodes) and ggplot is plot(encodes, glyph)
-\item can't handle figerative/semantic equivariance (this actually could be done as extension of fber)
+\item can't handle figerative/semantic equivariance (this actually could maybe be done as extension of fber)
+\item explicitely does not care about effectiveness - is domain specific 
 \item is only developed against Matplotlib
 \item is written against AGG/SVG
+\item layout
+\item text
+\item same as APT, is at gd level so cannot control render 
+\item keeping all data in sync if transform is changed in one place (so on multi composed graphs, changing x on does it propopgate - part of Bertin invariance)
 \end{itemize}
@@
 \end{itemize}
-
 Other things that may or may not play a big role would be working out the details of the S->K interaction for dashboards.

### 2021-02-14 draft/sections/next.tex
not totally coherant, halfway through what did we do list

@@
 By decomposing the mapping from data to visual representation as encoding \vchannel, assembly \vmark, and a mapping between data and graphic topologies \vindex and formalizing what equivariance each stage needs to preserved, we can derive constraints that visualization library authors could embed in their code to guaranteee visualizations that are equivariant transforms of the input data.
-While Mackinlay defined the visual encodings as equivariant maps, a new contribution is the generalization of topology preservation as an equivariant map from graphic space to data space.
+While Mackinlay discussed constraining visual encodings as equivariant maps, a new contribution is the generalization of topology preservation as a deformation retraction from graphic space to data space.
+By using a functional paradigm, we can deconstruct the graphic to the glyph associated with each data point or even to pieces of a glyph; therefore the renderer has full  flexibility in how to to generate the image, while the graphic to data topology maps provide a book keeping mechanism to keep track of which part of the image belongs to which data point.
 

### 2021-02-14 draft/sections/next.tex
n-dimensional encoders and more words

@@
 Combining Butler's fiber bundle model of data with Spivaks formalism of data schemas provides a way of decoupling topology from variability such that the model can support a very large variety of datasets, including discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks.
-Modeling the graphic as a fiber bundle provides a way to seperate the target display space from the topology of the graphic.
-By decomposing the mapping from data to visual representation as encoding \vchannel, assembly \vmark, and a mapping between data and graphic topologies \vindex and formalizing what equivariance each stage needs to preserved, we can derive constraints that visualization library authors could embed in their code to guaranteee visualizations that are equivariant transforms of the input data.
-While Mackinlay discussed constraining visual encodings as equivariant maps, a new contribution is the generalization of topology preservation as a deformation retraction from graphic space to data space.
+Modeling the graphic as a fiber bundle provides a way to separate the target display space from the topology of the graphic.
+By decomposing the mapping from data to visual representation as encoding \vchannel, assembly \vmark, and a mapping between data and graphic topologies \vindex and formalizing what equivariance each stage needs to preserved, this work derives constraints that visualization library authors could embed in their code to guarantee visualizations that are equivariant transforms of the input data.
+
+This work generalizes previous research constraining visual encodings from data components to graphic components as equivariant maps to components that are N-dimensional.
+Furthermore, it precisely defines the glyph as the visual element constructed from data on a simplex in a simplacial complex where the simplex is discrete or continuous.
+This is a restatement of for example Bertin's definition of a line that encapsulates all points on the continuous line.
+By modeling the data topology along with its variablity, this work also provides  a generalization of topology preservation as a deformation retraction from graphic space to data space.
 By using a functional paradigm, we can deconstruct the graphic to the glyph associated with each data point or even to pieces of a glyph; therefore the renderer has full  flexibility in how to to generate the image, while the graphic to data topology maps provide a book keeping mechanism to keep track of which part of the image belongs to which data point.
@@
  \begin{itemize}
-    \item Munzner glyph I can has parallisim
-    \item formalizing equivariance on both sides
-    \item supports multivariate components/encoders 
     \item prototype for scatter, line, bar

### 2021-02-14 draft/sections/next.tex
0.0.1 of what have we done done

@@
 In particular, it considers how this research helps to accomplish the motivating goal of this dissertation, which is the development of computer-based graphical communication." - Mackinlay
-"We have a unified mathematical description of understanding visualization.
-Lots of people have worked on this and each has identified and described important aspects, but each has specialized in some way and it is not clear how to align them.
-We can understand all of the prior work via our framework that pulls from {established math}"
-that way you make clear what your contribution is, gives you a context to bring up all of the citations, a frame to contrast them, and then a nice lead into the math
- "we validate the framework by using it to re-design artist and data access layer of Maptlotlib, a general purpose visualization tool"
 
@@
 By modeling the data topology along with its variablity, this work also provides  a generalization of topology preservation as a deformation retraction from graphic space to data space.
-By using a functional paradigm, we can deconstruct the graphic to the glyph associated with each data point or even to pieces of a glyph; therefore the renderer has full  flexibility in how to to generate the image, while the graphic to data topology maps provide a book keeping mechanism to keep track of which part of the image belongs to which data point.
+By using a functional paradigm, we can deconstruct the graphic to the glyph associated with each data point or even to pieces of a glyph; therefore the renderer has full  flexibility in how to to generate the image, while the graphic to data topology maps provide a way to keep track of which part of the image belongs to which data point.
 
- \begin{itemize}
-    \item prototype for scatter, line, bar
-    \item simple composition into multibar
-    \item commonly used container - dataframe
- \end{itemize}
-
+The toy prototype built using this model validates that is usable for a general purpose visualization tool since it can be iteratively integrated into the existing architecture rather than starting from scratch/ Factoring out glyph formation into assembly functions allows for much more clarity in how the glyphs differ.
+This prototype demonstrates that this framework can generate the fundemental marks, point (scatter plot), line (line chart), and area (bar chart).
+Furthermore, the grouped and stacked bar examples demonstrate that this model supports composition of glyphs into more complex graphics.
+These composite examples also rely on the fiber bundles section base book keeping to keep track of which components contribute to the attributes of the glyph.
+Implementing this example using a Pandas dataframe demonstrates the ease of incorporating existing widely used data containers rather than requiring users to conform to one stands.
 \subsection{What are the limitations?}

### 2021-02-14 draft/sections/intro.tex
trying to compile next + fixed some errors

@@
 that way you make clear what your contribution is, gives you a context to bring up all of the citations, a frame to contrast them, and then a nice lead into the math
-I like your framing
-and "we validate the framework by using it to re-design artist and data access layer of Maptlotlib, a general purpose visualization tool"
+ "we validate the framework by using it to re-design artist and data access layer of Maptlotlib, a general purpose visualization tool"
 
@@
 Bertin first classified these graphic components as retinal variables and discussed which types of data they can express \cite{bertinIIPropertiesGraphic2011} and how they are composited on point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond.
-Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information}.
+Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014}.
 Mackinlay's expressiveness criteria is well defined for the visual variables, such that he suggests the viability of a strict encoding relation that is a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.

### 2021-02-14 draft/sections/math.tex
trying to compile next + fixed some errors

@@
 As shown in figure~\ref{fig:graphic_rho_lookup}, the output space queries into the graphic bundle to render the image.
-We select a pixel $p$ in the output space, inverse map the region of the pixel into $\gbase_p \subset \gbase$, then compute the section $gsection$ over the region $\gbase_p$.
+We select a pixel $p$ in the output space, inverse map the region of the pixel into $\gbase_p \subset \gbase$, then compute the section $\gsection$ over the region $\gbase_p$.
 The section yields the set of elements in \gfiber\ that specify the $(r, g, b)$ values corresponding to the region $p$.

### 2021-02-15 draft/sections/next.tex
trying to compile next + fixed some errors
half of limitations para

@@
 Implementing this example using a Pandas dataframe demonstrates the ease of incorporating existing widely used data containers rather than requiring users to conform to one stands.
+
 \subsection{What are the limitations?}
@@
 If the rendering exchanges the horizontal and vertical axes of a bar chart, the text must also exchange its references to the axes.
+So far this model has only been worked out for a single data set tied to a primitive mark, but it should be extensible to compositing datasets and complex glyphs.
+The examples and prototype have so far only been implemented for the static 2D case, but nothing in the math limits to 2D and expansion to the animated case should be possible because the model is formalized in terms of the sheaf.
+Since it is a purely mathematical framework, this model does not have a way to model or check for semantic equivariance of figurative representations, but this could be an avenue of research for machine learning researchers.
+
 \begin{itemize}
-\item is at arc level
-\item is at one dataset level - needs invariants/composites
-\item static (what happens with interactive - but this is just update )
-\item is compose(encodes) and ggplot is plot(encodes, glyph)
+
 \item can't handle figerative/semantic equivariance (this actually could maybe be done as extension of fber)
@@
 \end{itemize}
+
 \subsection{What else needs to be implemented?}
 \begin{itemize}
+    \item is compose(encodes) and ggplot is plot(encodes, glyph)
     \item heatmaps/ proof of concept of face simplex
@@
 \end{itemize}
+
 Other things that may or may not play a big role would be working out the details of the S->K interaction for dashboards.

### 2021-02-15 draft/sections/next.tex
finished out limitations section

@@
 \subsection{What are the limitations?}
+
+So far this model has only been worked out for a single data set tied to a primitive mark, but it should be extensible to compositing datasets and complex glyphs.
+The examples and prototype have so far only been implemented for the static 2D case, but nothing in the math limits to 2D and expansion to the animated case should be possible because the model is formalized in terms of the sheaf.
+While this model supports equivariance of figurative glyphs generated form parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it currently does not have a way to evaluate the semantic accuracy of the figurative representation.
+This model also does not currently factor in effectiveness, but potentially effectiveness criteria could be incorporated into a scheme for assigning encoders and assembling glyphs.
+This model and the associated prototype is deeply tied to Matplotlib's existing architecture.
+While the model is expected to generalize to other libraries, such as those built on Mackinlay's APT framework, this has not been worked through.
+Even though the model is designed to be backend independent, it has only really been tested against the AGG backend.
+It is especially unknown how this framework interfaces with high performance rendering libraries such as opengl\cite{CarsonOpenGL1997}.
+
+=
+\subsection{What else needs to be implemented?}
 %%%An interesting control problem will be how to keep the text and graphics consistent.
 If the rendering exchanges the horizontal and vertical axes of a bar chart, the text must also exchange its references to the axes.
-So far this model has only been worked out for a single data set tied to a primitive mark, but it should be extensible to compositing datasets and complex glyphs.
-The examples and prototype have so far only been implemented for the static 2D case, but nothing in the math limits to 2D and expansion to the animated case should be possible because the model is formalized in terms of the sheaf.
-Since it is a purely mathematical framework, this model does not have a way to model or check for semantic equivariance of figurative representations, but this could be an avenue of research for machine learning researchers.
-
-\begin{itemize}
-
-\item can't handle figerative/semantic equivariance (this actually could maybe be done as extension of fber)
-\item explicitely does not care about effectiveness - is domain specific 
-\item is only developed against Matplotlib
-\item is written against AGG/SVG
-\item layout
-\item text
-\item same as APT, is at gd level so cannot control render 
-\item keeping all data in sync if transform is changed in one place (so on multi composed graphs, changing x on does it propopgate - part of Bertin invariance)
-\end{itemize}
-
-\subsection{What else needs to be implemented?}
 \begin{itemize}
@@
     \item numpy wrapper
+    \item layout
+    \item text
+    \item keeping all data in sync if transform is changed in one place (so on multi composed graphs, changing x on does it propopgate - part of Bertin invariance)
 \end{itemize}

### 2021-02-15 draft/sections/next.tex
half of whats' next and figures

@@
 
-%"This chapter steps back from the details and considers the larger issues that have been raised.
-In particular, it considers how this research helps to accomplish the motivating goal of this dissertation, which is the development of computer-based graphical communication." - Mackinlay
-
-\subsection{What have we done}
 This work contributes a mathematical description of the transformation from data to visual representation.
@@
 
-\subsection{What are the limitations?}
+\subsection{Limitations}
 
@@
 Even though the model is designed to be backend independent, it has only really been tested against the AGG backend.
-It is especially unknown how this framework interfaces with high performance rendering libraries such as opengl\cite{CarsonOpenGL1997}.
+It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}.
 
-=
-\subsection{What else needs to be implemented?}
+\subsection{Next Steps}
+While the model and prototype demonstrate that generation of simple marks from the data, there is a lot of work left to develop a model that underpins a minimally viable library.
+While the model supports simple composition of glyphs by overlaying glyphs at the same position, more work is needed to define an operator where the fiber bundles have shared $\gbase_2 \hookrightarrow \gbase_1$ such that fibers could be pulled back over the subset.
+This complex operator is necessary for building semantically meaningful components such as axes labels and interactive visualizations that update on shared \gbase, such as brush-linked views\cite{}
+
 %%%An interesting control problem will be how to keep the text and graphics consistent.

### 2021-02-15 draft/sections/next.tex
trying to sort out diff from mackinlay

@@
 Other things that may or may not play a big role would be working out the details of the S->K interaction for dashboards.
-By this I particularly mean when S is a composite object spanning all the figures shown in the dashboard and a component of K may have different compoents of S mapping into it slicing it in different ways.
+By this I particularly mean when S is a composite object spanning all the figures shown in the dashboard and a component of K may have different components of S mapping into it slicing it in different ways.
 \subsection{What are the lessons learned?}

### 2021-02-15 draft/sections/next.tex
finished out conclusion

--- 2021-02-15 draft/sections/next.tex
@@
 This model also does not currently factor in effectiveness, but potentially effectiveness criteria could be incorporated into a scheme for assigning encoders and assembling glyphs.
+Also, even though the model is designed to be backend independent, it has only really been tested against the AGG backend.
+It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}.
+Because this model has been limited to the graphic design space, it does not address the critical task of laying out the graphics in the image
+
 This model and the associated prototype is deeply tied to Matplotlib's existing architecture.
 While the model is expected to generalize to other libraries, such as those built on Mackinlay's APT framework, this has not been worked through.
-Even though the model is designed to be backend independent, it has only really been tested against the AGG backend.
-It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}.
+In particular, Mackinlay's formulation of graphics as a language with semantic and syntax lends itself a declarative interface\cite{loudenProgrammingLanguagesPrinciples2010}, which Heer and Bostock use to develop a domain specific visualization language that they argue makes it simpler for designers to construct graphics without sacrificing expressivity \cite{heerDeclarative2010}.
+Similarly, the model presented in this work formulates visualization as equivariant maps from data space to visual space, and is designed such that developers can build software libraries with data and graphic topologies tuned to specific domains.
 
-\subsection{Next Steps}
+\subsection{Future Work}
 While the model and prototype demonstrate that generation of simple marks from the data, there is a lot of work left to develop a model that underpins a minimally viable library.
-While the model supports simple composition of glyphs by overlaying glyphs at the same position, more work is needed to define an operator where the fiber bundles have shared $\gbase_2 \hookrightarrow \gbase_1$ such that fibers could be pulled back over the subset.
-This complex operator is necessary for building semantically meaningful components such as axes labels and interactive visualizations that update on shared \gbase, such as brush-linked views\cite{}
+Foremost is implementing a data object that encodes data with a 2D continous topology and an artist that can consume data with a 2D topology to visualize the image\cite{toryRethinkingVisualizationHighLevel2004,haber1990visualization,hansen2011visualization} and also encoding a separate heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas} artist that consumes 1D discrete data.
+A second important proof of concept artist is a boxplot\cite{wickham40YearsBoxplots2011} because it is a graphic that assumes computation on the data side and the glyph is built from semantically defined components and a list of outliers.
 
-%%%An interesting control problem will be how to keep the text and graphics consistent.
-If the rendering exchanges the horizontal and vertical axes of a bar chart, the text must also exchange its references to the axes.
+The model supports simple composition of glyphs by overlaying glyphs at the same position, but more work is needed to define an operator where the fiber bundles have shared $\gbase_2 \hookrightarrow \gbase_1$ such that fibers could be pulled back over the subset.
+This complex operator is necessary for building semantically meaningful components such as interactive visualizations that update on shared \gbase, such as brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991} and verifying constraints in multiple view systems \cite{hullmanKeeping2018}.
+The models simple addition supports axes as standalone artists with overlapping visual position encoding, but the complex operator would allow for keeping labels consistent with text when for example the horizontal and vertical positions of the data are changed.
+In summary, the proposed scope of work for the dissertation is
 \begin{itemize}
-    \item is compose(encodes) and ggplot is plot(encodes, glyph)
-    \item heatmaps/ proof of concept of face simplex
-    \item composition algebra
-    \item boxplot - multiglyph composition
-    \item debatable: non-trivial bundle like data on mobius backend
-    \item interactivity 
-    \item concurrent artist
-    \item numpy wrapper
-    \item layout
-    \item text
-    \item keeping all data in sync if transform is changed in one place (so on multi composed graphs, changing x on does it propopgate - part of Bertin invariance)
+    \item expansion of the mathematical framework to include complex addition
+    \item mathematical formulation of a graphic with axes labeling
+    \item implementation of data with 2D continuous topology and a compatible artist
+    \item provisional mathematics and implementation of user level composite artists
+    \item proof of concept domain specific user facing library 
 \end{itemize}
-\begin{itemize}
-    \item provisional implementation using new architecture in Matplotlib
-    \item developer-aimed documentation (API + narrative)
-    \item proof of concept 3rd party user-facing library with a biology partner
-    \item academic paper about building visualization tools on top of the core architecture
-\end{itemize}
+Additionally, implementing the complex addition operator would allow for developing a mathematical formalism and prototype of how interactivity would work in this model.
+Other potential tasks for future work is implementing a data object for a non-trivial fiber bundle and exploiting the models section level formalism to build distributed data source models and concurrent artists.
+This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
+While this paper formulates the math in terms of topology and groups, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers}.
 
-Other things that may or may not play a big role would be working out the details of the S->K interaction for dashboards.
-By this I particularly mean when S is a composite object spanning all the figures shown in the dashboard and a component of K may have different compoents of S mapping into it slicing it in different ways.
-\subsection{What are the lessons learned?}
-
+\section{Conclusion}
+Despite the alternative formalism to Mackinlay's APT framework, it is indebted to APT for providing a guideline for issues the model needs to address.
+As with APT, this works aims to develop a framework on which to build visualization tools.
+Unlike APT, our model is geared towards balancing the needs of the the scientific and information visualization communities; therefore we present a framework that can describe both how the components of the data are encoded and how the continuity is preserved in the graphic.
+We hope that this framework can distill the constraints on the data to graphic mapping such that it can be used to improve the architecture of a heavily used visualization tool and allow developers to more easily build domain specific tools.
 \end{document}

### 2021-02-15 draft/sections/intro.tex
rewrite opening paragraph

--- 2021-02-11 draft/sections/intro.tex
@@
 \label{sec:intro}
-In data visualization it is assumed the graphical representation of data match the properties of the data, and in this work we propose that the mathematical notion of equivariance formalizes this expectation.
-We propose a topological model that can encode heterogenous components of topologically complex datasets.
-To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
-In addition to providing a way to ensure the library preserves structure, we propose a functional approach to  improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
+
+ This work presents a mathematical description of how data is transformed into graphic representations.
+While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
+In this work, we present a framework for understanding visualization built on topology and group theory.
+Using this mathematical formalism, we can interpret and extend prior work and also develop new tools.
+We validate our model by using it to re-design artist and data access layer of Maptlotlib, a general purpose visualization tool.
 
@@
 \end{figure}
-%% find a way to work friendly in
+
 This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:intro_dubois}; to do so, we first discuss the criteria by which a visualization is evaluated.

### 2021-02-15 draft/sections/intro.tex
more specified about math:

@@
 While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
-In this work, we present a framework for understanding visualization built on topology and group theory.
+In this work, we present a framework for understanding visualization as equivariant maps between topological spaces.
 Using this mathematical formalism, we can interpret and extend prior work and also develop new tools.
-We validate our model by using it to re-design artist and data access layer of Maptlotlib, a general purpose visualization tool.
+We validate our model by using it to re-design artist and data access layer of Matplotlib, a general purpose visualization tool.
 

### 2021-02-15 draft/sections/math.tex
more specified about math:

@@
 
-%% whats the right model/abstraction for a visualization library? 
-%% why? the role this math is serving - static type checking/ reflection
-%% tracking/automatically inferring
-%% bookend w/ how this math does the why 
-%% if we deleted this, what would we lose? 
-%% can express continuous functions & not just discrete points
-\note{should this be in 3rd person passive?}\\
+can express continuous functions & not just discrete points
 Visualization is generally thought of as structure preserving maps from data into graphics, and in this section we formally define that structure and how it is preserved via equivariant maps.

### 2021-02-15 draft/sections/next.tex
more specified about math:

@@
 This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
-While this paper formulates the math in terms of topology and groups, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers}.
+While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
 

### 2021-02-15 draft/sections/math.tex
plus operator

@@
 
-can express continuous functions & not just discrete points
+can express continuous functions \& not just discrete points
 Visualization is generally thought of as structure preserving maps from data into graphics, and in this section we formally define that structure and how it is preserved via equivariant maps.
@@
 
-\subsubsection{Visual Channels}
+\subsubsection{Visual Channels \vchannel}
 \label{sec:artist_nu}
@@
 
-\subsubsection{Assembling Marks}
+\subsubsection{Assembling Glyphs \vmark}
 \label{sec:artist_q}
@@
 
-\note{DEGENERATE Q - drawing a blank if this is necessry and if so how\\
-Check a well defined map $\monoid\times Y \rightarrow Y$}
-
 To output a mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}, \vmark\ is called with all the regions \gbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$:
@@
 
-\subsubsection{Sample Qs}
-%rewrite as alpha/beta 
+\subsubsection{Sample Glyphs \vmark}
 In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and a heatmap. 
-\label{sec:artist_example_scatter}
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/scatter.png}
+    \includegraphics[width=.75\textwidth]{figures/math/scatter.png}
     \caption{The data is discrete points (temperature, time).
@@
 such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
-
-\label{sec:artist_example_line}
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/line.png}
+    \includegraphics[width=.75\textwidth]{figures/math/line.png}
     \caption{The line fiber $(time,\, temp)$ is thickend with the derivative $(time^{\prime},\, temperature^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a thick line.
@@
 
-\paragraph{Q: heatmap}\mbox{} \\
-\label{sec:artist_example_heatmap}
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/heatmap.png}
+    \includegraphics[width=.75\textwidth]{figures/math/heatmap.png}
     \caption{The only visual parameter a heatmap requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
@@
 
-\subsubsection{+ union}
+\subsubsection{Composition of Artists: +}
 \begin{figure}
-    \caption{components of axes + glyphs }
+    \includegraphics[width=1\textwidth]{figures/math/addition_operator.png}
+    \caption{$\dsection_1$ and $\dsection_2$ are distinct datasets passed through artists $\vartist_1$ and $\vartist_2$ to generate graphics $\gsection_1$ and $\gsection_2$.
+These graphics happen to be rendered to the same image, but otherwise have no intrinsinc link.}
+    \label{fig:artist_plus}
 \end{figure}
+In this paper we define a simple addition operator that is the disjoint union of fiber bundles \dtotal.
+For example, in figure~\ref{fig:artist_plus} the scatter plot $\dtotal_1$ and the line plot $\dtotal_2$ have different \dbase that are mapped to seperate \gbase.
+To fully display both graphics, the composite graphic $\vartist_1 + \vartist_2$ needs to include all records on both $\dbase_1$ and $\dbase_2$, which are the sections on the disjoint union $\dbase_1 \sqcup \dbase_2$.
+This in turn yields disjoint graphics $\gbase_1 \sqcup \gbase_2$ rendered to the same image.
+Constraints can be placed on the disjoint union such as that the fiber components need to have the same \vchannel position encodings or that the position \vsection need to be in a specified range.
+There are situations where $\dbase_2 \hookrightarrow \dbase_1$ that underpin more complex, especially interactive, visualizations; these cases require defining a more complex addition operator that is out of scope for this work.
+
 disjoin union of fibers so we can have an axes/real figure, - diagram operator w/o overplotting, library won't auto stack, have to bake that into data, haskell data/depth, 
@@
 Equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\, but act on different sections \vsection.
-To further simplify implementation, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data\note{needs citation, maybe friendlys history or acquired codes of meaning}.
-\begin{figure}
+To further simplify implementation, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data such that the graphic is identifiable as a given chart type.
+
+\begin{figure}[H]
     \caption{Each of these graphics is generated by a different artist \vartist which is the equivalance class of scatter plots $\vartist^{\prime}$\note{this is gonna be a whole bunch of scatter plots}}

### 2021-02-15 draft/sections/math.tex
qhat

@@
 The group action on these sets is permutation, so shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
-To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$, $\text{ if } delement_1 \leq delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$.
+To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$, $\text{ if } delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$.
 For interval scale data, \vchannel\ is equivariant under translation monoid actions if $\vchannel (x + c) = \vchannel(x) + \vchannel(c)$.
@@
 
+\subsubsection{Assembly factory \vmarkd}
+\begin{figure}
+    \includegraphics[width=.75\textwidth]{figures/math/q_hat.png}
+    \caption{The pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint.
+Because the \vsection\ is the same, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent set of \gbasepoint\ associated to that \dbasepoint}
+    \label{fig:artist_q_hat}
+\end{figure}
+As shown in eq~\ref{eq:artist_diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+Because $\vpreimg\ subset \gbase$ such that many \gbasepoint\ go to one \dbasepoint, by definition of the pull back $\vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber$.
+This means that the fiber \vtotal is copied over the preimage in \vpreimg.
+Given a section \vsectionpull\ pulled back from \vsection\ on $\pi:\vtotal\rightarrow\dbase$ and $\gbasepoint \in \vpreimg$, then the section $\vsectionpull(s) = \vindex^*(\vsection(\dbasepoint))$ is the image of $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$ for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
+This is illustrated in figure~\ref{fig:artist_q_hat}, where a \vsection\ over \dbasepath\ is copied over all  $\vindex(\gbasepoint) = \dbasepoint$.
+
+When \vmark\ maps sections $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull, then $\vmark(\vsectionpull) = \gsection$ depends on \gbasepoint.
+This means that $\gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)$, but since \vsectionpull\ is constant on \vpreimg, we can define a function $\vmarkd(\vsection(k))(s) \coloneqq \vmark((\vsectionpull)(\gbasepoint))$ where $\vpreimg = \dbasepoint$.
+
+In fact, \vmarkd\ is a map from visual to graphic $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
+This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
+The advantage of \vmarkd\ is that \gbase is part of the rendering stage of the pipeline and therefore hidden from the Matplotlib artists and therefore there is no direct way to construct a true \vmark.
+
 \subsubsection{Composition of Artists: +}
@@
 There are situations where $\dbase_2 \hookrightarrow \dbase_1$ that underpin more complex, especially interactive, visualizations; these cases require defining a more complex addition operator that is out of scope for this work.
-
-disjoin union of fibers so we can have an axes/real figure, - diagram operator w/o overplotting, library won't auto stack, have to bake that into data, haskell data/depth, 
-hover, mouse over curve, picked a point selecting a little k, + operator/disjoint union in different k, same D
-%% but what if it;s supposed to be Bertin invariant? will do w/ inclusion of K
 

### 2021-02-16 draft/sections/math.tex
notation call out + start of math revisions

@@
 
-can express continuous functions \& not just discrete points
-Visualization is generally thought of as structure preserving maps from data into graphics, and in this section we formally define that structure and how it is preserved via equivariant maps.
+As discussed in the introduction, visualization is generally thought of as structure preserving maps from data into graphic representation.
+In order to formalize this statement, we capture the connectivity of the records with topolology and the structure on the components of the records with the monoidal actions on the sets those values belong to.
+
+In this section we formally define that structure and how it is preserved via equivariant maps tha can express discrete and continuous topology.
+
 We can then specify that a faithful visual mapping is structure preserving, and apply these constraints to visualizations we may want to develop or implement.
+
 We model the data, visual characteristic, and graphic stages of visualization, shown in figure~\ref{fig:artist_stages}, as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.

### 2021-02-16 draft/sections/intro.tex
notation call out + start of math revisions
rework of data definitions

@@
 Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014}.
+
 Mackinlay's expressiveness criteria is well defined for the visual variables, such that he suggests the viability of a strict encoding relation that is a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
@@
 \end{figure}
-
 As shown in figure~\ref{fig:intro_data_format}, there are many types of continuity in data.
 A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph.
-In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector.
-Each \textit{component} of the record is a single object, such as a temperature, a color, or an image.
+
+\begin{mdframed}[roundcorner=10pt, frametitle= definitions, frametitlerule=true, frametitlebackgroundcolor=gray!10]
+ In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements.
+Each \textit{component} of the record is a single object, such as a temperature, a color, or image.
+We also generalize \textit{component} to mean all objects in the dataset of a given type, such as all temperatures or colors or images.
 The way in which these records are connected is the \textit{connectivity}, \textit{continuity}, or more generally \textit{topology}.
+\end{mdframed}
+
 \begin{figure}[H]

### 2021-02-16 draft/sections/math.tex
more math rewording

--- 2021-02-15 draft/sections/math.tex
@@
 
-can express continuous functions \& not just discrete points
-Visualization is generally thought of as structure preserving maps from data into graphics, and in this section we formally define that structure and how it is preserved via equivariant maps.
-We can then specify that a faithful visual mapping is structure preserving, and apply these constraints to visualizations we may want to develop or implement.
+As discussed in the introduction, visualization is generally defined as structure preserving maps from data into graphic representation.
+In order to formalize this statement, we describe the connectivity of the records using topology and define the structure on the components in terms of the monoid actions on the component types.
+By formalizing structure in this way, we can evaluate the extent to which a visualization preserves the structure of the data it is representing and build structure preserving visualization tools.
+
 We model the data, visual characteristic, and graphic stages of visualization, shown in figure~\ref{fig:artist_stages}, as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.

### 2021-02-16 draft/sections/intro.tex
moving figure, working through checklists

@@
 \label{sec:intro}
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/dar.png}
+    \caption{Visualization is equivariant maps between data and visual encoding of the variables and assembly of those encodings into a graphic.
+    \note{will replace w/ overarching figure w/ same structure}}
+    \label{fig:intro_artist_stages}
+\end{figure}
 
- This work presents a mathematical description of how data is transformed into graphic representations.
+ This work presents a mathematical description of how data is transformed into graphic representations, as shown in ref~\ref{fig:intro_artist_stages}.
 While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.

### 2021-02-16 draft/sections/math.tex
moving figure, working through checklists

@@
 \label{sec:tam}
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/dar.png}
-    \caption{Visualization is equivariant maps between data and visual encoding of the variables and assembly of those encodings into a graphic.
-    \note{not gonna name these bubbles tau, mu, rho, but might keep the same basic structure of different types of data and encodings }}
-    \label{fig:artist_stages}
-\end{figure}
 
@@
 
-We model the data, visual characteristic, and graphic stages of visualization, shown in figure~\ref{fig:artist_stages}, as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
+We model the data, visual characteristic, and graphic stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
 

### 2021-02-16 draft/sections/math.tex
rework of E and fiber

@@
 
-As discussed in the introduction, visualization is generally defined as structure preserving maps from data into graphic representation.
+As discussed in the introduction, visualization is generally defined as structure preserving maps from data to graphic representation.
 In order to formalize this statement, we describe the connectivity of the records using topology and define the structure on the components in terms of the monoid actions on the component types.
 By formalizing structure in this way, we can evaluate the extent to which a visualization preserves the structure of the data it is representing and build structure preserving visualization tools.
-
-We model the data, visual characteristic, and graphic stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
-
-We introduce a mathematical description of the visualization pipeline where artist $\mathscr{\vartist}$ functions transform data space $\mathscr{\dtotal}$ to an intermediate representation in a prerendered graphic space $\mathscr{\gtotal}$.
-
+We introduce the notion of an artist $\mathscr{\vartist}$ as a structure preserving map from data $\mathscr{\dtotal}$ to $\mathscr{\gtotal}$
 \begin{equation}
@@
 
-We first describe how we model data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles.
+We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
+To explain which structure the artist is preserving, we first describe how we model data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles.
 We then discuss the equivariant maps between data and visual characteristics (\ref{sec:artist_nu}) and visual characteristics and graphics (\ref{sec:artist_q}) that make up the artist.
+\note{xi should maybe be moved down to artist but is more readable in graphics}
 
@@
 \label{sec:data}
-We build on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989} because fiber bundles are mathematical structures that are flexible enough express all the types of data described in section~\ref{sec:intro_data_structure}.
-
-We model data as the fiber bundle $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$, where \dtotal\, \dfiber\, and \dbase\ are topological spaces that encode 
-\begin{description}
-\item[\dfiber] the properties of the variables in the fiber (\ref{sec:data_fiber})
-\item[\dbase] the continuity of the records in the base space (\ref{sec:data_base})
-\item[\dsection] collections of records (\ref{sec:data_section}). 
-\end{description}
-
-and \dtotal\ is the total space of data that \dfiber\ lives in. The bundle is the projection map $\pi$
+Building on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989},  fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
 \begin{equation}
@@
 
-that binds the variables \dfiber\ continuity \dbase.
-The fiber bundles mentioned in this work are assumed to be trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified, because the trivial bundle is $\dtotal=\dbase\times \dfiber$ such that extra structure in the total space \dtotal\ falls out and discussion can be focused on the fiber and base space.
+that binds the components of the data in \dfiber\ to the continuity represented in \dbase.
+We use fiber bundles as the data model because they are mathematical structures that are flexible enough express all the types of data described in section~\ref{sec:intro_data}.
+The fiber bundle models
+\begin{description}
+\item[\dfiber] properties of data component types (\ref{sec:data_fiber})
+\item[\dbase] continuity of records (\ref{sec:data_base})
+\item[\dsection] collections of records (\ref{sec:data_section}). 
+\end{description}
+where \dtotal\ is the space of all possible datasets with these components and continuity.
+The fiber bundles mentioned in this work are assumed to be \note{locally} trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified.
+We assume \note{local} triviality to dispenses with extra structure on $\dtotal$ and focus on the fiber and base space.
+\note{(and can do so b/c all bundles are locally trivial)}
 
@@
 \label{sec:data_fiber}
-%%% so we can do units, static + dynamic type checking, so that we can tak about variables 
-The fiber is a topological space that is the set of possible values of the data; the values themselves can be any dimension and type and have any continuity.
-We use Spivak's description of simplicial database schemas \cite{spivakSIMPLICIALDATABASES} as the basis of our fiber space because he binds the components of the fiber to variable names and types.
+To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names and types.
 Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_n\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.

### 2021-02-17 draft/sections/math.tex
fibers, start of monoids

@@
 where \dtotal\ is the space of all possible datasets with these components and continuity.
-The fiber bundles mentioned in this work are assumed to be \note{locally} trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, unless otherwise specified.
-We assume \note{local} triviality to dispenses with extra structure on $\dtotal$ and focus on the fiber and base space.
-\note{(and can do so b/c all bundles are locally trivial)}
+By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on $\dtotal$ and focus on the fiber and base spaces.
 
@@
 \label{sec:data_monoid}
-%% pin down the structure we're trying to preserve, this is what we mean by equivarance, most general useful formalism is a monoid, most specifically, more commonly it's a group but sometimes it's a mpnoid
-After specifying \dfiber\, we next describe the ways in which we can transform the values by identifying the monoid actions \monoid\ on the \dfiber.
-We use monoids as the abstraction because they encode composibilty, which maps well to the data transformation process in a software library \cite{yorgeyMonoidsThemeVariations}.
+Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define as the action of a monoid on the component.
+While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's scales \cite{stevensTheoryScalesMeasurement1946}, we generalize to monoids to support more component types.
+Monoids are also commonly found in functional programming because they specify compositions of transformations \cite{yorgeyMonoidsThemeVariations, stievenMonadJustMonoid2020}.
 
@@
 % scatter and line, heatmap express the continuity of the data, deeply implicit in every visualization system but equivariance of continuity of the records, explicit  topplogy/structure preserving - contuinity is the topology
+
 The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the base space \dbase\ without making assumptions as to what that continuity is.

### 2021-02-17 draft/sections/math.tex
why monoids

@@
 
-A monoid \cite{Monoid2021} $\monoid_i$ is a set with an associative binary operator $\ast:\monoid_i \times \monoid_i\rightarrow \monoid_i$.
-A monoid has an identity element $e\in \monoid_i$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid_i$.
-A left monoid action \cite{SemigroupAction2021,ActionNLab} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
+A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$.
+A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$.
+As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,ActionNLab} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
 \begin{align*}
@@
 Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-For example, given  the fiber $(\fname=temperature,\, \ftype=\texttt{float},\,\fttype=\reals)$ which is interval data:
+For example, given the interval scale fiber component $(\fname=temperature,\, \ftype=\texttt{float},\,\fttype=\reals)$:
 \begin{itemize}
@@
 
+While many component types will be one of the measurement scale types, we generalize to monoids specifically for the case of partially ordered set.
+Given a set $W=\{mist, \, drizzle, \, rain \}$, then the map $f: W\rightarrow W$ defined by
+\begin{enumerate}
+    \item $f(rain) = drizzle$,
+    \item  $f(drizzle) = mist$ 
+    \item $f(mist) = mist$
+\end{enumerate}
+is order preserving such that $mist \leq drizzle \leq rain$ but has no inverse since $drizzle$ and $mist$ go to the same value $mist$.
+Therefore order preserving maps do not form a group, and instead we generalize to monoids to support partial order component types.
+
 \subsubsection{Continuity: Base Space $K$} 
 \label{sec:data_base}
-% scatter and line, heatmap express the continuity of the data, deeply implicit in every visualization system but equivariance of continuity of the records, explicit  topplogy/structure preserving - contuinity is the topology
+The base space \dbase\ is way to represent the connectivity of the records.
+This is currently assumed in the choice of visualization, for example a line plot implies 1D continuous data and a scatter plot implies discrete records.
+Instead, this model can explicitly check if the visualization topology is equivalent to the
+
+Representing the data as a fiber bundle, allows one to express the continuity one assumes between observations.
+For instance if we want to consider data in a time series, with time implicitly a continuous variable, we could express the base space as a topological line.
+If we had cyclical data, then the base space could be represented as a circle.
+If the each observation was indpenedent then the data could be represented as discrete points
 
 The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the base space \dbase\ without making assumptions as to what that continuity is.
-In turn this representation of continuity can then be used to keep track of how the data fits together, for example if a visualization of a very large dataset calls for parallelization.
 

### 2021-02-17 draft/sections/math.tex
sheafs

@@
 \label{sec:data}
-Building on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989},  fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
+Building on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
 \begin{equation}
@@
 \label{sec:data_base}
-The base space \dbase\ is way to represent the connectivity of the records.
-This is currently assumed in the choice of visualization, for example a line plot implies 1D continuous data and a scatter plot implies discrete records.
-Instead, this model can explicitly check if the visualization topology is equivalent to the
-
-Representing the data as a fiber bundle, allows one to express the continuity one assumes between observations.
-For instance if we want to consider data in a time series, with time implicitly a continuous variable, we could express the base space as a topological line.
-If we had cyclical data, then the base space could be represented as a circle.
-If the each observation was indpenedent then the data could be represented as discrete points
-
-The advantage of fiber bundles is they provide a way to encode the continuity in a dataset as the base space \dbase\ without making assumptions as to what that continuity is.
+The base space \dbase\ is way to express the connectivity of the records.
+This is assumed in the choice of visualization, for example a line plot implies 1D continuous data and a scatter plot implies discrete records, but an explicit representation allows for verifying that the topology of the graphic representation is equivalent to the topology of the data.
+.
 
@@
 \label{sec:data_section}
-%% how do we represent data
-While the fiber and base space describe the general structure of all data that lives in the fiber bundle, the sections $\dsection: \dbase\rightarrow \dtotal$ define the datasets that live in the fiber.
-We generalize Spivak's description of the section as a table of records \cite{spivakSIMPLICIALDATABASES} to any sort of structured dataset such that
+A section $\dsection: \dbase\rightarrow \dtotal$ of the fiber bundle is a set of records.
+For example, in the special case of a table \cite{spivakSIMPLICIALDATABASES}, \dbase\ is a set of row ids, \dfiber\ is the columns, and the section \dsection\ returns the record at a given key in \dbase.
+In general, for any fiber bundle, there is always a map
 \begin{equation}
@@
 \end{equation}
-such that there is always a map $\pi(\dsection(\dbasepoint)) = \dbasepoint$.
-There can be many sections \dsection; the space of global sections is $\Gamma(\dtotal)$.
-For a trivial fiber bundle, the section is
+ such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$.
+There can be many sections \dsection\ and the space of all global sections is $\Gamma(\dtotal)$.
+Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is
 \begin{equation}
@@
 \end{equation}
-where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber. Because we can decompose the bundle and the fiber, we can formulate $\dsection$ as 
+where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber.
+This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle.
+Because we can decompose the bundle and the fiber, we can formulate $\dsection$ as
 \begin{align}
@@
 Section $\dsection_j$ returns a different timeseries of (time, temperature) values.
-Both sections are included in the global set of sections $\dsection_1, \dsection_2 \in \Gamma(\dtotal)$.
+Both sections are included in the global set of sections $\dsection_i, \dsection_j \in \Gamma(\dtotal)$.
 
@@
 \label{sec:data_sheaf_stalk}
-Often a graphic may need to be updated with live data or support zooming in on a segement of the dataset; to support working with a subset of data, we can use the sheaf $\mathcal{O}(\dtotal)$.
-All fiber bundles are locally trivial, which means that \dtotal\ restricted over a small enough neighborhood $U \subset \dbase$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
+To support working with a subset of data, we can take the sheaf $\mathcal{O}(\dtotal)$ as input into the artist.
+This facilitates interactions such as zooming or updating the graphic based on zooming data.
+\dtotal\ restricted over a small enough neighborhood $U \subset \dbase$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
 The sheaf $O(\dtotal)$ is the localized section of fibers $\iota^*\dsection: U \rightarrow \iota^*\dtotal$
@@
 The neighborhood of points $\dbasepoint_i$ surrounding the point \dbasepoint\ the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
-While \dtotal\ is only the fiber $\dfiber_k$ over a specific \dbasepoint, the stalk includes nearby records because the sheaf lies over the neighborhood $U$.
-While this can be useful for visual transforms, often the extra needed information can be found in the smaller jet bundle $\mathcal{J}$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016}.
-For example, line thickness requires the derivative of the given position to be rendered, which can be found in $\dtotal^\prime=\dtotal+\mathcal{J}(\dtotal)$
+
+While \dtotal\ is only the fiber $\dfiber_k$ over a specific \dbasepoint, the stalk includes the local behavior of the section at \dbasepoint\ which can include information about nearby records.
+In practice, often the only information needed from the stalk is some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
+In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives.
 

### 2021-02-17 draft/sections/math.tex
one issue down, three to go

@@
 \label{sec:graphic}  
-We can separate the structure of the graphic from the properties of the output format by modeling the space of graphics as a fiber bundle  $(\gtotal, \gbase, \pi, \gfiber)$.
-As with data, the fiber bundle is for a class of graphics with shared base space \gbase (~\ref{sec:graphic_fiber}) and fiber \gfiber (~\ref{sec:graphic_base}) and the sections \gsection (~\ref{sec:graphic_section}) encode a graphic where the visual characteristics are fully specified.
+We introduce a graphic bundle to hold the essential information necessary to render a graphical design constructed by the artist.
+As with the data, we can represent the target graphic as a section \gsection\ of a bundle  $(\gtotal, \gbase, \pi, \gfiber)$.
+The graphic bundle \gtotal\ consists of a base \gbase (~\ref{sec:graphic_fiber}) that is a thickened form of \dbase\,
+a fiber \gfiber (~\ref{sec:graphic_base}) that is an idealized display space, and sections \gsection (~\ref{sec:graphic_section}) that encode a graphic where the visual characteristics are fully specified.
 
@@
 \label{sec:graphic_fiber}
-%% bookends sentences
-%% first we need a  target output abstraction 
-%% this allows us to w/ some level of detail specify a target display
-The fiber \gfiber\ is an idealized infinite resolution version of the target display space, for example a 2D screen or 3D printer.
-In this work, we assume a 2D opaque image $\dfiber=\reals^5$ with elements
+To fully specify the visual characteristics of the image, we construct a fiber \gfiber\ that is an infinite resolution version of the target space.
+Typically \gtotal\ is trivial and therefore sections can be thought of as mappings into \gfiber.
+
+In this work, we assume a 2D opaque image $\gfiber=\reals^5$ with elements 
 \begin{equation}
@@
 such that a rendered graphic only consists of 2D position and color.
-To support overplotting and transparency, the fiber could be $\dfiber=\reals^{7}$ such that $(x, y, z, r, g, b, a) \in \gfiber$ specifies the target display.
+To support overplotting and transparency, the fiber could be $\gfiber=\reals^{7}$ such that $(x, y, z, r, g, b, a) \in \gfiber$ specifies the target display.
+By abstracting the target display space as \gfiber, the model can support different targets, such as a 2D screen or 3D printer.
 
@@
 \label{sec:graphic_base}
-%% structure preservation on the graphics side 
-An assumption of graphical representations is that they match the continuity of the data\cite{tufteVisualDisplayQuantitative2001,friendlyBriefHistoryData2008}, but the underlying topology \gbase\ of a graphic may need more dimensions than the data topology \dbase\ so that the glyph can be defined in \dfiber.
-Therefore we define the base space mapping from graphic \gbase to data \dbase
+Since we propose that data topology \dbase\ is by some measure preserved in the graphic, we introduce a graphic topology \gbase to encode the carried over structure.
+These topologies are not always identical because the underlying topology \gbase\ of a graphic may need more dimensions than the data topology \dbase\.
+In a typical 2D display (ignoring depth), visible elements need to have 2D extent so that for example a scatter marker has an area greater than 0 or a line is not infinitely thin.
+
+Since \dbase\ with fewer than two dimensions need to be thickened, we define the mapping from graphic \gbase\ to data \dbase 
 \begin{equation}
@@
 
-In figure~\ref{fig:graphic_retraction_map} each disk $\gbase_{\dbasepoint}$ indexes how elements in \gfiber\ are glued together to generate a single glyph that is the visual representation of a single record in $\dfiber_{\dbasepoint}$.
-For the line, the region \gy\ over a point $\gx_i$ specifies the thickness of the line in \gbase\ for the corresponding \dsection\ on \dbase.
-The heatmap has the same continuity in data space and graphic space such that no extra dimensions are needed.
+When \dbase\ is discrete points and the graphic is a scatter plot, each point $\dbasepoint \in \dbase$ corresponds to a 2D disk $\gbase_{\dbasepoint}$ as shown in figure~\ref{fig:graphic_retraction_map}.
+In the case of 1D continuous data and a line plot, the region \gy\ over a point $\gx_i$ specifies the thickness of the line in \gbase\ for the corresponding \dsection\ on \dbasepoint.
+The heatmap has the same continuity in data space and graphic space such that no extra dimensions are needed in \gbase.
+The mapping function \vindex\ provides a way to identify the part of the visual transformation that is specific to the the connectivity of the data rather than the values; for example it is common to flip a matrix when displaying an image.
 
@@
 \label{sec:graphic_section}
-% couple nu + Q , visual encodings 
-% pick a pixel which is the bounding box
-A section $\gsection: \gbase\rightarrow \gtotal$ defines a piece of the graphical representation of the data.
-Evaluated on a single \gbasepoint\, \gsection\ returns a single element in \gfiber\.
-For a 2D screen, the pixel is defined as a region $p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ of the rendered graphic.
-Since the x and y in $p$ are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box $\gbase_{p} ={\gsection_{xy}}^{-1}(p)$ is a region $\gbase_p \subset \gbase$.
-Integrating over this region on \gbase\
-\begin{align}
-    r_p &= \iint\limits_{S_p} \rho_r(s)ds^{2}\\
-    g_p &= \iint\limits_{S_p} \rho_g(s)ds^{2}\\
-    b_p &= \iint\limits_{S_p} \rho_b(s)ds^{2}
-\end{align}
-yields the color of the pixel $p$.
-
 \begin{figure}
@@
 \end{figure}
-As shown in figure~\ref{fig:graphic_rho_lookup}, the output space queries into the graphic bundle to render the image.
-We select a pixel $p$ in the output space, inverse map the region of the pixel into $\gbase_p \subset \gbase$, then compute the section $\gsection$ over the region $\gbase_p$.
-The section yields the set of elements in \gfiber\ that specify the $(r, g, b)$ values corresponding to the region $p$.
+
+This section describes how we go from a graphical design to a rendered image, where the graphical design is the section $\gsection: \gbase \rightarrow \gtotal$.
+It is sufficient to sketch out how an arbitrary pixel would be renderer, where a pixel $p$ in a real display corresponds to a region $\gbase_p$ in the idealized display.
+To determine the color of the pixel, we aggregate the color values over the region via integration.
+
+For a 2D screen, the pixel is defined as a region $p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ of the rendered graphic.
+Since the x and y in $p$ are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box $\gbase_{p} ={\gsection_{xy}}^{-1}(p)$ is a region $\gbase_p \subset \gbase$.
+To compute the color, we integrate on $\gbase_p$
+
+\begin{align}
+    r_p &= \iint\limits_{S_p} \rho_r(s)ds^{2}\\
+    g_p &= \iint\limits_{S_p} \rho_g(s)ds^{2}\\
+    b_p &= \iint\limits_{S_p} \rho_b(s)ds^{2}
+\end{align}
+
+As shown in figure~\ref{fig:graphic_rho_lookup}, a pixel $p$ in the output space is selected and inverse mapped into the corresponding region $\gbase_p \subset \gbase$.
+This triggers a lookup of the $\gsection$ over the region $\gbase_p$, which yields the set of elements in \gfiber\ that specify the $(r, g, b)$ values corresponding to the region $p$.
 The color of the pixel is then obtained by taking the integral of $\gsection_{rgb}(\gbase_p)$.
-
-%% pdf/svg/etc are different ways of implementing rhos
+In general, \gsection\ is an abstraction of rendering specifications, for example PDF\cite{bienz1993portable}, SVG\cite{quintScalable2003}, or an openGL scene graph\cite{CarsonOpenGL1997} or rendering engines such as cairo\cite{CairographicsOrg} and AGG\cite{AntiGrainGeometry}.
+Implementation of \gsection\ is out of scope for this work.
 

### 2021-02-17 draft/sections/intro.tex
rewrite ot intro, tools

@@
 
+ This work is motivated by a need for a visualization library that developers could build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.
+The core architecture also needs to be robust to the big data needs of many visualization practitioners, and therefore support distributed and streaming data needs.
+To support both exploratory and confirmatory visualization\cite{tukeyWeNeedBoth1980}, this tool needs to support 2D and 3D,  static, dynamic and interactive visualizations.
+Specifically, this work was driven by a rearchiture the Python visualization library Matplotlib\cite{hunterMatplotlib2DGraphics2007} to meet modern data visualization needs.
+We aim to take advantage of developments in software design, data structures, and visualization to improve the consistency, composibility, and discoverability of the API.
+
  This work presents a mathematical description of how data is transformed into graphic representations, as shown in ref~\ref{fig:intro_artist_stages}.
@@
 \section{Background}
+Visual algorithms that display information are designed such that structure of data is assumed according to Tory and Möller  \cite{toryRethinkingVisualizationHighLevel2004}.
+Specifically they note that discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}.
+This idea can be seen in modern general purpose visualization architecture,  where many libraries are either defined around specific data structures or make implicit assumptions about the data or the types of visualizations they will support.
+In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.
+
+\subsection{Tools}
+\label{sec:intro_data_tools}
+The library driving this rearchitecture, Matplotlib, aims to be flexible enough that developers can build domain specific libraries on top of it.
+To preserve this flexibility, Matplotlib enforces minimal constraints on what sorts of data the user is allowed to input and supports very many visual algorithms, from primative marks to computationally complex visualizations.
+Instead of enforcing constraints at the API level, Matplotlib carries implicit assumptions about data continuity in how each function interfaces with the input data.
+This has lead to poor API consistency and brittle code as every visual algorithm has a very different point of view on how the data is structured.
+
+A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+This framework is very popular in the data visualization community their declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create.
+It is also the wrong abstraction layer since the grammer exposes specific ways to compose parts together to build a chart, but not ways to build new parts\cite{wongsuphasawatNavigatingWideWorld2021}.
+
+A different class of user facing tools are those that support images, such as ImageJ\cite{schneiderNIHImageImageJ2012}.
+These tools mostly have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized.
+These tools are ill suited for general purpose library developers as the architure is to build plugins into the existing system \cite{WritingPlugins} rather than domain specific tools that are built on top.
+Even the digital humanties oriented macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supoorts some non-image aggregate reporting charts, is still built around image data as the primary input.
+
+There are also visualization tools designed around scientifically complex data that support explicit description of the data, such as vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview}.
+These libraries, generally speaking, are architectured more as graphics libraries than visualization libraries, meaning they lack a clear distinction between the graphic construction and the rendering of that graphic.
+This is very close to the existing Matplotlib architecture, where the data, visual encoding, and rendering are jumbled in ways that make it hard to figure out what the code is doing.
+We are proposing a functional framework instead in large part to clearly encapsulate these seperate responsibilities of a visualization tool.
+In turn, this should make it easier to reuse components in the building of new tools.
+
+\subsection{Visualization}
+
 \begin{figure}[H]
@@
 \label{sec:intro_data}
-Tory and Möller propose that assumptions about the structure of data are built into the visual algorithms that display that information \cite{toryRethinkingVisualizationHighLevel2004}.
-Specifically they note that discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}.
 
@@
 
-\subsection{Tools}
-\label{sec:intro_data_tools}
+\subsection{contribution}
+The contribition of this work is
 
-A motivator for this work is that currently Matplotlib carries implicit assumptions about data continuity in how each function interfaces with the input data.
-This work proposes a unified internal representation that encodes connectivity in a common interface.
-Matplotlib aims to natively support data of varying connectivities, so tools primarily concerned with visualizing relational data of the type found in databases are insufficient models.
-Many of these tools are built on top of Wilkenson's Grammar of Graphics \cite{wilkinsonGrammarGraphics2005} which itself incorporates Mackinlay's  A Presentation Tool (APT) display algebra; GoG derivative include ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
-While many of these tools support images, the first class data container is a table like object of discrete records.
-Tools that primarily support images are also insufficient.
-For example ImageJ\cite{schneiderNIHImageImageJ2012} and the ImagePlot\cite{studiesCulturevisImageplot2021}macro have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized.
-Plugins exist, but must work around the everything is an image data model\cite{WritingPlugins}.
-There are also visualization tools that do not carry implicit assumptions about structure, for example vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011}.
-\note{Not totally positive but I think VTK is deeply coupled to the renderer and that's why we're not using it as a model, but not positive and this is literally what Marc worked on...}
-
-\subsection{contribution}
-The contribition of this work is a model we call the topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles.
-This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
-Furthermore, this model supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
-To demonstrate the practical value of the model, we built a prototype where we represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
+\begin{enumerate}
+    \item functional paradigm visualization tool architecture
+    \item translating dataset connectivity into graphical objects via topology-preserving maps
+    \item defining transforms from data into graphical components as monoidal action equivariant maps
+    \item a model of data and graphics as fiber bundles that decouples components from connectivity
+    \item an algebraic sum operation such that more complex visualizations can be built from simple ones 
+    \item prototype where we represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering
+\end{enumerate}
 \end{document}

### 2021-02-17 draft/sections/intro.tex
intro: data rewrite

@@
 
- This work presents a mathematical description of how data is transformed into graphic representations, as shown in ref~\ref{fig:intro_artist_stages}.
+ This work presents a mathematical description of how data is transformed into graphic representations, as shown in figure~\ref{fig:intro_artist_stages}.
 While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
@@
 These tools mostly have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized.
-These tools are ill suited for general purpose library developers as the architure is to build plugins into the existing system \cite{WritingPlugins} rather than domain specific tools that are built on top.
-Even the digital humanties oriented macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supoorts some non-image aggregate reporting charts, is still built around image data as the primary input.
+These tools are ill suited for general purpose library developers as the architecture is to build plugins into the existing system \cite{WritingPlugins} rather than domain specific tools that are built on top.
+Even the digital humanities oriented macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supports some non-image aggregate reporting charts, is still built around image data as the primary input.
 
@@
 This is very close to the existing Matplotlib architecture, where the data, visual encoding, and rendering are jumbled in ways that make it hard to figure out what the code is doing.
-We are proposing a functional framework instead in large part to clearly encapsulate these seperate responsibilities of a visualization tool.
+We are proposing a functional framework instead in large part to clearly encapsulate these separate responsibilities of a visualization tool.
 In turn, this should make it easier to reuse components in the building of new tools.
+
+\subsection{Data}
+\label{sec:intro_data}
+In this work, we extend Butler's topology driven representation of visualization data \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Butler proposes that fiber bundles are a good model for visualization data because it allows for encoding the continuity of the data separately from the types of variables in the data.
+Fiber bundles are flexible enough to support discrete and ND continuous datasets.
+Since Butler's model lacks a robust way of describing the variables, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
+
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
+    \caption{One way to describe data is by the connectivity of the points in the dataset.
+A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid.
+This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
+    \label{fig:intro_data_format}
+\end{figure}
+As shown in figure~\ref{fig:intro_data_format}, there are many types of connectivity.
+A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph.
+In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements.
+Each \textit{component} of the record is a single object, such as a temperature measurement, a color value, or an image.
+We also generalize \textit{component} to mean all objects in the dataset of a given type, such as all temperatures or colors or images.
+The way in which these records are connected is the \textit{connectivity}, \textit{continuity}, or more generally \textit{topology}.
+
+\begin{mdframed}[roundcorner=10pt, frametitle= definitions, frametitlerule=true, frametitlebackgroundcolor=gray!10]
+\begin{definition}
+    \item[records] points, observations, entries 
+    \item[components] variables, attributes, fields 
+    \item[connectivity] how the records are connected to each other
+\end{definition}
+\end{mdframed}
+
+Often this topology has metadata associated with it, describing for example when or where the measurement was taken.
+Building on the idea of metadata as \textit{keys} and their associated \textit{value} proposed by Munzner \cite{munznerChDataAbstraction}, we propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids.
+In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model allows for renaming all the metadata, for example changing the coordinate systems or time resolution, without imposing new semantics on the underlying structure.
 
@@
 
-\subsection{Data}
-\label{sec:intro_data}
-
-\begin{figure}[H]
-    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
-    \caption{One way to describe data is by the connectivity of the points in the dataset.
-A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid.
-This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
-    \label{fig:intro_data_format}
-\end{figure}
-As shown in figure~\ref{fig:intro_data_format}, there are many types of continuity in data.
-A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph.
-
-\begin{mdframed}[roundcorner=10pt, frametitle= definitions, frametitlerule=true, frametitlebackgroundcolor=gray!10]
- In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements.
-Each \textit{component} of the record is a single object, such as a temperature, a color, or image.
-We also generalize \textit{component} to mean all objects in the dataset of a given type, such as all temperatures or colors or images.
-The way in which these records are connected is the \textit{connectivity}, \textit{continuity}, or more generally \textit{topology}.
-\end{mdframed}
-
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/intro/munzner_datatypes}
-   \caption{Keys are unique lookup values used to find individual observations in the dataset.
-Keys are positional references, and can be coordinates on a map or unique values such as a primary key in a database or a (time, latitude, longitude) index in a data cube.
-Image modified from a diagram from Munzner's \textit{Visualization Analysis and Design} \cite{munznerVisualizationAnalysisDesign2014}}
-   \label{fig:intro_munzner_datatypes}
-\end{figure}
-
-Often this topology has metadata associated with it, as shown in figure~\ref{fig:intro_munzner_datatypes}.
-Munzner denotes this metadata as \textit{keys} that can be used to locate the record, and further to locate the specific \textit{value} \cite{munznerChDataAbstraction}.
-In figure~\ref{fig:intro_munzner_datatypes} tables have keys identifying the row and column and sometimes also the depth, while maps have a location key.
-We propose that information rich keys such as location are additional components of the record, and that instead there are coordinate free structural keys that identify the location of the record within a dataset of any continuity.
-
-In this work, we extend Butler's topology driven representation of visualization data \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-Butler proposes that fiber bundles are a good model for visualization data because it allows for encoding the connectivity seperately from the records and supports discrete and ND continuous datasets.
-Since Butler's model lacks a robust way of describing the components of the record, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like discription of the data in the fiber bundle.
-We then propose criteria on expressivity that take into account both the components and the continuity of the data.
-
 \subsection{contribution}

### 2021-02-17 draft/sections/intro.tex
rewrite ot intro, visual

@@
 
+\note{Add somehow: Software Design Patterns for Information Visualization (heer, agrawala)}
+
 \subsection{Data}
 \label{sec:intro_data}
-In this work, we extend Butler's topology driven representation of visualization data \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-Butler proposes that fiber bundles are a good model for visualization data because it allows for encoding the continuity of the data separately from the types of variables in the data.
-Fiber bundles are flexible enough to support discrete and ND continuous datasets.
+As mentioned, one of the drivers of this work was to facilitate building libraries that could natively support domain specific data containers.
+Fiber bundles are one such way model to model containers, as Butler proposes because they encode the continuity of the data separately from the types of variables and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
 Since Butler's model lacks a robust way of describing the variables, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
@@
 \begin{definition}
-    \item[records] points, observations, entries 
-    \item[components] variables, attributes, fields 
-    \item[connectivity] how the records are connected to each other
+    \item[\textbf{records}] points, observations, entries 
+    \item[\textbf{components}] variables, attributes, fields 
+    \item[\textbf{connectivity}] how the records are connected to each other
 \end{definition}
@@
 \subsection{Visualization}
+A visualization tool produces a graphical design and an image rendered based on that design, as described by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
+He defines the graphical design as the set of encoding relations from data to visual representation\cite{mackinlayAutomatingDesignGraphical1986}, and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.
+In addition to the graphic representations, Byrne et al.
+describe how visualizations have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tool's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
 
@@
 
-This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:intro_dubois}; to do so, we first discuss the criteria by which a visualization is evaluated.
-Byrne et al.
-propose that visualizations have graphic representations that are mappings from data to visuals and figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+We propose that the Du Bois in figure~\ref{fig:intro_dubois} are representative of the expressivity a core architecture should allow a downstream library to express.
+This is because while the Du Bois figures are not the common scatter, bar, or line plot \cite{friendlyBriefHistoryData2008}, they conform to the constraint that a visualization is a mapping from data to visuals.
 In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph.
-Tufte specifies that visual representations must be in proportion to the quantitative data being represented for a chart to be faithful and that there should be no extra information in the graphic or figurative elements of the graph, but otherwise his notion of graphic integrity is heavily context dependent\cite{tufteVisualDisplayQuantitative2001}.
-As is Norman's Naturalness Principal, which states that visualizations are more understandable  when the properties of the representation match the properties of the information being represented\cite{norman_things_smart}.
-Bertin takes it as a given that data properties match visual properties, so much so that Munzner argues it is inherently built into his classification system \cite{munznerVisualizationAnalysisDesign2014} which is displayed in figure~\ref{fig:intro_retinal_variables}.
+The visual representations in the Du Bois figures are in proportion to the quantitative data being represented, so the a chart is faithful according to Tufte's Lie Principal\cite{tufteVisualDisplayQuantitative2001}.
+The properties of the representation match the properties of the information being represented, so they are fairly understandable according to Norman's Naturalness Principal\cite{norman_things_smart}.
+As with the Du Bois data portraits, it is fundemental that any architecture tool we build ensures that the data properties match the visual properties.
 
@@
 
-As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design.
-He defines the graphical design as the set encoding relations from data to visual representation\cite{mackinlayAutomatingDesignGraphical1986}, and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.
-Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tools effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
-Mackinlay's definition of expressiveness is formalized at the visual encoding level, which as shown in figure~\ref{fig:intro_retinal_variables} refers to the components of a graphic such as the color or position of a glyph.
-Bertin first classified these graphic components as retinal variables and discussed which types of data they can express \cite{bertinIIPropertiesGraphic2011} and how they are composited on point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond.
+The visual properties were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who also described the types of data that paired with the visual properties shown in figure~\ref{fig:intro_retinal_variables}.
+It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
+These visual encodings are composited into the point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond.
 Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014}.
-
-Mackinlay's expressiveness criteria is well defined for the visual variables, such that he suggests the viability of a strict encoding relation that is a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
-We expand on this suggestion by proposing that monoid action equivariance is a strict condition of building valid encoders.
-Mackinlay does not provide a generalized criteria for plot types, instead embedding the requirements within the definition of the charts.
+As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools.
 
 \subsection{contribution}
-The contribition of this work is
+This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, this work contributes
 
 \begin{enumerate}
-    \item functional paradigm visualization tool architecture
-    \item translating dataset connectivity into graphical objects via topology-preserving maps
-    \item defining transforms from data into graphical components as monoidal action equivariant maps
-    \item a model of data and graphics as fiber bundles that decouples components from connectivity
-    \item an algebraic sum operation such that more complex visualizations can be built from simple ones 
-    \item prototype where we represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering
+    \item a functional oriented visualization tool architecture
+    \item topology-preserving maps from data to graphic 
+    \item monoidal action equivariant maps from component to visual variable 
+    \item algebraic sum such that more complex visualizations can be built from simple ones 
+    \item prototype built on Matplotlib's existing infrastructure for the rendering 
 \end{enumerate}

### 2021-02-17 draft/sections/intro.tex
invariance, quality metrics

@@
 As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools.
+By building visualization concepts into the core architecture, developers can incorporate assessments of the visualization,  such as as quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmann2014algebraic} of visualizations with respect to graphical encoding choices.
 
-\subsection{contribution}
+\subsection{Contribution}
+\note{should I mention the categorical framework here too?}
 This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, this work contributes
@@
     \item algebraic sum such that more complex visualizations can be built from simple ones 
-    \item prototype built on Matplotlib's existing infrastructure for the rendering 
+    \item prototype built on Matplotlib's infrastructure
 \end{enumerate}
+In contrast to the algebraic basis for visualization design proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic} or Mackinlay's APT framework, we are proposing a topological framework for building tools to build visualizations.
+We defer judgement of expressivity and effectiveness to developers building domain specific tools, but provide them the framework to do so.
 \end{document}

### 2021-02-17 draft/sections/math.tex
finished out Tom on data

@@
 that binds the components of the data in \dfiber\ to the continuity represented in \dbase.
-We use fiber bundles as the data model because they are mathematical structures that are flexible enough express all the types of data described in section~\ref{sec:intro_data}.
-The fiber bundle models
-\begin{description}
-\item[\dfiber] properties of data component types (\ref{sec:data_fiber})
-\item[\dbase] continuity of records (\ref{sec:data_base})
-\item[\dsection] collections of records (\ref{sec:data_section}). 
-\end{description}
-where \dtotal\ is the space of all possible datasets with these components and continuity.
-By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on $\dtotal$ and focus on the fiber and base spaces.
+The fiber bundle models the properties of data component types \dfiber\ (\ref{sec:data_fiber}), the continuity of records \dbase\ (\ref{sec:data_base}), the collections of records \dsection\ (\ref{sec:data_section}), and the space \dtotal\ of all possible datasets with these components and continuity.
+
+By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on \dtotal\ and focus on the components and continuity.
+We use fiber bundles as the data model because they are inclusive enough to express all the types of data described in section~\ref{sec:intro_data}.
 
@@
 
-For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree C temperature measurements taken at those times.
-Time is a positive number of type \texttt{datetime} which can be resolved to positive floats $\ftotal_{\texttt{datetime}}= \realsp$.
-Temperature values are real numbers $\ftotal_{\texttt{float}} = \reals$.
+For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree K temperature measurements taken at those times.
+Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$.
+Temperature values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$.
 The fiber is
 \begin{equation}
-    \ftotal =  \realsp \times \reals 
+    \ftotal =  \reals \times \realsp 
 \end{equation} 
-where the first component $F_0$ is the set of values specified by $(\fname_0=time,\, \ftype_0=\texttt{datetime},\, \fttype=\mathbb{R}^+)$ and $F_1$ is specified by $(\fname_1=temperature,\, \ftype_1=\texttt{float},\, \fttype=\reals)$.
+where the first component $F_0$ is the set of values specified by $(\fname_0=time,\, \ftype_0=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname_1=temperature,\, \ftype_1=\texttt{float},\, \fttype=\realsp)$ and is the set of values $\fttype=\realsp$.
 In figure~\ref{fig:fiber_example_cube}, temperature is replaced with wind.
@@
 \end{equation} 
-such that $F_1$ is specified by $(\fname_1 = wind,\, \ftype_1=\texttt{wind},\, \fttype=\reals^{2})$
+such that $F_1$ is specified by $(\fname_1 = wind,\, \ftype_1=\texttt{wind},\, \fttype=\reals^{2})$.
+As illustrated in figure~\ref{fig:data}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
 
@@
 Therefore order preserving maps do not form a group, and instead we generalize to monoids to support partial order component types.
+Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmann2014algebraic} that must be preserved in the visual representation of the component.
 
@@
 This is assumed in the choice of visualization, for example a line plot implies 1D continuous data and a scatter plot implies discrete records, but an explicit representation allows for verifying that the topology of the graphic representation is equivalent to the topology of the data.
-.
 
@@
 The same fiber in figure~\ref{fig:base_example_continuous} lies over a continuous interval \dbase such that the records are samples from a continuous function defined on \dbase.
+By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.
+This in turn allows for building algorithms that can work with distributed or streaming data, since it has a common data access interface with a promise that the data exists.
 
@@
 \label{sec:data_section}
-A section $\dsection: \dbase\rightarrow \dtotal$ of the fiber bundle is a set of records.
+A set of records is a section $\dsection: \dbase\rightarrow \dtotal$ of the fiber bundle.
 For example, in the special case of a table \cite{spivakSIMPLICIALDATABASES}, \dbase\ is a set of row ids, \dfiber\ is the columns, and the section \dsection\ returns the record at a given key in \dbase.
@@
 
-%% here are some of our real data structures and here is how it works: dataframe, movies, etc...
-%% all the same, special versions of fiber bundles
+The section can be any instance of a data container, for example a numpy array\cite{harris2020array}, a pandas series or dataframe\cite{jeff_reback_2020_3715232}, or an xarray\cite{hoyer2017xarray}.
+A univariate numpy array that stores an image is a section of a fiber bundle where \dbase\ is a 2D continuous plane and the \dfiber\ is $(\reals^{3}, \reals, \reals)$ where $\reals^3$ is color, and the other two components are the x and y positions of the sampled data in the image.
+A series could store the values of $\dsection_i$ and a second series could be  $dsection_j$.
+We could also fatten the fiber to hold two temperature series, such that a section would be an instance of a dataframe with a time column and two temperature columns.
+While the series and dataframe explicitly have a time index column, they are components in our model and the index is always assumed to be random keys.
+An instance of an xarray would also be a \dsection\, for example if the data is a temperature field than the \dbase\ could be a continuous volume and the components would be the temperature and the time, latitude, and longitude the measurements were sampled at.
+As with the dataframe, the semantic index labels are considered components and the indicies are instead assumed to be random.
+A section can also be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.
+As with the other containers, \dbase\ and \dfiber\ are defined in terms of the index and dtypes of the components of the array.
+The section can remain curried until render since the model is functional, in keeping with dasks compute graph architecture.
+Our model provides a common interface to these widely used data containers without sacrificing the semantic structure embedded in each container.
 
@@
 To support working with a subset of data, we can take the sheaf $\mathcal{O}(\dtotal)$ as input into the artist.
-This facilitates interactions such as zooming or updating the graphic based on zooming data.
 \dtotal\ restricted over a small enough neighborhood $U \subset \dbase$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
@@
 In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives.
+The sheaf facilitates interactions such as zooming or updating the graphic based on zooming data; the jet ensures that the artist only needs a single record on \dbasepoint\ to render a piece of a graphic.
+This allows for the creation of highly concurrent and possibly distributed artists.
 

### 2021-02-17 draft/sections/math.tex
up to Q hat

@@
 \label{sec:artist}
-%%somewhere in this I need to add q \circ nu (\xi^{-1}\tau(s)) = rho (s)
-%%% line things up such that they're an equivariant maps 
-%%% make a little clearer what E' is 
+
 The artist is the function that converts data into graphics; its name is taken from the analogous part of Matplotlib\cite{hunterArchitectureOpenSource} that builds visual elements to pass off to the renderer.
-The artist \vartist\ is a mapping from \dtotal\ padded with data from $\mathcal{J}(\dtotal)$ to a graphic that is a section \gsection\ in  $\Gamma(\gtotal)$
-%%% \Gamma is total
-%%% H is point wise -
-%%% this map means we can do this pointwise 
-%%% explain the arrows, chase through here
-%%% new disagram w/ section?
-% so you aren't passing around \tau
-% thos maps make sense for \tau(k)\in E -> V
-% and then you have \mu(k) 
-% and then when you are in \xi*V .... well the arrow goes the WONG way so you didn't just pass \mu(k) inot \xi*V
-% so if you want to go from \xi*V to H via Q you need to do it over a PARTICULAR s
-% thats what I mean by Q (in this diagram) hiding a choice of s. 
-% and then IF you pick an s ...... its is just fine. because it returns \rho(s)
-% so types match \rho(s) is in D
+The artist \vartist\ is a mapping from \dtotal\ padded with data from $\dtotal^{\prime}=\mathcal{J}(\dtotal)$ to a graphic that is a section $\gsection \in \Gamma(\gtotal)$
 
@@
 \end{equation}
-%%% Q \circ \xi  \nu = \rho, and nu is on tau, and sections are weaker 
-with an intermediate fiber bundle \vtotal\ to hold visual representations and stages
-\begin{enumerate}
-    \item \vindex\ binding the continuity in the graphic to the continuity in the data (~\ref{sec:graphic_base})
-    \item \vchannel\ conversion of data into visual characteristics (~\ref{sec:artist_nu})
-    \item \vmark\ assembly of visual variables into a glyph (~\ref{sec:artist_q})
-\end{enumerate}
 
-of the visual transformation illustrated in figure~\ref{fig:artist_stages}.
-The functions \vindex\, \vchannel\ and \vmark\ are defined such that they can be evaluated on a single section \dsection, which allows the artist to be implemented such that it does not need all the data.
-This allows for artists tuned to distributed and streaming data.
+which due to using the jet is point wise such that the input can be $\dsection(\dbasepoint)$.
+The encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ converts the data components to visual components(\ref{sec:graphic_base}).
+The continuity map $\vindex:\gbase \dbase$ then pulls back the visual bundle \votal\ over \gbase (\ref{sec:artist_nu}).
+Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal (\ref{sec:artist_q}).
+This functional decomposition of the visualization artists allows us to specify what are the responsibilities of each function in a fully constrained way.
+In turn, this allows for building reusable components at each stage of the transformation.
 
 \subsubsection {Visual Fiber Bundle \vtotal}
-%% structure preserving on the visual side/target space, visualization library 
-%% equivariant, latent space of visual parameters, intermediate 
-%% don't formally need, let it more cleanly pin down the the fiber equivarance , expresisng the concept that is already in literature,
-
-The visual fiber bundle (\vtotal, \dbase, $\pi$, \vfiber) has section $\vsection: \vtotal \rightarrow \dbase$ that resolves to a visual variable \cite{bertinIIPropertiesGraphic2011,munznerMarksChannels} in fiber \vfiber.
-The fiber space \vfiber\ is defined in terms of the parameters of the visualization specification- for example aesthetics  in ggplot \cite{wickhamGgplot2ElegantGraphics2016a}, channels in vega\cite{satyanarayanDeclarativeInteractionDesign2014} or parameters in VTK\cite{hanwellVisualizationToolkitVTK2015} and Matplotlib.
+We introduce a visual bundle \vtotal\ to store the visual representations the artist needs to composite into a graphic.
+The visual bundle (\vtotal, \dbase, $\pi$, \vfiber) has section $\vsection: \vtotal \rightarrow \dbase$ that resolves to a visual variable in the fiber \vfiber.
+The visual bundle \vtotal\ is the latent space of possible parameters of a visualization type, such as a scatter or line plot.
+We define \vfiber\ in terms of the parameters of a visualization libraries compositing functions; for example table~\ref{tab:mpl_visual_variable_fiber} is a sample of the fiber space for Matplotlib \cite{hunterMatplotlib2DGraphics2007}.
 
@@
 
-Table~\ref{tab:mpl_visual_variable_fiber} is a sample of the fiber space for Matplotlib \cite{hunterMatplotlib2DGraphics2007}.
  A section \vsection is a tuple of visual values that specifies the visual characteristics of a part of the graphic.
@@
 
-\subsubsection{Visual Channels \vchannel}
+\subsubsection{Visual Encoders \vchannel}
 \label{sec:artist_nu}
@@
 where $\vchannel_i: \dsection_i \mapsto \vsection_i$ is an equivariant map such that there is a monoid homomorphism from $\dfiber_{i}$ to $vfiber_{i}$.
+As mentioned in section~\ref{sec:data_monoid}, we choose monoid actions as the basis for equivariance because they define the structure on the fiber components.
 A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$
@@
 For ratio data, there must be equivalent scaling $ \vchannel(xc) = \vchannel(x)\vchannel(c)$.
+These constraints can be embedded into our artist such that the \vchannel\ functions are equivariant; they also provide guidance on constructing new equivariant \vchannel\ functions.
 
-\subsubsection{Assembling Glyphs \vmark}
+\subsubsection{Graphic Assembler \vmark}
 \label{sec:artist_q}
@@
     \caption{\vchannel\ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.
-\gsection\ applied to a region of connected components $\gbase_{\dbasepathpoint}$  generates a graphical mark.}
+\gsection\ applied to a region of connected components $\gbase_{\dbasepathpoint}$  generates a part of a graphic, for example the point graphical mark.}
     \label{fig:artist_q}
@@
 
-As shown in figure~\ref{fig:artist_q}, the assembly function \vmark\ combines the fiber $\dfiber_i$ wise \vchannel\ transforms into a single glyph.
-Together, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings, reduce the encodings into a glyph.
-As with \vchannel\, the constraint on \vmark\ is that for every monoid actions on the input \vsection\ there is a corresponding monoid action on the output \gsection.
+As shown in figure~\ref{fig:artist_q}, the assembly function \vmark\ combines the fiber $\dfiber_i$ wise \vchannel\ transforms into a graphic in \gtotal.
+Together, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings, reduce the encodings into a graphic.
+As with \vchannel\, the constraint on \vmark\ is that for every monoid action on the input \vsection\, there is a corresponding monoid action on the output \gsection.
 
@@
 
-To output a mark  \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology}, \vmark\ is called with all the regions \gbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$:
+We can describe a mark \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology} as \vmark\ with input of all the regions \gbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$:
 \begin{equation}
@@
 such that for every mark there is at least one corresponding section on \dbase.
+We define a glyph as a composite of multiple \vmark\, in keeping with Munzner's definition of the glyph as an object with internal structure arising from multiple marks\cite{munznerVisualizationAnalysisDesign2014}.
 
-\subsubsection{Sample Glyphs \vmark}
+\subsubsection{Sample Graphics \vmark}
 In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and a heatmap. 

### 2021-02-17 draft/sections/math.tex
closed out math section

@@
 The encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ converts the data components to visual components(\ref{sec:graphic_base}).
-The continuity map $\vindex:\gbase \dbase$ then pulls back the visual bundle \votal\ over \gbase (\ref{sec:artist_nu}).
+The continuity map $\vindex:\gbase \dbase$ then pulls back the visual bundle \vtotal\ over \gbase (\ref{sec:artist_nu}).
 Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal (\ref{sec:artist_q}).
@@
 
-\subsubsection{Sample Graphics \vmark}
+\subsubsection{Assembly \vmark}
 In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and a heatmap. 
@@
     \includegraphics[width=.75\textwidth]{figures/math/heatmap.png}
-    \caption{The only visual parameter a heatmap requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
+    \caption{The only visual parameter a image requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
     \label{fig:artist_heatmap}
@@
 
-The heatmap $\vmark(color)$ in figure~\ref{fig:artist_heatmap} is a direct lookup $\vindex:\gbase\rightarrow\dbase$ such that
+The image $\vmark(color)$ in figure~\ref{fig:artist_heatmap} is a direct lookup $\vindex:\gbase\rightarrow\dbase$ such that
 
@@
 \subsubsection{Assembly factory \vmarkd}
-\begin{figure}
+The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib, because the rendering is tightly interlaced with the graphical design; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
+As shown in eq~\ref{eq:artist_diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+\begin{figure}[H]
     \includegraphics[width=.75\textwidth]{figures/math/q_hat.png}
@@
 \end{figure}
-As shown in eq~\ref{eq:artist_diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
-Because $\vpreimg\ subset \gbase$ such that many \gbasepoint\ go to one \dbasepoint, by definition of the pull back $\vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber$.
-This means that the fiber \vtotal is copied over the preimage in \vpreimg.
-Given a section \vsectionpull\ pulled back from \vsection\ on $\pi:\vtotal\rightarrow\dbase$ and $\gbasepoint \in \vpreimg$, then the section $\vsectionpull(s) = \vindex^*(\vsection(\dbasepoint))$ is the image of $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$ for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
-This is illustrated in figure~\ref{fig:artist_q_hat}, where a \vsection\ over \dbasepath\ is copied over all  $\vindex(\gbasepoint) = \dbasepoint$.
 
-When \vmark\ maps sections $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull, then $\vmark(\vsectionpull) = \gsection$ depends on \gbasepoint.
-This means that $\gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)$, but since \vsectionpull\ is constant on \vpreimg, we can define a function $\vmarkd(\vsection(k))(s) \coloneqq \vmark((\vsectionpull)(\gbasepoint))$ where $\vpreimg = \dbasepoint$.
+The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back $\vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber$ copies the visual fiber \vtotal over the preimage in \vpreimg.
+This is illustrated in figure~\ref{fig:artist_q_hat}, where the 1D fiber over \dbase\ is copied repeatedly to become the 2D fiber with identical components over \gbase.
+Given a section \vsectionpull\ pulled back from \vsection\ on $\pi:\vtotal\rightarrow\dbase$ and a point $\gbasepoint \in \vpreimg$ in the preimage of \dbasepoint\, the pulled back section $\vsectionpull(s) = \vindex^*(\vsection(\dbasepoint))$ is the image $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$.
+This means that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$, which is illustrated in figure~\ref{fig:artist_q_hat} as each dot on \vfiber is equivalent to the line intersection \vfiberpull.
 
-In fact, \vmarkd\ is a map from visual to graphic $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
+Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out.
+When \vmark\ maps visual sections $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull then $\gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)$.
+Since  $\vsectionpull(s) = \vindex^*(\vsection(\dbasepoint))$ and \vsectionpull\ is constant on \vpreimg, we can define a \vmark\ factory function $\vmarkd(\vsection(k))(s) \coloneqq \vmark((\vsectionpull)(\gbasepoint))$ where $\vpreimg = \dbasepoint$.
+
+Factoring out \gbasepoint\, $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ generates a curried \vmark.
+In fact, \vmarkd\ is a map from visua space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
 This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
-The advantage of \vmarkd\ is that \gbase is part of the rendering stage of the pipeline and therefore hidden from the Matplotlib artists and therefore there is no direct way to construct a true \vmark.
+The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing rendering pipeline.
 
@@
 \end{figure}
-In this paper we define a simple addition operator that is the disjoint union of fiber bundles \dtotal.
-For example, in figure~\ref{fig:artist_plus} the scatter plot $\dtotal_1$ and the line plot $\dtotal_2$ have different \dbase that are mapped to seperate \gbase.
+To build graphics that are the composites of multiple artists, we define a simple addition operator that is the disjoint union of fiber bundles \dtotal.
+For example, in figure~\ref{fig:artist_plus} the scatter plot $\dtotal_1$ and the line plot $\dtotal_2$ have different \dbase that are mapped to separate \gbase.
 To fully display both graphics, the composite graphic $\vartist_1 + \vartist_2$ needs to include all records on both $\dbase_1$ and $\dbase_2$, which are the sections on the disjoint union $\dbase_1 \sqcup \dbase_2$.
@@
 \label{sec:artist_equivalance}
-As formulated above, every artist function \vartist\ has fixed \vchannel\ and \vmark\ which  generates a distinct graphic \gsection.
-It is impractical to implement an artist for every single graphic; instead we implement the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$.
+It is impractical to implement an artist for every single graphic; instead we implement the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$ which is essentially an artist where the \vchannel\ functions have not yet been input.
 Equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\, but act on different sections \vsection.
@@
 \end{figure}
-For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another, as are the rest of the graphs in figure~\ref{fig:artist_equivalance}.
+
+For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another, as shown in figure~\ref{fig:artist_equivalance}.
 These two artists are equivalent since their only difference is in the literal visual encodings (color, shape).
@@
 We also use this criteria to identify derivative types, for example the bubble chart\cite{tufteVisualDisplayQuantitative2001} is a type of scatter where by definition the glyph size is mapped from the data.
+The criteria for equivalence class membership serves as the basis for evaluating invariance\cite{kindlmann2014algebraic}.
 
@@
         vertex  & discrete   & \dsection(\dbasepoint)                  \\ \hline
-        edge    & 1D         & $\dsection(\dbasepoint,\, \alpha)$        \\ \hline
-        face    & 2D         & $\dsection(\dbasepoint,\, \alpha,\, \beta)$\\ \hline
+        edge    & 1D         & \dsection(\dbasepoint,\, \dx)        \\ \hline
+        face    & 2D         & \dsection(\dbasepoint,\, \dx,\, \dy)\\ \hline
         \end{tabular}
@@
 such that each section is bound to a simplex $\dbasepoint \in \dbase$.
-As shown in table~\ref{tab:triangulization}, in a 1D continuous spaces each \dsection\ lies distance $\alpha$ along edge \dbasepoint, while in a 2D continuous space each \dsection\ lies at coordinate $\alpha, \beta$ on the face \dbasepoint.
+As shown in table~\ref{tab:triangulization}, in a 1D continuous spaces each \dsection\ lies distance \dx\ along edge \dbasepoint, while in a 2D continuous space each \dsection\ lies at coordinate \dx,\dy on the face \dbasepoint.
 This is directly analogous to indexing to express connectivity in N-D arrays, while also natively supporting graphs and trees as they are simplacial complices of nodes and edges.
@@
 \end{figure}
-One way of encoding the torus in figure~\ref{fig:triangle_torus} while retaining the continuity of both cross sections $a, b$ is to unravel it into a simplacial complex of two triangles with labeled edges.
+An advantage of fiber bundles is the ability to encode data continuity that is more complex than points, a line or a plane.
+For example, the tori in figure~\ref{fig:triangle_torus} can be unraveled into a simplacial complex of two triangles with labeled edges.
 Transition functions $\delta$ are defined on the edges such that $a$ can be glued to $a^\prime$ and $b$ to $b^\prime$ to reconstruct the torus.
@@
 As with the torus, the transition functions must preserve monoid commutativity.
+While we generally do not deal with complex or non-trivial bundles, the ability to implement them native to the model allows for a great deal of expressivity in encoding data.
+This is especially important when developing a tool that needs to be general enough to support any number of very domain specific types of data.
 

### 2021-02-17 draft/sections/intro.tex
closed out math section
functional framework + rewrite of contribution section

@@
 A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
-This framework is very popular in the data visualization community their declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create.
+This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create.
 It is also the wrong abstraction layer since the grammer exposes specific ways to compose parts together to build a chart, but not ways to build new parts\cite{wongsuphasawatNavigatingWideWorld2021}.
@@
 \begin{mdframed}[roundcorner=10pt, frametitle= definitions, frametitlerule=true, frametitlebackgroundcolor=gray!10]
-\begin{definition}
+    \begin{description}
         \item[\textbf{records}] points, observations, entries 
@@
         \item[\textbf{connectivity}] how the records are connected to each other
-\end{definition}
+    \end{description}
 \end{mdframed}
@@
 It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
+A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}.
 These visual encodings are composited into the point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond.
@@
 \end{enumerate}
-In contrast to the algebraic basis for visualization design proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic} or Mackinlay's APT framework, we are proposing a topological framework for building tools to build visualizations.
+In contrast to mathematical models of visualization that aim to evaluate visualization design, we propose a topological framework for building tools to build visualizations.
 We defer judgement of expressivity and effectiveness to developers building domain specific tools, but provide them the framework to do so.

### 2021-02-18 draft/sections/intro.tex
reworking intro + math with Z&K glyphs

@@
 \section{Background}
-Visual algorithms that display information are designed such that structure of data is assumed according to Tory and Möller  \cite{toryRethinkingVisualizationHighLevel2004}.
+Visual algorithms that display information are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighLevel2004}.
 Specifically they note that discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}.
-This idea can be seen in modern general purpose visualization architecture,  where many libraries are either defined around specific data structures or make implicit assumptions about the data or the types of visualizations they will support.
+This assumption is evident in the information visualization software design patterns categorized by Heer and Agrawala\cite{HeerSoftware2006} and can often be found in modern general purpose visualization libraries.
 In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.
@@
 This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create.
-It is also the wrong abstraction layer since the grammer exposes specific ways to compose parts together to build a chart, but not ways to build new parts\cite{wongsuphasawatNavigatingWideWorld2021}.
+Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building the elements.
 
@@
 
-There are also visualization tools designed around scientifically complex data that support explicit description of the data, such as vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview}.
+There are also visualization tools designed around scientifically complex data that support explicit description of the data, such as vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}.
 These libraries, generally speaking, are architectured more as graphics libraries than visualization libraries, meaning they lack a clear distinction between the graphic construction and the rendering of that graphic.
@@
 In turn, this should make it easier to reuse components in the building of new tools.
-
-\note{Add somehow: Software Design Patterns for Information Visualization (heer, agrawala)}
 
@@
 
-We propose that the Du Bois in figure~\ref{fig:intro_dubois} are representative of the expressivity a core architecture should allow a downstream library to express.
+We propose that the Du Bois visualizations in figure~\ref{fig:intro_dubois} are representative of the expressivity a core architecture should allow a downstream library to express.
 This is because while the Du Bois figures are not the common scatter, bar, or line plot \cite{friendlyBriefHistoryData2008}, they conform to the constraint that a visualization is a mapping from data to visuals.
@@
 The properties of the representation match the properties of the information being represented, so they are fairly understandable according to Norman's Naturalness Principal\cite{norman_things_smart}.
-As with the Du Bois data portraits, it is fundemental that any architecture tool we build ensures that the data properties match the visual properties.
+As with the Du Bois data portraits, it is fundamental that any architecture tool we build ensures that the data properties match the visual properties.
 
@@
 A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}.
-These visual encodings are composited into the point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond.
-Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014}.
+Vickers et al.
+propose a category theory framework\cite{vickersUnderstandingViz2013} that extends the equivariance checks to layout, but is focused strictly on the design layer like the other mathematical frameworks.
+
+The visual encodings are composited into the point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables}.
+Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014} and minimally need to be differentiable from other visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}.
 As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools.
@@
 \subsection{Contribution}
-\note{should I mention the categorical framework here too?}
 This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, this work contributes

### 2021-02-18 draft/sections/math.tex
reworking intro + math with Z&K glyphs

@@
 
-We can describe a mark \cite{bertinIIPropertiesGraphic2011,carpendaleVisualRepresentationSemiology} as \vmark\ with input of all the regions \gbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$:
+ We can formally describe a glyph as \vmark\, with the regions \dbasepoint that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
 \begin{equation}
@@
 where the path\cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1].
-We define the mark as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$
-
+We define the glyph as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$
 \begin{equation}
@@
 \end{equation}
-such that for every mark there is at least one corresponding section on \dbase.
-We define a glyph as a composite of multiple \vmark\, in keeping with Munzner's definition of the glyph as an object with internal structure arising from multiple marks\cite{munznerVisualizationAnalysisDesign2014}.
+such that for every glyph there is at least one corresponding section on \dbase.
+This is in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.
 
@@
 
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
-    \caption{The torus \dtotal is unraveled into a simplacial complex of 2 faces \dbase. Transition functions are defined on the edges of \dbase such that surface can be glued back into the torus.
-    \note{add cross sections a and b to ring and color same as edges in complex}}
-    \label{fig:triangle_torus}
-\end{figure}
-An advantage of fiber bundles is the ability to encode data continuity that is more complex than points, a line or a plane.
-For example, the tori in figure~\ref{fig:triangle_torus} can be unraveled into a simplacial complex of two triangles with labeled edges.
-Transition functions $\delta$ are defined on the edges such that $a$ can be glued to $a^\prime$ and $b$ to $b^\prime$ to reconstruct the torus.
-This simplacial complex is then used as the base space encoding the continuity of data that lies in the torus.
-A constraint on the transition functions is that the monoid actions on the fibers on the edges of \dtotal\ are commutative $\monoid*\dfiber\mapsto \delta(\monoid\dfiber)= \monoid*\delta(\dfiber)$
-
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
-    \caption{Many non-trivial spaces can be made locally trivial by dividing \dtotal into locally trivial subspaces and defining transition functions between the edges on \dbase for how to glue the two subspaces such that the \dsection are continuous.}
-    \label{fig:data_base_transition}
-\end{figure}
-Another advantages of triangulization is that it provides a way to encode non-trivial structures such as the mobius strip\cite{MobiusStripNLab}.
-As shown in figure ~\ref{fig:data_base_transition}, one way of making the mobius strip trivial is to seperate it into two spaces $\dtotal_1$ and $\dtotal_2$ and then define transition functions that specify that the edges of $\dtotal_1$ need to be reversed to line up with $\dtotal_2$ such that the sections along the edges meet.
-As with the torus, the transition functions must preserve monoid commutativity.
-While we generally do not deal with complex or non-trivial bundles, the ability to implement them native to the model allows for a great deal of expressivity in encoding data.
-This is especially important when developing a tool that needs to be general enough to support any number of very domain specific types of data.
+In general simplacial complexes, in these prototypes we implement graphs.
 

### 2021-02-18 draft/sections/topodata.tex
factor out traingulation examples

%%pull out cause not helping
\begin{figure}[H]
    \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
    \caption{The torus \dtotal is unraveled into a simplacial complex of 2 faces \dbase. Transition functions are defined on the edges of \dbase such that surface can be glued back into the torus.
    \note{add cross sections a and b to ring and color same as edges in complex}}
    \label{fig:triangle_torus}
\end{figure}
An advantage of fiber bundles is the ability to encode data continuity that is more complex than points, a line or a plane. For example, the tori in figure~\ref{fig:triangle_torus} can be unraveled into a simplacial complex of two triangles with labeled edges. Transition functions $\delta$ are defined on the edges such that $a$ can be glued to $a^\prime$ and $b$ to $b^\prime$ to reconstruct the torus. This simplacial complex is then used as the base space encoding the continuity of data that lies in the torus. A constraint on the transition functions is that the monoid actions on the fibers on the edges of \dtotal\ are commutative $\monoid*\dfiber\mapsto \delta(\monoid\dfiber)= \monoid*\delta(\dfiber)$

\begin{figure}[H]
    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
    \caption{Many non-trivial spaces can be made locally trivial by dividing \dtotal into locally trivial subspaces and defining transition functions between the edges on \dbase for how to glue the two subspaces such that the \dsection are continuous.}
    \label{fig:data_base_transition}
\end{figure}
Another advantages of triangulization is that it provides a way to encode non-trivial structures such as the mobius strip\cite{MobiusStripNLab}. As shown in figure ~\ref{fig:data_base_transition}, one way of making the mobius strip trivial is to seperate it into two spaces $\dtotal_1$ and $\dtotal_2$ and then define transition functions that specify that the edges of $\dtotal_1$ need to be reversed to line up with $\dtotal_2$ such that the sections along the edges meet. As with the torus, the transition functions must preserve monoid commutativity. While we generally do not deal with complex or non-trivial bundles, the ability to implement them native to the model allows for a great deal of expressivity in encoding data. This is especially important when developing a tool that needs to be general enough to support any number of very domain specific types of data.

### 2021-02-18 draft/sections/next.tex
new conclusion, make whatever plot you want

@@
 
-This work contributes a mathematical description of the transformation from data to visual representation.
-Combining Butler's fiber bundle model of data with Spivaks formalism of data schemas provides a way of decoupling topology from variability such that the model can support a very large variety of datasets, including discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks.
-Modeling the graphic as a fiber bundle provides a way to separate the target display space from the topology of the graphic.
-By decomposing the mapping from data to visual representation as encoding \vchannel, assembly \vmark, and a mapping between data and graphic topologies \vindex and formalizing what equivariance each stage needs to preserved, this work derives constraints that visualization library authors could embed in their code to guarantee visualizations that are equivariant transforms of the input data.
+This work contributes a mathematical description of the mapping \vartist\ from data to visual representation.
+Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this mode; support a variety of datasets, including  discrete relational tables,, multivariate high resolution spatio temporal datasets, and complex networks.
+Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications for producing visualization where the structure and properties match those of the input data.
+These specifications are that the graphic must have continuity equivalent to the data, and that the visual characteristics of the graphics are equivariant to their corresponding components under monoid actions.
+This model defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic.
+Encoding the graphic space as a fiber bundle provides a structure rich abstraction of the target graphical design in the target display space.
 
-This work generalizes previous research constraining visual encodings from data components to graphic components as equivariant maps to components that are N-dimensional.
-Furthermore, it precisely defines the glyph as the visual element constructed from data on a simplex in a simplacial complex where the simplex is discrete or continuous.
-This is a restatement of for example Bertin's definition of a line that encapsulates all points on the continuous line.
-By modeling the data topology along with its variablity, this work also provides  a generalization of topology preservation as a deformation retraction from graphic space to data space.
-By using a functional paradigm, we can deconstruct the graphic to the glyph associated with each data point or even to pieces of a glyph; therefore the renderer has full  flexibility in how to to generate the image, while the graphic to data topology maps provide a way to keep track of which part of the image belongs to which data point.
-
-The toy prototype built using this model validates that is usable for a general purpose visualization tool since it can be iteratively integrated into the existing architecture rather than starting from scratch/ Factoring out glyph formation into assembly functions allows for much more clarity in how the glyphs differ.
-This prototype demonstrates that this framework can generate the fundemental marks, point (scatter plot), line (line chart), and area (bar chart).
+The toy prototype built using this model validates that is usable for a general purpose visualization tool since it can be iteratively integrated into the existing architecture rather than starting from scratch.
+Factoring out glyph formation into assembly functions allows for much more clarity in how the glyphs differ.
+This prototype demonstrates that this framework can generate the fundemental marks: point (scatter plot), line (line chart), and area (bar chart).
 Furthermore, the grouped and stacked bar examples demonstrate that this model supports composition of glyphs into more complex graphics.
@@
 While this model supports equivariance of figurative glyphs generated form parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it currently does not have a way to evaluate the semantic accuracy of the figurative representation.
-This model also does not currently factor in effectiveness, but potentially effectiveness criteria could be incorporated into a scheme for assigning encoders and assembling glyphs.
+Effectiveness is out of scope for this model because it is not part of the structure being preserved, but potentially a developer building a domain specific library with this model could implement effectiveness criteria in the artists.
 Also, even though the model is designed to be backend independent, it has only really been tested against the AGG backend.
@@
 \section{Conclusion}
-Despite the alternative formalism to Mackinlay's APT framework, it is indebted to APT for providing a guideline for issues the model needs to address.
-As with APT, this works aims to develop a framework on which to build visualization tools.
-Unlike APT, our model is geared towards balancing the needs of the the scientific and information visualization communities; therefore we present a framework that can describe both how the components of the data are encoded and how the continuity is preserved in the graphic.
-We hope that this framework can distill the constraints on the data to graphic mapping such that it can be used to improve the architecture of a heavily used visualization tool and allow developers to more easily build domain specific tools.
+An unoffical philosophy of Matplotlib is to support making whatever kinds of plots a user may want, even if they seem nonsensical to the development team.
+The topological framework described in this work provides a way to facilitate this graph creation in a rigorous manner; any artist that meets the equivariance criteria described in this work by definition generates a graphic representation that matches the structure of the data being represented.
+We leave it to domain specialists to define the structure they need to preserve and the maps they want to make, and hopefully make the process easier by untangling these components into seperate constrained maps and providing a fairly general data and display model.
 \end{document}

### 2021-02-19 draft/sections/implementation.tex
intro, circles many circles
tables/

@@
 \begin{document}
-%what, why, how
 \section{Prototype Implementation: Matplottoy}
@@
 \end{multicols*}
+We then add the \mintinline{python}{Point} and \mintinline{python}{Line} artists that construct the scatter and line graphics.
+The arguments to the artist are the data \mintinline{python}{data} that is to be plotted and the aesthetic configuration \mintinline{python}{transforms}.
+In general, the way in which we implement the math is described as follows:
+\begin{tabular}
+    \begin{table}{|l|l|l|}
+        \hline
+        \textbf{Math} & \texbf{Code} & \textbf{Examples} \\
+        \textbf{\vartistEQ} & & \\
 
-We then add the \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{python}{Line} artists that construct the scatter and line graphics.
-The arguments to the artist are the data \dtotal=\mintinline{python}{data} that is to be plotted and the aesthetic configuration \vchannel=\mintinline{python}{transforms}.
+    \end{table}
+
+\end{tabular}
+
+\begin{mdframed}[roundcorner=10pt, frametitle= topology driven architecture, frametitlerule=true, frametitlebackgroundcolor=gray!10]
+    \begin{description}
+        \item[\textbf{\vartistEQ}] artist objects such as  \mintinline{python}{Point},  \mintinline{python}{Line}, \mintinline{python}{bar}
+        \item[\textbf{\vchannel}] 
+        \item[\textbf{\markd}] 
+        \item[\textbf{\vtotal}] 
+        \item[\textbf{\dtotal}] 
+    \end{description}
+\end{mdframed}
+Implementing \gtotal\ and \gsection\ are out of scope for this prototype because they are part of the rendering process.
+We do not implement any form of \vindex\ because the scatter, line, and bar plots prototyped here directly broadcast from \dbasepoint to \gbasepoint, unlike for example an image which may need to be rotated.
 We implement the artists as equivalence classes \vartisteq\ because it would be impractical to implement a new artist for every aesthetic setting, such as one artist for red lines and another for green.
@@
         self.fiber = fiber
-
-    def is_section(self, section):
-        """checks if a section is from a given fiber bundle:
-        are values in F, are keys in K"""
 \end{minted}

### 2021-02-19 draft/sections/implementation.tex
start of summary section

--- 2021-02-04 draft/sections/implementation.tex
@@
 
-We then add the \vartisteq=\mintinline{python}{Point} and  \vartisteq=\mintinline{python}{Line} artists that construct the scatter and line graphics.
-The arguments to the artist are the data \dtotal=\mintinline{python}{data} that is to be plotted and the aesthetic configuration \vchannel=\mintinline{python}{transforms}.
-We implement the artists as equivalence classes \vartisteq\ because it would be impractical to implement a new artist for every aesthetic setting, such as one artist for red lines and another for green.
+We then add the \mintinline{python}{Point} and \mintinline{python}{Line} artist that construct the scatter and line graphics.
+These artists are implemented as the equivalence class \vartisteq\ with the aesthetic configurations factored out into a \mintinline{python}{transforms} dictionary that specifies the visual bundle \vtotal\.
+The equivalence classes \vartisteq\ map well to Python classes since the functional aspects-\vchannel, \vmarkd, and \vindex - are completely reusable in a consistent composition, while the visual values in \vtotal\ are what change between different artists belonging to the same class \vartisteq.
+The \mintinline{python}{data} object is an abstraction of a data bundle \dtotal\ with a specified section \tau.
+Implementing \gtotal\ and \gsection\ are out of scope for this prototype because they are part of the rendering process.
+We also did not implement any form of \vindex\ because the scatter, line, and bar plots prototyped here directly broadcast from \dbasepoint to \gbasepoint, unlike for example an image which may need to be rotated.
 
@@
         # visual channel encoding applied fiberwise 
-        visual = {p: encoder(view.get(f, None)) for 
+        visual = {p: encoder(view[f] if f is not None else None) for 
                      p, (f, encoder) in self.transforms.items()}
@@
 
+\subsection{Summary}
+In general, the way in which we implement the artist is as follows: 
+\begin{description}
+    \item[\textbf{\vchannel}] encoder functions, can be \mintinline{python}{Encoder.__call__} method
+    \item[\textbf{\vtotal}]  \mintinline{python}|{parameter:(variable, encoder)}| curried until {Artist.draw}
+    \item[\textbf{\vartisteq}] objects that take in as input \dtotal\ and \vtotal\, for example \mintinline{python}{Point}, \mintinline{python}{Line}, \mintinline{python}{bar}
+    \item[\textbf{\vartist}] \mintinline{python}{Artist.draw}, which is \vartisteq\ parameterized by \vtotal
+
+    \item[\textbf{\vmarkd}] \mintinline{python}{Artist.assemble(visual)}  
+\end{description}
+
+with the data representation implemented as 
+\begin{description}
+    \item[\textbf{\dtotal}]
+    \item[\textbf{\dfiber}]
+    \item[\textbf{\d}]  
+\end{description}
+
 \note{This API may want to be redesigned such that there's a way to clearly couple the columns when doing multindex broadcasting}

### 2021-02-19 draft/sections/math.tex
rework next + math on xi and hook

@@
 In figure~\ref{fig:base_example_discrete} the fibers lie over discrete \dbase\ such that the records in the datasets in the fiber bundles are discrete.
-The same fiber in figure~\ref{fig:base_example_continuous} lies over a continuous interval \dbase such that the records are samples from a continuous function defined on \dbase.
+The same fiber in figure~\ref{fig:base_example_continuous} lies over a continuous interval \dbase\ such that the records are samples from a continuous function defined on \dbase.
 By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.
-This in turn allows for building algorithms that can work with distributed or streaming data, since it has a common data access interface with a promise that the data exists.
+This in turn allows for building visualizations that can work with distributed or streaming data, since it has a common data access interface with a promise that the data exists.
+The \vindex\ mapping is also used by interactive visualization components to look up the data associated with a region on screen.
+One example is to fill in details in a hover tooltip, another is to convert region selection (such as zooming) on \gbase\ to a query on the data to access the corresponding record components on \dbase.
 
@@
 Constraints can be placed on the disjoint union such as that the fiber components need to have the same \vchannel position encodings or that the position \vsection need to be in a specified range.
-There are situations where $\dbase_2 \hookrightarrow \dbase_1$ that underpin more complex, especially interactive, visualizations; these cases require defining a more complex addition operator that is out of scope for this work.
+There is a second type of composition where $\dtotal_1$ and $\dtotal_2$ share a base space $\dbase_2 \hookrightarrow \dbase_1$ such that the the artists can be considered to be acting on different components of the same section.
+This type of composition is important for creating visualizations where elements need to update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.
 

### 2021-02-19 draft/sections/next.tex
rework next + math on xi and hook

@@
 A second important proof of concept artist is a boxplot\cite{wickham40YearsBoxplots2011} because it is a graphic that assumes computation on the data side and the glyph is built from semantically defined components and a list of outliers.
-
 The model supports simple composition of glyphs by overlaying glyphs at the same position, but more work is needed to define an operator where the fiber bundles have shared $\gbase_2 \hookrightarrow \gbase_1$ such that fibers could be pulled back over the subset.
-This complex operator is necessary for building semantically meaningful components such as interactive visualizations that update on shared \gbase, such as brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991} and verifying constraints in multiple view systems \cite{hullmanKeeping2018}.
-The models simple addition supports axes as standalone artists with overlapping visual position encoding, but the complex operator would allow for keeping labels consistent with text when for example the horizontal and vertical positions of the data are changed.
+While the model's simple addition supports axes as standalone artists with overlapping visual position encoding, the complex operator would allow for binding together data that needs to be updated together.
 In summary, the proposed scope of work for the dissertation is
@@
     \item expansion of the mathematical framework to include complex addition
+    \item implementation of artist with explicit \vindex\
+    \item specification of interactive visualization
     \item mathematical formulation of a graphic with axes labeling
-    \item implementation of data with 2D continuous topology and a compatible artist
     \item provisional mathematics and implementation of user level composite artists
@@
 \end{itemize}
-Additionally, implementing the complex addition operator would allow for developing a mathematical formalism and prototype of how interactivity would work in this model.
+Additionally, implementing the complex addition operator and explicit graphic to data maps would allow for developing a mathematical formalism and prototype of how interactivity would work in this model.
 Other potential tasks for future work is implementing a data object for a non-trivial fiber bundle and exploiting the models section level formalism to build distributed data source models and concurrent artists.

### 2021-02-19 draft/sections/implementation.tex
rework of bar to incorporate make bar and simplify vert construction  + reordering around it

@@
     \end{subfigure}
-    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
+    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering. \note{update with bar}}
     \label{fig:code_scatter_line}
@@
 
-\begin{minted}{python}
+\begin{minted}[escapeinsize=||]{python}
 class ArtistClass(matplotlib.artist.Artist):
+    |\label{code:artist}|
     def __init__(self, data, transforms, *args, **kwargs):
@@
         # returns K, indexed on fiber then key 
-        view = self.data.view() 
+        # is passed the 
+        view = self.data.view(self.axes) 
         # visual channel encoding applied fiberwise 
-        visual = {p: encoder(view[f] if f is not None else None) for 
-                     p, (f, encoder) in self.transforms.items()}
+        visual = {p: encoder(view[f] if f is not None else None) 
+                     for p, (f, encoder) in self.transforms.items()}
         self.assemble(visual)
@@
 The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
-The return view object has a \mintinline{python}{get} method to support querying for components that are not in \dfiber\, which we exploit to support parameters in the visual fiber that are not bound to fiber components in \dfiber.
-The \vchannel\ functions are then applied to the data to generate the \vsection=\mintinline{python}{visual} input to \vmark.
-An explicit \vindex\ is not implemented since that would mean copying a single \vsection on \dbasepoint to all the associated \gbasepoint, as illustrated in figure~\ref{fig:graphic_retraction_map}, and that is unnecessary overhead for these scatter and line plots.
-In \vmarkd=\mintinline{python}{assemble} the artist generates instructions for the render by setting  the attributes that are related to the graphic.
-These are the settings that would have to be serialized in order to recreate a static version of the graphic.
+The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that we can use to query back into data to select a subset as discussed in section~\ref{sec:data_sheaf_stalk}.
+The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}.
+We allow for fixed visual parameter, such as setting a constant color for all sections, via setting in \mintinline{python}{None} as the data fiber \dfiber\ name in the \mintinline{python}{transform} dictionary.
+
+The visual object is then passed into the \mintinline{python}{assembly} function that is \vmarkd.
+This assembly function is responsible for generating a representation of the glyph such that it could be serialized to recreate a static version of the graphic.
 Although \mintinline{python}{assemble} could be implemented outside the class such that it returns an object the artist could then parse to set attributes, the attributes are directly set here to reduce indirection.
-The \vchannel\ functions could be evaluated in this function to avoid passing over \dbase twice but are not done so here to demonstrate the seperability of \vchannel\ and \vmarkd\.
+This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd.
 The last step in the artist function is handing itself off to the renderer.
+The extra \mintinline{python}{*arg, **kwargs} arguments in \mintinline{python}{__init__,draw} are artifacts of how these objects are currently implemnted in Matplotlib.
 
@@
 In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker.
+We only show the \mintinline{python}{assembly} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist shown in \ref{code:artist}.
 \begin{minted}{python}
-class Point(mcollections.Collection):
-    def __init__(self, data, transforms, *args, **kwargs):
-        super().__init__(*args, **kwargs)
-        self.data = data
-        self.transforms = transforms
-
+class Point(mcollections.Collection):=
     def assemble(self, visual):
@@
 
-    def draw(self, renderer, *args, **kwargs):
-        # query data for a vertex table K
-        view = self.data.view() 
-        visual = {p: encoder(view.get(f, None)) for 
-                     p, (f, encoder) in self.transforms.items()}
-        self.assemble(visual)
-        # call the renderer that will draw based on properties
-        super().draw(renderer, *args, **kwargs)
+\end{minted} 
+The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices, as described in section~\ref{sec:triangulization}.
+Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$.
+This means that all the data on one vertex maps to one glyph.
+To ensure the integrity of the section, \mintinline{python}{view} must be atomic.
+This means that the values cannot change after the method is called in draw until a new call in draw.
+We put this constraint on the return of the \mintedinline{python}{view} method so that we do not risk race conditions.
+Using triangulization provides a common interface to the data, one we will reuse for all the artists presented here.
 
-\end{minted} 
-The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices, as described in section~\ref{sec:triangulization}; even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$ such that all the data on one vertex maps to one marker.
-To ensure the integrity of the section, \mintinline{python}{view} must be atomic, meaning that the values cannot change after the method is called in draw until a new call in draw.
-This table is converted to a table of visual variables.
-It is then passed into \mintinline{python}{assemble}, where it is used to individually construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
-Since \mintinline{python}{view} returns a \dsection\, all these operations could be applied on a section on one \dbasepoint\ or a subset of \dbase.
+This table is converted to a table of visual variables and is then passed into \mintinline{python}{assemble}.
+In  \mintinline{python}{assemble}, the \vsection\ is used to individually construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
+This is done via the \mintinline{python}{Path.circle} object.
+As mentioned in sections ~\ref{sec:data_sheaf_stalk} and \ref{sec:artist_q}, this assembly function could as easily be implemented such that it was fed one $\dsection(\dbasepoint)$ at a time.
 
-The only difference between the \mintinline{python}{Point} and \mintinline{python}{Line} objects is in the \mintinline{python}{view} and \mintinline{python}{assemble} function because line has different continuity from scatter and is represented by a different type of graphical mark.
+The main difference between the \mintinline{python}{Point} and \mintinline{python}{Line} objects is in th3 \mintinline{python}{assemble} function because line has different continuity from scatter and is represented by a different type of graphical mark.
 
@@
         self.set_color(visual['color'])
+\end{minted}
+In the \mintinline{python}{Line} artist, \mintinline{python}{view} returns a table of edges.
+Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge.
+As with \mintinline{python}{Point}, the data is then converted into visual variables.
+In \mintinline{python}{assemble}, this visual representation is composed into a set of line segments, where each segement is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
+Then the colors of each line segment are set.
+The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on view.
+
+\begin{figure}[H]
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
+    \end{subfigure}
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
+    \end{subfigure}
+    \caption{Frequency of Penguin types visualized as discrete bars. }
+    \label{fig:code_bar_simple}
+\end{figure}
+The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist.
+The have the same required \vfiber\ components of (position, length).
+In \mintinline{pyton}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally and only applies holistically to the graphic and never to individual data parameters.
+This highlights another advantage of this model, that it encourages explicit differentiation between parameters in \vtotal\ and graphic parameters that specify \vmarkd\.
+
+\begin{minted}{python}
+class Bar(mcollections.Collection):
+    def __init__(self, data, transforms, orientation, *args, **kwargs):
+        # parameter of the graphic
+        self.orientation = orientation 
+        super().__init__(*args, **kwargs)
+        self.data = data
+        self.transforms = transforms
+
+    def assemble(self, visual):
+        #set some defaults
+        visual.setdefault('width', itertools.repeat(0.8))
+        visual.setdefault('floor', itertools.repeat(0))
+        visual.setdefault('facecolors', 'C0')
+        #build bar glyphs based on graphic parameter
+        if self.orientation in {'vertical', 'v'}:
+            order = ['position', 'width', 'floor', 'length']
+        elif self.orientation in {'horizontal', 'h'}:
+            order = ['floor', 'length', 'position', 'width']
+
+        verts = [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
+                for (x, xo, y, yo) in zip(*[visual[k] for k in order])]
+
+        self._paths = [mpath.Path(xy, closed=True) for xy in verts]
+        self.set_edgecolors('k')
+        self.set_facecolors(visual['facecolors'])
 
     def draw(self, renderer, *args, **kwargs):
-        # query data source for edge table
-        view = self.data.view()
-        visual = {p: encoder(view.get(f, None)) for 
-                     p, (f, encoder) in self.transforms.items()}
+        view = self.data.view(self.axes)
+        visual = utils.convert_transforms(view, self.transforms)  
         self.assemble(visual)
@@
 \end{minted}
-In the \mintinline{python}{Line} artist, \mintinline{python}{view} returns a table of edges.
-Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge.
-As with \mintinline{python}{Point}, the data is then converted into visual variables.
-In \mintinline{python}{assemble}, this visual representation is composed into a set of line segments and then the colors of each line segment are set.
-The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on view.
+The \mintinline{python}{draw} method is the same as the one in the artist example~\ref{code:artist}, but here the visual transformations are factored out into a separate function.
+This is because in bar we support a rudimentary version of the complex \plus operator such that we can specify multiple components to map to the same visual encoding without requiring th components to be individually specified in the \mintinline{python}{transforms} dictionary.
+The implementation of this support is slightly more complicated than the simple case, but not in a way that is useful for understanding the framework.
+The \mintinline{python}{assemble} function constructs bars and sets their edge color to black.
+Defaults are provided for 'width' and 'floor' to make this function more reusable.
+Typically the defaults are used for the type of chart shown in figure~\ref{fig:code_bar_simple}, but these visual variables are often set when building composite versions of this chart type as discussed in section~\ref{sec:code_case_study}.
 
@@
 More complex encoders can be implemented as callable classes, such as
+
 \begin{minted}{python}
@@
 Creating a callable class also provides a simple way to swap out the specific (data, value) mapping without having to reimplement the validation or conversion logic.
-
 A test for equivariance can be implemented trivially such that it is independent of data or encoder.
@@
 \label{sec:code_data}
-The data input into the \mintinline{pythin}{Artist} will often be a wrapper class around an existing data structure, but must meet the following criteria:
-\begin{enumerate}
-\item specify the fiber components \dfiber\ and connectivity \dbase\
-\item have a \mintinline{view} method that returns an atomic object that encapsulates \dsection\
-\item the view object must have \mintinline{get} method that returns a fiber component
-\end{enumerate}
-To support specifying the fiber bundle, we define an optional \mintinline{python}{FiberBundle} class
+The data input into the \mintinline{pythin}{Artist} will often be a wrapper class around an existing data structure.
+This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{view} method that returns an atomic object that encapsulates \dsection\.
+The object returned by the view must be key valued pairs of \mintinline{python}|{component name : component section}| where each section is a component as defined in equation~\ref{eq:data_section_decompose}.
+To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataclassesDataClasses}
 
 \begin{minted}{python}
+@dataclass
 class FiberBundle:
-    def __init__(self, base, fiber):
 """
-        base:  {'tables': ['vertex', 'edge', 'face']}
-        fiber: {'component name': {'type':, 'monoid':,  'range':}}
+Attributes
+----------
+K: {'tables': []}
+F: {variable name: {'type': type, 'moniod', 'range': []}}
 """
-        self.base = base
-        self.fiber = fiber
+    K: dict 
+    F: dict
 \end{minted}
@@
 that asks the user to specify how \dbase is triangulated and the attributes of \dfiber.
-The \mintinline{python}{assembly} functions expect tables that match the continuity of the graphic; scatter expects a vertex table because it is discontinuous, line expects an edge table because it is 1D continuous.
+Python dataclasses are a good abstraction for the fiber bundle class because the \mintinline{python}{FiberBundle} class only stores data.
+The \dbase\ is specified as tables because the \mintinline{python}{assembly} functions expect tables that match the continuity of the graphic; scatter expects a vertex table because it is discontinuous, line expects an edge table because it is 1D continuous.
 The fiber informs appropriate choice of \vchannel\, therefore it is a dictionary of attributes of the fiber components.
-\note{I've basically stripped this out of the artists above so should I just ditch this section?}
 
@@
 
-    def view(self):
+    def view(self, axes):
         table = defaultdict(list)
@@
 
-    def view(self, simplex):
+    def view(self, axes):
         table = defaultdict(list)
@@
 
-    def tau(self, k, simplex='edge'):
+    def tau(self, k):
         # evaluates functions defined in edge table
@@
 
-    def view(self, simplex='edge'):
+    def view(self, axes):
         """walk the edge_vertex table to return the edge function
@@
         self._view = dataframe
-    def view(self):
+    def view(self, axes):
         return self._view
@@
 The pandas indexer is a key valued set of discrete vertices, so there is no need to repackage for triangulation.
-As with the previous examples, there is no need to implement an explicit \mintinline{python}{get} method since the \mintinline{python}{dataframe} object has a get method.
-
-\begin{figure}[H]
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
-    \end{subfigure}
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
-    \end{subfigure}
-    \caption{Frequency of Penguin types visualized as discrete bars. }
-    \label{fig:code_bar_simple}
-\end{figure}
-The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist.
-The have the same required \vfiber\ components of (position, length).
-In \mintinline{pyton}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which only applies holistically to the graphic and never to individual data parameters.
-Explicitly differentiate between parameters in \vtotal\ and ones that are only in \vmarkd\ is another way this model allows for cleaner separation of roles in the code.
-
-\begin{minted}{python}
-class Bar(mcollections.Collection):
-    def __init__(self, data, transforms, *args, **kwargs):
-        # parameter of the graphic
-        self.orientation = kwargs.pop('orientation', 'v')
-
-        super().__init__(*args, **kwargs)
-        self.data = data
-        self.transforms = transforms
-
-    @staticmethod    
-    def _make_bars(orientation, position, width, floor, length):
-        if orientation in {'vertical', 'v'}:
-            xval, xoff, yval, yoff = position, width, floor, length
-        elif orientation in {'horizontal', 'h'}:
-            xval, xoff, yval, yoff = floor, length, position, width
-        return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
-                for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
-
-    def assemble(self, visual):
-        #set some defaults
-        visual['width'] = visual.get('width', itertools.repeat(0.8))
-        visual['floor'] = visual.get('floor', itertools.repeat(0))
-        visual['facecolors'] = visual.get('facecolors', 'C0')
-        #build bar glyphs based on graphic parameter
-        verts = self._make_bars(self.orientation, visual['position'], 
-                   visual['width'], visual['floor'], visual['length'])
-        self._paths = [mpath.Path(xy, closed=True) for xy in verts]
-        self.set_edgecolors('k')
-        self.set_facecolors(visual['facecolors'])
-
-    def draw(self, renderer, *args, **kwargs):
-        view = self.data.view()
-        visual = utils.convert_transforms(view, self.transforms)  
-        self.assemble(visual)
-        super().draw(renderer, *args, **kwargs)
-        return      
-\end{minted}
-The \mintinline{python}{draw} method identical to the ones above, but here the visual transformations are factored out into a separate function.
-The \mintinline{python}{assemble} function sets some defaults, constructs bars, and sets their edge color to black.
-The \mintinline{python}{_make_bars} function is somewhat factored out because this is an operation that may be used by other bar making functions that may not be able to make use of bars assemble or draw.
 
@@
 \subsection{Summary}
-In general, the way in which we implement the artist is as follows: 
+In general, the way in which we implemented the artist is as follows: 
 \begin{description}
-    \item[\textbf{\vchannel}] encoder functions, can be \mintinline{python}{Encoder.__call__} method
-    \item[\textbf{\vtotal}]  \mintinline{python}|{parameter:(variable, encoder)}| curried until {Artist.draw}
+    \item[\textbf{\vchannel}] encoder function converting data to library normalized form, can also be \mintinline{python}{Encoder.__call__} method
+    \item[\textbf{\vsection}]  dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| curried until {Artist.draw}
     \item[\textbf{\vartisteq}] objects that take in as input \dtotal\ and \vtotal\, for example \mintinline{python}{Point}, \mintinline{python}{Line}, \mintinline{python}{bar}
-    \item[\textbf{\vartist}] \mintinline{python}{Artist.draw}, which is \vartisteq\ parameterized by \vtotal
-
-    \item[\textbf{\vmarkd}] \mintinline{python}{Artist.assemble(visual)}  
+    \item[\textbf{\vartist}] \mintinline{python}{Artist.draw} method which is \vartisteq\ parameterized by \vtotal
+    \item[\textbf{\vmarkd}] \mintinline{python}{Artist.assemble(visual)} method that arranges components of \vtotal\ into glyphs  
 \end{description}
@@
 \begin{description}
-    \item[\textbf{\dtotal}]
-    \item[\textbf{\dfiber}]
-    \item[\textbf{\d}]  
+    \item[\textbf{\dtotal}] object with \mintinline{python}{view} method 
+    \item[\textbf{\dfiber}] named components of the data and their types
+    \item[\textbf{\dbase}] determines how the data is bound together when returned by view  
+    \item[\textbf{\dsection}] \mintinline{python}{Data.view()} method that returns component sections
 \end{description}
 
-\note{This API may want to be redesigned such that there's a way to clearly couple the columns when doing multindex broadcasting}
+While very rough, this API demonstrates that the ideas presented in the math framework are implementable.
+In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\ and assembly \vmarkd\.
+We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic.
 \end{document}

### 2021-02-19 code/matplottoy/artists/bar.py

@@ -11,17 +11,13 @@ import matplotlib.path as mpath
 
 from matplottoy.artists import utils
 
-
 class Bar(mcollections.Collection):
-    def __init__(self, data, transforms, *args, **kwargs):
+    def __init__(self, data, transforms, orientation, *args, **kwargs):
         """
         Parameters
         ----------
         datasource: matplottoy.datasources.DataSource like
             data object containing data to be plotted
-        stacked: bool, optional
-            True: bar chart is stacked (groups are added)
-            False: bar chart is unstacked (groups are interspersed)
         orientation: str, optional
             vertical: bars aligned along x axis, heights on y
             horizontal: bars aligned along y axis, heights on x
@@ -33,40 +29,36 @@ class Bar(mcollections.Collection):
         #utils.check_constraints(Bar, transforms)
         #utils.validate_transforms(data, transforms)
 
-        self.orientation = kwargs.pop('orientation', 'v')
+        self.orientation = orientation
      
         super().__init__(*args, **kwargs)
         self.data = data
         self.transforms = transforms
-
-    @staticmethod    
-    def _make_bars(orientation, position, width, floor, length):
-        if orientation in {'vertical', 'v'}:
-            xval, xoff, yval, yoff = position, width, floor, length
-        elif orientation in {'horizontal', 'h'}:
-            xval, xoff, yval, yoff = floor, length, position, width
-        return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
-                for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
-
     
-    def assemble(self, visual): 
-        visual['width'] = visual.get('width', itertools.repeat(0.8))
-        visual['floor'] = visual.get('floor', itertools.repeat(0))
-        visual['facecolors'] = visual.get('facecolors', 'C0')
-     
-        verts = self._make_bars(self.orientation, visual['position'], 
-                   visual['width'], visual['floor'], visual['length'])
+    def assemble(self, visual):
+        #set some defaults
+        visual.setdefault('width', itertools.repeat(0.8))
+        visual.setdefault('floor', itertools.repeat(0))
+        visual.setdefault('facecolors', 'C0')
+        #build bar glyphs based on graphic parameter
+        if self.orientation in {'vertical', 'v'}:
+            order = ['position', 'width', 'floor', 'length']
+        elif self.orientation in {'horizontal', 'h'}:
[…]

### 2021-02-19 code/matplottoy/artists/point.py

@@ -20,15 +20,13 @@ class Point(mcollections.Collection):
         data: sections of the fiber bundle
         transforms
         """
-        super().__init__(*args, **kwargs)
-        
         # check that the data you're trying to transform 
         # has a way to provide vertex data  
         # assert 'vertex' in data.FB.K['tables']
         # check that you've given the required parameters
         #utils.check_constraints(Point, transforms.keys())
         #utils.validate_transforms(data.FB.F, transforms)
-
+        super().__init__(*args, **kwargs)
         self.data = data
         self.transforms = transforms
 
@@ -42,8 +40,8 @@ class Point(mcollections.Collection):
         
     def draw(self, renderer, *args, **kwargs):
         # query data for a vertex table K
-        view = self.data.view('vertex') 
-        visual = {p: encoder(view.get(f, None)) for p, (f, encoder) in self.transforms.items()}
+        view = self.data.view(self.axes) 
+        visual = {p: encoder(view[f] if f is not None else None) for p, (f, encoder) in self.transforms.items()}
         self.assemble(visual)
         # call the renderer that will draw based on properties
         super().draw(renderer, *args, **kwargs)

### 2021-02-19 code/matplottoy/data/iris.py

@@ -6,7 +6,7 @@ class DataFrame:
         self._tau = dataframe.iloc
         self._view = dataframe
 
-    def view(self, simplex="vertex"):
+    def view(self, axes):
         return self._view
     
     

### 2021-02-19 code/matplottoy/data/simplex.py

@@ -1,31 +1,29 @@
 from collections import defaultdict
+from dataclasses import dataclass
 import itertools
-
 import numpy as np
-
 from matplottoy.encoders import mtypes
  
- 
+@dataclass
 class FiberBundle:
-    def __init__(self, K, F):
-        """
-        Parameters
-        ----------
-        K: dict
-            attributes of the base space of the fiber bundle, 
-            describes domain of base space + the type of simplexs
-            on the base space
-            {'tables': []}
+    """
+    Attributes
+    ----------
+    K: dict
+        attributes of the base space of the fiber bundle, 
+        describes domain of base space + the type of simplexs
+        on the base space
+        {'tables': []}
 
-        F: dict
-            the fiber, which is the space of all possible observations 
-            type is used here to describe the space 
-            {variable name: type}
+    F: dict
+        the fiber, which is the space of all possible observations 
+        type is used here to describe the space 
+        {variable name: type}
 
-            # rework as ('field name': (type, monoid action))
-        """
-        self.K = K 
-        self.F = F
+    # rework as ('field name': (type, monoid action))
+    """
+    K:dict 
+    F:dict
         # should maybe be folded into K
         # add check for consistency
 

### 2021-02-19 draft/sections/implementation.tex
stacked bar

@@
 class Bar(mcollections.Collection):
-    def __init__(self, data, transforms, orientation, *args, **kwargs):
+    def __init__(self, data, transforms, orientation = 'v', *args, **kwargs):
         # parameter of the graphic
@@
 
-    def draw(self, renderer, *args, **kwargs):
-        view = self.data.view(self.axes)
-        visual = utils.convert_transforms(view, self.transforms)  
-        self.assemble(visual)
-        super().draw(renderer, *args, **kwargs)
-
 \end{minted}
-The \mintinline{python}{draw} method is the same as the one in the artist example~\ref{code:artist}, but here the visual transformations are factored out into a separate function.
-This is because in bar we support a rudimentary version of the complex \plus operator such that we can specify multiple components to map to the same visual encoding without requiring th components to be individually specified in the \mintinline{python}{transforms} dictionary.
+The \mintinline{python}{draw} method is the same as the one in the artist example~\ref{code:artist} so is ommitted here.
+This bar underpins a rudimentary version of the complex \plus operator such that we can specify multiple components to map to the same visual components without requiring th components to be individually specified in the \mintinline{python}{transforms} dictionary.
 The implementation of this support is slightly more complicated than the simple case, but not in a way that is useful for understanding the framework.
@@
 Creating a callable class also provides a simple way to swap out the specific (data, value) mapping without having to reimplement the validation or conversion logic.
-A test for equivariance can be implemented trivially such that it is independent of data or encoder.
+A test for equivariance can be implemented trivially
 \begin{minted}{python}
@@
 \end{minted}
+but is currently factored out of the artist for clarity.
 In this example, \mintinline{python}{is_nominal} checks for equivariance of permutation group actions by applying the encoder to a set of values, shuffling values, and checking that (value, encoding) pairs remain the same.

### 2021-02-19 code/matplottoy/artists/bar.py

@@ -12,7 +12,7 @@ import matplotlib.path as mpath
 from matplottoy.artists import utils
 
 class Bar(mcollections.Collection):
-    def __init__(self, data, transforms, orientation, *args, **kwargs):
+    def __init__(self, data, transforms, orientation='v', *args, **kwargs):
         """
         Parameters
         ----------
@@ -55,11 +55,66 @@ class Bar(mcollections.Collection):
         
     def draw(self, renderer, *args, **kwargs):
         view = self.data.view(self.axes)
-        visual = utils.convert_transforms(view, self.transforms)  
+        visual = {p: encoder(view[f] if f is not None else None) for p, (f, encoder) in self.transforms.items()}
         self.assemble(visual)
         super().draw(renderer, *args, **kwargs)
         return
 
+class StackedBar(mcollections.Collection):
+    def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
+        """
+        Parameters
+        ----------
+   
+        orientation: str, optional
+            vertical: bars aligned along x axis, heights on y
+            horizontal: bars aligned along y axis, heights on x
+        **kwargs:
+            kwargs passed through         """
+  
+        #assert 'vertex' in data.FB.K['tables']
+        #utils.check_constraints(MultiBar, transforms)
+        #utils.validate_transforms(data, transforms)
+       
+        super().__init__(*args, **kwargs)
+        self.data = data
+        self.orientation = orientation
+        self.transforms = transforms.copy()
+        self.mtransforms = mtransforms.copy()
+        assert ('length' in mtransforms) # this is the criteria for stacked bar
+
+    def assemble(self, view):
+        ngroups = len(self.mtransforms['length'][0])
+        gtransforms = self.transforms.copy()
+        gtransforms.setdefault('floor', 0)
+        for gid in range(ngroups):    
+            # pull out the specific group transforms
+            gtransforms.update({k: (group[gid], encoder) for 
+                    (k, (group, encoder)) in self.mtransforms.items()})
+            Bar(self.data, gtransforms, self.orientation)
+            
+            #gtransforms['floor'] += view[gtransforms['length'][0]]
+        """
+        # convert lengths after all calculations are made
+        # here or in transform machinery?
+        if self.orientation in {'v', 'vertical'}:
+            tverts = [[(x, gencoder(y)) for (x, y) in vert] 
+                            for vert in verts]
[…]

### 2021-02-19 code/matplottoy/data/iris.py

@@ -6,7 +6,7 @@ class DataFrame:
         self._tau = dataframe.iloc
         self._view = dataframe
 
-    def view(self, axes):
+    def view(self, axes=None):
         return self._view
     
     

### 2021-02-19 draft/sections/implementation.tex
stripping out draw non used args + adding axes argument to view

@@
 
-    def draw(self, renderer, *args, **kwargs):
+    def draw(self, renderer):
         # returns K, indexed on fiber then key 
@@
         # pass configurations off to the renderer
-        super().draw(renderer, *args, **kwargs)
+        super().draw(renderer)
 \end{minted}

### 2021-02-19 code/matplottoy/artists/bar.py

@@ -8,6 +8,7 @@ from cycler import cycler
 from matplotlib import rcParams
 import matplotlib.collections as mcollections
 import matplotlib.path as mpath
+from matplotlib.artist import Artist
 
 from matplottoy.artists import utils
 
@@ -53,14 +54,14 @@ class Bar(mcollections.Collection):
         self.set_edgecolors('k')
         self.set_facecolors(visual['facecolors'])
         
-    def draw(self, renderer, *args, **kwargs):
+    def draw(self, renderer):
         view = self.data.view(self.axes)
         visual = {p: encoder(view[f] if f is not None else None) for p, (f, encoder) in self.transforms.items()}
         self.assemble(visual)
-        super().draw(renderer, *args, **kwargs)
-        return
+        super().draw(renderer)
 
-class StackedBar(mcollections.Collection):
+
+class StackedBar(Artist):
     def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
         """
         Parameters

### 2021-02-19 code/matplottoy/artists/line.py

@@ -35,11 +35,11 @@ class Line(mcollections.LineCollection):
         self.set_segments(segments)
         self.set_color(visual['color'])
         
-    def draw(self, renderer, *args, **kwargs):
+    def draw(self, renderer):
         # query data source for edge table
         view = self.data.view('edge')
         visual = {p: encoder(view.get(f, None)) for 
                      p, (f, encoder) in self.transforms.items()}
         self.assemble(visual)
-        super().draw(renderer, *args, **kwargs)
+        super().draw(renderer)
 

### 2021-02-19 code/matplottoy/artists/point.py

@@ -38,10 +38,10 @@ class Point(mcollections.Collection):
         # circles and facecolors are lists of the same size
         self.set_facecolors(visual['facecolors'])
         
-    def draw(self, renderer, *args, **kwargs):
+    def draw(self, renderer):
         # query data for a vertex table K
         view = self.data.view(self.axes) 
         visual = {p: encoder(view[f] if f is not None else None) for p, (f, encoder) in self.transforms.items()}
         self.assemble(visual)
         # call the renderer that will draw based on properties
-        super().draw(renderer, *args, **kwargs)
+        super().draw(renderer)

### 2021-02-19 code/matplottoy/data/simplex.py

@@ -67,7 +67,7 @@ class VertexSimplex: #maybe change name to something else
         e4 = rng.uniform(*self.FB.F['v4']['range'])
         return (k, (e1, e2, e3, e4))
 
-    def view(self, simplex="vertex"):
+    def view(self, axes=None):
         """"converts data into atomic column order for get method
         # maybe also pass in the columns?
         """
@@ -101,7 +101,7 @@ class EdgeSimplex: # ToDo: generalize to take list of functions as input
     def __init__(self, num_edges=4, num_samples=1000):
         """which simplex am I on and distance""" 
         # define the k and distance 
-        self.keys = range(num_)
+        self.keys = range(num_edges)
         self.num_samples = num_samples
         self.distances = np.linspace(0,1, num_samples)
         # concession to this is a half generlized representation of 
@@ -126,7 +126,7 @@ class EdgeSimplex: # ToDo: generalize to take list of functions as input
         return np.cos(angles), np.sin(angles)
 
         
-    def tau(self, k, simplex='edge'):
+    def tau(self, k):
         """arbitrary k, even between vertices, will give back
         an observation, so all functions must target back to vertice 
         table
@@ -137,20 +137,14 @@ class EdgeSimplex: # ToDo: generalize to take list of functions as input
         #distances along edge are stored as self.distances
         #self.angle_samples
         x, y = self._xy(k, self.distances, self.angle_samples[k], self.angle_samples[k+1]) 
-        color = self._color(k) if simplex=='edge' else np.repeat(self._color(k), self.num_samples) 
-        return (k, (x, y, color))
+        return (k, (x, y, self._color(k)))
 
-    def view(self, simplex):
+    def view(self, axes=None):
         table = defaultdict(list)
         for k in self.keys:
             table['index'].append(k)
-            for (name, value) in zip(self.FB.F.keys(), self.tau(k, simplex)[1]):
+            for (name, value) in zip(self.FB.F.keys(), self.tau(k)[-1]):
                 table[name].append(value)
-        
-        if simplex =='vertex':
-            table['index'] = [(e,d) for d in self.distances for e in table['index']]
-            for name in self.FB.F.keys():
-                table[name] = list(itertools.chain.from_iterable(table[name]))
         return table
 
 def is_continuous(vertices):
@@ -184,11 +178,11 @@ class GraphLine:
         self.edges = edge_table
         self.connect = connect
 
-    def tau(self, k, simplex='edge'):
+    def tau(self, k):
[…]

### 2021-02-19 code/matplottoy/artists/bar.py
paths not getting set

@@ -8,7 +8,8 @@ from cycler import cycler
 from matplotlib import rcParams
 import matplotlib.collections as mcollections
 import matplotlib.path as mpath
-from matplotlib.artist import Artist
+import matplotlib.patches as mpatches
+import matplotlib.artist as martist
 
 from matplottoy.artists import utils
 
@@ -56,12 +57,12 @@ class Bar(mcollections.Collection):
         
     def draw(self, renderer):
         view = self.data.view(self.axes)
-        visual = {p: encoder(view[f] if f is not None else None) for p, (f, encoder) in self.transforms.items()}
+        visual = {p: encoder(view[f] if f is not None else None) 
+                  for p, (f, encoder) in self.transforms.items()}
         self.assemble(visual)
         super().draw(renderer)
 
-
-class StackedBar(Artist):
+class StackedBar(mcollections.Collection):
     def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
         """
         Parameters
@@ -85,36 +86,29 @@ class StackedBar(Artist):
         assert ('length' in mtransforms) # this is the criteria for stacked bar
 
     def assemble(self, view):
+        self.children = [] # list of bars to be rendered
         ngroups = len(self.mtransforms['length'][0])
         gtransforms = self.transforms.copy()
-        gtransforms.setdefault('floor', 0)
+        floor = 0
         for gid in range(ngroups):    
             # pull out the specific group transforms
             gtransforms.update({k: (group[gid], encoder) for 
                     (k, (group, encoder)) in self.mtransforms.items()})
-            Bar(self.data, gtransforms, self.orientation)
+            self.children.append(Bar(self.data, gtransforms, self.orientation))
+            print(gtransforms)
+            break
             
-            #gtransforms['floor'] += view[gtransforms['length'][0]]
-        """
-        # convert lengths after all calculations are made
-        # here or in transform machinery?
-        if self.orientation in {'v', 'vertical'}:
-            tverts = [[(x, gencoder(y)) for (x, y) in vert] 
-                            for vert in verts]
-        elif self.orientation in {'h', 'horizontal'}:
-            tverts = [[(gencoder(x), y) for (x, y) in vert] 
-                            for vert in verts]
-        self._paths = [mpath.Path(xy, closed=True) for xy in tverts]
-
-        self.set_facecolor(list(itertools.chain.from_iterable(visual['facecolors'])))
-        self.set_edgecolors('k') 
-        """
[…]

### 2021-02-21 code/matplottoy/artists/bar.py
fighting to debug drawing bugs

@@ -50,7 +50,6 @@ class Bar(mcollections.Collection):
         
         verts = [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
                 for (x, xo, y, yo) in zip(*[visual[k] for k in order])]
-
         self._paths = [mpath.Path(xy, closed=True) for xy in verts]
         self.set_edgecolors('k')
         self.set_facecolors(visual['facecolors'])
@@ -62,7 +61,7 @@ class Bar(mcollections.Collection):
         self.assemble(visual)
         super().draw(renderer)
 
-class StackedBar(mcollections.Collection):
+class StackedBar(martist.Artist):
     def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
         """
         Parameters
@@ -72,11 +71,9 @@ class StackedBar(mcollections.Collection):
             vertical: bars aligned along x axis, heights on y
             horizontal: bars aligned along y axis, heights on x
         **kwargs:
-            kwargs passed through         """
-  
-        #assert 'vertex' in data.FB.K['tables']
-        #utils.check_constraints(MultiBar, transforms)
-        #utils.validate_transforms(data, transforms)
+            kwargs passed through         
+        """
+
        
         super().__init__(*args, **kwargs)
         self.data = data
@@ -92,20 +89,19 @@ class StackedBar(mcollections.Collection):
         floor = 0
         for gid in range(ngroups):    
             # pull out the specific group transforms
+            gtransforms['floor'] = (None, lambda _ : floor)
             gtransforms.update({k: (group[gid], encoder) for 
                     (k, (group, encoder)) in self.mtransforms.items()})
-            self.children.append(Bar(self.data, gtransforms, self.orientation))
-            print(gtransforms)
-            break
-            
-        self._path = list(itertools.chain.from_iterable(c._paths for c in self.children))
-        
+            bar = Bar(self.data, gtransforms, self.orientation)
+            self.children.append(bar)
+            (key, encoder) = self.mtransforms['length']
+            floor += encoder(view[key])
+      
+    
     def draw(self, renderer):
         view = self.data.view(self.axes)
         # all the visual conversion gets pushed to child artists
-        self.assemble(view)
-        super().draw(renderer)
-        
+        children = self.assemble(view)
         for artist in self.children:
             artist.draw(renderer)
 

### 2021-02-21 code/matplottoy/artists/bar.py
probably gonna have to replace dict w/ object - using last value of floor on draw

@@ -35,7 +35,7 @@ class Bar(mcollections.Collection):
      
         super().__init__(*args, **kwargs)
         self.data = data
-        self.transforms = transforms
+        self.transforms = transforms.copy()
     
     def assemble(self, visual):
         #set some defaults
@@ -56,8 +56,15 @@ class Bar(mcollections.Collection):
         
     def draw(self, renderer):
         view = self.data.view(self.axes)
-        visual = {p: encoder(view[f] if f is not None else None) 
-                  for p, (f, encoder) in self.transforms.items()}
+        visual = {}
+        for (p, trans) in self.transforms.items():
+            if 'encoder' in trans:
+                visual[p] = trans['encoder'](view[trans['name']])
+            elif 'name' in trans:
+                visual[p] = view[trans['name']]
+            else:
+                visual[p] = trans                              
+                                                
         self.assemble(visual)
         super().draw(renderer)
 
@@ -80,29 +87,26 @@ class StackedBar(martist.Artist):
         self.orientation = orientation
         self.transforms = transforms.copy()
         self.mtransforms = mtransforms.copy()
-        assert ('length' in mtransforms) # this is the criteria for stacked bar
 
     def assemble(self, view):
         self.children = [] # list of bars to be rendered
-        ngroups = len(self.mtransforms['length'][0])
-        gtransforms = self.transforms.copy()
-        floor = 0
-        for gid in range(ngroups):    
+        floor = np.zeros(len(view[self.transforms['position']['name']]))      
+        for group in self.mtransforms:
             # pull out the specific group transforms
-            gtransforms['floor'] = (None, lambda _ : floor)
-            gtransforms.update({k: (group[gid], encoder) for 
-                    (k, (group, encoder)) in self.mtransforms.items()})
+            gtransforms = self.transforms.copy()
+            gtransforms.update(group)
+            gtransforms['floor'] = floor
             bar = Bar(self.data, gtransforms, self.orientation)
             self.children.append(bar)
-            (key, encoder) = self.mtransforms['length']
-            floor += encoder(view[key])
-      
-    
+            floor += view[group['length']['name']]
+            
+            
     def draw(self, renderer):
         view = self.data.view(self.axes)
         # all the visual conversion gets pushed to child artists
-        children = self.assemble(view)
+        self.assemble(view)
         for artist in self.children:
+            print("DRAW")
             artist.draw(renderer)
 
         

### 2021-02-22 code/matplottoy/artists/bar.py
switched assemble to explicit args to maybe force variable binding

@@ -37,22 +37,23 @@ class Bar(mcollections.Collection):
         self.data = data
         self.transforms = transforms.copy()
     
-    def assemble(self, visual):
+    def assemble(self, position, length, floor=0, width=0.8, facecolors='C0', edgecolors='k'):
         #set some defaults
-        visual.setdefault('width', itertools.repeat(0.8))
-        visual.setdefault('floor', itertools.repeat(0))
-        visual.setdefault('facecolors', 'C0')
+        width = itertools.repeat(width) if np.isscalar(width) else width
+        floor = itertools.repeat(floor) if np.isscalar(floor) else (floor)
+        
+        def make_bars(xval, xoff, yval, yoff):
+             return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
+                for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
         #build bar glyphs based on graphic parameter
         if self.orientation in {'vertical', 'v'}:
-            order = ['position', 'width', 'floor', 'length']
+            verts = make_bars(position, width, floor, length)
         elif self.orientation in {'horizontal', 'h'}:
-            order = ['floor', 'length', 'position', 'width']
+            verts = make_bars(floor, length, position, width)
         
-        verts = [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
-                for (x, xo, y, yo) in zip(*[visual[k] for k in order])]
         self._paths = [mpath.Path(xy, closed=True) for xy in verts]
-        self.set_edgecolors('k')
-        self.set_facecolors(visual['facecolors'])
+        self.set_edgecolors(edgecolors)
+        self.set_facecolors(facecolors)
         
     def draw(self, renderer):
         view = self.data.view(self.axes)
@@ -65,7 +66,7 @@ class Bar(mcollections.Collection):
             else:
                 visual[p] = trans                              
                                                 
-        self.assemble(visual)
+        self.assemble(**visual)
         super().draw(renderer)
 
 class StackedBar(martist.Artist):

### 2021-02-22 code/matplottoy/artists/bar.py
floor works now

@@ -1,3 +1,4 @@
+import copy
 from collections import defaultdict
 import itertools
 
@@ -12,6 +13,8 @@ import matplotlib.patches as mpatches
 import matplotlib.artist as martist
 
 from matplottoy.artists import utils
+
+
 
 class Bar(mcollections.Collection):
     def __init__(self, data, transforms, orientation='v', *args, **kwargs):
@@ -35,7 +38,7 @@ class Bar(mcollections.Collection):
      
         super().__init__(*args, **kwargs)
         self.data = data
-        self.transforms = transforms.copy()
+        self.transforms = copy.deepcopy(transforms)
     
     def assemble(self, position, length, floor=0, width=0.8, facecolors='C0', edgecolors='k'):
         #set some defaults
@@ -59,13 +62,13 @@ class Bar(mcollections.Collection):
         view = self.data.view(self.axes)
         visual = {}
         for (p, trans) in self.transforms.items():
-            if 'encoder' in trans:
-                visual[p] = trans['encoder'](view[trans['name']])
-            elif 'name' in trans:
-                visual[p] = view[trans['name']]
+            if isinstance(trans, dict):
+                if "name" in trans:
+                    visual[p] =  view[trans['name']]
+                if "encoder" in trans:
+                    visual[p] = trans['encoder'](visual[p])
             else:
-                visual[p] = trans                              
-                                                
+                 visual[p] = trans
         self.assemble(**visual)
         super().draw(renderer)
 
@@ -86,18 +89,17 @@ class StackedBar(martist.Artist):
         super().__init__(*args, **kwargs)
         self.data = data
         self.orientation = orientation
-        self.transforms = transforms.copy()
-        self.mtransforms = mtransforms.copy()
+        self.transforms = copy.deepcopy(transforms)
+        self.mtransforms = copy.deepcopy(mtransforms)
 
     def assemble(self, view):
         self.children = [] # list of bars to be rendered
-        floor = np.zeros(len(view[self.transforms['position']['name']]))      
+        floor = 0
         for group in self.mtransforms:
             # pull out the specific group transforms
-            gtransforms = self.transforms.copy()
-            gtransforms.update(group)
-            gtransforms['floor'] = floor
-            bar = Bar(self.data, gtransforms, self.orientation)
+            group['floor'] = floor
+            group.update(self.transforms)
+            bar = Bar(self.data, group, self.orientation)
             self.children.append(bar)
             floor += view[group['length']['name']]
             

### 2021-02-22 code/matplottoy/artists/bar.py
grouped bar paths correct, transforms a mess

@@ -40,10 +40,13 @@ class Bar(mcollections.Collection):
         self.data = data
         self.transforms = copy.deepcopy(transforms)
     
-    def assemble(self, position, length, floor=0, width=0.8, facecolors='C0', edgecolors='k'):
+    def assemble(self, position, length, floor=0, width=0.8, facecolors='C0', edgecolors='k', offset=0):
         #set some defaults
         width = itertools.repeat(width) if np.isscalar(width) else width
         floor = itertools.repeat(floor) if np.isscalar(floor) else (floor)
+        
+        # offset is passed through via assemblers such as multigroup, not supposed to be directly tagged to position 
+        position = position + offset
         
         def make_bars(xval, xoff, yval, yoff):
              return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
@@ -58,7 +61,7 @@ class Bar(mcollections.Collection):
         self.set_edgecolors(edgecolors)
         self.set_facecolors(facecolors)
         
-    def draw(self, renderer):
+    def draw(self, renderer,  *args, **kwargs):
         view = self.data.view(self.axes)
         visual = {}
         for (p, trans) in self.transforms.items():
@@ -70,7 +73,7 @@ class Bar(mcollections.Collection):
             else:
                  visual[p] = trans
         self.assemble(**visual)
-        super().draw(renderer)
+        super().draw(renderer,  *args, **kwargs)
 
 class StackedBar(martist.Artist):
     def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
@@ -80,12 +83,8 @@ class StackedBar(martist.Artist):
    
         orientation: str, optional
             vertical: bars aligned along x axis, heights on y
-            horizontal: bars aligned along y axis, heights on x
-        **kwargs:
-            kwargs passed through         
+            horizontal: bars aligned along y axis, heights on x   
         """
-
-       
         super().__init__(*args, **kwargs)
         self.data = data
         self.orientation = orientation
@@ -95,6 +94,7 @@ class StackedBar(martist.Artist):
     def assemble(self, view):
         self.children = [] # list of bars to be rendered
         floor = 0
+  
         for group in self.mtransforms:
             # pull out the specific group transforms
             group['floor'] = floor
@@ -104,89 +104,49 @@ class StackedBar(martist.Artist):
             floor += view[group['length']['name']]
             
             
-    def draw(self, renderer):
+    def draw(self, renderer, *args, **kwargs):
         view = self.data.view(self.axes)
[…]

### 2021-02-22 draft/sections/math.tex
rework of bar to incorporate make bar and simplify vert construction  + reordering around it
remove line to remove indent

@@
 We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
-To explain which structure the artist is preserving, we first describe how we model data(\ref{sec:data}), graphics(\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles.
+To explain which structure the artist is preserving, we first describe how we model data (\ref{sec:data}), graphics (\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles.
 We then discuss the equivariant maps between data and visual characteristics (\ref{sec:artist_nu}) and visual characteristics and graphics (\ref{sec:artist_q}) that make up the artist.
-\note{xi should maybe be moved down to artist but is more readable in graphics}
 
@@
 \end{equation}
-
 that binds the components of the data in \dfiber\ to the continuity represented in \dbase.
@@
 Because we can decompose the bundle and the fiber, we can formulate $\dsection$ as
-\begin{align}
+\begin{equation}
+\label{eq:data_section_decompose}
 \dsection= (\dsection_0,\ldots, \dsection_i, \dots, \dsection_n) 
-\end{align}
+\end{equation}
 where each section $\dsection_i$ is a variable or set of variables.

### 2021-02-22 draft/sections/math.tex
fiber break down

@@
 To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names and types.
-Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_n\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
+Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_{j_0}, \ldots, \ftype_{j_m}\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
 He then defines the single variable set \fttype\
@@
 \end{equation}
-which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. Given \fsection, the fiber for a one variable dataset is
+which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname.
+The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype.
+Given \fsection, the fiber for a one variable dataset is
 \begin{equation}
@@
 \end{equation}
-
 which allows us to decouple \dfiber\ into components $\dfiber_i$.
@@
 \label{sec:data_monoid}
-Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define as the action of a monoid on the component.
+Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define by the action of a monoid on the component.
 While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's scales \cite{stevensTheoryScalesMeasurement1946}, we generalize to monoids to support more component types.
@@
 Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-For example, given the interval scale fiber component $(\fname=temperature,\, \ftype=\texttt{float},\,\fttype=\reals)$:
+For example, given an arbitrary interval scale fiber component $(\fname=temperature,\, \ftype=\texttt{float},\,\fttype=\reals)$ with arbitrarily chosen monoid translation actions actions
 \begin{itemize}
@@
 \end{itemize}
-then the translation monoid actions on temperature satisfy the condition
+By structure preservation, we mean that monoid actions are composable. For the translation actions described above on the temperature fiber, this means that they satisfy the condition
 \begin{equation}
@@
 where $1\degree$ and $2\degree$ are valid distances between two temperatures $x$.
+What this diagram means is that either the fiber could be shifted by 1 (vertical line) then by 2 (horizontal), or the two shifts could be combined such that in this case the fiber is shifted by 3 (diagonal) and these two paths yield the same temperature.
 
@@
 Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
-In figure~\ref{fig:base_space_div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the line.
-
+In figure~\ref{fig:base_space_div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the closed interval $\left[0,1\right]$.
 As with fibers and monoids, we can decompose the total space into components $\pi:\dtotal_i\rightarrow \dbase$ where
@@
 \end{equation}
-
 which is a decomposition of \dfiber. The \dbase\ remains the same because the connectivity of records does not change just because there are fewer elements in each record.

### 2021-02-22 code/matplottoy/artists/bar.py

@@ -110,7 +110,6 @@ class StackedBar(martist.Artist):
         self.assemble(view)
         #self._transform = self.children[0].get_transform()
         for artist in self.children:
-            print("DRAW")
             artist.draw(renderer, *args, **kwargs)
 
             
@@ -148,5 +147,4 @@ class GroupedBar(martist.Artist):
         self.assemble(view)
         #self._transform = self.children[0].get_transform()
         for artist in self.children:
-            print("DRAW")
             artist.draw(renderer, *args, **kwargs)

### 2021-02-22 draft/sections/math.tex
breaking down sheafs

@@
 To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names and types.
-Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_{j_0}, \ldots, \ftype_{j_m}\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
+Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m}\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
 He then defines the single variable set \fttype\
@@
 \begin{equation}
-F = \ftotal_{\ftype_1}\times \ldots \ftotal_{\ftype_i} \ldots\times \ftotal_{\ftype_n}
+F = \fttype_{(\fname_1)}\times \ldots \fttype_{(\fname_i)} \ldots\times \ftotal_{(\fname_n)}
 \end{equation}
@@
 \end{equation} 
-where the first component $F_0$ is the set of values specified by $(\fname_0=time,\, \ftype_0=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname_1=temperature,\, \ftype_1=\texttt{float},\, \fttype=\realsp)$ and is the set of values $\fttype=\realsp$.
+where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=temperature,\, \ftype=\texttt{float},\, \fttype=\realsp)$ and is the set of values $\fttype=\realsp$.
 In figure~\ref{fig:fiber_example_cube}, temperature is replaced with wind.
@@
 \end{equation} 
-such that $F_1$ is specified by $(\fname_1 = wind,\, \ftype_1=\texttt{wind},\, \fttype=\reals^{2})$.
+such that $F_1$ is specified by $(\fname = wind,\, \ftype = \texttt{wind},\, \fttype=\reals^{2})$.
 As illustrated in figure~\ref{fig:data}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
@@
 This in turn allows for building visualizations that can work with distributed or streaming data, since it has a common data access interface with a promise that the data exists.
-The \vindex\ mapping is also used by interactive visualization components to look up the data associated with a region on screen.
-One example is to fill in details in a hover tooltip, another is to convert region selection (such as zooming) on \gbase\ to a query on the data to access the corresponding record components on \dbase.
 
@@
 For example, in the special case of a table \cite{spivakSIMPLICIALDATABASES}, \dbase\ is a set of row ids, \dfiber\ is the columns, and the section \dsection\ returns the record at a given key in \dbase.
-In general, for any fiber bundle, there is always a map
+For any fiber bundle, there exists a map
 \begin{equation}
@@
  such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$.
-There can be many sections \dsection\ and the space of all global sections is $\Gamma(\dtotal)$.
+The set of all global sections is denoted as $\Gamma(\dtotal)$.
 Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is
@@
 This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle.
-Because we can decompose the bundle and the fiber, we can formulate $\dsection$ as
+Because we can decompose the bundle and the fiber, we can decompose \dsection\ as
 \begin{equation}
@@
 In the example in figure~\ref{fig:data_sections}, the fiber is $(time, \, temperature)$ as described in figure~\ref{fig:data_fiber_example} and the base space is the interval \dbase.
-The section $\dsection_i$ resolves to a series of monotonically increasing in time records of (time, temperature) values.
-Section $\dsection_j$ returns a different timeseries of (time, temperature) values.
-Both sections are included in the global set of sections $\dsection_i, \dsection_j \in \Gamma(\dtotal)$.
+The section $\dsection^{(1)}$ resolves to a series of monotonically increasing in time records of (time, temperature) values.
+Section $\dsection^{(2)}$ returns a different timeseries of (time, temperature) values.
+Both sections are included in the global set of sections $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$.
 
@@
 A univariate numpy array that stores an image is a section of a fiber bundle where \dbase\ is a 2D continuous plane and the \dfiber\ is $(\reals^{3}, \reals, \reals)$ where $\reals^3$ is color, and the other two components are the x and y positions of the sampled data in the image.
-A series could store the values of $\dsection_i$ and a second series could be  $dsection_j$.
+A series could store the values of $\dsection^{(1)}$ and a second series could be  $\dsection^{(2)}$.
 We could also fatten the fiber to hold two temperature series, such that a section would be an instance of a dataframe with a time column and two temperature columns.
@@
 As with the other containers, \dbase\ and \dfiber\ are defined in terms of the index and dtypes of the components of the array.
-The section can remain curried until render since the model is functional, in keeping with dasks compute graph architecture.
+Because our framework is defined in terms of the fiber, continuity, and sections, rather than the exact values of the data, our model does not need to know what the exact values are until the renderer needs to fill in the image.
 Our model provides a common interface to these widely used data containers without sacrificing the semantic structure embedded in each container.
@@
 \label{sec:data_sheaf_stalk}
+The section \dsection\ is defined globally on \dbase, including infinite \dbase\ for streaming data and total \dbase\ for chunked data, but often we want as input into the artist \vartist\ a subset of the data.
+For example sliding window\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} based visualizations of streaming data, or visualizations that support pan and zoom navigation\cite{nekaro}
+
 To support working with a subset of data, we can take the sheaf $\mathcal{O}(\dtotal)$ as input into the artist.
-\dtotal\ restricted over a small enough neighborhood $U \subset \dbase$ is a locally trivial bundle over $U$\cite{LocallyTrivialFibre}.
+
 The sheaf $O(\dtotal)$ is the localized section of fibers $\iota^*\dsection: U \rightarrow \iota^*\dtotal$
@@
     \begin{tikzcd}
-        \iota^*\dtotal \arrow[d, "\pi"']           & \dtotal \arrow[d, "\pi"'] \arrow[l, "\iota^*"']         \\
-        U \arrow[u, "\iota^*\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right] \arrow[l, "\iota"']
+        \iota^*\dtotal \arrow[d, "\pi"']           & \dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*"']         \\
+        U \arrow[u, "\iota^*\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right] \arrow[r, "\iota"']
     \end{tikzcd}
@@
 The mapping function \vindex\ provides a way to identify the part of the visual transformation that is specific to the the connectivity of the data rather than the values; for example it is common to flip a matrix when displaying an image.
+The \vindex\ mapping is also used by interactive visualization components to look up the data associated with a region on screen.
+One example is to fill in details in a hover tooltip, another is to convert region selection (such as zooming) on \gbase\ to a query on the data to access the corresponding record components on \dbase.
 

### 2021-02-23 draft/sections/math.tex
sheafs done for now

@@
 A set of records is a section $\dsection: \dbase\rightarrow \dtotal$ of the fiber bundle.
-For example, in the special case of a table \cite{spivakSIMPLICIALDATABASES}, \dbase\ is a set of row ids, \dfiber\ is the columns, and the section \dsection\ returns the record at a given key in \dbase.
+For example, in the special case of a table \cite{spivakSIMPLICIALDATABASES}, \dbase\ is a set of row ids, \dfiber\ is the columns, and the section \dsection\ returns the record \delement\ at a given key in \dbase.
 For any fiber bundle, there exists a map
@@
 \end{equation}
-where each section $\dsection_i$ is a variable or set of variables.
+where each section $\dsection_i$ is a variable or set of variables. This allows for accessing the data component wise rather than just as sections on \dbase.
 
@@
 The section \dsection\ is defined globally on \dbase, including infinite \dbase\ for streaming data and total \dbase\ for chunked data, but often we want as input into the artist \vartist\ a subset of the data.
-For example sliding window\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} based visualizations of streaming data, or visualizations that support pan and zoom navigation\cite{nekaro}
+For example sliding window\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} based visualizations of streaming data, or visualizations that support pan and zoom navigation\cite{NekrasovskiEvaluationPanZoom2006}.
+To make these types of visualizations, mathematical objects called sheafs provide a way to construct a \dsection\ on a subset of \dbase.
 
-To support working with a subset of data, we can take the sheaf $\mathcal{O}(\dtotal)$ as input into the artist.
-
-The sheaf $O(\dtotal)$ is the localized section of fibers $\iota^*\dsection: U \rightarrow \iota^*\dtotal$
-
+As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$.
+We can define the inclusion map $\iota: U \rightarrow \dbase$  which pulls \dtotal over $U$
 \begin{equation}
@@
     \begin{tikzcd}
-        \iota^*\dtotal \arrow[d, "\pi"']           & \dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*"']         \\
-        U \arrow[u, "\iota^*\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right] \arrow[r, "\iota"']
+        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
+        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
     \end{tikzcd}
 \end{equation}
-pulled back over the neighborhood $U$ via the inclusion map $\iota: U \rightarrow \dbase$.
-The localized section is the germ $\xi^*\dsection$.
-The neighborhood of points $\dbasepoint_i$ surrounding the point \dbasepoint\ the sheaf lies over is the stalk $\mathcal{F}_b$ \cite{StalkSheaf2019,spanier1989algebraic}.
+such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
+Restriction maps on the sheaf define how to glue together these localized sections $\iota^*\dsection$ into larger sections\cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018} over the subset of \dbase\ requested by the visualization.
 
-While \dtotal\ is only the fiber $\dfiber_k$ over a specific \dbasepoint, the stalk includes the local behavior of the section at \dbasepoint\ which can include information about nearby records.
-In practice, often the only information needed from the stalk is some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
+Another use for the sheaf is to provide access to all the records over the space on which the sheaf $\mathcal{O}(\dtotal)$ is defined.
+The artist \vartist\ could not generate any visualization that needed more information than was in a single record, such as a line plot, if it took as input \dsection, because $\dsection:\dbasepoint \mapsto \delement$ only knows about the record it returns.
+While taking  $\mathcal{O}(\dtotal)$ as input would solve this issue, in practice, often the only information \vartist\ needs is some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
 In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives.
-The sheaf facilitates interactions such as zooming or updating the graphic based on zooming data; the jet ensures that the artist only needs a single record on \dbasepoint\ to render a piece of a graphic.
+The sheaf facilitates interactions such as zooming or updating the graphic based on zooming data; the jet ensures that the artist only needs a single record on \dbasepoint\ extended with derivatives to render a piece of a graphic.
 This allows for the creation of highly concurrent and possibly distributed artists.

### 2021-02-23 draft/sections/math.tex
graphic section done again

@@
 To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names and types.
-Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m}\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
+Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
 He then defines the single variable set \fttype\
@@
 \begin{equation}
-F = \fttype_{(\fname_1)}\times \ldots \fttype_{(\fname_i)} \ldots\times \ftotal_{(\fname_n)}
+F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
 \end{equation}
@@
 Typically \gtotal\ is trivial and therefore sections can be thought of as mappings into \gfiber.
-
 In this work, we assume a 2D opaque image $\gfiber=\reals^5$ with elements
@@
 \label{sec:graphic_base}
-Since we propose that data topology \dbase\ is by some measure preserved in the graphic, we introduce a graphic topology \gbase to encode the carried over structure.
-These topologies are not always identical because the underlying topology \gbase\ of a graphic may need more dimensions than the data topology \dbase\.
-In a typical 2D display (ignoring depth), visible elements need to have 2D extent so that for example a scatter marker has an area greater than 0 or a line is not infinitely thin.
+Just as the \dbase\ encodes the connectivity of the records in the data, we propose an equivalent \gbase\ that encodes the connectivity of the rendered elements of the graphic.
+For some visualizations, \dbase may be lower dimension than \gbase.
+For example, in a typical 2D display (ignoring depth), a point that is 0D in \dbase cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the pixels that visually represent the point.
+This thickening is often not necessary when the dimensionality of \dbase\ matches the dimensionality of the target space, for example if \dbase is 2D and the display is a 2D screen.
+We introduce \gbase\ to thicken \dbase\ in a way which preserves the structure of \dbase.
 
-Since \dbase\ with fewer than two dimensions need to be thickened, we define the mapping from graphic \gbase\ to data \dbase 
+Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy.
+The surjective map $\vindex: \gbase \rightarrow \dbase$
 \begin{equation}
@@
 \end{equation}
-
- as the deformation retraction \cite{RetractionTopology2020} $\vindex: \gbase \rightarrow \dbase$ that goes from a region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$, such that when $\vindex(\gbasepoint) = \dbasepoint$,\, $\vindex^*\dsection(\gbasepoint) = \dsection(\gbasepoint)$.
-While dimensions can be added to \gbase, it retains the same continuity as \dbase.
+goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$.
+This means that if $\vindex(\gbasepoint) = \dbasepoint$, the record at \dbasepoint\ is copied over the region \gbasepoint\ such that $\dsection(\dbasepoint)=\vindex^*\dsection(\gbasepoint)$ where $\vindex^*\dsection(\gbasepoint)$  is \dsection\ pulled back over \gbase.
 
@@
     \caption{The scatter and line graphic base spaces have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
-The heatmap has the same continuity in the graphic \gbase\ as in the data \dbase.
+The image has the same dimension in \gbase\ as in \dbase.
 \note{add $\alpha, \beta$ coordinates to figures}}
@@
 In the case of 1D continuous data and a line plot, the region \gy\ over a point $\gx_i$ specifies the thickness of the line in \gbase\ for the corresponding \dsection\ on \dbasepoint.
-The heatmap has the same continuity in data space and graphic space such that no extra dimensions are needed in \gbase.
+The image has the same dimensions in data space and graphic space such that no extra dimensions are needed in \gbase.
+
 The mapping function \vindex\ provides a way to identify the part of the visual transformation that is specific to the the connectivity of the data rather than the values; for example it is common to flip a matrix when displaying an image.
@@
 
-This section describes how we go from a graphical design to a rendered image, where the graphical design is the section $\gsection: \gbase \rightarrow \gtotal$.
-It is sufficient to sketch out how an arbitrary pixel would be renderer, where a pixel $p$ in a real display corresponds to a region $\gbase_p$ in the idealized display.
+This section describes how we go from a graphic in an idealized prerender space to a rendered image, where the graphic is the section $\gsection: \gbase \rightarrow \gtotal$.
+It is sufficient to sketch out how an arbitrary pixel would be rendered, where a pixel $p$ in a real display corresponds to a region $\gbase_p$ in the idealized display.
 To determine the color of the pixel, we aggregate the color values over the region via integration.

### 2021-02-24 draft/sections/math.tex
pulled up jet to definition of artist since thats when is needed

@@
 
-\subsubsection{Sheaf and Stalk}
-\label{sec:data_sheaf_stalk}
-The section \dsection\ is defined globally on \dbase, including infinite \dbase\ for streaming data and total \dbase\ for chunked data, but often we want as input into the artist \vartist\ a subset of the data.
-For example sliding window\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} based visualizations of streaming data, or visualizations that support pan and zoom navigation\cite{NekrasovskiEvaluationPanZoom2006}.
-To make these types of visualizations, mathematical objects called sheafs provide a way to construct a \dsection\ on a subset of \dbase.
-
-As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$.
-We can define the inclusion map $\iota: U \rightarrow \dbase$  which pulls \dtotal over $U$
-\begin{equation}
-    \label{eq:sheaf}
-    \begin{tikzcd}
-        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
-        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
-    \end{tikzcd}
-\end{equation}
-such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
-Restriction maps on the sheaf define how to glue together these localized sections $\iota^*\dsection$ into larger sections\cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018} over the subset of \dbase\ requested by the visualization.
-
-Another use for the sheaf is to provide access to all the records over the space on which the sheaf $\mathcal{O}(\dtotal)$ is defined.
-The artist \vartist\ could not generate any visualization that needed more information than was in a single record, such as a line plot, if it took as input \dsection, because $\dsection:\dbasepoint \mapsto \delement$ only knows about the record it returns.
-While taking  $\mathcal{O}(\dtotal)$ as input would solve this issue, in practice, often the only information \vartist\ needs is some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
-In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives.
-The sheaf facilitates interactions such as zooming or updating the graphic based on zooming data; the jet ensures that the artist only needs a single record on \dbasepoint\ extended with derivatives to render a piece of a graphic.
-This allows for the creation of highly concurrent and possibly distributed artists.
-
 \subsection{Graphic: \gtotal}
@@
 The artist is the function that converts data into graphics; its name is taken from the analogous part of Matplotlib\cite{hunterArchitectureOpenSource} that builds visual elements to pass off to the renderer.
-The artist \vartist\ is a mapping from \dtotal\ padded with data from $\dtotal^{\prime}=\mathcal{J}(\dtotal)$ to a graphic that is a section $\gsection \in \Gamma(\gtotal)$
+The topological artist \vartist\  is a monoid equivariant sheaf map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
+\begin{equation}
+    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
+\end{equation}
+
+Usually \vartist can usually map into graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
+In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the data bundle padded with the jet bundle $\dtotal^{\prime} = \dtotal + \mathcal{J}^{2}(\dtotal)$.
+Specifically, \vartist is the map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$
 
@@
 \end{equation}
-
-which due to using the jet is point wise such that the input can be $\dsection(\dbasepoint)$.
-The encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ converts the data components to visual components(\ref{sec:graphic_base}).
-The continuity map $\vindex:\gbase \dbase$ then pulls back the visual bundle \vtotal\ over \gbase (\ref{sec:artist_nu}).
+where the input can be point wise $\dsecton(\dbasepoint)\mid \dbasepoint \in \dbase$.
+The encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components(\ref{sec:graphic_base}).
+The continuity map $\vindex:\gbase \rightarrow \dbase$ then pulls back the visual bundle \vtotal\ over \gbase (\ref{sec:artist_nu}).
 Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal (\ref{sec:artist_q}).
@@
 
+\subsubsection{Sheaf and Stalk}
+\label{sec:data_sheaf_stalk}
+\note{move to artost}
+Another use for the sheaf is to provide access to all the records over the space on which the sheaf $\mathcal{O}(\dtotal)$ is defined.
+The artist \vartist\ could not generate any visualization that needed more information than was in a single record, such as a line plot, if it took as input \dsection, because $\dsection:\dbasepoint \mapsto \delement$ only knows about the record it returns.
+The data bundle \dtotal\ is defined globally on \dbase, but often we want as
+
+ To make these types of visualizations, mathematical objects called sheafs provide a way to construct a \dsection\ on a subset of \dbase.
+
+As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$.
+We can define the inclusion map $\iota: U \rightarrow \dbase$  which pulls \dtotal over $U$
+\begin{equation}
+    \label{eq:sheaf}
+    \begin{tikzcd}
+        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
+        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
+    \end{tikzcd}
+\end{equation}
+such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
+Restriction maps on the sheaf define how to glue together these localized sections $\iota^*\dsection$ into larger sections\cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018} over the subset of \dbase\ requested by the visualization.
+
+The sheaf facilitates visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.; the jet ensures that the artist only needs a single record on \dbasepoint\ extended with derivatives to render a piece of a graphic.
+This allows for the creation of highly concurrent and possibly distributed artists.
+
 \subsubsection{Composition of Artists: +}

### 2021-02-24 draft/sections/math.tex
math issue

@@
 \end{equation}
-
-Usually \vartist can usually map into graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
+Sheafs are a mathematical object with restriction maps that define how to glue \dsection over local neighborhoods $U \subseteq \dbase$,  discussed in section~\ref{sec:artist_sec:data_sheaf_stalk}, such that the \vartist\ maps are consistent over continuous regions of \dbase.
+While \vartist can usually construct graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
 In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the data bundle padded with the jet bundle $\dtotal^{\prime} = \dtotal + \mathcal{J}^{2}(\dtotal)$.
@@
 \end{equation}
-where the input can be point wise $\dsecton(\dbasepoint)\mid \dbasepoint \in \dbase$.
+where the input can be point wise $\dsection(\dbasepoint)\mid \dbasepoint \in \dbase$.
 The encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components(\ref{sec:graphic_base}).
@@
 For example, given a fiber of $\{xpos, ypos, color\}$ one possible section could be  $\{.5, .5, (255, 20,147)\}$.
-The $codomain(\vchannel_i)$ determines the monoid actions on $\vsection_i$.
+The $codomain(\vchannel_i)$ determines the monoid actions on $\vfiber_i$.
 These fiber components are implicit in the library, by making them explicit as components of the fiber we can build consistent definitions and expectations of how these parameters behave.
@@
 \end{figure}
-As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers \vchannel\ as the set of independent conversion functions 
+As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers \vchannel\ 
 \begin{equation}
@@
 \end{equation}
-where $\vchannel_i: \dsection_i \mapsto \vsection_i$ is an equivariant map such that there is a monoid homomorphism from $\dfiber_{i}$ to $vfiber_{i}$.
+as the set of equivariant maps $\vchannel_i: \dsection_i \mapsto \vsection_i$.
+Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve.
 As mentioned in section~\ref{sec:data_monoid}, we choose monoid actions as the basis for equivariance because they define the structure on the fiber components.
+
 A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$
@@
 Together, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings, reduce the encodings into a graphic.
-As with \vchannel\, the constraint on \vmark\ is that for every monoid action on the input \vsection\, there is a corresponding monoid action on the output \gsection.
+As with \vchannel\, the constraint on \vmark\ is that for every monoid action on the input \vsection\, there is corresponding monoid action on the output \gsection.
 
@@
 
-The glyph in figure~\ref{fig:artist_mark_change} has the following characterstics \vfiber specified by  $(xpos,\, ypos,\, color,\, thickness)$ such that one section is $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
-The equivariance constraint on \vmark is that the action $m=(e, e, e, x+2)$, where e is identity, applied to \vsection\ such that $\vsection^{\prime}=(e,e,e,3)$ has an equivalent action on \gsection that causes $\vmark(\vsection^{\prime})$ to be equivalent to the thicker circle in figure~\ref{fig:artist_mark_change}.
+The glyph in figure~\ref{fig:artist_mark_change} has the following characteristics \vfiber specified by  $(xpos,\, ypos,\, color,\, thickness)$ such that one section is $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
+The equivariance constraint on \vmark\ is that the action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$.
+The corresponding action on \gsection causes $\vmark(\vsection^{\prime})$ to be the thicker circle in figure~\ref{fig:artist_mark_change}.
 
@@
 
-\subsubsection{Sheaf and Stalk}
-\label{sec:data_sheaf_stalk}
-\note{move to artost}
-Another use for the sheaf is to provide access to all the records over the space on which the sheaf $\mathcal{O}(\dtotal)$ is defined.
-The artist \vartist\ could not generate any visualization that needed more information than was in a single record, such as a line plot, if it took as input \dsection, because $\dsection:\dbasepoint \mapsto \delement$ only knows about the record it returns.
-The data bundle \dtotal\ is defined globally on \dbase, but often we want as
-
- To make these types of visualizations, mathematical objects called sheafs provide a way to construct a \dsection\ on a subset of \dbase.
-
+\subsubsection{Sheaf}
+\label{sec:artist_sheaf_stalk}
 As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$.
@@
 
-The sheaf facilitates visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.; the jet ensures that the artist only needs a single record on \dbasepoint\ extended with derivatives to render a piece of a graphic.
-This allows for the creation of highly concurrent and possibly distributed artists.
+The sheaf facilitates visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.;
 
@@
 \label{sec:artist_equivalance}
-It is impractical to implement an artist for every single graphic; instead we implement the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$ which is essentially an artist where the \vchannel\ functions have not yet been input.
-Equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\, but act on different sections \vsection.
-To further simplify implementation, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data such that the graphic is identifiable as a given chart type.
+It is impractical to implement an artist for every single graphic; instead we implement an approximation of an the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$.
+Roughly, equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\ but act on different sections \vsection, but we will formalize the definition of the equivalance class in future work.
+As a first pass for implementation purposes, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data such that the graphic is identifiable as a given chart type.
 

### 2021-02-24 draft/sections/topodata.tex
math issue

@@
-%%pull out cause not helping
+\subsection{Making the fiber bundle computable}
+\label{sec:triangulization}
+\begin{figure}[H]
+    \includegraphics{figures/math/simplex.png}
+    \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) records to all records are connected to at least 3 others}
+    \label{fig:triangle_simplex}
+\end{figure}
+%%big data
+One way of expressing the connectivity of records in a dataset is to implement \dbase\ as a simplacial complex, which is a set of simplicies such as those shown in figure~\ref{fig:triangle_simplex}.
+The advantage of triangulization is that it is general enough to work for more complex topology based visualization methods \cite{heineSurveyTopologybasedMethods2016} while also providing a consistent interface of vertices, edges, and faces for \vindex\ to map into.
+When triangulated, the simplices encode the continuity in the data
+
+\begin{table}[H]
+    \begin{center}
+        \begin{tabular}{|l|l|l|}\hline
+        \textbf{simplex} & \textbf{continuity} & \textbf{\dsection}   \\ \hline
+        vertex  & discrete   & \dsection(\dbasepoint)                  \\ \hline
+        edge    & 1D         & \dsection(\dbasepoint,\, \dx)        \\ \hline
+        face    & 2D         & \dsection(\dbasepoint,\, \dx,\, \dy)\\ \hline
+        \end{tabular}
+        \caption{}
+        \label{tab:triangulization}
+    \end{center}
+\end{table}
+
+such that each section is bound to a simplex $\dbasepoint \in \dbase$.
+As shown in table~\ref{tab:triangulization}, in a 1D continuous spaces each \dsection\ lies distance \dx\ along edge \dbasepoint, while in a 2D continuous space each \dsection\ lies at coordinate \dx,\dy on the face \dbasepoint.
+This is directly analogous to indexing to express connectivity in N-D arrays, while also natively supporting graphs and trees as they are simplacial complices of nodes and edges.
+Path connected components are then sections where edges or faces meet.
+
+In general simplacial complexes, in these prototypes we implement graphs. 
 \begin{figure}[H]

### 2021-02-24 draft/sections/intro.tex
tom's comments on 1

@@
 
- This work is motivated by a need for a visualization library that developers could build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.
+While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
+The work presented in this paper is motivated by a need for a visualization library that developers could build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.
 The core architecture also needs to be robust to the big data needs of many visualization practitioners, and therefore support distributed and streaming data needs.
 To support both exploratory and confirmatory visualization\cite{tukeyWeNeedBoth1980}, this tool needs to support 2D and 3D, static, dynamic and interactive visualizations.
-Specifically, this work was driven by a rearchiture the Python visualization library Matplotlib\cite{hunterMatplotlib2DGraphics2007} to meet modern data visualization needs.
+
+Specifically, this work was driven by a rearchitecture of the Python visualization library Matplotlib\cite{hunterMatplotlib2DGraphics2007} to meet modern data visualization needs.
 We aim to take advantage of developments in software design, data structures, and visualization to improve the consistency, composibility, and discoverability of the API.
+To do so, this work first presents a mathematical description of how data is transformed into graphic representations, as shown in figure~\ref{fig:intro_artist_stages}.
+As with other mathematical formalisms of visualization \cite{mackinlayAutomatingDesignGraphical1986,kindlmann2014algebraic,sugibuchiFramwork2009,vickersUnderstandingViz2013}, a mathematical framework provides a way to formalize the properties and structure of the visualization.
+In contrast to the other formalisms, the model presented here is focused on the components that build a visualization rather than the visualization itself.
 
- This work presents a mathematical description of how data is transformed into graphic representations, as shown in figure~\ref{fig:intro_artist_stages}.
-While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
+In other words this model is not intended to be evaluative, it is intended to be a reference specification for visualization library API.
+To make this model as implementation independent as possible, we propose fairly general mathematical abstractions of the data container such that we do not need to assume the data has any specific structure, such as a relational database.
+We reuse this structure for the graphic as that allows us to specifically discuss how structure is preserved.
+We take a functional approach because that gives implementers the flexibility to carry as little or as much state as they need and allows us to specify the constraints on these functions without having to know anything about the implementation or any state that these functions may be carrying.
+As with the other mathematical formalisms of visualization, we factor out the rendering into a separate stage; but our framework describes how these rendering instructions are generated.
+
  In this work, we present a framework for understanding visualization as equivariant maps between topological spaces.

### 2021-02-24 draft/sections/intro.tex
gog rework

@@
 \section{Background}
-Visual algorithms that display information are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighLevel2004}.
-Specifically they note that discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}.
-This assumption is evident in the information visualization software design patterns categorized by Heer and Agrawala\cite{HeerSoftware2006} and can often be found in modern general purpose visualization libraries.
+One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures.
+This in turn restricts the design space of visual algorithms that display information (the visualization types the library supports) since the algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighLevel2004}.
 In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.
@@
 The library driving this rearchitecture, Matplotlib, aims to be flexible enough that developers can build domain specific libraries on top of it.
-To preserve this flexibility, Matplotlib enforces minimal constraints on what sorts of data the user is allowed to input and supports very many visual algorithms, from primative marks to computationally complex visualizations.
+To preserve this flexibility, Matplotlib enforces minimal constraints on the types of data the user is allowed to input and supports very many types of graphics, from primitive marks to computationally complex visualizations.
 Instead of enforcing constraints at the API level, Matplotlib carries implicit assumptions about data continuity in how each function interfaces with the input data.
@@
 A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+The GoG family of tools are very strongly built on the assumption that data is a relational framework; this is explicit in Mackinlay's A Presentation Tool (APT)\cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987} formalism on which GoG is build.
 This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create.
 Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building the elements.
+A clear example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color.
+
+assumptions on each specific component of the visualization, which in turn allows for building composable
 
@@
 \label{sec:intro_data}
-As mentioned, one of the drivers of this work was to facilitate building libraries that could natively support domain specific data containers.
+Discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}, so one of the drivers of this work was to facilitate building libraries that could natively support domain specific data containers that do not make assumptions about data continuity.
 Fiber bundles are one such way model to model containers, as Butler proposes because they encode the continuity of the data separately from the types of variables and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.

### 2021-02-24 draft/sections/intro.tex
done for now w/ gog cause this took 2 hours

@@
 We reuse this structure for the graphic as that allows us to specifically discuss how structure is preserved.
-We take a functional approach because that gives implementers the flexibility to carry as little or as much state as they need and allows us to specify the constraints on these functions without having to know anything about the implementation or any state that these functions may be carrying.
-As with the other mathematical formalisms of visualization, we factor out the rendering into a separate stage; but our framework describes how these rendering instructions are generated.
+We take a functional approach because functional paradigms encourage writing APIs that are flexible, concise and predictable due to the lack of side effects \cite{loudenProgrammingLanguagesPrinciples2010}.
+Furthermore, by structuring the API in terms of composition of the smallest units of transformation for which we can define correctness, a functional paradigm naturally leads to a library of highly modular components that are composable in such a way that by definition the composition is also correct.
+This allows us to ensure that domain specific visualizations built on top of these components are also correct without needing knowledge of the domain.
+As with the other mathematical formalisms of visualization, we factor out the rendering into a separate stage; but, our framework describes how these rendering instructions are generated.
 
@@
 A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
-The GoG family of tools are very strongly built on the assumption that data is a relational framework; this is explicit in Mackinlay's A Presentation Tool (APT)\cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987} formalism on which GoG is build.
-This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create.
-Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building the elements.
-A clear example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color.
-
-assumptions on each specific component of the visualization, which in turn allows for building composable
+The GoG family of libraries are very strongly built on the assumption that data is a relational table, which restricts the native set of visualization types the libraries support.
+Since the data space and graphic space is very well defined in this grammar, it lends itself to a declarative interface \cite{heerDeclarative2010} that allows end users to describe the visualization they are trying to create.
+Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
+An example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that most computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color.
+Disentangling the computation from the visual transforms allows us to determine whether the visualization library needs to handle them or if they can be more efficiently computed by the data container.
 

### 2021-02-24 draft/sections/intro.tex
pushed mpl down to vtk, reworked gog again

@@
 \label{sec:intro_data_tools}
-The library driving this rearchitecture, Matplotlib, aims to be flexible enough that developers can build domain specific libraries on top of it.
-To preserve this flexibility, Matplotlib enforces minimal constraints on the types of data the user is allowed to input and supports very many types of graphics, from primitive marks to computationally complex visualizations.
-Instead of enforcing constraints at the API level, Matplotlib carries implicit assumptions about data continuity in how each function interfaces with the input data.
-This has lead to poor API consistency and brittle code as every visual algorithm has a very different point of view on how the data is structured.
-
-A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
-The GoG family of libraries are very strongly built on the assumption that data is a relational table, which restricts the native set of visualization types the libraries support.
-Since the data space and graphic space is very well defined in this grammar, it lends itself to a declarative interface \cite{heerDeclarative2010} that allows end users to describe the visualization they are trying to create.
-Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
+One extensive family of relational table based libraries are those based on Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+The restriction to tables in turn restricts the native design space to visualizations suited to tables.
+Since the data space and graphic space is very well defined in this grammar, it lends itself to a declarative interface \cite{heerDeclarative2010}.
+This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
 An example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that most computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color.
@@
 
-A different class of user facing tools are those that support images, such as ImageJ\cite{schneiderNIHImageImageJ2012}.
-These tools mostly have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized.
-These tools are ill suited for general purpose library developers as the architecture is to build plugins into the existing system \cite{WritingPlugins} rather than domain specific tools that are built on top.
-Even the digital humanities oriented macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supports some non-image aggregate reporting charts, is still built around image data as the primary input.
+A different class of user facing tools are those that support images, such as ImageJ\cite{schneiderNIHImageImageJ2012} or Napari\cite{nicholas_sofroniew_2021_4533308}.
+These tools often have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized.
+These tools are ill suited for general purpose libraries that need to support data other than images because the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure.
+Even the digital humanities oriented ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supports some non-image aggregate reporting charts, is still built around image data as the primary input.
 
-There are also visualization tools designed around scientifically complex data that support explicit description of the data, such as vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}.
-These libraries, generally speaking, are architectured more as graphics libraries than visualization libraries, meaning they lack a clear distinction between the graphic construction and the rendering of that graphic.
-This is very close to the existing Matplotlib architecture, where the data, visual encoding, and rendering are jumbled in ways that make it hard to figure out what the code is doing.
-We are proposing a functional framework instead in large part to clearly encapsulate these separate responsibilities of a visualization tool.
-In turn, this should make it easier to reuse components in the building of new tools.
+There are also visualization tools where there is no single core structure, and instead internally carry around many different representations of data.
+Matplotlib, has this structure, as does VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}.
+Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects.
+This in turn means that every new type of visualization must carry implicit assumptions about data structure in how it interfaces with the input data.
+This has lead to poor API consistency and brittle code as every visualization type has a very different point of view on how the data is structured.
+This API choice particularly breaks down when the same dataset is fed into visualizations with different assumptions about structure or into a dashboard consisting of different types of visualization\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007} because there is no consistent way to update the data and therefore no consistent way of guaranteeing that the views stay in sync.
+Our model is a structure dependent formalism, but then also provides a core representation of that structure that is abstract enough to provide a common interface for many different types of visualization.
 

### 2021-02-24 draft/sections/math.tex
pushed mpl down to vtk, reworked gog again

@@
 \label{sec:artist_sheaf_stalk}
+The restriction maps of a sheaf describe how local \dsections can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
 As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$.
@@
 such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
-Restriction maps on the sheaf define how to glue together these localized sections $\iota^*\dsection$ into larger sections\cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018} over the subset of \dbase\ requested by the visualization.
-
-The sheaf facilitates visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.;
+By gluing $\iota^*\dsection$ together, the sheaf is putting a continuous structure on local sections which allows for defining a section over a subset in \dbase.
+That section over subset \dbase\ maps to the graphic generated by \vartist\ for visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.
 
@@
 The criteria for equivalence class membership serves as the basis for evaluating invariance\cite{kindlmann2014algebraic}.
-
-\subsection{Making the fiber bundle computable}
-\label{sec:triangulization}
-\begin{figure}[H]
-    \includegraphics{figures/math/simplex.png}
-    \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) records to all records are connected to at least 3 others}
-    \label{fig:triangle_simplex}
-\end{figure}
-%%big data
-One way of expressing the connectivity of records in a dataset is to implement \dbase\ as a simplacial complex, which is a set of simplicies such as those shown in figure~\ref{fig:triangle_simplex}.
-The advantage of triangulization is that it is general enough to work for more complex topology based visualization methods \cite{heineSurveyTopologybasedMethods2016} while also providing a consistent interface of vertices, edges, and faces for \vindex\ to map into.
-When triangulated, the simplices encode the continuity in the data
-
-\begin{table}[H]
-    \begin{center}
-        \begin{tabular}{|l|l|l|}\hline
-        \textbf{simplex} & \textbf{continuity} & \textbf{\dsection}   \\ \hline
-        vertex  & discrete   & \dsection(\dbasepoint)                  \\ \hline
-        edge    & 1D         & \dsection(\dbasepoint,\, \dx)        \\ \hline
-        face    & 2D         & \dsection(\dbasepoint,\, \dx,\, \dy)\\ \hline
-        \end{tabular}
-        \caption{}
-        \label{tab:triangulization}
-    \end{center}
-\end{table}
-
-such that each section is bound to a simplex $\dbasepoint \in \dbase$.
-As shown in table~\ref{tab:triangulization}, in a 1D continuous spaces each \dsection\ lies distance \dx\ along edge \dbasepoint, while in a 2D continuous space each \dsection\ lies at coordinate \dx,\dy on the face \dbasepoint.
-This is directly analogous to indexing to express connectivity in N-D arrays, while also natively supporting graphs and trees as they are simplacial complices of nodes and edges.
-Path connected components are then sections where edges or faces meet.
-
-In general simplacial complexes, in these prototypes we implement graphs.
-
 \end{document}

### 2021-02-24 draft/sections/intro.tex
rework of data

@@
 Discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}, so one of the drivers of this work was to facilitate building libraries that could natively support domain specific data containers that do not make assumptions about data continuity.
-Fiber bundles are one such way model to model containers, as Butler proposes because they encode the continuity of the data separately from the types of variables and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-Since Butler's model lacks a robust way of describing the variables, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
+As shown in figure~\ref{fig:intro_data_format}, there are many types of connectivity.
+A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph.
+These data structures typically contain not only the measurements or values of the data, but also domain specific semantic information such as that the data is a map or an image that a modern visualization library could exploit if this information was exposed to the API.
 
@@
 \end{figure}
-As shown in figure~\ref{fig:intro_data_format}, there are many types of connectivity.
-A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph.
+As shown in figure~\ref{fig:intro_data_format}, there are many distinct ways of encoding each specific type of structure, while as mentioned in section~\ref{sec:intro_data_tools} APIs are clearer when structured around a common data representation.
+Fiber bundles were proposed by Butler as one such representation because they encode the continuity of the data separately from the types of variables and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
 In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements.
@@
 \subsection{Visualization}
+\label{sec:intro_viz}
 A visualization tool produces a graphical design and an image rendered based on that design, as described by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.

### 2021-02-24 draft/sections/intro.tex
rerework of intro

@@
 \end{figure}
-
+The work presented in this paper is motivated by a need for a library of visualization components that developers could use to build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.
 While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
-The work presented in this paper is motivated by a need for a visualization library that developers could build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.
 The core architecture also needs to be robust to the big data needs of many visualization practitioners, and therefore support distributed and streaming data needs.
@@
 \label{sec:intro_viz}
+
 A visualization tool produces a graphical design and an image rendered based on that design, as described by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.

### 2021-02-25 draft/sections/intro.tex
halfway rework of visual section

@@
 \label{sec:intro_viz}
+\begin{figure}[H]
+\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
+\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the components of a visualization.
+The best to show column describes which types of information can be expressed in the corresponding visual encoding.
+This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS}
+\cite{krygierMakingMapsVisual2005}}
+\label{fig:intro_retinal_variables}
+\end{figure}
+
+Visual representations of data have always tried to reflect something of the underlying structure of the data\cite{friendlyBriefHistoryData2008}.
+The components of the visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}.
+As shown in figure~\ref{fig:intro_retinal_variables}, he proposes that there are classes of visual encodings such as shape, color, and texture and which type of measurement, quantitative or qualitative, will be preserved when the data is mapped to that encoding.
+Furthermore he proposes that the visual encodings be composited into graphical marks that match the connectivity of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark.
+A general form of marks are glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014} and minimally need to be differentiable from other visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+The set of encoding relations from data to visual representation is termed the graphical design by Mackinlay \cite{aumackinlayAutomatingDesignGraphical1986} and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.
 
 A visualization tool produces a graphical design and an image rendered based on that design, as described by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
-He defines the graphical design as the set of encoding relations from data to visual representation\cite{mackinlayAutomatingDesignGraphical1986}, and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.
-In addition to the graphic representations, Byrne et al.
-describe how visualizations have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
 Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tool's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
+
+It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
+A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}.
+Vickers et al.
+propose a category theory framework\cite{vickersUnderstandingViz2013} that extends the equivariance checks to layout, but is focused strictly on the design layer like the other mathematical frameworks.
 
@@
 
+In addition to the graphic representations, Byrne et al.
+describe how visualizations have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+
 We propose that the Du Bois visualizations in figure~\ref{fig:intro_dubois} are representative of the expressivity a core architecture should allow a downstream library to express.
@@
 
-\begin{figure}[H]
-\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
-\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the components of a visualization.
-The best to show column describes which types of information can be expressed in the corresponding visual encoding.
-This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS}
-\cite{krygierMakingMapsVisual2005}}
-\label{fig:intro_retinal_variables}
-\end{figure}
-
-The visual properties were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who also described the types of data that paired with the visual properties shown in figure~\ref{fig:intro_retinal_variables}.
-It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
-A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}.
-Vickers et al.
-propose a category theory framework\cite{vickersUnderstandingViz2013} that extends the equivariance checks to layout, but is focused strictly on the design layer like the other mathematical frameworks.
-
-The visual encodings are composited into the point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables}.
-Marks can be generalized to glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014} and minimally need to be differentiable from other visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}.
  As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools.

### 2021-02-25 draft/sections/intro.tex
updated dubois, intro, conclusion

@@
 The best to show column describes which types of information can be expressed in the corresponding visual encoding.
-This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS}
+This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from Krygier and Wood's \textit{Making Maps: A Visual Guide to Map Design for GIS}
 \cite{krygierMakingMapsVisual2005}}
@@
 
-Visual representations of data have always tried to reflect something of the underlying structure of the data\cite{friendlyBriefHistoryData2008}.
-The components of the visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}.
-As shown in figure~\ref{fig:intro_retinal_variables}, he proposes that there are classes of visual encodings such as shape, color, and texture and which type of measurement, quantitative or qualitative, will be preserved when the data is mapped to that encoding.
+Visual representations of data, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}.
+As illustrated in figure~\ref{fig:intro_retinal_variables}, Bertin proposes that there are classes of visual encodings such as shape, color, and texture that when mapped to from specific types of measurement, quantitative or qualitative, will preserve the properties of that measurment type.
+For example, that nominal data mapped to hue preserves the selectivity of the nominal measurements.
 Furthermore he proposes that the visual encodings be composited into graphical marks that match the connectivity of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark.
 A general form of marks are glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014} and minimally need to be differentiable from other visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}.
-The set of encoding relations from data to visual representation is termed the graphical design by Mackinlay \cite{aumackinlayAutomatingDesignGraphical1986} and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.
+The set of encoding relations from data to visual representation is termed the graphical design by Mackinlay \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987} and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.
 
-A visualization tool produces a graphical design and an image rendered based on that design, as described by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
-Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tool's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
-
-It is at this encoding level that Mackinlay formalized expressiveness as a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
-A functional dependency framework for evaluating these equivariances was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algbraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}.
+The measure of how much of the structure of the data the graphic encodes is a concept Mackinlay termed expressiveness, while the graphic's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
+When the proporties of the representation match the properties of the data, then the visualization is easier to understand according to Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis}.
+These ideas are combined into Tufte's notion of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements \cite{tufteVisualDisplayQuantitative2001}.
+This notion of matching is explictely formalized by Mackinlay as a structure preserving mapping of a binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
+A functional dependency framework for evaluating visualizations was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algebraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}.
 Vickers et al.
-propose a category theory framework\cite{vickersUnderstandingViz2013} that extends the equivariance checks to layout, but is focused strictly on the design layer like the other mathematical frameworks.
+propose a category theory framework\cite{vickersUnderstandingViz2013} that extends structural preservation to layout, but is focused strictly on the design layer like the other mathematical frameworks.
 
@@
     \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{figures/intro/du_bois_bar.png}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_line.png}
         \caption{}
@@
     \caption{Du Bois' data portraits\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018} of post reconstruction Black American life exemplify that the fundemental characteristics of data visualization is that the visual elements vary in proportion to the source data.
-In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the bar charts are intersected to show the number of property owners and how much they own in a given year; in figure~\ref{fig:intro_dbc} the countries are scaled to population size; and figure~\ref{fig:intro_dbd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
+In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the line changes color to indicate a shift in the political environment; in figure~\ref{fig:intro_dbc} the countries are scaled to population size; and figure~\ref{fig:intro_dbd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
 The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroNegro1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
@@
 
-In addition to the graphic representations, Byrne et al.
-describe how visualizations have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+One example of highly expressive visualizations are the data portraits by Du Bois shown in figure~\ref{fig:intro_dubois}. While the Du Bois charts such as scatter, line, and plot-
 
-We propose that the Du Bois visualizations in figure~\ref{fig:intro_dubois} are representative of the expressivity a core architecture should allow a downstream library to express.
-This is because while the Du Bois figures are not the common scatter, bar, or line plot \cite{friendlyBriefHistoryData2008}, they conform to the constraint that a visualization is a mapping from data to visuals.
+they conform to the constraints that a visualization is a mapping from data to visuals.
+Figure~\ref{fig:intro_dpa} is semantically similar to a bar chart in that the lengths of the segments are mapped to the values, but in this chart the segments are stacked together.
+Figure~\ref{fig:intro_dpb} is a multicolored line chart where the color shifts are at periods of political significance.
 In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph.
-The visual representations in the Du Bois figures are in proportion to the quantitative data being represented, so the a chart is faithful according to Tufte's Lie Principal\cite{tufteVisualDisplayQuantitative2001}.
-The properties of the representation match the properties of the information being represented, so they are fairly understandable according to Norman's Naturalness Principal\cite{norman_things_smart}.
-As with the Du Bois data portraits, it is fundamental that any architecture tool we build ensures that the data properties match the visual properties.
-
- As retinal variables and marks are the building blocks of most visualizations, they must be fully expressible in a framework for building visualization tools.
-By building visualization concepts into the core architecture, developers can incorporate assessments of the visualization,  such as as quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmann2014algebraic} of visualizations with respect to graphical encoding choices.
+Figure~\ref{fig:intro_dpb} is simply a treemap\cite{heerTourVisualizationZoo2010} with space between the marks.
+Since the Du Bois data portraits meet the criteria of a faithful visual representation, we propose a mathematical framework and implementation that allows us to express the Du Bois charts and common charts with equal fidelity.
 

### 2021-02-25 draft/sections/next.tex
rework of bar to incorporate make bar and simplify vert construction  + reordering around it
updated dubois, intro, conclusion

@@
 While the model's simple addition supports axes as standalone artists with overlapping visual position encoding, the complex operator would allow for binding together data that needs to be updated together.
+Additionally, implementing the complex addition operator and explicit graphic to data maps would allow for developing a mathematical formalism and prototype of how interactivity would work in this model.
 In summary, the proposed scope of work for the dissertation is
@@
     \item mathematical formulation of a graphic with axes labeling
+    \item implementation of new prototype artists that do not inherit from Matplotlib artists
     \item provisional mathematics and implementation of user level composite artists
@@
 \end{itemize}
-Additionally, implementing the complex addition operator and explicit graphic to data maps would allow for developing a mathematical formalism and prototype of how interactivity would work in this model.
  Other potential tasks for future work is implementing a data object for a non-trivial fiber bundle and exploiting the models section level formalism to build distributed data source models and concurrent artists.
 This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
+Since this model formalizes notions of structure preservation, it can swerve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmann2014algebraic} of visualizations with respect to graphical encoding choices.
 While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.

### 2021-02-25 draft/sections/intro.tex
up to 3.3.4 in math

@@
 In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the line changes color to indicate a shift in the political environment; in figure~\ref{fig:intro_dbc} the countries are scaled to population size; and figure~\ref{fig:intro_dbd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
-The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroNegro1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
+The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroValuation1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
     \label{fig:intro_dubois}
@@
 
-One example of highly expressive visualizations are the data portraits by Du Bois shown in figure~\ref{fig:intro_dubois}. While the Du Bois charts such as scatter, line, and plot-
-
-they conform to the constraints that a visualization is a mapping from data to visuals.
+One example of highly expressive visualizations are the data portraits by Du Bois shown in figure~\ref{fig:intro_dubois}.
+While the Du Bois charts are different from the usual scatter, line, and plot charts, they conform to the constraint that a graphic is a structure preserving map from data to visual representation.
 Figure~\ref{fig:intro_dpa} is semantically similar to a bar chart in that the lengths of the segments are mapped to the values, but in this chart the segments are stacked together.
@@
 Figure~\ref{fig:intro_dpb} is simply a treemap\cite{heerTourVisualizationZoo2010} with space between the marks.
-Since the Du Bois data portraits meet the criteria of a faithful visual representation, we propose a mathematical framework and implementation that allows us to express the Du Bois charts and common charts with equal fidelity.
+Since the Du Bois data portraits meet the criteria of a faithful visual representation, we propose a mathematical framework and implementation that allows us to express the Du Bois charts and common chart types with equal fidelity.
 

### 2021-02-25 draft/sections/math.tex
up to 3.3.4 in math

@@
 By formalizing structure in this way, we can evaluate the extent to which a visualization preserves the structure of the data it is representing and build structure preserving visualization tools.
-We introduce the notion of an artist $\mathscr{\vartist}$ as a structure preserving map from data $\mathscr{\dtotal}$ to $\mathscr{\gtotal}$
+We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map from data to graphic
 \begin{equation}
@@
 \end{equation}
+that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}.
+Monoid actions are a set of operations and are discussed in detail in section~\ref{sec:data_monoid}.
+Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are symmetric
+\begin{equation}
+\mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
+\end{equation}
+such that applying a monoid $m \in \monoid$ transformation to the input $\delement \in \mathscr{\dtotal}$ to $\mathscr{\vartist}$ is equivalent to applying a monoid $\varphi(\monoid) \in \monoid^{\prime}$ to the output of the artist $\vartist(\delement) \in \mathscr{\gtotal}$.
 
@@
 such that $F_1$ is specified by $(\fname = wind,\, \ftype = \texttt{wind},\, \fttype=\reals^{2})$.
-As illustrated in figure~\ref{fig:data}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
+As illustrated in figure~\ref{fig:data_fiber_example}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
 
@@
 Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmann2014algebraic} that must be preserved in the visual representation of the component.
+We propose equivariance of monoid actions individually on the fiber to visual component maps and on the graphic as a whole.
 
@@
 \label{sec:data_base}
-The base space \dbase\ is way to express the connectivity of the records.
-This is assumed in the choice of visualization, for example a line plot implies 1D continuous data and a scatter plot implies discrete records, but an explicit representation allows for verifying that the topology of the graphic representation is equivalent to the topology of the data.
+The base space \dbase\ is way to express how the records in \dtotal\ are connected to each other, for example if they are discrete points or if they lie in a 2D continous surface.
+Connectivity type is assumed in the choice of visualization, for example a line plot implies 1D continuous data, but an explicit representation allows for verifying that the topology of the graphic representation is equivalent to the topology of the data.
 
@@
 By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.
-This in turn allows for building visualizations that can work with distributed or streaming data, since it has a common data access interface with a promise that the data exists.
+The explicit topology is a concise way of distinguishing visualizations that appear identical, for example heatmaps and images.
 
@@
 \label{sec:data_section}
-A set of records is a section $\dsection: \dbase\rightarrow \dtotal$ of the fiber bundle.
+The section $\dsection: \dbase\rightarrow \dtotal$ is what ties together the base space \dbase\ with the fiber \dfiber.
+A section is a function that takes as input location $\dbasepoint \in \dbase$ and returns a record $\delement \in \dtotal$.
 For example, in the special case of a table \cite{spivakSIMPLICIALDATABASES}, \dbase\ is a set of row ids, \dfiber\ is the columns, and the section \dsection\ returns the record \delement\ at a given key in \dbase.
@@
 
-The section can be any instance of a data container, for example a numpy array\cite{harris2020array}, a pandas series or dataframe\cite{jeff_reback_2020_3715232}, or an xarray\cite{hoyer2017xarray}.
-A univariate numpy array that stores an image is a section of a fiber bundle where \dbase\ is a 2D continuous plane and the \dfiber\ is $(\reals^{3}, \reals, \reals)$ where $\reals^3$ is color, and the other two components are the x and y positions of the sampled data in the image.
+This model provides a common interface to widely used data containers without sacrificing the semantic structure embedded in each container.
+For example, the section can be any instance of a univariate numpy array\cite{harris2020array} that stores an image.
+This could be a section of a fiber bundle where \dbase\ is a 2D continuous plane and the \dfiber\ is $(\reals^{3}, \reals, \reals)$ where $\reals^3$ is color, and the other two components are the x and y positions of the sampled data in the image.
+Instead of an image, the numpy array could also store a 2D discrete table.
+The fiber would not change, but the \dbase\ would now be 0D discrete points.
+These different choices in topology indicate, for example, what sorts of interpolation would be appropriate when visualizing the data.
+
+There are also many types of labeled containers that can richly be described in this framework because of the fiber.
+For example, a pandas series which stores a labeled list, or a dataframe\cite{jeff_reback_2020_3715232} which stores a relational table.
 A series could store the values of $\dsection^{(1)}$ and a second series could be  $\dsection^{(2)}$.
@@
 While the series and dataframe explicitly have a time index column, they are components in our model and the index is always assumed to be random keys.
-An instance of an xarray would also be a \dsection\, for example if the data is a temperature field than the \dbase\ could be a continuous volume and the components would be the temperature and the time, latitude, and longitude the measurements were sampled at.
-As with the dataframe, the semantic index labels are considered components and the indicies are instead assumed to be random.
+
+Where this model particularly shines are N dimensional labeled data structures.
+For example, an xarray\cite{hoyer2017xarray} data that stores temperature field could have a \dbase\ that is a continuous volume and the components would be the temperature and the time, latitude, and longitude the measurements were sampled at.
 A section can also be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.
@@
 Because our framework is defined in terms of the fiber, continuity, and sections, rather than the exact values of the data, our model does not need to know what the exact values are until the renderer needs to fill in the image.
-Our model provides a common interface to these widely used data containers without sacrificing the semantic structure embedded in each container.
 
@@
 The color of the pixel is then obtained by taking the integral of $\gsection_{rgb}(\gbase_p)$.
-In general, \gsection\ is an abstraction of rendering specifications, for example PDF\cite{bienz1993portable}, SVG\cite{quintScalable2003}, or an openGL scene graph\cite{CarsonOpenGL1997} or rendering engines such as cairo\cite{CairographicsOrg} and AGG\cite{AntiGrainGeometry}.
-Implementation of \gsection\ is out of scope for this work.
+
+In general, \gsection\ is an abstraction of rendering.
+In very broad strokes \gsection\  can be a specification such as PDF\cite{bienz1993portable}, SVG\cite{quintScalable2003}, or an openGL scene graph\cite{CarsonOpenGL1997}.
+Alternatively, \gsection can be a rendering engine such as cairo\cite{CairographicsOrg} or AGG\cite{AntiGrainGeometry}.
+Implementation of \gsection\ is out of scope for this work,
 
@@
 \label{sec:artist}
-
-The artist is the function that converts data into graphics; its name is taken from the analogous part of Matplotlib\cite{hunterArchitectureOpenSource} that builds visual elements to pass off to the renderer.
+We propose that the transformation from data to visual representation can be described as a structure preserving map from one topological space to another.
+We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements to pass off to the renderer.
 The topological artist \vartist\ is a monoid equivariant sheaf map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
@@
 In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the data bundle padded with the jet bundle $\dtotal^{\prime} = \dtotal + \mathcal{J}^{2}(\dtotal)$.
-Specifically, \vartist is the map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$
+
+Specifically, \vartist is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$
 
@@
 Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal (\ref{sec:artist_q}).
-This functional decomposition of the visualization artists allows us to specify what are the responsibilities of each function in a fully constrained way.
-In turn, this allows for building reusable components at each stage of the transformation.
+This functional decomposition of the visualization artist facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
 
 \subsubsection {Visual Fiber Bundle \vtotal}
-We introduce a visual bundle \vtotal\ to store the visual representations the artist needs to composite into a graphic.
+We introduce a visual bundle \vtotal\ to store the visual representations the artist needs to assemble into a graphic.
 The visual bundle (\vtotal, \dbase, $\pi$, \vfiber) has section $\vsection: \vtotal \rightarrow \dbase$ that resolves to a visual variable in the fiber \vfiber.
@@
 \end{figure}
-As introduced in section~\ref{sec:intro_visual_variables}, there are many ways to encode data visually. We define the visual transformers \vchannel\ 
+As introduced in section~\ref{sec:intro_viz}, there are many ways to visually represent data components. We define the visual transformers \vchannel\ 
 \begin{equation}
@@
 \end{equation}
-commutes such that $\vchannel_i(m_{\delement}(\dtotal_i)) = m_{\velement}(\vchannel_i(\dtotal_i))$.
+commutes such that 
+\begin{equation}
+\vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\velement})(\vchannel_i(\dtotal_i))\end{equation}
+
 This equivariance constraint yields guidance on what makes for an invalid transform.
-For example, the conversion $\vchannel_{i}(x) = .5$ does not commute under translation monoid action $t(x) = x+2$
+For example, given a visual fiber of the same type as the data fiber $\dfiber=\vfiber$, the transform $\vchannel_{i}(x) = .5$ does not commute under translation monoid action $t(x) = x+2$
 \begin{align}
@@
 The group action on these sets is permutation, so shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
-To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$, $\text{ if } delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$.
-For interval scale data, \vchannel\ is equivariant under translation monoid actions if $\vchannel (x + c) = \vchannel(x) + \vchannel(c)$.
-For ratio data, there must be equivalent scaling $ \vchannel(xc) = \vchannel(x)\vchannel(c)$.
-These constraints can be embedded into our artist such that the \vchannel\ functions are equivariant; they also provide guidance on constructing new equivariant \vchannel\ functions.
+Continuing with the assumption that $\dfiber=\vfiber$, we can describe the constraints on the other Steven's measurment group types.
+To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$
+\begin{equation}
+\text{ if } delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)
+\end{equation}
+the visual encodings must also have some sort of ordering. For interval scale data, \vchannel\ is equivariant under translation monoid actions if 
+\begin{equation}
+\vchannel (x + c) = \vchannel(x) + \vchannel(c)
+\end{equation}
+while for ratio data, there must be equivalent scaling 
+\begin{equation}
+\vchannel(xc) = \vchannel(x)\vchannel(c) 
+\end{equation}
+
+The assumption $\dfiber=\vfiber$ is made here so that we could omit  $\varphi: \monoid\rightarrow\monoid^{\prime}$ maps that convert the monoid action on \dfiber\ to the equivalent action on \vfiber.
+As shown in table~\ref{tab:mpl_visual_variable_fiber} though, it is not uncommon for data and visual variables to resolve to the same types.
+The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions are equivariant; they also provide guidance on constructing new equivariant \vchannel\ functions.
 
@@
 
-Lets call the visual encodings $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$.
-If for all monoids $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, the output is equivalent
-\begin{equation}
-\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
-\end{equation}
-then a group action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
-The transformed graphic $\gsection^{\prime}$ is equivariant to a transform on the visual bundle $\gsection^{\prime}=\vmark(m\circ \vsection)$ on a section that $\vsection \in \vmark^{-1}(\gsection)$ that must be part of generating \gsection.
-
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
-    \caption{These two glyphs are generated by the same \vmark\ function, but differ in the value of the edge thickness parameter $\vsection_i$.
-A valid \vmark\ is one where a shift in $\vsection_i$ is reflected in the glyph generated by \gsection.}
-    \label{fig:artist_mark_change}
-\end{figure}
-
-The glyph in figure~\ref{fig:artist_mark_change} has the following characteristics \vfiber specified by  $(xpos,\, ypos,\, color,\, thickness)$ such that one section is $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
-The equivariance constraint on \vmark\ is that the action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$.
-The corresponding action on \gsection causes $\vmark(\vsection^{\prime})$ to be the thicker circle in figure~\ref{fig:artist_mark_change}.
-
 We can formally describe a glyph as \vmark\, with the regions \dbasepoint that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
@@
 This is in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
-The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.
+The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs
+
+$\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$. If for all monoids $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, the output is equivalent 
+\begin{equation}
+\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
+\end{equation}
+then a group action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
+The transformed graphic $\gsection^{\prime}$ is equivariant to a transform on the visual bundle $\gsection^{\prime}=\vmark(m\circ \vsection)$ on a section that $\vsection \in \vmark^{-1}(\gsection)$ that must be part of generating \gsection.
+
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
+    \caption{These two glyphs are generated by the same \vmark\ function, but differ in the value of the edge thickness parameter $\vsection_i$.
+A valid \vmark\ is one where a shift in $\vsection_i$ is reflected in the glyph generated by \gsection.}
+    \label{fig:artist_mark_change}
+\end{figure}
+
+The glyph in figure~\ref{fig:artist_mark_change} has the following characteristics \vfiber\ specified by  $(xpos,\, ypos,\, color,\, thickness)$ such that one section is $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
+The equivariance constraint on \vmark\ is that the action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$.
+The corresponding action on \gsection causes $\vmark(\vsection^{\prime})$ to be the thicker circle in figure~\ref{fig:artist_mark_change}.
+
+.
 
@@
 \label{sec:artist_sheaf_stalk}
-The restriction maps of a sheaf describe how local \dsections can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
+The restriction maps of a sheaf describe how local \dsection can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
 As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$.

### 2021-02-25 draft/sections/math.tex
finished out math

@@
 
-Since we define the equivariant map as  $\vmark: \vsection \mapsto \gsection$, we define an action on the subset of graphics $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ that \vmark\ can generate.
-We then define the constraint on \vmark such that if \vmark\ is applied to $\vsection, \vsection^{\prime}$ that generate the same \gsection\, then the output of both sections acted on by the same monoid $m$ must be the same.
-
-We can formally describe a glyph as \vmark\, with the regions \dbasepoint that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
+While \gsection\ generates the entire graphic, we will restrict the discussion of \vmark\ to generation of sections of a glyph.
+We formally describe a glyph as \vmark\ applied to the regions \dbasepoint that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
 \begin{equation}
@@
 This is in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
-The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs
+The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.
 
-$\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$. If for all monoids $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, the output is equivalent 
+It is on sections of these glyphs that we define the equivariant map as  $\vmark: \vsection \mapsto \gsection$ and an action on the subset of graphics $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ that \vmark\ can generate.
+We then define the constraint on \vmark such that if \vmark\ is applied to $\vsection, \vsection^{\prime}$ that generate the same \gsection\, then the output of both sections acted on by the same monoid $m$ must be the same.
+While it may seem intuitive that visualizations that generate the same glyph should consistently generate the same glyph given the same input, we formalize this constraint such that it can be specified as part of the implementation of \vmark.
+
+Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$.
+If for all monoids $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, the output is equivalent
 \begin{equation}
@@
 
-.
-
 \subsubsection{Assembly \vmark}
-In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and a heatmap. 
+In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image. 
 \begin{figure}[H]
@@
 
-The line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet discussed in section~\ref{sec:sheaf_stalk}.
+The line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet.
 The line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
@@
 \begin{align}
- x = xpos(\vindex(\alpha)) &+ \beta\hat(n_1)(\vindex(\alpha)) \\
- y = ypos(\vindex(\alpha)) &+ \beta\hat(n_2)(\vindex(\alpha)) 
+ x = xpos(\vindex(\alpha)) &+ \beta\hat{n_1}(\vindex(\alpha)) \\
+ y = ypos(\vindex(\alpha)) &+ \beta\hat{n_2}(\vindex(\alpha)) 
 \end{align}
 
-where (x,y) look up the position $\vindex(\alpha)$ on the data and then apply thickness $\beta$ at that location.
+where (x,y) look up the position $\vindex(\alpha)$ on the data. At that point, we also look up the the derivatives $\hat{n_1}, \hat{n_2}$  which are then multiplied by data to specify the thickness.
 
@@
 \subsubsection{Assembly factory \vmarkd}
-The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib, because the rendering is tightly interlaced with the graphical design; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
+The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
 As shown in eq~\ref{eq:artist_diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
@@
 The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back $\vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber$ copies the visual fiber \vtotal over the preimage in \vpreimg.
-This is illustrated in figure~\ref{fig:artist_q_hat}, where the 1D fiber over \dbase\ is copied repeatedly to become the 2D fiber with identical components over \gbase.
-Given a section \vsectionpull\ pulled back from \vsection\ on $\pi:\vtotal\rightarrow\dbase$ and a point $\gbasepoint \in \vpreimg$ in the preimage of \dbasepoint\, the pulled back section $\vsectionpull(s) = \vindex^*(\vsection(\dbasepoint))$ is the image $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$.
+
+This is illustrated in figure~\ref{fig:artist_q_hat}, where the 1D fiber $\vfiber \hookrightarrow \vtotal$ over \dbase\ is copied repeatedly to become the 2D fiber $\vfiberpull \hookrightarrow \vtotalpull$ with identical components over \gbase.
+All the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase\ are in the preimage \vpreimg.
+Given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map from \vsection on a point \dbasepoint, there is a direct map from the visual section over data base space  $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$ to the visual section \vsectionpull\ over graphic base space.
+This map means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint.
 This means that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$, which is illustrated in figure~\ref{fig:artist_q_hat} as each dot on \vfiber is equivalent to the line intersection \vfiberpull.
@@
 Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out.
-When \vmark\ maps visual sections $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull then $\gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)$.
-Since  $\vsectionpull(s) = \vindex^*(\vsection(\dbasepoint))$ and \vsectionpull\ is constant on \vpreimg, we can define a \vmark\ factory function $\vmarkd(\vsection(k))(s) \coloneqq \vmark((\vsectionpull)(\gbasepoint))$ where $\vpreimg = \dbasepoint$.
+When \vmark\ maps visual sections  into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to the pulled back visual section \vsectionpull\ then
+\begin{equation}
+    $\gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)$
+\end{equation}
+the graphic section \rho\ evaluated on a visual region \gbasepoint\ is defined as the assembly function \vmark\ with input pulled back visual section \vsectionpull\ also evaluated on \gbasepoint.
+Since the pulled back visual section \vsectionpull\ is the visual section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function
+\begin{equation}
+\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
+\end{equation} 
+where the assembly function \vmarkd\ that takes as input the visual section on data \vsection\ is defined to be the assembly function \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gspace.
 
 Factoring out \gbasepoint\, $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ generates a curried \vmark.
-In fact, \vmarkd\ is a map from visua space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
+In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record   $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
 This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
@@
 Roughly, equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\ but act on different sections \vsection, but we will formalize the definition of the equivalance class in future work.
+
 As a first pass for implementation purposes, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data such that the graphic is identifiable as a given chart type.
-
 \begin{figure}[H]
@@
 
-For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another, as shown in figure~\ref{fig:artist_equivalance}.
+For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another, as shown in figure~\ref{fig:artist_equivalence}.
 These two artists are equivalent since their only difference is in the literal visual encodings (color, shape).

### 2021-02-25 draft/sections/math.tex
tweaking of monoid/fiber language

@@
 \end{equation}
-that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}.
-Monoid actions are a set of operations and are discussed in detail in section~\ref{sec:data_monoid}.
+that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}, which are discussed in detail in section~\ref{sec:data_monoid}.
 Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are symmetric
@@
 \end{equation}
-such that applying a monoid $m \in \monoid$ transformation to the input $\delement \in \mathscr{\dtotal}$ to $\mathscr{\vartist}$ is equivalent to applying a monoid $\varphi(\monoid) \in \monoid^{\prime}$ to the output of the artist $\vartist(\delement) \in \mathscr{\gtotal}$.
+such that applying a monoid action $m \in \monoid$ to the input $\delement \in \mathscr{\dtotal}$ to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the output of the artist $\vartist(\delement) \in \mathscr{\gtotal}$.
 
@@
 
-For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree K temperature measurements taken at those times.
+For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree^K temperature measurements taken at those times.
 Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$.
@@
     \begin{tabulary}{\textwidth}{|l|L|l|}\hline
-     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i})}$  \\ \hline
+     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i}) \subset \vfiber_{i}}$  \\ \hline
     position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
@@
 This equivariance constraint yields guidance on what makes for an invalid transform.
-For example, given a visual fiber of the same type as the data fiber $\dfiber=\vfiber$, the transform $\vchannel_{i}(x) = .5$ does not commute under translation monoid action $t(x) = x+2$
+For example, given a visual fiber of the same type as the data fiber $\dfiber_i=\vfiber_i$, the transform $\vchannel_{i}(x) = .5$ does not commute under translation $t(x) = x+2$
 \begin{align}
@@
 \end{equation}
-
-The assumption $\dfiber=\vfiber$ is made here so that we could omit  $\varphi: \monoid\rightarrow\monoid^{\prime}$ maps that convert the monoid action on \dfiber\ to the equivalent action on \vfiber.
-As shown in table~\ref{tab:mpl_visual_variable_fiber} though, it is not uncommon for data and visual variables to resolve to the same types.
-The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions are equivariant; they also provide guidance on constructing new equivariant \vchannel\ functions.
+The assumption $\dfiber_i=\vfiber_i$ is made here so that we could omit  $\varphi: \monoid\rightarrow\monoid^{\prime}$ maps that convert the monoid action on $\dfiber_{i}$ to the equivalent action on $\vfiber_{i}$.
+As shown in table~\ref{tab:mpl_visual_variable_fiber}, it is not uncommon for data and visual variables to resolve to the same types.
+The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions are equivariant and also provide guidance on constructing new equivariant \vchannel\ functions.
 
@@
 While \gsection\ generates the entire graphic, we will restrict the discussion of \vmark\ to generation of sections of a glyph.
-We formally describe a glyph as \vmark\ applied to the regions \dbasepoint that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
+We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
 \begin{equation}
@@
 
-The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back $\vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber$ copies the visual fiber \vtotal over the preimage in \vpreimg.
+The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back of \vsection\
+\begin{equation}
+    \vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber
+\end{equation}
+copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase\.
+This set of points \gbasepoint\ are the preimage \vpreimg\ of \dbasepoint.
 
-This is illustrated in figure~\ref{fig:artist_q_hat}, where the 1D fiber $\vfiber \hookrightarrow \vtotal$ over \dbase\ is copied repeatedly to become the 2D fiber $\vfiberpull \hookrightarrow \vtotalpull$ with identical components over \gbase.
-All the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase\ are in the preimage \vpreimg.
+This copying is illustrated in figure~\ref{fig:artist_q_hat}, where the 1D fiber $\vfiber \hookrightarrow \vtotal$ over \dbase\ is copied repeatedly to become the 2D fiber $\vfiberpull \hookrightarrow \vtotalpull$ with identical components over \gbase.
 Given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map from \vsection on a point \dbasepoint, there is a direct map from the visual section over data base space  $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$ to the visual section \vsectionpull\ over graphic base space.
@@
 \begin{equation}
-    $\gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)$
+    \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
 \end{equation}
-the graphic section \rho\ evaluated on a visual region \gbasepoint\ is defined as the assembly function \vmark\ with input pulled back visual section \vsectionpull\ also evaluated on \gbasepoint.
+the graphic section \gsection\ evaluated on a visual region \gbasepoint\ is defined as the assembly function \vmark\ with input pulled back visual section \vsectionpull\ also evaluated on \gbasepoint.
 Since the pulled back visual section \vsectionpull\ is the visual section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function

### 2021-02-25 code/matplottoy/artists/bar.py
bar transform fix, thanks @tacaswell

@@ -99,7 +99,7 @@ class StackedBar(martist.Artist):
             # pull out the specific group transforms
             group['floor'] = floor
             group.update(self.transforms)
-            bar = Bar(self.data, group, self.orientation)
+            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)
             self.children.append(bar)
             floor += view[group['length']['name']]
             
@@ -137,7 +137,7 @@ class GroupedBar(martist.Artist):
             width = group.get('width', .8)
             group['width'] = width/ngroups
             group['offset'] = gid/ngroups*width 
-            bar = Bar(self.data, group, self.orientation)     
+            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)     
             self.children.append(bar)
             
             

### 2021-02-25 draft/sections/implementation.tex
point & line assembly args

@@
     \end{subfigure}
-    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering. \note{update with bar}}
+    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
     \label{fig:code_scatter_line}
 \end{figure}
-
 To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.
@@
 The equivalence classes \vartisteq\ map well to Python classes since the functional aspects-\vchannel, \vmarkd, and \vindex - are completely reusable in a consistent composition, while the visual values in \vtotal\ are what change between different artists belonging to the same class \vartisteq.
-The \mintinline{python}{data} object is an abstraction of a data bundle \dtotal\ with a specified section \tau.
+The \mintinline{python}{data} object is an abstraction of a data bundle \dtotal\ with a specified section \dsection.
 Implementing \gtotal\ and \gsection\ are out of scope for this prototype because they are part of the rendering process.
-We also did not implement any form of \vindex\ because the scatter, line, and bar plots prototyped here directly broadcast from \dbasepoint to \gbasepoint, unlike for example an image which may need to be rotated.
+We also did not implement any form of \vindex\ because the scatter, line, and bar plots prototyped here directly broadcast from \dbasepoint\ to \gbasepoint, unlike for example an image which may need to be rotated.
 
@@
 
-    def assemble(self, visual):
+    def assemble(self, **args):
         # set the properties of the graphic
@@
         # visual channel encoding applied fiberwise 
-        visual = {p: encoder(view[f] if f is not None else None) 
+        visual = {p: encoder(view[f] if f in self.data.FB.F else f) 
                   for p, (f, encoder) in self.transforms.items()}
-        self.assemble(visual)
+        self.assemble(**visual)
         # pass configurations off to the renderer
@@
 The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}.
-We allow for fixed visual parameter, such as setting a constant color for all sections, via setting in \mintinline{python}{None} as the data fiber \dfiber\ name in the \mintinline{python}{transform} dictionary.
+We allow for fixed visual parameter, such as setting a constant color for all sections, by verifying that the named component is in \dfiber\ before accessing the data.
+If the data component name is not in \dfiber\, this is interpreted to mean this component is a thickening of \vtotal that could be pulled back to \dtotal\ via an inverse identity \vchannel.
+The conversion from data to visual space is simplified here to show that it is the encoding \vchannel\ applied to the component.
 
-The visual object is then passed into the \mintinline{python}{assembly} function that is \vmarkd.
+The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assembly} function that is \vmarkd.
 This assembly function is responsible for generating a representation of the glyph such that it could be serialized to recreate a static version of the graphic.
@@
 \begin{minted}{python}
-class Point(mcollections.Collection):=
-    def assemble(self, visual):
-        # construct geometries of the circle marks in visual coordinates
-        self._paths = [mpath.Path.circle(center=(x,y), radius=s) 
-                       for (x, y, s) in zip(visual['x'],visual['y'], visual['s'])] 
-        # set attributes of marks, these are vectorized 
+class Point(mcollections.Collection):
+    def assemble(self, x, y, s, facecolors='C0' ):
+        # construct geometries of the circle glyphs in visual coordinates
+        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
+                    for (xi, yi, si) in zip(x, y, s)] 
+        # set attributes of glyphs, these are vectorized 
         # circles and facecolors are lists of the same size
-        self.set_facecolors(visual['facecolors'])
-
+        self.set_facecolors(facecolors)
 \end{minted} 
@@
 We put this constraint on the return of the \mintedinline{python}{view} method so that we do not risk race conditions.
-Using triangulization provides a common interface to the data, one we will reuse for all the artists presented here.
 
 This table is converted to a table of visual variables and is then passed into \mintinline{python}{assemble}.
-In  \mintinline{python}{assemble}, the \vsection\ is used to individually construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
+In  \mintinline{python}{assemble}, the \vsection\ components are used to construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
 This is done via the \mintinline{python}{Path.circle} object.
@@
 
-The main difference between the \mintinline{python}{Point} and \mintinline{python}{Line} objects is in th3 \mintinline{python}{assemble} function because line has different continuity from scatter and is represented by a different type of graphical mark.
+The main difference between the \mintinline{python}{Point} and \mintinline{python}{Line} objects is in the \mintinline{python}{assemble} function because line has different continuity from scatter and is represented by a different type of graphical mark.
 
@@
 class Line(mcollections.LineCollection):
-    def assemble(self, visual):
+    def assemble(self, x, y, color='C0'):
             #assemble line marks as set of segments 
             segments = [np.vstack((vx, vy)).T for vx, vy 
-                    in zip(visual['x'], visual['y'])]
+                        in zip(x, y)]
             self.set_segments(segments)
-        self.set_color(visual['color'])
+            self.set_color(color)
+
 \end{minted}
@@
 The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist.
-The have the same required \vfiber\ components of (position, length).
-In \mintinline{pyton}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally and only applies holistically to the graphic and never to individual data parameters.
-This highlights another advantage of this model, that it encourages explicit differentiation between parameters in \vtotal\ and graphic parameters that specify \vmarkd\.
+The have the same required \vfiber\ components of (position, length), but in \mintinline{pyton}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally.
+This paramter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd\.
 

### 2021-02-25 draft/sections/implementation.tex
reworking of bar visual transforms a bit

@@
         # visual channel encoding applied fiberwise 
-        visual = {p: encoder(view[f] if f in self.data.FB.F else f) 
-                  for p, (f, encoder) in self.transforms.items()}
+        visual = {p: t['encoder'](view[t['name']])
+                  for p, t in self.transforms.items()}
         self.assemble(**visual)
@@
 The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}.
-We allow for fixed visual parameter, such as setting a constant color for all sections, by verifying that the named component is in \dfiber\ before accessing the data.
+The conversion from data to visual space is simplified here to directly show that it is the encoding \vchannel\ applied to the component.
+In the full implementation, we allow for fixed visual parameter, such as setting a constant color for all sections, by verifying that the named component is in \dfiber\ before accessing the data.
 If the data component name is not in \dfiber\, this is interpreted to mean this component is a thickening of \vtotal that could be pulled back to \dtotal\ via an inverse identity \vchannel.
-The conversion from data to visual space is simplified here to show that it is the encoding \vchannel\ applied to the component.
 
@@
             self.set_color(color)
-
 \end{minted}
@@
 The have the same required \vfiber\ components of (position, length), but in \mintinline{pyton}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally.
-This paramter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd\.
+This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd\.
 
 \begin{minted}{python}
-class Bar(mcollections.Collection):
-    def __init__(self, data, transforms, orientation = 'v', *args, **kwargs):
-        # parameter of the graphic
-        self.orientation = orientation 
-        super().__init__(*args, **kwargs)
-        self.data = data
-        self.transforms = transforms
-
-    def assemble(self, visual):
-        #set some defaults
-        visual.setdefault('width', itertools.repeat(0.8))
-        visual.setdefault('floor', itertools.repeat(0))
-        visual.setdefault('facecolors', 'C0')
-        #build bar glyphs based on graphic parameter
-        if self.orientation in {'vertical', 'v'}:
-            order = ['position', 'width', 'floor', 'length']
-        elif self.orientation in {'horizontal', 'h'}:
-            order = ['floor', 'length', 'position', 'width']
-
-        verts = [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
-                for (x, xo, y, yo) in zip(*[visual[k] for k in order])]
-
-        self._paths = [mpath.Path(xy, closed=True) for xy in verts]
-        self.set_edgecolors('k')
-        self.set_facecolors(visual['facecolors'])
 
 \end{minted}
-The \mintinline{python}{draw} method is the same as the one in the artist example~\ref{code:artist} so is ommitted here.
+The \mintinline{python}{draw} method here has a more complex unpacking of visual encodings to support passing in visual component data directly.
+This is necessary to build composite artists out ouf of this bar artists.
+
 This bar underpins a rudimentary version of the complex \plus operator such that we can specify multiple components to map to the same visual components without requiring th components to be individually specified in the \mintinline{python}{transforms} dictionary.

### 2021-02-25 code/matplottoy/artists/bar.py

@@ -64,14 +64,16 @@ class Bar(mcollections.Collection):
     def draw(self, renderer,  *args, **kwargs):
         view = self.data.view(self.axes)
         visual = {}
-        for (p, trans) in self.transforms.items():
-            if isinstance(trans, dict):
-                if "name" in trans:
-                    visual[p] =  view[trans['name']]
-                if "encoder" in trans:
-                    visual[p] = trans['encoder'](visual[p])
+        for (p, t) in self.transforms.items():
+            if isinstance(t, dict):
+                if t['name'] in self.data.FB.F and 'encoder' in t:
+                    visual[p] = t['encoder'](view[t['name']])
+                elif 'encoder' in t:
+                    visual[p] = t['encoder'](t['name'])
+                elif t['name'] in self.data.FB.F:
+                    visual[p] = view[t['name']]
             else:
-                 visual[p] = trans
+                 visual[p] = t
         self.assemble(**visual)
         super().draw(renderer,  *args, **kwargs)
 

### 2021-02-25 code/matplottoy/artists/line.py

@@ -28,18 +28,18 @@ class Line(mcollections.LineCollection):
         self.data = data
         self.transforms = transforms
 
-    def assemble(self, visual):
+    def assemble(self, x, y, color='C0'):
         #assemble line marks as set of segments 
         segments = [np.vstack((vx, vy)).T for vx, vy 
-                    in zip(visual['x'], visual['y'])]
+                    in zip(x, y)]
         self.set_segments(segments)
-        self.set_color(visual['color'])
+        self.set_color(color)
         
     def draw(self, renderer):
         # query data source for edge table
         view = self.data.view('edge')
-        visual = {p: encoder(view.get(f, None)) for 
-                     p, (f, encoder) in self.transforms.items()}
-        self.assemble(visual)
+        visual = {p: encoder(view[f] if f in self.data.FB.F else f) 
+                     for p, (f, encoder) in self.transforms.items()}
+        self.assemble(**visual)
         super().draw(renderer)
 

### 2021-02-25 code/matplottoy/artists/point.py

@@ -30,18 +30,20 @@ class Point(mcollections.Collection):
         self.data = data
         self.transforms = transforms
 
-    def assemble(self, visual):
+    def assemble(self, x, y, s, facecolors='C0' ):
         # construct geometries of the circle glyphs in visual coordinates
-        self._paths = [mpath.Path.circle(center=(x,y), radius=s) for (x, y, s) 
-                in zip(visual['x'],visual['y'], visual['s'])] 
+        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
+                       for (xi, yi, si) in zip(x, y, s)] 
         # set attributes of glyphs, these are vectorized 
         # circles and facecolors are lists of the same size
-        self.set_facecolors(visual['facecolors'])
+        self.set_facecolors(facecolors)
         
     def draw(self, renderer):
         # query data for a vertex table K
         view = self.data.view(self.axes) 
-        visual = {p: encoder(view[f] if f is not None else None) for p, (f, encoder) in self.transforms.items()}
-        self.assemble(visual)
+        visual = {p: t['encoder'](view[t['name']] 
+                  if t['name'] in self.data.FB.F else t['name']) 
+                  for p, t in self.transforms.items()}
+        self.assemble(**visual)
         # call the renderer that will draw based on properties
         super().draw(renderer)

### 2021-02-25 code/matplottoy/data/iris.py

@@ -3,6 +3,8 @@ from matplottoy.encoders import mtypes
 
 class DataFrame:
     def __init__(self, dataframe):
+        self.FB = FiberBundle(K = {'tables':['vertex']},
+                              F = dict(dataframe.dtypes))
         self._tau = dataframe.iloc
         self._view = dataframe
 

### 2021-02-25 draft/sections/implementation.tex
nominally done (no figure) but won't compile

@@
 This means that the values cannot change after the method is called in draw until a new call in draw.
-We put this constraint on the return of the \mintedinline{python}{view} method so that we do not risk race conditions.
+We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.
 
@@
 \begin{minted}{python}
+class Bar(mcollections.Collection):
+    def __init__(self, data, transforms, orientation='v', *args, **kwargs):
+        """
+        orientation: str, optional
+            v: bars aligned along x axis, heights on y
+            h: bars aligned along y axis, heights on x
+        """
+        self.orientation = orientation
+        super().__init__(*args, **kwargs)
+        self.data = data
+        self.transforms = copy.deepcopy(transforms)
+
+    def assemble(self, position, length, floor=0, width=0.8, facecolors='C0', edgecolors='k', offset=0):
+        #set some defaults
+        width = itertools.repeat(width) if np.isscalar(width) else width
+        floor = itertools.repeat(floor) if np.isscalar(floor) else (floor)
+
+        # offset is passed through via assemblers such as multigroup, not supposed to be directly tagged to position 
+        position = position + offset
+
+        def make_bars(xval, xoff, yval, yoff):
+             return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
+                for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
+        #build bar glyphs based on graphic parameter
+        if self.orientation in {'vertical', 'v'}:
+            verts = make_bars(position, width, floor, length)
+        elif self.orientation in {'horizontal', 'h'}:
+            verts = make_bars(floor, length, position, width)
+
+        self._paths = [mpath.Path(xy, closed=True) for xy in verts]
+        self.set_edgecolors(edgecolors)
+        self.set_facecolors(facecolors)
+
+    def draw(self, renderer,  *args, **kwargs):
+        view = self.data.view(self.axes)
+        visual = {}
+        for (p, t) in self.transforms.items():
+            if isinstance(t, dict):
+                if t['name'] in self.data.FB.F and 'encoder' in t:
+                    visual[p] = t['encoder'](view[t['name']])
+                elif 'encoder' in t: # constant value
+                    visual[p] = t['encoder'](t['name'])
+                elif t['name'] in self.data.FB.F: # identity 
+                    visual[p] = view[t['name']]
+            else: # no transform 
+                 visual[p] = t
+        self.assemble(**visual)
+        super().draw(renderer,  *args, **kwargs)
 
@@
 The \mintinline{python}{draw} method here has a more complex unpacking of visual encodings to support passing in visual component data directly.
-This is necessary to build composite artists out ouf of this bar artists.
-
-This bar underpins a rudimentary version of the complex \plus operator such that we can specify multiple components to map to the same visual components without requiring th components to be individually specified in the \mintinline{python}{transforms} dictionary.
-The implementation of this support is slightly more complicated than the simple case, but not in a way that is useful for understanding the framework.
-The \mintinline{python}{assemble} function constructs bars and sets their edge color to black.
+This is vastly simplifies building composite objects as the alternative would be higher order functions that take as input the transforms passed in by the user.
+This construction supports a constant visual parameter, an identity transform where the value is the same in \dtotal\ and \vtotal\, and setting the visual component directly.
+The \mintinline{python}{assemble} function constructs bars and sets their face and edge colors.
+The \mintinline{python}{make_bars} function converts the input position and length to the coordinates of a rectangle of the given width.
 Defaults are provided for 'width' and 'floor' to make this function more reusable.
@@
 cmap =  color.Categorical({'true':'deeppink', 'false':'deepskyblue'})
-transforms = {'y': ('v1', lambda x: x),       
-              'x': ('v3', lambda x: x),
-             'facecolors': ('v2', cmap), 
-             's':(None ,lambda _: itertools.repeat(.02))}
+transforms = {'x': {'name': 'v4', 'encoder': lambda x: x},
+              'y': {'name': 'v2', 'encoder': lambda x: x},
+              'facecolors': {'name':'v3', 'encoder': cmap}, 
+              's':{'name': None , 'encoder': lambda _: itertools.repeat(.02)}}
 \end{minted}
@@
 While size could be embedded inside the \mintinline{python}{assembly} function, it is added to the transformers to illustrate user configured visual parameters that could either be constant or mapped to a component in \dfiber.
-The identity and constant \vchannel\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they could be optimized away.
+The identity and constant \vchannel\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they are somewhat optimized away in \mintinline{python}{Bar}.
 More complex encoders can be implemented as callable classes, such as
-
 \begin{minted}{python}
@@
 In this example, \mintinline{python}{is_nominal} checks for equivariance of permutation group actions by applying the encoder to a set of values, shuffling values, and checking that (value, encoding) pairs remain the same.
-This equivariance test can be implemented as part of the artist or encoder, but for minimal overhead, the equivariant it is implemented as part of the library tests.
 
@@
 The view method returns a dictionary where the key is a fiber component name and the value is a list of values in the fiber component.
-The table is built one call to \mintinline{python}{tau} at a time, guaranteeing that all the fiber component values are over the same \dbasepoint.
+The table is built one call to the section method \mintinline{python}{tau} at a time, guaranteeing that all the fiber component values are over the same \dbasepoint.
 Table has a \mintinline{python}{get} method as it is a method on Python dictionaries.
@@
 \end{minted}
-Unlike scatter, the line \mintinline{python}{tau} method returns the functions on the edge evaluated on the interval [0,1].
+Unlike scatter, the line section method \mintinline{python}{tau} returns the functions on the edge evaluated on the interval [0,1].
 By default these means each \mintinline{python}{tau} returns a list of 1000 x and y points and the associated color.
@@
 For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020} since it is multivariate and has a varying number of penguins.
-We use a version of the data packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021, jeff_reback_2020_3715232} since that is a very commonly used Python labled data structure.
-The wrapper is very thin since here there is explicitly only one section.
+We use a version of the data packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.
+The wrapper is very thin because there is explicitly only one section.
 \begin{minted}{python}
-class DataFrameSection:
+class DataFrame:
     def __init__(self, dataframe):
+        self.FB = FiberBundle(K = {'tables':['vertex']},
+                              F = dict(dataframe.dtypes))
         self._tau = dataframe.iloc
         self._view = dataframe
-    def view(self, axes):
+
+    def view(self, axes=None):
         return self._view
 \end{minted}
-The pandas indexer is a key valued set of discrete vertices, so there is no need to repackage for triangulation.
+Since the aim for this wrapper is to be very generic, here the fiber is set by querying the dataframe for its metadata.
+The \mintinline{python}{dtypes} are a list of column names and the datatype of the values in each column; this is the minimal amount of information the model requires to verify constraints.
+The pandas indexer is a key valued set of discrete vertices, so there is no need to repackage for the data interface.
 
@@
 
-For example, the \mintinline{python}{MultiBar} artist that makes figure~\ref{fig:code_bar_multi} reuses \mintinline{python}{_make_bars} but does not reuse the assemble function because the composition of elements forces fundemental differences in glyph construction.
-As demonstrated in the \mintinline{python}{init}, the composite bar chart has orientation and whether it is stacked or not.
-While the stacked bar chart and the grouped bar chart could be seperate artists, as demonstrated they share so much overlapping code that it is far less redundant to implement them together.
-\note{looking at the mess that is this code, I'm a) not convinced these should be combined b) no longer convinced this provides anything over just bar if it isn't rewritten to use bar more}
+The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both out of \mintinline{python}{Bar} artists such that the difference between \mintinline{Python}{StackedBar} and \mintinline{Python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are arranged differently.
+These two artists have identical \mintinline{{python}{__init__} and \mintinline{python}{draw} methods.
+As with \mintinline{python}{Bar}, the orientation is set on instantiation.
+In both these artists, we separate the transforms applied to only one component \mintinline{python}{transforms} and the case \mintinline{python}{mtransforms} where the same transform is applied to multiple components such that \votal\ has multiple components that map to the same retinal variable.
+\begin{minted}{python}
+class StackedBar(martist.Artist):
+    def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
+        """
+        Parameters
+        ----------
+
+        orientation: str, optional
+            vertical: bars aligned along x axis, heights on y
+            horizontal: bars aligned along y axis, heights on x   
+        """
+        super().__init__(*args, **kwargs)
+        self.data = data
+        self.orientation = orientation
+        self.transforms = copy.deepcopy(transforms)
+        self.mtransforms = copy.deepcopy(mtransforms)
+
+    def assemble(self):
+        view = self.data.view(self.axes)
+        self.children = [] # list of bars to be rendered
+        floor = 0
+        for group in self.mtransforms:
+            # pull out the specific group transforms
+            group['floor'] = floor
+            group.update(self.transforms)
+            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)
+            self.children.append(bar)
+            floor += view[group['length']['name']]
+
+    def draw(self, renderer, *args, **kwargs):
+        # all the visual conversion gets pushed to child artists
+        self.assemble()
+        #self._transform = self.children[0].get_transform()
+        for artist in self.children:
+            artist.draw(renderer, *args, **kwargs)
+
+\end{minted}
+Since all the visual transformation is passed through to \mintinline{python}{Bar}, the \mintinline{python}{draw} method does not do any visual transformations.
+In \mintinline{Python}{StackedBar} the \mintinline{python}{view} is used to adjust the \mintinline{python}{floor} for every subsequent bar chart since a stacked bar chart is bar chart area marks concatenated together in the length parameter.
+In contrast, \mintinline{Python}{GroupedBar} does not even need the view, but instead keeps track of the relative position of each group of bars in the visual only variable \mintinline{python}{offset}.
 
 \begin{minted}{python}
-class MultiBar(mcollections.Collection):
-    def __init__(self, data, transforms, *args, **kwargs):
-        #set the orientation of the graphic  
-        self.orientation = kwargs.pop('orientation', 'v')
-        # set how the bar glyphs are put together to create the graphic
-        self.stacked = kwargs.pop('stacked', False)     
-        # rest is same as other artist __init__s
+class GroupedBar(martist.Artist):
+    def assemble(self):
+        self.children = [] # list of bars to be rendered
+        ngroups = len(self.mtransforms)
 
-        #this needs to be factored out but just want to finish now
-        self.width = kwargs.pop('width', .8)
+        for gid, group in enumerate(self.mtransforms):
+            group.update(self.transforms)
+            width = group.get('width', .8)
+            group['width'] = width/ngroups
+            group['offset'] = gid/ngroups*width 
+            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)     
+            self.children.append(bar)
+\end{minted}{python}
 
-    def assemble(self, visual, view):
-        (groups, gencoder) = self.transforms['length']
-        ngroups = len(np.atleast_1d(groups))
-        visual['floor'] = visual.get('floor', np.empty(len(view[groups[0]])))
-        visual['facecolors'] = visual.get('facecolors', 'C0')
-        # make equal width stacked columns 
-        if 'width' not in visual and self.stacked:
-            visual['width'] = itertools.repeat(self.width)
+Since the only difference between these two glyphs is in the composition of \mintinline{python}{Bar}, they take in the exact same transform specification dictionaries.
+\mintinline{python}{transform} dictates the position of the group, in this case by island the penguins are found on.
 
-        # make equal with groups
-        if not self.stacked:
-            visual['width'] = itertools.repeat(self.width/ngroups)
-            offset = (np.arange(ngroups) /ngroups) * self.width
-        else:
-            offset = itertools.repeat(0)
+\begin{minted}{python}
+transforms = {'position': {'name':'island',
+              'encoder': position.Nominal({'Biscoe':0.1, 'Dream':1.1, 'Torgersen':2.1})}} 
+group_transforms =  [{'length': {'name':'Adelie'},
+                      'facecolors': {'name':"Adelie_s", 'encoder':cmap}},
+                     {'length': {'name':'Chinstrap'},
+                      'facecolors': {'name':"Chinstrap_s", 'encoder':cmap}}, 
+                     {'length': {'name':'Gentoo'},
+                      'facecolors': {'name':"Gentoo_s", 'encoder':cmap}}]
+\end{minted}
+\mintinline{python}{group_transforms} describes the group, and takes a list of dictionaries where each dictionary is the aesthetics of each group.
+That \mintinline{python}{position} and \mintinline{python}{length} are required parameters is enforced in the creation of the \mintinline{python}{Bar} artist.
+These means that these two artists have identical function signatures
 
-        # make the bars and arrange them 
-        verts = []
-        for group, off in zip(groups, offset):
-            verts.extend(Bar._make_bars(self.orientation, visual['position'] + off, 
-                          visual['width'], visual['floor'], view[group]))
-            if self.stacked: #add stacked bar to previous bar
-                visual['floor'] += view[group]
-
-        # convert lengths after all calculations are made and reorient if needed
-        # here or in transform machinery?
-        if self.orientation in {'v', 'vertical'}:
-            tverts = [[(x, gencoder(y)) for (x, y) in vert] 
-                            for vert in verts]
-        elif self.orientation in {'h', 'horizontal'}:
-            tverts = [[(gencoder(x), y) for (x, y) in vert] 
-                            for vert in verts]
-        self._paths = [mpath.Path(xy, closed=True) for xy in tverts]
-        #flatted columns of colors to match list of bars
-        self.set_facecolor(list(itertools.chain.from_iterable(visual['facecolors'])))
-        self.set_edgecolors('k')
-
-    def draw(self, renderer, *args, **kwargs):
-        view = self.data.view()
-        #exclude converting the group visual length, special cased in assemble
-        visual = utils.convert_transforms(view, self.transforms, exclude=['length'])
-        # pass in view because nu is not distributable so may need to apply it 
-        # after visual assembly
-        self.assemble(visual, view)
-        super().draw(renderer, *args, **kwargs)
-        return      
+\begin{minted}{python}
+artistSB = bar.StackedBar(bt, ts, group_transforms)
+artistGB = bar.GroupedBar(bt, ts, group_transforms)
 \end{minted}
 
-In the \mintinline{python}{__draw__}, a utility function is used for conversions, but the length transforms are held until after assembly because the length is computed by adding the current length to the previous and many transforms are not distributable such that $\nu(x_0 + x_1 + x_2) = \nu(x_0) + \nu(x_1) + \nu(x_2)$.
-Inside \mintinline{python}{assemble}, the glyphs are either shifted vertically (\texttt{stacked}) or horizontally (\texttt{grouped}) such that the positions are recorded and added to with the next group.
-This function allows multiple columns to be mapped to a visual parameter, but it must be equal numbers of columns
-\begin{minted}{python}
-{'position': ('island', lambda x: {'Biscoe':0, 'Dream':1, 'Torgersen':2}[x]),
- 'length':(['Adelie', 'Chinstrap', 'Gentoo'], lambda x: x),
- 'facecolors': (['Adelie_s', 'Chinstrap_s', 'Gentoo_s'], 
-        color.Categorical({'Adelie':'#FF8C00',
-                           'Gentoo':'#159090', 
-                           'Chinstrap':'#A034F0'}))}
-\end{minted}
-such as in this example where for each column contributing to a segment of the bar there is a corresponding column of colors for this segment.
-The reason the multibar can work with such a trasnformer is because it is relying on the data model to do most of the bookkeeping of which values get mapped to which bars.
-This also yields a much simpler function call to the artist
-
-\begin{minted}{python}
-fig, ax = plt.subplots()
-artist = bar.MultiBar(table, trans, orientation='h', stacked=True)
-ax.add_artist(artist)
-\end{minted}
-
-where \mintinline{python}{trans} is the same dictionary for both stacked and grouped version, as is the \mintinline{python}{DataFrameSection} object \mintinline{python}{table}.
-The only difference between the two versions is the \texttt{stacked} flag, and the only difference between figures~\ref{fig:code_bar_simple} is the \texttt{orientation} argument.
+but differ in assembly \vmarkd.
 By decomposing the architecture into data, visual encoding, and assembly steps, we are able to build components that are more flexible and also more self contained than the existing code base.
-
-\subsection{Summary}
-In general, the way in which we implemented the artist is as follows: 
-\begin{description}
-    \item[\textbf{\vchannel}] encoder function converting data to library normalized form, can also be \mintinline{python}{Encoder.__call__} method
-    \item[\textbf{\vsection}]  dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| curried until {Artist.draw}
-    \item[\textbf{\vartisteq}] objects that take in as input \dtotal\ and \vtotal\, for example \mintinline{python}{Point}, \mintinline{python}{Line}, \mintinline{python}{bar}
-    \item[\textbf{\vartist}] \mintinline{python}{Artist.draw} method which is \vartisteq\ parameterized by \vtotal
-    \item[\textbf{\vmarkd}] \mintinline{python}{Artist.assemble(visual)} method that arranges components of \vtotal\ into glyphs  
-\end{description}
-
-with the data representation implemented as 
-\begin{description}
-    \item[\textbf{\dtotal}] object with \mintinline{python}{view} method 
-    \item[\textbf{\dfiber}] named components of the data and their types
-    \item[\textbf{\dbase}] determines how the data is bound together when returned by view  
-    \item[\textbf{\dsection}] \mintinline{python}{Data.view()} method that returns component sections
-\end{description}
-
 While very rough, this API demonstrates that the ideas presented in the math framework are implementable.
-In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\ and assembly \vmarkd\.
+In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\, assembly \vmarkd and artists \vartist.
 We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic.

### 2021-02-25 code/matplottoy/artists/bar.py

@@ -93,7 +93,8 @@ class StackedBar(martist.Artist):
         self.transforms = copy.deepcopy(transforms)
         self.mtransforms = copy.deepcopy(mtransforms)
 
-    def assemble(self, view):
+    def assemble(self):
+        view = self.data.view(self.axes)
         self.children = [] # list of bars to be rendered
         floor = 0
   
@@ -107,9 +108,9 @@ class StackedBar(martist.Artist):
             
             
     def draw(self, renderer, *args, **kwargs):
-        view = self.data.view(self.axes)
+
         # all the visual conversion gets pushed to child artists
-        self.assemble(view)
+        self.assemble()
         #self._transform = self.children[0].get_transform()
         for artist in self.children:
             artist.draw(renderer, *args, **kwargs)
@@ -130,13 +131,13 @@ class GroupedBar(martist.Artist):
         self.transforms = copy.deepcopy(transforms)
         self.mtransforms = copy.deepcopy(mtransforms)
 
-    def assemble(self, view):
+    def assemble(self):
         self.children = [] # list of bars to be rendered
         ngroups = len(self.mtransforms)
         
         for gid, group in enumerate(self.mtransforms):
             group.update(self.transforms)
-            width = group.get('width', .8)
+            width = group.get('width',6)
             group['width'] = width/ngroups
             group['offset'] = gid/ngroups*width 
             bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)     
@@ -146,7 +147,7 @@ class GroupedBar(martist.Artist):
     def draw(self, renderer, *args, **kwargs):
         view = self.data.view(self.axes)
         # all the visual conversion gets pushed to child artists
-        self.assemble(view)
+        self.assemble()
         #self._transform = self.children[0].get_transform()
         for artist in self.children:
             artist.draw(renderer, *args, **kwargs)

### 2021-02-26 draft/sections/implementation.tex
latex errors + cite tweaks

@@
 We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations.
-
 To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primitive objects we implemented based on our topology model.
@@
 
-\begin{minted}[escapeinsize=||]{python}
+\begin{minted}{python}
 class ArtistClass(matplotlib.artist.Artist):
-    |\label{code:artist}|
     def __init__(self, data, transforms, *args, **kwargs):
@@
 The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
-The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that we can use to query back into data to select a subset as discussed in section~\ref{sec:data_sheaf_stalk}.
+The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that we can use to query back into data to select a subset as discussed in section~\ref{sec:artist_sheaf_stalk}.
 The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}.
@@
 
-The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assembly} function that is \vmarkd.
+The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assemble} function that is \vmarkd.
 This assembly function is responsible for generating a representation of the glyph such that it could be serialized to recreate a static version of the graphic.
@@
 In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker.
-We only show the \mintinline{python}{assembly} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist shown in \ref{code:artist}.
+We only show the \mintinline{python}{assemble} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
 \begin{minted}{python}
@@
 \end{minted} 
-The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices, as described in section~\ref{sec:triangulization}.
+The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices.
 Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$.
@@
 The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist.
-The have the same required \vfiber\ components of (position, length), but in \mintinline{pyton}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally.
-This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd\.
+The have the same required \vfiber\ components of (position, length), but in \mintinline{python}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally.
+This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd.
 
@@
 This is vastly simplifies building composite objects as the alternative would be higher order functions that take as input the transforms passed in by the user.
-This construction supports a constant visual parameter, an identity transform where the value is the same in \dtotal\ and \vtotal\, and setting the visual component directly.
+This construction supports a constant visual parameter, an identity transform where the value is the same in \dtotal\ and \vtotal, and setting the visual component directly.
 The \mintinline{python}{assemble} function constructs bars and sets their face and edge colors.
@@
 The size \textit{s} transformer is not acting on a component of \dfiber, instead it is a \vchannel\ that returns a constant value.
-While size could be embedded inside the \mintinline{python}{assembly} function, it is added to the transformers to illustrate user configured visual parameters that could either be constant or mapped to a component in \dfiber.
+While size could be embedded inside the \mintinline{python}{assemble} function, it is added to the transformers to illustrate user configured visual parameters that could either be constant or mapped to a component in \dfiber.
 The identity and constant \vchannel\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they are somewhat optimized away in \mintinline{python}{Bar}.
@@
 In this example, \mintinline{python}{is_nominal} checks for equivariance of permutation group actions by applying the encoder to a set of values, shuffling values, and checking that (value, encoding) pairs remain the same.
-
 \subsection{Data \dtotal}
 \label{sec:code_data}
-The data input into the \mintinline{pythin}{Artist} will often be a wrapper class around an existing data structure.
-This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{view} method that returns an atomic object that encapsulates \dsection\.
+The data input into the \mintinline{python}{Artist} will often be a wrapper class around an existing data structure.
+This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{python}{view} method that returns an atomic object that encapsulates \dsection.
 The object returned by the view must be key valued pairs of \mintinline{python}|{component name : component section}| where each section is a component as defined in equation~\ref{eq:data_section_decompose}.
@@
 Python dataclasses are a good abstraction for the fiber bundle class because the \mintinline{python}{FiberBundle} class only stores data.
-The \dbase\ is specified as tables because the \mintinline{python}{assembly} functions expect tables that match the continuity of the graphic; scatter expects a vertex table because it is discontinuous, line expects an edge table because it is 1D continuous.
+The \dbase\ is specified as tables because the \mintinline{python}{assemble} functions expect tables that match the continuity of the graphic; scatter expects a vertex table because it is discontinuous, line expects an edge table because it is 1D continuous.
 The fiber informs appropriate choice of \vchannel\, therefore it is a dictionary of attributes of the fiber components.
@@
 To generate the scatter plot in figure~\ref{fig:code_scatter_line}, we fully specify a dataset with random keys and values in a section chosen at random form the corresponding fiber component.
-The fiberbundle \mintinline{python}{FB} is a class level attribute since all instances of \mintinline{python}{code}{VertexSimplex} come from the same fiberbundle.
+The fiberbundle \mintinline{python}{FB} is a class level attribute since all instances of \mintinline{python}{VertexSimplex} come from the same fiberbundle.
 \begin{minted}{python}
@@
     \end{subfigure}
-    \caption{Continuous and discontinuous lines as defined by different data models, but generated with the same \vartisteq=\mintinline{python}{artist}}
+    \caption{Continuous and discontinuous lines as defined by different data models, but generated with the same \vartisteq \mintinline{python}{Line}}
     \label{fig:circle_step}
@@
 \subsection{Case Study: Penguins}
+\label{sec:code_case_study}
 For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020} since it is multivariate and has a varying number of penguins.
@@
         \includegraphics[width=\textwidth]{figures/code/bar_stacked.png}
+        \caption{}
     \end{subfigure}
@@
         \includegraphics[width=\textwidth]{figures/code/bar_grouped.png}
+        \caption{}
     \end{subfigure}
@@
 \end{figure}
+The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both out of \mintinline{python}{Bar} artists such that the difference between \mintinline{python}{StackedBar} and \mintinline{python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are stitched together.
+These two artists have identical constructors and \mintinline{python}{draw} methods.
+As with \mintinline{python}{Bar}, the orientation is set in the constructor.
+In both these artists, we separate the transforms applied to only one component and the case \mintinline{python}{mtransforms} where the same transform is applied to multiple components such that \vtotal\ has multiple components that map to the same retinal variable.
 
-The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both out of \mintinline{python}{Bar} artists such that the difference between \mintinline{Python}{StackedBar} and \mintinline{Python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are arranged differently.
-These two artists have identical \mintinline{{python}{__init__} and \mintinline{python}{draw} methods.
-As with \mintinline{python}{Bar}, the orientation is set on instantiation.
-In both these artists, we separate the transforms applied to only one component \mintinline{python}{transforms} and the case \mintinline{python}{mtransforms} where the same transform is applied to multiple components such that \votal\ has multiple components that map to the same retinal variable.
 \begin{minted}{python}
@@
 \end{minted}
+
 Since all the visual transformation is passed through to \mintinline{python}{Bar}, the \mintinline{python}{draw} method does not do any visual transformations.
-In \mintinline{Python}{StackedBar} the \mintinline{python}{view} is used to adjust the \mintinline{python}{floor} for every subsequent bar chart since a stacked bar chart is bar chart area marks concatenated together in the length parameter.
-In contrast, \mintinline{Python}{GroupedBar} does not even need the view, but instead keeps track of the relative position of each group of bars in the visual only variable \mintinline{python}{offset}.
+In \mintinline{python}{StackedBar} the \mintinline{python}{view} is used to adjust the \mintinline{python}{floor} for every subsequent bar chart since a stacked bar chart is bar chart area marks concatenated together in the length parameter.
+In contrast, \mintinline{python}{GroupedBar} does not even need the view, but instead keeps track of the relative position of each group of bars in the visual only variable \mintinline{python}{offset}.
 
@@
             self.children.append(bar)
-\end{minted}{python}
+\end{minted}
 
 Since the only difference between these two glyphs is in the composition of \mintinline{python}{Bar}, they take in the exact same transform specification dictionaries.
-\mintinline{python}{transform} dictates the position of the group, in this case by island the penguins are found on.
+The \mintinline{python}{transform} dictionary dictates the position of the group, in this case by island the penguins are found on.
 
@@
 We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic.
+
 \end{document}

### 2021-02-26 draft/sections/intro.tex
latex errors + cite tweaks

@@
 One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures.
-This in turn restricts the design space of visual algorithms that display information (the visualization types the library supports) since the algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighLevel2004}.
+This in turn restricts the design space of visual algorithms that display information (the visualization types the library supports) since the algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{ToryRethinkingVisualization2004}.
 In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.
@@
 There are also visualization tools where there is no single core structure, and instead internally carry around many different representations of data.
-Matplotlib, has this structure, as does VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}.
+Matplotlib, has this structure, as does VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{RamachandranMayaVI2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}.
 Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects.

### 2021-02-26 draft/sections/math.tex
latex errors + cite tweaks

@@
 
-For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree^K temperature measurements taken at those times.
+For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree K temperature measurements taken at those times.
 Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$.
@@
 \end{equation} 
-where the assembly function \vmarkd\ that takes as input the visual section on data \vsection\ is defined to be the assembly function \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gspace.
+where the assembly function \vmarkd\ that takes as input the visual section on data \vsection\ is defined to be the assembly function \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.
 

### 2021-02-26 draft/sections/next.tex
finished out math
latex errors + cite tweaks

@@
 While the model and prototype demonstrate that generation of simple marks from the data, there is a lot of work left to develop a model that underpins a minimally viable library.
-Foremost is implementing a data object that encodes data with a 2D continous topology and an artist that can consume data with a 2D topology to visualize the image\cite{toryRethinkingVisualizationHighLevel2004,haber1990visualization,hansen2011visualization} and also encoding a separate heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas} artist that consumes 1D discrete data.
+Foremost is implementing a data object that encodes data with a 2D continous topology and an artist that can consume data with a 2D topology to visualize the image\cite{ToryRethinkingVisualization2004, haber1990visualization,hansen2011visualization} and also encoding a separate heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas} artist that consumes 1D discrete data.
 A second important proof of concept artist is a boxplot\cite{wickham40YearsBoxplots2011} because it is a graphic that assumes computation on the data side and the glyph is built from semantically defined components and a list of outliers.
@@
     \item expansion of the mathematical framework to include complex addition
+    \item formalization of definition of equivalance class \vartisteq
     \item implementation of artist with explicit \vindex\

### 2021-02-26 draft/sections/intro.tex
start of math section figs

@@
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/dar.png}
-    \caption{Visualization is equivariant maps between data and visual encoding of the variables and assembly of those encodings into a graphic.
-    \note{will replace w/ overarching figure w/ same structure}}
+    \includegraphics[width=\textwidth]{figures/intro/dar.png}
+    \caption{Visualization is a mapping from data into visual encodings that are then rendered into images. In our model, this visual encoding stage is called the artist.}
     \label{fig:intro_artist_stages}

### 2021-02-26 draft/sections/next.tex
start of math section figs

@@
 The examples and prototype have so far only been implemented for the static 2D case, but nothing in the math limits to 2D and expansion to the animated case should be possible because the model is formalized in terms of the sheaf.
-While this model supports equivariance of figurative glyphs generated form parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it currently does not have a way to evaluate the semantic accuracy of the figurative representation.
+While this model supports equivariance of figurative glyphs generated from parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it currently does not have a way to evaluate the semantic accuracy of the figurative representation.
 Effectiveness is out of scope for this model because it is not part of the structure being preserved, but potentially a developer building a domain specific library with this model could implement effectiveness criteria in the artists.

### 2021-02-26 draft/sections/math.tex
more figures

@@
     \includegraphics[width=.5\textwidth]{figures/math/k_different_types.png}
-    \caption{The topological base space \dbase\ encodes the connectivity of the data space, for example if the data is independent points or a map or on a sphere}
+    \caption{The topological base space \dbase\ encodes the connectivity of the data space, for example if the data is independent points or on a plane or a sphere}
     \label{fig:base_space_types}
@@
     \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
-    \caption{The base space \dtotal\ is divided into fiber segments \dfiber. The base space \dbase\ acts as an index into the records in the fibers.
-    \note{this figure might be good all the way up top to lay out the components of fb}}
+    \caption{The base space \dtotal\ is divided into fiber segments \dfiber. The base space \dbase\ acts as an index into the records in the fibers.}
     \label{fig:base_space_div}
@@
     \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_2k.png}
+        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
         \caption{}
@@
 In figure~\ref{fig:base_example_line} the total space \dtotal\ is discrete over points $\dbasepoint \in \dbase$, meaning the records in the fiber are also discrete.
-In figure~\ref{fig:base_example_plane} \dtotal\ lies over the continuous interval \dbase, meaning the records in the fiber are sampled from a continuous space.
-    \note{revamp figure: F=Plane, k1 = dots, k2=line}}
+In figure~\ref{fig:base_example_plane} \dtotal\ lies over the continuous interval \dbase, meaning the records in the fiber are sampled from a continuous space.}
     \label{fig:base_example}
@@
 As a first pass for implementation purposes, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data such that the graphic is identifiable as a given chart type.
-\begin{figure}[H]
-    \caption{Each of these graphics is generated by a different artist \vartist which is the equivalance class of scatter plots $\vartist^{\prime}$\note{this is gonna be a whole bunch of scatter plots}}
-    \label{fig:artist_equivalence}
-\end{figure}
+%%
+%%\begin{figure}[H]
+%%   \caption{Each of these graphics is generated by a different artist \vartist which is the equivalance class of scatter plots $\vartist^{\prime}$\note{this is gonna be a whole bunch of scatter plots}}
+%%    \label{fig:artist_equivalence}
+%%\end{figure}
 
-For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another, as shown in figure~\ref{fig:artist_equivalence}.
+For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another.
 These two artists are equivalent since their only difference is in the literal visual encodings (color, shape).

### 2021-02-26 draft/sections/math.tex
through retraction fig
render image, going w/ minor rework of old one

@@
 \begin{figure}[H]
-    \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
+    \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
     \caption{ Fiber (time, temperature) with an interval \dbase\ basespace.
@@
     \caption{The scatter and line graphic base spaces have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
-The image has the same dimension in \gbase\ as in \dbase.
-\note{add $\alpha, \beta$ coordinates to figures}}
+The image has the same dimension in \gbase\ as in \dbase.}
     \label{fig:graphic_retraction_map}
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{figures/math/graphic_render.png}
-    \caption{To render a graphic, a pixel $p$ is selected in the display space, which is defined in the same coordinates as the x and y components in \gfiber\.
+    \includegraphics[width=\textwidth]{figures/math/render.png}
+    \caption{To render a graphic, a pixel $p$ is selected in the display space, which is defined in the same coordinates as the x and y components in \gfiber.
 The inverse mapping ${\gsection_{xy}}^(p)$ returns a region $\gbase_{p} \subset \gbase$.

### 2021-02-28 draft/sections/math.tex
overthinking 3d

@@
     \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
+        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
         \caption{}
@@
     \end{subfigure}
-    \caption{These two datasets have the same base space \dbase\, but figure~\ref{fig:fiber_example_plane} has fiber  $\dfiber=\reals\times\reals$ which is (time, temperature) while figure~\ref{fig:fiber_example_cube} has fiber $\realsp\times\reals^2$ which is (time, wind=(speed, direction))}
+    \caption{These two datasets have the same base space \dbase\ of discrete points, but figure~\ref{fig:fiber_example_plane} has fiber  $\dfiber=\reals\times\reals$ which is (time, temperature) while figure~\ref{fig:fiber_example_cube} has fiber $\reals\times\realsp\times\reals$ which is (time, wind=(speed, direction))}
     \label{fig:data_fiber_example}

### 2021-02-28 draft/sections/math.tex
must stop touching the data + graphics figures

@@
 \begin{figure}[H]
-    \includegraphics[width=.5\textwidth]{figures/math/k_different_types.png}
+    \includegraphics[width=1\textwidth]{figures/math/k_different_types.png}
     \caption{The topological base space \dbase\ encodes the connectivity of the data space, for example if the data is independent points or on a plane or a sphere}
@@
 \begin{figure}[H]
-    \includegraphics[width=.5\linewidth]{figures/math/k_qspace.png}
+    \includegraphics[width=\linewidth]{figures/math/k_qspace.png}
     \caption{The base space \dtotal\ is divided into fiber segments \dfiber. The base space \dbase\ acts as an index into the records in the fibers.}
@@
         \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-        %% add box around neighboring P and Map
         \caption{}
@@
     \caption{These two datasets have the same (time, temperature) fiber.
-In figure~\ref{fig:base_example_line} the total space \dtotal\ is discrete over points $\dbasepoint \in \dbase$, meaning the records in the fiber are also discrete.
-In figure~\ref{fig:base_example_plane} \dtotal\ lies over the continuous interval \dbase, meaning the records in the fiber are sampled from a continuous space.}
+In figure~\ref{fig:base_example_discrete} the total space \dtotal\ is discrete over points $\dbasepoint \in \dbase$, meaning the records in the fiber are also discrete.
+In figure~\ref{fig:base_example_continuous} \dtotal\ lies over the continuous interval \dbase, meaning the records in the fiber are sampled from a continuous space.}
     \label{fig:base_example}

### 2021-03-01 draft/sections/math.tex
equivariant actions

@@
     \caption{In this artis, \vchannel\ maps the strings to the emojis.
-For \vchannel\ to be equivariant, a shuffle in the words should have an equivalent shuffle in the emojies, and a shuffle in the emojies should have an equivalent shuffle in the words.}
+This \vchannel\ is equivariant because the monoid actions (which are represented by the colored arrows) are the same on both the \dsection\ input and \vsection\ output sets.}
     \label{fig:artist_nu}

### 2021-03-01 draft/sections/math.tex
q hat

@@
 %% equivariant maps 
-\begin{figure}
+\begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/math/equivariance_nu.png}
@@
     \includegraphics[width=\textwidth]{figures/math/path_of_q}
-    \caption{\vchannel\ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.
+    \caption{$\vchannel_i$ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.
 \gsection\ applied to a region of connected components $\gbase_{\dbasepathpoint}$  generates a part of a graphic, for example the point graphical mark.}
@@
 Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$.
-If for all monoids $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, the output is equivalent
+If for all monoid actions $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, the output is equivalent
 \begin{equation}
@@
 \end{equation}
-then a group action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
+then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
 The transformed graphic $\gsection^{\prime}$ is equivariant to a transform on the visual bundle $\gsection^{\prime}=\vmark(m\circ \vsection)$ on a section that $\vsection \in \vmark^{-1}(\gsection)$ that must be part of generating \gsection.
@@
     \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
-    \caption{These two glyphs are generated by the same \vmark\ function, but differ in the value of the edge thickness parameter $\vsection_i$.
-A valid \vmark\ is one where a shift in $\vsection_i$ is reflected in the glyph generated by \gsection.}
+    \caption{These two glyphs are generated by the same \vmark\ function. The monoid 
+    action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
     \label{fig:artist_mark_change}
@@
 The equivariance constraint on \vmark\ is that the action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$.
-The corresponding action on \gsection causes $\vmark(\vsection^{\prime})$ to be the thicker circle in figure~\ref{fig:artist_mark_change}.
+The corresponding action on \gsection\ causes $\vmark(\vsection^{\prime})$ to be the thicker circle in figure~\ref{fig:artist_mark_change}.
 
@@
 \begin{figure}[H]
-    \includegraphics[width=.75\textwidth]{figures/math/q_hat.png}
+    \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
     \caption{The pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint.

### 2021-03-01 draft/sections/math.tex
line q

@@
 \begin{figure}[H]
-    \includegraphics[width=.75\textwidth]{figures/math/scatter.png}
+    \includegraphics[width=1\textwidth]{figures/math/scatter.png}
     \caption{The data is discrete points (temperature, time).
 Via \vchannel\ these are converted to (xpos, ypos) and pulled over discrete \gbase.
-These values are then used to parameterize \gsection which returns a color based on the parameters (xpos,ypos) and position $\alpha, \beta$ on $\gbase_{\dbasepoint}$ that \gsection\ is evaluated on.
+These values are then used to parameterize \gsection\ which returns a color based on the parameters (xpos,ypos) and position $\alpha, \beta$ on $\gbase_{\dbasepoint}$ that \gsection\ is evaluated on.
     }
@@
 \begin{figure}[H]
-    \includegraphics[width=.75\textwidth]{figures/math/line.png}
-    \caption{The line fiber $(time,\, temp)$ is thickend with the derivative $(time^{\prime},\, temperature^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a thick line.
+    \includegraphics[width=\textwidth]{figures/math/line.png}
+    \caption{The line fiber $(time,\, temp)$ is thickened with the derivative $(time^{\prime},\, temperature^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.
 This is because the line needs to be pushed perpendicular to the tangent of (xpos, ypos).
-\note{this is gonna move once this gets regenerated w/ labels} The data is converted to visual characteristics (xpos, ypos).
+The data is converted to visual characteristics (xpos, ypos).
 The $\alpha$ coordinates on \gbase\ specifies the position of the line, the $\beta$ coordinate specifies thickness.}
@@
 
-which yields components of \gsection
+which yields components of \gsection\
 \begin{align}
@@
 \subsubsection{Composition of Artists: +}
-\begin{figure}
+\begin{figure}[H]
     \includegraphics[width=1\textwidth]{figures/math/addition_operator.png}
@@
 Roughly, equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\ but act on different sections \vsection, but we will formalize the definition of the equivalance class in future work.
-
 As a first pass for implementation purposes, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data such that the graphic is identifiable as a given chart type.

### 2021-03-01 draft/sections/math.tex
image Q
pulling figure ref out of A+ section

@@
 which is a decomposition of \dfiber. The \dbase\ remains the same because the connectivity of records does not change just because there are fewer elements in each record.
-\begin{figure}[ht!]
+\begin{figure}[H]
     \begin{subfigure}{.5\textwidth}
@@
 
-\begin{figure}
+\begin{figure}[H]
     \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
@@
 \label{sec:graphic_section}
-\begin{figure}
+\begin{figure}[H]
     \includegraphics[width=\textwidth]{figures/math/render.png}
@@
 \begin{figure}[H]
-    \includegraphics[width=.75\textwidth]{figures/math/heatmap.png}
+    \includegraphics[width=\textwidth]{figures/math/image.png}
     \caption{The only visual parameter a image requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
@@
 \end{align}
-where \vindex\ may do some translating to a convention expected by \vmark\, for example reorientng the array such that the first row in the data is at the bottom of the graphic.
+where \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.
 
@@
 \subsubsection{Composition of Artists: +}
-\begin{figure}[H]
-    \includegraphics[width=1\textwidth]{figures/math/addition_operator.png}
-    \caption{$\dsection_1$ and $\dsection_2$ are distinct datasets passed through artists $\vartist_1$ and $\vartist_2$ to generate graphics $\gsection_1$ and $\gsection_2$.
+%%\begin{figure}[H]
+%%   \includegraphics[width=1\textwidth]{figures/math/addition_operator.png}
+%%  \caption{$\dsection_1$ and $\dsection_2$ are distinct datasets passed through artists $\vartist_1$ and $\vartist_2$ to generate graphics $\gsection_1$ and $\gsection_2$.
 These graphics happen to be rendered to the same image, but otherwise have no intrinsinc link.}
-    \label{fig:artist_plus}
-\end{figure}
+%%    \label{fig:artist_plus}
+%%\end{figure}
 To build graphics that are the composites of multiple artists, we define a simple addition operator that is the disjoint union of fiber bundles \dtotal.
-For example, in figure~\ref{fig:artist_plus} the scatter plot $\dtotal_1$ and the line plot $\dtotal_2$ have different \dbase that are mapped to separate \gbase.
+For example, a scatter plot $\dtotal_1$ and a line plot $\dtotal_2$ have different \dbase that are mapped to separate \gbase.
 To fully display both graphics, the composite graphic $\vartist_1 + \vartist_2$ needs to include all records on both $\dbase_1$ and $\dbase_2$, which are the sections on the disjoint union $\dbase_1 \sqcup \dbase_2$.

### 2021-03-01 draft/sections/abstract.tex
abstract edits

@@
- A critical aspect of data visualization is that the graphical representation of data match the properties of the data; this fails when order is not preserved in representations of ordinal data or scale for numerical data.
-In this work, we propose that the mathematical notion of equivariance formalizes the expectation that graphics match the data.
+ A critical aspect of data visualization is that the graphical representation of data is expected to match the properties of the data; this fails when order is not preserved in representations of ordinal data or scale for numerical data.
+In this work, we propose that the mathematical notions of equivariance and topology formalizes the expectation of matching properties.
 We developed a model we call the topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles.
@@
 To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
-We represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
-In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
+We can show that using triangulations we can concretely represent the base spaces, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
+In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library provides better support concurrency and interactivity.
 The thesis will follow through on this proposal to explore how to further develop our model, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.

### 2021-03-01 draft/sections/intro.tex
rework of contribution section

@@
 Fiber bundles were proposed by Butler as one such representation because they encode the continuity of the data separately from the types of variables and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
+Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplicial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
 In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements.
@@
 Often this topology has metadata associated with it, describing for example when or where the measurement was taken.
-Building on the idea of metadata as \textit{keys} and their associated \textit{value} proposed by Munzner \cite{munznerChDataAbstraction}, we propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids.
-In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model allows for renaming all the metadata, for example changing the coordinate systems or time resolution, without imposing new semantics on the underlying structure.
+Building on the idea of metadata as \textit{keys} and their associated \textit{values} proposed by Munzner \cite{munznerChDataAbstraction}, we propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids.
+In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology.
+This allows the metadata to be altered, for example by changing the coordinate systems or time resolution, without imposing new semantics on the underlying structure.
 
@@
 In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph.
-Figure~\ref{fig:intro_dpb} is simply a treemap\cite{heerTourVisualizationZoo2010} with space between the marks.
+Figure~\ref{fig:intro_dpd} is simply a treemap\cite{heerTourVisualizationZoo2010} with space between the marks.
 Since the Du Bois data portraits meet the criteria of a faithful visual representation, we propose a mathematical framework and implementation that allows us to express the Du Bois charts and common chart types with equal fidelity.

### 2021-03-01 draft/sections/intro.tex
rework of nu

@@
 \subsection{Contribution}
-This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, this work contributes
+This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, the contributions of this work are
 
 \begin{enumerate}
-    \item a functional oriented visualization tool architecture
-    \item topology-preserving maps from data to graphic 
-    \item monoidal action equivariant maps from component to visual variable 
-    \item algebraic sum such that more complex visualizations can be built from simple ones 
-    \item prototype built on Matplotlib's infrastructure
+    \item a formal description of the topology preserving relationship between data and graphic via continuous maps
+    \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
+    \item abstraction of data as a fiber bundle with schema like fibers to encode components and topology 
+    \item algebraic sum operator such that more complex visualizations can be built from simple ones 
+    \item a functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
+    \item a prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
 \end{enumerate}
+
 In contrast to mathematical models of visualization that aim to evaluate visualization design, we propose a topological framework for building tools to build visualizations.

### 2021-03-01 draft/sections/math.tex
rework of nu

@@
 
-As discussed in the introduction, visualization is generally defined as structure preserving maps from data to graphic representation.
+As discussed in the introduction, visualization is generally defined as structure preserving maps from a data object to a graphic object.
 In order to formalize this statement, we describe the connectivity of the records using topology and define the structure on the components in terms of the monoid actions on the component types.
@@
 that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}, which are discussed in detail in section~\ref{sec:data_monoid}.
-Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are symmetric
+Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps
 \begin{equation}
@@
 \end{equation}
-such that applying a monoid action $m \in \monoid$ to the input $\delement \in \mathscr{\dtotal}$ to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the output of the artist $\vartist(\delement) \in \mathscr{\gtotal}$.
+such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.
 
@@
 
-\subsubsection{Variables: Fiber Space \dfiber}
+\subsubsection{Variables in Fiber Space \dfiber}
 \label{sec:data_fiber}
@@
 
-\subsubsection{Continuity: Base Space $K$} 
+\subsubsection{Continuity of the Data $K$} 
 \label{sec:data_base}
@@
 
-\subsubsection{Data: Sections \dsection}
+\subsubsection{Data \dsection}
 \label{sec:data_section}
-The section $\dsection: \dbase\rightarrow \dtotal$ is what ties together the base space \dbase\ with the fiber \dfiber.
-A section is a function that takes as input location $\dbasepoint \in \dbase$ and returns a record $\delement \in \dtotal$.
+While the projection function $\pi:\dtotal \rightarrow\dbase$ ties together the base space \dbase\ with the fiber \dfiber, a section $\dsection: \dbase\rightarrow \dtotal$ encodes a dataset.
+A section function takes as input location $\dbasepoint \in \dbase$ and returns a record $\delement \in \dtotal$.
 For example, in the special case of a table \cite{spivakSIMPLICIALDATABASES}, \dbase\ is a set of row ids, \dfiber\ is the columns, and the section \dsection\ returns the record \delement\ at a given key in \dbase.
@@
 We could also fatten the fiber to hold two temperature series, such that a section would be an instance of a dataframe with a time column and two temperature columns.
-While the series and dataframe explicitly have a time index column, they are components in our model and the index is always assumed to be random keys.
+While the series and dataframe explicitly have a time index column, they are components in our model and the index is assumed to be data independent references such as hashvalues, virtual memory locations, or random number keys.
 
@@
 
-\subsection{Graphic: \gtotal}
+\subsection{Graphic Space \gtotal}
 \label{sec:graphic}  
@@
 Just as the \dbase\ encodes the connectivity of the records in the data, we propose an equivalent \gbase\ that encodes the connectivity of the rendered elements of the graphic.
-For some visualizations, \dbase may be lower dimension than \gbase.
-For example, in a typical 2D display (ignoring depth), a point that is 0D in \dbase cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the pixels that visually represent the point.
-This thickening is often not necessary when the dimensionality of \dbase\ matches the dimensionality of the target space, for example if \dbase is 2D and the display is a 2D screen.
+For example, consider a \gbase\ that is mapped to the region of a 2D display space that represents \dbase.
+For some visualizations, \dbase\ may be lower dimension than \gbase.
+For example, a point that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the pixels that visually represent the point.
+This thickening is often not necessary when the dimensionality of \dbase\ matches the dimensionality of the target space, for example if \dbase\ is 2D and the display is a 2D screen.
 We introduce \gbase\ to thicken \dbase\ in a way which preserves the structure of \dbase.
@@
 
-\subsubsection{Renderering \gsection}
+\subsubsection{Graphic \gsection}
 \label{sec:graphic_section}
@@
 While \vartist can usually construct graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
-In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the data bundle padded with the jet bundle $\dtotal^{\prime} = \dtotal + \mathcal{J}^{2}(\dtotal)$.
+In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the jet bundle $\dotal^{\prime}=\mathcal{J}^{2}(\dtotal)$.
 
@@
 as the set of equivariant maps $\vchannel_i: \dsection_i \mapsto \vsection_i$.
-Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve.
+Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve.
 As mentioned in section~\ref{sec:data_monoid}, we choose monoid actions as the basis for equivariance because they define the structure on the fiber components.
 
-A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$
+A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes such that 
 \begin{equation}
@@
 \end{equation}
-commutes such that 
+
+In general, the data fiber $\dfiber_{i}$ cannot be assumed to be of the same type as the visual fiber $\vfiber_{i}$; therefore the actions of \monoid\ on  $\dfiber_{i}$ are not the same as the actions of $\moniod^{prime}$ on \vfiber.
+This means that an equivariant \vchannel\ must commute
 \begin{equation}
-\vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\velement})(\vchannel_i(\dtotal_i))\end{equation}
+\vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))\end{equation} 
+such that $\varphi: m_{\delement} \mapsto m_{\velements}$ maps a monoid action on data to a monoid action on visual elements.
+However, we can construct a monoid action of \monoid\ on $\vfiber_{i}$ that is compatible with a monoid action of \monoid\ on $\dfiber_{i}$.
+We can compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi: \monoid \rightarrow \monoid^{\prime}$ that takes \monoid\ to $\monoid^{\prime}$.
+This allows us to define a monoid action on \vfiber\ of \monoid\ that is $(m, \velement) \rightarrow \varphi(m)\bullet\velement$.
+Therefore without a loss of generality, we can assume that \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.
 
-This equivariance constraint yields guidance on what makes for an invalid transform.
-For example, given a visual fiber of the same type as the data fiber $\dfiber_i=\vfiber_i$, the transform $\vchannel_{i}(x) = .5$ does not commute under translation $t(x) = x+2$
-\begin{align}
-    \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + \vchannel(2)\\
-    .5 &\neq .5 + .5
-\end{align}
-
-On the other hand figure~\ref{fig:artist_nu} illustrates a valid \vchannel\ mapping from \textbf{Strings} to symbols.
-The group action on these sets is permutation, so shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
-Continuing with the assumption that $\dfiber=\vfiber$, we can describe the constraints on the other Steven's measurment group types.
-To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$
+On example of an equivariant \vchannel\ is illustrated in figure~\ref{fig:artist_nu}, which is a mapping from \textbf{Strings} to symbols.
+This is also an example of a Steven's nominal measurement set, which is defined as having permutation group actions
 \begin{equation}
-\text{ if } delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)
+    \text{ if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)
+\end{equation}
+such that shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
+This is illustrated in the identical actions, represented by the colored arrows, on the words and emojis.
+To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$,
+\begin{equation}
+\text{ if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)
 \end{equation}
@@
 \begin{equation}
-\vchannel (x + c) = \vchannel(x) + \vchannel(c)
+\vchannel (x + c) = \vchannel(x) + c
 \end{equation}
@@
 \begin{equation}
-\vchannel(xc) = \vchannel(x)\vchannel(c) 
+\vchannel(xc) = \vchannel(x)*c 
 \end{equation}
-The assumption $\dfiber_i=\vfiber_i$ is made here so that we could omit  $\varphi: \monoid\rightarrow\monoid^{\prime}$ maps that convert the monoid action on $\dfiber_{i}$ to the equivalent action on $\vfiber_{i}$.
-As shown in table~\ref{tab:mpl_visual_variable_fiber}, it is not uncommon for data and visual variables to resolve to the same types.
+
+We therefore can test if a \vchannel\ is equivariant by testing the actions under which is must  commute.
+For example, we define a transform $\vchannel_{i}(x) = .5$ on interval data.
+This means it must commute under translation, for example $t(x) = x+2$.
+Testing this constraint
+\begin{align}
+    \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + \varphi(2)\\
+    .5 &\neq .5 + 2
+\end{align}
+we find that the \vchannel\ defined here does not commute and is therefore invalid.
 The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions are equivariant and also provide guidance on constructing new equivariant \vchannel\ functions.
@@
     \caption{The pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint.
-Because the \vsection\ is the same, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent set of \gbasepoint\ associated to that \dbasepoint}
+Because the \vsection\ is the same, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
     \label{fig:artist_q_hat}
@@
 
-\subsubsection{Sheaf}
+\subsubsection{Sheafs}
 \label{sec:artist_sheaf_stalk}

### 2021-03-02 draft/sections/math.tex
closed #17

@@
 While \vartist can usually construct graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
-In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the jet bundle $\dotal^{\prime}=\mathcal{J}^{2}(\dtotal)$.
+In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the jet bundle $\dtotal^{\prime}=\mathcal{J}^{2}(\dtotal)$.
 
@@
 
- A section \vsection is a tuple of visual values that specifies the visual characteristics of a part of the graphic.
+ A section \vsection\ is a tuple of visual values that specifies the visual characteristics of a part of the graphic.
 For example, given a fiber of $\{xpos, ypos, color\}$ one possible section could be  $\{.5, .5, (255, 20,147)\}$.
@@
     \includegraphics[width=\textwidth]{figures/math/equivariance_nu.png}
-    \caption{In this artis, \vchannel\ maps the strings to the emojis.
+    \caption{In this artist, \vchannel\ maps the strings to the emojis.
 This \vchannel\ is equivariant because the monoid actions (which are represented by the colored arrows) are the same on both the \dsection\ input and \vsection\ output sets.}
@@
 \end{equation}
-
-In general, the data fiber $\dfiber_{i}$ cannot be assumed to be of the same type as the visual fiber $\vfiber_{i}$; therefore the actions of \monoid\ on  $\dfiber_{i}$ are not the same as the actions of $\moniod^{prime}$ on \vfiber.
-This means that an equivariant \vchannel\ must commute
+In general, the data fiber $\dfiber_{i}$ cannot be assumed to be of the same type as the visual fiber $\vfiber_{i}$ and the actions of \monoid\ on $\dfiber_{i}$ cannot be assumed to be the same as the actions of $\monoid^{\prime}$ on \vfiber; therefore an equivariant $\vchannel_i$ must satisfy the constraint
 \begin{equation}
-\vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))\end{equation} 
-such that $\varphi: m_{\delement} \mapsto m_{\velements}$ maps a monoid action on data to a monoid action on visual elements.
+\vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
+\end{equation} 
+such that $\varphi$ maps a monoid action on data to a monoid action on visual elements.
 However, we can construct a monoid action of \monoid\ on $\vfiber_{i}$ that is compatible with a monoid action of \monoid\ on $\dfiber_{i}$.
-We can compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi: \monoid \rightarrow \monoid^{\prime}$ that takes \monoid\ to $\monoid^{\prime}$.
+We can compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi$ that takes \monoid\ to $\monoid^{\prime}$.
 This allows us to define a monoid action on \vfiber\ of \monoid\ that is $(m, \velement) \rightarrow \varphi(m)\bullet\velement$.
-Therefore without a loss of generality, we can assume that \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.
+Therefore, without a loss of generality, we can assume that an action of \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.
 
 On example of an equivariant \vchannel\ is illustrated in figure~\ref{fig:artist_nu}, which is a mapping from \textbf{Strings} to symbols.
-This is also an example of a Steven's nominal measurement set, which is defined as having permutation group actions
+The data is an example of a Steven's nominal measurement set, which is defined as having on it permutation group actions
 \begin{equation}
@@
 \end{equation}
-
 We therefore can test if a \vchannel\ is equivariant by testing the actions under which is must  commute.
@@
 \begin{align}
-    \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + \varphi(2)\\
+    \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + 2\\
     .5 &\neq .5 + 2
@@
 we find that the \vchannel\ defined here does not commute and is therefore invalid.
-The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions are equivariant and also provide guidance on constructing new equivariant \vchannel\ functions.
+The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions can test for equivariance and also provide guidance on constructing new \vchannel\ functions.
 
@@
 While \gsection\ generates the entire graphic, we will restrict the discussion of \vmark\ to generation of sections of a glyph.
-We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of connected components $\dbasepath \subset \dbase$ as input:
+We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of path connected components $\dbasepath \subset \dbase$ as input:
 \begin{equation}
@@
 Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$.
-If for all monoid actions $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, the output is equivalent
+If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
 \begin{equation}
@@
 \begin{equation}
+\label{eq:qhat_q_s}
 \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
@@
 
-Factoring out \gbasepoint\, $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ generates a curried \vmark.
+Factoring out \gbasepoint\ from equation~\ref{eq:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
 In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record   $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
 This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
-The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing rendering pipeline.
+The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delgate the construction of \gsection\ to a back end such as Matplotlib.
 

### 2021-03-02 draft/sections/abstract.tex
part of #18

@@
 To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
-We can show that using triangulations we can concretely represent the base spaces, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
+We can show that we can concretely represent the base spaces, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
 In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library provides better support concurrency and interactivity.

### 2021-03-02 draft/sections/implementation.tex
part of #18

@@
 We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations.
+We also exploit the Matplotlib transform stack to transform data coordinates into screen coordinates.
 To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primitive objects we implemented based on our topology model.
@@
 \label{sec:code_artist}
-The artist is the piece of the matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic.
+The artist is the piece of the Matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic.
 In the prototype artist, \mintinline{python}{transform} is a dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber,  and the \vchannel\ encoders are passed in as functions or callable objects.
@@
 The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assemble} function that is \vmarkd.
-This assembly function is responsible for generating a representation of the glyph such that it could be serialized to recreate a static version of the graphic.
+This assembly function is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
 Although \mintinline{python}{assemble} could be implemented outside the class such that it returns an object the artist could then parse to set attributes, the attributes are directly set here to reduce indirection.
@@
 The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist.
-The have the same required \vfiber\ components of (position, length), but in \mintinline{python}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally.
+The artist has required visual parameters \vfiber\ of (position, length), and an additional parameter \texttt{orientation} which controls whether the bars are arranged vertically or horizontally.
 This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd.
@@
 
-    def assemble(self, position, length, floor=0, width=0.8, facecolors='C0', edgecolors='k', offset=0):
+    def assemble(self, position, length, floor=0, width=0.8, 
+                    facecolors='C0', edgecolors='k', offset=0):
         #set some defaults
@@
 
-        # offset is passed through via assemblers such as multigroup, not supposed to be directly tagged to position 
+        # offset is passed through via assemblers such as multigroup, 
+        # not supposed to be directly tagged to position 
         position = position + offset
@@
 K: {'tables': []}
-F: {variable name: {'type': type, 'moniod', 'range': []}}
+F: {variable name: type}
 """
@@
     FB = FiberBundle({'tables': ['vertex']},  
-            {'v1': {'type': float,'monoid':'interval','range': [0,1]},
-             'v2': {'type': str, 'monoid':'nominal', 'range':['true', 'false']},
-             'v3': {'type': float, 'monoid':'interval', 'range':[2,3]}})
+            {'v1': float, 'v2': str, 'v3': float})
 
@@
 class EdgeSimplex:
-    # assign a class level FB attribute
+
+        FB = FiberBundle({'tables': ['vertex','edge']},
+                         {'x' : float, 'y':  float, 
+                         'color':mtypes.Color()}})
     def __init__(self, num_edges=4, num_samples=1000): 
@@
 While very rough, this API demonstrates that the ideas presented in the math framework are implementable.
+For example, the \mintinline{python}{draw} function that maps most closely to \vartist\ is functional, with state only being necessary for bookkeeping the many inputs that the function requires.
 In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\, assembly \vmarkd and artists \vartist.

### 2021-03-02 draft/sections/intro.tex
part of #18

@@
 
-Often this topology has metadata associated with it, describing for example when or where the measurement was taken.
+Often this topology has metadata associated with it, describing for example dependent variables and the independent variables they are dependent on, or information about the structure of the data such as when and where the measurement was taken.
 Building on the idea of metadata as \textit{keys} and their associated \textit{values} proposed by Munzner \cite{munznerChDataAbstraction}, we propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids.
@@
 This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, the contributions of this work are 
-
 \begin{enumerate}
@@
     \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
-    \item abstraction of data as a fiber bundle with schema like fibers to encode components and topology 
+    \item abstraction of data structure using fiber bundles with schema like fibers to encode components and topology 
     \item algebraic sum operator such that more complex visualizations can be built from simple ones 
@@
 \end{enumerate}
-
 In contrast to mathematical models of visualization that aim to evaluate visualization design, we propose a topological framework for building tools to build visualizations.

### 2021-03-02 draft/sections/math.tex
part of #18

@@
 \label{sec:data}
-Building on Butler's proposal of using fiber bundles as a common data representation format for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
+Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
 \begin{equation}
@@
 \label{sec:data_fiber}
-To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names and types.
+To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names.
+This allows us to describe the components in a schema like way.
 Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
@@
 \label{sec:data_monoid}
-Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define by the action of a monoid on the component.
+Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define by the actions of a monoid on the component.
+In doing so, we specify the properties of the component that must be preserved in a graphic representation.
 While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's scales \cite{stevensTheoryScalesMeasurement1946}, we generalize to monoids to support more component types.
@@
 Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-For example, given an arbitrary interval scale fiber component $(\fname=temperature,\, \ftype=\texttt{float},\,\fttype=\reals)$ with arbitrarily chosen monoid translation actions actions
+For example, given an arbitrary interval scale fiber component $(\fname=temperature,\, \ftype=\texttt{float},\,\fttype=\reals)$ with with arbitrary monoid translation actions chosen for this example:
 \begin{itemize}
     \item monoid operator addition $\ast = +$
-    \item monoid operations: $f: x\mapsto x + 1 $, $g: x\mapsto x + 2  $
+    \item monoid operations: $f: x\mapsto x + 1\degree C $, $g: x\mapsto x + 2\degree C$
     \item monoid action operator composition $\bullet = \circ$
@@
     \begin{tikzcd}
-        \reals \arrow[rd, "(x+ 1\degree)\circ(x+2\degree)"] \arrow[d, "x+ 1\degree"'] &            \\
-        \reals \arrow[r, "x+ 2\degree"']                                   & \reals
+        \reals \arrow[rd, "(x+ 1\degree C)\circ(x+2\degree C)"] \arrow[d, "x+ 1\degree"'] &            \\
+        \reals \arrow[r, "x+ 2\degree C"']                                   & \reals
     \end{tikzcd}
 \end{equation}
-where $1\degree$ and $2\degree$ are valid distances between two temperatures $x$.
-What this diagram means is that either the fiber could be shifted by 1 (vertical line) then by 2 (horizontal), or the two shifts could be combined such that in this case the fiber is shifted by 3 (diagonal) and these two paths yield the same temperature.
+where $1\degree C$ and $2\degree C$ are valid distances between two temperatures $x$.
+What this diagram means is that either the fiber could be shifted by $1 \degree C$ (vertical line) then by $2 \degree C$ (horizontal), or the two shifts could be combined such that in this case the fiber is shifted by $3 \degree C$ (diagonal) and these two paths yield the same temperature.
 
@@
 \end{equation}
-where each section $\dsection_i$ is a variable or set of variables. This allows for accessing the data component wise rather than just as sections on \dbase.
+where each section $\dsection_i$ is a variable or set of variables. This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.
 
@@
     \caption{ Fiber (time, temperature) with an interval \dbase\ basespace.
-The sections $\dsection_i$ and $\dsection_j$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of temperature values.
-They are included in the global set of sections  $\dsection_1, \dsection_2 \in \Gamma(\dtotal)$}
+The sections $\dsection^{(1)}$ and $\dsection^{(2)}$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of temperature values.
+They are included in the global set of sections  $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$}
     \label{fig:data_sections}
@@
 
-This model provides a common interface to widely used data containers without sacrificing the semantic structure embedded in each container.
+\subsubsection{Applications to Data Containers}
+This model provides a common formalism for widely used data containers without sacrificing the semantic structure embedded in each container.
 For example, the section can be any instance of a univariate numpy array\cite{harris2020array} that stores an image.
 This could be a section of a fiber bundle where \dbase\ is a 2D continuous plane and the \dfiber\ is $(\reals^{3}, \reals, \reals)$ where $\reals^3$ is color, and the other two components are the x and y positions of the sampled data in the image.
-Instead of an image, the numpy array could also store a 2D discrete table.
+This position information is already implicitely encoded in the array as the index and the resolution of the image being stored.Instead of an image, the numpy array could also store a 2D discrete table.
 The fiber would not change, but the \dbase\ would now be 0D discrete points.
@@
 
-There are also many types of labeled containers that can richly be described in this framework because of the fiber.
+There are also many types of labeled containers that can richly be described in this framework because of the schema like structure of the fiber.
 For example, a pandas series which stores a labeled list, or a dataframe\cite{jeff_reback_2020_3715232} which stores a relational table.
@@
 We propose that the transformation from data to visual representation can be described as a structure preserving map from one topological space to another.
-We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements to pass off to the renderer.
+We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.
 The topological artist \vartist\ is a monoid equivariant sheaf map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
@@
 \end{equation}
-such that for every glyph there is at least one corresponding section on \dbase.
+such that for every glyph there is at least one corresponding region on \dbase.
 This is in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
@@
 \end{figure}
-The scatter plot in figure~\ref{fig:scatter} can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
+The scatter plot in figure~\ref{fig:artist_scatter} can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
 The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in figure~\ref{fig:artist_scatter}:
 \begin{align}
-x &= size\bullet \alpha \bullet \cos(\beta) + xpos\\
-y &= size\bullet \alpha \bullet \sin(\beta) + ypos
+x &= size *\alpha \cos(\beta) + xpos \\
+y &= size *\alpha \sin(\beta) + ypos
 \end{align}
@@
 such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
+Here $size$ can either be defined inside \vmark\ or it could also be a parameter in \vtotal\ that is passed along with (xpos, vpos).
+As seen in figure~\ref{fig:artist_scatter}, a scatter has a direct mapping from a region on $\gbase_{k}$ to its corresponding \dbasepoint.
+
 \begin{figure}[H]
@@
 
-The line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ shown in fig~\ref{fig:artist_scatter} exemplifies the need for the jet.
-The line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
-The magnitude of the thickness is
+In contrast to the scatter, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ shown in fig~\ref{fig:artist_line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase\.
+The line also exemplifies the need for the jet since the line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
+The magnitude of the slope is
 \begin{equation}
@@
 \begin{align}
- x = xpos(\vindex(\alpha)) &+ \beta\hat{n_1}(\vindex(\alpha)) \\
- y = ypos(\vindex(\alpha)) &+ \beta\hat{n_2}(\vindex(\alpha)) 
+ x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
+ y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
 \end{align}
 
-where (x,y) look up the position $\vindex(\alpha)$ on the data. At that point, we also look up the the derivatives $\hat{n_1}, \hat{n_2}$  which are then multiplied by data to specify the thickness.
+where (x,y) look up the position $\vindex(\alpha)$ on the data.
+At that point, we also look up the the derivatives $\hat{n_1}, \hat{n_2}$  which are then multiplied by a $width$ parameter to specify the thickness.
+As with the $size$ parameter in scatter, $width$ can be defined in \vmark\ or as a component of \vtotal.
 
@@
     \includegraphics[width=\textwidth]{figures/math/image.png}
-    \caption{The only visual parameter a image requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
+    \caption{The only visual parameter an image requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
     \label{fig:artist_heatmap}
@@
 
-The image $\vmark(color)$ in figure~\ref{fig:artist_heatmap} is a direct lookup $\vindex:\gbase\rightarrow\dbase$ such that
-
+The image $\vmark(xpos, yposcolor)$ in figure~\ref{fig:artist_heatmap} is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
+Since \dbase\ is 2D continuous space, the indexing variables $(\alpha, \beta)$ define the distance along the space.
+This is then used by \vindex\ to map into \dbase\ to lookup the color values
 \begin{align}
@@
 \end{align}
-where \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.
+that the data values have been mapped into.
+In the case of an image, the indexing mapper \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.
 

### 2021-03-02 draft/sections/next.tex
part of #18

@@
 These composite examples also rely on the fiber bundles section base book keeping to keep track of which components contribute to the attributes of the glyph.
-Implementing this example using a Pandas dataframe demonstrates the ease of incorporating existing widely used data containers rather than requiring users to conform to one stands.
+Implementing this example using a Pandas dataframe demonstrates the ease of incorporating existing widely used data containers rather than requiring users to conform to one standard.
 
@@
 Effectiveness is out of scope for this model because it is not part of the structure being preserved, but potentially a developer building a domain specific library with this model could implement effectiveness criteria in the artists.
-Also, even though the model is designed to be backend independent, it has only really been tested against the AGG backend.
+Also, even though the model is designed to be backend and format independent, it has only really been tested against PNGs rendered with the AGG backend.
 It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}.

### 2021-03-05 slides/main.tex
slides up to graphics

\documentclass[xcolor={dvipsnames}, handout]{beamer}
\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
\usepackage{mathrsfs} % math script fonts
\usepackage{tikz-cd} % commutative diagrams
\usepackage{subcaption} % subfigure float captioning

\usepackage{tabularx}
\usepackage{multirow}
\usepackage{graphicx}

\usepackage{notation} % move this later

\graphicspath{.figures/}

\usepackage[backend=bibtex, style=numeric-comp, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
\bibliography{../draft/glasslab_viz.bib}

\usetheme{ccnycrest}

\newenvironment{changemargin}[2]{%
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{#1}%
\setlength{\rightmargin}{#2}%
\setlength{\listparindent}{\parindent}%
\setlength{\itemindent}{\parindent}%
\setleng{}th{\parsep}{\parskip}%
}%
\item[]}{\end{list}}

\begin{document}

\title{Topological Artist Model}

\begin{frame}
	\titlepage
    Hannah Aizenman\\
    Advisor: Dr. Michael Grossberg \\
    Committee: Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo\\
    External Member: Dr. Marcus Hanwell
\end{frame}

\section{Introduction}

\begin{frame}{What do visualization libraries do?}
    \begin{figure}
        \includegraphics[width=1\linewidth]{figures/intro/dar.png}
    \end{figure}
\end{frame}

\begin{frame}{Why are we doing this?}
    \begin{enumerate}
        \item visualization software design patterns are tuned to data structures\cite{HeerSoftware2006}
        \item visualization types tuned to data structures\cite{toryRethinkingVisualizationHighlevel2004}
        \item need to build a tool that supports domain specific structures and algorithms
        \item tool needs to be general purpose enough to support lots of domains
        \item tools needs to support 2D, 3D, static, dynamic, and interactive viz
    \end{enumerate}
\end{frame}

\subsection{Tools}
\begin{frame}{Everything is a table}
\begin{columns}
    \column{0.5\textwidth}
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/intro/grammar_example.png}
        \caption{Figure 1.5 in Wilkenson's Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}}
    \end{figure}
    \column{0.5\textwidth}
        \begin{enumerate}
            \item  ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, altair\cite{vanderplasAltairInteractiveStatistical2018}
            \item tends to be declarative\cite{heerDeclarative2010} 
            \item users describe how to compose visual elements \cite{wongsuphasawatNavigatingWideWorld2021}
        \end{enumerate}
    \end{columns}
    \pause
    \begin{center}
        \textbf{TAM: build the visual elements}
    \end{center}
\end{frame}

\begin{frame}{Everything is an image}
    \begin{columns}
        \column{0.5\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
        \end{figure}
        \column{0.5\textwidth}
        \begin{enumerate}
            \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}, Napari\cite{nicholas_sofroniew_2021_4533308}
            \item build plugins into existing system where image is primary input
        \end{enumerate}
    \end{columns}
    \pause
    \begin{center}
        \textbf{TAM: more general than tables and images}
    \end{center}
\end{frame}

\begin{frame}{Everything is ?}
    \begin{columns}
        \column{0.5\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
            \caption{Image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
        \end{figure}
        \column{0.5\textwidth}
        \begin{enumerate}
            \item Matplotlib\cite{hunterMatplotlib2DGraphics2007},  VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
            \item each visualization type API is tuned to the data structure
        \end{enumerate}
    \end{columns}
    \pause
    \begin{center}
        \textbf{TAM: common data abstraction interface between data and viz }
    \end{center}
\end{frame}

\subsection{Data}
\begin{frame}{Variables \& Topology}
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
        \caption{Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
    \end{figure}
    \begin{description}
        \item  metadata are \textit{keys} with associated \textit{values} (Munzner \cite{munznerVisualizationAnalysisDesign2014})
        \item[topology] Fiber bundles can be a common data abstraction (Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989})
        \item[variables] Fiber as a database schema (Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES})
    \end{description}
    \pause
    \begin{center}
        \textbf{Tam: variables are \textit{values} and \textit{keys} locate them in topology}
    \end{center}
\end{frame}

\subsection{Visualization}
\begin{frame}{Property Matching}
    \begin{figure}
        \includegraphics[width=.7\textwidth]{figures/intro/retinal_variables.png}
        \caption{This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from Krygier and Wood's \textit{Making Maps: A Visual Guide to Map Design for GIS}\cite{krygierMakingMapsVisual2005}}
    \end{figure}
\end{frame}

\begin{frame}{Evaluating visualizations}
    \begin{description}
        \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
        \item[Effectiveness] design choices made in deference to perceptual saliency \cite (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
        \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
        \item[Graphical Integrity] graphs show \textbf{only} the data (Tufte \cite{tufteVisualDisplayQuantitative2001})
    \end{description}
\end{frame}

\begin{frame}{Mathematical Frameworks for evaluating visualization}
    \begin{enumerate}
        \item APT: visualization has syntax and semantics like a language  (Mackinlay  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987})
        \item Visualization has functional dependencies that can be represented as graphs (Sugibuchi \cite{sugibuchiFramwork2009}) 
        \item the semiotics of visualization are commutative in a category theory framework (Vickers \cite{vickersUnderstandingViz2013})
    \end{enumerate}
    \pause
    \begin{center}
        \textbf{Tam: Framework for building visualizations}
    \end{center}
\end{frame}

\begin{frame}{Visualization is commutative maps}
    data ($\alpha$) and viz ($\omega$) symmetries (Kindlmann and Scheidegger \cite{kindlmann2014algebraic})
        \begin{block}{$v\circ r_2 \circ \alpha =\omega\circ v\circ r_1$}
            \begin{equation*}
            \begin{tikzcd}[ampersand replacement=\&]
                D \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& R \arrow[r, "\nu"]  \& V \arrow[d, "\omega"] \\
                D \arrow[r, "r_2"']                     \& R \arrow[r, "\nu"'] \& V                    
            \end{tikzcd}
        \end{equation*}
    \end{block}
    \pause
    \begin{center}
        \textbf{Tam: Add topology and make it functional}
    \end{center}
\end{frame}
\begin{frame}{So why?}
    \begin{figure}[H]
        \begin{subfigure}{.24\textwidth}
            \includegraphics[width=1\textwidth]{figures/intro/du_bois_spinny.png}
            \caption{}
            \label{fig:intro_dpa}
        \end{subfigure}
        \begin{subfigure}{.24\textwidth}
            \includegraphics[width=1\textwidth]{figures/intro/du_bois_line.png}
            \caption{}
            \label{fig:intro_dpb}
        \end{subfigure}
        \begin{subfigure}{.24\textwidth}
            \includegraphics[width=1\textwidth]{figures/intro/du_bois_country.png}
            \caption{}
            \label{fig:intro_dbc}
        \end{subfigure}
        \begin{subfigure}{.24\textwidth}
            \includegraphics[width=1\textwidth]{figures/intro/du_bois_heat.png}
            \caption{}
            \label{fig:intro_dbd}
        \end{subfigure}
        \caption{Du Bois' data portraits from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroValuation1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
        \label{fig:intro_dubois}
    \end{figure}
    \begin{center}
        \textbf{Framework for building any valid visualization}
    \end{center}
\end{frame}

\subsection{Contributions}
\begin{frame}{Contributions}
    \begin{enumerate}
        \item a formal description of the topology preserving relationship between data and graphic via continuous maps
        \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
        \item abstraction of data structure using fiber bundles with schema like fibers to encode components and topology 
        \item algebraic sum operator such that more complex visualizations can be built from simple ones 
        \item a functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
        \item a prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
    \end{enumerate}
\end{frame}

\section{Mathematical Framework}

\begin{frame}{Topological Artist Model}
    \begin{equation}
        \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
    \end{equation}
    \pause
    which is a map from data \dtotal\ to graphic \gtotal\ that carries a homomorphism of monoid actions 
    \pause 
    \begin{equation}
    \varphi: \monoid \rightarrow \monoid^{\prime}
    \end{equation}
    \pause
    such that artists are equivariant maps 
    \begin{equation}
    \mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
    \end{equation}
\end{frame}

\subsection{Data Model}

\begin{frame}{Data Bundle}
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/math/fiberbundle.png}
    \end{figure}
    a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
    \begin{equation}
        \label{eq:fiber_bundle}
        \begin{tikzcd}[ampersand replacement=\&]
            \dfiber \arrow[r, hook] \& \dtotal \arrow[r, "\pi"] \& \dbase
        \end{tikzcd}
    \end{equation}
\end{frame}

\begin{frame}{Variables: Fiber}
    \begin{equation}
        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
    \end{equation}
    where $\ftotal_{\fsection(\fname)}$ is 
    \begin{equation}
        \label{eq:data_types}
        \begin{tikzcd}[ampersand replacement=\&]
            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
        \end{tikzcd}
    \end{equation}
    \begin{description}
        \item[\ftypes] data types of the variables in the dataset 
        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
        \item[\fnames] variable names
        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
    \end{description}
\end{frame}

\begin{frame}{Sample Fiber}
    \begin{figure}[H]
        \begin{subfigure}{.4\textwidth}
            \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
            \caption{}
            \label{fig:fiber_example_plane}
        \end{subfigure}
        \begin{subfigure}{.4\textwidth}
            \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
            \caption{}
            \label{fig:fiber_example_cube}
        \end{subfigure}
        \caption{}
        \label{fig:data_fiber_example}
    \end{figure}
    \begin{description}
        \item[\ref{fig:fiber_example_plane}]  $\dfiber=\reals\times\reals$, for example (time, temperature)
        \item[\ref{fig:fiber_example_cube}]  $\reals\times\realsp\times\reals$, for example  (time, wind=(speed, direction))
    \end{description}
\end{frame}

\begin{frame}{Fiber Components}
    We can decouple decouple \dfiber\ into components $\dfiber_i$
    \begin{equation}
        \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
    \end{equation}
\end{frame}

\begin{frame}{Structure of Components: Monoids}
A monoid \monoid\ is a set with
\begin{description}
    \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
    \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
\end{description}
\end{frame}
\begin{frame}{Monoid Actions}
A left monoid action of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
    \begin{align*}
        \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
        \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
    \end{align*}
\end{frame}

\begin{frame}{Monoid Composition: Permutation}
    \begin{figure}
        \includegraphics[width=1\linewidth]{figures/math/monoid_emoji.png}
    \end{figure}
\end{frame}

\begin{frame}{Monoid Composition: Partial Order}
    MAKE THIS Hasse Diagram (borrow from theories of composablity)
\end{frame}

\begin{frame}{Continuity: base space}
    \begin{figure}[H]
    \includegraphics[width=\linewidth]{figures/math/k_qspace.png}
    \caption{}
    \label{fig:base_space_div}
\end{figure}
where the total space can be decomposed into components 
\begin{equation}
    \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
\end{equation}
\end{frame}

\begin{frame}{Sample: Base Space}
    \begin{figure}[H]
        \begin{subfigure}{.4\textwidth}
            \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
            \label{fig:base_example_discrete}
        \end{subfigure}
        \begin{subfigure}{.4\textwidth}
            \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
            \label{fig:base_example_continuous}
        \end{subfigure}
        \label{fig:base_example}
    \end{figure}
\end{frame}

\begin{frame}{Values: Section}
    For any fiber bundle, there exists a map
    \begin{equation}
        \begin{tikzcd}[ampersand replacement=\&]
            \dfiber \arrow[r, hook] \& \dtotal \arrow[d, "\pi"'] \\
                              \& \dbase \arrow[u, "\dsection"', bend right]
        \end{tikzcd}
    \end{equation}
     such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$.

     The set of all global sections is denoted as $\Gamma(\dtotal)$.
\end{frame}

\begin{frame}{Record}
    Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is 
\begin{equation}
    \label{eq:section_return}
    \dsection(\dbasepoint) = (\dbasepoint, (g_{\dfiber_{0}}(\dbasepoint), \ldots, g_{\dfiber_{n}}(\dbasepoint)))
\end{equation}
where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber.
\end{frame}

\begin{frame}{Sample dataset}
    \begin{figure}[H]
        \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
        \label{fig:data_sections}
    \end{figure}
\end{frame}

\subsection{Graphics}
\begin{frame}{Fiber Bundle}
    The graphics bundle is a tuple $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$ defined by the projection map $\pi$
    \begin{equation}
        \label{eq:fiber_bundle}
        \begin{tikzcd}[ampersand replacement=\&]
            \gfiber \arrow[r, hook] \& \gtotal \arrow[r, "\pi"] \& \gbase
        \end{tikzcd}
    \end{equation}
\end{frame}

\begin{frame}{Continuity: Base space}
    The surjective map $\vindex: \gbase \rightarrow \dbase$ 
    \begin{equation}
        \begin{tikzcd}[ampersand replacement=\&]
            \dtotal \arrow[d, "\pi"'] \& \gtotal \arrow[d, "\pi"'] \\
            \dbase                   \& \gbase \arrow[l, "\vindex"']
        \end{tikzcd}
    \end{equation}
    goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. 
\end{frame}

\begin{frame}{Sample Display Target: Fiber}
    Assume a 2D opaque image $\gfiber=\reals^5$ with elements 
    \begin{equation}
    (x,\, y,\, r,\, g,\, b) \in \gfiber
    \end{equation}
    such that a rendered graphic only has 2D position and color.
\end{frame}

\begin{frame}{Continuity Retraction Maps}   
    \begin{figure}[H]
        \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
        \label{fig:graphic_retraction_map}
    \end{figure}
\end{frame}

\begin{frame}{Rendering: Define a Pixel}
    \begin{columns}
        \column{0.5\textwidth}
        Given a pixel
        \begin{equation}
        p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]
        \end{equation}
        the inverse map of the bounding box 
        \begin{equation}
        \gbase_{p} ={\gsection_{xy}}^{-1}(p)
        \end{equation}
        is a region $\gbase_p \subset \gbase$ such that 
        \begin{align}
            \scriptstyle r_p &= \scriptstyle \iint\limits_{S_p} \rho_r(s)ds^{2}\\
            \scriptstyle g_p &= \scriptstyle \iint\limits_{S_p} \rho_g(s)ds^{2}\\
            \scriptstyle  b_p &= \scriptstyle \iint\limits_{S_p} \rho_b(s)ds^{2}
        \end{align}
        yields the color of the pixel. 
        \column{0.5\textwidth}
        \begin{figure}[H]
            \includegraphics[width=\textwidth]{figures/math/render.png}
        \end{figure}
    \end{columns}
\end{frame}

\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\end{document}

### 2021-03-05 slides/table.tex
slides up to graphics

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
\begin{table}[]
\centering
\begin{tabularx}{\textwidth}{lllllll}
\cline{1-4}
\multicolumn{2}{|l|}{\multirow{2}{*}{}}                                  & \multicolumn{2}{c|}{values}                                        &  &  &  \\ \cline{3-4}
\multicolumn{2}{|l|}{}                                                   & \multicolumn{1}{l|}{categorical} & \multicolumn{1}{l|}{quantitative} &  &  &  \\ \cline{1-4}
\multicolumn{1}{|l|}{\multirow{2}{*}{keys}} & \multicolumn{1}{l|}{categorical}   & \multicolumn{1}{l|}{\includegraphics[height=.15\textheight]{figs/catcat.png}}           & \multicolumn{1}{l|}{\includegraphics[height=.15\textheight]{figs/quantcat.png}}          &  &  &  \\ \cline{2-4}
\multicolumn{1}{|l|}{}                      & \multicolumn{1}{l|}{quantitative} & \multicolumn{1}{l|}{\includegraphics[height=.15\textheight]{figs/catquant.png}}           & \multicolumn{1}{l|}{\includegraphics[height=.15\textheight]{figs/quantquant.png}}          &  &  &  \\ \cline{1-4}
                                            &                            &                                  &                                 &  &  &  \\
                                            &                            &                                  &                                 &  &  &  \\
                                            &                            &                                  &                                 &  &  & 
\end{tabularx}
\caption{Bertin's Visualizations of Observations}

\end{table}

### 2021-03-08 slides/main.tex
shelve intro for now

@@
 \documentclass[xcolor={dvipsnames}, handout]{beamer}
+
 \usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
@@
 
-\title{Topological Artist Model}
+\title{Topological Equivariant Artist Model}
 
@@
 
-\begin{frame}{What do visualization libraries do?}
+\begin{frame}{Visualizations are structure preserving maps}
     \begin{figure}
@@
     \end{figure}
+    The aim of this work is to rearchitecture Matplotlib
+    \pause 
+    to take advantage of developments in software design, data structures, and visualization
+    \pause
+    to improve consistency, reusability, and discoverability,
+    \pause
+    so domain specific tool developers can build structure preserving visualization tools.
 \end{frame}
 
-\begin{frame}{Why are we doing this?}
-    \begin{enumerate}
-        \item visualization software design patterns are tuned to data structures\cite{HeerSoftware2006}
-        \item visualization types tuned to data structures\cite{toryRethinkingVisualizationHighlevel2004}
-        \item need to build a tool that supports domain specific structures and algorithms
-        \item tool needs to be general purpose enough to support lots of domains
-        \item tools needs to support 2D, 3D, static, dynamic, and interactive viz
-    \end{enumerate}
+\begin{frame}{Visualization component constraints}
+    \begin{description}
+        \item[equivariance] properties of data and visual encoding match
+        \item[continuity] connectivity of data and visual encoding match
+        \item[composibility] structure preserved by individual components is preserved in combined components
+    \end{description}
 \end{frame}
 
-\subsection{Tools}
-\begin{frame}{Everything is a table}
+\subsection{Background}
+\begin{frame}{Tools are tuned to the continuity of the date \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}}
     \begin{columns}
-    \column{0.5\textwidth}
+        \column{0.33\textwidth}
         \begin{figure}
-        \includegraphics[width=1\textwidth]{figures/intro/grammar_example.png}
-        \caption{Figure 1.5 in Wilkenson's Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}}
+            \includegraphics[width=1\textwidth]{figures/intro/data_formatting.png}
+            \caption{Based on fig~2.5 in Munzner's VAD\cite{VisualizationAnalysisDesign}}
         \end{figure}
-    \column{0.5\textwidth}
+
         \begin{enumerate}
-            \item  ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, altair\cite{vanderplasAltairInteractiveStatistical2018}
-            \item tends to be declarative\cite{heerDeclarative2010} 
-            \item users describe how to compose visual elements \cite{wongsuphasawatNavigatingWideWorld2021}
+            \item ggplot\cite{wickhamGgplot2ElegantGraphics2016a}
+            \item protovis\cite{bostockProtoviz2009}, D3 \cite{bostockDataDrivenDocuments2011}
+            \item vega\cite{satyanarayanDeclarativeInteractionDesign2014}, altair\cite{vanderplasAltairInteractiveStatistical2018}
         \end{enumerate}
-    \end{columns}
         \pause
-    \begin{center}
-        \textbf{TAM: build the visual elements}
-    \end{center}
-\end{frame}
-
-\begin{frame}{Everything is an image}
-    \begin{columns}
-        \column{0.5\textwidth}
+        \column{0.33\textwidth}
         \begin{figure}
@@
         \end{figure}
-        \column{0.5\textwidth}
         \begin{enumerate}
-            \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}, Napari\cite{nicholas_sofroniew_2021_4533308}
-            \item build plugins into existing system where image is primary input
+            \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}\item Napari\cite{nicholas_sofroniew_2021_4533308}
+        \end{enumerate}
+        \pause
+        \column{0.33\textwidth}
+        \begin{figure}
+            \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
+            \caption{Data Representation, MayaVi 4.7.2 docs\cite{DataRepresentationMayavi}}
+        \end{figure}
+        \begin{enumerate}
+            \item Matplotlib\cite{hunterMatplotlib2DGraphics2007}, 
+            \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
         \end{enumerate}
     \end{columns}
-    \pause
-    \begin{center}
-        \textbf{TAM: more general than tables and images}
-    \end{center}
 \end{frame}
 
-\begin{frame}{Everything is ?}
-    \begin{columns}
-        \column{0.5\textwidth}
-        \begin{figure}
-            \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
-            \caption{Image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
-        \end{figure}
-        \column{0.5\textwidth}
-        \begin{enumerate}
-            \item Matplotlib\cite{hunterMatplotlib2DGraphics2007},  VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
-            \item each visualization type API is tuned to the data structure
-        \end{enumerate}
-    \end{columns}
-    \pause
-    \begin{center}
-        \textbf{TAM: common data abstraction interface between data and viz }
-    \end{center}
-\end{frame}
-
-\subsection{Data}
-\begin{frame}{Variables \& Topology}
+\begin{frame}{Structure is encoded in variables and continuity}
     \begin{figure}
@@
     \begin{description}
-        \item  metadata are \textit{keys} with associated \textit{values} (Munzner \cite{munznerVisualizationAnalysisDesign2014})
-        \item[topology] Fiber bundles can be a common data abstraction (Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989})
-        \item[variables] Fiber as a database schema (Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES})
+        \item[binding] metadata are structural \textit{keys} with associated \textit{values} (Munzner \cite{munznerVisualizationAnalysisDesign2014})] 
+        \item[continuity] Fiber bundles can be a common data abstraction (Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989})
+        \item[variables] Fibers can hold schema like encodings of variables (Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES})
     \end{description}
-    \pause
-    \begin{center}
-        \textbf{Tam: variables are \textit{values} and \textit{keys} locate them in topology}
-    \end{center}
 \end{frame}
 
-\subsection{Visualization}
-\begin{frame}{Property Matching}
-    \begin{figure}
-        \includegraphics[width=.7\textwidth]{figures/intro/retinal_variables.png}
-        \caption{This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from Krygier and Wood's \textit{Making Maps: A Visual Guide to Map Design for GIS}\cite{krygierMakingMapsVisual2005}}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Evaluating visualizations}
+\begin{frame}{Visualizations are (mostly) evaluated on equivariance}
     \begin{description}
@@
 
-\begin{frame}{Mathematical Frameworks for evaluating visualization}
-    \begin{enumerate}
-        \item APT: visualization has syntax and semantics like a language  (Mackinlay  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987})
-        \item Visualization has functional dependencies that can be represented as graphs (Sugibuchi \cite{sugibuchiFramwork2009}) 
-        \item the semiotics of visualization are commutative in a category theory framework (Vickers \cite{vickersUnderstandingViz2013})
-    \end{enumerate}
-    \pause
-    \begin{center}
-        \textbf{Tam: Framework for building visualizations}
-    \end{center}
-\end{frame}
+\begin{frame}{Models describe composition}
+    \begin{description}
+        \item[language model] APT: syntax and semantics of visualization (Mackinlay  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987})
+        \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
+        \item[category theory] the semiotics of visualization are commutative (Vickers \cite{vickersUnderstandingViz2013})
+        \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
+        \begin{columns}
+            \column{.5\textwidth}
 
-\begin{frame}{Visualization is commutative maps}
-    data ($\alpha$) and viz ($\omega$) symmetries (Kindlmann and Scheidegger \cite{kindlmann2014algebraic})
-        \begin{block}{$v\circ r_2 \circ \alpha =\omega\circ v\circ r_1$}
+            \begin{description}
+                \item[D] data 
+                \item[R] representations
+                \item[V] visualizations
+            \end{description}
+            \column{.5\textwidth}
             \begin{equation*}
@@
                 \end{equation*}
-    \end{block}
-    \pause
-    \begin{center}
-        \textbf{Tam: Add topology and make it functional}
-    \end{center}
-\end{frame}
-\begin{frame}{So why?}
-    \begin{figure}[H]
-        \begin{subfigure}{.24\textwidth}
-            \includegraphics[width=1\textwidth]{figures/intro/du_bois_spinny.png}
-            \caption{}
-            \label{fig:intro_dpa}
-        \end{subfigure}
-        \begin{subfigure}{.24\textwidth}
-            \includegraphics[width=1\textwidth]{figures/intro/du_bois_line.png}
-            \caption{}
-            \label{fig:intro_dpb}
-        \end{subfigure}
-        \begin{subfigure}{.24\textwidth}
-            \includegraphics[width=1\textwidth]{figures/intro/du_bois_country.png}
-            \caption{}
-            \label{fig:intro_dbc}
-        \end{subfigure}
-        \begin{subfigure}{.24\textwidth}
-            \includegraphics[width=1\textwidth]{figures/intro/du_bois_heat.png}
-            \caption{}
-            \label{fig:intro_dbd}
-        \end{subfigure}
-        \caption{Du Bois' data portraits from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroValuation1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
-        \label{fig:intro_dubois}
-    \end{figure}
-    \begin{center}
-        \textbf{Framework for building any valid visualization}
-    \end{center}
+
+        \end{columns}
+
+    \end{description}
 \end{frame}
@@
 \begin{frame}{Contributions}
-    \begin{enumerate}
-        \item a formal description of the topology preserving relationship between data and graphic via continuous maps
-        \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
-        \item abstraction of data structure using fiber bundles with schema like fibers to encode components and topology 
-        \item algebraic sum operator such that more complex visualizations can be built from simple ones 
-        \item a functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
-        \item a prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
-    \end{enumerate}
+    \begin{description}
+        \item[Topological] topology preserving relationship between data and graphic via continuous maps
+        \item[Equivariant] property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
+        \item[Artist] functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
+        \item [Model] prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
+    \end{description}
 \end{frame}
@@
 
-\begin{frame}{Topological Artist Model}
+\begin{frame}{Topological Equivariant Artist Model}
+    The Artist $\mathscr{\vartist}$ is a map from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$
     \begin{equation}
@@
     \pause
-    which is a map from data \dtotal\ to graphic \gtotal\ that carries a homomorphism of monoid actions 
-    \pause 
+     that carries a homomorphism of monoid actions 
     \begin{equation}
@@
     \end{equation}
+    \pause 
+    with a deformation retraction from graphic to data space. 
+    \note{some kinda words about the continuity}
 \end{frame}
@@
 
-\begin{frame}{Fiber Components}
-    We can decouple decouple \dfiber\ into components $\dfiber_i$
-    \begin{equation}
-        \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
-    \end{equation}
-\end{frame}
-
 \begin{frame}{Structure of Components: Monoids}
@@
 \begin{frame}{Monoid Composition: Partial Order}
-    MAKE THIS Hasse Diagram (borrow from theories of composablity)
+    \begin{figure}
+        \begin{overprint}
+            \onslide<1|handout:0>\includegraphics[width=1\linewidth]{figures/math/monoid_hasse.png}
+            \onslide<2|handout:0>\includegraphics[width=1\linewidth]{figures/math/monoid_monotone.png}
+            \onslide<3>\includegraphics[width=1\linewidth]{figures/math/monoid_maps.png}
+        \end{overprint}
+    \caption{Inspired by definition 1.59 diagram in Spivak and Fong's An Invitation to Applied Category Theory \cite{fongInvitationAppliedCategory2019}}
+    \end{figure}
 \end{frame}
@@
 \subsection{Graphics}
-\begin{frame}{Fiber Bundle}
+\begin{frame}{Graphic Bundle}
     The graphics bundle is a tuple $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$ defined by the projection map $\pi$
@@
 \end{frame}
-
 \begin{frame}{Continuity: Base space}
@@
 
-\begin{frame}[allowframebreaks]{References}
-\printbibliography
-\end{frame}
+\subsection{Artist}
+
+\section{References}
+%\begin{frame}[allowframebreaks]{References}
+%\printbibliography
+%\end{frame}
+
 \end{document}

### 2021-03-08 slides/scraps.tex
shelve intro for now

--- 2021-03-05 slides/main.tex
@@
-\documentclass[xcolor={dvipsnames}, handout]{beamer}
-\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
-\usepackage{mathrsfs} % math script fonts
-\usepackage{tikz-cd} % commutative diagrams
-\usepackage{subcaption} % subfigure float captioning
+\subsection{Data}
 
-\usepackage{tabularx}
-\usepackage{multirow}
-\usepackage{graphicx}
-
-\usepackage{notation} % move this later
-
-\graphicspath{.figures/}
-
-\usepackage[backend=bibtex, style=numeric-comp, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
-\bibliography{../draft/glasslab_viz.bib}
-
-\usetheme{ccnycrest}
-
-\newenvironment{changemargin}[2]{%
-\begin{list}{}{%
-\setlength{\topsep}{0pt}%
-\setlength{\leftmargin}{#1}%
-\setlength{\rightmargin}{#2}%
-\setlength{\listparindent}{\parindent}%
-\setlength{\itemindent}{\parindent}%
-\setleng{}th{\parsep}{\parskip}%
-}%
-\item[]}{\end{list}}
-
-\begin{document}
-
-\title{Topological Artist Model}
-
-\begin{frame}
-	\titlepage
-    Hannah Aizenman\\
-    Advisor: Dr. Michael Grossberg \\
-    Committee: Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo\\
-    External Member: Dr. Marcus Hanwell
-\end{frame}
-
-\section{Introduction}
-
-\begin{frame}{What do visualization libraries do?}
-    \begin{figure}
-        \includegraphics[width=1\linewidth]{figures/intro/dar.png}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Why are we doing this?}
-    \begin{enumerate}
-        \item visualization software design patterns are tuned to data structures\cite{HeerSoftware2006}
-        \item visualization types tuned to data structures\cite{toryRethinkingVisualizationHighlevel2004}
-        \item need to build a tool that supports domain specific structures and algorithms
-        \item tool needs to be general purpose enough to support lots of domains
-        \item tools needs to support 2D, 3D, static, dynamic, and interactive viz
-    \end{enumerate}
-\end{frame}
-
-\subsection{Tools}
-\begin{frame}{Everything is a table}
-\begin{columns}
-    \column{0.5\textwidth}
-    \begin{figure}
-        \includegraphics[width=1\textwidth]{figures/intro/grammar_example.png}
-        \caption{Figure 1.5 in Wilkenson's Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}}
-    \end{figure}
-    \column{0.5\textwidth}
-        \begin{enumerate}
-            \item  ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, altair\cite{vanderplasAltairInteractiveStatistical2018}
-            \item tends to be declarative\cite{heerDeclarative2010} 
-            \item users describe how to compose visual elements \cite{wongsuphasawatNavigatingWideWorld2021}
-        \end{enumerate}
-    \end{columns}
-    \pause
-    \begin{center}
-        \textbf{TAM: build the visual elements}
-    \end{center}
-\end{frame}
-
-\begin{frame}{Everything is an image}
-    \begin{columns}
-        \column{0.5\textwidth}
-        \begin{figure}
-            \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
-        \end{figure}
-        \column{0.5\textwidth}
-        \begin{enumerate}
-            \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}, Napari\cite{nicholas_sofroniew_2021_4533308}
-            \item build plugins into existing system where image is primary input
-        \end{enumerate}
-    \end{columns}
-    \pause
-    \begin{center}
-        \textbf{TAM: more general than tables and images}
-    \end{center}
-\end{frame}
-
-\begin{frame}{Everything is ?}
-    \begin{columns}
-        \column{0.5\textwidth}
-        \begin{figure}
-            \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
-            \caption{Image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
-        \end{figure}
-        \column{0.5\textwidth}
-        \begin{enumerate}
-            \item Matplotlib\cite{hunterMatplotlib2DGraphics2007},  VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
-            \item each visualization type API is tuned to the data structure
-        \end{enumerate}
-    \end{columns}
-    \pause
-    \begin{center}
-        \textbf{TAM: common data abstraction interface between data and viz }
-    \end{center}
-\end{frame}
-
-\subsection{Data}
-\begin{frame}{Variables \& Topology}
-    \begin{figure}
-        \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
-        \caption{Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
-    \end{figure}
-    \begin{description}
-        \item  metadata are \textit{keys} with associated \textit{values} (Munzner \cite{munznerVisualizationAnalysisDesign2014})
-        \item[topology] Fiber bundles can be a common data abstraction (Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989})
-        \item[variables] Fiber as a database schema (Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES})
-    \end{description}
-    \pause
-    \begin{center}
-        \textbf{Tam: variables are \textit{values} and \textit{keys} locate them in topology}
-    \end{center}
-\end{frame}
-
-\subsection{Visualization}
 \begin{frame}{Property Matching}
@@
 
+\begin{frame}{Visualization is commutative maps}
+
+    \pause
+    %% deconstruct D\R\V
+    \begin{center}
+        \textbf{Tam: Add topology and make it functional}
+    \end{center}
+\end{frame}
 \begin{frame}{Evaluating visualizations}
@@
         \item APT: visualization has syntax and semantics like a language  (Mackinlay  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987})
-        \item Visualization has functional dependencies that can be represented as graphs (Sugibuchi \cite{sugibuchiFramwork2009}) 
+        \item visualization has functional dependencies that can be represented as graphs (Sugibuchi \cite{sugibuchiFramwork2009}) 
         \item the semiotics of visualization are commutative in a category theory framework (Vickers \cite{vickersUnderstandingViz2013})
-    \end{enumerate}
-    \pause
-    \begin{center}
-        \textbf{Tam: Framework for building visualizations}
-    \end{center}
-\end{frame}
-
-\begin{frame}{Visualization is commutative maps}
-    data ($\alpha$) and viz ($\omega$) symmetries (Kindlmann and Scheidegger \cite{kindlmann2014algebraic})
+        \item data ($\alpha$) and viz ($\omega$) symmetries (Kindlmann and Scheidegger \cite{kindlmann2014algebraic})
         \begin{block}{$v\circ r_2 \circ \alpha =\omega\circ v\circ r_1$}
@@
         \end{block}
-    \pause
-    \begin{center}
-        \textbf{Tam: Add topology and make it functional}
-    \end{center}
 \end{frame}
-\begin{frame}{So why?}
-    \begin{figure}[H]
-        \begin{subfigure}{.24\textwidth}
-            \includegraphics[width=1\textwidth]{figures/intro/du_bois_spinny.png}
-            \caption{}
-            \label{fig:intro_dpa}
-        \end{subfigure}
-        \begin{subfigure}{.24\textwidth}
-            \includegraphics[width=1\textwidth]{figures/intro/du_bois_line.png}
-            \caption{}
-            \label{fig:intro_dpb}
-        \end{subfigure}
-        \begin{subfigure}{.24\textwidth}
-            \includegraphics[width=1\textwidth]{figures/intro/du_bois_country.png}
-            \caption{}
-            \label{fig:intro_dbc}
-        \end{subfigure}
-        \begin{subfigure}{.24\textwidth}
-            \includegraphics[width=1\textwidth]{figures/intro/du_bois_heat.png}
-            \caption{}
-            \label{fig:intro_dbd}
-        \end{subfigure}
-        \caption{Du Bois' data portraits from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroValuation1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
-        \label{fig:intro_dubois}
-    \end{figure}
-    \begin{center}
-        \textbf{Framework for building any valid visualization}
-    \end{center}
-\end{frame}
-
-\subsection{Contributions}
-\begin{frame}{Contributions}
-    \begin{enumerate}
-        \item a formal description of the topology preserving relationship between data and graphic via continuous maps
-        \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
-        \item abstraction of data structure using fiber bundles with schema like fibers to encode components and topology 
-        \item algebraic sum operator such that more complex visualizations can be built from simple ones 
-        \item a functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
-        \item a prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
-    \end{enumerate}
-\end{frame}
-
-\section{Mathematical Framework}
-
-\begin{frame}{Topological Artist Model}
-    \begin{equation}
-        \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
-    \end{equation}
-    \pause
-    which is a map from data \dtotal\ to graphic \gtotal\ that carries a homomorphism of monoid actions 
-    \pause 
-    \begin{equation}
-    \varphi: \monoid \rightarrow \monoid^{\prime}
-    \end{equation}
-    \pause
-    such that artists are equivariant maps 
-    \begin{equation}
-    \mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
-    \end{equation}
-\end{frame}
-
-\subsection{Data Model}
-
-\begin{frame}{Data Bundle}
-    \begin{figure}
-        \includegraphics[width=1\textwidth]{figures/math/fiberbundle.png}
-    \end{figure}
-    a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
-    \begin{equation}
-        \label{eq:fiber_bundle}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \dfiber \arrow[r, hook] \& \dtotal \arrow[r, "\pi"] \& \dbase
-        \end{tikzcd}
-    \end{equation}
-\end{frame}
-
-\begin{frame}{Variables: Fiber}
-    \begin{equation}
-        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
-    \end{equation}
-    where $\ftotal_{\fsection(\fname)}$ is 
-    \begin{equation}
-        \label{eq:data_types}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
-            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
-        \end{tikzcd}
-    \end{equation}
-    \begin{description}
-        \item[\ftypes] data types of the variables in the dataset 
-        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
-        \item[\fnames] variable names
-        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
-    \end{description}
-\end{frame}
-
-\begin{frame}{Sample Fiber}
-    \begin{figure}[H]
-        \begin{subfigure}{.4\textwidth}
-            \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-            \caption{}
-            \label{fig:fiber_example_plane}
-        \end{subfigure}
-        \begin{subfigure}{.4\textwidth}
-            \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
-            \caption{}
-            \label{fig:fiber_example_cube}
-        \end{subfigure}
-        \caption{}
-        \label{fig:data_fiber_example}
-    \end{figure}
-    \begin{description}
-        \item[\ref{fig:fiber_example_plane}]  $\dfiber=\reals\times\reals$, for example (time, temperature)
-        \item[\ref{fig:fiber_example_cube}]  $\reals\times\realsp\times\reals$, for example  (time, wind=(speed, direction))
-    \end{description}
-\end{frame}
-
-\begin{frame}{Fiber Components}
-    We can decouple decouple \dfiber\ into components $\dfiber_i$
-    \begin{equation}
-        \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
-    \end{equation}
-\end{frame}
-
-\begin{frame}{Structure of Components: Monoids}
-A monoid \monoid\ is a set with
-\begin{description}
-    \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
-    \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
-\end{description}
-\end{frame}
-\begin{frame}{Monoid Actions}
-A left monoid action of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
-    \begin{align*}
-        \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-        \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
-    \end{align*}
-\end{frame}
-
-\begin{frame}{Monoid Composition: Permutation}
-    \begin{figure}
-        \includegraphics[width=1\linewidth]{figures/math/monoid_emoji.png}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Monoid Composition: Partial Order}
-    MAKE THIS Hasse Diagram (borrow from theories of composablity)
-\end{frame}
-
-\begin{frame}{Continuity: base space}
-    \begin{figure}[H]
-    \includegraphics[width=\linewidth]{figures/math/k_qspace.png}
-    \caption{}
-    \label{fig:base_space_div}
-\end{figure}
-where the total space can be decomposed into components 
-\begin{equation}
-    \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
-\end{equation}
-\end{frame}
-
-\begin{frame}{Sample: Base Space}
-    \begin{figure}[H]
-        \begin{subfigure}{.4\textwidth}
-            \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-            \label{fig:base_example_discrete}
-        \end{subfigure}
-        \begin{subfigure}{.4\textwidth}
-            \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
-            \label{fig:base_example_continuous}
-        \end{subfigure}
-        \label{fig:base_example}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Values: Section}
-    For any fiber bundle, there exists a map
-    \begin{equation}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \dfiber \arrow[r, hook] \& \dtotal \arrow[d, "\pi"'] \\
-                              \& \dbase \arrow[u, "\dsection"', bend right]
-        \end{tikzcd}
-    \end{equation}
-     such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$.
-
-     The set of all global sections is denoted as $\Gamma(\dtotal)$.
-\end{frame}
-
-\begin{frame}{Record}
-    Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is 
-\begin{equation}
-    \label{eq:section_return}
-    \dsection(\dbasepoint) = (\dbasepoint, (g_{\dfiber_{0}}(\dbasepoint), \ldots, g_{\dfiber_{n}}(\dbasepoint)))
-\end{equation}
-where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber.
-\end{frame}
-
-\begin{frame}{Sample dataset}
-    \begin{figure}[H]
-        \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
-        \label{fig:data_sections}
-    \end{figure}
-\end{frame}
-
-\subsection{Graphics}
-\begin{frame}{Fiber Bundle}
-    The graphics bundle is a tuple $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$ defined by the projection map $\pi$
-    \begin{equation}
-        \label{eq:fiber_bundle}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \gfiber \arrow[r, hook] \& \gtotal \arrow[r, "\pi"] \& \gbase
-        \end{tikzcd}
-    \end{equation}
-\end{frame}
-
-\begin{frame}{Continuity: Base space}
-    The surjective map $\vindex: \gbase \rightarrow \dbase$ 
-    \begin{equation}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \dtotal \arrow[d, "\pi"'] \& \gtotal \arrow[d, "\pi"'] \\
-            \dbase                   \& \gbase \arrow[l, "\vindex"']
-        \end{tikzcd}
-    \end{equation}
-    goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. 
-\end{frame}
-
-\begin{frame}{Sample Display Target: Fiber}
-    Assume a 2D opaque image $\gfiber=\reals^5$ with elements 
-    \begin{equation}
-    (x,\, y,\, r,\, g,\, b) \in \gfiber
-    \end{equation}
-    such that a rendered graphic only has 2D position and color.
-\end{frame}
-
-\begin{frame}{Continuity Retraction Maps}   
-    \begin{figure}[H]
-        \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
-        \label{fig:graphic_retraction_map}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Rendering: Define a Pixel}
-    \begin{columns}
-        \column{0.5\textwidth}
-        Given a pixel
-        \begin{equation}
-        p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]
-        \end{equation}
-        the inverse map of the bounding box 
-        \begin{equation}
-        \gbase_{p} ={\gsection_{xy}}^{-1}(p)
-        \end{equation}
-        is a region $\gbase_p \subset \gbase$ such that 
-        \begin{align}
-            \scriptstyle r_p &= \scriptstyle \iint\limits_{S_p} \rho_r(s)ds^{2}\\
-            \scriptstyle g_p &= \scriptstyle \iint\limits_{S_p} \rho_g(s)ds^{2}\\
-            \scriptstyle  b_p &= \scriptstyle \iint\limits_{S_p} \rho_b(s)ds^{2}
-        \end{align}
-        yields the color of the pixel. 
-        \column{0.5\textwidth}
-        \begin{figure}[H]
-            \includegraphics[width=\textwidth]{figures/math/render.png}
-        \end{figure}
-    \end{columns}
-\end{frame}
-
-\begin{frame}[allowframebreaks]{References}
-\printbibliography
-\end{frame}
-\end{document}

### 2021-03-08 slides/main.tex
up to Q

@@
 \usepackage{mathrsfs} % math script fonts
-\usepackage{tikz-cd} % commutative diagrams
+\usepackage{tikz}
+\usetikzlibrary{calc} %boxes
+\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
+\usetikzlibrary{cd} % commutative diagrams
 \usepackage{subcaption} % subfigure float captioning
@@
 \usepackage{graphicx}
+\usepackage{appendixnumberbeamer}
 
@@
     \end{figure}
-    a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
+    A fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
     \begin{equation}
@@
     \end{equation}
+    where \dtotal\ is the total data space, \dfiber\ is the variable space, and \dbase\ encodes the continuity.
 \end{frame}
@@
 \begin{frame}{Variables: Fiber}
-    \begin{equation}
-        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
-    \end{equation}
-    where $\ftotal_{\fsection(\fname)}$ is 
+    Given a space of all possible values \ftotal\
     \begin{equation}
@@
     \end{equation}
+    a fiber component is the restricted space $\ftotal_{\fsection(\fname)}$. 
+    \begin{equation}
+        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
+    \end{equation}
     \begin{description}
@@
         \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
-        \item[\fnames] variable names
+        \item[\fnames] variable names, $\fname \in \fnames$
         \item[\fttype] \ftotal\ restricted to the data type of a named variable   
@@
 
-\begin{frame}{Sample Fiber}
+\begin{frame}{Variable types are dimensions of the fiber}
     \begin{figure}[H]
-        \begin{subfigure}{.4\textwidth}
+        \begin{subfigure}{.49\textwidth}
             \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
@@
         \end{subfigure}
-        \begin{subfigure}{.4\textwidth}
+        \begin{subfigure}{.49\textwidth}
             \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
@@
     \begin{description}
-        \item[\ref{fig:fiber_example_plane}]  $\dfiber=\reals\times\reals$, for example (time, temperature)
-        \item[\ref{fig:fiber_example_cube}]  $\reals\times\realsp\times\reals$, for example  (time, wind=(speed, direction))
+        \item[\ref{fig:fiber_example_plane}]  $\dfiber=\reals\times\reals$, (time, temperature)
+        \item[\ref{fig:fiber_example_cube}]  $\reals\times\realsp\times\reals$, (time, wind=(speed, direction))
     \end{description}
@@
 
-\begin{frame}{Structure of Components: Monoids}
+\begin{frame}{Structure of Components: Monoid \& Monoid Actions}
 A monoid \monoid\ is a set with
@@
 \end{description}
-\end{frame}
-\begin{frame}{Monoid Actions}
-A left monoid action of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
+\pause
+\begin{block}{left monoid action}
+A set \dfiber\ with an action $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
     \begin{align*}
-        \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-        \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
+        \textbf{associativity}\;& \text{for all } f,g \in \monoid \text{ and } x\in \dfiber,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
+        \textbf{identity}\;& \text{for all } x\in \dfiber, e\in \monoid,\,  e\bullet x = x 
     \end{align*}
+\end{block}
 \end{frame}
 
-\begin{frame}{Monoid Composition: Permutation}
+\begin{frame}{Monoid Actions: Permutation}
     \begin{figure}
@@
 
-\begin{frame}{Monoid Composition: Partial Order}
+\begin{frame}{Why monoids? partial orders}
     \begin{figure}
@@
 
-\begin{frame}{Continuity: base space}
+\begin{frame}{Data Continuity: Base space}
     \begin{figure}[H]
-    \includegraphics[width=\linewidth]{figures/math/k_qspace.png}
-    \caption{}
+    \includegraphics[height=.5\textheight]{figures/math/k_qspace.png}
     \label{fig:base_space_div}
@@
 
-\begin{frame}{Sample: Base Space}
+\begin{frame}{Data connectivity is encoded as the base space}
     \begin{figure}[H]
-        \begin{subfigure}{.4\textwidth}
+        \begin{subfigure}{.49\textwidth}
             \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
+            \caption{}
             \label{fig:base_example_discrete}
         \end{subfigure}
-        \begin{subfigure}{.4\textwidth}
+        \begin{subfigure}{.49\textwidth}
             \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
+            \caption{}
             \label{fig:base_example_continuous}
         \end{subfigure}
+        \caption{}
         \label{fig:base_example}
     \end{figure}
+    \begin{description}
+        \item[\ref{fig:base_example_discrete}] data is discrete 0D points
+        \item[\ref{fig:base_example_continuous}]  data is lies on the 1D continuous interval \dbase\
+    \end{description}
 \end{frame}
@@
     \end{equation}
-     such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$.
-
-     The set of all global sections is denoted as $\Gamma(\dtotal)$.
-\end{frame}
-
-\begin{frame}{Record}
+     s.t. $\pi(\dsection(\dbasepoint)) = \dbasepoint$.  Set of all global sections is denoted $\Gamma(\dtotal)$.
+     \pause
+     \begin{block}{Record}
      Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is 
@@
 where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber.
+\end{block}
 \end{frame}
@@
     \end{figure}
+    \begin{itemize}
+        \item \dfiber\ is $\reals\times\reals$
+        \item \dbase\ is interval $\left[0,1\right]$
+        \item $\dsection^{(1)}$ is a $sin$ function
+        \item $\dsection^{(2)}$ is a $cos$ function
+        \item $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$
+    \end{itemize}
+
 \end{frame}
@@
     \begin{equation}
-        \label{eq:fiber_bundle}
         \begin{tikzcd}[ampersand replacement=\&]
-            \gfiber \arrow[r, hook] \& \gtotal \arrow[r, "\pi"] \& \gbase
+            \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
+                              \& \gbase \arrow[u, "\gsection"', bend right]
         \end{tikzcd}
     \end{equation}
+    where \gsection\ is the fully encoded graphic. 
+    \pause
+    \begin{block}{Example: 2D opaque image}
+    The target display is $\gfiber=\reals^5$ with elements 
+    \begin{equation*}
+    (x,\, y,\, r,\, g,\, b) \in \gfiber
+    \end{equation*}
+    returned by \gsection\ such that a graphic has color and 2D position.
+    \end{block}
+
 \end{frame}
-\begin{frame}{Continuity: Base space}
+\begin{frame}{Graphic Continuity}
     The surjective map $\vindex: \gbase \rightarrow \dbase$
@@
     \end{equation}
-    goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. 
-\end{frame}
-
-\begin{frame}{Sample Display Target: Fiber}
-    Assume a 2D opaque image $\gfiber=\reals^5$ with elements 
-    \begin{equation}
-    (x,\, y,\, r,\, g,\, b) \in \gfiber
-    \end{equation}
-    such that a rendered graphic only has 2D position and color.
-\end{frame}
-
-\begin{frame}{Continuity Retraction Maps}   
+    goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\dbasepoint$ in data space. 
+\pause
     \begin{figure}[H]
@@
 
+\subsection{Artist}
+\begin{frame}{Topological Equivariant Artist Model}
+The topological artist \vartist\ is a monoid equivariant sheaf map 
+\begin{equation}
+    \label{eq:artist_diagram}
+    \begin{tikzcd}[ampersand replacement=\&]
+        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] \& \vtotal \arrow[d, "\pi"] \& \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] \& \gtotal \arrow[ld, "\pi"] \\
+                                              \& \dbase                  \& \gbase \arrow[l, "\vindex"']                                              \&                    
+        \end{tikzcd}
+\end{equation}
+
+where the artist $\vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)$ takes as input $\dtotal^{\prime}=\mathcal{J}^{2}(\dtotal)$. 
+\end{frame}
+
+\begin{frame}{Visual Bundle}
+    The visual bundle is a tuple $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ defined by the projection map $\pi$
+    \begin{equation}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
+                              \& \dbase \arrow[u, "\vsection"', bend right]
+        \end{tikzcd}
+    \end{equation}
+    where \vsection\ is the visual variable encoding\cite{bertinIIPropertiesGraphic2011} of the representation of the data section \dsection. 
+    \begin{block}{Example: position and color}
+        Given an artist with parameters $\{xpos, ypos, color\}$, a sample visual section \vsection\ could be $\{.5, .5, (255, 20,147)\}$
+    \end{block}
+\end{frame}
+
+\begin{frame}{Visual Channel Encoders}
+We define the visual transformers \vchannel\ on components of the data bundle $\dsection_{i}$
+\begin{equation}
+    \label{eq:nu_expanded}
+    \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
+\end{equation}
+as the set of equivariant maps with the constraint 
+\begin{equation}
+    \vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
+\end{equation} 
+where $\varphi:\monoid\rightarrow \monoid^{\prime}$ carries a homomorphism of monoid actions. 
+\end{frame}
+
+\begin{frame}{Example: Nominal Equivariance}
+    \begin{figure}
+        \includegraphics[width=1\textwidth]{figures/math/equivariance_nu.png}
+        \caption{The actions on the text data are the same as the actions on the visual representation of that data as emojis.}
+    \end{figure}
+\end{frame}
+
+\begin{frame}{Visualization Assembly Function}
+    \begin{figure}
+        \includegraphics[width=\textwidth]{figures/math/path_of_q.png}
+    \end{figure}
+\end{frame}
+
+\begin{frame}{Glyph}
+    The glyph is the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$ where the path connected components $\dbasepath \subset \dbase$ are defined 
+    \begin{equation}
+    \dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
+    \end{equation}
+    such that the  path $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1] and $\gbase_{\dbasepathpoint}$ is the region
+    \begin{equation}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \gtotal \arrow[r, shift left] \& \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] \& \& \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
+            \end{tikzcd}
+        \label{eq:mark}
+    \end{equation}
+    such that the glyph is differentiable, in keeping with Ziemkiewicz and Kosara's description of a glyph\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+\end{frame}
+
+\begin{frame}{Visualization Equivariance}
+    If \vmark\ is applied to $\vsection, \vsection^{\prime}$ that generate the same \gsection\, 
+    \begin{equation}
+    \vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
+    \end{equation}
+    then the output of both sections acted on by the same monoid $m$ must be the same. 
+    \pause
+    \begin{figure}[H]
+        \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
+    \end{figure}
+\end{frame}
+
+\begin{frame}{Scatter: $\vmark(xpos, ypos)(\alpha, \beta)$}
+    \begin{figure}[H]
+        \includegraphics[width=1\textwidth]{figures/math/scatter.png}
+    \end{figure}
+
+    \begin{align*}
+        x &= size *\alpha \cos(\beta) + xpos \\
+        y &= size *\alpha \sin(\beta) + ypos
+    \end{align*}    
+\end{frame}
+
+\begin{frame}{Line: $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ }
+    \begin{figure}[H]
+        \includegraphics[width=1\textwidth]{figures/math/line.png}
+    \end{figure}
+        \begin{equation*}
+            \lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2},\; 
+            \hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}
+        \end{equation*}
+    \begin{align*}
+     x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
+     y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
+    \end{align*}
+\end{frame}
+
+\begin{frame}{Image $\vmark(xpos, yposcolor)$}
+    \begin{figure}[H]
+        \includegraphics[width=1\textwidth]{figures/math/image.png}
+    \end{figure}
+    \begin{align*}
+        R &= R(\vindex(\alpha, \beta))\\
+        G &= G(\vindex(\alpha, \beta))\\
+        B &= B(\vindex(\alpha, \beta))
+    \end{align*}
+\end{frame}
+
+\section{References}
+\begin{frame}[allowframebreaks]{References}
+%\printbibliography
+\end{frame}
+\appendix 
 \begin{frame}{Rendering: Define a Pixel}
@@
 
-\subsection{Artist}
-
-\section{References}
-%\begin{frame}[allowframebreaks]{References}
-%\printbibliography
-%\end{frame}
-
 \end{document}

### 2021-03-09 slides/main.tex
commo
math section

@@
 
+\begin{frame}{Sheafs}
+    Restriction maps of a sheaf describe how local $\iota^*\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
+
+    \begin{equation}
+        \label{eq:sheaf}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             \& \dtotal \arrow[d, "\pi"']                  \\
+            U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] \& \dbase \arrow[u, "\dsection"', bend right]
+        \end{tikzcd}
+    \end{equation}
+    The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$ such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
+\end{frame}
+
 \subsection{Graphics}
@@
 
-\begin{frame}{Image $\vmark(xpos, yposcolor)$}
+\begin{frame}{Image $\vmark(xpos, ypos, color)$}
     \begin{figure}[H]
@@
 
+\begin{frame}{Assembly Function Factory}
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
+\end{figure}
+\begin{equation}
+\label{eq:qhat_q_s}
+\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
+\end{equation} 
+such \gbasepoint\ can be factored out when $\vpreimg = \gbasepoint$
+\end{frame}
+
+\begin{frame}{Composition of artists}
+    Given the family of artists $(\dtotal_i: i\in I)$ on the same image
+    \begin{equation}
+        + \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
+    \end{equation}
+    the + operator defines a simple composition of artists.
+
+    When artists share a base space 
+    \begin{equation}
+        \dbase_2 \hookrightarrow \dbase_1
+    \end{equation}
+    a composition operator can be defined such that the the artists can be considered to be acting on different components of the same section. 
+\end{frame}
+\begin{frame}{Equivalance class of artists}
+    An approximation of the equivalence class of artists $\vartist^{\prime}$
+    \begin{equation}
+    \vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}
+    \end{equation}
+    roughly treats two artists as equivalent if they 
+    \begin{itemize}
+        \item act on the same visual bundle \vtotal
+        \item have the same assembly function \vmarkd 
+        \item have the same continuity map \vindex 
+    \end{itemize}
+\end{frame}
+
+\section{Prototype}
+
 \section{References}
 \begin{frame}[allowframebreaks]{References}
-%\printbibliography
+\printbibliography
 \end{frame}

### 2021-03-09 draft/sections/next.tex
slides first draft

@@
 This work contributes a mathematical description of the mapping \vartist\ from data to visual representation.
-Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this mode; support a variety of datasets, including  discrete relational tables,, multivariate high resolution spatio temporal datasets, and complex networks.
+Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables,, multivariate high resolution spatio temporal datasets, and complex networks.
 Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications for producing visualization where the structure and properties match those of the input data.

### 2021-03-09 slides/main.tex
slides first draft

@@
 \usepackage{appendixnumberbeamer} 
-
+\usepackage{minted}
+\setminted[python]{fontsize=\scriptsize, 
+                   linenos,
+                   numbersep=8pt,
+                   gobble=0,
+                   frame=lines,
+                   bgcolor=bg,
+                   framesep=3mm} 
 \usepackage{notation} % move this later
@@
 
-\usepackage[backend=bibtex, style=numeric-comp, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
+\usepackage[backend=bibtex, style=numeric-comp, sorting=none, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
 \bibliography{../draft/glasslab_viz.bib}
@@
             \includegraphics[width=1\textwidth]{figures/intro/data_formatting.png}
-            \caption{Based on fig~2.5 in Munzner's VAD\cite{VisualizationAnalysisDesign}}
+            \caption{Based on fig~2.5 in Munzner's VAD\cite{munznerVisualizationAnalysisDesign2014}}
         \end{figure}
@@
         \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
-        \item[Effectiveness] design choices made in deference to perceptual saliency \cite (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
+        \item[Effectiveness] design choices made in deference to perceptual saliency (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
         \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
@@
     \begin{description}
-        \item[language model] APT: syntax and semantics of visualization (Mackinlay  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987})
+        \item[language model] APT, GoG: syntax, semantics, and grammar of graphics (Mackinlay, Wilkenson  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005})
         \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
@@
 \section{Prototype}
+\subsection{Artist}
+\begin{frame}[fragile]{Artist}
+    \begin{minted}{python}
+        class ArtistClass(matplotlib.artist.Artist):
+            def __init__(self, data, transforms, *args, **kwargs):
+                # properties that are specific to the graphic
+                self.data = data 
+                self.transforms = transforms
+                super().__init__(*args, **kwargs)
+
+            def assemble(self, **args):
+                # set the properties of the graphic
+
+            def draw(self, renderer):
+                # returns K, indexed on fiber then key 
+                view = self.data.view(self.axes) 
+                # visual channel encoding applied fiberwise 
+                visual = {p: t['encoder'](view[t['name']])
+                          for p, t in self.transforms.items()}
+                self.assemble(**visual)
+                # pass configurations off to the renderer
+                super().draw(renderer)
+        \end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{Artists: Scatter \& Line}
+\begin{figure}[H]
+    \begin{subfigure}{0.49\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
+    \end{subfigure}
+    \begin{subfigure}{0.49\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/line_1.png}
+    \end{subfigure}
+\end{figure}
+\begin{columns}
+\column{,49\textwidth}
+\begin{minted}{python}
+fig, ax = plt.subplots()
+artist = Point(data, transforms)
+ax.add_artist(artist)
+\end{minted}
+\column{.49\textwidth}
+\begin{minted}{python}
+fig, ax = plt.subplots()
+artist = Line(data, transforms)
+ax.add_artist(artist)
+\end{minted}
+\end{columns}
+\end{frame}
+
+\begin{frame}[fragile]{Artists: Scatter \& Line}
+\begin{minted}{python}
+class Point(mcollections.Collection):
+    def assemble(self, x, y, s, facecolors='C0' ):
+        # construct geometries of the circle glyphs in visual coordinates
+        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
+                    for (xi, yi, si) in zip(x, y, s)] 
+        # set attributes of glyphs, these are vectorized 
+        # circles and facecolors are lists of the same size
+        self.set_facecolors(facecolors)
+\end{minted} 
+    \pause
+\begin{minted}{python}
+class Line(mcollections.LineCollection):
+    def assemble(self, x, y, color='C0'):
+        #assemble line marks as set of segments 
+        segments = [np.vstack((vx, vy)).T for vx, vy in zip(x, y)]
+        self.set_segments(segments)
+        self.set_color(color)
+\end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{Visual Transformations}
+\begin{minted}{python}
+cmap =  color.Categorical({'true':'deeppink', 'false':'deepskyblue'})
+transforms = {'x': {'name': 'v4', 'encoder': lambda x: x},
+                'y': {'name': 'v2', 'encoder': lambda x: x},
+                'facecolors': {'name':'v3', 'encoder': cmap}, 
+                's':{'name': None , 
+                    'encoder': lambda _: itertools.repeat(.02)}}
+\end{minted}
+    \begin{itemize}
+        \item \mintinline{python}{lambda x: x} is identity \vchannel\
+        \item \mintinline{python}|{'name':None}| map into \vfiber\ without corresponding \dsection\
+        \item \mintinline{python}{color.Categorical} is custom \vchannel
+    \end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Custom Complex \vchannel}
+\begin{minted}{python}
+class Categorical:
+    def __init__(self, mapping):
+        # check that the conversion is to valid colors
+        assert(mcolors.is_color_like(color) for color in mapping.values())
+        self._mapping = mapping
+
+    def __call__(self, value):
+        # convert value to a color
+        return [mcolors.to_rgba(self._mapping[v]) for v in values]
+\end{minted}
+\pause
+That we can test for action equivariance
+\begin{minted}{python}
+def test_nominal(values, encoder):
+    m1 = list(zip(values, encoder(values)))
+    random.shuffle(values)
+    m2 = list(zip(values, encoder(values)))
+    assert sorted(m1) == sorted(m2)
+\end{minted}
+\end{frame}
+\subsection{Data}
+\begin{frame}[fragile]{Fiber Bundle}
+\begin{minted}{python}
+@dataclass
+class FiberBundle:
+"""
+Attributes
+----------
+K: {'tables': []}
+F: {variable name: type}
+"""
+    K: dict 
+    F: dict
+\end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{Discrete Connectivity}
+\begin{minted}{python}
+class VertexSimplex: #maybe change name to something else
+    """Fiberbundle is consistent across all sections
+    """
+    FB = FiberBundle({'tables': ['vertex']},  
+            {'v1': float, 'v2': str, 'v3': float})
+
+    def __init__(self, sid = 45, size=1000, max_key=10**10):
+        # create random list of keys
+    def tau(self, k):
+        # e1 is sampled from F1, e2 from F2, etc...
+        return (k, (e1, e2, e3, e4))
+\end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{1D Continuous Connectivity}
+\begin{minted}{python}
+class EdgeSimplex: 
+    FB = FiberBundle({'tables': ['vertex','edge']},
+                         {'x' : float, 'y':  float, 
+                         'color':mtypes.Color()}})
+    def __init__(self, num_edges=4, num_samples=1000): 
+        self.keys = range(num_edge) #edge id
+        self.distances = np.linspace(0,1, num_samples)
+        # half generlized representation of arcs on a circle
+        self.angle_samples = np.linspace(0, 2*np.pi, len(self.keys)+1)
+    @staticmethod
+    def _color(edge):
+        return ['red','orange', 'green','blue'][edge%4]
+    @staticmethod
+    def _xy(edge, distances, start=0, end=2*np.pi):
+        # start and end are parameterizations b/c really there is 
+        angles = (distances *(end-start)) + start
+        return np.cos(angles), np.sin(angles)
+    def tau(self, k): #will fix location on page on revision
+        x, y = self._xy(k, self.distances, 
+                        self.angle_samples[k], self.angle_samples[k+1]) 
+        color = self._color(k) 
+        return (k, (x, y, color))
+\end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{View}
+\begin{minted}{python}
+def view(self, axes):
+    table = defaultdict(list)
+    for k in self.keys:
+        table['index'].append(k)
+        for (name, value) in zip(self.FB.fiber.keys(), self.tau(k)[1]):
+            table[name].append(value)
+    return table
+\end{minted}
+\begin{description}
+    \item [\mintinline{python}{VertexSimplex}] (name, value), value is scaler
+    \item [\mintinline{python}{EdgeSimplex}]  (name, value), value is [x0, ..., xn]
+\end{description}
+\end{frame}
+\begin{frame}[fragile]{Same Artist, Different Data Configurations}
+    \begin{figure}[H]
+        \begin{subfigure}{0.49\textwidth}
+            \includegraphics[width=\textwidth]{figures/code/linec_1.png}
+        \end{subfigure}
+        \begin{subfigure}{0.49\textwidth}
+            \includegraphics[width=\textwidth]{figures/code/lined_1.png}
+        \end{subfigure}
+    \end{figure}
+\begin{minted}{Python}
+simplex.GraphLine(FB, edge_table, vertex_table, 
+                      connect=True)
+simplex.GraphLine(FB, edge_table, vertex_table, 
+                      num_samples=2, connect=False)
+\end{minted}
+\end{frame}
+
+\section{Conclusion}
+\begin{frame}{Summary}
+    \begin{itemize}
+        \item structure preserving maps from data to visual representation:
+            \begin{itemize}
+                \item data and graphics have equivalent continuity
+                \item properties are equivariant under monoid actions
+            \end{itemize}  
+        \item fiber bundles with a schema are structure rich abstractions of 
+            \begin{itemize}
+                \item topologically complex heterogenous data 
+                \item target display spaces
+            \end{itemize}
+        \item model can be iteratively integrated into existing Matplotlib architecture
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Proposed Dissertation}
+\begin{itemize}
+    \item expansion of the mathematical framework to include worked out simple and complex addition
+    \item formalization of definition of equivalance class \vartisteq
+    \item implementation of artist with explicit \vindex\
+    \item specification of interactive visualization
+    \item mathematical formulation of a graphic with axes labeling
+    \item implementation of new prototype artists that do not inherit from Matplotlib artists
+    \item provisional mathematics and implementation of user level composite artists
+    \item proof of concept domain specific user facing library 
+\end{itemize}
+\end{frame}
 
@@
 \end{frame}
+\begin{frame}[fragile]{GraphLine Data Model}
+\begin{minted}{python}
+class GraphLine:
+    def __init__(self, FB, edge_table, vertex_table, num_samples=1000,
+                        connect=False):
+        #set args as attributes and generate distance
+        if connect: # test connectivity if edges are continuous
+            assert edge_table.keys() == self.FB.F.keys()
+            assert is_continuous(vertex_table)
+
+    def tau(self, k):
+        # evaluates functions defined in edge table
+        return(k, (self.edges[c][k](self.distances) 
+                        for c in self.FB.F.keys()))
+
+    def view(self, axes):
+        # walk the edge_vertex table to return the edge function
+        table = defaultdict(list)
+        for (i, (start, end)) in sorted(zip(self.ids, self.vertices), 
+                                            key=lambda v:v[1][0]):
+            table['index'].append(i)
+            # same as view for line, returns nested list
+            for (name, value) in zip(self.FB.F.keys(), self.tau(i)[1]):
+                table[name].append(value)
+        return table
+\end{minted}
+\end{frame}
 

### 2021-03-10 draft/sections/next.tex
cleaned up some reference quirks

@@
 While the model and prototype demonstrate that generation of simple marks from the data, there is a lot of work left to develop a model that underpins a minimally viable library.
-Foremost is implementing a data object that encodes data with a 2D continous topology and an artist that can consume data with a 2D topology to visualize the image\cite{ToryRethinkingVisualization2004, haber1990visualization,hansen2011visualization} and also encoding a separate heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas} artist that consumes 1D discrete data.
+Foremost is implementing a data object that encodes data with a 2D continous topology and an artist that can consume data with a 2D topology to visualize the image\cite{toryRethinkingVisualizationHighlevel2004, haber1990visualization,hansen2011visualization} and also encoding a separate heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas} artist that consumes 1D discrete data.
 A second important proof of concept artist is a boxplot\cite{wickham40YearsBoxplots2011} because it is a graphic that assumes computation on the data side and the glyph is built from semantically defined components and a list of outliers.
@@
 This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
-Since this model formalizes notions of structure preservation, it can swerve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmann2014algebraic} of visualizations with respect to graphical encoding choices.
+Since this model formalizes notions of structure preservation, it can swerve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
 While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.

### 2021-03-10 draft/sections/next.tex
more small fixes

@@
 This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
-Since this model formalizes notions of structure preservation, it can swerve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
+Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
 While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.

### 2021-03-11 paper/paper.tex
paper checkins

\documentclass[journal]{vgtc}                % final (journal style)
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% Please use the ``preprint''  option when producing a preprint version
%% for sharing your article on an open access repository

%% Please note that the use of figures other than the optional teaser is not permitted on the first page
%% of the journal version.  Figures should begin on the second page and be
%% in CMYK or Grey scale format, otherwise, colour shifting may occur
%% during the printing process.  Papers submitted with figures other than the optional teaser on the
%% first page will be refused. Also, the teaser figure should only have the
%% width of the abstract as the template enforces it.

%% These few lines make a distinction between latex and pdflatex calls and they
%% bring in essential packages for graphics and font handling.
%% Note that due to the \DeclareGraphicsExtensions{} call it is no longer necessary
%% to provide the the path and extension of a graphics file:
%% \includegraphics{diamondrule} is completely sufficient.
%%
\ifpdf%                                % if we use pdflatex
  \pdfoutput=1\relax                   % create PDFs from pdfLaTeX
  \pdfcompresslevel=9                  % PDF Compression
  \pdfoptionpdfminorversion=7          % create PDF 1.7
  \ExecuteOptions{pdftex}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.pdf,.png,.jpg,.jpeg} % for pdflatex we expect .pdf, .png, or .jpg files
\else%                                 % else we use pure latex
  \ExecuteOptions{dvips}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.eps}     % for pure latex we expect eps files
\fi%

%% it is recomended to use ``\autoref{sec:bla}'' instead of ``Fig.~\ref{sec:bla}''
\graphicspath{{figures/}{pictures/}{images/}{./}} % where to search for the images

\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
\usepackage{tabu}                      % only used for the table example
\usepackage{booktabs}                  % only used for the table example
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

%% In preprint mode you may define your own headline. If not, the default IEEE copyright message will appear in preprint mode.
%\preprinttext{To appear in IEEE Transactions on Visualization and Computer Graphics.}

%% In preprint mode, this adds a link to the version of the paper on IEEEXplore
%% Uncomment this line when you produce a preprint version of the article 
%% after the article receives a DOI for the paper from IEEE
%\ieeedoi{xx.xxxx/TVCG.201x.xxxxxxx}

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{0}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}
%% please declare the paper type of your paper to help reviewers, only shown in review mode
%% choices:
%% * algorithm/technique
%% * application/design study
%% * evaluation
%% * system
%% * theory/model
\vgtcpapertype{Theory/Model}

%% Paper title.
\title{Topological Equivariance Artist Model}

%% This is how authors are specified in the journal style

%% indicate IEEE Member or Student Member in form indicated below
\author{A, B, C}
\authorfooter{
%% insert punctuation at end of each item
\item
 A B are with Computer Science, City College of New York. E-mail: roy.g.biv@aol.com.
\item
 Thomas Caswell is with Brookhaven National Lab E-mail: ed.grimley@aol.com.
}

%other entries to be set up for journal
\shortauthortitle{ \MakeLowercase{\textit{et al.}}: Topological Equivariant Artist Model}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
\abstract{ A critical aspect of data visualization is that the graphical representation of data is expected to match the properties of the data; this fails when order is not preserved in representations of ordinal data or scale for numerical data. In this work, we propose that the mathematical notions of equivariance and topology formalizes the expectation of matching properties. We developed a model we call the topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles. This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements. Furthermore, our model supports an algebraic sum operation such that more complex visualizations can be built from simple ones. We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap. We show that this model can be implemented with a small prototype.} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{}

%% ACM Computing Classification System (CCS). 
%% See <http://www.acm.org/class/1998/> for details.
%% The ``\CCScat'' command takes four arguments.

\CCScatlist{ % not used in journal version
 \CCScat{K.6.1}{Management of Computing and Information Systems}%
{Project and People Management}{Life Cycle};
 \CCScat{K.7.m}{The Computing Profession}{Miscellaneous}{Ethics}
}

%% A teaser figure can be included as follows
\teaser{
  \centering
  \includegraphics[width=\linewidth]{teaser.png}
  \caption{this is our model basically...}
  \label{fig:teaser}
}

%% Uncomment below to disable the manuscript note
%\renewcommand{\manuscriptnotetxt}{}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

\vgtcinsertpkg

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}

\maketitle

%% \section{Introduction} %for journal use above \firstsection{..} instead
This template is for papers of VGTC-sponsored conferences such as IEEE VIS, IEEE VR, and ISMAR which are published as special issues of TVCG. The template does not contain the respective dates of the conference/journal issue, these will be entered by IEEE as part of the publication production process. Therefore, \textbf{please leave the copyright statement at the bottom-left of this first page untouched}.

\section{Background}
\begin{figure}[tb]
  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
  \includegraphics[width=\columnwidth]{paper-count-w-2015-new}
  \caption{A visualization of the 1990--2015 data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
  \label{fig:sample}
 \end{figure}

\section{Model}

\section{Prototype}

\begin{figure}[tb]
 \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
 \includegraphics[width=\columnwidth]{paper-count-w-2015-new}
 \caption{A visualization of the 1990--2015 data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
 \label{fig:sample}
\end{figure}

\subsubsection{Ejector Seat Reservation}

Duis autem~\cite{Lorensen:1987:MCA}\footnote{The algorithm behind
Marching Cubes \cite{Lorensen:1987:MCA} had already been
described by Wyvill et al. \cite{Wyvill:1986:DSS} a year
earlier.} vel eum iriure dolor in hendrerit
in vulputate velit esse molestie consequat,\footnote{Footnotes
appear at the bottom of the column.} vel illum dolore eu
feugiat nulla facilisis at vero eros et accumsan et iusto odio
dignissim qui blandit praesent luptatum zzril delenit augue duis
dolore te feugait nulla facilisi. Lorem ipsum dolor sit amet,
consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt
ut laoreet dolore magna aliquam erat volutpat.

\paragraph{Confirmed Ejector Seat Reservation}

Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
suscipit lobortis nisl ut aliquip ex ea commodo
consequat~\cite{Nielson:1991:TAD}. Duis autem vel eum iriure dolor in
hendrerit in vulputate velit esse molestie consequat, vel illum dolore
eu feugiat nulla facilisis at vero eros et accumsan et iusto odio
dignissim qui blandit praesent luptatum zzril delenit augue duis
dolore te feugait nulla facilisi.

\paragraph{Rejected Ejector Seat Reservation}

Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
suscipit lobortis nisl ut aliquip ex ea commodo consequat. Duis autem
vel eum iriure dolor in hendrerit in vulputate velit esse molestie

\subsection{Vestibulum}

Vestibulum ut est libero. Suspendisse non libero id massa congue egestas nec at ligula. Donec nibh lorem, ornare et odio eu, cursus accumsan felis. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Donec dapibus quam vel eros mattis, id ornare dolor convallis. Donec at nisl sapien. Integer fringilla laoreet tempor. Fusce accumsan ante vel augue euismod, sit amet maximus turpis mattis. Nam accumsan vestibulum rhoncus. Aenean quis pellentesque augue. Suspendisse sed augue et velit consequat bibendum id nec est. Quisque erat purus, ullamcorper ut ex vel, dapibus dignissim erat.

Quisque sit amet orci quam. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Aliquam pharetra, nunc non efficitur convallis, tellus purus iaculis lorem, nec ultricies dolor ligula in metus. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Aenean eu ex nulla. Morbi id ex interdum, scelerisque lorem nec, iaculis dui. Phasellus ultricies nunc vitae quam fringilla laoreet. Quisque sed dolor at sem vestibulum fringilla nec ac augue. Ut consequat, velit ac mattis ornare, eros arcu pellentesque erat, non ultricies libero metus nec mi. Sed eget elit sed quam malesuada viverra. Quisque ullamcorper, felis ut convallis fermentum, purus ligula varius ligula, sit amet tempor neque dui non neque. Donec vulputate ultricies tortor in mollis.

Integer sit amet dolor sit amet turpis ullamcorper varius. Cras volutpat bibendum scelerisque. Maecenas mauris dolor, gravida eu elit et, sodales consequat tortor. Integer id commodo elit. Pellentesque sollicitudin ex non nulla molestie eleifend. Mauris sagittis metus nec turpis imperdiet, vel ullamcorper nibh tincidunt. Sed semper tempus ex, ut aliquet erat hendrerit id. Maecenas sit amet dolor sollicitudin, luctus nunc sit amet, malesuada justo.

Mauris ut sapien non ipsum imperdiet sodales sit amet ac diam. Nulla vel convallis est. Etiam dapibus augue urna. Aenean enim leo, fermentum quis pulvinar at, ultrices quis enim. Sed placerat porta libero et feugiat. Phasellus ullamcorper, felis id porta sollicitudin, dolor dui venenatis augue, vel fringilla risus massa non risus. Maecenas ut nulla vitae ligula pharetra feugiat non eu ante. Donec quis neque quis lorem cursus pretium ac vulputate quam. Cras viverra tellus vitae sapien pretium laoreet. Pellentesque fringilla odio venenatis ex viverra, quis eleifend tortor ornare. Ut ut enim nunc. Vivamus id ligula nec est dignissim eleifend.

Nunc ac velit tellus. Donec et venenatis mauris. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Ut vitae lectus vel ante mollis congue. Vestibulum at cursus velit. Curabitur in facilisis enim. Vestibulum eget dui aliquet risus laoreet laoreet. Phasellus et est id magna interdum venenatis. Donec luctus vehicula justo sed laoreet. Quisque tincidunt suscipit augue, in molestie sem accumsan sed.
\section{Conclusion}

Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
invidunt ut labore et dolore magna aliquyam erat, sed diam
voluptua. At vero eos et accusam et justo duo dolores et ea
rebum.

%% if specified like this the section will be committed in review mode
\acknowledgments{
The authors wish to thank A, B, and C. This work was supported in part by
a grant from XYZ (\# 12345-67890).}

%\bibliographystyle{abbrv}
\bibliographystyle{abbrv-doi}
%\bibliographystyle{abbrv-doi-narrow}
%\bibliographystyle{abbrv-doi-hyperref}
%\bibliographystyle{abbrv-doi-hyperref-narrow}

\bibliography{template}
\end{document}

### 2021-03-11 paper/paper.tex
intro

@@
 %% please capitalize first letter and insert punctuation after last keyword
-\keywords{}
+\keywords{Taxonomy, Models, Frameworks, Theory }
 
@@
 \firstsection{Introduction}
-
 \maketitle
-
-%% \section{Introduction} %for journal use above \firstsection{..} instead
-This template is for papers of VGTC-sponsored conferences such as IEEE VIS, IEEE VR, and ISMAR which are published as special issues of TVCG.
-The template does not contain the respective dates of the conference/journal issue, these will be entered by IEEE as part of the publication production process.
-Therefore, \textbf{please leave the copyright statement at the bottom-left of this first page untouched}.
+The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization, 
+to improve consistency, reusability, and discoverability, so that domain specific tool developers can build structure preserving visualization tools.
+We present a visualization model that formalizes constraints on componenents of a visualization in terms of equivaraiance, continuity, and composibility.
 

### 2021-03-11 paper/paper.tex
what even is an intro?

@@
 to improve consistency, reusability, and discoverability, so that domain specific tool developers can build structure preserving visualization tools.
-We present a visualization model that formalizes constraints on componenents of a visualization in terms of equivaraiance, continuity, and composibility.
+To enable this rearchitecture, we first present a visualization model that formalizes the structure that a visualization preserves in terms of
+\begin{enumerate}
+  \item equivaraiance of data and visual properties
+  \item continuity of data and the graphic
+  \item composibility of components developed in ths model
+\end{enumerate}
 

### 2021-03-12 paper/paper.tex
gonna delete all this but whatever

--- 2021-03-11 paper/paper.tex
@@
 %% please capitalize first letter and insert punctuation after last keyword
+% http://ieeevis.org/year/2021/info/call-participation/paper-keywords
 \keywords{Taxonomy, Models, Frameworks, Theory }
@@
 %% The ``\CCScat'' command takes four arguments.
-
+%https://www.acm.org/publications/computing-classification-system/1998/top
 \CCScatlist{ % not used in journal version
- \CCScat{K.6.1}{Management of Computing and Information Systems}%
-{Project and People Management}{Life Cycle};
- \CCScat{K.7.m}{The Computing Profession}{Miscellaneous}{Ethics}
-}
+ \CCScat{I.3.6}{Computer Graphics}%
+{Methodology and Techniques}{Graphics data structures and data types};}
 
@@
 \maketitle
-The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization, 
-to improve consistency, reusability, and discoverability, so that domain specific tool developers can build structure preserving visualization tools.
-We present a visualization model that formalizes constraints on componenents of a visualization in terms of equivaraiance, continuity, and composibility.
+The work presented in this paper is motivated by a need for a library of visualization components that developers could use to build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.
+While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
+The core architecture also needs to be robust to the big data needs of many visualization practitioners, and therefore support distributed and streaming data needs.
+To support both exploratory and confirmatory visualization\cite{tukeyWeNeedBoth1980}, this tool needs to support 2D and 3D, static, dynamic and interactive visualizations.
+
+Specifically, this work was driven by a rearchitecture of the Python visualization library Matplotlib\cite{hunterMatplotlib2DGraphics2007} to meet modern data visualization needs.
+We aim to take advantage of developments in software design, data structures, and visualization to improve the consistency, composibility, and discoverability of the API.
+We take a functional approach because functional paradigms encourage writing APIs that are flexible, concise and predictable due to the lack of side effects \cite{loudenProgrammingLanguagesPrinciples2010}.
+Furthermore, by structuring the API in terms of composition of the smallest units of transformation for which we can define correctness, a functional paradigm naturally leads to a library of highly modular components that are composable in such a way that by definition the composition is also correct.
+This allows us to ensure that domain specific visualizations built on top of these components are also correct without needing knowledge of the domain.
+
+In this work, we present a framework for understanding visualization as equivariant maps between topological spaces.
+Using this mathematical formalism, we can interpret and extend prior work and also develop new tools.
+We validate our model by using it to re-design artist and data access layer of Matplotlib, a general purpose visualization tool.
 
@@
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
-  \includegraphics[width=\columnwidth]{paper-count-w-2015-new}
-  \caption{A visualization of the 1990--2015 data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
+  \includegraphics[width=\columnwidth]{}
   \label{fig:sample}
@@
 
-\bibliography{template}
+\bibliography{../draft/glasslab_viz.bib}
 \end{document}

### 2021-03-12 paper/paper.tex
intro + related

--- 2021-03-11 paper/paper.tex
@@
 \maketitle
-The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization, 
-to improve consistency, reusability, and discoverability, so that domain specific tool developers can build structure preserving visualization tools.
-We present a visualization model that formalizes constraints on componenents of a visualization in terms of equivaraiance, continuity, and composibility.
+The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization to improve consistency, reusability, and discoverability, so domain specific tool developers can build structure preserving visualization tools.
 
-\section{Background}
+The contribution of this work is
+\begin{description}
+  \item[Topological] topology preserving relationship between data and graphic via continuous maps
+  \item[Equivariant] property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
+  \item[Artist] functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
+  \item [Model] prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
+\end{description}
+
+\section{Related Work}
+Most information visualization tools are tuned to data continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}.
+For example, the relational database is core to libraries built on top of Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and  the digital humanaties oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro.
+Neither the table nor image model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} explicitly carry around different data representations for all the different types of data they encode.
+Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects.
+This has lead to poor API consistency and brittle code as every visualization type has a very different point of view on how the data is structured.
+We present a model that has a single core data abstraction for the API consistency benefits, but the abstraction is general t enough to provide a common interface for many different types of visualization.
+
+The structure visualizations are expected to preserve is encoded in the connectivity of the records and in the properties of the variables in the dataset.
+Fiber bundles were proposed by Butler as one way to encode the continuity of the data separately from the types of variables \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Our model builds on this work because the abstraction of data and flexible enough to support discrete and ND continuous datasets.
+Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplicial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
+One way of describing the binding between the schame and the continuinity is using the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}.
+Unlike Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology.
+This allows the metadata to be altered, for example by changing the coordinate systems or time resolution, without imposing new semantics on the underlying structure.
+Visual representations of data, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization.
+Expressiveness, as defined by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987, mackinlayAutomatingDesignGraphical1986} is a measure of how much of the structure any map can encode.
+A fully expressive component is one that is equivariant since it has preserved all the structure in the data.
+Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations.
+Mackinlay's A Presentation Tool introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}.
+.
+This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
+This same limitation is explicitly stated in the functional dependency notion of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative.
+The algebraic process model by Kindlmann and Scheideggar proposes that the data and visualization transformations are commutative; while similar to our framework, it is missing any explicit mention of continuity.
+
+\section{Topological Equivariant Artist Model}
 \begin{figure}[tb]
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
-  \includegraphics[width=\columnwidth]{paper-count-w-2015-new}
-  \caption{A visualization of the 1990--2015 data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
+  \includegraphics[width=\columnwidth]{}
   \label{fig:sample}
  \end{figure}
-
-\section{Model}
-
 \section{Prototype}
@@
 
-\bibliography{template}
+\bibliography{../draft/glasslab_viz.tex}
 \end{document}

### 2021-03-12 paper/paper.tex
trying to get the theme working:
fixed bib errors, added start of artist

@@
 %% with other math-related packages, you may want to disable it.
-
+\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
+\usepackage{mathrsfs} % math script fonts
+\usepackage{tikz}
+\usetikzlibrary{cd} % commutative diagrams
+\usepackage{minted}
+\setminted[python]{fontsize=\scriptsize, 
+                   linenos,
+                   numbersep=8pt,
+                   autogobble, 
+                   frame=lines,
+                   bgcolor=bg,
+                   framesep=3mm} 
+\usepackage{notation} % move this later
 %% In preprint mode you may define your own headline. If not, the default IEEE copyright message will appear in preprint mode.
@@
 This has lead to poor API consistency and brittle code as every visualization type has a very different point of view on how the data is structured.
-We present a model that has a single core data abstraction for the API consistency benefits, but the abstraction is general t enough to provide a common interface for many different types of visualization.
+We present a model that has a single core data abstraction for the API consistency benefits, but the abstraction is general enough to provide a common interface for many different types of visualization.
 
@@
 A fully expressive component is one that is equivariant since it has preserved all the structure in the data.
+
 Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations.
 Mackinlay's A Presentation Tool introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}.
-.
 This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
@@
 \section{Topological Equivariant Artist Model}
+We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map from data to graphic
+\begin{equation}
+    \label{eq:artist}
+    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
+\end{equation}
+that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}, which are discussed in detail in section~\ref{sec:data_monoid}.
+Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps
+\begin{equation}
+\mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
+\end{equation}
+such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.
+
+We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
+\subsection{Data Bundle}
 \begin{figure}[tb]
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
-  \includegraphics[width=\columnwidth]{}
+  \includegraphics[width=\columnwidth]{fiberbundle.png}
   \label{fig:sample}
@@
  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
- \includegraphics[width=\columnwidth]{paper-count-w-2015-new}
+ %\includegraphics[width=\columnwidth]{}
  \caption{A visualization of the 1990--2015 data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
@@
 %\bibliographystyle{abbrv-doi-hyperref-narrow}
+\bibliography{../draft/glasslab_viz.bib}
 
-\bibliography{../draft/glasslab_viz.tex}
 \end{document}

### 2021-03-12 draft/sections/math.tex
up to Q
up to monoids

@@
 \end{equation}
-while for ratio data, there must be equivalent scaling 
+while for ratio data, there must be equivalent scaling\cite{weissteinSimilarityTransformation} 
 \begin{equation}
@@
 
-The image $\vmark(xpos, yposcolor)$ in figure~\ref{fig:artist_heatmap} is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
+The image $\vmark(xpos, ypos, color)$ in figure~\ref{fig:artist_heatmap} is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
 Since \dbase\ is 2D continuous space, the indexing variables $(\alpha, \beta)$ define the distance along the space.

### 2021-03-12 paper/paper.tex
up to monoids

@@
 \subsection{Data Bundle}
+Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
+\begin{equation}
+    \label{eq:fiber_bundle}
+    \begin{tikzcd}
+        \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
+    \end{tikzcd}
+\end{equation}
+that binds the components of the data in \dfiber\ to the continuity represented in \dbase.
+The fiber bundle models the properties of data component types \dfiber\ (\ref{sec:data_fiber}), the continuity of records \dbase\ (\ref{sec:data_base}), the collections of records \dsection\ (\ref{sec:data_section}), and the space \dtotal\ of all possible datasets with these components and continuity.
+
+By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on \dtotal\ and focus on the components and continuity.
+\subsubsection{Fiber: Variables}
+
 \begin{figure}[tb]
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
-  \includegraphics[width=\columnwidth]{fiberbundle.png}
-  \label{fig:sample}
+  \includegraphics[width=\columnwidth]{fiber.png}
+  \caption{These two datasets have the same base space \dbase\ of discrete points, but the square is a visual representation of the fiber space  $\dfiber=\reals\times\reals$, while the cube is a representation of $\dfiber=\reals\times\realsp\times\reals$}
+  \label{fig:data_fiber}
 \end{figure}
+
+To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names.
+Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
+He then defines the single variable set \fttype\
+\begin{equation}
+    \label{eq:data_types}
+    \begin{tikzcd}
+        \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] & \ftotal \arrow[d, "\pi"] \\
+        \fnames \arrow[r, "\fsection"']                          & \ftypes       
+    \end{tikzcd}
+\end{equation}
+which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname.
+The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype.
+Given \fsection, the fiber for a one variable dataset is
+\begin{equation}
+    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
+\end{equation}
+where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype.
+A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
+\begin{equation}
+F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
+\end{equation}
+which is equivalent to 
+\begin{equation}
+    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
+\end{equation}
+which allows us to decouple \dfiber\ into components $\dfiber_i$.
+Each component of \dfiber\ is a dimension of the topological fiber space, as illustrated in \autoref{fig:data_fiber}.
+The space $\dfiber=\reals\times\reals$ encodes that the dataset has two quantative continuous variables, while $\dfiber=\reals\times\reals\realsp$ encodes a dataset with three quantative contunous variables, one of which must be positive.
+\subsubsection{Equivariance of Properties: Monoid Actions}
+
+\begin{figure}[tb]
+  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
+ %% \includegraphics[width=\columnwidth]{}
+  \caption{}
+  \label{fig:data_partial_order}
+\end{figure}
+
+While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's scales \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale}, we generalize to monoids to support partial orderings.
+As shown in \autoref{fig:data_partial_order}, a partial ordering allows for multiple values to have the same rank in an ordering.
+
+A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$.
+A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$.
+As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,ActionNLab} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
+\begin{align*}
+    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
+    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
+\end{align*}
+As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
+\begin{equation}
+\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
+\end{equation}
+of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$
+
+Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmann2014algebraic} that must be preserved in the visual representation of the component.
+A secondary advantage of defining structure in terms of monoids is that they are commonly found in functional programming because they specify compositions of transformations \cite{yorgeyMonoidsThemeVariations, stievenMonadJustMonoid2020}.
+
 \section{Prototype}

### 2021-03-14 paper/paper.tex
monoids

@@
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
- %% \includegraphics[width=\columnwidth]{}
-  \caption{}
+  \includegraphics[width=\columnwidth]{monoid_maps.png}
+  \caption{Monoids support partial ordering, such as the two emojis with tilted mouths being at the same rank above and below emojis with different mouth positions.
+The monotonic map on the right is a meta ordering of the partial order.
+Figure inspired by definition 1.59 diagram in Spivak and Fong's An Invitation to Applied Category Theory \cite{fongInvitationAppliedCategory2019}}
   \label{fig:data_partial_order}

### 2021-03-14 paper/paper.tex
base space

--- 2021-03-12 paper/paper.tex
@@
 \subsubsection{Equivariance of Properties: Monoid Actions}
-
-\begin{figure}[tb]
-  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
- %% \includegraphics[width=\columnwidth]{}
-  \caption{}
-  \label{fig:data_partial_order}
-\end{figure}
-
-While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's scales \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale}, we generalize to monoids to support partial orderings.
-As shown in \autoref{fig:data_partial_order}, a partial ordering allows for multiple values to have the same rank in an ordering.
+While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale}, we generalize to monoids to support partial orderings.
+A partial ordering allows for multiple measurement values to have the same rank\cite{fongInvitationAppliedCategory2019}, which is useful for visualizing many types of multi indicator systems\cite{bruggemannRankingPrioritizationMultiindicator2011}.
 
@@
 
-Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmann2014algebraic} that must be preserved in the visual representation of the component.
+Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmannAlgebraicProcessVisualization2014} that must be preserved in the visual representation of the component.
 A secondary advantage of defining structure in terms of monoids is that they are commonly found in functional programming because they specify compositions of transformations \cite{yorgeyMonoidsThemeVariations, stievenMonadJustMonoid2020}.
+\subsubsection{}
 

### 2021-03-14 paper/scraps.tex
base space

\begin{figure}[tb]
    \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
    \includegraphics[width=\columnwidth]{monoid_maps.png}
    \caption{Monoids support partial ordering, such as the two emojis with tilted mouths being at the same rank above and below emojis with different mouth positions. The monotonic map on the right is a meta ordering of the partial order.  Figure inspired by definition 1.59 diagram in Spivak and Fong's An Invitation to Applied Category Theory \cite{fongInvitationAppliedCategory2019}}
    \label{fig:data_partial_order}
  \end{figure}

### 2021-03-14 paper/paper.tex
sections + sheafs

@@
 \section{Topological Equivariant Artist Model}
+\label{sec:math}
 We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map from data to graphic
@@
 such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.
-
 We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
 \subsection{Data Bundle}
+\label{sec:math:data}
 Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
@@
 By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on \dtotal\ and focus on the components and continuity.
-\subsubsection{Fiber: Variables}
-
+\subsubsection{Fiber Space: Variables}
+\label{sec:math:data:fiber}
 \begin{figure}[tb]
@@
 \begin{equation}
+    \label{eq:math:data:fiber:decompose}
     \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
@@
 The space $\dfiber=\reals\times\reals$ encodes that the dataset has two quantative continuous variables, while $\dfiber=\reals\times\reals\realsp$ encodes a dataset with three quantative contunous variables, one of which must be positive.
-\subsubsection{Equivariance of Properties: Monoid Actions}
+
+\subsubsection{Equivariance: Monoid Actions}
+\label{sec:math:data:monoid}
 While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946, leaFormalizationMeasurementScale}, we generalize to monoids to support partial orderings.
@@
 A secondary advantage of defining structure in terms of monoids is that they are commonly found in functional programming because they specify compositions of transformations \cite{yorgeyMonoidsThemeVariations, stievenMonadJustMonoid2020}.
-\subsubsection{}
 
-\section{Prototype}
+\subsubsection{Base Space: Continuity}
+\label{sec:math:data:base}
+ The base space \dbase\ is way to express how the records in \dtotal\ are connected to each other, for example if they are discrete points or if they lie in a 2D continous surface.
+Formally \dbase\ is the quotient space of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
+Our model formulates \dbase\ as akin to an indexing space into \dtotal\ that describes the structure of \dtotal.
 
@@
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
- %\includegraphics[width=\columnwidth]{}
- \caption{A visualization of the 1990--2015 data from \autoref{tab:vis_papers}. The image is from \cite{Isenberg:2017:VMC} and is in the public domain.}
- \label{fig:sample}
+  \includegraphics[width=\columnwidth]{base.png}
+  \caption{These two datasets have the same fiber space  $\dfiber=\reals\times\reals$, but the left dataset has a discrete continuity while the right is 1D continuous over the interval $\left[0,1\right]$.}
+  \label{fig:data_continuity}
  \end{figure}
+
+ As illustrated in ~\autoref{fig:data_continuity}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber.
+As with fibers and monoids, we can decompose the total space into components $\pi:\dtotal_i\rightarrow \dbase$ where
+ \begin{equation}
+    \label{eq:math:data:base:decompose}
+     \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
+ \end{equation}
+ which is a decomposition of \dfiber.
+The \dbase\ remains the same because the connectivity of records does not change just because there are fewer elements in each record.
+By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.
+The explicit topology is a concise way of distinguishing visualizations that appear identical, for example heatmaps and images.
+
+ \subsection{Section: Values}
+ \label{sec:math:data:section}
+ \begin{figure}[tb]
+  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
+  \includegraphics[width=\columnwidth]{fiberbundle.png}
+  \caption{Each section in the fiber bundle is a unique continuous map from base space to fiber encoding the set of records in the dataset.
+The two sections, $\dsection^{(1)}$ and $\dsection^{(2)}$, shown here are in the set of global sections $\Gamma(\dtotal)$.}
+  \label{fig:data_section}
+ \end{figure}
+
+ While the projection function $\pi:\dtotal \rightarrow\dbase$ ties together the base space \dbase\ with the fiber \dfiber, a section $\dsection: \dbase\rightarrow \dtotal$ encodes a dataset.
+A section function takes as input location $\dbasepoint \in \dbase$ and returns a record $\delement \in \dtotal$.
+For any fiber bundle, there exists a map
+ \begin{equation}
+     \begin{tikzcd}
+         \dfiber \arrow[r, hook] & \dtotal \arrow[d, "\pi"'] \\
+                           & \dbase \arrow[u, "\dsection"', bend right]
+     \end{tikzcd}
+ \end{equation}
+  such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$.
+The set of all global sections is denoted as $\Gamma(\dtotal)$.
+As illustrated in \autoref{fig:data_section}, the section is a continuous mapping from a location $\dbasepoint \in \dbase$ on the base space to a record $\delement \in \dfiber$ in the fiber.
+Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is
+ \begin{equation}
+     \label{eq:section_return}
+     \dsection(\dbasepoint) = (\dbasepoint, (g_{\dfiber_{0}}(\dbasepoint), \ldots, g_{\dfiber_{n}}(\dbasepoint)))
+ \end{equation}
+ where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber component.
+This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle.
+As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:base:decompose}, we can decompose \dsection\ as
+ \begin{equation}
+ \label{eq:math:data:section:decompose}
+ \dsection= (\dsection_0,\ldots, \dsection_i, \dots, \dsection_n) 
+ \end{equation}
+ where each section $\dsection_i$ is a variable or set of variables. This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.
+
+ \subsubsection{Sheafs: Gluing Sections }
+ The restriction maps of a sheaf describe how local sections $\iota^{*}\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
+This puts a continuous structure on local sections, which allows for defining a section over a subset in \dbase.
+The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over local neighborhood $U$
+ \begin{equation}
+     \label{eq:sheaf}
+     \begin{tikzcd}
+         \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
+         U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
+     \end{tikzcd}
+ \end{equation}
+ such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
+The glued together  $\iota^*\dsection$  sections over subset \dbase\ map to the graphic generated by \vartist\ for visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.
+
+\section{Prototype}
 

### 2021-03-14 paper/paper.tex
visual bundle + artist

@@
 that binds the components of the data in \dfiber\ to the continuity represented in \dbase.
-The fiber bundle models the properties of data component types \dfiber\ (\ref{sec:data_fiber}), the continuity of records \dbase\ (\ref{sec:data_base}), the collections of records \dsection\ (\ref{sec:data_section}), and the space \dtotal\ of all possible datasets with these components and continuity.
+By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on \dtotal\ and focus on the components and continuity.
 
-By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on \dtotal\ and focus on the components and continuity.
 \subsubsection{Fiber Space: Variables}
@@
 
-\subsubsection{Equivariance: Monoid Actions}
+\subsubsection{Equivariant Variable Properties: Monoid Actions}
 \label{sec:math:data:monoid}
@@
 
- \subsection{Section: Values}
+ \subsubsection{Section: Values}
  \label{sec:math:data:section}
@@
 This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle.
-As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:base:decompose}, we can decompose \dsection\ as
+As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:base:decompose}, \dsection\ can be decomposed into components
  \begin{equation}
@@
  \end{equation}
- where each section $\dsection_i$ is a variable or set of variables. This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.
+ where each section $\dsection_i$ maps into a record on a component $\dfiber_i \in \dfiber$.
+This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.
 
  \subsubsection{Sheafs: Gluing Sections }
- The restriction maps of a sheaf describe how local sections $\iota^{*}\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
+ The restriction maps of a sheaf describe how local sections $\iota^{*}\dsection$ over a neighborhood $U$ around \dbasepoint\ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
 This puts a continuous structure on local sections, which allows for defining a section over a subset in \dbase.
-The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over local neighborhood $U$
+The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$
  \begin{equation}
@@
  such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
-The glued together  $\iota^*\dsection$  sections over subset \dbase\ map to the graphic generated by \vartist\ for visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.
+Gluing together $\iota^*\dsection$ sections is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.
+
+ \subsection{Graphic Bundle}
+ \label{sec:math:graphic}  
+We introduce a graphic bundle to hold the essential information necessary to render a graphical design constructed by the artist.
+As with the data, we can represent the target graphic as a section \gsection\ of a bundle  $(\gtotal, \gbase, \pi, \gfiber)$
+\begin{equation}
+  \begin{tikzcd}[ampersand replacement=\&]
+      \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
+                        \& \gbase \arrow[u, "\gsection"', bend right]
+  \end{tikzcd}
+\end{equation}
+where \gsection\ is the fully encoded graphic.
+To fully specify the visual characteristics of the image, we construct a fiber \gfiber\ that is an infinite resolution version of the target space.
+Typically \gtotal\ is trivial and therefore sections can be thought of as mappings into \gfiber.
+In this work, we assume a 2D opaque image $\gfiber=\reals^5$ with elements
+\begin{equation}
+(x,\, y,\, r,\, g,\, b) \in \gfiber
+\end{equation}
+such that a rendered graphic only consists of 2D position and color. By abstracting the target display space as \gfiber, the model can support different targets, such as a 2D screen or 3D printer.
+
+\subsubsection{Equivariant Topology: Graphic Base Space}
+\label{sec:math:graphic:base}
+\begin{figure}[tb]
+  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
+  \includegraphics[width=\columnwidth]{retraction_maps.png}
+  \caption{The 0D scatter \dbasepoint\ and 1D line \dbasepoint\ are thickened into \gbase\ with coordinates $\gbasepoint=(\gx, \gy)$ that are a region in an idealized 2D screen.
+The image has the same dimension in \gbase\ as in \dbase.}
+  \label{fig:graphic_base}
+ \end{figure}
+
+Just as the \dbase\ encodes the connectivity of the records in the data, we propose an equivalent \gbase\ that encodes the connectivity of the rendered elements of the graphic.
+Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy.
+The surjective map $\vindex: \gbase \rightarrow \dbase$
+\begin{equation}
+    \begin{tikzcd}
+        \dtotal \arrow[d, "\pi"'] & \gtotal \arrow[d, "\pi"'] \\
+        \dbase                   & \gbase \arrow[l, "\vindex"']
+    \end{tikzcd}
+\end{equation}
+goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$.
+While \gbase\ must have the same continuity as \dbase\, it is sometimes the  thickened version shown in \autoref{fig:graphic_base}.
+This thickening is necessary when the dimensionality of \dbase\ is less than the dimensionality of the target display.
+For example, a \dbasepoint\ that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the points in \gfiber\ that visually represent the record at \dbasepoint.
+
+\subsection{Artist}
+The topological artist \vartist\ is a monoid equivariant sheaf map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
+\begin{equation}
+    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
+\end{equation}
+While \vartist\ can usually construct graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
+In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the jet bundle $\dtotal^{\prime}=\mathcal{J}^{2}(\dtotal)$.
+
+Specifically, \vartist is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$ 
+\begin{equation}
+    \label{eq:math:artist:diagram}
+    \begin{tikzcd}
+        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \gtotal \arrow[ld, "\pi"] \\
+                                              & \dbase                  & \gbase \arrow[l, "\vindex"']                                              &                    
+        \end{tikzcd}
+\end{equation}
+where the input can be point wise $\dsection(\dbasepoint)\mid \dbasepoint \in \dbase$.
+The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the latent space of possible parameters of a visualization type, such as a scatter or line plot.
+As with the data and graphic bundles, the visual bundle is  defined by the projection map $\pi$
+\begin{equation}
+    \begin{tikzcd}[ampersand replacement=\&]
+        \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
+                          \& \dbase \arrow[u, "\vsection"', bend right]
+    \end{tikzcd}
+\end{equation}
+where \vsection\ is the visual variable encoding\cite{bertinSemiologyGraphicsDiagrams2011a} of the data section \dsection.
+The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave.
+
+In \autoref{eq:math:artist:diagram}, the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components.
+The continuity map $\vindex:\gbase \rightarrow \dbase$ then pulls back the visual bundle \vtotal\ over \gbase.
+Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
+This functional decomposition of the visualization artist facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
+We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.
 
@@
 The authors wish to thank A, B, and C. This work was supported in part by
-a grant from XYZ (\# 12345-67890).}
+a grant from Chan Zuckerberg Institute  (\# 12345-67890).}
 

### 2021-03-14 slides/main.tex
bibtex thing
fixed bib errors, added start of artist
sections + sheafs
visual bundle + artist

@@
 \usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
+\usepackage{bm}
 \usepackage{mathrsfs} % math script fonts
 \usepackage{tikz}
-\usetikzlibrary{calc} %boxes
-\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
+
 \usetikzlibrary{cd} % commutative diagrams
@@
 
-\usepackage{tabularx}
+\usepackage{tabulary}
 \usepackage{multirow}
 \usepackage{graphicx}
+
 \usepackage{appendixnumberbeamer} 
@@
                    numbersep=8pt,
-                   gobble=0,
+                   autogobble, 
                    frame=lines,
@@
     \end{equation}
-    where \vsection\ is the visual variable encoding\cite{bertinIIPropertiesGraphic2011} of the representation of the data section \dsection. 
+    where \vsection\ is the visual variable encoding\cite{bertinIIPropertiesGraphic2011} of the data section \dsection. 
     \begin{block}{Example: position and color}
@@
 \end{frame}
+\begin{frame}{\vfiber\ Components}
+
+\begin{table}[H]
+    \renewcommand{\arraystretch}{2}
+    \begin{tabulary}{\textwidth}{|l|L|l|}\hline
+     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i}) \subset \vfiber_{i}}$  \\ \hline
+    position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
+    size                        & linewidth, markersize                                            & $\mathbb{R}^{+}$   \\ \hline
+    shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
+    color                       & color, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
+    \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
+                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
+    \end{tabulary}
+    \label{tab:mpl_visual_variable_fiber}
+\end{table}
+\end{frame}
 \begin{frame}[fragile]{GraphLine Data Model}

### 2021-03-14 paper/paper.tex
nu

@@
 \end{equation}
-that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}, which are discussed in detail in section~\ref{sec:data_monoid}.
+that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}, which are discussed in detail in \autoref{sec:math:data:monoid}.
 Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps
@@
 
+\subsubsection{Equivariant Visual Component Maps}
+\label{sec:math:artist:nu}
+\begin{figure}[tb]
+  \centering
+  \includegraphics[width=\columnwidth]{equivariance_nu.png}
+  \caption{In this artist, \vchannel\ maps the strings to the emojis.
+This \vchannel\ is equivariant because the monoid actions, represented by the colored arrows, are the same on the input \dsection\ and output \vsection\ sets.}
+  \label{fig:math:artist:nu}
+\end{figure}
+ We define the visual transformers \vchannel\ 
+\begin{equation}
+  \label{eq:math:artist:nu}
+  \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
+\end{equation}
+as the set of equivariant maps $\vchannel_i: \dsection_i \mapsto \vsection_i$.
+Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve.
+As mentioned in \autoref{sec:math:data:monoid}, monoid actions define the structure on the fiber components and are therefore the basis for equivariance.
+A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes such that 
+\begin{equation}
+  \label{eq:math:artist:nu_commute}
+\begin{tikzcd}
+  \dtotal_i \arrow[r] \arrow[r, "\vchannel_i"] \arrow[d, "m_{\delement}"'] & \vtotal_i \arrow[d, "m_{\velement}"] \\
+  \dtotal_i \arrow[r, "\vchannel_i"]                           & \vtotal_i               
+\end{tikzcd}
+\end{equation}
+In general, the data fiber $\dfiber_{i}$ cannot be assumed to be of the same type as the visual fiber $\vfiber_{i}$ and the actions of \monoid\ on $\dfiber_{i}$ cannot be assumed to be the same as the actions of $\monoid^{\prime}$ on \vfiber; therefore an equivariant $\vchannel_i$ must satisfy the constraint
+\begin{equation}
+\vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
+\end{equation} 
+such that $\varphi$ maps a monoid action on data to a monoid action on visual elements.
+However, we can construct a monoid action of \monoid\ on $\vfiber_{i}$ that is compatible with a monoid action of \monoid\ on $\dfiber_{i}$.
+We can compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi$ that takes \monoid\ to $\monoid^{\prime}$.
+This allows us to define a monoid action on \vfiber\ of \monoid\ that is $(m, \velement) \rightarrow \varphi(m)\bullet\velement$.
+Therefore, without a loss of generality, we can assume that an action of \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.
+
+One example of an equivariant \vchannel\ is illustrated in \autoref{fig:math:artist:nu}, which is a mapping from \textbf{Strings} to symbols.
+The data is an example of a Steven's nominal measurement set, which is defined as having on it permutation group actions
+\begin{equation}
+  \text{ if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)
+\end{equation}
+such that shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
+This is illustrated in the identical actions, represented by the colored arrows, on the words and emojis.
+To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$,
+\begin{equation}
+\text{ if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)
+\end{equation}
+the visual encodings must also have some sort of ordering. For interval scale data, \vchannel\ is equivariant under translation monoid actions if 
+\begin{equation}
+\vchannel (x + c) = \vchannel(x) + c
+\end{equation}
+while for ratio data, there must be equivalent scaling\cite{weissteinSimilarityTransformation} 
+\begin{equation}
+\vchannel(xc) = \vchannel(x)*c 
+\end{equation}
+We therefore can test if a \vchannel\ is equivariant by testing the actions under which it must  commute.
+For example, given a a transform $\vchannel_{i}(x) = .5$ on interval data, it must commute under translation.
+Testing this constraint with a translation action $t(x) = x+2$
+\begin{align}
+  \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + 2\\
+  .5 &\neq .5 + 2
+\end{align}
+\vchannel\ does not commute and is therefore invalid.
+The constraints on \vchannel\ can be embedded into the artist such that the \vchannel\ functions can test for equivariance and also provide guidance on constructing new \vchannel\ functions.
+
 \section{Prototype}

### 2021-03-15 paper/paper.tex
q

@@
   \includegraphics[width=\linewidth]{teaser.png}
-  \caption{this is our model basically...}
+  \caption{Visualizations consist of topologically equivariant maps.
+There is a set of monoid action equivariant maps from data components to visual components \vchannel\ that are then reduced via \vmark\ into a single graphic, and there is a deformation retraction \vindex\ from graphic continuity to data continuity.}
   \label{fig:teaser}
@@
 The contribution of this work is
-\begin{description}
-  \item[Topological] topology preserving relationship between data and graphic via continuous maps
-  \item[Equivariant] property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
-  \item[Artist] functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
-  \item [Model] prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
-\end{description}
+\begin{enumerate}
+  \item topology preserving relationship between data and graphic via continuous maps
+  \item property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
+  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
+  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
+\end{enumerate}
 
@@
 \label{sec:math:data:fiber}
-\begin{figure}[tb]
+\begin{figure}[htb]
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
@@
   \caption{These two datasets have the same base space \dbase\ of discrete points, but the square is a visual representation of the fiber space  $\dfiber=\reals\times\reals$, while the cube is a representation of $\dfiber=\reals\times\realsp\times\reals$}
-  \label{fig:data_fiber}
+  \label{fig:math:data:fiber}
 \end{figure}
@@
 which allows us to decouple \dfiber\ into components $\dfiber_i$.
-Each component of \dfiber\ is a dimension of the topological fiber space, as illustrated in \autoref{fig:data_fiber}.
+Each component of \dfiber\ is a dimension of the topological fiber space, as illustrated in \autoref{fig:math:data:fiber}.
 The space $\dfiber=\reals\times\reals$ encodes that the dataset has two quantative continuous variables, while $\dfiber=\reals\times\reals\realsp$ encodes a dataset with three quantative contunous variables, one of which must be positive.
@@
 
- \begin{figure}[tb]
+ \begin{figure}[htb]
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
@@
   \caption{These two datasets have the same fiber space  $\dfiber=\reals\times\reals$, but the left dataset has a discrete continuity while the right is 1D continuous over the interval $\left[0,1\right]$.}
-  \label{fig:data_continuity}
+  \label{fig:math:data:base}
  \end{figure}
 
- As illustrated in ~\autoref{fig:data_continuity}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber.
+ As illustrated in ~\autoref{fig:math:data:base}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber.
 As with fibers and monoids, we can decompose the total space into components $\pi:\dtotal_i\rightarrow \dbase$ where
@@
  \label{sec:math:data:section}
- \begin{figure}[tb]
+ \begin{figure}[htb]
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
@@
 The two sections, $\dsection^{(1)}$ and $\dsection^{(2)}$, shown here are in the set of global sections $\Gamma(\dtotal)$.}
-  \label{fig:data_section}
+  \label{fig:math:data:section}
  \end{figure}
@@
 The set of all global sections is denoted as $\Gamma(\dtotal)$.
-As illustrated in \autoref{fig:data_section}, the section is a continuous mapping from a location $\dbasepoint \in \dbase$ on the base space to a record $\delement \in \dfiber$ in the fiber.
+As illustrated in \autoref{fig:math:data:section}, the section is a continuous mapping from a location $\dbasepoint \in \dbase$ on the base space to a record $\delement \in \dfiber$ in the fiber.
 Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is
@@
 \label{sec:math:graphic:base}
-\begin{figure}[tb]
+\begin{figure}[htb]
   \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
@@
 The image has the same dimension in \gbase\ as in \dbase.}
-  \label{fig:graphic_base}
+  \label{fig:math:graphic:base}
  \end{figure}
@@
 goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$.
-While \gbase\ must have the same continuity as \dbase\, it is sometimes the  thickened version shown in \autoref{fig:graphic_base}.
+While \gbase\ must have the same continuity as \dbase\, it is sometimes the  thickened version shown in \autoref{fig:math:graphic:base}.
 This thickening is necessary when the dimensionality of \dbase\ is less than the dimensionality of the target display.
@@
 
-\subsubsection{Equivariant Visual Component Maps}
+\subsubsection{Visual Component Maps}
 \label{sec:math:artist:nu}
-\begin{figure}[tb]
+\begin{figure}[htb]
   \centering
@@
 The constraints on \vchannel\ can be embedded into the artist such that the \vchannel\ functions can test for equivariance and also provide guidance on constructing new \vchannel\ functions.
+
+\subsubsection{Visualization Assembly}
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=\columnwidth]{path_of_q}
+  \caption{$\vchannel_i$ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.}
+  \label{fig:math:artist:q}
+\end{figure}p
+
+As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\, reduce the encodings into a graphic \gsection.
+While it may seem intuitive that visualizations that generate the same graphic should do so consistently given the same input, we formalize this constraint such that it can be specified as a constraint of \vmark.
+We define the equivariant map as  $\vmark: \vsection \mapsto \gsection$ and an action on the subset of graphics $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ that \vmark\ can generate.
+We then define the constraint on \vmark such that if \vmark\ is applied to two visual sections $\vsection,\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection,\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.
+
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=\columnwidth]{diff_type_q.png}
+  \caption{These two glyphs are generated by the same \vmark\ function. The monoid 
+  action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
+  \label{fig:math:arist:graphic}
+\end{figure}
+
+Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$.
+If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
+\begin{equation}
+\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
+\end{equation}
+then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
+The transformed graphic $\gsection^{\prime}$ is equivariant to a transform on the visual bundle $\gsection^{\prime}=\vmark(m\circ \vsection)$ on a section that $\vsection \in \vmark^{-1}(\gsection)$ that must be part of generating \gsection.
+This equivariance is illustrated in figure \autoref{fig:math:artist:graphic}.
+Given fiber $\vfiber=(xpos,\, ypos,\, color,\, thickness)$, then sections $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
+The action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$ and the corresponding action on \gsection\ causes $\vmark(\vsection^{\prime})$ to be the thicker circle in \autoref{fig:math:artist:graphic}.
+
+We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of path connected components $\dbasepath \subset \dbase$ as input:
+\begin{equation}
+\dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
+\end{equation}
+where the path\cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1].
+We define the glyph as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$
+\begin{equation}
+  \begin{tikzcd}
+      \gtotal \arrow[r, shift left] & \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] &  & \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
+      \end{tikzcd}
+  \label{eq:mark}
+\end{equation}
+such that for every glyph there is at least one corresponding region on \dbase, in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.
+
+\begin{figure}
+  \centering
+  \includegraphics{}
+\end{figure}
+
+In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image.
+
+The scatter plot can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
+
+The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in figure~\ref{fig:artist_scatter}:
+\begin{align}
+x &= size *\alpha \cos(\beta) + xpos \\
+y &= size *\alpha \sin(\beta) + ypos
+\end{align}
+
+such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
+Here $size$ can either be defined inside \vmark\ or it could also be a parameter in \vtotal\ that is passed along with (xpos, vpos).
+As seen in figure~\ref{fig:artist_scatter}, a scatter has a direct mapping from a region on $\gbase_{k}$ to its corresponding \dbasepoint.
 

### 2021-03-15 paper/paper.tex
q hat

@@
   \centering
-  \includegraphics{}
+  \includegraphics[width=\columnwidth]{base_q}
+  \caption{The coordinates $(\gx, \gy)$ dictate the color of the region in prerender space \gbase\, which over the whole disk generates the graphical mark.
+The line fiber is thickened with the derivative because the tangent the line needs to be pushed perpendicular to the tangent of (xpos, ypos) in order to have visible thickness.}
+  \label{fig:math:artist:base}
 \end{figure}
@@
 In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image.
-
 The scatter plot can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
-
-The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in figure~\ref{fig:artist_scatter}:
-\begin{align}
+The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:artist:base}
+\begin{align*}
 x &= size *\alpha \cos(\beta) + xpos \\
 y &= size *\alpha \sin(\beta) + ypos
-\end{align}
+\end{align*}
 
 such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
-Here $size$ can either be defined inside \vmark\ or it could also be a parameter in \vtotal\ that is passed along with (xpos, vpos).
-As seen in figure~\ref{fig:artist_scatter}, a scatter has a direct mapping from a region on $\gbase_{k}$ to its corresponding \dbasepoint.
+In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase\.
+As shown in \ref{fig:math:arist:base} line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
+The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
+such that the normal is  $\hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
+\begin{align*}
+ x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
+ y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
+\end{align*}
+
+where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n_1}, \hat{n_2}$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
+
+The image $\vmark(xpos, ypos, color)$ in \autoref{fig:teaser} is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
+The indexing variables $(\alpha, \beta)$ define the distance along the space, which is then used by \vindex\ to map into \dbase\ to lookup the color values
+\begin{equation*}
+R = R(\vindex(\alpha, \beta)), \; G = G(\vindex(\alpha, \beta)), \; B = B(\vindex(\alpha, \beta))
+\end{equation*}
+In the case of an image, the indexing mapper \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.
+
+\subsubsection{Assembly Factory}
+The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
+As shown in eq~\ref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+\begin{figure}[htb]
+  \centering
+    \includegraphics[width=1\columnwidth]{q_hat.png}
+    \caption{Because the pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
+    \label{fig:math:artist:qhat}
+\end{figure}
+
+The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back of \vsection\
+\begin{equation}
+    \vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber
+\end{equation}
+copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase\.
+This set of points \gbasepoint\ are the preimage \vpreimg\ of \dbasepoint.
+
+As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$  from \vsection\ over \dbasepoint\ to the section \vsectionpull\ over \gbasepoint.
+This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that\vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
+In \autoref{fig:math:artist:qhat} each dot on \vfiber is equivalent to the line on \vfiberpull.
+
+Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out.
+When \vmark\ maps visual sections into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to the pulled back visual section \vsectionpull\ then the graphic section \gsection\ evaluated on a visual region \gbasepoint\
+\begin{equation}
+    \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
+    \label{eq:math:artist:qhat}
+\end{equation}
+ is defined as the assembly function \vmark\ with input \vsectionpull\ evaluated on \gbasepoint.Since the pulled back section \vsectionpull\ is the section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function
+\begin{equation}
+\label{eq:math:artist:qhat_q_s}
+\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
+\end{equation} 
+where \vmarkd\ with input \vsection\ is defined to \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.
+
+Factoring out \gbasepoint\ from equation~\ref{eq:math:artist:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
+In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
+This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
+The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
 

### 2021-03-15 paper/paper.tex
End of math section :)

@@
 
+\subsection{Composite and Reusable Artists}
+Given the family of artists $(\dtotal_i: i\in I)$ on the same image $+ \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}$, the + operator defines a simple composition of artists.
+When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the the artists can be considered to be acting on different components of the same section.
+This type of composition is important for creating visualizations where elements need to update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.
+It is impractical to implement an artist for every single graphic; instead we implement an approximation of an the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$.
+Roughly, equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\ but act on different sections \vsection, but we will formalize the definition of the equivalance class in future work.
+
 \section{Prototype}

### 2021-03-15 paper/paper.tex
rework intro

--- 2021-03-15 paper/paper.tex
@@
 %% Paper title.
-\title{Topological Equivariance Artist Model}
+\title{Topological Equivariant Artist Model}
 
@@
 \item
- A B are with Computer Science, City College of New York. E-mail: roy.g.biv@aol.com.
+ A B are with the Computer Science department, City College of New York. E-mail: roy.g.biv@aol.com.
 \item
@@
 The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization to improve consistency, reusability, and discoverability, so domain specific tool developers can build structure preserving visualization tools.
-
-The contribution of this work is
-\begin{enumerate}
-  \item topology preserving relationship between data and graphic via continuous maps
-  \item property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
-  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
-  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
-\end{enumerate}
+The contribution of this work is (a) topology preserving relationship between data and graphic via continuous maps, (b) property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions, (c) functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model, (d) prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
 
 \section{Related Work}
-Most information visualization tools are tuned to data continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}.
-For example, the relational database is core to libraries built on top of Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
-Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and  the digital humanaties oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro.
-Neither the table nor image model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} explicitly carry around different data representations for all the different types of data they encode.
-Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects.
-This has lead to poor API consistency and brittle code as every visualization type has a very different point of view on how the data is structured.
-We present a model that has a single core data abstraction for the API consistency benefits, but the abstraction is general enough to provide a common interface for many different types of visualization.
-
-The structure visualizations are expected to preserve is encoded in the connectivity of the records and in the properties of the variables in the dataset.
-Fiber bundles were proposed by Butler as one way to encode the continuity of the data separately from the types of variables \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-Our model builds on this work because the abstraction of data and flexible enough to support discrete and ND continuous datasets.
-Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplicial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
-One way of describing the binding between the schame and the continuinity is using the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}.
-Unlike Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology.
-This allows the metadata to be altered, for example by changing the coordinate systems or time resolution, without imposing new semantics on the underlying structure.
-Visual representations of data, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}.
 The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization.
@@
 A fully expressive component is one that is equivariant since it has preserved all the structure in the data.
-
 Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations.
@@
 This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
-This same limitation is explicitly stated in the functional dependency notion of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative.
+This same limitation is explicitly stated in the functional dependency model of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative.
 The algebraic process model by Kindlmann and Scheideggar proposes that the data and visualization transformations are commutative; while similar to our framework, it is missing any explicit mention of continuity.
+
+On the other hand, most information visualization tools are explicitly tuned to specific continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}.
+For example, the relational database is core to libraries built on top of Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and  the digital humanaties oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro.
+Neither the table nor image model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} explicitly carry around different data representations for all the different types of visualizations they support.
+Where libraries with a single core data structure have very consistent APIs, VTK and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.
+
+Fiber bundles are a generalized abstraction proposed by Butler to encode the continuity of the data separately from the variable information \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplicial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} to incorporate a schema like description of the variables.
+One way of describing the binding between the schame and the continuinity is using the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}.
+Unlike Munzner's model where the semantic meaning of the key is tightly coupled to the index of the value, our model considers keys to be a reference to topology.
+This allows the metadata to be altered, for example by changing the coordinate system or time resolution, without imposing new semantics on the underlying structure.
 
@@
 Typically \gtotal\ is trivial and therefore sections can be thought of as mappings into \gfiber.
-In this work, we assume a 2D opaque image $\gfiber=\reals^5$ with elements
-\begin{equation}
-(x,\, y,\, r,\, g,\, b) \in \gfiber
-\end{equation}
-such that a rendered graphic only consists of 2D position and color. By abstracting the target display space as \gfiber, the model can support different targets, such as a 2D screen or 3D printer.
+In this work, we assume a 2D opaque image $\gfiber=\reals^5$ with elements $(x,\, y,\, r,\, g,\, b) \in \gfiber$ such that a rendered graphic only consists of 2D position and color.
+By abstracting the target display space as \gfiber, the model can support different targets, such as a 2D screen or 3D printer.
 
@@
 The data is an example of a Steven's nominal measurement set, which is defined as having on it permutation group actions
-\begin{equation}
+\begin{equation*}
   \text{ if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)
-\end{equation}
+\end{equation*}
 such that shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
@@
 To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$,
-\begin{equation}
-\text{ if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)
-\end{equation}
+\begin{equation*}
+\text{ if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2*)
+\end{equation*}
 the visual encodings must also have some sort of ordering. For interval scale data, \vchannel\ is equivariant under translation monoid actions if 
-\begin{equation}
+\begin{equation*}
 \vchannel (x + c) = \vchannel(x) + c
-\end{equation}
+\end{equation*}
 while for ratio data, there must be equivalent scaling\cite{weissteinSimilarityTransformation} 
-\begin{equation}
+\begin{equation*}
 \vchannel(xc) = \vchannel(x)*c 
-\end{equation}
+\end{equation*}
 We therefore can test if a \vchannel\ is equivariant by testing the actions under which it must  commute.
@@
 Testing this constraint with a translation action $t(x) = x+2$
-\begin{align}
+\begin{align*}
   \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + 2\\
   .5 &\neq .5 + 2
-\end{align}
+\end{align*}
 \vchannel\ does not commute and is therefore invalid.
@@
   \label{fig:math:artist:q}
-\end{figure}p
-
+\end{figure}
 As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\, reduce the encodings into a graphic \gsection.
@@
   action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
-  \label{fig:math:arist:graphic}
+  \label{fig:math:artist:graphic}
 \end{figure}
@@
 The transformed graphic $\gsection^{\prime}$ is equivariant to a transform on the visual bundle $\gsection^{\prime}=\vmark(m\circ \vsection)$ on a section that $\vsection \in \vmark^{-1}(\gsection)$ that must be part of generating \gsection.
-This equivariance is illustrated in figure \autoref{fig:math:artist:graphic}.
 Given fiber $\vfiber=(xpos,\, ypos,\, color,\, thickness)$, then sections $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
@@
 
-We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of path connected components $\dbasepath \subset \dbase$ as input:
+We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of path connected components $\dbasepath \subset \dbase$ as input 
 \begin{equation}
@@
 \end{align*}
-
 such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
@@
 \end{align*}
-
 where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n_1}, \hat{n_2}$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
@@
 \end{equation}
-copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase\.
+copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase.
 This set of points \gbasepoint\ are the preimage \vpreimg\ of \dbasepoint.
@@
 As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$  from \vsection\ over \dbasepoint\ to the section \vsectionpull\ over \gbasepoint.
-This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that\vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
+This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
 In \autoref{fig:math:artist:qhat} each dot on \vfiber is equivalent to the line on \vfiberpull.
@@
 Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out.
-When \vmark\ maps visual sections into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to the pulled back visual section \vsectionpull\ then the graphic section \gsection\ evaluated on a visual region \gbasepoint\
+When \vmark\ maps visual sections into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull\ then the graphic section \gsection\ evaluated on a visual region \gbasepoint\
 \begin{equation}
@@
 \end{equation}
- is defined as the assembly function \vmark\ with input \vsectionpull\ evaluated on \gbasepoint.Since the pulled back section \vsectionpull\ is the section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function
+ is defined as the assembly function \vmark\ with input \vsectionpull\ evaluated on \gbasepoint.
+Since the pulled back section \vsectionpull\ is the section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function
 \begin{equation}
@@
 The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
+
+\subsubsection{Composite and Reusable Artists}
+Given the family of artists $(\dtotal_i: i\in I)$ on the same image, the + operator 
+\begin{equation}
++ \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
+\end{equation}
+defines a simple composition of artists.
+When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the artists are acting on different components of the same section.
+This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.
+It is impractical to implement an artist for every single graphic; instead we implement an approximation of the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$.
+Roughly, two artists are equivalent if they have the same visual fiber \vfiber\, assembly function \vmark\, and continuity map \vindex.
 

### 2021-03-15 paper/paper.tex
rewrote channels to remove words

@@
 \usepackage{mathrsfs} % math script fonts
+\usepackage{tabulary}
 \usepackage{tikz}
@@
 
-One example of an equivariant \vchannel\ is illustrated in \autoref{fig:math:artist:nu}, which is a mapping from \textbf{Strings} to symbols.
-The data is an example of a Steven's nominal measurement set, which is defined as having on it permutation group actions
-\begin{equation*}
-  \text{ if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)
-\end{equation*}
-such that shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
-This is illustrated in the identical actions, represented by the colored arrows, on the words and emojis.
-To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$,
-\begin{equation*}
-\text{ if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2*)
-\end{equation*}
-the visual encodings must also have some sort of ordering. For interval scale data, \vchannel\ is equivariant under translation monoid actions if 
-\begin{equation*}
-\vchannel (x + c) = \vchannel(x) + c
-\end{equation*}
-while for ratio data, there must be equivalent scaling\cite{weissteinSimilarityTransformation} 
-\begin{equation*}
-\vchannel(xc) = \vchannel(x)*c 
-\end{equation*}
+The equivariant \vchannel\ illustrated in \autoref{fig:math:artist:nu} is a mapping from \textbf{Strings} to symbols, which is an example of a Steven's nominal measurement.
+More generally, for the Steven's scales an equivariant \vchannel\ has the following properties
+\begin{table}[H]
+  \begin{tabulary}{\columnwidth}{llL}
+      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
+      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
+      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
+      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\
+  \end{tabulary}
+\end{table}
+\noindent where the middle column is the group structure.
 We therefore can test if a \vchannel\ is equivariant by testing the actions under which it must  commute.

### 2021-03-15 paper/paper.tex
bulk update of everything + draft 0

@@
 \usepackage{cite}                      % needed to automatically sort the references
-\usepackage{tabu}                      % only used for the table example
-\usepackage{booktabs}                  % only used for the table example
 %% We encourage the use of mathptmx for consistent usage of times font
@@
 \usepackage{tabulary}
+\usepackage{subfig}
+\usepackage{multicol}
 \usepackage{tikz}
@@
                    frame=lines,
-                   bgcolor=bg,
                    framesep=3mm} 
@@
  \subsubsection{Sheafs: Gluing Sections }
+ \label{sec:math:data:sheaf}
  The restriction maps of a sheaf describe how local sections $\iota^{*}\dsection$ over a neighborhood $U$ around \dbasepoint\ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
@@
 
-The equivariant \vchannel\ illustrated in \autoref{fig:math:artist:nu} is a mapping from \textbf{Strings} to symbols, which is an example of a Steven's nominal measurement.
-More generally, for the Steven's scales an equivariant \vchannel\ has the following properties
+The \vchannel\ illustrated in \autoref{fig:math:artist:nu} is an equivariant mapping from \textbf{Strings} to emojis, as seen in the identical actions, represented as arrows, on the data and emojis.
+More generally, for the Steven's scales\cite{stevensTheoryScalesMeasurement1946} an equivariant \vchannel\ has the following properties
 \begin{table}[H]
@@
 \noindent where the middle column is the group structure.
-We therefore can test if a \vchannel\ is equivariant by testing the actions under which it must  commute.
-For example, given a a transform $\vchannel_{i}(x) = .5$ on interval data, it must commute under translation.
+Therefore, a \vchannel\ is equivariant if it satisfies the listed condition.
+For example, given a transform $\vchannel_{i}(x) = .5$ on interval data, it must commute under translation.
 Testing this constraint with a translation action $t(x) = x+2$
@@
 \section{Prototype}
+To build a prototype, we make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to transform data coordinates into screen coordinates.
+\begin{minted}{python}
+  fig, ax = plt.subplots()
+  artist = Artist(data, transforms)
+  ax.add_artist(artist)
+\end{minted}
+Building on the current Matplotlib artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalance class artist \vartisteq.
+The visual bundle \vtotal\ is specified as the \mintinline{python}{transform}dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber,  and the \vchannel\ encoders are passed in as functions or callable objects.
+The data bundle \dtotal\ is passed in as a \mintinline{python}{data} object.
+By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist.
+\begin{minted}{python}
+  class ArtistClass(matplotlib.artist.Artist):
+      def __init__(self, data, transforms, *args, **kwargs):
+          # properties that are specific to the graphic
+          self.data = data 
+          self.transforms = transforms
+          super().__init__(*args, **kwargs)
 
-\subsubsection{Ejector Seat Reservation}
+      def assemble(self, **args):
+          # set the properties of the graphic
 
-Duis autem~\cite{Lorensen:1987:MCA}\footnote{The algorithm behind
-Marching Cubes \cite{Lorensen:1987:MCA} had already been
-described by Wyvill et al. \cite{Wyvill:1986:DSS} a year
-earlier.} vel eum iriure dolor in hendrerit
-in vulputate velit esse molestie consequat,\footnote{Footnotes
-appear at the bottom of the column.} vel illum dolore eu
-feugiat nulla facilisis at vero eros et accumsan et iusto odio
-dignissim qui blandit praesent luptatum zzril delenit augue duis
-dolore te feugait nulla facilisi. Lorem ipsum dolor sit amet,
-consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt
-ut laoreet dolore magna aliquam erat volutpat.
+      def draw(self, renderer):
+          # returns K, indexed on fiber then key 
+          view = self.data.view(self.axes) 
+          # visual channel encoding applied fiberwise 
+          visual = {p: t['encoder'](view[t['name']])
+                    for p, t in self.transforms.items()}
+          self.assemble(**visual)
+          # pass configurations off to the renderer
+          super().draw(renderer)
+  \end{minted}
+ The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
+The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset.
+To ensure the integrity of the section, \mintinline{python}{view} must be atomic, which means that the values cannot change after the method is called in draw until a new call in draw.
+We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.
 
-\paragraph{Confirmed Ejector Seat Reservation}
+ The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}.
+The conversion from data to visual space is simplified here to directly show that it is the encoding \vchannel\ applied to the component.
+The \mintinline{python}{assemble} function that is \vmarkd is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
+This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd.
+The last step in the artist function is handing itself off to the renderer.
+The extra \mintinline{python}{*arg, **kwargs} arguments in \mintinline{python}{__init__,draw} are artifacts of how these objects are currently implemented.
 
-Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
-suscipit lobortis nisl ut aliquip ex ea commodo
-consequat~\cite{Nielson:1991:TAD}. Duis autem vel eum iriure dolor in
-hendrerit in vulputate velit esse molestie consequat, vel illum dolore
-eu feugiat nulla facilisis at vero eros et accumsan et iusto odio
-dignissim qui blandit praesent luptatum zzril delenit augue duis
-dolore te feugait nulla facilisi.
+ \subsection{Scatter and Line Artists}
+ \begin{figure}[H]
+    \centering 
+    \subfloat[\label{fig:code:scatter}]{%
+      \includegraphics[width=0.49\columnwidth]{scatter_0.png}}
+    \subfloat[\label{fig:code:line}]{%
+      \includegraphics[width=0.49\columnwidth]{line_1.png}}
+  \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
+  \label{fig:code_scatter_line}
+\end{figure}
 
-\paragraph{Rejected Ejector Seat Reservation}
+To generate the figure in \autoref{fig:code:scatter}, the \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of primitive point and area marks.
+In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker.
+We only show the \mintinline{python}{assemble} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
+The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices.
+Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$.
+This means that all the data on one vertex maps to one glyph.
+\begin{minted}{python}
+class Point(mcollections.Collection):
+  def assemble(self, x, y, s, facecolors='C0' ):
+    # construct geometries of circle glyphs
+    self._paths = [mpath.Path.circle((xi,yi), radius=si) 
+                for (xi, yi, si) in zip(x, y, s)] 
+    # set attributes of glyphs, these are vectorized 
+    # circles and facecolors are lists of the same size
+    self.set_facecolors(facecolors)
+\end{minted} 
+In \mintinline{python}{assemble}, the \vsection\ components are used to construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
+This is done via the \mintinline{python}{Path.circle} object.
+\begin{minted}{python}
+class Line(mcollections.LineCollection):
+    def assemble(self, x, y, color='C0'):
+            #assemble line marks as set of segments 
+            segments = [np.vstack((vx, vy)).T for vx, vy 
+                        in zip(x, y)]
+            self.set_segments(segments)
+            self.set_color(color)
+\end{minted}
+To generate \autoref{fig:code:line}, the \mintinline{python}{Line} artist \mintinline{python}{view} method returns a table of edges.
+Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge.
+As with \mintinline{python}{Point}, the data is then converted into visual variables.
+In \mintinline{python}{assemble}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
+Then the colors of each line segment are set.
+The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on view.
 
-Ut wisi enim ad minim veniam, quis nostrud exerci tation ullamcorper
-suscipit lobortis nisl ut aliquip ex ea commodo consequat. Duis autem
-vel eum iriure dolor in hendrerit in vulputate velit esse molestie
+\subsubsection{Visual Encoders}
+The visual parameter serves as the dictionary key because the visual representation is constructed from the encoding applied to the data  $\vsection = \vchannel \circ \dsection$.
+For the scatter plot, the mappings for the visual fiber components $\vfiber=(x,y, facecolors, s)$ are defined as
+\begin{minted}{python}
+cmap =  color.Categorical({'true':'deeppink', 
+                           'false':'deepskyblue'})
+transforms = {'x': {'name': 'v4', 'encoder': lambda x: x},
+              'y': {'name': 'v2', 'encoder': lambda x: x},
+              'facecolors': {'name':'v3', 'encoder': cmap}, 
+              's':{'name': None , 
+                   'encoder': lambda _: itertools.repeat(.02)}}
+\end{minted}
 
-\subsection{Vestibulum}
+where \mintinline{python}{lambda x: x} is an identity \vchannel\, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.
+A test for equivariance can be implemented trivially
+\begin{minted}{python}
+def test_nominal(values, encoder):
+    m1 = list(zip(values, encoder(values)))
+    random.shuffle(values)
+    m2 = list(zip(values, encoder(values)))
+    assert sorted(m1) == sorted(m2)
+\end{minted}
+but is currently factored out of the artist for clarity.
 
-Vestibulum ut est libero.
-Suspendisse non libero id massa congue egestas nec at ligula.
-Donec nibh lorem, ornare et odio eu, cursus accumsan felis.
-Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
-Donec dapibus quam vel eros mattis, id ornare dolor convallis.
-Donec at nisl sapien.
-Integer fringilla laoreet tempor.
-Fusce accumsan ante vel augue euismod, sit amet maximus turpis mattis.
-Nam accumsan vestibulum rhoncus.
-Aenean quis pellentesque augue.
-Suspendisse sed augue et velit consequat bibendum id nec est.
-Quisque erat purus, ullamcorper ut ex vel, dapibus dignissim erat.
+\subsubsection{Data Model}
+The data input into the \mintinline{python}{Artist} will often be a wrapper class around an existing data structure.
+This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{python}{view} method that returns an atomic object that encapsulates \dsection.
+To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataclassesDataClasses}
 
-Quisque sit amet orci quam.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Aliquam pharetra, nunc non efficitur convallis, tellus purus iaculis lorem, nec ultricies dolor ligula in metus.
-Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos.
-Aenean eu ex nulla.
-Morbi id ex interdum, scelerisque lorem nec, iaculis dui.
-Phasellus ultricies nunc vitae quam fringilla laoreet.
-Quisque sed dolor at sem vestibulum fringilla nec ac augue.
-Ut consequat, velit ac mattis ornare, eros arcu pellentesque erat, non ultricies libero metus nec mi.
-Sed eget elit sed quam malesuada viverra.
-Quisque ullamcorper, felis ut convallis fermentum, purus ligula varius ligula, sit amet tempor neque dui non neque.
-Donec vulputate ultricies tortor in mollis.
+\begin{minted}{python}
+@dataclass
+class FiberBundle:
+    K: dict #{'tables': []}
+    F: dict #  {variable name: type}
+\end{minted}
+that asks the user to specify the the properties of \dfiber\ and the \dbase\ connectivity in terms of a  simplicial triangulation scheme \cite{geometricaSimplicialComplexes}.
+To generate the scatter plot and the line plot, the distinction is in the \mintinline{python}{tau} method that is the section.
+\begin{minted}{python}
+class VertexSimplex: 
+    """Fiberbundle is consistent across all sections
+    """
+    FB = FiberBundle({'tables': ['vertex']},  
+            {'v1': float, 'v2': str, 'v3': float})
 
-Integer sit amet dolor sit amet turpis ullamcorper varius.
-Cras volutpat bibendum scelerisque.
-Maecenas mauris dolor, gravida eu elit et, sodales consequat tortor.
-Integer id commodo elit.
-Pellentesque sollicitudin ex non nulla molestie eleifend.
-Mauris sagittis metus nec turpis imperdiet, vel ullamcorper nibh tincidunt.
-Sed semper tempus ex, ut aliquet erat hendrerit id.
-Maecenas sit amet dolor sollicitudin, luctus nunc sit amet, malesuada justo.
+    def __init__(self, sid = 45, size=1000, max_key=10**10):
+        # create random list of keys
+    def tau(self, k):
+      # e1 is sampled from F1, e2 from F2, etc...
+      return (k, (e1, e2, e3, e4))
+\end{minted}
+The discrete \mintinline{python}{tau} method returns a record of discrete points, while the line\mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint
+\begin{minted}{python}
+def tau(self, k): #will fix location on page on revision
+    x, y = self._xy(k, self.distances, 
+               self.angle_samples[k], self.angle_samples[k+1]) 
+    color = self._color(k) 
+    return (k, (x, y, color))
+\end{minted}
+In both cases the \mintinline{python}{view} method packages the data
+\begin{minted}{python}
+  def view(self, axes):
+      table = defaultdict(list)
+      for k in self.keys:
+          table['index'].append(k)
+          for (name, value) in zip(self.FB.fiber.keys(), 
+                                   self.tau(k)[1]):
+              table[name].append(value)
+      return table
+\end{minted}
+into a data structure that the artist can unpack via data component name.
 
-Mauris ut sapien non ipsum imperdiet sodales sit amet ac diam.
-Nulla vel convallis est.
-Etiam dapibus augue urna.
-Aenean enim leo, fermentum quis pulvinar at, ultrices quis enim.
-Sed placerat porta libero et feugiat.
-Phasellus ullamcorper, felis id porta sollicitudin, dolor dui venenatis augue, vel fringilla risus massa non risus.
-Maecenas ut nulla vitae ligula pharetra feugiat non eu ante.
-Donec quis neque quis lorem cursus pretium ac vulputate quam.
-Cras viverra tellus vitae sapien pretium laoreet.
-Pellentesque fringilla odio venenatis ex viverra, quis eleifend tortor ornare.
-Ut ut enim nunc.
-Vivamus id ligula nec est dignissim eleifend.
+\begin{figure}[htb]
+  \centering 
+  \subfloat[\label{fig:code:line:arc}]{%
+    \includegraphics[width=0.49\columnwidth]{linec_1.png}}
+  \subfloat[\label{fig:code:line:step}]{%
+    \includegraphics[width=0.49\columnwidth]{lined_1.png}}
+\caption{Continuous and discontinuous lines as defined via the same data model, and generated with the same \vartisteq \mintinline{python}{Line}}
+\label{fig:code:multilines}
+\end{figure}
+The graphics in figure~\ref{fig:code:multilines} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{Graphline} data source where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix.
+The multicolored line is a connected graph of edges with each edge function evaluated on 1000 samples,
+\begin{minted}{python}
+simplex.GraphLine(FB, edges, verticies, 
+                      num_samples=1000, connect=True)
+\end{minted}
+while the stair chart is discontinuous and only needs to be evaluated at the edges of the interval 
+\begin{minted}{python}
+simplex.GraphLine(FB, edges, verticies, 
+                      num_samples=2, connect=False)
+\end{minted}
+such that one advantage of this model is it helps differentiate graphics that have different artists from graphics that have the same artist but make different assumptions about the source data.
 
-Nunc ac velit tellus.
-Donec et venenatis mauris.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Ut vitae lectus vel ante mollis congue.
-Vestibulum at cursus velit.
-Curabitur in facilisis enim.
-Vestibulum eget dui aliquet risus laoreet laoreet.
-Phasellus et est id magna interdum venenatis.
-Donec luctus vehicula justo sed laoreet.
-Quisque tincidunt suscipit augue, in molestie sem accumsan sed.
+\section{Discussion}
+This work contributes a mathematical description of the mapping \vartist\ from data to visual representation.
+Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables,, multivariate high resolution spatio temporal datasets, and complex networks.
+Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications that the graphic must have continuity equivalent to the data, and that the visual characteristics of the graphics are equivariant to their corresponding components under monoid actions.
+This model defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic.
+Encoding the graphic space as a fiber bundle provides a structure rich abstraction of the target graphical design in the target display space.
+The toy prototype built using this model validates that is usable for a general purpose visualization tool since it can be iteratively integrated into the existing architecture rather than starting from scratch.
+Factoring out graphic formation into assembly functions allows for much more clarity in how they differ.
+This prototype demonstrates that this framework can generate the fundemental point (scatter plot) and line (line chart) marks.
+
+\subsection{Limitations}
+So far this model has only been worked out for a single data set tied to a primitive mark.
+The examples and prototype have so far only been implemented for the static 2D case, but nothing in the math limits to 2D and expansion to the animated case should be possible because the model is formalized in terms of the sheaf.
+While this model supports equivariance of figurative glyphs generated from parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it does not have a way to evaluate the semantic accuracy of the figurative representation.
+Even though the model is designed to be backend and format independent, it has only really been tested against PNGs rendered with the AGG backend.
+It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}.
+This model and the associated prototype is deeply tied to Matplotlib's existing architecture, so it has not been worked through how the model generalizes to other libraries, such as those built on Mackinlay's APT framework.
+
+\subsection{Future Work}
+More work is needed to formalize the composition operators and equivalence class \vartisteq.
+More artists need to be implemented to demonstrate that the model can underpin a minimally viable library,  foremost an image\cite{toryRethinkingVisualizationHighlevel2004, haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an imherently computational artist such as the boxplot\cite{wickham40YearsBoxplots2011}.
+This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
+Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
+While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
 \section{Conclusion}
-
-Lorem ipsum dolor sit amet, consetetur sadipscing elitr, sed diam
-nonumy eirmod tempor invidunt ut labore et dolore magna aliquyam erat,
-sed diam voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum. Stet clita kasd gubergren, no sea takimata sanctus est Lorem
-ipsum dolor sit amet. Lorem ipsum dolor sit amet, consetetur
-sadipscing elitr, sed diam nonumy eirmod tempor invidunt ut labore et
-dolore magna aliquyam erat, sed diam voluptua. At vero eos et accusam
-et justo duo dolores et ea rebum. Stet clita kasd gubergren, no sea
-takimata sanctus est Lorem ipsum dolor sit amet. Lorem ipsum dolor sit
-amet, consetetur sadipscing elitr, sed diam nonumy eirmod tempor
-invidunt ut labore et dolore magna aliquyam erat, sed diam
-voluptua. At vero eos et accusam et justo duo dolores et ea
-rebum.
+An unoffical philosophy of Matplotlib is to support making whatever kinds of plots a user may want, even if they seem nonsensical to the development team.
+The topological framework described in this work provides a way to facilitate this graph creation in a rigorous manner; any artist that meets the equivariance criteria described in this work by definition generates a graphic representation that matches the structure of the data being represented.
+We leave it to domain specialists to define the structure they need to preserve and the maps they want to make, and hopefully make the process easier by untangling these components into seperate constrained maps and providing a fairly general data and display model.
 
@@
 \acknowledgments{
-The authors wish to thank A, B, and C. This work was supported in part by
-a grant from Chan Zuckerberg Institute  (\# 12345-67890).}
+The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran for discussions about the visual fiber space.
+This work was supported in part by a grant from the Chan Zuckerberg Institute (EOSS cycle 1).}
 

### 2021-03-15 slides/main.tex
bulk update of everything + draft 0

@@
 \end{frame}
+\begin{frame}{Measurement Scale Groups}
+
+\begin{table}[H]
+    \begin{tabulary}{\textwidth}{llL}
+        nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
+        ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
+        interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
+        ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $
+    \end{tabulary}
+\end{table}
+
+    \pause
+    \begin{block}{Invalid \vchannel}
+     Given $\vchannel_{i}(x) = .5$ and $t(x) = x+2$,
+    \pause    
+    \begin{align*}
+            \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + 2\\
+            .5 &\neq .5 + 2
+    \end{align*}
+    \end{block}
+\end{frame}
 
@@
         \item act on the same visual bundle \vtotal
-        \item have the same assembly function \vmarkd 
+        \item have the same assembly function \vmark 
         \item have the same continuity map \vindex 
@@
 \begin{columns}
-\column{,49\textwidth}
+\column{0.49\textwidth}
 \begin{minted}{python}

### 2021-03-15 draft/sections/math.tex
fixed nlab cite + ackwnolegemtn

@@
 A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$.
-As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,ActionNLab} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
+As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
 \begin{align*}

### 2021-03-15 paper/paper.tex
fixed nlab cite + ackwnolegemtn

@@
 The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran for discussions about the visual fiber space.
-This work was supported in part by a grant from the Chan Zuckerberg Institute (EOSS cycle 1).}
+This project has been made possible in part by grant number 2019-207333 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.}
 

### 2021-03-16 paper/paper.tex
line numbetrs

@@
 \usetikzlibrary{cd} % commutative diagrams
+\usepackage[switch]{lineno}
 \usepackage{minted}
@@
 \usepackage{notation} % move this later
+
 %% In preprint mode you may define your own headline. If not, the default IEEE copyright message will appear in preprint mode.
@@
 %% indicate IEEE Member or Student Member in form indicated below
-\author{A, B, C}
+\author{Hannah Aizenman, Thomas Caswell, Michael Grossberg}
 \authorfooter{
@@
 \item
- A B are with the Computer Science department, City College of New York. E-mail: roy.g.biv@aol.com.
+ Hannah Aizenman and Michael Grossberg are with the Computer Science department, City College of New York. E-mail: roy.g.biv@aol.com.
 \item
@@
 \begin{document}
-
+\linenumbers %linenumbers for comments [make sure to remove later]
 %% The ``\maketitle'' command must be the first command after the

### 2021-03-16 paper/paper.tex
abstract

@@
 \usepackage{tabulary}
-\usepackage{subfig}
+\usepackage{subfig} %ieee does not like subfigure
 \usepackage{multicol}
@@
 \item
- Hannah Aizenman and Michael Grossberg are with the Computer Science department, City College of New York. E-mail: roy.g.biv@aol.com.
+ Hannah Aizenman and Michael Grossberg are with the Computer Science department, City College of New York. E-mail: haizenman@ccny.cuny.edu, mgrossberg@ccny.cuny.edu.
 \item
- Thomas Caswell is with Brookhaven National Lab E-mail: ed.grimley@aol.com.
+ Thomas Caswell is with National Synchrotron Light Source II, Brookhaven National Lab E-mail: tcaswell@bnl.gov.
 }
@@
 %% Abstract section.
-\abstract{ A critical aspect of data visualization is that the graphical representation of data is expected to match the properties of the data; this fails when order is not preserved in representations of ordinal data or scale for numerical data.
-In this work, we propose that the mathematical notions of equivariance and topology formalizes the expectation of matching properties.
-We developed a model we call the topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles.
-This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
-Furthermore, our model supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
-We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap.
-We show that this model can be implemented with a small prototype.} % end of abstract
+\abstract{ A critical aspect of data visualization is that the graphical representation of data is expected to match the structure of the data; this fails, for example, when order is not preserved in representations of ordinal data or proportional distance for numerical data or when discrete data is plotted on a line plot or interpolated as an image.
+In this work, we propose the topological equivariant artist model (TEAM), wherein we formalize the expectations of attributes and continuity matching using the mathematical notion of equivariance and by modeling data and graphics as fiber bundles.
+Specifically, our model introduces measurement scale preserving maps on the fiber that transform data components (variables) into visual channels and a continuity preserving map on the base space that transforms dataset connectivity into graphical elements.
+Furthermore, TEAM has a sum operator to allow for building composite visualizations.
+We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap and show that this model can be implemented with a small prototype.}
+
+% end of abstract
 
@@
 The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization to improve consistency, reusability, and discoverability, so domain specific tool developers can build structure preserving visualization tools.
-The contribution of this work is (a) topology preserving relationship between data and graphic via continuous maps, (b) property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions, (c) functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model, (d) prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
+The contribution of this work is
+\begin{enumerate}
+  \item topology preserving relationship between data and graphic via continuous maps
+  \item property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
+  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
+  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
+\end{enumerate}
 
@@
 
-Specifically, \vartist is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$ 
+Specifically, \vartist\ is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$ 
 \begin{equation}
@@
 \end{figure}
-As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\, reduce the encodings into a graphic \gsection.
+As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\ reduce the encodings into a graphic \gsection.
 While it may seem intuitive that visualizations that generate the same graphic should do so consistently given the same input, we formalize this constraint such that it can be specified as a constraint of \vmark.
@@
 such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
-In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase\.
+In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
 As shown in \ref{fig:math:arist:base} line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
@@
 This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
-In \autoref{fig:math:artist:qhat} each dot on \vfiber is equivalent to the line on \vfiberpull.
+In \autoref{fig:math:artist:qhat} each dot on \vfiber\ is equivalent to the line on \vfiberpull.
 
@@
 
-where \mintinline{python}{lambda x: x} is an identity \vchannel\, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.
+where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.
 A test for equivariance can be implemented trivially

### 2021-03-16 paper/paper.tex
D3 is explicit data structure

@@
 Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations.
-Mackinlay's A Presentation Tool introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}.
+Mackinlay's \textit{A Presentation Tool} (APT) introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}.
 This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
@@
 On the other hand, most information visualization tools are explicitly tuned to specific continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}.
-For example, the relational database is core to libraries built on top of Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+For example, the relational database is core to libraries built on top of Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}, and Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}.
+
 Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and  the digital humanaties oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro.
-Neither the table nor image model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} explicitly carry around different data representations for all the different types of visualizations they support.
+
+Neither the table nor image model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.
 Where libraries with a single core data structure have very consistent APIs, VTK and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.

### 2021-03-16 slides/main.tex
fixed nlab cite + ackwnolegemtn
D3 is explicit data structure
rework tools slides to add graphs

@@
 
-\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
+%\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
+\usepackage{amsmath, amsfonts, amssymb, mathtools, eulervm, xspace}
 \usepackage{bm}
@@
 \subsection{Background}
-\begin{frame}{Tools are tuned to the continuity of the date \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}}
+\begin{frame}{Tools are tuned to date continuity \cite{HeerSoftware2006}}
     \begin{columns}
@@
             \end{figure}
-
             \begin{enumerate}
+                \item Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}
                 \item ggplot\cite{wickhamGgplot2ElegantGraphics2016a}
-            \item protovis\cite{bostockProtoviz2009}, D3 \cite{bostockDataDrivenDocuments2011}
-            \item vega\cite{satyanarayanDeclarativeInteractionDesign2014}, altair\cite{vanderplasAltairInteractiveStatistical2018}
+                \item Vega\cite{satyanarayanDeclarativeInteractionDesign2014}, Altair\cite{vanderplasAltairInteractiveStatistical2018}
             \end{enumerate}
@@
             \begin{enumerate}
-            \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}\item Napari\cite{nicholas_sofroniew_2021_4533308}
+                \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}
+                \item Napari\cite{nicholas_sofroniew_2021_4533308}
             \end{enumerate}
@@
             \begin{enumerate}
+                \item Gephi\cite{bastianGephiOpenSource2009}
+                \item Graphviz\cite{ellsonGraphvizOpenSource2002}
+                \item Networkx\cite{HagbergExploringNetwork2008}
+            \end{enumerate}
+    \end{columns}
+\end{frame}
+
+\begin{frame}{Visualizations are tuned to data continuinity\cite{toryRethinkingVisualizationHighlevel2004}}
+    \begin{figure}
+        \includegraphics[height=.5\textheight]{figures/intro/dataset_diagram.png}
+        \caption{Data Representation, MayaVi 4.7.2 docs\cite{DataRepresentationMayavi}}
+    \end{figure}
+    \begin{enumerate}
         \item Matplotlib\cite{hunterMatplotlib2DGraphics2007}, 
+        \item D3 \cite{bostockDataDrivenDocuments2011}
         \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
     \end{enumerate}
-    \end{columns}
 \end{frame}
@@
 \begin{block}{left monoid action}
-A set \dfiber\ with an action $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
+A set \dfiber\ with an action\cite{nlab:action} $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
     \begin{align*}
@@
 \end{figure}
-\begin{equation}
-\label{eq:qhat_q_s}
+\begin{equation}\label{eq:qhat_q_s}
     \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
@@
 \end{frame}
+
 \begin{frame}{Equivalance class of artists}

### 2021-03-16 paper/paper.tex
intro revisions

@@
 On the other hand, most information visualization tools are explicitly tuned to specific continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}.
-For example, the relational database is core to libraries built on top of Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}, and Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}.
-
-Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and  the digital humanaties oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro.
-
-Neither the table nor image model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.
-Where libraries with a single core data structure have very consistent APIs, VTK and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.
+For example, the relational database is core to tools influenced by APT, such as
+Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}.
+Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}.
+Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.
+Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.
 

### 2021-03-17 paper/paper.tex
math intro sections

@@
 In this work, we propose the topological equivariant artist model (TEAM), wherein we formalize the expectations of attributes and continuity matching using the mathematical notion of equivariance and by modeling data and graphics as fiber bundles.
-Specifically, our model introduces measurement scale preserving maps on the fiber that transform data components (variables) into visual channels and a continuity preserving map on the base space that transforms dataset connectivity into graphical elements.
+Specifically, our model introduces measurement scale preserving maps on the fiber that transform components of the data into visual channels and a continuity preserving map on the base space that transforms dataset connectivity into graphical elements.
 Furthermore, TEAM has a sum operator to allow for building composite visualizations.
@@
 
-Fiber bundles are a generalized abstraction proposed by Butler to encode the continuity of the data separately from the variable information \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Fiber bundle and sheafs are generalized abstractions proposed by Butler to encode the continuity of the data separately from the properties of each variable  \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
 Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplicial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} to incorporate a schema like description of the variables.
-One way of describing the binding between the schame and the continuinity is using the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}.
+One way of describing the binding between the scheme and the continuinity is using the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}.
 Unlike Munzner's model where the semantic meaning of the key is tightly coupled to the index of the value, our model considers keys to be a reference to topology.
@@
 \label{sec:math}
-We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map from data to graphic
+We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map
 \begin{equation}
@@
 \end{equation}
-that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}, which are discussed in detail in \autoref{sec:math:data:monoid}.
-Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps
-\begin{equation}
-\mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
-\end{equation}
-such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.
-We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
+from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$.
+To formalize the structure the artist is preserving, we describe how we model data (\ref{sec:math:data}) and graphics (\ref{sec:math:graphic}) as fiber bundles.
+We then discuss the equivariant maps that make up $\mathscr{\vartist}$, which are a map from  graphic continuity to data continuity (\ref{sec:math:graphic:base}), data components to visual components (\ref{sec:math:artist:nu}) and visual components to graphic (\ref{sec:math:artist:q}).
+
 \subsection{Data Bundle}
@@
 that binds the components of the data in \dfiber\ to the continuity represented in \dbase.
-By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on \dtotal\ and focus on the components and continuity.
+By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartestian product $\dbase\times\dfiber$.
 
@@
 \end{figure}
-
 To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names.
@@
 \subsubsection{Visualization Assembly}
+\label{sec:math:artist:q}
 \begin{figure}[htb]
@@
 \subsubsection{Assembly Factory}
+\label{sec:math:artis:qhat}
 The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.

### 2021-03-17 draft/sections/math.tex
cleaned up fibers

@@
 
-We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity; by doing so we can develop implementations that keep track of both in ways that let us distribute computation while still allowing assembly and dynamic update of the graphic.
+We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity.
 To explain which structure the artist is preserving, we first describe how we model data (\ref{sec:data}), graphics (\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles.

### 2021-03-17 paper/paper.tex
cleaned up fibers

@@
 from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$.
-To formalize the structure the artist is preserving, we describe how we model data (\ref{sec:math:data}) and graphics (\ref{sec:math:graphic}) as fiber bundles.
-We then discuss the equivariant maps that make up $\mathscr{\vartist}$, which are a map from  graphic continuity to data continuity (\ref{sec:math:graphic:base}), data components to visual components (\ref{sec:math:artist:nu}) and visual components to graphic (\ref{sec:math:artist:q}).
+To formalize the structure the artist is preserving, we describe how we model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as fiber bundles.
+We then discuss the equivariant maps that make up $\mathscr{\vartist}$, which are a map from  graphic continuity to data continuity (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}) and visual components to graphic (\autoref{sec:math:artist:q}).
 
@@
 that binds the components of the data in \dfiber\ to the continuity represented in \dbase.
-By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartestian product $\dbase\times\dfiber$.
+By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartesian product $\dbase\times\dfiber$.
 
@@
 \label{sec:math:data:fiber}
-\begin{figure}[htb]
-  \centering % avoid the use of \begin{center}...\end{center} and use \centering instead (more compact)
-  \includegraphics[width=\columnwidth]{fiber.png}
-  \caption{These two datasets have the same base space \dbase\ of discrete points, but the square is a visual representation of the fiber space  $\dfiber=\reals\times\reals$, while the cube is a representation of $\dfiber=\reals\times\realsp\times\reals$}
-  \label{fig:math:data:fiber}
-\end{figure}
 To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names.
@@
 which allows us to decouple \dfiber\ into components $\dfiber_i$.
-Each component of \dfiber\ is a dimension of the topological fiber space, as illustrated in \autoref{fig:math:data:fiber}.
-The space $\dfiber=\reals\times\reals$ encodes that the dataset has two quantative continuous variables, while $\dfiber=\reals\times\reals\realsp$ encodes a dataset with three quantative contunous variables, one of which must be positive.
+Each component of \dfiber\ is a dimension of the topological fiber space and is specified by a tuple of the form (\fname, \ftype, \(\ftotal_{\fsection(\fname)}\)).
+\begin{figure}[htb]
+  \centering 
+  \includegraphics[width=\columnwidth]{fiber.png}
+  \caption{These two datasets have the same base space \dbase.
+The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature),  while the cube is the fiber $\reals\times\realsp\times\reals$ associated with (time, wind=(speed, direction))}
+  \label{fig:math:data:fiber}
+\end{figure}
+In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\realsp\times\reals)\)) which encodes (\emph{speed}, \emph{direction}).
 
@@
 As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
-\begin{equation}
+\begin{equation}\label{eq:math:data:monoid:decompose}
 \monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
@@
 \label{sec:math:data:base}
- The base space \dbase\ is way to express how the records in \dtotal\ are connected to each other, for example if they are discrete points or if they lie in a 2D continous surface.
-Formally \dbase\ is the quotient space of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
-Our model formulates \dbase\ as akin to an indexing space into \dtotal\ that describes the structure of \dtotal.
+ The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other and does not itself store any of the components of the data.
+For example, \dbase\ can encode that the timeseries data described in \autoref{sec:math:data:fiber} is continuous, but is doing so independent of the timestamps.
+Every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$ because \dbase\ is the quotient space\cite{QuotientSpaceTopology2020,aurouxMath131Introduction} of \dtotal.
 
@@
  As illustrated in ~\autoref{fig:math:data:base}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber.
-As with fibers and monoids, we can decompose the total space into components $\pi:\dtotal_i\rightarrow \dbase$ where
- \begin{equation}
-    \label{eq:math:data:base:decompose}
+As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose},  we can decompose the total space into component bundles
+ \begin{equation}\label{eq:math:data:base:decompose}
      \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
  \end{equation}
- which is a decomposition of \dfiber.
-The \dbase\ remains the same because the connectivity of records does not change just because there are fewer elements in each record.
+ such that \(\monoid_{i}\) acts on component bundle \(\dtotal_i\).
+The \dbase\ remains the same because the connectivity of records does not change just because there are fewer components in each record.
 By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.
@@
 \end{equation}
-where \gsection\ is the fully encoded graphic.
+where \gsection\ is a fully specified graphic such that it is an abstraction of rendering.
 To fully specify the visual characteristics of the image, we construct a fiber \gfiber\ that is an infinite resolution version of the target space.
@@
 For example, a \dbasepoint\ that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the points in \gfiber\ that visually represent the record at \dbasepoint.
+The \vindex\ mapping is critical to interactive visualizations as it is the map from a region on screen to the data associated with that region.
+One example is to fill in details in a hover tooltip, another is to convert region selection on \gbase\ to a query on the data to access the corresponding record components on \dbase.
 
 \subsection{Artist}
-The topological artist \vartist\ is a monoid equivariant sheaf map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
+The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$. 
 \begin{equation}
@@
 \end{equation}
-While \vartist\ can usually construct graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
-In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the jet bundle $\dtotal^{\prime}=\mathcal{J}^{2}(\dtotal)$.
+that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}.
+Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps
+\begin{equation}
+\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
+\end{equation}
+such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.
 
-Specifically, \vartist\ is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$ 
+The monoid equivariant map has two stages: the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components, and the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
 \begin{equation}
@@
     \begin{tikzcd}
-        \dtotal^{\prime} \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \gtotal \arrow[ld, "\pi"] \\
+        \dtotal \arrow[r, "\vchannel"] \arrow[rd, "\pi"'] & \vtotal \arrow[d, "\pi"] & \vindex^*\vtotal \arrow[r, "\vmark"] \arrow[d, "\vindex^*\pi"'] \arrow[l, "\vindex^*"'] & \gtotal \arrow[ld, "\pi"] \\
                                               & \dbase                  & \gbase \arrow[l, "\vindex"']                                              &                    
@@
 \end{equation}
-where the input can be point wise $\dsection(\dbasepoint)\mid \dbasepoint \in \dbase$.
+\vtotalpull\ is the visual bundle \vtotal\ pulled back over \gbase\ via the equivariant continuity map $\vindex:\gbase \rightarrow \dbase$ introduced in \autoref{sec:math:graphic:base}.
 The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the latent space of possible parameters of a visualization type, such as a scatter or line plot.
@@
 The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave.
-
-In \autoref{eq:math:artist:diagram}, the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components.
-The continuity map $\vindex:\gbase \rightarrow \dbase$ then pulls back the visual bundle \vtotal\ over \gbase.
-Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
-This functional decomposition of the visualization artist facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
+The functional decomposition of the visualization artist facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
 We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.
@@
 In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
-As shown in \ref{fig:math:arist:base} line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
+As shown in \autoref{fig:math:artist:base}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal.
 The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
@@
 \subsubsection{Assembly Factory}
-\label{sec:math:artis:qhat}
+\label{sec:math:artist:qhat}
 The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
-As shown in eq~\ref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
 \begin{figure}[htb]
@@
 
-Factoring out \gbasepoint\ from equation~\ref{eq:math:artist:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
+Factoring out \gbasepoint\ from \autoref{eq:math:artist:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
 In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
@@
 \end{figure}
-The graphics in figure~\ref{fig:code:multilines} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{Graphline} data source where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix.
+The graphics in figure \autoref{fig:code:multilines} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{Graphline} data source where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix.
 The multicolored line is a connected graph of edges with each edge function evaluated on 1000 samples,

### 2021-03-18 paper/paper.tex
equivariant monoid figure

@@
 
- \subsubsection{Sheafs: Gluing Sections }
+ \subsubsection{Sheafs}
  \label{sec:math:data:sheaf}
@@
 \label{sec:math:artist:nu}
-\begin{figure}[htb]
-  \centering
-  \includegraphics[width=\columnwidth]{equivariance_nu.png}
-  \caption{In this artist, \vchannel\ maps the strings to the emojis.
-This \vchannel\ is equivariant because the monoid actions, represented by the colored arrows, are the same on the input \dsection\ and output \vsection\ sets.}
-  \label{fig:math:artist:nu}
-\end{figure}
  We define the visual transformers \vchannel\ 
@@
 
-The \vchannel\ illustrated in \autoref{fig:math:artist:nu} is an equivariant mapping from \textbf{Strings} to emojis, as seen in the identical actions, represented as arrows, on the data and emojis.
-More generally, for the Steven's scales\cite{stevensTheoryScalesMeasurement1946} an equivariant \vchannel\ has the following properties
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
+  \caption{fill in this caption tomorrow}
+  \label{fig:math:artist:nu}
+\end{figure}
+
+\autoref{fig:math:artist:nu} illustrates a \vchannel\ that preserves partial order relations because 
+ou would have to state that the monoid of monitonic actions is equivalent to the partial order realationship
+ok
+so that the order is preserved IF and only IF it is an equivariant map.
+I think that is true but it would be nice to have a reference.
+but then you don't have to really plot out m.
+
+Generally, for the Steven's scales\cite{stevensTheoryScalesMeasurement1946} an equivariant \vchannel\ has the following properties
 \begin{table}[H]
@@
 \end{table}
-\noindent where the middle column is the group structure.
-Therefore, a \vchannel\ is equivariant if it satisfies the listed condition.
-For example, given a transform $\vchannel_{i}(x) = .5$ on interval data, it must commute under translation.
-Testing this constraint with a translation action $t(x) = x+2$
-\begin{align*}
-  \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + 2\\
-  .5 &\neq .5 + 2
-\end{align*}
-\vchannel\ does not commute and is therefore invalid.
-The constraints on \vchannel\ can be embedded into the artist such that the \vchannel\ functions can test for equivariance and also provide guidance on constructing new \vchannel\ functions.
+\noindent where the middle column is the group structure. Therefore, a \vchannel\ is equivariant if it satisfies the listed condition.
 
@@
   \centering
-  \includegraphics[width=\columnwidth]{base_q}
+  \includegraphics[width=.5\columnwidth]{base_q}
   \caption{The coordinates $(\gx, \gy)$ dictate the color of the region in prerender space \gbase\, which over the whole disk generates the graphical mark.

### 2021-03-18 slides/main.tex
equivariant monoid figure

@@
     \begin{figure}[H]
-        \begin{subfigure}{.49\textwidth}
-            \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-            \caption{}
-            \label{fig:fiber_example_plane}
-        \end{subfigure}
-        \begin{subfigure}{.49\textwidth}
-            \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
-            \caption{}
-            \label{fig:fiber_example_cube}
-        \end{subfigure}
+        \includegraphics[width=\textwidth]{figures/math/fiber.png}
         \caption{}
@@
     \begin{description}
-        \item[\ref{fig:fiber_example_plane}]  $\dfiber=\reals\times\reals$, (time, temperature)
-        \item[\ref{fig:fiber_example_cube}]  $\reals\times\realsp\times\reals$, (time, wind=(speed, direction))
+        \item[plane]  $\dfiber=\reals\times\reals$, (time, temperature)
+        \item[cube]  $\reals\times\realsp\times\reals$, (time, wind=(speed, direction))
     \end{description}
@@
     \begin{figure}[H]
-        \begin{subfigure}{.49\textwidth}
-            \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-            \caption{}
-            \label{fig:base_example_discrete}
-        \end{subfigure}
-        \begin{subfigure}{.49\textwidth}
-            \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
-            \caption{}
-            \label{fig:base_example_continuous}
-        \end{subfigure}
+        \includegraphics[width=1\textwidth]{figures/math/base.png}
         \caption{}
@@
     \begin{description}
-        \item[\ref{fig:base_example_discrete}] data is discrete 0D points
-        \item[\ref{fig:base_example_continuous}]  data is lies on the 1D continuous interval \dbase\
+        \item[points] data is 0D discrete
+        \item[line]  data is lies on the 1D continuous interval \dbase\
     \end{description}
@@
     \begin{figure}
-        \includegraphics[width=1\textwidth]{figures/math/equivariance_nu.png}
-        \caption{The actions on the text data are the same as the actions on the visual representation of that data as emojis.}
+        \includegraphics[width=1\textwidth]{figures/math/monoid_equivariant.png}
     \end{figure}

### 2021-03-22 paper/paper.tex
abstract

@@
 %% Abstract section.
-\abstract{ A critical aspect of data visualization is that the graphical representation of data is expected to match the structure of the data; this fails, for example, when order is not preserved in representations of ordinal data or proportional distance for numerical data or when discrete data is plotted on a line plot or interpolated as an image.
-In this work, we propose the topological equivariant artist model (TEAM), wherein we formalize the expectations of attributes and continuity matching using the mathematical notion of equivariance and by modeling data and graphics as fiber bundles.
-Specifically, our model introduces measurement scale preserving maps on the fiber that transform components of the data into visual channels and a continuity preserving map on the base space that transforms dataset connectivity into graphical elements.
-Furthermore, TEAM has a sum operator to allow for building composite visualizations.
-We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap and show that this model can be implemented with a small prototype.}
+\abstract{This work presents a functional model of the structure-preserving maps from data to visual representation to guide the development of visualization libraries.
+Our model, which we call the  topological equivariant artist model (TEAM), provides a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables.
+We formalize these transformations as actions on sections of topological fiber bundles, which are mathematical structures that allow us to encode continuity as a base space, variable properties as a fiber space, and data as binding maps, called sections, between the base and fiber spaces.
+This abstraction allows us to generalize to any type of data structure, rather than assuming, for example, that the data is a relational table, image, data cube, or network-graph.
+Moreover, we extend the fiber bundle abstraction to the graphic objects that the data is mapped to.
+By doing so, we can track the preservation of data continuity in terms of continuous maps from the base space of the data bundle to the base space of the graphic bundle.
+Equivariant maps on the fiber spaces preserve the structure of the variables; this structure can be represented in terms of monoid actions, which are a generalization of the mathematical structure of Stevens’ theory of measurement scales.
+We briefly sketch that these transformations have an algebraic structure which lets us build complex components for visualization from simple ones.
+We demonstrate the utility of this model through case studies of a scatter plot, line plot, and image.
+To demonstrate the feasibility of the model, we implement a prototype of a scatter and line plot in the context of the Matplotlib Python visualization library.
+We propose that the functional architecture derived from a TEAM based design specification can provide a basis for a more consistent API and better modularity, extendability, scaling and support for concurrency}
 

### 2021-03-22 paper/paper.tex
reworked monoid, q, and xi figures

@@
 %other entries to be set up for journal
-\shortauthortitle{ \MakeLowercase{\textit{et al.}}: Topological Equivariant Artist Model}
+\shortauthortitle{Aizenman \MakeLowercase{\textit{et al.}}: Topological Equivariant Artist Model}
 %\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}
@@
 % http://ieeevis.org/year/2021/info/call-participation/paper-keywords
-\keywords{Taxonomy, Models, Frameworks, Theory }
+\keywords{Taxonomy, Models, Frameworks, Theory}
 
@@
 
-\autoref{fig:math:artist:nu} illustrates a \vchannel\ that preserves partial order relations because 
-ou would have to state that the monoid of monitonic actions is equivalent to the partial order realationship
-ok
-so that the order is preserved IF and only IF it is an equivariant map.
-I think that is true but it would be nice to have a reference.
-but then you don't have to really plot out m.
+\autoref{fig:math:artist:nu} illustrates a \vchannel\ that preserves partial order relations.
 
-Generally, for the Steven's scales\cite{stevensTheoryScalesMeasurement1946} an equivariant \vchannel\ has the following properties
+Stevens scales\cite{stevensTheoryScalesMeasurement1946} are groups, which are monoids with invertible elements,
+
 \begin{table}[H]
-  \begin{tabulary}{\columnwidth}{llL}
+  \begin{tabulary}{\columnwidth}{|llL|}\hline
+      \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
       nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
@@
       interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
-      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\
+      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
   \end{tabulary}
 \end{table}
-\noindent where the middle column is the group structure. Therefore, a \vchannel\ is equivariant if it satisfies the listed condition.
+
+\noindent where the group is a monoid with invertible elements. Therefore, a \vchannel\ is equivariant if it satisfies
 

### 2021-03-23 paper/paper.tex
nu and q

@@
 \usetikzlibrary{cd} % commutative diagrams
+\newtheorem{prop}{Proposition} %math?
 \usepackage[switch]{lineno}
@@
 As mentioned in \autoref{sec:math:data:monoid}, monoid actions define the structure on the fiber components and are therefore the basis for equivariance.
-A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes such that 
+A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes
 \begin{equation}
@@
 \end{equation}
+such that applying equivariant monoid actions to $\dtotal_{i}$ and $\vtotal_{i}$ preserves the map $\vchannel_{i}: \dtotal_{i}\rightarrow \vtotal_{i}$.
 In general, the data fiber $\dfiber_{i}$ cannot be assumed to be of the same type as the visual fiber $\vfiber_{i}$ and the actions of \monoid\ on $\dfiber_{i}$ cannot be assumed to be the same as the actions of $\monoid^{\prime}$ on \vfiber; therefore an equivariant $\vchannel_i$ must satisfy the constraint
@@
 
-\begin{figure}[htb]
-  \centering
-  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
-  \caption{fill in this caption tomorrow}
-  \label{fig:math:artist:nu}
-\end{figure}
-
-\autoref{fig:math:artist:nu} illustrates a \vchannel\ that preserves partial order relations.
-
-Stevens scales\cite{stevensTheoryScalesMeasurement1946} are groups, which are monoids with invertible elements,
-
 \begin{table}[H]
+  %%\renewcommand{\arraystretch}{2}
   \begin{tabulary}{\columnwidth}{|llL|}\hline
@@
   \end{tabulary}
+  \caption{}
+  \label{tab:math:artist:nu}
 \end{table}
+The equivariance constraints on the Stevens measurement scales \cite{stevensTheoryScalesMeasurement1946} are listed in \autoref{tab:math:artist:nu}.
+These are derived from the group actions on the scales, as defined by Stevens.
 
-\noindent where the group is a monoid with invertible elements. Therefore, a \vchannel\ is equivariant if it satisfies
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
+  \caption{In the top two panels \vchannel\ maps to visual elements such that the relative order of the visual elements is the same as the data.
+In the bottom right panel, \textit{rain} is mapped to a visual element that is less than the visual element \textit{storm} is mapped to, which is not an allowed \vchannel.
+Changing the map so that \textit{storm} goes to the same visual element as \textit{rain} yields an order preserving map and is therefore a valid \vchannel.}
+  \label{fig:math:artist:nu}
+\end{figure}
+
+The equivariance constraint for monoid actions is that relative order is preserved, as shown in the top two panels of \autoref{fig:math:artist:nu}.
+In the bottom right panel the ordering of \textit{rain} and \textit{storm} is swapped in the visual elements; in the bottom left panel, \textit{storm} is instead mapped to the same visual element as \textit{rain} so that the ordering is preserved.
 
@@
 As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\ reduce the encodings into a graphic \gsection.
-While it may seem intuitive that visualizations that generate the same graphic should do so consistently given the same input, we formalize this constraint such that it can be specified as a constraint of \vmark.
-We define the equivariant map as  $\vmark: \vsection \mapsto \gsection$ and an action on the subset of graphics $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ that \vmark\ can generate.
-We then define the constraint on \vmark such that if \vmark\ is applied to two visual sections $\vsection,\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection,\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.
-
+The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$.
+We formalize the expectation that visualization generation functions parameterized in the same way should generate the same functions as the equivariant map $\vmark: \vsection \mapsto \gsection$.
+We then define the constraint on \vmark\ such that if \vmark\ is applied to two visual sections $\vsection$ and $\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection$ and $\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.
+We cannot define monoid actions on all of $\Gamma(\gtotal)$ because there are graphics $\gsection \in \Gamma(\gtotal)$ for which we cannot construct a valid mapping from \vtotal.
 \begin{figure}[htb]
@@
   \includegraphics[width=\columnwidth]{diff_type_q.png}
-  \caption{These two glyphs are generated by the same \vmark\ function. The monoid 
-  action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
+  \caption{These two glyphs are generated by the same annulus \vmark\ function.
+The monoid action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
   \label{fig:math:artist:graphic}
 \end{figure}
-
-Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$.
+Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$
+\begin{prop}
 If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
@@
 \vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
+\label{eq:math:artist:q}
 \end{equation}
 then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
-The transformed graphic $\gsection^{\prime}$ is equivariant to a transform on the visual bundle $\gsection^{\prime}=\vmark(m\circ \vsection)$ on a section that $\vsection \in \vmark^{-1}(\gsection)$ that must be part of generating \gsection.
-Given fiber $\vfiber=(xpos,\, ypos,\, color,\, thickness)$, then sections $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
+If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection.
+\end{prop}
+
+For example, given fiber $\vfiber=(xpos,\, ypos,\, color,\, thickness)$, then sections $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
 The action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$ and the corresponding action on \gsection\ causes $\vmark(\vsection^{\prime})$ to be the thicker circle in \autoref{fig:math:artist:graphic}.
@@
   \centering
-  \includegraphics[width=.5\columnwidth]{base_q}
+  \includegraphics[width=\columnwidth]{base_q}
   \caption{The coordinates $(\gx, \gy)$ dictate the color of the region in prerender space \gbase\, which over the whole disk generates the graphical mark.
@@
 More work is needed to formalize the composition operators and equivalence class \vartisteq.
-More artists need to be implemented to demonstrate that the model can underpin a minimally viable library,  foremost an image\cite{toryRethinkingVisualizationHighlevel2004, haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an imherently computational artist such as the boxplot\cite{wickham40YearsBoxplots2011}.
+More artists need to be implemented to demonstrate that the model can underpin a minimally viable library,  foremost an image\cite{toryRethinkingVisualizationHighlevel2004, haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as the boxplot\cite{wickham40YearsBoxplots2011}.
 This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.

### 2021-03-23 paper/paper.tex
paper draft

@@
 \usepackage{tabulary}
+\usepackage[utf8]{inputenc}
 \usepackage{subfig} %ieee does not like subfigure
@@
 \maketitle
-The aim of this work is to rearchitecture Matplotlib to take advantage of developments in software design, data structures, and visualization to improve consistency, reusability, and discoverability, so domain specific tool developers can build structure preserving visualization tools.
-The contribution of this work is
+The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+Visualization tools allow practioners to automate these transformations, but most visualization tools are either explicitly tuned to specific continuity \cite{HeerSoftware2006} or define every interface relative to its expected data structure.
+The former approach is often too restrictive for a general purpose tool, while the latter often results in a library that can feel inconsistent or disjointed.
+Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation.
+The contribution of TEAM is
 \begin{enumerate}
@@
 \section{Related Work}
-The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}.
 The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization.
@@
 
-On the other hand, most information visualization tools are explicitly tuned to specific continuity \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}.
-For example, the relational database is core to tools influenced by APT, such as
-Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}.
+Much of visualization algorithm design is oriented around continuity, as described in Tory and Möller's taxonomy\cite{toryRethinkingVisualizationHighlevel2004}.
+For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}.
 Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}.
@@
  \label{sec:math:data:sheaf}
- The restriction maps of a sheaf describe how local sections $\iota^{*}\dsection$ over a neighborhood $U$ around \dbasepoint\ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
-This puts a continuous structure on local sections, which allows for defining a section over a subset in \dbase.
-The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$
+A sheaf is a mathematical structure for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.
+On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections $\iota^{*}\dsection$
  \begin{equation}
@@
  \end{equation}
- such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
-Gluing together $\iota^*\dsection$ sections is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.
+which are sections of \dtotal\ pulled back over local neighborhood $U \subset \dtotal$ via the inclusion map \(\iota:\dtotal\rightarrow U\).
+The bookkeeping enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}
 
@@
     \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
+    \label{eq:math:artist:artist}
 \end{equation}
@@
 \vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
+\label{eq:math:artist:nu:equivariance}
 \end{equation} 
@@
 
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
+  \caption{Equivariant \vchannel\ maps the weather data to corresponding symbols such that the relative order of the symbols is the same as the data as shown in the top two panels.
+In the bottom right panel the ordering of \textit{rain} and \textit{storm} is swapped in the symbols, meaning that the \vchannel\ is invalid; in the bottom left panel, \textit{storm} is instead mapped to the same symbol as \textit{rain} so that the ordering is preserved.}
+  \label{fig:math:artist:nu}
+\end{figure}
+
 \begin{table}[H]
-  %%\renewcommand{\arraystretch}{2}
+  \renewcommand{\arraystretch}{2}
   \begin{tabulary}{\columnwidth}{|llL|}\hline
@@
   \end{tabulary}
-  \caption{}
   \label{tab:math:artist:nu}
 \end{table}
-The equivariance constraints on the Stevens measurement scales \cite{stevensTheoryScalesMeasurement1946} are listed in \autoref{tab:math:artist:nu}.
-These are derived from the group actions on the scales, as defined by Stevens.
-
-\begin{figure}[htb]
-  \centering
-  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
-  \caption{In the top two panels \vchannel\ maps to visual elements such that the relative order of the visual elements is the same as the data.
-In the bottom right panel, \textit{rain} is mapped to a visual element that is less than the visual element \textit{storm} is mapped to, which is not an allowed \vchannel.
-Changing the map so that \textit{storm} goes to the same visual element as \textit{rain} yields an order preserving map and is therefore a valid \vchannel.}
-  \label{fig:math:artist:nu}
-\end{figure}
-
-The equivariance constraint for monoid actions is that relative order is preserved, as shown in the top two panels of \autoref{fig:math:artist:nu}.
-In the bottom right panel the ordering of \textit{rain} and \textit{storm} is swapped in the visual elements; in the bottom left panel, \textit{storm} is instead mapped to the same visual element as \textit{rain} so that the ordering is preserved.
+The equivariance constraints for the Steven measurement scales\cite{stevensTheoryScalesMeasurement1946} are derived from the mathematical group structure of the scales.
+For monoid actions, the equivariance constraint is that relative order is preserved, as shown in the top two panels of \autoref{fig:math:artist:nu}.
 
@@
 \end{figure}
-As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings \vsection\ reduce the encodings into a graphic \gsection.
+As shown in \autoref{fig:math:artist:q}, \vchannel\ and \vmark\ are analogous to a  map-reduce operation: data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal.
 The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$.
@@
 In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image.
-The scatter plot can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
+The scatter plot can be defined as
+\begin{equation}
+\vmark(xpos, ypos)(\alpha, \beta)
+\label{eq:math:artist:q:scatter}
+\end{equation}
+where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
 The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:artist:base}
@@
 such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black. 
-In contrast, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
+In contrast, the line plot 
+\begin{equation}
+\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)
+\label{eq:math:artist:q:line}
+\end{equation}
+in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
 As shown in \autoref{fig:math:artist:base}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal.
@@
 where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n_1}, \hat{n_2}$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
-
-The image $\vmark(xpos, ypos, color)$ in \autoref{fig:teaser} is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
+In \autoref{fig:teaser}, the image 
+\begin{equation}
+\vmark(xpos, ypos, color)
+\label{eq:math:artist:q:image}
+\end{equation}
+ is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
 The indexing variables $(\alpha, \beta)$ define the distance along the space, which is then used by \vindex\ to map into \dbase\ to lookup the color values
@@
 where \vmarkd\ with input \vsection\ is defined to \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.
-
-Factoring out \gbasepoint\ from \autoref{eq:math:artist:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
+Factoring out \gbasepoint\ from \autoref{eq:math:artist:qhat_q_s} yields
+\begin{equation}
+\vmarkd(\vsection(k)) = \vmark(\vsectionpull)
+\label{eq:math:artist:qhat_k}
+\end{equation}
+where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
 In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
@@
 This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.
-It is impractical to implement an artist for every single graphic; instead we implement an approximation of the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$.
+It is impractical to implement an artist for every single graphic; instead we implement an approximation of the equivalence class of artists
+\begin{equation}
+\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}
+\label{eq:math:artist:aprime}
+\end{equation}
 Roughly, two artists are equivalent if they have the same visual fiber \vfiber\, assembly function \vmark\, and continuity map \vindex.
@@
 \section{Prototype}
+\label{sec:code}
 To build a prototype, we make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to transform data coordinates into screen coordinates.
+While the artist is specified in a fully functional manner in \autoref{eq:math:artist:diagram}, we implement our prototype in a heavily object oriented manner.
+This is a concession to existing Matplotlib architecture and so that the object can do bookkeeping of parameter information.
+But, in keeping with functional norms, we attempt to pass around as little information as necessary.
 \begin{minted}{python}
@@
 \end{minted}
-Building on the current Matplotlib artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalance class artist \vartisteq.
+Building on the current Matplotlib artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalance class artist \vartisteq\ as described in \autoref{eq:math:artist:aprime} .
 The visual bundle \vtotal\ is specified as the \mintinline{python}{transform}dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber, and the \vchannel\ encoders are passed in as functions or callable objects.
 The data bundle \dtotal\ is passed in as a \mintinline{python}{data} object.
-By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist.
+By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist\ as defined in \autoref{eq:math:artist:artist}.
 \begin{minted}{python}
-  class ArtistClass(matplotlib.artist.Artist):
+  class ArtistClass(matplotlib.artist.Artist): #A'
       def __init__(self, data, transforms, *args, **kwargs):
           # properties that are specific to the graphic
-          self.data = data 
-          self.transforms = transforms
+          self.data = data #E
+          self.transforms = transforms #V
           super().__init__(*args, **kwargs)
 
-      def assemble(self, **args):
+      def assemble(self, **args): #\hat{Q}
           # set the properties of the graphic
 
-      def draw(self, renderer):
+      def draw(self, renderer): #A
           # returns K, indexed on fiber then key 
-          view = self.data.view(self.axes) 
+          view = self.data.view(self.axes) #\tau
           # visual channel encoding applied fiberwise 
+          # \nu_i(\tau_i)
           visual = {p: t['encoder'](view[t['name']])
@@
 
- The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}.
+ The \vchannel\ functions are then applied to the data, as describe in \autoref{eq:math:artist:nu}, to generate the visual section \vsection that here is the object \mintinline{python}{visual}.
 The conversion from data to visual space is simplified here to directly show that it is the encoding \vchannel\ applied to the component.
-The \mintinline{python}{assemble} function that is \vmarkd is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
+The \mintinline{python}{assemble} function that is \vmarkd,  as defined in \autoref{eq:math:artist:qhat_k}, is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
 This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd.
@@
     \subfloat[\label{fig:code:scatter}]{%
-      \includegraphics[width=0.49\columnwidth]{scatter_0.png}}
+      \includegraphics[width=0.5\columnwidth]{scatter_0.png}}
     \subfloat[\label{fig:code:line}]{%
-      \includegraphics[width=0.49\columnwidth]{line_1.png}}
+      \includegraphics[width=0.5\columnwidth]{line_1.png}}
   \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
@@
 To generate the figure in \autoref{fig:code:scatter}, the \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of primitive point and area marks.
-In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker.
+In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker, as described  by \autoref{eq:math:artist:q:scatter}.
 We only show the \mintinline{python}{assemble} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
@@
 class Point(mcollections.Collection):
-  def assemble(self, x, y, s, facecolors='C0' ):
+  def assemble(self, x, y, s, facecolors='C0' ): #\hat{Q}
     # construct geometries of circle glyphs
@@
 class Line(mcollections.LineCollection):
-    def assemble(self, x, y, color='C0'):
+    def assemble(self, x, y, color='C0'): #\hat{Q}
             #assemble line marks as set of segments 
@@
 As with \mintinline{python}{Point}, the data is then converted into visual variables.
-In \mintinline{python}{assemble}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
+In \mintinline{python}{assemble}, described by \autoref{eq:math:artist:q:line}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
 Then the colors of each line segment are set.
@@
                            'false':'deepskyblue'}) 
+# {P_i name:{'name':c_i, 'encoder:\nu_i}}
 transforms = {'x': {'name': 'v4', 'encoder': lambda x: x}, 
@@
 \end{minted}
-
 where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.
-A test for equivariance can be implemented trivially
+A test for equivariance, as described in \autoref{eq:math:artist:nu:equivariance}, can be implemented trivially
 \begin{minted}{python}
+#\nu_i(m_r(E_i)) = \varphi(m_r)(\nu_i((E_i))
 def test_nominal(values, encoder):
@@
   \subfloat[\label{fig:code:line:arc}]{%
-    \includegraphics[width=0.49\columnwidth]{linec_1.png}}
+    \includegraphics[width=0.5\columnwidth]{linec_1.png}}
   \subfloat[\label{fig:code:line:step}]{%
-    \includegraphics[width=0.49\columnwidth]{lined_1.png}}
+    \includegraphics[width=0.5\columnwidth]{lined_1.png}}
 \caption{Continuous and discontinuous lines as defined via the same data model, and generated with the same \vartisteq \mintinline{python}{Line}}
@@
 \acknowledgments{
-The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran for discussions about the visual fiber space.
+The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran.
 This project has been made possible in part by grant number 2019-207333 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.}

### 2021-03-24 slides/main.tex
figure tweaks
moved figures creation code into a mostly central folder, added k labels to fiber bundle fig
bulk changes, everything but timeline

@@
 \documentclass[xcolor={dvipsnames}, handout]{beamer}
-
+%\documentclass[xcolor={dvipsnames}]{beamer}
 %\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
@@
 \usepackage{tikz}
+\newtheorem{prop}{Proposition} %math?
 
@@
 \usepackage{tabulary}
+\usepackage{booktabs}
 \usepackage{multirow}
+
 \usepackage{graphicx}
@@
 
-\usepackage[backend=bibtex, style=numeric-comp, sorting=none, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
+\usepackage[backend=bibtex, style=alphabetic, sorting=none, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
 \bibliography{../draft/glasslab_viz.bib}
@@
     \begin{figure}
-        \includegraphics[width=1\linewidth]{figures/intro/dar.png}
+        \includegraphics[width=\textwidth]{figures/intro/viz_same.png}
     \end{figure}
-    The aim of this work is to rearchitecture Matplotlib
-    \pause 
-    to take advantage of developments in software design, data structures, and visualization
-    \pause
-    to improve consistency, reusability, and discoverability,
-    \pause
-    so domain specific tool developers can build structure preserving visualization tools.
-\end{frame}
-
-\begin{frame}{Visualization component constraints}
     \begin{description}
@@
 
-\subsection{Background}
-\begin{frame}{Tools are tuned to date continuity \cite{HeerSoftware2006}}
-    \begin{columns}
-            \column{0.33\textwidth}
-            \begin{figure}
-                \includegraphics[width=1\textwidth]{figures/intro/data_formatting.png}
-                \caption{Based on fig~2.5 in Munzner's VAD\cite{munznerVisualizationAnalysisDesign2014}}
-            \end{figure}
-            \begin{enumerate}
-                \item Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}
-                \item ggplot\cite{wickhamGgplot2ElegantGraphics2016a}
-                \item Vega\cite{satyanarayanDeclarativeInteractionDesign2014}, Altair\cite{vanderplasAltairInteractiveStatistical2018}
-            \end{enumerate}
-            \pause
-            \column{0.33\textwidth}
-            \begin{figure}
-                \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
-            \end{figure}
-            \begin{enumerate}
-                \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}
-                \item Napari\cite{nicholas_sofroniew_2021_4533308}
-            \end{enumerate}
-            \pause
-            \column{0.33\textwidth}
-            \begin{figure}
-                \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
-                \caption{Data Representation, MayaVi 4.7.2 docs\cite{DataRepresentationMayavi}}
-            \end{figure}
-            \begin{enumerate}
-                \item Gephi\cite{bastianGephiOpenSource2009}
-                \item Graphviz\cite{ellsonGraphvizOpenSource2002}
-                \item Networkx\cite{HagbergExploringNetwork2008}
-            \end{enumerate}
-    \end{columns}
+\begin{frame}{Tools are tuned to continuity \cite{HeerSoftware2006}}
+\begin{table}
+    %\renewcommand{\arraystretch}{2}
+    \begin{tabulary}{\textwidth}{|L|l|l|}\toprule
+        \includegraphics[width=.24\textwidth]{figures/intro/table.png} & \includegraphics[width=.3\textwidth]{figures/intro/landsat.png} & \includegraphics[width=.33\textwidth]{figures/math/graph.png} \\
+        ggplot\cite{wickhamGgplot2ElegantGraphics2016a}  & ImageJ\cite{schneiderNIHImageImageJ2012}& Gephi\cite{bastianGephiOpenSource2009}\\
+        Vega\cite{satyanarayanDeclarativeInteractionDesign2014} & ImagePlot\cite{studiesCulturevisImageplot2021} & Graphviz\cite{ellsonGraphvizOpenSource2002}\\
+        Altair\cite{vanderplasAltairInteractiveStatistical2018}& Napari\cite{nicholas_sofroniew_2021_4533308} & Networkx\cite{HagbergExploringNetwork2008}\\
+         Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} & &\\
+         \bottomrule
+    \end{tabulary}
+\end{table}
 \end{frame}
 
-\begin{frame}{Visualizations are tuned to data continuinity\cite{toryRethinkingVisualizationHighlevel2004}}
+\begin{frame}{Visualizations are tuned to continuinity\cite{toryRethinkingVisualizationHighlevel2004}}
     \begin{figure}
@@
 
-\begin{frame}{Structure is encoded in variables and continuity}
-    \begin{figure}
-        \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
-        \caption{Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
-    \end{figure}
+\begin{frame}{Tools are evaluated on the structure they preserve}
     \begin{description}
-        \item[binding] metadata are structural \textit{keys} with associated \textit{values} (Munzner \cite{munznerVisualizationAnalysisDesign2014})] 
-        \item[continuity] Fiber bundles can be a common data abstraction (Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989})
-        \item[variables] Fibers can hold schema like encodings of variables (Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES})
-    \end{description}
-\end{frame}
-
-\begin{frame}{Visualizations are (mostly) evaluated on equivariance}
-    \begin{description}
-        \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
-        \item[Effectiveness] design choices made in deference to perceptual saliency (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
-        \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
-        \item[Graphical Integrity] graphs show \textbf{only} the data (Tufte \cite{tufteVisualDisplayQuantitative2001})
-    \end{description}
-\end{frame}
-
-\begin{frame}{Models describe composition}
-    \begin{description}
-        \item[language model] APT, GoG: syntax, semantics, and grammar of graphics (Mackinlay, Wilkenson  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005})
-        \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
-        \item[category theory] the semiotics of visualization are commutative (Vickers \cite{vickersUnderstandingViz2013})
-        \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
-        \begin{columns}
-            \column{.5\textwidth}
-
-            \begin{description}
-                \item[D] data 
-                \item[R] representations
-                \item[V] visualizations
-            \end{description}
-            \column{.5\textwidth}
-            \begin{equation*}
-                \begin{tikzcd}[ampersand replacement=\&]
-                    D \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& R \arrow[r, "\nu"]  \& V \arrow[d, "\omega"] \\
-                    D \arrow[r, "r_2"']                     \& R \arrow[r, "\nu"'] \& V                    
-                \end{tikzcd}
-                \end{equation*}
-
-        \end{columns}
-
+        \item[Naturalness] easier to understand when properties match \cite{norman_things_smart}
+        \item[Graphical Integrity] graphs show \textbf{only} the data \cite{tufteVisualDisplayQuantitative2001}
+        \item[Expressiveness] structure preserving mappings from data to graphic \cite{mackinlayAutomatingDesignGraphical1986}
+        \item[Effectiveness] design choices made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}
     \end{description}
@@
 \begin{frame}{Topological Equivariant Artist Model}
-    The Artist $\mathscr{\vartist}$ is a map from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$
-    \begin{equation}
+    \centering
+    An artist $\mathscr{\vartist}$ is an equivariant map 
+    \begin{equation}\Huge
         \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
     \end{equation}
-    \pause
-     that carries a homomorphism of monoid actions 
-    \begin{equation}
-    \varphi: \monoid \rightarrow \monoid^{\prime}
-    \end{equation}
-    \pause
-    such that artists are equivariant maps 
-    \begin{equation}
-    \mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
-    \end{equation}
-    \pause 
-    with a deformation retraction from graphic to data space. 
-    \note{some kinda words about the continuity}
+    from data $\mathscr{\dtotal}$ space to graphic $\mathscr{\gtotal}$ space.
+
 \end{frame}
@@
 
-\begin{frame}{Data Bundle}
-    \begin{figure}
-        \includegraphics[width=1\textwidth]{figures/math/fiberbundle.png}
-    \end{figure}
-    A fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
+\begin{frame}{Model data as a fiber bundle \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}}
+    A fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the map $\pi$
     \begin{equation}
@@
     \end{equation}
-    where \dtotal\ is the total data space, \dfiber\ is the variable space, and \dbase\ encodes the continuity.
+    \begin{description}
+        \item[base space \dbase] coordinate system agnostic \textit{keys}
+        \item[fiber space \dfiber] contains \textit{values}    
+    \end{description}
+    \begin{block}{Structural \textit{keys} with associated \textit{values} \cite{munznerVisualizationAnalysisDesign2014}}
+        \begin{figure}
+            \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
+            \caption{Figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
+        \end{figure}
+    \end{block}
+
 \end{frame}
 
-\begin{frame}{Variables: Fiber}
-    Given a space of all possible values \ftotal\
-    \begin{equation}
-        \label{eq:data_types}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
-            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
-        \end{tikzcd}
-    \end{equation}
-    a fiber component is the restricted space $\ftotal_{\fsection(\fname)}$. 
-    \begin{equation}
-        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
-    \end{equation}
-    \begin{description}
-        \item[\ftypes] data types of the variables in the dataset 
-        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
-        \item[\fnames] variable names, $\fname \in \fnames$
-        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
-    \end{description}
-\end{frame}
-
-\begin{frame}{Variable types are dimensions of the fiber}
+\begin{frame}{Encode variable types in a schema like fiber \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
     \begin{figure}[H]
-        \includegraphics[width=\textwidth]{figures/math/fiber.png}
-        \caption{}
+        \centering
+        \includegraphics[height=.5\textwidth]{figures/math/fiber.png}
         \label{fig:data_fiber_example}
     \end{figure}
-    \begin{description}
-        \item[plane]  $\dfiber=\reals\times\reals$, (time, temperature)
-        \item[cube]  $\reals\times\realsp\times\reals$, (time, wind=(speed, direction))
-    \end{description}
-\end{frame}
-
-\begin{frame}{Structure of Components: Monoid \& Monoid Actions}
-A monoid \monoid\ is a set with
-\begin{description}
-    \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
-    \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
-\end{description}
-\pause
-\begin{block}{left monoid action}
-A set \dfiber\ with an action\cite{nlab:action} $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
-    \begin{align*}
-        \textbf{associativity}\;& \text{for all } f,g \in \monoid \text{ and } x\in \dfiber,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-        \textbf{identity}\;& \text{for all } x\in \dfiber, e\in \monoid,\,  e\bullet x = x 
-    \end{align*}
+    \begin{block}{Fiber components can be acted on independently}
+        \begin{equation}
+            \dfiber = \dfiber_0 \times \ldots \times \dfiber_i \times \ldots \times \dfiber_n
+        \end{equation}
     \end{block}
@@
 
-\begin{frame}{Monoid Actions: Permutation}
-    \begin{figure}
-        \includegraphics[width=1\linewidth]{figures/math/monoid_emoji.png}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Why monoids? partial orders}
-    \begin{figure}
-        \begin{overprint}
-            \onslide<1|handout:0>\includegraphics[width=1\linewidth]{figures/math/monoid_hasse.png}
-            \onslide<2|handout:0>\includegraphics[width=1\linewidth]{figures/math/monoid_monotone.png}
-            \onslide<3>\includegraphics[width=1\linewidth]{figures/math/monoid_maps.png}
-        \end{overprint}
-    \caption{Inspired by definition 1.59 diagram in Spivak and Fong's An Invitation to Applied Category Theory \cite{fongInvitationAppliedCategory2019}}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Data Continuity: Base space}
-    \begin{figure}[H]
-    \includegraphics[height=.5\textheight]{figures/math/k_qspace.png}
-    \label{fig:base_space_div}
-\end{figure}
-where the total space can be decomposed into components 
-\begin{equation}
-    \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
-\end{equation}
-\end{frame}
-
-\begin{frame}{Data connectivity is encoded as the base space}
+\begin{frame}{}
     \begin{figure}[H]
@@
     \end{figure}
-    \begin{description}
-        \item[points] data is 0D discrete
-        \item[line]  data is lies on the 1D continuous interval \dbase\
-    \end{description}
+    \begin{block}{\dbase\ is an indexing space regardless of components}
+        \begin{equation}
+            \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
+        \end{equation}
+    \end{block}
 \end{frame}
 
-\begin{frame}{Values: Section}
+\begin{frame}{Data are sections\dsection\ on the \dtotal}
     For any fiber bundle, there exists a map
@@
     \end{equation}
-     s.t. $\pi(\dsection(\dbasepoint)) = \dbasepoint$.  Set of all global sections is denoted $\Gamma(\dtotal)$.
-     \pause
-     \begin{block}{Record}
-     Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is 
-\begin{equation}
-    \label{eq:section_return}
-    \dsection(\dbasepoint) = (\dbasepoint, (g_{\dfiber_{0}}(\dbasepoint), \ldots, g_{\dfiber_{n}}(\dbasepoint)))
-\end{equation}
-where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber.
-\end{block}
-\end{frame}
-
-\begin{frame}{Sample dataset}
+     s.t. $\pi(\dsection(\dbasepoint)) = \dbasepoint$. $\Gamma(\dtotal)$ is the set of all global sections.
      \begin{figure}[H]
@@
     \end{figure}
-    \begin{itemize}
-        \item \dfiber\ is $\reals\times\reals$
-        \item \dbase\ is interval $\left[0,1\right]$
-        \item $\dsection^{(1)}$ is a $sin$ function
-        \item $\dsection^{(2)}$ is a $cos$ function
-        \item $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Sheafs}
-    Restriction maps of a sheaf describe how local $\iota^*\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
-
-    \begin{equation}
-        \label{eq:sheaf}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             \& \dtotal \arrow[d, "\pi"']                  \\
-            U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] \& \dbase \arrow[u, "\dsection"', bend right]
-        \end{tikzcd}
-    \end{equation}
-    The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$ such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
 \end{frame}
@@
 \begin{frame}{Topological Equivariant Artist Model}
-The topological artist \vartist\ is a monoid equivariant sheaf map 
+The topological artist $\vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)$ is sheaf map
 \begin{equation}
@@
 \end{equation}
+that carries homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019} 
+\begin{equation}\huge
+\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
+\end{equation} 
+\end{frame}
 
-where the artist $\vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)$ takes as input $\dtotal^{\prime}=\mathcal{J}^{2}(\dtotal)$. 
+\begin{frame}{Existing models don't have continuity}
+        \begin{description}
+            \item[language model] APT, GoG: syntax, semantics, and grammar of graphics (Mackinlay, Wilkenson  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005})
+            \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
+            \item[category theory] data, representation, and understanding are commutative (Vickers \cite{vickersUnderstandingViz2013})
+            \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
+            \begin{columns}
+                \column{.5\textwidth}
+                \begin{description}
+                    \item[\vtotal] visual components
+                    \item[$\alpha$] \monoid
+                    \item[$\omega$] $\varphi(\monoid)$
+                \end{description}
+                \column{.5\textwidth}
+                \begin{equation*}
+                    \begin{tikzcd}[ampersand replacement=\&]
+                        \dtotal \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& \vtotal \arrow[r, "\nu"]  \& \gtotal \arrow[d, "\omega"] \\
+                        \dtotal \arrow[r, "r_2"']                     \& \vtotal \arrow[r, "\nu"'] \& \gtotal                  
+                    \end{tikzcd}
+                    \end{equation*}
+
+            \end{columns}
+
+        \end{description}
 \end{frame}
@@
 
-\begin{frame}{Example: Nominal Equivariance}
+\begin{frame}{Partial Order Equivariance and Stevens Scales \cite{stevensTheoryScalesMeasurement1946}}
     \begin{figure}
@@
     \end{figure}
-\end{frame}
-\begin{frame}{Measurement Scale Groups}
 
@@
 \end{table}
-
-    \pause
-    \begin{block}{Invalid \vchannel}
-     Given $\vchannel_{i}(x) = .5$ and $t(x) = x+2$,
-    \pause    
-    \begin{align*}
-            \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + 2\\
-            .5 &\neq .5 + 2
-    \end{align*}
-    \end{block}
 \end{frame}
@@
 
-\begin{frame}{Glyph}
-    The glyph is the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$ where the path connected components $\dbasepath \subset \dbase$ are defined 
-    \begin{equation}
-    \dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
-    \end{equation}
-    such that the  path $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1] and $\gbase_{\dbasepathpoint}$ is the region
-    \begin{equation}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \gtotal \arrow[r, shift left] \& \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] \& \& \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
-            \end{tikzcd}
-        \label{eq:mark}
-    \end{equation}
-    such that the glyph is differentiable, in keeping with Ziemkiewicz and Kosara's description of a glyph\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
-\end{frame}
-
 \begin{frame}{Visualization Equivariance}
-    If \vmark\ is applied to $\vsection, \vsection^{\prime}$ that generate the same \gsection\, 
-    \begin{equation}
-    \vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
-    \end{equation}
-    then the output of both sections acted on by the same monoid $m$ must be the same. 
-    \pause
     \begin{figure}[H]
@@
     \end{figure}
+    \begin{prop}
+        If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
+        \begin{equation}
+        \vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
+        \label{eq:math:artist:q}
+        \end{equation}
+        then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
+If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection.
+    \end{prop}
 \end{frame}
@@
 \appendix 
+\begin{frame}{Fiber is all possible values a variable can be \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
+    Given a space of all possible values \ftotal\
+    \begin{equation}
+        \label{eq:data_types}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
+            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
+        \end{tikzcd}
+    \end{equation}
+    a fiber component is the restricted space $\ftotal_{\fsection(\fname)}$. 
+    \begin{equation}
+        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
+    \end{equation}
+    \begin{description}
+        \item[\ftypes] data types of the variables in the dataset 
+        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
+        \item[\fnames] variable names, $\fname \in \fnames$
+        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
+    \end{description}
+\end{frame}
+\begin{frame}{Monoid actions}
+    A monoid \monoid\ is a set with
+    \begin{description}
+        \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
+        \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
+    \end{description}
+    \pause
+    \begin{block}{left monoid action}
+    A set \dfiber\ with an action\cite{nlab:action} $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
+        \begin{align*}
+            \textbf{associativity}\;& \text{for all } f,g \in \monoid \text{ and } x\in \dfiber,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
+            \textbf{identity}\;& \text{for all } x\in \dfiber, e\in \monoid,\,  e\bullet x = x 
+        \end{align*}
+    \end{block}
+\end{frame}
+\begin{frame}{Keeping track of sections with sheafs}
+    Restriction maps of a sheaf describe how local $\iota^*\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
+
+    \begin{equation}
+        \label{eq:sheaf}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             \& \dtotal \arrow[d, "\pi"']                  \\
+            U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] \& \dbase \arrow[u, "\dsection"', bend right]
+        \end{tikzcd}
+    \end{equation}
+    The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$ such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
+\end{frame}
+
 \begin{frame}{Rendering: Define a Pixel}
@@
 \end{frame}
+
+\begin{frame}{Glyph}
+    The glyph is the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$ where the path connected components $\dbasepath \subset \dbase$ are defined 
+    \begin{equation}
+    \dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
+    \end{equation}
+    such that the  path $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1] and $\gbase_{\dbasepathpoint}$ is the region
+    \begin{equation}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \gtotal \arrow[r, shift left] \& \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] \& \& \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
+            \end{tikzcd}
+        \label{eq:mark}
+    \end{equation}
+    such that the glyph is differentiable, in keeping with Ziemkiewicz and Kosara's description of a glyph\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+\end{frame}
+
 \begin{frame}[fragile]{GraphLine Data Model}

### 2021-03-24 slides/scraps.tex
bulk changes, everything but timeline

@@
     \end{figure}
+\end{frame}
+
+\begin{frame}{Visualizations are (mostly) evaluated on equivariance}
+    \begin{description}
+        \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
+        \item[Effectiveness] design choices made in deference to perceptual saliency (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
+        \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
+        \item[Graphical Integrity] graphs show \textbf{only} the data (Tufte \cite{tufteVisualDisplayQuantitative2001})
+    \end{description}
 \end{frame}
@@
 \end{frame}
-\begin{frame}{Evaluating visualizations}
+
+\begin{frame}{Models describe composition}
     \begin{description}
-        \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
-        \item[Effectiveness] design choices made in deference to perceptual saliency \cite (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
-        \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
-        \item[Graphical Integrity] graphs show \textbf{only} the data (Tufte \cite{tufteVisualDisplayQuantitative2001})
+        \item[language model] APT, GoG: syntax, semantics, and grammar of graphics (Mackinlay, Wilkenson  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005})
+        \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
+        \item[category theory] the semiotics of visualization are commutative (Vickers \cite{vickersUnderstandingViz2013})
+        \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
+        \begin{columns}
+            \column{.5\textwidth}
+
+            \begin{description}
+                \item[D] data 
+                \item[R] representations
+                \item[V] visualizations
+            \end{description}
+            \column{.5\textwidth}
+            \begin{equation*}
+                \begin{tikzcd}[ampersand replacement=\&]
+                    D \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& R \arrow[r, "\nu"]  \& V \arrow[d, "\omega"] \\
+                    D \arrow[r, "r_2"']                     \& R \arrow[r, "\nu"'] \& V                    
+                \end{tikzcd}
+                \end{equation*}
+
+        \end{columns}
+
     \end{description}
@@
 \end{frame}
+
+\begin{frame}{Structure is encoded in variables and continuity}
+    \begin{figure}
+        \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
+        \caption{Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
+    \end{figure}
+    \begin{description}
+        \item[binding] metadata are structural \textit{keys} with associated \textit{values} (Munzner \cite{munznerVisualizationAnalysisDesign2014})] 
+        \item[continuity])
+        \item[variables] Fibers can hold schema like encodings of variables (Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES})
+    \end{description}
+\end{frame}
+
+begin{frame}{Monoid Actions: Permutation}
+    \begin{figure}
+        \includegraphics[width=1\linewidth]{figures/math/monoid_emoji.png}
+    \end{figure}
+\end{frame}
+
+\begin{frame}{Why monoids? partial orders}
+    \begin{figure}
+        \begin{overprint}
+            \onslide<1|handout:0>\includegraphics[width=1\linewidth]{figures/math/monoid_hasse.png}
+            \onslide<2|handout:0>\includegraphics[width=1\linewidth]{figures/math/monoid_monotone.png}
+            \onslide<3>\includegraphics[width=1\linewidth]{figures/math/monoid_maps.png}
+        \end{overprint}
+    \caption{Inspired by definition 1.59 diagram in Spivak and Fong's An Invitation to Applied Category Theory \cite{fongInvitationAppliedCategory2019}}
+    \end{figure}
+\end{frame}

### 2021-03-24 slides/main.tex
still too many slides probably

@@
 \usepackage{tabulary}
+\usepackage{tabularx}
 \usepackage{booktabs}
@@
     \begin{enumerate}
-        \item Matplotlib\cite{hunterMatplotlib2DGraphics2007}, 
+        \item Matplotlib\cite{hunterMatplotlib2DGraphics2007} $\rightarrow$Seaborn\cite{waskom2020seaborn}, xarray \cite{hoyer2017xarray} 
         \item D3 \cite{bostockDataDrivenDocuments2011}
-        \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
+        \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}(MayaVi\cite{RamachandranMayaVI2011})$\rightarrow$ Titan\cite{brianwylieUnifiedToolkitInformation2009}, ParaView\cite{ahrens2005paraview}
     \end{enumerate}
@@
         \centering
-        \includegraphics[height=.5\textwidth]{figures/math/fiber.png}
+        \includegraphics[height=.3\textwidth]{figures/math/fiber.png}
         \label{fig:data_fiber_example}
     \end{figure}
-    \begin{block}{Fiber components can be acted on independently}
+    \begin{block}{Monoid actions define the structure on $\dfiber_i$}
         \begin{equation}
@@
         \end{equation}
+
+        $\monoid_{i}$ actions (e.g. rotation, partial ordering) 
+        \begin{equation}
+            \bullet: \monoid_i\times \dfiber_i \rightarrow \dfiber_i
+        \end{equation}
+        must be associative and have an identity element.
+
     \end{block}
@@
         \includegraphics[width=1\textwidth]{figures/math/base.png}
-        \caption{}
         \label{fig:base_example}
@@
 \subsection{Graphics}
-\begin{frame}{Graphic Bundle}
-    The graphics bundle is a tuple $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$ defined by the projection map $\pi$
+\begin{frame}{Graphic Bundle $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$}
+    Continuity is preserved via the deformation retraction $\vindex: \gbase\rightarrow\dbase$
     \begin{equation}
         \begin{tikzcd}[ampersand replacement=\&]
-            \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
-                              \& \gbase \arrow[u, "\gsection"', bend right]
+            F \arrow[r, hook] \& E \arrow[d, "\pi"']              \& D \arrow[r, hook] \& H \arrow[d, "\pi"']                                 \\
+                              \& K \arrow[u, "\tau"', bend right] \&                   \& S \arrow[ll, "\xi"'] \arrow[u, "\rho"', bend right]
             \end{tikzcd}
         \end{equation}
-    where \gsection\ is the fully encoded graphic. 
-    \pause
-    \begin{block}{Example: 2D opaque image}
-    The target display is $\gfiber=\reals^5$ with elements 
-    \begin{equation*}
-    (x,\, y,\, r,\, g,\, b) \in \gfiber
-    \end{equation*}
-    returned by \gsection\ such that a graphic has color and 2D position.
-    \end{block}
-
-\end{frame}
-\begin{frame}{Graphic Continuity}
-    The surjective map $\vindex: \gbase \rightarrow \dbase$
-    \begin{equation}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \dtotal \arrow[d, "\pi"'] \& \gtotal \arrow[d, "\pi"'] \\
-            \dbase                   \& \gbase \arrow[l, "\vindex"']
-        \end{tikzcd}
-    \end{equation}
-    goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\dbasepoint$ in data space. 
-\pause
+    \gfiber\ is a proxy for the target display, for example $(x,\, y,\, r,\, g,\, b) \in \gfiber$
     \begin{figure}[H]
@@
 \begin{frame}{Topological Equivariant Artist Model}
-The topological artist $\vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)$ is sheaf map
+The topological artist is a sheaf map
+\begin{equation}
+\vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
+\end{equation}
+that carries homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019} 
+\begin{equation}
+\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
+\end{equation} 
+and is implemented via an intermediate visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$
 \begin{equation}
@@
 \end{equation}
-that carries homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019} 
-\begin{equation}\huge
-\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
-\end{equation} 
 \end{frame}
 
-\begin{frame}{Existing models don't have continuity}
+\begin{frame}{Existing models don't have continuity, many evaluative}
         \begin{description}
-            \item[language model] APT, GoG: syntax, semantics, and grammar of graphics (Mackinlay, Wilkenson  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005})
-            \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
-            \item[category theory] data, representation, and understanding are commutative (Vickers \cite{vickersUnderstandingViz2013})
-            \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
-            \begin{columns}
-                \column{.5\textwidth}
-                \begin{description}
-                    \item[\vtotal] visual components
-                    \item[$\alpha$] \monoid
-                    \item[$\omega$] $\varphi(\monoid)$
-                \end{description}
-                \column{.5\textwidth}
+            \item[algebraic process] data and viz transforms are symmetric \cite{kindlmannAlgebraicProcessVisualization2014}
             \begin{equation*}
                 \begin{tikzcd}[ampersand replacement=\&]
-                        \dtotal \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& \vtotal \arrow[r, "\nu"]  \& \gtotal \arrow[d, "\omega"] \\
+                    \dtotal \arrow[d, "m"'] \arrow[r, "r_1"] \& \vtotal \arrow[r, "\nu"]  \& \gtotal \arrow[d, "\varphi(m)"] \\
                     \dtotal \arrow[r, "r_2"']                     \& \vtotal \arrow[r, "\nu"'] \& \gtotal                  
@@
 
-            \end{columns}
+            \item[language] APT and GoG: syntax, semantics, and grammar\\ 
+            \cite{mackinlayAutomatingDesignGraphical1986,mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005}
+            \item[functional dependencies] relationship between components \cite{sugibuchiFramwork2009}
+            \item[category theory]  \textit{understanding} = \textit{read} $\circ$ \textit{render} \cite{vickersUnderstandingVisualizationFormal2013}
 
         \end{description}
-\end{frame}
-
-\begin{frame}{Visual Bundle}
-    The visual bundle is a tuple $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ defined by the projection map $\pi$
-    \begin{equation}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
-                              \& \dbase \arrow[u, "\vsection"', bend right]
-        \end{tikzcd}
-    \end{equation}
-    where \vsection\ is the visual variable encoding\cite{bertinIIPropertiesGraphic2011} of the data section \dsection. 
-    \begin{block}{Example: position and color}
-        Given an artist with parameters $\{xpos, ypos, color\}$, a sample visual section \vsection\ could be $\{.5, .5, (255, 20,147)\}$
-    \end{block}
 \end{frame}
@@
 
-\begin{frame}{Partial Order Equivariance and Stevens Scales \cite{stevensTheoryScalesMeasurement1946}}
+\begin{frame}{Equivariance: Partial Order \& and Stevens' Scales \cite{stevensTheoryScalesMeasurement1946}}
     \begin{figure}
@@
 \begin{table}[H]
-    \begin{tabulary}{\textwidth}{llL}
+    \begin{tabularx}{\textwidth}{|l|l|X|}\toprule
+        \textbf{scale} & \textbf{group} & \textbf{constraint} \\\midrule
         nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
@@
         interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
-        ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $
-    \end{tabulary}
+        ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \bottomrule
+    \end{tabularx}
 \end{table}
@@
 \begin{frame}{Assembly Function Factory}
+\begin{equation}\label{eq:qhat_q_s}
+    \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
+\end{equation} 
 \begin{figure}[H]
@@
 \end{figure}
-\begin{equation}\label{eq:qhat_q_s}
-    \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
-\end{equation} 
 such \gbasepoint\ can be factored out when $\vpreimg = \gbasepoint$
@@
 \end{minted}
-\pause
 That we can test for action equivariance
@@
     \end{description}
-    \pause
     \begin{block}{left monoid action}

### 2021-03-24 slides/main.tex
transition into code

@@
     \begin{description}
-        \item[base space \dbase] coordinate system agnostic \textit{keys}
-        \item[fiber space \dfiber] contains \textit{values}    
+        \item[base space \dbase] continuity
+        \item[fiber space \dfiber] variable characteristics
+        \item[total space \dtotal] dataset topology
     \end{description}
-    \begin{block}{Structural \textit{keys} with associated \textit{values} \cite{munznerVisualizationAnalysisDesign2014}}
-        \begin{figure}
-            \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
-            \caption{Figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
-        \end{figure}
-    \end{block}
-
 \end{frame}
@@
 
-\begin{frame}{}
+\begin{frame}{Structural \textit{keys} with associated \textit{values} \cite{munznerVisualizationAnalysisDesign2014}}
     \begin{figure}[H]
@@
     \end{figure}
-    \begin{block}{\dbase\ is an indexing space regardless of components}
+
+    \begin{block}{\dbase\ is an indexing space independent of component semantics}
         \begin{equation}
@@
 
-\begin{frame}{Data are sections\dsection\ on the \dtotal}
+\begin{frame}{Data are sections \dsection\ on \dtotal}
     For any fiber bundle, there exists a map
@@
 \end{frame}
+\section{Prototype}
 
-\section{Prototype}
-\subsection{Artist}
-\begin{frame}[fragile]{Artist}
-    \begin{minted}{python}
-        class ArtistClass(matplotlib.artist.Artist):
-            def __init__(self, data, transforms, *args, **kwargs):
-                # properties that are specific to the graphic
-                self.data = data 
-                self.transforms = transforms
-                super().__init__(*args, **kwargs)
-
-            def assemble(self, **args):
-                # set the properties of the graphic
-
-            def draw(self, renderer):
-                # returns K, indexed on fiber then key 
-                view = self.data.view(self.axes) 
-                # visual channel encoding applied fiberwise 
-                visual = {p: t['encoder'](view[t['name']])
-                          for p, t in self.transforms.items()}
-                self.assemble(**visual)
-                # pass configurations off to the renderer
-                super().draw(renderer)
-        \end{minted}
+\begin{frame}{TEAM driven rearchitecture of Matplotlib}
+    \begin{itemize}
+        \item fiber bundle with schema is a structure rich abstraction
+        \begin{itemize}
+            \item topologically complex heterogenous data 
+            \item target display spaces
+        \end{itemize}
+        \item structure preserving maps from data to visual:
+            \begin{itemize}
+                \item data and graphics have equivalent continuity
+                \item properties are equivariant under monoid actions
+            \end{itemize}  
+        \item composable into complex visualizations
+    \end{itemize}
 \end{frame}
@@
 
-\begin{frame}[fragile]{Artists: Scatter \& Line}
-\begin{minted}{python}
-class Point(mcollections.Collection):
-    def assemble(self, x, y, s, facecolors='C0' ):
-        # construct geometries of the circle glyphs in visual coordinates
-        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
-                    for (xi, yi, si) in zip(x, y, s)] 
-        # set attributes of glyphs, these are vectorized 
-        # circles and facecolors are lists of the same size
-        self.set_facecolors(facecolors)
-\end{minted} 
-    \pause
-\begin{minted}{python}
-class Line(mcollections.LineCollection):
-    def assemble(self, x, y, color='C0'):
-        #assemble line marks as set of segments 
-        segments = [np.vstack((vx, vy)).T for vx, vy in zip(x, y)]
-        self.set_segments(segments)
-        self.set_color(color)
-\end{minted}
-\end{frame}
-
+\subsection{Visual Transforms}
 \begin{frame}[fragile]{Visual Transformations}
@@
 
-\begin{frame}[fragile]{Custom Complex \vchannel}
+    \begin{frame}[fragile]{Complex \vchannel}
     \begin{minted}{python}
@@
 \end{frame}
-\subsection{Data}
-\begin{frame}[fragile]{Fiber Bundle}
+
+\subsection{Artist}
+\begin{frame}[fragile]{Artist}
     \begin{minted}{python}
-@dataclass
-class FiberBundle:
-"""
-Attributes
-----------
-K: {'tables': []}
-F: {variable name: type}
-"""
-    K: dict 
-    F: dict
+        class ArtistClass(matplotlib.artist.Artist):
+            def __init__(self, data, transforms, *args, **kwargs):
+                # properties that are specific to the graphic
+                self.data = data 
+                self.transforms = transforms
+                super().__init__(*args, **kwargs)
+
+            def assemble(self, **args):
+                # set the properties of the graphic
+
+            def draw(self, renderer):
+                # returns K, indexed on fiber then key 
+                view = self.data.view(self.axes) 
+                # visual channel encoding applied fiberwise 
+                visual = {p: t['encoder'](view[t['name']])
+                          for p, t in self.transforms.items()}
+                self.assemble(**visual)
+                # pass configurations off to the renderer
+                super().draw(renderer)
         \end{minted}
@@
 
-\begin{frame}[fragile]{Discrete Connectivity}
+\begin{frame}[fragile]{Artists: Scatter \& Line}
+\begin{minted}{python}
+class Point(mcollections.Collection):
+    def assemble(self, x, y, s, facecolors='C0' ):
+        # construct geometries of the circle glyphs in visual coordinates
+        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
+                    for (xi, yi, si) in zip(x, y, s)] 
+        # set attributes of glyphs, these are vectorized 
+        # circles and facecolors are lists of the same size
+        self.set_facecolors(facecolors)
+
+class Line(mcollections.LineCollection):
+    def assemble(self, x, y, color='C0'):
+        #assemble line marks as set of segments 
+        segments = [np.vstack((vx, vy)).T for vx, vy in zip(x, y)]
+        self.set_segments(segments)
+        self.set_color(color)
+\end{minted}
+\end{frame}
+\subsection{Data}
+\begin{frame}[fragile]{Continuity}
 \begin{minted}{python}
@@
             {'v1': float, 'v2': str, 'v3': float})
-
     def __init__(self, sid = 45, size=1000, max_key=10**10):
@@
         return (k, (e1, e2, e3, e4))
-\end{minted}
-\end{frame}
 
-\begin{frame}[fragile]{1D Continuous Connectivity}
-\begin{minted}{python}
 class EdgeSimplex: 
     FB = FiberBundle({'tables': ['vertex','edge']},
-                         {'x' : float, 'y':  float, 
-                         'color':mtypes.Color()}})
-    def __init__(self, num_edges=4, num_samples=1000): 
-        self.keys = range(num_edge) #edge id
-        self.distances = np.linspace(0,1, num_samples)
-        # half generlized representation of arcs on a circle
-        self.angle_samples = np.linspace(0, 2*np.pi, len(self.keys)+1)
-    @staticmethod
-    def _color(edge):
-        return ['red','orange', 'green','blue'][edge%4]
-    @staticmethod
-    def _xy(edge, distances, start=0, end=2*np.pi):
-        # start and end are parameterizations b/c really there is 
-        angles = (distances *(end-start)) + start
-        return np.cos(angles), np.sin(angles)
+                     {'x' : float, 'y':  float, 'color':mtypes.Color()}})
     def tau(self, k): #will fix location on page on revision
@@
 \end{minted}
-\end{frame}
-
-\section{Conclusion}
-\begin{frame}{Summary}
-    \begin{itemize}
-        \item structure preserving maps from data to visual representation:
-            \begin{itemize}
-                \item data and graphics have equivalent continuity
-                \item properties are equivariant under monoid actions
-            \end{itemize}  
-        \item fiber bundles with a schema are structure rich abstractions of 
-            \begin{itemize}
-                \item topologically complex heterogenous data 
-                \item target display spaces
-            \end{itemize}
-        \item model can be iteratively integrated into existing Matplotlib architecture
-    \end{itemize}
 \end{frame}
@@
 \end{frame}
+\begin{frame}[fragile]{Fiber Bundle}
+    \begin{minted}{python}
+    @dataclass
+    class FiberBundle:
+    """
+    Attributes
+    ----------
+    K: {'tables': []}
+    F: {variable name: type}
+    """
+        K: dict 
+        F: dict
+    \end{minted}
+\end{frame}
 

### 2021-03-26 slides/main.tex
lets not delete my appendix

@@
 \usepackage{appendixnumberbeamer} 
+\usepackage{comment}
 \usepackage{minted}
@@
 \begin{frame}{Visualizations are structure preserving maps}
+    %add arrow between data and vizualizations
+    % we're focused on building good arrows 
     \begin{figure}
@@
     \end{figure}
+    \pause 
     \begin{description}
         \item[equivariance] properties of data and visual encoding match
+    \pause 
         \item[continuity] connectivity of data and visual encoding match
-        \item[composibility] structure preserved by individual components is preserved in combined components
     \end{description}
@@
 
-\begin{frame}{Tools are tuned to continuity \cite{HeerSoftware2006}}
+\begin{frame}{Domain specific libraries assume data structure\cite{HeerSoftware2006}}
     \begin{table}
@@
 
-\begin{frame}{Visualizations are tuned to continuinity\cite{toryRethinkingVisualizationHighlevel2004}}
+\begin{frame}{General purpose libraries can't\cite{toryRethinkingVisualizationHighlevel2004}}
     \begin{figure}
@@
 
-\begin{frame}{Tools are evaluated on the structure they preserve}
+\begin{frame}{Best practices in visualization design}
     \begin{description}
-        \item[Naturalness] easier to understand when properties match \cite{norman_things_smart}
-        \item[Graphical Integrity] graphs show \textbf{only} the data \cite{tufteVisualDisplayQuantitative2001}
-        \item[Expressiveness] structure preserving mappings from data to graphic \cite{mackinlayAutomatingDesignGraphical1986}
-        \item[Effectiveness] design choices made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}
+        \item[Expressiveness] structure preserving mappings\cite{mackinlayAutomatingDesignGraphical1986}
+        \pause
+        \item[Naturalness] easier to understand when properties match\cite{norman_things_smart}
+        \pause
+       \item[Graphical Integrity] graphs show \textbf{only} the data\cite{tufteVisualDisplayQuantitative2001}
     \end{description}
@@
     \begin{description}
-        \item[Topological] topology preserving relationship between data and graphic via continuous maps
-        \item[Equivariant] property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
-        \item[Artist] functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
-        \item [Model] prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
+        \item[Topological] continuity  
+        \item[Equivariant] monoid action 
+        \item[Artist] Matplotlib $\textit{artist}: \textit{data} \rightarrow \textit{graphic}$ 
+        \item [Model] 
     \end{description}
@@
     An artist $\mathscr{\vartist}$ is an equivariant map 
-    \begin{equation}\Huge
+    \begin{equation*}\Huge
         \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
-    \end{equation}
+    \end{equation*}
     from data $\mathscr{\dtotal}$ space to graphic $\mathscr{\gtotal}$ space. 
-
 \end{frame}
@@
 \subsection{Data Model}
-
 \begin{frame}{Model data as a fiber bundle \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}}
     A fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the map $\pi$
-    \begin{equation}
+    \begin{equation*}
         \label{eq:fiber_bundle}
@@
         \end{tikzcd}
-    \end{equation}
+    \end{equation*}
     \begin{description}
+        \pause
+        \item[total space \dtotal] dataset topology
+        \pause
+        \item[fiber space \dfiber] variable schema
+        \pause
         \item[base space \dbase] continuity
-        \item[fiber space \dfiber] variable characteristics
-        \item[total space \dtotal] dataset topology
     \end{description}
@@
         \centering
-        \includegraphics[height=.3\textwidth]{figures/math/fiber.png}
+        \includegraphics[width=\textwidth]{figures/math/fiber.png}
     \label{fig:data_fiber_example}
     \end{figure}
-    \begin{block}{Monoid actions define the structure on $\dfiber_i$}
-        \begin{equation}
-            \dfiber = \dfiber_0 \times \ldots \times \dfiber_i \times \ldots \times \dfiber_n
-        \end{equation}
-
-        $\monoid_{i}$ actions (e.g. rotation, partial ordering) 
-        \begin{equation}
-            \bullet: \monoid_i\times \dfiber_i \rightarrow \dfiber_i
-        \end{equation}
-        must be associative and have an identity element.
-
-    \end{block}
 \end{frame}
 
-\begin{frame}{Structural \textit{keys} with associated \textit{values} \cite{munznerVisualizationAnalysisDesign2014}}
+\begin{frame}{Structure of components of \dfiber}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=.3\textheight]{figures/math/fiber.png}
+    \label{fig:data_fiber_example}
+    \end{figure}
+        \begin{equation*}
+            \dfiber = \dfiber_0 \times \ldots \times \dfiber_i \times \ldots \times \dfiber_n
+        \end{equation*}
+        \pause
+        Monoid actions $\monoid_{i}$ (e.g. rotation, partial ordering) define the structure on $\dfiber_i$
+        \begin{equation*}
+            \bullet: \monoid_i\times \dfiber_i \rightarrow \dfiber_i
+        \end{equation*}
+        They must be associative and have an identity element. 
+\end{frame}
+
+\begin{frame}{\dbase\ is an indexing space}
     \begin{figure}[H]
@@
     \end{figure}
+\end{frame}
 
-    \begin{block}{\dbase\ is an indexing space independent of component semantics}
-        \begin{equation}
-            \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
-        \end{equation}
-    \end{block}
+\begin{frame}{Structural \textit{keys} with associated \textit{values} \cite{munznerVisualizationAnalysisDesign2014}}
+\begin{figure}
+    \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
+    \caption{Figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
+\end{figure}
+\pause
+\begin{figure}[H]
+    \includegraphics[width=.5\textheight]{figures/math/base.png}
+    \label{fig:base_example}
+\end{figure}
 \end{frame}
@@
     For any fiber bundle, there exists a map
-    \begin{equation}
+    \begin{equation*}
         \begin{tikzcd}[ampersand replacement=\&]
@@
         \end{tikzcd}
-    \end{equation}
+    \end{equation*}
      s.t. $\pi(\dsection(\dbasepoint)) = \dbasepoint$. $\Gamma(\dtotal)$ is the set of all global sections.
@@
 \subsection{Graphics}
-\begin{frame}{Graphic Bundle $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$}
-    Continuity is preserved via the deformation retraction $\vindex: \gbase\rightarrow\dbase$
-    \begin{equation}
+\begin{frame}{Graphic Bundle $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$} % watch dr explaination
+    Continuity is preserved via the many \gbasepoint\ to one \dbasepoint\ map $\vindex: \gbase\rightarrow\dbase$
+    \begin{equation*} 
         \begin{tikzcd}[ampersand replacement=\&]
@@
             \end{tikzcd}
-        \end{equation}
+        \end{equation*}
+    \pause
     \gfiber\ is a proxy for the target display, for example $(x,\, y,\, r,\, g,\, b) \in \gfiber$
-    \begin{figure}[H]
+    \pause
+    \begin{figure}[H] %%%bring up one at a time
         \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
@@
 
-\subsection{Artist}
-\begin{frame}{Topological Equivariant Artist Model}
-The topological artist is a sheaf map
-\begin{equation}
-\vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
-\end{equation}
-that carries homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019} 
-\begin{equation}
-\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
-\end{equation} 
-and is implemented via an intermediate visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$
-\begin{equation}
+\subsection{Artist}%formalize the structure
+\begin{frame}{Visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$}
+\begin{equation*}
     \label{eq:artist_diagram}
@@
         \end{tikzcd}
-\end{equation}
+\end{equation*}
+\pause
+\begin{equation*}
+    \mathcal{A}:\mathcal{E}\rightarrow\mathcal{E}
+\end{equation*}
+\pause
+The topological artist is a sheaf map
+\begin{equation*}
+    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
+\end{equation*}
 \end{frame}
 
-\begin{frame}{Existing models don't have continuity, many evaluative}
-        \begin{description}
-            \item[algebraic process] data and viz transforms are symmetric \cite{kindlmannAlgebraicProcessVisualization2014}
+\begin{frame}{Visualization Assembly Function}
+    \begin{figure}
+        \includegraphics[width=\textwidth]{figures/math/path_of_q.png}
+    \end{figure}
+    \pause
     \begin{equation*}
-                \begin{tikzcd}[ampersand replacement=\&]
-                    \dtotal \arrow[d, "m"'] \arrow[r, "r_1"] \& \vtotal \arrow[r, "\nu"]  \& \gtotal \arrow[d, "\varphi(m)"] \\
-                    \dtotal \arrow[r, "r_2"']                     \& \vtotal \arrow[r, "\nu"'] \& \gtotal                  
-                \end{tikzcd}
+        \label{eq:nu_expanded}
+        \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
     \end{equation*}
-
-            \item[language] APT and GoG: syntax, semantics, and grammar\\ 
-            \cite{mackinlayAutomatingDesignGraphical1986,mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005}
-            \item[functional dependencies] relationship between components \cite{sugibuchiFramwork2009}
-            \item[category theory]  \textit{understanding} = \textit{read} $\circ$ \textit{render} \cite{vickersUnderstandingVisualizationFormal2013}
-
-        \end{description}
+    \pause
+    \begin{equation*}
+        Q = \nu \circ \tau
+    \end{equation*}
 \end{frame}
 
-\begin{frame}{Visual Channel Encoders}
-We define the visual transformers \vchannel\ on components of the data bundle $\dsection_{i}$
-\begin{equation}
-    \label{eq:nu_expanded}
-    \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
-\end{equation}
-as the set of equivariant maps with the constraint 
-\begin{equation}
-    \vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
-\end{equation} 
-where $\varphi:\monoid\rightarrow \monoid^{\prime}$ carries a homomorphism of monoid actions. 
-\end{frame}
-
-\begin{frame}{Equivariance: Partial Order \& and Stevens' Scales \cite{stevensTheoryScalesMeasurement1946}}
-    \begin{figure}
-        \includegraphics[width=1\textwidth]{figures/math/monoid_equivariant.png}
-    \end{figure}
-
+\begin{frame}{Stevens' Scales \cite{stevensTheoryScalesMeasurement1946}}
 \begin{table}[H]
@@
 
-\begin{frame}{Visualization Assembly Function}
-    \begin{figure}
-        \includegraphics[width=\textwidth]{figures/math/path_of_q.png}
+\begin{frame}{Equivariance: Partial Order}
+    \begin{figure} %break out steps/reveal
+        \includegraphics[width=1\textwidth]{figures/math/monoid_equivariant.png}
     \end{figure}
@@
     \end{figure}
-    \begin{prop}
-        If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
-        \begin{equation}
-        \vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
-        \label{eq:math:artist:q}
-        \end{equation}
-        then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
-If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection.
-    \end{prop}
+
+    %put equation w/ rho, label rho, rho'
 \end{frame}
@@
     \end{figure}
-        \begin{equation*}
-            \lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2},\; 
-            \hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}
-        \end{equation*}
-    \begin{align*}
-     x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
-     y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
-    \end{align*}
 \end{frame}
@@
     \end{figure}
-    \begin{align*}
-        R &= R(\vindex(\alpha, \beta))\\
-        G &= G(\vindex(\alpha, \beta))\\
-        B &= B(\vindex(\alpha, \beta))
-    \end{align*}
 \end{frame}
 
-\begin{frame}{Assembly Function Factory}
-\begin{equation}\label{eq:qhat_q_s}
-    \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
-\end{equation} 
+\begin{frame}{Build \vmark over \dbase:\vmarkd}
 \begin{figure}[H]
@@
 \end{figure}
+\begin{equation*}\label{eq:qhat_q_s}
+    \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
+\end{equation*} 
 such \gbasepoint\ can be factored out when $\vpreimg = \gbasepoint$
@@
 \begin{frame}{Composition of artists} 
+    % make disjointed figure or scatter + line
     Given the family of artists $(\dtotal_i: i\in I)$ on the same image
-    \begin{equation}
+    \begin{equation*}
         + \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
-    \end{equation}
+    \end{equation*}
     the + operator defines a simple composition of artists. 
-
-    When artists share a base space 
-    \begin{equation}
-        \dbase_2 \hookrightarrow \dbase_1
-    \end{equation}
-    a composition operator can be defined such that the the artists can be considered to be acting on different components of the same section. 
 \end{frame}
 
-\begin{frame}{Equivalance class of artists}
-    An approximation of the equivalence class of artists $\vartist^{\prime}$
-    \begin{equation}
-    \vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}
-    \end{equation}
-    roughly treats two artists as equivalent if they 
-    \begin{itemize}
-        \item act on the same visual bundle \vtotal
-        \item have the same assembly function \vmark 
-        \item have the same continuity map \vindex 
-    \end{itemize}
-\end{frame}
 \section{Prototype}
-
 \begin{frame}{TEAM driven rearchitecture of Matplotlib}
     \begin{itemize}
-        \item fiber bundle with schema is a structure rich abstraction
         \begin{itemize}
@@
         \end{itemize}
-        \item structure preserving maps from data to visual:
+        \pause
+        \item structure preserving maps from data to visual
             \begin{itemize}
@@
             \end{itemize}  
+        \pause
         \item composable into complex visualizations
@@
 
-\begin{frame}[fragile]{Artists: Scatter \& Line}
+\begin{frame}[fragile]{How do we make things with this model?}
     \begin{figure}[H]
@@
 \subsection{Visual Transforms}
-\begin{frame}[fragile]{Visual Transformations}
+\begin{frame}[fragile]{How do we implement \vchannel?}
     \begin{minted}{python}
@@
 
-    \begin{frame}[fragile]{Complex \vchannel}
-    \begin{minted}{python}
-    class Categorical:
-        def __init__(self, mapping):
-            # check that the conversion is to valid colors
-            assert(mcolors.is_color_like(color) for color in mapping.values())
-            self._mapping = mapping
-
-        def __call__(self, value):
-            # convert value to a color
-            return [mcolors.to_rgba(self._mapping[v]) for v in values]
-    \end{minted}
-    That we can test for action equivariance
-    \begin{minted}{python}
-    def test_nominal(values, encoder):
-        m1 = list(zip(values, encoder(values)))
-        random.shuffle(values)
-        m2 = list(zip(values, encoder(values)))
-        assert sorted(m1) == sorted(m2)
-    \end{minted}
-\end{frame}
-
 \subsection{Artist}
-\begin{frame}[fragile]{Artist}
+\begin{frame}[fragile]{Artist} %% rewrite w/ letters in talk
     \begin{minted}{python}
         class ArtistClass(matplotlib.artist.Artist):
-            def __init__(self, data, transforms, *args, **kwargs):
-                # properties that are specific to the graphic
-                self.data = data 
-                self.transforms = transforms
+            def __init__(self, E, V, *args, **kwargs):
+                # set properties that are specific to the artist
+                # stash the input E and V
                 super().__init__(*args, **kwargs)
 
-            def assemble(self, **args):
+            def qhat(self, **args):
                 # set the properties of the graphic
@@
             def draw(self, renderer):
-                # returns K, indexed on fiber then key 
-                view = self.data.view(self.axes) 
+                # returns tau, indexed on fiber then key 
+                tau = self.E.view(self.axes) 
                 # visual channel encoding applied fiberwise 
-                visual = {p: t['encoder'](view[t['name']])
-                          for p, t in self.transforms.items()}
-                self.assemble(**visual)
+                visual = {p_i: nu_i(tau_i)
+                          for p_i, nu_i, tau)i 
+                          in zip(self.V, tau)}
+                self.qhat(**visual)
                 # pass configurations off to the renderer
@@
 \begin{frame}[fragile]{Artists: Scatter \& Line}
+%%keep the signatures, comments, change names to math
 \begin{minted}{python}
@@
         def assemble(self, x, y, color='C0'):
-        #assemble line marks as set of segments 
+            # assemble line marks as set of segments 
             segments = [np.vstack((vx, vy)).T for vx, vy in zip(x, y)]
@@
 \end{frame}
+
 \subsection{Data}
@@
     \begin{minted}{python}
-class VertexSimplex: #maybe change name to something else
-    """Fiberbundle is consistent across all sections
-    """
+class PointData: 
+    # Fiberbundle is consistent across all sections
     FB = FiberBundle({'tables': ['vertex']},  
         {'v1': float, 'v2': str, 'v3': float})
-    def __init__(self, sid = 45, size=1000, max_key=10**10):
-        # create random list of keys
     def tau(self, k):
-        # e1 is sampled from F1, e2 from F2, etc...
-        return (k, (e1, e2, e3, e4))
+        return # tau evaluated at one point k
 
-class EdgeSimplex: 
-    FB = FiberBundle({'tables': ['vertex','edge']},
-                     {'x' : float, 'y':  float, 'color':mtypes.Color()}})
-    def tau(self, k): #will fix location on page on revision
-        x, y = self._xy(k, self.distances, 
-                        self.angle_samples[k], self.angle_samples[k+1]) 
-        color = self._color(k) 
-        return (k, (x, y, color))
+class LineData: 
+    FB = FiberBundle({'tables': ['edge']},
+                {'x' : float, 'y':  float, 'color':mtypes.Color()})
+    def tau(self, k): 
+        return # tau evaluated on interval k
     \end{minted}
@@
 
-\begin{frame}[fragile]{View}
-\begin{minted}{python}
-def view(self, axes):
-    table = defaultdict(list)
-    for k in self.keys:
-        table['index'].append(k)
-        for (name, value) in zip(self.FB.fiber.keys(), self.tau(k)[1]):
-            table[name].append(value)
-    return table
-\end{minted}
-\begin{description}
-    \item [\mintinline{python}{VertexSimplex}] (name, value), value is scaler
-    \item [\mintinline{python}{EdgeSimplex}]  (name, value), value is [x0, ..., xn]
-\end{description}
-\end{frame}
 \begin{frame}[fragile]{Same Artist, Different Data Configurations}
@@
     \end{figure}
-\begin{minted}{Python}
-simplex.GraphLine(FB, edge_table, vertex_table, 
-                      connect=True)
-simplex.GraphLine(FB, edge_table, vertex_table, 
-                      num_samples=2, connect=False)
+\begin{minted}{python}
+LineData(FB, edge_table, vertex_table, connect=True)
+LineData(FB, edge_table, vertex_table, num_samples=2, connect=False)
 \end{minted}
@@
 \begin{frame}{Proposed Dissertation}
+% math + prototypes, organize it
+% papers + contributions to matplotlib
 \begin{itemize}
@@
 \appendix 
-\begin{frame}{Fiber is all possible values a variable can be \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
-    Given a space of all possible values \ftotal\
-    \begin{equation}
-        \label{eq:data_types}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
-            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
-        \end{tikzcd}
-    \end{equation}
-    a fiber component is the restricted space $\ftotal_{\fsection(\fname)}$. 
-    \begin{equation}
-        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
-    \end{equation}
-    \begin{description}
-        \item[\ftypes] data types of the variables in the dataset 
-        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
-        \item[\fnames] variable names, $\fname \in \fnames$
-        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
-    \end{description}
-\end{frame}
-\begin{frame}{Monoid actions}
-    A monoid \monoid\ is a set with
-    \begin{description}
-        \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
-        \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
-    \end{description}
-    \begin{block}{left monoid action}
-    A set \dfiber\ with an action\cite{nlab:action} $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
-        \begin{align*}
-            \textbf{associativity}\;& \text{for all } f,g \in \monoid \text{ and } x\in \dfiber,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-            \textbf{identity}\;& \text{for all } x\in \dfiber, e\in \monoid,\,  e\bullet x = x 
-        \end{align*}
-    \end{block}
-\end{frame}
-\begin{frame}{Keeping track of sections with sheafs}
-    Restriction maps of a sheaf describe how local $\iota^*\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
-
-    \begin{equation}
-        \label{eq:sheaf}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             \& \dtotal \arrow[d, "\pi"']                  \\
-            U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] \& \dbase \arrow[u, "\dsection"', bend right]
-        \end{tikzcd}
-    \end{equation}
-    The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$ such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
-\end{frame}
-
-\begin{frame}{Rendering: Define a Pixel}
-    \begin{columns}
-        \column{0.5\textwidth}
-        Given a pixel
-        \begin{equation}
-        p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]
-        \end{equation}
-        the inverse map of the bounding box 
-        \begin{equation}
-        \gbase_{p} ={\gsection_{xy}}^{-1}(p)
-        \end{equation}
-        is a region $\gbase_p \subset \gbase$ such that 
-        \begin{align}
-            \scriptstyle r_p &= \scriptstyle \iint\limits_{S_p} \rho_r(s)ds^{2}\\
-            \scriptstyle g_p &= \scriptstyle \iint\limits_{S_p} \rho_g(s)ds^{2}\\
-            \scriptstyle  b_p &= \scriptstyle \iint\limits_{S_p} \rho_b(s)ds^{2}
-        \end{align}
-        yields the color of the pixel. 
-        \column{0.5\textwidth}
-        \begin{figure}[H]
-            \includegraphics[width=\textwidth]{figures/math/render.png}
-        \end{figure}
-    \end{columns}
-\end{frame}
-\begin{frame}{\vfiber\ Components}
-
-\begin{table}[H]
-    \renewcommand{\arraystretch}{2}
-    \begin{tabulary}{\textwidth}{|l|L|l|}\hline
-     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i}) \subset \vfiber_{i}}$  \\ \hline
-    position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
-    size                        & linewidth, markersize                                            & $\mathbb{R}^{+}$   \\ \hline
-    shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
-    color                       & color, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
-    \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
-                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
-    \end{tabulary}
-    \label{tab:mpl_visual_variable_fiber}
-\end{table}
-\end{frame}
-
-\begin{frame}{Glyph}
-    The glyph is the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$ where the path connected components $\dbasepath \subset \dbase$ are defined 
-    \begin{equation}
-    \dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
-    \end{equation}
-    such that the  path $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1] and $\gbase_{\dbasepathpoint}$ is the region
-    \begin{equation}
-        \begin{tikzcd}[ampersand replacement=\&]
-            \gtotal \arrow[r, shift left] \& \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] \& \& \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
-            \end{tikzcd}
-        \label{eq:mark}
-    \end{equation}
-    such that the glyph is differentiable, in keeping with Ziemkiewicz and Kosara's description of a glyph\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
-\end{frame}
-\begin{frame}[fragile]{Fiber Bundle}
-    \begin{minted}{python}
-    @dataclass
-    class FiberBundle:
-    """
-    Attributes
-    ----------
-    K: {'tables': []}
-    F: {variable name: type}
-    """
-        K: dict 
-        F: dict
-    \end{minted}
-\end{frame}
-
-\begin{frame}[fragile]{GraphLine Data Model}
-\begin{minted}{python}
-class GraphLine:
-    def __init__(self, FB, edge_table, vertex_table, num_samples=1000,
-                        connect=False):
-        #set args as attributes and generate distance
-        if connect: # test connectivity if edges are continuous
-            assert edge_table.keys() == self.FB.F.keys()
-            assert is_continuous(vertex_table)
-
-    def tau(self, k):
-        # evaluates functions defined in edge table
-        return(k, (self.edges[c][k](self.distances) 
-                        for c in self.FB.F.keys()))
-
-    def view(self, axes):
-        # walk the edge_vertex table to return the edge function
-        table = defaultdict(list)
-        for (i, (start, end)) in sorted(zip(self.ids, self.vertices), 
-                                            key=lambda v:v[1][0]):
-            table['index'].append(i)
-            # same as view for line, returns nested list
-            for (name, value) in zip(self.FB.F.keys(), self.tau(i)[1]):
-                table[name].append(value)
-        return table
-\end{minted}
-\end{frame}
+\subsection{Appendix}
 

### 2021-03-26 slides/main.tex
lets not delete my appendix

@@
 \end{frame}
-
 \section{Introduction}
@@
     \begin{itemize}
-        \begin{itemize}
-            \item topologically complex heterogenous data 
-            \item target display spaces
-        \end{itemize}
+        \item complex visualizations
         \pause 
@@
         \pause
-        \item composable into complex visualizations
+        \item fiber bundles are an abstraction
+        \begin{itemize}
+            \item topologically complex heterogenous data 
+            \item target display spaces
+        \end{itemize}
     \end{itemize}
@@
 \appendix
+
 \subsection{Appendix}
 
+\begin{frame}{Mathematical Models of Vizualization}
+    \begin{description}
+        \item[algebraic process] data and viz transforms are symmetric \cite{kindlmannAlgebraicProcessVisualization2014}
+        \begin{equation*}
+            \begin{tikzcd}[ampersand replacement=\&]
+                \dtotal \arrow[d, "m"'] \arrow[r, "\nu"] \& \vtotal \arrow[r, "Q"]  \& \gtotal \arrow[d, "\varphi(m)"] \\
+                \dtotal \arrow[r, "\nu"']                     \& \vtotal \arrow[r, "Q"'] \& \gtotal                  
+            \end{tikzcd}
+        \end{equation*}
+
+        \item[language] APT and GoG: syntax, semantics, and grammar\\ 
+        \cite{mackinlayAutomatingDesignGraphical1986,mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005}
+        \item[functional dependencies] relationship between components \cite{sugibuchiFramwork2009}
+        \item[category theory]  \textit{understanding} = \textit{read} $\circ$ \textit{render} \cite{vickersUnderstandingVisualizationFormal2013}    
+    \end{description}
+\end{frame}
+
+\begin{frame}{Fiber is all possible values a variable can be \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}}
+    Given a space of all possible values \ftotal\
+    \begin{equation*}
+        \label{eq:data_types}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] \& \ftotal \arrow[d, "\pi"] \\
+            \fnames \arrow[r, "\fsection"']                  \& \ftypes       
+        \end{tikzcd}
+    \end{equation*}
+    a fiber component is the restricted space $\ftotal_{\fsection(\fname)}$. 
+    \begin{equation*}
+        \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
+    \end{equation*}
+    \begin{description}
+        \item[\ftypes] data types of the variables in the dataset 
+        \item[\ftotal] disjoint union of all values of type $\ftype \in \ftypes$ 
+        \item[\fnames] variable names, $\fname \in \fnames$
+        \item[\fttype] \ftotal\ restricted to the data type of a named variable   
+    \end{description}
+\end{frame}
+
+\begin{frame}{Monoid actions}
+    A monoid \monoid\ is a set with
+    \begin{description}
+        \item[associative binary operator] $\ast:\monoid \times \monoid\rightarrow \monoid$
+        \item[identity element] $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. 
+    \end{description}
+    \begin{block}{left monoid action}
+    A set \dfiber\ with an action\cite{nlab:action} $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
+        \begin{align*}
+            \textbf{associativity}\;& \text{for all } f,g \in \monoid \text{ and } x\in \dfiber,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
+            \textbf{identity}\;& \text{for all } x\in \dfiber, e\in \monoid,\,  e\bullet x = x 
+        \end{align*}
+    \end{block}
+\end{frame}
+
+\begin{frame}{Keeping track of sections with sheafs}
+    Restriction maps of a sheaf describe how local $\iota^*\dsection$ can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}
+    \begin{equation*}
+        \label{eq:sheaf}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             \& \dtotal \arrow[d, "\pi"']                  \\
+            U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] \& \dbase \arrow[u, "\dsection"', bend right]
+        \end{tikzcd}
+    \end{equation*}
+    The inclusion map $\iota: U \rightarrow \dbase$ pulls \dtotal\ over $U$ such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
+\end{frame}
+
+\begin{frame}{Rendering: Define a Pixel}
+    \begin{columns}
+        \column{0.5\textwidth}
+        Given a pixel
+        \begin{equation*}
+        p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]
+        \end{equation*}
+        the inverse map of the bounding box 
+        \begin{equation*}
+        \gbase_{p} ={\gsection_{xy}}^{-1}(p)
+        \end{equation*}
+        is a region $\gbase_p \subset \gbase$ such that 
+        \begin{align}
+            \scriptstyle r_p &= \scriptstyle \iint\limits_{S_p} \rho_r(s)ds^{2}\\
+            \scriptstyle g_p &= \scriptstyle \iint\limits_{S_p} \rho_g(s)ds^{2}\\
+            \scriptstyle  b_p &= \scriptstyle \iint\limits_{S_p} \rho_b(s)ds^{2}
+        \end{align}
+        yields the color of the pixel. 
+        \column{0.5\textwidth}
+        \begin{figure}[H]
+            \includegraphics[width=\textwidth]{figures/math/render.png}
+        \end{figure}
+    \end{columns}
+\end{frame}
+
+\begin{frame}{$\mathcal{A}:\mathcal{E}\rightarrow\mathcal{E}$}
+    The topological artist is a sheaf map
+    \begin{equation*}
+    \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
+    \end{equation*}
+    that carries homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ 
+    \cite{cegarraCohomologyMonoidsOperators2019} 
+    \begin{equation*}
+    \vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
+    \end{equation*}
+\end{frame}
+
+\begin{frame}{Visual Channel Encoders}
+    We define the visual transformers \vchannel\ on components of the data bundle $\dsection_{i}$
+    \begin{equation*}
+        \label{eq:nu_expanded}
+        \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
+    \end{equation*}
+    as the set of equivariant maps with the constraint 
+    \begin{equation*}
+        \vchannel_i(m_{\delement}(\dtotal_i)) = \varphi(m_{\delement})(\vchannel_i(\dtotal_i))
+    \end{equation*} 
+    where $\varphi:\monoid\rightarrow \monoid^{\prime}$ carries a homomorphism of monoid actions. 
+\end{frame}
+
+\begin{frame}{\vfiber\ Components}    
+\begin{table}[H]
+    \renewcommand{\arraystretch}{2}
+    \begin{tabulary}{\textwidth}{|l|L|l|}\hline
+     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i}) \subset \vfiber_{i}}$  \\ \hline
+    position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
+    size                        & linewidth, markersize                                            & $\mathbb{R}^{+}$   \\ \hline
+    shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
+    color                       & color, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
+    \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
+                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
+    \end{tabulary}
+    \label{tab:mpl_visual_variable_fiber}
+\end{table}
+\end{frame}
+
+\begin{frame}{Glyph}
+    The glyph is the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$ where the path connected components $\dbasepath \subset \dbase$ are defined 
+    \begin{equation*}
+    \dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
+    \end{equation*}
+    such that the  path $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1] and $\gbase_{\dbasepathpoint}$ is the region
+    \begin{equation*}
+        \begin{tikzcd}[ampersand replacement=\&]
+            \gtotal \arrow[r, shift left] \& \gbase_\dbasepathpoint \arrow[rr, "\vindex(\gbasepoint)", shift left] \arrow[l, "\gsection(\gbase_\dbasepathpoint)"] \& \& \dbasepath_{\dbasepoint} \arrow[ll, "\vindex^{-1}(\dbasepath)"]
+            \end{tikzcd}
+        \label{eq:mark}
+    \end{equation*}
+    such that the glyph is differentiable, in keeping with Ziemkiewicz and Kosara's description of a glyph\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+\end{frame}
+
+\begin{frame}{Artist Equivalance class}
+    When artists share a base space 
+    \begin{equation*}
+        \dbase_2 \hookrightarrow \dbase_1
+    \end{equation*}
+    a composition operator can be defined such that the the artists can be considered to be acting on different components of the same section. 
+\end{frame}
+
+\begin{frame}[fragile]{Complex \vchannel}
+    \begin{minted}{python}
+    class Categorical:
+        def __init__(self, mapping):
+            # check that the conversion is to valid colors
+            assert(mcolors.is_color_like(color) for color in mapping.values())
+            self._mapping = mapping
+
+        def __call__(self, value):
+            # convert value to a color
+            return [mcolors.to_rgba(self._mapping[v]) for v in values]
+    \end{minted}
+    That we can test for action equivariance
+    \begin{minted}{python}
+    def test_nominal(values, encoder):
+        m1 = list(zip(values, encoder(values)))
+        random.shuffle(values)
+        m2 = list(zip(values, encoder(values)))
+        assert sorted(m1) == sorted(m2)
+    \end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{Artist}
+    \begin{minted}{python}
+        class ArtistClass(matplotlib.artist.Artist):
+            def __init__(self, data, transforms, *args, **kwargs):
+                # properties that are specific to the graphic
+                self.data = data 
+                self.transforms = transforms
+                super().__init__(*args, **kwargs)
+
+            def assemble(self, **args):
+                # set the properties of the graphic
+
+            def draw(self, renderer):
+                # returns K, indexed on fiber then key 
+                view = self.data.view(self.axes) 
+                # visual channel encoding applied fiberwise 
+                visual = {p: t['encoder'](view[t['name']])
+                          for p, t in self.transforms.items()}
+                self.assemble(**visual)
+                # pass configurations off to the renderer
+                super().draw(renderer)
+        \end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{Artists: Scatter \& Line}
+\begin{minted}{python}
+class Point(mcollections.Collection):
+    def assemble(self, x, y, s, facecolors='C0' ):
+        # construct geometries of the circle glyphs in visual coordinates
+        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
+                    for (xi, yi, si) in zip(x, y, s)] 
+        # set attributes of glyphs, these are vectorized 
+        # circles and facecolors are lists of the same size
+        self.set_facecolors(facecolors)
+
+class Line(mcollections.LineCollection):
+    def assemble(self, x, y, color='C0'):
+        #assemble line marks as set of segments 
+        segments = [np.vstack((vx, vy)).T for vx, vy in zip(x, y)]
+        self.set_segments(segments)
+        self.set_color(color)
+\end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{View}
+\begin{minted}{python}
+def view(self, axes):
+    table = defaultdict(list)
+    for k in self.keys:
+        table['index'].append(k)
+        for (name, value) in zip(self.FB.fiber.keys(), self.tau(k)[1]):
+            table[name].append(value)
+    return table
+\end{minted}
+\begin{description}
+    \item [\mintinline{python}{VertexSimplex}] (name, value), value is scaler
+    \item [\mintinline{python}{EdgeSimplex}]  (name, value), value is [x0, ..., xn]
+\end{description}
+\end{frame}
+
+\begin{frame}[fragile]{Fiber Bundle}
+    \begin{minted}{python}
+    @dataclass
+    class FiberBundle:
+    """
+    Attributes
+    ----------
+    K: {'tables': []}
+    F: {variable name: type}
+    """
+        K: dict 
+        F: dict
+    \end{minted}
+\end{frame}
+
+\begin{frame}[fragile]{GraphLine Data Model}
+\begin{minted}{python}
+class GraphLine:
+    def __init__(self, FB, edge_table, vertex_table, num_samples=1000,
+                        connect=False):
+        #set args as attributes and generate distance
+        if connect: # test connectivity if edges are continuous
+            assert edge_table.keys() == self.FB.F.keys()
+            assert is_continuous(vertex_table)
+
+    def tau(self, k):
+        # evaluates functions defined in edge table
+        return(k, (self.edges[c][k](self.distances) 
+                        for c in self.FB.F.keys()))
+
+    def view(self, axes):
+        # walk the edge_vertex table to return the edge function
+        table = defaultdict(list)
+        for (i, (start, end)) in sorted(zip(self.ids, self.vertices), 
+                                            key=lambda v:v[1][0]):
+            table['index'].append(i)
+            # same as view for line, returns nested list
+            for (name, value) in zip(self.FB.F.keys(), self.tau(i)[1]):
+                table[name].append(value)
+        return table
+\end{minted}
+\end{frame}
+
 \end{document}

### 2021-03-26 slides/main.tex
everything but animated slides

@@
         \pause
+        \item[Graphical Integrity] graphs show \textbf{only} the data\cite{tufteVisualDisplayQuantitative2001}
+        \pause 
         \item[Naturalness] easier to understand when properties match\cite{norman_things_smart}
-        \pause
-       \item[Graphical Integrity] graphs show \textbf{only} the data\cite{tufteVisualDisplayQuantitative2001}
     \end{description}
@@
     An artist $\mathscr{\vartist}$ is an equivariant map 
-    \begin{equation*}\Huge
+    \begin{equation*}\huge
         \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
@@
         \pause
-        \item[total space \dtotal] dataset topology
+        \item[total space \dtotal] topology
         \pause
-        \item[fiber space \dfiber] variable schema
+        \item[fiber space \dfiber] schema
         \pause
@@
         \end{equation*}
-        They must be associative and have an identity element. 
+        where $\bullet$ is associative and has an identity action. 
 \end{frame}
@@
 
-\subsection{Artist}%formalize the structure
+\subsection{Artist}
 \begin{frame}{Visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$}
+    \begin{equation*}
+        \mathcal{A}:\mathcal{E}\rightarrow\mathcal{H}
+    \end{equation*}
+\pause
 \begin{equation*}
@@
 \pause
-\begin{equation*}
-    \mathcal{A}:\mathcal{E}\rightarrow\mathcal{E}
-\end{equation*}
-\pause
-The topological artist is a sheaf map
 \begin{equation*}
@@
 
-\begin{frame}{Stevens' Scales \cite{stevensTheoryScalesMeasurement1946}}
+\begin{frame}{Group Equivariance: Stevens' Scales \cite{stevensTheoryScalesMeasurement1946}}
 \begin{table}[H]
@@
 
-\begin{frame}{Equivariance: Partial Order}
+\begin{frame}{Monoid Equivariance: Partial Orders}
     \begin{figure} %break out steps/reveal
@@
     \end{figure}
+    \begin{equation*}
+        \vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
+    \end{equation*}
 
-    %put equation w/ rho, label rho, rho'
 \end{frame}
@@
     \end{figure}   
-
-    \begin{align*}
-        x &= size *\alpha \cos(\beta) + xpos \\
-        y &= size *\alpha \sin(\beta) + ypos
-    \end{align*}    
 \end{frame}
@@
 
-\begin{frame}{Build \vmark over \dbase:\vmarkd}
+\begin{frame}{Build \vmark\ over \dbase:\vmarkd}
 \begin{figure}[H]
@@
 \end{equation*} 
-such \gbasepoint\ can be factored out when $\vpreimg = \gbasepoint$
 \end{frame}
 
-\begin{frame}{Composition of artists} 
-    % make disjointed figure or scatter + line
-    Given the family of artists $(\dtotal_i: i\in I)$ on the same image
-    \begin{equation*}
-        + \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
-    \end{equation*}
-    the + operator defines a simple composition of artists. 
+\begin{frame}{Composition of artists $+ \coloneqq \sqcup \dtotal_{i}$} 
+    \begin{figure}
+        \includegraphics[width=1\textwidth]{figures/math/exploding_artist.png}
+    \end{figure}
 \end{frame}
@@
 
-\begin{frame}[fragile]{Artists: Scatter \& Line}
-%%keep the signatures, comments, change names to math
+\begin{frame}[fragile]{Artists: Scatter \& Line \vmarkd}
 \begin{minted}{python}
@@
             # construct geometries of the circle glyphs in visual coordinates
-            self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
-                        for (xi, yi, si) in zip(x, y, s)] 
-            # set attributes of glyphs, these are vectorized 
-            # circles and facecolors are lists of the same size
-            self.set_facecolors(facecolors)
+            # set attributes of glyphs
 
@@
             # assemble line marks as set of segments 
-            segments = [np.vstack((vx, vy)).T for vx, vy in zip(x, y)]
-            self.set_segments(segments)
-            self.set_color(color)
     \end{minted}
@@
 
-\begin{frame}{Proposed Dissertation}
-% math + prototypes, organize it
-% papers + contributions to matplotlib
-\begin{itemize}
-    \item expansion of the mathematical framework to include worked out simple and complex addition
-    \item formalization of definition of equivalance class \vartisteq
-    \item implementation of artist with explicit \vindex\
-    \item specification of interactive visualization
-    \item mathematical formulation of a graphic with axes labeling
-    \item implementation of new prototype artists that do not inherit from Matplotlib artists
-    \item provisional mathematics and implementation of user level composite artists
-    \item proof of concept domain specific user facing library 
-\end{itemize}
+\begin{frame}{Proposed Work}
+\begin{figure}
+    \includegraphics[width=1\textwidth]{figures/intro/gantt.png}
+\end{figure}
 \end{frame}

### 2021-03-26 slides/main.tex
new figure code + new figures + animated pdf

@@
-\documentclass[xcolor={dvipsnames}, handout]{beamer}
-%\documentclass[xcolor={dvipsnames}]{beamer}
+%\documentclass[xcolor={dvipsnames}, handout]{beamer}
+\documentclass[xcolor={dvipsnames}]{beamer}
 %\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
@@
 \usepackage{multirow}
+\usepackage{array}
 
@@
 \begin{frame}{Visualizations are structure preserving maps}
-    %add arrow between data and vizualizations
-    % we're focused on building good arrows 
     \begin{figure}
-        \includegraphics[width=\textwidth]{figures/intro/viz_same.png}
+        \begin{overprint}
+            \onslide<1|handout:0>\includegraphics[width=1\linewidth]{figures/intro/table_only.png}
+            \onslide<2|handout:0>\includegraphics[width=1\linewidth]{figures/intro/map.png}
+            \onslide<3|handout:0>\includegraphics[width=1\linewidth]{figures/intro/time.png}
+            \onslide<4->\includegraphics[width=1\linewidth]{figures/intro/artist.png}
+        \end{overprint}
     \end{figure}
@@
     \begin{description}
-        \item[equivariance] properties of data and visual encoding match
-    \pause 
-        \item[continuity] connectivity of data and visual encoding match
+        \uncover<5->{\item[equivariance] properties of data and visual encoding match}
+        \uncover<6>{\item[continuity] connectivity of data and visual encoding match}
     \end{description}
@@
         %\renewcommand{\arraystretch}{2}
-        \begin{tabulary}{\textwidth}{|L|l|l|}\toprule
+        %
+        \begin{tabular}{>{\onslide<1->}l>{\onslide<2->}l>{\onslide<3->}l}
             \includegraphics[width=.24\textwidth]{figures/intro/table.png} & \includegraphics[width=.3\textwidth]{figures/intro/landsat.png} & \includegraphics[width=.33\textwidth]{figures/math/graph.png} \\
@@
             Altair\cite{vanderplasAltairInteractiveStatistical2018}& Napari\cite{nicholas_sofroniew_2021_4533308} & Networkx\cite{HagbergExploringNetwork2008}\\
-             Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} & &\\
-             \bottomrule
-        \end{tabulary}
+             Tableau \cite{StoltePolaris2002}& &\\
+            \cite{hanrahanVizQL2006,MackinlayShowme2007}&&\\
+
+        \end{tabular}
     \end{table}
@@
     An artist $\mathscr{\vartist}$ is an equivariant map 
-    \begin{equation*}\huge
+    \begingroup 
+    \Huge
+    \begin{equation*}
         \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
     \end{equation*}
+    \endgroup
     from data $\mathscr{\dtotal}$ space to graphic $\mathscr{\gtotal}$ space. 
@@
 
-\begin{frame}{Structure of components of \dfiber}
+\begin{frame}{Monoids are the structure of the components of \dfiber}
     \begin{figure}[H]
@@
 
-\begin{frame}{Structural \textit{keys} with associated \textit{values} \cite{munznerVisualizationAnalysisDesign2014}}
+\begin{frame}{\dbase\ is the space of \textit{keys} into data in \dtotal \cite{munznerVisualizationAnalysisDesign2014}}
 \begin{figure}
@@
 \end{figure}
-\pause
 \begin{figure}[H]
@@
 \begin{frame}{Graphic Bundle $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$} % watch dr explaination
-    Continuity is preserved via the many \gbasepoint\ to one \dbasepoint\ map $\vindex: \gbase\rightarrow\dbase$
+    \only<2->{Continuity is preserved via the many \gbasepoint\ to one \dbasepoint\ map $\vindex: \gbase\rightarrow\dbase$}
+
+    \only<1>{
+        \begin{equation*} 
+            \begin{tikzcd}[ampersand replacement=\&]
+                F \arrow[r, hook] \& E \arrow[d, "\pi"']              \& D \arrow[r, hook] \& H \arrow[d, "\pi"']              \\
+                                  \& K \arrow[u, "\tau"', bend right] \&                   \& S \arrow[u, "\rho"', bend right]
+                \end{tikzcd}
+            \end{equation*}}
+    \only<2->{
     \begin{equation*} 
@@
             \end{tikzcd}
-        \end{equation*}
-    \pause
-    \gfiber\ is a proxy for the target display, for example $(x,\, y,\, r,\, g,\, b) \in \gfiber$
+        \end{equation*}}
+    \only<3->{   
+    \gfiber\ is a proxy for the target display, for example $(x,\, y,\, r,\, g,\, b) \in \gfiber$}
     \pause
     \begin{figure}[H] %%%bring up one at a time
-        \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
-        \label{fig:graphic_retraction_map}
+        \begin{overprint}
+            \onslide<4|handout:0>\includegraphics[width=1\linewidth]{figures/math/scatter_s.png}
+            \onslide<5|handout:0>\includegraphics[width=1\linewidth]{figures/math/line_s.png}
+            \onslide<6>\includegraphics[width=1\linewidth]{figures/math/image_s.png}
+        \end{overprint}
     \end{figure}
@@
 \begin{frame}{Monoid Equivariance: Partial Orders}
-    \begin{figure} %break out steps/reveal
-        \includegraphics[width=1\textwidth]{figures/math/monoid_equivariant.png}
+    \begin{figure} 
+        \begin{overprint}
+            \onslide<1|handout:0>\includegraphics[width=1\linewidth]{figures/math/partial.png}
+            \onslide<2|handout:0>\includegraphics[width=1\linewidth]{figures/math/partial_invalid.png}
+            \onslide<3>\includegraphics[width=1\linewidth]{figures/math/partial_fixed.png}
+        \end{overprint}
     \end{figure}
@@
     \begin{figure}[H]
-        \includegraphics[width=1\textwidth]{figures/math/scatter.png}
+        \begin{overprint}
+            \onslide<1|handout:0>\includegraphics[width=1\linewidth]{figures/math/scatter_q.png}
+            \onslide<2|handout:0>\includegraphics[width=1\linewidth]{figures/math/scatter_with_s.png}
+        \end{overprint}
     \end{figure}   
@@
     \begin{figure}[H]
-        \includegraphics[width=1\textwidth]{figures/math/line.png}
+        \begin{overprint}
+            \onslide<1|handout:0>\includegraphics[width=1\linewidth]{figures/math/line_q.png}
+            \onslide<2|handout:0>\includegraphics[width=1\linewidth]{figures/math/line_with_s.png}
+        \end{overprint}
     \end{figure}   
@@
 
-\begin{frame}{Build \vmark\ over \dbase:\vmarkd}
+\begin{frame}{Build \vmark\ over \dbase: \vmarkd}
 \begin{figure}[H]
@@
 
-\begin{frame}[fragile]{How do we make things with this model?}
+\begin{frame}[fragile]{How do we make things?}
     \begin{figure}[H]
@@
 \subsection{Visual Transforms}
-\begin{frame}[fragile]{How do we implement \vchannel?}
+\begin{frame}[fragile]{ \vchannel}
     \begin{minted}{python}
@@
 \subsection{Artist}
-\begin{frame}[fragile]{Artist} %% rewrite w/ letters in talk
+\begin{frame}[fragile]{\vartist} %% rewrite w/ letters in talk
     \begin{minted}{python}
@@
 
-\begin{frame}[fragile]{Artists: Scatter \& Line \vmarkd}
+\begin{frame}[fragile]{\vmarkd}
 \begin{minted}{python}
@@
 
-\begin{frame}[fragile]{Same Artist, Different Data Configurations}
+\begin{frame}[fragile]{Same Artist, Different \dtotal}
     \begin{figure}[H]
@@
 \end{figure}
+\end{frame}
+
+\begin{frame}{Acknowledgments}
+    \begin{itemize}
+        \item Professor Michael Grossberg and Dr. Thomas Caswell
+        \item Professor Haralick, Professor Vo, Professor Manovich, Dr. Hanwell
+        \item Matplotlib development team 
+        \item CZI EOSS (grant number 2019-207333) from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation
+    \end{itemize}
+
 \end{frame}
@@
 \end{table}
+\end{frame}
+\begin{frame}{Monoid Equivariance: Partial Orders}
+    \begin{figure} %break out steps/reveal
+        \includegraphics[width=1\textwidth]{figures/math/monoid_equivariant.png}
+    \end{figure}
 \end{frame}

### 2021-03-31 paper/paper.tex
still need to flesh out intro

@@
 \maketitle
-The underlying structure and semantics of visualization are by definition reflected in visual representations \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity to external concepts \cite{byrneAcquiredCodesMeaning2016}.
-Visualization tools allow practioners to automate these transformations, but most visualization tools are either explicitly tuned to specific continuity \cite{HeerSoftware2006} or define every interface relative to its expected data structure.
-The former approach is often too restrictive for a general purpose tool, while the latter often results in a library that can feel inconsistent or disjointed.
+%% this is still a mess, come back to it
+Models for describing visualizations and visualization tools are deeply tied to to the domains in which the visualizations are made \cite{friendlyBriefHistoryData2006a,pousmanCasualInformation2007}, and these tools are often made from building block libraries \cite{wongsuphasawatNavigatingWideWorld2021} such as the Python visualization library Matplotlib.
+Because building block libraries often support so many domains, the API can grow organically in ways that are incoherant and brittle.
+To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software needs to satisfy so that a visualization, however it is specified in the library built out of these blocks, is guaranteed to be a faithful representation of the data.
+Our model, which we call the  topological equivariant artist model (TEAM), provides a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables.
+
+\begin{figure}
+  \includegraphics[width=\columnwidth]{figures/continuity.png}
+  \caption{Continuity describes how the records in the dataset, in this case the rows in the table, are connected to each other.
+The line plot is an invalid visual representation of the table because it implies that the records are connected to each other in some way.
+On the other hand the bar plot is valid because it preserves the records as independent data points.}
+  \label{fig:intro:continuity}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\columnwidth]{figures/equivariance.png}
+  \caption{Equivariance means that transformations on the data side and on the visual side are equal.
+In this example, the temperature is doubled via the arrows representing the doubling action, and the plot of the temperature is doubled via the arrows on the visual side.}
+  \label{fig:intro:equivariance}
+\end{figure}
+
 Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation.
@@
   \item topology preserving relationship between data and graphic via continuous maps
-  \item property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
+  \item property preservation from data component to visual representation as monoid action equivariant maps
   \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
@@
 \section{Related Work}
-The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, and the notion that the properties of the data and visual representation match is the basis of most evaluations of visualization.
-Expressiveness, as defined by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987, mackinlayAutomatingDesignGraphical1986} is a measure of how much of the structure any map can encode.
-A fully expressive component is one that is equivariant since it has preserved all the structure in the data.
-Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations.
-Mackinlay's \textit{A Presentation Tool} (APT) introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}.
-This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
-This same limitation is explicitly stated in the functional dependency model of visualization developed by Sugibuchi\cite{sugibuchiFramwork2009} and evident in Vickers' category theory oriented framework in which semiotics are commutative.
-The algebraic process model by Kindlmann and Scheideggar proposes that the data and visualization transformations are commutative; while similar to our framework, it is missing any explicit mention of continuity.
-
-Much of visualization algorithm design is oriented around continuity, as described in Tory and Möller's taxonomy\cite{toryRethinkingVisualizationHighlevel2004}.
-For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}.
-Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}.
-Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.
-Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.
-
-Fiber bundle and sheafs are generalized abstractions proposed by Butler to encode the continuity of the data separately from the properties of each variable  \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplicial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} to incorporate a schema like description of the variables.
-One way of describing the binding between the scheme and the continuinity is using the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}.
-Unlike Munzner's model where the semantic meaning of the key is tightly coupled to the index of the value, our model considers keys to be a reference to topology.
-This allows the metadata to be altered, for example by changing the coordinate system or time resolution, without imposing new semantics on the underlying structure.
+%
+% My best paraphrasing of the main idea would be that this paper takes the ideas of going from data to graphic along with some important constraints that people generally agree are important and formalizes them into a topological model, then shows how that model can map into a concrete implementation.
+The important part of this model is that it models the data sufficiently abstractly that it's able to work with basically any form or type of data.
+The sentence starting at line 489 seems to be the big "aha" to me, basically saying something like "if you implement things such that the code equivalents of these terms are separated in some specific way, then you get these important formally-describable properties".
+Is that right?
+If so, I'm left very curious about what alternative (possibly already-existing?) Artist implementations there might be that seem to work, but don't decompose things in this way, and therefore don't have these important properties.
+Or maybe the existing implementations do have these properties but aren't general enough?
+I know there are space constraints etc but if there was some way of showing how the current Matplotlib artists work and how they fall short of this model, that would be very enlightening I think.
 
@@
 \end{equation}
-from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$.
-To formalize the structure the artist is preserving, we describe how we model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as fiber bundles.
-We then discuss the equivariant maps that make up $\mathscr{\vartist}$, which are a map from  graphic continuity to data continuity (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}) and visual components to graphic (\autoref{sec:math:artist:q}).
+from data $\mathscr{\dtotal}$ (\autoref{sec:math:data}) to graphic $\mathscr{\gtotal}$ (\autoref{sec:math:graphic}) fiber bundles.
+We decompose the artist into an indexing map from graphic to data (\autoref{sec:math:graphic:base}), a map from data components to visual components (\autoref{sec:math:artist:nu}), and map from visual components to graphic (\autoref{sec:math:artist:q}).
+In this section we discuss the formal properties of these fiber bundles and maps such that they can be used to specify an implementation, for example the one prototyped in \autoref{sec:code}.
 
@@
 \label{sec:math:data:fiber}
-To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names.
+To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010} that binds the components of the fiber to variable names.
 Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
-He then defines the single variable set \fttype\
-\begin{equation}
-    \label{eq:data_types}
-    \begin{tikzcd}
-        \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] & \ftotal \arrow[d, "\pi"] \\
-        \fnames \arrow[r, "\fsection"']                          & \ftypes       
-    \end{tikzcd}
-\end{equation}
-which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname.
+He then defines the single variable set \fttype, which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname.
 The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype.
@@
   \caption{These two datasets have the same base space \dbase.
-The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature),  while the cube is the fiber $\reals\times\realsp\times\reals$ associated with (time, wind=(speed, direction))}
+The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature) from \autoref{fig:intro:equivariance}, while the cube is the fiber $\reals\times\realsp\times\reals$ associated with (time, wind=(speed, direction))}
   \label{fig:math:data:fiber}
 \end{figure}
+
 In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\realsp\times\reals)\)) which encodes (\emph{speed}, \emph{direction}).
@@
 A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$.
-As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
-\begin{align*}
-    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
-\end{align*}
+As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties of associativity and identity.
 As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
@@
 Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmannAlgebraicProcessVisualization2014} that must be preserved in the visual representation of the component.
-A secondary advantage of defining structure in terms of monoids is that they are commonly found in functional programming because they specify compositions of transformations \cite{yorgeyMonoidsThemeVariations, stievenMonadJustMonoid2020}.
+Monoids are commonly found in functional programming because a property of monoids is that the components can be composed into complex transformation pipelines \cite{yorgeyMonoidsThemeVariations}.
+For example, a component that converts \(\dfiber_i=time\) with \(\monoid_i\) to hours can be combined with a component that shifts \(time\) by 2 hours to yield \(hours_2\) with the same \(monoid_i\).
 
@@
 \label{sec:math:data:base}
- The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other and does not itself store any of the components of the data.
-For example, \dbase\ can encode that the timeseries data described in \autoref{sec:math:data:fiber} is continuous, but is doing so independent of the timestamps.
-Every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$ because \dbase\ is the quotient space\cite{QuotientSpaceTopology2020,aurouxMath131Introduction} of \dtotal.
+
+ The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other.
+This is similar the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, but our model treats keys as a pure reference to topology.
+Decoupling the keys from their semantics allows the metadata to be altered facilitates encoding of data where the independent variable may not be clear; for example the growth of a plant is dependent on both time and sunlight, and changing the changing the coordinate system or time resolution should have no effect on how the records are connected to each other.
 
@@
     \includegraphics[width=\columnwidth]{base.png}
-  \caption{These two datasets have the same fiber space  $\dfiber=\reals\times\reals$, but the left dataset has a discrete continuity while the right is 1D continuous over the interval $\left[0,1\right]$.}
+    \caption{These two datasets have the same fiber space $\dfiber=\reals\times\reals$.
+For example, the left dataset is a set of discrete (\textit{time}, \textit{temperature}) records while the right is a 1D continuous function over the interval $\left[0,1\right]$ with return values of the form (\textit{time}, \textit{temperature}).}
     \label{fig:math:data:base}
    \end{figure}
-
  As illustrated in ~\autoref{fig:math:data:base}, \dbase\ can have any number of dimensions, can be continuous or discrete, and is somewhat independent of the dimensions of the fiber.
+Every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$ because \dbase\ is the quotient space\cite{QuotientSpaceTopology2020,aurouxMath131Introduction} of \dtotal.
 As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose}, we can decompose the total space into component bundles
@@
  \label{sec:math:data:section}
+
  \begin{figure}[htb]
@@
   \includegraphics[width=\columnwidth]{fiberbundle.png}
-  \caption{Each section in the fiber bundle is a unique continuous map from base space to fiber encoding the set of records in the dataset.
-The two sections, $\dsection^{(1)}$ and $\dsection^{(2)}$, shown here are in the set of global sections $\Gamma(\dtotal)$.}
+  \caption{Each section in the fiber bundle is a unique continuous map from base space to fiber encoding a set of records in the dataset.
+For example, the two sections $\dsection^{(1)}$ and $\dsection^{(2)}$ each encode a timeseries from a different weather station.}
   \label{fig:math:data:section}
@@
 As illustrated in \autoref{fig:math:data:section}, the section is a continuous mapping from a location $\dbasepoint \in \dbase$ on the base space to a record $\delement \in \dfiber$ in the fiber.
-Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is
+Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, a section
  \begin{equation}
@@
  \end{equation}
- where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber component.
+ returns a record for each \dbasepoint.
+The index function $g: \dbase \rightarrow \dfiber$ into each fiber component returns a value in the component.
 This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle.
@@
      \label{eq:sheaf}
-     \begin{tikzcd}
+     \begin{tikzcd}[column sep=huge]
          \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
@@
 which are sections of \dtotal\ pulled back over local neighborhood $U \subset \dtotal$ via the inclusion map \(\iota:\dtotal\rightarrow U\).
-The bookkeeping enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}
+The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}
 
@@
 This thickening is necessary when the dimensionality of \dbase\ is less than the dimensionality of the target display.
-For example, a \dbasepoint\ that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the points in \gfiber\ that visually represent the record at \dbasepoint.
+For example, a \dbasepoint\ that is a point in 0D \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the points in \gfiber\ that visually represent the record at \dbasepoint.
 The \vindex\ mapping is critical to interactive visualizations as it is the map from a region on screen to the data associated with that region.
@@
 \vtotalpull\ is the visual bundle \vtotal\ pulled back over \gbase\ via the equivariant continuity map $\vindex:\gbase \rightarrow \dbase$ introduced in \autoref{sec:math:graphic:base}.
-The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the latent space of possible parameters of a visualization type, such as a scatter or line plot.
+The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the space of possible parameters of a visualization type, such as a scatter or line plot.
 As with the data and graphic bundles, the visual bundle is defined by the projection map $\pi$
@@
 \end{equation}
-where \vsection\ is the visual variable encoding\cite{bertinSemiologyGraphicsDiagrams2011a} of the data section \dsection.
+where \vsection\ is the visual variable encoding, as described by Bertin \cite{bertinSemiologyGraphicsDiagrams2011a}, of the data section \dsection.
 The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave.
@@
   \centering
-  \includegraphics[width=1\columnwidth]{monoid_equivariant.png}
-  \caption{Equivariant \vchannel\ maps the weather data to corresponding symbols such that the relative order of the symbols is the same as the data as shown in the top two panels.
-In the bottom right panel the ordering of \textit{rain} and \textit{storm} is swapped in the symbols, meaning that the \vchannel\ is invalid; in the bottom left panel, \textit{storm} is instead mapped to the same symbol as \textit{rain} so that the ordering is preserved.}
+  \includegraphics[width=1\columnwidth]{figures/partial_fixed.png}
+  \caption{The equivariance constraint is not met in the first figure because \(rain\geq storm\), but \(\vchannel(rain) \leq \vchannel(storm)\) as shown by the crossed arrows.
+This is fixed by mapping \textit{storm} to the same element as \textit{rain}, such that \(\vchannel(rain) \geq \vchannel(storm)\).}
   \label{fig:math:artist:nu}
@@
 \end{table}
-The equivariance constraints for the Steven measurement scales\cite{stevensTheoryScalesMeasurement1946} are derived from the mathematical group structure of the scales.
-For monoid actions, the equivariance constraint is that relative order is preserved, as shown in the top two panels of \autoref{fig:math:artist:nu}.
+We can state the conditions on \vchannel\ such that they cover the Stevens measurement scales\cite{stevensTheoryScalesMeasurement1946}, as defined in \autoref{tab:math:artist:nu}.
+This is because Stevens’ defined the measurement scales in terms of their mathematical group structure and a group is a monoid with inverses\cite{remlingAlgebraMath5353}.
+An example of monoid action equivariance is the preservation of partial ordering illustrated in \autoref{fig:math:artist:nu}.
 
@@
   \includegraphics[width=\columnwidth]{path_of_q}
-  \caption{$\vchannel_i$ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.}
+  \caption{The transform functions $\vchannel_i$ convert data $\dsection_i\in \dtotal$ to visual characteristics $\vsection_i \in \vtotal$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection \in \gtotal$.}
   \label{fig:math:artist:q}
@@
 We then define the constraint on \vmark\ such that if \vmark\ is applied to two visual sections $\vsection$ and $\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection$ and $\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.
-We cannot define monoid actions on all of $\Gamma(\gtotal)$ because there are graphics $\gsection \in \Gamma(\gtotal)$ for which we cannot construct a valid mapping from \vtotal.
+We do not define monoid actions on all of $\Gamma(\gtotal)$ because there may be graphics $\gsection \in \Gamma(\gtotal)$ for which we cannot construct a valid mapping from \vtotal.
 \begin{figure}[htb]
@@
   \includegraphics[width=\columnwidth]{base_q}
-  \caption{The coordinates $(\gx, \gy)$ dictate the color of the region in prerender space \gbase\, which over the whole disk generates the graphical mark.
+  \caption{The coordinates \(\gbasepoint=(\gx, \gy)\) dictate the color of the region in prerender space \gbase.
+When \vmark\ is applied over the whole disk \(\gbase_{\dbasepoint}\), it generates the graphical point mark.
 The line fiber is thickened with the derivative because the tangent the line needs to be pushed perpendicular to the tangent of (xpos, ypos) in order to have visible thickness.}
@@
 \end{equation}
-where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
-The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:artist:base}
+with a constant \textit{size} and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of \vmark.
+The position of this swatch of color can be computed relative to the location on the disc \((\alpha, \beta) \in \gbase_{\dbasepoint}\) as shown in \autoref{fig:math:artist:base}
 \begin{align*}
@@
 \begin{equation}
-\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)
+\vmark(xpos, \hat{n}_{1}, ypos, \hat{n}_{2})(\alpha, \beta)
 \label{eq:math:artist:q:line}
@@
 The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
-such that the normal is  $\hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
+such that the normal is  $\hat{n}_{1} = \frac{n_1}{\lvert n \rvert}, \; \hat{n}_{2} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
 \begin{align*}
- x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
- y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
+ x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n}_1(\vindex(\alpha)) \\
+ y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n}_2(\vindex(\alpha)) 
 \end{align*}
-where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n_1}, \hat{n_2}$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
+where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n}_1, \hat{n}_2$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
 In \autoref{fig:teaser}, the image 
@@
 While the artist is specified in a fully functional manner in \autoref{eq:math:artist:diagram}, we implement our prototype in a heavily object oriented manner.
-This is a concession to existing Matplotlib architecture and so that the object can do bookkeeping of parameter information.
-But, in keeping with functional norms, we attempt to pass around as little information as necessary.
+We do so mostly to more easily manage function inputs, especially parameters that are passed through to the structurally functional transform and draw methods.
+
 \begin{minted}{python}
   fig, ax = plt.subplots()
-  artist = Artist(data, transforms)
+  artist = Artist(E, V)
   ax.add_artist(artist)
 \end{minted}
-Building on the current Matplotlib artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalance class artist \vartisteq\ as described in \autoref{eq:math:artist:aprime} .
-The visual bundle \vtotal\ is specified as the \mintinline{python}{transform}dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber, and the \vchannel\ encoders are passed in as functions or callable objects.
-The data bundle \dtotal\ is passed in as a \mintinline{python}{data} object.
+Building on the current Matplotlib artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalence class artist \vartisteq\ as described in \autoref{eq:math:artist:aprime}.
+The visual bundle \vtotal\ is specified as the \mintinline{python}{V} dictionary of the form \mintinline{python}|{parameter:(variable name, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber, and the \vchannel\ encoders are passed in as functions or callable objects.
+The data bundle \dtotal\ is passed in as a \mintinline{python}{E} object.
 By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist\ as defined in \autoref{eq:math:artist:artist}.
@@
   class ArtistClass(matplotlib.artist.Artist): #A'
-      def __init__(self, data, transforms, *args, **kwargs):
+      def __init__(self, E, V, *args, **kwargs):
           # properties that are specific to the graphic
-          self.data = data #E
-          self.transforms = transforms #V
+          self.E = E
+          self.V = V 
           super().__init__(*args, **kwargs)
 
-      def assemble(self, **args): #\hat{Q}
+      def hat_Q(self, **args): 
           # set the properties of the graphic
 
-      def draw(self, renderer): #A
+      def draw(self, renderer):
           # returns K, indexed on fiber then key 
-          view = self.data.view(self.axes) #\tau
+          tau = self.E.view(self.axes) 
           # visual channel encoding applied fiberwise 
-          # \nu_i(\tau_i)
-          visual = {p: t['encoder'](view[t['name']])
-                    for p, t in self.transforms.items()} 
-          self.assemble(**visual)
+          mu = {p: nu(tau(c))
+                   for p, (c, nu) in self.V.items()} 
+          self.hat_q(**mu)
           # pass configurations off to the renderer
@@
  The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
-The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset.
+The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}.
 To ensure the integrity of the section, \mintinline{python}{view} must be atomic, which means that the values cannot change after the method is called in draw until a new call in draw.
@@
 
- The \vchannel\ functions are then applied to the data, as describe in \autoref{eq:math:artist:nu}, to generate the visual section \vsection that here is the object \mintinline{python}{visual}.
+ The \vchannel\ functions are then applied to the data, as describe in \autoref{eq:math:artist:nu}, to generate the visual section \vsection that here is the object \mintinline{python}{V}.
 The conversion from data to visual space is simplified here to directly show that it is the encoding \vchannel\ applied to the component.
-The \mintinline{python}{assemble} function that is \vmarkd,  as defined in \autoref{eq:math:artist:qhat_k}, is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
+The \mintinline{python}{q_hat} function that is \vmarkd,  as defined in \autoref{eq:math:artist:qhat_k}, is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
 This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd.
@@
 \end{figure}
+The figure in \autoref{fig:code:scatter} is described by \autoref{eq:math:artist:q:scatter}.
+This is implemented via a \mintinline{python}{Line} object where the scatter marker shape is fixed as a circle, and the visual fiber components are x and y position and the facecolor and size of the marker.
+We only show the \mintinline{python}{q_hat} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
 
-To generate the figure in \autoref{fig:code:scatter}, the \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of primitive point and area marks.
-In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker, as described  by \autoref{eq:math:artist:q:scatter}.
-We only show the \mintinline{python}{assemble} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
 The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices.
@@
 class Point(mcollections.Collection):
-  def assemble(self, x, y, s, facecolors='C0' ): #\hat{Q}
+  def q_hat(self, x, y, s, facecolors): #\hat{Q}
     # construct geometries of circle glyphs
@@
 \end{minted} 
-In \mintinline{python}{assemble}, the \vsection\ components are used to construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
+In \mintinline{python}{q_hat}, the \vsection\ components are used to construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
 This is done via the \mintinline{python}{Path.circle} object.
@@
 class Line(mcollections.LineCollection):
-    def assemble(self, x, y, color='C0'): #\hat{Q}
+  def q_hat(self, x, y, color): #\hat{Q}
     #assemble line marks as set of segments 
@@
 As with \mintinline{python}{Point}, the data is then converted into visual variables.
-In \mintinline{python}{assemble}, described by \autoref{eq:math:artist:q:line}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
+In \mintinline{python}{q_hat}, described by \autoref{eq:math:artist:q:line}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
 Then the colors of each line segment are set.
@@
 # {P_i name:{'name':c_i, 'encoder:\nu_i}}
-transforms = {'x': {'name': 'v4', 'encoder': lambda x: x}, 
+V = {'x': {'name': 'v4', 'encoder': lambda x: x}, 
      'y': {'name': 'v2', 'encoder': lambda x: x},
@@
 \end{minted}
-that asks the user to specify the the properties of \dfiber\ and the \dbase\ connectivity in terms of a  simplicial triangulation scheme \cite{geometricaSimplicialComplexes}.
+that asks the user to specify the the properties of \dfiber\ and the \dbase\ connectivity as either discrete vertices or continuous data along edges.
 To generate the scatter plot and the line plot, the distinction is in the \mintinline{python}{tau} method that is the section.
 \begin{minted}{python}
-class VertexSimplex: 
-    """Fiberbundle is consistent across all sections
-    """
+class PointData:     
+    def __init__(self):
       FB = FiberBundle({'tables': ['vertex']},  
                        {'v1': float, 'v2': str, 'v3': float})
-
-    def __init__(self, sid = 45, size=1000, max_key=10**10):
-        # create random list of keys
     def tau(self, k):
-      # e1 is sampled from F1, e2 from F2, etc...
-      return (k, (e1, e2, e3, e4))
+      return # tau evaluated at one point k
 \end{minted}
-The discrete \mintinline{python}{tau} method returns a record of discrete points, while the line\mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint
+The discrete \mintinline{python}{tau} method returns a record of discrete points, akin to a row in a table, 
 \begin{minted}{python}
-def tau(self, k): #will fix location on page on revision
-    x, y = self._xy(k, self.distances, 
-               self.angle_samples[k], self.angle_samples[k+1]) 
-    color = self._color(k) 
-    return (k, (x, y, color))
+class LineData:
+  def __init__(self):
+    FB = FiberBundle({'tables': ['edge']},  
+                       {'x': float, 'y': float, 'color': str})
+  def tau(self, k):
+    return # tau evaluated on interval k
 \end{minted}
-In both cases the \mintinline{python}{view} method packages the data
+while the line\mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint.
 \begin{minted}{python}
@@
 \end{minted}
-into a data structure that the artist can unpack via data component name.
+In both cases the \mintinline{python}{view} method packages the data into a data structure that the artist can unpack via data component name, akin to a table with column names when \dbase\ is 0 or 1 D.
 
@@
 \begin{minted}{python}
-simplex.GraphLine(FB, edges, verticies, 
-                      num_samples=1000, connect=True)
+LineData(FB, edges, verticies, num_samples=1000, connect=True)
 \end{minted}
@@
 \begin{minted}{python}
-simplex.GraphLine(FB, edges, verticies, 
-                      num_samples=2, connect=False)
+LineData(FB, edges, verticies, num_samples=2, connect=False)
 \end{minted}
@@
 \section{Discussion}
-This work contributes a mathematical description of the mapping \vartist\ from data to visual representation.
-Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables,, multivariate high resolution spatio temporal datasets, and complex networks.
+This work contributes a functional model of the structure-preserving maps from data to visual representation to guide the development of visualization libraries, thereby providing a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables.
+Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks.
 Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications that the graphic must have continuity equivalent to the data, and that the visual characteristics of the graphics are equivariant to their corresponding components under monoid actions.
@@
 \subsection{Limitations}
-So far this model has only been worked out for a single data set tied to a primitive mark.
-The examples and prototype have so far only been implemented for the static 2D case, but nothing in the math limits to 2D and expansion to the animated case should be possible because the model is formalized in terms of the sheaf.
-While this model supports equivariance of figurative glyphs generated from parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it does not have a way to evaluate the semantic accuracy of the figurative representation.
+This model and the associated prototype is deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as VTK or D3.
 Even though the model is designed to be backend and format independent, it has only really been tested against PNGs rendered with the AGG backend.
 It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}.
-This model and the associated prototype is deeply tied to Matplotlib's existing architecture, so it has not been worked through how the model generalizes to other libraries, such as those built on Mackinlay's APT framework.
+While this model supports equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it cannot evaluate the semantic accuracy of the figurative representation.
+Effectiveness criteria\cite{mackinlayAutomaticDesignGraphical1987,chambersGraphicalMethodsData1983a} are out of scope for this model.
 
@@
 More artists need to be implemented to demonstrate that the model can underpin a minimally viable library, foremost an image\cite{toryRethinkingVisualizationHighlevel2004, haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as the boxplot\cite{wickham40YearsBoxplots2011}.
-This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
 Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
 While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
+
 \section{Conclusion}
-An unoffical philosophy of Matplotlib is to support making whatever kinds of plots a user may want, even if they seem nonsensical to the development team.
-The topological framework described in this work provides a way to facilitate this graph creation in a rigorous manner; any artist that meets the equivariance criteria described in this work by definition generates a graphic representation that matches the structure of the data being represented.
-We leave it to domain specialists to define the structure they need to preserve and the maps they want to make, and hopefully make the process easier by untangling these components into seperate constrained maps and providing a fairly general data and display model.
+The data model and functional transform refactor presented here allows bulding block libraries to better support domain specific libraries without having to explicitly take in the specific data structure and visualization needs of those domains.
+Adopting this model would induce a separation of data representation and visual representation that for example in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base.
+Because of the model presented here, this refactor can be done in consistent and robust manner such that the resulting components can be be guaranteed to preserve continuity and structure as defined by the domain specific library developers without having to fold those definitions back into the base library.
 
@@
 The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran.
+Also much appreciation to Nicolas Kruchten, Eric Schles, and Chana Tilevitz idenfiying the confusing bits and offering lots of suggestions.
+Thank you Marc Hanwell and Lev Manovich for helping refine the motivation, and Robert Haralick and Huy Vo for invaluable feedback on various iterations of this work.
 This project has been made possible in part by grant number 2019-207333 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.}

### 2021-03-31 paper/paper.tex
paper paper

@@
 %% this is still a mess, come back to it
-Models for describing visualizations and visualization tools are deeply tied to to the domains in which the visualizations are made \cite{friendlyBriefHistoryData2006a,pousmanCasualInformation2007}, and these tools are often made from building block libraries \cite{wongsuphasawatNavigatingWideWorld2021} such as the Python visualization library Matplotlib.
-Because building block libraries often support so many domains, the API can grow organically in ways that are incoherant and brittle.
-To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software needs to satisfy so that a visualization, however it is specified in the library built out of these blocks, is guaranteed to be a faithful representation of the data.
-Our model, which we call the  topological equivariant artist model (TEAM), provides a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables.
+
+Building block libraries underpin the visualization ecosystem; they are not the automated tools or high level libraries many practitioners use to build their visualizations, rather they are the libraries on which these tools are built \cite{wongsuphasawatNavigatingWideWorld2021}.
+These building block libraries provide specific components and utilities that are composed by domain specific visualization practitioners into reusable visualization tools; because building block libraries often support many domains, the API can grow organically in ways that are incoherent and brittle.
+To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software needs to satisfy so that a visualization--however it is specified in the library built out of these blocks--is guaranteed to be a structure preserving map from data to visual representation.
 
@@
   \includegraphics[width=\columnwidth]{figures/continuity.png}
-  \caption{Continuity describes how the records in the dataset, in this case the rows in the table, are connected to each other.
-The line plot is an invalid visual representation of the table because it implies that the records are connected to each other in some way.
-On the other hand the bar plot is valid because it preserves the records as independent data points.}
+  \caption{The line plot does not preserve continuity because it implies that the discrete records are connected to each other, while the bar plot is continuity preserving because it visually represents the records as independent data points.}
   \label{fig:intro:continuity}
@@
 
+The structure that visualizations must preserve is \textit{continuity} and \textit{equivariance}.
+As illustrated in \autoref{fig:intro:continuity}, \textit{continuity} is the way in which the records in the dataset are connected to each other.
+While the visualization tool can generate the line plot, it does not preserve continuity structure because it implies that the discrete categories are somehow continuous.
 \begin{figure}
   \includegraphics[width=\columnwidth]{figures/equivariance.png}
-  \caption{Equivariance means that transformations on the data side and on the visual side are equal.
-In this example, the temperature is doubled via the arrows representing the doubling action, and the plot of the temperature is doubled via the arrows on the visual side.}
+  \caption{The data in blue is scaled by a factor of two, yielding the data in orange.
+To preserve equivariance, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
   \label{fig:intro:equivariance}
 \end{figure}
+Equivariance means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side.
+In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.
 
 Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation.
-The contribution of TEAM is
+The contribution of this work is a
 \begin{enumerate}
-  \item topology preserving relationship between data and graphic via continuous maps
-  \item property preservation from data component to visual representation as monoid action equivariant maps
-  \item functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
-  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model.
+  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
+  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist}
+  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
+  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code}
 \end{enumerate}
@@
 \section{Related Work}
-%
-% My best paraphrasing of the main idea would be that this paper takes the ideas of going from data to graphic along with some important constraints that people generally agree are important and formalizes them into a topological model, then shows how that model can map into a concrete implementation.
-The important part of this model is that it models the data sufficiently abstractly that it's able to work with basically any form or type of data.
-The sentence starting at line 489 seems to be the big "aha" to me, basically saying something like "if you implement things such that the code equivalents of these terms are separated in some specific way, then you get these important formally-describable properties".
-Is that right?
-If so, I'm left very curious about what alternative (possibly already-existing?) Artist implementations there might be that seem to work, but don't decompose things in this way, and therefore don't have these important properties.
-Or maybe the existing implementations do have these properties but aren't general enough?
-I know there are space constraints etc but if there was some way of showing how the current Matplotlib artists work and how they fall short of this model, that would be very enlightening I think.
+The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal, but this work bridges the formalism and implementation in a functional manner at a building building blocks library level.
+When introducing the retinal variables, Bertin informally introduces the notion that continuity is preserved in the mark and  defines equivariance constraints (selective, associative, ordered, quantitative)\cite{bertinSemiologyGraphicsDiagrams2011a}.
+In the \textit{A Presentation Tool}(APT) model, Mackinlay embeds the continuity constraint in the choice of visualization type and generalizes the equivariance constraint to preserving a binary operator from one domain to another.
+The algebraic model of visualization proposed by Kindlmann and Scheidegger restrict equivarance to group actions, but explicitly define composable symmetric mappings from data space to representation space to graphic space.
+Our model targets monoid actions as they are more restrictive than all binary operations but more general than groups, and explicitly formalizes the way in which continuity is preserved as topologically maps.
+APT and the algebraic model are also targeted at visual designs and tools that automatically generate those designs, while our model targets the tools that build those tools.
+
+Current visualization tools tend to either be architectured around a core data structure\cite{HeerSoftware2006} or explicitly implement the visual algorithm in terms of the data continuity the algorithm expects \cite{toryRethinkingVisualizationHighlevel2004}.
+For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}.
+Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}.
+Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.
+Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.
+Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects\cite{loudenProgrammingLanguagesPrinciples2010}, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.
 
@@
 \subsection{Artist}
+\label{sec:math:artist}
 The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$. 
@@
 \subsection{Limitations}
-This model and the associated prototype is deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as VTK or D3.
-Even though the model is designed to be backend and format independent, it has only really been tested against PNGs rendered with the AGG backend.
-It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}.
-While this model supports equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it cannot evaluate the semantic accuracy of the figurative representation.
-Effectiveness criteria\cite{mackinlayAutomaticDesignGraphical1987,chambersGraphicalMethodsData1983a} are out of scope for this model.
+Our model and prototype are deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as VTK or D3.
+Even though the model is designed to be backend and format independent, it has only been tested against PNGs rendered with AGG\cite{shemanarevAntiGrainGeometry}.
+It is unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997} that implement different models of \gsection.
+While our model supports equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from data components\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it cannot evaluate the semantic accuracy of the figurative representation.
+Effectiveness criteria\cite{mackinlayAutomaticDesignGraphical1987,chambersGraphicalMethodsData1983a} are out of scope.
 
 \subsection{Future Work}
-More work is needed to formalize the composition operators and equivalence class \vartisteq.
-More artists need to be implemented to demonstrate that the model can underpin a minimally viable library, foremost an image\cite{toryRethinkingVisualizationHighlevel2004, haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as the boxplot\cite{wickham40YearsBoxplots2011}.
-Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
-While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
+More work is needed to formalize the composition operators and equivalence class \vartisteq and  we need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}.
+Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014}.
+While this paper formulates visualization in terms of monoidal action homomorphisms between fiber bundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
 
@@
 The data model and functional transform refactor presented here allows bulding block libraries to better support domain specific libraries without having to explicitly take in the specific data structure and visualization needs of those domains.
-Adopting this model would induce a separation of data representation and visual representation that for example in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base.
-Because of the model presented here, this refactor can be done in consistent and robust manner such that the resulting components can be be guaranteed to preserve continuity and structure as defined by the domain specific library developers without having to fold those definitions back into the base library.
+Adopting this model would induce a separation of data representation and visual representation that, for example, in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base.
+A refactor driven by TEAM would result in components that can be be guaranteed to preserve continuity and structure, as defined by the domain specific library developers, without having to fold domain specific assumptions back into the base library.
 
@@
 \acknowledgments{
-The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran.
-Also much appreciation to Nicolas Kruchten, Eric Schles, and Chana Tilevitz idenfiying the confusing bits and offering lots of suggestions.
-Thank you Marc Hanwell and Lev Manovich for helping refine the motivation, and Robert Haralick and Huy Vo for invaluable feedback on various iterations of this work.
+The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran, Eric Schles and Chana Tilevitz for idenfiying the confusing bits, Nicolas Kruchten for articulating the framing and Marc Hanwell, Lev Manovich, Robert Haralick and Huy Vo for invaluable feedback on various iterations.
+This paper is indebted to Kindlmann and Scheidegger's \cite{kindlmannAlgebraicProcessVisualization2014} for serving as a model and Munzner's guide to writing visualization papers.
 This project has been made possible in part by grant number 2019-207333 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.}

### 2021-03-31 paper/paper.tex
cleaned up intro

@@
 
-Building block libraries underpin the visualization ecosystem; they are not the automated tools or high level libraries many practitioners use to build their visualizations, rather they are the libraries on which these tools are built \cite{wongsuphasawatNavigatingWideWorld2021}.
-These building block libraries provide specific components and utilities that are composed by domain specific visualization practitioners into reusable visualization tools; because building block libraries often support many domains, the API can grow organically in ways that are incoherent and brittle.
-To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software needs to satisfy so that a visualization--however it is specified in the library built out of these blocks--is guaranteed to be a structure preserving map from data to visual representation.
+Building block libraries underpin the visualization ecosystem; they are not the automated tools or high level libraries many practitioners use to build their visualizations, rather they are the libraries on which these tools are built.
+These building block libraries, as described by Wongsuphasawat, provide specific components and utilities that are composed by domain specific visualization practitioners into reusable visualization tools \cite{wongsuphasawatNavigatingWideWorld2021}.
+Because building block libraries often support many domains, the API can grow organically in ways that are incoherent and brittle.
+Motivated by the challenge of rearchitecturing a building blocks library, Matplotlib\cite{hunterMatplotlib2DGraphics2007}, we developed a functional model, which we call the Topological Equivariant Artist Model (TEAM), to formalize the constraints building block components must satisfy to be considered structure preserving maps from data to visual representation.
 
-\begin{figure}
-  \includegraphics[width=\columnwidth]{figures/continuity.png}
-  \caption{The line plot does not preserve continuity because it implies that the discrete records are connected to each other, while the bar plot is continuity preserving because it visually represents the records as independent data points.}
-  \label{fig:intro:continuity}
+\begin{figure}[H]
+  \centering 
+  \subfloat[\label{fig:intro:continuity}]{%
+    \includegraphics[width=1\columnwidth]{continuity.png}}\\
+  \subfloat[\label{fig:intro:equivariance}]{%
+    \includegraphics[width=1\columnwidth]{equivariance.png}}
+\caption{In \autoref{fig:intro:continuity}, the line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points.
+In \autoref{fig:intro:equivariance}, the data in blue is scaled by a factor of two, yielding the data in orange.
+To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
+\label{fig:intro:structure}
 \end{figure}
-
-The structure that visualizations must preserve is \textit{continuity} and \textit{equivariance}.
-As illustrated in \autoref{fig:intro:continuity}, \textit{continuity} is the way in which the records in the dataset are connected to each other.
-While the visualization tool can generate the line plot, it does not preserve continuity structure because it implies that the discrete categories are somehow continuous.
-\begin{figure}
-  \includegraphics[width=\columnwidth]{figures/equivariance.png}
-  \caption{The data in blue is scaled by a factor of two, yielding the data in orange.
-To preserve equivariance, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
-  \label{fig:intro:equivariance}
-\end{figure}
+The structure that visualizations must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}.
+\textit{continuity} is the way in which the records in the dataset are connected to each other.
+In \autoref{fig:intro:continuity} the line plot does not preserve continuity because the lines connecting the discrete categories implies that weather events and frequency of those events is continuous.
 Equivariance means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side.
@@
 
-Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation.
-The contribution of this work is a
+To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software must satisfy so that a visualization--however it is specified in the library built out of these blocks--is guaranteed to be an equivariant continuity preserving mapping from data to visualization.
+The contribution of this work is
 \begin{enumerate}

### 2021-03-31 paper/paper.tex
more rework of intro

@@
 \item
- Hannah Aizenman and Michael Grossberg are with the Computer Science department, City College of New York. E-mail: haizenman@ccny.cuny.edu, mgrossberg@ccny.cuny.edu.
+ Hannah Aizenman and Michael Grossberg are with the department of Computer Science, City College of New York. E-mail: haizenman@ccny.cuny.edu, mgrossberg@ccny.cuny.edu.
 \item
@@
 
-Building block libraries underpin the visualization ecosystem; they are not the automated tools or high level libraries many practitioners use to build their visualizations, rather they are the libraries on which these tools are built.
-These building block libraries, as described by Wongsuphasawat, provide specific components and utilities that are composed by domain specific visualization practitioners into reusable visualization tools \cite{wongsuphasawatNavigatingWideWorld2021}.
-Because building block libraries often support many domains, the API can grow organically in ways that are incoherent and brittle.
-Motivated by the challenge of rearchitecturing a building blocks library, Matplotlib\cite{hunterMatplotlib2DGraphics2007}, we developed a functional model, which we call the Topological Equivariant Artist Model (TEAM), to formalize the constraints building block components must satisfy to be considered structure preserving maps from data to visual representation.
-
+General purpose visualization libraries are often tasked with supporting many domains, leading the libraries to grow organically in ways that can lead to incoherent interfaces and brittle components.
+To adapt to modern library development needs, this paper presents a functional model of the how components provided by lower level libraries\cite{wongsuphasawatNavigatingWideWorld2021} can be guaranteed to be structure preserving maps from data to graphic.
+We target this level of library architecture because our aim is the libraries developers use to build automated and user facing tools, not those tools themselves; we are agnostic to what makes for a good visualization so long as the structure preserving constraints are satisfied.
 \begin{figure}[H]
@@
 \end{figure}
-The structure that visualizations must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}.
-\textit{continuity} is the way in which the records in the dataset are connected to each other.
-In \autoref{fig:intro:continuity} the line plot does not preserve continuity because the lines connecting the discrete categories implies that weather events and frequency of those events is continuous.
-Equivariance means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side.
+The structure that components of a visualization library must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}.
+\textit{Continuity} is the way in which the records in the dataset are connected to each other.
+In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval.
+\textit{Equivariance} means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side.
 In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.
 
-To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software must satisfy so that a visualization--however it is specified in the library built out of these blocks--is guaranteed to be an equivariant continuity preserving mapping from data to visualization.
+Motivated by the challenge of re-architecturing the Python visualization library Matplotlib, we developed a functional model of library architecture, which we call the Topological Equivariant Artist Model (TEAM), to specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation.
+A functional architecture derived from our model would gain a generalized data abstraction and well specified reusable components with minimal side effects\cite{loudenProgrammingLanguagesPrinciples2010}, in turn yielding improvements in maintainability, extendability, scaling, and support for concurrency.
 The contribution of this work is
@@
   \includegraphics[width=1\columnwidth]{figures/partial_fixed.png}
-  \caption{The equivariance constraint is not met in the first figure because \(rain\geq storm\), but \(\vchannel(rain) \leq \vchannel(storm)\) as shown by the crossed arrows.
-This is fixed by mapping \textit{storm} to the same element as \textit{rain}, such that \(\vchannel(rain) \geq \vchannel(storm)\).}
+  \caption{As highlighted by the red arrows, the equivariance constraint is not met because \(rain\geq storm\), but \(\vchannel(rain) \leq \vchannel(storm)\).
+Instead, mapping \textit{storm} to the same element as \textit{rain} yields an equivariant \vchannel\ such that \(\vchannel(rain) \geq \vchannel(storm)\).}
   \label{fig:math:artist:nu}
@@
       return # tau evaluated at one point k
-\end{minted}
-The discrete \mintinline{python}{tau} method returns a record of discrete points, akin to a row in a table, 
-\begin{minted}{python}
+
 class LineData:
@@
 \end{minted}
-while the line\mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint.
+The discrete \mintinline{python}{tau} method returns a record of discrete points, akin to a row in a table, while the line\mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint.
 \begin{minted}{python}
@@
 \end{minted}
-In both cases the \mintinline{python}{view} method packages the data into a data structure that the artist can unpack via data component name, akin to a table with column names when \dbase\ is 0 or 1 D.
+In both cases, the \mintinline{python}{view} method packages the data into a data structure that the artist can unpack via data component name, akin to a table with column names when \dbase\ is 0 or 1 D.
 
@@
 \end{figure}
-The graphics in figure \autoref{fig:code:multilines} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{Graphline} data source where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix.
+The graphics in figure \autoref{fig:code:multilines} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{GraphData} data source where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix.
 The multicolored line is a connected graph of edges with each edge function evaluated on 1000 samples,
 \begin{minted}{python}
-LineData(FB, edges, verticies, num_samples=1000, connect=True)
+GraphData(FB, edges, verticies, num_samples=1000, connect=True)
 \end{minted}
@@
 \begin{minted}{python}
-LineData(FB, edges, verticies, num_samples=2, connect=False)
+GraphData(FB, edges, verticies, num_samples=2, connect=False)
 \end{minted}

### 2021-03-31 paper/paper.tex
currently submitted version

@@
 General purpose visualization libraries are often tasked with supporting many domains, leading the libraries to grow organically in ways that can lead to incoherent interfaces and brittle components.
-To adapt to modern library development needs, this paper presents a functional model of the how components provided by lower level libraries\cite{wongsuphasawatNavigatingWideWorld2021} can be guaranteed to be structure preserving maps from data to graphic.
-We target this level of library architecture because our aim is the libraries developers use to build automated and user facing tools, not those tools themselves; we are agnostic to what makes for a good visualization so long as the structure preserving constraints are satisfied.
+To adapt to modern library development needs, this paper presents a functional model of the how components provided by lower level libraries can be guaranteed to be structure preserving maps from data to graphic.
+We target this level of library architecture because our aim is the libraries developers use to build automated and user facing tools, not those tools themselves; our model is agnostic as to what makes for a good visualization so long as the structure preserving constraints are satisfied.
 \begin{figure}[H]
@@
 The structure that components of a visualization library must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}.
-\textit{Continuity} is the way in which the records in the dataset are connected to each other.
+\textit{Continuity} is the way in which the records in the dataset are connected to each other and \textit{equivariance} means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side.
 In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval.
-\textit{Equivariance} means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side.
 In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.
 
-Motivated by the challenge of re-architecturing the Python visualization library Matplotlib, we developed a functional model of library architecture, which we call the Topological Equivariant Artist Model (TEAM), to specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation.
-A functional architecture derived from our model would gain a generalized data abstraction and well specified reusable components with minimal side effects\cite{loudenProgrammingLanguagesPrinciples2010}, in turn yielding improvements in maintainability, extendability, scaling, and support for concurrency.
+To specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation, we developed a functional model of library architecture, which we call the Topological Equivariant Artist Model (TEAM).
+Motivated by the challenge of re-architecturing the Python visualization library Matplotlib, a functional architecture derived from our model would gain a generalized data abstraction and well specified reusable components with minimal side effects\cite{loudenProgrammingLanguagesPrinciples2010}, in turn yielding improvements in maintainability, extendability, scaling, and support for concurrency.
 The contribution of this work is
@@
 \section{Related Work}
-The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal, but this work bridges the formalism and implementation in a functional manner at a building building blocks library level.
-When introducing the retinal variables, Bertin informally introduces the notion that continuity is preserved in the mark and  defines equivariance constraints (selective, associative, ordered, quantitative)\cite{bertinSemiologyGraphicsDiagrams2011a}.
+The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal, but this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library\cite{wongsuphasawatNavigatingWideWorld2021} level.
+When introducing the retinal variables, Bertin informally specifies that continuity is preserved in the mark and defines equivariance constraints in terms of data and visual variables being selective, associative, ordered, or quantitative\cite{bertinSemiologyGraphicsDiagrams2011a}.
 In the \textit{A Presentation Tool}(APT) model, Mackinlay embeds the continuity constraint in the choice of visualization type and generalizes the equivariance constraint to preserving a binary operator from one domain to another.
-The algebraic model of visualization proposed by Kindlmann and Scheidegger restrict equivarance to group actions, but explicitly define composable symmetric mappings from data space to representation space to graphic space.
-Our model targets monoid actions as they are more restrictive than all binary operations but more general than groups, and explicitly formalizes the way in which continuity is preserved as topologically maps.
-APT and the algebraic model are also targeted at visual designs and tools that automatically generate those designs, while our model targets the tools that build those tools.
+The algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}, proposed by Kindlmann and Scheidegger, restricts equivariance to invertible transformations.
+Our model defines equivariance in terms of monoid actions, which are more restrictive than all binary operations but more general than invertible transformations.
+Their algebraic model also describes visualization as commutative mappings from data space to representation space to graphic space.
+Our model has the same notion of commutative mappings across analogous spaces, but uses topology to explicitly include continuity.
 
-Current visualization tools tend to either be architectured around a core data structure\cite{HeerSoftware2006} or explicitly implement the visual algorithm in terms of the data continuity the algorithm expects \cite{toryRethinkingVisualizationHighlevel2004}.
+We emphasize continuity because visualization tools tend to either be architectured around a core data structure\cite{HeerSoftware2006} or explicitly implement the visual algorithm in terms of the data continuity the algorithm expects \cite{toryRethinkingVisualizationHighlevel2004}.
 For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}.
@@
 Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.
-Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects\cite{loudenProgrammingLanguagesPrinciples2010}, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.
+Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.
 
@@
 We decompose the artist into an indexing map from graphic to data (\autoref{sec:math:graphic:base}), a map from data components to visual components (\autoref{sec:math:artist:nu}), and map from visual components to graphic (\autoref{sec:math:artist:q}).
-In this section we discuss the formal properties of these fiber bundles and maps such that they can be used to specify an implementation, for example the one prototyped in \autoref{sec:code}.
 
@@
   \caption{These two datasets have the same base space \dbase.
-The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature) from \autoref{fig:intro:equivariance}, while the cube is the fiber $\reals\times\realsp\times\reals$ associated with (time, wind=(speed, direction))}
+The plane is a representation of the fiber $\dfiber=\reals\times\reals$ for the variables (time, temperature) from \autoref{fig:intro:equivariance}, while the cube is the fiber $\reals\times\reals\times\reals$ associated with (time, wind=(speed, direction))}
   \label{fig:math:data:fiber}
@@
 
-In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\realsp\times\reals)\)) which encodes (\emph{speed}, \emph{direction}).
+In \autoref{fig:math:data:fiber} the plane fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{temperature}, \texttt{float}, \reals), while the cube fiber has components (\emph{time}, \texttt{datetime}, \reals) and (\emph{wind}, \texttt{wind}, \(\reals\times\reals\)) which encodes (\emph{speed}, \emph{direction}).
 
@@
 \end{equation}
-of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$
+of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)
 
@@
 Monoids are commonly found in functional programming because a property of monoids is that the components can be composed into complex transformation pipelines \cite{yorgeyMonoidsThemeVariations}.
-For example, a component that converts \(\dfiber_i=time\) with \(\monoid_i\) to hours can be combined with a component that shifts \(time\) by 2 hours to yield \(hours_2\) with the same \(monoid_i\).
 
@@
 \label{sec:math:data:base}
-
  The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other.
@@
 
-\subsubsection{Equivariant Topology: Graphic Base Space}
+\subsubsection{Equivalant Continuity: Graphic Base Space}
 \label{sec:math:graphic:base}
@@
   \end{tabulary}
+  \caption{}
   \label{tab:math:artist:nu}
@@
 \label{sec:code}
-To build a prototype, we make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to transform data coordinates into screen coordinates.
+To build a prototype, we make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to convert data coordinates into screen coordinates.
 While the artist is specified in a fully functional manner in \autoref{eq:math:artist:diagram}, we implement our prototype in a heavily object oriented manner.
@@
 The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}.
-To ensure the integrity of the section, \mintinline{python}{view} must be atomic, which means that the values cannot change after the method is called in draw until a new call in draw.
+To ensure the integrity of the section, \mintinline{python}{view} must be atomic, which means that the values cannot change after the method is called in draw until a new call to draw.
 We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.

### 2021-04-06 paper/paper.tex
post passover housekeeping

@@
 %% * theory/model
-\vgtcpapertype{Theory/Model}
+\vgtcpapertype{theory/model}
 
@@
 To adapt to modern library development needs, this paper presents a functional model of the how components provided by lower level libraries can be guaranteed to be structure preserving maps from data to graphic.
-We target this level of library architecture because our aim is the libraries developers use to build automated and user facing tools, not those tools themselves; our model is agnostic as to what makes for a good visualization so long as the structure preserving constraints are satisfied.
+The target for this model are the libraries developers use to build automated and user facing tools rather than those tools.
+The model presented here is agnostic towards what makes for a good visualization, so long as the specified constraints for building those visualizations are satisfied.
 \begin{figure}[H]
@@
 In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval.
-In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.
+In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from blue to orange data and lines.
 
-To specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation, we developed a functional model of library architecture, which we call the Topological Equivariant Artist Model (TEAM).
+To specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation, we developed a functional model, which we call the Topological Equivariant Artist Model (TEAM).
 Motivated by the challenge of re-architecturing the Python visualization library Matplotlib, a functional architecture derived from our model would gain a generalized data abstraction and well specified reusable components with minimal side effects\cite{loudenProgrammingLanguagesPrinciples2010}, in turn yielding improvements in maintainability, extendability, scaling, and support for concurrency.
@@
 \section{Related Work}
-The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal, but this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library\cite{wongsuphasawatNavigatingWideWorld2021} level.
+The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal; this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library\cite{wongsuphasawatNavigatingWideWorld2021} level.
 When introducing the retinal variables, Bertin informally specifies that continuity is preserved in the mark and defines equivariance constraints in terms of data and visual variables being selective, associative, ordered, or quantitative\cite{bertinSemiologyGraphicsDiagrams2011a}.
@@
 Our model defines equivariance in terms of monoid actions, which are more restrictive than all binary operations but more general than invertible transformations.
-Their algebraic model also describes visualization as commutative mappings from data space to representation space to graphic space.
+The Kindlmann and Scheidegger algebraic model also describes visualization as commutative mappings from data space to representation space to graphic space.
 Our model has the same notion of commutative mappings across analogous spaces, but uses topology to explicitly include continuity.
@@
 Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.
-Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured.
+Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can seem incoherent as every visualization has a different notion of how the data is structured.
 Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.
@@
 
-In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image.
+In \autoref{fig:teaser}, we illustrate the output of a minimal \vmark\ that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image.
 The scatter plot can be defined as
@@
 where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.
-A test for equivariance, as described in \autoref{eq:math:artist:nu:equivariance}, can be implemented trivially
 \begin{minted}{python}
@@
 \end{minted}
-but is currently factored out of the artist for clarity.
+As described in \autoref{eq:math:artist:nu:equivariance}, a test for equivariance can be implemented trivially. It is currently factored out of the artist for clarity.
 
@@
 Factoring out graphic formation into assembly functions allows for much more clarity in how they differ.
-This prototype demonstrates that this framework can generate the fundemental point (scatter plot) and line (line chart) marks.
+This prototype demonstrates that this framework can generate the fundamental point (scatter plot) and line (line chart) marks.
 
@@
 \subsection{Future Work}
-More work is needed to formalize the composition operators and equivalence class \vartisteq and  we need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}.
+More work is needed to formalize the composition operators and equivalence class \vartisteq.
+We also need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}.
 Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014}.
@@
 \section{Conclusion}
-The data model and functional transform refactor presented here allows bulding block libraries to better support domain specific libraries without having to explicitly take in the specific data structure and visualization needs of those domains.
+The data model and functional transform refactor presented here allows building block libraries to better support downstream, including domain specific, libraries without having to explicitly incorporate the
+specific data structure and visualization needs of those domains without having to fold domain specific assumptions back into the base library.
 Adopting this model would induce a separation of data representation and visual representation that, for example, in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base.
-A refactor driven by TEAM would result in components that can be be guaranteed to preserve continuity and structure, as defined by the domain specific library developers, without having to fold domain specific assumptions back into the base library.
+A TEAM driven refactor would produce more maintainable, reusable, and extensible code, leading to better building blocks for the ecosystem of tools built on top of Matplotlib.
 
@@
 \acknowledgments{
-The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran, Eric Schles and Chana Tilevitz for idenfiying the confusing bits, Nicolas Kruchten for articulating the framing and Marc Hanwell, Lev Manovich, Robert Haralick and Huy Vo for invaluable feedback on various iterations.
+The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran, Eric Schles and Chana Tilevitz for idenfiying the confusing bits, Nicolas Kruchten for articulating the framing, and Marc Hanwell, Lev Manovich, Robert Haralick and Huy Vo for very helpful feedback on various iterations.
 This paper is indebted to Kindlmann and Scheidegger's \cite{kindlmannAlgebraicProcessVisualization2014} for serving as a model and Munzner's guide to writing visualization papers.

### 2021-04-08 draft/sections/abstract.tex
dropping in paper abstract

--- 2021-01-04 draft/sections/abstract.tex
@@
- A critical aspect of data visualization is that the graphical representation of data match the properties of the data; this fails when order is not preserved in representations of ordinal data or scale for numerical data.
-In this work, we propose that the mathematical notion of equivariance formalizes the expectation that graphics match the data.
-We developed a model we call the topological artist model (TAM) in which data and graphics can be viewed as sections of fiber bundles.
-This model allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that translates the dataset connectivity into graphical elements.
-Furthermore, our model supports an algebraic sum operation such that more complex visualizations can be built from simple ones.
-We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap.
-We show that this model can be implemented with a small prototype.
-
-To demonstrate the practical value of our model, we propose a model driven re-architecture of the artist layer of the Python visualization library Matplotlib.
-We represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering.
-In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity.
-The thesis will follow through on this proposal to explore how to further develop our model, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.
+This work presents a functional model of the structure-preserving maps from data to visual representation to guide the development of visualization libraries.
+Our model, which we call the  topological equivariant artist model (TEAM), provides a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables.
+We formalize these transformations as actions on sections of topological fiber bundles, which are mathematical structures that allow us to encode continuity as a base space, variable properties as a fiber space, and data as binding maps, called sections, between the base and fiber spaces.
+This abstraction allows us to generalize to any type of data structure, rather than assuming, for example, that the data is a relational table, image, data cube, or network-graph.
+Moreover, we extend the fiber bundle abstraction to the graphic objects that the data is mapped to.
+By doing so, we can track the preservation of data continuity in terms of continuous maps from the base space of the data bundle to the base space of the graphic bundle.
+Equivariant maps on the fiber spaces preserve the structure of the variables; this structure can be represented in terms of monoid actions, which are a generalization of the mathematical structure of Stevens’ theory of measurement scales.
+We briefly sketch that these transformations have an algebraic structure which lets us build complex components for visualization from simple ones.
+We demonstrate the utility of this model through case studies of a scatter plot, line plot, and image.
+To demonstrate the feasibility of the model, we implement a prototype of a scatter and line plot in the context of the Matplotlib Python visualization library.
+We propose that the functional architecture derived from a TEAM based design specification can provide a basis for a more consistent API and better modularity, extendability, scaling and support for concurrency.

### 2021-04-08 draft/sections/intro.tex
moving around intro outline + more references

@@
 \end{figure}
+
+- [ ] (1) Audience: lower-level visualization library developers, domain specific library developers (continuing the architecture choices mpl made)
+- [ ] (2) Need: unified model: continuity + equivariance 
+        - [ ] mpl works, unmaintainable, grew organically, this stuff is buried implicitly, gives a way to make it clear
+        - [ ] interactivity screen space back to data space, different implicit data than API incoherancy than bug?
+            - [ ] currently done in adhoc manner, want to enforce contracts for the interactivity 
+        - [ ] redesign that reliably enforces these constraints 
+        - [ ] library that supports all the data 
+            - [ ] relational tables, and images, and data cubes (with coordinates), and networks, and other data structures
+            - [ ] streaming and concurrency 
+            - [ ] blob the same? - model explains why this should be true
+        - [ ] framework to enforce invariance at the architecture level
+        - [ ] match engineering w/ math
+- [ ] (3) Proposed solution: 
+    - [ ] functional math model, structure/type preserving 
+- [ ] (4) Novelty:
+    - [ ] functional architecture
+    - [ ] explicitly expressing continuity in the model
+    - [ ] model is constraint oriented rather than evaluative
+- [ ] (5) Benefits: 
+    - [ ] functional: robustness, concurrency, modularity, proves that reliably, backfit needs to this 
+- [ ] (6) Evaluation: The method of evaluation to verify that the proposed solution delivers the stated benefits must be given.
+        The best justifications explicitly discuss particular choices in the context of several possible alternatives.
+        1. Case studies in the form of implementations (is this feasible) 
+        2. Comparison w/ existing implementation          
+- [ ] (7) More specific discussion of what will be done and on what timeline.
+        [quarterly gantt charts?]
+[] close out board https://github.com/story645/proposal/projects/3
+
 The work presented in this paper is motivated by a need for a library of visualization components that developers could use to build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.

### 2021-04-08 draft/sections/intro.tex
moving stuff around + regenerate figures + some words

@@
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/intro/dar.png}
-    \caption{Visualization is a mapping from data into visual encodings that are then rendered into images. In our model, this visual encoding stage is called the artist.}
+    \includegraphics[width=\textwidth]{figures/intro/artist.png}
+    \caption{This table encodes information that here we choose to represent as a map or time series to try to match the visual encoding of the data to the structure inferred from the date and latitute, longitude fields of the dataset.}
     \label{fig:intro_artist_stages}
@@
 We validate our model by using it to re-design artist and data access layer of Matplotlib, a general purpose visualization tool.
+
+\begin{figure}[H]
+    \begin{subfigure}{1\textwidth}
+        \includegraphics[width=1\columnwidth]{figures/intro/continuity.png}
+        \caption{The line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points.
+}
+        \label{fig:intro:continuity}
+    \end{subfigure}
+    \begin{subfigure}{1\textwidth}
+        \includegraphics[width=1\columnwidth]{figures/intro/equivariance.png}
+        \caption{The data in blue is scaled by a factor of two, yielding the data in orange.
+To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
+        \label{fig:intro:equivariance}
+    \end{subfigure}
+  \caption{}
+  \label{fig:intro:structure}
+  \end{figure}
+  The structure that the transformations that make up a visualization library must preserve are \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}.
+\begin{mdframed}[roundcorner=10pt, frametitle= definitions, frametitlerule=true, frametitlebackgroundcolor=gray!10]
+    \begin{description}
+        \item[\textbf{continuity}] How records in the dataset are connected to each other, e.g. discrete rows, neworked nodes, points on a continuous surface
+        \item[\textbf{equivariance}] if an action is applied to the data or the graphic--e.g.
+a rotation, permutation, translation, or rescaling-- there must be an equivalent action applied on the other side of the transformation.
+    \end{description}
+\end{mdframed}
+
+  In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval.
+In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from blue to orange data and lines.
 

### 2021-04-09 draft/sections/intro.tex
reworked intro, start of reshuffling of background

--- 2021-03-02 draft/sections/intro.tex
@@
 \label{sec:intro}
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/intro/dar.png}
-    \caption{Visualization is a mapping from data into visual encodings that are then rendered into images. In our model, this visual encoding stage is called the artist.}
-    \label{fig:intro_artist_stages}
-\end{figure}
-The work presented in this paper is motivated by a need for a library of visualization components that developers could use to build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.
-While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
-The core architecture also needs to be robust to the big data needs of many visualization practitioners, and therefore support distributed and streaming data needs.
-To support both exploratory and confirmatory visualization\cite{tukeyWeNeedBoth1980}, this tool needs to support 2D and 3D, static, dynamic and interactive visualizations.
+Building block level visualization libraries implement the functions that transform data into some component of the visual representation, providing a collection of components and utilities that can be combined to create a visualization \cite{wongsuphasawatNavigatingWideWorld2021}.
+Often domain specific tools are built out of these building block libraries, meaning the building block libraries must provide components general enough to satisfy a wide variety of data and visual representation needs.
+Specifically they must satisfy confirmatory and exploratory visualization needs \cite{tukeyWeNeedBoth1980}, in scientific and information (formal and informal\cite{pousmanCasualInformation2007}) visualization domains.
+To satisfy ever developing visualization needs, new components can be incorporated into the library in sometimes ad-hoc ways that can lead to API incoherancy and inconsistent behavior.
+We propose that like physical building blocks, building block libraries should provide a collection of well defined pieces that can be composed in whichever ways the blocks fit together.\note{figure of what we mean by building block?} We specify that for a valid visualization block is a structure preserving transformation from data to visual space, and we define structure in terms of \textit{continuity} and \textit{equivariance}.
+We then use this model to develop a design specification for the components of a building block visualization library.
+The notion of self contained, inherently modular, building blocks lends itself naturally to a functional paradigm of visualization \cite{hughesWhyFunctionalProgramming1989}.
+We adopt a functional model for a redesign because the lack of side effects means functional architecture can be evaluated for correctness, functional programs tend to be shorter and clearer, and are well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.
 
-Specifically, this work was driven by a rearchitecture of the Python visualization library Matplotlib\cite{hunterMatplotlib2DGraphics2007} to meet modern data visualization needs.
-We aim to take advantage of developments in software design, data structures, and visualization to improve the consistency, composibility, and discoverability of the API.
-To do so, this work first presents a mathematical description of how data is transformed into graphic representations, as shown in figure~\ref{fig:intro_artist_stages}.
-As with other mathematical formalisms of visualization \cite{mackinlayAutomatingDesignGraphical1986,kindlmann2014algebraic,sugibuchiFramwork2009,vickersUnderstandingViz2013}, a mathematical framework provides a way to formalize the properties and structure of the visualization.
-In contrast to the other formalisms, the model presented here is focused on the components that build a visualization rather than the visualization itself.
-
-In other words this model is not intended to be evaluative, it is intended to be a reference specification for visualization library API.
-To make this model as implementation independent as possible, we propose fairly general mathematical abstractions of the data container such that we do not need to assume the data has any specific structure, such as a relational database.
-We reuse this structure for the graphic as that allows us to specifically discuss how structure is preserved.
-We take a functional approach because functional paradigms encourage writing APIs that are flexible, concise and predictable due to the lack of side effects \cite{loudenProgrammingLanguagesPrinciples2010}.
-Furthermore, by structuring the API in terms of composition of the smallest units of transformation for which we can define correctness, a functional paradigm naturally leads to a library of highly modular components that are composable in such a way that by definition the composition is also correct.
-This allows us to ensure that domain specific visualizations built on top of these components are also correct without needing knowledge of the domain.
-As with the other mathematical formalisms of visualization, we factor out the rendering into a separate stage; but, our framework describes how these rendering instructions are generated.
-
-In this work, we present a framework for understanding visualization as equivariant maps between topological spaces.
-Using this mathematical formalism, we can interpret and extend prior work and also develop new tools.
-We validate our model by using it to re-design artist and data access layer of Matplotlib, a general purpose visualization tool.
+This work is strongly motivated by the needs of the Matplotlib\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library.
+One of the most widely used visualization libraries in Python, since 2002 new components and features have been added in a some what adhoc, sometimes hard to maintain, manner.
+Particularly, each new component carries its own implicit notion of how it believes the data is structured-for example if the data is a table, cube, image, or network - that is then expressed in the API for that component.
+In turn, this yields an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards\cite{a.sarikayaWhatWeTalk2019}.
+This entangling of data model with visual transform also yields inconsistencies in how visual component transforms, e.g.
+shape or color, are supported.
+We propose that these issues can be ameliorated via a redesign of the functions that convert data to graphics, named \textit{Artists} in Matplotlib, in a manner that reliably enforces \textit{continuity} and \textit{equivariance} constraints.
+We evaluate our functional model by implementing new artists in Matplotlib that are specified via \textit{equivariance} and \textit{continuity} constraints.
+We then use the common data model introduced by the model to demonstrate how plotting functions can be consolidated in a way that makes clear whether the difference is in expected data structure, visual component encoding, or the resulting graphic.
 
 \section{Background}
+The notion that visualization is structure preserving maps from data to visual representation is neither a new formalism nor a new implementation goal; this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library level to propose a new model that can be used to guide library development.
+
+\subsection{Structure: \textit{equivariance} \& \textit{continuity}}
+\label{sec:intro_viz}
+Visual representations of data, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.
+
+\begin{figure}[H]
+        \includegraphics[width=1\columnwidth]{figures/intro/equivariance.png}
+        \caption{The data in blue is scaled by a factor of two, yielding the data in orange.
+To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
+        \label{fig:intro:equivariance}
+\end{figure}
+Bertin proposes that there are classes of visual encodings-such as position, shape, color, and texture-that when mapped to from specific types of measurement, quantitative or qualitative, will preserve the properties of that measurement type.
+For example, in \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from blue to orange data and lines.
+The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}.
+The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations\cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}.
+Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups.
+As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitely include continuity.
+
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/intro/continuity.png}
+    \caption{The line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points.
+}
+    \label{fig:intro:continuity}
+\end{figure}
+Bertin proposes that the visual encodings be composited into graphical marks that match the \textit{continuity} of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark.
+In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval.
+\note{transition statement}
+
+\begin{mdframed}[roundcorner=10pt, frametitle=Structure, frametitlerule=true, frametitlebackgroundcolor=gray!10]
+    \begin{description}
+        \item[\textbf{continuity}] How records in the dataset are connected to each other, e.g. discrete rows, neworked nodes, points on a continuous surface
+        \item[\textbf{equivariance}] if an action is applied to the data or the graphic--e.g.
+a rotation, permutation, translation, or rescaling-- there must be an equivalent action applied on the other side of the transformation.
+    \end{description}
+\end{mdframed}
+
+The measure of how much of the structure of the data the graphic encodes is a concept Mackinlay termed expressiveness, while the graphic's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
+When the properties of the representation match the properties of the data, then the visualization is easier to understand according to Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis}.
+These ideas are combined into Tufte's notion of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements \cite{tufteVisualDisplayQuantitative2001}.
+
+\subsection{Tools}
+\label{sec:intro_data_tools}
+
 One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures.
@@
 
-\subsection{Tools}
-\label{sec:intro_data_tools}
+\begin{figure}[H]
+    \begin{subfigure}{.3\textwidth}
+        \includegraphics[width=2.75\textwidth]{figures/intro/table_only.png}
+    \end{subfigure}
+    \begin{subfigure}{.3\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
+    \end{subfigure}
+    \begin{subfigure}{.3\textwidth}
+        \includegraphics[width=1\textwidth]{figures/math/graph.png}
+    \end{subfigure}
+    \caption{Visualization libraries, especially ones tied to specific domains, tend to be architectured around a core data structure, such as tables, images, or networks. }
+\end{figure}
+
 One extensive family of relational table based libraries are those based on Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
@@
 
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
+    \caption{One way to describe data is by the connectivity of the points in the dataset.
+A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid.
+This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
+    \label{fig:intro_data_format}
+\end{figure}
+
 There are also visualization tools where there is no single core structure, and instead internally carry around many different representations of data.
@@
 Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects.
+VTK has explicitely codified this in terms of continuity based data representations, as illustrated in figure~\ref{fig:intro_data_format}.
 This in turn means that every new type of visualization must carry implicit assumptions about data structure in how it interfaces with the input data.
@@
 \begin{figure}[H]
-    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
-    \caption{One way to describe data is by the connectivity of the points in the dataset.
-A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid.
-This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
-    \label{fig:intro_data_format}
+    \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
+    \caption{Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
 \end{figure}
+
 As shown in figure~\ref{fig:intro_data_format}, there are many distinct ways of encoding each specific type of structure, while as mentioned in section~\ref{sec:intro_data_tools} APIs are clearer when structured around a common data representation.
@@
 
-\subsection{Visualization}
-\label{sec:intro_viz}
-\begin{figure}[H]
-\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
-\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the components of a visualization.
-The best to show column describes which types of information can be expressed in the corresponding visual encoding.
-This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from Krygier and Wood's \textit{Making Maps: A Visual Guide to Map Design for GIS}
-\cite{krygierMakingMapsVisual2005}}
-\label{fig:intro_retinal_variables}
-\end{figure}
-
-Visual representations of data, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
-The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}.
-As illustrated in figure~\ref{fig:intro_retinal_variables}, Bertin proposes that there are classes of visual encodings such as shape, color, and texture that when mapped to from specific types of measurement, quantitative or qualitative, will preserve the properties of that measurment type.
-For example, that nominal data mapped to hue preserves the selectivity of the nominal measurements.
-Furthermore he proposes that the visual encodings be composited into graphical marks that match the connectivity of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark.
-A general form of marks are glyphs, which are graphical objects that convey one or more attributes of the data entity mapped to it\cite{ware2019information, munznerVisualizationAnalysisDesign2014} and minimally need to be differentiable from other visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}.
-The set of encoding relations from data to visual representation is termed the graphical design by Mackinlay \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987} and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic.
-
-The measure of how much of the structure of the data the graphic encodes is a concept Mackinlay termed expressiveness, while the graphic's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
-When the proporties of the representation match the properties of the data, then the visualization is easier to understand according to Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis}.
-These ideas are combined into Tufte's notion of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements \cite{tufteVisualDisplayQuantitative2001}.
-This notion of matching is explictely formalized by Mackinlay as a structure preserving mapping of a binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}.
-A functional dependency framework for evaluating visualizations was proposed by Sugibuchi et al \cite{sugibuchiFramwork2009}, and an algebraic basis for visualization design and evaluation was proposed by Kindlmann and Scheideggar\cite{kindlmann2014algebraic}.
-Vickers et al.
-propose a category theory framework\cite{vickersUnderstandingViz2013} that extends structural preservation to layout, but is focused strictly on the design layer like the other mathematical frameworks.
-
-\begin{figure}[H]
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{figures/intro/du_bois_spinny.png}
-        \caption{}
-        \label{fig:intro_dpa}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{figures/intro/du_bois_line.png}
-        \caption{}
-        \label{fig:intro_dpb}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{figures/intro/du_bois_country.png}
-        \caption{}
-        \label{fig:intro_dbc}
-    \end{subfigure}
-    \begin{subfigure}{.24\textwidth}
-        \includegraphics[width=1\textwidth]{figures/intro/du_bois_heat.png}
-        \caption{}
-        \label{fig:intro_dbd}
-    \end{subfigure}
-    \caption{Du Bois' data portraits\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018} of post reconstruction Black American life exemplify that the fundemental characteristics of data visualization is that the visual elements vary in proportion to the source data.
-In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the line changes color to indicate a shift in the political environment; in figure~\ref{fig:intro_dbc} the countries are scaled to population size; and figure~\ref{fig:intro_dbd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
-The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroValuation1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
-    \label{fig:intro_dubois}
-\end{figure}
-
-One example of highly expressive visualizations are the data portraits by Du Bois shown in figure~\ref{fig:intro_dubois}.
-While the Du Bois charts are different from the usual scatter, line, and plot charts, they conform to the constraint that a graphic is a structure preserving map from data to visual representation.
-Figure~\ref{fig:intro_dpa} is semantically similar to a bar chart in that the lengths of the segments are mapped to the values, but in this chart the segments are stacked together.
-Figure~\ref{fig:intro_dpb} is a multicolored line chart where the color shifts are at periods of political significance.
-In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph.
-Figure~\ref{fig:intro_dpd} is simply a treemap\cite{heerTourVisualizationZoo2010} with space between the marks.
-Since the Du Bois data portraits meet the criteria of a faithful visual representation, we propose a mathematical framework and implementation that allows us to express the Du Bois charts and common chart types with equal fidelity.
-
 \subsection{Contribution}

### 2021-04-09 draft/sections/scraps.tex
reworked intro, start of reshuffling of background

--- 2021-01-25 draft/sections/scraps.tex
@@
-\begin{figure}[ht!]
-    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
-    \label{fig:data_monoid_rotation}
-    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.
-    %% dataset that is symmetric, fiber is x,y, rotatation action
-    %% rotate + plot or plot + rotate
-    %%emoji shuffle 
-    }  
+- [x] (1) Audience: lower-level visualization library developers, domain specific library developers (continuing the architecture choices mpl made)
+
+- [x] (3) Proposed solution: 
+    - [x] functional math model, structure/type preserving 
+-  [x] (4) Novelty:
+    - [x] functional architecture
+    - [x] explicitly expressing continuity in the model
+    - [x] model is constraint oriented rather than evaluative
+- [x] (5) Benefits: 
+    - [x] functional: robustness, concurrency, modularity, proves that reliably, backfit needs to this
+
+- [x] (2) Need: unified model: continuity + equivariance 
+    - [x] mpl works, unmaintainable, grew organically, this stuff is buried implicitly, gives a way to make it clear
+    - [x] interactivity screen space back to data space, different implicit data than API incoherancy than bug?
+        - [x] currently done in adhoc manner, want to enforce contracts for the interactivity 
+    - [x] redesign that reliably enforces these constraints 
+    - [x] library that supports all the data 
+        - [x] relational tables, and images, and data cubes (with coordinates), and networks, and other data structures
+        - [x] streaming and concurrency 
+        - x blob the same? - model explains why this should be true
+    - [x] framework to enforce invariance at the architecture level
+    - [x] match engineering w/ math
+
+- [ ] (6) Evaluation: The method of evaluation to verify that the proposed solution delivers the stated benefits must be given.
+        The best justifications explicitly discuss particular choices in the context of several possible alternatives.
+        1. Case studies in the form of implementations (is this feasible) 
+        2. Comparison w/ existing implementation
+
+- [ ] (7) More specific discussion of what will be done and on what timeline.
+        [quarterly gantt charts?]
+[] close out board https://github.com/story645/proposal/projects/3
+
+The work presented in this paper is motivated by a need for a library of visualization components that developers could use to build complex, domain specific tools tuned to the semantics and structure carried in domain specific data.
+While many researchers have identified and described important aspects of visualization, they have specialized in such different ways as to not provide a model general enough to natively support the full range of data and visualization types many general purpose modern visualization tools may need to support.
+The core architecture also needs to be robust to the big data needs of many visualization practitioners, and therefore support distributed and streaming data needs.
+To support both exploratory and confirmatory visualization\cite{tukeyWeNeedBoth1980}, this tool needs to support 2D and 3D, static, dynamic and interactive visualizations.
+
+Specifically, this work was driven by a rearchitecture of the Python visualization library Matplotlib\cite{hunterMatplotlib2DGraphics2007} to meet modern data visualization needs.
+We aim to take advantage of developments in software design, data structures, and visualization to improve the consistency, composibility, and discoverability of the API.
+To do so, this work first presents a mathematical description of how data is transformed into graphic representations, as shown in figure~\ref{fig:intro_artist_stages}.
+As with other mathematical formalisms of visualization \cite{mackinlayAutomatingDesignGraphical1986,kindlmann2014algebraic,sugibuchiFramwork2009,vickersUnderstandingViz2013}, a mathematical framework provides a way to formalize the properties and structure of the visualization.
+In contrast to the other formalisms, the model presented here is focused on the components that build a visualization rather than the visualization itself.
+
+In other words this model is not intended to be evaluative, it is intended to be a reference specification for visualization library API.
+To make this model as implementation independent as possible, we propose fairly general mathematical abstractions of the data container such that we do not need to assume the data has any specific structure, such as a relational database.
+We reuse this structure for the graphic as that allows us to specifically discuss how structure is preserved.
+We take a functional approach because functional paradigms encourage writing APIs that are flexible, concise and predictable due to the lack of side effects \cite{loudenProgrammingLanguagesPrinciples2010}.
+Furthermore, by structuring the API in terms of composition of the smallest units of transformation for which we can define correctness, a functional paradigm naturally leads to a library of highly modular components that are composable in such a way that by definition the composition is also correct.
+This allows us to ensure that domain specific visualizations built on top of these components are also correct without needing knowledge of the domain.
+As with the other mathematical formalisms of visualization, we factor out the rendering into a separate stage; but, our framework describes how these rendering instructions are generated.
+
+In this work, we present a framework for understanding visualization as equivariant maps between topological spaces.
+Using this mathematical formalism, we can interpret and extend prior work and also develop new tools.
+We validate our model by using it to re-design artist and data access layer of Matplotlib, a general purpose visualization tool.
+
+\subsection{Visualization}
+
+\begin{figure}[H]
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_spinny.png}
+        \caption{}
+        \label{fig:intro_dpa}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_line.png}
+        \caption{}
+        \label{fig:intro_dpb}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_country.png}
+        \caption{}
+        \label{fig:intro_dbc}
+    \end{subfigure}
+    \begin{subfigure}{.24\textwidth}
+        \includegraphics[width=1\textwidth]{figures/intro/du_bois_heat.png}
+        \caption{}
+        \label{fig:intro_dbd}
+    \end{subfigure}
+    \caption{Du Bois' data portraits\cite{duboiscenterattheuniversityofmassachusettsBoisDataPortraits2018} of post reconstruction Black American life exemplify that the fundemental characteristics of data visualization is that the visual elements vary in proportion to the source data.
+In figure~\ref{fig:intro_dpa}, the length of each segment maps to population; in figure~\ref{fig:intro_dpb}, the line changes color to indicate a shift in the political environment; in figure~\ref{fig:intro_dbc} the countries are scaled to population size; and figure~\ref{fig:intro_dbd} is a treemap where the area of the rectangle is representative of the number of businesses in each field.
+The images here are from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroValuation1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
+    \label{fig:intro_dubois}
 \end{figure}
 
-\begin{figure}[ht!]
-    \includegraphics[width=.5\linewidth]{figures/math/fiberbundle.png}
-
-    \caption{ The fiber $F$ is a continuous plane encoding the set of all possible (time, temperature) values.
-The base space $K$ expresses that the fibers lie in continous space and therefore the data is continuous.
-The mapping $\pi:E\rightarrow K$ from total space to base space is what defines this structure as a fiber bundle.
-The continuous function $\tau_1$ returns a record in each fiber for each point on $K$; the set of all values $\tau_1$ returns is a dataset that lies in this fiberbundle.
-\note{remove one of the sections from this figure/maybe kill this example}}
-    \label{fig:data_base_space} 
-\end{figure}
-The fiber bundle in figure~\ref{fig:data_base_space} is a representation of continuous (temperature, time) data.
-The fiber bundle abstraction facilitates seperately representing the time and temperature variables in the fiber $F$ and how these variables are connected in the base space $K$.
-The function $\tau_1$ returns a record (temperature, time) for ever point on the interval $K$ and there can be many $\tau$ functions in the fiber bundle.
+One example of highly expressive visualizations are the data portraits by Du Bois shown in figure~\ref{fig:intro_dubois}.
+While the Du Bois charts are different from the usual scatter, line, and plot charts, they conform to the constraint that a graphic is a structure preserving map from data to visual representation.
+Figure~\ref{fig:intro_dpa} is semantically similar to a bar chart in that the lengths of the segments are mapped to the values, but in this chart the segments are stacked together.
+Figure~\ref{fig:intro_dpb} is a multicolored line chart where the color shifts are at periods of political significance.
+In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph.
+Figure~\ref{fig:intro_dpd} is simply a treemap\cite{heerTourVisualizationZoo2010} with space between the marks.
+Since the Du Bois data portraits meet the criteria of a faithful visual representation, we propose a mathematical framework and implementation that allows us to express the Du Bois charts and common chart types with equal fidelity.

### 2021-04-12 draft/sections/intro.tex
intro + background

@@
 \label{sec:intro}
-Building block level visualization libraries implement the functions that transform data into some component of the visual representation, providing a collection of components and utilities that can be combined to create a visualization \cite{wongsuphasawatNavigatingWideWorld2021}.
-Often domain specific tools are built out of these building block libraries, meaning the building block libraries must provide components general enough to satisfy a wide variety of data and visual representation needs.
-Specifically they must satisfy confirmatory and exploratory visualization needs \cite{tukeyWeNeedBoth1980}, in scientific and information (formal and informal\cite{pousmanCasualInformation2007}) visualization domains.
-To satisfy ever developing visualization needs, new components can be incorporated into the library in sometimes ad-hoc ways that can lead to API incoherancy and inconsistent behavior.
-We propose that like physical building blocks, building block libraries should provide a collection of well defined pieces that can be composed in whichever ways the blocks fit together.\note{figure of what we mean by building block?} We specify that for a valid visualization block is a structure preserving transformation from data to visual space, and we define structure in terms of \textit{continuity} and \textit{equivariance}.
+\begin{figure}[h]
+    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
+    \caption{Visualizations are made up of transformations from data into visual representation.
+These functions are are both transformations of individual data values to visual representation, such as date to x position or latitude to y position, and the assembly of all these transformations into a visual mark, such as a line or point.
+The same variable can be mapped in different ways, for example line is mapped to a color in the scatter plot and to y position in the line plot.}
+    \label{fig:intro_artist}
+\end{figure}
+Visualization is the transformation of data into visual representation.
+As illustrated by \autoref{fig:intro_artist}, these translations are both at the level of the individual variable and the entire record.
+In the case of the scatter plot, the latitude and longitude are encoded as the x and y position, respectively, while the temperature and station are represented by the face and edge colors.
+A row in the table is collectively encoded as a point mark.
+None of these encodings are fixed, as evidenced by temperature being translated into the y value in the case of the line plot.
+The station is now the source of the color of the entire line, and the date is the x position.
+As with scatter, the encodings of the individual transformations, which again are on values from the same record in the table, are composited into a line mark.
+It is these raw transformations from data space to visualization space that are implemented by building block level visualization libraries, named as such because the functions provided by the library can be composited in any number of ways to yield visualizations \cite{wongsuphasawatNavigatingWideWorld2021}.
+We propose that like physical building blocks, building block libraries must provide a collection of well defined pieces that can be composed in whichever ways the blocks fit together.
+We specify that a valid visualization block is a structure preserving transformation from data to visual space, and we define structure in terms of \textit{continuity} and \textit{equivariance}.
 We then use this model to develop a design specification for the components of a building block visualization library.
@@
 \section{Background}
-The notion that visualization is structure preserving maps from data to visual representation is neither a new formalism nor a new implementation goal; this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library level to propose a new model that can be used to guide library development.
+There are many formalisms of the notion that visualization is structure preserving maps from data to visual representation, and many visualization libraries that attempt to preserve structure in some manner; this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library level to propose a new model of the constraints visual transformations must satisfy such that they can be composed to produce visualize representations that can be considered equivalent to the data being represented.
 
-\subsection{Structure: \textit{equivariance} \& \textit{continuity}}
+\subsection{Structure:}
 \label{sec:intro_viz}
@@
 \begin{figure}[H]
-        \includegraphics[width=1\columnwidth]{figures/intro/equivariance.png}
+        \includegraphics[width=1\columnwidth]{figures/intro/equivariant.png}
         \caption{The data in blue is scaled by a factor of two, yielding the data in orange.
@@
     \includegraphics[width=1\textwidth]{figures/intro/continuity.png}
-    \caption{The line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points.
-}
+    \caption{The line plot does not preserve \textit{continuity} because it implies that the average temperature at each station lie along a 1D continuous line, while the bar plot preserves \textit{continuity} by representing the average temperatures at each station as the discrete values they are.}
     \label{fig:intro:continuity}
@@
 In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval.
-\note{transition statement}
+But, when the continuity is preserved, as in the bar chart, then the graphic has not introduced new structure into the data.
 
@@
 
-The measure of how much of the structure of the data the graphic encodes is a concept Mackinlay termed expressiveness, while the graphic's effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}.
-When the properties of the representation match the properties of the data, then the visualization is easier to understand according to Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis}.
-These ideas are combined into Tufte's notion of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements \cite{tufteVisualDisplayQuantitative2001}.
+The notion that a graphic should be equivalent to the data has been expressed in a variety of ways.
+Informally, Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis} states that a visualization is easier to understand when the properties of the visualization match the properties of the data.
+This principal is made more concrete in Tufte's concept of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements\cite{tufteVisualDisplayQuantitative2001}.
+expressing, as defined by Mackinlay, is a measure how much of the mathematical structure in the data that can be expressed in the visualizations; for example that ordered variables can be mapped into ordered visual elements.
+We propose that a graphic is an equivalent representation of the data when \textit{continuity} and \textit{equivariance} are preserved.
 
@@
 
-One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures.
-This in turn restricts the design space of visual algorithms that display information (the visualization types the library supports) since the algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{ToryRethinkingVisualization2004}.
-In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.
-
 \begin{figure}[H]
     \begin{subfigure}{.3\textwidth}
-        \includegraphics[width=2.75\textwidth]{figures/intro/table_only.png}
+        \includegraphics[width=1\textwidth]{figures/intro/table.png}
     \end{subfigure}
@@
     \caption{Visualization libraries, especially ones tied to specific domains, tend to be architectured around a core data structure, such as tables, images, or networks. }
+    \label{fig:intro:data}
 \end{figure}
 
-One extensive family of relational table based libraries are those based on Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
-The restriction to tables in turn restricts the native design space to visualizations suited to tables.
-Since the data space and graphic space is very well defined in this grammar, it lends itself to a declarative interface \cite{heerDeclarative2010}.
-This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
-An example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that most computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color.
+One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures.
+These libraries can often assume the expected data structure because they are domain specific, and that is the common data structure in that domain.
+For users who generally work in one domain, such as the data, networks, or graphs shown in \autoref{fig:intro:data}, this well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a very coherant user experience\cite{heerDeclarative2010}.
+But, for developers who want to build new visualizations on top of these libraries, they must work around the existing assumptions, sometimes in ways that break the model the libraries are developed around.
+
+For example, many domain specific libraries integrate computation into the visualization, for example libraries based that assume all data is a relational database.
+This assumption is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, such as ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+Since these libraries represent all data as a table, and computations on tables are fairly well defined\cite{ullmanFirstCourseDatabase2008}, they can include computations on the table with a fair bit of confidence that the computation is accurate.
+Since most computations are specific to domains, general purpose block libraries can not make this assumption; instead a goal of this model is to identify which computations are specifically part of the visual encoding - for example mapping data to a color-and which are manipulations on the data.
 Disentangling the computation from the visual transforms allows us to determine whether the visualization library needs to handle them or if they can be more efficiently computed by the data container.
@@
 Even the digital humanities oriented ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supports some non-image aggregate reporting charts, is still built around image data as the primary input.
+The need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}.
+As with tables and images, extending network libraries to work with other types of data either require breaking their internal model of how data is structured and what transformations of the data are allowable or growing a model for other types of data structures alongside the network model.
 
@@
 \end{figure}
-
-There are also visualization tools where there is no single core structure, and instead internally carry around many different representations of data.
-Matplotlib, has this structure, as does VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{RamachandranMayaVI2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}.
-Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects.
-VTK has explicitely codified this in terms of continuity based data representations, as illustrated in figure~\ref{fig:intro_data_format}.
-This in turn means that every new type of visualization must carry implicit assumptions about data structure in how it interfaces with the input data.
-This has lead to poor API consistency and brittle code as every visualization type has a very different point of view on how the data is structured.
-This API choice particularly breaks down when the same dataset is fed into visualizations with different assumptions about structure or into a dashboard consisting of different types of visualization\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007} because there is no consistent way to update the data and therefore no consistent way of guaranteeing that the views stay in sync.
-Our model is a structure dependent formalism, but then also provides a core representation of that structure that is abstract enough to provide a common interface for many different types of visualization.
+Many building block libraries carry multiple models of data internally because they cannot assume a data structure.
+Algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{ToryRethinkingVisualization2004}, and by definition building block libraries try to  provide the components to build any sort of visualization.
+Matplotlib, D3\cite{bostockDataDrivenDocuments2011}, and VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{RamachandranMayaVI2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}.
+Where GoG and ImageJ type libraries have coherant APIs for their visualization tools because the data structure is the same, the APIs for visualizations in Matplotlib, D3, and VTK are significantly dependent on the structure of the data it expects.
+VTK has codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro_data_format}.
+This API choice can lead to visualizations that break \textit{continuity} when fed into visualizations with different assumptions about structure.
+The lack of consistent data model can also mean no consistent way of updating the data and therefore no way of guaranteeing that the views are in sync, in visualizations that consistent of multiple views of the same datasource, such as dashboards\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007}.
+To resolve this issue, our functional model takes as input a structure aware data abstraction general enough to provide a common interface for many different types of visualization.
 
@@
 \label{sec:intro_data}
-Discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}, so one of the drivers of this work was to facilitate building libraries that could natively support domain specific data containers that do not make assumptions about data continuity.
-As shown in figure~\ref{fig:intro_data_format}, there are many types of connectivity.
-A database typically consists of unconnected records, while an image is an implicit 2D grid and a network is some sort of explicitly connected graph.
-These data structures typically contain not only the measurements or values of the data, but also domain specific semantic information such as that the data is a map or an image that a modern visualization library could exploit if this information was exposed to the API.
-
-\begin{figure}[H]
-    \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
-    \caption{Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
-\end{figure}
-
-As shown in figure~\ref{fig:intro_data_format}, there are many distinct ways of encoding each specific type of structure, while as mentioned in section~\ref{sec:intro_data_tools} APIs are clearer when structured around a common data representation.
-Fiber bundles were proposed by Butler as one such representation because they encode the continuity of the data separately from the types of variables and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-Since Butler's model lacks a robust way of describing variables, we fold in Spivak's Simplicial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
+One such general abstraction are fiber bundles, which Butler proposed as a core data structure for visualization because they encode data continuity separately from the variable properties and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Since Butler's model lacks a robust way of describing variables, we can encode a schema like description of the data in the fiber bundle by folding in Spivak's topological description of data types \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}.
 In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements.
@@
 
-\begin{mdframed}[roundcorner=10pt, frametitle= definitions, frametitlerule=true, frametitlebackgroundcolor=gray!10]
-    \begin{description}
-        \item[\textbf{records}] points, observations, entries 
-        \item[\textbf{components}] variables, attributes, fields 
-        \item[\textbf{connectivity}] how the records are connected to each other
-    \end{description}
-\end{mdframed}
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
+    \caption{Values in a dataset have keys associated with them that describe where the value is in the dataset.
+These keys can be indexers or semantically meaningful;for example, in a table the keys are the variable name and the row ID.
+In the data cube,  the keys is the row, column, and cell ID, and in the map the key is the position in the grid.
+Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
+    \label{fig:intro:keys_values}
+\end{figure}
 
-Often this topology has metadata associated with it, describing for example dependent variables and the independent variables they are dependent on, or information about the structure of the data such as when and where the measurement was taken.
-Building on the idea of metadata as \textit{keys} and their associated \textit{values} proposed by Munzner \cite{munznerChDataAbstraction}, we propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids.
+The \textit{continuity} can often be described by some variables in the dataset; this is formalized Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}.
+As shown in \autoref{fig:intro:keys_values}, keys can be labeled indexes, such as the attribute name and row ID, or physical entities such as locations on a map.
+We propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids.
 In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology.
-This allows the metadata to be altered, for example by changing the coordinate systems or time resolution, without imposing new semantics on the underlying structure.
+This allows the metadata to be altered, without imposing new semantics on the underlying structure, for example by changing the coordinate systems or time resolution.
+This value agnostic model also supports encoding datasets where there may be multiple independent variables - such as a measure of plant growth given variations in water, sunlight, and time - without having to assume any one variable is inducing the change in growth.
+For building block library developers, this means the components are able to fully traverse the data structures without having to know anything about the values or the semantic meaning of the structure.
+Since these components are by design \textit{equivariant} and \textit{continuity} preserving, domain specific library developers in different domains that both rely on the same continuity, for example 2D continuity, can then safely reuse the components to build tools that can safely make domain specific assumptions.
 
 \subsection{Contribution}
-This work presents a mathematical model of the transformation from data to graphic representation and a proof of concept implementation. Specifically, the contributions of this work are 
+The contribution of this work is 
 \begin{enumerate}
-    \item a formal description of the topology preserving relationship between data and graphic via continuous maps
-    \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
-    \item abstraction of data structure using fiber bundles with schema like fibers to encode components and topology 
-    \item algebraic sum operator such that more complex visualizations can be built from simple ones 
-    \item a functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
-    \item a prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
+  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:graphic_base}
+  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:artist_nu}
+  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:artist_q}
+  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code_artist}
 \end{enumerate}
-In contrast to mathematical models of visualization that aim to evaluate visualization design, we propose a topological framework for building tools to build visualizations.
-We defer judgement of expressivity and effectiveness to developers building domain specific tools, but provide them the framework to do so.
 \end{document}

### 2021-04-13 draft/sections/intro.tex
figure caption

@@
     \caption{Visualizations are made up of transformations from data into visual representation.
-These functions are are both transformations of individual data values to visual representation, such as date to x position or latitude to y position, and the assembly of all these transformations into a visual mark, such as a line or point.
+These functions transform individual data values to visual representation, such as date to x position or latitude to y position.
+These functions are composed into the assembly of all these transformations into a visual mark, such as a line or point.
+%%% break this out into seperate sentences to break
     The same variable can be mapped in different ways, for example line is mapped to a color in the scatter plot and to y position in the line plot.}

### 2021-04-13 draft/sections/math.tex
reworking intro sentence lots of going back to feedback & trying to assimilate it

@@
 \label{sec:tam}
-
-As discussed in the introduction, visualization is generally defined as structure preserving maps from a data object to a graphic object.
-In order to formalize this statement, we describe the connectivity of the records using topology and define the structure on the components in terms of the monoid actions on the component types.
-By formalizing structure in this way, we can evaluate the extent to which a visualization preserves the structure of the data it is representing and build structure preserving visualization tools.
+To guide the implementation of structure preserving building block components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}.
 We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map from data to graphic

### 2021-04-13 draft/sections/math.tex
reworked intro paragraph

@@
 To guide the implementation of structure preserving building block components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}.
-We introduce the notion of an artist $\mathscr{\vartist}$ as an equivariant map from data to graphic
+Inspired by the somewhat analogous component in Matplotlib\cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic that these building block components implement the \textit{artist}.
 \begin{equation}
@@
 \end{equation}
-that carries a homomorphism of monoid actions $\varphi: \monoid \rightarrow \monoid^{\prime}$ \cite{cegarraCohomologyMonoidsOperators2019}, which are discussed in detail in section~\ref{sec:data_monoid}.
-Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps
-\begin{equation}
-\mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
-\end{equation}
-such that applying a monoid action $m \in \monoid$ to the data $\delement \in \mathscr{\dtotal}$ input to $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.
-
-We model the data $\mathscr{\dtotal}$, graphic $\mathscr{\gtotal}$, and intermediate visual encoding $\mathscr{\vtotal}$ stages of visualization as topological structures that encapsulate types of variables and continuity.
-To explain which structure the artist is preserving, we first describe how we model data (\ref{sec:data}), graphics (\ref{sec:graphic}), and intermediate visual characteristics (\ref{sec:artist}) as fiber bundles.
-We then discuss the equivariant maps between data and visual characteristics (\ref{sec:artist_nu}) and visual characteristics and graphics (\ref{sec:artist_q}) that make up the artist.
+The \textit{artist} $\mathscr{\vartist}$ is a map from the data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$ fiber bundles.
+To explain how the \textit{artist} is a structure preserving map from data to graphic, we first describe how we model data (\ref{sec:data}) and graphics (\ref{sec:graphic}) as topological structures that encapsulate component types and continuity.
+We then discuss the maps from graphic to data (\ref{sec:graphic_base}, data components to visual components (\ref{sec:artist_nu}), and visual components into graphic (\ref{sec:artist_q}) that make up the artist.
 

### 2021-04-14 draft/sections/math.tex
dbase figure

@@
     \includegraphics[width=1\textwidth]{figures/math/k_different_types.png}
-    \caption{The topological base space \dbase\ encodes the connectivity of the data space, for example if the data is independent points or on a plane or a sphere}
+    \caption{The topological base space \dbase\ encodes the continuity of the data space, for example if the data is discrete points or lies on a plane or a sphere}
     \label{fig:base_space_types}
@@
 \begin{figure}[H]
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-        \caption{}
-        \label{fig:base_example_discrete}
-    \end{subfigure}
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_2f.png}
-        \caption{}
-        \label{fig:base_example_continuous}
-    \end{subfigure}
-    \caption{These two datasets have the same (time, temperature) fiber.
-In figure~\ref{fig:base_example_discrete} the total space \dtotal\ is discrete over points $\dbasepoint \in \dbase$, meaning the records in the fiber are also discrete.
-In figure~\ref{fig:base_example_continuous} \dtotal\ lies over the continuous interval \dbase, meaning the records in the fiber are sampled from a continuous space.}
+    \includegraphics[width=1\textwidth]{figures/math/base.png}
+    \caption{These two datasets have the same (time, temperature) fiber, but different continuities.
+The dataset on the left consists of discrete records, while the records in the dataset on the right sampled from a continuous space.}
     \label{fig:base_example}
@@
 The datasets in figure~\ref{fig:base_example} have the same fiber of (temperature, time).
-In figure~\ref{fig:base_example_discrete} the fibers lie over discrete \dbase\ such that the records in the datasets in the fiber bundles are discrete.
-The same fiber in figure~\ref{fig:base_example_continuous} lies over a continuous interval \dbase\ such that the records are samples from a continuous function defined on \dbase.
+The dot represents a discrete base space \dbase, meaning that every dataset encoded in the fiber bundle has discrete continuity.
+The line is a representation of a 1D continuity, meaning that every dataset in the fiber bundle is 1D continous.
 By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.

### 2021-04-14 draft/sections/math.tex
fiber

@@
 \begin{figure}[H]
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_1k.png}
-        \caption{}
-        \label{fig:fiber_example_plane}
-    \end{subfigure}
-    \begin{subfigure}{.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/math/temp_3f.png}
-        \caption{}
-        \label{fig:fiber_example_cube}
-    \end{subfigure}
-    \caption{These two datasets have the same base space \dbase\ of discrete points, but figure~\ref{fig:fiber_example_plane} has fiber  $\dfiber=\reals\times\reals$ which is (time, temperature) while figure~\ref{fig:fiber_example_cube} has fiber $\reals\times\realsp\times\reals$ which is (time, wind=(speed, direction))}
+    \includegraphics[width=1\textwidth]{figures/math/fiber.png}
+    \caption{These two datasets have the same continuity, but different components and therefore different fibers.
+The 2D fiber \(\dfiber=\reals\times\reals\) encodes a dataset consisting of \textit{time} and \textit{temperature} components.
+One dimension of the fiber encodes the range of possible values for the time component of the dataset, which is a subset of the \reals, while the other dimension encodes the range of possible values  \reals\ for the temperature component.
+The 3D fiber encodes a dataset consisting of\textit{temperature}, \textit{latitude}, and \textit{longitude} components.}
     \label{fig:data_fiber_example}
@@
 
-For example, the data in figure~\ref{fig:fiber_example_plane} is a pair of times and \textdegree K temperature measurements taken at those times.
+For example, the records in the 2D fiber in \autoref{fig:data_fiber_example} are a pair of times and \textdegree K temperature measurements taken at those times.
 Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$.
@@
 The fiber is
-\begin{equation}
-    \ftotal =  \reals \times \realsp 
-\end{equation} 
-where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=temperature,\, \ftype=\texttt{float},\, \fttype=\realsp)$ and is the set of values $\fttype=\realsp$.
-In figure~\ref{fig:fiber_example_cube}, temperature is replaced with wind.
-This wind variable is of type \texttt{wind} and has two components speed and direction $\{(s,d) \in \reals^{2} \mid  0\leq s,\, 0 \leq d \leq 360\}$ .
-Therefore, the fiber is
-\begin{equation}
-    \dfiber = \realsp \times \reals^{2}
-\end{equation} 
-such that $F_1$ is specified by $(\fname = wind,\, \ftype = \texttt{wind},\, \fttype=\reals^{2})$.
+\begin{equation*}
+    \dfiber =  \reals \times \realsp 
+\end{equation*} 
+where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=temperature,\, \ftype=\texttt{float},\, \fttype=\reals)$ and is the set of values $\fttype=\reals$.
+In the 3D fiber in \autoref{fig:fiber_example_cube}, time is replaced with location.
+This location variable is of type \texttt{point} and has two components \textit{latitude} and \textit{longitude} $\{(lat,lon) \in \reals^{2} \mid  -90\leq lat \leq 90,\, 0 \leq lon \leq 360\}$.
+The fiber for this dataset is
+\begin{equation*}
+    \dfiber = \reals \times \reals^{2} = \reals \times \reals \times \reals
+\end{equation*} 
+where the dimensionality of the fiber does not change, but the components of the fiber can be coupled.
+For example, \textit{location} can can either be specified as \((\fname = location,\, \ftype = \texttt{point},\, \fttype=\reals^{2})\)or \((\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\reals)\) and \((\fname=longitude,\, \ftype=\texttt{float},\,\fttype=\reals)\).
+
 As illustrated in figure~\ref{fig:data_fiber_example}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.

### 2021-04-14 draft/sections/intro.tex
action figures, graphic space figures, updating all the q figures, plus figure + words in monoid section

@@
 \subsection{Structure:}
-\label{sec:intro_viz}
+\label{sec:intro:structure}
 Visual representations of data, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.

### 2021-04-14 draft/sections/math.tex
action figures, graphic space figures, updating all the q figures, plus figure + words in monoid section

@@
 Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-For example, given an arbitrary interval scale fiber component $(\fname=temperature,\, \ftype=\texttt{float},\,\fttype=\reals)$ with with arbitrary monoid translation actions chosen for this example:
-\begin{itemize}
-    \item monoid operator addition $\ast = +$
-    \item monoid operations: $f: x\mapsto x + 1\degree C $, $g: x\mapsto x + 2\degree C$
-    \item monoid action operator composition $\bullet = \circ$
-\end{itemize}
-By structure preservation, we mean that monoid actions are composable. For the translation actions described above on the temperature fiber, this means that they satisfy the condition
-\begin{equation}
-    \begin{tikzcd}
-        \reals \arrow[rd, "(x+ 1\degree C)\circ(x+2\degree C)"] \arrow[d, "x+ 1\degree"'] &            \\
-        \reals \arrow[r, "x+ 2\degree C"']                                   & \reals
-    \end{tikzcd}
-\end{equation}
-where $1\degree C$ and $2\degree C$ are valid distances between two temperatures $x$.
-What this diagram means is that either the fiber could be shifted by $1 \degree C$ (vertical line) then by $2 \degree C$ (horizontal), or the two shifts could be combined such that in this case the fiber is shifted by $3 \degree C$ (diagonal) and these two paths yield the same temperature.
+For example, lets say one component of the data is weather states, such that the fiber component is $\dfiber_i=\{sun,\, clouds,\, rain,\, lightning,\, storm\}$.
+If the data is ordinal, that means some ordering action is applied to the set.
 
-While many component types will be one of the measurement scale types, we generalize to monoids specifically for the case of partially ordered set.
-Given a set $W=\{mist, \, drizzle, \, rain \}$, then the map $f: W\rightarrow W$ defined by
-\begin{enumerate}
-    \item $f(rain) = drizzle$,
-    \item  $f(drizzle) = mist$ 
-    \item $f(mist) = mist$
-\end{enumerate}
-is order preserving such that $mist \leq drizzle \leq rain$ but has no inverse since $drizzle$ and $mist$ go to the same value $mist$.
-Therefore order preserving maps do not form a group, and instead we generalize to monoids to support partial order component types.
-Defining the monoid actions on the components serves as the basis for identifying the invariance\cite{kindlmann2014algebraic} that must be preserved in the visual representation of the component.
-We propose equivariance of monoid actions individually on the fiber to visual component maps and on the graphic as a whole.
+\begin{figure}[H]
+\begin{subfigure}{.5\textwidth}
+    \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
+    \caption{The identity action $\bullet$ maps the weather states to themselves.}
+    \label{fig:monoid:total}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
+    \caption{The action $\bullet$ is a monotonic map, meaning that it preserves the relative ordering despite mapping multiple elements to the same element.}
+    \label{fig:monoid:partial}
+\end{subfigure}
+\end{figure}
+\note{I think there's something about associativity here, but I'm not sure what}
+ In \autoref{fig:monoid:partial}, the action $\bullet$ acts on the weather states such that the states that they are mapped into have the same relative ordering as the original states.
+Unlike group actions, the action $\bullet$ is not invertable and therefore a monoid action.
+We generalize to monoids specifically for the case of partially ordered scales are common in indicator rankings \cite{bruggemannRankingPrioritizationMultiindicator2011}.
 
@@
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/equivariance_nu.png}
+    \includegraphics[width=\textwidth]{figures/math/partial_fixed.png}
     \caption{In this artist, \vchannel\ maps the strings to the emojis.
@@
 \begin{figure}[H]
-    \includegraphics[width=1\textwidth]{figures/math/scatter.png}
+    \includegraphics[width=1\textwidth]{figures/math/scatter_with_s.png}
     \caption{The data is discrete points (temperature, time).
@@
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/line.png}
+    \includegraphics[width=\textwidth]{figures/math/line_with_s.png}
     \caption{The line fiber $(time,\, temp)$ is thickened with the derivative $(time^{\prime},\, temperature^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.
@@
 \subsubsection{Composition of Artists: +}
-%%\begin{figure}[H]
-%%   \includegraphics[width=1\textwidth]{figures/math/addition_operator.png}
-%%  \caption{$\dsection_1$ and $\dsection_2$ are distinct datasets passed through artists $\vartist_1$ and $\vartist_2$ to generate graphics $\gsection_1$ and $\gsection_2$.
-These graphics happen to be rendered to the same image, but otherwise have no intrinsinc link.}
-%%    \label{fig:artist_plus}
-%%\end{figure}
+\begin{figure}[H]
+\begin{subfigure}{.5\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/exploding_artist.png}
+    \caption{Each component of a visualization is generated via a different artist.}
+    \label{fig:seperate_artists}    
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/combined_artist.png}
+    \caption{This visualization is the disjoint union of the artists in \autoref{fig:seperate_artists}.}
+    \label{fig:combined_artists}
+\end{subfigure}
+\caption{}
+\label{fig:artist_plus}
+\end{figure}
 To build graphics that are the composites of multiple artists, we define a simple addition operator that is the disjoint union of fiber bundles \dtotal.

### 2021-04-15 draft/sections/scraps.tex
intro + background
figure overhaul

@@
 
-- [ ] (6) Evaluation: The method of evaluation to verify that the proposed solution delivers the stated benefits must be given.
+- [x] (6) Evaluation: The method of evaluation to verify that the proposed solution delivers the stated benefits must be given.
         The best justifications explicitly discuss particular choices in the context of several possible alternatives.
@@
 Since the Du Bois data portraits meet the criteria of a faithful visual representation, we propose a mathematical framework and implementation that allows us to express the Du Bois charts and common chart types with equal fidelity.
+
+Associativitly means (a*b)*c = a*(b*c), as you know.
+Replace * with whatever binary operator you like.
+Same thing.
+Key one for CS/cat theory is when a, b, c are functions and * is composition.
+When you have something that seems like it always holds (like associativity does for function composition and most operations you can think of), it can be confusing because it is like ....
+why do you need to check this?
+it always has to be true.
+It turns out that it isn't for some binary operators that can be useful but we just don't see unless we are making general statements.
+But when you are proving something, it is important step in the proof.
+Part of the fact that the composition of equivariant functions eg.
+\nu and Q is equivariant, relies on the associativity.
+It just seems so obvious that you don't notice it, but if things wheren't associative ....
+suddenly all kinds of equations you are writing down and moving parenthesis, would either NOT be true, or be much more complicated.
+one way to write equivariance (for a group at least) as invariance is
+g^-1*\nu(g*\tau) = \nu(\tau)
+just imagine if order of operations between function call and g* mattered... it is hard because function composition IS associative but if it wasn't you couldn't just write that.
+There is an algebra called the octonians which is NOT associative but has a more complicated function so you can relate a*(b*c) with (a*b)*c with a isomorphic map (just isn't equal).
+Every equation becomes a mess.
+So the bottom line is that in a definition or a proof you might have to whip out associativity, but most of the time it does "just hold".
+It is mentioned when needed, but it isn't a serious constrain like continutity or equivariance ...
+it mostly "just holds".
+
+monoid -> set w/ associativity and identity
+
+set = R, f: R->R, 
+composotion of functions from a space to itself, is associative (\compo)

### 2021-04-16 draft/sections/math.tex
compose + equiv sections

@@
     \includegraphics[width=1\textwidth]{figures/math/exploding_artist.png}
-    \caption{Each component of a visualization is generated via a different artist.}
-    \label{fig:seperate_artists}    
+    \caption{}
+    \label{fig:math:seperate}    
 \end{subfigure}
@@
     \includegraphics[width=1\textwidth]{figures/math/combined_artist.png}
-    \caption{This visualization is the disjoint union of the artists in \autoref{fig:seperate_artists}.}
-    \label{fig:combined_artists}
+    \caption{}
+    \label{fig:math:combined}
 \end{subfigure}
-\caption{}
+\caption{Each of the visual elements in \autoref{fig:math:seprate} is generated via a unique artist \vartist.
+In \autoref{fig:math:seperate}, they are added to the image independent of the other elements, creating an incoherant visualization.
+In \autoref{fig:math:combined}, these artists are composited before being added to the image.
+Disjoint union of \dtotal\ aligns the two timeseries with the x and y axis so all these elements use a shared coordinate system.
+A more complex composition dictates that the legend is connected to the \dtotal\ such that it must use the same color as the data it is identifying.}
 \label{fig:artist_plus}
 \end{figure}
-To build graphics that are the composites of multiple artists, we define a simple addition operator that is the disjoint union of fiber bundles \dtotal.
-For example, a scatter plot $\dtotal_1$ and a line plot $\dtotal_2$ have different \dbase that are mapped to separate \gbase.
-To fully display both graphics, the composite graphic $\vartist_1 + \vartist_2$ needs to include all records on both $\dbase_1$ and $\dbase_2$, which are the sections on the disjoint union $\dbase_1 \sqcup \dbase_2$.
-This in turn yields disjoint graphics $\gbase_1 \sqcup \gbase_2$ rendered to the same image.
-Constraints can be placed on the disjoint union such as that the fiber components need to have the same \vchannel position encodings or that the position \vsection need to be in a specified range.
-There is a second type of composition where $\dtotal_1$ and $\dtotal_2$ share a base space $\dbase_2 \hookrightarrow \dbase_1$ such that the the artists can be considered to be acting on different components of the same section.
-This type of composition is important for creating visualizations where elements need to update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.
+Visualizations with a single artist do not provide much information, so we define addition operators for generating more complex visualizations.
+Given the family of artists $(\dtotal_i: i\in I)$ on the same image, the + operator
+\begin{equation}
++ \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
+\end{equation}
+defines a simple composition of artists.
+For example, the components in \autoref{fig:math:seperate} are each generated by different artists, and a visualization of solely the x axis is rarely all that useful.
+In \autoref{fig:math:seperate}, these artists are all added to the image independently of the other and therefore there are no constraints on how they are generated in conjunction with each other.
+In \autoref{fig:math:combined}, the data is joined via disjoint union; doing so aligns the components in \dfiber\ such the \vchannel\ to the same component in \vfiber\ targets the same coordinate system.
+When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the artists are acting on different components of the same section.
+This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.
 
@@
 \label{sec:artist_equivalance}
-It is impractical to implement an artist for every single graphic; instead we implement an approximation of an the equivalence class of artists $\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}$.
-Roughly, equivalent artists have the same fiber bundle \vtotal\ and same assembly function \vmark\ but act on different sections \vsection, but we will formalize the definition of the equivalance class in future work.
-As a first pass for implementation purposes, we identify a minimal \vfiber\ associated with each $\vartist^{\prime}$ that defines what visual characteristics of the graphic must originate in the data such that the graphic is identifiable as a given chart type.
-%%
-%%\begin{figure}[H]
-%%   \caption{Each of these graphics is generated by a different artist \vartist which is the equivalance class of scatter plots $\vartist^{\prime}$\note{this is gonna be a whole bunch of scatter plots}}
-%%    \label{fig:artist_equivalence}
-%%\end{figure}
-
-For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another.
-These two artists are equivalent since their only difference is in the literal visual encodings (color, shape).
-Shape and color could also be defined in \vmark\, but the position must come from the fiber $\vfiber=(xpos,ypos)$ since fundementally a scatter plot is the plotting of one position against another\cite{friendlyBriefHistoryData2008}.
-We also use this criteria to identify derivative types, for example the bubble chart\cite{tufteVisualDisplayQuantitative2001} is a type of scatter where by definition the glyph size is mapped from the data.
-The criteria for equivalence class membership serves as the basis for evaluating invariance\cite{kindlmann2014algebraic}.
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/math/equivalent_artists.png}
+   \caption{Each scatter plot is generated via a unique artist function $\vartist_{i}$, but they only differ in aesthetic styling.
+Therefore, these artists are all members of an equivalence class $\vartist_{i} \in \vartist{\prime}$}
+    \label{fig:math:artist:equivalence}
+\end{figure}
+Representational invariance, as defined by Kindlmann and Scheidegger, is the notion that visualizations are equivalent if changing the visual representation, such as colors or shapes, does not change the meaning of the visualization\cite{kindlmannAlgebraicProcessVisualization2014}.
+We propose that visualizations are invariant if they are generated by artists that are members of an equivalence class
+\begin{equation*}
+\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}
+\end{equation*}
+For example, every scatter plot in \autoref{fig:math:artist:equivalence} is a scatter of the same datasets mapped to the \textit{x position} and \textit{y position} in the same way.
+The scatter plots only differ in the choice of constant visual literals, differing in color and marker shape.
+Each scatter is generated by an artist $\vartist_{i}$, and every scatter is generated by a member of the equivalence class $\vartist_{i} \in \vartist{\prime}$.
+Since it is impractical to implement a new artist for every single graphic, the equivalance class provides a way to evaluate an implementation of a generalized artist.
+Given equivalent, but no necessarily identical, \vchannel, \vmark, and \vindex, two artists are equivalent.
+This criteria also allows for comparing artists across libraries.
 \end{document}

### 2021-04-16 draft/sections/intro.tex
data->monoid sections + reference renaming

@@
     The same variable can be mapped in different ways, for example line is mapped to a color in the scatter plot and to y position in the line plot.}
-    \label{fig:intro_artist}
+    \label{fig:intro:artist}
 \end{figure}
 Visualization is the transformation of data into visual representation.
-As illustrated by \autoref{fig:intro_artist}, these translations are both at the level of the individual variable and the entire record.
+As illustrated by \autoref{fig:intro:artist}, these translations are both at the level of the individual variable and the entire record.
 In the case of the scatter plot, the latitude and longitude are encoded as the x and y position, respectively, while the temperature and station are represented by the face and edge colors.
@@
 \subsection{Tools}
-\label{sec:intro_data_tools}
+\label{sec:intro:data:tools}
 
@@
 This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
-    \label{fig:intro_data_format}
+    \label{fig:intro:data:format}
 \end{figure}
@@
 Where GoG and ImageJ type libraries have coherant APIs for their visualization tools because the data structure is the same, the APIs for visualizations in Matplotlib, D3, and VTK are significantly dependent on the structure of the data it expects.
-VTK has codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro_data_format}.
+VTK has codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}.
 This API choice can lead to visualizations that break \textit{continuity} when fed into visualizations with different assumptions about structure.
@@
 Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
-    \label{fig:intro:keys_values}
+    \label{fig:intro:keys:values}
 \end{figure}
@@
 The \textit{continuity} can often be described by some variables in the dataset; this is formalized Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}.
-As shown in \autoref{fig:intro:keys_values}, keys can be labeled indexes, such as the attribute name and row ID, or physical entities such as locations on a map.
+As shown in \autoref{fig:intro:keys:values}, keys can be labeled indexes, such as the attribute name and row ID, or physical entities such as locations on a map.
 We propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids.
@@
 \begin{enumerate}
-  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:graphic_base}
-  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:artist_nu}
-  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:artist_q}
-  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code_artist}
+  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
+  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist:nu}
+  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
+  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:math:code:artist}
 \end{enumerate}

### 2021-04-16 draft/sections/math.tex
data->monoid sections + reference renaming

@@
 The \textit{artist} $\mathscr{\vartist}$ is a map from the data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$ fiber bundles.
-To explain how the \textit{artist} is a structure preserving map from data to graphic, we first describe how we model data (\ref{sec:data}) and graphics (\ref{sec:graphic}) as topological structures that encapsulate component types and continuity.
-We then discuss the maps from graphic to data (\ref{sec:graphic_base}, data components to visual components (\ref{sec:artist_nu}), and visual components into graphic (\ref{sec:artist_q}) that make up the artist.
+To explain how the \textit{artist} is a structure preserving map from data to graphic, we first describe how we model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as topological structures that encapsulate component types and continuity.
+We then discuss the maps from graphic to data (\autoref{sec:math:graphic:base}, data components to visual components (\autoref{sec:math:artist:nu}), and visual components into graphic (\autoref{sec:math:artist:q}) that make up the artist.
 
 \subsection{Data Space \dtotal}
-\label{sec:data}
+\label{sec:math:data}
 Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
@@
 \end{equation}
-that binds the components of the data in \dfiber\ to the continuity represented in \dbase.
-The fiber bundle models the properties of data component types \dfiber\ (\ref{sec:data_fiber}), the continuity of records \dbase\ (\ref{sec:data_base}), the collections of records \dsection\ (\ref{sec:data_section}), and the space \dtotal\ of all possible datasets with these components and continuity.
-
-By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on \dtotal\ and focus on the components and continuity.
-We use fiber bundles as the data model because they are inclusive enough to express all the types of data described in section~\ref{sec:intro_data}.
+that binds the components of the data in \dfiber\ to the continuity of the data encoded in \dbase.
+The fiber bundle models the properties of data component types \dfiber\ (\autoref{sec:math:data:fiber}), the continuity of records \dbase\ (\autoref{sec:math:data:base}), the collections of records \dsection\ (\autoref{sec:math:data:section}), and the space \dtotal\ of all possible datasets with these components and continuity.
+By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartesian product $\dbase\times\dfiber$.
+We use fiber bundles as the data model because they are inclusive enough to express all the types of structures of data described in \autoref{sec:intro:data:tools}
 
 \subsubsection{Variables in Fiber Space \dfiber}
-\label{sec:data_fiber}
+\label{sec:math:data:fiber}
 To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names.
@@
 \end{equation}
-where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype.
-A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
+where \fsection\ is the schema that binds a variable name \fname\ to its datatype \ftype.
+A dataset with multiple components has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
 \begin{equation}
@@
     \includegraphics[width=1\textwidth]{figures/math/fiber.png}
-    \caption{These two datasets have the same continuity, but different components and therefore different fibers.
-The 2D fiber \(\dfiber=\reals\times\reals\) encodes a dataset consisting of \textit{time} and \textit{temperature} components.
+    \caption{The fiber space is the set of all combinations of all theoretically possible values of the components.
+The 2D fiber \(\dfiber=\reals\times\reals\) encodes the properties of \textit{time} and \textit{temperature} components.
 One dimension of the fiber encodes the range of possible values for the time component of the dataset, which is a subset of the \reals, while the other dimension encodes the range of possible values  \reals\ for the temperature component.
-The 3D fiber encodes a dataset consisting of\textit{temperature}, \textit{latitude}, and \textit{longitude} components.}
-    \label{fig:data_fiber_example}
+This means the fiber is the set of points \textit{(temperature, time)}that are all the combinations of \textit{temperatature} $\times$ \textit{time}.
+The 3D fiber encodes points at all possible combinations of \textit{temperature}, \textit{latitude}, and \textit{longitude}.}
+    \label{fig:data:fiber:example}
 \end{figure}
 
-For example, the records in the 2D fiber in \autoref{fig:data_fiber_example} are a pair of times and \textdegree K temperature measurements taken at those times.
+For example, the records in the 2D fiber in \autoref{fig:data:fiber:example} are a pair of \textit{times} and \textdegree K \textit{temperature} measurements taken at those times.
 Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$.
@@
 In doing so, we specify the properties of the component that must be preserved in a graphic representation.
-While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's scales \cite{stevensTheoryScalesMeasurement1946}, we generalize to monoids to support more component types.
-Monoids are also commonly found in functional programming because they specify compositions of transformations \cite{yorgeyMonoidsThemeVariations, stievenMonadJustMonoid2020}.
-
-A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$.
-A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$.
+A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
+\begin{align*}
+    \textbf{associativity}\; & \text{for all} a, b, c \in \monoid \, (a \bullet b) \bullet c = a \bullet (b \bullet c)\\
+    \textbf{identity}\; & \text{for all } a\in \monoid,\,  e\bullet a = a 
+\end{align*} 
 As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
@@
 \end{align*}
+The identity and associativity properties of the action denote that the action is a monoid homomorphism, which means that the group operation is preserved on both sides of the action\cite{weissteinGroupHomomorphism}.
+
+\begin{figure}[H]
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
+        \caption{}
+        \label{fig:monoid:total}
+    \end{subfigure}
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
+        \caption{}
+        \label{fig:monoid:partial}
+    \end{subfigure}
+    \caption{The action $\bullet$ in \autoref{fig:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right.
+Since the action maps the weather states to themselves, the ordering defined by the monoid $\ast$ is preserved on both sides of the action.
+The action in \autoref{fig:monoid:partial} is monoid homomorphism because the ordering of the weather states is the same as the ordering of the elements they are mapped to.
+Given $sun \geq clouds\geq rain\, lightining$ on the right, the action  $sun\, clouds \rightarrow clouds$, and $rain\, lightining \rightarrow rain$ is structure preserving because on the left $cloud \geq rain$ so the relative ordering of elements is the same as the elements they are mapped to.}
+    \label{fig:monoid}
+\end{figure}
+
+One example of monoids are partial orderings on a set, such as seen in \label{fig:monoid}.
+Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one.
+For example, $storm \leq rain$.
+In \autoref{fig:monoid:total}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states.
+The action in  \autoref{fig:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action and it is therefore homomorphic.
+The action $\bullet$ in \autoref{fig:monoid:partial} is a monotone map\cite{fongInvitationAppliedCategory2019}, meaning that for  $a,b \in \dfiber_{i}$
+\begin{equation}
+    if\, a \leq b \; then \; \bullet(a) \leq \bullet(b)
+\end{equation}
+the structure the action must preserve is the relative, rather than exact, ordering.
+Since groups are monoids with invertible operations, this definition of structure is also broad enough to include the Steven's measurment scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
+Monoids are also commonly found in functional programming because a property of monoids is that the components can be composed into complex transformation pipelines \cite{yorgeyMonoidsThemeVariations}.
+
 As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
@@
 of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$
-
-Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.
-For example, lets say one component of the data is weather states, such that the fiber component is $\dfiber_i=\{sun,\, clouds,\, rain,\, lightning,\, storm\}$.
-If the data is ordinal, that means some ordering action is applied to the set.
-
-\begin{figure}[H]
-\begin{subfigure}{.5\textwidth}
-    \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
-    \caption{The identity action $\bullet$ maps the weather states to themselves.}
-    \label{fig:monoid:total}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-    \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
-    \caption{The action $\bullet$ is a monotonic map, meaning that it preserves the relative ordering despite mapping multiple elements to the same element.}
-    \label{fig:monoid:partial}
-\end{subfigure}
-\end{figure}
-\note{I think there's something about associativity here, but I'm not sure what}
- In \autoref{fig:monoid:partial}, the action $\bullet$ acts on the weather states such that the states that they are mapped into have the same relative ordering as the original states.
-Unlike group actions, the action $\bullet$ is not invertable and therefore a monoid action.
-We generalize to monoids specifically for the case of partially ordered scales are common in indicator rankings \cite{bruggemannRankingPrioritizationMultiindicator2011}.
 

### 2021-04-16 draft/sections/math.tex
finished out math section

@@
 The 3D fiber encodes points at all possible combinations of \textit{temperature}, \textit{latitude}, and \textit{longitude}.}
-    \label{fig:data:fiber:example}
+    \label{fig:math:data:fiber}
 \end{figure}
@@
 
-As illustrated in figure~\ref{fig:data_fiber_example}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
+As illustrated in figure~\ref{fig:math:data:fiber}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
 
 \subsubsection{Measurement Scales: Monoid Actions}
-\label{sec:data_monoid}
+\label{sec:math:data:monoid}
 Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define by the actions of a monoid on the component.
@@
         \caption{}
-        \label{fig:monoid:total}
+        \label{fig:math:data:monoid:total}
     \end{subfigure}
@@
         \caption{}
-        \label{fig:monoid:partial}
+        \label{fig:math:data:monoid:partial}
     \end{subfigure}
@@
 Given $sun \geq clouds\geq rain\, lightining$ on the right, the action  $sun\, clouds \rightarrow clouds$, and $rain\, lightining \rightarrow rain$ is structure preserving because on the left $cloud \geq rain$ so the relative ordering of elements is the same as the elements they are mapped to.}
-    \label{fig:monoid}
+    \label{fig:math:data:monoid}
 \end{figure}
 
-One example of monoids are partial orderings on a set, such as seen in \label{fig:monoid}.
+One example of monoids are partial orderings on a set, such as seen in \label{fig:math:data:monoid}.
 Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one.
@@
 In \autoref{fig:monoid:total}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states.
-The action in  \autoref{fig:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action and it is therefore homomorphic.
-The action $\bullet$ in \autoref{fig:monoid:partial} is a monotone map\cite{fongInvitationAppliedCategory2019}, meaning that for  $a,b \in \dfiber_{i}$
-\begin{equation}
-    if\, a \leq b \; then \; \bullet(a) \leq \bullet(b)
-\end{equation}
-the structure the action must preserve is the relative, rather than exact, ordering.
+The action in  \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action and it is therefore homomorphic.
+The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map\cite{fongInvitationAppliedCategory2019}
+\begin{equation*}
+    if\, a \leq b \; then \; \bullet(a) \leq \bullet(b) \mid a,b \in \dfiber_{i}
+\end{equation*}
+where the structure the action preserves is the relative, rather than exact, ordering.
 Since groups are monoids with invertible operations, this definition of structure is also broad enough to include the Steven's measurment scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
-Monoids are also commonly found in functional programming because a property of monoids is that the components can be composed into complex transformation pipelines \cite{yorgeyMonoidsThemeVariations}.
+Monoids are also commonly found in functional programming since the core property of monoids is composability \cite{yorgeyMonoidsThemeVariations}.
 
@@
 \begin{equation}
+    \label{eq:math:data:monoid:decompose}
 \monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
@@
 \subsubsection{Continuity of the Data $K$} 
-\label{sec:data_base}
-The base space \dbase\ is way to express how the records in \dtotal\ are connected to each other, for example if they are discrete points or if they lie in a 2D continous surface.
-Connectivity type is assumed in the choice of visualization, for example a line plot implies 1D continuous data, but an explicit representation allows for verifying that the topology of the graphic representation is equivalent to the topology of the data.
-
+\label{sec:math:data:base}
 \begin{figure}[H]
@@
     \caption{The topological base space \dbase\ encodes the continuity of the data space, for example if the data is discrete points or lies on a plane or a sphere}
-    \label{fig:base_space_types}
+    \label{fig:math:data:base:types}
 \end{figure}
-As illustrated in figure~\ref{fig:base_space_types}, \dbase\ is akin to an indexing space into \dtotal\ that describes the structure of \dtotal.
+The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other.
+As shown in \autoref{fig:math:data:base:types},
 \dbase\ can have any number of dimensions and can be continuous or discrete.
+The base space also does not describe anything about the dataset besides the continuity.
+While the base space may have components to identify the continuity, such as \textit{time, latitutde, longitude}, these labels are indexed into from \dbase\ the same as any other component.
+This is similar to the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, but our model treats keys as a pure reference to topology.
+Decoupling the keys from their semantics allows the metadata to be altered; this provides for coordinate agnostic representation of the continuity and facilitates encoding of data where the independent variable may not be clear.
+For example the amount of snow on the ground is dependent on time of day and how much snow has fallen, and changing the coordinate system or time resolution should have no effect on how the records are connected to each other.
 
@@
     \caption{The base space \dtotal\ is divided into fiber segments \dfiber. The base space \dbase\ acts as an index into the records in the fibers.}
-    \label{fig:base_space_div}
+    \label{fig:math:data:base:div}
 \end{figure}
@@
 Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
-In figure~\ref{fig:base_space_div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the closed interval $\left[0,1\right]$.
-As with fibers and monoids, we can decompose the total space into components $\pi:\dtotal_i\rightarrow \dbase$ where
+In \autoref{fig:math:data:base:div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the closed interval $\left[0,1\right]$.
+
+As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose}, we can decompose the total space into component bundles$\pi:\dtotal_i\rightarrow \dbase$ where
 \begin{equation}
+    \label{eq:math:data:base:decompose}
     \pi:\dtotal_1\oplus\ldots\oplus \dtotal_i \oplus\ldots \oplus \dtotal_n \rightarrow \dbase
 \end{equation}
-which is a decomposition of \dfiber. The \dbase\ remains the same because the connectivity of records does not change just because there are fewer elements in each record.
+such that \(\monoid_{i}\) acts on component bundle \(\dtotal_i\).
+The \dbase\ remains the same because the connectivity of records does not change just because there are fewer components in each record.
+By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.
 \begin{figure}[H]
     \includegraphics[width=1\textwidth]{figures/math/base.png}
-    \caption{These two datasets have the same (time, temperature) fiber, but different continuities.
+    \caption{These two datasets have the same \textit{(time, temperature)} fiber, but different continuities.
 The dataset on the left consists of discrete records, while the records in the dataset on the right sampled from a continuous space.}
-    \label{fig:base_example}
+    \label{fig:math:data:base}
 \end{figure}
 
-The datasets in figure~\ref{fig:base_example} have the same fiber of (temperature, time).
+The datasets in figure~\ref{fig:math:data:base} have the same fiber of (temperature, time).
 The dot represents a discrete base space \dbase, meaning that every dataset encoded in the fiber bundle has discrete continuity.
@@
 This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle.
-Because we can decompose the bundle and the fiber, we can decompose \dsection\ as
+Because we can decompose the bundle and the fiber (\autoref{eq:math:data:base:decompose}, \autoref{eq:math:data:fiber:decompose}), we can decompose \dsection\ as
 \begin{equation}
-\label{eq:data_section_decompose}
+\label{eq:math:data:section:decompose}
 \dsection= (\dsection_0,\ldots, \dsection_i, \dots, \dsection_n) 
 \end{equation}
-where each section $\dsection_i$ is a variable or set of variables. This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.
+where each section $\dsection_i$ maps into a record on a component $\dfiber_i \in \dfiber$.
+This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.
 
@@
 They are included in the global set of sections  $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$}
-    \label{fig:data_sections}
+    \label{fig:math:data:sections}
 \end{figure}
 
-In the example in figure~\ref{fig:data_sections}, the fiber is $(time, \, temperature)$ as described in figure~\ref{fig:data_fiber_example} and the base space is the interval \dbase.
-The section $\dsection^{(1)}$ resolves to a series of monotonically increasing in time records of (time, temperature) values.
-Section $\dsection^{(2)}$ returns a different timeseries of (time, temperature) values.
+In \autoref{fig:math:data:sections}, the fiber is the same encoding of $(time, \, temperature)$ illustrated in \autoref{fig:math:data:fiber}, and the base space is the interval \dbase\ shown in \autoref{fig:math:data:base}.
+The section $\dsection^{(1)}$ is a function that for a point \dbasepoint\ returns a record in the fiber \dfiber.
+The section applied to a set of points in \dbase\ resolves to a series of monotonically increasing in time records of \textit{(time, temperature)} values.
+Section $\dsection^{(2)}$ returns a different timeseries of \textit{(time, temperature)} values.
 Both sections are included in the global set of sections $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$.

### 2021-04-16 draft/sections/math.tex
graphics section

@@
 
-As illustrated in figure~\ref{fig:math:data:fiber}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
+As illustrated in \autoref{fig:math:data:fiber}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
 
@@
 
-The datasets in figure~\ref{fig:math:data:base} have the same fiber of (temperature, time).
+The datasets in \autoref{fig:math:data:base} have the same fiber of (temperature, time).
 The dot represents a discrete base space \dbase, meaning that every dataset encoded in the fiber bundle has discrete continuity.
@@
 
+\subsubsection{Sheafs}
+\label{sec:math:data:sheaf}
+Many types of dynamic visualizations require evaluating sections on different subspaces of \dbase, an a sheaf, denoted $\mathcal{O}$ provides a way to do so.
+A sheaf is a mathematical structure for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.
+On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections $\iota^{*}\dsection$
+\begin{equation}
+    \label{eq:sheaf}
+    \begin{tikzcd}[column sep=huge]
+        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
+        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
+    \end{tikzcd}
+\end{equation}
+which are sections of \dtotal\ pulled back over local neighborhood $U \subset \dtotal$ via the inclusion map \(\iota:\dtotal\rightarrow U\).
+The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.
+
 \subsubsection{Applications to Data Containers}
@@
 \subsection{Graphic Space \gtotal}
-\label{sec:graphic}  
-We introduce a graphic bundle to hold the essential information necessary to render a graphical design constructed by the artist.
+\label{sec:math:graphic} 
+To establish that the artist is structure preserving map from data \dtotal\ to graphic \gtotal\, we construct a graphic bundle so that we can define \textit{equivariance} in terms of maps on the fiber spaces and \textit{continuity} in terms of maps on the base space.
 As with the data, we can represent the target graphic as a section \gsection\ of a bundle  $(\gtotal, \gbase, \pi, \gfiber)$.
-The graphic bundle \gtotal\ consists of a base \gbase (~\ref{sec:graphic_fiber}) that is a thickened form of \dbase\,
-a fiber \gfiber (~\ref{sec:graphic_base}) that is an idealized display space, and sections \gsection (~\ref{sec:graphic_section}) that encode a graphic where the visual characteristics are fully specified.
+\begin{equation}
+    \begin{tikzcd}[ampersand replacement=\&]
+        \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
+                          \& \gbase \arrow[u, "\gsection"', bend right]
+    \end{tikzcd}
+  \end{equation}
+The graphic bundle \gtotal\ consists of a base \gbase (\autoref{sec:math:graphic:fiber}) that is a thickened form of \dbase\, a fiber \gfiber (\autoref{sec:math:graphic:base}) that is an idealized display space, and sections \gsection (\autoref{sec:math:graphic:section}) that encode a graphic where the visual characteristics are fully specified.
 
 \subsubsection{Idealized Display \gfiber}
-\label{sec:graphic_fiber}
+\label{sec:math:graphic:fiber}
 To fully specify the visual characteristics of the image, we construct a fiber \gfiber\ that is an infinite resolution version of the target space.
@@
 In this work, we assume a 2D opaque image $\gfiber=\reals^5$ with elements
-\begin{equation}
+\begin{equation*}
 (x,\, y,\, r,\, g,\, b) \in \gfiber
-\end{equation}
+\end{equation*}
 such that a rendered graphic only consists of 2D position and color.
@@
 \subsubsection{Continuity of the Graphic \gbase} 
-\label{sec:graphic_base}
-Just as the \dbase\ encodes the connectivity of the records in the data, we propose an equivalent \gbase\ that encodes the connectivity of the rendered elements of the graphic.
+\label{sec:math:graphic:base}
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
+    \caption{For a visualization component to preserve continuity, it must have a continuous surjective map $\vindex: \gbase\rightarrow\dbase$ from graphic continuity to data continuity.
+The scatter and line graphic base spaces \gbase\ have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
+The image has the same dimension in \gbase\ as in \dbase.}
+    \label{fig:math:graphic:retraction:map}
+\end{figure}
+To establish that a visualization component preserves continuity, we propose that their must be a continuous map $\vindex: \gbase\rightarrow\dbase$ from the graphic base space to the data space.
 For example, consider a \gbase\ that is mapped to the region of a 2D display space that represents \dbase.
@@
 
-Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy.
+Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy, meaning there is a continuous map from \gbase\ to \dbase\cite{weissteinHomotopy}.
 The surjective map $\vindex: \gbase \rightarrow \dbase$
@@
 This means that if $\vindex(\gbasepoint) = \dbasepoint$, the record at \dbasepoint\ is copied over the region \gbasepoint\ such that $\dsection(\dbasepoint)=\vindex^*\dsection(\gbasepoint)$ where $\vindex^*\dsection(\gbasepoint)$  is \dsection\ pulled back over \gbase.
-
-\begin{figure}[H]
-    \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
-    \caption{The scatter and line graphic base spaces have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
-The image has the same dimension in \gbase\ as in \dbase.}
-    \label{fig:graphic_retraction_map}
-\end{figure}
-
-When \dbase\ is discrete points and the graphic is a scatter plot, each point $\dbasepoint \in \dbase$ corresponds to a 2D disk $\gbase_{\dbasepoint}$ as shown in figure~\ref{fig:graphic_retraction_map}.
+When \dbase\ is discrete points and the graphic is a scatter plot, each point $\dbasepoint \in \dbase$ corresponds to a 2D disk $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:graphic:retraction:map}.
 In the case of 1D continuous data and a line plot, the region \gy\ over a point $\gx_i$ specifies the thickness of the line in \gbase\ for the corresponding \dsection\ on \dbasepoint.
@@
 \label{sec:graphic_section}
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/render.png}
-    \caption{To render a graphic, a pixel $p$ is selected in the display space, which is defined in the same coordinates as the x and y components in \gfiber.
-The inverse mapping ${\gsection_{xy}}^(p)$ returns a region $\gbase_{p} \subset \gbase$.
-$\gsection(\gbase_{p})$ returns the list of elements $(x,\,y,\,r,\,g,\,b) \in \gfiber$ that lie over $\gbase_{p}$.
-The integral over the $(r,\,g,\,b)$ elements is the color of the pixel.}
-    \label{fig:graphic_rho_lookup}
-\end{figure}
-
-This section describes how we go from a graphic in an idealized prerender space to a rendered image, where the graphic is the section $\gsection: \gbase \rightarrow \gtotal$.
+The section $\gsection: \gbase \rightarrow \gtotal$ is the graphic in an idealizes prerender space and also acts as a specification for rendering the graphic to an image.
 It is sufficient to sketch out how an arbitrary pixel would be rendered, where a pixel $p$ in a real display corresponds to a region $\gbase_p$ in the idealized display.
-To determine the color of the pixel, we aggregate the color values over the region via integration.
-
-For a 2D screen, the pixel is defined as a region $p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ of the rendered graphic.
-Since the x and y in $p$ are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box $\gbase_{p} ={\gsection_{xy}}^{-1}(p)$ is a region $\gbase_p \subset \gbase$.
-To compute the color, we integrate on $\gbase_p$
-
-\begin{align}
+To determine the color of the pixel, we aggregate the color values over the region via integration:
+\begin{align*}
+    \label{eq:math:graphic:section:color}
     r_p &= \iint\limits_{S_p} \rho_r(s)ds^{2}\\
@@
     b_p &= \iint\limits_{S_p} \rho_b(s)ds^{2}
-\end{align}
+\end{align*}
+For a 2D screen, the pixel is defined as a region $p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ of the rendered graphic.
+Since the x and y in $p$ are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box $\gbase_{p} ={\gsection_{xy}}^{-1}(p)$ is a region $\gbase_p \subset \gbase$.
+The color is the result of the integration over  $\gbase_p$.
 
-As shown in figure~\ref{fig:graphic_rho_lookup}, a pixel $p$ in the output space is selected and inverse mapped into the corresponding region $\gbase_p \subset \gbase$.
-This triggers a lookup of the $\gsection$ over the region $\gbase_p$, which yields the set of elements in \gfiber\ that specify the $(r, g, b)$ values corresponding to the region $p$.
-The color of the pixel is then obtained by taking the integral of $\gsection_{rgb}(\gbase_p)$.
+\begin{figure}[H]
+    \includegraphics[width=\textwidth]{figures/math/render.png}
+    \caption{To render a graphic, a pixel $p$ is selected in the display space, which is defined in the same coordinates as the x and y components in \gfiber\ via the renderer.
+In \gtotal\, the inverse mapping ${\gsection_{xy}}^(p)$ returns a region $\gbase_{p} \subset \gbase$.
+$\gsection(\gbase_{p})$ returns a set of points $(x,\,y,\,r,\,g,\,b) \in \gfiber$ that lie over $\gbase_{p}$.
+The integral over the $(r,\,g,\,b)$ pixels specifies that the pixel should be green}
+    \label{fig:math:graphic:rho:lookup}
+\end{figure}
 
+As shown in \autoref{fig:math:graphic:rho:lookup}, a pixel $p$ in the output space, drawn in yellow, is selected and mapped, via the renderer, into a region on \gtotal.
+The region on \gtotal\ corresponds to a region $\gbase_p \subset \gbase$ via the inverse mapping  ${\gsection_{xy}}^(p)$.
+The base space \gbase\ is an annulus to match the topology of the graphic idealized in \gtotal.
+The section  $\gsection(\gbase_{p})$  then maps into the fiber \gfiber\ over $\gbase_p$ to obtain the set of points in \gfiber, here represented as a table, that correspond to that section.
+The integral over the pixel components of this set of points in the fiber yields \{3, 135, 60\} the actual color of the pixel.
 In general, \gsection\ is an abstraction of rendering.
 In very broad strokes \gsection\  can be a specification such as PDF\cite{bienz1993portable}, SVG\cite{quintScalable2003}, or an openGL scene graph\cite{CarsonOpenGL1997}.
-Alternatively, \gsection can be a rendering engine such as cairo\cite{CairographicsOrg} or AGG\cite{AntiGrainGeometry}.
+Alternatively, \gsection\ can be a rendering engine such as cairo\cite{CairographicsOrg} or AGG\cite{shemanarevAntiGrainGeometry}.
 Implementation of \gsection\ is out of scope for this work,
@@
 \end{equation}
-Sheafs are a mathematical object with restriction maps that define how to glue \dsection over local neighborhoods $U \subseteq \dbase$,  discussed in section~\ref{sec:artist_sec:data_sheaf_stalk}, such that the \vartist\ maps are consistent over continuous regions of \dbase.
+Sheafs are a mathematical object with restriction maps that define how to glue \dsection over local neighborhoods $U \subseteq \dbase$,  discussed in \autoref{sec:artist_sec:data_sheaf_stalk}, such that the \vartist\ maps are consistent over continuous regions of \dbase.
 While \vartist can usually construct graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
@@
 where the input can be point wise $\dsection(\dbasepoint)\mid \dbasepoint \in \dbase$.
-The encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components(\ref{sec:graphic_base}).
-The continuity map $\vindex:\gbase \rightarrow \dbase$ then pulls back the visual bundle \vtotal\ over \gbase (\ref{sec:artist_nu}).
-Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal (\ref{sec:artist_q}).
+The encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components(\autoref{sec:graphic_base}).
+The continuity map $\vindex:\gbase \rightarrow \dbase$ then pulls back the visual bundle \vtotal\ over \gbase (\autoref{sec:artist_nu}).
+Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal (\autoref{sec:artist_q}).
 This functional decomposition of the visualization artist facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
@@
 The visual bundle \vtotal\ is the latent space of possible parameters of a visualization type, such as a scatter or line plot.
-We define \vfiber\ in terms of the parameters of a visualization libraries compositing functions; for example table~\ref{tab:mpl_visual_variable_fiber} is a sample of the fiber space for Matplotlib \cite{hunterMatplotlib2DGraphics2007}.
+We define \vfiber\ in terms of the parameters of a visualization libraries compositing functions; for example \autoref{tab:mpl_visual_variable_fiber} is a sample of the fiber space for Matplotlib \cite{hunterMatplotlib2DGraphics2007}.
 
@@
 \end{figure}
-As introduced in section~\ref{sec:intro_viz}, there are many ways to visually represent data components. We define the visual transformers \vchannel\ 
+As introduced in \autoref{sec:intro_viz}, there are many ways to visually represent data components. We define the visual transformers \vchannel\ 
 \begin{equation}
@@
 Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve.
-As mentioned in section~\ref{sec:data_monoid}, we choose monoid actions as the basis for equivariance because they define the structure on the fiber components.
+As mentioned in \autoref{sec:data_monoid}, we choose monoid actions as the basis for equivariance because they define the structure on the fiber components.
 
@@
 
-On example of an equivariant \vchannel\ is illustrated in figure~\ref{fig:artist_nu}, which is a mapping from \textbf{Strings} to symbols.
+On example of an equivariant \vchannel\ is illustrated in \autoref{fig:artist_nu}, which is a mapping from \textbf{Strings} to symbols.
 The data is an example of a Steven's nominal measurement set, which is defined as having on it permutation group actions
@@
 
-As shown in figure~\ref{fig:artist_q}, the assembly function \vmark\ combines the fiber $\dfiber_i$ wise \vchannel\ transforms into a graphic in \gtotal.
+As shown in \autoref{fig:artist_q}, the assembly function \vmark\ combines the fiber $\dfiber_i$ wise \vchannel\ transforms into a graphic in \gtotal.
 Together, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings, reduce the encodings into a graphic.
@@
 
-The glyph in figure~\ref{fig:artist_mark_change} has the following characteristics \vfiber\ specified by  $(xpos,\, ypos,\, color,\, thickness)$ such that one section is $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
+The glyph in \autoref{fig:artist_mark_change} has the following characteristics \vfiber\ specified by  $(xpos,\, ypos,\, color,\, thickness)$ such that one section is $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
 The equivariance constraint on \vmark\ is that the action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$.
-The corresponding action on \gsection\ causes $\vmark(\vsection^{\prime})$ to be the thicker circle in figure~\ref{fig:artist_mark_change}.
+The corresponding action on \gsection\ causes $\vmark(\vsection^{\prime})$ to be the thicker circle in \autoref{fig:artist_mark_change}.
 
@@
 \end{figure}
-The scatter plot in figure~\ref{fig:artist_scatter} can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
-The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in figure~\ref{fig:artist_scatter}:
+The scatter plot in \autoref{fig:artist_scatter} can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
+The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:artist_scatter}:
 \begin{align}
@@
 Here $size$ can either be defined inside \vmark\ or it could also be a parameter in \vtotal\ that is passed along with (xpos, vpos).
-As seen in figure~\ref{fig:artist_scatter}, a scatter has a direct mapping from a region on $\gbase_{k}$ to its corresponding \dbasepoint.
+As seen in \autoref{fig:artist_scatter}, a scatter has a direct mapping from a region on $\gbase_{k}$ to its corresponding \dbasepoint.
 
@@
 
-In contrast to the scatter, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ shown in fig~\ref{fig:artist_line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase\.
+In contrast to the scatter, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ shown in \autoref{fig:artist_line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase\.
 The line also exemplifies the need for the jet since the line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
@@
 
-The image $\vmark(xpos, ypos, color)$ in figure~\ref{fig:artist_heatmap} is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
+The image $\vmark(xpos, ypos, color)$ in \autoref{fig:artist_heatmap} is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
 Since \dbase\ is 2D continuous space, the indexing variables $(\alpha, \beta)$ define the distance along the space.
@@
 The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
-As shown in eq~\ref{eq:artist_diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+As shown in \autoref{eq:artist_diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
 \begin{figure}[H]
@@
 
-This copying is illustrated in figure~\ref{fig:artist_q_hat}, where the 1D fiber $\vfiber \hookrightarrow \vtotal$ over \dbase\ is copied repeatedly to become the 2D fiber $\vfiberpull \hookrightarrow \vtotalpull$ with identical components over \gbase.
+This copying is illustrated in \autoref{fig:artist_q_hat}, where the 1D fiber $\vfiber \hookrightarrow \vtotal$ over \dbase\ is copied repeatedly to become the 2D fiber $\vfiberpull \hookrightarrow \vtotalpull$ with identical components over \gbase.
 Given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map from \vsection on a point \dbasepoint, there is a direct map from the visual section over data base space  $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$ to the visual section \vsectionpull\ over graphic base space.
 This map means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint.
-This means that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$, which is illustrated in figure~\ref{fig:artist_q_hat} as each dot on \vfiber is equivalent to the line intersection \vfiberpull.
+This means that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$, which is illustrated in \autoref{fig:artist_q_hat} as each dot on \vfiber is equivalent to the line intersection \vfiberpull.
 
@@
 
-Factoring out \gbasepoint\ from equation~\ref{eq:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
+Factoring out \gbasepoint\ from \autoref{eq:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
 In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record   $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
@@
 The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delgate the construction of \gsection\ to a back end such as Matplotlib.
-
-\subsubsection{Sheafs}
-\label{sec:artist_sheaf_stalk}
-The restriction maps of a sheaf describe how local \dsection can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}.
-As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$.
-We can define the inclusion map $\iota: U \rightarrow \dbase$  which pulls \dtotal over $U$
-\begin{equation}
-    \label{eq:sheaf}
-    \begin{tikzcd}
-        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
-        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
-    \end{tikzcd}
-\end{equation}
-such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$.
-By gluing $\iota^*\dsection$ together, the sheaf is putting a continuous structure on local sections which allows for defining a section over a subset in \dbase.
-That section over subset \dbase\ maps to the graphic generated by \vartist\ for visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.
 

### 2021-04-16 draft/sections/math.tex
nu section rework

@@
 \label{sec:artist}
-We propose that the transformation from data to visual representation can be described as a structure preserving map from one topological space to another.
-We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.
-The topological artist \vartist\ is a monoid equivariant sheaf map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
+The topological artist \vartist\ is how we model the building block component that transforms data into a graphic.
+The artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
 \begin{equation}
     \vartist: \mathcal{O}(\dtotal) \rightarrow \mathcal{O}(\gtotal)
+    \label{eq:math:artist:artist}
 \end{equation}
-Sheafs are a mathematical object with restriction maps that define how to glue \dsection over local neighborhoods $U \subseteq \dbase$,  discussed in \autoref{sec:artist_sec:data_sheaf_stalk}, such that the \vartist\ maps are consistent over continuous regions of \dbase.
-While \vartist can usually construct graphical elements solely with the data in \dsection, some visualizations, such as line, may also need some finite number $n$ of derivatives, which is captured by the jet bundle $\mathcal{J}^n$ \cite{JetBundle2020,musilovaCalculusVariationsJet2016} with $\mathcal{J}^{0}(\dtotal)=\dtotal$.
-In this work, we at most need $\mathcal{J}^{2}(\dtotal)$ which is the value at \dsection\ and its first and second derivatives; therefore the artist takes as input the jet bundle $\dtotal^{\prime}=\mathcal{J}^{2}(\dtotal)$.
+The artist preserves \textit{continuity} through the \vindex\ map discussed in \autoref{sec:math:graphic:base} and is an \textit{equivariant} map because it carries a homomorphism of monoid actions \cite{cegarraCohomologyMonoidsOperators2019}
+\begin{equation}
+    \varphi: \monoid \rightarrow \monoid^{\prime}
+\end{equation}
 
-Specifically, \vartist is the equivariant map from $\dtotal^{\prime}$ to a specific graphic $\gsection \in \Gamma(H)$
+Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps 
+\begin{equation}
+\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
+\end{equation}
+such that applying a monoid action $m \in \monoid$ to the data input $\delement \in \mathscr{\dtotal}$ of the artist $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.
 
+The monoid equivariant map has two stages: the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components, and the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
 \begin{equation}
-    \label{eq:artist_diagram}
+    \label{eq:math:artist:diagram}
     \begin{tikzcd}
@@
 \end{equation}
-where the input can be point wise $\dsection(\dbasepoint)\mid \dbasepoint \in \dbase$.
-The encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components(\autoref{sec:graphic_base}).
-The continuity map $\vindex:\gbase \rightarrow \dbase$ then pulls back the visual bundle \vtotal\ over \gbase (\autoref{sec:artist_nu}).
-Then the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal (\autoref{sec:artist_q}).
-This functional decomposition of the visualization artist facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
+\vtotalpull\ is the visual bundle \vtotal\ pulled back over \gbase\ via the equivariant continuity map $\vindex:\gbase \rightarrow \dbase$ introduced in \autoref{sec:math:graphic:base}.
+The functional decomposition of the visualization artist in \autoref{eq:math:artist:diagram}facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
+We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.
 
 \subsubsection {Visual Fiber Bundle \vtotal}
-We introduce a visual bundle \vtotal\ to store the visual representations the artist needs to assemble into a graphic.
-The visual bundle (\vtotal, \dbase, $\pi$, \vfiber) has section $\vsection: \vtotal \rightarrow \dbase$ that resolves to a visual variable in the fiber \vfiber.
-The visual bundle \vtotal\ is the latent space of possible parameters of a visualization type, such as a scatter or line plot.
-We define \vfiber\ in terms of the parameters of a visualization libraries compositing functions; for example \autoref{tab:mpl_visual_variable_fiber} is a sample of the fiber space for Matplotlib \cite{hunterMatplotlib2DGraphics2007}.
-
+\label{sec:math:visual}
+We introduce a visual bundle \vtotal\ to store the mappings of the data components into components of the graphic.
+The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the space of possible parameters of a visualization type, such as a scatter or line plot.
+As with the data and graphic bundles, the visual bundle is defined by the projection map $\pi$
+\begin{equation}
+    \begin{tikzcd}[ampersand replacement=\&]
+        \vfiber \arrow[r, hook] \& \vtotal \arrow[d, "\pi"'] \\
+                          \& \dbase \arrow[u, "\vsection"', bend right]
+    \end{tikzcd}
+\end{equation}
+where \vsection\ is the visual variable encoding, as described by Bertin \cite{bertinSemiologyGraphicsDiagrams2011a}, of the data section \dsection.
+The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave.
 \begin{table}[H]
@@
     \caption{Some possible components of the fiber \vfiber\ for a visualization function implemented in Matplotlib}
-    \label{tab:mpl_visual_variable_fiber}
+    \label{tab:math:artist:mpl:fiber}
 \end{table}
-
  A section \vsection\ is a tuple of visual values that specifies the visual characteristics of a part of the graphic.
-For example, given a fiber of $\{xpos, ypos, color\}$ one possible section could be  $\{.5, .5, (255, 20,147)\}$.
-The $codomain(\vchannel_i)$ determines the monoid actions on $\vfiber_i$.
-These fiber components are implicit in the library, by making them explicit as components of the fiber we can build consistent definitions and expectations of how these parameters behave.
+For example, given a fiber of \(\{x, y, color\}\) one possible section could be  \(\{.5, .5, (255, 20,147)\}\).
+The \(codomain(\vchannel_i)\) determines which monoids can act on $\vfiber_i$.
+These fiber components are implicit in the library, as seen in \autoref{tab:math:artist:mpl:fiber}, and by making them explicit as components of the fiber we can build consistent definitions and expectations of how these parameters behave.
 
 \subsubsection{Visual Encoders \vchannel}
-\label{sec:artist_nu}
-%% equivariant maps 
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/partial_fixed.png}
-    \caption{In this artist, \vchannel\ maps the strings to the emojis.
-This \vchannel\ is equivariant because the monoid actions (which are represented by the colored arrows) are the same on both the \dsection\ input and \vsection\ output sets.}
-    \label{fig:artist_nu}
-\end{figure}
-As introduced in \autoref{sec:intro_viz}, there are many ways to visually represent data components. We define the visual transformers \vchannel\ 
+\label{sec:math:artist:nu}
+We define the visual transformers \vchannel\ 
 \begin{equation}
-    \label{eq:nu_expanded}
+  \label{eq:math:artist:nu}
   \{\vchannel_{0}, \ldots, \vchannel_{n}\}: \{\dsection_{0}, \ldots, \dsection_{n}\} \mapsto \{\vsection_{0}, \ldots, \vsection_{n}\}
@@
 Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve.
-As mentioned in \autoref{sec:data_monoid}, we choose monoid actions as the basis for equivariance because they define the structure on the fiber components.
-
-A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes such that 
+As mentioned in \autoref{sec:math:data:monoid}, monoid actions define the structure on the fiber components and are therefore the basis for equivariance.
+A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes
 \begin{equation}
-    \label{eq:nu_categorical}
+  \label{eq:math:artist:nu_commute}
 \begin{tikzcd}
@@
 
-On example of an equivariant \vchannel\ is illustrated in \autoref{fig:artist_nu}, which is a mapping from \textbf{Strings} to symbols.
-The data is an example of a Steven's nominal measurement set, which is defined as having on it permutation group actions
-\begin{equation}
-    \text{ if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)
-\end{equation}
-such that shuffling the words must have an equivalent shuffle of the symbols they are mapped to.
-This is illustrated in the identical actions, represented by the colored arrows, on the words and emojis.
-To preserve ordinal and partial order monoid actions, \vchannel\ must be a monotonic function such that given $\delement_1, \delement_2 \in \dtotal_{i}$,
-\begin{equation}
-\text{ if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)
-\end{equation}
-the visual encodings must also have some sort of ordering. For interval scale data, \vchannel\ is equivariant under translation monoid actions if 
-\begin{equation}
-\vchannel (x + c) = \vchannel(x) + c
-\end{equation}
-while for ratio data, there must be equivalent scaling\cite{weissteinSimilarityTransformation} 
-\begin{equation}
-\vchannel(xc) = \vchannel(x)*c 
-\end{equation}
-We therefore can test if a \vchannel\ is equivariant by testing the actions under which is must  commute.
-For example, we define a transform $\vchannel_{i}(x) = .5$ on interval data.
-This means it must commute under translation, for example $t(x) = x+2$.
-Testing this constraint
-\begin{align}
-    \vchannel(t(\delement + 2)) & \overset{?}{=} \vchannel(\delement) + 2\\
-    .5 &\neq .5 + 2
-\end{align}
-we find that the \vchannel\ defined here does not commute and is therefore invalid.
-The constraints on \vchannel\ can be embedded into our artist such that the \vchannel\ functions can test for equivariance and also provide guidance on constructing new \vchannel\ functions.
+\begin{figure}[htb]
+  \centering
+  \begin{subfigure}{.49\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/partial_invalid.png}
+    \caption{}
+    \label{fig:math:artist:nu:invalid}
+  \end{subfigure}
+  \begin{subfigure}{.49\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/partial_fixed.png}
+    \caption{}
+    \label{fig:math:artist:nu:valid}
+  \end{subfigure}
+  \caption{The map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not homomorphic, and therefore invalid, because $rain \geq storm$ is mapped to elements with the reverse ordering $\nu(storm) \geq \nu(storm)$.
+In contrast, the mapping in \autoref{fig:math:artist:nu:valid} is valud since $\nu(storm)=\nu(rain)$ satisfies the condition $\nu(storm) \geq \nu(storm)$}
+  \label{fig:math:artist:nu}
+\end{figure}
+
+The translation from weather state data to visual representation as umbrella emoji in \autoref{fig:math:artist:nu:invalid} is an invalid visual encoding map \vchannel\ because it is not homomorphic.
+This is because the monotonic condition \(rain \geq storm \implies \vchannel(rain) \geq  \vchannel(storm)\) is not met since \(\vchannel(rain) \leq\vchannel(storm)\).
+To satisfy the monotonic condition for $rain \geq storm$, either red arrow in \autoref{fig:math:artist:nu:invalid} would have to go in a different direction.
+On the other hand, the mapping from weather state to umbrellla in \autoref{fig:math:artist:nu:valid} is a homomorphism since $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$.
+
+\begin{table}[H]
+  \renewcommand{\arraystretch}{2}
+  \begin{tabulary}{\columnwidth}{|llL|}\hline
+      \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
+      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
+      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
+      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
+      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
+  \end{tabulary}
+  \caption{}
+  \label{tab:math:artist:nu}
+\end{table}
+An example of monoid action equivariance is the preservation of partial ordering illustrated in
+
+\autoref{fig:math:artist:nu} is an example of a valid visual representation of partially ordered data components.
+\autoref{tab:math:artist:nu} describes the conditions on \vchannel\ for components that are one of the Stevens measurement types\cite{stevensTheoryScalesMeasurement1946}, which he defined in terms of their mathematical group structure.
+By specifying the equivariance constraints on \vchannel\, we can guarantee that the stage of the artist that transforms data components into visual representations is equivariant.
+These constraints guide the implementation of reusable component transformers \vchannel\ that are composed when generating the graphic.
 

### 2021-04-16 draft/sections/math.tex
q and q hat

@@
 \begin{table}[H]
+    \centering
     \renewcommand{\arraystretch}{2}
@@
 On the other hand, the mapping from weather state to umbrellla in \autoref{fig:math:artist:nu:valid} is a homomorphism since $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$.
+\autoref{fig:math:artist:nu} is an example of how the model supports partially ordered data components, which was a motivation for defining equivariance as monoid homomorphisms.
 
 \begin{table}[H]
+\centering
   \renewcommand{\arraystretch}{2}
@@
 \end{table}
-An example of monoid action equivariance is the preservation of partial ordering illustrated in
 
-\autoref{fig:math:artist:nu} is an example of a valid visual representation of partially ordered data components.
-\autoref{tab:math:artist:nu} describes the conditions on \vchannel\ for components that are one of the Stevens measurement types\cite{stevensTheoryScalesMeasurement1946}, which he defined in terms of their mathematical group structure.
+The Stevens measurement types\cite{stevensTheoryScalesMeasurement1946}, listed in \autoref{tab:math:artist:nu}, are specified in terms of groups, which are monoids with invertible operations\cite{remlingAlgebraMath5353}.
+Despite critiques of the scales\cite{johnsonPseudoMathematicsMentalSocial1936,thomasMathematizationNotMeasurement2014}, we believe it is critical for the model to include the measurement scales since they are commonly used in visualization to classify components \cite{munznerVisualizationAnalysisDesign2014,wilkinsonGrammarGraphics2005}.
 By specifying the equivariance constraints on \vchannel\, we can guarantee that the stage of the artist that transforms data components into visual representations is equivariant.
@@
 
-\subsubsection{Graphic Assembler \vmark}
-\label{sec:artist_q}
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/path_of_q}
-    \caption{$\vchannel_i$ functions convert data $\dsection_i$ to visual characteristics $\vsection_i$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection$ such that there is a map \vindex\ preserving the continuity of the data.
-\gsection\ applied to a region of connected components $\gbase_{\dbasepathpoint}$  generates a part of a graphic, for example the point graphical mark.}
-    \label{fig:artist_q}
+\subsubsection{Visualization Assembly}
+\label{sec:math:artist:q}
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=\textwidth]{figures/math/path_of_q.png}
+  \caption{The transform functions $\vchannel_i$ convert data $\dsection_i\in \dtotal$ to visual characteristics $\vsection_i \in \vtotal$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection \in \gtotal$.}
+  \label{fig:math:artist:q}
 \end{figure}
+The transformation from data into graphic is analagous to a map-reduce operation; as illustrated in \autoref{fig:math:artist}, data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal.
+The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$.
+The full space of graphics is not necessarily equivariant; therefore we formalize the constraints on \vmark\ such that it produces structure preserving graphics.
 
-As shown in \autoref{fig:artist_q}, the assembly function \vmark\ combines the fiber $\dfiber_i$ wise \vchannel\ transforms into a graphic in \gtotal.
-Together, \vchannel\ and \vmark\ are a map-reduce operation: map the data into their visual encodings, reduce the encodings into a graphic.
-As with \vchannel\, the constraint on \vmark\ is that for every monoid action on the input \vsection\, there is corresponding monoid action on the output \gsection.
+We formalize the expectation that visualization generation functions parameterized in the same way should generate the same functions as the equivariant map $\vmark: \vsection \mapsto \gsection$.
+We then define the constraint on \vmark\ such that if \vmark\ is applied to two visual sections $\vsection$ and $\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection$ and $\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.
+We do not define monoid actions on all of $\Gamma(\gtotal)$ because there may be graphics $\gsection \in \Gamma(\gtotal)$ for which we cannot construct a valid mapping from \vtotal.
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
+  \caption{These two glyphs are generated by the same annulus \vmark\ function.
+The monoid action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
+  \label{fig:math:artist:graphic}
+\end{figure}
+Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$
+\begin{prop}
+If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
+\begin{equation}
+\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
+\label{eq:math:artist:q}
+\end{equation}
+then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
+If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection.
+\end{prop}
 
-While \gsection\ generates the entire graphic, we will restrict the discussion of \vmark\ to generation of sections of a glyph.
-We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of path connected components $\dbasepath \subset \dbase$ as input:
+For example, given fiber $\vfiber=(xpos,\, ypos,\, color,\, thickness)$, then sections $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
+The action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$ and the corresponding action on \gsection\ causes $\vmark(\vsection^{\prime})$ to be the thicker circle in \autoref{fig:math:artist:graphic}.
+
+We formally describe a glyph as \vmark\ applied to the regions \dbasepoint\ that map back to a set of path connected components $\dbasepath \subset \dbase$ as input 
 \begin{equation}
@@
 \end{equation}
-such that for every glyph there is at least one corresponding region on \dbase.
-This is in keeping with the definition of glyph as any differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+such that for every glyph there is at least one corresponding region on \dbase, in keeping with the definition of glyph as any visually differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
 The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.
 
-It is on sections of these glyphs that we define the equivariant map as  $\vmark: \vsection \mapsto \gsection$ and an action on the subset of graphics $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ that \vmark\ can generate.
-We then define the constraint on \vmark such that if \vmark\ is applied to $\vsection, \vsection^{\prime}$ that generate the same \gsection\, then the output of both sections acted on by the same monoid $m$ must be the same.
-While it may seem intuitive that visualizations that generate the same glyph should consistently generate the same glyph given the same input, we formalize this constraint such that it can be specified as part of the implementation of \vmark.
-
-Lets call the visual representations of the components $\Gamma(\vtotal)=X$ and the graphic $\vmark(\Gamma(\vtotal))=Y$.
-If for elements of the monoid $m \in \monoid$ and for all $\vsection, \vsection^{\prime} \in X$, we define the monoid action on $X$ so that it is by definition equivariant
-\begin{equation}
-\vmark(\vsection) = \vmark(\vsection^{\prime})\implies \vmark(m\circ\vsection) = \vmark(m\circ\vsection^{\prime})
-\end{equation}
-then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
-The transformed graphic $\gsection^{\prime}$ is equivariant to a transform on the visual bundle $\gsection^{\prime}=\vmark(m\circ \vsection)$ on a section that $\vsection \in \vmark^{-1}(\gsection)$ that must be part of generating \gsection.
-
-\begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/diff_type_q.png}
-    \caption{These two glyphs are generated by the same \vmark\ function. The monoid 
-    action $m_i$ on edge thickness $\vsection_i$ of the first glyph yields the thicker edge ${\vsection_i}^{\prime}$ in the second glyph.}
-    \label{fig:artist_mark_change}
-\end{figure}
-
-The glyph in \autoref{fig:artist_mark_change} has the following characteristics \vfiber\ specified by  $(xpos,\, ypos,\, color,\, thickness)$ such that one section is $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
-The equivariance constraint on \vmark\ is that the action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$.
-The corresponding action on \gsection\ causes $\vmark(\vsection^{\prime})$ to be the thicker circle in \autoref{fig:artist_mark_change}.
-
 \subsubsection{Assembly \vmark}
-In this section we formulate the minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image. 
+Given the continuities described in \ref{fig:math:graphic:retraction:map}, 
+we illustrate a minimal Q that will generate the most minimal visualizations associated with those continuities: non-overlapping scatter points, a non-infinitely thin line, and an image. 
 \begin{figure}[H]
@@
     }
-    \label{fig:artist_scatter}
+    \label{fig:math:artist:q:scatter}
 \end{figure}
-The scatter plot in \autoref{fig:artist_scatter} can be defined as $\vmark(xpos, ypos)(\alpha, \beta)$ where color $\gsection_{RGB} = (0,0,0)$ is defined as part of \vmark\ and $\gbasepoint=(\alpha, \beta)$ defines the region on \gbase.
-The position of this swatch of color can be computed relative to the location on the disc $\gbase_{\dbasepoint}$ as shown in \autoref{fig:artist_scatter}:
-\begin{align}
+The scatter plot in \autoref{fig:math:artist:q:scatter} can be defined as 
+\begin{equation}
+\vmark(xpos, ypos)(\alpha, \beta)
+\label{eq:math:artist:q:scatter}
+\end{equation}
+with a constant \textit{size} and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of \vmark.
+The position of this swatch of color can be computed relative to the location on the disc \((\alpha, \beta) \in \gbase_{\dbasepoint}\) as shown in \autoref{fig:math:artist:q:scatter}
+\begin{align*}
 x &= size *\alpha \cos(\beta) + xpos \\
 y &= size *\alpha \sin(\beta) + ypos
-\end{align}
-
+\end{align*}
 such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
-Here $size$ can either be defined inside \vmark\ or it could also be a parameter in \vtotal\ that is passed along with (xpos, vpos).
-As seen in \autoref{fig:artist_scatter}, a scatter has a direct mapping from a region on $\gbase_{k}$ to its corresponding \dbasepoint.
 
@@
 \end{figure}
-
-In contrast to the scatter, the line plot $\vmark(xpos, \hat{n_{1}}, ypos, \hat{n_{2}})(\alpha, \beta)$ shown in \autoref{fig:artist_line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase\.
-The line also exemplifies the need for the jet since the line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness.
-The magnitude of the slope is
+In contrast, the line plot 
 \begin{equation}
-    \lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}
+\vmark(xpos, \hat{n}_{1}, ypos, \hat{n}_{2})(\alpha, \beta)
+\label{eq:math:artist:q:line}
 \end{equation}
-such that the normal is  
-\begin{equation}
-    \hat{n_{1}} = \frac{n_1}{\lvert n \rvert}, \; \hat{n_{2}} = \frac{n_2}{\lvert n \rvert}
-\end{equation}
-
-which yields components of \gsection\
-\begin{align}
- x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n_1}(\vindex(\alpha)) \\
- y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n_2}(\vindex(\alpha)) 
-\end{align}
-
-where (x,y) look up the position $\vindex(\alpha)$ on the data.
-At that point, we also look up the the derivatives $\hat{n_1}, \hat{n_2}$  which are then multiplied by a $width$ parameter to specify the thickness.
-As with the $size$ parameter in scatter, $width$ can be defined in \vmark\ or as a component of \vtotal.
+in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
+As shown in \autoref{fig:math:artist:q:line}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal.
+The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
+such that the normal is  $\hat{n}_{1} = \frac{n_1}{\lvert n \rvert}, \; \hat{n}_{2} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
+\begin{align*}
+ x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n}_1(\vindex(\alpha)) \\
+ y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n}_2(\vindex(\alpha)) 
+\end{align*}
+where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n}_1, \hat{n}_2$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
 
@@
     \caption{The only visual parameter an image requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
-    \label{fig:artist_heatmap}
+    \label{fig:math:artist:q:heatmap}
 \end{figure}
-
-The image $\vmark(xpos, ypos, color)$ in \autoref{fig:artist_heatmap} is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
-Since \dbase\ is 2D continuous space, the indexing variables $(\alpha, \beta)$ define the distance along the space.
-This is then used by \vindex\ to map into \dbase\ to lookup the color values
-\begin{align}
-R &= R(\vindex(\alpha, \beta))\\
-G &= G(\vindex(\alpha, \beta))\\
-B &= B(\vindex(\alpha, \beta))
-\end{align}
-that the data values have been mapped into.
+In \autoref{fig:math:artist:q:heatmap}, the image 
+\begin{equation}
+\vmark(xpos, ypos, color)
+\label{eq:math:artist:q:image}
+\end{equation}
+ is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
+The indexing variables $(\alpha, \beta)$ define the distance along the space, which is then used by \vindex\ to map into \dbase\ to lookup the color values
+\begin{equation*}
+R = R(\vindex(\alpha, \beta)), \; G = G(\vindex(\alpha, \beta)), \; B = B(\vindex(\alpha, \beta))
+\end{equation*}
 In the case of an image, the indexing mapper \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.

### 2021-04-16 draft/sections/math.tex
throught q, most of new Q figure

@@
 
-\subsubsection{Assembly factory \vmarkd}
+\label{sec:math:artist:qhat}
 The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
-As shown in \autoref{eq:artist_diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
-\begin{figure}[H]
+As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+\begin{figure}[htb]
+  \centering
     \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
-    \caption{The pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint.
-Because the \vsection\ is the same, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
-    \label{fig:artist_q_hat}
+    \caption{Because the pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
+    \label{fig:math:artist:qhat}
 \end{figure}
@@
 \end{equation}
-copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase\.
+copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase.
 This set of points \gbasepoint\ are the preimage \vpreimg\ of \dbasepoint.
 
-This copying is illustrated in \autoref{fig:artist_q_hat}, where the 1D fiber $\vfiber \hookrightarrow \vtotal$ over \dbase\ is copied repeatedly to become the 2D fiber $\vfiberpull \hookrightarrow \vtotalpull$ with identical components over \gbase.
-Given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map from \vsection on a point \dbasepoint, there is a direct map from the visual section over data base space  $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$ to the visual section \vsectionpull\ over graphic base space.
-This map means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint.
-This means that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$, which is illustrated in \autoref{fig:artist_q_hat} as each dot on \vfiber is equivalent to the line intersection \vfiberpull.
+As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$  from \vsection\ over \dbasepoint\ to the section \vsectionpull\ over \gbasepoint.
+This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
+In \autoref{fig:math:artist:qhat} each dot on \vfiber\ is equivalent to the line on \vfiberpull.
 
 Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out.
-When \vmark\ maps visual sections  into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to the pulled back visual section \vsectionpull\ then
+When \vmark\ maps visual sections into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull\ then the graphic section \gsection\ evaluated on a visual region \gbasepoint\
 \begin{equation}
     \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
+    \label{eq:math:artist:qhat}
 \end{equation}
-the graphic section \gsection\ evaluated on a visual region \gbasepoint\ is defined as the assembly function \vmark\ with input pulled back visual section \vsectionpull\ also evaluated on \gbasepoint.
-Since the pulled back visual section \vsectionpull\ is the visual section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function
+ is defined as the assembly function \vmark\ with input \vsectionpull\ evaluated on \gbasepoint.
+Since the pulled back section \vsectionpull\ is the section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function
 \begin{equation}
-\label{eq:qhat_q_s}
+\label{eq:math:artist:qhat_q_s}
 \vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
 \end{equation} 
-where the assembly function \vmarkd\ that takes as input the visual section on data \vsection\ is defined to be the assembly function \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.
-
-Factoring out \gbasepoint\ from \autoref{eq:qhat_q_s} yields $\vmarkd(\vsection(k)) = \vmark(\vsectionpull)$ where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
+where \vmarkd\ with input \vsection\ is defined to \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.
+Factoring out \gbasepoint\ from \autoref{eq:math:artist:qhat_q_s} yields
+\begin{equation}
+\vmarkd(\vsection(k)) = \vmark(\vsectionpull)
+\label{eq:math:artist:qhat_k}
+\end{equation}
+where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
 In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
 This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
-The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delgate the construction of \gsection\ to a back end such as Matplotlib.
+The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
 

### 2021-04-20 draft/sections/intro.tex
housekeeping

@@
 These libraries can often assume the expected data structure because they are domain specific, and that is the common data structure in that domain.
-For users who generally work in one domain, such as the data, networks, or graphs shown in \autoref{fig:intro:data}, this well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a very coherant user experience\cite{heerDeclarative2010}.
+For users who generally work in one domain, such as the data, networks, or graphs shown in \autoref{fig:intro:data}, this well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a very coherent user experience\cite{heerDeclarative2010}.
 But, for developers who want to build new visualizations on top of these libraries, they must work around the existing assumptions, sometimes in ways that break the model the libraries are developed around.

### 2021-04-20 draft/sections/math.tex
housekeeping

@@
 \end{figure}
-Visualizations with a single artist do not provide much information, so we define addition operators for generating more complex visualizations.
-Given the family of artists $(\dtotal_i: i\in I)$ on the same image, the + operator
+We describe the constraints for compositing artists by defining addition operators. Given the family of artists $(\dtotal_i: i\in I)$ that are rendered to the same image, the + operator 
 \begin{equation}

### 2021-04-21 draft/sections/implementation.tex
slides first draft
housekeeping
code section

@@
 \begin{document}
-\section{Prototype Implementation: Matplottoy}
+\section{Prototype: Matplottoy}
 \label{sec:implementation}
-\begin{figure}[H]
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
-    \end{subfigure}
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/code/line_1.png}
-    \end{subfigure}
-    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
-    \label{fig:code_scatter_line}
-\end{figure}
-To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.
-We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations.
-We also exploit the Matplotlib transform stack to transform data coordinates into screen coordinates.
-To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primitive objects we implemented based on our topology model.
+To evaluate the feasibility of the model described in ~\autoref{sec:math:}, we built prototypes of a \texttt{point}, \texttt{line}, and \texttt{bar} artist.
+We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to convert data coordinates into screen coordinates.
+While the artist is specified in a fully functional manner in \autoref{eq:math:artist:diagram}, we implement the  prototype in a heavily object oriented manner.
+We do so mostly to more easily manage function inputs, especially parameters that are passed through to methods that are structurally functional.
 
-\begin{multicols*}{2}
 \begin{minted}{python}
   fig, ax = plt.subplots()
-    artist = Point(data, transforms)
+  artist = Artist(E, V)
   ax.add_artist(artist)
 \end{minted}
-\columnbreak
+Building on the curr
+ent Matplotlib artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalence class artist \vartisteq\ as described in \autoref{eq:math:artist:aprime}.
+The visual bundle \vtotal\ is specified as the \mintinline{python}{V} dictionary of the form \mintinline{python}|{parameter:(variable name, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber, and the \vchannel\ encoders are passed in as functions or callable objects.
+The data bundle \dtotal\ is passed in as a \mintinline{python}{E} object.
+By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist\ as defined in \autoref{eq:math:artist:artist}.
 \begin{minted}{python}
-    fig, ax = plt.subplots()
-    artist = Line(data, transforms)
-    ax.add_artist(artist)
-\end{minted}
-\end{multicols*}
-
-We then add the \mintinline{python}{Point} and \mintinline{python}{Line} artist that construct the scatter and line graphics.
-These artists are implemented as the equivalence class \vartisteq\ with the aesthetic configurations factored out into a \mintinline{python}{transforms} dictionary that specifies the visual bundle \vtotal\.
-The equivalence classes \vartisteq\ map well to Python classes since the functional aspects-\vchannel, \vmarkd, and \vindex - are completely reusable in a consistent composition, while the visual values in \vtotal\ are what change between different artists belonging to the same class \vartisteq.
-The \mintinline{python}{data} object is an abstraction of a data bundle \dtotal\ with a specified section \dsection.
-Implementing \gtotal\ and \gsection\ are out of scope for this prototype because they are part of the rendering process.
-We also did not implement any form of \vindex\ because the scatter, line, and bar plots prototyped here directly broadcast from \dbasepoint\ to \gbasepoint, unlike for example an image which may need to be rotated.
-
-\subsection{Artist Class $\vartist^{\prime}$}
-\label{sec:code_artist}
-The artist is the piece of the Matplotlib architecture that constructs an internal representation of the graphic that the render then uses to draw the graphic.
-In the prototype artist, \mintinline{python}{transform} is a dictionary of the form \mintinline{python}|{parameter:(variable, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber,  and the \vchannel\ encoders are passed in as functions or callable objects.
-The data bundle \dtotal\ is passed in as a \mintinline{python}{data} object.
-By binding data and transforms to \vartisteq\ inside \mintinline{python}{__init__}, the \mintinline{python}{draw} method is a fully specified artist \vartist.
-
-\begin{minted}{python}
-class ArtistClass(matplotlib.artist.Artist):
-    def __init__(self, data, transforms, *args, **kwargs):
-        # properties that are specific to the graphic but not the channels
-        self.data = data 
-        self.transforms = transforms
+  class ArtistClass(matplotlib.artist.Artist): #A'
+      def __init__(self, E, V, *args, **kwargs):
+          # properties that are specific to the graphic
+          self.E = E
+          self.V = V 
           super().__init__(*args, **kwargs)
 
-    def assemble(self, **args):
+      def hat_Q(self, **args): 
           # set the properties of the graphic
@@
           # returns K, indexed on fiber then key 
-        # is passed the 
-        view = self.data.view(self.axes) 
+          tau = self.E.view(self.axes) 
           # visual channel encoding applied fiberwise 
-        visual = {p: t['encoder'](view[t['name']])
-                  for p, t in self.transforms.items()}
-        self.assemble(**visual)
+          mu = {p: nu(tau(c))
+                   for p, (c, nu) in self.V.items()} 
+          self.hat_q(**mu)
           # pass configurations off to the renderer
@@
   \end{minted}
+ The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
+The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}.
+To ensure the integrity of the section, \mintinline{python}{view} must be atomic, which means that the values cannot change after the method is called in draw until a new call to draw\cite{ullmanFirstCourseDatabase2008}.
+We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.
 
-The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
-The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that we can use to query back into data to select a subset as discussed in section~\ref{sec:artist_sheaf_stalk}.
-The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}.
+ The \vchannel\ functions are then applied to the data, as describe in \autoref{eq:math:artist:nu}, to generate the visual section \vsection that here is the object \mintinline{python}{V}.
 The conversion from data to visual space is simplified here to directly show that it is the encoding \vchannel\ applied to the component.
-In the full implementation, we allow for fixed visual parameter, such as setting a constant color for all sections, by verifying that the named component is in \dfiber\ before accessing the data.
-If the data component name is not in \dfiber\, this is interpreted to mean this component is a thickening of \vtotal that could be pulled back to \dtotal\ via an inverse identity \vchannel.
-
-The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assemble} function that is \vmarkd.
-This assembly function is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
-Although \mintinline{python}{assemble} could be implemented outside the class such that it returns an object the artist could then parse to set attributes, the attributes are directly set here to reduce indirection.
+The \mintinline{python}{q_hat} function that is \vmarkd,  as defined in \autoref{eq:math:artist:qhat_k}, is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
 This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd.
 The last step in the artist function is handing itself off to the renderer.
-The extra \mintinline{python}{*arg, **kwargs} arguments in \mintinline{python}{__init__,draw} are artifacts of how these objects are currently implemnted in Matplotlib.
+The extra \mintinline{python}{*arg, **kwargs} arguments in \mintinline{python}{__init__,draw} are artifacts of how these objects are currently implemented.
 
-The \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of primitive point and area marks.
-In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker.
-We only show the \mintinline{python}{assemble} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
+ \subsection{Scatter, Line, and Bar Artists}
+ \label{sec:code:artists}
+ \begin{figure}[H]
+    \centering 
+    \begin{subfigure}{.49\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
+        \caption{}
+        \label{fig:code:scatter}
+    \end{subfigure}
+    \begin{subfigure}{.48\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/line_1.png}
+        \caption{}
+        \label{fig:code:line}
+    \end{subfigure}
+  \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
+  \label{fig:code:scatter_line}
+\end{figure}
+The figure in \autoref{fig:code:scatter} is described by \autoref{eq:math:artist:q:scatter}.
+This is implemented via a \mintinline{python}{Line} object where the scatter marker shape is fixed as a circle, and the visual fiber components are x and y position and the facecolor and size of the marker.
+We only show the \mintinline{python}{q_hat} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
+
+The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices.
+Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$.
+This means that all the data on one vertex maps to one glyph.
 \begin{minted}{python}
-class Point(mcollections.Collection):
-    def assemble(self, x, y, s, facecolors='C0' ):
-        # construct geometries of the circle glyphs in visual coordinates
-        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
+class Point(ArtistClass, mcollections.Collection):
+  def q_hat(self, x, y, s, facecolors): #\hat{Q}
+    # construct geometries of circle glyphs
+    self._paths = [mpath.Path.circle((xi,yi), radius=si) 
                 for (xi, yi, si) in zip(x, y, s)] 
@@
 \end{minted} 
-The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices.
-Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$.
-This means that all the data on one vertex maps to one glyph.
-To ensure the integrity of the section, \mintinline{python}{view} must be atomic.
-This means that the values cannot change after the method is called in draw until a new call in draw.
-We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.
-
-This table is converted to a table of visual variables and is then passed into \mintinline{python}{assemble}.
-In  \mintinline{python}{assemble}, the \vsection\ components are used to construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
+In \mintinline{python}{q_hat}, the \vsection\ components are used to construct the vector path of each circular marker with center \texttt{(x,y)} and size \texttt{x} and set the colors of each circle.
 This is done via the \mintinline{python}{Path.circle} object.
-As mentioned in sections ~\ref{sec:data_sheaf_stalk} and \ref{sec:artist_q}, this assembly function could as easily be implemented such that it was fed one $\dsection(\dbasepoint)$ at a time.
-
-The main difference between the \mintinline{python}{Point} and \mintinline{python}{Line} objects is in the \mintinline{python}{assemble} function because line has different continuity from scatter and is represented by a different type of graphical mark.
-
 \begin{minted}{python}
-class Line(mcollections.LineCollection):
-    def assemble(self, x, y, color='C0'):
+class Line(ArtistClass, mcollections.LineCollection):
+  def q_hat(self, x, y, color): #\hat{Q}
     #assemble line marks as set of segments 
@@
 \end{minted}
-In the \mintinline{python}{Line} artist, \mintinline{python}{view} returns a table of edges.
+To generate \autoref{fig:code:line}, the \mintinline{python}{Line} artist \mintinline{python}{view} method returns a table of edges.
 Each edge consists of (x,y) points sampled along the line defined by the edge and information such as the color of the edge.
 As with \mintinline{python}{Point}, the data is then converted into visual variables.
-In \mintinline{python}{assemble}, this visual representation is composed into a set of line segments, where each segement is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
+In \mintinline{python}{q_hat}, described by \autoref{eq:math:artist:q:line}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
 Then the colors of each line segment are set.
-The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on view.
+The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on the view.
+The current implementation of line in Matplotlib does not have this functionality because it has no notion of an atomic view \dsection, instead the user keeps track of which values are input into a single call of \vmarkd\ by creating a loop where every color is a new call to the line plotting function.
 
@@
 \begin{minted}{python}
-class Bar(mcollections.Collection):
+class Bar(ArtistClass, mcollections.Collection):
     def __init__(self, data, transforms, orientation='v', *args, **kwargs):
@@
         self.orientation = orientation
-        super().__init__(*args, **kwargs)
-        self.data = data
-        self.transforms = copy.deepcopy(transforms)
+        # set E and V
 
-    def assemble(self, position, length, floor=0, width=0.8, 
+    def q_hat(self, position, length, floor=0, width=0.8, 
                     facecolors='C0', edgecolors='k', offset=0):
@@
 
-    def draw(self, renderer,  *args, **kwargs):
-        view = self.data.view(self.axes)
-        visual = {}
-        for (p, t) in self.transforms.items():
-            if isinstance(t, dict):
-                if t['name'] in self.data.FB.F and 'encoder' in t:
-                    visual[p] = t['encoder'](view[t['name']])
-                elif 'encoder' in t: # constant value
-                    visual[p] = t['encoder'](t['name'])
-                elif t['name'] in self.data.FB.F: # identity 
-                    visual[p] = view[t['name']]
-            else: # no transform 
-                 visual[p] = t
-        self.assemble(**visual)
-        super().draw(renderer,  *args, **kwargs)
-
 \end{minted}
-The \mintinline{python}{draw} method here has a more complex unpacking of visual encodings to support passing in visual component data directly.
-This is vastly simplifies building composite objects as the alternative would be higher order functions that take as input the transforms passed in by the user.
-This construction supports a constant visual parameter, an identity transform where the value is the same in \dtotal\ and \vtotal, and setting the visual component directly.
-The \mintinline{python}{assemble} function constructs bars and sets their face and edge colors.
+ As with \mintinline{python}{Point} and \mintinline{python}{scatter}, \mintinline{python}{q_hat} function constructs bars and sets their properties, face and edge colors.
 The \mintinline{python}{make_bars} function converts the input position and length to the coordinates of a rectangle of the given width.
@@
 
-\subsection{Encoders \vchannel}
-\label{sec:code_channels}
-As mentioned above, the encoding dictionary is specified by the visual fiber component, the corresponding data fiber component, and the mapping function.
+\subsection{Visual Encoders}
+\label{sec:code:nu}
 The visual parameter serves as the dictionary key because the visual representation is constructed from the encoding applied to the data  $\vsection = \vchannel \circ \dsection$.
@@
 \begin{minted}{python}
-cmap =  color.Categorical({'true':'deeppink', 'false':'deepskyblue'})
-transforms = {'x': {'name': 'v4', 'encoder': lambda x: x},
+cmap =  color.Categorical({'true':'deeppink', 
+                           'false':'deepskyblue'}) 
+# {P_i name:{'name':c_i, 'encoder:\nu_i}}
+V = {'x': {'name': 'v4', 'encoder': lambda x: x}, 
      'y': {'name': 'v2', 'encoder': lambda x: x},
      'facecolors': {'name':'v3', 'encoder': cmap}, 
-              's':{'name': None , 'encoder': lambda _: itertools.repeat(.02)}}
+     's':{'name': None , 
+          'encoder': lambda _: itertools.repeat(.02)}}
 \end{minted}
-where the position \textit{(x,y)} \vchannel\ transformers are identity functions.
-The size \textit{s} transformer is not acting on a component of \dfiber, instead it is a \vchannel\ that returns a constant value.
-While size could be embedded inside the \mintinline{python}{assemble} function, it is added to the transformers to illustrate user configured visual parameters that could either be constant or mapped to a component in \dfiber.
-The identity and constant \vchannel\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they are somewhat optimized away in \mintinline{python}{Bar}.
-More complex encoders can be implemented as callable classes, such as
+where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.
 \begin{minted}{python}
-class Categorical:
-    def __init__(self, mapping):
-        # check that the conversion is to valid colors
-        assert(mcolors.is_color_like(color) for color in mapping.values())
-        self._mapping = mapping
-
-    def __call__(self, value):
-        # convert value to a color
-        return [mcolors.to_rgba(self._mapping[v]) for v in values]
-\end{minted}
-
-where \mintinline{python}{__init__} can validate that the output of the \vchannel\ is a valid element of the \vfiber\ component the \vchannel\ function is targeting.
-Creating a callable class also provides a simple way to swap out the specific (data, value) mapping without having to reimplement the validation or conversion logic.
-A test for equivariance can be implemented trivially
-\begin{minted}{python}
+#\nu_i(m_r(E_i)) = \varphi(m_r)(\nu_i((E_i))
 def test_nominal(values, encoder):
@@
 \end{minted}
-but is currently factored out of the artist for clarity.
-In this example, \mintinline{python}{is_nominal} checks for equivariance of permutation group actions by applying the encoder to a set of values, shuffling values, and checking that (value, encoding) pairs remain the same.
-\subsection{Data \dtotal}
-\label{sec:code_data}
+As described in \autoref{eq:math:artist:nu:equivariance}, a test for equivariance can be implemented trivially. It is currently factored out of the artist for clarity.
+
+\subsection{Data Model}
+\label{sec:code:data}
 The data input into the \mintinline{python}{Artist} will often be a wrapper class around an existing data structure.
 This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{python}{view} method that returns an atomic object that encapsulates \dsection.
-The object returned by the view must be key valued pairs of \mintinline{python}|{component name : component section}| where each section is a component as defined in equation~\ref{eq:data_section_decompose}.
 To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataclassesDataClasses}
@@
 class FiberBundle:
-"""
-Attributes
-----------
-K: {'tables': []}
-F: {variable name: type}
-"""
-    K: dict 
-    F: dict
+    K: dict #{'tables': []}
+    F: dict #  {variable name: type}
 \end{minted}
-
-that asks the user to specify how \dbase is triangulated and the attributes of \dfiber.
-Python dataclasses are a good abstraction for the fiber bundle class because the \mintinline{python}{FiberBundle} class only stores data.
-The \dbase\ is specified as tables because the \mintinline{python}{assemble} functions expect tables that match the continuity of the graphic; scatter expects a vertex table because it is discontinuous, line expects an edge table because it is 1D continuous.
-The fiber informs appropriate choice of \vchannel\, therefore it is a dictionary of attributes of the fiber components.
-
-To generate the scatter plot in figure~\ref{fig:code_scatter_line}, we fully specify a dataset with random keys and values in a section chosen at random form the corresponding fiber component.
-The fiberbundle \mintinline{python}{FB} is a class level attribute since all instances of \mintinline{python}{VertexSimplex} come from the same fiberbundle.
+that asks the user to specify the the properties of \dfiber\ and the \dbase\ connectivity as either discrete vertices or continuous data along edges.
+To generate the scatter plot and the line plot, the distinction is in the \mintinline{python}{tau} method that is the section.
 \begin{minted}{python}
-class VertexSimplex: #maybe change name to something else
-    """Fiberbundle is consistent across all sections
-    """
+class PointData:     
+    def __init__(self):
       FB = FiberBundle({'tables': ['vertex']},  
                        {'v1': float, 'v2': str, 'v3': float})
+    def tau(self, k):
+      return # tau evaluated at one point k
 
-    def __init__(self, sid = 45, size=1000, max_key=10**10):
-        # create random list of keys
+class LineData:
+  def __init__(self):
+    FB = FiberBundle({'tables': ['edge']},  
+                     {'x': float, 'y': float, 'color': str})
   def tau(self, k):
-        # e1 is sampled from F1, e2 from F2, etc...
-        return (k, (e1, e2, e3, e4))
-
-    def view(self, axes):
-        table = defaultdict(list)
-        for k in self.keys:
-            table['index'] = k
-            # on each iteration, add one (name, value) pair per component
-            for (name, value) in zip(self.FB.fiber.keys(), self.tau(k)[1]):
-                table[name].append(value)
-        return table
+    return # tau evaluated on interval k
 \end{minted}
-The view method returns a dictionary where the key is a fiber component name and the value is a list of values in the fiber component.
-The table is built one call to the section method \mintinline{python}{tau} at a time, guaranteeing that all the fiber component values are over the same \dbasepoint.
-Table has a \mintinline{python}{get} method as it is a method on Python dictionaries.
-In contrast, the line in \mintinline{python}{EdgeSimplex} is defined as the functions \mintinline{python}{_color,_xy} on each edge.
-
+The discrete \mintinline{python}{tau} method returns a record of discrete points, akin to a row in a table, while the line\mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint.
 \begin{minted}{python}
-class EdgeSimplex:
-
-        FB = FiberBundle({'tables': ['vertex','edge']},
-                         {'x' : float, 'y':  float, 
-                         'color':mtypes.Color()}})
-    def __init__(self, num_edges=4, num_samples=1000): 
-        self.keys = range(num_edge) #edge id
-        # distance along edge
-        self.distances = np.linspace(0,1, num_samples)
-        # half generlized representation of arcs on a circle
-        self.angle_samples = np.linspace(0, 2*np.pi, len(self.keys)+1)
-
-    @staticmethod
-    def _color(edge):
-        colors = ['red','orange', 'green','blue']
-        return colors[edge%len(colors)]
-
-    @staticmethod
-    def _xy(edge, distances, start=0, end=2*np.pi):
-        # start and end are parameterizations b/c really there is 
-        angles = (distances *(end-start)) + start
-        return np.cos(angles), np.sin(angles)
-
-    def tau(self, k): #will fix location on page on revision
-        x, y = self._xy(k, self.distances, 
-                        self.angle_samples[k], self.angle_samples[k+1]) 
-        color = self._color(k) 
-        return (k, (x, y, color))
-
   def view(self, axes):
@@
           table['index'].append(k)
-            # (name, value) pair, value is [x0, ..., xn] for x, y
-            for (name, value) in zip(self.FB.fiber.keys(), self.tau(k, simplex)[1]):
-                table[name].append(value)
-\end{minted}
-Unlike scatter, the line section method \mintinline{python}{tau} returns the functions on the edge evaluated on the interval [0,1].
-By default these means each \mintinline{python}{tau} returns a list of 1000 x and y points and the associated color.
-As with scatter, \mintinline{python}{view} builds a table by calling \mintinline{python}{tau} for each \dbasepoint\.
-Unlike scatter, the line table is a list where each item contains a list of points.
-This bookkeeping of which data is on an edge is used by the \mintinline{python}{assembly} functions to bind segments to their visual properties.
-
-\begin{figure}[H]
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/code/linec_1.png}
-    \end{subfigure}
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{figures/code/lined_1.png}
-    \end{subfigure}
-    \caption{Continuous and discontinuous lines as defined by different data models, but generated with the same \vartisteq \mintinline{python}{Line}}
-    \label{fig:circle_step}
-\end{figure}
-
-The graphics in figure~\ref{fig:circle_step} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{Graphline} data source
-
-\begin{minted}{python}
-class GraphLine:
-    def __init__(self, FB, edge_table, vertex_table, num_samples=1000, connect=False):
-        #s set args as attributes and generate distance
-        if connect: # test connectivity if edges are continuous
-            assert edge_table.keys() == self.FB.F.keys()
-            assert is_continuous(vertex_table)
-
-    def tau(self, k):
-        # evaluates functions defined in edge table
-        return(k, (self.edges[c][k](self.distances) for c in self.FB.F.keys()))
-
-    def view(self, axes):
-        """walk the edge_vertex table to return the edge function
-        """
-        table = defaultdict(list)
-        #sort since intervals lie along number line and are ordered pair neighbors
-        for (i, (start, end)) in sorted(zip(self.ids, self.vertices), key=lambda v:v[1][0]):
-            table['index'].append(i)
-            # same as view for line, returns nested list
-            for (name, value) in zip(self.FB.F.keys(), self.tau(i, simplex)[1]):
+          for (name, value) in zip(self.FB.fiber.keys(), 
+                                   self.tau(k)[1]):
               table[name].append(value)
@@
 \end{minted}
-where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix.
-The multicolored line is a connected graph of edges with each edge function evaluated on 1000 samples
+In both cases, the \mintinline{python}{view} method packages the data into a data structure that the artist can unpack via data component name, akin to a table with column names when \dbase\ is 0 or 1 D.
+
+\begin{figure}[htb]
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/linec_1.png}
+    \end{subfigure}
+    \begin{subfigure}{.5\textwidth}
+        \includegraphics[width=\textwidth]{figures/code/lined_1.png}
+    \end{subfigure}
+\caption{Continuous and discontinuous lines as defined via the same data model, and generated with the same \vartisteq \mintinline{python}{Line}}
+\label{fig:code:multilines}
+\end{figure}
+The graphics in figure \autoref{fig:code:multilines} are made using the \mintinline{python}{Line} artist and the \mintinline{python}{GraphData} data source where if told that the data is connected, the data source will check for that connectivity by constructing an adjacency matrix.
+The multicolored line is a connected graph of edges with each edge function evaluated on 100 samples,
 \begin{minted}{python}
-simplex.GraphLine(FB, edge_table, vertex_table, connect=True)
+GraphData(FB, edges, verticies, num_samples=100, connect=True)
 \end{minted}
-while the stair chart is discontinuous and only needs to be evaluated at the edges of the interval 
+which is an arbitrary choice made to display a smooth curve.
+The axes can also be used to choose an appropriate number of samples.
+In contrast, the stair chart is discontinuous and only needs to be evaluated at the edges of the interval
 \begin{minted}{python}
-simplex.GraphLine(FB, edge_table, vertex_table, num_samples=2, connect=False)
+GraphData(FB, edges, verticies, num_samples=2, connect=False)
 \end{minted}
@@
 \label{sec:code_case_study}
-For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020} since it is multivariate and has a varying number of penguins.
-We use a version of the data packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.
+Building on the \mintinline{python}{Bar} artist in \autoref{sec:code:artists}, we implement grouped bar charts as these do not exist out of the box in the current version of Matplotlib.Instead, grouped bar charts are often achived via looping over an implementation of bar, which forces the user to keep track which values are mapped to a single visual element and how that is achieved.
+For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020}, packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.
 The wrapper is very thin because there is explicitly only one section.
@@
 \end{figure}
-The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both out of \mintinline{python}{Bar} artists such that the difference between \mintinline{python}{StackedBar} and \mintinline{python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are stitched together.
+The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both composites of \mintinline{python}{Bar} artists such that the difference between \mintinline{python}{StackedBar} and \mintinline{python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are stitched together.
 These two artists have identical constructors and \mintinline{python}{draw} methods.
 As with \mintinline{python}{Bar}, the orientation is set in the constructor.
-In both these artists, we separate the transforms applied to only one component and the case \mintinline{python}{mtransforms} where the same transform is applied to multiple components such that \vtotal\ has multiple components that map to the same retinal variable.
+In both these artists, we separate the transforms that are applied to only one component (column) from transforms applied to multiple components (columns).
+We encode the one component case as \mintinline{python}{V} dictionary, and the multiple components case as \mintinline{python}{MV}.
+This convention allows us to, for example, map one column to position, but multiple to length.
 
@@
 class StackedBar(martist.Artist):
-    def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
+    def __init__(self, E, V, MV, orientation='v', *args, **kwargs):
         """
@@
         super().__init__(*args, **kwargs)
-        self.data = data
+        self.E = E
         self.orientation = orientation
-        self.transforms = copy.deepcopy(transforms)
-        self.mtransforms = copy.deepcopy(mtransforms)
+        self.V = V
+        self.MV = MV
 
     def assemble(self):
-        view = self.data.view(self.axes)
+        tau = self.data.view(self.axes)
         self.children = [] # list of bars to be rendered
         floor = 0
-        for group in self.mtransforms:
+        for group in self.MV:
             # pull out the specific group transforms
-            group['floor'] = floor
-            group.update(self.transforms)
-            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)
+            bar = Bar(self.E, {**group, **self.V, 'floor':floor}, self.orientation, transform=self.axes.transData)
             self.children.append(bar)
@@
 Since all the visual transformation is passed through to \mintinline{python}{Bar}, the \mintinline{python}{draw} method does not do any visual transformations.
-In \mintinline{python}{StackedBar} the \mintinline{python}{view} is used to adjust the \mintinline{python}{floor} for every subsequent bar chart since a stacked bar chart is bar chart area marks concatenated together in the length parameter.
+In \mintinline{python}{StackedBar} the \mintinline{python}{view} is used to adjust the \mintinline{python}{floor} for every subsequent bar chart, since a stacked bar chart is bar chart area marks concatenated together in the length parameter.
 In contrast, \mintinline{python}{GroupedBar} does not even need the view, but instead keeps track of the relative position of each group of bars in the visual only variable \mintinline{python}{offset}.
@@
             width = group.get('width', .8)
-            group['width'] = width/ngroups
-            group['offset'] = gid/ngroups*width 
-            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)     
+            gwidth = width/ngroups
+            offset = gid/ngroups*width 
+            bar = Bar(self.E, {**group, **self.V, 'width':gwidth, 'offset':offset}, 
+                                self.orientation, transform=self.axes.transData)     
             self.children.append(bar)
@@
 We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic.
-
 \end{document}

### 2021-04-21 draft/sections/math.tex
code section

@@
 
-\section{Topological Artist Model}
-\label{sec:tam}
+\section{Topological Equivariant Artist Model}
+\label{sec:math}
 To guide the implementation of structure preserving building block components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}.

### 2021-04-21 code/matplottoy/artists/bar.py

@@ -100,9 +100,7 @@ class StackedBar(martist.Artist):
   
         for group in self.mtransforms:
             # pull out the specific group transforms
-            group['floor'] = floor
-            group.update(self.transforms)
-            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)
+            bar = Bar(self.data, {**group, **self.transforms, 'floor':floor}, self.orientation, transform=self.axes.transData)
             self.children.append(bar)
             floor += view[group['length']['name']]
             
@@ -137,10 +135,11 @@ class GroupedBar(martist.Artist):
         
         for gid, group in enumerate(self.mtransforms):
             group.update(self.transforms)
-            width = group.get('width',6)
-            group['width'] = width/ngroups
-            group['offset'] = gid/ngroups*width 
-            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)     
+            width = group.get('width',.8)
+            gwidth = width/ngroups
+            offset = gid/ngroups*width 
+            bar = Bar(self.data, {**group, **self.transforms,'width':gwidth, 'offset':offset},
+                        self.orientation, transform=self.axes.transData)     
             self.children.append(bar)
             
             

### 2021-04-21 draft/sections/next.tex
schedule

@@
 Additionally, implementing the complex addition operator and explicit graphic to data maps would allow for developing a mathematical formalism and prototype of how interactivity would work in this model.
-In summary, the proposed scope of work for the dissertation is
-\begin{itemize}
-    \item expansion of the mathematical framework to include complex addition
-    \item formalization of definition of equivalance class \vartisteq
-    \item implementation of artist with explicit \vindex\
-    \item specification of interactive visualization
-    \item mathematical formulation of a graphic with axes labeling
-    \item implementation of new prototype artists that do not inherit from Matplotlib artists
-    \item provisional mathematics and implementation of user level composite artists
-    \item proof of concept domain specific user facing library 
-\end{itemize}
- Other potential tasks for future work is implementing a data object for a non-trivial fiber bundle and exploiting the models section level formalism to build distributed data source models and concurrent artists.
-This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
+
+In summary, the optimistic proposed scope of work is
+
+\begin{table}[H]
+    \centering
+    \renewcommand{\arraystretch}{2}
+    \begin{tabulary}{\textwidth}{|l|L|}\hline
+    \textbf{work period} & \textbf{milestones \& tasks} \\ \hline
+    April-July 2021 & prepare \textbf{conference presentation} on new functionality enabled by model for \textit{SciPy}: artists that do not inherit from existing Matplotlib artists, computational artists such as histograms, non tabular data, composite interactive artist \\ \hline
+    June-September 2021  & prepare \textbf{theory paper} on interactivity for submission to \textit{TCVG or Eurovis 2022}: fully work out and describe math of addition operators and lookups from graphic to data space, implement brush linked artist (shared base space) and artist that exploits sheafs \\ \hline
+    May - November 2021 & prepare \textbf{applications paper} on high dimensional data for submission to \textit{TCVG}: math and implementation of computational artists, concurrent artists and data sources, non-trivial data bundles \\ \hline
+    August 2020-March 2022 & prepare \textbf{systems paper} on building domain specific libraries based on this model for \textit{Infoviz 2022}: domain specific structured data, composite artists, inference of meta data components, mathematical notion of a visualization (labeled, multiple artists, etc)\\ \hline
+    April 2022 & \textbf{dissertation} writing: synthesize previous work on climate data, compile topological equivariant artist model work \\ \hline
+    May 2022 & \textbf{defense} \\ \hline
+    \end{tabulary}
+    \caption{}
+    \label{tab:code:schedule}
+\end{table}
+In acknowledgement that the proposed scope is optimistic, the data applications of this work could be revisited after the defense as they are not integral to visualization library architecture.
+The data applications could be further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
 Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.

### 2021-04-21 draft/sections/next.tex
conclusion + bar refactor

--- 2021-03-10 draft/sections/next.tex
@@
 \section{Discussion}
-
-This work contributes a mathematical description of the mapping \vartist\ from data to visual representation.
-Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables,, multivariate high resolution spatio temporal datasets, and complex networks.
-Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications for producing visualization where the structure and properties match those of the input data.
-These specifications are that the graphic must have continuity equivalent to the data, and that the visual characteristics of the graphics are equivariant to their corresponding components under monoid actions.
-This model defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic.
-Encoding the graphic space as a fiber bundle provides a structure rich abstraction of the target graphical design in the target display space.
-
-The toy prototype built using this model validates that is usable for a general purpose visualization tool since it can be iteratively integrated into the existing architecture rather than starting from scratch.
-Factoring out glyph formation into assembly functions allows for much more clarity in how the glyphs differ.
-This prototype demonstrates that this framework can generate the fundemental marks: point (scatter plot), line (line chart), and area (bar chart).
-Furthermore, the grouped and stacked bar examples demonstrate that this model supports composition of glyphs into more complex graphics.
-These composite examples also rely on the fiber bundles section base book keeping to keep track of which components contribute to the attributes of the glyph.
-Implementing this example using a Pandas dataframe demonstrates the ease of incorporating existing widely used data containers rather than requiring users to conform to one standard.
+This work contributes a functional model of the structure-preserving maps from data to visual representation to guide the development of visualization libraries, thereby providing a means to express the constraints that visualization must preserve continuity and faithfully translate the properties of the data variables into visual variables.
+Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks.
+By decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ maps, the model expresses the specifications that graphic and data must have equivalent \textit{continuity} equivalent to the data, and that the visual characteristics of the graphics are \textit{equivariant} to their corresponding components under monoid actions.
+TEAM defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic.
+The toy prototype built using this model validates that it is usable for a general purpose visualization tool since it can be iteratively integrated into existing architecture rather than starting from scratch.
+Factoring out graphic formation into assembly functions allows for much more clarity in how they differ.
+This prototype demonstrates that this framework can generate the fundamental point (scatter plot) and line (line chart) marks.
 
 \subsection{Limitations}
-
-So far this model has only been worked out for a single data set tied to a primitive mark, but it should be extensible to compositing datasets and complex glyphs.
-The examples and prototype have so far only been implemented for the static 2D case, but nothing in the math limits to 2D and expansion to the animated case should be possible because the model is formalized in terms of the sheaf.
-While this model supports equivariance of figurative glyphs generated from parameters of the data\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it currently does not have a way to evaluate the semantic accuracy of the figurative representation.
-Effectiveness is out of scope for this model because it is not part of the structure being preserved, but potentially a developer building a domain specific library with this model could implement effectiveness criteria in the artists.
-Also, even though the model is designed to be backend and format independent, it has only really been tested against PNGs rendered with the AGG backend.
-It is especially unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997}.
-Because this model has been limited to the graphic design space, it does not address the critical task of laying out the graphics in the image
-
-This model and the associated prototype is deeply tied to Matplotlib's existing architecture.
-While the model is expected to generalize to other libraries, such as those built on Mackinlay's APT framework, this has not been worked through.
-In particular, Mackinlay's formulation of graphics as a language with semantic and syntax lends itself a declarative interface\cite{loudenProgrammingLanguagesPrinciples2010}, which Heer and Bostock use to develop a domain specific visualization language that they argue makes it simpler for designers to construct graphics without sacrificing expressivity \cite{heerDeclarative2010}.
-Similarly, the model presented in this work formulates visualization as equivariant maps from data space to visual space, and is designed such that developers can build software libraries with data and graphic topologies tuned to specific domains.
+Our model and prototype are deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as VTK or D3.
+Even though the model is designed to be backend and format independent, it has only been tested using PNGs rendered with AGG\cite{shemanarevAntiGrainGeometry}.
+It is unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997} that implement different models of \gsection.
+While our model supports equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from data components\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it cannot evaluate the semantic accuracy of the figurative representation.
+Effectiveness criteria\cite{mackinlayAutomaticDesignGraphical1987,chambersGraphicalMethodsData1983a} are out of scope.
 
 \subsection{Future Work}
-While the model and prototype demonstrate that generation of simple marks from the data, there is a lot of work left to develop a model that underpins a minimally viable library.
-Foremost is implementing a data object that encodes data with a 2D continous topology and an artist that can consume data with a 2D topology to visualize the image\cite{toryRethinkingVisualizationHighlevel2004, haber1990visualization,hansen2011visualization} and also encoding a separate heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas} artist that consumes 1D discrete data.
-A second important proof of concept artist is a boxplot\cite{wickham40YearsBoxplots2011} because it is a graphic that assumes computation on the data side and the glyph is built from semantically defined components and a list of outliers.
-The model supports simple composition of glyphs by overlaying glyphs at the same position, but more work is needed to define an operator where the fiber bundles have shared $\gbase_2 \hookrightarrow \gbase_1$ such that fibers could be pulled back over the subset.
-While the model's simple addition supports axes as standalone artists with overlapping visual position encoding, the complex operator would allow for binding together data that needs to be updated together.
-Additionally, implementing the complex addition operator and explicit graphic to data maps would allow for developing a mathematical formalism and prototype of how interactivity would work in this model.
-In summary, the proposed scope of work for the dissertation is
-\begin{itemize}
-    \item expansion of the mathematical framework to include complex addition
-    \item formalization of definition of equivalance class \vartisteq
-    \item implementation of artist with explicit \vindex\
-    \item specification of interactive visualization
-    \item mathematical formulation of a graphic with axes labeling
-    \item implementation of new prototype artists that do not inherit from Matplotlib artists
-    \item provisional mathematics and implementation of user level composite artists
-    \item proof of concept domain specific user facing library 
-\end{itemize}
- Other potential tasks for future work is implementing a data object for a non-trivial fiber bundle and exploiting the models section level formalism to build distributed data source models and concurrent artists.
-This could be pushed further to integrate with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
+More work is needed to formalize the composition operators, equivalence class \vartisteq, and the mathematical model of interactivity.
+We also need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}.
+In summary, the optimistic proposed scope of work is
+
+\begin{table}[H]
+    \centering
+    \renewcommand{\arraystretch}{2}
+    \begin{tabulary}{\textwidth}{|l|L|}\hline
+    \textbf{work period} & \textbf{milestones \& tasks} \\ \hline
+    April-July 2021 & prepare \textbf{conference presentation} on new functionality enabled by model for \textit{SciPy}: artists that do not inherit from existing Matplotlib artists, computational artists such as histograms, non tabular data, composite interactive artist \\ \hline
+    June-September 2021  & prepare \textbf{theory paper} on interactivity for submission to \textit{TCVG or Eurovis 2022}: fully work out and describe math of addition operators and lookups from graphic to data space, implement brush linked artist (shared base space) and artist that exploits sheafs \\ \hline
+    May - November 2021 & prepare \textbf{applications paper} on high dimensional data for submission to \textit{TCVG}: math and implementation of computational artists, concurrent artists and data sources, non-trivial data bundles \\ \hline
+    August 2020-February 2022 & prepare \textbf{systems paper} on building domain specific libraries based on this model for \textit{Infoviz 2022}: domain specific structured data, composite artists, inference of meta data components, mathematical notion of a visualization (labeled, multiple artists, etc)\\ \hline
+    March 2022 & \textbf{dissertation} writing: synthesize previous work on climate data, compile topological equivariant artist model work \\ \hline
+    April 2022 & \textbf{defense} \\ \hline
+    \end{tabulary}
+    \caption{}
+    \label{tab:code:schedule}
+\end{table}
+In acknowledgement that the schedule is optimistic, the data applications of this work could instead be revisited after the defense as that work is not integral to visualization library architecture.
+The data applications could be further integrated with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
 Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
@@
 \section{Conclusion}
-An unoffical philosophy of Matplotlib is to support making whatever kinds of plots a user may want, even if they seem nonsensical to the development team.
-The topological framework described in this work provides a way to facilitate this graph creation in a rigorous manner; any artist that meets the equivariance criteria described in this work by definition generates a graphic representation that matches the structure of the data being represented.
-We leave it to domain specialists to define the structure they need to preserve and the maps they want to make, and hopefully make the process easier by untangling these components into seperate constrained maps and providing a fairly general data and display model.
+A TEAM driven refactor of visualizations libraries could produce more maintainable, reusable, and extensible code, leading to better building blocks for the ecosystem of tools built on top of TEAM architectured libraries.
+Building block libraries could better support downstream, including domain specific, libraries without having to explicitly incorporate the specific data structure and visualization needs of those domains back into the base library.
+Adopting this model would induce a separation of data representation and visual representation that, for example, in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base.
+A refactor that incorporated the generalized data model and functional transforms presented in TEAM would lead to building block libraries that provide a more consistent, reusable, flexible, collection of blocks.
 \end{document}

### 2021-04-22 draft/sections/implementation.tex
fixed code

@@
     def __init__(self):
-      FB = FiberBundle({'tables': ['vertex']},  
+        self.FB = FiberBundle({'tables': ['vertex']},  
                        {'v1': float, 'v2': str, 'v3': float})
@@
     def __init__(self):
-    FB = FiberBundle({'tables': ['edge']},  
+        self.FB = FiberBundle({'tables': ['edge']},  
                      {'x': float, 'y': float, 'color': str})
@@
             # pull out the specific group transforms
-            bar = Bar(self.E, {**group, **self.V, 'floor':floor}, self.orientation, transform=self.axes.transData)
+            bar = Bar(self.E, {**group, **self.V, 'floor':floor}, 
+                      self.orientation, transform=self.axes.transData)
             self.children.append(bar)

### 2021-04-22 draft/sections/next.tex
updated schedule

@@
 We also need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}.
-In summary, the optimistic proposed scope of work is
+In summary, the proposed scope of work is
 
@@
     \renewcommand{\arraystretch}{2}
-    \begin{tabulary}{\textwidth}{|l|L|}\hline
+    \begin{tabularx}{\textwidth}{|l|X|}\hline
     \textbf{work period} & \textbf{milestones \& tasks} \\ \hline
-    April-July 2021 & prepare \textbf{conference presentation} on new functionality enabled by model for \textit{SciPy}: artists that do not inherit from existing Matplotlib artists, computational artists such as histograms, non tabular data, composite interactive artist \\ \hline
-    June-September 2021  & prepare \textbf{theory paper} on interactivity for submission to \textit{TCVG or Eurovis 2022}: fully work out and describe math of addition operators and lookups from graphic to data space, implement brush linked artist (shared base space) and artist that exploits sheafs \\ \hline
-    May - November 2021 & prepare \textbf{applications paper} on high dimensional data for submission to \textit{TCVG}: math and implementation of computational artists, concurrent artists and data sources, non-trivial data bundles \\ \hline
-    August 2020-February 2022 & prepare \textbf{systems paper} on building domain specific libraries based on this model for \textit{Infoviz 2022}: domain specific structured data, composite artists, inference of meta data components, mathematical notion of a visualization (labeled, multiple artists, etc)\\ \hline
-    March 2022 & \textbf{dissertation} writing: synthesize previous work on climate data, compile topological equivariant artist model work \\ \hline
+    April - July 2021  & prepare and submit \textbf{conference presentation} on new functionality enabled by model for \textit{SciPy}: \newline artists that do not inherit from existing Matplotlib artists, computational artists such as histograms, non tabular data, composite interactive artist \\ \hline
+    June - Sept 2021  & prepare and submit \textbf{theory paper} on interactivity to \textit{TCVG or Eurovis 2022}: \newline fully work out and describe math of addition operators and lookups from graphic to data space, implement brush linked artist (shared base space) and artist that exploits sheafs \\ \hline
+    May - Nov 2021 & prepare and submit \textbf{applications paper} on high dimensional to \textit{TCVG}: \newline math and implementation of computational artists, concurrent artists and data sources, non-trivial data bundles \\ \hline
+    Aug 2021 - Feb 2022 & prepare and submit \textbf{systems paper} on building domain specific libraries based on this model to \textit{Infoviz 2022}: \newline domain specific structured data, composite artists, inference of meta data components, mathematical notion of a visualization (labeled, multiple artists, etc)\\ \hline
+    March 2022 & \textbf{dissertation} writing: \newline synthesize previous work on climate data, compile topological equivariant artist model work \\ \hline
     April 2022 & \textbf{defense} \\ \hline
-    \end{tabulary}
+    \end{tabularx}
     \caption{}

### 2021-04-23 figures/intro/intro_plots.py
reworking figures

@@ -1,0 +1,74 @@
+import numpy as np
+
+import matplotlib as mpl
+
+import matplotlib.colors as mcolors
+import matplotlib.dates as mdates
+import matplotlib.table as mtable
+from mpl_toolkits.axes_grid1.axes_divider import make_axes_locatable
+
+
+cdict = {'GANG MILLS NEW YORK': '#e6194B',
+         'SARA NEW YORK': '#42d4f4',
+         'SCHROON LAKE NEW YORK': '#9A6324',
+         'SHERBURNE NEW YORK': '#4363d8',
+         'STONYKILL NEW YORK': '#469990',
+         'BINGHAMTON': '#3cb44b',
+         'ISLIP LI MACARTHUR AP': '#ffe119',
+         'NEW YORK LAGUARDIA AP': '#f58231',
+         'BUFFALO': '#f032e6',
+         'ALBANY AP': '#fabed4',
+         'GLENS FALLS AP': '#dcbeff',
+         'ROCHESTER GTR INTL AP': '#800000',
+         'SYRACUSE HANCOCK INTL AP': '#aaffc3',
+         'NEW YORK JFK INTL AP': '#000075'}
+
+def plot_table(ax, dfs, ccolors, edgecolor='k', textcolor='k'):
+    tab = mtable.table(ax, cellText=dfs.astype('str').values, cellLoc='center',
+                   colLabels=dfs.columns, loc='center')
+    #bbox=(0, 0, 1, 1))
+    tab.auto_set_font_size(False)
+    tab.set_fontsize('medium')
+    tab.scale(1.25, 1.5)
+    tab.auto_set_column_width(np.arange(len(dfs.columns)))
+    for i, color in enumerate(ccolors):
+        for j in range(len(dfs)+1):
+            if textcolor is None:
+                tab[(j,i)].get_text().set_color(color)
+            else:
+                tab[(j,i)].get_text().set_color(textcolor)
+            tab[(j,i)].set_edgecolor(edgecolor)    
+    
+    ax.set(xticks=[], yticks=[], aspect='equal')
+    ax.axis('off')
+    ax.add_table(tab)
+    return tab
+
+def plot_map(ax, gdf, nystate, cmap, norm, fade='k'):
+    for name in cdict:
+        df = gdf.loc[[name], ['TAVG','geometry']]
+        df.plot(ax=ax, facecolor=cmap(norm(df['TAVG'])), edgecolor=cdict[name], linewidth=2.5, alpha=.75, markersize=100)
+    
[…]

### 2021-04-23 figures/intro/intro_plots.py
anchoring to cell

@@ -72,3 +72,8 @@ def plot_time(ax, df, fade='k'):
     ax.spines[:].set_color(fade)
    
     
+def source_cell(cell, color='k', xr=.5, yr=.5):
+    cell.get_text().set_color(color)
+    x = cell.get_x() + cell.get_width()*xr
+    y = cell.get_y() + cell.get_height()*yr
+    return x, y

### 2021-04-25 figures/intro/intro_plots.py
still have weird scoping bug

@@ -5,8 +5,10 @@ import matplotlib as mpl
 import matplotlib.colors as mcolors
 import matplotlib.dates as mdates
 import matplotlib.table as mtable
+import matplotlib.patches as mpatches
 from mpl_toolkits.axes_grid1.axes_divider import make_axes_locatable
 
+#scatter position: 
 
 cdict = {'GANG MILLS NEW YORK': '#e6194B',
          'SARA NEW YORK': '#42d4f4',
@@ -34,43 +36,46 @@ def plot_table(ax, dfs, ccolors, edgecolor='k', textcolor='k'):
     for i, color in enumerate(ccolors):
         for j in range(len(dfs)+1):
             if textcolor is None:
-                tab[(j,i)].get_text().set_color(color)
+                tab[j,i].get_text().set_color(color)
             else:
-                tab[(j,i)].get_text().set_color(textcolor)
-            tab[(j,i)].set_edgecolor(edgecolor)    
+                tab[j,i].get_text().set_color(textcolor)
+            tab[j,i].set_edgecolor(edgecolor)    
     
     ax.set(xticks=[], yticks=[], aspect='equal')
     ax.axis('off')
     ax.add_table(tab)
     return tab
 
-def plot_map(ax, gdf, nystate, cmap, norm, fade='k'):
+def plot_map(ax, gdf, nystate, cmap, norm, fade='k', alpha=.5, station=None):
     for name in cdict:
+        a = 1 if name == station else alpha
         df = gdf.loc[[name], ['TAVG','geometry']]
-        df.plot(ax=ax, facecolor=cmap(norm(df['TAVG'])), edgecolor=cdict[name], linewidth=2.5, alpha=.75, markersize=100)
+        df.plot(ax=ax, label=name, facecolor=cmap(norm(df['TAVG'])), 
+                                edgecolor=mcolors.to_rgba(cdict[name], alpha=a), legend=False,
+                                linewidth=2.5, alpha=a, markersize=100)
     
-    mapplot= nystate[nystate['postal'].str.match('NY')].plot(ax=ax, 
-                                                             facecolor='white', 
-                                                             edgecolor=fade, zorder=-1)
+    nystate[nystate['postal'].str.match('NY')].plot(ax=ax,
+                                facecolor='white', edgecolor=fade, zorder=-1)
     
     ax1_divider = make_axes_locatable(ax)
[…]

### 2021-04-26 figures/intro_plots.py
moving around

@@ -14,16 +14,22 @@ cdict = {'GANG MILLS NEW YORK': '#e6194B',
          'SARA NEW YORK': '#42d4f4',
          'SCHROON LAKE NEW YORK': '#9A6324',
          'SHERBURNE NEW YORK': '#4363d8',
-         'STONYKILL NEW YORK': '#469990',
+         'STONYKILL NEW YORK': '#fabed4',
          'BINGHAMTON': '#3cb44b',
          'ISLIP LI MACARTHUR AP': '#ffe119',
          'NEW YORK LAGUARDIA AP': '#f58231',
          'BUFFALO': '#f032e6',
-         'ALBANY AP': '#fabed4',
+         'ALBANY AP':'#469990',
          'GLENS FALLS AP': '#dcbeff',
          'ROCHESTER GTR INTL AP': '#800000',
          'SYRACUSE HANCOCK INTL AP': '#aaffc3',
          'NEW YORK JFK INTL AP': '#000075'}
+
+ssubset = ['ROCHESTER GTR INTL AP','STONYKILL NEW YORK', 
+           'ALBANY AP', 'SCHROON LAKE NEW YORK', 
+           'SARA NEW YORK','NEW YORK LAGUARDIA AP',
+           'NEW YORK JFK INTL AP', 'SYRACUSE HANCOCK INTL AP', 
+           'ISLIP LI MACARTHUR AP', 'GLENS FALLS AP']
 
 def plot_table(ax, dfs, ccolors, edgecolor='k', textcolor='k'):
     tab = mtable.table(ax, cellText=dfs.astype('str').values, cellLoc='center',
@@ -49,17 +55,18 @@ def plot_table(ax, dfs, ccolors, edgecolor='k', textcolor='k'):
 def plot_map(ax, gdf, nystate, cmap, norm, fade='k', alpha=.5, station=None):
     for name in cdict:
         a = 1 if name == station else alpha
-        df = gdf.loc[[name], ['TAVG','geometry']]
-        df.plot(ax=ax, label=name, facecolor=cmap(norm(df['TAVG'])), 
-                                edgecolor=mcolors.to_rgba(cdict[name], alpha=a), legend=False,
-                                linewidth=2.5, alpha=a, markersize=100)
+        df = gdf.loc[[name], ['PRCPI','geometry']]
+        df.plot(ax=ax, label=name, facecolor=cmap(norm(df['PRCPI'])), 
[…]

### 2021-04-28 draft/sections/intro.tex
new intro

@@
     \includegraphics[width=1\textwidth]{figures/intro/functions.png}
-    \caption{Visualizations are made up of transformations from data into visual representation.
-These functions transform individual data values to visual representation, such as date to x position or latitude to y position.
-These functions are composed into the assembly of all these transformations into a visual mark, such as a line or point.
-%%% break this out into seperate sentences to break
-    The same variable can be mapped in different ways, for example line is mapped to a color in the scatter plot and to y position in the line plot.}
+    \caption{Building block visualization libraries implement the functions that encode data as visual representations.
+For example, functions that map latitude and longitude into positions on a graphical representation of a map and functions that map precipitation and station name to colors.
+Visualization libraries also provide functions that composite these encoding functions into attributes of a visual element, such as a point, and we call these functions \textit{artists}.
+The encoders and \textit{artists} are reusable, so position encoders can also be used to to translate date and time to positions, which can then be composited into a line by a \texttt{line} function.
+    }
     \label{fig:intro:artist}
 \end{figure}
-Visualization is the transformation of data into visual representation.
-As illustrated by \autoref{fig:intro:artist}, these translations are both at the level of the individual variable and the entire record.
-In the case of the scatter plot, the latitude and longitude are encoded as the x and y position, respectively, while the temperature and station are represented by the face and edge colors.
-A row in the table is collectively encoded as a point mark.
-None of these encodings are fixed, as evidenced by temperature being translated into the y value in the case of the line plot.
-The station is now the source of the color of the entire line, and the date is the x position.
-As with scatter, the encodings of the individual transformations, which again are on values from the same record in the table, are composited into a line mark.
-It is these raw transformations from data space to visualization space that are implemented by building block level visualization libraries, named as such because the functions provided by the library can be composited in any number of ways to yield visualizations \cite{wongsuphasawatNavigatingWideWorld2021}.
-We propose that like physical building blocks, building block libraries must provide a collection of well defined pieces that can be composed in whichever ways the blocks fit together.
-We specify that a valid visualization block is a structure preserving transformation from data to visual space, and we define structure in terms of \textit{continuity} and \textit{equivariance}.
-We then use this model to develop a design specification for the components of a building block visualization library.
-The notion of self contained, inherently modular, building blocks lends itself naturally to a functional paradigm of visualization \cite{hughesWhyFunctionalProgramming1989}.
-We adopt a functional model for a redesign because the lack of side effects means functional architecture can be evaluated for correctness, functional programs tend to be shorter and clearer, and are well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.
+
+Visualizations, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008} of the data, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+We define visualization components to be structure preserving maps from data to visual representations.
+As illustrated in \autoref{fig:intro:artist}, visualizations map data into graphical elements such as points and lines and map individual data components (columns) into components of the visualization such as position or color.
+For example, the gray position encoder function converts the latitude and longitude to x and y positions, and the color encoders map temperature and station name to colors.
+A \texttt{point} function composites these encodings into attributes of a point on screen.
+The \texttt{line} function composites position and color encodings into a line, specifically encoders that convert date and precipitation to x and y positions and a map from station name to color.
+We identify the structure these maps must preserved as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.
+
+We introduce a model of visualization components based on these \textit{continuity} and \textit{equivariance} constraints and use this model to develop a design specification.
+Our specification is targeted at building block libraries, as defined by Wongsuphasawat\cite{wongsuphasawatNavigatingWideWorld2021}, because they are the visualization tools that provide independent functions that map components of data to visual representations.
+Just as physical building blocks can be arranged in an almost infinite number of ways, solely constrained by the size and shape of the blocks, we propose that the only restriction on the composition of building block library components are \textit{continuity} and \textit{equivariance}.
+We adopt a functional model for our proposed architecture because independent, modular, components are inherently functional\cite{hughesWhyFunctionalProgramming1989}.
+Doing so means the functional architecture can be evaluated for correctness, the resulting code is likely to be shorter and clearer, and the architecture is well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.
 
 This work is strongly motivated by the needs of the Matplotlib\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library.
+We even name these compositing functions \texttt{point} and \texttt{line} in \autoref{fig:intro:artist} \textit{artists} as a nod to the somewhat analogous object in the Matplotlib architecture.
 One of the most widely used visualization libraries in Python, since 2002 new components and features have been added in a some what adhoc, sometimes hard to maintain, manner.
@@
 \label{sec:intro:structure}
-Visual representations of data, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
  The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.

### 2021-04-29 draft/sections/intro.tex
reworked viz section

@@
     \includegraphics[width=1\textwidth]{figures/intro/functions.png}
-    \caption{Building block visualization libraries implement the functions that encode data as visual representations.
-For example, functions that map latitude and longitude into positions on a graphical representation of a map and functions that map precipitation and station name to colors.
-Visualization libraries also provide functions that composite these encoding functions into attributes of a visual element, such as a point, and we call these functions \textit{artists}.
-The encoders and \textit{artists} are reusable, so position encoders can also be used to to translate date and time to positions, which can then be composited into a line by a \texttt{line} function.
-    }
+    \caption{Building block visualization libraries implement independent reusable functions that map data to visual representations.
+For example, position encoding functions map the latitude and longitude values in the table (a) to locations on the map (b).
+These same encoding functions map dates and precipitation (a) to x and y positions in the timeseries plot (c).
+Color encoding functions map the precipitation (a) to the color of the point in the map (a) and color encoding functions map names of weather stations (a) to colors in both the map (a) and timeseries (b).
+Functions, which we call \textit{artists}, compose these encoding functions into attributes of visual elements.
+The \texttt{point} artist transforms the output of the position, facecolor, and edgecolor maps into attributes of each point in the map (a), while the \texttt{line} artist transforms the output of the position and color maps into attributes of each line in the timeseries plot (c).}
     \label{fig:intro:artist}
@@
 We define visualization components to be structure preserving maps from data to visual representations.
-As illustrated in \autoref{fig:intro:artist}, visualizations map data into graphical elements such as points and lines and map individual data components (columns) into components of the visualization such as position or color.
+As illustrated in \autoref{fig:intro:artist}, visualizations map weather station precipitation data (a) into graphical elements such as points (b) and lines (c) and map individual data components (columns) into components of the visualization such as position or color.
 For example, the gray position encoder function converts the latitude and longitude to x and y positions, and the color encoders map temperature and station name to colors.
-A \texttt{point} function composites these encodings into attributes of a point on screen.
-The \texttt{line} function composites position and color encodings into a line, specifically encoders that convert date and precipitation to x and y positions and a map from station name to color.
-We identify the structure these maps must preserved as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.
+A \texttt{point} function composites these encodings into attributes of a point in the map (a).
+The \texttt{line} function composites encoders that convert station name to color and date and precipitation to x and y positions into a piece of a line in the timeseries plot (c).
+We identify the structure these maps must preserve as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.
 
@@
 Our specification is targeted at building block libraries, as defined by Wongsuphasawat\cite{wongsuphasawatNavigatingWideWorld2021}, because they are the visualization tools that provide independent functions that map components of data to visual representations.
-Just as physical building blocks can be arranged in an almost infinite number of ways, solely constrained by the size and shape of the blocks, we propose that the only restriction on the composition of building block library components are \textit{continuity} and \textit{equivariance}.
-We adopt a functional model for our proposed architecture because independent, modular, components are inherently functional\cite{hughesWhyFunctionalProgramming1989}.
+Just as the number of ways to arrange physical building blocks is solely constrained by the size and shape of the blocks, we propose that the only restriction on how building block library components can be composed are that the compositions preserve \textit{continuity} and \textit{equivariance}.
+Independent, modular, components are inherently functional\cite{hughesWhyFunctionalProgramming1989}, so we propose a functional architecture for our model.
 Doing so means the functional architecture can be evaluated for correctness, the resulting code is likely to be shorter and clearer, and the architecture is well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.
@@
 This work is strongly motivated by the needs of the Matplotlib\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library.
-We even name these compositing functions \texttt{point} and \texttt{line} in \autoref{fig:intro:artist} \textit{artists} as a nod to the somewhat analogous object in the Matplotlib architecture.
-One of the most widely used visualization libraries in Python, since 2002 new components and features have been added in a some what adhoc, sometimes hard to maintain, manner.
-Particularly, each new component carries its own implicit notion of how it believes the data is structured-for example if the data is a table, cube, image, or network - that is then expressed in the API for that component.
-In turn, this yields an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards\cite{a.sarikayaWhatWeTalk2019}.
+One of the most widely used visualization libraries in Python, new components and features have been added in an organic, sometimes hard to maintain, manner since 2002.
+In Matplotlib, every component carries its own implicit notion of how it believes the data is structured-for example if the data is a table, cube, image, or network - that is expressed in the API for that component.
+This leads to an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards\cite{a.sarikayaWhatWeTalk2019}.
 This entangling of data model with visual transform also yields inconsistencies in how visual component transforms, e.g.
 shape or color, are supported.
-We propose that these issues can be ameliorated via a redesign of the functions that convert data to graphics, named \textit{Artists} in Matplotlib, in a manner that reliably enforces \textit{continuity} and \textit{equivariance} constraints.
+We propose a redesign of the functions that convert data to graphics, named \textit{Artists} in Matplotlib, in a manner that reliably enforces \textit{continuity} and \textit{equivariance} constraints.
 We evaluate our functional model by implementing new artists in Matplotlib that are specified via \textit{equivariance} and \textit{continuity} constraints.
@@
 \section{Background}
-There are many formalisms of the notion that visualization is structure preserving maps from data to visual representation, and many visualization libraries that attempt to preserve structure in some manner; this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library level to propose a new model of the constraints visual transformations must satisfy such that they can be composed to produce visualize representations that can be considered equivalent to the data being represented.
+There are many formal models of visualization as structure preserving maps from data to visual representation, and many implementations of visualization libraries that preserve this structure in some manner.
+This work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library level.
+We propose a data structure agnostic model of the constraints visual transformations must satisfy such that they can be composed to produce \textit{equivariant} and \textit{continuity preserving} visual representations.
 
@@
 \label{sec:intro:structure}
- The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.
 
 \begin{figure}[H]
-        \includegraphics[width=1\columnwidth]{figures/intro/equivariant.png}
-        \caption{The data in blue is scaled by a factor of two, yielding the data in orange.
-To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
+        \includegraphics[width=1\textwidth]{figures/intro/equivariant.png}
+        \caption{The precipitation in blue is scaled by a factor of 10, yielding the precipitation in orange in the table (a).
+This scaling factor is represented by the green arrows in (a).
+The \textit{equivariant} action is the scaling in the timeseries (b), which is represented by the green fill.
+Because the scalings are equivariant, the orange precipitation timeseries is 10x the blue precipitation timeseries.
+The red line is an example of \textit{equivariance} not being preserved, as it is not 10x the the y value of the blue precipitation timeseries.}
       \label{fig:intro:equivariance}
 \end{figure}
-Bertin proposes that there are classes of visual encodings-such as position, shape, color, and texture-that when mapped to from specific types of measurement, quantitative or qualitative, will preserve the properties of that measurement type.
-For example, in \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from blue to orange data and lines.
+The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.
+Bertin proposes that there are categories of visual encodings-such as position, shape, color, and texture-that preserve the properties of the measurement type, quantitative or qualitative, of the encoded data.
+For example, in \autoref{fig:intro:equivariance}, the blue precipitation data in the table (a) is scaled by a factor of 10.
+This scaling action is represented by the green arrows that translate the precipitation into the scaled values in orange.
+For this visualization to be equivariant, this same scaling factor must be present in the timeseries representation of the data (b).
+The precipitation in orange is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the orange line is 10x the y-values of the blue precipitation timeseries.
+The green fill in the timeseries (b) is the scaling equivalent to the arrows in the table (a).
+In contrast, the red line is not an \textit{equivariant} representation to the 10x scaling because the red y-values are not 10x the blue y values.
 The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}.
@@
     \includegraphics[width=1\textwidth]{figures/intro/continuity.png}
-    \caption{The line plot does not preserve \textit{continuity} because it implies that the average temperature at each station lie along a 1D continuous line, while the bar plot preserves \textit{continuity} by representing the average temperatures at each station as the discrete values they are.}
+    \caption{The bar chart (b) and line plot (c) are different visual encodings of the precipitation data in the table (a).
+The bar chart preserves the \textit{continuity} of the stations by encoding the discrete station data as discrete bars (b).
+In contrast, the line plot (c) does not preserve this discrete \textit{continuity} because it connects the total temperature at each station by drawing a line through each point.
+Doing so implies that the total temperatures are points along a 1D continuous line, whereas there is no connectivity between the stations or their corresponding rows in the table.}
     \label{fig:intro:continuity}
 \end{figure}
-Bertin proposes that the visual encodings be composited into graphical marks that match the \textit{continuity} of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark.
-In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval.
-But, when the continuity is preserved, as in the bar chart, then the graphic has not introduced new structure into the data.
+The notion that data is equivalent to visual representations when structure is preserved serves as the basis for visualization best practices.
+When \textit{continuity} is preserved, as in the bar chart (b) in \autoref{fig:intro:continuity}, then the graphic has not introduced new structure into the data.
+In \autoref{fig:intro:continuity}, the line plot (c) does not preserve \textit{continuity} because the line connecting the total precipitation of the stations implies that the stations are connected to each other along a 1D continuous line.
+Bertin asserted that \textit{continuity} was preserved by choosing graphical marks that match the \textit{continuity} of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark.
+Informally, Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis} states that a visualization is easier to understand when the properties of the visualization match the properties of the data.
+This principal is made more concrete in Tufte's concept of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements\cite{tufteVisualDisplayQuantitative2001}.
+Expressivity, as defined by Mackinlay, is a measure how much of the mathematical structure in the data can be expressed in the visualizations; for example that ordered variables can be mapped into ordered visual elements.
+We generalize these different codifications of structure preserving encodings, proposing that a graphic is an equivalent representation of the data when \textit{continuity} and \textit{equivariance} are preserved.
 
@@
     \begin{description}
-        \item[\textbf{continuity}] How records in the dataset are connected to each other, e.g. discrete rows, neworked nodes, points on a continuous surface
+        \item[\textbf{continuity}] How elements in the dataset are connected to each other, e.g. discrete points, networked nodes, points on a continuous surface
         \item[\textbf{equivariance}] if an action is applied to the data or the graphic--e.g.
@@
 \end{mdframed}
-
-The notion that a graphic should be equivalent to the data has been expressed in a variety of ways.
-Informally, Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis} states that a visualization is easier to understand when the properties of the visualization match the properties of the data.
-This principal is made more concrete in Tufte's concept of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements\cite{tufteVisualDisplayQuantitative2001}.
-expressing, as defined by Mackinlay, is a measure how much of the mathematical structure in the data that can be expressed in the visualizations; for example that ordered variables can be mapped into ordered visual elements.
-We propose that a graphic is an equivalent representation of the data when \textit{continuity} and \textit{equivariance} are preserved.
 

### 2021-04-29 draft/sections/intro.tex
tools

@@
         \includegraphics[width=1\textwidth]{figures/intro/table.png}
+        \caption{}
+        \label{fig:intro:table}
     \end{subfigure}
@@
         \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
+        \caption{}
+        \label{fig:intro:image}
     \end{subfigure}
@@
         \includegraphics[width=1\textwidth]{figures/math/graph.png}
+        \caption{}
+        \label{fig:intro:graph}
     \end{subfigure}
-    \caption{Visualization libraries, especially ones tied to specific domains, tend to be architectured around a core data structure, such as tables, images, or networks. }
+    \caption{Visualization libraries, especially ones tied to specific domains, tend to be architectured around a core data structure, such as tables \autoref{fig:intro:table}, images \autoref{fig:intro:image}, or networks \autoref{fig:intro:graph}.
+    %%revise as combined fig?
+    }
     \label{fig:intro:data}
@@
 
-One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures.
-These libraries can often assume the expected data structure because they are domain specific, and that is the common data structure in that domain.
-For users who generally work in one domain, such as the data, networks, or graphs shown in \autoref{fig:intro:data}, this well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a very coherent user experience\cite{heerDeclarative2010}.
-But, for developers who want to build new visualizations on top of these libraries, they must work around the existing assumptions, sometimes in ways that break the model the libraries are developed around.
-
-For example, many domain specific libraries integrate computation into the visualization, for example libraries based that assume all data is a relational database.
-This assumption is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, such as ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
-Since these libraries represent all data as a table, and computations on tables are fairly well defined\cite{ullmanFirstCourseDatabase2008}, they can include computations on the table with a fair bit of confidence that the computation is accurate.
-Since most computations are specific to domains, general purpose block libraries can not make this assumption; instead a goal of this model is to identify which computations are specifically part of the visual encoding - for example mapping data to a color-and which are manipulations on the data.
-Disentangling the computation from the visual transforms allows us to determine whether the visualization library needs to handle them or if they can be more efficiently computed by the data container.
-
-A different class of user facing tools are those that support images, such as ImageJ\cite{schneiderNIHImageImageJ2012} or Napari\cite{nicholas_sofroniew_2021_4533308}.
-These tools often have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized.
-These tools are ill suited for general purpose libraries that need to support data other than images because the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure.
-Even the digital humanities oriented ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supports some non-image aggregate reporting charts, is still built around image data as the primary input.
-The need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}.
-As with tables and images, extending network libraries to work with other types of data either require breaking their internal model of how data is structured and what transformations of the data are allowable or growing a model for other types of data structures alongside the network model.
+Most information visualization software design patterns are tuned to specific data structures and domains, as categorized by Heer and Agrawala\cite{HeerSoftware2006}.
+For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience\cite{heerDeclarative2010}.
+This coherant experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries.
+For example tools influenced by APT that assume that data is a relational table integrate computation into the visualization pipeline.
+This is a wide array of tools, including Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
+Since these libraries represent data as a table, and computations on tables are well defined\cite{ullmanFirstCourseDatabase2008}, these libraries implement computation as part of the visualization process.
+This is also true of tools that support images, such as the biology oriented ImageJ\cite{schneiderNIHImageImageJ2012} and  Napari\cite{nicholas_sofroniew_2021_4533308} or the digital humanities ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}.
+While these tools often have some support for visualizing non image components of the data, the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure.
+Tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs.
+As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model.
+Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distrobution statistics\cite{wickham40YearsBoxplots2011}.
+Disentangling the computation from the visual transforms allows us to determine whether the visualization library should implement computations or if they should be implemented in data space.
 
@@
     \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
-    \caption{One way to describe data is by the connectivity of the points in the dataset.
-A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid.
+    \caption{The \textit{continuity} of the data is often used to classify data structures.
+For example, a relational database often consists of discrete unconnected points, while an image is an implicitly connected 2D grid.
 This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
@@
 \end{figure}
-Many building block libraries carry multiple models of data internally because they cannot assume a data structure.
-Algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{ToryRethinkingVisualization2004}, and by definition building block libraries try to  provide the components to build any sort of visualization.
-Matplotlib, D3\cite{bostockDataDrivenDocuments2011}, and VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{RamachandranMayaVI2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}.
-Where GoG and ImageJ type libraries have coherant APIs for their visualization tools because the data structure is the same, the APIs for visualizations in Matplotlib, D3, and VTK are significantly dependent on the structure of the data it expects.
-VTK has codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}.
-This API choice can lead to visualizations that break \textit{continuity} when fed into visualizations with different assumptions about structure.
-The lack of consistent data model can also mean no consistent way of updating the data and therefore no way of guaranteeing that the views are in sync, in visualizations that consistent of multiple views of the same datasource, such as dashboards\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007}.
-To resolve this issue, our functional model takes as input a structure aware data abstraction general enough to provide a common interface for many different types of visualization.
+
+Visualizations assume the structure of the input data, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighlevel2004}, which leads to many building block libraries implementing multiple models of data to support these different visualizations.
+For example, in Matplotlib, D3\cite{bostockDataDrivenDocuments2011} VTK \cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} and MayaVi\cite{RamachandranMayaVI2011}, every plot is defined in terms of the \textit{continuity} of the data it expects as input.
+VTK has explicitly codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}.
+Downstream library developers impose some coherency by writing domain specific libraries with assumed data structures on top of the building block libraries--for example Seaborn\cite{waskom2020seaborn} and Titan\cite{brianwylieUnifiedToolkitInformation2009} assume a relational database, xarray\cite{hoyer2017xarray} and ParaView\cite{ahrens2005paraview} assume a data cube--but must work around the incoherencies in the building block libraries to do so.
+Our model navigates the tradeoff between coherency and extensibility by proposing functional composable well constrained visual components that take as input a structure aware data abstraction general enough to provide a common interface for many different types of visualizations.
 

### 2021-04-29 draft/sections/intro.tex
data, closes out background

@@
 Downstream library developers impose some coherency by writing domain specific libraries with assumed data structures on top of the building block libraries--for example Seaborn\cite{waskom2020seaborn} and Titan\cite{brianwylieUnifiedToolkitInformation2009} assume a relational database, xarray\cite{hoyer2017xarray} and ParaView\cite{ahrens2005paraview} assume a data cube--but must work around the incoherencies in the building block libraries to do so.
-Our model navigates the tradeoff between coherency and extensibility by proposing functional composable well constrained visual components that take as input a structure aware data abstraction general enough to provide a common interface for many different types of visualizations.
+Our model navigates the tradeoff between coherency and extensibility by proposing functional composable well constrained visual components that take as input a structure aware data abstraction general enough to provide a common interface for many different types of data continuities.
 
@@
 \label{sec:intro_data}
-One such general abstraction are fiber bundles, which Butler proposed as a core data structure for visualization because they encode data continuity separately from the variable properties and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
-Since Butler's model lacks a robust way of describing variables, we can encode a schema like description of the data in the fiber bundle by folding in Spivak's topological description of data types \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}.
-In this work we will refer to the points of the dataset as \textit{records} to indicate that a point can be a vector of heterogenous elements.
-Each \textit{component} of the record is a single object, such as a temperature measurement, a color value, or an image.
-We also generalize \textit{component} to mean all objects in the dataset of a given type, such as all temperatures or colors or images.
-The way in which these records are connected is the \textit{connectivity}, \textit{continuity}, or more generally \textit{topology}.
+Fiber bundles were proposed by Butler as a core data structure for visualization because they encode data \textit{continuity} separately from the components of the dataset\cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Butler's model lacks a robust way of describing variables; therefore we encode a schema like description of the data in the fiber bundle using Spivak's topological description of data types \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}.
+In this work, we refer to the points of the dataset as \textit{records}, as defined by Spivak.
+Each \textit{component} of the record is a single object, such as a precipitation measurement, a station name, or an image.
+We generalize \textit{component} to mean all objects in the dataset of a given type, such as all precipitation values or station names or images.
+The way in which these records are connected is the \textit{continuity} or more generally topology.
 
@@
 
-The \textit{continuity} can often be described by some variables in the dataset; this is formalized Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}.
-As shown in \autoref{fig:intro:keys:values}, keys can be labeled indexes, such as the attribute name and row ID, or physical entities such as locations on a map.
-We propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids.
-In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology.
-This allows the metadata to be altered, without imposing new semantics on the underlying structure, for example by changing the coordinate systems or time resolution.
-This value agnostic model also supports encoding datasets where there may be multiple independent variables - such as a measure of plant growth given variations in water, sunlight, and time - without having to assume any one variable is inducing the change in growth.
-For building block library developers, this means the components are able to fully traverse the data structures without having to know anything about the values or the semantic meaning of the structure.
-Since these components are by design \textit{equivariant} and \textit{continuity} preserving, domain specific library developers in different domains that both rely on the same continuity, for example 2D continuity, can then safely reuse the components to build tools that can safely make domain specific assumptions.
+The \textit{continuity} can be described in some datasets by components of the dataset.
+This is formalized by Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}.
+As shown in \autoref{fig:intro:keys:values}, keys can be labeled indexes, such as the attribute name and row ID, or semantically significant physical entities such as locations on a map.
+In contrast to Munzner's model, in out model components may describe the keys but are never themselves the keys; instead we propose that keys are points on a topological space encoding the continuity of the data.
+This allows the metadata to be altered without imposing new semantics on the underlying structure, for example by changing the coordinate systems or time resolution.
+This value agnostic model also supports encoding datasets where there may be multiple independent variables without having to assume any one variable is inducing the change , for example measures of plant growth given variations in water, sunlight, and time.
+
+For building block library developers, a non-semantic model of data continuity allows for the implementation of components that can traverse data structures without having to know the semantics of the data.
+Since these building block components are by design \textit{equivariant} and \textit{continuity preserving}, domain specific library developers in different domains that rely on the same continuity, for example 2D continuity, can use the same components to build tools that can make domain specific assumptions.
+One way we evaluate our model is by reproducing fully constrained domain specific library functionality using our proposed architecture.
 
@@
   \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
-  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:math:code:artist}
+  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code:artists}
 \end{enumerate}

### 2021-04-29 draft/sections/intro.tex
contribution sectuib as summary of answers to committee questions

@@
 This value agnostic model also supports encoding datasets where there may be multiple independent variables without having to assume any one variable is inducing the change , for example measures of plant growth given variations in water, sunlight, and time.
-
 For building block library developers, a non-semantic model of data continuity allows for the implementation of components that can traverse data structures without having to know the semantics of the data.
 Since these building block components are by design \textit{equivariant} and \textit{continuity preserving}, domain specific library developers in different domains that rely on the same continuity, for example 2D continuity, can use the same components to build tools that can make domain specific assumptions.
-One way we evaluate our model is by reproducing fully constrained domain specific library functionality using our proposed architecture.
 
 \subsection{Contribution}
-The contribution of this work is 
+In this work, we present a framework for understanding visualization as equivariant continuity preserving maps between topological spaces.
+Using this mathematical formalism, we develop an architecture specification developers can use to implement components in building block visualization libraries that domain specific library developers can carry through in the tools they build.
+Our work diverges from previous models of visualization and implementations of those models in that it contributes
+
 \begin{enumerate}
   \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
-  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist:nu}
+  \item formalization of property preservation from data component to visual representation as equivariant maps \autoref{sec:math:artist:nu}
   \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
@@
 \end{enumerate}
+We validate our model by using it to re-design artist and data access layer of Matplotlib, a general purpose visualization tool.
+We evaluate whether the redesign is successful by comparing it to existing implementation, recreating existing domain specific functionality with the new components, and by implementing components that provide new functionality in Matplotlib.
+While much of this functionality is currently possible in Matplotlib, a functional approach allows us to implement components in a more robust, modular, reliable way.
 \end{document}

### 2021-04-30 draft/sections/math.tex
closing out fig redos + labeling + consistently precip

@@
 \end{equation}
-which is equivalent to 
+which can also be written as 
 \begin{equation}
@@
 \end{equation}
-which allows us to decouple \dfiber\ into components $\dfiber_i$.
+which allows us to decouple \dfiber\ into components $\dfiber_i = \ftotal_{\fsection{(\fname_{i})}}$.
 
@@
     \caption{The fiber space is the set of all combinations of all theoretically possible values of the components.
-The 2D fiber \(\dfiber=\reals\times\reals\) encodes the properties of \textit{time} and \textit{temperature} components.
-One dimension of the fiber encodes the range of possible values for the time component of the dataset, which is a subset of the \reals, while the other dimension encodes the range of possible values  \reals\ for the temperature component.
-This means the fiber is the set of points \textit{(temperature, time)}that are all the combinations of \textit{temperatature} $\times$ \textit{time}.
-The 3D fiber encodes points at all possible combinations of \textit{temperature}, \textit{latitude}, and \textit{longitude}.}
+The 2D fiber \(\dfiber=\reals\times\realsp\) (a) encodes the properties of \textit{time} and \textit{precipitation} components.
+One dimension of the fiber encodes the range of possible values for the time component of the dataset, which is a subset of the \reals, while the other dimension encodes the range of possible values  \realsp\ for the precipitation component.
+This means the fiber is the set of points \textit{(precipitation, time)} that are all the combinations of \textit{precipitation} $\times$ \textit{time}.
+The 3D fiber (b) encodes points at all possible combinations of \textit{precipitation}, \textit{latitude}, and \textit{longitude}.}
     \label{fig:math:data:fiber}
@@
 
-For example, the records in the 2D fiber in \autoref{fig:data:fiber:example} are a pair of \textit{times} and \textdegree K \textit{temperature} measurements taken at those times.
+For example, the records in the 2D fiber (a) in \autoref{fig:math:data:fiber} are a pair of \textit{times} and \textdegree K \textit{precipitation} measurements taken at those times.
 Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$.
@@
 \end{equation*} 
-where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=temperature,\, \ftype=\texttt{float},\, \fttype=\reals)$ and is the set of values $\fttype=\reals$.
-In the 3D fiber in \autoref{fig:fiber_example_cube}, time is replaced with location.
+where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals)$ and is the set of values $\fttype=\reals$.
+In the 3D fiber (b) in \autoref{fig:math:data:fiber}, time is replaced with location.
 This location variable is of type \texttt{point} and has two components \textit{latitude} and \textit{longitude} $\{(lat,lon) \in \reals^{2} \mid  -90\leq lat \leq 90,\, 0 \leq lon \leq 360\}$.
@@
 \begin{equation*}
-    \dfiber = \reals \times \reals^{2} = \reals \times \reals \times \reals
+    \dfiber = \reals \times [0, 360] \times [-90, 90]
 \end{equation*} 
@@
 For example, \textit{location} can can either be specified as \((\fname = location,\, \ftype = \texttt{point},\, \fttype=\reals^{2})\)or \((\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\reals)\) and \((\fname=longitude,\, \ftype=\texttt{float},\,\fttype=\reals)\).
-
-As illustrated in \autoref{fig:math:data:fiber}, Spivak's framework provides a consistent way to describe potentially complex components of the input data.
 
@@
 
-One example of monoids are partial orderings on a set, such as seen in \label{fig:math:data:monoid}.
+One example of monoids are partial orderings on a set, such as seen in \autoref{fig:math:data:monoid}.
 Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one.
@@
     \includegraphics[width=\linewidth]{figures/math/k_qspace.png}
-    \caption{The base space \dtotal\ is divided into fiber segments \dfiber. The base space \dbase\ acts as an index into the records in the fibers.}
+    \caption{The total space \dtotal\ is divided into fiber segments \dfiber.
+The base space \dbase\ acts as an index into the records in the fibers, such that every point \dbasepoint\ has a corresponding fiber $\dfiber_{\dbasepoint}$.}
     \label{fig:math:data:base:div}
@@
 
-As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose}, we can decompose the total space into component bundles$\pi:\dtotal_i\rightarrow \dbase$ where
+As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose}, we can decompose the total space into component bundles $\pi:\dtotal_i\rightarrow \dbase$ where
 \begin{equation}
@@
     \includegraphics[width=1\textwidth]{figures/math/base.png}
-    \caption{These two datasets have the same \textit{(time, temperature)} fiber, but different continuities.
-The dataset on the left consists of discrete records, while the records in the dataset on the right sampled from a continuous space.}
+    \caption{The fiber bundles in (a) and (b) encode the two component dataset from \autoref{fig:math:data:fiber}, with \textit{(time, precipitation)} components, as having different continuities.
+The fiber bundle with discrete continuity (a) encodes the dataset as being a set of discrete records.
+The fiber bundle over the continuous interval \dbase\ (b) encodes the records as if they were sampled from a 1D continuous space.}
     \label{fig:math:data:base}
@@
 
-The datasets in \autoref{fig:math:data:base} have the same fiber of (temperature, time).
+The datasets in \autoref{fig:math:data:base} have the same fiber of (precipitation, time).
 The dot represents a discrete base space \dbase, meaning that every dataset encoded in the fiber bundle has discrete continuity.
@@
     \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
-    \caption{ Fiber (time, temperature) with an interval \dbase\ basespace.
-The sections $\dsection^{(1)}$ and $\dsection^{(2)}$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of temperature values.
+    \caption{ Fiber (time, precipitation) with an interval \dbase\ basespace.
+The sections $\dsection^{(1)}$ and $\dsection^{(2)}$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of precipitation values.
 They are included in the global set of sections  $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$}
@@
 
-In \autoref{fig:math:data:sections}, the fiber is the same encoding of $(time, \, temperature)$ illustrated in \autoref{fig:math:data:fiber}, and the base space is the interval \dbase\ shown in \autoref{fig:math:data:base}.
+In \autoref{fig:math:data:sections}, the fiber is the same encoding of $(time, \, precipitation)$ illustrated in \autoref{fig:math:data:fiber}, and the base space is the interval \dbase\ shown in \autoref{fig:math:data:base}.
 The section $\dsection^{(1)}$ is a function that for a point \dbasepoint\ returns a record in the fiber \dfiber.
-The section applied to a set of points in \dbase\ resolves to a series of monotonically increasing in time records of \textit{(time, temperature)} values.
-Section $\dsection^{(2)}$ returns a different timeseries of \textit{(time, temperature)} values.
+The section applied to a set of points in \dbase\ resolves to a series of monotonically increasing in time records of \textit{(time, precipitation)} values.
+Section $\dsection^{(2)}$ returns a different timeseries of \textit{(time, precipitation)} values.
 Both sections are included in the global set of sections $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$.
@@
 A series could store the values of $\dsection^{(1)}$ and a second series could be  $\dsection^{(2)}$.
-We could also fatten the fiber to hold two temperature series, such that a section would be an instance of a dataframe with a time column and two temperature columns.
+We could also fatten the fiber to hold two precipitationure series, such that a section would be an instance of a dataframe with a time column and two precipitation columns.
 While the series and dataframe explicitly have a time index column, they are components in our model and the index is assumed to be data independent references such as hashvalues, virtual memory locations, or random number keys.
@@
 Where this model particularly shines are N dimensional labeled data structures.
-For example, an xarray\cite{hoyer2017xarray} data that stores temperature field could have a \dbase\ that is a continuous volume and the components would be the temperature and the time, latitude, and longitude the measurements were sampled at.
+For example, an xarray\cite{hoyer2017xarray} data that stores precipitation field could have a \dbase\ that is a continuous volume and the components would be the precipitation and the time, latitude, and longitude the measurements were sampled at.
 A section can also be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.
@@
 \label{sec:math:graphic} 
-To establish that the artist is structure preserving map from data \dtotal\ to graphic \gtotal\, we construct a graphic bundle so that we can define \textit{equivariance} in terms of maps on the fiber spaces and \textit{continuity} in terms of maps on the base space.
+To establish that the artist is a structure preserving map from data \dtotal\ to graphic \gtotal\, we construct a graphic bundle so that we can define \textit{equivariance} in terms of maps on the fiber spaces and \textit{continuity} in terms of maps on the base space.
 As with the data, we can represent the target graphic as a section \gsection\ of a bundle  $(\gtotal, \gbase, \pi, \gfiber)$.
@@
 \label{sec:math:graphic:fiber}
-To fully specify the visual characteristics of the image, we construct a fiber \gfiber\ that is an infinite resolution version of the target space.
+To fully specify the visual characteristics of the image, we construct a fiber \gfiber\ that is a non-pixelated version of the target space.
 Typically \gtotal\ is trivial and therefore sections can be thought of as mappings into \gfiber.
@@
     \caption{For a visualization component to preserve continuity, it must have a continuous surjective map $\vindex: \gbase\rightarrow\dbase$ from graphic continuity to data continuity.
-The scatter and line graphic base spaces \gbase\ have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
-The image has the same dimension in \gbase\ as in \dbase.}
+The scatter (a) and line (b) graphic base spaces \gbase\ have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
+The image (c) has the same dimension in \gbase\ as in \dbase\ because \dbase\ is already 2D and therefore can directly map into screen space.}
     \label{fig:math:graphic:retraction:map}
@@
     \end{tikzcd}
+    \label{eq:math:graphic:vindex}
 \end{equation}
@@
 This means that if $\vindex(\gbasepoint) = \dbasepoint$, the record at \dbasepoint\ is copied over the region \gbasepoint\ such that $\dsection(\dbasepoint)=\vindex^*\dsection(\gbasepoint)$ where $\vindex^*\dsection(\gbasepoint)$  is \dsection\ pulled back over \gbase.
-When \dbase\ is discrete points and the graphic is a scatter plot, each point $\dbasepoint \in \dbase$ corresponds to a 2D disk $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:graphic:retraction:map}.
+The map \vindex\ is part of the implementation of the artist $\mathcal{\vartist}$ and therefore is not defined in terms of the data; instead it is how we specify the constraint that the type of the graphic \textit{continuity} must be able to map to the type of the data \textit{continuity}.
+When \dbase\ is discrete points and the graphic is a scatter plot, each point $\dbasepoint \in \dbase$ can correspond to a 2D disk $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:graphic:retraction:map}.
 In the case of 1D continuous data and a line plot, the region \gy\ over a point $\gx_i$ specifies the thickness of the line in \gbase\ for the corresponding \dsection\ on \dbasepoint.
@@
 \end{equation}
-The artist preserves \textit{continuity} through the \vindex\ map discussed in \autoref{sec:math:graphic:base} and is an \textit{equivariant} map because it carries a homomorphism of monoid actions \cite{cegarraCohomologyMonoidsOperators2019}
+The artist preserves \textit{continuity} through the \vindex\ map discussed in \autoref{sec:math:graphic:base}.
+We conjecture that it is an \textit{equivariant} map because it carries a homomorphism of monoid actions \cite{cegarraCohomologyMonoidsOperators2019}
 \begin{equation}
@@
     \includegraphics[width=1\textwidth]{figures/math/scatter_with_s.png}
-    \caption{The data is discrete points (temperature, time).
+    \caption{The data is discrete points (precipitation, time).
 Via \vchannel\ these are converted to (xpos, ypos) and pulled over discrete \gbase.
@@
     \includegraphics[width=\textwidth]{figures/math/line_with_s.png}
-    \caption{The line fiber $(time,\, temp)$ is thickened with the derivative $(time^{\prime},\, temperature^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.
+    \caption{The line fiber $(time,\, temp)$ is thickened with the derivative $(time^{\prime},\, precipitation^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.
 This is because the line needs to be pushed perpendicular to the tangent of (xpos, ypos).
@@
 
+\subsubsection{ \vmarkd}
 \label{sec:math:artist:qhat}
-The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
-As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+
 \begin{figure}[htb]
@@
   \end{figure}
+
+The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
+As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+
+\begin{equation}
+    \begin{tikzcd}
+        V \arrow[d, "\pi"']             & \xi^{*}V \arrow[d, "\pi"'] \arrow[l, "\xi^{*}"']         \\
+        K \arrow[u, "\mu"', bend right] & S \arrow[l, "\xi"'] \arrow[u, "\xi^{*}\mu"', bend right]
+    \end{tikzcd}
+    \label{eq:math:artist:qhat}
+\end{equation}
 
@@
 \begin{subfigure}{.5\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/combined_artist.png}
+    \caption{}
+    \label{fig:math:combined}    
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
     \includegraphics[width=1\textwidth]{figures/math/exploding_artist.png}
@@
     \label{fig:math:seperate}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-    \includegraphics[width=1\textwidth]{figures/math/combined_artist.png}
-    \caption{}
-    \label{fig:math:combined}
 \end{subfigure}

### 2021-04-30 draft/sections/scraps.tex
code section
closing out fig redos + labeling + consistently precip

@@
 composotion of functions from a space to itself, is associative (\compo)
+
+S
+This API choice can lead to visualizations that break \textit{continuity} when fed into visualizations with different assumptions about structure.
+The lack of consistent data model can also mean no consistent way of updating the data and therefore no way of guaranteeing that the views are in sync, in visualizations that consistent of multiple views of the same datasource, such as dashboards\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007}.

### 2021-04-30 draft/sections/math.tex
tminipage for Q/S map

@@
 \end{figure}
-The scatter plot in \autoref{fig:math:artist:q:scatter} can be defined as 
-\begin{equation}
-\vmark(xpos, ypos)(\alpha, \beta)
-\label{eq:math:artist:q:scatter}
-\end{equation}
-with a constant \textit{size} and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of \vmark.
-The position of this swatch of color can be computed relative to the location on the disc \((\alpha, \beta) \in \gbase_{\dbasepoint}\) as shown in \autoref{fig:math:artist:q:scatter}
+The scatter plot in \autoref{fig:math:artist:q:scatter} has a constant \textit{size} and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of the point assembly function $\vmark(xpos, ypos)(\alpha, \beta)$.
+
+\begin{figure} [H]
+\begin{minipage}{.5\textwidth}
+    \centering
+    \includegraphics{figures/math/scatter_s.png}
+\end{minipage}
+\begin{minipage}{.5\textwidth}
     \begin{align*}
-x &= size *\alpha \cos(\beta) + xpos \\
+        x &= size *\alpha \cos(\beta) + xpos \\[.5\baselineskip]
         y &= size *\alpha \sin(\beta) + ypos
     \end{align*}
-such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
+\end{minipage}
+\label{fig:math:artist:q:scatter:s}
+\caption{lookup for scattter glyphs}
+\end{figure}
+
+The position of this swatch of color can be computed relative to the location on the disc \((\alpha, \beta) \in \gbase_{\dbasepoint}\) as shown in \autoref{fig:math:artist:q:scatter:s}.
+\note{EXPLAIN WHAT IS HAPPING IN THE FIGURE THAT GETS BACK TO gsection}
+
+$\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
 
@@
 The $\alpha$ coordinates on \gbase\ specifies the position of the line, the $\beta$ coordinate specifies thickness.}
-    \label{fig:artist_line}
+    \label{fig:math:artist:q:line}
 \end{figure}
-In contrast, the line plot 
-\begin{equation}
-\vmark(xpos, \hat{n}_{1}, ypos, \hat{n}_{2})(\alpha, \beta)
-\label{eq:math:artist:q:line}
-\end{equation}
-in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
-As shown in \autoref{fig:math:artist:q:line}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal.
-The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
-such that the normal is  $\hat{n}_{1} = \frac{n_1}{\lvert n \rvert}, \; \hat{n}_{2} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
+In contrast, the line plot $\vmark(xpos, \hat{n}_{1}, ypos, \hat{n}_{2})(\alpha, \beta)$ in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
+
+\begin{figure} [H]
+    \begin{minipage}{.5\textwidth}
+        \centering
+        \includegraphics{figures/math/line_s.png}
+    \end{minipage}
+    \begin{minipage}{.5\textwidth}
         \begin{align*}
- x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n}_1(\vindex(\alpha)) \\
- y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n}_2(\vindex(\alpha)) 
+            \lvert n \rvert &=\sqrt{{n_{1}}^2 + {n_{2}}^2},\,
+            \hat{n}_{1} = \frac{n_1}{\lvert n \rvert}, \,\hat{n}_{2} = \frac{n_2}{\lvert n \rvert} \\[.5\baselineskip]
+            x &= xpos(\vindex(\alpha)) + width*\beta\hat{n}_1(\vindex(\alpha))  \\[.5\baselineskip]
+            y &= ypos(\vindex(\alpha)) + width*\beta\hat{n}_2(\vindex(\alpha)) 
            \end{align*}
-where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n}_1, \hat{n}_2$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
+    \end{minipage}
+    \label{fig:math:artist:q:line:s}
+    \caption{lookup for line glyphs}
+\end{figure}
+
+As shown in \autoref{fig:math:artist:q:line:s}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal.
+The indexing map $\vindex(\alpha)$ finds the point in \dbase\ corresponding to the region in \gbase\ at coordinate $\alpha$.
+The section \dsection\ on the \dbasepoint\ that corresponds to the region in \gbase\ returns  the position \textit{xpos, ypos} and the derivatives $\hat{n}_1, \hat{n}_2$ .
+The derivatives are then multiplied by a \textit{width} parameter to specify the thickness.
+\note{lil bit messy}
 

### 2021-05-03 draft/sections/math.tex
reworking q and q hat

@@
 \end{figure}
-The scatter plot in \autoref{fig:math:artist:q:scatter} has a constant \textit{size} and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of the point assembly function $\vmark(xpos, ypos)(\alpha, \beta)$.
+The scatter plot in \autoref{fig:math:artist:q:scatter} has a constant size and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of the point assembly function.
 
 \begin{figure} [H]
+
+\begin{minipage}{.5\textwidth}
+    \begingroup \leqnomode
+    \begin{equation}
+        \vmark(xpos, ypos)(\alpha, \beta)
+        \label{eq:eq:math:artist:q:scatter}
+    \end{equation}
+    \endgroup
+    \begin{align*}
+        x &= \text{size} *\alpha \cos(\beta) + xpos \\[.5\baselineskip]
+        y &= \text{size} *\alpha \sin(\beta) + ypos
+    \end{align*}
+\end{minipage}
 \begin{minipage}{.5\textwidth}
@@
 \end{minipage}
-\begin{minipage}{.5\textwidth}
-    \begin{align*}
-        x &= size *\alpha \cos(\beta) + xpos \\[.5\baselineskip]
-        y &= size *\alpha \sin(\beta) + ypos
-    \end{align*}
-\end{minipage}
+\caption{The simplest form of the scatter plot takes as input the expected position of the marker in visual space \textit{(xpos, ypos)}.
+The marker shape is determined by the polar coordinates $(\alpha, \beta)$ on the disc; these coordinates dictate whether anything is drawn at that region of \gbase.
+To obtain the color of the pixel at \textit{(x,y)}, the region on \gbase\ is scaled by a constant size and shifted by the \textit{xpos} and \textit{ypos}.}
 \label{fig:math:artist:q:scatter:s}
-\caption{lookup for scattter glyphs}
 \end{figure}
-
-The position of this swatch of color can be computed relative to the location on the disc \((\alpha, \beta) \in \gbase_{\dbasepoint}\) as shown in \autoref{fig:math:artist:q:scatter:s}.
-\note{EXPLAIN WHAT IS HAPPING IN THE FIGURE THAT GETS BACK TO gsection}
-
-$\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black.
+The position of this swatch of color is computed relative to the location on the disc \((\alpha, \beta) \in \gbase_{\dbasepoint}\) as shown in \autoref{fig:math:artist:q:scatter:s}.
+The region $\alpha, \beta)$ is scaled by a constant size and shifted by \textit{xpos} and \textit{ypos}.
+This computation yields the values \textit{(x,y)} that map into \gfiber\ and have a corresponding function $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ which colors the point \textit{(x,y)} black.
 
@@
 \end{figure}
-In contrast, the line plot $\vmark(xpos, \hat{n}_{1}, ypos, \hat{n}_{2})(\alpha, \beta)$ in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
+In contrast, the line plot in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
 
 \begin{figure} [H]
+    \begin{minipage}{.5\textwidth}
+        \begingroup \leqnomode
+        \begin{equation}
+            \vmark(xpos, n_1, ypos, n_2 )(\alpha, \beta) 
+            \label{eq:eq:math:artist:q:line}
+        \end{equation} \endgroup
+        \begin{align*}
+            \lvert n \rvert &=\sqrt{{n_{1}}^2(\vindex(\alpha)) + {n_{2}}^2(\vindex(\alpha))} \\
+            \hat{n}_{1} &= \frac{n_1(\vindex(\alpha))}{\lvert n \rvert}, \,\hat{n}_{2} = \frac{n_2(\vindex(\alpha))}{\lvert n \rvert} \\[.5\baselineskip]
+            x &= xpos(\vindex(\alpha)) + \text{width}*\beta\hat{n}_1(\vindex(\alpha))  \\
+            y &= ypos(\vindex(\alpha)) + \text{width}*\beta\hat{n}_2(\vindex(\alpha)) 
+           \end{align*}
+    \end{minipage}
     \begin{minipage}{.5\textwidth}
@@
     \end{minipage}
-    \begin{minipage}{.5\textwidth}
-        \begin{align*}
-            \lvert n \rvert &=\sqrt{{n_{1}}^2 + {n_{2}}^2},\,
-            \hat{n}_{1} = \frac{n_1}{\lvert n \rvert}, \,\hat{n}_{2} = \frac{n_2}{\lvert n \rvert} \\[.5\baselineskip]
-            x &= xpos(\vindex(\alpha)) + width*\beta\hat{n}_1(\vindex(\alpha))  \\[.5\baselineskip]
-            y &= ypos(\vindex(\alpha)) + width*\beta\hat{n}_2(\vindex(\alpha)) 
-           \end{align*}
-    \end{minipage}
+    \caption{The \textit{xpos} and \textit{ypos} variables give the position of the line in screen space, but render an infinitely thin line.
+To draw equidistant lines parallel to(\textit{xpos}, \textit{ypos}), defined by the distance  $(n_1, n_2)$, requires the derivatives $(n_{1} = xpos^{\prime}, n_{2} = ypos^{\prime})$.
+The position (\textit{xpos}, \textit{ypos}) and width of the line is then used to determine whether a pixel is colored at the position (\textit{x}, \textit{y}).
+The values in data space are only looked up via the $\alpha$ coordinate of \gbase\ because it maps to a location on \dbase.
+The $\beta$ parameter is used to specify how thick the line is in conjunction with the constant width.
+}
     \label{fig:math:artist:q:line:s}
-    \caption{lookup for line glyphs}
 \end{figure}
-
 As shown in \autoref{fig:math:artist:q:line:s}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal.
@@
 The section \dsection\ on the \dbasepoint\ that corresponds to the region in \gbase\ returns the position \textit{xpos, ypos} and the derivatives $\hat{n}_1, \hat{n}_2$ .
-The derivatives are then multiplied by a \textit{width} parameter to specify the thickness.
-\note{lil bit messy}
+The derivatives are then multiplied by a width parameter to specify the thickness of the line.
+This is then used to determine the color of the pixel at \textit{(x,y)}.
 
@@
     \includegraphics[width=\textwidth]{figures/math/image.png}
-    \caption{The only visual parameter an image requires is color since \vindex\ encodes the mapping between position in data and position in graphic. }
+    \caption{Via \vindex\, the artist maps from a point \textit{(x,y)} on the screen to a corresponding point on \dbase.
+This maps into \dfiber\ via \dsection.
+These data points are converted to visual points via \vchannel\, and then \vmark\ assembles the \textit{(xpos, ypos, color)} parameters into attributes of each pixel.
+}
     \label{fig:math:artist:q:heatmap}
 \end{figure}
-In \autoref{fig:math:artist:q:heatmap}, the image 
+In \autoref{fig:math:artist:q:heatmap}, the image is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
+The indexing variables $(\alpha, \beta)$ define the distance along the space, which is then used by \vindex\ to map into \dbase\ to lookup the color values.
 \begin{equation}
-\vmark(xpos, ypos, color)
+    \vmark(xpos, ypos, color)(\alpha, \beta)
     \label{eq:math:artist:q:image}
     \end{equation}
- is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$.
-The indexing variables $(\alpha, \beta)$ define the distance along the space, which is then used by \vindex\ to map into \dbase\ to lookup the color values
-\begin{equation*}
-R = R(\vindex(\alpha, \beta)), \; G = G(\vindex(\alpha, \beta)), \; B = B(\vindex(\alpha, \beta))
-\end{equation*}
+\begin{align*}
+x &= xpos(\vindex(\alpha))\\
+y &= ypos(\vindex(\beta))\\
+R, G, B &= color(\vindex(\alpha, \beta))
+\end{align*}
 In the case of an image, the indexing mapper \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.
@@
 As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
-
+%[column sep=huge]
 \begin{equation}
     \begin{tikzcd}
-        V \arrow[d, "\pi"']             & \xi^{*}V \arrow[d, "\pi"'] \arrow[l, "\xi^{*}"']         \\
-        K \arrow[u, "\mu"', bend right] & S \arrow[l, "\xi"'] \arrow[u, "\xi^{*}\mu"', bend right]
+        \textcolor{gray!50}{\dtotal} \arrow[r, "\vchannel", color=gray!50] \arrow[rd, "\pi"', color=gray!50] & \vtotal \arrow[d, "\pi"']                  & \vindex^*\vtotal \arrow[r, "\vmark", color=gray!50] \arrow[d, "\vindex^*\pi"'] \arrow[l,  "\vindex^*"'] & \textcolor{gray!50}{\gtotal} \arrow[ld, "\pi", color=gray!50] \\
+                                                          & \dbase \arrow[u, "\vsection"', bend right] & \gbase \arrow[l, "\vindex"'] \arrow[u, "\vindex^*\vsection"', bend right]               &
         \end{tikzcd}
@@
 \end{equation}
-
 The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back of \vsection\

### 2021-05-03 draft/sections/next.tex
discussion section

@@
 \section{Discussion}
-This work contributes a functional model of the structure-preserving maps from data to visual representation to guide the development of visualization libraries, thereby providing a means to express the constraints that visualization must preserve continuity and faithfully translate the properties of the data variables into visual variables.
-Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks.
-By decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ maps, the model expresses the specifications that graphic and data must have equivalent \textit{continuity} equivalent to the data, and that the visual characteristics of the graphics are \textit{equivariant} to their corresponding components under monoid actions.
-TEAM defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic.
-The toy prototype built using this model validates that it is usable for a general purpose visualization tool since it can be iteratively integrated into existing architecture rather than starting from scratch.
-Factoring out graphic formation into assembly functions allows for much more clarity in how they differ.
+The Topological Equivariant Artist Model (TEAM) is a functional model of the structure preserving maps from data to visual representation.
+TEAM expresses the specifications that graphic and data must have equivalent \textit{continuity} equivalent to the data, and that the visual characteristics of the graphics are \textit{equivariant} to their corresponding components.
+TEAM expresses these constraints in the encoding \vchannel, assembly \vmark, and reindexing \vindex\ functions that make up the artist \vartist.
+This decomposition of the artist into smaller components functional pieces allows TEAM to provide well specified guidance on implementing visualization components based on this architecture.
+The proof of concept prototype built using this model validates that it is usable for a general purpose visualization tool.
+Additionally, the decomposition facilitates iteratively integrating TEAM into existing architecture rather than starting from scratch, since \vchannel, \vmark\ and \vindex\ functions can be implemented independently.
 This prototype demonstrates that this framework can generate the fundamental point (scatter plot) and line (line chart) marks.
+Furthermore, combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets TEAM support a variety of data continuities, including  discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks.
+Although the prototype currently only implements 0D and 1D continuity, we expected it to generalize to the other continuities.
 

### 2021-05-03 draft/sections/next.tex
discussion/conclusion sorted out

@@
 \subsection{Limitations}
-Our model and prototype are deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as VTK or D3.
-Even though the model is designed to be backend and format independent, it has only been tested using PNGs rendered with AGG\cite{shemanarevAntiGrainGeometry}.
-It is unknown how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997} that implement different models of \gsection.
-While our model supports equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from data components\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}, it cannot evaluate the semantic accuracy of the figurative representation.
-Effectiveness criteria\cite{mackinlayAutomaticDesignGraphical1987,chambersGraphicalMethodsData1983a} are out of scope.
+The TEAM model is a specification visualization library developers can use to implement structure preserving library components.
+Implementing a TEAM based architecture involves developers explicitly describing the structure and continuity of the data and the structure and continuity the artist expect.
+TEAM does not provide a framework for recommending visualizations to the user,  therefore effectiveness \cite{mackinlayAutomaticDesignGraphical1987,chambersGraphicalMethodsData1983a} is out of scope.
+But, automatic recommendation tools could be built using TEAM components.
+
+While TEAM specifies the components, the developers building libraries using TEAM components decide which compositions of components are semantically correct for the domain.
+For this reason, TEAM does not include data space transforms, as are incorporated into libraries like Tableau or ggplot, instead leaving choice of computations to implementors of the data object.
+TEAM's intentional ignorance of semantics also means it cannot evaluate whether a figurative glyph \cite{byrneAcquiredCodesMeaning2016} is a semantically correct choice, but it can enforce equivariance constraints of glyphs generated from data components enforcing equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from data components\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}.
+TEAM also allows to graphics to have a lower dimensional continuity than the source data when a retraction map from one continuity to the other exists.
+For example, TEAM components could transform 1D continuous data into 0D discrete graphics, e.g.
+bar charts or scatter plots.
+As with computations, it is the role of the domain specific library to determine which figurative glyphs and continuity downgrades are appropriate.
+
+The prototype is deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as R graphics\cite{murrellGraphicsThirdEdition2018}, VTK, and D3.
+TEAM has only been tested using PNGs rendered with AGG\cite{shemanarevAntiGrainGeometry}, but is expected to work with all the file types Matplotlib currently supports, including svg, pdf, and eps.
+We have not yet addressed how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997} that implement different models of \gsection.
 
@@
 \end{table}
-In acknowledgement that the schedule is optimistic, the data applications of this work could instead be revisited after the defense as that work is not integral to visualization library architecture.
+In acknowledgement that the schedule is optimistic, this work has various scales of data applications.
+We plan to apply this model to datasets with complex continuities, such as the trajectories of rats running around a maze and the positions of their limbs.
+We also potentially can look at large scale biology or climate datasets.
 The data applications could be further integrated with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
 Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
+This specification of structure could also be used to develop a serialization structure tht could then be use to allow Matplotlib to interface with other visualization libraries such as open GL via shared serialization protocol.
 While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.

### 2021-05-03 draft/sections/implementation.tex
implementation section

@@
 \label{sec:implementation}
-To evaluate the feasibility of the model described in ~\autoref{sec:math:}, we built prototypes of a \texttt{point}, \texttt{line}, and \texttt{bar} artist.
-We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to convert data coordinates into screen coordinates.
+To evaluate the feasibility of the model described in ~\autoref{sec:math}, we built prototypes of a \texttt{point}, \texttt{line}, and \texttt{bar} artist.
+We make use of the Matplotlib Figure and Axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations and exploit the Matplotlib transform stack to convert data coordinates into screen coordinates.
 While the artist is specified in a fully functional manner in \autoref{eq:math:artist:diagram}, we implement the  prototype in a heavily object oriented manner.
-We do so mostly to more easily manage function inputs, especially parameters that are passed through to methods that are structurally functional.
+This is done to manage function inputs, especially parameters that are passed through to methods that are structurally functional.
 
@@
   fig, ax = plt.subplots()
-  artist = Artist(E, V)
+  artist = ArtistClass(E, V)
   ax.add_artist(artist)
 \end{minted}
-Building on the curr
-ent Matplotlib artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalence class artist \vartisteq\ as described in \autoref{eq:math:artist:aprime}.
+Building on the current Matplotlib Artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalence class artist \vartisteq\ as described in \autoref{eq:math:artist:aprime}.
 The visual bundle \vtotal\ is specified as the \mintinline{python}{V} dictionary of the form \mintinline{python}|{parameter:(variable name, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber, and the \vchannel\ encoders are passed in as functions or callable objects.
@@
 
-      def hat_Q(self, **args): 
+      def q_hat(self, **args): 
           # set the properties of the graphic
@@
                    for p, (c, nu) in self.V.items()} 
-          self.hat_q(**mu)
+          self.q_hat(**mu)
           # pass configurations off to the renderer
@@
 The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}.
-To ensure the integrity of the section, \mintinline{python}{view} must be atomic, which means that the values cannot change after the method is called in draw until a new call to draw\cite{ullmanFirstCourseDatabase2008}.
-We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.
+We require that the \mintinline{python}{view} method be atomic so that we do not risk race conditions.
+Atomicity means that the values cannot change after the method is called in draw until a new call to draw\cite{ullmanFirstCourseDatabase2008}, which ensures the integrity of the section.
 
@@
 The \mintinline{python}{q_hat} function that is \vmarkd,  as defined in \autoref{eq:math:artist:qhat_k}, is responsible for generating a representation such that it could be serialized to recreate a static version of the graphic.
-This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd.
 The last step in the artist function is handing itself off to the renderer.
@@
     \end{subfigure}
-  \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
+  \caption{Scatter plot and line plot implemented using \mintinline{python}{Point} and \mintinline{python}{Line} artists and fiber bundle inspired data models. Matplotlib is used for the rendering.}
   \label{fig:code:scatter_line}
@@
 This is implemented via a \mintinline{python}{Line} object where the scatter marker shape is fixed as a circle, and the visual fiber components are x and y position and the facecolor and size of the marker.
-We only show the \mintinline{python}{q_hat} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
+We only show the \mintinline{python}{q_hat} function here because the \mintinline{python}{__init__, draw} are inherited from the prototype artist \mintinline{Python}{ArtistClass}.
 
 The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices.
-Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$.
+Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint\ has corresponding values in section $\dsection(\dbasepoint_{i})$.
 This means that all the data on one vertex maps to one glyph.
@@
 The colors are guaranteed to correspond to the correct segment because of the atomicity constraint on the view.
-The current implementation of line in Matplotlib does not have this functionality because it has no notion of an atomic view \dsection, instead the user keeps track of which values are input into a single call of \vmarkd\ by creating a loop where every color is a new call to the line plotting function.
+The implementation of line in Matplotlib does not have this functionality because it has no notion of rows and columns of a table, and therefore no notion of a \dsection.
+Instead, line is assumed to be points along one edge and therefore has only one color, and the user is responsible for aligning the x and y components and colors along the implicit \dbase\ over which they are plotted.
 
@@
 class Bar(ArtistClass, mcollections.Collection):
-    def __init__(self, data, transforms, orientation='v', *args, **kwargs):
+    def __init__(self, E, V, orientation, *args, **kwargs):
         """
-        orientation: str, optional
+        orientation: str
             v: bars aligned along x axis, heights on y
@@
         self.orientation = orientation
-        # set E and V
+        super().__init__(*args, **kwargs) # set E & V
 
-    def q_hat(self, position, length, floor=0, width=0.8, 
-                    facecolors='C0', edgecolors='k', offset=0):
-        #set some defaults
-        width = itertools.repeat(width) if np.isscalar(width) else width
-        floor = itertools.repeat(floor) if np.isscalar(floor) else (floor)
-
+    def q_hat(self, position, length, floor, width, 
+                    facecolors, edgecolors, offset):
         # offset is passed through via assemblers such as multigroup, 
@@
 The \mintinline{python}{make_bars} function converts the input position and length to the coordinates of a rectangle of the given width.
-Defaults are provided for 'width' and 'floor' to make this function more reusable.
-Typically the defaults are used for the type of chart shown in figure~\ref{fig:code_bar_simple}, but these visual variables are often set when building composite versions of this chart type as discussed in section~\ref{sec:code_case_study}.
+Typically defaults are used for the type of chart shown in figure~\ref{fig:code_bar_simple}, but these visual variables are often set when building composite versions of this chart type as discussed in section~\ref{sec:code_case_study}.
 
@@
 \end{minted}
-where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.
+where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class.
+
 \begin{minted}{python}
@@
 The discrete \mintinline{python}{tau} method returns a record of discrete points, akin to a row in a table, while the line\mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint.
+These continuities are also specified in the \mintinline{python}|k={'tables':[]}| dictionary.
 \begin{minted}{python}
@@
     \end{subfigure}
-\caption{Continuous and discontinuous lines as defined via the same data model, and generated with the same \vartisteq \mintinline{python}{Line}}
+\caption{Continuous and discontinuous lines as defined via the same data model, and generated with the same \vartisteq\ \mintinline{python}{Line}}
 \label{fig:code:multilines}
@@
 The axes can also be used to choose an appropriate number of samples.
-In contrast, the stair chart is discontinuous and only needs to be evaluated at the edges of the interval
+In contrast, the stair chart only needs to be evaluated at the edges of the interval
 \begin{minted}{python}
@@
 \label{sec:code_case_study}
-Building on the \mintinline{python}{Bar} artist in \autoref{sec:code:artists}, we implement grouped bar charts as these do not exist out of the box in the current version of Matplotlib.Instead, grouped bar charts are often achived via looping over an implementation of bar, which forces the user to keep track which values are mapped to a single visual element and how that is achieved.
+Building on the \mintinline{python}{Bar} artist in \autoref{sec:code:artists}, we implement grouped bar charts as these do not exist out of the box in the current version of Matplotlib.
+Instead, grouped bar charts are often achieved via looping over an implementation of bar, which forces the user to keep track which values are mapped to a single visual element and how that is achieved.
 For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020}, packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.
+
+\begin{table}[H]
+    \centering
+\begin{tabular}{|l|r|r|r|l|l|l|}
+    \toprule
+       sex &  Adelie &  Chinstrap &  Gentoo & Adelie\_c & Chinstrap\_c & Gentoo\_c \\
+    \midrule
+     female &      73 &         34 &      58 &   Adelie &   Chinstrap &   Gentoo \\
+     male &      73 &         34 &      61 &   Adelie &   Chinstrap &   Gentoo \\
+    \bottomrule
+\end{tabular}
+\caption{Palmer Penguins dataset that is processed to become input into the grouped bar chart.
+This data is a count of penguin species.
+The columns with suffix c are used to specify the color of the corresponding visual element.}
+\label{tab:code:penguins}
+\end{table}
+
 The wrapper is very thin because there is explicitly only one section.
@@
 As with \mintinline{python}{Bar}, the orientation is set in the constructor.
-In both these artists, we separate the transforms that are applied to only one component (column) from transforms applied to multiple components (columns).
-We encode the one component case as \mintinline{python}{V} dictionary, and the multiple components case as \mintinline{python}{MV}.
+In both these artists, we separate the transforms \mintinline{python}{V}  that are applied to only one component (column) from transforms  \mintinline{python}{MV} applied to multiple components (columns).
 This convention allows us to, for example, map one column to position, but multiple to length.
+In effect, we are decomposing \dtotal\ into $\dtotal_1, \cdots, \dtotal_i, \cdots, \dtotal_n$ via specifications in \mintinline{python}{V} rather than by directly taking subsections of the table.
+This allows us to ensure shared \dbase\ and coherant \dsection.
 
@@
 
-    def assemble(self):
+    def q_hat(self):
         tau = self.data.view(self.axes)
@@
 class GroupedBar(martist.Artist):
-    def assemble(self):
+    def q_hat(self):
         self.children = [] # list of bars to be rendered
@@
 \begin{minted}{python}
-    transforms = {'position': {'name':'island',
-                  'encoder': position.Nominal({'Biscoe':0.1, 'Dream':1.1, 'Torgersen':2.1})}} 
+    transforms = {'position': {'name':'sex',
+                  'encoder': position.Nominal({'female':0, 'male':1})}} 
     group_transforms =  [{'length': {'name':'Adelie'},
@@
 For example, the \mintinline{python}{draw} function that maps most closely to \vartist\ is functional, with state only being necessary for bookkeeping the many inputs that the function requires.
-In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\, assembly \vmarkd and artists \vartist.
+In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\, assembly \vmarkd\ and artists \vartist.
 We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic.

### 2021-05-05 draft/sections/intro.tex
closing out fig redos + labeling + consistently precip
cleaned up figures/new figures

@@
 
-\subsection{Structure:}
+\subsection{Structure Visualizations Preserve}
 \label{sec:intro:structure}
@@
 \label{sec:intro:data:tools}
-
 \begin{figure}[H]
@@
 \subsection{Data}
-\label{sec:intro_data}
+\label{sec:intro:data:continuity}
 Fiber bundles were proposed by Butler as a core data structure for visualization because they encode data \textit{continuity} separately from the components of the dataset\cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.

### 2021-05-05 draft/sections/math.tex
implementation section
cleaned up figures/new figures

@@
 \label{sec:math}
-To guide the implementation of structure preserving building block components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}.
-Inspired by the somewhat analogous component in Matplotlib\cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic that these building block components implement the \textit{artist}.
+To guide the implementation of structure preserving visualization components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}.
+Inspired by the analogous classes in Matplotlib\cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic space that these building block components implement the \textit{artist}.
 \begin{equation}
@@
 \end{equation}
-The \textit{artist} $\mathscr{\vartist}$ is a map from the data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$ fiber bundles.
-To explain how the \textit{artist} is a structure preserving map from data to graphic, we first describe how we model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as topological structures that encapsulate component types and continuity.
-We then discuss the maps from graphic to data (\autoref{sec:math:graphic:base}, data components to visual components (\autoref{sec:math:artist:nu}), and visual components into graphic (\autoref{sec:math:artist:q}) that make up the artist.
+The \textit{artist} $\mathscr{\vartist}$ is a map from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$ fiber bundles.
+To explain how the \textit{artist} is a structure preserving map from data to graphic, we first model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as topological structures that encapsulate component types and continuity.
+We then discuss the funtional maps from graphic to data (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}), and visual components into graphic (\autoref{sec:math:artist:q}) that make up the artist.
 
@@
 \label{sec:math:data}
-Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
+We use fiber bundles as the data model because they are inclusive enough to express all the types of structures of data described in \autoref{sec:intro:data:tools}.
+A fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
 \begin{equation}
@@
 that binds the components of the data in \dfiber\ to the continuity of the data encoded in \dbase.
-The fiber bundle models the properties of data component types \dfiber\ (\autoref{sec:math:data:fiber}), the continuity of records \dbase\ (\autoref{sec:math:data:base}), the collections of records \dsection\ (\autoref{sec:math:data:section}), and the space \dtotal\ of all possible datasets with these components and continuity.
+Our use of fiber bundles builds on Butler's work proposing that fiber bundles should be the common data abstraction for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}.
+The fiber bundle models the properties of data component types \dfiber\ (\autoref{sec:math:data:fiber}), the continuity of records \dbase\ (\autoref{sec:math:data:base}), the collections of records (\autoref{sec:math:data:section}), and the space \dtotal\ of all possible datasets with these components and continuity.
 By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartesian product $\dbase\times\dfiber$.
-We use fiber bundles as the data model because they are inclusive enough to express all the types of structures of data described in \autoref{sec:intro:data:tools}
 
@@
 \label{sec:math:data:fiber}
-To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names.
-This allows us to describe the components in a schema like way.
+To formalize the structure of the data components, we use Spivak's description of the schema \cite{spivakSIMPLICIALDATABASES} as a fiber bundle to bind the components of the fiber to variable names and data types.
 Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
@@
     \includegraphics[width=1\textwidth]{figures/math/fiber.png}
-    \caption{The fiber space is the set of all combinations of all theoretically possible values of the components.
+    \caption{The fiber space is the cartesian product of the components.
 The 2D fiber \(\dfiber=\reals\times\realsp\) (a) encodes the properties of \textit{time} and \textit{precipitation} components.
@@
 
-For example, the records in the 2D fiber (a) in \autoref{fig:math:data:fiber} are a pair of \textit{times} and \textdegree K \textit{precipitation} measurements taken at those times.
-Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$.
-Temperature values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$.
+For example, the records in the 2D fiber (a) in \autoref{fig:math:data:fiber} are a pair of \textit{times} and \textit{precipitation} measurements taken at those times.
+Time is a positive number of type \texttt{datetime} which can be resolved to $\ftotal_{\texttt{datetime}}= \reals$.
+Precipitation values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$.
 The fiber is
@@
 \end{equation*} 
-where the dimensionality of the fiber does not change, but the components of the fiber can be coupled.
-For example, \textit{location} can can either be specified as \((\fname = location,\, \ftype = \texttt{point},\, \fttype=\reals^{2})\)or \((\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\reals)\) and \((\fname=longitude,\, \ftype=\texttt{float},\,\fttype=\reals)\).
+and can either be thought of as having two components  
+\begin{equation}
+    \{(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals), ((\fname = location,\, \ftype = \texttt{point},\, \fttype=\reals^{2})\}
+\end{equation}
+ or three
+ \begin{equation}
+    \{(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals), (\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\reals), (\fname=longitude,\, \ftype=\texttt{float},\,\fttype=\reals)\}
+ \end{equation}
 
@@
 \subsubsection{Data \dsection}
-\label{sec:data_section}
+\label{sec:math:data:section}
 While the projection function $\pi:\dtotal \rightarrow\dbase$ ties together the base space \dbase\ with the fiber \dfiber, a section $\dsection: \dbase\rightarrow \dtotal$ encodes a dataset.
@@
 \{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}
+\label{eq:math:artist:aprime}
 \end{equation*}

### 2021-05-05 draft/sections/next.tex
cleaned up figures/new figures

@@
 TEAM's intentional ignorance of semantics also means it cannot evaluate whether a figurative glyph \cite{byrneAcquiredCodesMeaning2016} is a semantically correct choice, but it can enforce equivariance constraints of glyphs generated from data components enforcing equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from data components\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}.
-TEAM also allows to graphics to have a lower dimensional continuity than the source data when a retraction map from one continuity to the other exists.
-For example, TEAM components could transform 1D continuous data into 0D discrete graphics, e.g.
+TEAM also allows graphics to have a lower dimensional continuity than the source data when a retraction map from one continuity to the other exists.
+For example, TEAM components could transform 1D continuous segments into 0D discrete elements, e.g.
 bar charts or scatter plots.

### 2021-05-05 draft/sections/intro.tex
conjectures, no homomorphisms

@@
 
-\subsection{Structure Visualizations Preserve}
+\subsection{Structure}
 \label{sec:intro:structure}
@@
         \includegraphics[width=1\textwidth]{figures/intro/equivariant.png}
-        \caption{The precipitation in blue is scaled by a factor of 10, yielding the precipitation in orange in the table (a).
-This scaling factor is represented by the green arrows in (a).
-The \textit{equivariant} action is the scaling in the timeseries (b), which is represented by the green fill.
-Because the scalings are equivariant, the orange precipitation timeseries is 10x the blue precipitation timeseries.
-The red line is an example of \textit{equivariance} not being preserved, as it is not 10x the the y value of the blue precipitation timeseries.}
+        \caption{The table (a) lists precipitation in millimeters in blue.
+These values are converted to  inches, in green, by dividing the millimeter values by 25.4.
+This conversion is a scaling action, represented by the orange arrows in (a).
+The \textit{equivariant} action is the scaling in the timeseries (b), which is represented by the orange fill between the green and blue lines.
+In (c), the scaling is illustrated in the different millimeter and inches y axis labeling of data that appears identical since the distances were scaled by a constant factor that is now incorporated in the labeling.}
       \label{fig:intro:equivariance}
@@
 Bertin proposes that there are categories of visual encodings-such as position, shape, color, and texture-that preserve the properties of the measurement type, quantitative or qualitative, of the encoded data.
-For example, in \autoref{fig:intro:equivariance}, the blue precipitation data in the table (a) is scaled by a factor of 10.
-This scaling action is represented by the green arrows that translate the precipitation into the scaled values in orange.
+For example, in \autoref{fig:intro:equivariance}, the blue precipitation data in millimeters in the table (a) is converted to inches.
+This scaling action is represented by the green arrows that translate the precipitation into the scaled values in green.
 For this visualization to be equivariant, this same scaling factor must be present in the timeseries representation of the data (b).
-The precipitation in orange is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the orange line is 10x the y-values of the blue precipitation timeseries.
-The green fill in the timeseries (b) is the scaling equivalent to the arrows in the table (a).
-In contrast, the red line is not an \textit{equivariant} representation to the 10x scaling because the red y-values are not 10x the blue y values.
+The precipitation in green is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the green line is $\frac{1}{25.4}$ that of the y-values of the blue precipitation timeseries.
+The orange fill in the timeseries (b) is the scaling equivalent to the arrows in the table (a).
+By definition, the distances in the millimeter data were scaled equally \cite{stothersSimilarityGroup} when converted to inches, which means the shapes of the graphs are equivalent when plotted against y-axes that preserve relative distance (c).
+This visualization is also equivariant to the table (a), but this is indicated through labeling rather than the line plots.
 The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}.
@@
 Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups.
-As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitely include continuity.
+As with the algebraic model, our model also defines structure preservation as commutative mappings from data space to representation space to graphic space, but our model uses topology to explicitly include continuity.
 

### 2021-05-05 draft/sections/math.tex
conjectures, no homomorphisms

@@
 \begin{equation*}
-    \dfiber = \reals \times [0, 360] \times [-90, 90]
+    \dfiber = \reals \times \left[0, 360\right] \times \left[-90, 90\right]
 \end{equation*} 
-and can either be thought of as having two components  
-\begin{equation}
-    \{(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals), ((\fname = location,\, \ftype = \texttt{point},\, \fttype=\reals^{2})\}
-\end{equation}
- or three
- \begin{equation}
-    \{(\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals), (\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\reals), (\fname=longitude,\, \ftype=\texttt{float},\,\fttype=\reals)\}
- \end{equation}
+with components \((\fname=precipitation,\, \ftype=\texttt{float},\, \fttype=\reals)\), \((\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\left[0, 360\right])\), and \((\fname=longitude,\, \ftype=\texttt{float},\,\fttype= \left[-90, 90\right]\).
+By adapting Spivak's framework, our model has a consistent way to describe the components of the data, no matter their complexity.
 
@@
 \label{sec:math:data:monoid}
-Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define by the actions of a monoid on the component.
-In doing so, we specify the properties of the component that must be preserved in a graphic representation.
+Implementing expressive visual encodings requires formally describing the structure on the components of the fiber.
+We conjecture that the structure can be formally defined as the actions of a monoid on the component.
+We can also discuss complex structures since a core property of monoids is composability \cite{yorgeyMonoidsThemeVariations}.
+In doing so, we can specify the properties of the component that must be preserved in a graphic representation.
+
 A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
@@
 \end{align*}
-The identity and associativity properties of the action denote that the action is a monoid homomorphism, which means that the group operation is preserved on both sides of the action\cite{weissteinGroupHomomorphism}.
+We conjecture that given these properties, if there is a partial ordering on both sides of the action then the action is equivariant and monotonic.
 
@@
     \end{subfigure}
-    \caption{The action $\bullet$ in \autoref{fig:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right.
-Since the action maps the weather states to themselves, the ordering defined by the monoid $\ast$ is preserved on both sides of the action.
-The action in \autoref{fig:monoid:partial} is monoid homomorphism because the ordering of the weather states is the same as the ordering of the elements they are mapped to.
-Given $sun \geq clouds\geq rain\, lightining$ on the right, the action  $sun\, clouds \rightarrow clouds$, and $rain\, lightining \rightarrow rain$ is structure preserving because on the left $cloud \geq rain$ so the relative ordering of elements is the same as the elements they are mapped to.}
+    \caption{The action $\bullet$ in \autoref{fig:math:data:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right.
+The action maps the weather states to themselves, thereby preserving the ordering defined by the monoid $\ast$ on both sides of the action.
+We conjecture that the action in \autoref{fig:math:data:monoid:partial} is monotonic when the partial ordering of the weather states is the same as the partial order of the elements they are mapped to.
+Given $sun \geq clouds\geq rain,\, lightining$ on the right, we conjecture that the action  $sun,\, clouds \rightarrow clouds$, and $rain\, lightining \rightarrow rain$ is structure preserving such that the relative ordering of elements is the same as the elements they are mapped to.}
     \label{fig:math:data:monoid}
@@
 For example, $storm \leq rain$.
-In \autoref{fig:monoid:total}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states.
-The action in  \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action and it is therefore homomorphic.
+In \autoref{fig:math:data:monoid}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states.
+The action in  \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action.
 The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map\cite{fongInvitationAppliedCategory2019}
@@
 where the structure the action preserves is the relative, rather than exact, ordering.
-Since groups are monoids with invertible operations, this definition of structure is also broad enough to include the Steven's measurment scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
-Monoids are also commonly found in functional programming since the core property of monoids is composability \cite{yorgeyMonoidsThemeVariations}.
+Since groups are monoids with invertible operations, this definition of structure is broad enough to include the Steven's measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
 
@@
 \begin{figure}[H]
-    \includegraphics[width=1\textwidth]{figures/math/scatter_with_s.png}
+    \includegraphics[width=1\textwidth]{figures/math/scatter.png}
     \caption{The data is discrete points (precipitation, time).
@@
 \begin{figure}[H]
-    \includegraphics[width=\textwidth]{figures/math/line_with_s.png}
+    \includegraphics[width=\textwidth]{figures/math/line.png}
     \caption{The line fiber $(time,\, temp)$ is thickened with the derivative $(time^{\prime},\, precipitation^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.

### 2021-05-05 draft/sections/intro.tex
data section revise

@@
 \label{sec:intro:structure}
-
 \begin{figure}[H]
@@
 As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model.
-Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distrobution statistics\cite{wickham40YearsBoxplots2011}.
+Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distribution statistics\cite{wickham40YearsBoxplots2011}.
 Disentangling the computation from the visual transforms allows us to determine whether the visualization library should implement computations or if they should be implemented in data space.

### 2021-05-05 draft/sections/math.tex
data section revise

@@
 We conjecture that the action in \autoref{fig:math:data:monoid:partial} is monotonic when the partial ordering of the weather states is the same as the partial order of the elements they are mapped to.
-Given $sun \geq clouds\geq rain,\, lightining$ on the right, we conjecture that the action  $sun,\, clouds \rightarrow clouds$, and $rain\, lightining \rightarrow rain$ is structure preserving such that the relative ordering of elements is the same as the elements they are mapped to.}
+Given $sun \geq clouds\geq rain,\, lightining$ on the right, we conjecture that the action  $sun,\, clouds \rightarrow clouds$, and $rain,\, lightining \rightarrow rain$ is structure preserving such that the relative ordering of elements is the same as the elements they are mapped to.}
     \label{fig:math:data:monoid}
@@
 \end{equation}
-of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$
+of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$
 
@@
     \includegraphics[width=1\textwidth]{figures/math/k_different_types.png}
-    \caption{The topological base space \dbase\ encodes the continuity of the data space, for example if the data is discrete points or lies on a plane or a sphere}
+    \caption{The topological base space \dbase\ encodes the continuity of the data space.
+The table of discrete weather station records has discrete continuity such that each record maps to a single point (a).
+A gaussian has a value at all points along the interval x is sampled from and therefore has a 1D continuity (b).
+The globe has a value at all points (latitude, longitude) on the globe and therefore has 2D continuity (c).
+}
     \label{fig:math:data:base:types}
 \end{figure}
+
+The base space \dbase\ provides a way to explicitly encode the continuity of the data, as described in \autoref{sec:intro:data:continuity}.
+This explicit topology is a concise way of distinguishing between visualizations that appear identical but assume different continuity, for example heatmaps and images.
 The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other.
-As shown in \autoref{fig:math:data:base:types},
-\dbase\ can have any number of dimensions and can be continuous or discrete.
-The base space also does not describe anything about the dataset besides the continuity.
-While the base space may have components to identify the continuity, such as \textit{time, latitutde, longitude}, these labels are indexed into from \dbase\ the same as any other component.
-This is similar to the notion of structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, but our model treats keys as a pure reference to topology.
-Decoupling the keys from their semantics allows the metadata to be altered; this provides for coordinate agnostic representation of the continuity and facilitates encoding of data where the independent variable may not be clear.
-For example the amount of snow on the ground is dependent on time of day and how much snow has fallen, and changing the coordinate system or time resolution should have no effect on how the records are connected to each other.
+As shown in \autoref{fig:math:data:base:types}, \dbase\ can have any number of dimensions and can be continuous or discrete.
+Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
 
@@
     \caption{The total space \dtotal\ is divided into fiber segments \dfiber.
-The base space \dbase\ acts as an index into the records in the fibers, such that every point \dbasepoint\ has a corresponding fiber $\dfiber_{\dbasepoint}$.}
+The base space \dbase\ acts as an index into the records in the fibers, such that every point \dbasepoint\ has a corresponding fiber $\dfiber_{\dbasepoint}$.
+The projection map $\pi$ maps every fiber $\dfiber_{\dbasepoint}$ to a point $\dbasepoint \in \dbase$ in the base space.}
     \label{fig:math:data:base:div}
@@
 
-Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
 In \autoref{fig:math:data:base:div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the closed interval $\left[0,1\right]$.
+While the total space \dtotal\ may have components in \dfiber\ that describe any given point $\dbasepoint \in \dbase$, such as \textit{time, latitutde, longitude}, these labels are indexed into from \dbase\ the same as any other components.
+In contrast to the structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, our model treats keys \dbasepoint\ as a pure reference to topology.
+Decoupling the keys from their semantics allows the components identifying the keys to be altered, which provides for coordinate agnostic representation of the continuity and facilitates encoding of data where the independent variable may not be clear.
+For example total rainfall is dependent on time of day and how much rain has already fallen, and changing the coordinate system or time resolution should have no effect on how the records are connected to each other.
 
@@
 \end{equation}
-such that \(\monoid_{i}\) acts on component bundle \(\dtotal_i\).
-The \dbase\ remains the same because the connectivity of records does not change just because there are fewer components in each record.
-By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.
+such that the monoid \(\monoid_{i}\) acts on component bundle \(\dtotal_i\).
+The \dbase\ remains the same because the continuity of the data does not change just because there are fewer components in each record.
+
 \begin{figure}[H]
@@
 The datasets in \autoref{fig:math:data:base} have the same fiber of (precipitation, time).
-The dot represents a discrete base space \dbase, meaning that every dataset encoded in the fiber bundle has discrete continuity.
-The line is a representation of a 1D continuity, meaning that every dataset in the fiber bundle is 1D continous.
-By encoding this continuity in the model as \dbase\, the data model now explicitly carries information about its structure such that the implicit assumptions of the visualization algorithms are now explicit.
-The explicit topology is a concise way of distinguishing visualizations that appear identical, for example heatmaps and images.
+The points (a) represent a discrete base space \dbase, meaning that every dataset encoded in the fiber bundle has discrete continuity.
+The line (b) is a representation of a 1D continuity, meaning that every dataset in the fiber bundle is 1D continuous.
+Explicitly encoding data continuity, for example that (a) has discrete continuity and (b) is 1D continuous, provides a means to explicitly specify the continuities visualization components must preserve.
 
@@
 The set of all global sections is denoted as $\Gamma(\dtotal)$.
-Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section is
+Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section can be decomposed as
 \begin{equation}
@@
     \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
-    \caption{ Fiber (time, precipitation) with an interval \dbase\ basespace.
+    \caption{ Fiber (time, precipitation) with a 1D continuous \dbase\  defined on an interval $\left[0,n\right]$.
 The sections $\dsection^{(1)}$ and $\dsection^{(2)}$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of precipitation values.
@@
 \label{sec:math:data:sheaf}
-Many types of dynamic visualizations require evaluating sections on different subspaces of \dbase, an a sheaf, denoted $\mathcal{O}$ provides a way to do so.
-A sheaf is a mathematical structure for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.
+Dynamic visualizations require evaluating sections on different subspaces of \dbase; this can be achieved using a mathematical structure, called a sheaf $\mathcal{O}$, for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.
 On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections $\iota^{*}\dsection$
@@
 
-\subsubsection{Applications to Data Containers}
-This model provides a common formalism for widely used data containers without sacrificing the semantic structure embedded in each container.
-For example, the section can be any instance of a univariate numpy array\cite{harris2020array} that stores an image.
-This could be a section of a fiber bundle where \dbase\ is a 2D continuous plane and the \dfiber\ is $(\reals^{3}, \reals, \reals)$ where $\reals^3$ is color, and the other two components are the x and y positions of the sampled data in the image.
-This position information is already implicitely encoded in the array as the index and the resolution of the image being stored.Instead of an image, the numpy array could also store a 2D discrete table.
-The fiber would not change, but the \dbase\ would now be 0D discrete points.
+\subsubsection{Applications}
+Using fiber bundles as our data abstraction allows the model to be applicable to a wide variety of fields and use cases.
+For example, the section can be any instance of a numpy array\cite{harris2020array} that stores an image, such as an image where the \dbase\ is a 2D continuous plane and the \dfiber\ is $(\reals^{3}, \reals, \reals)$.
+In this fiber, the $\reals^3$ components encode color, and the other two components are the x and y positions of the sampled data in the image.
+The continuity of the image is implicitly encoded in the array as the index, so the position components encode the resolution.
+Instead of an image, the numpy array could also store a 2D discrete table.
+The fiber may not change, but the \dbase\ would now be 0D discrete points.
 These different choices in topology indicate, for example, what sorts of interpolation would be appropriate when visualizing the data.
-
-There are also many types of labeled containers that can richly be described in this framework because of the schema like structure of the fiber.
-For example, a pandas series which stores a labeled list, or a dataframe\cite{jeff_reback_2020_3715232} which stores a relational table.
-A series could store the values of $\dsection^{(1)}$ and a second series could be  $\dsection^{(2)}$.
-We could also fatten the fiber to hold two precipitationure series, such that a section would be an instance of a dataframe with a time column and two precipitation columns.
-While the series and dataframe explicitly have a time index column, they are components in our model and the index is assumed to be data independent references such as hashvalues, virtual memory locations, or random number keys.
-
-Where this model particularly shines are N dimensional labeled data structures.
-For example, an xarray\cite{hoyer2017xarray} data that stores precipitation field could have a \dbase\ that is a continuous volume and the components would be the precipitation and the time, latitude, and longitude the measurements were sampled at.
-A section can also be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.
-As with the other containers, \dbase\ and \dfiber\ are defined in terms of the index and dtypes of the components of the array.
-Because our framework is defined in terms of the fiber, continuity, and sections, rather than the exact values of the data, our model does not need to know what the exact values are until the renderer needs to fill in the image.
+Labeled containers can also be described in this framework because of the schema like structure of the fiber.
+One such example is a pandas series which stores a labeled list, another is a  dataframe\cite{jeff_reback_2020_3715232} which has the structure of a relational table.
+A series could store the values of $\dsection^{(1)}$ and a second series could be  $\dsection^{(2)}$, while a dataframe would have multiple components and each data frame would be a unique section \dsection.
+The ability to encode complexity in continuity and components is particulary beneficial when working with  N dimensional labeled data containers.
+For example, an xarray\cite{hoyer2017xarray} data cube that stores precipitation would be a section of a fiber bundle with a \dbase\ that is a continuous volume and components \textit{(time, latitude, longitude, precipitation)}.
+This section does not need to resolve to values immediately and instead can be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.
+Our model provides a common formalism for describing widely used data containers without sacrificing the semantic structure embedded in each container.
 

### 2021-05-05 draft/sections/math.tex
graphics section

@@
 \label{sec:math:graphic:base}
-\begin{figure}[H]
-    \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
-    \caption{For a visualization component to preserve continuity, it must have a continuous surjective map $\vindex: \gbase\rightarrow\dbase$ from graphic continuity to data continuity.
-The scatter (a) and line (b) graphic base spaces \gbase\ have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
-The image (c) has the same dimension in \gbase\ as in \dbase\ because \dbase\ is already 2D and therefore can directly map into screen space.}
-    \label{fig:math:graphic:retraction:map}
-\end{figure}
-To establish that a visualization component preserves continuity, we propose that their must be a continuous map $\vindex: \gbase\rightarrow\dbase$ from the graphic base space to the data space.
-For example, consider a \gbase\ that is mapped to the region of a 2D display space that represents \dbase.
-For some visualizations, \dbase\ may be lower dimension than \gbase.
-For example, a point that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D to encode the connectivity of the pixels that visually represent the point.
-This thickening is often not necessary when the dimensionality of \dbase\ matches the dimensionality of the target space, for example if \dbase\ is 2D and the display is a 2D screen.
-We introduce \gbase\ to thicken \dbase\ in a way which preserves the structure of \dbase.
 
-Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ so that \dbase\ and \gbase\ have the same homotopy, meaning there is a continuous map from \gbase\ to \dbase\cite{weissteinHomotopy}.
+For a visualization component to preserve continuity, we propose that there must exist a structure preserving surjective map $\vindex: \gbase\rightarrow\dbase$ from the data base space \dbase\ to the graphic base space \gbase.
+Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ such that \dbase\ and \gbase\ have the same homotopy, meaning there is a continuous map from \gbase\ to \dbase\cite{weissteinHomotopy}.
 The surjective map $\vindex: \gbase \rightarrow \dbase$
@@
 \end{equation}
-goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$.
+goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\dbasepoint \in \dbase$.
 This means that if $\vindex(\gbasepoint) = \dbasepoint$, the record at \dbasepoint\ is copied over the region \gbasepoint\ such that $\dsection(\dbasepoint)=\vindex^*\dsection(\gbasepoint)$ where $\vindex^*\dsection(\gbasepoint)$  is \dsection\ pulled back over \gbase.
 The map \vindex\ is part of the implementation of the artist $\mathcal{\vartist}$ and therefore is not defined in terms of the data; instead it is how we specify the constraint that the type of the graphic \textit{continuity} must be able to map to the type of the data \textit{continuity}.
-When \dbase\ is discrete points and the graphic is a scatter plot, each point $\dbasepoint \in \dbase$ can correspond to a 2D disk $\gbase_{\dbasepoint}$ as shown in \autoref{fig:math:graphic:retraction:map}.
-In the case of 1D continuous data and a line plot, the region \gy\ over a point $\gx_i$ specifies the thickness of the line in \gbase\ for the corresponding \dsection\ on \dbasepoint.
-The image has the same dimensions in data space and graphic space such that no extra dimensions are needed in \gbase.
 
-The mapping function \vindex\ provides a way to identify the part of the visual transformation that is specific to the the connectivity of the data rather than the values; for example it is common to flip a matrix when displaying an image.
-The \vindex\ mapping is also used by interactive visualization components to look up the data associated with a region on screen.
+\begin{figure}[H]
+    \includegraphics[width=1\textwidth]{figures/math/retraction_maps.png}
+    \caption{For a visualization component to preserve continuity, it must have a continuous surjective map $\vindex: \gbase\rightarrow\dbase$ from graphic continuity to data continuity.
+The scatter (a) and line (b) graphic base spaces \gbase\ have one more dimension of continuity than \dbase\ so that \gbase\ can encode physical aspects of the glyph, such as shape (a circle) or thickness.
+The image (c) has the same dimension in \gbase\ as in \dbase\ because \dbase\ is already 2D and therefore can directly map into screen space.}
+    \label{fig:math:graphic:retraction:map}
+\end{figure}
+
+To encode the continuity of the elements in the display fiber \gfiber, the graphic base space \gbase\ has the same dimensionality as the target output space.
+For example, in \autoref{fig:math:graphic:retraction:map} the base space \gbase\ is a representation of a region of a 2D display space.
+Since \gbase\ must have the same dimensionality as the output graphic, it is allowed to add dimensions to \dbase\ to make \dbase\ renderable.
+A point that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D (a) to encode the connectivity of the pixels that visually represent the point.
+This is also the case with the line (b), which would be infinitely think on screen if \gbase\ was not thickened to 2D.
+This thickening is often not necessary when the dimensionality of \dbase\ matches the dimensionality of the target space, for example if \dbase\ is 2D and the display is a 2D screen (c).
+Since the mapping function \vindex\ binds the graphic base space to the data base space, it can be used by interactive visualization components to look up the data associated with a region on screen.
 One example is to fill in details in a hover tooltip, another is to convert region selection (such as zooming) on \gbase\ to a query on the data to access the corresponding record components on \dbase.
@@
 \subsubsection{Graphic \gsection}
-\label{sec:graphic_section}
-The section $\gsection: \gbase \rightarrow \gtotal$ is the graphic in an idealizes prerender space and also acts as a specification for rendering the graphic to an image.
-It is sufficient to sketch out how an arbitrary pixel would be rendered, where a pixel $p$ in a real display corresponds to a region $\gbase_p$ in the idealized display.
+\label{sec:math:graphic:section}
+The section $\gsection: \gbase \rightarrow \gtotal$ is the graphic in an idealized prerender space and also acts as a specification for rendering the graphic to target display format.
+To demonstrate the role of \gsection\, it is sufficient to sketch out how an arbitrary pixel would be rendered, where a pixel $p$ in a real display corresponds to a region $\gbase_p$ in the idealized display.
 To determine the color of the pixel, we aggregate the color values over the region via integration:
@@
 For a 2D screen, the pixel is defined as a region $p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ of the rendered graphic.
-Since the x and y in $p$ are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box $\gbase_{p} ={\gsection_{xy}}^{-1}(p)$ is a region $\gbase_p \subset \gbase$.
+Since the x and y in $p$ are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box $\gbase_{p} ={\gsection_{x,y}}^{-1}(p)$ is a region $\gbase_p \subset \gbase$.
 The color is the result of the integration over  $\gbase_p$.
@@
     \caption{To render a graphic, a pixel $p$ is selected in the display space, which is defined in the same coordinates as the x and y components in \gfiber\ via the renderer.
-In \gtotal\, the inverse mapping ${\gsection_{xy}}^(p)$ returns a region $\gbase_{p} \subset \gbase$.
+In \gtotal\, the inverse mapping ${\gsection_{x,y}}^(p)$ returns a region $\gbase_{p} \subset \gbase$.
 $\gsection(\gbase_{p})$ returns a set of points $(x,\,y,\,r,\,g,\,b) \in \gfiber$ that lie over $\gbase_{p}$.
@@
 The section  $\gsection(\gbase_{p})$  then maps into the fiber \gfiber\ over $\gbase_p$ to obtain the set of points in \gfiber, here represented as a table, that correspond to that section.
-The integral over the pixel components of this set of points in the fiber yields \{3, 135, 60\} the actual color of the pixel.
+The integral over the pixel components of this set of points in the fiber yields the color of the pixel.
 In general, \gsection\ is an abstraction of rendering.
-In very broad strokes \gsection\  can be a specification such as PDF\cite{bienz1993portable}, SVG\cite{quintScalable2003}, or an openGL scene graph\cite{CarsonOpenGL1997}.
-Alternatively, \gsection\ can be a rendering engine such as cairo\cite{CairographicsOrg} or AGG\cite{shemanarevAntiGrainGeometry}.
-Implementation of \gsection\ is out of scope for this work,
+In very broad strokes \gsection\ can be a specification such as PDF\cite{bienz1993portable}, SVG\cite{quintScalable2003}, or an openGL scene graph\cite{CarsonOpenGL1997} or a rendering engine such as cairo\cite{CairographicsOrg} or AGG\cite{shemanarevAntiGrainGeometry}.
+Implementation of \gsection\ is out of scope for this proposal.
 

### 2021-05-05 draft/sections/math.tex
q & qhat

@@
 \label{sec:artist}
-The topological artist \vartist\ is how we model the building block component that transforms data into a graphic.
-The artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
+We propose that visualization is structure preserving maps from data \dtotal\ to graphic \gtotal; having described \dtotal in \autoref{sec:math:data} and \gtotal in \autoref{sec:math:graphic}, we now define the visual transformations from \dtotal\ to \gtotal\ that formalize the components that visualization libraries implement.
+The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
 \begin{equation}
@@
 The artist preserves \textit{continuity} through the \vindex\ map discussed in \autoref{sec:math:graphic:base}.
-We conjecture that it is an \textit{equivariant} map because it carries a homomorphism of monoid actions \cite{cegarraCohomologyMonoidsOperators2019}
+We propose that the artist $\mathscr{\vartist}$ is an \textit{equivariant} map of monoid action  $m \in \monoid$
+\begin{equation}
+\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
+\end{equation}
+between data element $\delement \in \mathscr{\dtotal}$ and graphic element $\vartist(\delement) \in \mathscr{\gtotal}$.
+To be equivariant with respect to monoids action, we conjecture that an  artist carries a monoid homomorphism $\varphi$
 \begin{equation}
@@
 \end{equation}
+such that an action in data space $m \in \monoid$ is equivalent to an action in graphic space $\varphi(\monoid) \in \monoid^{\prime}$.
 
-Given \monoid\ on data $\mathscr{\dtotal}$ and $\monoid^{\prime}$ on graphic $\mathscr{\gtotal}$, we propose that artists $\mathscr{\vartist}$ are equivariant maps 
-\begin{equation}
-\vartist(m\cdot \delement) = \varphi(m)\cdot \vartist(\delement) 
-\end{equation}
-such that applying a monoid action $m \in \monoid$ to the data input $\delement \in \mathscr{\dtotal}$ of the artist $\mathscr{\vartist}$ is equivalent to applying a monoid action $\varphi(\monoid) \in \monoid^{\prime}$ to the graphic $\vartist(\delement) \in \mathscr{\gtotal}$ output of the artist.
-
-The monoid equivariant map has two stages: the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components, and the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
+The artist \vartist\ has two stages: the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components, and the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
 \begin{equation}
@@
 \vtotalpull\ is the visual bundle \vtotal\ pulled back over \gbase\ via the equivariant continuity map $\vindex:\gbase \rightarrow \dbase$ introduced in \autoref{sec:math:graphic:base}.
-The functional decomposition of the visualization artist in \autoref{eq:math:artist:diagram}facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
+The functional decomposition of the visualization artist in \autoref{eq:math:artist:diagram} facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
 We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.
@@
 We introduce a visual bundle \vtotal\ to store the mappings of the data components into components of the graphic.
-The visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is the space of possible parameters of a visualization type, such as a scatter or line plot.
-As with the data and graphic bundles, the visual bundle is defined by the projection map $\pi$
+These graphic components are implicit visualization library APIS; by making them explicit as components of the fiber we can define expectations of how these parameters behave.
+As with the data and graphic bundles, the visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is defined by the projection map $\pi$
 \begin{equation}
@@
 where \vsection\ is the visual variable encoding, as described by Bertin \cite{bertinSemiologyGraphicsDiagrams2011a}, of the data section \dsection.
-The visual fiber \vfiber\ is defined in terms of the input parameters of the visualization library's plotting functions; by making these parameters explicit components of the fiber, we can build consistent definitions and expectations of how these parameters behave.
+The visual bundle \vtotal\ is the full design space\cite{cardStructureInformationVisualization1997} of possible parameters of a visualization type, such as a scatter plot or line plot.
+For example, one section \vsection\ of \vtotal\ is a tuple of visual values that specifies the visual characteristics of a part of a graphic.
+
 \begin{table}[H]
@@
 \end{table}
- A section \vsection\ is a tuple of visual values that specifies the visual characteristics of a part of the graphic.
-For example, given a fiber of \(\{x, y, color\}\) one possible section could be  \(\{.5, .5, (255, 20,147)\}\).
-The \(codomain(\vchannel_i)\) determines which monoids can act on $\vfiber_i$.
-These fiber components are implicit in the library, as seen in \autoref{tab:math:artist:mpl:fiber}, and by making them explicit as components of the fiber we can build consistent definitions and expectations of how these parameters behave.
+In \autoref{tab:math:artist:mpl:fiber}, the fiber components are specified by the visual parameter they are encoding.
+Multiple parameters can be encoded with the same transformation from data space to graphic space, for example x and y are both positions on a screen.
+Given a fiber of \(\{x, y, color\}\) one possible section could be  \(\{.5, .5, (255, 20,147)\}\).
+The \(codomain(\vchannel_i)\) in \autoref{tab:math:artist:mpl:fiber} specifies the libraries internal representation of visual variables and can be used to determine which monoids can act on $\vfiber_i$.
 
@@
 \label{sec:math:artist:nu}
+We propose that the map from data components to graphic components $\vchannel: \dsection \mapsto \vsection$ is a monoid \textit{equivariant} map.
+By specifying this constraints, we can guarantee that the stage of the artist that transforms data components into graphic representations is equivariant.
+These constraints then guide the implementation of reusable component transformers \vchannel\ that are composed when generating the graphic.
 We define the visual transformers \vchannel\
@@
 As mentioned in \autoref{sec:math:data:monoid}, monoid actions define the structure on the fiber components and are therefore the basis for equivariance.
-A validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes
+Therefore, a validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes
 \begin{equation}
@@
 \end{equation} 
-such that $\varphi$ maps a monoid action on data to a monoid action on visual elements.
-However, we can construct a monoid action of \monoid\ on $\vfiber_{i}$ that is compatible with a monoid action of \monoid\ on $\dfiber_{i}$.
-We can compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi$ that takes \monoid\ to $\monoid^{\prime}$.
-This allows us to define a monoid action on \vfiber\ of \monoid\ that is $(m, \velement) \rightarrow \varphi(m)\bullet\velement$.
-Therefore, without a loss of generality, we can assume that an action of \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.
+such that $\varphi$ maps a monoid action on data to a monoid action on visual elements. However, 
+without a loss of generality we can assume that an action of \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.
+We can make this assumption because we can construct a monoid action of \monoid\ on $\vfiber_{i}$ that is compatible with a monoid action of \monoid\ on $\dfiber_{i}$.
+We can then compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi$ that takes \monoid\ to $\monoid^{\prime}$.
+This allows us to define a monoid action on \vfiber\ of \monoid\ that is $(m, \velement) \rightarrow \varphi(m)\bullet\velement$, which lets us incorporate $\varphi$ into the action $\bullet$ such that $\varphi$ does not need to be explicitly defined in the constraints.
 
@@
   \centering
+  \begin{subfigure}{.49\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/partial_fixed.png}
+    \caption{}
+    \label{fig:math:artist:nu:valid}
+  \end{subfigure}
   \begin{subfigure}{.49\textwidth}
@@
   \end{subfigure}
-  \begin{subfigure}{.49\textwidth}
-    \includegraphics[width=1\textwidth]{figures/math/partial_fixed.png}
-    \caption{}
-    \label{fig:math:artist:nu:valid}
-  \end{subfigure}
-  \caption{The map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not homomorphic, and therefore invalid, because $rain \geq storm$ is mapped to elements with the reverse ordering $\nu(storm) \geq \nu(storm)$.
-In contrast, the mapping in \autoref{fig:math:artist:nu:valid} is valud since $\nu(storm)=\nu(rain)$ satisfies the condition $\nu(storm) \geq \nu(storm)$}
+  \caption{ The \vchannel\ mapping in \autoref{fig:math:artist:nu:valid} represented by the colored arrows is monotonic, and therefore monoid equivariant,  since $\nu(storm)=\nu(rain)$ satisfies the condition $\nu(storm) \geq \nu(storm)$.
+In contrast, the map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not monotonic, and therefore not monoid equivariant, because $rain \geq storm$ is mapped to elements with the reverse ordering $\nu(storm) \geq \nu(storm)$.}
   \label{fig:math:artist:nu}
 \end{figure}
-
-The translation from weather state data to visual representation as umbrella emoji in \autoref{fig:math:artist:nu:invalid} is an invalid visual encoding map \vchannel\ because it is not homomorphic.
-This is because the monotonic condition \(rain \geq storm \implies \vchannel(rain) \geq  \vchannel(storm)\) is not met since \(\vchannel(rain) \leq\vchannel(storm)\).
+The mapping from weather state to umbrellla in \autoref{fig:math:artist:nu:valid} is monotonic, and therefore we conjecture equivariant, because $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$.
+\autoref{fig:math:artist:nu} is an example of how the model supports partially ordered data components, which was a motivation for defining equivariance as monoid homomorphisms.
+In contrast, the translation from weather state data to visual representation as umbrella emoji in \autoref{fig:math:artist:nu:invalid} is an invalid visual encoding map \vchannel\ because it is not monotonic and therefore not equivariant This is because the monotonic condition \(rain \geq storm \implies \vchannel(rain) \geq  \vchannel(storm)\) is not met since \(\vchannel(rain) \leq\vchannel(storm)\).
 To satisfy the monotonic condition for $rain \geq storm$, either red arrow in \autoref{fig:math:artist:nu:invalid} would have to go in a different direction.
-On the other hand, the mapping from weather state to umbrellla in \autoref{fig:math:artist:nu:valid} is a homomorphism since $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$.
-\autoref{fig:math:artist:nu} is an example of how the model supports partially ordered data components, which was a motivation for defining equivariance as monoid homomorphisms.
 
@@
   \end{tabulary}
-  \caption{}
+  \caption{Equivariance constraints for the Stevens' measurement scales\cite{MeasurementScalesStatistics}}
   \label{tab:math:artist:nu}
@@
 The Stevens measurement types\cite{stevensTheoryScalesMeasurement1946}, listed in \autoref{tab:math:artist:nu}, are specified in terms of groups, which are monoids with invertible operations\cite{remlingAlgebraMath5353}.
-Despite critiques of the scales\cite{johnsonPseudoMathematicsMentalSocial1936,thomasMathematizationNotMeasurement2014}, we believe it is critical for the model to include the measurement scales since they are commonly used in visualization to classify components \cite{munznerVisualizationAnalysisDesign2014,wilkinsonGrammarGraphics2005}.
-By specifying the equivariance constraints on \vchannel\, we can guarantee that the stage of the artist that transforms data components into visual representations is equivariant.
-These constraints guide the implementation of reusable component transformers \vchannel\ that are composed when generating the graphic.
+We generalize to monoids to account for limitations in the types of data that can be described with the Stevens' scales \cite{johnsonPseudoMathematicsMentalSocial1936,thomasMathematizationNotMeasurement2014}
 
 \subsubsection{Visualization Assembly}
+Having described the maps to components in \autoref{sec:math:artist:nu}, we now specify the assembly function \vmarkd\ that composites components in \vtotal\ into a graphic in \gtotal.
+Since the component transformes \vchannel\ are equivariant, the equivariance constraints carry through to \vmarkd.
+We specify these constraints to guide the implementation of library components responsible for generating graphics.
+
 \label{sec:math:artist:q}
@@
   \includegraphics[width=\textwidth]{figures/math/path_of_q.png}
-  \caption{The transform functions $\vchannel_i$ convert data $\dsection_i\in \dtotal$ to visual characteristics $\vsection_i \in \vtotal$, then \vmark\ assembles $\vsection_i$ into a graphic $\gsection \in \gtotal$.}
+  \caption{The transform functions $\vchannel_i$ convert data $\dsection_i\in \dtotal$ to visual characteristics $\vsection_i \in \vtotal$.
+These visual components $\vsection_i$ are then assembled by \vmark\ into a graphic $\gsection \in \gtotal$.}
   \label{fig:math:artist:q}
 \end{figure}
-The transformation from data into graphic is analagous to a map-reduce operation; as illustrated in \autoref{fig:math:artist}, data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal.
+The transformation from data into graphic is analagous to a map-reduce operation; as illustrated in \autoref{fig:math:artist:q}, data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal.
 The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$.
@@
 
-For example, given fiber $\vfiber=(xpos,\, ypos,\, color,\, thickness)$, then sections $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin hollow circle.
+For example, given fiber $\vfiber=(xpos,\, ypos,\, color,\, thickness)$, then sections $\vsection=(0,0,0,1)$ and $\vmark(\vsection) = \gsection$ generates a piece of the thin circle.
 The action $m=(e, e, e, x+2)$, where e is identity, translates \vsection\ to  $\vsection^{\prime}=(e,e,e,3)$ and the corresponding action on \gsection\ causes $\vmark(\vsection^{\prime})$ to be the thicker circle in \autoref{fig:math:artist:graphic}.
@@
 \begin{equation}
-\dbasepath = \{\dbasepathpoint \in \dbase \text{ s. t. } \exists \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
+\dbasepath = \{\dbasepathpoint \in \dbase \text{ exists } \gamma \text{ s.t. } \gamma(0)=\dbasepoint \text{ and }\gamma(1)=\dbasepathpoint\}
 \end{equation}
@@
 \end{equation}
-such that for every glyph there is at least one corresponding region on \dbase, in keeping with the definition of glyph as any visually differentiable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+such that for every glyph there is at least one corresponding region on \dbase, in keeping with the definition of glyph as any visually distinguishable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
 The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.
@@
     \includegraphics[width=1\textwidth]{figures/math/scatter.png}
-    \caption{The data is discrete points (precipitation, time).
-Via \vchannel\ these are converted to (xpos, ypos) and pulled over discrete \gbase.
-These values are then used to parameterize \gsection\ which returns a color based on the parameters (xpos,ypos) and position $\alpha, \beta$ on $\gbase_{\dbasepoint}$ that \gsection\ is evaluated on.
-    }
+    \caption{The data is discrete points (time, precipitation).
+Via \vchannel\ these are converted to (xpos, ypos) and pulled over discrete \gbase\ via $\vindex^*$.
+The pulled back visual secion $\vchannel$ is composited with the assembly function $\vmarkd \circ  \vchannel=\gsection$ to produce the instructions to make the graphic \gsection.
+The graphic section fills in the pixels in the screen via lookup on \gbase.}
     \label{fig:math:artist:q:scatter}
@@
     \includegraphics[width=\textwidth]{figures/math/line.png}
-    \caption{The line fiber $(time,\, temp)$ is thickened with the derivative $(time^{\prime},\, precipitation^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.
+    \caption{The line fiber $(time,\, precipitation)$ is thickened with the derivative $(time^{\prime},\, precipitation^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.
 This is because the line needs to be pushed perpendicular to the tangent of (xpos, ypos).
@@
 \end{figure}
-In contrast, the line plot in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase.
+In contrast, the line plot in \autoref{fig:math:artist:q:line} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along the interval \dbasepoint\ and corresponding region in \gbase.
 
@@
 
-\subsubsection{ \vmarkd}
+\subsubsection{Assembly Template\vmarkd}
 \label{sec:math:artist:qhat}
-
-\begin{figure}[htb]
-    \centering
-      \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
-      \caption{Because the pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint, we can constract a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
-      \label{fig:math:artist:qhat}
-  \end{figure}
-
 The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
@@
 \end{equation}
-The preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back of \vsection\
+
+The map from graphic base space $\vindex:\gbase \rightarrow \dbase$ (\autoref{sec:math:graphic:base}) to data space maps many points in \gbase\ to a single point in \dbase.
+This means that the preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back of \vsection\
 \begin{equation}
@@
 
-As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is a direct map $(\dbasepoint, \vsection(\dbasepoint)) \mapsto (\gbasepoint, \vsectionpull(\gbasepoint))$  from \vsection\ over \dbasepoint\ to the section \vsectionpull\ over \gbasepoint.
+\begin{figure}[H]
+    \centering
+      \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
+      \caption{Because the pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint, we can construct a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
+      \label{fig:math:artist:qhat}
+  \end{figure}
+
+As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is mapping from section \vsectionpull\ over \gbasepoint to \vsection\ over \dbasepoint.
 This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
@@
 where \vmarkd\ with input \vsection\ is defined to \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.
-Factoring out \gbasepoint\ from \autoref{eq:math:artist:qhat_q_s} yields
+We can then factor \gbasepoint\ out of \autoref{eq:math:artist:qhat_q_s}, which yields
 \begin{equation}
@@
 This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
-The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to majorly restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
+The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
 
@@
 \end{subfigure}
-\caption{Each of the visual elements in \autoref{fig:math:seprate} is generated via a unique artist \vartist.
-In \autoref{fig:math:seperate}, they are added to the image independent of the other elements, creating an incoherant visualization.
-In \autoref{fig:math:combined}, these artists are composited before being added to the image.
+\caption{ In \autoref{fig:math:combined}, these artists are composited before being added to the image.
 Disjoint union of \dtotal\ aligns the two timeseries with the x and y axis so all these elements use a shared coordinate system.
-A more complex composition dictates that the legend is connected to the \dtotal\ such that it must use the same color as the data it is identifying.}
+A more complex composition dictates that the legend is connected to the \dtotal\ such that it must use the same color as the data it is identifying.
+None of this machinary exists in \autoref{fig:math:seperate}, therefore each artist is added to the page independented of the other elements.}
 \label{fig:artist_plus}
 \end{figure}
+
 We describe the constraints for compositing artists by defining addition operators. Given the family of artists $(\dtotal_i: i\in I)$ that are rendered to the same image, the + operator 

### 2021-05-06 draft/sections/math.tex
artist equivalence

@@
 \end{figure}
-The mapping from weather state to umbrellla in \autoref{fig:math:artist:nu:valid} is monotonic, and therefore we conjecture equivariant, because $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$.
+The mapping from weather state to umbrella in \autoref{fig:math:artist:nu:valid} is monotonic, and therefore we conjecture equivariant, because $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$.
 \autoref{fig:math:artist:nu} is an example of how the model supports partially ordered data components, which was a motivation for defining equivariance as monoid homomorphisms.
@@
 \end{figure}
-The transformation from data into graphic is analagous to a map-reduce operation; as illustrated in \autoref{fig:math:artist:q}, data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal.
+The transformation from data into graphic is analogous to a map-reduce operation; as illustrated in \autoref{fig:math:artist:q}, data components $\dtotal_{i}$ are mapped into visual components $\vtotal_{i}$ that are reduced into a graphic in \gtotal.
 The space of all graphics that \vmark\ can generate is the subset of graphics reachable via applying the reduction function $\vmark(\Gamma(\vtotal)) \in \Gamma(\gtotal)$ to the visual section $\vsection \in \Gamma(\vtotal)$.
@@
 
-We formalize the expectation that visualization generation functions parameterized in the same way should generate the same functions as the equivariant map $\vmark: \vsection \mapsto \gsection$.
+We define the visualization assembly function $\vmark: \vsection \mapsto \gsection$ as an equivariant map to formalize the expectation that two \vmark\ functions parameterized in the same way should generate the same graphic.
 We then define the constraint on \vmark\ such that if \vmark\ is applied to two visual sections $\vsection$ and $\vsection^{\prime}$ that generate the same \gsection\, then the output of $\vsection$ and $\vsection^{\prime}$ acted on by the same monoid $m$ must be the same.
@@
 
-\subsubsection{Assembly Template\vmarkd}
+\subsubsection{Assembly Template \vmarkd}
 \label{sec:math:artist:qhat}
@@
 \end{figure}
-
-We describe the constraints for compositing artists by defining addition operators. Given the family of artists $(\dtotal_i: i\in I)$ that are rendered to the same image, the + operator 
+Visualizations generally consist of more than one artist, commonly having visual elements such as the plot and axis labels and maybe legends.
+To generate these composite images, we define addition operators and specify the constraints for compositing artists.
+Given the family of artists $(\dtotal_i: i\in I)$ that are rendered to the same image, the + operator
 \begin{equation}
@@
 defines a simple composition of artists.
-For example, the components in \autoref{fig:math:seperate} are each generated by different artists, and a visualization of solely the x axis is rarely all that useful.
-In \autoref{fig:math:seperate}, these artists are all added to the image independently of the other and therefore there are no constraints on how they are generated in conjunction with each other.
-In \autoref{fig:math:combined}, the data is joined via disjoint union; doing so aligns the components in \dfiber\ such the \vchannel\ to the same component in \vfiber\ targets the same coordinate system.
+For example, in \autoref{fig:math:combined} the data is joined via disjoint union; doing so aligns the components in \dfiber\ such the \vchannel\ to the same component in \vfiber\ targets the same coordinate system.
+In \autoref{fig:math:seperate}, these artists are all added to the image independently of the other and therefore there are no constraints on where they are placed in the image.
 When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the artists are acting on different components of the same section.
@@
    \caption{Each scatter plot is generated via a unique artist function $\vartist_{i}$, but they only differ in aesthetic styling.
-Therefore, these artists are all members of an equivalence class $\vartist_{i} \in \vartist{\prime}$}
+Therefore, these artists are all members of an equivalence class $\vartist_{i} \in \vartisteq$}
     \label{fig:math:artist:equivalence}
@@
 Representational invariance, as defined by Kindlmann and Scheidegger, is the notion that visualizations are equivalent if changing the visual representation, such as colors or shapes, does not change the meaning of the visualization\cite{kindlmannAlgebraicProcessVisualization2014}.
+By defining a criteria for invariance, we can evaluate whether two artists generate the same type of graphic and compare artists across libraries.
 We propose that visualizations are invariant if they are generated by artists that are members of an equivalence class
 \begin{equation*}
-\{\vartist \in \vartist^{\prime}: \vartist_{1} \equiv \vartist_{2}\}
+\{\vartist \in \vartisteq: \vartist_{1} \equiv \vartist_{2}\}
 \label{eq:math:artist:aprime}
@@
 The scatter plots only differ in the choice of constant visual literals, differing in color and marker shape.
-Each scatter is generated by an artist $\vartist_{i}$, and every scatter is generated by a member of the equivalence class $\vartist_{i} \in \vartist{\prime}$.
+Each scatter is generated by an artist $\vartist_{i}$, and every scatter is generated by a member of the equivalence class $\vartist_{i} \in \vartisteq$.
 Since it is impractical to implement a new artist for every single graphic, the equivalance class provides a way to evaluate an implementation of a generalized artist.
-Given equivalent, but no necessarily identical, \vchannel, \vmark, and \vindex, two artists are equivalent.
-This criteria also allows for comparing artists across libraries.
 \end{document}

### 2021-05-09 draft/sections/implementation.tex
clean ups

@@
           tau = self.E.view(self.axes) 
-          # visual channel encoding applied fiberwise 
+          # visual channel encoding applied component wise 
           mu = {p: nu(tau(c))
@@
 \end{figure}
-The figure in \autoref{fig:code:scatter} is described by \autoref{eq:math:artist:q:scatter}.
+The figure in \autoref{fig:code:scatter} is described by \autoref{eq:eq:math:artist:q:scatter}.
 This is implemented via a \mintinline{python}{Line} object where the scatter marker shape is fixed as a circle, and the visual fiber components are x and y position and the facecolor and size of the marker.
@@
 As with \mintinline{python}{Point}, the data is then converted into visual variables.
-In \mintinline{python}{q_hat}, described by \autoref{eq:math:artist:q:line}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
+In \mintinline{python}{q_hat}, described by \autoref{eq:eq:math:artist:q:line}, this visual representation is composed into a set of line segments, where each segment is the array generated by \mintinline{python}{np.vstack((vx, vy))}.
 Then the colors of each line segment are set.
@@
 \begin{figure}[H]
-    \begin{subfigure}{0.5\textwidth}
+    \begin{subfigure}{0.49\textwidth}
         \includegraphics[width=\textwidth]{figures/code/bar_v.png}
     \end{subfigure}
-    \begin{subfigure}{0.5\textwidth}
+    \begin{subfigure}{0.49\textwidth}
         \includegraphics[width=\textwidth]{figures/code/bar_h.png}
@@
                     facecolors, edgecolors, offset): 
-        # offset is passed through via assemblers such as multigroup, 
-        # not supposed to be directly tagged to position 
         position = position + offset
@@
 This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{python}{view} method that returns an atomic object that encapsulates \dsection.
-To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataclassesDataClasses}
+To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataClasses}
 
@@
 \begin{figure}[htb]
-    \begin{subfigure}{.5\textwidth}
+    \begin{subfigure}{.49\textwidth}
         \includegraphics[width=\textwidth]{figures/code/linec_1.png}
     \end{subfigure}
-    \begin{subfigure}{.5\textwidth}
+    \begin{subfigure}{.49\textwidth}
         \includegraphics[width=\textwidth]{figures/code/lined_1.png}
@@
 \begin{minted}{python}
-GraphData(FB, edges, verticies, num_samples=100, connect=True)
+GraphData(FB, edges, vertices, num_samples=100, connect=True)
 \end{minted}
@@
 \begin{minted}{python}
-GraphData(FB, edges, verticies, num_samples=2, connect=False)
+GraphData(FB, edges, vertices, num_samples=2, connect=False)
 \end{minted}
@@
 Instead, grouped bar charts are often achieved via looping over an implementation of bar, which forces the user to keep track which values are mapped to a single visual element and how that is achieved.
-For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020}, packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.
+For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020}, packaged as a pandas data frame\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.
 
@@
 \begin{figure}[H]
-    \begin{subfigure}{0.5\textwidth}
+    \begin{subfigure}{0.49\textwidth}
         \includegraphics[width=\textwidth]{figures/code/bar_stacked.png}
@@
     \end{subfigure}
-    \begin{subfigure}{0.5\textwidth}
+    \begin{subfigure}{0.49\textwidth}
         \includegraphics[width=\textwidth]{figures/code/bar_grouped.png}
@@
 In effect, we are decomposing \dtotal\ into $\dtotal_1, \cdots, \dtotal_i, \cdots, \dtotal_n$ via specifications in \mintinline{python}{V} rather than by directly taking subsections of the table.
-This allows us to ensure shared \dbase\ and coherant \dsection.
+This allows us to ensure shared \dbase\ and coherent \dsection.
 

### 2021-05-09 draft/sections/intro.tex
clean ups

@@
 For example, position encoding functions map the latitude and longitude values in the table (a) to locations on the map (b).
-These same encoding functions map dates and precipitation (a) to x and y positions in the timeseries plot (c).
-Color encoding functions map the precipitation (a) to the color of the point in the map (a) and color encoding functions map names of weather stations (a) to colors in both the map (a) and timeseries (b).
+These same encoding functions map dates and precipitation (a) to x and y positions in the time series plot (c).
+Color encoding functions map the precipitation (a) to the color of the point in the map (a) and color encoding functions map names of weather stations (a) to colors in both the map (a) and time series (b).
 Functions, which we call \textit{artists}, compose these encoding functions into attributes of visual elements.
-The \texttt{point} artist transforms the output of the position, facecolor, and edgecolor maps into attributes of each point in the map (a), while the \texttt{line} artist transforms the output of the position and color maps into attributes of each line in the timeseries plot (c).}
+The \texttt{point} artist transforms the output of the position, face color, and edge color maps into attributes of each point in the map (a), while the \texttt{line} artist transforms the output of the position and color maps into attributes of each line in the time series plot (c).}
     \label{fig:intro:artist}
@@
 A \texttt{point} function composites these encodings into attributes of a point in the map (a).
-The \texttt{line} function composites encoders that convert station name to color and date and precipitation to x and y positions into a piece of a line in the timeseries plot (c).
+The \texttt{line} function composites encoders that convert station name to color and date and precipitation to x and y positions into a piece of a line in the time series plot (c).
 We identify the structure these maps must preserve as the \textit{continuity} of the data and the \textit{equivariance} of the data and visual components.
@@
 This conversion is a scaling action, represented by the orange arrows in (a).
-The \textit{equivariant} action is the scaling in the timeseries (b), which is represented by the orange fill between the green and blue lines.
+The \textit{equivariant} action is the scaling in the time series (b), which is represented by the orange fill between the green and blue lines.
 In (c), the scaling is illustrated in the different millimeter and inches y axis labeling of data that appears identical since the distances were scaled by a constant factor that is now incorporated in the labeling.}
@@
 This scaling action is represented by the green arrows that translate the precipitation into the scaled values in green.
-For this visualization to be equivariant, this same scaling factor must be present in the timeseries representation of the data (b).
-The precipitation in green is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the green line is $\frac{1}{25.4}$ that of the y-values of the blue precipitation timeseries.
-The orange fill in the timeseries (b) is the scaling equivalent to the arrows in the table (a).
-By definition, the distances in the millimeter data were scaled equally \cite{stothersSimilarityGroup} when converted to inches, which means the shapes of the graphs are equivalent when plotted against y-axes that preserve relative distance (c).
+For this visualization to be equivariant, this same scaling factor must be present in the time series representation of the data (b).
+The precipitation in green is an \textit{equivariant} scaling of the precipitation in blue, meaning that the y-values of the green line is $\frac{1}{25.4}$ that of the y-values of the blue precipitation time series.
+The orange fill in the time series (b) is the scaling equivalent to the arrows in the table (a).
+By definition, the distances in the millimeter data were scaled equally \cite{stothersKleinViewGeometry} when converted to inches, which means the shapes of the graphs are equivalent when plotted against y-axes that preserve relative distance (c).
 This visualization is also equivariant to the table (a), but this is indicated through labeling rather than the line plots.
@@
     \end{subfigure}
-    \caption{Visualization libraries, especially ones tied to specific domains, tend to be architectured around a core data structure, such as tables \autoref{fig:intro:table}, images \autoref{fig:intro:image}, or networks \autoref{fig:intro:graph}.
+    \caption{Visualization libraries, especially ones tied to specific domains, often are designed around a core data structure, such as tables \autoref{fig:intro:table}, images \autoref{fig:intro:image}, or networks \autoref{fig:intro:graph}.
     %%revise as combined fig?
@@
 For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience\cite{heerDeclarative2010}.
-This coherant experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries.
+This coherent experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries.
 For example tools influenced by APT that assume that data is a relational table integrate computation into the visualization pipeline.

### 2021-05-09 draft/sections/math.tex
clean ups

@@
 To explain how the \textit{artist} is a structure preserving map from data to graphic, we first model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as topological structures that encapsulate component types and continuity.
-We then discuss the funtional maps from graphic to data (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}), and visual components into graphic (\autoref{sec:math:artist:q}) that make up the artist.
+We then discuss the functional maps from graphic to data (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}), and visual components into graphic (\autoref{sec:math:artist:q}) that make up the artist.
 
@@
 \end{align*} 
-As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
+As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlabauthorsAction2021} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
 \begin{align*}
@@
 \begin{figure}[H]
-    \begin{subfigure}{.5\textwidth}
+    \begin{subfigure}{.49\textwidth}
         \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
@@
     \end{subfigure}
-    \begin{subfigure}{.5\textwidth}
+    \begin{subfigure}{.49\textwidth}
         \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
@@
 \end{equation}
-of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$
+of each monoid \(\monoid_{i}\) on \(\dfiber_{i}\).  The monoid is added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)
 
-\subsubsection{Continuity of the Data $K$} 
+\subsubsection{Continuity of the Data} 
 \label{sec:math:data:base}
@@
 The base space \dbase\ provides a way to explicitly encode the continuity of the data, as described in \autoref{sec:intro:data:continuity}.
-This explicit topology is a concise way of distinguishing between visualizations that appear identical but assume different continuity, for example heatmaps and images.
+This explicit topology is a concise way of distinguishing between visualizations that appear identical but assume different continuity, for example heat maps and images.
 The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other.
@@
 In \autoref{fig:math:data:base:div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the closed interval $\left[0,1\right]$.
-While the total space \dtotal\ may have components in \dfiber\ that describe any given point $\dbasepoint \in \dbase$, such as \textit{time, latitutde, longitude}, these labels are indexed into from \dbase\ the same as any other components.
+While the total space \dtotal\ may have components in \dfiber\ that describe any given point $\dbasepoint \in \dbase$, such as \textit{time, latitude, longitude}, these labels are indexed into from \dbase\ the same as any other components.
 In contrast to the structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, our model treats keys \dbasepoint\ as a pure reference to topology.
@@
 
-\subsubsection{Data \dsection}
+\subsubsection{Data Values}
 \label{sec:math:data:section}
@@
     \caption{ Fiber (time, precipitation) with a 1D continuous \dbase\  defined on an interval $\left[0,n\right]$.
-The sections $\dsection^{(1)}$ and $\dsection^{(2)}$ are constrained such that the time variable must be monotonic, which means each section is a timeseries of precipitation values.
+The sections $\dsection^{(1)}$ and $\dsection^{(2)}$ are constrained such that the time variable must be monotonic, which means each section is a time series of precipitation values.
 They are included in the global set of sections  $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$}
@@
 The section applied to a set of points in \dbase\ resolves to a series of monotonically increasing in time records of \textit{(time, precipitation)} values.
-Section $\dsection^{(2)}$ returns a different timeseries of \textit{(time, precipitation)} values.
+Section $\dsection^{(2)}$ returns a different time series of \textit{(time, precipitation)} values.
 Both sections are included in the global set of sections $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$.
@@
 A series could store the values of $\dsection^{(1)}$ and a second series could be  $\dsection^{(2)}$, while a dataframe would have multiple components and each data frame would be a unique section \dsection.
-The ability to encode complexity in continuity and components is particulary beneficial when working with  N dimensional labeled data containers.
+The ability to encode complexity in continuity and components is particularly beneficial when working with  N dimensional labeled data containers.
 For example, an xarray\cite{hoyer2017xarray} data cube that stores precipitation would be a section of a fiber bundle with a \dbase\ that is a continuous volume and components \textit{(time, latitude, longitude, precipitation)}.
@@
 
-\subsubsection{Graphic \gsection}
+\subsubsection{Graphic}
 \label{sec:math:graphic:section}
@@
 
-\subsubsection{Visual Encoders \vchannel}
+\subsubsection{Visual Encoders}
 \label{sec:math:artist:nu}
@@
 Having described the maps to components in \autoref{sec:math:artist:nu}, we now specify the assembly function \vmarkd\ that composites components in \vtotal\ into a graphic in \gtotal.
-Since the component transformes \vchannel\ are equivariant, the equivariance constraints carry through to \vmarkd.
+Since the component transforms \vchannel\ are equivariant, the equivariance constraints carry through to \vmarkd.
 We specify these constraints to guide the implementation of library components responsible for generating graphics.
@@
 then a monoid action on $Y$ can be defined as $m\circ \gsection = \gsection^{\prime}$.
-If and only if \vmark\ satistfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection.
+If and only if \vmark\ satisfies \autoref{eq:math:artist:q}, we can state that the transformed graphic $\gsection^{\prime}=\vmark(m\circ \vsection)$ is equivariant to a monoid action applied on \vmark\ with input $\vsection \in \vmark^{-1}(\gsection)$ that must generate valid \gsection.
 \end{prop}
@@
 Via \vchannel\ these are converted to (xpos, ypos) and pulled over discrete \gbase\ via $\vindex^*$.
-The pulled back visual secion $\vchannel$ is composited with the assembly function $\vmarkd \circ  \vchannel=\gsection$ to produce the instructions to make the graphic \gsection.
+The pulled back visual section $\vchannel$ is composited with the assembly function $\vmarkd \circ  \vchannel=\gsection$ to produce the instructions to make the graphic \gsection.
 The graphic section fills in the pixels in the screen via lookup on \gbase.}
@@
 
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.49\textwidth}
     \begingroup \leqnomode
@@
 \end{minipage}
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.49\textwidth}
     \centering
@@
     \includegraphics[width=\textwidth]{figures/math/line.png}
-    \caption{The line fiber $(time,\, precipitation)$ is thickened with the derivative $(time^{\prime},\, precipitation^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.
+    \caption{The line fiber \((time,\, precipitation)\) is thickened with the derivative $(time^{\prime},\, precipitation^{\prime}$ because that information will be necessary to figure out the tangent to the point to draw a line.
 This is because the line needs to be pushed perpendicular to the tangent of (xpos, ypos).
@@
 \begin{figure} [H]
-    \begin{minipage}{.5\textwidth}
+    \begin{minipage}{.45\textwidth}
         \begingroup \leqnomode
@@
     \end{minipage}
-    \begin{minipage}{.5\textwidth}
+    \begin{minipage}{.45\textwidth}
         \centering
@@
 \end{align*}
-In the case of an image, the indexing mapper \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.
+In the case of an image, the indexing mapper \vindex\ may do some translating to a convention expected by \vmark, for example reorienting the array such that the first row in the data is at the bottom of the graphic.
 
@@
     \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
-    \label{eq:math:artist:qhat}
+    \label{eq:math:artist:qhat_factor}
 \end{equation}
@@
 \caption{ In \autoref{fig:math:combined}, these artists are composited before being added to the image.
-Disjoint union of \dtotal\ aligns the two timeseries with the x and y axis so all these elements use a shared coordinate system.
+Disjoint union of \dtotal\ aligns the two time series with the x and y axis so all these elements use a shared coordinate system.
 A more complex composition dictates that the legend is connected to the \dtotal\ such that it must use the same color as the data it is identifying.
-None of this machinary exists in \autoref{fig:math:seperate}, therefore each artist is added to the page independented of the other elements.}
+None of this machinery exists in \autoref{fig:math:seperate}, therefore each artist is added to the page independent of the other elements.}
 \label{fig:artist_plus}
@@
 
-\subsubsection{Equivalance class of artists $\vartist^{\prime}$}
+\subsubsection{Equivalance class of artists}
 \label{sec:artist_equivalance}
@@
 Each scatter is generated by an artist $\vartist_{i}$, and every scatter is generated by a member of the equivalence class $\vartist_{i} \in \vartisteq$.
-Since it is impractical to implement a new artist for every single graphic, the equivalance class provides a way to evaluate an implementation of a generalized artist.
+Since it is impractical to implement a new artist for every single graphic, the equivalence class provides a way to evaluate an implementation of a generalized artist.
 \end{document}

### 2021-05-09 draft/sections/next.tex
clean ups

@@
 TEAM expresses the specifications that graphic and data must have equivalent \textit{continuity} equivalent to the data, and that the visual characteristics of the graphics are \textit{equivariant} to their corresponding components.
-TEAM expresses these constraints in the encoding \vchannel, assembly \vmark, and reindexing \vindex\ functions that make up the artist \vartist.
+TEAM expresses these constraints in the encoding \vchannel, assembly \vmark, and indexing \vindex\ functions that make up the artist \vartist.
 This decomposition of the artist into smaller components functional pieces allows TEAM to provide well specified guidance on implementing visualization components based on this architecture.
@@
 This prototype demonstrates that this framework can generate the fundamental point (scatter plot) and line (line chart) marks.
-Furthermore, combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets TEAM support a variety of data continuities, including  discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks.
+Furthermore, combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets TEAM support a variety of data continuities, including  discrete relational tables, multivariate high resolution spatio-temporal datasets, and complex networks.
 Although the prototype currently only implements 0D and 1D continuity, we expected it to generalize to the other continuities.
@@
 The prototype is deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as R graphics\cite{murrellGraphicsThirdEdition2018}, VTK, and D3.
-TEAM has only been tested using PNGs rendered with AGG\cite{shemanarevAntiGrainGeometry}, but is expected to work with all the file types Matplotlib currently supports, including svg, pdf, and eps.
+TEAM has only been tested using PNG files rendered with AGG\cite{shemanarevAntiGrainGeometry}, but is expected to work with all the file types Matplotlib currently supports, including svg, pdf, and eps.
 We have not yet addressed how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997} that implement different models of \gsection.
@@
 More work is needed to formalize the composition operators, equivalence class \vartisteq, and the mathematical model of interactivity.
-We also need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}.
+We also need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heat map\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}.
 In summary, the proposed scope of work is
@@
 This specification of structure could also be used to develop a serialization structure tht could then be use to allow Matplotlib to interface with other visualization libraries such as open GL via shared serialization protocol.
-While this paper formulates visualization in terms of monoidal action homomorphisms between fiberbundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
+While this paper formulates visualization in terms of monoidal action homomorphisms between fiber bundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
 
 \section{Conclusion}
-A TEAM driven refactor of visualizations libraries could produce more maintainable, reusable, and extensible code, leading to better building blocks for the ecosystem of tools built on top of TEAM architectured libraries.
+A TEAM driven refactor of visualizations libraries could produce more maintainable, reusable, and extensible code, leading to better building blocks for the ecosystem of tools built on top of libraries with a TEAM driven architecture.
 Building block libraries could better support downstream, including domain specific, libraries without having to explicitly incorporate the specific data structure and visualization needs of those domains back into the base library.

### 2021-05-09 figures/intro_plots.py
more clean up + regen figures

@@ -66,7 +66,8 @@ def plot_map(ax, gdf, nystate, cmap, norm, fade='k', alpha=.5, station=None):
     ax1_divider = make_axes_locatable(ax)
     cax = ax1_divider.append_axes("right", size="5%", pad="2%")
     cb = mpl.colorbar.ColorbarBase(cax, cmap=cmap,norm=norm, orientation='vertical', drawedges=False, alpha=alpha)
-    cb.set_label('total precipitation (in.)', color=fade)
+    cb.set_label('total precipitation (in.)', 
+                 rotation=270, color=fade, labelpad=12)
     for axes in [ax, cax]:
         axes.tick_params(color=fade, labelcolor=fade)
         axes.spines[:].set_color(fade)
@@ -82,7 +83,8 @@ def plot_time(ax, df, fade='k', alpha=.5, station=None):
     ax.xaxis.set_major_formatter(mdates.ConciseDateFormatter(mdates.AutoDateLocator()))
     ax.tick_params(color=fade, labelcolor=fade)
     ax.spines[:].set_color(fade)
-    ax.set_ylabel('precipitation (in.)', color=fade)
+    ax.set_ylabel('precipitation (in.)', 
+                  rotation=270,  color=fade, labelpad=12)
     ax.yaxis.set_label_position('right')
     ax.tick_params('y', labelright=True, labelleft=False, right=False, left=False)
     return 

### 2021-05-10 draft/sections/intro.tex
intro read through

@@
 
-Visualizations, by definition, reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008} of the data, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+Visualizations are representations of data and therefore reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
 We define visualization components to be structure preserving maps from data to visual representations.
@@
 We introduce a model of visualization components based on these \textit{continuity} and \textit{equivariance} constraints and use this model to develop a design specification.
-Our specification is targeted at building block libraries, as defined by Wongsuphasawat\cite{wongsuphasawatNavigatingWideWorld2021}, because they are the visualization tools that provide independent functions that map components of data to visual representations.
+Our specification is targeted at building block libraries, which Wongsuphasawat defines as visualization tools that provide independent functions that map components of data to visual representations\cite{wongsuphasawatNavigatingWideWorld2021}.
 Just as the number of ways to arrange physical building blocks is solely constrained by the size and shape of the blocks, we propose that the only restriction on how building block library components can be composed are that the compositions preserve \textit{continuity} and \textit{equivariance}.
@@
 The bar chart preserves the \textit{continuity} of the stations by encoding the discrete station data as discrete bars (b).
-In contrast, the line plot (c) does not preserve this discrete \textit{continuity} because it connects the total temperature at each station by drawing a line through each point.
+In contrast, the line plot (c) does not preserve this discrete \textit{continuity} because it connects the total temperature at each station with a line.
 Doing so implies that the total temperatures are points along a 1D continuous line, whereas there is no connectivity between the stations or their corresponding rows in the table.}
@@
 Bertin asserted that \textit{continuity} was preserved by choosing graphical marks that match the \textit{continuity} of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark.
-Informally, Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis} states that a visualization is easier to understand when the properties of the visualization match the properties of the data.
+Informally, Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis,norman_things_smart} states that a visualization is easier to understand when the properties of the visualization match the properties of the data.
 This principal is made more concrete in Tufte's concept of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements\cite{tufteVisualDisplayQuantitative2001}.
@@
 \end{enumerate}
-We validate our model by using it to re-design artist and data access layer of Matplotlib, a general purpose visualization tool.
+We validate our model by using it to re-design the artist and data access layer of Matplotlib.
 We evaluate whether the redesign is successful by comparing it to existing implementation, recreating existing domain specific functionality with the new components, and by implementing components that provide new functionality in Matplotlib.

### 2021-05-10 draft/sections/math.tex
data section

@@
 \end{equation}
-The \textit{artist} $\mathscr{\vartist}$ is a map from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$ fiber bundles.
+The \textit{artist} \(\mathscr{\vartist}\) is a map from data \(\mathscr{\dtotal}\) to graphic \(\mathscr{\gtotal}\) fiber bundles.
 To explain how the \textit{artist} is a structure preserving map from data to graphic, we first model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as topological structures that encapsulate component types and continuity.
@@
 We use fiber bundles as the data model because they are inclusive enough to express all the types of structures of data described in \autoref{sec:intro:data:tools}.
-A fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
+A fiber bundle is a tuple \((\dtotal,\,\dbase,\,\pi ,\,\dfiber)\) defined by the projection map \(\pi\)
 \begin{equation}
@@
 The fiber bundle models the properties of data component types \dfiber\ (\autoref{sec:math:data:fiber}), the continuity of records \dbase\ (\autoref{sec:math:data:base}), the collections of records (\autoref{sec:math:data:section}), and the space \dtotal\ of all possible datasets with these components and continuity.
-By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartesian product $\dbase\times\dfiber$.
+By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood \(U\) the total space is the cartesian product \(\dbase\times\dfiber\).
 
@@
 To formalize the structure of the data components, we use Spivak's description of the schema \cite{spivakSIMPLICIALDATABASES} as a fiber bundle to bind the components of the fiber to variable names and data types.
-Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset.
+Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types \(\{\ftype_0, \ldots, \ftype_m\} \in \ftypes\), where \ftypes\ are the data types of the variables in the dataset.
 He then defines the single variable set \fttype\
@@
 where \fsection\ is the schema that binds a variable name \fname\ to its datatype \ftype.
-A dataset with multiple components has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
+A dataset with multiple components has a fiber that is the cartesian cross product of \(\ftotal_{\fsection}\) applied to all the columns:
 \begin{equation}
@@
 As shown in \autoref{fig:math:data:base:types}, \dbase\ can have any number of dimensions and can be continuous or discrete.
-Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every $\dbasepoint \in \dbase$ has a corresponding fiber $\dfiber_k$\cite{QuotientSpaceTopology2020}.
+Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every \(\dbasepoint \in \dbase\) has a corresponding fiber \(\dfiber_k\)\cite{QuotientSpaceTopology2020}.
 
@@
     \caption{The total space \dtotal\ is divided into fiber segments \dfiber.
-The base space \dbase\ acts as an index into the records in the fibers, such that every point \dbasepoint\ has a corresponding fiber $\dfiber_{\dbasepoint}$.
-The projection map $\pi$ maps every fiber $\dfiber_{\dbasepoint}$ to a point $\dbasepoint \in \dbase$ in the base space.}
+The base space \dbase\ acts as an index into the records in the fibers, such that every point \dbasepoint\ has a corresponding fiber \(\dfiber_{\dbasepoint}\).
+The projection map \(\pi\) maps every fiber \(\dfiber_{\dbasepoint}\) to a point \(\dbasepoint \in \dbase\) in the base space.}
     \label{fig:math:data:base:div}
@@
 
-In \autoref{fig:math:data:base:div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping $\pi: \dtotal\rightarrow \dbase$ is the closed interval $\left[0,1\right]$.
-While the total space \dtotal\ may have components in \dfiber\ that describe any given point $\dbasepoint \in \dbase$, such as \textit{time, latitude, longitude}, these labels are indexed into from \dbase\ the same as any other components.
+In \autoref{fig:math:data:base:div}, \dtotal\ is a rectangle divided by vertical fibers \dfiber, so the minimal \dbase\ for which there is always a mapping \(\pi: \dtotal\rightarrow \dbase\) is the closed interval \(\left[0,1\right]\).
+While the total space \dtotal\ may have components in \dfiber\ that describe any given point \(\dbasepoint \in \dbase\), such as \textit{time, latitude, longitude}, these labels are indexed into from \dbase\ the same as any other components.
 In contrast to the structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, our model treats keys \dbasepoint\ as a pure reference to topology.
-Decoupling the keys from their semantics allows the components identifying the keys to be altered, which provides for coordinate agnostic representation of the continuity and facilitates encoding of data where the independent variable may not be clear.
-For example total rainfall is dependent on time of day and how much rain has already fallen, and changing the coordinate system or time resolution should have no effect on how the records are connected to each other.
+Decoupling the keys from their semantics allows the components identifying the keys to be altered, which provides for a coordinate agnostic representation of the continuity and facilitates encoding of data where the independent variable may not be clear.
+For example total rainfall is dependent on time of day and how much rain has already fallen; therefore changing the coordinate system should have no effect on how the records are connected to each other, as illustrated in \autoref{fig:intro:equivariance} where precipitation in inches and millimeters yield equivalent line plots.
 
-As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose}, we can decompose the total space into component bundles $\pi:\dtotal_i\rightarrow \dbase$ where
+As with \autoref{eq:math:data:fiber:decompose} and \autoref{eq:math:data:monoid:decompose}, we can decompose the total space into component bundles \(\pi:\dtotal_i\rightarrow \dbase\) where
 \begin{equation}
@@
 \label{sec:math:data:section}
-While the projection function $\pi:\dtotal \rightarrow\dbase$ ties together the base space \dbase\ with the fiber \dfiber, a section $\dsection: \dbase\rightarrow \dtotal$ encodes a dataset.
-A section function takes as input location $\dbasepoint \in \dbase$ and returns a record $\delement \in \dtotal$.
+While the projection function \(\pi:\dtotal \rightarrow\dbase\) ties together the base space \dbase\ with the fiber \dfiber, a section \(\dsection: \dbase\rightarrow \dtotal\) encodes a dataset.
+A section function takes as input location \(\dbasepoint \in \dbase\) and returns a record \(\delement \in \dtotal\).
 For example, in the special case of a table \cite{spivakSIMPLICIALDATABASES}, \dbase\ is a set of row ids, \dfiber\ is the columns, and the section \dsection\ returns the record \delement\ at a given key in \dbase.
@@
 \end{equation}
- such that $\pi(\dsection(\dbasepoint)) = \dbasepoint$.
-The set of all global sections is denoted as $\Gamma(\dtotal)$.
-Assuming a trivial fiber bundle $\dtotal = \dbase \times \dfiber$, the section can be decomposed as
+ such that \(\pi(\dsection(\dbasepoint)) = \dbasepoint\).
+The set of all global sections is denoted as \(\Gamma(\dtotal)\).
+Assuming a trivial fiber bundle \(\dtotal = \dbase \times \dfiber\), the section can be decomposed as
 \begin{equation}
@@
 \end{equation}
-where $g: \dbase \rightarrow \dfiber$ is the index function into the fiber.
+where \(g: \dbase \rightarrow \dfiber\) is the index function into the fiber.
 This formulation of the section also holds on locally trivial sections of a non-trivial fiber bundle.
@@
 \end{equation}
-where each section $\dsection_i$ maps into a record on a component $\dfiber_i \in \dfiber$.
+where each section \(\dsection_i\) maps into a record on a component \(\dfiber_i \in \dfiber\).
 This allows for accessing the data component wise in addition to accessing the data in terms of its location over \dbase.
@@
     \includegraphics[width=1\linewidth]{figures/math/fiberbundle.png}
-    \caption{ Fiber (time, precipitation) with a 1D continuous \dbase\  defined on an interval $\left[0,n\right]$.
-The sections $\dsection^{(1)}$ and $\dsection^{(2)}$ are constrained such that the time variable must be monotonic, which means each section is a time series of precipitation values.
-They are included in the global set of sections  $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$}
+    \caption{ Fiber (time, precipitation) with a 1D continuous \dbase\  defined on an interval \(\left[0,n\right]\).
+The sections \(\dsection^{(1)}\) and \(\dsection^{(2)}\) are constrained such that the time variable must be monotonic, which means each section is a time series of precipitation values.
+They are included in the global set of sections  \(\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)\)}
     \label{fig:math:data:sections}
@@
 
-In \autoref{fig:math:data:sections}, the fiber is the same encoding of $(time, \, precipitation)$ illustrated in \autoref{fig:math:data:fiber}, and the base space is the interval \dbase\ shown in \autoref{fig:math:data:base}.
-The section $\dsection^{(1)}$ is a function that for a point \dbasepoint\ returns a record in the fiber \dfiber.
+In \autoref{fig:math:data:sections}, the fiber is the same encoding of \((time, \, precipitation)\) illustrated in \autoref{fig:math:data:fiber}, and the base space is the interval \dbase\ shown in \autoref{fig:math:data:base}.
+The section \(\dsection^{(1)}\) is a function that for a point \dbasepoint\ returns a record in the fiber \dfiber.
 The section applied to a set of points in \dbase\ resolves to a series of monotonically increasing in time records of \textit{(time, precipitation)} values.
-Section $\dsection^{(2)}$ returns a different time series of \textit{(time, precipitation)} values.
-Both sections are included in the global set of sections $\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)$.
+Section \(\dsection^{(2)}\) returns a different time series of \textit{(time, precipitation)} values.
+Both sections are included in the global set of sections \(\dsection^{(1)}, \dsection^{(2)} \in \Gamma(\dtotal)\).
 
@@
 \label{sec:math:data:sheaf}
-Dynamic visualizations require evaluating sections on different subspaces of \dbase; this can be achieved using a mathematical structure, called a sheaf $\mathcal{O}$, for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.
-On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections $\iota^{*}\dsection$
+Dynamic visualizations require evaluating sections on different subspaces of \dbase; this can be achieved using a mathematical structure, called a sheaf \(\mathcal{O}\), for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.
+On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections \(\iota^{*}\dsection\)
 \begin{equation}
@@
 \end{equation}
-which are sections of \dtotal\ pulled back over local neighborhood $U \subset \dtotal$ via the inclusion map \(\iota:\dtotal\rightarrow U\).
+which are sections of \dtotal\ pulled back over local neighborhood \(U \subset \dtotal\) via the inclusion map \(\iota:\dtotal\rightarrow U\).
 The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.
@@
 \subsubsection{Applications}
-Using fiber bundles as our data abstraction allows the model to be applicable to a wide variety of fields and use cases.
-For example, the section can be any instance of a numpy array\cite{harris2020array} that stores an image, such as an image where the \dbase\ is a 2D continuous plane and the \dfiber\ is $(\reals^{3}, \reals, \reals)$.
-In this fiber, the $\reals^3$ components encode color, and the other two components are the x and y positions of the sampled data in the image.
+Using fiber bundles as the data abstraction allows the model to describe widely used data containers without sacrificing the semantic structure embedded in each container.
+For example, the section can be any instance of a numpy array\cite{harris2020array} that stores an image, such as an image where the \dbase\ is a 2D continuous plane and the \dfiber\ is \((\reals^{3}, \reals, \reals)\).
+In this fiber, the \(\reals^3\) components encode color, and the other two components are the x and y positions of the sampled data in the image.
 The continuity of the image is implicitly encoded in the array as the index, so the position components encode the resolution.
@@
 One such example is a pandas series which stores a labeled list, another is a  dataframe\cite{jeff_reback_2020_3715232} which has the structure of a relational table.
-A series could store the values of $\dsection^{(1)}$ and a second series could be  $\dsection^{(2)}$, while a dataframe would have multiple components and each data frame would be a unique section \dsection.
+A series could store the values of \(\dsection^{(1)}\) and a second series could be  \(\dsection^{(2)}\), while a dataframe would have multiple components and each data frame would be a unique section \dsection.
 The ability to encode complexity in continuity and components is particularly beneficial when working with  N dimensional labeled data containers.
@@
 This section does not need to resolve to values immediately and instead can be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.
-Our model provides a common formalism for describing widely used data containers without sacrificing the semantic structure embedded in each container.
 

### 2021-05-10 draft/sections/math.tex
graphics

@@
 To establish that the artist is a structure preserving map from data \dtotal\ to graphic \gtotal\, we construct a graphic bundle so that we can define \textit{equivariance} in terms of maps on the fiber spaces and \textit{continuity} in terms of maps on the base space.
-As with the data, we can represent the target graphic as a section \gsection\ of a bundle  $(\gtotal, \gbase, \pi, \gfiber)$.
+As with the data \dsection, we can represent the target graphic as a section \gsection\ of a bundle  \((\gtotal, \gbase, \pi, \gfiber)\).
 \begin{equation}
@@
 such that a rendered graphic only consists of 2D position and color.
-To support overplotting and transparency, the fiber could be $\gfiber=\reals^{7}$ such that $(x, y, z, r, g, b, a) \in \gfiber$ specifies the target display.
+To support overplotting and transparency, the fiber could be \(\gfiber=\reals^{7}\) such that \((x, y, z, r, g, b, a) \in \gfiber\) specifies the target display.
 By abstracting the target display space as \gfiber, the model can support different targets, such as a 2D screen or 3D printer.
@@
 
-For a visualization component to preserve continuity, we propose that there must exist a structure preserving surjective map $\vindex: \gbase\rightarrow\dbase$ from the data base space \dbase\ to the graphic base space \gbase.
+For a visualization component to preserve continuity, we propose that there must exist a structure preserving surjective map \(\vindex: \gbase\rightarrow\dbase\) from the data base space \dbase\ to the graphic base space \gbase.
 Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ such that \dbase\ and \gbase\ have the same homotopy, meaning there is a continuous map from \gbase\ to \dbase\cite{weissteinHomotopy}.
-The surjective map $\vindex: \gbase \rightarrow \dbase$
+The surjective map \(\vindex: \gbase \rightarrow \dbase\)
 \begin{equation}
@@
 \end{equation}
-goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\dbasepoint \in \dbase$.
-This means that if $\vindex(\gbasepoint) = \dbasepoint$, the record at \dbasepoint\ is copied over the region \gbasepoint\ such that $\dsection(\dbasepoint)=\vindex^*\dsection(\gbasepoint)$ where $\vindex^*\dsection(\gbasepoint)$  is \dsection\ pulled back over \gbase.
-The map \vindex\ is part of the implementation of the artist $\mathcal{\vartist}$ and therefore is not defined in terms of the data; instead it is how we specify the constraint that the type of the graphic \textit{continuity} must be able to map to the type of the data \textit{continuity}.
+goes from region \(\gbasepoint \in \gbase_{\dbasepoint}\) to its associated point \(\dbasepoint \in \dbase\).
+This means that if \(\vindex(\gbasepoint) = \dbasepoint\), the record at \dbasepoint\ is copied over the region \gbasepoint\ such that \(\dsection(\dbasepoint)=\vindex^*\dsection(\gbasepoint)\) where \(\vindex^*\dsection(\gbasepoint)\)  is \dsection\ pulled back over \gbase.
+The map \vindex\ is part of the implementation of the artist \(\mathcal{\vartist}\) and therefore is not defined in terms of the data; instead it is how we specify the constraint that the type of the graphic \textit{continuity} must be able to map to the type of the data \textit{continuity}.
 
@@
 A point that is 0D in \dbase\ cannot be represented on screen unless it is thickened to 2D (a) to encode the connectivity of the pixels that visually represent the point.
-This is also the case with the line (b), which would be infinitely think on screen if \gbase\ was not thickened to 2D.
+This is also the case with the line (b), which would be infinitely thin on screen if \gbase\ was not thickened to 2D.
 This thickening is often not necessary when the dimensionality of \dbase\ matches the dimensionality of the target space, for example if \dbase\ is 2D and the display is a 2D screen (c).
@@
 \label{sec:math:graphic:section}
-The section $\gsection: \gbase \rightarrow \gtotal$ is the graphic in an idealized prerender space and also acts as a specification for rendering the graphic to target display format.
-To demonstrate the role of \gsection\, it is sufficient to sketch out how an arbitrary pixel would be rendered, where a pixel $p$ in a real display corresponds to a region $\gbase_p$ in the idealized display.
+The section \(\gsection: \gbase \rightarrow \gtotal\) is the graphic in an idealized prerender space and also acts as a specification for rendering the graphic to target display format.
+To demonstrate the role of \gsection\, it is sufficient to sketch out how an arbitrary pixel would be rendered, where a pixel \(p\) in a real display corresponds to a region \(\gbase_p\) in the idealized display.
 To determine the color of the pixel, we aggregate the color values over the region via integration:
@@
 \end{align*}
-For a 2D screen, the pixel is defined as a region $p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]$ of the rendered graphic.
-Since the x and y in $p$ are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box $\gbase_{p} ={\gsection_{x,y}}^{-1}(p)$ is a region $\gbase_p \subset \gbase$.
-The color is the result of the integration over  $\gbase_p$.
+For a 2D screen, the pixel is defined as a region \(p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]\) of the rendered graphic.
+Since the x and y in \(p\) are in the same coordinate system as the x and y components of \gfiber\,  the inverse map of the bounding box \(\gbase_{p} ={\gsection_{x,y}}^{-1}(p)\) is a region \(\gbase_p \subset \gbase\).
+The color is the result of the integration over  \(\gbase_p\).
 
@@
     \includegraphics[width=\textwidth]{figures/math/render.png}
-    \caption{To render a graphic, a pixel $p$ is selected in the display space, which is defined in the same coordinates as the x and y components in \gfiber\ via the renderer.
-In \gtotal\, the inverse mapping ${\gsection_{x,y}}^(p)$ returns a region $\gbase_{p} \subset \gbase$.
-$\gsection(\gbase_{p})$ returns a set of points $(x,\,y,\,r,\,g,\,b) \in \gfiber$ that lie over $\gbase_{p}$.
-The integral over the $(r,\,g,\,b)$ pixels specifies that the pixel should be green}
+    \caption{To render a graphic, a pixel \(p\) is selected in the display space, which is defined in the same coordinates as the x and y components in \gfiber\ via the renderer.
+Therefore, the pixel \(p\) maps to a region on \gtotal.
+In \gtotal\, the inverse mapping \({\gsection_{x,y}}^(p)\) returns a region \(\gbase_{p} \subset \gbase\).
+\(\gsection(\gbase_{p})\) returns a set of points \((x,\,y,\,r,\,g,\,b) \in \gfiber\) that lie over \(\gbase_{p}\).
+The integral over the \((r,\,g,\,b)\) pixels specifies that the pixel should be green}
     \label{fig:math:graphic:rho:lookup}
@@
 
-As shown in \autoref{fig:math:graphic:rho:lookup}, a pixel $p$ in the output space, drawn in yellow, is selected and mapped, via the renderer, into a region on \gtotal.
-The region on \gtotal\ corresponds to a region $\gbase_p \subset \gbase$ via the inverse mapping  ${\gsection_{xy}}^(p)$.
+As shown in \autoref{fig:math:graphic:rho:lookup}, a pixel \(p\) in the output space, drawn in yellow, is selected and mapped, via the renderer, into a region on \gtotal.
+The region on \gtotal\ corresponds to a region \(\gbase_p \subset \gbase\) via the inverse mapping  \({\gsection_{xy}}^(p)\).
 The base space \gbase\ is an annulus to match the topology of the graphic idealized in \gtotal.
-The section  $\gsection(\gbase_{p})$  then maps into the fiber \gfiber\ over $\gbase_p$ to obtain the set of points in \gfiber, here represented as a table, that correspond to that section.
+The section  \(\gsection(\gbase_{p})\) then maps into the fiber \gfiber\ over \(\gbase_p\)to obtain the set of points in \gfiber, here represented as a table, that correspond to that section.
 The integral over the pixel components of this set of points in the fiber yields the color of the pixel.

### 2021-05-10 draft/sections/math.tex
graphic section

@@
 \label{sec:artist}
-We propose that visualization is structure preserving maps from data \dtotal\ to graphic \gtotal; having described \dtotal in \autoref{sec:math:data} and \gtotal in \autoref{sec:math:graphic}, we now define the visual transformations from \dtotal\ to \gtotal\ that formalize the components that visualization libraries implement.
-The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is $\mathcal{O}(\dtotal)$ to the sheaf on the graphic bundle \gtotal, $\mathcal{O}(\gtotal)$.
+We propose that visualization is structure preserving maps from data \dtotal\ to graphic \gtotal; having described \dtotal\ in \autoref{sec:math:data} and \gtotal\ in \autoref{sec:math:graphic}, we now define the visual transformations from \dtotal\ to \gtotal\ that formalize the components that visualization libraries implement.
+The topological artist \vartist\ is a map from the sheaf on a data bundle \dtotal\ which is \(\mathcal{O}(\dtotal)\) to the sheaf on the graphic bundle \gtotal, \(\mathcal{O}(\gtotal)\).
 \begin{equation}
@@
 The artist preserves \textit{continuity} through the \vindex\ map discussed in \autoref{sec:math:graphic:base}.
-We propose that the artist $\mathscr{\vartist}$ is an \textit{equivariant} map of monoid action  $m \in \monoid$
+We propose that the artist \(\mathscr{\vartist}\) is an \textit{equivariant} map of monoid action  \(m \in \monoid\)
 \begin{equation}
@@
 \end{equation}
-between data element $\delement \in \mathscr{\dtotal}$ and graphic element $\vartist(\delement) \in \mathscr{\gtotal}$.
-To be equivariant with respect to monoids action, we conjecture that an  artist carries a monoid homomorphism $\varphi$
+between data element \(\delement \in \mathscr{\dtotal}\) and graphic element \(\vartist(\delement) \in \mathscr{\gtotal}\).
+To be equivariant with respect to monoids action, we conjecture that an artist carries a monoid homomorphism \(\varphi\)
 \begin{equation}
@@
 \end{equation}
-such that an action in data space $m \in \monoid$ is equivalent to an action in graphic space $\varphi(\monoid) \in \monoid^{\prime}$.
+such that an action in data space \(m \in \monoid\) is equivalent to an action in graphic space \(\varphi(\monoid) \in \monoid^{\prime}\).
 
-The artist \vartist\ has two stages: the encoders $\vchannel:\dtotal^{\prime} \rightarrow \vtotal$ convert the data components to visual components, and the assembly function $\vmark: \vtotalpull \rightarrow \gtotal$ composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
+The artist \vartist\ has two stages: the encoders \(\vchannel:\dtotal \rightarrow \vtotal\) convert the data components to visual components, and the assembly function \(\vmark: \vtotalpull \rightarrow \gtotal\) composites the fiber components of \vtotalpull\ into a graphic in \gtotal.
 \begin{equation}
@@
 \end{equation}
-\vtotalpull\ is the visual bundle \vtotal\ pulled back over \gbase\ via the equivariant continuity map $\vindex:\gbase \rightarrow \dbase$ introduced in \autoref{sec:math:graphic:base}.
+\vtotalpull\ is the visual bundle \vtotal\ pulled back over \gbase\ via the equivariant continuity map \(\vindex:\gbase \rightarrow \dbase\) introduced in \autoref{sec:math:graphic:base}.
 The functional decomposition of the visualization artist in \autoref{eq:math:artist:diagram} facilitates building reusable components at each stage of the transformation because the equivariance constraints are defined on \vchannel, \vmark, and \vindex.
-We name this map the artist as that is the analogous part of the  Matplotlib\cite{hunterArchitectureOpenSource} architecture that builds visual elements.
+We name this map the artist as that is the analogous part of the  Matplotlib \cite{hunterArchitectureOpenSource} architecture that builds visual elements.
 
@@
 We introduce a visual bundle \vtotal\ to store the mappings of the data components into components of the graphic.
-These graphic components are implicit visualization library APIS; by making them explicit as components of the fiber we can define expectations of how these parameters behave.
-As with the data and graphic bundles, the visual bundle $(\vtotal,\,\dbase,\,\pi ,\,\vfiber)$ is defined by the projection map $\pi$
+These graphic components are implicit visualization library APIs; by making them explicit as components of the fiber we can define expectations of how these parameters behave.
+As with the data and graphic bundles, the visual bundle \((\vtotal,\,\dbase,\,\pi ,\,\vfiber)\) is defined by the projection map \(\pi\)
 \begin{equation}
@@
     \begin{tabulary}{\textwidth}{|l|L|l|}\hline
-     $\bm{\vchannel_{i}}$                      & $\bm{\vsection_{i}}$                                                            & $\bm{codomain(\vchannel_{i}) \subset \vfiber_{i}}$  \\ \hline
-    position                    & x, y, z, theta, r                                                          & $\mathbb{R}$   \\ \hline
-    size                        & linewidth, markersize                                            & $\mathbb{R}^{+}$   \\ \hline
-    shape                       & markerstyle                                                      & $\{f_{0}, \ldots, f_{n}\}$ \\ \hline
-    color                       & color, facecolor, markerfacecolor, edgecolor  & $\mathbb{R}^{4}$ \\ \hline
-    \multirow{2}{*}{texture}    & hatch                                                            & $\mathbb{N}^{10}$\\\cline{2-3}
-                                & linestyle                                                        & $(\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})$ \\ \hline              
+     \(\bm{\vchannel_{i}}\)    & \(\bm{\vsection_{i}}\)  & \(\bm{codomain(\vchannel_{i}) \subset \vfiber_{i}}\)  \\ \hline                                              
+    position                    & x, y, z, theta, r      & \(\mathbb{R}\)   \\ \hline
+    size                        & linewidth, markersize  & \(\mathbb{R}^{+}\)  \\ \hline
+    shape                       & markerstyle            & \(\{f_{0}, \ldots, f_{n}\}\)\\ \hline
+    color                       & color, facecolor, markerfacecolor, edgecolor  & \(\mathbb{R}^{4}\) \\ \hline
+    \multirow{2}{*}{texture}    & hatch      & \(\mathbb{N}^{10}\)\\\cline{2-3}
+                                & linestyle    & \((\mathbb{R}, \mathbb{R^+}^{n, n\%2=0})\) \\ \hline              
     \end{tabulary}
@@
 Given a fiber of \(\{x, y, color\}\) one possible section could be  \(\{.5, .5, (255, 20,147)\}\).
-The \(codomain(\vchannel_i)\) in \autoref{tab:math:artist:mpl:fiber} specifies the libraries internal representation of visual variables and can be used to determine which monoids can act on $\vfiber_i$.
+The \(codomain(\vchannel_i)\) in \autoref{tab:math:artist:mpl:fiber} specifies the libraries internal representation of visual variables and can be used to determine which monoids can act on \(\vfiber_i\).
 
@@
 \label{sec:math:artist:nu}
-We propose that the map from data components to graphic components $\vchannel: \dsection \mapsto \vsection$ is a monoid \textit{equivariant} map.
+We propose that the map from data components to graphic components \(\vchannel: \dsection \mapsto \vsection\) is a monoid \textit{equivariant} map.
 By specifying this constraints, we can guarantee that the stage of the artist that transforms data components into graphic representations is equivariant.
@@
 \end{equation}
-as the set of equivariant maps $\vchannel_i: \dsection_i \mapsto \vsection_i$.
-Given $\monoid_i$ is the monoid action on $\dtotal_i$ and that there is a monoid ${\monoid_i}^{\prime}$ on $\vtotal_i$, then there is a monoid homomorphism from $\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}$ that \vchannel\ must preserve.
+as the set of equivariant maps \(\vchannel_i: \dsection_i \mapsto \vsection_i\).
+Given \(\monoid_i\) is the monoid action on \(\dtotal_i\) and that there is a monoid \({\monoid_i}^{\prime}\) on \(\vtotal\), then there is a monoid homomorphism from \(\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}\) that \vchannel\ must preserve.
 As mentioned in \autoref{sec:math:data:monoid}, monoid actions define the structure on the fiber components and are therefore the basis for equivariance.
-Therefore, a validly constructed \vchannel\ is one where the diagram of the monoid transform $m$ commutes
+Therefore, a validly constructed \vchannel\ is one where the diagram of the monoid transform \(m\) commutes
 \begin{equation}
@@
 \end{equation}
-such that applying equivariant monoid actions to $\dtotal_{i}$ and $\vtotal_{i}$ preserves the map $\vchannel_{i}: \dtotal_{i}\rightarrow \vtotal_{i}$.
-In general, the data fiber $\dfiber_{i}$ cannot be assumed to be of the same type as the visual fiber $\vfiber_{i}$ and the actions of \monoid\ on $\dfiber_{i}$ cannot be assumed to be the same as the actions of $\monoid^{\prime}$ on \vfiber; therefore an equivariant $\vchannel_i$ must satisfy the constraint
+such that applying equivariant monoid actions to \(\dtotal_{i}\) and \(\vtotal_{i}\) preserves the map \(\vchannel_{i}: \dtotal_{i}\rightarrow \vtotal_{i}\).
+In general, the data fiber \(\dfiber_{i}\) cannot be assumed to be of the same type as the visual fiber \(\vfiber_{i}\) and the actions of \monoid\ on \(\dfiber_{i}\) cannot be assumed to be the same as the actions of \(\monoid^{\prime}\) on \vfiber; therefore an equivariant \(\vchannel_i\) must satisfy the constraint
 \begin{equation}
@@
 \end{equation} 
-such that $\varphi$ maps a monoid action on data to a monoid action on visual elements. However, 
-without a loss of generality we can assume that an action of \monoid\ acts on $\dfiber_{i}$ and on $\vfiber_{i}$ compatibly such that $\varphi$ is the identity function.
-We can make this assumption because we can construct a monoid action of \monoid\ on $\vfiber_{i}$ that is compatible with a monoid action of \monoid\ on $\dfiber_{i}$.
-We can then compose the monoid actions on the visual fiber $\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi$ that takes \monoid\ to $\monoid^{\prime}$.
-This allows us to define a monoid action on \vfiber\ of \monoid\ that is $(m, \velement) \rightarrow \varphi(m)\bullet\velement$, which lets us incorporate $\varphi$ into the action $\bullet$ such that $\varphi$ does not need to be explicitly defined in the constraints.
+such that \(\varphi\) maps a monoid action on data to a monoid action on visual elements. However, 
+without a loss of generality we can assume that an action of \monoid\ acts on \(\dfiber_{i}\) and on \(\vfiber_{i}\) compatibly such that $\varphi$ is the identity function.
+We can make this assumption because we can construct a monoid action of \monoid\ on \(\vfiber_{i}\) that is compatible with a monoid action of \monoid\ on \(\dfiber_{i}\).
+We can then compose the monoid actions on the visual fiber \(\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi\) that takes \monoid\ to \(\monoid^{\prime}\).
+This allows us to define a monoid action on \vfiber\ of \monoid\ that is \((m, \velement) \rightarrow \varphi(m)\bullet\velement\), which lets us incorporate \(\varphi\) into the action \(\bullet\) such that \(\varphi\) does not need to be explicitly defined in the constraints.
 
@@
   \end{subfigure}
-  \caption{ The \vchannel\ mapping in \autoref{fig:math:artist:nu:valid} represented by the colored arrows is monotonic, and therefore monoid equivariant,  since $\nu(storm)=\nu(rain)$ satisfies the condition $\nu(storm) \geq \nu(storm)$.
-In contrast, the map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not monotonic, and therefore not monoid equivariant, because $rain \geq storm$ is mapped to elements with the reverse ordering $\nu(storm) \geq \nu(storm)$.}
+  \caption{ The \vchannel\ mapping in \autoref{fig:math:artist:nu:valid} represented by the colored arrows is monotonic, and therefore monoid equivariant,  since \(\nu(storm)=\nu(rain)\) satisfies the condition \(\nu(storm) \geq \nu(storm)\).
+In contrast, the map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not monotonic, and therefore not monoid equivariant, because \(rain \geq storm\) is mapped to elements with the reverse ordering \(\nu(storm) \geq \nu(storm)\).}
   \label{fig:math:artist:nu}
 \end{figure}
-The mapping from weather state to umbrella in \autoref{fig:math:artist:nu:valid} is monotonic, and therefore we conjecture equivariant, because $\vchannel(rain) = \vchannel(storm)$ satisfies the monotonic condition of $rain \geq storm$.
+The mapping from weather state to umbrella in \autoref{fig:math:artist:nu:valid} is monotonic, and therefore we conjecture equivariant, because \(\vchannel(rain) = \vchannel(storm)\) satisfies the monotonic condition of \(rain \geq storm\).
 \autoref{fig:math:artist:nu} is an example of how the model supports partially ordered data components, which was a motivation for defining equivariance as monoid homomorphisms.
 In contrast, the translation from weather state data to visual representation as umbrella emoji in \autoref{fig:math:artist:nu:invalid} is an invalid visual encoding map \vchannel\ because it is not monotonic and therefore not equivariant This is because the monotonic condition \(rain \geq storm \implies \vchannel(rain) \geq  \vchannel(storm)\) is not met since \(\vchannel(rain) \leq\vchannel(storm)\).
-To satisfy the monotonic condition for $rain \geq storm$, either red arrow in \autoref{fig:math:artist:nu:invalid} would have to go in a different direction.
+To satisfy the monotonic condition for \(rain \geq storm\), either red arrow in \autoref{fig:math:artist:nu:invalid} would have to go in a different direction.
 
@@
       \textbf{scale} & \textbf{group} & \textbf{constraint}\\ \hline
-      nominal & permutation &  $\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)$\\
-      ordinal &  monotonic & $\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
-      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c$ \\
-      ratio &  scaling &  $\vchannel(xc) = \vchannel(x)*c $\\ \hline
+      nominal & permutation &  \(\text{if } \delement_1 \neq \delement_2 \text{ then } \vchannel (\delement_1) \neq\vchannel(\delement_2)\)\\
+      ordinal &  monotonic & \(\text{if } \delement_1 \leq \delement_2 \text{ then } \vchannel (\delement_1) \leq \vchannel(\delement_2)$\\
+      interval &  translation &  $\vchannel (x + c) = \vchannel(x) + c\) \\
+      ratio &  scaling &  \(\vchannel(xc) = \vchannel(x)*c \)\\ \hline
   \end{tabulary}
@@
 
-As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is mapping from section \vsectionpull\ over \gbasepoint to \vsection\ over \dbasepoint.
+As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is mapping from section \vsectionpull\ over \gbasepoint\ to \vsection\ over \dbasepoint.
 This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.

### 2021-05-10 draft/sections/math.tex
math

@@
 To guide the implementation of structure preserving visualization components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}.
-Inspired by the analogous classes in Matplotlib\cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic space that these building block components implement the \textit{artist}.
+Inspired by the analogous classes in Matplotlib \cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic space that these building block components implement the \textit{artist}.
 \begin{equation}
@@
 that binds the components of the data in \dfiber\ to the continuity of the data encoded in \dbase.
-Our use of fiber bundles builds on Butler's work proposing that fiber bundles should be the common data abstraction for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}.
+Our use of fiber bundles builds on Butler's work proposing that fiber bundles should be the common data abstraction for visualization data \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}.
 The fiber bundle models the properties of data component types \dfiber\ (\autoref{sec:math:data:fiber}), the continuity of records \dbase\ (\autoref{sec:math:data:base}), the collections of records (\autoref{sec:math:data:section}), and the space \dtotal\ of all possible datasets with these components and continuity.
-By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood \(U\) the total space is the cartesian product \(\dbase\times\dfiber\).
+By definition fiber bundles are locally trivial \cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood \(U\) the total space is the cartesian product \(\dbase\times\dfiber\).
 
@@
 which are sections of \dtotal\ pulled back over local neighborhood \(U \subset \dtotal\) via the inclusion map \(\iota:\dtotal\rightarrow U\).
-The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.
+The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom \cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.
 
@@
 Using fiber bundles as the data abstraction allows the model to describe widely used data containers without sacrificing the semantic structure embedded in each container.
-For example, the section can be any instance of a numpy array\cite{harris2020array} that stores an image, such as an image where the \dbase\ is a 2D continuous plane and the \dfiber\ is \((\reals^{3}, \reals, \reals)\).
+For example, the section can be any instance of a numpy array \cite{harris2020array} that stores an image, such as an image where the \dbase\ is a 2D continuous plane and the \dfiber\ is \((\reals^{3}, \reals, \reals)\).
 In this fiber, the \(\reals^3\) components encode color, and the other two components are the x and y positions of the sampled data in the image.
@@
 Labeled containers can also be described in this framework because of the schema like structure of the fiber.
-One such example is a pandas series which stores a labeled list, another is a  dataframe\cite{jeff_reback_2020_3715232} which has the structure of a relational table.
+One such example is a pandas series which stores a labeled list, another is a  dataframe \cite{jeff_reback_2020_3715232} which has the structure of a relational table.
 A series could store the values of \(\dsection^{(1)}\) and a second series could be  \(\dsection^{(2)}\), while a dataframe would have multiple components and each data frame would be a unique section \dsection.
-The ability to encode complexity in continuity and components is particularly beneficial when working with  N dimensional labeled data containers.
-For example, an xarray\cite{hoyer2017xarray} data cube that stores precipitation would be a section of a fiber bundle with a \dbase\ that is a continuous volume and components \textit{(time, latitude, longitude, precipitation)}.
+The ability to encode complexity in continuity and components is particularly beneficial when working with  N d imensional labeled data containers.
+For example, an xarray \cite{hoyer2017xarray} data cube that stores precipitation would be a section of a fiber bundle with a \dbase\ that is a continuous volume and components \textit{(time, latitude, longitude, precipitation)}.
 This section does not need to resolve to values immediately and instead can be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.
@@
 For a visualization component to preserve continuity, we propose that there must exist a structure preserving surjective map \(\vindex: \gbase\rightarrow\dbase\) from the data base space \dbase\ to the graphic base space \gbase.
-Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ such that \dbase\ and \gbase\ have the same homotopy, meaning there is a continuous map from \gbase\ to \dbase\cite{weissteinHomotopy}.
+Formally, we require that \dbase\ be a deformation retract \cite{RetractionTopology2020} of \gbase\ such that \dbase\ and \gbase\ have the same homotopy, meaning there is a continuous map from \gbase\ to \dbase \cite{weissteinHomotopy}.
 The surjective map \(\vindex: \gbase \rightarrow \dbase\)
@@
 When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the artists are acting on different components of the same section.
-This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.
+This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views \cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.
 
@@
 \end{figure}
-Representational invariance, as defined by Kindlmann and Scheidegger, is the notion that visualizations are equivalent if changing the visual representation, such as colors or shapes, does not change the meaning of the visualization\cite{kindlmannAlgebraicProcessVisualization2014}.
+Representational invariance, as defined by Kindlmann and Scheidegger, is the notion that visualizations are equivalent if changing the visual representation, such as colors or shapes, does not change the meaning of the visualization \cite{kindlmannAlgebraicProcessVisualization2014}.
 By defining a criteria for invariance, we can evaluate whether two artists generate the same type of graphic and compare artists across libraries.

### 2021-05-10 draft/sections/implementation.tex
closing out cite/spacing cleanups

@@
 \end{minted}
-Building on the current Matplotlib Artists which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalence class artist \vartisteq\ as described in \autoref{eq:math:artist:aprime}.
+Building on the current Matplotlib Artists, which construct an internal representation of the graphic, \mintinline{python}{ArtistClass} acts as an equivalence class artist \vartisteq\ as described in \autoref{eq:math:artist:aprime}.
 The visual bundle \vtotal\ is specified as the \mintinline{python}{V} dictionary of the form \mintinline{python}|{parameter:(variable name, encoder)}| where parameter is a component in \vfiber, variable is a component in \dfiber, and the \vchannel\ encoders are passed in as functions or callable objects.
@@
   \end{minted}
- The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
+ The data in section \dsection\ is fetched via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal.
 The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that can be used to query back into data to select a subset as described in \autoref{sec:math:data:sheaf}.
@@
 \end{figure}
-The figure in \autoref{fig:code:scatter} is described by \autoref{eq:eq:math:artist:q:scatter}.
+The figure in \autoref{fig:code:scatter} is a scatter plot, as described by \autoref{eq:eq:math:artist:q:scatter}.
 This is implemented via a \mintinline{python}{Line} object where the scatter marker shape is fixed as a circle, and the visual fiber components are x and y position and the facecolor and size of the marker.
@@
 \end{minted}
-As described in \autoref{eq:math:artist:nu:equivariance}, a test for equivariance can be implemented trivially. It is currently factored out of the artist for clarity.
+As described in \autoref{eq:math:artist:nu:equivariance}, a test for equivariance can be implemented trivially.
+The test shown here is the nominal test described in \autoref{tab:math:artist:nu} because \mintinline{python}{Categorical} is intended to encode nominal data.
+Equivariance tests are currently factored out of the artist for clarity.
 
@@
 This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{python}{view} method that returns an atomic object that encapsulates \dsection.
-To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataClasses}
+To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class \cite{DataClasses}
 
@@
 \end{minted}
-The discrete \mintinline{python}{tau} method returns a record of discrete points, akin to a row in a table, while the line\mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint.
+The discrete \mintinline{python}{tau} method returns a record of discrete points, akin to a row in a table, while the line \mintinline{python}{tau} returns a sampling of points along an edge \dbasepoint.
 These continuities are also specified in the \mintinline{python}|k={'tables':[]}| dictionary.
@@
 Instead, grouped bar charts are often achieved via looping over an implementation of bar, which forces the user to keep track which values are mapped to a single visual element and how that is achieved.
-For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020}, packaged as a pandas data frame\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.
+For this case study, we use the Palmer Penguins dataset \cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020}, packaged as a pandas data frame \cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure.
 
@@
 \end{table}
-
-The wrapper is very thin because there is explicitly only one section.
+The wrapper is very thin because there is explicitly only one section, which is the penguins dataframe. 
 \begin{minted}{python}
@@
 
-Since all the visual transformation is passed through to \mintinline{python}{Bar}, the \mintinline{python}{draw} method does not do any visual transformations.
+The visual transformations are passed through to \mintinline{python}{Bar}; therefore, the \mintinline{python}{draw} method does not do any visual transformations.
 In \mintinline{python}{StackedBar} the \mintinline{python}{view} is used to adjust the \mintinline{python}{floor} for every subsequent bar chart, since a stacked bar chart is bar chart area marks concatenated together in the length parameter.

### 2021-05-10 draft/sections/intro.tex
closing out cite/spacing cleanups

@@
 
-Visualizations are representations of data and therefore reflect something of the underlying structure and semantics\cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
+Visualizations are representations of data and therefore reflect something of the underlying structure and semantics \cite{friendlyBriefHistoryData2008}, whether through direct mappings from data into visual elements or via figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}.
 We define visualization components to be structure preserving maps from data to visual representations.
@@
 We introduce a model of visualization components based on these \textit{continuity} and \textit{equivariance} constraints and use this model to develop a design specification.
-Our specification is targeted at building block libraries, which Wongsuphasawat defines as visualization tools that provide independent functions that map components of data to visual representations\cite{wongsuphasawatNavigatingWideWorld2021}.
+Our specification is targeted at building block libraries, which Wongsuphasawat defines as visualization tools that provide independent functions that map components of data to visual representations \cite{wongsuphasawatNavigatingWideWorld2021}.
 Just as the number of ways to arrange physical building blocks is solely constrained by the size and shape of the blocks, we propose that the only restriction on how building block library components can be composed are that the compositions preserve \textit{continuity} and \textit{equivariance}.
-Independent, modular, components are inherently functional\cite{hughesWhyFunctionalProgramming1989}, so we propose a functional architecture for our model.
+Independent, modular, components are inherently functional \cite{hughesWhyFunctionalProgramming1989}, so we propose a functional architecture for our model.
 Doing so means the functional architecture can be evaluated for correctness, the resulting code is likely to be shorter and clearer, and the architecture is well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.
 
-This work is strongly motivated by the needs of the Matplotlib\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library.
+This work is strongly motivated by the needs of the Matplotlib \cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} visualization library.
 One of the most widely used visualization libraries in Python, new components and features have been added in an organic, sometimes hard to maintain, manner since 2002.
 In Matplotlib, every component carries its own implicit notion of how it believes the data is structured-for example if the data is a table, cube, image, or network - that is expressed in the API for that component.
-This leads to an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards\cite{a.sarikayaWhatWeTalk2019}.
+This leads to an inconsistent API for interfacing with the data, for example when  updating streaming visualizations or constructing dashboards  \cite{a.sarikayaWhatWeTalk2019}.
 This entangling of data model with visual transform also yields inconsistencies in how visual component transforms, e.g.
@@
 \end{figure}
-The components of a visual representation were first codified by Bertin\cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.
+The components of a visual representation were first codified by Bertin \cite{bertinSemiologyGraphicsDiagrams2011a}, who introduced a notion of structure preservation that we formally describe in terms of \textit{equivariance} and \textit{continuity}.
 Bertin proposes that there are categories of visual encodings-such as position, shape, color, and texture-that preserve the properties of the measurement type, quantitative or qualitative, of the encoded data.
@@
 This visualization is also equivariant to the table (a), but this is indicated through labeling rather than the line plots.
-The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}.
-The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations\cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}.
+The idea of equivariance is formally defined as the mapping of a binary operator from the data domain to the visual domain in Mackinlay's \textit{A Presentation Tool}(APT) model \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomaticDesignGraphical1987}.
+The algebraic model of visualization proposed by Kindlmann and Scheidegger uses equivariance to refer generally to invertible binary transformations \cite{kindlmannAlgebraicProcessVisualization2014}, which are mathematical groups \cite{shadrachIntroductionGroups2017}.
 Our model defines \textit{equivariance} in terms of monoid actions, which are a more restrictive set than all binary operations and more general than groups.
@@
 Bertin asserted that \textit{continuity} was preserved by choosing graphical marks that match the \textit{continuity} of the data - for example discrete data is a point, 1D continuous is the line, and 2D data is the area mark.
-Informally, Norman's Naturalness Principal\cite{NaturalnessPrincipleInfoVis,norman_things_smart} states that a visualization is easier to understand when the properties of the visualization match the properties of the data.
-This principal is made more concrete in Tufte's concept of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements\cite{tufteVisualDisplayQuantitative2001}.
+Informally, Norman's Naturalness Principal  \cite{NaturalnessPrincipleInfoVis,norman_things_smart} states that a visualization is easier to understand when the properties of the visualization match the properties of the data.
+This principal is made more concrete in Tufte's concept of graphical integrity, which is that a visual representation of quantitative data must be directly proportional to the numerical quantities it represents (Lie Principal), must have the same number of visual dimensions as the data, and should be well labeled and contextualized, and not have any extraneous visual elements  \cite{tufteVisualDisplayQuantitative2001}.
 Expressivity, as defined by Mackinlay, is a measure how much of the mathematical structure in the data can be expressed in the visualizations; for example that ordered variables can be mapped into ordered visual elements.
@@
 
-Most information visualization software design patterns are tuned to specific data structures and domains, as categorized by Heer and Agrawala\cite{HeerSoftware2006}.
-For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience\cite{heerDeclarative2010}.
+Most information visualization software design patterns are tuned to specific data structures and domains, as categorized by Heer and Agrawala  \cite{HeerSoftware2006}.
+For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space  \cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience \cite{heerDeclarative2010}.
 This coherent experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries.
 For example tools influenced by APT that assume that data is a relational table integrate computation into the visualization pipeline.
-This is a wide array of tools, including Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}.
-Since these libraries represent data as a table, and computations on tables are well defined\cite{ullmanFirstCourseDatabase2008}, these libraries implement computation as part of the visualization process.
-This is also true of tools that support images, such as the biology oriented ImageJ\cite{schneiderNIHImageImageJ2012} and  Napari\cite{nicholas_sofroniew_2021_4533308} or the digital humanities ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}.
+This is a wide array of tools, including Tableau \cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics \cite{wilkinsonGrammarGraphics2005} inspired ggplot \cite{wickhamGgplot2ElegantGraphics2016a}, protovis \cite{bostockProtoviz2009}, vega \cite{satyanarayanDeclarativeInteractionDesign2014} and altair \cite{vanderplasAltairInteractiveStatistical2018}.
+Since these libraries represent data as a table, and computations on tables are well defined \cite{ullmanFirstCourseDatabase2008}, these libraries implement computation as part of the visualization process.
+This is also true of tools that support images, such as the biology oriented ImageJ \cite{schneiderNIHImageImageJ2012} and  Napari \cite{nicholas_sofroniew_2021_4533308} or the digital humanities ImageJ macro ImagePlot \cite{studiesCulturevisImageplot2021}.
 While these tools often have some support for visualizing non image components of the data, the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure.
-Tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs.
+Tools like Gephi \cite{bastianGephiOpenSource2009}, Graphviz \cite{ellsonGraphvizOpenSource2002}, and Networkx \cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs.
 As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model.
-Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distribution statistics\cite{wickham40YearsBoxplots2011}.
+Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distribution statistics \cite{wickham40YearsBoxplots2011}.
 Disentangling the computation from the visual transforms allows us to determine whether the visualization library should implement computations or if they should be implemented in data space.
@@
 For example, a relational database often consists of discrete unconnected points, while an image is an implicitly connected 2D grid.
-This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
+This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.
+\cite{DataRepresentationMayavi}}
     \label{fig:intro:data:format}
@@
 Visualizations assume the structure of the input data, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighlevel2004}, which leads to many building block libraries implementing multiple models of data to support these different visualizations.
-For example, in Matplotlib, D3\cite{bostockDataDrivenDocuments2011} VTK \cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} and MayaVi\cite{RamachandranMayaVI2011}, every plot is defined in terms of the \textit{continuity} of the data it expects as input.
+For example, in Matplotlib, D3 \cite{bostockDataDrivenDocuments2011} VTK \cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} and MayaVi \cite{RamachandranMayaVI2011}, every plot is defined in terms of the \textit{continuity} of the data it expects as input.
 VTK has explicitly codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}.
-Downstream library developers impose some coherency by writing domain specific libraries with assumed data structures on top of the building block libraries--for example Seaborn\cite{waskom2020seaborn} and Titan\cite{brianwylieUnifiedToolkitInformation2009} assume a relational database, xarray\cite{hoyer2017xarray} and ParaView\cite{ahrens2005paraview} assume a data cube--but must work around the incoherencies in the building block libraries to do so.
+Downstream library developers impose some coherency by writing domain specific libraries with assumed data structures on top of the building block libraries--for example Seaborn \cite{waskom2020seaborn} and Titan \cite{brianwylieUnifiedToolkitInformation2009} assume a relational database, xarray \cite{hoyer2017xarray} and ParaView \cite{ahrens2005paraview} assume a data cube--but must work around the incoherencies in the building block libraries to do so.
 Our model navigates the tradeoff between coherency and extensibility by proposing functional composable well constrained visual components that take as input a structure aware data abstraction general enough to provide a common interface for many different types of data continuities.
@@
 \label{sec:intro:data:continuity}
-Fiber bundles were proposed by Butler as a core data structure for visualization because they encode data \textit{continuity} separately from the components of the dataset\cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
+Fiber bundles were proposed by Butler as a core data structure for visualization because they encode data \textit{continuity} separately from the components of the dataset \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}.
 Butler's model lacks a robust way of describing variables; therefore we encode a schema like description of the data in the fiber bundle using Spivak's topological description of data types \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}.
@@
 In the data cube,  the keys is the row, column, and cell ID, and in the map the key is the position in the grid.
-Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
+Image is figure 2.8 in Munzner's Visualization Analysis and Design \cite{munznerVisualizationAnalysisDesign2014}}
     \label{fig:intro:keys:values}
@@
 The \textit{continuity} can be described in some datasets by components of the dataset.
-This is formalized by Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}.
+This is formalized by Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values} \cite{munznerChDataAbstraction}.
 As shown in \autoref{fig:intro:keys:values}, keys can be labeled indexes, such as the attribute name and row ID, or semantically significant physical entities such as locations on a map.

### 2021-05-10 draft/sections/math.tex
closing out cite/spacing cleanups

@@
 The action in \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action.
-The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map\cite{fongInvitationAppliedCategory2019}
+The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map \cite{fongInvitationAppliedCategory2019}
 \begin{equation*}
@@
 where the structure the action preserves is the relative, rather than exact, ordering.
-Since groups are monoids with invertible operations, this definition of structure is broad enough to include the Steven's measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
+Since groups are monoids with invertible operations, this definition of structure is broad enough to include the Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
 
@@
 This explicit topology is a concise way of distinguishing between visualizations that appear identical but assume different continuity, for example heat maps and images.
-The base space \dbase\ acts as an indexing space, as emphasized by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other.
+The base space \dbase\ acts as an indexing space, as emphasized by Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, to express how the records in \dtotal\ are connected to each other.
 As shown in \autoref{fig:math:data:base:types}, \dbase\ can have any number of dimensions and can be continuous or discrete.
-Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space\cite{aurouxMath131Introduction} such that every \(\dbasepoint \in \dbase\) has a corresponding fiber \(\dfiber_k\)\cite{QuotientSpaceTopology2020}.
+Formally \dbase\ is the quotient space \cite{QuotientSpaceTopology2020} of \dtotal\, meaning it is the finest space \cite{aurouxMath131Introduction} such that every \(\dbasepoint \in \dbase\) has a corresponding fiber \(\dfiber_k\) \cite{QuotientSpaceTopology2020}.
 
@@
 While the total space \dtotal\ may have components in \dfiber\ that describe any given point \(\dbasepoint \in \dbase\), such as \textit{time, latitude, longitude}, these labels are indexed into from \dbase\ the same as any other components.
-In contrast to the structural \textit{keys} with associated \textit{values} proposed by Munzner\cite{munznerVisualizationAnalysisDesign2014}, our model treats keys \dbasepoint\ as a pure reference to topology.
+In contrast to the structural \textit{keys} with associated \textit{values} proposed by Munzner \cite{munznerVisualizationAnalysisDesign2014}, our model treats keys \dbasepoint\ as a pure reference to topology.
 Decoupling the keys from their semantics allows the components identifying the keys to be altered, which provides for a coordinate agnostic representation of the continuity and facilitates encoding of data where the independent variable may not be clear.
@@
 \label{sec:math:data:sheaf}
-Dynamic visualizations require evaluating sections on different subspaces of \dbase; this can be achieved using a mathematical structure, called a sheaf \(\mathcal{O}\), for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.
+Dynamic visualizations require evaluating sections on different subspaces of \dbase; this can be achieved using a mathematical structure, called a sheaf \(\mathcal{O}\), for defining collections of objects \cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.
 On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections \(\iota^{*}\dsection\)
@@
 which are sections of \dtotal\ pulled back over local neighborhood \(U \subset \dtotal\) via the inclusion map \(\iota:\dtotal\rightarrow U\).
-The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom \cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.
+The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom \cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows \cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.
 
@@
 In general, \gsection\ is an abstraction of rendering.
-In very broad strokes \gsection\ can be a specification such as PDF\cite{bienz1993portable}, SVG\cite{quintScalable2003}, or an openGL scene graph\cite{CarsonOpenGL1997} or a rendering engine such as cairo\cite{CairographicsOrg} or AGG\cite{shemanarevAntiGrainGeometry}.
+In very broad strokes \gsection\ can be a specification such as PDF \cite{bienz1993portable}, SVG \cite{quintScalable2003}, or an openGL scene graph \cite{CarsonOpenGL1997} or a rendering engine such as cairo \cite{CairographicsOrg} or AGG \cite{shemanarevAntiGrainGeometry}.
 Implementation of \gsection\ is out of scope for this proposal.
@@
 where \vsection\ is the visual variable encoding, as described by Bertin \cite{bertinSemiologyGraphicsDiagrams2011a}, of the data section \dsection.
-The visual bundle \vtotal\ is the full design space\cite{cardStructureInformationVisualization1997} of possible parameters of a visualization type, such as a scatter plot or line plot.
+The visual bundle \vtotal\ is the full design space \cite{cardStructureInformationVisualization1997} of possible parameters of a visualization type, such as a scatter plot or line plot.
 For example, one section \vsection\ of \vtotal\ is a tuple of visual values that specifies the visual characteristics of a part of a graphic.
@@
 
-The Stevens measurement types\cite{stevensTheoryScalesMeasurement1946}, listed in \autoref{tab:math:artist:nu}, are specified in terms of groups, which are monoids with invertible operations\cite{remlingAlgebraMath5353}.
+The Stevens measurement types \cite{stevensTheoryScalesMeasurement1946}, listed in \autoref{tab:math:artist:nu}, are specified in terms of groups, which are monoids with invertible operations\cite{remlingAlgebraMath5353}.
 We generalize to monoids to account for limitations in the types of data that can be described with the Stevens' scales \cite{johnsonPseudoMathematicsMentalSocial1936,thomasMathematizationNotMeasurement2014}
@@
 \end{equation}
-where the path\cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1].
+where the path \cite{ConnectedSpace2020}  $\gamma$ from \dbasepoint\ to \dbasepathpoint\ is a continuous function from the interval [0,1].
 We define the glyph as the graphic generated by $\vmark(\gbase_{\dbasepathpoint})$
@@
 \end{equation}
-such that for every glyph there is at least one corresponding region on \dbase, in keeping with the definition of glyph as any visually distinguishable element put forth by Ziemkiewicz and Kosara\cite{ziemkiewiczEmbeddingInformationVisualization2009}.
-The primitive point, line, and area marks\cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.
+such that for every glyph there is at least one corresponding region on \dbase, in keeping with the definition of glyph as any visually distinguishable element put forth by Ziemkiewicz and Kosara \cite{ziemkiewiczEmbeddingInformationVisualization2009}.
+The primitive point, line, and area marks \cite{bertinSemiologyGraphicsDiagrams2011a,carpendaleVisualRepresentationSemiology} are specially cased glyphs.
 

### 2021-05-10 draft/sections/next.tex
closing out cite/spacing cleanups

@@
 For this reason, TEAM does not include data space transforms, as are incorporated into libraries like Tableau or ggplot, instead leaving choice of computations to implementors of the data object.
-TEAM's intentional ignorance of semantics also means it cannot evaluate whether a figurative glyph \cite{byrneAcquiredCodesMeaning2016} is a semantically correct choice, but it can enforce equivariance constraints of glyphs generated from data components enforcing equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from data components\cite{beckfeathers2014,byrneFigurativeFramesCritical2017}.
+TEAM's intentional ignorance of semantics also means it cannot evaluate whether a figurative glyph \cite{byrneAcquiredCodesMeaning2016} is a semantically correct choice, but it can enforce equivariance constraints of glyphs generated from data components enforcing equivariance of figurative glyphs \cite{byrneAcquiredCodesMeaning2016} generated from data components \cite{beckfeathers2014,byrneFigurativeFramesCritical2017}.
 TEAM also allows graphics to have a lower dimensional continuity than the source data when a retraction map from one continuity to the other exists.
@@
 
-The prototype is deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as R graphics\cite{murrellGraphicsThirdEdition2018}, VTK, and D3.
-TEAM has only been tested using PNG files rendered with AGG\cite{shemanarevAntiGrainGeometry}, but is expected to work with all the file types Matplotlib currently supports, including svg, pdf, and eps.
-We have not yet addressed how this framework interfaces with high performance rendering libraries such as openGL\cite{CarsonOpenGL1997} that implement different models of \gsection.
+The prototype is deeply tied to Matplotlib's existing architecture, so it has not yet been worked through how the model generalizes to libraries such as R graphics \cite{murrellGraphicsThirdEdition2018}, VTK, and D3.
+TEAM has only been tested using PNG files rendered with AGG \cite{shemanarevAntiGrainGeometry}, but is expected to work with all the file types Matplotlib currently supports, including svg, pdf, and eps.
+We have not yet addressed how this framework interfaces with high performance rendering libraries such as openGL \cite{CarsonOpenGL1997} that implement different models of \gsection.
 
@@
 More work is needed to formalize the composition operators, equivalence class \vartisteq, and the mathematical model of interactivity.
-We also need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heat map\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}.
+We also need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image \cite{haber1990visualization,hansen2011visualization}, a heat map \cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot \cite{wickham40YearsBoxplots2011}.
 In summary, the proposed scope of work is
@@
 We also potentially can look at large scale biology or climate datasets.
-The data applications could be further integrated with topological\cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
-Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
-This specification of structure could also be used to develop a serialization structure tht could then be use to allow Matplotlib to interface with other visualization libraries such as open GL via shared serialization protocol.
-While this paper formulates visualization in terms of monoidal action homomorphisms between fiber bundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
+The data applications could be further integrated with topological \cite{heineSurveyTopologybasedMethods2016} and functional \cite{ramsayFunctionalDataAnalysis2006a} data analysis methods.
+Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics \cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014} of visualizations with respect to graphical encoding choices.
+This specification of structure could also be used to develop a serialization structure that could then be use to allow Matplotlib to interface with other visualization libraries such as open GL via shared serialization protocol.
+While this paper formulates visualization in terms of monoidal action homomorphisms between fiber bundles, the model lends itself to a categorical formulation \cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.
 

### 2021-05-14 draft/sections/intro.tex
color coding to clean up explainations

@@
 While these tools often have some support for visualizing non image components of the data, the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure.
-Tools like Gephi \cite{bastianGephiOpenSource2009}, Graphviz \cite{ellsonGraphvizOpenSource2002}, and Networkx \cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs.
+Tools like Gephi \cite{bastianGephiOpenSource2009}, Graphviz \cite{ellsonGraphvizOpenSource2002}, cytoscape \cite{shannonCytoscapeSoftwareEnvironment2003}, and Networkx \cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs.
 As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model.

### 2021-05-16 draft/sections/math.tex
middle of reworking monoid section

@@
 \label{sec:math:data:monoid}
-Implementing expressive visual encodings requires formally describing the structure on the components of the fiber.
-We conjecture that the structure can be formally defined as the actions of a monoid on the component.
-We can also discuss complex structures since a core property of monoids is composability \cite{yorgeyMonoidsThemeVariations}.
-In doing so, we can specify the properties of the component that must be preserved in a graphic representation.
+Expressiveness of visual encodings is defined as encoding the relations of the data in the visual space \cite{mackinlayAutomatingDesignGraphical1986}; we formally describe these relations as monoid actions on the data component.
+We propose that an expressive visual transformaton is equivariant with respect to actions on the data components and visual component.
+We define relations as monoids because they generalize to monoids because they are composable \cite{yorgeyMonoidsThemeVariations} and can be used to encode partial orders, such as multi-ranked indicators \cite{bruggemannRankingPrioritizationMultiindicator2011}.
 
@@
 \end{align*}
-We conjecture that given these properties, if there is a partial ordering on both sides of the action then the action is equivariant and monotonic.
 

### 2021-05-16 draft/sections/math.tex
redo of monoid section

@@
 We propose that an expressive visual transformaton is equivariant with respect to actions on the data components and visual component.
-We define relations as monoids because they generalize to monoids because they are composable \cite{yorgeyMonoidsThemeVariations} and can be used to encode partial orders, such as multi-ranked indicators \cite{bruggemannRankingPrioritizationMultiindicator2011}.
+We define relations as monoids because they generalize to monoids because they are composable \cite{yorgeyMonoidsThemeVariations} and can be used to describe partial order relations, such as multi-ranked indicators \cite{bruggemannRankingPrioritizationMultiindicator2011}.
 
 A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
-\begin{align*}
-    \textbf{associativity}\; & \text{for all} a, b, c \in \monoid \, (a \bullet b) \bullet c = a \bullet (b \bullet c)\\
-    \textbf{identity}\; & \text{for all } a\in \monoid,\,  e\bullet a = a 
-\end{align*} 
-As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlabauthorsAction2021} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
-\begin{align*}
-    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
-\end{align*}
+\begin{align}
+    \textbf{associativity}\; & \text{for all } a, b, c \in \monoid \, (a \ast b) \ast c = a \ast (b \ast c)\\
+    \textbf{identity}\; & \text{for all } a\in \monoid,\,  e\ast a = a 
+\end{align} 
+As defined on data components $\dfiber$, a left monoid action \cite{SemigroupAction2021,nlabauthorsAction2021} of $\monoid$ is a set $\dfiber$ with an action $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
+\begin{align}
+    \textbf{associativity}\;& \text{for all } m_j, m_k \in \monoid \text{ and } x\in \dfiber,\, m_j\bullet(m_k\bullet x) = (m_j\ast m_k) \bullet x\label{eq:math:data:monoid:asso}\\
+    \textbf{identity}\;& \text{for all } x\in \dfiber,\; e\in \monoid,\,  e\bullet x = x \label{eq:math:data:monoid:identity}
+\end{align}
+As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
+\begin{equation}
+    \label{eq:math:data:monoid:decompose}
+\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
+\end{equation}
+of each monoid \(\monoid_{i}\) on \(\dfiber_{i}\).  The monoid is added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)
 
@@
     \end{subfigure}
-    \caption{The action $\bullet$ in \autoref{fig:math:data:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right.
-The action maps the weather states to themselves, thereby preserving the ordering defined by the monoid $\ast$ on both sides of the action.
-We conjecture that the action in \autoref{fig:math:data:monoid:partial} is monotonic when the partial ordering of the weather states is the same as the partial order of the elements they are mapped to.
-Given $sun \geq clouds\geq rain,\, lightining$ on the right, we conjecture that the action  $sun,\, clouds \rightarrow clouds$, and $rain,\, lightining \rightarrow rain$ is structure preserving such that the relative ordering of elements is the same as the elements they are mapped to.}
+    \caption{In this example, the partially ordered set of weather values are a component of the data $F_1$.
+In \autoref{fig:math:data:monoid:total} the monoid $m_1$ is the identity $e \in \monoid$ and takes every element $x \in \dfiber_i$ to itself.
+In \autoref{fig:math:data:monoid:partial}, the monoid $m_2$ maps the weather elements on the left to the elements on the right such that multiple weather elements on the right are mapped into the same element on the left.
+The monoid $m_2$ is order preserving, meaning for example that given $sun \geq clouds$ is true, then $m_2(sun) \geq m_2(clouds)$ must be true.
+The monoids are composable, meaning that whether they are applied in stages, such as the output of $m_1$ \autoref{fig:math:data:monoid:total} is the input to \autoref{fig:math:data:monoid:partial}, or the monoids (arrows) are first composed $m_1 \circ m_2$, the result will be the same.}
     \label{fig:math:data:monoid}
 \end{figure}
-
-One example of monoids are partial orderings on a set, such as seen in \autoref{fig:math:data:monoid}.
-Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one.
-For example, $storm \leq rain$.
-In \autoref{fig:math:data:monoid}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states.
-The action in \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action.
-The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map \cite{fongInvitationAppliedCategory2019}
+In \autoref{fig:math:data:monoid}, the monoids $m_1, m_2\in \monoid_i$ act on the partially ordered set $\dfiber_i=weather$.
+In this example, we define the actions $\bullet$ that compose $m_j \in \monoid_i$ with the set \textit{weather} to be monotone maps \cite{fongInvitationAppliedCategory2019}
 \begin{equation*}
+    \label{eq:math:data:monoid:monotone}
     if\, a \leq b \; then \; \bullet(a) \leq \bullet(b) \mid a,b \in \dfiber_{i}
 \end{equation*}
-where the structure the action preserves is the relative, rather than exact, ordering.
+In \autoref{fig:math:data:monoid:total}, the monoid $m_1$ takes every element in the set \textit{weather} to itself and is therefore the identity $m_1 = e \in \monoid_i$.
+The monoid $m_2$ in \autoref{fig:math:data:monoid:partial} maps the elements of weather on the left to the subset on the right in a way that satisfies the monotonicity condition in \autoref{eq:math:data:monoid:monotone}.
+For example, on the right $sun \geq clouds$ is true.
+Applying the monoid, the result  $m_2(sun) \geq m_2(clouds)$ is true because $m_2(sun)=clouds$ and $m_2(clouds)=clouds$.
+Composable actions, as defined in \autoref{eq:math:data:monoid:asso}, means that either applying $m_1$ to the weather elements in $F_1$ and then $m_2$ to the results
+\begin{equation}
+\begin{tikzcd}
+    \dfiber_i \arrow[d, "m_1"'] \arrow[rd, "m_1\;\circ\; m_2"] &         \\
+    \dfiber_i \arrow[r, "m_2"']                            & \dfiber_i
+\end{tikzcd}
+\end{equation}
+or composing $m_1\circ m_2$ and applying the composition to the elements in $F_i$ will yield the same result.
 Since groups are monoids with invertible operations, this definition of structure is broad enough to include the Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
-
-As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
-\begin{equation}
-    \label{eq:math:data:monoid:decompose}
-\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
-\end{equation}
-of each monoid \(\monoid_{i}\) on \(\dfiber_{i}\).  The monoid is added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)
 

### 2021-05-19 draft/sections/math.tex
monoid section

--- 2021-05-10 draft/sections/math.tex
@@
 \label{sec:math:data:monoid}
-Implementing expressive visual encodings requires formally describing the structure on the components of the fiber.
-We conjecture that the structure can be formally defined as the actions of a monoid on the component.
-We can also discuss complex structures since a core property of monoids is composability \cite{yorgeyMonoidsThemeVariations}.
-In doing so, we can specify the properties of the component that must be preserved in a graphic representation.
+Expressiveness of visual encodings is defined as encoding the relations of the data in the visual space \cite{mackinlayAutomatingDesignGraphical1986} and we formally describe these relations as monoid actions on the data component.
+We describe relations using monoids because they encompass the Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}, partial order relations, such as multi-ranked indicators \cite{bruggemannRankingPrioritizationMultiindicator2011}, and are composable \cite{yorgeyMonoidsThemeVariations}.
 
 A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
-\begin{align*}
-    \textbf{associativity}\; & \text{for all} a, b, c \in \monoid \, (a \bullet b) \bullet c = a \bullet (b \bullet c)\\
-    \textbf{identity}\; & \text{for all } a\in \monoid,\,  e\bullet a = a 
-\end{align*} 
-As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlabauthorsAction2021} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
-\begin{align*}
-    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
-    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
-\end{align*}
-We conjecture that given these properties, if there is a partial ordering on both sides of the action then the action is equivariant and monotonic.
+\begin{align}
+    \textbf{associativity}\; & \text{for all } m_j, m_k, m_l \in \monoid \, (m_j \ast m_k) \ast m_l = m_j \ast (m_k \ast m_l)\\
+    \textbf{identity}\; & \text{for all } m_j\in \monoid,\,  e\ast m_j = m_j 
+\end{align} 
+As defined on data components $\dfiber$, a left monoid action \cite{SemigroupAction2021,nlabauthorsAction2021} is a set $\dfiber$ with an action $\bullet: \monoid\times \dfiber \rightarrow \dfiber$ with the properties:
+\begin{align}
+    \textbf{associativity}\;& \text{for all } m_j, m_k \in \monoid \text{ and } x\in \dfiber,\, m_j\bullet(m_k\bullet x) = (m_j\ast m_k) \bullet x\label{eq:math:data:monoid:asso}\\
+    \textbf{identity}\;& \text{for all } x\in \dfiber,\; e\in \monoid,\,  e\bullet x = x \label{eq:math:data:monoid:identity}
+\end{align}
+
+As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
+\begin{equation}
+    \label{eq:math:data:monoid:decompose}
+\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
+\end{equation}
+of each monoid \(\monoid_{i}\) on each component \(\dfiber_{i}\).  The monoid is added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)
+
+As defined in \autoref{eq:math:data:monoid:asso}, the functions $m_j, m_k \in \monoid_i$ are composable
+\begin{equation}
+\begin{tikzcd}
+    \dfiber_i \arrow[d, "m_j"'] \arrow[rd, "m_j\;\circ\; m_k"] &         \\
+    \dfiber_i \arrow[r, "m_k"']  & \dfiber_i
+\end{tikzcd}
+\end{equation}
+This means either applying $m_k$ to the elements in $F_i$ and then $m_j$ to the results or composing $m_j\circ m_k$ and applying the composition to the elements in $F_i$ will yield the same result.
 
@@
     \end{subfigure}
-    \caption{The action $\bullet$ in \autoref{fig:math:data:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right.
-The action maps the weather states to themselves, thereby preserving the ordering defined by the monoid $\ast$ on both sides of the action.
-We conjecture that the action in \autoref{fig:math:data:monoid:partial} is monotonic when the partial ordering of the weather states is the same as the partial order of the elements they are mapped to.
-Given $sun \geq clouds\geq rain,\, lightining$ on the right, we conjecture that the action  $sun,\, clouds \rightarrow clouds$, and $rain,\, lightining \rightarrow rain$ is structure preserving such that the relative ordering of elements is the same as the elements they are mapped to.}
+    \caption{In this example, the partially ordered set of weather values are the elements of the fiber component $F_i$.
+In \autoref{fig:math:data:monoid:total} the function $m_j$ is the identity function $e \in \monoid$ that takes every element $x \in \dfiber_i$ to itself.
+In \autoref{fig:math:data:monoid:partial}, the function $m_k$ maps the weather elements on the left to the elements on the right such that multiple weather elements on the right are mapped into the same element on the left.
+The function $m_k$ is order preserving, meaning for example that $sun \geq clouds$ and $m_k(sun) \geq m_k(clouds)$ are both true.
+The functions $m_j, m_k$ are composable, meaning that whether they are applied in stages, such as the output of $e$ \autoref{fig:math:data:monoid:total} is the input to \autoref{fig:math:data:monoid:partial}, or the functions (the arrows) are first composed $m_j \circ m_k$, the result will be the same.}
     \label{fig:math:data:monoid}
 \end{figure}
-
-One example of monoids are partial orderings on a set, such as seen in \autoref{fig:math:data:monoid}.
-Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one.
-For example, $storm \leq rain$.
-In \autoref{fig:math:data:monoid}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states.
-The action in \autoref{fig:math:data:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action.
-The action $\bullet$ in \autoref{fig:math:data:monoid:partial} is a monotone map \cite{fongInvitationAppliedCategory2019}
+In \autoref{fig:math:data:monoid}, the arbitrarily chosen functions $m_j, m_k\in \monoid_i$ act on the partially ordered set $\dfiber_i=weather$.
+In this example, we define the functions $m_j, m_k \in \monoid_i$ applied to elements of  $\dfiber_i$ to be monotone maps \cite{fongInvitationAppliedCategory2019}
 \begin{equation*}
-    if\, a \leq b \; then \; \bullet(a) \leq \bullet(b) \mid a,b \in \dfiber_{i}
+    \label{eq:math:data:monoid:monotone}
+    if\, a \leq b \; then \; m_j(a) \leq m_j(b) \mid a,b \in \dfiber_{i}
 \end{equation*}
-where the structure the action preserves is the relative, rather than exact, ordering.
-Since groups are monoids with invertible operations, this definition of structure is broad enough to include the Steven's measurement scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}.
-
-As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
-\begin{equation}
-    \label{eq:math:data:monoid:decompose}
-\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
-\end{equation}
-of each monoid \(\monoid_{i}\) on \(\dfiber_{i}\).  The monoid is added to the specification of the fiber \((\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)\)
+In \autoref{fig:math:data:monoid:total}, the function $m_j$ is the identity function $e$ which takes every element in $\dfiber_i$ to itself.
+The function $m_k$ in \autoref{fig:math:data:monoid:partial} maps the elements of weather on the left to the subset on the right in a way that satisfies the monotonicity condition in \autoref{eq:math:data:monoid:monotone}.
+For example, on the right $sun \geq clouds$ is true and on the left $m_k(sun) \geq m_k(clouds)$ is true since $m_k(sun)=clouds$ and $m_k(clouds)=rains$ and $clouds\geq rain$.
+Since the functions $m_j, m_k$ are composable, the result is the same whether the input to $m_k$ is the output of $m_j$ or the functions (arrows) are combined before being applied to the elements in $F_i$.
 

### 2021-05-19 draft/sections/math.tex
color coding, function labels

@@
 between data element \(\delement \in \mathscr{\dtotal}\) and graphic element \(\vartist(\delement) \in \mathscr{\gtotal}\).
-To be equivariant with respect to monoids action, we conjecture that an artist carries a monoid homomorphism \(\varphi\)
+To be equivariant with respect to monoids action, we propose that an artist carries a monoid homomorphism \(\varphi\)
 \begin{equation}
@@
 \end{equation}
-such that an action in data space \(m \in \monoid\) is equivalent to an action in graphic space \(\varphi(\monoid) \in \monoid^{\prime}\).
+such that an action in data space \(m \in \monoid\) is equivariant to an action in graphic space \(\varphi(\monoid) \in \monoid^{\prime}\).
 
@@
 We propose that the map from data components to graphic components \(\vchannel: \dsection \mapsto \vsection\) is a monoid \textit{equivariant} map.
-By specifying this constraints, we can guarantee that the stage of the artist that transforms data components into graphic representations is equivariant.
+We conjecture that if the data space or visual space is partially ordered and the map \vchannel\ between the two spaces is equivariant, then it must also be monotonic.
+By specifying this constraint, we can guarantee that the stage of the artist that transforms data components into graphic representations is equivariant.
 These constraints then guide the implementation of reusable component transformers \vchannel\ that are composed when generating the graphic.
@@
 as the set of equivariant maps \(\vchannel_i: \dsection_i \mapsto \vsection_i\).
-Given \(\monoid_i\) is the monoid action on \(\dtotal_i\) and that there is a monoid \({\monoid_i}^{\prime}\) on \(\vtotal\), then there is a monoid homomorphism from \(\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}\) that \vchannel\ must preserve.
+Given \(\monoid_i\) is the monoid action on \(\dtotal_i\) and that there is a monoid \({\monoid_i}^{\prime}\) on \(\vtotal_i\), then there is a monoid homomorphism from \(\varphi:\monoid_i \rightarrow {\monoid_i}^{\prime}\) that \vchannel\ must preserve.
 As mentioned in \autoref{sec:math:data:monoid}, monoid actions define the structure on the fiber components and are therefore the basis for equivariance.
-Therefore, a validly constructed \vchannel\ is one where the diagram of the monoid transform \(m\) commutes
+Therefore, a validly constructed \vchannel\ is one where the diagram of the monoid action \(m\) commutes
 \begin{equation}
@@
 such that applying equivariant monoid actions to \(\dtotal_{i}\) and \(\vtotal_{i}\) preserves the map \(\vchannel_{i}: \dtotal_{i}\rightarrow \vtotal_{i}\).
-In general, the data fiber \(\dfiber_{i}\) cannot be assumed to be of the same type as the visual fiber \(\vfiber_{i}\) and the actions of \monoid\ on \(\dfiber_{i}\) cannot be assumed to be the same as the actions of \(\monoid^{\prime}\) on \vfiber; therefore an equivariant \(\vchannel_i\) must satisfy the constraint
+In general, the data fiber \(\dfiber_{i}\) cannot be assumed to be of the same type as the visual fiber \(\vfiber_{i}\) and the actions of \(\monoid_i\) on \(\dfiber_{i}\) cannot be assumed to be the same as the actions of \({\monoid_i}^{\prime}\) on \(\vfiber_i\); therefore an equivariant \(\vchannel_i\) must satisfy the constraint
 \begin{equation}
@@
 such that \(\varphi\) maps a monoid action on data to a monoid action on visual elements. However, 
-without a loss of generality we can assume that an action of \monoid\ acts on \(\dfiber_{i}\) and on \(\vfiber_{i}\) compatibly such that $\varphi$ is the identity function.
-We can make this assumption because we can construct a monoid action of \monoid\ on \(\vfiber_{i}\) that is compatible with a monoid action of \monoid\ on \(\dfiber_{i}\).
-We can then compose the monoid actions on the visual fiber \(\monoid^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi\) that takes \monoid\ to \(\monoid^{\prime}\).
-This allows us to define a monoid action on \vfiber\ of \monoid\ that is \((m, \velement) \rightarrow \varphi(m)\bullet\velement\), which lets us incorporate \(\varphi\) into the action \(\bullet\) such that \(\varphi\) does not need to be explicitly defined in the constraints.
-
-\begin{figure}[htb]
-  \centering
-  \begin{subfigure}{.49\textwidth}
-    \includegraphics[width=1\textwidth]{figures/math/partial_fixed.png}
-    \caption{}
-    \label{fig:math:artist:nu:valid}
-  \end{subfigure}
-  \begin{subfigure}{.49\textwidth}
-    \includegraphics[width=1\textwidth]{figures/math/partial_invalid.png}
-    \caption{}
-    \label{fig:math:artist:nu:invalid}
-  \end{subfigure}
-  \caption{ The \vchannel\ mapping in \autoref{fig:math:artist:nu:valid} represented by the colored arrows is monotonic, and therefore monoid equivariant,  since \(\nu(storm)=\nu(rain)\) satisfies the condition \(\nu(storm) \geq \nu(storm)\).
-In contrast, the map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not monotonic, and therefore not monoid equivariant, because \(rain \geq storm\) is mapped to elements with the reverse ordering \(\nu(storm) \geq \nu(storm)\).}
-  \label{fig:math:artist:nu}
-\end{figure}
-The mapping from weather state to umbrella in \autoref{fig:math:artist:nu:valid} is monotonic, and therefore we conjecture equivariant, because \(\vchannel(rain) = \vchannel(storm)\) satisfies the monotonic condition of \(rain \geq storm\).
-\autoref{fig:math:artist:nu} is an example of how the model supports partially ordered data components, which was a motivation for defining equivariance as monoid homomorphisms.
-In contrast, the translation from weather state data to visual representation as umbrella emoji in \autoref{fig:math:artist:nu:invalid} is an invalid visual encoding map \vchannel\ because it is not monotonic and therefore not equivariant This is because the monotonic condition \(rain \geq storm \implies \vchannel(rain) \geq  \vchannel(storm)\) is not met since \(\vchannel(rain) \leq\vchannel(storm)\).
-To satisfy the monotonic condition for \(rain \geq storm\), either red arrow in \autoref{fig:math:artist:nu:invalid} would have to go in a different direction.
+without a loss of generality we can assume that an action of\(\monoid_i\) acts on \(\dfiber_{i}\) and on \(\vfiber_{i}\) compatibly such that \(\varphi\) is the identity function.
+We can make this assumption because we can construct a monoid action of \({monoid_i}^{\prime}\) on \(\vfiber_{i}\) that is compatible with a monoid action of \(monoid_i\) on \(\dfiber_{i}\).
+We can then compose the monoid actions on the visual fiber \({\monoid_i}^{\prime} \times \vfiber_{i} \rightarrow \vfiber_{i}$ with the homomorphism $\varphi\) that takes \(monoid_i\) to \({\monoid_i}^{\prime}\).
+This allows us to define a monoid action on \(vfiber_i\) of \(monoid_i\) that is \((m, \velement) \rightarrow \varphi(m)\bullet\velement\), which lets us incorporate \(\varphi\) into the action \(\bullet\) such that \(\varphi\) does not need to be explicitly defined in the constraints.
 
@@
 \end{table}
+We generalize equivariance constraints to monoid action equivariance to account for limitations in the types of data that can be described with the Stevens' scales \cite{johnsonPseudoMathematicsMentalSocial1936,thomasMathematizationNotMeasurement2014}.
+The Stevens measurement types \cite{stevensTheoryScalesMeasurement1946}, listed in \autoref{tab:math:artist:nu}, are specified in terms of groups, which are monoids with invertible operations\cite{remlingAlgebraMath5353}.
 
-The Stevens measurement types \cite{stevensTheoryScalesMeasurement1946}, listed in \autoref{tab:math:artist:nu}, are specified in terms of groups, which are monoids with invertible operations\cite{remlingAlgebraMath5353}.
-We generalize to monoids to account for limitations in the types of data that can be described with the Stevens' scales \cite{johnsonPseudoMathematicsMentalSocial1936,thomasMathematizationNotMeasurement2014}
+\begin{figure}[H]
+  \centering
+  \begin{subfigure}{.49\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/partial_fixed.png}
+    \caption{}
+    \label{fig:math:artist:nu:valid}
+  \end{subfigure}
+  \begin{subfigure}{.49\textwidth}
+    \includegraphics[width=1\textwidth]{figures/math/partial_invalid.png}
+    \caption{}
+    \label{fig:math:artist:nu:invalid}
+  \end{subfigure}
+  \caption{ The data component \(\dfiber_i\) is the partially ordered set of words describing weather and the visual component \(\vfiber_i\) is a partially ordered set of emojis; the hasse diagrams in the figure are a subset of each fiber component.
+The \vchannel\ mapping in \autoref{fig:math:artist:nu:valid} from \(\dfiber_i\) to \(\vfiber_i\) is monotonic, and therefore equivariant.
+One example of this monotonicity is that \(rain \geq storm \) and \(\nu(rain) \geq \nu(storm)\).
+In contrast, the map from data component to visual component in \autoref{fig:math:artist:nu:invalid} is not monotonic, and therefore not monoid equivariant, because \(rain \geq storm\) is mapped to elements with the reverse ordering such that \(\nu(rain) \geq \nu(storm)\) is false.}
+  \label{fig:math:artist:nu}
+\end{figure}
+\autoref{fig:math:artist:nu} illustrates the encoding of weather data as umbrella emojis.
+The weather data is a subset of the partially ordered data component \(\dfiber_i\), while the umbrellas are a subset of the partially ordered visual component \(vfiber_i\).
+In \autoref{fig:math:artist:nu}, \vchannel\ is equivariant and therefore monotonic, meaning it maps elements from \(\dfiber\) to elements in \(\vfiber_i\) such that the monotonicity condition in \autoref{tab:math:artist:nu} is satisfied.
+In contrast, the \vchannel\ in \autoref{fig:math:artist:nu:invalid} is not equivariant and therefore not monotonic, which also means it is invalid.
+To satisfy the monotonic condition for \(rain \geq storm\), either the arrow between \textit{storm} and \textit{rain} or the arrow between the \textit{closed} and \textit{wet} umbrellas in \autoref{fig:math:artist:nu:invalid} would have to go in a different direction.
+This is because the monotonic condition \(rain \geq storm \implies \vchannel(rain) \geq  \vchannel(storm)\) is not met since \(\vchannel(rain) \leq\vchannel(storm)\).
 
-\subsubsection{Visualization Assembly}
+\subsubsection{Visualization Assembly \vmark}
 Having described the maps to components in \autoref{sec:math:artist:nu}, we now specify the assembly function \vmarkd\ that composites components in \vtotal\ into a graphic in \gtotal.
@@
 
-\subsubsection{Assembly \vmark}
+\subsubsection{Assembly Template \vmarkd}
+\label{sec:math:artist:qhat}
+The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
+As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
+%[column sep=huge]
+\begin{equation}
+    \begin{tikzcd}
+        \textcolor{gray!50}{\dtotal} \arrow[r, "\vchannel", color=gray!50] \arrow[rd, "\pi"', color=gray!50] & \vtotal \arrow[d, "\pi"']                  & \vindex^*\vtotal \arrow[r, "\vmark", color=gray!50] \arrow[d, "\vindex^*\pi"'] \arrow[l,  "\vindex^*"'] & \textcolor{gray!50}{\gtotal} \arrow[ld, "\pi", color=gray!50] \\
+                                                          & \dbase \arrow[u, "\vsection"', bend right] & \gbase \arrow[l, "\vindex"'] \arrow[u, "\vindex^*\vsection"', bend right]               &
+        \end{tikzcd}
+        \label{eq:math:artist:qhat}
+\end{equation}
+
+The map from graphic base space $\vindex:\gbase \rightarrow \dbase$ (\autoref{sec:math:graphic:base}) to data space maps many points in \gbase\ to a single point in \dbase.
+This means that the preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back of \vsection\
+\begin{equation}
+    \vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber
+\end{equation}
+copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase.
+This set of points \gbasepoint\ are the preimage \vpreimg\ of \dbasepoint.
+
+\begin{figure}[H]
+    \centering
+      \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
+      \caption{Because the pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint, we can construct a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
+      \label{fig:math:artist:qhat}
+  \end{figure}
+
+As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is mapping from section \vsectionpull\ over \gbasepoint\ to \vsection\ over \dbasepoint.
+This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
+In \autoref{fig:math:artist:qhat} each dot on \vfiber\ is equivalent to the line on \vfiberpull.
+
+Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out.
+When \vmark\ maps visual sections into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull\ then the graphic section \gsection\ evaluated on a visual region \gbasepoint\
+\begin{equation}
+    \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
+    \label{eq:math:artist:qhat_factor}
+\end{equation}
+ is defined as the assembly function \vmark\ with input \vsectionpull\ evaluated on \gbasepoint.
+Since the pulled back section \vsectionpull\ is the section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function
+\begin{equation}
+\label{eq:math:artist:qhat_q_s}
+\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
+\end{equation} 
+where \vmarkd\ with input \vsection\ is defined to \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.
+We can then factor \gbasepoint\ out of \autoref{eq:math:artist:qhat_q_s}, which yields
+\begin{equation}
+\vmarkd(\vsection(k)) = \vmark(\vsectionpull)
+\label{eq:math:artist:qhat_k}
+\end{equation}
+where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
+In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
+This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
+The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
+
+\subsubsection{Case Studies: Scatter, Line, Image}
 Given the continuities described in \ref{fig:math:graphic:retraction:map}, 
-we illustrate a minimal Q that will generate the most minimal visualizations associated with those continuities: non-overlapping scatter points, a non-infinitely thin line, and an image. 
+we illustrate a minimal \vmark (\vmarkd) that will generate the most minimal visualizations associated with those continuities: non-overlapping scatter points, a non-infinitely thin line, and an image.
 \begin{figure}[H]
@@
 
-\subsubsection{Assembly Template \vmarkd}
-\label{sec:math:artist:qhat}
-The graphic base space \gbase\ is not accessible in many architectures, including Matplotlib; instead we can construct a factory function \vmarkd\ over \dbase\ that can build a \vmark.
-As shown in \autoref{eq:math:artist:diagram}, \vmark\ is a bundle map $\vmark: \vtotalpull\rightarrow \gtotal$ where \vtotalpull\ and \gtotal\ are both bundles over \gbase.
-%[column sep=huge]
-\begin{equation}
-    \begin{tikzcd}
-        \textcolor{gray!50}{\dtotal} \arrow[r, "\vchannel", color=gray!50] \arrow[rd, "\pi"', color=gray!50] & \vtotal \arrow[d, "\pi"']                  & \vindex^*\vtotal \arrow[r, "\vmark", color=gray!50] \arrow[d, "\vindex^*\pi"'] \arrow[l,  "\vindex^*"'] & \textcolor{gray!50}{\gtotal} \arrow[ld, "\pi", color=gray!50] \\
-                                                          & \dbase \arrow[u, "\vsection"', bend right] & \gbase \arrow[l, "\vindex"'] \arrow[u, "\vindex^*\vsection"', bend right]               &
-        \end{tikzcd}
-        \label{eq:math:artist:qhat}
-\end{equation}
-
-The map from graphic base space $\vindex:\gbase \rightarrow \dbase$ (\autoref{sec:math:graphic:base}) to data space maps many points in \gbase\ to a single point in \dbase.
-This means that the preimage of the continuity map $\vpreimg \subset \gbase$ is such that many graphic continuity points $\gbasepoint \in \gbase_{\dbase}$ go to one data continuity point \dbasepoint; therefore, by definition the pull back of \vsection\
-\begin{equation}
-    \vtotalpull \mid_{\vpreimg} = \vpreimg \times \vfiber
-\end{equation}
-copies the visual fiber \vfiber\ over the the points \gbasepoint\ in graphic space \gbase\ that correspond to one \dbasepoint\ in data space \dbase.
-This set of points \gbasepoint\ are the preimage \vpreimg\ of \dbasepoint.
-
-\begin{figure}[H]
-    \centering
-      \includegraphics[width=1\textwidth]{figures/math/q_hat.png}
-      \caption{Because the pullback of the visual bundle \vtotalpull\ is the replication of a \vsection\ over all points \gbasepoint\ that map back to a single \dbasepoint, we can construct a \vmarkd\ on \vsection\ over \dbasepoint\ that will fabricate the \vmark\ for the equivalent region of \gbasepoint\ associated to that \dbasepoint}
-      \label{fig:math:artist:qhat}
-  \end{figure}
-
-As shown in \autoref{fig:math:artist:qhat}, given the section \vsectionpull\ pulled back from \vsection\ and the point $\gbasepoint \in \vpreimg$, there is mapping from section \vsectionpull\ over \gbasepoint\ to \vsection\ over \dbasepoint.
-This means that the pulled back section $\vsectionpull(\gbasepoint) = \vindex^*(\vsection(\dbasepoint))$ is the section \vsection\ copied over all \gbasepoint such that \vsectionpull\ is identical for all \gbasepoint\ where $\vindex(\gbasepoint) = \dbasepoint$.
-In \autoref{fig:math:artist:qhat} each dot on \vfiber\ is equivalent to the line on \vfiberpull.
-
-Given the equivalence between \vsection\ and \vsectionpull\ defined above, the reliance on \gbase\ can be factored out.
-When \vmark\ maps visual sections into graphics $\vmark: \Gamma(\vtotalpull) \rightarrow \Gamma(\gtotal)$, if we restrict \vmark\ input to \vsectionpull\ then the graphic section \gsection\ evaluated on a visual region \gbasepoint\
-\begin{equation}
-    \gsection(\gbasepoint) \coloneqq \vmark(\vsectionpull)(s)
-    \label{eq:math:artist:qhat_factor}
-\end{equation}
- is defined as the assembly function \vmark\ with input \vsectionpull\ evaluated on \gbasepoint.
-Since the pulled back section \vsectionpull\ is the section \vsection\ copied over every graphic region $\gbasepoint \in \vpreimg$, we can define a \vmark\ factory function
-\begin{equation}
-\label{eq:math:artist:qhat_q_s}
-\vmarkd(\vsection(\dbasepoint))(\gbasepoint) \coloneqq \vmark((\vsectionpull)(\gbasepoint))
-\end{equation} 
-where \vmarkd\ with input \vsection\ is defined to \vmark\ that takes as input the copied section \vsectionpull\ such that both functions are evaluated over the same location $\vpreimg = \gbasepoint$ in the base space \gbase.
-We can then factor \gbasepoint\ out of \autoref{eq:math:artist:qhat_q_s}, which yields
-\begin{equation}
-\vmarkd(\vsection(k)) = \vmark(\vsectionpull)
-\label{eq:math:artist:qhat_k}
-\end{equation}
-where \vmark\ is no longer bound to input but \vmarkd\ is still defined in terms of \dbase.
-In fact, \vmarkd\ is a map from visual space to graphic space $\vmarkd:\Gamma(\vtotal) \rightarrow \Gamma(\gtotal)$ locally over \dbasepoint\ such that it can be evaluated on a single visual record  $\vmarkd:\Gamma(\vtotal_{\dbasepoint}) \rightarrow \Gamma(\gtotal\mid_{\vpreimg})$.
-This allows us to construct a \vmarkd\ that only depends on \dbase, such that for each $\vsection(\dbasepoint)$ there is part of $\gsection\mid_{\vpreimg}$.
-The construction of \vmarkd\ allows us to retain the functional map reduce benefits of \vmark\ without having to restructure the existing pipeline for libraries that delegate the construction of \gsection\ to a back end such as Matplotlib.
-
 \subsubsection{Composition of Artists: +}

### 2021-05-19 draft/sections/next.tex
revision

@@
     \textbf{work period} & \textbf{milestones \& tasks} \\ \hline
-    April - July 2021  & prepare and submit \textbf{conference presentation} on new functionality enabled by model for \textit{SciPy}: \newline artists that do not inherit from existing Matplotlib artists, computational artists such as histograms, non tabular data, composite interactive artist \\ \hline
+    April - July 2021  & prepare and submit \textbf{conference presentation} on new functionality enabled by model for \textit{SciPy}: \newline artists that do not inherit from existing Matplotlib artists, computational artists such as histograms, non-tabular data, composite interactive artist \\ \hline
     June - Sept 2021  & prepare and submit \textbf{theory paper} on interactivity to \textit{TCVG or Eurovis 2022}: \newline fully work out and describe math of addition operators and lookups from graphic to data space, implement brush linked artist (shared base space) and artist that exploits sheafs \\ \hline
-    May - Nov 2021 & prepare and submit \textbf{applications paper} on high dimensional to \textit{TCVG}: \newline math and implementation of computational artists, concurrent artists and data sources, non-trivial data bundles \\ \hline
+    May - Nov 2021 & prepare and submit \textbf{applications paper} on high dimensional data to \textit{TCVG}: \newline math and implementation of computational artists, concurrent artists and data sources, non-trivial data bundles \\ \hline
     Aug 2021 - Feb 2022 & prepare and submit \textbf{systems paper} on building domain specific libraries based on this model to \textit{Infoviz 2022}: \newline domain specific structured data, composite artists, inference of meta data components, mathematical notion of a visualization (labeled, multiple artists, etc)\\ \hline
-    March 2022 & \textbf{dissertation} writing: \newline synthesize previous work on climate data, compile topological equivariant artist model work \\ \hline
+    March 2022 & \textbf{dissertation} writing: \newline incorporate previous work on climate data, compile topological equivariant artist model work \\ \hline
     April 2022 & \textbf{defense} \\ \hline

### 2021-05-19 code/matplottoy/artists/bar.py

@@ -14,6 +14,66 @@ import matplotlib.artist as martist
 
 from matplottoy.artists import utils
 
+class BarArtist(martist.Artist):
+    def __init__(self, data, transforms, orientation='v', *args, **kwargs):
+        """
+        Parameters
+        ----------
+        datasource: matplottoy.datasources.DataSource like
+            data object containing data to be plotted
+        orientation: str, optional
+            vertical: bars aligned along x axis, heights on y
+            horizontal: bars aligned along y axis, heights on x
+        **kwargs:
+            kwargs passed through 
+        """
+  
+        #assert 'vertex' in data.FB.K['tables']
+        #utils.check_constraints(Bar, transforms)
+        #utils.validate_transforms(data, transforms)
+
+        self.orientation = orientation
+     
+        super().__init__(*args, **kwargs)
+        self.data = data
+        self.transforms = copy.deepcopy(transforms)
+    
+    def assemble(self, position, length, floor=0, width=0.8, facecolors='C0', edgecolors='k', offset=0):
+        #set some defaults
+        width = itertools.repeat(width) if np.isscalar(width) else width
+        floor = itertools.repeat(floor) if np.isscalar(floor) else (floor)
+        
+        # offset is passed through via assemblers such as multigroup, not supposed to be directly tagged to position 
+        position = position + offset
+        
+        def make_bars(xval, xoff, yval, yoff):
+             return [[(x, y), (x, y+yo), (x+xo, y+yo), (x+xo, y), (x, y)] 
+                for (x, xo, y, yo) in zip(xval, xoff, yval, yoff)]
+        #build bar glyphs based on graphic parameter
+        if self.orientation in {'vertical', 'v'}:
+            verts = make_bars(position, width, floor, length)
+        elif self.orientation in {'horizontal', 'h'}:
+            verts = make_bars(floor, length, position, width)
+        
+        #self._paths = [mpath.Path(xy, closed=True) for xy in verts]
+        #self.set_edgecolors(edgecolors)
+        #self.set_facecolors(facecolors)
+        
+    def draw(self, renderer,  *args, **kwargs):
+        view = self.data.view(self.axes)
+        visual = {}
+        for (p, t) in self.transforms.items():
+            if isinstance(t, dict):
+                if t['name'] in self.data.FB.F and 'encoder' in t:
+                    visual[p] = t['encoder'](view[t['name']])
+                elif 'encoder' in t:
[…]

### 2021-05-21 code/matplottoy/artists/bar.py
dataclasses

@@ -1,5 +1,7 @@
 import copy
 from collections import defaultdict
+from dataclasses import dataclass
+
 import itertools
 
 import numpy as np 
@@ -7,13 +9,20 @@ import numpy as np
 from cycler import cycler
 
 from matplotlib import rcParams
+
+import matplotlib.artist as martist
 import matplotlib.collections as mcollections
 import matplotlib.path as mpath
 import matplotlib.patches as mpatches
-import matplotlib.artist as martist
+import matplotlib.transforms as mtransforms
 
 from matplottoy.artists import utils
 
+@dataclass
+class Graphic:
+    paths: list[mpath.Path]
+    edgecolors: list[]
+    facecolors: list[]
 class BarArtist(martist.Artist):
     def __init__(self, data, transforms, orientation='v', *args, **kwargs):
         """
@@ -55,9 +64,9 @@ class BarArtist(martist.Artist):
         elif self.orientation in {'horizontal', 'h'}:
             verts = make_bars(floor, length, position, width)
         
-        #self._paths = [mpath.Path(xy, closed=True) for xy in verts]
-        #self.set_edgecolors(edgecolors)
-        #self.set_facecolors(facecolors)
+        return Graphic([mpath.Path(xy, closed=True) for xy in verts], 
+                        edgecolors, facecolors)
+
         
     def draw(self, renderer,  *args, **kwargs):
         view = self.data.view(self.axes)
@@ -72,8 +81,10 @@ class BarArtist(martist.Artist):
                     visual[p] = view[t['name']]
             else:
                  visual[p] = t
-        self.assemble(**visual)
-        super().draw(renderer,  *args, **kwargs)
+        graphic = self.assemble(**visual)
+        gc = renderer.new_gc()
+        renderer.draw_path_collection(gc, master_transform=mtransforms.IdentityTransform(), **graphic)
+        #super().draw(renderer,  *args, **kwargs)
 
 
 class Bar(mcollections.Collection):

### 2021-05-21 code/matplottoy/encoders/mtypes.py

@@ -3,6 +3,7 @@ to check against visual channels
 """
 # if validate overwrites is instance, then we can 
 # use float/int/etc...
+from dataclasses import dataclass
 
 import matplotlib.colors as mcolors
 class Nominal: 
@@ -13,6 +14,12 @@ class Nominal:
     def validate(self, value):
         return value in self.categories
 
+@dataclass
+class RGBA:
+    r: float
+    g: float
+    b: float
+    a: float
 class Color:
     mtype = 'nominal'
     @staticmethod

### 2021-08-23 code/matplottoy/artists/bar.py
horizontal code + vertical code

@@ -1,6 +1,7 @@
+from __future__ import annotations
 from collections import namedtuple
 import copy
-from typing import Tuple
+from typing import Tuple, Optional
 
 
 import numpy as np 
@@ -20,106 +21,97 @@ from matplottoy.encoders import mtypes
 # mu = \nu \circ \tau 
 #\tau = data.query().projection()
 Mu = namedtuple('Mu', ['nu_i', 'F_i'])
-
-
-class QHatData: 
-# also maybe should be owned by data library? 
-#is bad name for this since this is more than just qhat
-    def __init__(self, graphic):
-        self.graphic = graphic
-        self.section = None
-
-    def compose_with_tau(self, section): #doesn't need to be there
-        # do I care about bounds yet? or only at draw time?
-        new = copy.copy(self)
-        new.section = section
-        return new
-
-    def query(self, data_bounds=None):
-        return self.section.query(data_bounds)
-        # section.view, #xi seperates nicly 
-
-
-class QHatGraphic:
-    def make_mu(self, bounds):
+class Rho: # this is really rho = \qhat \circ \mu
+    def mu(self, tau_restricted:dict):
         pass
     @staticmethod
     def qhat():
         pass
 
-class Bar(QHatGraphic): 
+class TopologicalArtist: 
+# also maybe should be owned by data library? 
+#is bad name for this since this is more than just qhat
+    def __init__(self, graphic:Rho):
+        # is parameterized by the rho
+        # only input is \tau (E)
+        self.graphic = graphic
+        self.data = None
+
+    def compose_with_tau(self, section): #doesn't need to be there
+        # do I care about bounds yet? or only at draw time?
+        new = copy.copy(self)
+        new.data = section
+        return new
+
+    def query(self, data_bounds:Optional[dict]=None, sampling_rate:Optional[int]=None)->dict:
+        return self.data.query(data_bounds, sampling_rate)
+        # section.view, #xi seperates nicely 
+
+class Bar(Rho): 
     #what are the 
     P = {'position', 'height', 'floor', 'width', 
          'facecolor', 'edgecolor', 'linewidth', 'linestyle'}
     
-    def __init__(self, orientation='v', V=None):
+    def __init__(self, orientation='v', V:Optional[dict]=None):
         # F->V maps need to be fixed here, but use fiber name instead of function
[…]

### 2021-08-23 code/matplottoy/data/fruit.py

@@ -1,4 +1,5 @@
 from dataclasses import dataclass
+from typing import Tuple, Optional, List
 
 import numpy as np
 import pandas as pd
@@ -6,95 +7,20 @@ import pandas as pd
 # index is like the sheaf, stalk is either the fiber or the values? 
 # check the applied category theory book 
 # index is subset on K
-# technically this can be wrapped even more 
-
+# technically this can be wrapped even more
+# 
+# maybe make a Record type Record = {Fiber_Name: values} 
 class DataFrameWrapper:
-    def __init__(self, section):
+    def __init__(self, section: pd.DataFrame):
         self.global_section = section 
 
-    def query(self, data_bounds=None, sampling_rate=None):
-        #can do tests of condition 
-        data_bounds = {} if data_bounds is None else data_bounds
-        mask = '&'.join(f'{k}>={vmin} & {k}<={vmax}' for (k, (vmin, vmax)) in data_bounds.items())
-        local_section = self.global_section.query(mask)
-        return {Fi:local_section[Fi] for Fi in local_section.columns}
-
-        #return [dict{'F_i': f()/np.array/etc}]
-        
-
-@dataclass
-class FruitSection:
-    K = 'integer'
-    F = {'fruit': np.dtype('O'), 
-         'calories': np.dtype('int64'), 
-         'juice': np.dtype('bool'), 
-         'None': np.dtype('float')}
-    section: pd.DataFrame
-
-    def __post_init__(self):
-        #check the fiber is a subset
-        # section: K \rightarrow F
-        # checks that output is in F
-        if not (self.section.dtypes.to_dict().items() <= self.F.items()):
-            raise TypeError(f"{self.section.dtypes.to_dict()} not in fiber {self.F}")
-        # checking on type since section doesn't have to be global
-        # check that input is on K
-        if self.section.index.inferred_type != 'integer':
-            raise TypeError(f"{self.section.index.inferred_type} is not type {self.K}")
-
-    #local_section F over K->F over {U}
-    def local_section(self, data_bounds=None): #query
-        """
-        # this is BAD!
-        data_bounds: {'fiber_key', (data_min, data_max)}
-        """
-        if data_bounds is None:
-            return FruitSection(self.section)
-        data_bounds = {} if data_bounds is None else data_bounds
-        mask = '&'.join(f'{k}>={vmin} & {k}<={vmax}' for (k, (vmin, vmax)) in data_bounds.items())
[…]

### 2021-08-25 code/matplottoy/artists/bar.py
more on case study

@@ -47,7 +47,8 @@ class TopologicalArtist:
         return self.data.query(data_bounds, sampling_rate)
         # section.view, #xi seperates nicely 
 
-class Bar(Rho): 
+# BarH and BarV are different Q^{hat} so BarFactory returns the right Bar?
+class Bar(RhoFactory): 
     #what are the 
     P = {'position', 'height', 'floor', 'width', 
          'facecolor', 'edgecolor', 'linewidth', 'linestyle'}
@@ -73,17 +74,29 @@ class Bar(Rho):
             new_key = {'position': 'x', 'height':'y'}
         elif orientation == 'h': 
             # needs to be a different artist 'cause data_bounds won't work here
-
-            new_key = {'floor':'x', 'height':'width', 'width':'y', 'position':'floor'}
+            #data bounds need position y, height x
+            new_key = {'floor':'x', 'height':'width', 
+                       'width':'y', 'position':'floor'}
         return {(new_key[k] if k in new_key else k):v for (k, v) in V.items()}
 
     def mu(self, tau_restricted: dict)->dict: #draw
-        return {p_i : mu_i.nu_i(*(tau_restricted[f] for f in mu_i.F_i)) 
-                for (p_i, mu_i) in self.rename_P(self.V, self.orientation).items()} 
+        V = self.rename_P(self.V, self.orientation)
+        mus = {}
+        mus['x'] = V['x'].nu_i(**(tau_restricted[f] for f in V['x'].F_i))
+        mus['y'] = V['y'].nu_i(**(tau_restricted[f] for f in V['y'].F_i))
+        V.pop('x')
+        V.pop('y')
+
+        for (p_i, mu_i) in V.items():
+            if mu_i.F_i is None:
+                pass
+            else:
+                mus[p_i] = mu_i.nu_i(*(tau_restricted[f] for f in mu_i.F_i)) 
+       return mus
+              
 
     @staticmethod
     def qhat(x, width, y, floor, facecolor, edgecolor, linewidth, linestyle):
-
         def fake_draw(render, transform=mtransforms.IdentityTransform()):
             for (xi, xw, yi, yf, fc, ec, lw, ls) in zip(x, width, y, floor, facecolor, edgecolor, linewidth, linestyle):
                 gc = render.new_gc()

### 2021-08-25 code/matplottoy/data/fruit.py

@@ -12,15 +12,8 @@ import pandas as pd
 # maybe make a Record type Record = {Fiber_Name: values} 
 class DataFrameWrapper:
     def __init__(self, section: pd.DataFrame):
-        self.global_section = section 
+        self.global_section = section
 
     def query(self, data_bounds:Optional[dict] = None, sampling_rate:int=None)->List[dict]:
-        if data_bounds is not None:
-        #  in theory this should be smarter to account for categorical...
-            query_str = '&'.join(f'{k}>={vmin} & {k}<={vmax}' for (k, (vmin, vmax)) in data_bounds.items())
-            local_section = self.global_section.query(query_str)
-        else:
-            local_section = self.global_section
-        #theory would return list of local sections
-        local_section = local_section.assign(default=np.nan)
+        local_section = self.global_section
         return [{Fi:local_section[Fi].values.squeeze() for Fi in local_section.columns}]

### 2021-08-25 leg.py
clean up accidental inclusion

@@ -1,8 +1,0 @@
-import matplotlib
-matplotlib.use('TkAgg')
-
-import matplotlib.pyplot as plt
-fig, ax = plt.subplots(figsize=(6,10))
-l, = ax.plot([1,2,3], [1,2,3], label="hello")
-ax.legend(bbox_to_anchor=(1, 1))
-plt.show()

### 2021-08-25 paper/scraps.tex
cleaned up fibers
more cleanups/adding to bib

@@
   \end{figure}
+
+  Expressiveness, as defined by Mackinlay \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987, mackinlayAutomatingDesignGraphical1986} is a measure of how much of the structure any map can encode.
+A fully expressive component is one that is equivariant since it has preserved all the structure in the data.
+Models of visualization evaluate this equivariance and how elements built in the model can be composed to build more complex visualizations.
+Mackinlay's \textit{A Presentation Tool} (APT) introduced the notion of visualizations having syntax and semantics \cite{mackinlayAutomatingDesignGraphical1986} and Wilkenson described the grammar of this language \cite{wilkinsonGrammarGraphics2005}.
+This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements.
